### 2025-03-13 thesis/acronyms.tex
thesis documents

%%%%%%%%%%%%%%% Acronyms %%%%%%%%%%%%%%%%%%
\usepackage[acronym,automake,symbols,
nogroupskip,
accsupp,
counter=section,
stylemods={longextra}
]{glossaries-extra}
\makeglossaries

\setabbreviationstyle[acronym]{long-short-sc}
\GlsXtrEnablePreLocationTag{\S }{\S }

\newacronym
[ description ={Core collapse supernovae. Massive star explosions. {\sc Ccsne} produce many elements including \gls{alpha}, \gls{fepeak}, and r and \gls{sproc}.}
]
{cc}{ccsne}{core collapse supernovae}

\newacronym
[description={Asymptotic giant branch stars. The {\sc agb} phase is the last phase of \glspl{lowmass}, before stars become white dwarfs. Produces C, N, and heavy \gls{sproc} elements. See \ref{sec:stel_evo}.}.]
{agb}{agb}{asymptotic giant branch}

\newacronym
[description={Type Ia Supernovae. Exploding white dwarfs. Produces \gls{fepeak} and has a long delay time.}]
{ia}{sne ia}{supernovae type Ia}

\newacronym
[description={Galactic chemical evolution.}]
{gce}{gce}{galactic chemical evolution}

\newacronym
[description={Star formation history.}]
{sfh}{sfh}{star formation history}

\newacronym
[description={Single stellar population. A group of stars formed all at the same time.}]
{ssp}{ssp}{single stellar population}

\newacronym
[description={Initial mass function. A function describing the mass distribution of newly formed stars. I use a \citet{kroupa01} {\sc imf}, which is described as a piecewise power-law function of $M$.}]
{imf}{imf}{initial mass function}

\newacronym
[description={
Apache Point Observatory Galactic Evolution Experiment. 
A large near-infrared spectroscopic survey of stars in the Milky Way. \citep{apogee17.}}]
{apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

\newacronym
[
description={Damped Lyman-alpha system. {\sc dla}s are clouds of gas from the 
early universe which are observed through their absorption of quasar spectra. The name 
comes from the strong Lyman-alpha lines ($1216\AA$) due to H absorption.}
]
{dla}{dla}{damped Lyman-alpha systems}

\newcommand{\cc}{\gls{cc}}
\newcommand{\Cc}{\Gls{cc}}
\newcommand{\agb}{\gls{agb}}
\newcommand{\ia}{\gls{ia}}
\newcommand{\sfh}{\gls{sfh}}
\newcommand{\dla}{\gls{dla}}
\newcommand{\ssp}{\gls{ssp}}
\newcommand{\imf}{\gls{imf}}
\newcommand{\gce}{\gls{gce}}
\newcommand{\Gce}{\Gls{gce}}
\newcommand{\apogee}{\gls{apogee}}

%%%%%%%%%%%%%%%%%%%%%% terms %%%%%%%%%%%%%%%
\newglossaryentry{metallicity}{name={metallicity},
    description={the (mass) fraction of a star or gas which is not made of either H or He. For the sun, the metallicity is $\Zo = 0.014$}
}

\newglossaryentry{yield}{name={yield},
    description={The net production of a new element during a star's lifecycle divided by the star's mass (including winds and supernovae ejecta). }
}

\newglossaryentry{nucleosynthesis}{name={nucleosynthesis},
    description={The synthesis of new elements through fusion inside stars. See section \ref{sec:stel_evo}.}
}

\newglossaryentry{multizone}{name={multi-zone},
    description={A chemical evolution model where a galaxy is divided into several zones, each with different stars, gas, and properties. }
}

\newglossaryentry{onezone}{name={one-zone},
    description={A chemical evolution model where the gas is all the same composition, i.e. neglecting spatial variations.}
}

\newglossaryentry{alpha}{name={$\alpha$-element},
    description={Elements which are made up of $\alpha$-particles (He nuclei) throuch the triple-$\alpha$ process (see Eq.~\ref{eq:triple_alpha}). Essentially light, even-numbered elements like O, Mg, and Na.}
}

\newglossaryentry{fepeak}{name={Fe-peak elements},
    description={Fe and nearby elements, produced in \gls{cc} and \gls{ia}.}
}

\newglossaryentry{sproc}{name={s-process elements},
    description={Elements produced through slow neutron captures, typically in \agb\ stars.}
}

\newglossaryentry{rgb}{name={red giant branch},
    description={Red giant branch stars are stars that have completed hydrogen core burning and have expanded in size. See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{tdu}{name={third dredge up},
    description={Third dredge up occurs inside \gls{agb} stars. During each thermal pulse, material is {\it dredged up} from the core, changing the chemical abundances of the stellar atmosphere. (While nominally called {\it third dredge up}, there are typically several third dredge ups.). See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{fdu}{name={first dredge up},
    description={First dredge up occurs when a \gls{lowmass} enters the \gls{rgb} phase. Material from the core is brought to the surface, increasing N and decreasing C abundances. See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{hbb}{name={hot bottom burning},
    description={Hot bottom burning occurs inside \agb\ stars. The base of the convective envelope becomes hot enough for CNO burning to initiate.}
}

\newglossaryentry{subgiant}{name={subgiant},
    description={A star in the process of leaving the main sequence and becoming a \gls{rgb}.}
}

\newglossaryentry{imfave}{name={\gls{imf}-averaged},
    description={Averaged over the initial-mass function ({\sc imf}). The {\imf}-weighted yield is the mass of the newly produced element divided by the mass of star formation for a single stellar population (see \ssp).}
}

\newglossaryentry{dtd}{name={delay time distribution},
    description={The distribution in time of when an element is produced 
    after a star formation event.}
}

\newglossaryentry{massloading}{name={mass loading},
    description={The strength of outflows relative to star formation. See also $\eta$. }
}

\newglossaryentry{lowalpha}{name={low-$\alpha$},
    description={The low-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
}

\newglossaryentry{highalpha}{name={high-$\alpha$},
    description={The high-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
}

\newglossaryentry{lowmass}{name={low-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$ which end life as a white dwarf. }
}

\newglossaryentry{highmass}{name={high-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$, which end as a neutron star, black hole, or supernovae.}
}

\newglossaryentry{insideout}{name={{\it insideout}},
    description={Our fiducial star formation history. The rate of star formation is highest towards the center of the galaxy and at earlier times. See Eq.~\ref{eq:inside_out}.}
}

\newglossaryentry{cno}{name={CNO cycle},
    description = { A proton-fusion cycle which occurs in \gls{rgb} stars 
        consisting of a chain of proton captures releasing a He nucleus ($\alpha$-particle) and energy. See Eq.~\ref{eq:cno_cycle}.
}
}

### 2025-03-13 thesis/introduction.tex
thesis documents

\subsection{Background \& Past Work}\label{background-past-work}

\begin{itemize}
\tightlist
\item
  What is dark matter? Why do we look at dwarfs?
\item
  Forms of dark matter, lambda-CDM, and dwarf galaxies
\item
  How does gravity affect dwarfs, theory of tidal perturbations

  \begin{itemize}
  \tightlist
  \item
    EN21, peñarrubia+09, etc.
  \end{itemize}
\item
  Instances of dwarfs undergoing weird processess
\item
  Alternative processes and uncertainties in the evolution of dwarfs
\end{itemize}

\subsection{Gaia Membership Selection}\label{gaia-membership-selection}

\begin{figure}
\centering
\includegraphics{figures/scl_selection.png}
\caption{Sculptor selection criteria.}\label{fig:sculptor_selection}
\end{figure}

We use J+24 data, updated PM from the MV 2020a catalogue.

J+24 select members using a multi-component Baysian algorithm.

\begin{itemize}
\tightlist
\item
  First, stars with poor Gaia astrometry, color excess nearby
  parallaxes, and proper motions \textgreater{} 10 mas/yr are removed
\item
  Stars are assigned a likelihood based on the location on the CMD
  (assuming)
\end{itemize}

Simple cuts, such as selecting PM, CMD based on purely geometric cuts
reveal a similar density profile.

One caveat is the original algorithm takes spatial position into
account. When deriving a density profile, this assumption may influence
the derived density profile, especially when the galaxy density is
fainter than the background of similar appearing stars. To remidy this
and estimate where the background begins to take over, we also explore a
cut based on the likelihood ratio of only the CMD and PM components.

\subsection{Searches for tidal tails}\label{searches-for-tidal-tails}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313132102118.png}
\caption{image-20250313132102118}
\end{figure}

\subsection{Density Profile Reliability and
Uncertainties}\label{density-profile-reliability-and-uncertainties}

Given a set of probable members, the density profile is given by
EQUATION, where we use \ldots{}

\begin{itemize}
\tightlist
\item
  Using J+24 data, we validate

  \begin{itemize}
  \tightlist
  \item
    Check that PSAT, magnitude, no-space do not affect density profile
    shape too significantly
  \end{itemize}
\item
  Our ``high quality'' members all have \textgreater{} 50 member stars
  and do not depend too highly on the spatial component, mostly
  corresponding to the classical dwarfs
\item
  We fit Sérsic profiles to each galaxy

  \begin{itemize}
  \tightlist
  \item
    The Sérsic index, \(n\), is a measure of the deviation from an
    exponential. Exponentials have \(n=1\), whereas more extended dwarf
    galaxies will have higher \(n\)
  \end{itemize}
\item
  To better estimate the uncertanties due to unknown galaxy properties
  and flexibility in the likelihood cut, we can
\end{itemize}

\begin{figure}
\centering
\includegraphics{figures/density_profiles_medley.png}
\caption{Density profiles}\label{fig:sculptor_observed_profiles}
\end{figure}

Dark Matter is one of the biggest open questions in modern astronomy.
Hello fig.~\ref{fig:sculptor}.

\subsection{Comparison of the Classical
dwarfs}\label{comparison-of-the-classical-dwarfs}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130050775.png}
\caption{image-20250313130050775}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130110550.png}
\caption{image-20250313130110550}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130043114.png}
\caption{image-20250313130043114}
\end{figure}

\subsection{Summary}\label{summary}

\begin{itemize}
\tightlist
\item
  Of the classical dwarfs, UMi \& Scl stand out statistically, with high
  \(n\) given their luminosity
\item
  Including fainter dwarf galaxies, Boo 3 and Boo 1 appear to also have
  extended density profiles

  \begin{itemize}
  \tightlist
  \item
    Deeper data would be required to robustly measure this
  \end{itemize}
\end{itemize}

### 2025-03-13 thesis/main.tex
thesis documents

 %! TeX program = xelatex

\DocumentMetadata{lang=en-CA}

\documentclass[12pt,oneside,letterpaper]{report}

\input{formatting}
\input{acronyms}
\usepackage[single]{accents}

% useful commands
% below are for statement
\newcommand{\schwa}{ə}
\newcommand{\eng}{\ng{}}
%\newcommand{\overcomma}[1]{${\rm \accentset{\mbox{\normalfont ,}}{#1}}$}
\newcommand{\Lekwnen}{L\schwa{}\overcomma{k}\textsuperscript{w}\schwa{}\ng{}\schwa{}n}
\newcommand{\skipline}{\vspace{\baselineskip}}

\def\overcomma#1{%
  \setbox0=\hbox{#1}% Save the base letter in box0.
  \dimen0=\wd0       % Get its width.
  \vbox{%
    \hbox to \dimen0{\hfil,\hfil}% Create an hbox of the same width with a centered comma.
    \nointerlineskip            % Prevent extra vertical space.
    \box0                       % Place the base letter.
  }%
}

% for pandoc
\newcommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\newcommand{\about}[1]{${\sim} #1$}

\title{Tidal to tell: \\
    Does the Milky Way affect Sculptor and Ursa Minor?
}
\author{Daniel A. Boyea}

\date{\today}

\begin{document}
% \linespread{1.25}

\pagenumbering{roman}

\makeatletter
\begin{titlepage}
   \begin{center}
       \textbf{\@title}\\
        \skipline

        by\\
        \skipline

       \@author\\
       B.Sc. The Ohio State University, 2023\\
       \vspace*{3\baselineskip}
    A Thesis Submitted in Partial Fulfillment of the Requirements for the Degree of\\
    \skipline

    MASTER OF SCIENCE \\
    \skipline

    in the Department of Physics and Astronomy\\
       \vfill
       ©\@author, 2025\\
       University of Victoria\\
       \vspace*{3\baselineskip}

       All rights reserved. This thesis may not be reproduced in whole or in part,
by photocopy or other means, without the permission of the author. \\
\skipline
   \end{center}
We acknowledge and respect the \Lekwnen{} (Songhees and X\textsuperscript{w}seps\schwa{}m/
Esquimalt) Peoples on whose territory the university stands, and the
\Lekwnen{} and ${\underaccent{\bar}{\rm W}}$S\'ANE\'C Peoples whose historical relationships with the
land continue to this day. 
\end{titlepage}

\addtocounter{page}{1}

\addcontentsline{toc}{chapter}{Supervisory Committee}
\begin{centering}
\textbf{\@title}
\skipline
by\\
\skipline
\@author\\
B.Sc. The Ohio State University, 2023\\
\vspace*{3\baselineskip}
\end{centering}

\subsection*{Supervisory Committee}

Dr. Julio Navarro, Supervisor\\
Department of Astronomy \\
\skipline

\noindent Dr. Kim Venn, Co-Supervisor\\
Department of Astronomy

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Scl and UMi are weird bois.

\newlength{\myw}
\newlength{\myl}

---\rule[2.4pt]{12pt}{0.6pt}
\settowidth{\myw}{---}\the\myw

{\bf ---}\rule[2.2pt]{12pt}{1.1pt}

\settoheight{\myl}{---}\the\myl

-\rule[2.4pt]{3pt}{0.75pt}
I\rule{1.2pt}{8pt}
{\bf I}\rule{2.0pt}{8pt}
|\rule[-2pt]{0.6pt}{12pt}
% \chapter*{Acknowledgements}
% \addcontentsline{toc}{chapter}{Acknowledgements}
% 
% 
% I hope that the people who helped me complete this project will appreciate how much they have helped me. This year has been an incredibly challenging journey---and perhaps I am not open enough about what it has been like. I undoubtedly would not have reached this day without everyone's help.
% 
% James Johnson---your mentorship and guidance throughout this project was critical. 
% You helped me learn an incredible amount about being an astronomer, and I am forever grateful for how you have guided this project, through late nights and last-minute comments.  Wayne, for everything you do for everyone---you are the glue holding together our department and nobody can say thank you enough. 
% David and Jennifer for helping this project move forward, by providing guidance at critical junctions and being part of my committee. Rolando, for being the non-astronomer thrown into this work, and pointing out the flaw of incomprehensibility in this work.
% 
% Even though you may have not been able to help me with the science, dear friends and family, you have been no less important in the completion of this work. 
% Eric, even across the country, you have always been a point of support and stability, and I am always happy to be around you. Anya, you are a wonderful friend and I cannot thank you enough for your unconditional support this year, 
% All my friends who are just wonderful, inspiring, and incredible people---Kaia, Maria, Autumn, Simon, Bailee, Aaliyah, Sanskruti, Alyssa, Harrison, Denis, Keith, and so many others. 
% And my family (including Arya, my most lovable doodle)---I can not even being to thank you enough for simply helping me recover my health and make it to graduation. 
% 
% 
%

%% Lists

\tableofcontents
\listoffigures
\listoftables

\chapter*{Symbols}

\setlength{\tabcolsep}{0pt}
\begin{longtable}{p{0.2\textwidth} p{0.8\textwidth}}
$\rho$ & 3-dimensional mass density. \\

$\Sigma$ & 2-dimensional density \\
\end{longtable}

\setlength{\tabcolsep}{6pt}
\renewcommand*{\arraystretch}{1}

\printglossary[type=\acronymtype,nonumberlist]
\printglossary

\newpage

\pagenumbering{arabic}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

\chapter{Introduction}
\input{introduction}

\the\textwidth

%%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%

\newpage
\bibliographystyle{aasjournal}
\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{main}

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

\appendix
% \chapter*{Appendix}
% \addcontentsline{toc}{chapter}{Appendix}
% \renewcommand{\thesection}{A.\arabic{section}}
% \renewcommand\thefigure{A.\arabic{figure}}    
% \renewcommand\theequation{A.\arabic{equation}}    
% \setcounter{figure}{0}
% \setcounter{equation}{0}
% 
% \renewcommand*{\theHsection}{A.\arabic{section}}%
% \renewcommand*{\theHsubsection}{A.\arabic{subsection}}%
% \renewcommand*{\theHfigure}{A.\arabic{figure}}%
% \renewcommand*{\theHequation}{A.\arabic{equation}}%

\chapter{Model parameters}\label{sec:jack}

⁄\end{document}

### 2025-03-13 thesis/compile.sh

@@ -1,0 +1,7 @@
+set -e 
+rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
+
+pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib"
+
+pandoc $pandocoptions -o introduction.tex introduction.md
+latexmk -xelatex -f main.tex

### 2025-03-13 thesis/sync.sh

@@ -1,0 +1,2 @@
+rsync -av ~/thesis/ .
+rsync -av ~/dwarfs_plots/thesis/plots/figures ~/thesis/figures

### 2025-03-14 thesis/introduction.tex
markdown filters

@@
-\subsection{Background \& Past Work}\label{background-past-work}
+\section{Background \& Past Work}\label{background-past-work}
 
@@
 
+\subsubsection{Sculptor}\label{sculptor}
+
+\begin{itemize}
+\tightlist
+\item
+  sestito+23a
+\item
+  tolstoy+22
+\item
+  battaglia+09
+\item
+  arroyo-polonio+22
+\end{itemize}
+
+\begin{table*}[t]
+\centering
+\caption{Measured properties of Sculptor}
+\label{tbl:Measured-properties-of-Sculptor}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $15.0183 \pm 0.0012$˚ & M+18\\
+$\delta$ & $-33.7186 \pm 0.00072$˚ & M+18\\
+distance & $83.2 \pm 2$ kpc & Tran+22\\
+$\mu_\alpha \cos \delta$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
+RV & $111.03 \pm 0.23$ & This work\\
+$\sigma_v$ & $9.61\pm0.16$ & This work\\
+$r_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
+ell & $0.36 \pm 0.01$ & M+18\\
+PA & $92\pm1$ & M+18\\
+$M_V$ & $-10.82\pm0.14$ & M+18\\
+$\Upsilon_\star$ & $1.5 \pm 0.3$ & assumed\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
 \subsection{Gaia Membership Selection}\label{gaia-membership-selection}
@@
 \includegraphics{figures/scl_selection.png}
-\caption{Sculptor selection criteria.}\label{fig:sculptor_selection}
+\caption{Sculptor selection criteria. This will be a super long caption
+to test the figure list.}\label{fig:sculptor_selection}
 \end{figure}
 
-We use J+24 data, updated PM from the MV 2020a catalogue.
+We use J+24 data. J+24 select members using a multi-component Baysian
+algorithm:
 
-J+24 select members using a multi-component Baysian algorithm.
+\begin{itemize}
+\item
+  Remove stars with poor astrometry or photometry, no colour excess
+  (\citet{lindegren+2018} equation C.2), 3\(\sigma\) consistency of
+  measured parallax with dwarf distance (near zero with
+  \citet{lindegren+2018} zero-point correction), and absolute RA and Dec
+  proper motions less than 10\(\,{\rm mas\ yr^{-1}}\).
+\item
+  Spatial likelihood based on a double exponential
+  \(\Sigma_\star \propto e^{-r/r_s} + B\,e^{-r/r_{\rm outer}}\) where
+  the inner scale radius is fixed.
+\item
+  Stars are assigned a likelihood based on the location on the CMD
+  (using Padova isochrones including an intrinsic 0.1 CMD width in
+  colour convolved with colour and distance modulus)
+\item
+  Background KDE density maps for the CMD and PM are constructed using
+  the other quality-selected stars outside of \(5R_h\), where the
+  satellite density would be orders of magnitude less than the
+  background (even in the presence of extended tidal features).
+\item
+  Likelihoods normalized to unity to represnt a PDF
+
+  Membership probabilities are then given by
+\end{itemize}
+
+\[
+P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}} = \frac{1}{1 + \frac{(1-f_{\rm sat}){\cal L}_{\rm bg}}{f_{\rm sat}{\cal L}_{\rm sat} }}
+\]
+
+where \(f_{\rm sat}\) is the fraction of stars belonging to the system
+inside the given field, \({\cal L}_{\rm sat}\) is the likelihood of a
+star belonging to the satellite, and \({\cal L}_{\rm bg}\) is the
+likelihood of the star belonging to the background. Each likelihood is
+calculated as a product of the CMD, PM, and spatial likelihoods: \[
+{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}
+\]
+
+Not shown here, we explore simple cuts of the stars, using absolute cuts
+in parallax, proper motions, and the CMD. The results are similar to the
+nospace model.
+
+The above formula suggests that a cut in \(P_{\rm sat}\) is equivalent
+to the cut in likelihoods \[
+\frac{{\cal L}_{\rm sat}}{{\cal L}_{\rm bg}} > \frac{(1-f_{\rm sat})/f_{\rm sat}}{1/P_{\rm sat}- 1}
+\]
+
+Note that if we remove the spatial component of the likelihood, then
+\(f_{\rm sat}\) represents a global normalization
+
+\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}
+
+\begin{figure}
+\centering
+\includegraphics{figures/scl_tidal_tails.png}
+\caption{Density profiles}\label{fig:sculptor_tidal_tails}
+\end{figure}
+
+\begin{itemize}
+\item
+  The above figure shows the distribution of member stars (orange), PM
+  \& CMD only selected stars (blue) and all stars (passing quality cuts,
+  black)
+\item
+  There are no apparent overdensities in the PM \& CMD only selected
+  stars to suggest the presence of a tidal tail
+\item
+  This means that at least at the level of where the background density
+  dominates, we can exclude models which produce tidal tails brighter
+  than a density of
+  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
+  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
+\end{itemize}
+
+\subsection{Density Profile Reliability and
+Uncertainties}\label{density-profile-reliability-and-uncertainties}
 
@@
 \item
-  First, stars with poor Gaia astrometry, color excess nearby
-  parallaxes, and proper motions \textgreater{} 10 mas/yr are removed
+  How well do we know the density profiles?
 \item
-  Stars are assigned a likelihood based on the location on the CMD
-  (assuming)
+  What uncertainties affect derived density profiles?
+\item
+  Can we determine if Gaia, structural, or algorithmic systematics
+  introduce important errors in derived density profiles?
 \end{itemize}
 
-Simple cuts, such as selecting PM, CMD based on purely geometric cuts
-reveal a similar density profile.
+J+24's algorithm takes spatial position into account, assuming either a
+one or two component exponential density profile. When deriving a
+density profile, this assumption may influence the derived density
+profile, especially when the galaxy density is fainter than the
+background of similar appearing stars. To remedy this and estimate where
+the background begins to take over, we also explore a cut based on the
+likelihood ratio of only the CMD and PM components. This is in essence
+assuming that the spatial position of a star contains no information on
+it's membership probability (a uniform distribution like the background)
 
-One caveat is the original algorithm takes spatial position into
-account. When deriving a density profile, this assumption may influence
-the derived density profile, especially when the galaxy density is
-fainter than the background of similar appearing stars. To remidy this
-and estimate where the background begins to take over, we also explore a
-cut based on the likelihood ratio of only the CMD and PM components.
+To incorporate the structural uncertainties and robustly model the
+sampling uncertainty, we construct the following bootstrap model
 
-\subsection{Searches for tidal tails}\label{searches-for-tidal-tails}
+\begin{itemize}
+\item
+  Centre is varied by a centring error, estimated from the standard
+  normal error of the positions plus the systematic shift of the mean
+\item
+  Position angle and ellipticity are sampled from a normal distribution
+  given the reported uncertainties
+\item
+  \(f_{\rm sat}\) is sampled from \ldots{}
+
+  \textbf{TODO}: Look into normalization of Likelihoods and check how
+  \(f_{\rm sat}\) matters. If \(f_{\rm sat}\) is related to the
+  normalizations of fg / bg densities, and other likelihoods are
+  area-normalized to 1, then this makes life much easier.
+
+  Test if psat weighted density profiles are similar
+
+  Save MC density profile outputs
+\end{itemize}
 
@@
 \centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313132102118.png}
-\caption{image-20250313132102118}
+\includegraphics{figures/scl_density_methods.png}
+\caption[Density profiles]{add simple cuts \& delve to convince
+more\ldots{}}\label{fig:sculptor_observed_profiles}
 \end{figure}
 
-\subsection{Density Profile Reliability and
-Uncertainties}\label{density-profile-reliability-and-uncertainties}
-
-Given a set of probable members, the density profile is given by
-EQUATION, where we use \ldots{}
+\section{Comparison of the Classical
+dwarfs}\label{comparison-of-the-classical-dwarfs}
 
@@
 \item
-  To better estimate the uncertanties due to unknown galaxy properties
+  To better estimate the uncertainties due to unknown galaxy properties
   and flexibility in the likelihood cut, we can
 \end{itemize}
-
-\begin{figure}
-\centering
-\includegraphics{figures/density_profiles_medley.png}
-\caption{Density profiles}\label{fig:sculptor_observed_profiles}
-\end{figure}
-
-Dark Matter is one of the biggest open questions in modern astronomy.
-Hello fig.~\ref{fig:sculptor}.
-
-\subsection{Comparison of the Classical
-dwarfs}\label{comparison-of-the-classical-dwarfs}
 
@@
 
+\subsection{Radial Velocity
+Measurements}\label{radial-velocity-measurements}
+
 \subsection{Summary}\label{summary}
@@
 \end{itemize}
+
+\textbf{TODO}: Use updated density profiles (nospace) with uncertainties
+included to MCMC fit Sérsic profiles to every dwarf galaxy.

### 2025-03-14 thesis/compile.sh

@@ -1,7 +1,7 @@
 set -e 
 rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
-pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib"
+pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib --filter ./table.py --filter ./md_figure.py"
 
 pandoc $pandocoptions -o introduction.tex introduction.md
-latexmk -xelatex -f main.tex
+# latexmk -xelatex -f main.tex

### 2025-03-14 thesis/md_figure.py

@@ -1,0 +1,58 @@
+from panflute import *
+import sys
+
+def process_figures(doc):
+    blocks = list(doc.content)
+    i = 0
+    while i < len(blocks):
+        if isinstance(blocks[i], Figure):
+            print("Figure found!", file=sys.stderr) 
+            if i + 1 < len(blocks) and isinstance(blocks[i+1], Para):
+                next_para = blocks[i+1]
+                if next_para.content:
+                    first_part = next_para.content[0]
+                    if isinstance(first_part, Str) and first_part.text == "Figure:" and isinstance(next_para.content[1], Space):
+
+                        print("Figure caption found!", file=sys.stderr) 
+                         # Get original caption from the Figure
+                        original_caption = blocks[i].caption
+                        print(original_caption.content, file=sys.stderr)
+                        print(original_caption.content[0], file=sys.stderr)
+                        print(type(original_caption.content[0]), file=sys.stderr)
+                        # Create new caption: [original] + LineBreak + [new content]
+                        inl = original_caption.content[0].content
+
+                        new_caption = Caption(
+                                Plain(*next_para.content[2:]),
+                                short_caption = [x for x in inl]
+                                )
+                        blocks[i].caption = new_caption
+                        
+                        # Remove the caption paragraph
+                        del blocks[i+1]
+
+        i += 1
+    doc.content = blocks
+
+def action(elem, doc):
+    if doc.format != 'latex':
+        return None
+
+    return None
+
+
+def prepare(doc):
+    pass
+
+def finalize(doc):
+    pass
+
+def main(doc=None):
+    return run_filter(action,
+        prepare=prepare, 
+        finalize=process_figures,
+        doc=doc, 
+    )
+
+if __name__ == "__main__":
+    main()

### 2025-03-14 thesis/sync.sh

@@ -1,2 +1,2 @@
 rsync -av ~/thesis/ .
-rsync -av ~/dwarfs_plots/thesis/plots/figures ~/thesis/figures
+rsync -av ~/dwarfs_figures/thesis/plots/figures ~/thesis/

### 2025-03-14 thesis/table.py

@@ -1,0 +1,90 @@
+# from discussion https://gist.github.com/wgroeneveld/9dbeb0d0b60c6cb5d8dfe9b938c5e94e
+import sys
+
+from panflute import *
+
+def replace_longtables_with_tabular(elem, doc):
+    try:
+        def get_text(item):
+            if isinstance(item, Str):
+                return item.text.replace('_', r'\_')
+            elif isinstance(item, Plain):
+                return ''.join([get_text(i) for i in item.content])
+            elif isinstance(item, ListContainer):
+                return ''.join([get_text(i) for i in item])
+            elif isinstance(item, Code):
+                return '\\verb|' + (item.text) + '|'
+            elif isinstance(item, Math):
+                return '$' + (item.text) + '$'
+            elif isinstance(item, RawInline):
+                return item.text
+            elif isinstance(item, Cite):
+                return ''.join(stringify(elem) for elem in item.content)
+            elif str(item) == 'Space':
+                return ' '
+            else:
+                return str(item)
+
+        def caption():
+            if elem.caption and elem.caption.content:
+                return '\\caption{' + get_text(elem.caption.content) + '}\n' + \
+                    '\\label{tbl:' + get_text(elem.caption.content).replace(' ', r'-').translate(str.maketrans("", "", "();,.")) + '}\n'
+            return ''
+
+        def tabular():
+            return '\\begin{tabular}{' + 'l' * elem.cols + '}\n\\toprule\n'
+
+        def headers():
+            if elem.head and elem.head.content:
+                return ' & '.join([get_text(cell.content) 
+                                   for cell in elem.head.content[0].content]
+                                  ) + '\\\\\n\\midrule\n'
+            return ''
+
+        def items():
+            rows = []
+            for body in elem.content:
+                for row in body.content:
+                    rows.append(' & '.join([get_text(cell.content) for cell in row.content]) + '\\\\')
+            return '\n'.join(rows) + '\n'
+
+        result = '\\begin{table*}[t]\n\\centering\n' + \
+                 caption() + \
+                 tabular() + \
+                 headers() + \
+                 items() + \
+                 '\\bottomrule\n\\end{tabular}\n\\end{table*}'
+
+        print("Table processed successfully", file=sys.stderr)
+        return RawBlock(result, 'latex')
+    except Exception as e:
+        print(f"Error processing table: {str(e)}", file=sys.stderr)
+        return elem
+
+
+def prepare(doc):
[…]

### 2025-03-21 thesis/introduction.tex
thesis updates

@@
   \item
-    EN21, peñarrubia+09, etc.
+    \citet{EN2021}, \citet{PNM2008}, etc.
   \end{itemize}
@@
 \item
-  sestito+23a
+  \citet{sestito+2023a}
 \item
-  tolstoy+22
+  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
+  \citet{arroyo-polonio+2024}
 \item
-  battaglia+09
+  \citet{battaglia+2008}
 \item
-  arroyo-polonio+22
+  \citet{iorio+2019}
 \end{itemize}
@@
 \centering
-\includegraphics{figures/scl_selection.png}
-\caption{Sculptor selection criteria. This will be a super long caption
-to test the figure list.}\label{fig:sculptor_selection}
+\includegraphics{figures/scl_selection.pdf}
+\caption[Sculptor selection criteria]{The selection criteria for Scl
+members. Members are red and all field stars (satisfying quality
+criteria) are in light grey. \textbf{Top left:} tangent plane.
+\textbf{Top Right:} Colour magnitude
+diagram.}\label{fig:sculptor_selection}
 \end{figure}
@@
 
-Not shown here, we explore simple cuts of the stars, using absolute cuts
-in parallax, proper motions, and the CMD. The results are similar to the
-nospace model.
-
 The above formula suggests that a cut in \(P_{\rm sat}\) is equivalent
@@
 Note that if we remove the spatial component of the likelihood, then
-\(f_{\rm sat}\) represents a global normalization
+\(f_{\rm sat}\) represents a global normalization.
+
+Not shown here, we explore simple cuts of the stars, using absolute cuts
+in parallax, proper motions, and the CMD. The results are similar to the
+nospace model.
 
@@
 \centering
-\includegraphics{figures/scl_tidal_tails.png}
-\caption{Density profiles}\label{fig:sculptor_tidal_tails}
+\includegraphics{figures/scl_tidal_tails.pdf}
+\caption[Tidal tails]{The distribution of member stars (orange), PM \&
+CMD only selected stars (blue) and all stars (passing quality cuts,
+black).}\label{fig:sculptor_tidal_tails}
 \end{figure}
@@
 \begin{itemize}
-\item
-  The above figure shows the distribution of member stars (orange), PM
-  \& CMD only selected stars (blue) and all stars (passing quality cuts,
-  black)
 \item
@@
 \centering
-\includegraphics{figures/scl_density_methods.png}
-\caption[Density profiles]{add simple cuts \& delve to convince
-more\ldots{}}\label{fig:sculptor_observed_profiles}
+\includegraphics{figures/scl_density_methods.pdf}
+\caption{Density profiles}\label{fig:sculptor_observed_profiles}
 \end{figure}
+
+Figure:
 
@@
 \centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130050775.png}
-\caption{image-20250313130050775}
+\includegraphics{figures/classical_dwarf_profiles.pdf}
+\caption[Sculptor and UMi versus classical dwarfs]{Density profiles for
+each dwarf galaxy.}
+\end{figure}
+
+\textbf{TODO}: Use updated density profiles (nospace) with uncertainties
+included to MCMC fit Sérsic profiles to every dwarf galaxy.
+
+\begin{table*}[t]
+\centering
+\caption{Sérsic profile fits and basic structural parameters for classical and bright dwarfs around the MW.}
+\label{tbl:Sérsic-profile-fits-and-basic-structural-parameters-for-classical-and-bright-dwarfs-around-the-MW}
+\begin{tabular}{llllll}
+\toprule
+galaxy & Ell. & Sersic Rh & sersic n & Sersic mass & \\
+\midrule
+Fornax &  &  &  &  & \\
+Leo I &  &  &  &  & \\
+Sculptor &  &  &  &  & \\
+Leo II &  &  &  &  & \\
+Carina &  &  &  &  & \\
+Sextans I &  &  &  &  & \\
+Ursa Minor &  &  &  &  & \\
+Draco &  &  &  &  & \\
+Canes Venatici I &  &  &  &  & \\
+Crater II &  &  &  &  & \\
+Bootes I &  &  &  &  & \\
+Bootes III &  &  &  &  & \\
+Antlia II &  &  &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+Cannot resolve
+
+\begin{itemize}
+\tightlist
+\item
+  reticulum II yyy
+\end{itemize}
+
+\subsection{Radial Velocity
+Measurements}\label{radial-velocity-measurements}
+
+\section{Simulation-based
+motivations}\label{simulation-based-motivations}
+
+To motivate why a tidal interaction may give rise to the observed
+density profiles, we create a toy simulation following \citet{PNM2008}.
+
+\begin{itemize}
+\item
+  NFW initial conditions (sculptor like, vcm, rcm)
+\item
+  Evolved in x-y plane using \citet{EP2020} potential for
+  \textasciitilde{} 5Gyr with pericentre of 15 kpc and apocentre of 100
+  kpc.
+\item
+  Exponential initial stellar profile.
+\end{itemize}
+
+As a dark matter halo is perturbed on a pericentric passage with the
+milky way,
+
+\begin{itemize}
+\tightlist
+\item
+  Tidal stress heats halo slightly
+\item
+  Mass loss, particularly of loosely bound particles
+\end{itemize}
+
+The stellar component tracers will similarly follow the behaviour of the
+dark matter.
+
+An emperical estimate of where the simulation's stars are becoming
+unbound is, as stated in \citet{PNM2008}, the break radius \[
+R_b = C\,\sigma_{v}\,\Delta t
+\] where \(\sigma_v\) is the present line of sight velocity dispersion ,
+\(\delta t\) is the time since pericentre, and \(C \approx 0.55\) is a
+fit. The idea motivating this equation is stars in the inner regions
+will have dynamically equilibriated to the new potential (phase mixed),
+however the outer regions are no longer in steady state, so we have to
+wait until the crossing time reaches them as well.
+
+As illustrated in fig.~\ref{fig:toy_profiles}, the density profile
+initially stars off exponential. At increasing times since the first
+pericentric passage, the break radius, appearing as an apparent
+separation between the slopes of the inner and outer profile, increases.
+
+\begin{figure}
+\centering
+\includegraphics{figures/scl_umi_vs_idealized.pdf}
+\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
+profiles are well-matched to an idealized
+simulation}\label{fig:toy_profiles}
 \end{figure}
@@
 \centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130110550.png}
-\caption{image-20250313130110550}
+\includegraphics{figures/idealized_break_radius.pdf}
+\caption[Break radius validation]{The break radius of the simulations is
+set by}\label{fig:idealized_break_radius}
 \end{figure}
 
-\begin{figure}
-\centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130043114.png}
-\caption{image-20250313130043114}
-\end{figure}
+From this argument, we note that the following properties must be
+approximately true for tides to occur:
 
-\subsection{Radial Velocity
-Measurements}\label{radial-velocity-measurements}
+\begin{itemize}
+\tightlist
+\item
+  Close enough pericentre. The other break radius \(r_J\) implies that
+  if the host density is 3x the satellite, stars will be lost
+\item
+  Corresponding time since last pericentre: If the time since last
+  pericentre is not \(\sim\) consistent with an observed break in the
+  density profile, then tides
+\item
+  Halo evolution. As found in \citet{EN2021}, galaxies evolve along well
+  defined tidal tracks (assuming spherical, isotropic, NFW halo, which
+  may not be true, see \ldots). These tracks tend to ``puff up'' the
+  stellar component while also removing dark matter mass, leaving a
+  smaller, compacter DM halo with a more extended stellar component.
 
-\subsection{Summary}\label{summary}
+  \begin{itemize}
+  \tightlist
+  \item
+    This information is mostly related to the statistical initial
+    distribution of satellites from cosmology {[}ludlow+2016;
+    \citet{fattahi+2018}{]}
+  \end{itemize}
+\end{itemize}
+
+\section{Summary}\label{summary}
 
@@
 \end{itemize}
-
-\textbf{TODO}: Use updated density profiles (nospace) with uncertainties
-included to MCMC fit Sérsic profiles to every dwarf galaxy.

### 2025-03-21 thesis/main.tex
thesis updates

@@
 
-\title{Tidal to tell: \\
+\title{From tides to tails: \\
     Does the Milky Way affect Sculptor and Ursa Minor?
@@
    \begin{center}
-       \textbf{\@title}\\
+       \textbf{\large \@title}\\
         \skipline
@@
        \vfill
+       {\small
        ©\@author, 2025\\
        University of Victoria\\
-       \vspace*{3\baselineskip}
-
+   }
+   \skipline
+       {\small
        All rights reserved. This thesis may not be reproduced in whole or in part,
-by photocopy or other means, without the permission of the author. \\
+   by photocopy or other means, without the permission of the author.} \\
 \skipline
@@
 % context
-Scl and UMi are weird bois.
+\input{abstract.tex}
 
-\newlength{\myw}
-\newlength{\myl}
-
----\rule[2.4pt]{12pt}{0.6pt}
-\settowidth{\myw}{---}\the\myw
-
-{\bf ---}\rule[2.2pt]{12pt}{1.1pt}
-
-\settoheight{\myl}{---}\the\myl
-
--\rule[2.4pt]{3pt}{0.75pt}
-I\rule{1.2pt}{8pt}
-{\bf I}\rule{2.0pt}{8pt}
-|\rule[-2pt]{0.6pt}{12pt}
 % \chapter*{Acknowledgements}
 % \addcontentsline{toc}{chapter}{Acknowledgements}
-% 
-% 
-% I hope that the people who helped me complete this project will appreciate how much they have helped me.
-This year has been an incredibly challenging journey---and perhaps I am not open enough about what it has been like.
-I undoubtedly would not have reached this day without everyone's help.
-% 
-% James Johnson---your mentorship and guidance throughout this project was critical. 
-% You helped me learn an incredible amount about being an astronomer, and I am forever grateful for how you have guided this project, through late nights and last-minute comments.
-Wayne, for everything you do for everyone---you are the glue holding together our department and nobody can say thank you enough.
-% David and Jennifer for helping this project move forward, by providing guidance at critical junctions and being part of my committee.
-Rolando, for being the non-astronomer thrown into this work, and pointing out the flaw of incomprehensibility in this work.
-% 
-% Even though you may have not been able to help me with the science, dear friends and family, you have been no less important in the completion of this work. 
-% Eric, even across the country, you have always been a point of support and stability, and I am always happy to be around you.
-Anya, you are a wonderful friend and I cannot thank you enough for your unconditional support this year,
-% All my friends who are just wonderful, inspiring, and incredible people---Kaia, Maria, Autumn, Simon, Bailee, Aaliyah, Sanskruti, Alyssa, Harrison, Denis, Keith, and so many others. 
-% And my family (including Arya, my most lovable doodle)---I can not even being to thank you enough for simply helping me recover my health and make it to graduation. 
-% 
-% 
 %
@@
 
-\chapter*{Symbols}
-
-\setlength{\tabcolsep}{0pt}
-\begin{longtable}{p{0.2\textwidth} p{0.8\textwidth}}
-$\rho$ & 3-dimensional mass density. \\
-
-$\Sigma$ & 2-dimensional density \\
-\end{longtable}
-
-\setlength{\tabcolsep}{6pt}
-\renewcommand*{\arraystretch}{1}
-
-\printglossary[type=\acronymtype,nonumberlist]
-\printglossary
-
+%\chapter*{Symbols}
+%
+%
+%\setlength{\tabcolsep}{0pt}
+%\begin{longtable}{p{0.2\textwidth} p{0.8\textwidth}}
+%$\rho$ & 3-dimensional mass density. \\
+%
+%$\Sigma$ & 2-dimensional density \\
+%\end{longtable}
+%
+%\setlength{\tabcolsep}{6pt}
+%\renewcommand*{\arraystretch}{1}
+%
+%
+%\printglossary[type=\acronymtype,nonumberlist]
+%\printglossary
+%
+%
 \newpage
@@
 \input{introduction}
-
-\the\textwidth
 
@@
 
+\chapter{Font properties}
+
+textwidth: \the\textwidth
+
+\newlength{\myw}
+\newlength{\myl}
+
+---\rule[2.4pt]{12pt}{0.6pt}
+\settowidth{\myw}{---}\the\myw
+
+{\bf ---}\rule[2.2pt]{12pt}{1.1pt}
+
+\settoheight{\myl}{---}\the\myl
+
+-\rule[2.4pt]{3pt}{0.75pt}
+I\rule{1.2pt}{8pt}
+{\bf I}\rule{2.0pt}{8pt}
+|\rule[-2pt]{0.6pt}{12pt}
+
 ⁄\end{document}

### 2025-03-21 thesis/compile.sh

@@ -3,5 +3,6 @@ rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
 pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib --filter ./table.py --filter ./md_figure.py"
 
+pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
-# latexmk -xelatex -f main.tex
+latexmk -xelatex -f main.tex

### 2025-03-21 thesis/md_figure.py

@@ -1,5 +1,20 @@
 from panflute import *
 import sys
+from os import path
+
+def rename_url(blocks, i):
+    content = blocks[i].content
+    for j in range(len(content)):
+        if isinstance(content[j], Plain):
+            for con2 in content[j].content:
+                if isinstance(con2, Image):
+                    url = con2.url
+                    print("Found image at ", url, file=sys.stderr)
+                    if url.startswith("figures"):
+                        url_new = path.splitext(url)[0] + ".pdf"
+                        con2.url = url_new
+                        print("Renaming ", url,"=>", con2.url, file=sys.stderr)
+
 
 def process_figures(doc):
     blocks = list(doc.content)
@@ -7,18 +22,21 @@ def process_figures(doc):
     while i < len(blocks):
         if isinstance(blocks[i], Figure):
             print("Figure found!", file=sys.stderr) 
+            url = rename_url(blocks, i)
+
             if i + 1 < len(blocks) and isinstance(blocks[i+1], Para):
                 next_para = blocks[i+1]
                 if next_para.content:
                     first_part = next_para.content[0]
-                    if isinstance(first_part, Str) and first_part.text == "Figure:" and isinstance(next_para.content[1], Space):
+                    if (len(next_para.content) > 2 
+                            and isinstance(first_part, Str) 
+                            and first_part.text == "Figure:"
+                            and isinstance(next_para.content[1], Space)
+                        ):
 
                         print("Figure caption found!", file=sys.stderr) 
                          # Get original caption from the Figure
                         original_caption = blocks[i].caption
-                        print(original_caption.content, file=sys.stderr)
-                        print(original_caption.content[0], file=sys.stderr)
-                        print(type(original_caption.content[0]), file=sys.stderr)
                         # Create new caption: [original] + LineBreak + [new content]
                         inl = original_caption.content[0].content
 

### 2025-04-07 thesis/data.tex
thesis updates (most of data chapter)

\section{Gaia Membership Selection}\label{gaia-membership-selection}

Gaia provides unprecedented accuracy in proper motions and magnitudes.
As such, Gaia data is uniquely excellent to produce low-contamination
samples of likely member stars belonging to satellites. Here, we breifly
describe J+24's membership estimation and discuss how this informs our
observational knoledge of each galaxies density profile. In general,
J+24 use a Bayesian framework incorporating proper motion (PM,
colour-magnitude diagram (CMD), and spatial information to determine the
probability that a given star belongs to the satellite. J+24 extends
\citet{MV2020a} (see also \citet{pace+li2019}, etc.).

J+24 select stars initially from Gaia satisfying:

\begin{itemize}
\tightlist
\item
  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3})
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance +
  uncertainty (near zero with \citet{lindegren+2018} zero-point
  correction)
\item
  Absolute RA and Dec proper motions less than
  10\(\,{\rm mas\ yr^{-1}}\)
\item
  Good photometry
\item
  No colour excess (\citet{lindegren+2018} equation C.2)
\item
  G \textgreater{} 22 and less than 5\(\sigma\) above TRGB, and between
  -0.5 and 2.5 in Bp - Rp.
\end{itemize}

J+24 calculate the probability that any star belongs to either the
satellite or the MW background as \[
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\]

where the satellite (sat) and background (bg) likelihoods are simply the
product of the PM, CMD, and spatial components: \[
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}
\]

The satellite likelihood is constructed as

\begin{itemize}
\tightlist
\item
  CMD: The CMD is from Padova \citep{girardi+2002} isochrone at 12 Gyr
  with a colour width of 0.1 mag plus the Gaia colour uncertanty and the
  observed metallicity of the dwarf. The HB is modelled as a constant
  magnitude extending blue of the CMD with a 0.1 mag width plus the mean
  colour error. A likelihood map is constructed by sampling the distance
  modulus in addition to the CMD width, taking the maximum of RGB and HB
  likelihoods.
\item
  Spatial: A single exponential
  (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)) using mean values of the
  structural parameters (\textbf{verify}). For Scl and UMi, this is
  instead a double exponential
  \(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\) where
  the inner exponential remains fixed and structural parameters are not
  accounted for. Structural parameter uncertainties are not accounted
  for.
\item
  PM. A bivariate gaussian with variance and covariance equal to each
  star's proper motions. Each star's proper motions uncertainty are
  assumed to be the dominant uncertainty.
\end{itemize}

The background likelihood is constructed as:

\begin{itemize}
\tightlist
\item
  CMD : Constructed as a KDE using the other quality-selected stars
  outside of \(5R_h\) in the catalogue. The width of each star is
  projected as a bivariate Gaussian.
\item
  PM: same as CMD except in PM space.
\item
  Spatial: a constant likelihood.
\end{itemize}

Note that each likelihood map is normalized over the respective
parameter space. In order to represent the difference in frequency of
background and forground stars, \(f_{\rm sat}\) represents the field
fraction of member stars.

In J+24, a MCMC simulation is ran using the above likelihood to solve
for the following parameters

\begin{itemize}
\tightlist
\item
  Systemic proper motions \(\mu_\alpha\), \(\mu_\delta\). Prior: within
  5\(\sigma\) of single component case w/ systematic uncertainties.
  Single component prior is based on MV2020 (RV members / expected
  velocity dispersion of halo at distance?)
\item
  \(f_{\rm sat}\) density normalization (uniform between 0 and 1)
\item
  Spatial component parameters \(B\) and \(R_{\rm outer}\) for extended
  profiles (Scl and UMi here.)
\end{itemize}

The median parameters from the simulation are then used to calculate the
final \(P_{\rm sat}\) values we use here.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to 0 or 1, so the
choice of probability threshhold is not too significant. Additionally,
even for a probability cut of 0.2, the purity of the resulting sample
with RV measurements is very high (\textasciitilde90\%, J+24). (Note:
there is likely a high systematic bias in using stars with RV
measurements to measure purity. Fainter stars have poorer measurements
and distant stars are less likely to have been targeted. )

\subsection{Resulting Samples}\label{resulting-samples}

\begin{figure}
\centering
\includegraphics{figures/scl_selection.pdf}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Probable members (2-component) are orange, and all field stars
(satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
plane. \textbf{Bottom left:} Colour magnitude diagram. \textbf{Bottom
right:} Proper motion.}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/umi_selection.pdf}
\caption[Ursa Minor Selection]{Similar to
fig.~\ref{fig:sculptor_selection} except for Ursa Minor. UMi features a
very extended density profile with some stars \textasciitilde{} 6\(R_h\)
including a RV member. UMi is also highly elliptical compared to other
classical dwarfs.}\label{fig:umi_selection}
\end{figure}

Fornax(figures/fornax\_selection.png)\{\#fig:fornax\_selection\}

Figure: Similar to fig.~\ref{fig:sculptor_selection} except for Fornax.
Note that we only use one-component probabilities here. Fornax is the
quintessential unperturbed dwarf galaxy.

\subsection{Density Profiles}\label{density-profiles}

Our primary observational constraint is the density profile of a dwarf
galaxy.

To derive density profiles, we use 0.05 dex bins in log radius (i.e.~the
bins are derived from 10\^{}(minimum(logR):0.05:maximum(logR))). The
density in each bin is then (from Poisson statistics) \[
\Sigma_b = N_{\rm memb} / A_{\rm bin} \pm \sqrt{N_{\rm memb}} / A_{\rm bin}
\] where \(N_{\rm memb}\) is the number of members in the bin and
\(A_{\rm bin}\) is the area of the bin. As discussed below, these
uncertainties underrepresent the true uncertainty on multiple accounts.
We retain Poisson errors for simplicity here.

In Figures fig.~\ref{fig:scl_observed_profiles},
fig.~\ref{fig:umi_observed_profiles},
\citet{fig_fornax_observed_profiles}, we show the derived density
profiles for each galaxy for samples similar to in the selection plots
above. In each case, all samples are the same towards the inner regions
of the satellite, illustrating that these density profiles are dominated
by satellite stars in the centre. The sample containing all stars
reaches a plateau at the total background in the field. However,
restricting stars to being most likely satellite members by CMD + PM,
the background is much lower. This plateau likely represents the real
background of background stars which could be mistaken as members.
Finally, we have the probable members and bg-subtracted densities. BG
subtracted is based on the \texttt{all} density profile, subtracting the
mean background density for stars beyond the last point of the BG
subtracted profile.

Note that the probable members (fiducial) density profile continues to
confidently estimate the density profile below the CMD+PM likely star
background. These points are likely unreliable (see discussion below).
However, before this point, both the BG subtracted and probable members
density profiles are strikingly similar. Assumptions about the details
of the likelihood and spatial dependence have marginal influence on the
resulting density profile when the satellite is higher density than the
background.

\begin{figure}
\centering
\includegraphics{figures/scl_density_methods.pdf}
\caption[Sculptor density profiles]{The density profile of Sculptor for
different selection criteria. \emph{probable members} selects stars with
PSAT \textgreater{} 0.2 considering PM, CMD, and spatial, \emph{CMD+PM}
select stars more likely to be members according to CMD and PM only,
\emph{all} selects any high quality star, and \emph{BG subtracted} is
the background-subtracted density derived from high-quality
stars.}\label{fig:scl_observed_profiles}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/umi_density_methods.pdf}
\caption[Ursa Minor density profiles]{Similar to
fig.~\ref{fig:scl_observed_profiles} except for Ursa
Minor.}\label{fig:umi_observed_profiles}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/fornax_density_methods.pdf}
\caption[Fornax density profiles]{Similar to
fig.~\ref{fig:scl_observed_profiles} except for
Fornax.}\label{fig:fornax_observed_profiles}
\end{figure}

\section{Comparison of the Classical
dwarfs}\label{comparison-of-the-classical-dwarfs}

Classical dwarfs are often the brightest dwarfs in the sky.c The density
profiles of classical dwarf galaxies is thus well measured, enabling
detailed comparisons.

\begin{table*}[t]
\centering
\begin{tabular}{lll}
\toprule
galaxy & R\_h (exp inner 3Rs) & num cand\\
\midrule
Fornax & $17.8\pm0.6$ & 23,154\\
Leo I & $3.7\pm0.2$ & 1,242\\
Sculptor & $9.4\pm0.3$ & 6,888\\
Leo II & $2.4\pm0.3$ & 347\\
Carina & $8.7\pm0.4$ & 2,389\\
Sextans I & $20.2 \pm0.9$ & 1,830\\
Ursa Minor & $11.7 \pm 0.5$ & 2,122\\
Draco & $7.3\pm0.3$ & 1,781\\
\bottomrule
\end{tabular}
\end{table*}

Using the same methods above, we select members from J+24's stellar
probabilities. We use the one-component exponential density profiles
with

Figure: Density profiles for each dwarf galaxy. Here, we use the
1-component exponential stellar probabilities from J+24. Dwarf galaxies
are scaled by our derived \(R_h\) values.

\subsection{Density Profile Reliability and
Uncertainties}\label{density-profile-reliability-and-uncertainties}

\begin{itemize}
\tightlist
\item
  How well do we know the density profiles?
\item
  What uncertainties affect derived density profiles?
\item
  Can we determine if Gaia, structural, or algorithmic systematics
  introduce important errors in derived density profiles?
\item
  Using J+24 data, we validate

  \begin{itemize}
  \tightlist
  \item
    Check that PSAT, magnitude, no-space do not affect density profile
    shape too significantly
  \end{itemize}
\item
  Our ``high quality'' members all have \textgreater{} 50 member stars
  and do not depend too highly on the spatial component, mostly
  corresponding to the classical dwarfs
\end{itemize}

J+24's algorithm takes spatial position into account, assuming either a
one or two component exponential density profile. When deriving a
density profile, this assumption may influence the derived density
profile, especially when the galaxy density is fainter than the
background of similar appearing stars. To remedy this and estimate where
the background begins to take over, we also explore a cut based on the
likelihood ratio of only the CMD and PM components. This is in essence
assuming that the spatial position of a star contains no information on
it's membership probability (a uniform distribution like the background)

\subsection{Caveats}\label{caveats}

The J+24 method was designed to determine high probability members for
spectroscopic followup in particular. Note that we instead care about
retrieving a reliable density profile.

In particular, in fig.~\ref{fig:umi_observed_profiles}, notice that the
PSAT method produces artifically small errorbars even when the density
is \textgreater1dex below the local background. These stars are likely
selecting stars from the statistical MW background consistent with UMi
PM / CMD, recovering the assumed density profile. As a result, the
reliability of faint features in these density profiles is questionable
and a more robust analysis, removing this particular density assumtion,
would be required to more appropriately represent the knowledge of the
density profile as the background begins to dominate.

J+24 do not account for structural uncertainties in dwarfs. This is a
not insignificant source of uncertainty in the derived density profile

We assume constant ellipticity and position angle. Dwarf galaxies, in
reality, are not necessarily smooth and constant.

\section{Comparison and conclusions}\label{comparison-and-conclusions}

To illustrate the differences between each dwarf galaxy, in
fig.~\ref{fig:classical_dwarfs_densities}, we compare Scl, UMi, and Fnx
against exponential and plummer density profiles (\textbf{TODO: state
these somewhere}). While all dwarfs have marginal differences in the
inner regions, each dwarf diverges in the outer regions relative to an
exponential. In particular, while Fnx is underdense, Scl and UMi are
both overdense, approximately fitting a Plummer density profile instead.

In summary, we have used J+24 data to derive the density profiles for
Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
robust against different selection criteria. Both Sculptor (Ursa Minor)
show strong (weak) evidence for deviations from an exponential profile.
We will explore a tidal explanation for these features in this work.

\begin{figure}
\centering
\includegraphics{figures/scl_umi_fornax_exp_fit.pdf}
\caption[Classical dwarf density profiles]{The density profiles of
Sculptor, Ursa Minor, and Fornax compared to Exp2D and Plummer density
profiles. Dwarf galaxies are scaled to the same half-light radius and
density at half-light radius (fit from the inner 3 scale radii
exponential recursively. )}\label{fig:classical_dwarfs_densities}
\end{figure}

\section{Appendix / Extra Notes}\label{appendix-extra-notes}

\subsection{Does Scl show tidal tails}\label{does-scl-show-tidal-tails}

\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}

\begin{itemize}
\item
  There are no apparent overdensities in the PM \& CMD only selected
  stars to suggest the presence of a tidal tail
\item
  We have tried selective matched filter
\item
  This means that at least at the level of where the background density
  dominates, we can exclude models which produce tidal tails brighter
  than a density of
  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
\end{itemize}

\subsection{Radial velocity modeling}\label{radial-velocity-modeling}

For Scl and UMi, to derive robust RV measurements, we compile a variety
of literature radial velocity data and compute the radial velocity and
velocity dispersion.

For Scl, we compile and cross match data from

We present a combined sample crossmatching the studies
\citep{sestito+23}, apogee, \citep{walker+2009}, and \citet{tolstoy+23},
containing a total of 2466 stars with measurements. Note that
\citet{battaglia+2009} is superceded by \citet{tolstoy+23}. If multiple
samples measure the same star, then we take the weighted mean of the
measurements. We flag stars with a measurement dispersion greater than
5\(\sigma\) of the combined measurement error and stars outside of the
range 60-150 km/s as these are likely too far off to be members (and the
intrinsic velocity dispersion and systematic errors are the dominant
contributors to dispersion.)

Note that there is, in detail, small projection effects to the radial
velocities, especially as we move towards 1 degree from the centre of
the galaxy. We define the perpendicular tangent plane velocity \(v_z\)
to be \[
v_z = v_{\rm los,\ gsr} \cos\theta + v_{\rm tan} \sin\theta
\] where \(v_{\rm tan}\) is the tangental systematic proper motion of
the system. Note that variations in proper motions due to the intrinsic
velocity dispersion of the system (of order 10 km/s) would introduce
only a 0.17 km/s error in the particular velocity. Typically, proper
motions are not measured well enough for individual stars for this to
matter (requiring \(\mu {\rm as}\) precision :/ )

To fit the systematic radial velocity, we use the Bayesian model \[
v_z \sim N(v_0, \sigma_v + \delta v_z)\\
v_0 \sim U() \\
\sigma_v \sim N?
\] for Sculptor, and only changing the priors above for Ursa Minor \[
v_0 = \\
\sigma_v = \\
\]

To test if there is a spatial gradient in velocity, we construct a new
model \[
v_z = N(v_0 + a \xi + b \eta, \sigma_v + \delta v_z)
\]

\begin{figure}
\centering
\includegraphics{figures/scl_v_z_obs_scatter.pdf}
\caption{Sculptor radial velocity observations}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/umi_v_z_obs_scatter.pdf}
\caption{Sculptor radial velocity observations}
\end{figure}

\subsection{Additional density profile
tests}\label{additional-density-profile-tests}

\begin{figure}
\centering
\includegraphics{figures/scl_density_methods_extra.pdf}
\caption[Density profiles]{Density profiles for various assumptions for
Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
2-component bayesian model assuming circular radii, simple is the series
of simple cuts described in Appendix ?, bright is the sample of the
brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
(background subtracted and rescaled to
match).}\label{fig:sculptor_observed_profiles}
\end{figure}

Note that a full rigorous statistical analysis would require a
simulation study of injecting dwarfs into Gaia and assessing the
reliability of various methods of membership and density profiles. This
is beyond the scope of this thesis.

\begin{verbatim}
SELECT TOP 1000
       *
FROM delve_dr2.objects
WHERE 11 < ra
and ra < 19
and -37.7 < dec
and dec < -29.7
\end{verbatim}

### 2025-04-07 thesis/introduction.tex
thesis updates (most of data chapter)

--- 2025-03-13 thesis/introduction.tex
@@
 
-\subsection{Gaia Membership Selection}\label{gaia-membership-selection}
+The classical dwarfs are some of the earliest discovered systems,
+begining with \citet{shapley1938}
+
+\subsection{Sculptor}\label{sculptor}
 
@@
 \centering
-\includegraphics{figures/scl_selection.png}
-\caption{Sculptor selection criteria.}\label{fig:sculptor_selection}
+\includegraphics[width=5.41667in,height=5.41667in]{figures/scl_des_dr2.png}
+\caption[scl\_image]{Sculptor image from DES DR 2 via
+HiPS2FITS.}\label{fig:scl_image}
 \end{figure}
 
-We use J+24 data, updated PM from the MV 2020a catalogue.
+\begin{figure}
+\centering
+\includegraphics{figures/scl_umi_vs_penarrubia.pdf}
+\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
+profiles are well-matched to \citet{PNM2008}.}\label{fig:toy_profiles}
+\end{figure}
 
-J+24 select members using a multi-component Baysian algorithm.
+\section{}\label{section}
+
+Sculptor (Scl) is one of the first discovered dwarf galaxies of the
+Milky Way (Shapley 1938; only preceded by the SMC and LMC!). As a
+classical dwarf spheroidal, Scl is relatively bright and compact.
+
+Since the initial discovery of Scl, most authors have noted that Scl has
+a slight ellipticity (\(0.36\)), often attributed to tidal effects.
+However, this does not align with the absolute proper motion or orbital
+path of the galaxy.
+
+While Scl has a relatively large pericentre (greater than 50kpc),
+\citet{Sestito+23} detect that the galaxy likely has an excess of stars
+in the outskirts (past about 60 arcminutes). This excess is perhaps one
+of the clearest indications that Scl may be affected by tides. Here, our
+goal is to determine if under a \(\Lambda\) CMD paradigm with DM-only
+simulations, tidal effects of the Milky Way (or LMC) may indeed be
+consistent with these observations, or if these observations may reveal
+instead a extended stellar ``halo'' or second component of the
+galaxy---illustrating a complex formation for the galaxy.
+
+Figure: Image of Sculptor from DES.
+
+Theoretical work on Sculptor
 
@@
 \item
-  First, stars with poor Gaia astrometry, color excess nearby
-  parallaxes, and proper motions \textgreater{} 10 mas/yr are removed
+  \citet{battaglia+2008}
 \item
-  Stars are assigned a likelihood based on the location on the CMD
-  (assuming)
+  \citet{iorio+2019}
 \end{itemize}
 
-Simple cuts, such as selecting PM, CMD based on purely geometric cuts
-reveal a similar density profile.
-
-One caveat is the original algorithm takes spatial position into
-account. When deriving a density profile, this assumption may influence
-the derived density profile, especially when the galaxy density is
-fainter than the background of similar appearing stars. To remidy this
-and estimate where the background begins to take over, we also explore a
-cut based on the likelihood ratio of only the CMD and PM components.
-
-\subsection{Searches for tidal tails}\label{searches-for-tidal-tails}
-
-\begin{figure}
-\centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313132102118.png}
-\caption{image-20250313132102118}
-\end{figure}
-
-\subsection{Density Profile Reliability and
-Uncertainties}\label{density-profile-reliability-and-uncertainties}
-
-Given a set of probable members, the density profile is given by
-EQUATION, where we use \ldots{}
+Observational work on Scl
 
@@
 \item
-  Using J+24 data, we validate
+  \citet{sestito+2023a}
+\item
+  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
+  \citet{arroyo-polonio+2024}
+\item
+  \citet{eskridge1988}
+\end{itemize}
+
+\begin{table*}[t]
+\centering
+\caption{Measured properties of Sculptor}
+\label{tbl:Measured-properties-of-Sculptor}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $15.0183 \pm 0.0012$˚ & M+18\\
+$\delta$ & $-33.7186 \pm 0.00072$˚ & M+18\\
+distance & $83.2 \pm 2$ kpc & Tran+22\\
+$\mu_\alpha \cos \delta$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
+RV & $111.4 \pm 0.2$ & This work\\
+$\sigma_v$ & $9.61\pm0.16$ & This work\\
+$R_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
+$R_h$ & $8.69 \pm 0.2$ arcmin & this work?\\
+ell & $0.36 \pm 0.01$ & M+18\\
+PA & $92\pm1$ & M+18\\
+$M_V$ & $-10.82\pm0.14$ & M+18\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+Can I just cite \citet{arroyo-polonio+2024} for velocity information for
+Sculptor? They correct for binary and do a more detailed discussion of
+rotation.
+
+\subsection{Ursa Minor}\label{ursa-minor}
+
+\begin{itemize}
+\item
+  \citet{sestito+2023b}
+\item
+  \citet{pace+2020}
+\end{itemize}
+
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
+\caption[Ursa Minor Image]{Ursa Minor Dwarf. Composite Public
+data/Giuseppe Donatiello. Updated image in 2022.
+https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
+\end{figure}
+
+\begin{table*}[t]
+\centering
+\caption{Measured properties of Ursa Minor}
+\label{tbl:Measured-properties-of-Ursa-Minor}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $ 227.2420 \pm 0.0045$˚ & M+18\\
+$\delta$ & $67.2221 \pm 0.00158$˚ & M+18\\
+distance & $70.1 \pm 2.9$ kpc & \\
+$\mu_\alpha \cos \delta$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
+$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
+RV & $-245 \pm 1$ km s$^{-1}$ & This work\\
+$\sigma_v$ & $9 \pm 0.6$ & This work\\
+$r_h$ & $18.2 \pm 0.1$ arcmin & M+18\\
+ell & $0.55 \pm 0.01$ & M+18\\
+PA & $50 \pm 1^\circ$ & M+18\\
+$M_V$ & $-9.03 \pm 0.05$ & M+18\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Theoretical Background}\label{theoretical-background}
 
@@
 \item
-    Check that PSAT, magnitude, no-space do not affect density profile
-    shape too significantly
-  \end{itemize}
+  Cosmology foundations
 \item
-  Our ``high quality'' members all have \textgreater{} 50 member stars
-  and do not depend too highly on the spatial component, mostly
-  corresponding to the classical dwarfs
-\item
-  We fit Sérsic profiles to each galaxy
+  NFW plot (density or energy), maybe borrow from paper?
 
@@
   \item
-    The Sérsic index, \(n\), is a measure of the deviation from an
-    exponential. Exponentials have \(n=1\), whereas more extended dwarf
-    galaxies will have higher \(n\)
+    Explain origin of NFW
   \end{itemize}
 \item
-  To better estimate the uncertanties due to unknown galaxy properties
-  and flexibility in the likelihood cut, we can
+  Dwarf galaxy formation, halos
+\item
+  Evolution under tidal field
+\item
 \end{itemize}
 
-\begin{figure}
-\centering
-\includegraphics{figures/density_profiles_medley.png}
-\caption{Density profiles}\label{fig:sculptor_observed_profiles}
-\end{figure}
+To motivate why a tidal interaction may give rise to the observed
+density profiles, we create a toy simulation following \citet{PNM2008}.
 
-Dark Matter is one of the biggest open questions in modern astronomy.
-Hello fig.~\ref{fig:sculptor}.
+\begin{itemize}
+\item
+  NFW initial conditions (sculptor like, vcm, rcm)
+\item
+  Evolved in x-y plane using \citet{EP2020} potential for
+  \textasciitilde{} 5Gyr with pericentre of 15 kpc and apocentre of 100
+  kpc.
+\item
+  Exponential initial stellar profile.
+\end{itemize}
 
-\subsection{Comparison of the Classical
-dwarfs}\label{comparison-of-the-classical-dwarfs}
-
-\begin{figure}
-\centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130050775.png}
-\caption{image-20250313130050775}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130110550.png}
-\caption{image-20250313130110550}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130043114.png}
-\caption{image-20250313130043114}
-\end{figure}
-
-\subsection{Summary}\label{summary}
+As a dark matter halo is perturbed on a pericentric passage with the
+milky way,
 
@@
 \item
-  Of the classical dwarfs, UMi \& Scl stand out statistically, with high
-  \(n\) given their luminosity
+  Tidal stress heats halo slightly
 \item
-  Including fainter dwarf galaxies, Boo 3 and Boo 1 appear to also have
-  extended density profiles
+  Mass loss, particularly of loosely bound particles
+\end{itemize}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Deeper data would be required to robustly measure this
-  \end{itemize}
-\end{itemize}
+The stellar component tracers will similarly follow the behaviour of the
+dark matter.
+
+An emperical estimate of where the simulation's stars are becoming
+unbound is, as stated in \citet{PNM2008}, the break radius \[
+R_b = C\,\sigma_{v}\,\Delta t
+\] where \(\sigma_v\) is the present line of sight velocity dispersion ,
+\(\delta t\) is the time since pericentre, and \(C \approx 0.55\) is a
+fit. The idea motivating this equation is stars in the inner regions
+will have dynamically equilibriated to the new potential (phase mixed),
+however the outer regions are no longer in steady state, so we have to
+wait until the crossing time reaches them as well.
+
+As illustrated in fig.~\ref{fig:toy_profiles}, the density profile
+initially stars off exponential. At increasing times since the first
+pericentric passage, the break radius, appearing as an apparent
+separation between the slopes of the inner and outer profile, increases.

### 2025-04-07 thesis/methods.tex
thesis updates (most of data chapter)

\section{Simulation Methods}\label{simulation-methods}

In this section, we discuss our simulation methods from orbital
estimation to initial conditions to simulation parameters and software.

We use Agama \citep{vasiliev} for potential specification and initial
conditions, and Gadget-4 \citep{springle2021} for orbit integration and
dark matter simulations (with a custom patch to integrate with Agama).

\subsection{Potential}\label{potential}

Our fiducial potential is as described in \citet{EP2020}. This potential
is an analytic approximation of \citet{McMillan2011}, consisting of a
stellar Bulge, disk, and a dark matter NFW halo. The specific parameters
are

\subsubsection{Bulge}\label{bulge}

Hernquist potential with \(r_s=1.3\), mass = 2.1.

The \citet{hernquist1990} density profile for the galactic bulge is
parameterized in terms of a characteristic mass, \(M\), and radius
\(a\).

Density Profile \[
\rho(r) = \frac{M}{2\pi} \frac{a}{r} \frac{1}{(r+a)^3}
\]

\subsubsection{Disk}\label{disk}

\begin{itemize}
\tightlist
\item
  Thin disk. miyamoto+nagai disk with parameters \(a=3.944\),
  \(b=0.311\), \(M=3.944\),
\item
  Thick disk. miyamoto+nagai disk with parameters \(a=4.4\), \(b=0.92\),
  and \(M=2\)
\end{itemize}

\[
\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
\]

for convenience, we define the helper variables,
\(S = \sqrt{z^2 + b^2}\) and \(D = \sqrt{R^2 + (a+S)^2}\), so the
potential simplifies to \[
\Phi(R, z) = \frac{-GM}{D}
\]

\[
\rho = \frac{b^2 M}{4\pi} \frac{aR^2 + (a+3r)(a + r)^2}{S^3 D^5}
\]

\subsubsection{Halo}\label{halo}

Mvir=115, r=20.2, c=9.545

\begin{itemize}
\tightlist
\item
  NFW Dark matter halo. \(M_s=79.5\) (double check). and \(r_s = 20.2\).
\end{itemize}

Variations to the potential of the inner disk (exclusion of a bar)
should minimally affect our results as no orbit we consider reaches less
than \textasciitilde15 kpc of the MW centre.

\textbf{Figure:} Circular velocity profile of \citet{EP2020} potential.

\subsection{Orbital Estimation}\label{orbital-estimation}

To estimate the orbits of a dwarf galaxy, we perform a Monte Carlo
sampling of the present-day observables. We integrate each sampled
observable back in time for 10 Gyr in Gadget as massless point
particles.

We use the astropy v4 galactocentric frame. This frame assumes
\ldots\ldots{}

\subsection{Initial conditions}\label{initial-conditions}

We use Agama (cite) to generate initial conditions. We initially assume
galaxies are described by an NFW dark matter potential and the stars are
merely collisionless tracers embedded in this potential

\subsubsection{Stellar Probabilities}\label{stellar-probabilities}

We assign stellar probabilities via Eddington inversion using the
following method:

\begin{itemize}
\tightlist
\item
  \(\Psi\) is taken as known from the underlying assumed analytic dark
  matter potential.
\item
  \(\rho_\star\) is
\end{itemize}

\subsubsection{Stellar profiles}\label{stellar-profiles}

We consider several different stellar profiles in this work.

\paragraph{Exp2D}\label{exp2d}

The 2D exponential stellar profile is given by \[
\Sigma_\star(R) = A e^{-R / R_s}
\] where \(A\) is some normalization and \(R_s\) is the exponential
scale radius. The 3D deprojected profile is found through abel inversion
(e.g.~Rapha\ldots), i.e. \[
\rho_\star (r) = \int dR = xxx
\]

\paragraph{Plummer}\label{plummer}

\subsection{Dark Matter simulations}\label{dark-matter-simulations}

See \citet{power+2003} for a discussion of this.

The gravitational softening is based on: \[
h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
\]

I in fact find that dividing the softening by a factor of sqrt(10) gives
slightly more precise results with a negligible increase in compute
time.

\section{Appendix: Convergence tests and Simulation
Parameters}\label{appendix-convergence-tests-and-simulation-parameters}

Here, we describe some convergence tests to ensure our methods and
results are minimally impacted by numerical limitations and assumptions.
See \citet{power2003} for a detailed discussion of various assumptions
and parameters used in N-body simulations.

\subsubsection{Particle Number.}\label{particle-number.}

\subsubsection{Timestepping}\label{timestepping}

\subsubsection{Gravitational methods}\label{gravitational-methods}

\subsubsection{Gravitational accuracy}\label{gravitational-accuracy}

\subsubsection{Fiducial Parameters}\label{fiducial-parameters}

\begin{verbatim}
# -----------------------------------IO---------------------------------------

#---Filenames
InitCondFile                initial
OutputDir                   ./out
SnapshotFileBase            snapshot
OutputListFilename          outputs.txt

#---File formats 
ICFormat                    3       # use HDF5
SnapFormat                  3

#---Mem & CPU limits
TimeLimitCPU                86400   # TUNE
CpuTimeBetRestartFile       7200    # TUNE 
MaxMemSize                  2400    # TUNE to system (MB)

#---Time
TimeBegin                   0
TimeMax                       2120      # 10 Gyr

#---Output frequency
OutputListOn                0
TimeBetSnapshot             10      # change as needed
TimeOfFirstSnapshot         0       # MATCH to TimeBegin
TimeBetStatistics           10      # change as needed 
NumFilesPerSnapshot         1
MaxFilesWithConcurrentIO    1

# ----------------------------------Gravity-----------------------------------

# MATCH all the below if continuing run

#---Timestep accuracy
ErrTolIntAccuracy           0.01    # MATCH
CourantFac                  0.1     # ignored; for SPH
MaxSizeTimestep             0.5     # MATCH
MinSizeTimestep             0.0

#---Tree algorithm
TypeOfOpeningCriterion      1       # 0: Barnes-Hut, 1: Relative
ErrTolTheta                 0.5     # mostly used for Barnes-Hut
ErrTolThetaMax              1.0     # (used only for relative)
ErrTolForceAcc              0.005   # (used only for relative)

#---Domain decomposition: should only affect performance
TopNodeFactor                       3.0
ActivePartFracForNewDomainDecomp    0.02

#---Gravitational Softening
SofteningComovingClass0      # FILL IN
SofteningMaxPhysClass0       0      # ignored; for cosmological
SofteningClassOfPartType0    0
SofteningClassOfPartType1    0

#--------------------------------Misc-------------------------------------

# probably do not need to change the options below

#---Unit System
UnitLength_in_cm            1 
UnitMass_in_g               1
UnitVelocity_in_cm_per_s    1 
GravityConstantInternal     1

#---Cosmological Parameters 
ComovingIntegrationOn     0 # no cosmology
Omega0                    0
OmegaLambda                 0 
OmegaBaryon                 0
HubbleParam                 1
Hubble                      100
BoxSize                     0

#---SPH
ArtBulkViscConst             0.8
MinEgySpec                   0
InitGasTemp                  100

#---Initial density estimate (SPH)
DesNumNgb                   64
MaxNumNgbDeviation          1
\end{verbatim}

### 2025-04-07 thesis/compile.sh

@@ -5,4 +5,6 @@ pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=secti
 
 pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
+pandoc $pandocoptions -o data.tex data.md
+pandoc $pandocoptions -o methods.tex methods.md
 latexmk -xelatex -f main.tex

### 2025-04-07 thesis/md_figure.py

@@ -12,8 +12,9 @@ def rename_url(blocks, i):
                     print("Found image at ", url, file=sys.stderr)
                     if url.startswith("figures"):
                         url_new = path.splitext(url)[0] + ".pdf"
-                        con2.url = url_new
-                        print("Renaming ", url,"=>", con2.url, file=sys.stderr)
+                        if path.isfile(url_new):
+                            con2.url = url_new
+                            print("Renaming ", url,"=>", con2.url, file=sys.stderr)
 
 
 def process_figures(doc):

### 2025-04-08 thesis/data.tex
thesis updates to methods

@@
 quintessential unperturbed dwarf galaxy.
+
+\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}
+
+\begin{itemize}
+\item
+  There are no apparent overdensities in the PM \& CMD only selected
+  stars to suggest the presence of a tidal tail
+\item
+  We have tried selective matched filter
+\item
+  This means that at least at the level of where the background density
+  dominates, we can exclude models which produce tidal tails brighter
+  than a density of
+  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
+  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
+\end{itemize}
 
@@
 
-\subsection{Does Scl show tidal tails}\label{does-scl-show-tidal-tails}
-
-\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}
-
-\begin{itemize}
-\item
-  There are no apparent overdensities in the PM \& CMD only selected
-  stars to suggest the presence of a tidal tail
-\item
-  We have tried selective matched filter
-\item
-  This means that at least at the level of where the background density
-  dominates, we can exclude models which produce tidal tails brighter
-  than a density of
-  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
-  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
-\end{itemize}
-
-\subsection{Radial velocity modeling}\label{radial-velocity-modeling}
-
-For Scl and UMi, to derive robust RV measurements, we compile a variety
-of literature radial velocity data and compute the radial velocity and
-velocity dispersion.
-
-For Scl, we compile and cross match data from
-
-We present a combined sample crossmatching the studies
-\citep{sestito+23}, apogee, \citep{walker+2009}, and \citet{tolstoy+23},
-containing a total of 2466 stars with measurements. Note that
-\citet{battaglia+2009} is superceded by \citet{tolstoy+23}. If multiple
-samples measure the same star, then we take the weighted mean of the
-measurements. We flag stars with a measurement dispersion greater than
-5\(\sigma\) of the combined measurement error and stars outside of the
-range 60-150 km/s as these are likely too far off to be members (and the
-intrinsic velocity dispersion and systematic errors are the dominant
-contributors to dispersion.)
-
-Note that there is, in detail, small projection effects to the radial
-velocities, especially as we move towards 1 degree from the centre of
-the galaxy. We define the perpendicular tangent plane velocity \(v_z\)
-to be \[
-v_z = v_{\rm los,\ gsr} \cos\theta + v_{\rm tan} \sin\theta
-\] where \(v_{\rm tan}\) is the tangental systematic proper motion of
-the system. Note that variations in proper motions due to the intrinsic
-velocity dispersion of the system (of order 10 km/s) would introduce
-only a 0.17 km/s error in the particular velocity. Typically, proper
-motions are not measured well enough for individual stars for this to
-matter (requiring \(\mu {\rm as}\) precision :/ )
-
-To fit the systematic radial velocity, we use the Bayesian model \[
-v_z \sim N(v_0, \sigma_v + \delta v_z)\\
-v_0 \sim U() \\
-\sigma_v \sim N?
-\] for Sculptor, and only changing the priors above for Ursa Minor \[
-v_0 = \\
-\sigma_v = \\
-\]
-
-To test if there is a spatial gradient in velocity, we construct a new
-model \[
-v_z = N(v_0 + a \xi + b \eta, \sigma_v + \delta v_z)
-\]
-
-\begin{figure}
-\centering
-\includegraphics{figures/scl_v_z_obs_scatter.pdf}
-\caption{Sculptor radial velocity observations}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{figures/umi_v_z_obs_scatter.pdf}
-\caption{Sculptor radial velocity observations}
-\end{figure}
-
 \subsection{Additional density profile

### 2025-04-08 thesis/introduction.tex
thesis updates to methods

@@
 
+\subsection{Cosmological context}\label{cosmological-context}
+
+\begin{figure}
+\centering
+\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250408091622850.png}
+\caption[Cosmological Power Spectrum]{figure 1 from
+\citet{bechtol+2022}.}
+\end{figure}
+
+Figure: Density profiles of comological simulated halos, matching
+approximantly the NFW formula.
+
 \subsection{Sculptor}\label{sculptor}
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/scl_des_dr2.png}
-\caption[scl\_image]{Sculptor image from DES DR 2 via
+\caption[Picture of Sculptor]{Sculptor image from DES DR 2 via
 HiPS2FITS.}\label{fig:scl_image}
@@
 While Scl has a relatively large pericentre (greater than 50kpc),
-\citet{Sestito+23} detect that the galaxy likely has an excess of stars
-in the outskirts (past about 60 arcminutes). This excess is perhaps one
-of the clearest indications that Scl may be affected by tides. Here, our
-goal is to determine if under a \(\Lambda\) CMD paradigm with DM-only
-simulations, tidal effects of the Milky Way (or LMC) may indeed be
-consistent with these observations, or if these observations may reveal
-instead a extended stellar ``halo'' or second component of the
-galaxy---illustrating a complex formation for the galaxy.
-
-Figure: Image of Sculptor from DES.
+\citet{sestito+2023a} detect that the galaxy likely has an excess of
+stars in the outskirts (past about 60 arcminutes). This excess is
+perhaps one of the clearest indications that Scl may be affected by
+tides. Here, our goal is to determine if under a \(\Lambda\) CMD
+paradigm with DM-only simulations, tidal effects of the Milky Way (or
+LMC) may indeed be consistent with these observations, or if these
+observations may reveal instead a extended stellar ``halo'' or second
+component of the galaxy---illustrating a complex formation for the
+galaxy.
 
@@
 $\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
-RV & $111.4 \pm 0.2$ & This work\\
-$\sigma_v$ & $9.61\pm0.16$ & This work\\
+RV & $111.2 \pm 0.2$ & arroyo-polonio+24\\
+$\sigma_v$ & $9.7\pm0.2$ & arroyo-polonio+24\\
 $R_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
-$R_h$ & $8.69 \pm 0.2$ arcmin & this work?\\
+$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
 ell & $0.36 \pm 0.01$ & M+18\\
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
-\caption[Ursa Minor Image]{Ursa Minor Dwarf. Composite Public
+\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. Composite Public
 data/Giuseppe Donatiello. Updated image in 2022.
@@
   Cosmology foundations
+
+  \begin{itemize}
+  \tightlist
+  \item
+    power spectrum plot
+  \end{itemize}
 \item
@@
   Dwarf galaxy formation, halos
+\end{itemize}
+
+N-body DM simulations
+
+\begin{itemize}
+\tightlist
+\item
+  Collisionless Boltzmann equation and meaning of such simulations
+\item
+  Assumptions \& context \& past work
 \item
   Evolution under tidal field
-\item
 \end{itemize}
@@
 separation between the slopes of the inner and outer profile, increases.
+
+\subsection{Introduction to Dark Matter
+simulations}\label{introduction-to-dark-matter-simulations}
+
+In this section, we will cover
+
+\begin{itemize}
+\tightlist
+\item
+  How are dark matter simulations conducted
+\item
+  Interpretations and uncertainties
+\item
+  Methods
+\end{itemize}
+
+\begin{figure}
+\centering
+\includegraphics{/Users/daniel/thesis/figures/idealized_break_radius.png}
+\caption[Break radius validation]{The break radius of the simulations is
+set by}\label{fig:idealized_break_radius}
+\end{figure}
+
+From this argument, we note that the following properties must be
+approximately true for tides to occur:
+
+\begin{itemize}
+\tightlist
+\item
+  Close enough pericentre. The other break radius \(r_J\) implies that
+  if the host density is 3x the satellite, stars will be lost
+\item
+  Corresponding time since last pericentre: If the time since last
+  pericentre is not \(\sim\) consistent with an observed break in the
+  density profile, then tides
+\item
+  Halo evolution. As found in \citet{EN2021}, galaxies evolve along well
+  defined tidal tracks (assuming spherical, isotropic, NFW halo, which
+  may not be true, see \ldots). These tracks tend to ``puff up'' the
+  stellar component while also removing dark matter mass, leaving a
+  smaller, compacter DM halo with a more extended stellar component.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    This information is mostly related to the statistical initial
+    distribution of satellites from cosmology {[}ludlow+2016;
+    \citet{fattahi+2018}{]}
+  \end{itemize}
+\end{itemize}
+
+\subsection{Potentials}\label{potentials}
+
+From the \citet{nfw1996} paper, eqns. 3 \& 4 \[
+\frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
+\] where \[
+\delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
+\] \(c\) is concentration parameter, \(\rho_c\) is the critical density
+of the universe, and \(r_s\) is the characteristic scale length of the
+halo.
+
+The NFW halo is sometimes described by \(M_{200}\). \(r_{200}\) is the
+radius at which the mean density of the halo interior to \(r_{200}\) is
+200 times the critical density of the universe, and \(M_{200}\) is the
+mass contained inside \(r_{200}\). As equations: \[
+\rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
+\]
+
+\[
+r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
+}}}
+\]
+
+\[
+M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
+\]
+
+\(M_{200}\) is also sometimes called the virial mass of the halo.
+\(r_{200}\) is directly related to \(r_s\) by \[
+r_{200} = c\,r_s
+\]
+
+Another useful definition is \[
+A(x) \equiv \log (1+x) - \frac{x}{1+x}.
+\]
+
+We will also define a dimensionless radius \[
+x \equiv r/r_s.
+\]
+
+A simple substitution to the definition gives
+
+\[
+\rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
+\] where \[
+\rho_s = \frac{c^3}{A(c)} \rho_{200}
+\] and \(A(c)\) is as above. The characteristic density can also be
+written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see below),
+giving \[
+\rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
+\] Note that the NFW density profile is the same as an alpha-beta-gamma
+profile where \(\alpha=\gamma=1\) and \(\beta =3\).
+
+\subsubsection{Circular velocity}\label{circular-velocity}
+
+The circular velocity in terms of
+\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \[
+\left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
+\]
+
+or in terms of \(M_s\) and \(r_s\), \[
+v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
+\] Another parameterization of the NFW profile is in terms of the
+maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the radius at
+which it is reached \(r_{\rm circ}^{\rm max}\). Given the scale radius,
+\[
+r_{\rm circ}^{\rm max} = \alpha\ r_s
+\]
+
+where \(\alpha\approx2.16258\), and \(v_{\rm circ}^{\rm max}\) can be
+found from either of the equations above.

### 2025-04-08 thesis/methods.tex
thesis updates to methods

@@
 
-We use Agama \citep{vasiliev} for potential specification and initial
-conditions, and Gadget-4 \citep{springle2021} for orbit integration and
-dark matter simulations (with a custom patch to integrate with Agama).
+We use Agama \citep{agama} for potential specification and initial
+conditions, and Gadget-4 \citep{gadget4} for orbit integration and dark
+matter simulations (with a custom patch to integrate with Agama).
 
@@
 is an analytic approximation of \citet{McMillan2011}, consisting of a
-stellar Bulge, disk, and a dark matter NFW halo. The specific parameters
-are
-
-\subsubsection{Bulge}\label{bulge}
-
-Hernquist potential with \(r_s=1.3\), mass = 2.1.
+stellar Bulge, disk, and a dark matter NFW halo.
+fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles.
 
@@
 
-Density Profile \[
+\[
 \rho(r) = \frac{M}{2\pi} \frac{a}{r} \frac{1}{(r+a)^3}
-\]
-
-\subsubsection{Disk}\label{disk}
+\] Hernquist potential with \(r_s=1.3\), mass = 2.1.
 
@@
 
-for convenience, we define the helper variables,
-\(S = \sqrt{z^2 + b^2}\) and \(D = \sqrt{R^2 + (a+S)^2}\), so the
-potential simplifies to \[
-\Phi(R, z) = \frac{-GM}{D}
-\]
-
-\[
-\rho = \frac{b^2 M}{4\pi} \frac{aR^2 + (a+3r)(a + r)^2}{S^3 D^5}
-\]
-
-\subsubsection{Halo}\label{halo}
-
-Mvir=115, r=20.2, c=9.545
-
 \begin{itemize}
@@
 should minimally affect our results as no orbit we consider reaches less
-than \textasciitilde15 kpc of the MW centre.
+than \textasciitilde15 kpc of the MW centre. We exclude the mass
+evolution of the halo from this analysis. Over 10 Gyr, this would be
+fairly significant (\textasciitilde2x in MW mass) but since we want to
+determine the upper limit of tidal effects, it is safe to neglect this.
+
+\begin{figure}
+\centering
+\includegraphics{figures/v_circ_potential.pdf}
+\caption{Circular velocity of potential}\label{fig:v_circ_potential}
+\end{figure}
 
 \textbf{Figure:} Circular velocity profile of \citet{EP2020} potential.
+We also show the MW and LMC potential from \citet{vasiliev2024} for
+their L3M11 model.
 
@@
 sampling of the present-day observables. We integrate each sampled
-observable back in time for 10 Gyr in Gadget as massless point
-particles.
+observable back in time for 10 Gyr in Gadget as massless point particles
+outputing every 0.5Myr (with otherwise similar parameters to n-body runs
+below.)
 
-We use the astropy v4 galactocentric frame. This frame assumes
-\ldots\ldots{}
+To convert from Gaia to galactocentric coordinates, we use the astropy
+v4 galactocentric frame. This frame assumes an
+
+\begin{itemize}
+\tightlist
+\item
+  J2000 `ra' = 17h45m37.224s, `dec' = -28°56'10.23'' (appendix),
+  `pmracosdec=-3.151±0.018 mas/yr', `pmdec=-5.547±0.026' (Table 2). from
+  \citet{reid+brunthaler2004}
+\item
+  distance = 8.122 ± 0.033 kpc, solar radial velocity = 11 + 1.9 ± 3
+  km/s. from \citet{gravitycollaboration+2018}.
+\item
+  `z\_sun' = 20.8±0.3pc from bennett+bovy2019
+\item
+  `v\_sun' is slightly more complicated, relying on parameters from
+  above and using the procedure described in \citet{drimmel+poggio2018}.
+  This results in
+
+  \begin{itemize}
+  \tightlist
+  \item
+    `v\_sun' = {[}-12.9 ± 3.0, 245.6 ± 1.4, 7.78 ± 0.08{]} km/s
+  \end{itemize}
+\end{itemize}
 
@@
 
-We use Agama (cite) to generate initial conditions. We initially assume
-galaxies are described by an NFW dark matter potential and the stars are
-merely collisionless tracers embedded in this potential
+We use Agama \citep{agama} to generate initial conditions. We initially
+assume galaxies are described by an NFW dark matter potential (REF) and
+the stars are merely collisionless tracers embedded in this potential.
+We adopt a truncation radius of 100 times the scale radius (likely
+larger than necessary, but should not affect results.)
 
@@
 \item
-  \(\rho_\star\) is
+  \(\rho_\star\) is pre-specified or calculated via abel integration /
+  deprojection
+\item
+  \(f(\epsilon)\) (the energy distribution function) is determined by
+  Edington inversion
 \end{itemize}
 
-\subsubsection{Stellar profiles}\label{stellar-profiles}
+To find the distribution function, we use Eddington inversion (eq.
+4-140b in BT87) \[
+f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
+\]
 
-We consider several different stellar profiles in this work.
+where \({\cal E}\) is the binding energy, and \(\Psi\) is the potential
+normalized to go to zero at infinity. In practice the right, boundary
+term is zero as \(\Psi \to 0\) as \(r\to\infty\), and if
+\(\rho \propto r^{-n}\) at large \(r\) and \(\Psi \sim r^{-1}\) then
+\(d\rho / d\Psi \sim r^{-n+1}\) which goes to zero provided that
+\(n > 1\).
+
+We consider both a Plummer and 2-dimensional exponential (Exp2D) stellar
+profile:
 
@@
 (e.g.~Rapha\ldots), i.e. \[
-\rho_\star (r) = \int dR = xxx
-\]
+\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
+\] where \(K_0\) is the 0th order modified Bessel function of the second
+type.
 
@@
 
+A Plummer profile is defined by \[
+\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
+\] where \(M\) is the total mass and \(R_s\) is the characteristic scale
+radius. The density is \[
+\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
+\]
+
 \subsection{Dark Matter simulations}\label{dark-matter-simulations}
 
-See \citet{power+2003} for a discussion of this.
+We adopt a softening length of \[
+h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}
+\] see appendix for a discussion of this choice, which is similar to the
+\citet{power+2003} suggested softening.
 
-The gravitational softening is based on: \[
-h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
-\]
+\subsubsection{Analysis}\label{analysis}
 
-I in fact find that dividing the softening by a factor of sqrt(10) gives
-slightly more precise results with a negligible increase in compute
-time.
+Shrinking spheres centres inspired by \citet{power+2003}
+
+\begin{itemize}
+\tightlist
+\item
+  Recursively shrink radius by 0.975 quantile and recalculating centroid
+  until radius is less than \textasciitilde1kpc
+\item
+  Remove bound particles (using instantanious potential).
+\item
+  Use previous snapshot centre and acceleration to predict new centre
+  for next snapshot.
+\end{itemize}
 
@@
 results are minimally impacted by numerical limitations and assumptions.
-See \citet{power2003} for a detailed discussion of various assumptions
+See \citet{power+2003} for a detailed discussion of various assumptions
 and parameters used in N-body simulations.
 
-\subsubsection{Particle Number.}\label{particle-number.}
+\subsubsection{Softening}\label{softening}
 
-\subsubsection{Timestepping}\label{timestepping}
+\citet{power+2003} suggest the empirical rule that the ideal softening
+(balancing integration time and only compromising resolution in
+collisional regime) is \[
+h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
+\]
 
-\subsubsection{Gravitational methods}\label{gravitational-methods}
+For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
+particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
+slightly smaller softening which was reduced by a factor of
+\(\sqrt{10}\) based on initial numerical experiments. However, this
+choice is likely too small and a larger softening would minimally impact
+our results but improve computation speed.
 
-\subsubsection{Gravitational accuracy}\label{gravitational-accuracy}
+\subsubsection{Particle Number}\label{particle-number}
+
+As discussed in \citet{power+2003}, the region where density is
+converged is related to the region which becomes collisionally relaxed
+over the time of the universe (so about 5-10 Gyr). (i.e.~where the
+collisionless assumption breaks down). The relaxation timescale is given
+by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise, \[
+t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
+= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
+\] This works out to be about 6-10 times (increasing with particle
+number) our adopted softening length for NFW halos given our
+assumptions. As such, at full resolution, we can only trust density
+profiles down to \(10\epsilon\) or about
+\(0.14\,{\rm kpc} (r_s/2.76\,{\rm kpc})\), just enough to resolve
+stellar density profiles.
+
+Note that by decreasing the truncation radius, the effective particle
+number is improved, so future experiments could be less generous with
+this parameter to improve computational performance (order 30\% better
+resolution).
+
+\begin{figure}
+\centering
+\includegraphics{figures/iso_converg_num.pdf}
+\caption[num convergence]{Numerical convergence test for circular
+velocity as a function of log radius for simulations with different
+total numbers of particles in isolation. Residuals in lower panel are
+relative to NFW. The initial conditions are dotted and the converged
+radius is marked by arrows (REF).}
+\end{figure}
+
+\subsubsection{Timestepping and force
+accuracy}\label{timestepping-and-force-accuracy}
+
+In general, we use adaptive timestepping and relative opening criteria
+for gravitational force computations. To verify that these choices and
+associated accuracy parameters minimally impact convergence or speed, we
+show a few more isolation runs (using only 1e5 particles)
+
+\begin{itemize}
+\tightlist
+\item
+  constant timestep (\ldots), approximantly half of minimum timestep
+  with adaptive timestepping
+\item
+  geometric opening, with \(\theta = 0.5\).
+\item
+  strict integration accuracy, (facc = \ldots.)
+\end{itemize}
+
+\subsubsection{Alternative methods}\label{alternative-methods}
+
+\begin{itemize}
+\tightlist
+\item
+  FMM
+\item
+  PMM-tree
+\item
+  Gadget2
+\item
+  etc.
+\end{itemize}
 
@@
 
+Note that we use code units which assume that \(G=1\) for convenience
+and numerical stability. The conversion between code units to physical
+units is (for our convention):
+
+\begin{itemize}
+\tightlist
+\item
+  1 length = 1 kpc
+\item
+  1 mass unit = \(10^{10}\) Msun
+\item
+  1 velocity unit = 207.4 km/s
+\item
+  1 time unit = 4.715 Myr
+\end{itemize}
+
+Most parameters below are not too relevant or have been discussed or are
+merely dealing with cpu and IO details. The changes between simulation
+runs primarily affect the integration time, output frequency, and
+softening. Otherwise, we leave all other parameters fixed.
+
 \begin{verbatim}
-# -----------------------------------IO---------------------------------------
+#======IO parameters======
 
@@
 #---Mem & CPU limits
-TimeLimitCPU                86400   # TUNE
-CpuTimeBetRestartFile       7200    # TUNE 
-MaxMemSize                  2400    # TUNE to system (MB)
+TimeLimitCPU                86400
+CpuTimeBetRestartFile       7200
+MaxMemSize                  2400
 
@@
 OutputListOn                0
-TimeBetSnapshot             10      # change as needed
-TimeOfFirstSnapshot         0       # MATCH to TimeBegin
-TimeBetStatistics           10      # change as needed 
+TimeBetSnapshot             10
+TimeOfFirstSnapshot         0 
+TimeBetStatistics           10
 NumFilesPerSnapshot         1
@@
 
-# ----------------------------------Gravity-----------------------------------
-
-# MATCH all the below if continuing run
+#=======Gravity======
 
 #---Timestep accuracy
-ErrTolIntAccuracy           0.01    # MATCH
+ErrTolIntAccuracy           0.01
 CourantFac                  0.1     # ignored; for SPH
-MaxSizeTimestep             0.5     # MATCH
+MaxSizeTimestep             0.5
 MinSizeTimestep             0.0
@@
 #---Tree algorithm
-TypeOfOpeningCriterion      1       # 0: Barnes-Hut, 1: Relative
+TypeOfOpeningCriterion      1       # Relative
 ErrTolTheta                 0.5     # mostly used for Barnes-Hut
@@
 #---Gravitational Softening
-SofteningComovingClass0      # FILL IN
+SofteningComovingClass0      0.044  # HALO dependent
 SofteningMaxPhysClass0       0      # ignored; for cosmological
@@
 
-#--------------------------------Misc-------------------------------------
+#=======Miscellanius=======
 

### 2025-04-08 thesis/compile.sh

@@ -7,4 +7,4 @@ pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
-latexmk -xelatex -f main.tex
+latexmk -xelatex -f main.tex -output-directory=out

### 2025-04-08 thesis/data.tex
fixes compiling and warnings

@@
 
-Fornax(figures/fornax\_selection.png)\{\#fig:fornax\_selection\}
-
 Figure: Similar to fig.~\ref{fig:sculptor_selection} except for Fornax.
@@
 In Figures fig.~\ref{fig:scl_observed_profiles},
-fig.~\ref{fig:umi_observed_profiles},
-\citet{fig_fornax_observed_profiles}, we show the derived density
+fig.~\ref{fig:umi_observed_profiles}, we show the derived density
 profiles for each galaxy for samples similar to in the selection plots

### 2025-04-08 thesis/compile.sh

@@ -7,4 +7,4 @@ pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
-latexmk -xelatex -f main.tex -output-directory=out
+latexmk -f

### 2025-04-15 thesis/data.tex
thesis updates

@@
 observational knoledge of each galaxies density profile. In general,
-J+24 use a Bayesian framework incorporating proper motion (PM,
+J+24 use a Bayesian framework incorporating proper motion (PM),
 colour-magnitude diagram (CMD), and spatial information to determine the
 probability that a given star belongs to the satellite. J+24 extends
-\citet{MV2020a} (see also \citet{pace+li2019}, etc.).
+\citet{MV2020a} (see also \citet{pace+li2019}, \citet{battaglia+2022},
+\citet{pace+erkal+li2022}, etc.).
 
-J+24 select stars initially from Gaia satisfying:
+J+24 select stars initially from Gaia within a 2--4 degree circular
+region centred on the dwarf satisfying:
 
@@
 \item
+  Solved parallax, proper motions, colour, and magnitudes.
+\item
   High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3})
@@
   3\(\sigma\) consistency of measured parallax with dwarf distance +
-  uncertainty (near zero with \citet{lindegren+2018} zero-point
-  correction)
+  uncertainty (typically near zero; with \citet{lindegren+2018}
+  zero-point correction).
 \item
   Absolute RA and Dec proper motions less than
-  10\(\,{\rm mas\ yr^{-1}}\)
-\item
-  Good photometry
+  10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
+  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
 \item
@@
 \item
-  G \textgreater{} 22 and less than 5\(\sigma\) above TRGB, and between
+  G \textless{} 22 and less than 5\(\sigma\) above TRGB, and between
   -0.5 and 2.5 in Bp - Rp.
 \end{itemize}
+
+Photometry is dereddened using \citet{schlegel+1988} extenction.
 
@@
 \item
-  CMD: The CMD is from Padova \citep{girardi+2002} isochrone at 12 Gyr
-  with a colour width of 0.1 mag plus the Gaia colour uncertanty and the
-  observed metallicity of the dwarf. The HB is modelled as a constant
-  magnitude extending blue of the CMD with a 0.1 mag width plus the mean
-  colour error. A likelihood map is constructed by sampling the distance
-  modulus in addition to the CMD width, taking the maximum of RGB and HB
-  likelihoods.
+  CMD: The CMD is the lowest metallicity isochrone from Padova
+  \citep{girardi+2002} with age 12 Gyr with a colour width of 0.1 mag
+  plus the Gaia colour uncertainty at each magnitude. The HB is modelled
+  as a constant magnitude extending blue of the CMD. The HB magnitude is
+  the mean magnitude of HB stars from most metal poor isochrone with a
+  0.1 mag width plus the mean colour error. A likelihood map is
+  constructed by sampling the distance modulus in addition to the CMD
+  width, taking the maximum of RGB and HB likelihoods.
 \item
   Spatial: A single exponential
-  (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)) using mean values of the
-  structural parameters (\textbf{verify}). For Scl and UMi, this is
-  instead a double exponential
+  (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)) accounting for structural
+  uncertainty (sampled over position angle, ellipticity, and half light
+  radius).
+\item
+  Alternative spatial: For Scl and UMi, this is instead a double
+  exponential
   \(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\) where
-  the inner exponential remains fixed and structural parameters are not
-  accounted for. Structural parameter uncertainties are not accounted
-  for.
+  the inner exponential remains fixed. Structural parameter
+  uncertainties are not accounted for.
 \item
@@
 \item
-  CMD : Constructed as a KDE using the other quality-selected stars
-  outside of \(5R_h\) in the catalogue. The width of each star is
-  projected as a bivariate Gaussian.
+  CMD : Constructed as a density map using the other quality-selected
+  stars outside of \(5R_h\) in the catalogue. The map is a sum of
+  bivariate gaussians for each star with standard deviations based on
+  the Gaia uncertainties.
 \item
@@
 
-Note that each likelihood map is normalized over the respective
+Note that each likelihood map is normalized over the respective 2D
 parameter space. In order to represent the difference in frequency of
@@
 \item
-  Systemic proper motions \(\mu_\alpha\), \(\mu_\delta\). Prior: within
-  5\(\sigma\) of single component case w/ systematic uncertainties.
-  Single component prior is based on MV2020 (RV members / expected
-  velocity dispersion of halo at distance?)
+  Systemic proper motions \(\mu_\alpha\), \(\mu_\delta\). Single
+  component prior is same as \citet{MV2020}: a normal distribution with
+  mean 0 and standard deviation 100 km/s. If 2-component spatial,
+  instead is a uniform distribution spanning 5\(\sigma\) of single
+  component case w/ systematic uncertainties.
 \item
-  \(f_{\rm sat}\) density normalization (uniform between 0 and 1)
+  \(f_{\rm sat}\) density normalization. Prior is a uniform distribution
+  between 0 and 1.
 \item
-  Spatial component parameters \(B\) and \(R_{\rm outer}\) for extended
+  Spatial component parameters \(B\) is uniform from 0-1 and
+  \(R_{\rm outer}\) is uniform and greater than \(R_s\) for extended
   profiles (Scl and UMi here.)
@@
 
-The median parameters from the simulation are then used to calculate the
+The mode of each parameter from the MCMC are then used to calculate the
 final \(P_{\rm sat}\) values we use here.
@@
 
+In figures fig.~\ref{fig:sculptor_selection},
+fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
+the algorithm. In each case, each criteria plays a roll: proper motions
+are centred around the dwarf systemic motion, the CMD is well defined,
+and stars only within a few \(R_h\) are included. We also plot the RV
+members found in general and in \citet{sestito+2024},
+\citet{sestito+2024b}.
+
+The tangent plane, \(\xi\), \(\eta\), is the projection that
+
+We also illustrate the approximate result of removing the spatial
+component from the likelihood. We define the CMD+PM selection as stars
+satisfying \[
+{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
+\]
+
+These stars appear similar to the fiducial (probable members) sample,
+but instead also appear as an approximately uniform distribution across
+the entire field. This illustrates the approximate background of stars
+which may be confused as members. Additionally, since there is no clear
+spatial structure in the CMD+PM sample, it is unlikely that there are
+tidal tails detectable with Gaia. Not shown here, we also try a variety
+of simpler, absolute cuts and thresholds, finding no extended structure
+beyond what is detected in J+24.
+
+This means that at least at the level of where the background density
+dominates, we can exclude models which produce tidal tails brighter than
+a density of
+\(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
+(TODO assuming a distance of \ldots{} and stellar mass of \ldots).
+
 \begin{figure}
 \centering
-\includegraphics{figures/scl_selection.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
 \caption[Sculptor selection criteria]{The selection criteria for Scl
@@
 \centering
-\includegraphics{figures/umi_selection.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
 \caption[Ursa Minor Selection]{Similar to
@@
 \end{figure}
-
-Figure: Similar to fig.~\ref{fig:sculptor_selection} except for Fornax.
-Note that we only use one-component probabilities here. Fornax is the
-quintessential unperturbed dwarf galaxy.
-
-\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}
-
-\begin{itemize}
-\item
-  There are no apparent overdensities in the PM \& CMD only selected
-  stars to suggest the presence of a tidal tail
-\item
-  We have tried selective matched filter
-\item
-  This means that at least at the level of where the background density
-  dominates, we can exclude models which produce tidal tails brighter
-  than a density of
-  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
-  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
-\end{itemize}
 
@@
 \] where \(N_{\rm memb}\) is the number of members in the bin and
-\(A_{\rm bin}\) is the area of the bin. As discussed below, these
-uncertainties underrepresent the true uncertainty on multiple accounts.
-We retain Poisson errors for simplicity here.
+\(A_{\rm bin}\) is the area of the bin's annulus in 2D. As discussed
+below, these uncertainties underrepresent the true uncertainty on
+multiple accounts. We retain Poisson errors for simplicity here.
 
@@
 \centering
-\includegraphics{figures/scl_density_methods.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods.pdf}}
 \caption[Sculptor density profiles]{The density profile of Sculptor for
@@
 \centering
-\includegraphics{figures/umi_density_methods.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods.pdf}}
 \caption[Ursa Minor density profiles]{Similar to
@@
 \centering
-\includegraphics{figures/fornax_density_methods.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_density_methods.pdf}}
 \caption[Fornax density profiles]{Similar to
@@
 \centering
-\includegraphics{figures/scl_umi_fornax_exp_fit.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fornax_exp_fit.pdf}}
 \caption[Classical dwarf density profiles]{The density profiles of
@@
 \centering
-\includegraphics{figures/scl_density_methods_extra.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
 \caption[Density profiles]{Density profiles for various assumptions for
@@
 \end{verbatim}
+
+\section{Radial velocity modeling}\label{radial-velocity-modeling}
+
+For both Sculptor and Ursa Minor, we construct samples of radial
+velocity measurements and cross match each sample to produce estimates
+of the total velocity dispersion and line of sight velocity for each
+galaxy.
+
+\subsection{Methodology}\label{methodology}
+
+Our RV methodology is
+
+\begin{itemize}
+\tightlist
+\item
+  Cross match all catalogues to J+24 Gaia members. If Gaia ID's are
+  included, we use these. Otherwise, we match to the nearest star within
+  2 arcseconds.
+\item
+  We combine the mean RV measurement from each study using the
+  inverse-variance weighted mean and standard uncertainty.
+\end{itemize}
+
+\[
+v_i = \frac{1}{\sum w}\sum_j w\ v_{i, j} \\
+\delta v_i = \sqrt{\frac{1}{\sum w^2}}
+\]
+
+where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
+deviation \[
+s_{v}^2 = \frac{1}{\sum w} \sum w (v_{i, j} - v_i)^2
+\] Note that we do not include the standard deviation if a source does
+measure the velocity multiple times.
+
+If the total standard deviation is greater than five times the velocity
+standard uncertainty (reduced), then we remove the star since either the
+star is a binary or contains erroneous measurements or other
+uncertainties. Note that since stars typically only have very few
+measurements, these choices are highly approximate and likely limited.
+
+The combined RV likelihood is then \[
+{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
+\] where \[
+{\cal L}_{\rm RV} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
+\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
+dispersion, and \(\delta v_i\) is the individual measurement
+uncertainty. Typically, the velocity dispersion will dominate the
+uncertainty budget here.
+
+Similar to above, we retain stars with the resulting membership
+probability of greater than 0.2.
+
+Finally, we need to correct the coordinate frames. We need a proper
+motion estimate of each star. Since most stars have Gaia measured proper
+motions, but the uncertainties on each proper motion measurement are
+large, we combine the system and individual star's proper motion by
+weighting the systemic proper motion by the z-score of the consistency
+of the measurement with the systemic proper motion. \[
+\mu_{*}' = \frac{1}{w_\star + w_{\rm sys}}\left(\mu_{\star}w_\star + \mu_{\rm sys} w_{\rm sys}\right)
+\] where we model \[
+{w_{\rm sys}} = P_{\rm sat}(\delta \mu ^2 + \sigma_{v, tan}^2)^{-1}
+\] and the weight of the star is just the inverse variance weight plus
+the gaia systematics.
+
+The corrected radial velocities are then \[
+v_z = v_{\rm los, gsr}\cos\theta  + v_{\rm tan, R} \sin\theta
+\] where
+\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
+is the radial component of the proper motion with respect to the centre
+of the galaxy. Since \(\theta \lesssim 1^\circ\), this correction is
+small (\(\sim 1\)km,s\(^{-1}\)), but can prevent the introduction of a
+spurious gradient on the sky \citep{WMO2008}. The uncertainty is then
+the velocity uncertainty plus the distance uncertainties times the PM
+uncertainty from above. We then use the \(v_z\) values for the following
+modelling, however repeating with plain, solar-frame velocities does not
+substantially affect the results too much .
+
+\subsection{Sculptor}\label{sculptor}
+
+For Sculptor, we combine radial velocity measurements
+
+\begin{itemize}
+\item
+  APOGEE (crossmatched to stars meeting basic quality cuts in field)
+\item
+  \citet{sestito+2024a}
+\item
+  \citet{tolstoy+2024}
+\end{itemize}
+
+We adopt a very loose prior of \[
+\mu_{v} = N(111, 10) \\
+\log\sigma_{v} = N(1, 1)
+\] Our resulting velocity dispersions are \$\$
+
+\$\$ Comparing to past work, note that our values are reasonable.
+
+\subsection{Ursa Minor}\label{ursa-minor}
+
+\begin{itemize}
+\tightlist
+\item
+  APOGEE
+\item
+  \citet{sestito+2024b}
+\item
+  \citet{pace+2022}
+\item
+  \citet{spencer+2018}
+\end{itemize}
+
+\subsection{Discussion and
+limitations}\label{discussion-and-limitations}
+
+Our model here is relatively simple. Some things which we do not account
+for include
+
+\begin{itemize}
+\tightlist
+\item
+  interstudy systematics and biases
+\item
+  Inappropriate uncertainty reporting.
+\item
+  Binarity. Binay stars are expected to induce an average velocity
+  dispersion of xxx on an individual star.
+\item
+  Selection effects. RV studies each have their own targeted program and
+\end{itemize}
+
+We partially address several of these by also performing the analysis on
+large RV samples for each galaxy. This allows for individual comparison
+illustrating the differences in selection effects.

### 2025-04-15 thesis/introduction.tex
fixes compiling and warnings
thesis updates

@@
 \centering
-\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250408091622850.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/power_spectrum.png}
 \caption[Cosmological Power Spectrum]{figure 1 from
-\citet{bechtol+2022}.}
+\citet{bechtol+2022}.}\label{fig:cosmological_power_spectrum}
 \end{figure}
@@
 \centering
-\includegraphics{figures/scl_umi_vs_penarrubia.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_vs_penarrubia.pdf}}
 \caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
@@
 \end{figure}
-
-\section{}\label{section}
 
@@
   \citet{iorio+2019}
+\item
+  \citet{amorisco+zavala+deboer2014}
+\item
+  \citet{battaglia+2008}
+\item
+  \citet{breddels+2013}
+\item
+  \citet{breddels+helmi2013}
+\item
+  \citet{richardson+fairbairn2014}
+\item
+  \citet{SFW2017}
+\item
+  \citet{innanen+papp1979}
 \end{itemize}
@@
 \item
-  \citet{eskridge1988}
+  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
+\item
+  \citet{coleman+dacosta+bland-hawthorn2005}
+\item
+  \citet{DQ1994}
+\item
+  \citet{WMO2009}
+\item
+  \citet{IH1995}, \citet{munoz+2018}
+\item
+  \citet{kirby+2009}
+\item
+  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
 \end{itemize}
@@
   \citet{pace+2020}
+\item
+  \citet{bellazzini+2002}
+\item
+  \citet{hargreaves+1994}
+\item
+  \citet{martinez-delgado+2001}
+\item
+  \citet{munoz+2005}
+\item
+  \citet{pace+2020}
+\item
+  \citet{palma+2003}
+\item
+  \citet{spencer+2018}
+\item
+  \citet{vitral+2023}
 \end{itemize}
@@
 \centering
-\caption{Measured properties of Ursa Minor}
-\label{tbl:Measured-properties-of-Ursa-Minor}
+\caption{Measured properties of Ursa Minor.
+References are: (1) Muñoz et al.
+(2018), (2) Garofalo et al.
+(2025), (3) McConnachie and Venn (2020), (4) average of Pace et al.
+(2020) and Spencer et al.
+(2018).}
+\label{tbl:Measured-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
 \begin{tabular}{lll}
@@
 \midrule
-$\alpha$ & $ 227.2420 \pm 0.0045$˚ & M+18\\
-$\delta$ & $67.2221 \pm 0.00158$˚ & M+18\\
-distance & $70.1 \pm 2.9$ kpc & \\
-$\mu_\alpha \cos \delta$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
-$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
-RV & $-245 \pm 1$ km s$^{-1}$ & This work\\
-$\sigma_v$ & $9 \pm 0.6$ & This work\\
-$r_h$ & $18.2 \pm 0.1$ arcmin & M+18\\
-ell & $0.55 \pm 0.01$ & M+18\\
-PA & $50 \pm 1^\circ$ & M+18\\
-$M_V$ & $-9.03 \pm 0.05$ & M+18\\
+$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
+$\delta$ & $67.2221 \pm 0.00158$˚ & 1\\
+distance modulus & $19.23 \pm 0.11$ & 2\\
+distance & $70.1 \pm 3.6$ kpc & \\
+$\mu_\alpha \cos \delta$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
+RV & $-244.9 \pm 0.3_{\rm stat} \pm 1.7_{\rm sys}$ km s$^{-1}$ & 4\\
+$\sigma_v$ & $8.3 \pm 0.3_{\rm stat} \pm 0.4_{\rm sys}$ & 4\\
+$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
+ell & $0.55 \pm 0.01$ & 1\\
+PA & $50 \pm 1^\circ$ & 1\\
+$M_V$ & $-9.03 \pm 0.05$ & 1\\
 \bottomrule
@@
 \centering
-\includegraphics{/Users/daniel/thesis/figures/idealized_break_radius.png}
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
 \caption[Break radius validation]{The break radius of the simulations is

### 2025-04-15 thesis/main.tex
thesis updates (most of data chapter)
fixes compiling and warnings
thesis updates

@@
 
-\title{From tides to tails: \\
-    Does the Milky Way affect Sculptor and Ursa Minor?
+\title{Timid tides or tall tails? \\
+    Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies
 }
@@
 \begin{centering}
-\textbf{\@title}
+\textbf{\@title}\\
 \skipline
@@
 \subsection*{Supervisory Committee}
+\skipline
 
@@
 \noindent Dr. Kim Venn, Co-Supervisor\\
-Department of Astronomy
+Department of Astronomy\\
+
+\skipline
+
+\noindent External Examiner \\
+Outside Department
 
@@
 
-% \chapter*{Acknowledgements}
-% \addcontentsline{toc}{chapter}{Acknowledgements}
-%
+\chapter*{Acknowledgements}
+\addcontentsline{toc}{chapter}{Acknowledgements}
+
+%Julio Navarro, my supervisor. Rapha and other collegues
+
+%Friends and family and collegues (fill in later).
+
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement. 
+\citep{gaiacollaboration+2016, gaiacollaboration+2023}.
+
+This research made use of hips2fits,\footnote{https://alasky.cds.unistra.fr/hips-image-services/hips2fits} a service provided by CDS.
 
@@
 
+\chapter{Data Selection}
+\input{data}
+
+\chapter{Methods}
+\input{methods}
+
 %%%%%%%%%%%%%%%%%%%% REFERENCES %%%%%%%%%%%%%%%%%%
@@
 
-\chapter{Model parameters}\label{sec:jack}
+\chapter{Model parameters}
 
-\chapter{Font properties}
-
-textwidth: \the\textwidth
-
-\newlength{\myw}
-\newlength{\myl}
-
----\rule[2.4pt]{12pt}{0.6pt}
-\settowidth{\myw}{---}\the\myw
-
-{\bf ---}\rule[2.2pt]{12pt}{1.1pt}
-
-\settoheight{\myl}{---}\the\myl
-
--\rule[2.4pt]{3pt}{0.75pt}
-I\rule{1.2pt}{8pt}
-{\bf I}\rule{2.0pt}{8pt}
-|\rule[-2pt]{0.6pt}{12pt}
+\input{numerical_convergence}
 

### 2025-04-15 thesis/methods.tex
fixes compiling and warnings
thesis updates

--- 2025-04-07 thesis/methods.tex
@@
-\section{Simulation Methods}\label{simulation-methods}
+In this section, we discuss our simulation methods from orbital
+estimation and initial conditions to post-processing and software.
 
-In this section, we discuss our simulation methods from orbital
-estimation to initial conditions to simulation parameters and software.
+We use Agama \citep{agama} for potential specification and initial
+conditions, and Gadget-4 \citep{gadget4} for orbit integration and dark
+matter simulations (with a custom patch to integrate with Agama).
 
-We use Agama \citep{vasiliev} for potential specification and initial
-conditions, and Gadget-4 \citep{springle2021} for orbit integration and
-dark matter simulations (with a custom patch to integrate with Agama).
+\section{Milky Way Potential}\label{milky-way-potential}
 
-\subsection{Potential}\label{potential}
+fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
+each component of our fiducially potential. We adopt the potential
+described in \citet{EP2020}, an analytic approximation of
+\citet{mcmillan2011}. The potential consisting of a stellar Bulge, a
+thin and thick disk, and a dark matter NFW halo. Especially in the
+regime where Sculptor and Ursa Minor orbit, the dark matter halo is the
+dominant component of the potential. The details of the stellar disk and
+bulge are less important.
 
-Our fiducial potential is as described in \citet{EP2020}. This potential
-is an analytic approximation of \citet{McMillan2011}, consisting of a
-stellar Bulge, disk, and a dark matter NFW halo. The specific parameters
-are
+The galactic bulge is described by a \citet{hernquist1990} potential.
+The potential is
 
-\subsubsection{Bulge}\label{bulge}
+\[
+\Phi(r) = - \frac{GM}{r + a}
+\] where \(a=1.3\,{\rm kpc}\) is the scale radius and
+\(M=2.1 \times 10^{10}\,\Mo\) is the total mass.
 
-Hernquist potential with \(r_s=1.3\), mass = 2.1.
-
-The \citet{hernquist1990} density profile for the galactic bulge is
-parameterized in terms of a characteristic mass, \(M\), and radius
-\(a\).
-
-Density Profile \[
-\rho(r) = \frac{M}{2\pi} \frac{a}{r} \frac{1}{(r+a)^3}
+The thin and thick disks are represented with the
+\citet{miyamoto+nagai1975} cylindrical potential: \[
+\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
 \]
 
-\subsubsection{Disk}\label{disk}
+where \(a\) is the disc radial scale length, \(b\) is the scale height,
+and \(M\) is the total mass of the disk. For the thin disk,
+\(a=3.944\,\)kpc, \(b=0.311\,\)kpc,
+\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
+\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
+
+The halo is a NFW dark matter halo (REF) with
+\(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.
+
+Variations to the potential of the inner disk (exclusion of a bar)
+should minimally affect our results as no orbit we consider reaches less
+than \textasciitilde15 kpc of the MW centre. We exclude the mass
+evolution of the halo from this analysis. Over \(10\,\)Gyr, this would
+be fairly significant (factor of \(\sim 2\)in MW mass, REF) but since we
+want to determine the upper limit of tidal effects, it is safe to
+neglect this.
+
+Finally, in chapter REF, we consider the influence of the large Milky
+Cloud on the orbits and evolution of Scl. We adopt the
+\citet{vasiliev2024} multipole approximation of an N-body simulation of
+the LMC and MW. Their initial conditions are
 
@@
 \item
-  Thin disk. miyamoto+nagai disk with parameters \(a=3.944\),
-  \(b=0.311\), \(M=3.944\),
+  MW halo:
 \item
-  Thick disk. miyamoto+nagai disk with parameters \(a=4.4\), \(b=0.92\),
-  and \(M=2\)
+  MW bulge (static):
+\item
+  MW disk (static):
+\item
+  LMC halo:
 \end{itemize}
 
-\[
-\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
-\]
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
+\caption[Circular velocity of potential]{Circular velocity profile of
+\citet{EP2020} potential. We also show the MW and LMC potential from
+\citet{vasiliev2024} for their L3M11 model.}\label{fig:v_circ_potential}
+\end{figure}
 
-for convenience, we define the helper variables,
-\(S = \sqrt{z^2 + b^2}\) and \(D = \sqrt{R^2 + (a+S)^2}\), so the
-potential simplifies to \[
-\Phi(R, z) = \frac{-GM}{D}
-\]
+\section{Orbital Estimation}\label{orbital-estimation}
 
-\[
-\rho = \frac{b^2 M}{4\pi} \frac{aR^2 + (a+3r)(a + r)^2}{S^3 D^5}
-\]
+To estimate the orbits of a dwarf galaxy, we perform a Monte Carlo
+sampling of the present-day observables. We integrate each sampled
+observable back in time for 10 Gyr in Gadget as massless point particles
+outputting positions every 5Myr (with otherwise similar parameters to
+n-body runs below.)
 
-\subsubsection{Halo}\label{halo}
-
-Mvir=115, r=20.2, c=9.545
+To convert from Gaia to galactocentric coordinates, we use the
+\citet{astropy} v4 galactocentric frame. This frame assumes the galactic
+centre (Sagittarius A*) appears:
 
@@
 \item
-  NFW Dark matter halo. \(M_s=79.5\) (double check). and \(r_s = 20.2\).
+  J2000 `ra' = 17h45m37.224s, `dec' = -28°56'10.23'',
+  \(\mu_{\alpha*}=-3.151\pm0.018\) mas/yr, \(\mu_\delta=-5.547\pm0.026\)
+  from the appendix and Table 2 of \citet{reid+brunthaler2004}.
+\item
+  distance = \(8.122\pm0.033\,\)kpc, solar radial velocity =
+  \(11 + 1.9 \pm 3\) km/s. from \citet{gravitycollaboration+2018}.
 \end{itemize}
 
-Variations to the potential of the inner disk (exclusion of a bar)
-should minimally affect our results as no orbit we consider reaches less
-than \textasciitilde15 kpc of the MW centre.
-
-\textbf{Figure:} Circular velocity profile of \citet{EP2020} potential.
-
-\subsection{Orbital Estimation}\label{orbital-estimation}
-
-To estimate the orbits of a dwarf galaxy, we perform a Monte Carlo
-sampling of the present-day observables. We integrate each sampled
-observable back in time for 10 Gyr in Gadget as massless point
-particles.
-
-We use the astropy v4 galactocentric frame. This frame assumes
-\ldots\ldots{}
-
-\subsection{Initial conditions}\label{initial-conditions}
-
-We use Agama (cite) to generate initial conditions. We initially assume
-galaxies are described by an NFW dark matter potential and the stars are
-merely collisionless tracers embedded in this potential
-
-\subsubsection{Stellar Probabilities}\label{stellar-probabilities}
-
-We assign stellar probabilities via Eddington inversion using the
-following method:
+Finally, adding that the sun is \(20.8\pm0.3\,\)pc above the disk from
+bennett+bovy2019, and using the procedure described in
+\citet{drimmel+poggio2018}, the solar velocity relative to the galactic
+rest frame is
 
@@
 \item
-  \(\Psi\) is taken as known from the underlying assumed analytic dark
-  matter potential.
-\item
-  \(\rho_\star\) is
+  `v\_sun' = \([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s
 \end{itemize}
 
-\subsubsection{Stellar profiles}\label{stellar-profiles}
+We select the initial conditions from the location of the first
+apocentre of a selected orbit.
 
-We consider several different stellar profiles in this work.
+While we assume that the galaxy can be described as a point particle,
+not subject to (internal) dynamical friction, this approximation works
+well if the galaxy is not strongly affected by tides (like Scl).
+Analytic approximations of dynamical friction tend to be inadequate to
+more accurately model the N-body trajectory.
 
-\paragraph{Exp2D}\label{exp2d}
+\section{N-Body Modelling}\label{n-body-modelling}
+
+\subsection{Initial conditions}\label{initial-conditions}
+
+We use Agama \citep{agama} to generate initial conditions. We initially
+assume galaxies are described by an NFW dark matter potential (REF) and
+the stars are merely collisionless tracers embedded in this potential
+(added on in post-processing). The density is exponentially truncated
+with a profile
+\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
+\rho_{\rm tNFW} = e^{-r/r_t}\ \rho_{\rm NFW}(t)
+}\end{equation} where we adopt \(r_t = 100 r_s\) or approximately 10
+times \(r_{200}\).
+
+\subsection{Isolation runs and simulation
+parameters}\label{isolation-runs-and-simulation-parameters}
+
+To ensure that the initial conditionss of the simulation are dynamically
+relaxed and well-converged, we run the simulation in isolation (no
+external potential) for 5 Gyr (or approximately the crossing timescale).
+Our fiducial isolation halo uses \(r_s=2.76\) kpc and
+\(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for any
+length or mass scale.
+
+We adopt a softening length of \[
+h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
+\] See appendix REF for a discussion of this choice, which is similar to
+the \citet{power+2003} suggested softening.
+
+We use the relative tree opening criterion with the accuracy parameter
+set to 0.005, and adaptive time stepping with integration accuracy set
+to 0.01.
+
+\subsubsection{Numerical convergence}\label{numerical-convergence}
+
+As discussed in \citet{power+2003}, the region where density is
+converged is related to the region which becomes collisionally relaxed
+over the time of the universe (so about 5-10 Gyr). (i.e.~where the
+collisionless assumption breaks down). The relaxation timescale is given
+by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise, \[
+t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
+= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
+\] This works out to be about 6-10 times (increasing with particle
+number) our adopted softening length for NFW halos given our
+assumptions. As such, at full resolution, we can only trust density
+profiles down to \(\sim10\epsilon\), just enough to resolve stellar
+density profiles.
+
+Note that by decreasing the truncation radius, the effective particle
+number is improved, so future experiments could be less generous with
+this parameter to improve computational performance (order 30\% better
+resolution).
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
+\caption[Numerical halo convergence]{Numerical convergence test for
+circular velocity as a function of log radius for simulations with
+different total numbers of particles in isolation. Residuals in lower
+panel are relative to NFW. The initial conditions are dotted and the
+converged radius is marked by arrows
+(REF).}\label{fig:numerical_convergance}
+\end{figure}
+
+\subsection{Orbital runs}\label{orbital-runs}
+
+To perform the simulations of a given galaxy in a given potential, we
+centre the isolation run's final snapshot and place the dwarf galaxy in
+the specified orbit in the given potential. We typically run the
+simulation for 10 Gyr, which allows us to orbit slightly past the
+expected initial conditions.
+
+\section{Post Processing}\label{post-processing}
+
+\subsection{Centring}\label{centring}
+
+Shrinking spheres centres inspired by \citet{power+2003}
+
+\begin{itemize}
+\tightlist
+\item
+  Recursively shrink radius by 0.975 quantile and recalculating centroid
+  until radius is less than \textasciitilde1kpc or fewer than 0.1\% of
+  particles remain.
+\item
+  Remove bound particles (using instantanious potential).
+\item
+  Use previous snapshot centre and acceleration to predict new centre
+  for next snapshot and only include particles included in previous
+  centring timestep.
+\end{itemize}
+
+The statistical centring uncertainty for the 1e7 particle isolation run
+is of order 0.003 kpc, however oscillations in the centre are of order
+0.03 kpc. This is about three times the softening length but is less
+than the numerically converged radius scale.
+
+\subsection{Velocity profiles}\label{velocity-profiles}
+
+\begin{itemize}
+\tightlist
+\item
+  Circular velocity is computed assuming spherical symmetry and only
+  shown for every 200th particle (ranked from the centre outwards)
+\end{itemize}
+
+\[
+v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
+\]
+
+\begin{itemize}
+\tightlist
+\item
+  \(v_{\rm circ\ max}\) and \(r_{\rm circ,\ max}\) is found by
+  least-squares fitting the NFW functional velocity form to the points
+  of the velocity profile that have the 10\% highest velocities. This is
+  not a necessarily good fit, especially as the halo becomes stripped,
+  but accurate enough to find a reliable maximum.
+\end{itemize}
+
+\subsection{Stellar Probabilities}\label{stellar-probabilities}
+
+We assign stellar probabilities (assuming spherical, isotropic symmetry)
+via Eddington inversion.
+
+Let \(\Psi\) be the potential (normalized to vanish at infinity) and
+\({\cal E}\) is the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
+know \(f({\cal E})\), the distribution function (phase-space density in
+energy), then we assign the stellar weight for a given particle with
+energy \({\cal E}\) is \[
+P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
+\] While \(f({\cal E})\) is a phase-space density, the differential
+energy distribution includes an additional \(g({\cal E})\) occupation
+term (BTXXX).
+
+We use Eddington inversion to find the distribution function, (eq.
+4-140b in BT87) \[
+f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
+\]
+
+In practice the right, boundary term is zero as \(\Psi \to 0\) as
+\(r\to\infty\), and if \(\rho \propto r^{-n}\) at large \(r\) and
+\(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\) which goes to
+zero provided that \(n > 1\).
+
+\(\Psi\) is taken as known from the underlying assumed analytic dark
+matter potential. \(\rho_\star\) is pre-specified or calculated via Abel
+integration / deprojection
+
+We consider both a Plummer and 2-dimensional exponential (Exp2D) stellar
+models:
+
+\subsubsection{Exp2D}\label{exp2d}
 
@@
 (e.g.~Rapha\ldots), i.e. \[
-\rho_\star (r) = \int dR = xxx
+\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
+\] where \(K_0\) is the 0th order modified Bessel function of the second
+type.
+
+\subsubsection{Plummer}\label{plummer}
+
+A Plummer profile is defined by \[
+\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
+\] where \(M\) is the total mass and \(R_s\) is the characteristic scale
+radius. The density is \[
+\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
 \]
 
-\paragraph{Plummer}\label{plummer}
+\subsection{Stellar density profiles and velocity
+dispersion}\label{stellar-density-profiles-and-velocity-dispersion}
 
-\subsection{Dark Matter simulations}\label{dark-matter-simulations}
-
-See \citet{power+2003} for a discussion of this.
-
-The gravitational softening is based on: \[
-h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
-\]
-
-I in fact find that dividing the softening by a factor of sqrt(10) gives
-slightly more precise results with a negligible increase in compute
-time.
-
-\section{Appendix: Convergence tests and Simulation
-Parameters}\label{appendix-convergence-tests-and-simulation-parameters}
-
-Here, we describe some convergence tests to ensure our methods and
-results are minimally impacted by numerical limitations and assumptions.
-See \citet{power2003} for a detailed discussion of various assumptions
-and parameters used in N-body simulations.
-
-\subsubsection{Particle Number.}\label{particle-number.}
-
-\subsubsection{Timestepping}\label{timestepping}
-
-\subsubsection{Gravitational methods}\label{gravitational-methods}
-
-\subsubsection{Gravitational accuracy}\label{gravitational-accuracy}
-
-\subsubsection{Fiducial Parameters}\label{fiducial-parameters}
-
-\begin{verbatim}
-# -----------------------------------IO---------------------------------------
-
-#---Filenames
-InitCondFile                initial
-OutputDir                   ./out
-SnapshotFileBase            snapshot
-OutputListFilename          outputs.txt
-
-#---File formats 
-ICFormat                    3       # use HDF5
-SnapFormat                  3
-
-#---Mem & CPU limits
-TimeLimitCPU                86400   # TUNE
-CpuTimeBetRestartFile       7200    # TUNE 
-MaxMemSize                  2400    # TUNE to system (MB)
-
-#---Time
-TimeBegin                   0
-TimeMax                       2120      # 10 Gyr
-
-#---Output frequency
-OutputListOn                0
-TimeBetSnapshot             10      # change as needed
-TimeOfFirstSnapshot         0       # MATCH to TimeBegin
-TimeBetStatistics           10      # change as needed 
-NumFilesPerSnapshot         1
-MaxFilesWithConcurrentIO    1
-
-# ----------------------------------Gravity-----------------------------------
-
-# MATCH all the below if continuing run
-
-#---Timestep accuracy
-ErrTolIntAccuracy           0.01    # MATCH
-CourantFac                  0.1     # ignored; for SPH
-MaxSizeTimestep             0.5     # MATCH
-MinSizeTimestep             0.0
-
-#---Tree algorithm
-TypeOfOpeningCriterion      1       # 0: Barnes-Hut, 1: Relative
-ErrTolTheta                 0.5     # mostly used for Barnes-Hut
-ErrTolThetaMax              1.0     # (used only for relative)
-ErrTolForceAcc              0.005   # (used only for relative)
-
-#---Domain decomposition: should only affect performance
-TopNodeFactor                       3.0
-ActivePartFracForNewDomainDecomp    0.02
-
-#---Gravitational Softening
-SofteningComovingClass0      # FILL IN
-SofteningMaxPhysClass0       0      # ignored; for cosmological
-SofteningClassOfPartType0    0
-SofteningClassOfPartType1    0
-
-#--------------------------------Misc-------------------------------------
-
-# probably do not need to change the options below
-
-#---Unit System
-UnitLength_in_cm            1 
-UnitMass_in_g               1
-UnitVelocity_in_cm_per_s    1 
-GravityConstantInternal     1
-
-#---Cosmological Parameters 
-ComovingIntegrationOn     0 # no cosmology
-Omega0                    0
-OmegaLambda                 0 
-OmegaBaryon                 0
-HubbleParam                 1
-Hubble                      100
-BoxSize                     0
-
-#---SPH
-ArtBulkViscConst             0.8
-MinEgySpec                   0
-InitGasTemp                  100
-
-#---Initial density estimate (SPH)
-DesNumNgb                   64
-MaxNumNgbDeviation          1
-\end{verbatim}
+\begin{itemize}
+\tightlist
+\item
+  Stellar velocity dispersion is calculated for all stars within 1kpc
+  (3D) of the centre by assuming
+  \(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
+  i.e.~isotropic velocity dispersion
+\item
+  We calculate density profiles similar to stars and assume a constant
+  bin width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\)
+  (Freedman-Diaconis prescription). \(R\) may be either the cylendrical
+  radius in \(x-y\) or the on-sky project \(\xi, \eta\) tangent plane
+  coordinates.
+\item
+  Because our dwarfs are assumed to be spherical/isotropic, we retain
+  this assumption when calculating predicted density profiles.
+\end{itemize}

### 2025-04-15 thesis/numerical_convergence.tex
thesis updates

Here, we describe some convergence tests to ensure our methods and
results are minimally impacted by numerical limitations and assumptions.
See \citet{power+2003} for a detailed discussion of various assumptions
and parameters used in N-body simulations.

\subsubsection{Softening}\label{softening}

\citet{power+2003} suggest the empirical rule that the ideal softening
(balancing integration time and only compromising resolution in
collisional regime) is \[
h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
\]

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions. However, this choice likely unnecessarily increases
computation time for the relative gain in accuracy.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_softening.png}}
\caption{Softening convergence}\label{fig:softening_convergence}
\end{figure}

\subsubsection{Timestepping and force
accuracy}\label{timestepping-and-force-accuracy}

In general, we use adaptive timestepping and relative opening criteria
for gravitational force computations. To verify that these choices and
associated accuracy parameters minimally impact convergence or speed, we
show a few more isolation runs (using only 1e5 particles)

\begin{itemize}
\tightlist
\item
  constant timestep (\ldots), approximantly half of minimum timestep
  with adaptive timestepping
\item
  geometric opening, with \(\theta = 0.5\).
\item
  strict integration accuracy, (facc = \ldots.)
\end{itemize}

\subsubsection{Alternative methods}\label{alternative-methods}

\begin{itemize}
\tightlist
\item
  FMM
\item
  PMM-tree
\item
  Gadget2
\item
  etc.
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_methods.png}}
\caption{Isolation method convergence}\label{fig:methods_convergence}
\end{figure}

\subsubsection{Fiducial Parameters}\label{fiducial-parameters}

Note that we use code units which assume that \(G=1\) for convenience
and numerical stability. The conversion between code units to physical
units is (for our convention):

\begin{itemize}
\tightlist
\item
  1 length = 1 kpc
\item
  1 mass unit = \(10^{10}\) Msun
\item
  1 velocity unit = 207.4 km/s
\item
  1 time unit = 4.715 Myr
\end{itemize}

Most parameters below are not too relevant or have been discussed or are
merely dealing with cpu and IO details. The changes between simulation
runs primarily affect the integration time, output frequency, and
softening. Otherwise, we leave all other parameters fixed.

\begin{verbatim}
#======IO parameters======

#---Filenames
InitCondFile                initial
OutputDir                   ./out
SnapshotFileBase            snapshot
OutputListFilename          outputs.txt

#---File formats 
ICFormat                    3       # use HDF5
SnapFormat                  3

#---Mem & CPU limits
TimeLimitCPU                86400
CpuTimeBetRestartFile       7200
MaxMemSize                  2400

#---Time
TimeBegin                   0
TimeMax                    2120      # 10 Gyr

#---Output frequency
OutputListOn                0
TimeBetSnapshot             10
TimeOfFirstSnapshot         0 
TimeBetStatistics           10
NumFilesPerSnapshot         1
MaxFilesWithConcurrentIO    1

#=======Gravity======

#---Timestep accuracy
ErrTolIntAccuracy           0.01
CourantFac                  0.1     # ignored; for SPH
MaxSizeTimestep             0.5
MinSizeTimestep             0.0

#---Tree algorithm
TypeOfOpeningCriterion      1       # Relative
ErrTolTheta                 0.5     # mostly used for Barnes-Hut
ErrTolThetaMax              1.0     # (used only for relative)
ErrTolForceAcc              0.005   # (used only for relative)

#---Domain decomposition: should only affect performance
TopNodeFactor                       3.0
ActivePartFracForNewDomainDecomp    0.02

#---Gravitational Softening
SofteningComovingClass0      0.044  # HALO dependent
SofteningMaxPhysClass0       0      # ignored; for cosmological
SofteningClassOfPartType0    0
SofteningClassOfPartType1    0

#=======Miscellanius=======

# probably do not need to change the options below

#---Unit System
UnitLength_in_cm            1 
UnitMass_in_g               1
UnitVelocity_in_cm_per_s    1 
GravityConstantInternal     1

#---Cosmological Parameters 
ComovingIntegrationOn      0 # no cosmology
Omega0                     0
OmegaLambda                 0 
OmegaBaryon                 0
HubbleParam                 1
Hubble                      100
BoxSize                     0

#---SPH
ArtBulkViscConst             0.8
MinEgySpec                   0
InitGasTemp                  100

#---Initial density estimate (SPH)
DesNumNgb                   64
MaxNumNgbDeviation          1
\end{verbatim}

### 2025-04-15 thesis/compile.sh

@@ -1,10 +1,13 @@
 set -e 
 rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
-pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib --filter ./table.py --filter ./md_figure.py"
+pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py"
 
 pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
+pandoc $pandocoptions -o numerical_convergence.tex numerical_convergence.md
+
+
 latexmk -f

### 2025-04-17 thesis/data.tex
updated data!

@@
 Gaia provides unprecedented accuracy in proper motions and magnitudes.
-As such, Gaia data is uniquely excellent to produce low-contamination
-samples of likely member stars belonging to satellites. Here, we breifly
-describe J+24's membership estimation and discuss how this informs our
+Gaia data is uniquely excellent to produce low-contamination samples of
+likely member stars belonging to satellites. Here, we breifly describe
+J+24's membership estimation and discuss how this informs our
 observational knoledge of each galaxies density profile. In general,
@@
 
+\section{Radial velocity modeling}\label{radial-velocity-modeling}
+
+For both Sculptor and Ursa Minor, we construct samples of radial
+velocity measurements and cross match each sample to produce estimates
+of the total velocity dispersion and line of sight velocity for each
+galaxy.
+
+\subsection{Methodology}\label{methodology}
+
+First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
+not report GaiaDR3 source ID's, we match to the nearest star within 2
+arcseconds. We exclude stars not matched to Gaia for simplicity.
+
+We combine the mean RV measurement from each study using the
+inverse-variance weighted mean and standard uncertainty. \[
+\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} \\
+\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
+\] where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
+deviation \[
+s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
+\] We remove stars with significant velocity dispersions as measured
+between or within a study: \[
+s < 5\,\delta v\,\sqrt{n}
+\] where \(s, \delta v, n\) are the standard deviation, standard, error,
+and number of measurements for both a study (with multiple epochs) and
+inter-study comparisons.
+
+The combined RV likelihood is then \[
+{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
+\] where \[
+{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2) \\
+{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
+\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
+dispersion, and \(\delta v_i\) is the individual measurement
+uncertainty. Typically, the velocity dispersion will dominate the
+uncertainty budget here. We assume a halo/background velocity dispersion
+of a constant \$\sigma\_\{\rm halo\} = 100 \$ km/s
+\citep[e.g.][]{brown+2010}.
+
+Similar to above, we retain stars with the resulting membership
+probability of greater than 0.2.
+
+Finally, we need to correct the coordinate frames for the solar motion
+and on-sky size of the galaxy. The first step is to subtract out the
+solar motion from each radial velocity, corresponding to a typical
+gradient of \textasciitilde3 km/s across the field. The next step is to
+account for the slight differences in the direction of each radial
+velocity. Define the \(\hat z\) direction to point parallel to the
+direction from the sun to the centre of the dwarf galaxy. Then if
+\(\phi\) is the angular distance between the centre of the galaxy and
+the individual star, the corrected radial velocity is then \[
+v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
+\] where
+\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
+is the radial component of the proper motion with respect to the centre
+of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
+induces a gradient of about \(1 km/s/degree\) for sculptor
+\citep[see][]{WMO2008}. The uncertainty is then the velocity uncertainty
+plus the distance uncertainties times the PM uncertainty from above. We
+then use the \(v_z\) values for the following modelling, however
+repeating with plain, solar-frame velocities does not substantially
+affect the results too much .
+
+For the priors on the satellite velocity dispersion and systematic
+velocity, we use \[
+\mu_{v} = N(0, \sigma_{\rm halo}^2) \\
+\sigma_{v} = U(0, 20\,{\rm km\,s^{-1}})
+\] where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is the velocity
+dispersion of the MW halo adopted above, a reasonable assumption for
+dwarfs in orbit around the MW.
+
+\subsection{Sculptor}\label{sculptor}
+
+\begin{table*}[t]
+\centering
+\caption{Summary of velocity measurements and derived properties.}
+\label{tbl:Summary-of-velocity-measurements-and-derived-properties}
+\begin{tabular}{llllllll}
+\toprule
+ & Study & Instrument & Memb & Rep. & N$\sigma > s$ & $\bar v$ & $\sigma_v$\\
+\midrule
+Scl &  &  &  &  &  &  & \\
+ & tolstoy+23 & VLT/FLAMES & 200 (1604) & 500 &  &  & 9.6\\
+ & sestito+23a & GMOS & 2000 &  &  &  & -\\
+ & walker+09 & MMFS & 2000 &  &  &  & 9.5\\
+ & APOGEE & APOGEE & 200 & 200 &  & 8 & 9\\
+UMi &  &  &  &  &  &  & \\
+ & sestito+23b & GRACES & 6 &  &  &  & \\
+ & pace+20 & Keck/DEIMOS &  &  &  & -245 & 8.6\\
+ & spencer+18 & MMT/Hectoshell &  &  &  & -247 & \\
+ & APOGEE & APOGEE & 100 & 100 &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+For Sculptor, we combine radial velocity measurements from APOGEE,
+\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
+\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
+measurements. We find that there is no significant velocity shift in
+crossmatched stars between catalogues. After crossmatching to Gaia and
+excluding significant inter-study dispersions, we have a sample of XXXX
+members.
+
+We derive a systemic velocity of \(111.2\pm0.2\) km/s with velocity
+dispersion \(9.67\pm0.16\) km/s. Our values are very consistent with
+previous work \citep[e.g.][\citet{arroyo-polonio+2024},
+\citet{tolstoy+2023}]{WMO2009}.
+
+Finally, we attempt to fit a linear velocity gradient to Scl by adding
+parameters for the gradient in \(\xi\) ande \(\eta\). We derive a
+gradient of xxx and xxx (see Figure\textasciitilde FIG.), noting that
+this direction is different and higher in magnitude than the proper
+motion of Scl. Comparing the density difference, we find a bayes factor
+of -15. Compared to past work, \citet{battaglia+2008},
+\citet{arroyo-polonio+2021}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vlos_xi_eta.pdf}}
+\caption[Scl velocity sample]{A plot of the corrected los velocities for
+Scl binned in tangent plane coordinates. We detect a slight rotational
+gradient towards the bottom right. \textbf{TODO} add a histogram of
+velocities here too with fits, maybe with respect to distance.}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient.pdf}}
+\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
+arrow marks the gradient induced by Scl's proper motion on the sky.
+\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
+\end{figure}
+
+\emph{Is it interesting that the velocity dispersion of Scl seems to
+increase significantly with Rell?}
+
+\subsection{Ursa Minor}\label{ursa-minor}
+
+For UMi, we collect radial velocities from, APOGEE,
+\citet{sestito+2024b}, \citet{pace+2020}, and \citet{spencer+2018}.
+
+We shifted the velocities of \citet{spencer+2018} (\(-0.9\) km/s) and
+\citet{pace+2020} (\(+1.0\) km/s) to reach the same scale based on a
+crossmatch of about 200 common stars. Since the mean difference in
+velocities in this crossmatch is of order 1.9 km/s, we adopt this as the
+systematic LOS velocity error.
+
+UMi interestingly has a more structured observational pattern, but does
+not appear to have any significant velocity substructure.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_vlos_xi_eta.pdf}}
+\caption[UMi velocity sample]{A plot of the corrected los velocities for
+UMi binned in tangent plane coordinates. There is no clear rotation or
+velocity gradient here. Interestingly, many velocity members are as far
+as 100 arcmin away.}
+\end{figure}
+
+\subsection{Discussion and
+limitations}\label{discussion-and-limitations}
+
+Our model here is relatively simple. Some things which we note as
+systematics and are challenging to account fully for are
+
+\begin{itemize}
+\tightlist
+\item
+  Inter-study systematics and biases. While basic crossmatches and a
+  simple velocity shift, combining data from multiple instruments is
+  challenging.
+\item
+  Inappropriate uncertainty reporting. Inspection of the variances
+  compared to the standard deviations within a study seems to imply that
+  errors are accurately reported. APOGEE notes that their RV
+  uncertainties are known to be underestimates.
+\item
+  Binarity. While not too large of a change for classical dwarfs, this
+  could inflate velocity dispersions of about 9 km/s by about 1 km/s
+  \citet{spencer+2017}. Thus, our measurement is likely slightly
+  inflated given the high binarity fractions measured in classical
+  dwarfs \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
+\item
+  Selection effects. RV studies each have their own selection effects. I
+  do not know how to correct for this.
+\end{itemize}
+
+Because the derived parameters are similar for the two different larger
+surveys we consider for UMi and Scl, we note that many of these effects
+are likely not too significant (with the exception of the systemic
+motion of UMi.)
+
 \section{Appendix / Extra Notes}\label{appendix-extra-notes}
@@
 \end{verbatim}
-
-\section{Radial velocity modeling}\label{radial-velocity-modeling}
-
-For both Sculptor and Ursa Minor, we construct samples of radial
-velocity measurements and cross match each sample to produce estimates
-of the total velocity dispersion and line of sight velocity for each
-galaxy.
-
-\subsection{Methodology}\label{methodology}
-
-Our RV methodology is
-
-\begin{itemize}
-\tightlist
-\item
-  Cross match all catalogues to J+24 Gaia members. If Gaia ID's are
-  included, we use these. Otherwise, we match to the nearest star within
-  2 arcseconds.
-\item
-  We combine the mean RV measurement from each study using the
-  inverse-variance weighted mean and standard uncertainty.
-\end{itemize}
-
-\[
-v_i = \frac{1}{\sum w}\sum_j w\ v_{i, j} \\
-\delta v_i = \sqrt{\frac{1}{\sum w^2}}
-\]
-
-where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
-deviation \[
-s_{v}^2 = \frac{1}{\sum w} \sum w (v_{i, j} - v_i)^2
-\] Note that we do not include the standard deviation if a source does
-measure the velocity multiple times.
-
-If the total standard deviation is greater than five times the velocity
-standard uncertainty (reduced), then we remove the star since either the
-star is a binary or contains erroneous measurements or other
-uncertainties. Note that since stars typically only have very few
-measurements, these choices are highly approximate and likely limited.
-
-The combined RV likelihood is then \[
-{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
-\] where \[
-{\cal L}_{\rm RV} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
-\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
-dispersion, and \(\delta v_i\) is the individual measurement
-uncertainty. Typically, the velocity dispersion will dominate the
-uncertainty budget here.
-
-Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2.
-
-Finally, we need to correct the coordinate frames. We need a proper
-motion estimate of each star. Since most stars have Gaia measured proper
-motions, but the uncertainties on each proper motion measurement are
-large, we combine the system and individual star's proper motion by
-weighting the systemic proper motion by the z-score of the consistency
-of the measurement with the systemic proper motion. \[
-\mu_{*}' = \frac{1}{w_\star + w_{\rm sys}}\left(\mu_{\star}w_\star + \mu_{\rm sys} w_{\rm sys}\right)
-\] where we model \[
-{w_{\rm sys}} = P_{\rm sat}(\delta \mu ^2 + \sigma_{v, tan}^2)^{-1}
-\] and the weight of the star is just the inverse variance weight plus
-the gaia systematics.
-
-The corrected radial velocities are then \[
-v_z = v_{\rm los, gsr}\cos\theta  + v_{\rm tan, R} \sin\theta
-\] where
-\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
-is the radial component of the proper motion with respect to the centre
-of the galaxy. Since \(\theta \lesssim 1^\circ\), this correction is
-small (\(\sim 1\)km,s\(^{-1}\)), but can prevent the introduction of a
-spurious gradient on the sky \citep{WMO2008}. The uncertainty is then
-the velocity uncertainty plus the distance uncertainties times the PM
-uncertainty from above. We then use the \(v_z\) values for the following
-modelling, however repeating with plain, solar-frame velocities does not
-substantially affect the results too much .
-
-\subsection{Sculptor}\label{sculptor}
-
-For Sculptor, we combine radial velocity measurements
-
-\begin{itemize}
-\item
-  APOGEE (crossmatched to stars meeting basic quality cuts in field)
-\item
-  \citet{sestito+2024a}
-\item
-  \citet{tolstoy+2024}
-\end{itemize}
-
-We adopt a very loose prior of \[
-\mu_{v} = N(111, 10) \\
-\log\sigma_{v} = N(1, 1)
-\] Our resulting velocity dispersions are \$\$
-
-\$\$ Comparing to past work, note that our values are reasonable.
-
-\subsection{Ursa Minor}\label{ursa-minor}
-
-\begin{itemize}
-\tightlist
-\item
-  APOGEE
-\item
-  \citet{sestito+2024b}
-\item
-  \citet{pace+2022}
-\item
-  \citet{spencer+2018}
-\end{itemize}
-
-\subsection{Discussion and
-limitations}\label{discussion-and-limitations}
-
-Our model here is relatively simple. Some things which we do not account
-for include
-
-\begin{itemize}
-\tightlist
-\item
-  interstudy systematics and biases
-\item
-  Inappropriate uncertainty reporting.
-\item
-  Binarity. Binay stars are expected to induce an average velocity
-  dispersion of xxx on an individual star.
-\item
-  Selection effects. RV studies each have their own targeted program and
-\end{itemize}
-
-We partially address several of these by also performing the analysis on
-large RV samples for each galaxy. This allows for individual comparison
-illustrating the differences in selection effects.

### 2025-04-17 thesis/introduction.tex
updated data!

@@
 \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/power_spectrum.png}
-\caption[Cosmological Power Spectrum]{figure 1 from
+\caption[Cosmological Power Spectrum]{The matter power spectrum under
+different assumptions for dark matter. Dwarf galaxies occupy the middle
+and low end of the blue region (10\^{}10 - 10\^{}8 solar masses),
+enabling a unique window into properties of dark matter on small scales.
+The smaller scales we can understand dark matter, the better we are able
+to test different models of dark matter. figure 1 from
 \citet{bechtol+2022}.}\label{fig:cosmological_power_spectrum}
@@
 \centering
-\caption{Measured properties of Sculptor}
-\label{tbl:Measured-properties-of-Sculptor}
+\caption{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018), 2. Tran et al. (2022), 3. (MV2020b?), 4 McConnachie and Venn (2020).}
+\label{tbl:Observed-properties-of-Sculptor-References-are:-1-Muñoz-et-al-2018-2-Tran-et-al-2022-3-MV2020b?-4-McConnachie-and-Venn-2020}
 \begin{tabular}{lll}
@@
 \midrule
-$\alpha$ & $15.0183 \pm 0.0012$˚ & M+18\\
-$\delta$ & $-33.7186 \pm 0.00072$˚ & M+18\\
-distance & $83.2 \pm 2$ kpc & Tran+22\\
-$\mu_\alpha \cos \delta$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
-$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
-RV & $111.2 \pm 0.2$ & arroyo-polonio+24\\
-$\sigma_v$ & $9.7\pm0.2$ & arroyo-polonio+24\\
-$R_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
+$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
+$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
+distance modulus & $19.60 \pm 0.05$ (RRL) & 2\\
+distance & $83.2 \pm 2$ kpc & 2\\
+$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
+radial velocity & $111.2 \pm 0.2\ {\rm km\,s^{-1}}$ & this work\\
+$\sigma_v$ & $9.67\pm0.16\ {\rm km\,s^{-1}}$ & this work\\
+$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
 $R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
-ell & $0.36 \pm 0.01$ & M+18\\
-PA & $92\pm1$ & M+18\\
-$M_V$ & $-10.82\pm0.14$ & M+18\\
+ellipticity & $0.36 \pm 0.01$ & 1\\
+position angle & $92\pm1^\circ$ & 1\\
+$M_V$ & $-10.82\pm0.14$ & 1\\
 \bottomrule
@@
 \end{table*}
-
-Can I just cite \citet{arroyo-polonio+2024} for velocity information for
-Sculptor? They correct for binary and do a more detailed discussion of
-rotation.
 
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
-\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. Composite Public
-data/Giuseppe Donatiello. Updated image in 2022.
+\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. UMi appears as a
+diagonal/elliptical haze of stars from the top left to the bottom right.
+Even as as classical dwarf, Ursa minor is fairly diffuse and does not
+stand obviously out from the background. Composite Public data/Giuseppe
+Donatiello. Updated image in 2022.
 https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
@@
 \centering
-\caption{Measured properties of Ursa Minor.
+\caption{Observed properties of Ursa Minor.
 References are: (1) Muñoz et al.
@@
 (2018).}
-\label{tbl:Measured-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
+\label{tbl:Observed-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
 \begin{tabular}{lll}
@@
 $\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
-$\delta$ & $67.2221 \pm 0.00158$˚ & 1\\
-distance modulus & $19.23 \pm 0.11$ & 2\\
-distance & $70.1 \pm 3.6$ kpc & \\
-$\mu_\alpha \cos \delta$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
+distance modulus & $19.23 \pm 0.11$ (RRL) & 2\\
+distance & $70.1 \pm 3.6$ kpc & -\\
+$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
 $\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-RV & $-244.9 \pm 0.3_{\rm stat} \pm 1.7_{\rm sys}$ km s$^{-1}$ & 4\\
-$\sigma_v$ & $8.3 \pm 0.3_{\rm stat} \pm 0.4_{\rm sys}$ & 4\\
+radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
+$\sigma_v$ & $8.8 \pm 0.2$ & this work\\
 $r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
-ell & $0.55 \pm 0.01$ & 1\\
-PA & $50 \pm 1^\circ$ & 1\\
+ellipticity & $0.55 \pm 0.01$ & 1\\
+position angle & $50 \pm 1^\circ$ & 1\\
 $M_V$ & $-9.03 \pm 0.05$ & 1\\
@@
 \caption[Break radius validation]{The break radius of the simulations is
-set by}\label{fig:idealized_break_radius}
+set by the time since pericentre.}\label{fig:idealized_break_radius}
 \end{figure}
@@
 found from either of the equations above.
+
+\section{What is up with
+probabilities}\label{what-is-up-with-probabilities}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250416112038183.png}
+\caption{image-20250416112038183}\label{fig:test}
+\end{figure}

### 2025-04-24 thesis/data.tex
tests to validate gaia filters and thesis updates

@@
 J+24 select stars initially from Gaia within a 2--4 degree circular
-region centred on the dwarf satisfying:
+region centred on the dwarf. J+24 only consider stars with:
 
@@
 \item
-  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3})
+  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3}).
 \item
@@
   10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
-  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
+  \(\gtrsim 500\) km/s at distances larger than 10 kpc.).
 \item
-  No colour excess (\citet{lindegren+2018} equation C.2)
+  No colour excess (\citet{lindegren+2018} equation C.2).
 \item
-  G \textless{} 22 and less than 5\(\sigma\) above TRGB, and between
-  -0.5 and 2.5 in Bp - Rp.
+  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB (in
+  practice astrometry ruwe =\textgreater{} 21 \textgreater{} G).
+\item
+  Between -0.5 and 2.5 in Bp - Rp.
 \end{itemize}
 
-Photometry is dereddened using \citet{schlegel+1988} extenction.
+Photometry is dereddened using \citet{schlegel+1988} extinction maps.
 
 J+24 calculate the probability that any star belongs to either the
-satellite or the MW background as \[
+satellite or the MW background with \[
 P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
@@
 
-where the satellite (sat) and background (bg) likelihoods are simply the
-product of the PM, CMD, and spatial components: \[
-{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}
+where \(f_{\rm sat}\) is the fraction of candidate members in the field,
+and \({\cal L}_{\rm sat}\) and \({\cal L}\)\_\{\rm bg\}\$ are the
+satellite (sat) and background (bg) likelihoods respectively. Each
+likelihood is the product of a spatial, proper motion, and CMD term: \[
+{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
 \]
 
-The satellite likelihood is constructed as
+The satellite likelihoods are specified as
 
@@
 
-The background likelihood is constructed as:
+The background likelihoods are constructed as:
 
@@
 \item
-  PM: same as CMD except in PM space.
+  PM: same as CMD except in PM space (\(\mu_{\alpha*}\) and
+  \(\mu_\delta\) with respecive uncertainties and covariance.)
 \item
@@
 
-Note that each likelihood map is normalized over the respective 2D
-parameter space. In order to represent the difference in frequency of
-background and forground stars, \(f_{\rm sat}\) represents the field
-fraction of member stars.
+Each likelihood map is normalized over the respective 2D parameter
+space. \(f_{\rm sat}\) thus is the only term controlling the total
+abundance of satellite stars relative to the background.
 
-In J+24, a MCMC simulation is ran using the above likelihood to solve
-for the following parameters
+In J+24, a MCMC simulation is ran using the above total likelihood to
+solve for the following parameters:
 
@@
 even for a probability cut of 0.2, the purity of the resulting sample
-with RV measurements is very high (\textasciitilde90\%, J+24). (Note:
-there is likely a high systematic bias in using stars with RV
-measurements to measure purity. Fainter stars have poorer measurements
-and distant stars are less likely to have been targeted. )
+with RV measurements is very high (\textasciitilde90\%, J+24). However,
+there is likely a systematic bias in using stars with RV measurements to
+measure purity. Fainter stars have poorer astrometry and are less likely
+to have been targeted. We find no difference in the resulting density
+distributions when restricting stars to be brighter than a specific
+magnitude.
 
-\subsection{Resulting Samples}\label{resulting-samples}
+\subsection{Selected samples}\label{selected-samples}
 
@@
 fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
-the algorithm. In each case, each criteria plays a roll: proper motions
-are centred around the dwarf systemic motion, the CMD is well defined,
-and stars only within a few \(R_h\) are included. We also plot the RV
-members found in general and in \citet{sestito+2024},
-\citet{sestito+2024b}.
+the algorithm in the tangent plane (\(\xi\), \(\eta\), \emph{does this
+need defined?}), CMD, and proper motion space. For both galaxies, each
+criteria plays a commensurate role in sifting out members. Proper
+motions are clustered around the dwarf systemic motion, the CMD is well
+defined including the horizontal branch, and stars only within a few
+\(R_h\) are included. We also mark stars with consistent radial
+velocities (RVs) (see below), which trace each feature albeit more
+cautiously.
 
-The tangent plane, \(\xi\), \(\eta\), is the projection that
-
-We also illustrate the approximate result of removing the spatial
-component from the likelihood. We define the CMD+PM selection as stars
-satisfying \[
+Figure REF also illustrats the distribution of stars selected without a
+spatial criterion. We define the CMD+PM selection as stars satisfying \[
 {\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
-\]
-
-These stars appear similar to the fiducial (probable members) sample,
-but instead also appear as an approximately uniform distribution across
-the entire field. This illustrates the approximate background of stars
-which may be confused as members. Additionally, since there is no clear
+\] These stars are distributed similar to the fiducial (probable
+members) sample but with a fainter uniform distribution across the
+entire field. This illustrates the approximate background of stars which
+may be confused as members. Additionally, since there is no clear
 spatial structure in the CMD+PM sample, it is unlikely that there are
-tidal tails detectable with Gaia. Not shown here, we also try a variety
-of simpler, absolute cuts and thresholds, finding no extended structure
-beyond what is detected in J+24.
-
-This means that at least at the level of where the background density
-dominates, we can exclude models which produce tidal tails brighter than
-a density of
-\(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
-(TODO assuming a distance of \ldots{} and stellar mass of \ldots).
+additional faint, tidal features detectable with Gaia. Not shown here,
+we also try a variety of simpler, absolute cuts and thresholds, finding
+no extended structure beyond what is detected in J+24. (\emph{Is it
+worth using this to calculate an OOM upper limit on the density of tidal
+tails?}).
 
@@
 (satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
-plane. \textbf{Bottom left:} Colour magnitude diagram. \textbf{Bottom
-right:} Proper motion.}\label{fig:sculptor_selection}
+plane. We outline in star symbols the two stars from
+\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram.
+\textbf{Bottom right:} Proper motion.}\label{fig:sculptor_selection}
 \end{figure}
@@
 \caption[Ursa Minor Selection]{Similar to
-fig.~\ref{fig:sculptor_selection} except for Ursa Minor. UMi features a
-very extended density profile with some stars \textasciitilde{} 6\(R_h\)
-including a RV member. UMi is also highly elliptical compared to other
-classical dwarfs.}\label{fig:umi_selection}
+fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
+members outside of \(3R_h\) in black stars (from \citet{sestito+2023b}
+and \citet{pace+2020} and
+\citet{spencer+2018}).}\label{fig:umi_selection}
 \end{figure}
@@
 
-Our primary observational constraint is the density profile of a dwarf
-galaxy.
+Density profiles are an essential observational constraint for our later
+simulations. To derive density profiles, we use 0.05 dex bins in log
+radius (i.e.~the bins are derived from
+10\^{}(minimum(logR):0.05:maximum(logR))). We remove bins at smaller
+(larger) radii than the first bin to contain no stars, working outwards.
+We use poisson uncertainties
+(\(\delta \Sigma_i / \Sigma_i = 1/\sqrt{N_i}\) for a derived density
+\(\Sigma_i\) in a bin with \(N_i\) stars). As discussed below, these
+uncertainties are straightforward but are likely under-represented.
 
-To derive density profiles, we use 0.05 dex bins in log radius (i.e.~the
-bins are derived from 10\^{}(minimum(logR):0.05:maximum(logR))). The
-density in each bin is then (from Poisson statistics) \[
-\Sigma_b = N_{\rm memb} / A_{\rm bin} \pm \sqrt{N_{\rm memb}} / A_{\rm bin}
-\] where \(N_{\rm memb}\) is the number of members in the bin and
-\(A_{\rm bin}\) is the area of the bin's annulus in 2D. As discussed
-below, these uncertainties underrepresent the true uncertainty on
-multiple accounts. We retain Poisson errors for simplicity here.
-
-In Figures fig.~\ref{fig:scl_observed_profiles},
-fig.~\ref{fig:umi_observed_profiles}, we show the derived density
-profiles for each galaxy for samples similar to in the selection plots
-above. In each case, all samples are the same towards the inner regions
-of the satellite, illustrating that these density profiles are dominated
-by satellite stars in the centre. The sample containing all stars
-reaches a plateau at the total background in the field. However,
+fig.~\ref{fig:scl_observed_profiles},
+fig.~\ref{fig:umi_observed_profiles} show the derived density profiles
+for Scl and UMi. We calculate density profiles for different selections
+of stars from above: all quality stars, CMD+PM only, and the probable
+member samples. In each case, all samples are the same towards the inner
+regions of the satellite. Classical dwarfs dominate the stellar density
+in their cores. However, the density profile \emph{all stars} plateaus
+at the total background in the field at radii of 30-60 arcminutes. By
 restricting stars to being most likely satellite members by CMD + PM,
-the background is much lower. This plateau likely represents the real
-background of background stars which could be mistaken as members.
-Finally, we have the probable members and bg-subtracted densities. BG
-subtracted is based on the \texttt{all} density profile, subtracting the
-mean background density for stars beyond the last point of the BG
-subtracted profile.
+the background is reduced by 1-2 dex. The CMD+PM background plateau
+likely represents the density of background stars which could be
+mistaken as members. Finally, the background (BG)-subtracted profile
+results from subtracting the apparent background in the \emph{all}
+profile.
 
 Note that the probable members (fiducial) density profile continues to
-confidently estimate the density profile below the CMD+PM likely star
-background. These points are likely unreliable (see discussion below).
-However, before this point, both the BG subtracted and probable members
-density profiles are strikingly similar. Assumptions about the details
-of the likelihood and spatial dependence have marginal influence on the
+confidently estimate the density profile below the CMD+PM background.
+These points are likely unreliable (see discussion below). However,
+before this point, both the BG subtracted and probable members density
+profiles are strikingly similar. Assumptions about the details of the
+likelihood and spatial dependence have marginal influence on the
 resulting density profile when the satellite is higher density than the
-background.
+background. Thus the detection of excesses of stars in Sculptor (past 20
+arcmin) and UMi (past 15-20 arcmin) are robust.
+
+One interesting note is for UMi, there is a small ultrafaint star
+cluster at a position of (-100, -15) arcmin \citet{munoz+2012}. However
+this cluster does not have a bright RGB so has few stars brighter than g
+mag of 22. While munoz 1 may contribute about 4 stars to the density
+profile of UMi, this would only apply to the bin at XXX because of the
+small half light radius. We do not subtract this out.
 
@@
 
-\section{Comparison of the Classical
-dwarfs}\label{comparison-of-the-classical-dwarfs}
-
-Classical dwarfs are often the brightest dwarfs in the sky.c The density
-profiles of classical dwarf galaxies is thus well measured, enabling
-detailed comparisons.
-
-\begin{table*}[t]
-\centering
-\begin{tabular}{lll}
-\toprule
-galaxy & R\_h (exp inner 3Rs) & num cand\\
-\midrule
-Fornax & $17.8\pm0.6$ & 23,154\\
-Leo I & $3.7\pm0.2$ & 1,242\\
-Sculptor & $9.4\pm0.3$ & 6,888\\
-Leo II & $2.4\pm0.3$ & 347\\
-Carina & $8.7\pm0.4$ & 2,389\\
-Sextans I & $20.2 \pm0.9$ & 1,830\\
-Ursa Minor & $11.7 \pm 0.5$ & 2,122\\
-Draco & $7.3\pm0.3$ & 1,781\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-Using the same methods above, we select members from J+24's stellar
-probabilities. We use the one-component exponential density profiles
-with
-
-Figure: Density profiles for each dwarf galaxy. Here, we use the
-1-component exponential stellar probabilities from J+24. Dwarf galaxies
-are scaled by our derived \(R_h\) values.
-
-\subsection{Density Profile Reliability and
-Uncertainties}\label{density-profile-reliability-and-uncertainties}
-
-\begin{itemize}
-\tightlist
-\item
-  How well do we know the density profiles?
-\item
-  What uncertainties affect derived density profiles?
-\item
-  Can we determine if Gaia, structural, or algorithmic systematics
-  introduce important errors in derived density profiles?
-\item
-  Using J+24 data, we validate
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Check that PSAT, magnitude, no-space do not affect density profile
-    shape too significantly
-  \end{itemize}
-\item
-  Our ``high quality'' members all have \textgreater{} 50 member stars
-  and do not depend too highly on the spatial component, mostly
-  corresponding to the classical dwarfs
-\end{itemize}
-
-J+24's algorithm takes spatial position into account, assuming either a
-one or two component exponential density profile. When deriving a
-density profile, this assumption may influence the derived density
-profile, especially when the galaxy density is fainter than the
-background of similar appearing stars. To remedy this and estimate where
-the background begins to take over, we also explore a cut based on the
-likelihood ratio of only the CMD and PM components. This is in essence
-assuming that the spatial position of a star contains no information on
-it's membership probability (a uniform distribution like the background)
-
 \subsection{Caveats}\label{caveats}
 
-The J+24 method was designed to determine high probability members for
-spectroscopic followup in particular. Note that we instead care about
-retrieving a reliable density profile.
+The J+24 method was designed in particular to detect the presence of a
+density excess and find individual stars at large radii to be followed
+up. We are more interested in accurately quantifying the density profile
+and size of any perturbations. One potential problem with using J+24's
+candidate members is that the algorithm assumes the density is either
+described by a single or double exponential. If this model does not
+accurately match the actual density profile of the dwarf galaxy, we
+would like to understand how strongly influenced the density profile is
+by this assumption.
 
@@
 PM / CMD, recovering the assumed density profile. As a result, the
-reliability of faint features in these density profiles is questionable
-and a more robust analysis, removing this particular density assumtion,
-would be required to more appropriately represent the knowledge of the
-density profile as the background begins to dominate.
+reliability of these density profiles below the CMD+PM background may be
+questionable. A more robust analysis, removing this particular density
+assumption, would be required to more appropriately represent the
+knowledge of the density profile as the background begins to dominate.
 
-J+24 do not account for structural uncertainties in dwarfs. This is a
-not insignificant source of uncertainty in the derived density profile
-
-We assume constant ellipticity and position angle. Dwarf galaxies, in
-reality, are not necessarily smooth and constant.
-
-\section{Comparison and conclusions}\label{comparison-and-conclusions}
-
-To illustrate the differences between each dwarf galaxy, in
-fig.~\ref{fig:classical_dwarfs_densities}, we compare Scl, UMi, and Fnx
-against exponential and plummer density profiles (\textbf{TODO: state
-these somewhere}). While all dwarfs have marginal differences in the
-inner regions, each dwarf diverges in the outer regions relative to an
-exponential. In particular, while Fnx is underdense, Scl and UMi are
-both overdense, approximately fitting a Plummer density profile instead.
-
-In summary, we have used J+24 data to derive the density profiles for
-Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
-robust against different selection criteria. Both Sculptor (Ursa Minor)
-show strong (weak) evidence for deviations from an exponential profile.
-We will explore a tidal explanation for these features in this work.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fornax_exp_fit.pdf}}
-\caption[Classical dwarf density profiles]{The density profiles of
-Sculptor, Ursa Minor, and Fornax compared to Exp2D and Plummer density
-profiles. Dwarf galaxies are scaled to the same half-light radius and
-density at half-light radius (fit from the inner 3 scale radii
-exponential recursively. )}\label{fig:classical_dwarfs_densities}
-\end{figure}
+J+24 do not account for structural uncertainties in dwarfs for the two
+component case. We assume constant ellipticity and position angle. Dwarf
+galaxies, in reality, are not necessarily smooth and constant. They do
+test an alternative method using circular radii for the extended density
+component, and we find these density profiles are very similar. We also
+assume a constant ellipticity and position angle here.
 
@@
 
-For both Sculptor and Ursa Minor, we construct samples of radial
-velocity measurements and cross match each sample to produce estimates
-of the total velocity dispersion and line of sight velocity for each
-galaxy.
-
-\subsection{Methodology}\label{methodology}
+For both Sculptor and Ursa Minor, we construct literature samples of
+radial velocity measurements. We combine these samples with J+24's
+members to produce RV consistent stars and to compute velocity
+dispersion, systematic velocities, and test for the appearance of
+velocity gradients.
 
@@
 inverse-variance weighted mean and standard uncertainty. \[
-\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} \\
+\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} 
+\] \[
 \delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
-\] where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
+\]
+
+where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
 deviation \[
@@
 between or within a study: \[
-s < 5\,\delta v\,\sqrt{n}
-\] where \(s, \delta v, n\) are the standard deviation, standard, error,
-and number of measurements for both a study (with multiple epochs) and
-inter-study comparisons.
+P\left(\chi2(N-1) < \frac{s^2}{\delta v^2}\right) > 0.001
+\] where \(s, \delta v, N\) are the weighted standard deviation,
+weighted standard error, and number of observations. We apply this cut
+to both within a study and between studies, which typically removes
+stars with reduced \(\chi^2 \gtrsim 2\).
 
@@
 \] where \[
-{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2) \\
+{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
+\] \[
 {\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
-\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
-dispersion, and \(\delta v_i\) is the individual measurement
-uncertainty. Typically, the velocity dispersion will dominate the
-uncertainty budget here. We assume a halo/background velocity dispersion
-of a constant \$\sigma\_\{\rm halo\} = 100 \$ km/s
+\]
+
+where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
+dispersion of the satellite, and \(\delta v_i\) is the individual
+measurement uncertainty. Typically, the velocity dispersion will
+dominate the uncertainty budget here. We assume a halo/background
+velocity dispersion of a constant \(\sigma_{\rm halo} = 100\) km/s
 \citep[e.g.][]{brown+2010}.
@@
 then use the \(v_z\) values for the following modelling, however
-repeating with plain, solar-frame velocities does not substantially
-affect the results too much .
+repeating with uncorrected, heliocentric velocities does not
+significantly affect the results .
 
@@
 
-\subsection{Sculptor}\label{sculptor}
+\subsection{}\label{section}
 
@@
 \centering
-\caption{Summary of velocity measurements and derived properties.}
-\label{tbl:Summary-of-velocity-measurements-and-derived-properties}
+\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
+\label{tbl:Summary-of-velocity-measurements-and-derived-properties-sestito+2023a-number-of-members-depends-on-spatial-model-used}
 \begin{tabular}{llllllll}
 \toprule
- & Study & Instrument & Memb & Rep. & N$\sigma > s$ & $\bar v$ & $\sigma_v$\\
+ & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$\\
 \midrule
-Scl &  &  &  &  &  &  & \\
- & tolstoy+23 & VLT/FLAMES & 200 (1604) & 500 &  &  & 9.6\\
- & sestito+23a & GMOS & 2000 &  &  &  & -\\
- & walker+09 & MMFS & 2000 &  &  &  & 9.5\\
- & APOGEE & APOGEE & 200 & 200 &  & 8 & 9\\
-UMi &  &  &  &  &  &  & \\
- & sestito+23b & GRACES & 6 &  &  &  & \\
- & pace+20 & Keck/DEIMOS &  &  &  & -245 & 8.6\\
- & spencer+18 & MMT/Hectoshell &  &  &  & -247 & \\
- & APOGEE & APOGEE & 100 & 100 &  &  & \\
+Scl & combined &  & 8945 & 2280 & 2034 & 1920 & 0.9\\
+ & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1481 & 0.65\\
+ & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13\\
+ & walker+09 & MMFS & 1818 & 1522 & 1417 & 1330 & 1.8\\
+ & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6\\
+UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3\\
+ & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8\\
+ & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5\\
+ & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9\\
+ & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9\\
 \bottomrule
@@
 \end{table*}
+
+\subsection{Sculptor}\label{sculptor}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_2dhist.pdf}}
+\caption[Scl velocity sample]{A plot of the corrected los velocities for
+Scl binned in tangent plane coordinates. We detect a slight rotational
+gradient towards the bottom right.}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_binned.pdf}}
+\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
+arrow marks the gradient induced by Scl's proper motion on the sky.
+\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vlos_xi_eta.pdf}}
-\caption[Scl velocity sample]{A plot of the corrected los velocities for
-Scl binned in tangent plane coordinates. We detect a slight rotational
-gradient towards the bottom right. \textbf{TODO} add a histogram of
-velocities here too with fits, maybe with respect to distance.}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient.pdf}}
-\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
-arrow marks the gradient induced by Scl's proper motion on the sky.
-\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
-\end{figure}
+Bayes factor found using relative density via KDE with Silverman's
+bandwith to prior (XXX theorem).
 
@@
 
+\begin{table*}[t]
+\centering
+\caption{MCMC fits for different RV datasets for Scl amoung 3 different models.}
+\label{tbl:MCMC-fits-for-different-RV-datasets-for-Scl-amoung-3-different-models}
+\begin{tabular}{lllllll}
+\toprule
+study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log bf$\\
+\midrule
+all &  &  &  &  &  & \\
+ & $111.17\pm0.22$ & $9.68\pm0.17$ & - & - & - & 0\\
+ & $111.18 \pm 0.23$ & $9.64\pm0.16$ & - & $4.8\pm1.3$ & $-147_{-12}^{+15}$ & -3.7*\\
+ & $111.15\pm0.22$ & $9.70\pm0.16$ & $0.055\pm0.021$ & - & - & -0.8\\
+tolstoy+23 &  &  &  &  &  & \\
+ & $111.2 \pm 0.3$ & $9.77 \pm 0.18$ & - & - & - & 0\\
+ & $111.2\pm0.3$ & $9.73\pm0.18$ & – & $4.8\pm1.4$ & $-154_{-12}^{+16}$ & -2.5\\
+ & $111.2 \pm 0.3$ & $9.71\pm0.18$ & $0.081 \pm 0.023$ & – & – & -3.3\\
+walker+09 &  &  &  &  &  & \\
+ & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – & 0\\
+ & $111.1\pm0.3$ & $9.54\pm0.21$ & - & $5.3_{-1.6}^{+1.8}$ & $-134_{-16}^{+22}$ & -2.0\\
+ & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – & +1.6\\
+apogee &  &  &  &  &  & \\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – & –\\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ & +0.3\\
+ & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – & +1.1\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
 \subsection{Ursa Minor}\label{ursa-minor}
 
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_rv_2dhist.pdf}}
+\caption[UMi velocity sample]{UMi velocities}
+\end{figure}
+
+Figure: A plot of the corrected los velocities for UMi binned in tangent
+plane coordinates. There is no clear rotation or velocity gradient here.
+Interestingly, many velocity members are as far as 100 arcmin away.
+
 For UMi, we collect radial velocities from, APOGEE,
-\citet{sestito+2024b}, \citet{pace+2020}, and \citet{spencer+2018}.
+\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}.
 
-We shifted the velocities of \citet{spencer+2018} (\(-0.9\) km/s) and
-\citet{pace+2020} (\(+1.0\) km/s) to reach the same scale based on a
-crossmatch of about 200 common stars. Since the mean difference in
-velocities in this crossmatch is of order 1.9 km/s, we adopt this as the
-systematic LOS velocity error.
+We shifted the velocities of \citet{spencer+2018} (\(-1.1\) km/s) and
+\citet{pace+2020} (\(+1.1\) km/s) to reach the same scale. We found 183
+crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
+dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
+the median difference in velocities in this crossmatch is about 2.2
+km/s, we adopt 1 km/s as the approximate systematic error here.
 
@@
 
-\begin{figure}
+\begin{table*}[t]
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_vlos_xi_eta.pdf}}
-\caption[UMi velocity sample]{A plot of the corrected los velocities for
-UMi binned in tangent plane coordinates. There is no clear rotation or
-velocity gradient here. Interestingly, many velocity members are as far
-as 100 arcmin away.}
-\end{figure}
+\caption{MCMC fits for UMi velocity dispersion.}
+\label{tbl:MCMC-fits-for-UMi-velocity-dispersion}
+\begin{tabular}{lllll}
+\toprule
+study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
+\midrule
+all & $-245.9\pm0.3$ & $8.76\pm0.24$ & +1.5 & +2.2\\
+pace & $-244.5\pm0.4$ & $9.1\pm0.3$ & +0.2 & +1.1\\
+spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.3\\
+apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ & +0.8 & +0.8\\
+\bottomrule
+\end{tabular}
+\end{table*}
 
@@
   errors are accurately reported. APOGEE notes that their RV
-  uncertainties are known to be underestimates.
+  uncertainties are known to be underestimates but are a small
+  proportion of our sample.
 \item
@@
 motion of UMi.)
+
+\section{Comparison and conclusions}\label{comparison-and-conclusions}
+
+To illustrate the differences between each dwarf galaxy, in
+fig.~\ref{fig:classical_dwarfs_densities}, we compare Scl, UMi, and Fnx
+against exponential and plummer density profiles (\textbf{TODO: state
+these somewhere}). While all dwarfs have marginal differences in the
+inner regions, each dwarf diverges in the outer regions relative to an
+exponential. In particular, while Fnx is underdense, Scl and UMi are
+both overdense, approximately fitting a Plummer density profile instead.
+
+In summary, we have used J+24 data to derive the density profiles for
+Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
+robust against different selection criteria. Both Sculptor (Ursa Minor)
+show strong (weak) evidence for deviations from an exponential profile.
+We also compile velocity measurements to derive the systemic motions and
+velocity dispersions of each galaxy. We find evidence for a velocity
+gradient in Scl of XXX. We find no evidence of additional (velocity or
+stellar) substructure in either galaxy. Our goal is thus to explain why
+Scl and UMi have an excess of stars in their outer regions and why Scl
+may have a velocity gradient.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fornax_exp_fit.pdf}}
+\caption[Classical dwarf density profiles]{The density profiles of
+Sculptor, Ursa Minor, and Fornax compared to Exp2D and Plummer density
+profiles. Dwarf galaxies are scaled to the same half-light radius and
+density at half-light radius (fit from the inner 3 scale radii
+exponential recursively. )}\label{fig:classical_dwarfs_densities}
+\end{figure}
 

### 2025-04-24 thesis/introduction.tex
tests to validate gaia filters and thesis updates

@@
 $\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
+radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
 $\sigma_v$ & $8.8 \pm 0.2$ & this work\\
@@
 found from either of the equations above.
-
-\section{What is up with
-probabilities}\label{what-is-up-with-probabilities}
-
-\begin{figure}
-\centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250416112038183.png}
-\caption{image-20250416112038183}\label{fig:test}
-\end{figure}

### 2025-04-24 observations/test_fbest.jl

@@ -1,0 +1,340 @@
+### A Pluto.jl notebook ###
+# v0.20.6
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 944a654e-d955-4ea8-94ab-8aa5b73f9ea6
+import Pkg; Pkg.activate()
+
+# ╔═╡ 8cff9c1e-1a19-11f0-2dd6-7bddd47fe438
+using PyFITS
+
+# ╔═╡ 3b0e285e-cefd-42ff-8c0e-dbe71ef2b720
+using CairoMakie
+
+# ╔═╡ c8f9d3d9-8d91-4e63-8ba1-dc2139f9dca4
+using DustExtinction
+
+# ╔═╡ 061b5be8-fc49-4fc4-9c5d-239079740900
+import TOML
+
+# ╔═╡ 0653d025-c0e1-4b98-9701-f3db2f6ebe8d
+galaxy = "ursa_minor"
+
+# ╔═╡ c3933f11-ec53-49b0-a7c6-fc34ed8ef5c6
+df = read_fits("$galaxy.fits")
+
+# ╔═╡ 9beeec68-ca34-49a8-8f5b-6be279bacd13
+obs_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "observations", galaxy, "observed_properties.toml"))
+
+# ╔═╡ 8515cbfb-1385-40f6-a53a-36183be38375
+md"""
+# Total
+"""
+
+# ╔═╡ ead04430-df91-42f2-9a66-9e47563411db
+md"""
+Combination of all filters, positive satellite CMD likelihood, and within 2 arcmin.
+"""
+
+# ╔═╡ f35de603-cf15-4a1b-9c00-f4c70308af60
+filt_missed = df.F_BEST .!= (df.F_NOTANAN .* df.F_CPAR .* df.F_FLUXEXCESS .* df.F_INGRID .* df.F_ASTROMETRIC .* (df.L_CMD_SAT .> 0) .* (df.L_S_SAT .> 0))
+
+# ╔═╡ 2d407500-5ed6-4f14-9a19-db68616202b8
+df_missed = df[filt_missed, :]
+
+# ╔═╡ dd6a2a61-910f-4a14-afeb-ea6016125478
[…]

### 2025-04-25 thesis/appendix.tex
thesis updates and completed test fbest notebook

--- 2025-04-15 thesis/numerical_convergence.tex
@@
+\section{Appendix / Extra Notes}\label{appendix-extra-notes}
+
+\subsection{Additional density profile
+tests}\label{additional-density-profile-tests}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
+\caption[Density profiles]{Density profiles for various assumptions for
+Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
+2-component bayesian model assuming circular radii, simple is the series
+of simple cuts described in Appendix ?, bright is the sample of the
+brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
+(background subtracted and rescaled to
+match).}\label{fig:sculptor_observed_profiles}
+\end{figure}
+
+Note that a full rigorous statistical analysis would require a
+simulation study of injecting dwarfs into Gaia and assessing the
+reliability of various methods of membership and density profiles. This
+is beyond the scope of this thesis.
+
+\begin{verbatim}
+SELECT TOP 1000
+       *
+FROM delve_dr2.objects
+WHERE 11 < ra
+and ra < 19
+and -37.7 < dec
+and dec < -29.7
+\end{verbatim}
+
+\subsection{Velocity fit parameters.}\label{velocity-fit-parameters.}
+
+Savage-Dickey calculated bayes factor using Silverman-bandwidth KDE
+smoothed samples from posterior/prior.
+
+\emph{Is it interesting that the velocity dispersion of Scl seems to
+increase significantly with Rell?}
+
+\begin{table*}[t]
+\centering
+\caption{MCMC fits for different RV datasets for Scl amoung 3 different models.}
+\label{tbl:MCMC-fits-for-different-RV-datasets-for-Scl-amoung-3-different-models}
+\begin{tabular}{lllllll}
+\toprule
+study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log bf$\\
+\midrule
+all &  &  &  &  &  & \\
+ & $111.18\pm0.23$ & $9.71\pm0.17$ & - & - & - & 0\\
+ & $111.19 \pm 0.23$ & $9.68\pm0.17$ & - & $4.3\pm1.3$ & $-149_{-13}^{+17}$ & -1.7\\
+ & $111.16\pm0.23$ & $9.73\pm0.17$ & $0.056\pm0.021$ & - & - & -0.9\\
+tolstoy+23 &  &  &  &  &  & \\
+ & $111.3 \pm 0.3$ & $9.80 \pm 0.18$ & - & - & - & 0\\
+ & $111.3\pm0.3$ & $9.78\pm0.18$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ & -1.4\\
+ & $111.2 \pm 0.3$ & $9.74\pm0.19$ & $0.085 \pm 0.023$ & – & – & -4.5\\
+walker+09 &  &  &  &  &  & \\
+ & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – & 0\\
+ & $111.1\pm0.3$ & $9.53\pm0.21$ & - & $5.2_{-1.6}^{+1.8}$ & $-134_{-16}^{+23}$ & -2.4\\
+ & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – & +1.6\\
+apogee &  &  &  &  &  & \\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – & –\\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ & +0.4\\
+ & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – & +1.1\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption{MCMC fits for UMi velocity dispersion.}
+\label{tbl:MCMC-fits-for-UMi-velocity-dispersion}
+\begin{tabular}{lllll}
+\toprule
+study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
+\midrule
+all & $-245.9\pm0.3$ & $8.76\pm0.24$ & +1.5 & +2.2\\
+pace & $-244.5\pm0.4$ & $9.1\pm0.3$ & +0.2 & +1.1\\
+spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.3\\
+apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ & +0.8 & +0.8\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
+\label{tbl:Summary-of-velocity-measurements-and-derived-properties-sestito+2023a-number-of-members-depends-on-spatial-model-used}
+\begin{tabular}{llllllll}
+\toprule
+ & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$\\
+\midrule
+Scl & combined &  & 8945 & 2280 & 2034 & 1918 & 0.9\\
+ & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1480 & 0.65\\
+ & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13\\
+ & walker+09 & MMFS & 1818 & 1522 & 1417 & 1329 & 1.8\\
+ & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6\\
+UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3\\
+ & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8\\
+ & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5\\
+ & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9\\
+ & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\section{Numerical Convergence}\label{numerical-convergence}
+
 Here, we describe some convergence tests to ensure our methods and

### 2025-04-25 thesis/data.tex
thesis updates and completed test fbest notebook

@@
 \item
-  Solved parallax, proper motions, colour, and magnitudes.
+  Solved astrometry, magnitude, and colour.
 \item
@@
 \item
-  3\(\sigma\) consistency of measured parallax with dwarf distance +
-  uncertainty (typically near zero; with \citet{lindegren+2018}
-  zero-point correction).
+  3\(\sigma\) consistency of measured parallax with dwarf distance
+  (typically near zero; with \citet{lindegren+2021} zero-point
+  correction).
 \item
@@
 \item
-  No colour excess (\citet{lindegren+2018} equation C.2).
+  small colour excess factor \(|C*| \leq 3\sigma_{C*}(G)\)
+  (\citet{riello+2021}).
 \item
-  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB (in
-  practice astrometry ruwe =\textgreater{} 21 \textgreater{} G).
+  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB
+  (dereddened). In practice astrometry cuts limit 21 \textgreater{} G.
 \item
-  Between -0.5 and 2.5 in Bp - Rp.
+  Between -0.5 and 2.5 in Bp - Rp (dereddened).
 \end{itemize}
 
-Photometry is dereddened using \citet{schlegel+1988} extinction maps.
+Photometry is dereddened using \citet{schlegel+finkbeiner+davis1998}
+extinction maps.
 
@@
 induces a gradient of about \(1 km/s/degree\) for sculptor
-\citep[see][]{WMO2008}. The uncertainty is then the velocity uncertainty
-plus the distance uncertainties times the PM uncertainty from above. We
-then use the \(v_z\) values for the following modelling, however
-repeating with uncorrected, heliocentric velocities does not
-significantly affect the results .
+\citep[see][]{walker+mateo+olszewski2008}. The uncertainty is then the
+velocity uncertainty plus the distance uncertainties times the PM
+uncertainty from above. We then use the \(v_z\) values for the following
+modelling, however repeating with uncorrected, heliocentric velocities
+does not significantly affect the results .
 
@@
 
-\subsection{}\label{section}
-
-\begin{table*}[t]
-\centering
-\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
-\label{tbl:Summary-of-velocity-measurements-and-derived-properties-sestito+2023a-number-of-members-depends-on-spatial-model-used}
-\begin{tabular}{llllllll}
-\toprule
- & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$\\
-\midrule
-Scl & combined &  & 8945 & 2280 & 2034 & 1920 & 0.9\\
- & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1481 & 0.65\\
- & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13\\
- & walker+09 & MMFS & 1818 & 1522 & 1417 & 1330 & 1.8\\
- & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6\\
-UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3\\
- & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8\\
- & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5\\
- & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9\\
- & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
 \subsection{Sculptor}\label{sculptor}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_2dhist.pdf}}
-\caption[Scl velocity sample]{A plot of the corrected los velocities for
-Scl binned in tangent plane coordinates. We detect a slight rotational
-gradient towards the bottom right.}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
+\caption{LOS velocity fit to Scl.}
+\end{figure}
+
+\textbf{Figure:} Velocity histogram of Scl and UMi in terms of \(v_z\)
+(REF). Orange points are from our crossmatched RV membership sample.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
+\caption[Scl velocity sample]{RV members of Scl plotted in the tangent
+plane coloured by corrected velocity difference from mean
+\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
+fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
+motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
+scale).}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_binned.pdf}}
-\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
-arrow marks the gradient induced by Scl's proper motion on the sky.
-\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
+\caption[Scl velocity gradient]{The corrected LOS velocity along the
+best fit rotational axis. RV members are black points, the systematic
+\(v_z\) is the horizontal grey line, blue lines represent the
+(projected) gradient from MCMC samples, and the orange line is a rolling
+median (with a window size of 50).}
 \end{figure}
@@
 
-We derive a systemic velocity of \(111.2\pm0.2\) km/s with velocity
-dispersion \(9.67\pm0.16\) km/s. Our values are very consistent with
-previous work \citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{tolstoy+2023}]{WMO2009}.
+We additionally add a velocity gradient for Sculptor (adding parameters
+A, B \(\sim N(0, 6)\)km/s/deg). We derive a systemic velocity of
+\(111.2\pm0.2\) km/s with velocity dispersion \(9.67\pm0.16\) km/s. Our
+values are very consistent with previous work
+\citep[e.g.][\citet{arroyo-polonio+2024},
+\citet{tolstoy+2023}]{WMO2009}. See Appendix for a more detailed
+comparison and inter-study tests.
 
-Finally, we attempt to fit a linear velocity gradient to Scl by adding
-parameters for the gradient in \(\xi\) ande \(\eta\). We derive a
-gradient of xxx and xxx (see Figure\textasciitilde FIG.), noting that
-this direction is different and higher in magnitude than the proper
-motion of Scl. Comparing the density difference, we find a bayes factor
-of -15. Compared to past work, \citet{battaglia+2008},
-\citet{arroyo-polonio+2021}.
-
-Bayes factor found using relative density via KDE with Silverman's
-bandwith to prior (XXX theorem).
-
-\emph{Is it interesting that the velocity dispersion of Scl seems to
-increase significantly with Rell?}
-
-\begin{table*}[t]
-\centering
-\caption{MCMC fits for different RV datasets for Scl amoung 3 different models.}
-\label{tbl:MCMC-fits-for-different-RV-datasets-for-Scl-amoung-3-different-models}
-\begin{tabular}{lllllll}
-\toprule
-study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log bf$\\
-\midrule
-all &  &  &  &  &  & \\
- & $111.17\pm0.22$ & $9.68\pm0.17$ & - & - & - & 0\\
- & $111.18 \pm 0.23$ & $9.64\pm0.16$ & - & $4.8\pm1.3$ & $-147_{-12}^{+15}$ & -3.7*\\
- & $111.15\pm0.22$ & $9.70\pm0.16$ & $0.055\pm0.021$ & - & - & -0.8\\
-tolstoy+23 &  &  &  &  &  & \\
- & $111.2 \pm 0.3$ & $9.77 \pm 0.18$ & - & - & - & 0\\
- & $111.2\pm0.3$ & $9.73\pm0.18$ & – & $4.8\pm1.4$ & $-154_{-12}^{+16}$ & -2.5\\
- & $111.2 \pm 0.3$ & $9.71\pm0.18$ & $0.081 \pm 0.023$ & – & – & -3.3\\
-walker+09 &  &  &  &  &  & \\
- & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – & 0\\
- & $111.1\pm0.3$ & $9.54\pm0.21$ & - & $5.3_{-1.6}^{+1.8}$ & $-134_{-16}^{+22}$ & -2.0\\
- & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – & +1.6\\
-apogee &  &  &  &  &  & \\
- & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – & –\\
- & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ & +0.3\\
- & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – & +1.1\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\subsection{Ursa Minor}\label{ursa-minor}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_rv_2dhist.pdf}}
-\caption[UMi velocity sample]{UMi velocities}
-\end{figure}
-
-Figure: A plot of the corrected los velocities for UMi binned in tangent
-plane coordinates. There is no clear rotation or velocity gradient here.
-Interestingly, many velocity members are as far as 100 arcmin away.
+We detect a gradient of \(4.8\pm1.3\) km/s/deg at a position angle of
+\(-147_{-12}^{+15}\) degrees. Compared to past work,
+\citet{battaglia+2008}, \citet{arroyo-polonio+2024}, our model is XX
 
 For UMi, we collect radial velocities from, APOGEE,
-\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}.
-
-We shifted the velocities of \citet{spencer+2018} (\(-1.1\) km/s) and
+\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
+shifted the velocities of \citet{spencer+2018} (\(-1.1\) km/s) and
 \citet{pace+2020} (\(+1.1\) km/s) to reach the same scale. We found 183
@@
 
-UMi interestingly has a more structured observational pattern, but does
-not appear to have any significant velocity substructure.
-
-\begin{table*}[t]
-\centering
-\caption{MCMC fits for UMi velocity dispersion.}
-\label{tbl:MCMC-fits-for-UMi-velocity-dispersion}
-\begin{tabular}{lllll}
-\toprule
-study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
-\midrule
-all & $-245.9\pm0.3$ & $8.76\pm0.24$ & +1.5 & +2.2\\
-pace & $-244.5\pm0.4$ & $9.1\pm0.3$ & +0.2 & +1.1\\
-spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.3\\
-apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ & +0.8 & +0.8\\
-\bottomrule
-\end{tabular}
-\end{table*}
+We derive a mean \(-245.9\pm0.3\) and velocity dispersion of
+\(8.76\pm0.24\). However, since the velocity means of
+\citet{spencer+2018} and \citet{pace+2020} differ by 2 km/s, we add a 1
+km/s systematic error to our measurement.
 
@@
 density at half-light radius (fit from the inner 3 scale radii
-exponential recursively. )}\label{fig:classical_dwarfs_densities}
+exponential recursively. ). Scl and UMi have an excess of stars in the
+outer regions (past \(\log R/R_h \sim 0.3\)) compared with
+Scl.}\label{fig:classical_dwarfs_densities}
 \end{figure}
-
-\section{Appendix / Extra Notes}\label{appendix-extra-notes}
-
-\subsection{Additional density profile
-tests}\label{additional-density-profile-tests}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
-\caption[Density profiles]{Density profiles for various assumptions for
-Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
-2-component bayesian model assuming circular radii, simple is the series
-of simple cuts described in Appendix ?, bright is the sample of the
-brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
-(background subtracted and rescaled to
-match).}\label{fig:sculptor_observed_profiles}
-\end{figure}
-
-Note that a full rigorous statistical analysis would require a
-simulation study of injecting dwarfs into Gaia and assessing the
-reliability of various methods of membership and density profiles. This
-is beyond the scope of this thesis.
-
-\begin{verbatim}
-SELECT TOP 1000
-       *
-FROM delve_dr2.objects
-WHERE 11 < ra
-and ra < 19
-and -37.7 < dec
-and dec < -29.7
-\end{verbatim}

### 2025-04-25 thesis/introduction.tex
thesis updates and completed test fbest notebook

@@
 $\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $111.2 \pm 0.2\ {\rm km\,s^{-1}}$ & this work\\
-$\sigma_v$ & $9.67\pm0.16\ {\rm km\,s^{-1}}$ & this work\\
+radial velocity & $111.18 \pm 0.23\ {\rm km\,s^{-1}}$ & this work\\
+$\sigma_v$ & $9.71\pm0.17\ {\rm km\,s^{-1}}$ & this work\\
 $R_h$ & $12.33 \pm 0.05$ arcmin & 4\\

### 2025-04-25 thesis/sculptor.tex
thesis updates and completed test fbest notebook

\section{Sculptor (MW only)}\label{sculptor-mw-only}

\subsection{Orbital estimation}\label{orbital-estimation}

fig.~\ref{fig:unref} contains example orbits

Given the parameters in table@REF, we

\subsection{Simulations}\label{simulations}

\begin{itemize}
\tightlist
\item
  dark matter loss
\item
  little influence to stars
\item
  smallest pericentre does not help
\end{itemize}

\section{Sculptor (LMC orbits)}\label{sculptor-lmc-orbits}

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC. Figure
fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
Scl's pericentre with the MW and implies that Scl may be on its first
infall with the MW, in contrast to the discussion above. Thus, the LMC
has a critical impact on the evolution of Scl. Additionally, Scl passes
rapidly but relatively close to the LMC in the past 100 Myr.

We now consider 3 orbits under the L3M11 potential model from
\citet{vasiliev2024}. We focus on this LMC potential as a lighter LMC or
MW should only reduce tidal impacts and the recent orbit of Scl is
minimally affected by the LMC potential.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}}
\caption{image-20250415112706417}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}}
\caption{image-20250415113122385}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113133353.png}}
\caption{image-20250415113133353}
\end{figure}

### 2025-04-25 thesis/compile.sh

@@ -1,13 +1,19 @@
 set -e 
 rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
-pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py"
+pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
 
 pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
-pandoc $pandocoptions -o numerical_convergence.tex numerical_convergence.md
+pandoc $pandocoptions -o sculptor.tex sculptor.md
+# pandoc $pandocoptions -o ursa_minor.tex ursa_minor.md
+# pandoc $pandocoptions -o conclusion.tex conclusion.md
+
+
+pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=chapter --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
+pandoc $pandocoptions -o appendix.tex appendix.md
 
 
 latexmk -f

### 2025-04-30 thesis/compile.sh
updated structure and table shortcaptions

@@ -1,7 +1,7 @@
 set -e 
 rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
-pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
+pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections --filter equation.py"
 
 pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
@@ -12,8 +12,6 @@ pandoc $pandocoptions -o sculptor.tex sculptor.md
 # pandoc $pandocoptions -o conclusion.tex conclusion.md
 
 
-pandocoptions="--filter pandoc-crossref --citeproc -M --top-level-division=chapter --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
+pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=chapter --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
 pandoc $pandocoptions -o appendix.tex appendix.md
 
-
-latexmk -f

### 2025-04-30 thesis/equation.py

@@ -1,0 +1,15 @@
+import panflute as pf
+
+def equation_environment(math_elem, doc):
+    # Check for paragraphs containing only a DisplayMath element
+    if isinstance(math_elem, pf.Math) and math_elem.format == "DisplayMath":
+        # Replace with raw LaTeX equation environment
+        tex = f"\\begin{{equation}}{math_elem.text}\\end{{equation}}"
+        return pf.RawInline(tex, format="latex")
+
+
+def main(doc=None):
+    return pf.run_filter(equation_environment, doc=doc)
+
+if __name__ == "__main__":
+    main()

### 2025-04-30 thesis/table_shortcaption.py

@@ -1,0 +1,117 @@
+import panflute as pf
+import re
+
+
+def extract_attributes(inlines):
+    # Reverse search for attribute block
+    attr_text = []
+    indices = []
+    
+    # Find closing '}'
+    end_pos = None
+    for i in reversed(range(len(inlines))):
+        if isinstance(inlines[i], pf.Str) and '}' in inlines[i].text:
+            end_pos = i
+            break
+    
+    if end_pos is None:
+        return None, None, inlines
+    
+    # Find opening '{'
+    start_pos = None
+    for i in reversed(range(end_pos + 1)):
+        if isinstance(inlines[i], pf.Str) and '{' in inlines[i].text:
+            start_pos = i
+            break
+
+    if start_pos is None:
+        return None, None, inlines
+    
+
+    not_id_pos = start_pos
+    for i in range(start_pos, end_pos):
+        if isinstance(inlines[i], pf.Space):
+            not_id_pos = i
+            break
+
+    # Extract attribute content
+    attr_elements = inlines[not_id_pos:end_pos]
+    id_str = pf.stringify(inlines[start_pos:not_id_pos])[2:]
+    remaining_inlines = pf.ListContainer(*inlines[:start_pos],
+                                      *inlines[end_pos+1:])
+    
+    attrs_dict = parse_attributes(attr_elements)
+
+    return id_str, attrs_dict, remaining_inlines
+
+
+def parse_attributes(attr_elements):
+    if len(attr_elements) == 0:
+        return {}
+    # Convert attribute elements to text
+    splits = [0]
+
+    for i in range(len(attr_elements)):
+        if isinstance(attr_elements[i], pf.Space):
+            splits.append(i)
+    splits.append(len(attr_elements))
+
+    attrs_dict = {}
+    for i in range(len(splits)-1):
+        attr_str = pf.stringify(attr_elements[splits[i]:splits[i+1]])
+        if attr_str:
+            key, val = attr_str.split("=")
+            if val[0] == '"' and val[-1] == '"':
+                val = val[1:-1]
+            elif val[0] == "'" and val[-1] == "'":
+                val = val[1:-1]
+            else:
+                val = val.strip()
+
+            attrs_dict[key.strip()] = val
+
+    return attrs_dict
+
+
+
+
+def process_table(element, doc):
+    if isinstance(element, pf.Table) and element.caption:
+        caption = element.caption
+        if caption.content:
+            pf.debug("table caption: ")
+            # Get first paragraph of caption
[…]

### 2025-05-02 observations/sculptor/velocities/clean_apogee.jl
cleaning up all the data

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.6
+# v0.20.8
 
 using Markdown
 using InteractiveUtils
@@ -85,7 +85,7 @@ data_dir = "../data/"
 obs_properties = TOML.parsefile("../observed_properties.toml")
 
 # ╔═╡ 7a50a176-96a5-4098-88d6-0fa2874d0f90
-j24 = read_fits("..//data/jensen+24_wide.fits")
+j24 = read_fits("..//data/jensen+24_wide_2c.fits")
 
 # ╔═╡ d4d0a488-1c0e-4bc6-9a88-94b27d84e8ce
 apogee_raw = read_fits("processed/apogee_xmatch.fits")

### 2025-05-02 observations/sculptor/velocities/clean_tolstoy23.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.6
+# v0.20.8
 
 using Markdown
 using InteractiveUtils
@@ -84,7 +84,7 @@ No need to filter further...
 obs_properties = TOML.parsefile("../observed_properties.toml")
 
 # ╔═╡ c4649b4e-f8b3-4b2c-8595-5dc9a3a302da
-j24 = read_fits("../data/jensen+24_wide.fits")
+j24 = read_fits("../data/jensen+24_wide_2c.fits")
 
 # ╔═╡ bb7f6769-ec92-460d-8423-449029175f79
 
@@ -190,23 +190,6 @@ md"""
 # ╔═╡ 91208c77-3664-48fd-a659-1151a4dc0bf4
 (tolstoy23_all.RV_sigma ./ tolstoy23_all.RV_err)[.!F_scatter] # number of sigma for filtered stars
 
-# ╔═╡ 28e07273-2177-408c-aabf-2f9ecf16b155
-hist(filter(isfinite, p_chi2))
-
-# ╔═╡ 67e33e9f-fab3-4fe3-9418-992b50ef8431
-hist(tolstoy23_all.RV_sigma ./ tolstoy23_all.RV_err)
-
-# ╔═╡ de4788f4-d446-4e04-8d43-6a8308f62a97
-md"""
-## Membership
-"""
-
-# ╔═╡ 33d34e7e-0c0a-4c49-b0c1-0116c8ef7a38
-filt_tolstoy = RVUtils.sigma_clip(tolstoy23_all.RV, 3) .& (tolstoy23_all.Mem .== "m")
-
-# ╔═╡ 344f3c29-0873-4180-9025-51aeaeb2c681
-tolstoy23 = tolstoy23_all[filt_tolstoy, :]
-
 # ╔═╡ 2eebfca4-e364-467f-a1b3-c1cfab2822e6
 chi2 = @. tolstoy23_all.RV_sigma^2 ./ tolstoy23_all.RV_err^2 
 
@@ -223,6 +206,23 @@ import Distributions
 md"""
 If errors are representative, then the below plot should be uniform with only an excess at zero.
 """
+
+# ╔═╡ 28e07273-2177-408c-aabf-2f9ecf16b155
+hist(filter(isfinite, p_chi2))
+
[…]

### 2025-05-02 observations/sculptor/velocities/clean_walker09.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.6
+# v0.20.8
 
 using Markdown
 using InteractiveUtils
@@ -99,7 +99,7 @@ data_dir = "../data/"
 obs_properties = TOML.parsefile("../observed_properties.toml")
 
 # ╔═╡ 7a50a176-96a5-4098-88d6-0fa2874d0f90
-j24 = read_fits("../data/jensen+24_wide.fits")
+j24 = read_fits("../data/jensen+24_wide_2c.fits")
 
 # ╔═╡ ea398623-ebc2-420a-8377-65bef43c0154
 begin 
@@ -164,6 +164,7 @@ begin
 	j24_filt, j24_idx = RVUtils.xmatch(walker09_all, j24, 3) # catches 1 missed star
 
 	walker09_all[!, :source_id] = j24.source_id[j24_idx]
+	allowmissing!(walker09_all, :source_id)
 	walker09_all[.!j24_filt, :source_id] .= missing
 
 	walker09_all

### 2025-05-02 observations/sculptor/velocities/select_members.jl

@@ -1,0 +1,287 @@
+### A Pluto.jl notebook ###
+# v0.20.8
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 04bbc735-e0b4-4f0a-9a83-e50c8b923caf
+begin 
+	import Pkg; Pkg.activate()
+	using Arya
+	using CairoMakie
+
+	import LilGuys as lguys
+end
+
+# ╔═╡ 93838644-cad6-4df3-b554-208b7afeb3b8
+using PyFITS
+
+# ╔═╡ bd6dfd17-02ee-4855-be37-fecfdab6776f
+using LilGuys; FIGDIR = "figures"
+
+# ╔═╡ 72f1febc-c6ea-449a-8cec-cd0e49c4e20c
+using DataFrames
+
+# ╔═╡ 9070c811-550c-4c49-9c58-0943b0f808b2
+using Turing
+
+# ╔═╡ e1cdc7ac-b1a4-45db-a363-2ea5b5ad9990
+using PairPlots
+
+# ╔═╡ 035cdedb-da23-4cfd-aa19-a3aff089d3ac
+using Measurements
+
+# ╔═╡ 3ed8c28f-5908-42dc-a56b-24a9b2685a07
+md"""
+## Inputs
+"""
+
+# ╔═╡ 50488b8f-6886-4191-8778-af66929f1445
+rv_file = "rv_tolstoy+23.fits"
+
+# ╔═╡ 4827c99d-6326-42fb-a329-d56992f6d20d
+j24_sample = "1c"
+
+# ╔═╡ cfadb58a-5569-41f5-a72b-04eba70cc89e
+psat_min = 0.2
+
+# ╔═╡ 680e7f76-cb4d-40d6-9a9f-d4672427a633
+md"""
+## derived
+"""
+
+# ╔═╡ 3e0eb6d1-6be4-41ec-98a5-5e9167506e61
+data_dir = "processed"
+
+# ╔═╡ 86fe351f-ef12-474a-85cc-c10c22a65e77
+outname = splitext(basename(rv_file))[1] * "_x_" * j24_sample * "_psat_$psat_min"
+
+# ╔═╡ 7330c75e-1bf9-476a-8274-ebc86d555e6f
+md"""
+# RV sample models
+"""
+
+# ╔═╡ 6bec7416-40c8-4e2b-9d3d-14aa19e5642d
+md"""
[…]

### 2025-05-02 observations/sculptor/velocities/velocity_single_models.jl

@@ -1,949 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.20.6
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ 04bbc735-e0b4-4f0a-9a83-e50c8b923caf
-begin 
-	import Pkg; Pkg.activate()
-	using Arya
-	using CairoMakie
-
-	import LilGuys as lguys
-end
-
-# ╔═╡ 93838644-cad6-4df3-b554-208b7afeb3b8
-using PyFITS
-
-# ╔═╡ 72f1febc-c6ea-449a-8cec-cd0e49c4e20c
-using DataFrames
-
-# ╔═╡ 9070c811-550c-4c49-9c58-0943b0f808b2
-using Turing
-
-# ╔═╡ e1cdc7ac-b1a4-45db-a363-2ea5b5ad9990
-using PairPlots
-
-# ╔═╡ 35c87efe-6899-444d-952f-94e5e5342846
-using Measurements
-
-# ╔═╡ 4cc490ba-db14-4ead-9c9d-ffe02a301f57
-using OrderedCollections
-
-# ╔═╡ c05a4e8e-c6d2-43e6-9b87-679f4ecee84c
-study = "apogee"
-
-# ╔═╡ b3ad64bc-7dbb-4c27-a91d-a9f2342f98da
-begin 
-	FIGDIR = "./figures/"
-	FIGSUFFIX = ".$study"
-	using LilGuys
-end
-
-# ╔═╡ 6bec7416-40c8-4e2b-9d3d-14aa19e5642d
-md"""
-This notebook takes the dataset created from velocity_xmatch.jl and analyzes it using MCMC to estimate the velocity dispersion and search for any possible velocity gradients.
-"""
-
-# ╔═╡ 030d4c9a-5660-4a3c-8133-6341ebdb421d
-md"""
-Creates:
-- `figures/vlos_xi_eta_hist`.study.*
-
-"""
-
-# ╔═╡ 9e9ba645-b780-4afa-b305-a2b1d8a97220
-import StatsBase: quantile, mean, std, median, sem
-
-# ╔═╡ d2888213-61e3-4a6f-872b-48a075640ef5
-import TOML
-
-# ╔═╡ c3298b45-7c5d-4937-8e4c-c87de36a1354
[…]

### 2025-05-02 observations/sculptor/velocities/xmatch_all_studies.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.6
+# v0.20.8
 
 using Markdown
 using InteractiveUtils
@@ -66,7 +66,7 @@ data_dir = "../data/"
 obs_properties = TOML.parsefile("../observed_properties.toml")
 
 # ╔═╡ 7a50a176-96a5-4098-88d6-0fa2874d0f90
-j24 = read_fits("../processed/best_fullfield_sample.fits")
+j24 = read_fits("../processed/best_sample.fits")
 
 # ╔═╡ d4d0a488-1c0e-4bc6-9a88-94b27d84e8ce
 apogee_all = read_fits("processed/rv_apogee.fits")
@@ -134,7 +134,7 @@ end
 # ╔═╡ e472cbb6-258e-4303-85e2-56f26358c97b
 all_stars = let
 
-	all_stars = copy(j24)
+	all_stars = copy(j24)[:, [:source_id, :ra, :dec, :dr2_radial_velocity, :dr2_radial_velocity_error]]
 	add_xmatch!(all_stars, apogee_all, "apogee")
 	add_xmatch!(all_stars, walker09_all, "w09")
 	add_xmatch!(all_stars, tolstoy23_all, "t23")
@@ -225,6 +225,21 @@ p_chi2 = RVUtils.prob_chi2.(rv_meas.RV_sigma, rv_meas.RV_err, rv_meas.RV_nstudy 
 
 # ╔═╡ e7c0edd6-770d-40db-8619-b5c471569fb6
 F_scatter = @. isnan(p_chi2) || (p_chi2 > 0.001)
+
+# ╔═╡ ce420c18-b09d-4094-bc06-5537c3d756e8
+chi2 = @. rv_meas.RV_sigma^2 / rv_meas.RV_err^2 / (rv_meas.RV_nstudy - 1)
+
+# ╔═╡ e146c4c6-4843-42b3-aa9e-35c8e0615d70
+hist(filter(isfinite, chi2[F_scatter]))
+
+# ╔═╡ cc44c8c1-6eaa-404b-9470-12c0e6616605
+hist(log10.(chi2[.!F_scatter]))
+
+# ╔═╡ a7fad724-ed91-4ae8-8a9b-0ed05774d2de
+minimum(chi2[.!F_scatter])
+
[…]

### 2025-05-02 observations/sculptor/velocities/xmatch_apogee.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.6
+# v0.20.8
 
 using Markdown
 using InteractiveUtils
@@ -56,7 +56,7 @@ md"""
 """
 
 # ╔═╡ f26e6eb1-640b-40fa-8cbc-e87092f132f7
-j24 = read_fits("../data/jensen+24_wide.fits")
+j24 = read_fits("../data/jensen+24_wide_2c.fits")
 
 # ╔═╡ 5a9c76ed-eccf-42d7-96f1-814471f66d89
  apogee_f = read_fits("../../all/data/allStarLite-dr17-synspec_rev1.fits")

### 2025-05-02 thesis/data.tex

@@
+\section{Abstract}\label{abstract}
+
+To provide an empirical benchmark for simulations in the next chapters,
+we would like to understand and determine the observed properties of
+Sculptor and Ursa Minor. In this section, we first review past
+measurements of each galaxy and the observed properties of each dwarf
+galaxy. We describe the \citet{jensen+2024} Bayesian methodology to
+select high probability member stars from \emph{Gaia}. We test the
+derived density profiles for Sculptor, Ursa Minor, and Fornax, finding
+that \citet{jensen+2024}'s methodology successfully reduces the
+background and produces similar results to a simple
+background-subtracted density model. We show that Fornax is
+approximately represented by a 2-D exponential profile. Instead,
+Sculptor and Ursa Minor show an excess of stars relative to an 2-D
+exponential in the outer regions. A Plummer profile provides a better
+fit instead. Finally, we combine various radial velocity samples with
+\citet{jensen+2024}'s members to derive systemic velocities and
+dispersions for Sculptor and Ursa Minor.
+
+\section{Introduction}\label{introduction}
+
+The determination of basic properties of astronomical objects is an
+ongoing and challenging endeavour.
+
+\subsection{Gaia}\label{gaia}
+
+Some of the most fundamental properties of astronomical objects are
+their position and velocity. Unfortunately, determining positions to
+stars is nontrivial, . Additionally, while line-of-sight or radial
+velocities (RVs) are easily determined from spectroscopy, the tangental
+velocities, perpendicular to RV, are only measurable through proper
+motions, typically requiring precise astrometry as well. \emph{Gaia}'s
+mission is to produce extremely precise astrometry enabling measurements
+of unprecedented accuracy and scale for proper motions, parallaxes, and
+magnitudes.
+
+\emph{Gaia} was designed to revolutionize proper motion and parallax
+measurements. \emph{Gaia} is a space-based, all-sky survey telescope
+with two primary 1.3x0.75m mirrors situated at the Sun-Earth L2 lagrange
+point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
+completing its mission in 2025 (but with two more data releases
+planned). By imagining two patches of sky on the same focal plane,
+separated by a fixed angle of 106.5degrees, Gaia is able to measure
+absolute proper motions by comparing the apparent shifts of stars in
+different regions of the sky. In addition to precise astrometric
+information, \emph{Gaia} measures the magnitude of stars in the very
+wide \emph{G} band (330-1050nm), blue and red colours using the blue and
+red photometers (BP and RP), and takes low resolution BP-RP spectra and
+radial velocity measurements of bright stars (magnitudes \textgreater{}
+16?).
+
+\emph{Gaia} has revolutionized many astronomical disciplines, the least
+of which is local group and Milky Way science. While proper motions of a
+dwarf galaxies has been measured in a case by case basis by the Hubble
+Space Telescope, a full systematic determination of proper motions for
+most dwarf galaxies was unavailable until \emph{Gaia}
+\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
+the detection of many substructures of the halo, streams,
+
+While \emph{Gaia} has performed excellently, a few limitations and
+systematic uncertainties could become problematic. Gaia does show
+systematic variations in its derived proper motions and parallaxes, as
+measured by background quasars \citep{lindegren+2021}. These errors
+limit the ability to determine absolute proper motions to some degree,
+but since we would like to measure if a star is \emph{consistent} with a
+dwarf galaxy, these only factor into the uncertainty budget for orbital
+motion and are included in J+24. Additionally, while \emph{Gaia} was
+designed to be complete down to about a magnitude of \(G=20\),
+overcrowding or dense fields can limit this. \emph{Gaia} can only
+successfully separate stars at a distance of about \textbf{X} (a few
+times the pixel length of \ldots; pixel width is in scanning direction).
+For the galaxies Ursa Minor, Fornax, and Sculptor, the highest density
+(in FOrnax), is only \textbf{xxx} which appears to not significantly
+affect \emph{Gaia}'s survey completeness. We test for brightness affects
+in \textbf{REF}.
+
+For our purposes, \emph{Gaia}
+
+\citet[hereafter J+24]{jensen+2024} utilize \emph{Gaia} data with a
+Bayesian framework to determine samples of high-probability members for
+many Milky Way group dwarf galaxies.
+
+Here, we breifly describe J+24's membership estimation and discuss how
+this informs our observational knoledge of each galaxies density
+profile. In general, J+24 use a Bayesian framework incorporating proper
+motion (PM), colour-magnitude diagram (CMD), and spatial information to
+determine the probability that a given star belongs to the satellite.
+J+24 extends \citet{MV2020a}; \citet{MV2020b}. \citep[See also similar
+work by][]{pace+li2019, battaglia+2022, pace+erkal+li2022, qi+2022}
+
+\subsection{Summary of properties}\label{summary-of-properties}
+
+In tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} we present
+adopted observed properties for each galaxy.
+
+\begin{table*}[t]
+\centering
+\caption[Observed Properties of Sculptor]{\label{tbl:scl_obs_props}Observed properties of Sculptor.
+References are: 1.
+Ricardo R.
+Muñoz et al.
+(2018), 2.
+Tran et al.
+(2022), 3.
+McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a).
+}
+\label{tbl:scl_obs_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
+$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
+distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
+distance & $83.2 \pm 2$ kpc & 2\\
+$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
+radial velocity & $111.18 \pm 0.23\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
+$\sigma_v$ & $9.71\pm0.17\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
+$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
+$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & XREF\\
+ellipticity & $0.36 \pm 0.01$ & 1\\
+position angle & $92\pm1^\circ$ & 1\\
+$M_V$ & $-10.82\pm0.14$ & 1\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Observed Properties of Ursa Minor]{\label{tbl:umi_obs_props}Observed properties of Ursa Minor.
+References are: (1) Ricardo R.
+Muñoz et al.
+(2018), (2) Garofalo et al.
+(2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al.
+(2020) and Spencer et al.
+(2018).
+}
+\label{tbl:umi_obs_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
+$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
+distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
+distance & $70.1 \pm 3.6$ kpc & 2\\
+$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
+radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
+$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
+$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
+ellipticity & $0.55 \pm 0.01$ & 1\\
+position angle & $50 \pm 1^\circ$ & 1\\
+$M_V$ & $-9.03 \pm 0.05$ & 1\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
 \section{Gaia Membership Selection}\label{gaia-membership-selection}
 
-Gaia provides unprecedented accuracy in proper motions and magnitudes.
-Gaia data is uniquely excellent to produce low-contamination samples of
-likely member stars belonging to satellites. Here, we breifly describe
-J+24's membership estimation and discuss how this informs our
-observational knoledge of each galaxies density profile. In general,
-J+24 use a Bayesian framework incorporating proper motion (PM),
-colour-magnitude diagram (CMD), and spatial information to determine the
-probability that a given star belongs to the satellite. J+24 extends
-\citet{MV2020a} (see also \citet{pace+li2019}, \citet{battaglia+2022},
-\citet{pace+erkal+li2022}, etc.).
+\subsection{Data}\label{data}
 
@@
 \item
-  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3}).
+  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), so astrometry
+  is high quality.
 \item
@@
 \item
-  Absolute RA and Dec proper motions less than
-  10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
-  \(\gtrsim 500\) km/s at distances larger than 10 kpc.).
+  Absolute RA and Dec proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\),
+  less than 10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental
+  velocities of \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
 \item
-  small colour excess factor \(|C*| \leq 3\sigma_{C*}(G)\)
-  (\citet{riello+2021}).
+  Corrected colour excess is within expectations:
+  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
+  from \citet{riello+2021}. Removes stars with unreliable photometry.
 \item
-  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB
-  (dereddened). In practice astrometry cuts limit 21 \textgreater{} G.
+  \(G\) magnitude is between \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\)
+  (dereddened magnitudes). Removes very faint stars and stars
+  significantly brighter than the tip of the red giant branch (TRGB)
+  magnitude plus the distance modulus uncertainty \(\delta {\rm DM}\).
 \item
-  Between -0.5 and 2.5 in Bp - Rp (dereddened).
+  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
 \end{itemize}
@@
 J+24 calculate the probability that any star belongs to either the
-satellite or the MW background with \[
+satellite or the MW background with \begin{equation}
 P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
-\]
+\end{equation}
 
 where \(f_{\rm sat}\) is the fraction of candidate members in the field,
-and \({\cal L}_{\rm sat}\) and \({\cal L}\)\_\{\rm bg\}\$ are the
-satellite (sat) and background (bg) likelihoods respectively. Each
-likelihood is the product of a spatial, proper motion, and CMD term: \[
+and \({\cal L}_{\rm sat}\) and \({\cal L}_{\rm bg}\) are the satellite
+(sat) and background (bg) likelihoods respectively. The likelihoods are
+the product of a spatial, proper motion, and CMD term: \begin{equation}
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
-\]
-
-The satellite likelihoods are specified as
-
-\begin{itemize}
-\tightlist
-\item
-  CMD: The CMD is the lowest metallicity isochrone from Padova
-  \citep{girardi+2002} with age 12 Gyr with a colour width of 0.1 mag
-  plus the Gaia colour uncertainty at each magnitude. The HB is modelled
-  as a constant magnitude extending blue of the CMD. The HB magnitude is
-  the mean magnitude of HB stars from most metal poor isochrone with a
-  0.1 mag width plus the mean colour error. A likelihood map is
-  constructed by sampling the distance modulus in addition to the CMD
-  width, taking the maximum of RGB and HB likelihoods.
-\item
-  Spatial: A single exponential
-  (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)) accounting for structural
-  uncertainty (sampled over position angle, ellipticity, and half light
-  radius).
-\item
-  Alternative spatial: For Scl and UMi, this is instead a double
-  exponential
-  \(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\) where
-  the inner exponential remains fixed. Structural parameter
-  uncertainties are not accounted for.
-\item
-  PM. A bivariate gaussian with variance and covariance equal to each
-  star's proper motions. Each star's proper motions uncertainty are
-  assumed to be the dominant uncertainty.
-\end{itemize}
-
-The background likelihoods are constructed as:
-
-\begin{itemize}
-\tightlist
-\item
-  CMD : Constructed as a density map using the other quality-selected
-  stars outside of \(5R_h\) in the catalogue. The map is a sum of
-  bivariate gaussians for each star with standard deviations based on
-  the Gaia uncertainties.
-\item
-  PM: same as CMD except in PM space (\(\mu_{\alpha*}\) and
-  \(\mu_\delta\) with respecive uncertainties and covariance.)
-\item
-  Spatial: a constant likelihood.
-\end{itemize}
+\end{equation}
 
 Each likelihood map is normalized over the respective 2D parameter
-space. \(f_{\rm sat}\) thus is the only term controlling the total
-abundance of satellite stars relative to the background.
+space. \(f_{\rm sat}\) scales the likelihoods by the abundance of stars
+in the satellite compared to the background.
+
+J+24 consider both a one-component and a two-component spatial
+likelihood. The one component model is constructed as a single
+exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with
+\(R_s\) fixed to the value in the table in \citet{MV2020a}.
+Additionally, structural uncertainties (for position angle, ellipticity,
+and scale radius) are sampled over to construct the final likelihood
+map. The two-component model instead adds a second exponential,
+\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
+scale radius is fixed, and the outer scale radius and magnitude of the
+second component \(R_{\rm outer}\), \(B\) are solved for. Structural
+properties are not accounted for. The PM likelihood is a bivariate
+gaussian with variance and covariance equal to each star's proper
+motions. (The stellar PM uncertainties is assumed and typically is the
+dominant PM uncertainty.)
+
+The satellite's CMD likelihood is based on a Padova isochrone
+\citep{girardi+2002}. The isochrone has a matching metallicity and 12
+Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
+Gaia colour uncertainty at each magnitude. The HB is modelled as a
+constant magnitude extending blue of the CMD (mean magnitude of
+isochone's HB stars and a 0.1 mag width plus the mean colour error). A
+likelihood map is constructed by sampling the distance modulus in
+addition to the CMD width, taking the maximum of RGB and HB likelihoods.
+
+The background likelihoods are instead empirically constructed. Stars
+stars outside of 5\(R_h\) passing the quality cuts estimate the
+background density in PM and CMD space. The density is a sum of
+bivariate gaussians with variances based on Gaia uncertainties (and
+covariance for proper motions). The spatial background likelihood is
+assumed to be constant.
 
 In J+24, a MCMC simulation is ran using the above total likelihood to
-solve for the following parameters:
-
-\begin{itemize}
-\tightlist
-\item
-  Systemic proper motions \(\mu_\alpha\), \(\mu_\delta\). Single
-  component prior is same as \citet{MV2020}: a normal distribution with
-  mean 0 and standard deviation 100 km/s. If 2-component spatial,
-  instead is a uniform distribution spanning 5\(\sigma\) of single
-  component case w/ systematic uncertainties.
-\item
-  \(f_{\rm sat}\) density normalization. Prior is a uniform distribution
-  between 0 and 1.
-\item
-  Spatial component parameters \(B\) is uniform from 0-1 and
-  \(R_{\rm outer}\) is uniform and greater than \(R_s\) for extended
-  profiles (Scl and UMi here.)
-\end{itemize}
-
-The mode of each parameter from the MCMC are then used to calculate the
-final \(P_{\rm sat}\) values we use here.
+solve \(\mu_{\alpha*}\), \(\mu_\delta\), and \(f_{\rm sat}\). For the
+two component models, the amplitude and scale radius of the excented
+compopnet is also solved for. The proper motion single component prior
+is same as \citet{MV2020a}: a normal distribution with mean 0 and
+standard deviation 100 km/s. If 2-component spatial, instead is a
+uniform distribution spanning 5\(\sigma\) of single component case w/
+systematic uncertainties. Priors on \(f_{\rm sat}\) and \(B\) (if 2
+component) are uniform from 0--1. \(R_{\rm outer}\) has a uniform prior
+only restricting \(R_{\rm outer} > R_s\). The mode of each parameter
+from the MCMC are then used to calculate the final \(P_{\rm sat}\)
+values we use here.
 
 We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
-sample. Most stars are assigned probabilities close to 0 or 1, so the
-choice of probability threshhold is not too significant. Additionally,
-even for a probability cut of 0.2, the purity of the resulting sample
-with RV measurements is very high (\textasciitilde90\%, J+24). However,
-there is likely a systematic bias in using stars with RV measurements to
-measure purity. Fainter stars have poorer astrometry and are less likely
-to have been targeted. We find no difference in the resulting density
-distributions when restricting stars to be brighter than a specific
-magnitude.
+sample. Most stars are assigned probabilities close to either 0 or 1, so
+the choice of probability threshold is only marginally significant.
+Additionally, even for a probability cut of 0.2, the purity of the
+resulting sample with RV measurements is very high (\textasciitilde90\%,
+J+24). Note there is likely a systematic bias in using stars with RV
+measurements to measure purity. Fainter stars are less likely to have
+been targeted and have poorer astrometry and colour information. We find
+no difference in the resulting density distributions when restricting
+stars to be brighter than a specific magnitude.
 
@@
 
-In figures fig.~\ref{fig:sculptor_selection},
-fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
-the algorithm in the tangent plane (\(\xi\), \(\eta\), \emph{does this
-need defined?}), CMD, and proper motion space. For both galaxies, each
-criteria plays a commensurate role in sifting out members. Proper
-motions are clustered around the dwarf systemic motion, the CMD is well
-defined including the horizontal branch, and stars only within a few
-\(R_h\) are included. We also mark stars with consistent radial
-velocities (RVs) (see below), which trace each feature albeit more
-cautiously.
+In Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, we
+illustrate the resulting samples from the algorithm in the tangent plane
+(\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and proper
+motion space. For both galaxies, each criteria plays a commensurate role
+in sifting out nonmembers. Proper motions are clustered around the dwarf
+systemic motion, the CMD is well defined including the horizontal
+branch, and stars only within a few \(R_h\) are included. We also mark
+stars with consistent radial velocities (RVs) (see below), which trace
+each feature albeit more cautiously. We additionally mark the furthest
+confirmed RV members of each galaxy, illustrating the large extent of
+each galaxy past \(R_h\).
 
-Figure REF also illustrats the distribution of stars selected without a
-spatial criterion. We define the CMD+PM selection as stars satisfying \[
+Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection} also
+illustrates the distribution of stars selected without a spatial
+criterion. We define the CMD+PM selection as stars satisfying
+\begin{equation}
 {\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
-\] These stars are distributed similar to the fiducial (probable
-members) sample but with a fainter uniform distribution across the
-entire field. This illustrates the approximate background of stars which
-may be confused as members. Additionally, since there is no clear
-spatial structure in the CMD+PM sample, it is unlikely that there are
-additional faint, tidal features detectable with Gaia. Not shown here,
-we also try a variety of simpler, absolute cuts and thresholds, finding
-no extended structure beyond what is detected in J+24. (\emph{Is it
-worth using this to calculate an OOM upper limit on the density of tidal
-tails?}).
+\end{equation} These stars are distributed similar to the fiducial
+(probable members) sample but with a fainter uniform distribution across
+the entire field. This illustrates the approximate background of stars
+which may be confused as members. Additionally, since there is no clear
+spatial structure in the CMD+PM sample outside several \(R_h\), it is
+unlikely that there are additional faint, tidal features detectable with
+Gaia. Not shown here, we also try a variety of simpler, absolute cuts
+and thresholds, finding no extended structure beyond what is detected in
+J+24. (\emph{Is it worth using this to calculate an OOM upper limit on
+the density of tidal tails?}).
 
@@
 \caption[Sculptor selection criteria]{The selection criteria for Scl
-members. Probable members (2-component) are orange, and all field stars
-(satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
-plane. We outline in star symbols the two stars from
-\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram.
-\textbf{Bottom right:} Proper motion.}\label{fig:sculptor_selection}
+members. Field stars (satesfying quality criteria) are light grey,
+CMD+PM selected stars are blue, probable members (2-component) are
+orange, and RV confirmed members are red. \textbf{Top:} Tangent plane
+\(\xi, \eta\). We outline in star symbols the two stars from
+\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram in
+Gaia G versus BP - RP. \textbf{Bottom right:} Proper motion in
+declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
+(corrected).}\label{fig:sculptor_selection}
 \end{figure}
@@
 \caption[Ursa Minor Selection]{Similar to
-fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
-members outside of \(3R_h\) in black stars (from \citet{sestito+2023b}
-and \citet{pace+2020} and
-\citet{spencer+2018}).}\label{fig:umi_selection}
+Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
+members outside of \(3R_h\) in black stars (from \citet{sestito+2023b},
+\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
 \end{figure}
@@
 simulations. To derive density profiles, we use 0.05 dex bins in log
-radius (i.e.~the bins are derived from
-10\^{}(minimum(logR):0.05:maximum(logR))). We remove bins at smaller
-(larger) radii than the first bin to contain no stars, working outwards.
-We use poisson uncertainties
-(\(\delta \Sigma_i / \Sigma_i = 1/\sqrt{N_i}\) for a derived density
-\(\Sigma_i\) in a bin with \(N_i\) stars). As discussed below, these
-uncertainties are straightforward but are likely under-represented.
+radius. We remove bins at smaller (larger) radii than the first bin to
+contain no stars, working outwards. We use symmetric poisson
+uncertainties. As discussed below, these uncertainties are
+straightforward but are likely under-representative.
 
-fig.~\ref{fig:scl_observed_profiles},
-fig.~\ref{fig:umi_observed_profiles} show the derived density profiles
-for Scl and UMi. We calculate density profiles for different selections
-of stars from above: all quality stars, CMD+PM only, and the probable
-member samples. In each case, all samples are the same towards the inner
-regions of the satellite. Classical dwarfs dominate the stellar density
-in their cores. However, the density profile \emph{all stars} plateaus
-at the total background in the field at radii of 30-60 arcminutes. By
-restricting stars to being most likely satellite members by CMD + PM,
-the background is reduced by 1-2 dex. The CMD+PM background plateau
-likely represents the density of background stars which could be
-mistaken as members. Finally, the background (BG)-subtracted profile
-results from subtracting the apparent background in the \emph{all}
-profile.
+Figs.~\ref{fig:scl_observed_profiles}, \ref{fig:umi_observed_profiles}
+show the derived density profiles for Scl and UMi. We calculate density
+profiles for different selections of stars from above: all quality
+stars, CMD+PM only, and the probable member samples. In each case, all
+samples are the same towards the inner regions of the satellite.
+Classical dwarfs dominate the stellar density in their cores. However,
+the density profile \emph{all stars} plateaus at the total background in
+the field at radii of 30-60 arcminutes. By restricting stars to being
+most likely satellite members by CMD + PM, the background is reduced by
+1-2 dex. The CMD+PM background plateau likely represents the density of
+background stars which could be mistaken as members. Finally, the
+background (BG)-subtracted profile results from subtracting the apparent
+background in the \emph{all} profile.
 
@@
 
-One interesting note is for UMi, there is a small ultrafaint star
-cluster at a position of (-100, -15) arcmin \citet{munoz+2012}. However
-this cluster does not have a bright RGB so has few stars brighter than g
-mag of 22. While munoz 1 may contribute about 4 stars to the density
-profile of UMi, this would only apply to the bin at XXX because of the
-small half light radius. We do not subtract this out.
+Nearby to UMi, there is a small, likely unassociated, ultrafaint star
+cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
+position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
+to an elliptical radius of 36 arcminutes. However this cluster does not
+have a bright RGB, so has few stars brighter than a \(G\) mag of 22.
+While Muñoz 1 may contribute \(\sim\) 5 stars to the density profile of
+UMi, this would possibly only affect 1-2 bins due to its compact size.
 
@@
 \caption[Sculptor density profiles]{The density profile of Sculptor for
-different selection criteria. \emph{probable members} selects stars with
-PSAT \textgreater{} 0.2 considering PM, CMD, and spatial, \emph{CMD+PM}
-select stars more likely to be members according to CMD and PM only,
-\emph{all} selects any high quality star, and \emph{BG subtracted} is
-the background-subtracted density derived from high-quality
-stars.}\label{fig:scl_observed_profiles}
+different selection criteria, ploted as log surface dencity versus log
+elliptical radius. Residuals are with respect to th interpolated
+\emph{probable members} density. \emph{Probable members} selects stars
+with PSAT \textgreater{} 0.2 considering PM, CMD, and spatial,
+\emph{CMD+PM} select stars more likely to be members according to CMD
+and PM only, \emph{all} selects any high quality star, and \emph{BG
+subtracted} is the background-subtracted density derived from
+high-quality stars. We mark the half-light radius ( vertical dashed
+line) and the break radius (black arrow,
+REF).}\label{fig:scl_observed_profiles}
 \end{figure}
@@
 
-In particular, in fig.~\ref{fig:umi_observed_profiles}, notice that the
-PSAT method produces artifically small errorbars even when the density
-is \textgreater1dex below the local background. These stars are likely
-selecting stars from the statistical MW background consistent with UMi
-PM / CMD, recovering the assumed density profile. As a result, the
-reliability of these density profiles below the CMD+PM background may be
-questionable. A more robust analysis, removing this particular density
-assumption, would be required to more appropriately represent the
-knowledge of the density profile as the background begins to dominate.
+In particular, in Fig.~\ref{fig:umi_observed_profiles}, notice that the
+PSAT method produces small errorbars, even when the density is \(>1\)dex
+below the local background. These stars are likely selecting stars from
+the statistical MW background consistent with UMi PM / CMD, recovering
+the assumed density profile. As a result, the reliability of these
+density profiles below the CMD+PM background may be questionable. A more
+robust analysis, removing this particular density assumption, would be
+required to more appropriately represent the knowledge of the density
+profile as the background begins to dominate.
 
@@
 
+Finally, our approximation to the poisson uncertainties Additionally, a
+more self consistent model would fit the density profile to the entire
+field at once, eliminating possible misrepresentation of the
+uncertainties.
+
 \section{Radial velocity modeling}\label{radial-velocity-modeling}
+
+\subsection{Data selection}\label{data-selection}
 
@@
 First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
-not report GaiaDR3 source ID's, we match to the nearest star within 2
-arcseconds. We exclude stars not matched to Gaia for simplicity.
+not report GaiaDR3 source ID's, we match to the nearest star within 1-3
+arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
+measurement from each study using the inverse-variance weighted mean
+\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
+variance \(s^2\). We remove stars with significant velocity dispersions
+as measured between observations in a study or between studies. By using
+that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
+\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
+distribution with \(N-1\) measurements. This cut typically removes stars
+with reduced chi-squared values
+\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
+number of measurements is 1-3 typically).
 
-We combine the mean RV measurement from each study using the
-inverse-variance weighted mean and standard uncertainty. \[
-\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} 
-\] \[
-\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
-\]
+Next, we need to correct the coordinate frames for the solar motion and
+on-sky size of the galaxy. We transform the frame into the galactic
+standard of rest (GSR). The next step is to account for the slight
+differences in the direction of each radial velocity. Let the \(\hat z\)
+be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
+the angular distance between the centre of the galaxy and the individual
+star, the corrected radial velocity is then \begin{equation}
+v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
+\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
+in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
+velocities in RA and Dec, and \(\theta\) is the position angle of the
+star with respect to the centre of the dwarf. The correction from both
+effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
+Sculptor and less for Ursa Minor \citep[see
+also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
+the distance uncertainty and velocity dispersion in quadrature to the RV
+uncertainties for each star. We then use the \(v_z\) values for the
+following modelling, however repeating with uncorrected, heliocentric
+velocities does not significantly affect the results.
 
-where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
-deviation \[
-s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
-\] We remove stars with significant velocity dispersions as measured
-between or within a study: \[
-P\left(\chi2(N-1) < \frac{s^2}{\delta v^2}\right) > 0.001
-\] where \(s, \delta v, N\) are the weighted standard deviation,
-weighted standard error, and number of observations. We apply this cut
-to both within a study and between studies, which typically removes
-stars with reduced \(\chi^2 \gtrsim 2\).
-
-The combined RV likelihood is then \[
+The combined likelihood, including RV information, becomes
+\begin{equation}
 {\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
-\] where \[
-{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
-\] \[
-{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
-\]
-
-where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
-dispersion of the satellite, and \(\delta v_i\) is the individual
-measurement uncertainty. Typically, the velocity dispersion will
-dominate the uncertainty budget here. We assume a halo/background
-velocity dispersion of a constant \(\sigma_{\rm halo} = 100\) km/s
+\end{equation} where we assume that the satellite and background
+distributions are Gaussian. Specifically, \begin{equation}
+\begin{split}
+{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
+{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
+\end{split}
+\end{equation} where \(f\) is the probability density of a standard
+normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
+velocity and dispersion of the satellite, and \(\delta v_i\) is the
+individual measurement uncertainty. Typically, the velocity dispersion
+will dominate the uncertainty budget here. We assume a halo/background
+velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
 \citep[e.g.][]{brown+2010}.
@@
 Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2.
+probability of greater than 0.2. Because of the additional information
+from radial velocities, most stars have probabilities close to 1 or 0 so
+the probability cut is not too significant.
 
-Finally, we need to correct the coordinate frames for the solar motion
-and on-sky size of the galaxy. The first step is to subtract out the
-solar motion from each radial velocity, corresponding to a typical
-gradient of \textasciitilde3 km/s across the field. The next step is to
-account for the slight differences in the direction of each radial
-velocity. Define the \(\hat z\) direction to point parallel to the
-direction from the sun to the centre of the dwarf galaxy. Then if
-\(\phi\) is the angular distance between the centre of the galaxy and
-the individual star, the corrected radial velocity is then \[
-v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
-\] where
-\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
-is the radial component of the proper motion with respect to the centre
-of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
-induces a gradient of about \(1 km/s/degree\) for sculptor
-\citep[see][]{walker+mateo+olszewski2008}. The uncertainty is then the
-velocity uncertainty plus the distance uncertainties times the PM
-uncertainty from above. We then use the \(v_z\) values for the following
-modelling, however repeating with uncorrected, heliocentric velocities
-does not significantly affect the results .
+We assume priors on systematic velocity and velocity dispersion of
+\begin{equation}
+\begin{split}
+\mu_{v} &= N(0\,\kms, \sigma_{\rm halo}^2) \\ 
+\sigma_{v} &= U(0, 20\,\kms)
+\end{split}
+\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
+the velocity dispersion of the MW halo adopted above, a reasonable
+assumption for dwarfs in orbit around the MW.
 
-For the priors on the satellite velocity dispersion and systematic
-velocity, we use \[
-\mu_{v} = N(0, \sigma_{\rm halo}^2) \\
-\sigma_{v} = U(0, 20\,{\rm km\,s^{-1}})
-\] where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is the velocity
-dispersion of the MW halo adopted above, a reasonable assumption for
-dwarfs in orbit around the MW.
-
-\subsection{Sculptor}\label{sculptor}
+\subsection{Results}\label{sec:rv_results}
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption{LOS velocity fit to Scl.}
-\end{figure}
-
-\textbf{Figure:} Velocity histogram of Scl and UMi in terms of \(v_z\)
-(REF). Orange points are from our crossmatched RV membership sample.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
-\caption[Scl velocity sample]{RV members of Scl plotted in the tangent
-plane coloured by corrected velocity difference from mean
-\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
-fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
-motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
-scale).}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
-\caption[Scl velocity gradient]{The corrected LOS velocity along the
-best fit rotational axis. RV members are black points, the systematic
-\(v_z\) is the horizontal grey line, blue lines represent the
-(projected) gradient from MCMC samples, and the orange line is a rolling
-median (with a window size of 50).}
+\caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
+terms of \(v_z\) (REF). Orange points are from our crossmatched RV
+membership sample.}
 \end{figure}
@@
 measurements. We find that there is no significant velocity shift in
-crossmatched stars between catalogues. After crossmatching to Gaia and
-excluding significant inter-study dispersions, we have a sample of XXXX
-members.
+crossmatched stars between catalogues. After crossmatching to high
+quality Gaia stars and excluding significant stellar velocity
+dispersions, we have a sample of 1918 members.
 
-We additionally add a velocity gradient for Sculptor (adding parameters
-A, B \(\sim N(0, 6)\)km/s/deg). We derive a systemic velocity of
-\(111.2\pm0.2\) km/s with velocity dispersion \(9.67\pm0.16\) km/s. Our
-values are very consistent with previous work
+We derive a systemic velocity for Sculptor of
+\(111.18\pm0.23\,\kms\)with velocity dispersion \(9.71\pm0.17\,\kms\).
+Our values are very consistent with previous work
 \citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{tolstoy+2023}]{WMO2009}. See Appendix for a more detailed
-comparison and inter-study tests.
+\citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
+detailed comparison between different samples and additional tests.
 
-We detect a gradient of \(4.8\pm1.3\) km/s/deg at a position angle of
-\(-147_{-12}^{+15}\) degrees. Compared to past work,
-\citet{battaglia+2008}, \citet{arroyo-polonio+2024}, our model is XX
+We detect a moderately significant gradient of \(4.3\pm1.4\,\kmsdeg\) at
+a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
+Several past work has attempted to detect a gradient in Sculptor, but no
+consensus has been reached. \citet{arroyo-polonio+2024} detect a
+velocity gradient of \(4\pm1.5\,\kmsdeg\) in a similar direction using
+the \citet{tolstoy+2023} sample, finding inconclusive statistical
+evidence. They additionally suggest a third chemodynamical component of
+the galaxy which may bias rotation measurements. \citet{battaglia+2008}
+also detect a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along
+the major axis, approximately the same direction. Instead,
+\citet{strigari2010}; \citet{martinez-garcia+2023} detect no significant
+gradient in Sculptor using \citet{WMO2009} sample. Note that
+pre-\emph{Gaia} work did not have as strong of a constraint on the
+proper motion of Scl, which limits conclusions of the intrinsic velocity
+gradient in Scl.
 
@@
 \citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
-shifted the velocities of \citet{spencer+2018} (\(-1.1\) km/s) and
-\citet{pace+2020} (\(+1.1\) km/s) to reach the same scale. We found 183
-crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
+shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
+\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
+183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
 dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
 the median difference in velocities in this crossmatch is about 2.2
-km/s, we adopt 1 km/s as the approximate systematic error here.
+km/s, we adopt 1 km/s as the approximate systematic error here. Our
+final sample includes 831 members.
 
-We derive a mean \(-245.9\pm0.3\) and velocity dispersion of
-\(8.76\pm0.24\). However, since the velocity means of
-\citet{spencer+2018} and \citet{pace+2020} differ by 2 km/s, we add a 1
-km/s systematic error to our measurement.
+We derive a mean \(-245.9\pm0.3_{\rm stat}\,\kms\) and velocity
+dispersion of \(8.76\pm0.24\,\kms\) for UMi. This is consistent with
+\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
+do not find evidence for a velocity gradient, consistent with past work
+\citep{pace+2020, martinez-garcia+2023}.
 
@@
 Our model here is relatively simple. Some things which we note as
-systematics and are challenging to account fully for are
+systematics or limitations:
 
@@
   simple velocity shift, combining data from multiple instruments is
-  challenging.
+  challenging. This appears to be a minor issue (Sculptor) or is
+  corrected for (Ursa Minor).
 \item
-  Inappropriate uncertainty reporting. Inspection of the variances
-  compared to the standard deviations within a study seems to imply that
-  errors are accurately reported. APOGEE notes that their RV
-  uncertainties are known to be underestimates but are a small
-  proportion of our sample.
+  Misrepresentative uncertainties. Inspection of the variances compared
+  to the standard deviations within a study seems to imply that errors
+  are accurately reported. APOGEE notes that their RV uncertainties are
+  known to be underestimates but are a small proportion of our sample.
 \item
   Binarity. While not too large of a change for classical dwarfs, this
-  could inflate velocity dispersions of about 9 km/s by about 1 km/s
-  \citet{spencer+2017}. Thus, our measurement is likely slightly
-  inflated given the high binarity fractions measured in classical
-  dwarfs \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
+  could inflate velocity dispersions of about \(9\,\kms\) by about
+  \(1\,\kms\)\citep{spencer+2017}. Thus, our measurement is likely
+  slightly inflated given the high binarity fractions measured in these
+  systems \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
 \item
-  Selection effects. RV studies each have their own selection effects. I
-  do not know how to correct for this.
+  Multiple populations. Both Sculptor and Ursa Minor likely contain
+  multiple populations \citep[\citet{pace+2020},
+  \citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a
+  single population, and each population may have a different extent and
+  velocity dispersion, this could result in biased velocity dispersions.
+  However, it is unclear how to uniquely determine an overall velocity
+  dispersion in a multi-population system.
+\item
+  Selection effects. RV studies each have their own selection effects,
+  which may affect the resulting dispersion, especially if different
+  populations or regions of the galaxy have different velocities or
+  velocity dispersions. We do not attempt to correct for this.
 \end{itemize}
 
-Because the derived parameters are similar for the two different larger
-surveys we consider for UMi and Scl, we note that many of these effects
-are likely not too significant (with the exception of the systemic
-motion of UMi.)
+For both Ursa Minor and Sculptor, we also fit models to only data from
+individual surveys (see REF). Since the resulting parameters are very
+similar, we conclude that many of the systematic uncertainties are
+likely smaller than the present errors or that each large survey has
+similar biases.
 
@@
 To illustrate the differences between each dwarf galaxy, in
-fig.~\ref{fig:classical_dwarfs_densities}, we compare Scl, UMi, and Fnx
-against exponential and plummer density profiles (\textbf{TODO: state
-these somewhere}). While all dwarfs have marginal differences in the
-inner regions, each dwarf diverges in the outer regions relative to an
-exponential. In particular, while Fnx is underdense, Scl and UMi are
-both overdense, approximately fitting a Plummer density profile instead.
+Fig.~\ref{fig:classical_dwarfs_densities}, we compare Sculptor, Ursa
+Minor, and Fornax against 2D-exponential and Plummer density profiles
+(REF). While all dwarfs appear similar in the inner regions, each dwarf
+diverges in the outer regions relative to an exponential. Relative to an
+exponential, Fornax is underdense but Sculptor and Ursa Minor are both
+overdense. A Plummer profile instead provides a more reasonable fit to
+Scl and UMi.
 
@@
 Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
-robust against different selection criteria. Both Sculptor (Ursa Minor)
-show strong (weak) evidence for deviations from an exponential profile.
+robust against different selection criteria. Both Sculptor and Ursa
+Minor show strong evidence for deviations from an exponential profile.
 We also compile velocity measurements to derive the systemic motions and
 velocity dispersions of each galaxy. We find evidence for a velocity
-gradient in Scl of XXX. We find no evidence of additional (velocity or
-stellar) substructure in either galaxy. Our goal is thus to explain why
-Scl and UMi have an excess of stars in their outer regions and why Scl
-may have a velocity gradient.
+gradient in Sculptor of \(4.3\pm1.3\,\kmsdeg\). We find no evidence of
+additional (velocity or stellar) substructure in either galaxy. Our goal
+in the following chapters is to test if tides provide a viable
+explanation for the observed properties of Sculptor and Ursa Minor.
 
@@
 density at half-light radius (fit from the inner 3 scale radii
-exponential recursively. ). Scl and UMi have an excess of stars in the
-outer regions (past \(\log R/R_h \sim 0.3\)) compared with
-Scl.}\label{fig:classical_dwarfs_densities}
+exponential recursively. ). Sculptor and Ursa Minor have an excess of
+stars in the outer regions (past \(\log R/R_h \sim 0.3\)) compared with
+Sculptor. Sculptor and Ursa Minor's density profiles are created with
+the 2-component J+24 model, but the excess does not change significantly
+for the 1-component model.}\label{fig:classical_dwarfs_densities}
 \end{figure}

### 2025-05-02 thesis/introduction.tex

--- 2025-04-24 thesis/introduction.tex
@@
 
+\begin{itemize}
+\tightlist
+\item
+  \citet{fattahi+2013}, \citet{fattahi+2018}
+\item
+  \citet{sanchez-salcedo+hernandez2007}: mond in dsph
+\item
+  \citet{mayer+2001} theory of tidal stripping
+\item
+  \citet{IH1995} structural parameters
+\end{itemize}
+
 \subsection{Cosmological context}\label{cosmological-context}
@@
 
-\subsection{Sculptor}\label{sculptor}
+\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}Figure:
+Image of the Sculptor dwarf spheroidal galaxy from Dark Energy Survey
+Data Release 2 {[}\citet{abbott+2021}; image created with HiPS2FITS{]}.
+Sculptor appears as a fairly prominent, extended over density of
+predominantly faint, red stars. 0.5 degree field of view centred on
+Sculptor.
 
-\begin{figure}
-\centering
-\includegraphics[width=5.41667in,height=5.41667in]{figures/scl_des_dr2.png}
-\caption[Picture of Sculptor]{Sculptor image from DES DR 2 via
-HiPS2FITS.}\label{fig:scl_image}
-\end{figure}
+\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}Figure:
+Image of Ursa Minor dwarf galaxy from the Digitized Sky Survey 2 (0.75
+deg field of view tangent plane). UMi appears as a diagonal/elliptical
+haze of faint, reddish stars from the top left to the bottom right. Even
+as classical dwarf, Ursa minor is fairly diffuse and does not stand
+obviously out from the background.
 
@@
 \end{figure}
-
-Sculptor (Scl) is one of the first discovered dwarf galaxies of the
-Milky Way (Shapley 1938; only preceded by the SMC and LMC!). As a
-classical dwarf spheroidal, Scl is relatively bright and compact.
-
-Since the initial discovery of Scl, most authors have noted that Scl has
-a slight ellipticity (\(0.36\)), often attributed to tidal effects.
-However, this does not align with the absolute proper motion or orbital
-path of the galaxy.
-
-While Scl has a relatively large pericentre (greater than 50kpc),
-\citet{sestito+2023a} detect that the galaxy likely has an excess of
-stars in the outskirts (past about 60 arcminutes). This excess is
-perhaps one of the clearest indications that Scl may be affected by
-tides. Here, our goal is to determine if under a \(\Lambda\) CMD
-paradigm with DM-only simulations, tidal effects of the Milky Way (or
-LMC) may indeed be consistent with these observations, or if these
-observations may reveal instead a extended stellar ``halo'' or second
-component of the galaxy---illustrating a complex formation for the
-galaxy.
-
-Theoretical work on Sculptor
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{battaglia+2008}
-\item
-  \citet{iorio+2019}
-\item
-  \citet{amorisco+zavala+deboer2014}
-\item
-  \citet{battaglia+2008}
-\item
-  \citet{breddels+2013}
-\item
-  \citet{breddels+helmi2013}
-\item
-  \citet{richardson+fairbairn2014}
-\item
-  \citet{SFW2017}
-\item
-  \citet{innanen+papp1979}
-\end{itemize}
-
-Observational work on Scl
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{sestito+2023a}
-\item
-  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
-  \citet{arroyo-polonio+2024}
-\item
-  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
-\item
-  \citet{coleman+dacosta+bland-hawthorn2005}
-\item
-  \citet{DQ1994}
-\item
-  \citet{WMO2009}
-\item
-  \citet{IH1995}, \citet{munoz+2018}
-\item
-  \citet{kirby+2009}
-\item
-  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
-\end{itemize}
-
-\begin{table*}[t]
-\centering
-\caption{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018), 2. Tran et al. (2022), 3. (MV2020b?), 4 McConnachie and Venn (2020).}
-\label{tbl:Observed-properties-of-Sculptor-References-are:-1-Muñoz-et-al-2018-2-Tran-et-al-2022-3-MV2020b?-4-McConnachie-and-Venn-2020}
-\begin{tabular}{lll}
-\toprule
-parameter & value & Source\\
-\midrule
-$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
-$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
-distance modulus & $19.60 \pm 0.05$ (RRL) & 2\\
-distance & $83.2 \pm 2$ kpc & 2\\
-$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $111.2 \pm 0.2\ {\rm km\,s^{-1}}$ & this work\\
-$\sigma_v$ & $9.67\pm0.16\ {\rm km\,s^{-1}}$ & this work\\
-$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
-$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
-ellipticity & $0.36 \pm 0.01$ & 1\\
-position angle & $92\pm1^\circ$ & 1\\
-$M_V$ & $-10.82\pm0.14$ & 1\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\subsection{Ursa Minor}\label{ursa-minor}
-
-\begin{itemize}
-\item
-  \citet{sestito+2023b}
-\item
-  \citet{pace+2020}
-\item
-  \citet{bellazzini+2002}
-\item
-  \citet{hargreaves+1994}
-\item
-  \citet{martinez-delgado+2001}
-\item
-  \citet{munoz+2005}
-\item
-  \citet{pace+2020}
-\item
-  \citet{palma+2003}
-\item
-  \citet{spencer+2018}
-\item
-  \citet{vitral+2023}
-\end{itemize}
-
-\begin{figure}
-\centering
-\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
-\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. UMi appears as a
-diagonal/elliptical haze of stars from the top left to the bottom right.
-Even as as classical dwarf, Ursa minor is fairly diffuse and does not
-stand obviously out from the background. Composite Public data/Giuseppe
-Donatiello. Updated image in 2022.
-https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
-\end{figure}
-
-\begin{table*}[t]
-\centering
-\caption{Observed properties of Ursa Minor.
-References are: (1) Muñoz et al.
-(2018), (2) Garofalo et al.
-(2025), (3) McConnachie and Venn (2020), (4) average of Pace et al.
-(2020) and Spencer et al.
-(2018).}
-\label{tbl:Observed-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
-\begin{tabular}{lll}
-\toprule
-parameter & value & Source\\
-\midrule
-$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
-$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
-distance modulus & $19.23 \pm 0.11$ (RRL) & 2\\
-distance & $70.1 \pm 3.6$ kpc & -\\
-$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
-$\sigma_v$ & $8.8 \pm 0.2$ & this work\\
-$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
-ellipticity & $0.55 \pm 0.01$ & 1\\
-position angle & $50 \pm 1^\circ$ & 1\\
-$M_V$ & $-9.03 \pm 0.05$ & 1\\
-\bottomrule
-\end{tabular}
-\end{table*}
 
@@
 An emperical estimate of where the simulation's stars are becoming
-unbound is, as stated in \citet{PNM2008}, the break radius \[
+unbound is, as stated in \citet{PNM2008}, the break radius
+\begin{equation}
 R_b = C\,\sigma_{v}\,\Delta t
-\] where \(\sigma_v\) is the present line of sight velocity dispersion ,
-\(\delta t\) is the time since pericentre, and \(C \approx 0.55\) is a
-fit. The idea motivating this equation is stars in the inner regions
-will have dynamically equilibriated to the new potential (phase mixed),
-however the outer regions are no longer in steady state, so we have to
-wait until the crossing time reaches them as well.
+\end{equation} where \(\sigma_v\) is the present line of sight velocity
+dispersion , \(\delta t\) is the time since pericentre, and
+\(C \approx 0.55\) is a fit. The idea motivating this equation is stars
+in the inner regions will have dynamically equilibriated to the new
+potential (phase mixed), however the outer regions are no longer in
+steady state, so we have to wait until the crossing time reaches them as
+well.
 
@@
 
-From the \citet{nfw1996} paper, eqns. 3 \& 4 \[
+From the \citet{nfw1996} paper, eqns. 3 \& 4 \begin{equation}
 \frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
-\] where \[
+\end{equation} where \begin{equation}
 \delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
-\] \(c\) is concentration parameter, \(\rho_c\) is the critical density
-of the universe, and \(r_s\) is the characteristic scale length of the
-halo.
+\end{equation} \(c\) is concentration parameter, \(\rho_c\) is the
+critical density of the universe, and \(r_s\) is the characteristic
+scale length of the halo.
 
@@
 200 times the critical density of the universe, and \(M_{200}\) is the
-mass contained inside \(r_{200}\). As equations: \[
+mass contained inside \(r_{200}\). As equations: \begin{equation}
 \rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
-\]
+\end{equation}
 
-\[
+\begin{equation}
 r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
 }}}
-\]
+\end{equation}
 
-\[
+\begin{equation}
 M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
-\]
+\end{equation}
 
 \(M_{200}\) is also sometimes called the virial mass of the halo.
-\(r_{200}\) is directly related to \(r_s\) by \[
+\(r_{200}\) is directly related to \(r_s\) by \begin{equation}
 r_{200} = c\,r_s
-\]
+\end{equation}
 
-Another useful definition is \[
+Another useful definition is \begin{equation}
 A(x) \equiv \log (1+x) - \frac{x}{1+x}.
-\]
+\end{equation}
 
-We will also define a dimensionless radius \[
+We will also define a dimensionless radius \begin{equation}
 x \equiv r/r_s.
-\]
+\end{equation}
 
@@
 
-\[
+\begin{equation}
 \rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
-\] where \[
+\end{equation} where \begin{equation}
 \rho_s = \frac{c^3}{A(c)} \rho_{200}
-\] and \(A(c)\) is as above. The characteristic density can also be
-written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see below),
-giving \[
+\end{equation} and \(A(c)\) is as above. The characteristic density can
+also be written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see
+below), giving \begin{equation}
 \rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
-\] Note that the NFW density profile is the same as an alpha-beta-gamma
-profile where \(\alpha=\gamma=1\) and \(\beta =3\).
+\end{equation} Note that the NFW density profile is the same as an
+alpha-beta-gamma profile where \(\alpha=\gamma=1\) and \(\beta =3\).
 
@@
 The circular velocity in terms of
-\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \[
+\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \begin{equation}
 \left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
-\]
+\end{equation}
 
-or in terms of \(M_s\) and \(r_s\), \[
+or in terms of \(M_s\) and \(r_s\), \begin{equation}
 v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
-\] Another parameterization of the NFW profile is in terms of the
-maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the radius at
-which it is reached \(r_{\rm circ}^{\rm max}\). Given the scale radius,
-\[
+\end{equation} Another parameterization of the NFW profile is in terms
+of the maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the
+radius at which it is reached \(r_{\rm circ}^{\rm max}\). Given the
+scale radius, \begin{equation}
 r_{\rm circ}^{\rm max} = \alpha\ r_s
-\]
+\end{equation}
 

### 2025-05-02 thesis/methods.tex

@@
 
-\[
+\begin{equation}
 \Phi(r) = - \frac{GM}{r + a}
-\] where \(a=1.3\,{\rm kpc}\) is the scale radius and
+\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
 \(M=2.1 \times 10^{10}\,\Mo\) is the total mass.
@@
 The thin and thick disks are represented with the
-\citet{miyamoto+nagai1975} cylindrical potential: \[
+\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}
 \Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
-\]
+\end{equation}
 
@@
 
-We adopt a softening length of \[
+We adopt a softening length of \begin{equation}
 h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-\] See appendix REF for a discussion of this choice, which is similar to
-the \citet{power+2003} suggested softening.
+\end{equation} See appendix REF for a discussion of this choice, which
+is similar to the \citet{power+2003} suggested softening.
 
@@
 collisionless assumption breaks down). The relaxation timescale is given
-by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise, \[
+by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise,
+\begin{equation}
 t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
 = {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
-\] This works out to be about 6-10 times (increasing with particle
-number) our adopted softening length for NFW halos given our
+\end{equation} This works out to be about 6-10 times (increasing with
+particle number) our adopted softening length for NFW halos given our
 assumptions. As such, at full resolution, we can only trust density
@@
 
-\[
+\begin{equation}
 v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
-\]
+\end{equation}
 
@@
 energy), then we assign the stellar weight for a given particle with
-energy \({\cal E}\) is \[
+energy \({\cal E}\) is \begin{equation}
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-\] While \(f({\cal E})\) is a phase-space density, the differential
-energy distribution includes an additional \(g({\cal E})\) occupation
-term (BTXXX).
+\end{equation} While \(f({\cal E})\) is a phase-space density, the
+differential energy distribution includes an additional \(g({\cal E})\)
+occupation term (BTXXX).
 
 We use Eddington inversion to find the distribution function, (eq.
-4-140b in BT87) \[
+4-140b in BT87) \begin{equation}
 f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-\]
+\end{equation}
 
@@
 
-The 2D exponential stellar profile is given by \[
+The 2D exponential stellar profile is given by \begin{equation}
 \Sigma_\star(R) = A e^{-R / R_s}
-\] where \(A\) is some normalization and \(R_s\) is the exponential
-scale radius. The 3D deprojected profile is found through abel inversion
-(e.g.~Rapha\ldots), i.e. \[
+\end{equation} where \(A\) is some normalization and \(R_s\) is the
+exponential scale radius. The 3D deprojected profile is found through
+abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
 \rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
-\] where \(K_0\) is the 0th order modified Bessel function of the second
-type.
+\end{equation} where \(K_0\) is the 0th order modified Bessel function
+of the second type.
 
@@
 
-A Plummer profile is defined by \[
+A Plummer profile is defined by \begin{equation}
 \Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
-\] where \(M\) is the total mass and \(R_s\) is the characteristic scale
-radius. The density is \[
+\end{equation} where \(M\) is the total mass and \(R_s\) is the
+characteristic scale radius. The density is \begin{equation}
 \rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
-\]
+\end{equation}
 

### 2025-05-19 thesis/acronyms.tex

@@
 
+% Ordinary abbreviations
+\newcommand{\kms}{{\rm km\,s^{-1}}}
+\newcommand{\kmsdeg}{{\rm km\,s^{-1}\,deg^{-1}}}
+
 \setabbreviationstyle[acronym]{long-short-sc}
 \GlsXtrEnablePreLocationTag{\S }{\S }
-
-\newacronym
-[ description ={Core collapse supernovae. Massive star explosions. {\sc Ccsne} produce many elements including \gls{alpha}, \gls{fepeak}, and r and \gls{sproc}.}
-]
-{cc}{ccsne}{core collapse supernovae}
-
-\newacronym
-[description={Asymptotic giant branch stars.
-The {\sc agb} phase is the last phase of \glspl{lowmass}, before stars become white dwarfs.
-Produces C, N, and heavy \gls{sproc} elements.
-See \ref{sec:stel_evo}.}.]
-{agb}{agb}{asymptotic giant branch}
-
-\newacronym
-[description={Type Ia Supernovae. Exploding white dwarfs. Produces \gls{fepeak} and has a long delay time.}]
-{ia}{sne ia}{supernovae type Ia}
-
-\newacronym
-[description={Galactic chemical evolution.}]
-{gce}{gce}{galactic chemical evolution}
-
-\newacronym
-[description={Star formation history.}]
-{sfh}{sfh}{star formation history}
-
-\newacronym
-[description={Single stellar population. A group of stars formed all at the same time.}]
-{ssp}{ssp}{single stellar population}
 
@@
 
-\newacronym
-[
-description={Damped Lyman-alpha system. {\sc dla}s are clouds of gas from the 
-early universe which are observed through their absorption of quasar spectra. The name 
-comes from the strong Lyman-alpha lines ($1216\AA$) due to H absorption.}
-]
-{dla}{dla}{damped Lyman-alpha systems}
-
-\newcommand{\cc}{\gls{cc}}
-\newcommand{\Cc}{\Gls{cc}}
-\newcommand{\agb}{\gls{agb}}
-\newcommand{\ia}{\gls{ia}}
-\newcommand{\sfh}{\gls{sfh}}
-\newcommand{\dla}{\gls{dla}}
-\newcommand{\ssp}{\gls{ssp}}
-\newcommand{\imf}{\gls{imf}}
-\newcommand{\gce}{\gls{gce}}
-\newcommand{\Gce}{\Gls{gce}}
-\newcommand{\apogee}{\gls{apogee}}
-
 %%%%%%%%%%%%%%%%%%%%%% terms %%%%%%%%%%%%%%%
@@
 }
-
-\newglossaryentry{multizone}{name={multi-zone},
-    description={A chemical evolution model where a galaxy is divided into several zones, each with different stars, gas, and properties. }
-}
-
-\newglossaryentry{onezone}{name={one-zone},
-    description={A chemical evolution model where the gas is all the same composition, i.e. neglecting spatial variations.}
-}
-
-\newglossaryentry{alpha}{name={$\alpha$-element},
-    description={Elements which are made up of $\alpha$-particles (He nuclei) throuch the triple-$\alpha$ process (see Eq.~\ref{eq:triple_alpha}).
-Essentially light, even-numbered elements like O, Mg, and Na.}
-}
-
-\newglossaryentry{fepeak}{name={Fe-peak elements},
-    description={Fe and nearby elements, produced in \gls{cc} and \gls{ia}.}
-}
-
-\newglossaryentry{sproc}{name={s-process elements},
-    description={Elements produced through slow neutron captures, typically in \agb\ stars.}
-}
-
-\newglossaryentry{rgb}{name={red giant branch},
-    description={Red giant branch stars are stars that have completed hydrogen core burning and have expanded in size. See Appendix \ref{sec:stel_evo}.}
-}
-
-\newglossaryentry{tdu}{name={third dredge up},
-    description={Third dredge up occurs inside \gls{agb} stars.
-During each thermal pulse, material is {\it dredged up} from the core, changing the chemical abundances of the stellar atmosphere.
-(While nominally called {\it third dredge up}, there are typically several third dredge ups.).
-See Appendix \ref{sec:stel_evo}.}
-}
-
-\newglossaryentry{fdu}{name={first dredge up},
-    description={First dredge up occurs when a \gls{lowmass} enters the \gls{rgb} phase.
-Material from the core is brought to the surface, increasing N and decreasing C abundances.
-See Appendix \ref{sec:stel_evo}.}
-}
-
-\newglossaryentry{hbb}{name={hot bottom burning},
-    description={Hot bottom burning occurs inside \agb\ stars. The base of the convective envelope becomes hot enough for CNO burning to initiate.}
-}
-
-\newglossaryentry{subgiant}{name={subgiant},
-    description={A star in the process of leaving the main sequence and becoming a \gls{rgb}.}
-}
-
-\newglossaryentry{imfave}{name={\gls{imf}-averaged},
-    description={Averaged over the initial-mass function ({\sc imf}).
-The {\imf}-weighted yield is the mass of the newly produced element divided by the mass of star formation for a single stellar population (see \ssp).}
-}
-
-\newglossaryentry{dtd}{name={delay time distribution},
-    description={The distribution in time of when an element is produced 
-    after a star formation event.}
-}
-
-\newglossaryentry{massloading}{name={mass loading},
-    description={The strength of outflows relative to star formation. See also $\eta$. }
-}
-
-\newglossaryentry{lowalpha}{name={low-$\alpha$},
-    description={The low-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
-}
-
-\newglossaryentry{highalpha}{name={high-$\alpha$},
-    description={The high-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
-}
-
-\newglossaryentry{lowmass}{name={low-mass star},
-    description={Stars with masses $\lesssim 8\,M_\odot$ which end life as a white dwarf. }
-}
-
-\newglossaryentry{highmass}{name={high-mass star},
-    description={Stars with masses $\lesssim 8\,M_\odot$, which end as a neutron star, black hole, or supernovae.}
-}
-
-\newglossaryentry{insideout}{name={{\it insideout}},
-    description={Our fiducial star formation history. The rate of star formation is highest towards the center of the galaxy and at earlier times. See Eq.~\ref{eq:inside_out}.}
-}
-
-\newglossaryentry{cno}{name={CNO cycle},
-    description = { A proton-fusion cycle which occurs in \gls{rgb} stars 
-        consisting of a chain of proton captures releasing a He nucleus ($\alpha$-particle) and energy. See Eq.~\ref{eq:cno_cycle}.
-}
-}

### 2025-05-19 thesis/appendix.tex

--- 2025-04-15 thesis/numerical_convergence.tex
@@
+\section{Density profile tests}\label{density-profile-tests}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
+\caption[Density profiles]{Density profiles for various assumptions for
+Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
+2-component bayesian model assuming circular radii, simple is the series
+of simple cuts described, bright is the sample of the brightest half of
+stars (scaled by 2), DELVE is a sample of RGB stars (background
+subtracted and rescaled to
+match).}\label{fig:sculptor_observed_profiles}
+\end{figure}
+
+Note that a full rigorous statistical analysis would require a
+simulation study of injecting dwarfs into Gaia and assessing the
+reliability of various methods of membership and density profiles. This
+is beyond the scope of this thesis.
+
+\begin{verbatim}
+SELECT TOP 1000
+       *
+FROM delve_dr2.objects
+WHERE 11 < ra
+and ra < 19
+and -37.7 < dec
+and dec < -29.7
+\end{verbatim}
+
+\subsection{An MCMC model}\label{an-mcmc-model}
+
+We additionally consider a model which solves for the density relative
+to the background in each annulus. This model is specified as \[
+\log\Sigma_i \sim {\rm U}(-12, 6)
+\] where the stars are statically placed into each bin. This model
+solves for the density in each bin independently.
+
+\section{Velocity modelling and
+comparisons}\label{velocity-modelling-and-comparisons}
+
+Here, we describe in additional detail, our methods and comparisons for
+RV modelling between studies.
+
+Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
+smoothed samples from posterior/prior.
+
+\begin{table*}[t]
+\centering
+\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
+\begin{tabular}{lllllllll}
+\toprule
+ & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
+\midrule
+Scl & combined &  & 8945 & 2280 & 2034 & 1918 & 0.9 & \\
+ & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1480 & 0.65 & –\\
+ & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
+ & walker+09 & MMFS & 1818 & 1522 & 1417 & 1329 & 1.8 & 3\\
+ & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6 & –\\
+UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3 & \\
+ & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
+ & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5 & 1\\
+ & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & ?\\
+ & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9 & –\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Sculptor RV fits]{\label{tbl:scl_rv_mcmc}MCMC fits for different RV datasets for Sculptor among 3 different models. }
+\label{tbl:scl_rv_mcmc}
+\begin{tabular}{lllllllllll}
+\toprule
+study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\hat R$ & $n_{\rm eff}$ & $\log B_2/B_1$ & WAIC & LOO\\
+\midrule
+all &  &  &  &  &  &  &  &  &  & \\
+ & $111.18\pm0.23$ & $9.71\pm0.17$ & - & - & - &  &  & 0 &  & \\
+ & $111.19 \pm 0.23$ & $9.68\pm0.17$ & - & $4.3\pm1.3$ & $-149_{-13}^{+17}$ &  &  & -1.7 &  & \\
+ & $111.16\pm0.23$ & $9.73\pm0.17$ & $0.056\pm0.021$ & - & - &  &  & -0.9 &  & \\
+tolstoy+23 &  &  &  &  &  &  &  &  &  & \\
+ & $111.3 \pm 0.3$ & $9.80 \pm 0.18$ & - & - & - &  &  & 0 &  & \\
+ & $111.3\pm0.3$ & $9.78\pm0.18$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ &  &  & -1.4 &  & \\
+ & $111.2 \pm 0.3$ & $9.74\pm0.19$ & $0.085 \pm 0.023$ & – & – &  &  & -4.5 &  & \\
+walker+09 &  &  &  &  &  &  &  &  &  & \\
+ & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – &  &  & 0 &  & \\
+ & $111.1\pm0.3$ & $9.53\pm0.21$ & - & $5.2_{-1.6}^{+1.8}$ & $-134_{-16}^{+23}$ &  &  & -2.4 &  & \\
+ & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – &  &  & +1.6 &  & \\
+apogee &  &  &  &  &  &  &  &  &  & \\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – &  &  & – &  & \\
+ & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ &  &  & +0.4 &  & \\
+ & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – &  &  & +1.1 &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Ursa Minor RV fits]{\label{tbl:umi_rv_mcmc}MCMC fits for UMi velocity dispersion. }
+\label{tbl:umi_rv_mcmc}
+\begin{tabular}{lllllll}
+\toprule
+study & mean & sigma & $\hat R$ & $n_{\rm eff}$ & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
+\midrule
+all & $-245.9\pm0.3$ & $8.76\pm0.24$ &  &  & +1.5 & +2.2\\
+pace & $-244.5\pm0.4$ & $9.1\pm0.3$ &  &  & +0.2 & +1.1\\
+spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ &  &  & +1.8 & -0.3\\
+apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ &  &  & +0.8 & +0.8\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_rv_scatter.pdf}}
+\caption[Scl velocity sample]{RV members of Sculptor plotted in the
+tangent plane coloured by corrected velocity difference from mean
+\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
+fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
+motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
+scale).}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
+\caption[Scl velocity gradient]{The corrected LOS velocity along the
+best fit rotational axis. RV members are black points, the systematic
+\(v_z\) is the horizontal grey line, blue lines represent the
+(projected) gradient from MCMC samples, and the orange line is a rolling
+median (with a window size of 50).}
+\end{figure}
+
+\section{Numerical Convergence}\label{numerical-convergence}
+
 Here, we describe some convergence tests to ensure our methods and

### 2025-05-19 thesis/sculptor.tex

@@
+\section{Abstract}\label{abstract}
+
+\section{Introduction}\label{introduction}
+
+Sculptor (Scl) is one of the first discovered dwarf galaxies of the
+Milky Way (Shapley 1938; only preceded by the SMC and LMC!). As a
+classical dwarf spheroidal, Scl is relatively bright and compact.
+
+Since the initial discovery of Scl, most authors have noted that Scl has
+a slight ellipticity (\(0.36\)), often attributed to tidal effects.
+However, this does not align with the absolute proper motion or orbital
+path of the galaxy.
+
+While Scl has a relatively large pericentre (greater than 50kpc),
+\citet{sestito+2023a} detect that the galaxy likely has an excess of
+stars in the outskirts (past about 60 arcminutes). This excess is
+perhaps one of the clearest indications that Scl may be affected by
+tides. Here, our goal is to determine if under a \(\Lambda\) CMD
+paradigm with DM-only simulations, tidal effects of the Milky Way (or
+LMC) may indeed be consistent with these observations, or if these
+observations may reveal instead a extended stellar ``halo'' or second
+component of the galaxy---illustrating a complex formation for the
+galaxy.
+
+Theoretical work on Sculptor
+
+\begin{itemize}
+\tightlist
+\item
+  \citet{battaglia+2008}
+\item
+  \citet{iorio+2019}
+\item
+  \citet{amorisco+zavala+deboer2014}
+\item
+  \citet{battaglia+2008}
+\item
+  \citet{breddels+2013}
+\item
+  \citet{breddels+helmi2013}
+\item
+  \citet{richardson+fairbairn2014}
+\item
+  \citet{SFW2017}
+\item
+  \citet{innanen+papp1979}
+\item
+  \citet{wilkinson+2002}
+\item
+  \citet{yang+2025}: chemical evolution in Scl.
+\item
+  \citet{skuladottir+2024}; \citet{skuladottir+2021};
+  \citet{lee+jeon+bromm2024}, pop III high res spectro?
+\item
+  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
+\item
+  \citet{tang+2023}: apogee modeling of Scl.
+\item
+  \citet{pandey+west2022} chem evo OMEGA and isotopes.
+\item
+  \citet{an+koposov2022}: distance gradients?
+\item
+  \citet{kawata+2006}, 2pop origins?
+\end{itemize}
+
+Observational work on Scl
+
+\begin{itemize}
+\tightlist
+\item
+  \citet{sestito+2023a}
+\item
+  \citet{westfall+2006} wide degree survey for extended structure.
+\item
+  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
+  \citet{arroyo-polonio+2024}
+\item
+  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
+\item
+  \citet{coleman+dacosta+bland-hawthorn2005}
+\item
+  \citet{DQ1994}
+\item
+  \citet{WMO2009}
+\item
+  \citet{IH1995},
+\item
+  \citet{munoz+2018}: Using Megacam to derive density profiles and
+  structural properties of many dwarf spheroidal galaxies.
+\item
+  \citet{kirby+2009}
+\item
+  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
+\item
+  \citet{grebel1996}
+\item
+  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
+  metallicity gradient and search for metal poor stars in Scl.
+\end{itemize}
+
+Future ideas:
+
+\begin{itemize}
+\tightlist
+\item
+  \citet{evslin2016}: measuring ellipticities of halos w TMT.
+\end{itemize}
+
 \section{Sculptor (MW only)}\label{sculptor-mw-only}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}}
-\caption{image-20250415112706417}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}
+\caption[Effect of LMC on Sculptor's Orbit]{Orbits of Sculptor with and
+without the LMC. The potential is that of \citet{vasiliev+2021} in each
+case.}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}}
-\caption{image-20250415113122385}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}
+\caption[Selected orbits of Sculptor with an LMC]{Selected orbits of
+Sculptor in the MW frame. \textbf{TODO}: include LMC distance or frame,
+maybe xyz instead?}
 \end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113133353.png}}
-\caption{image-20250415113133353}
-\end{figure}

### 2025-05-31 thesis/data.tex
thesis updates / reorganization of RV modeling

@@
 To provide an empirical benchmark for simulations in the next chapters,
-we would like to understand and determine the observed properties of
-Sculptor and Ursa Minor. In this section, we first review past
-measurements of each galaxy and the observed properties of each dwarf
-galaxy. We describe the \citet{jensen+2024} Bayesian methodology to
-select high probability member stars from \emph{Gaia}. We test the
-derived density profiles for Sculptor, Ursa Minor, and Fornax, finding
-that \citet{jensen+2024}'s methodology successfully reduces the
-background and produces similar results to a simple
-background-subtracted density model. We show that Fornax is
-approximately represented by a 2-D exponential profile. Instead,
-Sculptor and Ursa Minor show an excess of stars relative to an 2-D
-exponential in the outer regions. A Plummer profile provides a better
-fit instead. Finally, we combine various radial velocity samples with
-\citet{jensen+2024}'s members to derive systemic velocities and
-dispersions for Sculptor and Ursa Minor.
+we aim to understand the observed properties of Sculptor and Ursa Minor.
+In this section, we first compile past measurements of each galaxy. We
+describe the \citet{jensen+2024} Bayesian methodology to select high
+probability member stars from \emph{Gaia}. The derived density profiles
+are robust to changes in assumptions. Fornax is well represented by a 2D
+exponential profile. Instead, Sculptor and Ursa Minor show an excess of
+stars relative to an 2D exponential in the outer regions, better
+described my a Plummer profile. Finally, we combine various radial
+velocity samples with \citet{jensen+2024}'s members to derive systemic
+velocities and dispersions for Sculptor and Ursa Minor.
 
@@
 
-The determination of basic properties of astronomical objects is an
-ongoing and challenging endeavour.
+Since the discovery of dwarf galaxies around the Milky Way,
+observational work has attempted to measure and refine the basic
+properties of these objects. While the Milky Way's satellites are close
+(by extragalactic standards), their low numbers of stars and large areas
+on the sky present challenges for observational work. The classical
+systems discussed here each extend about 1-2 degrees across the sky
+(figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}).
 
-\subsection{Gaia}\label{gaia}
+In particular, \emph{Gaia} has provided a wealth of high quality data,
+including the distances to nearby stars and proper motions. This allows
+for some of the first measurements on the motions for many local group
+dwarf galaxies {[}e.g.; \citet{MV2020}{]}. Accurate distances and
+velocities are invaluable for understanding the orbital history of
+satellites.
 
-Some of the most fundamental properties of astronomical objects are
-their position and velocity. Unfortunately, determining positions to
-stars is nontrivial, . Additionally, while line-of-sight or radial
-velocities (RVs) are easily determined from spectroscopy, the tangental
-velocities, perpendicular to RV, are only measurable through proper
-motions, typically requiring precise astrometry as well. \emph{Gaia}'s
-mission is to produce extremely precise astrometry enabling measurements
-of unprecedented accuracy and scale for proper motions, parallaxes, and
-magnitudes.
-
-\emph{Gaia} was designed to revolutionize proper motion and parallax
-measurements. \emph{Gaia} is a space-based, all-sky survey telescope
-with two primary 1.3x0.75m mirrors situated at the Sun-Earth L2 lagrange
-point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
-completing its mission in 2025 (but with two more data releases
-planned). By imagining two patches of sky on the same focal plane,
-separated by a fixed angle of 106.5degrees, Gaia is able to measure
-absolute proper motions by comparing the apparent shifts of stars in
-different regions of the sky. In addition to precise astrometric
-information, \emph{Gaia} measures the magnitude of stars in the very
-wide \emph{G} band (330-1050nm), blue and red colours using the blue and
-red photometers (BP and RP), and takes low resolution BP-RP spectra and
-radial velocity measurements of bright stars (magnitudes \textgreater{}
-16?).
-
-\emph{Gaia} has revolutionized many astronomical disciplines, the least
-of which is local group and Milky Way science. While proper motions of a
-dwarf galaxies has been measured in a case by case basis by the Hubble
-Space Telescope, a full systematic determination of proper motions for
-most dwarf galaxies was unavailable until \emph{Gaia}
-\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
-the detection of many substructures of the halo, streams,
-
-While \emph{Gaia} has performed excellently, a few limitations and
-systematic uncertainties could become problematic. Gaia does show
-systematic variations in its derived proper motions and parallaxes, as
-measured by background quasars \citep{lindegren+2021}. These errors
-limit the ability to determine absolute proper motions to some degree,
-but since we would like to measure if a star is \emph{consistent} with a
-dwarf galaxy, these only factor into the uncertainty budget for orbital
-motion and are included in J+24. Additionally, while \emph{Gaia} was
-designed to be complete down to about a magnitude of \(G=20\),
-overcrowding or dense fields can limit this. \emph{Gaia} can only
-successfully separate stars at a distance of about \textbf{X} (a few
-times the pixel length of \ldots; pixel width is in scanning direction).
-For the galaxies Ursa Minor, Fornax, and Sculptor, the highest density
-(in FOrnax), is only \textbf{xxx} which appears to not significantly
-affect \emph{Gaia}'s survey completeness. We test for brightness affects
-in \textbf{REF}.
-
-For our purposes, \emph{Gaia}
-
-\citet[hereafter J+24]{jensen+2024} utilize \emph{Gaia} data with a
-Bayesian framework to determine samples of high-probability members for
-many Milky Way group dwarf galaxies.
-
-Here, we breifly describe J+24's membership estimation and discuss how
-this informs our observational knoledge of each galaxies density
-profile. In general, J+24 use a Bayesian framework incorporating proper
-motion (PM), colour-magnitude diagram (CMD), and spatial information to
-determine the probability that a given star belongs to the satellite.
-J+24 extends \citet{MV2020a}; \citet{MV2020b}. \citep[See also similar
-work by][]{pace+li2019, battaglia+2022, pace+erkal+li2022, qi+2022}
-
-\subsection{Summary of properties}\label{summary-of-properties}
+High-quality large spectroscopic samples combined with \emph{Gaia} has
+opened new windows into understanding the internal kinematics and
+history of dwarf galaxies. \textless Discuss battaglia, pace, etc. and
+conclusions about dwarf galaxies.\textgreater{}
 
@@
 \centering
-\caption[Observed Properties of Sculptor]{\label{tbl:scl_obs_props}Observed properties of Sculptor.
+\caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
 References are: 1.
@@
 $\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $111.18 \pm 0.23\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
-$\sigma_v$ & $9.71\pm0.17\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
-$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
-$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & XREF\\
+radial velocity & $111.3 \pm 0.2\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
+$\sigma_v$ & $9.64\pm0.16\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
+$R_h$ & $9.79 \pm 0.04$ arcmin & 4\\
+$R_{h,inner}$ & $9.46 \pm 0.26$ arcmin & XREF\\
 ellipticity & $0.36 \pm 0.01$ & 1\\
@@
 \centering
-\caption[Observed Properties of Ursa Minor]{\label{tbl:umi_obs_props}Observed properties of Ursa Minor.
+\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor.
 References are: (1) Ricardo R.
@@
 $\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
+radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
 $\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
-$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
+$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
+$R_{h, \rm inner}$ & $11.7\pm0.5$ & \\
 ellipticity & $0.55 \pm 0.01$ & 1\\
@@
 
+\subsection{\texorpdfstring{The \emph{Gaia}
+mission}{The Gaia mission}}\label{the-gaia-mission}
+
+Some of the most fundamental properties of astronomical objects are
+their position and velocity. Unfortunately, determining distances to
+stars is nontrivial. Additionally, while line-of-sight or radial
+velocities (RVs) are easily determined from spectroscopy, the tangental
+velocities, perpendicular to RV, are only measurable through proper
+motions, typically requiring precise astrometry as well. \emph{Gaia}'s
+mission is to produce extremely precise astrometry enabling measurements
+of unprecedented accuracy and scale for proper motions, parallaxes, and
+magnitudes.
+
+\emph{Gaia} was designed to revolutionize proper motion and parallax
+measurements. \emph{Gaia} is a space-based, all-sky survey telescope
+with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 lagrange
+point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
+completing its mission in 2025 (but with two more data releases
+planned). By imagining two patches of sky on the same focal plane,
+separated by a fixed angle of 106.5degrees, \emph{Gaia} is able to
+measure absolute proper motions by comparing the apparent shifts of
+stars in different regions of the sky. In addition to precise
+astrometric information, \emph{Gaia} measures the magnitude of stars in
+the very wide \emph{G} band (330-1050nm), blue and red colours using the
+blue and red photometers (BP and RP, 330-680, 640-1050 respectively),
+and takes low resolution BP-RP spectra and radial velocity measurements
+of bright stars (magnitudes \textless16?).
+
+\emph{Gaia} has revolutionized many astronomical disciplines, the least
+of which is local group and Milky Way science. While proper motions of a
+dwarf galaxies has been measured in a case by case basis by the Hubble
+Space Telescope, a full systematic determination of proper motions for
+most dwarf galaxies was unavailable until \emph{Gaia}
+\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
+the detection of many substructures of the halo, streams,
+
 \section{Gaia Membership Selection}\label{gaia-membership-selection}
 
-\subsection{Data}\label{data}
+Here, we briefly describe J+24's membership estimation method. J+24 use
+a Bayesian framework incorporating proper motion (PM), colour-magnitude
+diagram (CMD), and spatial information to determine the probability that
+a given star belongs to the satellite. By accounting for PM in
+particular, J+24 produces low contamination samples of candidate member
+stars. J+24 extends the algorithm presented in \citet{MV2020a};
+\citet{MV2020b} but additionally includes an optional secondary,
+extended spatial component to find possible members as far as
+\textasciitilde10 half-light radii \(R_h\) from some dwarf galaxies. See
+also similar work by \citet{pace+li2019}; \citet{battaglia+2022};
+\citet{pace+erkal+li2022}; \citet{qi+2022}.
 
-J+24 select stars initially from Gaia within a 2--4 degree circular
-region centred on the dwarf. J+24 only consider stars with:
+To create a high-quality sample, J+24 select stars initially from Gaia
+within a 2--4 degree circular region centred on the dwarf satisfying:
 
@@
 \item
-  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), so astrometry
-  is high quality.
+  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
+  quality astrometry. \texttt{ruwe} is a measure of the excess
+  astrometric noise on fitting a consistent parallax-proper motion
+  solution \citep[see][]{lindegrenXXX}.
 \item
-  3\(\sigma\) consistency of measured parallax with dwarf distance
-  (typically near zero; with \citet{lindegren+2021} zero-point
+  3\(\sigma\) consistency of measured parallax with dwarf's distance
+  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
   correction).
 \item
-  Absolute RA and Dec proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\),
-  less than 10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental
-  velocities of \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
+  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
+  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
+  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
 \item
@@
 \item
-  \(G\) magnitude is between \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\)
-  (dereddened magnitudes). Removes very faint stars and stars
-  significantly brighter than the tip of the red giant branch (TRGB)
-  magnitude plus the distance modulus uncertainty \(\delta {\rm DM}\).
+  De-reddened \(G\) magnitude is between
+  \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\). Removes very faint stars
+  and stars significantly brighter than the tip of the red giant branch
+  (TRGB) magnitude plus the distance modulus uncertainty
+  \(\delta {\rm DM}\).
 \item
@@
 
-Photometry is dereddened using \citet{schlegel+finkbeiner+davis1998}
+Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
 extinction maps.
 
-J+24 calculate the probability that any star belongs to either the
-satellite or the MW background with \begin{equation}
-P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
-\end{equation}
-
-where \(f_{\rm sat}\) is the fraction of candidate members in the field,
-and \({\cal L}_{\rm sat}\) and \({\cal L}_{\rm bg}\) are the satellite
-(sat) and background (bg) likelihoods respectively. The likelihoods are
-the product of a spatial, proper motion, and CMD term: \begin{equation}
+J+24 define likelihoods \({\cal L}\) representing the probability
+density that a star is consistent with either the MW stellar background
+(\({\cal L}_{\rm bg}\)) or the satellite galaxy
+(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
+product of a spatial, PM, and CMD term: \begin{equation}
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
@@
 
-Each likelihood map is normalized over the respective 2D parameter
-space. \(f_{\rm sat}\) scales the likelihoods by the abundance of stars
-in the satellite compared to the background.
+Each likelihoods is normalized over their respective 2D parameter space
+for both the satellite. Since the likelihoods are normalized,
+\(f_{\rm sat}\), representing the fraction of member stars in the field,
+controls the relative scaling between the satellite and background
+likelihoods. The total likelihood for any star in this model is then
+\begin{equation}\protect\phantomsection\label{eq:Ltot}{
+{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}
+}\end{equation} The probability that any star belongs to the satellite
+is then given by \begin{equation}
+P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
+\end{equation}
 
-J+24 consider both a one-component and a two-component spatial
-likelihood. The one component model is constructed as a single
-exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with
-\(R_s\) fixed to the value in the table in \citet{MV2020a}.
+For the satellite's spatial likelihood, J+24 consider both one-component
+and a two-component cases. The one component model is constructed as a
+single exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)),
+with \(R_s\) fixed to the value in the table in \citet{MV2020a}.
 Additionally, structural uncertainties (for position angle, ellipticity,
@@
 second component \(R_{\rm outer}\), \(B\) are solved for. Structural
-properties are not accounted for. The PM likelihood is a bivariate
-gaussian with variance and covariance equal to each star's proper
-motions. (The stellar PM uncertainties is assumed and typically is the
-dominant PM uncertainty.)
+properties are not accounted for.
+
+The PM likelihood is a bivariate gaussian with variance and covariance
+equal to each star's proper motions. J+24 assume the stellar PM errors
+are the main source of uncertainty.
 
@@
 Gaia colour uncertainty at each magnitude. The HB is modelled as a
-constant magnitude extending blue of the CMD (mean magnitude of
-isochone's HB stars and a 0.1 mag width plus the mean colour error). A
+constant magnitude extending blue of the CMD (mean magnitude of -2.2, 12
+Gyr HB stars and a 0.1 mag width plus the mean colour error). A
 likelihood map is constructed by sampling the distance modulus in
@@
 
-In J+24, a MCMC simulation is ran using the above total likelihood to
-solve \(\mu_{\alpha*}\), \(\mu_\delta\), and \(f_{\rm sat}\). For the
-two component models, the amplitude and scale radius of the excented
-compopnet is also solved for. The proper motion single component prior
-is same as \citet{MV2020a}: a normal distribution with mean 0 and
-standard deviation 100 km/s. If 2-component spatial, instead is a
-uniform distribution spanning 5\(\sigma\) of single component case w/
-systematic uncertainties. Priors on \(f_{\rm sat}\) and \(B\) (if 2
-component) are uniform from 0--1. \(R_{\rm outer}\) has a uniform prior
-only restricting \(R_{\rm outer} > R_s\). The mode of each parameter
-from the MCMC are then used to calculate the final \(P_{\rm sat}\)
-values we use here.
+J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
+\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
+with likelihood from Eq.~\ref{eq:Ltot}. Priors are weakly informative or
+uniform. The proper motion single component prior is same as
+\citet{MV2020a}: a normal distribution with mean 0 and standard
+deviation \(100\,\kms\). If 2-component spatial, instead is a uniform
+distribution spanning 5\(\sigma\) of single component case w/ systematic
+uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
+\(R_{\rm outer}\) has a uniform prior only restricting
+\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
+then reported and used to calculate the final \(P_{\rm sat}\) values.
 
@@
 resulting sample with RV measurements is very high (\textasciitilde90\%,
-J+24). Note there is likely a systematic bias in using stars with RV
+J+24). Note there is likely a systematic bias in using stars with RV
 measurements to measure purity. Fainter stars are less likely to have
@@
 
-In Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, we
-illustrate the resulting samples from the algorithm in the tangent plane
-(\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and proper
-motion space. For both galaxies, each criteria plays a commensurate role
-in sifting out nonmembers. Proper motions are clustered around the dwarf
-systemic motion, the CMD is well defined including the horizontal
-branch, and stars only within a few \(R_h\) are included. We also mark
-stars with consistent radial velocities (RVs) (see below), which trace
-each feature albeit more cautiously. We additionally mark the furthest
-confirmed RV members of each galaxy, illustrating the large extent of
-each galaxy past \(R_h\).
+In
+Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
+we illustrate the resulting samples from the algorithm in the tangent
+plane (\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and
+proper motion space. For all galaxies, each criteria plays a
+commensurate role in sifting out nonmembers. The CMD is well defined and
+probable members only extend a few times the colour uncertainty from the
+CMD. In proper motion space, selected stars are within \(\masyr\) from
+the systematic proper motion. The stars furthest away in proper motion
+space typically have large uncertainties \(\sim 1\masyr\), so most
+members are reasonably consistent. In any case, the algorithm does
+enable removing many stars which do not have consistent proper motions
+with the dwarf. Finally, the spatial likelihood reduces the
+probabilities of stars distant from the dwarfs centres. Member stars are
+rare outside of \(\sim5R_h\), resulting from both the likelihood
+specification and the lower density of background stars consistent with
+the CMD and PM of the satellite.
 
-Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection} also
-illustrates the distribution of stars selected without a spatial
-criterion. We define the CMD+PM selection as stars satisfying
-\begin{equation}
+One limitation of the J+24 method is that it assumes a spatial form of
+the dwarf galaxy, possibly limiting the detection of very distant stars
+or features. Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}
+also illustrates the distribution of stars selected without a spatial
+criterion. We define the \emph{CMD+PM} selection as
+\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
 {\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
-\end{equation} These stars are distributed similar to the fiducial
-(probable members) sample but with a fainter uniform distribution across
-the entire field. This illustrates the approximate background of stars
-which may be confused as members. Additionally, since there is no clear
-spatial structure in the CMD+PM sample outside several \(R_h\), it is
+}\end{equation} with the likelihoods from J+24 as described above. These
+stars are distributed similar to the fiducial (probable members) sample
+but with a fainter uniform distribution across the entire field. This
+illustrates the approximate background of stars which may be confused as
+members. Additionally, since there is no clear spatial structure in the
+\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
 unlikely that there are additional faint, tidal features detectable with
-Gaia. Not shown here, we also try a variety of simpler, absolute cuts
-and thresholds, finding no extended structure beyond what is detected in
-J+24. (\emph{Is it worth using this to calculate an OOM upper limit on
-the density of tidal tails?}).
+\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
+cuts and thresholds, finding no evidence of other structure in both
+\emph{Gaia} and DELVE/UNIONS data.
+
+Finally, we illustrate the location of RV-confirmed members from
+sec.~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
+brighter than a typical \emph{Gaia} candidate, RV members typically have
+more precise PMs. However, the RV members fill about the same area on
+the CMD down to the magnitude limit. Additionally, while the number of
+stars with RV measurements decreases at large radii, there are still
+confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
+stars illustrate the extended profiles of each galaxy. Note that for an
+exponential profile, 99.95\% of stars should be within \(6R_h\), so the
+discovery of any stars beyond \(6R_h\) hints at a deviation from an
+exponential profile. These \emph{extratidal} stars hint at something
+tidal in nature, and our goal is to explore possible interpretations.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
-\caption[Sculptor selection criteria]{The selection criteria for Scl
-members. Field stars (satesfying quality criteria) are light grey,
-CMD+PM selected stars are blue, probable members (2-component) are
-orange, and RV confirmed members are red. \textbf{Top:} Tangent plane
-\(\xi, \eta\). We outline in star symbols the two stars from
-\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram in
-Gaia G versus BP - RP. \textbf{Bottom right:} Proper motion in
-declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
-(corrected).}\label{fig:sculptor_selection}
+\caption[Sculptor sample selection]{The selection criteria for Scl
+members. Field stars (satisfying quality criteria) are light grey,
+CMD+PM selected stars are turqoise (eq.~\ref{eq:sel_cmd_pm}, probable
+members (2-component) are blue diamonds, and RV confirmed members are
+indigo crosses. We mark the two stars from \citet{sestito+2023a} with
+orange-outlined indigo stars. \textbf{Top:} Tangent plane \(\xi, \eta\).
+The ellipse represents 3 half-light radii. \textbf{Bottom left:} Colour
+magnitude diagram in Gaia G versus BP - RP. We plot the Padova 12Gyr
+MH=-1.68 isochone in orange. The black bar in the top left represents
+the median colour error. \textbf{Bottom right:} Proper motion in
+declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
+orange circle represents the \citet{MV2020b} proper motion. The black
+cross represents the median proper motion
+error.}\label{fig:sculptor_selection}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
-\caption[Ursa Minor Selection]{Similar to
+\caption[Ursa Minor sample selection]{Similar to
 Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
-members outside of \(3R_h\) in black stars (from \citet{sestito+2023b},
+members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
 \citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.pdf}}
+\caption[Fornax sample selection]{Similar to
+Fig.~\ref{fig:sculptor_selection} except for
+Fornax.}\label{fig:fornax_selection}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods.pdf}}
-\caption[Sculptor density profiles]{The density profile of Sculptor for
-different selection criteria, ploted as log surface dencity versus log
-elliptical radius. Residuals are with respect to th interpolated
-\emph{probable members} density. \emph{Probable members} selects stars
-with PSAT \textgreater{} 0.2 considering PM, CMD, and spatial,
-\emph{CMD+PM} select stars more likely to be members according to CMD
-and PM only, \emph{all} selects any high quality star, and \emph{BG
-subtracted} is the background-subtracted density derived from
-high-quality stars. We mark the half-light radius ( vertical dashed
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
+\caption[Sculptor density profiles]{The density profile of Sculptor,
+Ursa Minor, and Fornax for different selection criteria, ploted as log
+surface dencity versus log elliptical radius. Residuals are with respect
+to th interpolated \emph{probable members} density. \emph{Probable
+members} selects stars with PSAT \textgreater{} 0.2 considering PM, CMD,
+and spatial, \emph{CMD+PM} select stars more likely to be members
+according to CMD and PM only, \emph{all} selects any high quality star,
+and \emph{BG subtracted} is the background-subtracted density derived
+from high-quality stars. We mark the half-light radius ( vertical dashed
 line) and the break radius (black arrow,
 REF).}\label{fig:scl_observed_profiles}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods.pdf}}
-\caption[Ursa Minor density profiles]{Similar to
-fig.~\ref{fig:scl_observed_profiles} except for Ursa
-Minor.}\label{fig:umi_observed_profiles}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_density_methods.pdf}}
-\caption[Fornax density profiles]{Similar to
-fig.~\ref{fig:scl_observed_profiles} except for
-Fornax.}\label{fig:fornax_observed_profiles}
 \end{figure}
@@
 
-\section{Radial velocity modeling}\label{radial-velocity-modeling}
+While \emph{Gaia} has shown excellent performance, some notable
+limitations may introduce problems in our interpretation and reliability
+of density profiles.
+
+Gaia systematics in proper motions and parallaxes are typically smaller
+than the values for sources of magnitudes \(G\in[18,20]\). Since we use
+proper motions and parallaxes as general consistency with the dwarf, and
+factor in systematic uncertainties in each case, these effects should
+not be too significant. However, the systematic proper motion
+uncertainties becomes the dominant source of uncertainty in the derived
+systemic proper motions of each galaxy (see
+tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
+
+\emph{Gaia} shows high but imperfect completeness, particularly showing
+limitations in crowded fields and for faint sources (\(G\gtrapprox20\)).
+As discussed in \citet{fabricius+2021}, fir the high stellar densities
+in globular clusters, the completeness relative to HST varies
+significantly with the stellar density. However, the typical stellar
+densities of dwarf galaxies are much lower, at about 20 stars/arcmin =
+90,000 stars / degree, lower than the lowest globular cluster densities
+and safely below the crowding limit of 750,000 objects/degree for BP/RP
+photometry. , where the completeness down to \(G\approx 20\) is
+\(\sim 80\%\). Closely separated stars pose problems for Gaia's on-board
+processing, as the pixel size is 59x177 mas on the sky. This results in
+a reduction of stars separated by less than 1.5'' and especially for
+stars separated by less than 0.6 arcseconds. The astrometric parameters
+of closely separated stars furthermore tends to be of lower quality
+\citep{fabricius+2021}. However, even for the denser field of Fornax,
+only about 3\% of stars have a neighbour within 2 arc seconds, so
+multiplicity should not affect completeness too much (except for
+unresolved binaries). One potential issue is that the previous analysis
+do not account for our cuts on quality and number of astrometric
+parameters. These could worsen completeness, particularly since the
+BP-RP spectra are more sensitive to dense fields. In \textbf{REF}, we
+test if magnitude cuts impact the resulting density profiles, finding
+that this is likely not an issue.
+
+\section{Radial velocity modeling}\label{sec:rv_obs}
 
@@
 
-We derive a systemic velocity for Sculptor of
-\(111.18\pm0.23\,\kms\)with velocity dispersion \(9.71\pm0.17\,\kms\).
-Our values are very consistent with previous work
-\citep[e.g.][\citet{arroyo-polonio+2024},
+We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
+velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
+consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
 \citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
@@
 
-We detect a moderately significant gradient of \(4.3\pm1.4\,\kmsdeg\) at
+We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
 a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
@@
 
-We derive a mean \(-245.9\pm0.3_{\rm stat}\,\kms\) and velocity
-dispersion of \(8.76\pm0.24\,\kms\) for UMi. This is consistent with
+We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
+dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
 \citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
@@
 \end{figure}
+
+\section{Appendix}\label{appendix}
+
+\subsection{Density profile tests}\label{density-profile-tests}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_density_methods_extra.pdf}}
+\caption[Density profiles]{Density profiles for various assumptions for
+Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
+2-component bayesian model assuming circular radii, simple is the series
+of simple cuts described, bright is the sample of the brightest half of
+stars (scaled by 2), DELVE is a sample of RGB stars (background
+subtracted and rescaled to match). \textbf{TODO} plot exponential fit
+and add better labels}\label{fig:sculptor_observed_profiles}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
+\caption[UMi Density profiles]{Similar to
+fig.~\ref{fig:sculptor_observed_profiles} except for Ursa
+Minor}\label{fig:umi_observed_profiles}
+\end{figure}
+
+Note that a full rigorous statistical analysis would require a
+simulation study of injecting dwarfs into Gaia and assessing the
+reliability of various methods of membership and density profiles. This
+is beyond the scope of this thesis.
+
+\begin{verbatim}
+SELECT TOP 1000
+       *
+FROM delve_dr2.objects
+WHERE 11 < ra
+and ra < 19
+and -37.7 < dec
+and dec < -29.7
+\end{verbatim}
+
+\section{Velocity modelling and
+comparisons}\label{velocity-modelling-and-comparisons}
+
+Here, we describe in additional detail, our methods and comparisons for
+RV modelling between studies.
+
+Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
+smoothed samples from posterior/prior.
+
+\begin{table*}[t]
+\centering
+\caption{Summary of velocity measurements and derived properties.}
+\begin{tabular}{lllllllll}
+\toprule
+ & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
+\midrule
+Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
+ & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
+ & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
+ & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
+ & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
+UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
+ & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
+ & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
+ & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
+ & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Sculptor RV fits]{MCMC fits for different RV datasets for Sculptor among 3 different models. }
+\label{tbl:scl_rv_mcmc}
+\begin{tabular}{lllllll}
+\toprule
+study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log B_2/B_1$\\
+\midrule
+all &  &  &  &  &  & \\
+ & $111.3\pm0.2$ & $9.64\pm0.16$ & - & - & - & 0\\
+ & $111.3 \pm 0.2$ & $9.61 \pm0.16$ & - & $4.3\pm1.3$ & $-147_{-13}^{+17}$ & -1.9\\
+ & $111.2\pm0.2$ & $9.66\pm0.16$ & $0.07\pm0.02$ & - & - & -3.4\\
+tolstoy+23 &  &  &  &  &  & \\
+ & $111.3 \pm 0.3$ & $9.79 \pm 0.18$ & - & - & - & 0\\
+ & $111.3\pm0.3$ & $9.77\pm0.19$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ & -1.3\\
+ & $111.2 \pm 0.3$ & $9.73\pm0.19$ & $0.085 \pm 0.023$ & – & – & -4.6\\
+walker+09 &  &  &  &  &  & \\
+ & $111.1\pm0.3$ & $9.5\pm0.2$ & – & – & – & 0\\
+ & $111.1\pm0.3$ & $9.5\pm0.2$ & - & $5.2_{-1.6}^{+1.7}$ & $-134_{-16}^{+22}$ & -1.9\\
+ & $111.1\pm0.3$ & $9.6\pm0.2$ & $0.06\pm0.03$ & – & – & +0.3\\
+apogee &  &  &  &  &  & \\
+ & $111.2\pm0.7$ & $8.6\pm0.5$ & – & – & – & –\\
+ & $111.2\pm0.7$ & $8.5\pm0.5$ & – & $6\pm3$ & $-126_{-33}^{+45}$ & +0.1\\
+ & $111.1\pm0.7$ & $8.5\pm0.5$ & $0.07\pm0.06$ & – & – & +0.9\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Ursa Minor RV fits]{MCMC fits for UMi velocity dispersion. }
+\label{tbl:umi_rv_mcmc}
+\begin{tabular}{lllll}
+\toprule
+study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
+\midrule
+all & $-245.8\pm0.3$ & $8.8\pm0.2$ & +1.3 & +0.9\\
+pace & $-244.6\pm0.4$ & $9.0\pm0.3$ & +0.3 & +0.5\\
+spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.06\\
+apogee & $-245.6\pm1.2$ & $10.0_{-0.8}^{+1.0}$ & +1.0 & +0.5\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
+\caption[Scl velocity sample]{RV members of Sculptor plotted in the
+tangent plane coloured by corrected velocity difference from mean
+\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
+fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
+motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
+scale).}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
+\caption[Scl velocity gradient]{The corrected LOS velocity along the
+best fit rotational axis. RV members are black points, the systematic
+\(v_z\) is the horizontal grey line, blue lines represent the
+(projected) gradient from MCMC samples, and the orange line is a rolling
+median (with a window size of 50).}
+\end{figure}

### 2025-05-31 thesis/introduction.tex
thesis updates / reorganization of RV modeling

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}
+\caption[Picture of Sculptor]{Image of the Sculptor dwarf spheroidal
+galaxy from Dark Energy Survey Data Release 2 {[}\citet{abbott+2021};
+image created with HiPS2FITS{]}. Sculptor appears as a fairly prominent,
+extended over density of predominantly faint, red stars. 0.5 degree
+field of view centred on Sculptor. \textbf{TODO: combine with UMi, maybe
+add LMC and Fornax?}}\label{fig:scl_image}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}
+\caption[Picture of Ursa Minor]{Image of Ursa Minor dwarf galaxy from
+the Digitized Sky Survey 2 (0.75 deg field of view tangent plane). UMi
+appears as a diagonal/elliptical haze of faint, reddish stars from the
+top left to the bottom right. Even as classical dwarf, Ursa minor is
+fairly diffuse and does not stand obviously out from the
+background.}\label{fig:umi_image}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/mw_satellites_ongaia.pdf}}
+\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
+on the sky. Dwarf galaxies are taken as confirmed dwarfs with MW or LMC
+hosts from the \citet{pace2024} catalogue (version 1.0.3. We label all
+classical dwarfs and the Magellanic clouds, changing symbols based on
+the relevance to this work (classical dwarfs, Magellanic clouds REF, and
+the systems in question or comparison). \textbf{TODO: use
+abbreviations}}
+\end{figure}
+
 \subsection{Cosmological context}\label{cosmological-context}
@@
 approximantly the NFW formula.
-
-\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}Figure:
-Image of the Sculptor dwarf spheroidal galaxy from Dark Energy Survey
-Data Release 2 {[}\citet{abbott+2021}; image created with HiPS2FITS{]}.
-Sculptor appears as a fairly prominent, extended over density of
-predominantly faint, red stars. 0.5 degree field of view centred on
-Sculptor.
-
-\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}Figure:
-Image of Ursa Minor dwarf galaxy from the Digitized Sky Survey 2 (0.75
-deg field of view tangent plane). UMi appears as a diagonal/elliptical
-haze of faint, reddish stars from the top left to the bottom right. Even
-as classical dwarf, Ursa minor is fairly diffuse and does not stand
-obviously out from the background.
 

### 2025-05-31 thesis/methods.tex
thesis updates / reorganization of RV modeling

@@
 
-The 2D exponential stellar profile is given by \begin{equation}
+Perhaps the simplest in form, the 2D exponential is commonly used to
+describe dwarf galaxy's density profiles as well as stellar disks. The
+stellar profile is given by \begin{equation}
 \Sigma_\star(R) = A e^{-R / R_s}
-\end{equation} where \(A\) is some normalization and \(R_s\) is the
+\end{equation} where \(A\) is the normalization and \(R_s\) is the
 exponential scale radius. The 3D deprojected profile is found through
-abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
+Abel inversion \citep[e.g.][]{errani+2024}: \begin{equation}
 \rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
@@
 
-A Plummer profile is defined by \begin{equation}
+The Plummer density profile is generally considered to be a reasonable
+empirical fit to Globular Clusters, where it was first proposed
+\citep{sadfjk}. Plummer profiles are also sometimes used to describe
+dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
+profile is defined by \begin{equation}
 \Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
 \end{equation} where \(M\) is the total mass and \(R_s\) is the
-characteristic scale radius. The density is \begin{equation}
+characteristic scale radius. The 3D density is \begin{equation}
 \rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
@@
 \end{itemize}
+
+\section{Appendix: Numerical convergence and
+parameters}\label{appendix-numerical-convergence-and-parameters}
+
+Here, we describe some convergence tests to ensure our methods and
+results are minimally impacted by numerical limitations and assumptions.
+See \citet{power+2003} for a detailed discussion of various assumptions
+and parameters used in N-body simulations.
+
+\subsubsection{Softening}\label{softening}
+
+\citet{power+2003} suggest the empirical rule that the ideal softening
+(balancing integration time and only compromising resolution in
+collisional regime) is \begin{equation}
+h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
+\end{equation}
+
+For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
+particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
+slightly smaller softening which was reduced by a factor of
+\(\sqrt{10}\) which appears to improve agreement slightly in the
+innermost regions. However, this choice likely unnecessarily increases
+computation time for the relative gain in accuracy.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_softening.png}}
+\caption{Softening convergence}\label{fig:softening_convergence}
+\end{figure}
+
+\subsubsection{Timestepping and force
+accuracy}\label{timestepping-and-force-accuracy}
+
+In general, we use adaptive timestepping and relative opening criteria
+for gravitational force computations. To verify that these choices and
+associated accuracy parameters minimally impact convergence or speed, we
+show a few more isolation runs (using only 1e5 particles)
+
+\begin{itemize}
+\tightlist
+\item
+  constant timestep (\ldots), approximantly half of minimum timestep
+  with adaptive timestepping
+\item
+  geometric opening, with \(\theta = 0.5\).
+\item
+  strict integration accuracy, (facc = \ldots.)
+\end{itemize}
+
+\subsubsection{Alternative methods}\label{alternative-methods}
+
+\begin{itemize}
+\tightlist
+\item
+  FMM
+\item
+  PMM-tree
+\item
+  Gadget2
+\item
+  etc.
+\end{itemize}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_methods.png}}
+\caption{Isolation method convergence}\label{fig:methods_convergence}
+\end{figure}
+
+\subsubsection{Fiducial Parameters}\label{fiducial-parameters}
+
+Note that we use code units which assume that \(G=1\) for convenience
+and numerical stability. The conversion between code units to physical
+units is (for our convention):
+
+\begin{itemize}
+\tightlist
+\item
+  1 length = 1 kpc
+\item
+  1 mass unit = \(10^{10}\) Msun
+\item
+  1 velocity unit = 207.4 km/s
+\item
+  1 time unit = 4.715 Myr
+\end{itemize}
+
+Most parameters below are not too relevant or have been discussed or are
+merely dealing with cpu and IO details. The changes between simulation
+runs primarily affect the integration time, output frequency, and
+softening. Otherwise, we leave all other parameters fixed.
+
+\begin{verbatim}
+#======IO parameters======
+
+#---Filenames
+InitCondFile                initial
+OutputDir                   ./out
+SnapshotFileBase            snapshot
+OutputListFilename          outputs.txt
+
+#---File formats 
+ICFormat                    3       # use HDF5
+SnapFormat                  3
+
+#---Mem & CPU limits
+TimeLimitCPU                86400
+CpuTimeBetRestartFile       7200
+MaxMemSize                  2400
+
+#---Time
+TimeBegin                   0
+TimeMax                    2120      # 10 Gyr
+
+#---Output frequency
+OutputListOn                0
+TimeBetSnapshot             10
+TimeOfFirstSnapshot         0 
+TimeBetStatistics           10
+NumFilesPerSnapshot         1
+MaxFilesWithConcurrentIO    1
+
+#=======Gravity======
+
+#---Timestep accuracy
+ErrTolIntAccuracy           0.01
+CourantFac                  0.1     # ignored; for SPH
+MaxSizeTimestep             0.5
+MinSizeTimestep             0.0
+
+#---Tree algorithm
+TypeOfOpeningCriterion      1       # Relative
+ErrTolTheta                 0.5     # mostly used for Barnes-Hut
+ErrTolThetaMax              1.0     # (used only for relative)
+ErrTolForceAcc              0.005   # (used only for relative)
+
+#---Domain decomposition: should only affect performance
+TopNodeFactor                       3.0
+ActivePartFracForNewDomainDecomp    0.02
+
+#---Gravitational Softening
+SofteningComovingClass0      0.044  # HALO dependent
+SofteningMaxPhysClass0       0      # ignored; for cosmological
+SofteningClassOfPartType0    0
+SofteningClassOfPartType1    0
+
+#=======Miscellanius=======
+
+# probably do not need to change the options below
+
+#---Unit System
+UnitLength_in_cm            1 
+UnitMass_in_g               1
+UnitVelocity_in_cm_per_s    1 
+GravityConstantInternal     1
+
+#---Cosmological Parameters 
+ComovingIntegrationOn      0 # no cosmology
+Omega0                     0
+OmegaLambda                 0 
+OmegaBaryon                 0
+HubbleParam                 1
+Hubble                      100
+BoxSize                     0
+
+#---SPH
+ArtBulkViscConst             0.8
+MinEgySpec                   0
+InitGasTemp                  100
+
+#---Initial density estimate (SPH)
+DesNumNgb                   64
+MaxNumNgbDeviation          1
+\end{verbatim}

### 2025-05-31 thesis/table.py

@@ -3,97 +3,108 @@ import re
 
 from panflute import *
 
+
+def get_text(item):
+    if isinstance(item, Str):
+        return item.text.replace('_', r'\_')
+    elif isinstance(item, Plain):
+        return ''.join([get_text(i) for i in item.content])
+    elif isinstance(item, ListContainer):
+        return ''.join([get_text(i) for i in item])
+    elif isinstance(item, Code):
+        return '\\verb|' + (item.text) + '|'
+    elif isinstance(item, Math):
+        return '$' + (item.text) + '$'
+    elif isinstance(item, Quoted):
+        # Handle quoted text with proper quote characters
+        quote = '"' if item.quote_type == 'DoubleQuote' else "'"
+        content = ''.join(get_text(c) for c in item.content)
+        return f'{quote}{content}{quote}'
+    elif isinstance(item, RawInline):
+        return item.text
+    elif isinstance(item, Cite):
+        return ''.join(stringify(elem) for elem in item.content)
+    elif str(item) == 'Space':
+        return ' '
+    else:
+        return str(item)
+
+
+
+def table_caption(elem):
+    if elem.caption and elem.caption.content:
+        raw_caption = get_text(elem.caption.content)
+    else:
+        return ""
+
+    # Extract components using regex
+    label = None
+    short = None
+    caption_text = raw_caption
+    labels = re.findall(r'\\label\{([^}]+)\}', raw_caption)
+
+    # Extract short caption (e.g., short="Short caption")
+    if elem.caption.short_caption:
+        short = stringify(elem.caption.short_caption)
+
+    if len(labels) > 1:
+        debug("multiple labels")
+
+    if len(labels):
+        label = labels[0]
+        # should only replace once if only 1 label
+        caption_text = caption_text.replace(f"\\label{{{label}}}", "")
+    elif elem.identifier:
+        debug("using ID instead of label")
+        label = get_text(elem.identifier)
+    else:
+        debug("no label found")
+
+    # Build LaTeX components
+    parts = []
+    if short:
+        parts.append(f'\\caption[{short}]{{{caption_text}}}')
+    else:
+        parts.append(f'\\caption{{{caption_text}}}')
+    
+    if label:
+        parts.append(f'\\label{{{label}}}')
+
+    return '\n'.join(parts) + '\n'
+   
+    return '\n'.join(parts) + '\n'
+    return ''
+
+
+
+def table_headers(elem):
+    if elem.head and elem.head.content:
[…]

### 2025-05-31 thesis/table_shortcaption.py

@@ -79,31 +79,31 @@ def process_table(element, doc):
     if isinstance(element, pf.Table) and element.caption:
         caption = element.caption
         if caption.content:
-            pf.debug("table caption: ")
+            #pf.debug("table caption: ")
             # Get first paragraph of caption
             if isinstance(caption.content[0], pf.Plain):
                 inlines = caption.content[0].content
             
-                pf.debug(caption)
+                #pf.debug(caption)
                 # Extract attributes from inlines
                 id_str, attrs, new_inlines = extract_attributes(inlines)
-                pf.debug("attributes", attrs)
-                pf.debug("new inlines", pf.stringify(new_inlines))
+                #pf.debug("attributes", attrs)
+                #pf.debug("new inlines", pf.stringify(new_inlines))
 
                 if attrs:
                     # Set short caption
                     if 'short' in attrs.keys():
                         short_text = attrs['short']
-                        pf.debug("short caption", short_text)
+                        #pf.debug("short caption", short_text)
                     
                         element.caption = pf.Caption(pf.Plain(*new_inlines),
                                    short_caption=[pf.Str(short_text)],
                                                      )
                         element.identifier = id_str
-                        pf.debug(id_str)
-                        pf.debug("final : ", (element.caption))
-                        pf.debug("id : ", (element.identifier))
-                pf.debug()
+                        #pf.debug(id_str)
+                        #pf.debug("final : ", (element.caption))
+                        #pf.debug("id : ", (element.identifier))
+                #pf.debug()
 
     return element
 

### 2025-06-26 thesis/data.tex
thesis w full ch2 draft

@@
 To provide an empirical benchmark for simulations in the next chapters,
-we aim to understand the observed properties of Sculptor and Ursa Minor.
-In this section, we first compile past measurements of each galaxy. We
-describe the \citet{jensen+2024} Bayesian methodology to select high
-probability member stars from \emph{Gaia}. The derived density profiles
-are robust to changes in assumptions. Fornax is well represented by a 2D
-exponential profile. Instead, Sculptor and Ursa Minor show an excess of
-stars relative to an 2D exponential in the outer regions, better
-described my a Plummer profile. Finally, we combine various radial
-velocity samples with \citet{jensen+2024}'s members to derive systemic
-velocities and dispersions for Sculptor and Ursa Minor.
+we aim to understand the observed properties and density profiles of
+Sculptor and Ursa Minor. In this section, we first compile past
+measurements of each galaxy. We then describe the \citet{jensen+2024}
+Bayesian methodology to select high probability member stars from
+\emph{Gaia}. We show the derived density profiles are robust to changes
+in assumptions. While Fornax is well represented by a 2D exponential
+profile, Sculptor and Ursa Minor show an excess of stars relative to an
+2D exponential in the outer regions, better described by a Plummer
+profile. Future chapters assess possible explanations for these excess.
 
@@
 Since the discovery of dwarf galaxies around the Milky Way,
-observational work has attempted to measure and refine the basic
-properties of these objects. While the Milky Way's satellites are close
-(by extragalactic standards), their low numbers of stars and large areas
-on the sky present challenges for observational work. The classical
-systems discussed here each extend about 1-2 degrees across the sky
-(figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}).
+observational work has attempted to measure and refine their basic
+properties. While the Milky Way's satellites are close (by extragalactic
+standards), their low numbers of stars and large apparent sizes present
+challenges for observational work. In particular, the properties of the
+faintest dwarf galaxy systems and especially \emph{ambiguous} systems
+remains a pressing challenge as a unique probe into the nature of the
+early universe and dark matter.
 
-In particular, \emph{Gaia} has provided a wealth of high quality data,
-including the distances to nearby stars and proper motions. This allows
-for some of the first measurements on the motions for many local group
-dwarf galaxies {[}e.g.; \citet{MV2020}{]}. Accurate distances and
-velocities are invaluable for understanding the orbital history of
-satellites.
+Among local dwarf galaxies, classical systems are perhaps the systems
+best studied and with the best derived properties.
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} present our
+adopted observed properties for Sculptor and Ursa Minor. While extending
+1-2 degrees across the sky (e.g.
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
+classical dwarfs larger number of bright (giant) stars allow for
+thousands of stars to be observed with deep photometry and spectroscopy
+(e.g. \citet{tolstoy+2023}, \citet{pace+2020}). As a result, the
+determination of fundamental properties such as the position, size,
+orientation, distance (from 100s of RRL stars, e.g. \citet{tran+2022};
+\citet{garofalo+2025}), proper motions (from Gaia, described in next
+section), line-of-sight (LOS) velocity, and velocity dispersion
+\(\sigma_v\) are relatively well constrained today and show excellent
+convergence among different works. However, ongoing research still is
+refining our understanding of the detailed structure of Milky Way
+satellites, hinting that these objects may be more complex than at first
+glance.
 
-High-quality large spectroscopic samples combined with \emph{Gaia} has
-opened new windows into understanding the internal kinematics and
-history of dwarf galaxies. \textless Discuss battaglia, pace, etc. and
-conclusions about dwarf galaxies.\textgreater{}
+A number of works have previously speculated that dwarf galaxies such as
+Sculptor and Ursa Minor may have signatures of tidal perturbations,
+however not without contention. While \citet{hodge1961} and
+\citet{demers+krautter+kunkel1980} were some of the earliest work to
+derive the density profile of Sculptor, \citet{innanen+papp1979} was
+perhaps the first to speculate that Sculptor may harbour a substantial
+population of ``extratidal stars'' (stars beyond the tidal radius),
+finding candidate members out to 180' in an elongated distribution from
+\citet{vanagt1978}`s catalogue of variable stars.\footnote{Interestingly,
+  \citet{innanen+papp1979} also speculate about Uranus's satellite
+  distribution, covering gravitational dynamics at two very different
+  scales.} Additionally, \citet{eskridge1988} showed a possible excess
+of stars in Sculptor relative to a \citet{king1962} and Exponential
+density profile beginning around 50', but suggest that this excess may
+not be unusual. Later work by \citet{IH1995}, \citet{walcher+2003}, and
+\citet{westfall+2006} also showed evidence of an extended component of
+Sculptor's density profile, interpreting these stars as either evidence
+of tidal debris or a dwarf galaxy halo. On the other hand,
+\citet{coleman+dacosta+bland-hawthorn2005} show that Sculptor is well
+described by a two component density profile, they additionally mention
+that it is unlikely many of Sculptor's stars (less than 2\%) are
+extratidal. While \citet{munoz+2018} is the most recent and deepest
+photometric study of Sculptor, they only cover an area of out to radii
+of about 30', so they are unable to study the possible extended
+components of Sculptor detected in past works. In addition to the
+presence of ``extratidal stars'', many studies note that Sculptor's
+stars appear to become more elliptical with radius, consistent with
+tidal effects (\citet{IH1995}; \citet{westfall+2006}).
 
-In tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} we present
-adopted observed properties for each galaxy.
+Ursa Minor was often noted not for a density excess but for unusual
+features. Starting with the first density profile from
+\citet{hodge1964}, Ursa Minor is known to be highly elliptical. Studies
+by \citet{olszewski+aaronson1985}, \citet{demers+1995}, \citet{IH1995},
+\citet{kleyna+1998}, and \citet{bellazzini+2002} note that Ursa Minor
+appears to contain substructure along its major axis, but without strong
+interpretation for the causes. As one interpretation \citet{kleyna+2003}
+suggest that this feature is a long-lived star cluster residing in a
+cored dark matter halo to be dynamically stable. One of the first direct
+suggestions of tidal features came from \citet{martinez-delgado+2001},
+who find that stars extend far beyond the nominal tidal radius for Ursa
+Minor in the direction of the galaxies elongation, interpreting this as
+evidence for tidal effects. \citet{palma+2003} corroborate many of these
+earlier findings, showing further evidence for Ursa Minor's peculiar
+morphology including S-shaped contours, a possible extended extratidal
+population of stars, and a failure for a King density profile to
+adequately capture the actual stellar distribution. Each of these
+characteristics appear to indicate strong tidal disruption. Using
+velocity observations, \citet{pace+2014} additionally show that there
+are multiple components in spatial-velocity information. However,
+\citet{munoz+2018}'s more modern photometry show a more regular
+distribution of stars. Irregardless, Ursa Minor has had enough work
+suggesting peculiarities that a deeper investigation into the
+possibility of tidal effects is worthwhile.
+
+However, everything discussed so far is before \emph{Gaia}. Thus, the
+knowledge of the orbits of these systems is largely unknown and most
+surveys could only filter out stars using photometry. As a result, the
+tidal radius could only be guessed based on either the density profile
+or current distance of the dwarf galaxy. With \emph{Gaia} (discussed
+more below), recent work has used Bayesian frameworks to derive
+systematic proper motions of many dwarf galaxies (e.g. \citet{MV2020a})
+and filter away foreground stars using proper motions and parallax to
+detect distant members, and study the 6D internal kinematics of these
+galaxies (e.g. \citet{tolstoy+2023}). Most relevant here,
+\citet{jensen+2024} present a bayesian algorithm to determine likely
+members in \emph{Gaia} (described below), and detect extended secondary
+components for 9 dwarf galaxies, including Sculptor and Ursa Minor.
+\citet{sestito+2023a} and \citet{sestito+2023b} followed up a few of the
+most distant stars detected in \citet{jensen+2024} confirming that
+members exist in each galaxy as far ass 9-12 \(R_h\) from the centres.
+In this chapter, we discuss the origin and reliability of these
+detections, confirming that the density profiles are indeed robust.
+These density profiles then provide the scientific motivation to explore
+the tidal interpretation in the next chapters.
+
+\emph{Maybe define \(R_{\rm ell}\)} \emph{and} \(\xi, \eta\)
+\emph{here}? \emph{We will discuss multiple populations in future
+chapters}
 
@@
 Muñoz et al.
-(2018), 2.
+(2018) Sérsic fits, 2.
 Tran et al.
 (2022), 3.
-McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a).
+McConnachie and Venn (2020b), 4.
+Arroyo-Polonio et al.
+(2024), 5.
+McConnachie and Venn (2020a).
 }
@@
 $\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $111.3 \pm 0.2\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
-$\sigma_v$ & $9.64\pm0.16\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
-$R_h$ & $9.79 \pm 0.04$ arcmin & 4\\
-$R_{h,inner}$ & $9.46 \pm 0.26$ arcmin & XREF\\
-ellipticity & $0.36 \pm 0.01$ & 1\\
-position angle & $92\pm1^\circ$ & 1\\
+LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
+$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & 4\\
+$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
+ellipticity & $0.37 \pm 0.01$ & 1\\
+position angle & $94\pm1^\circ$ & 1\\
 $M_V$ & $-10.82\pm0.14$ & 1\\
@@
 Muñoz et al.
-(2018), (2) Garofalo et al.
-(2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al.
-(2020) and Spencer et al.
-(2018).
+(2018) Sérsic fits, (2) Garofalo et al.
+(2025), (3) McConnachie and Venn (2020a), (4) Pace et al.
+(2020), average of MMT and Keck results.
 }
@@
 $\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
-$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
+LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
+$\sigma_v$ & $8.6 \pm 0.3$ & 4\\
 $R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
-$R_{h, \rm inner}$ & $11.7\pm0.5$ & \\
 ellipticity & $0.55 \pm 0.01$ & 1\\
@@
 Some of the most fundamental properties of astronomical objects are
-their position and velocity. Unfortunately, determining distances to
-stars is nontrivial. Additionally, while line-of-sight or radial
-velocities (RVs) are easily determined from spectroscopy, the tangental
-velocities, perpendicular to RV, are only measurable through proper
-motions, typically requiring precise astrometry as well. \emph{Gaia}'s
-mission is to produce extremely precise astrometry enabling measurements
-of unprecedented accuracy and scale for proper motions, parallaxes, and
-magnitudes.
+their positions and velocities. Unfortunately, determining distances to
+stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
+are easily determined from spectroscopy, the tangental velocities,
+perpendicular to the LOS, are best measured through proper motions,
+typically requiring accurate determinations of small changes in a star's
+position. \emph{Gaia}'s mission is to produce extremely precise
+astrometry enabling measurements of unprecedented accuracy and scale for
+proper motions, parallaxes, and magnitudes.
 
@@
 measurements. \emph{Gaia} is a space-based, all-sky survey telescope
-with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 lagrange
-point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
+with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
+point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
 completing its mission in 2025 (but with two more data releases
 planned). By imagining two patches of sky on the same focal plane,
-separated by a fixed angle of 106.5degrees, \emph{Gaia} is able to
-measure absolute proper motions by comparing the apparent shifts of
-stars in different regions of the sky. In addition to precise
-astrometric information, \emph{Gaia} measures the magnitude of stars in
-the very wide \emph{G} band (330-1050nm), blue and red colours using the
-blue and red photometers (BP and RP, 330-680, 640-1050 respectively),
-and takes low resolution BP-RP spectra and radial velocity measurements
-of bright stars (magnitudes \textless16?).
+separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
+measure absolute proper motions. Stars in different regions of the sky
+are affected by parallax differently, so by observing the relative
+positions of stars separated by large angles during multiple epochs over
+a year, an absolute all-sky reference frame can be derived
+\citep{gaiacollaboration+2016}. In addition to precise astrometric
+information, \emph{Gaia} measures the magnitude of stars in the very
+wide \emph{G} band (330-1050nm), blue and red colours using the blue and
+red photometers (BP and RP, 330-680, 640-1050 respectively), and takes
+low resolution BP-RP spectra and radial velocity measurements of bright
+stars (\emph{Gaia} radial velocity magnitudes \textless16)
+\citep{gaiacollaboration+2016}. For our purposes, most useful are
+\emph{Gaia}'s measurements of \(G\) magnitude,
+\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
+the proper motions in RA and declination \(\mu_{\alpha*}\) and
+\(\mu_\delta\) (where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\)
+corrects for projection effects).
 
@@
 \citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
-the detection of many substructures of the halo, streams,
+the detection and improved measurements of halo substructures
+(CITATIONS), streams (CITAITONS), and Milky Way structure and
+kinematics, among many other applications.
 
@@
 
-Here, we briefly describe J+24's membership estimation method. J+24 use
-a Bayesian framework incorporating proper motion (PM), colour-magnitude
-diagram (CMD), and spatial information to determine the probability that
-a given star belongs to the satellite. By accounting for PM in
-particular, J+24 produces low contamination samples of candidate member
-stars. J+24 extends the algorithm presented in \citet{MV2020a};
-\citet{MV2020b} but additionally includes an optional secondary,
-extended spatial component to find possible members as far as
-\textasciitilde10 half-light radii \(R_h\) from some dwarf galaxies. See
-also similar work by \citet{pace+li2019}; \citet{battaglia+2022};
-\citet{pace+erkal+li2022}; \citet{qi+2022}.
+\subsection{Quality cuts and membership
+model}\label{quality-cuts-and-membership-model}
+
+Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
+membership estimation method. J+24 use a Bayesian framework
+incorporating proper motion (PM), colour-magnitude diagram (CMD), and
+spatial information to determine the probability that a given star
+belongs to the satellite. By accounting for PM in particular, J+24
+produces low contamination samples of candidate member stars out to
+large distances from a dwarf galaxy. J+24 extends the algorithm
+presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
+an optional secondary, extended spatial component to find possible
+members as far as \textasciitilde10 half-light radii \(R_h\) from
+several dwarf galaxies. See also similar work by \citet{pace+li2019};
+\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.
 
@@
   astrometric noise on fitting a consistent parallax-proper motion
-  solution \citep[see][]{lindegrenXXX}.
+  solution \citep[see][]{lindegren+2021}.\\
 \item
@@
 \item
-  Corrected colour excess is within expectations:
-  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
-  from \citet{riello+2021}. Removes stars with unreliable photometry.
+  Corrected colour excess is within 3\(\sigma\) of the expected
+  distribution from \citet{riello+2021}. Removes stars with unreliable
+  photometry.
 \item
   De-reddened \(G\) magnitude is between
-  \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\). Removes very faint stars
+  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
   and stars significantly brighter than the tip of the red giant branch
   (TRGB) magnitude plus the distance modulus uncertainty
-  \(\delta {\rm DM}\).
+  \(\sigma_{\rm DM}\).
 \item
   Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
+  Removes stars substantially outside the expected CMD.
 \end{itemize}
@@
 (\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
-product of a spatial, PM, and CMD term: \begin{equation}
+product of a spatial, PM, and CMD term: \begin{equation}{
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
-\end{equation}
+}\end{equation}
 
-Each likelihoods is normalized over their respective 2D parameter space
-for both the satellite. Since the likelihoods are normalized,
-\(f_{\rm sat}\), representing the fraction of member stars in the field,
-controls the relative scaling between the satellite and background
-likelihoods. The total likelihood for any star in this model is then
+Each likelihood is normalized over their respective 2D parameter space
+for both the satellite. To control the relative frequency of member and
+background stars, \(f_{\rm sat}\) representing the fraction of member
+stars in the field. The total likelihood for any star in this model is
+the sum of the satellite and background likelihoods, weighted by their
+relative frequencies,
 \begin{equation}\protect\phantomsection\label{eq:Ltot}{
-{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}
+{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
 }\end{equation} The probability that any star belongs to the satellite
-is then given by \begin{equation}
-P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
-\end{equation}
+is then given by \begin{equation}{
+P_{\rm sat} = 
+\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
+= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
+}\end{equation}
 
 For the satellite's spatial likelihood, J+24 consider both one-component
-and a two-component cases. The one component model is constructed as a
-single exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)),
-with \(R_s\) fixed to the value in the table in \citet{MV2020a}.
-Additionally, structural uncertainties (for position angle, ellipticity,
-and scale radius) are sampled over to construct the final likelihood
-map. The two-component model instead adds a second exponential,
+and a two-component density models. The one component model is
+constructed as a single exponential profile ( surface density
+\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\)
+fixed to the value in table 1 of \citet{MV2020a} from \citet{munoz+2018}
+(for a Sérsic fit). Additionally, structural uncertainties (for position
+angle, ellipticity, and scale radius) are sampled over to construct the
+final likelihood map. The two-component model instead adds a second
+exponential,
 \(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
 scale radius is fixed, and the outer scale radius and magnitude of the
-second component \(R_{\rm outer}\), \(B\) are solved for. Structural
-properties are not accounted for.
+second component \(R_{\rm outer}\), \(B\) are free parameters.
+Structural property uncertainties are not included in the two-component
+model.
 
@@
 \citep{girardi+2002}. The isochrone has a matching metallicity and 12
-Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
-Gaia colour uncertainty at each magnitude. The HB is modelled as a
-constant magnitude extending blue of the CMD (mean magnitude of -2.2, 12
-Gyr HB stars and a 0.1 mag width plus the mean colour error). A
-likelihood map is constructed by sampling the distance modulus in
-addition to the CMD width, taking the maximum of RGB and HB likelihoods.
+Gyr age (except 2 Gyr is used for Fornax). The (gaussian) colour width
+is assumed to be 0.1 mag plus the Gaia colour uncertainty at each
+magnitude. The horizontal branch is modelled as a constant magnitude
+extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB stars and a
+0.1 mag width plus the mean colour error). A likelihood map is
+constructed by sampling the distance modulus in addition to the CMD
+width, taking the maximum of RGB and HB likelihoods.
 
@@
 \(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
-with likelihood from Eq.~\ref{eq:Ltot}. Priors are weakly informative or
-uniform. The proper motion single component prior is same as
+with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
+informative. The proper motion single component prior is same as
 \citet{MV2020a}: a normal distribution with mean 0 and standard
-deviation \(100\,\kms\). If 2-component spatial, instead is a uniform
+deviation \(100\ \kms\). If 2-component spatial, instead is a uniform
 distribution spanning 5\(\sigma\) of single component case w/ systematic
@@
 
-We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
-sample. Most stars are assigned probabilities close to either 0 or 1, so
-the choice of probability threshold is only marginally significant.
-Additionally, even for a probability cut of 0.2, the purity of the
-resulting sample with RV measurements is very high (\textasciitilde90\%,
-J+24). Note there is likely a systematic bias in using stars with RV
-measurements to measure purity. Fainter stars are less likely to have
-been targeted and have poorer astrometry and colour information. We find
-no difference in the resulting density distributions when restricting
-stars to be brighter than a specific magnitude.
-
 \subsection{Selected samples}\label{selected-samples}
@@
 In
-Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
 we illustrate the resulting samples from the algorithm in the tangent
-plane (\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and
-proper motion space. For all galaxies, each criteria plays a
-commensurate role in sifting out nonmembers. The CMD is well defined and
-probable members only extend a few times the colour uncertainty from the
-CMD. In proper motion space, selected stars are within \(\masyr\) from
-the systematic proper motion. The stars furthest away in proper motion
-space typically have large uncertainties \(\sim 1\masyr\), so most
-members are reasonably consistent. In any case, the algorithm does
-enable removing many stars which do not have consistent proper motions
-with the dwarf. Finally, the spatial likelihood reduces the
-probabilities of stars distant from the dwarfs centres. Member stars are
-rare outside of \(\sim5R_h\), resulting from both the likelihood
-specification and the lower density of background stars consistent with
-the CMD and PM of the satellite.
+plane, CMD, and proper motion space. The tangent plane coordinates
+\(\xi\) and \(\eta\) are the distances in RA and declination as measured
+on a plane tangent to the dwarf galaxies centre.
+
+For our fiducial sample, we adopt a probability cut of
+\(P_{\rm sat} = 0.2\). Most stars are assigned probabilities close to
+either 0 or 1, so the choice of probability threshold is only marginally
+significant. Additionally, even for a probability cut of 0.2, the purity
+of the resulting sample with RV measurements is very high
+(\textasciitilde90\%, J+24). Note there is likely a systematic bias in
+using stars with RV measurements to measure purity. Fainter stars are
+less likely to have been targeted and have poorer astrometry and colour
+information. We find no difference in the resulting density
+distributions when restricting stars to be brighter than a specific
+magnitude.
+
+For all galaxies, each criteria is commensurate in sifting out
+nonmembers. The CMD is well defined, and probable members only extend a
+few times the colour uncertainty from the CMD. In proper motion space,
+selected stars are within \(\masyr\) from the systematic proper motion.
+The stars furthest away in proper motion space typically have large
+uncertainties \(\sim 1\masyr\), so most members are reasonably
+consistent. In any case, the algorithm removes stars with inconsistent
+PMs. Finally, the spatial likelihood reduces the probabilities of stars
+distant from the dwarfs centres. Member stars are rare outside of
+\(\sim5R_h\), resulting from both the likelihood specification and the
+lower density of background stars consistent with the CMD and PM of the
+satellite.
 
@@
 the dwarf galaxy, possibly limiting the detection of very distant stars
-or features. Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}
-also illustrates the distribution of stars selected without a spatial
+or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
+illustrates the distribution of stars selected without a spatial
 criterion. We define the \emph{CMD+PM} selection as
@@
 stars are distributed similar to the fiducial (probable members) sample
-but with a fainter uniform distribution across the entire field. This
+but with a background uniform distribution across the entire field. This
 illustrates the approximate background of stars which may be confused as
@@
 Finally, we illustrate the location of RV-confirmed members from
-sec.~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
+Section~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
 brighter than a typical \emph{Gaia} candidate, RV members typically have
@@
 \caption[Sculptor sample selection]{The selection criteria for Scl
-members. Field stars (satisfying quality criteria) are light grey,
-CMD+PM selected stars are turqoise (eq.~\ref{eq:sel_cmd_pm}, probable
-members (2-component) are blue diamonds, and RV confirmed members are
-indigo crosses. We mark the two stars from \citet{sestito+2023a} with
-orange-outlined indigo stars. \textbf{Top:} Tangent plane \(\xi, \eta\).
-The ellipse represents 3 half-light radii. \textbf{Bottom left:} Colour
-magnitude diagram in Gaia G versus BP - RP. We plot the Padova 12Gyr
-MH=-1.68 isochone in orange. The black bar in the top left represents
-the median colour error. \textbf{Bottom right:} Proper motion in
-declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
+members. Light grey points represent field stars (satisfying quality
+criteria), turquoise points CMD+PM selected stars
+(Eq.~\ref{eq:sel_cmd_pm}), blue xs probable members (2-component), and
+RV confirmed members indigo diamonds. We mark the two stars from
+\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
+Tangent plane \(\xi, \eta\). The orange ellipse represents 3 half-light
+radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\)
+magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot the Padova
+12Gyr {[}Fe/H{]}=-1.68 isochone in orange. The black bar in the top left
+represents the median colour error. \textbf{Bottom right:} Proper motion
+in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
 orange circle represents the \citet{MV2020b} proper motion. The black
 cross represents the median proper motion
-error.}\label{fig:sculptor_selection}
+error.}\label{fig:scl_selection}
 \end{figure}
@@
 \caption[Ursa Minor sample selection]{Similar to
-Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
+Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
 members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
@@
 \caption[Fornax sample selection]{Similar to
-Fig.~\ref{fig:sculptor_selection} except for
-Fornax.}\label{fig:fornax_selection}
+Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
+\citet{WMO2009}. While possibly a limitation of RV sample selection,
+Fornax does not show the same extended outer halo of probable members as
+Sculptor or Ursa Minor despite having many more
+stars.}\label{fig:fornax_selection}
 \end{figure}
@@
 simulations. To derive density profiles, we use 0.05 dex bins in log
-radius. We remove bins at smaller (larger) radii than the first bin to
-contain no stars, working outwards. We use symmetric poisson
-uncertainties. As discussed below, these uncertainties are
-straightforward but are likely under-representative.
+radius. We remove bins at smaller (larger) radii than the first empty
+bin working outwards. We use symmetric poisson uncertainties. As
+discussed below, these uncertainties are straightforward but are likely
+under-representative. We use the values from \citet{munoz+2018}'s Sérsic
+maximum likelihood fits for \(R_h\), which are more precisely derived
+given their deeper photometry.
 
-Figs.~\ref{fig:scl_observed_profiles}, \ref{fig:umi_observed_profiles}
-show the derived density profiles for Scl and UMi. We calculate density
-profiles for different selections of stars from above: all quality
-stars, CMD+PM only, and the probable member samples. In each case, all
-samples are the same towards the inner regions of the satellite.
-Classical dwarfs dominate the stellar density in their cores. However,
-the density profile \emph{all stars} plateaus at the total background in
-the field at radii of 30-60 arcminutes. By restricting stars to being
-most likely satellite members by CMD + PM, the background is reduced by
-1-2 dex. The CMD+PM background plateau likely represents the density of
-background stars which could be mistaken as members. Finally, the
-background (BG)-subtracted profile results from subtracting the apparent
-background in the \emph{all} profile.
+Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
+for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
+different selections of stars from above: all quality stars, CMD+PM
+only, and the probable member samples. In each case, all samples are the
+same towards the inner regions of the satellite. Classical dwarfs
+dominate the stellar density in their cores. However, the density
+profile \emph{all stars} plateaus at the total background in the field
+at radii of 30-60 arcminutes. By restricting stars to being most likely
+satellite members by CMD + PM, the background is reduced by 1-2 dex. The
+CMD+PM background plateau likely represents the density of background
+stars which could be mistaken as members. Finally, the background
+(BG)-subtracted profile results from subtracting the apparent background
+in the \emph{all} profile.
 
@@
 
+To illustrate the differences between each dwarf galaxy, in
+Fig.~\ref{fig:classical_dwarfs_densities}, we compare Sculptor, Ursa
+Minor, and other classical dwarfs against 2D-exponential and Plummer
+density profiles (REF). All dwarf galaxies appear to be well described
+by an exponential profile in the inner regions. However, while an
+exponential continues to be a good description for other dwarf galaxies,
+Sculptor and Ursa Minor diverge, showing a substantial excess over an
+exponential, better fit by a Plummer instead. In fact, Sculptor and Ursa
+Minor show excesses of between 2-3 orders of magnitude beyond what is
+expected of an exponential at the very outer regions!
+
 \begin{figure}
@@
 Ursa Minor, and Fornax for different selection criteria, ploted as log
-surface dencity versus log elliptical radius. Residuals are with respect
-to th interpolated \emph{probable members} density. \emph{Probable
-members} selects stars with PSAT \textgreater{} 0.2 considering PM, CMD,
-and spatial, \emph{CMD+PM} select stars more likely to be members
-according to CMD and PM only, \emph{all} selects any high quality star,
-and \emph{BG subtracted} is the background-subtracted density derived
-from high-quality stars. We mark the half-light radius ( vertical dashed
-line) and the break radius (black arrow,
-REF).}\label{fig:scl_observed_profiles}
+surface dencity versus log elliptical radius. \emph{CMD+PM} select stars
+more likely to be members according to CMD and PM only, \emph{all}
+selects any high quality star, and \emph{BG subtracted} is the
+background-subtracted density derived from \emph{all} stars. We mark the
+half-light radius (vertical dashed line) and the break radius (black
+arrow, REF).}\label{fig:scl_observed_profiles}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/classical_dwarf_profiles.pdf}}
+\caption[Classical dwarf density profiles]{The density profiles of
+Sculptor, Ursa Minor, and other classical dwarfs compared to Exp2D and
+Plummer density profiles. Dwarf galaxies are scaled to the same
+half-light radius and density at half-light radius. Sculptor and Ursa
+Minor have an excess of stars in the outer regions (past
+\(\log R/R_h \sim 0.3\)) compared with other classical
+dwarfs.}\label{fig:classical_dwarfs_densities}
 \end{figure}
@@
 described by a single or double exponential. If this model does not
-accurately match the actual density profile of the dwarf galaxy, we
-would like to understand how strongly influenced the density profile is
-by this assumption.
+accurately match the actual density profile of the dwarf galaxy, we want
+to understand the impact of this assumption.
 
-In particular, in Fig.~\ref{fig:umi_observed_profiles}, notice that the
-PSAT method produces small errorbars, even when the density is \(>1\)dex
-below the local background. These stars are likely selecting stars from
-the statistical MW background consistent with UMi PM / CMD, recovering
-the assumed density profile. As a result, the reliability of these
-density profiles below the CMD+PM background may be questionable. A more
-robust analysis, removing this particular density assumption, would be
-required to more appropriately represent the knowledge of the density
-profile as the background begins to dominate.
+In particular, in Fig.~\ref{fig:scl_observed_profiles}, notice that the
+\(P_{\rm sat}\) selection method produces small errorbars, even when the
+density is more than 1 dex below the local background. These stars are
+likely selecting stars from the statistical MW background consistent
+with UMi PM / CMD, recovering the assumed density profile. As a result,
+the reliability of these density profiles below the CMD+PM background
+may be questionable. A more robust analysis, removing this particular
+density assumption, would be required to more appropriately represent
+the knowledge of the density profile as the background begins to
+dominate.
 
@@
 component case. We assume constant ellipticity and position angle. Dwarf
-galaxies, in reality, are not necessarily smooth and constant. They do
-test an alternative method using circular radii for the extended density
-component, and we find these density profiles are very similar. We also
-assume a constant ellipticity and position angle here.
+galaxies, in reality, are not necessarily smooth or have constant
+ellipticity. J+24 test an alternative method using circular radii for
+the extended density component, and we find these density profiles are
+very similar to the fully elliptical case, even when assuming circular
+bins for the circular outer component. As such, even reducing the
+assumed ellipticity from \(0.37-0.55\) to 0 does not substantially
+impact the density profiles.
 
-Finally, our approximation to the poisson uncertainties Additionally, a
-more self consistent model would fit the density profile to the entire
-field at once, eliminating possible misrepresentation of the
-uncertainties.
+Finally, our uncertainties are likely underrepresented. A more self
+consistent model would fit the density profile to the entire field at
+once, eliminating possible misrepresentation of the uncertainties. In
+the Appendix to this section, we \emph{will} also discuss an alternative
+method which runs a MCMC model using the likelihoods above to solve for
+the density in each elliptical bin. From these tests, we note that the
+density profile and uncertainties derived from the J+24 sample are
+reliable insofar as the dwarf density is above the background from MW
+CMD+PM-consistent interlopers. We estimate that this effect comprimises
+the density profiles past \(\log R / {\rm arcmin} = 1.8\) for Sculptor
+and Ursa Minor, but agreement is good before then.
 
@@
 limitations may introduce problems in our interpretation and reliability
-of density profiles.
-
-Gaia systematics in proper motions and parallaxes are typically smaller
-than the values for sources of magnitudes \(G\in[18,20]\). Since we use
-proper motions and parallaxes as general consistency with the dwarf, and
-factor in systematic uncertainties in each case, these effects should
-not be too significant. However, the systematic proper motion
-uncertainties becomes the dominant source of uncertainty in the derived
-systemic proper motions of each galaxy (see
-tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
+of density profiles. Gaia systematics in proper motions and parallaxes
+are typically smaller than the values for sources of magnitudes
+\(G\in[18,20]\). Since we use proper motions and parallaxes as general
+consistency with the dwarf, and factor in systematic uncertainties in
+each case, these effects should not be too significant. However, the
+systematic proper motion uncertainties becomes the dominant source of
+uncertainty in the derived systemic proper motions of each galaxy (see
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
 
@@
 limitations in crowded fields and for faint sources (\(G\gtrapprox20\)).
-As discussed in \citet{fabricius+2021}, fir the high stellar densities
+As discussed in \citet{fabricius+2021}, for the high stellar densities
 in globular clusters, the completeness relative to HST varies
@@
 and safely below the crowding limit of 750,000 objects/degree for BP/RP
-photometry. , where the completeness down to \(G\approx 20\) is
-\(\sim 80\%\). Closely separated stars pose problems for Gaia's on-board
-processing, as the pixel size is 59x177 mas on the sky. This results in
-a reduction of stars separated by less than 1.5'' and especially for
-stars separated by less than 0.6 arcseconds. The astrometric parameters
-of closely separated stars furthermore tends to be of lower quality
+photometry. In \citet{fabricius+2021}, for the lowest density globular
+clusters, the completeness down to \(G\approx 20\) is \(\sim 80\%\).
+Closely separated stars pose problems for Gaia's on-board processing, as
+the pixel size is 59x177 mas on the sky. This results in a reduction of
+stars separated by less than 1.5'' and especially for stars separated by
+less than 0.6 arcseconds. The astrometric parameters of closely
+separated stars furthermore tends to be of lower quality
 \citep{fabricius+2021}. However, even for the denser field of Fornax,
@@
 multiplicity should not affect completeness too much (except for
-unresolved binaries). One potential issue is that the previous analysis
+unresolved binaries). One potential issue is that the previous analyses
 do not account for our cuts on quality and number of astrometric
 parameters. These could worsen completeness, particularly since the
-BP-RP spectra are more sensitive to dense fields. In \textbf{REF}, we
-test if magnitude cuts impact the resulting density profiles, finding
-that this is likely not an issue.
+BP-RP spectra are more sensitive to dense fields. In the appendix to
+this section, we test if magnitude cuts impact the resulting density
+profiles, finding that this is likely not an issue.
 
-\section{Radial velocity modeling}\label{sec:rv_obs}
+Finally, an excellent test of systematics and methods in \emph{Gaia} is
+to compare against another survey. In the appendix, we show that
+profiles derived from J+24 agree with DELVE or UNIONS data within
+uncertainties. Systematics, completion effects, and selection methods
+are unlikely to substantially change the density profiles presented
+here.
 
-\subsection{Data selection}\label{data-selection}
+\section{Discussion and Conclusions}\label{discussion-and-conclusions}
 
-For both Sculptor and Ursa Minor, we construct literature samples of
-radial velocity measurements. We combine these samples with J+24's
-members to produce RV consistent stars and to compute velocity
-dispersion, systematic velocities, and test for the appearance of
-velocity gradients.
+In summary, we have used J+24 data to derive the density profiles for
+the classical dwarf galaxies. In each case, the density profile is
+robust against different selection criteria. Both Sculptor and Ursa
+Minor show strong evidence for deviations from an exponential profile.
+We find no evidence of additional (velocity or stellar) substructure in
+either galaxy. Our density profiles are consistent with many previous
+works, all indicating divergence from a King or Exponential density
+profile with a break between 30 and 60 arcminutes. Our goal in the
+following chapters is to test if tides provide a viable explanation for
+the observed properties of Sculptor and Ursa Minor.
 
-First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
-not report GaiaDR3 source ID's, we match to the nearest star within 1-3
-arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
-measurement from each study using the inverse-variance weighted mean
-\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
-variance \(s^2\). We remove stars with significant velocity dispersions
-as measured between observations in a study or between studies. By using
-that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
-\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
-distribution with \(N-1\) measurements. This cut typically removes stars
-with reduced chi-squared values
-\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
-number of measurements is 1-3 typically).
+Are dwarf galaxies indeed expected to be one-component exponential-like
+density profiles? The suggestion of exponential density profiles dates
+back to \citet{faber+lin1983}. Expand\ldots{}
 
-Next, we need to correct the coordinate frames for the solar motion and
-on-sky size of the galaxy. We transform the frame into the galactic
-standard of rest (GSR). The next step is to account for the slight
-differences in the direction of each radial velocity. Let the \(\hat z\)
-be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
-the angular distance between the centre of the galaxy and the individual
-star, the corrected radial velocity is then \begin{equation}
-v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
-\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
-in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
-velocities in RA and Dec, and \(\theta\) is the position angle of the
-star with respect to the centre of the dwarf. The correction from both
-effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
-Sculptor and less for Ursa Minor \citep[see
-also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
-the distance uncertainty and velocity dispersion in quadrature to the RV
-uncertainties for each star. We then use the \(v_z\) values for the
-following modelling, however repeating with uncorrected, heliocentric
-velocities does not significantly affect the results.
+\section{Additional density profile
+tests}\label{additional-density-profile-tests}
 
-The combined likelihood, including RV information, becomes
-\begin{equation}
-{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
-\end{equation} where we assume that the satellite and background
-distributions are Gaussian. Specifically, \begin{equation}
-\begin{split}
-{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
-{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
-\end{split}
-\end{equation} where \(f\) is the probability density of a standard
-normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
-velocity and dispersion of the satellite, and \(\delta v_i\) is the
-individual measurement uncertainty. Typically, the velocity dispersion
-will dominate the uncertainty budget here. We assume a halo/background
-velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
-\citep[e.g.][]{brown+2010}.
+In this section, we discuss additional tests and verification of the
+derived density profiles. In particular, we check that methodology
+(simpler cuts, circularized radii, algorithm version) do not
+substantially affect the density profile. We also compile density
+profiles presented in the literature as reference. In all cases, the
+density profiles appear to have excellent convergence out to
+\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
+where the background dominates.
 
-Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2. Because of the additional information
-from radial velocities, most stars have probabilities close to 1 or 0 so
-the probability cut is not too significant.
-
-We assume priors on systematic velocity and velocity dispersion of
-\begin{equation}
-\begin{split}
-\mu_{v} &= N(0\,\kms, \sigma_{\rm halo}^2) \\ 
-\sigma_{v} &= U(0, 20\,\kms)
-\end{split}
-\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
-the velocity dispersion of the MW halo adopted above, a reasonable
-assumption for dwarfs in orbit around the MW.
-
-\subsection{Results}\label{sec:rv_results}
+Discuss selection criteria for DELVE and UNIONS samples, literature
+comparison, simple selection criteria, MCMC density profiles and when
+\citet{jensen+2024} becomes background-limited.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
-terms of \(v_z\) (REF). Orange points are from our crossmatched RV
-membership sample.}
-\end{figure}
-
-For Sculptor, we combine radial velocity measurements from APOGEE,
-\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
-\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
-measurements. We find that there is no significant velocity shift in
-crossmatched stars between catalogues. After crossmatching to high
-quality Gaia stars and excluding significant stellar velocity
-dispersions, we have a sample of 1918 members.
-
-We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
-velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
-consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
-detailed comparison between different samples and additional tests.
-
-We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
-a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
-Several past work has attempted to detect a gradient in Sculptor, but no
-consensus has been reached. \citet{arroyo-polonio+2024} detect a
-velocity gradient of \(4\pm1.5\,\kmsdeg\) in a similar direction using
-the \citet{tolstoy+2023} sample, finding inconclusive statistical
-evidence. They additionally suggest a third chemodynamical component of
-the galaxy which may bias rotation measurements. \citet{battaglia+2008}
-also detect a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along
-the major axis, approximately the same direction. Instead,
-\citet{strigari2010}; \citet{martinez-garcia+2023} detect no significant
-gradient in Sculptor using \citet{WMO2009} sample. Note that
-pre-\emph{Gaia} work did not have as strong of a constraint on the
-proper motion of Scl, which limits conclusions of the intrinsic velocity
-gradient in Scl.
-
-For UMi, we collect radial velocities from, APOGEE,
-\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
-shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
-\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
-183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
-dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
-the median difference in velocities in this crossmatch is about 2.2
-km/s, we adopt 1 km/s as the approximate systematic error here. Our
-final sample includes 831 members.
-
-We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
-dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
-\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
-do not find evidence for a velocity gradient, consistent with past work
-\citep{pace+2020, martinez-garcia+2023}.
-
-\subsection{Discussion and
-limitations}\label{discussion-and-limitations}
-
-Our model here is relatively simple. Some things which we note as
-systematics or limitations:
-
-\begin{itemize}
-\tightlist
-\item
-  Inter-study systematics and biases. While basic crossmatches and a
-  simple velocity shift, combining data from multiple instruments is
-  challenging. This appears to be a minor issue (Sculptor) or is
-  corrected for (Ursa Minor).
-\item
-  Misrepresentative uncertainties. Inspection of the variances compared
-  to the standard deviations within a study seems to imply that errors
-  are accurately reported. APOGEE notes that their RV uncertainties are
-  known to be underestimates but are a small proportion of our sample.
-\item
-  Binarity. While not too large of a change for classical dwarfs, this
-  could inflate velocity dispersions of about \(9\,\kms\) by about
-  \(1\,\kms\)\citep{spencer+2017}. Thus, our measurement is likely
-  slightly inflated given the high binarity fractions measured in these
-  systems \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
-\item
-  Multiple populations. Both Sculptor and Ursa Minor likely contain
-  multiple populations \citep[\citet{pace+2020},
-  \citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a
-  single population, and each population may have a different extent and
-  velocity dispersion, this could result in biased velocity dispersions.
-  However, it is unclear how to uniquely determine an overall velocity
-  dispersion in a multi-population system.
-\item
-  Selection effects. RV studies each have their own selection effects,
-  which may affect the resulting dispersion, especially if different
-  populations or regions of the galaxy have different velocities or
-  velocity dispersions. We do not attempt to correct for this.
-\end{itemize}
-
-For both Ursa Minor and Sculptor, we also fit models to only data from
-individual surveys (see REF). Since the resulting parameters are very
-similar, we conclude that many of the systematic uncertainties are
-likely smaller than the present errors or that each large survey has
-similar biases.
-
-\section{Comparison and conclusions}\label{comparison-and-conclusions}
-
-To illustrate the differences between each dwarf galaxy, in
-Fig.~\ref{fig:classical_dwarfs_densities}, we compare Sculptor, Ursa
-Minor, and Fornax against 2D-exponential and Plummer density profiles
-(REF). While all dwarfs appear similar in the inner regions, each dwarf
-diverges in the outer regions relative to an exponential. Relative to an
-exponential, Fornax is underdense but Sculptor and Ursa Minor are both
-overdense. A Plummer profile instead provides a more reasonable fit to
-Scl and UMi.
-
-In summary, we have used J+24 data to derive the density profiles for
-Fornax, Sculptor, and Ursa Minor. In each case, the density profile is
-robust against different selection criteria. Both Sculptor and Ursa
-Minor show strong evidence for deviations from an exponential profile.
-We also compile velocity measurements to derive the systemic motions and
-velocity dispersions of each galaxy. We find evidence for a velocity
-gradient in Sculptor of \(4.3\pm1.3\,\kmsdeg\). We find no evidence of
-additional (velocity or stellar) substructure in either galaxy. Our goal
-in the following chapters is to test if tides provide a viable
-explanation for the observed properties of Sculptor and Ursa Minor.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fornax_exp_fit.pdf}}
-\caption[Classical dwarf density profiles]{The density profiles of
-Sculptor, Ursa Minor, and Fornax compared to Exp2D and Plummer density
-profiles. Dwarf galaxies are scaled to the same half-light radius and
-density at half-light radius (fit from the inner 3 scale radii
-exponential recursively. ). Sculptor and Ursa Minor have an excess of
-stars in the outer regions (past \(\log R/R_h \sim 0.3\)) compared with
-Sculptor. Sculptor and Ursa Minor's density profiles are created with
-the 2-component J+24 model, but the excess does not change significantly
-for the 1-component model.}\label{fig:classical_dwarfs_densities}
-\end{figure}
-
-\section{Appendix}\label{appendix}
-
-\subsection{Density profile tests}\label{density-profile-tests}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_density_methods_extra.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
 \caption[Density profiles]{Density profiles for various assumptions for
@@
 stars (scaled by 2), DELVE is a sample of RGB stars (background
-subtracted and rescaled to match). \textbf{TODO} plot exponential fit
-and add better labels}\label{fig:sculptor_observed_profiles}
+subtracted and rescaled to match).}\label{fig:scl_density_extras}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_j24.pdf}}
+\caption[Density profiles]{Comparison of density profiles for each J+24
+method. The fiducial is a 2-component elliptical model. However, the
+1-component is still elliptical but only contains 1 component and the
+circular model assumes a circular outer density profile and bins in
+circular bins instead of elliptical
+bins.}\label{fig:scl_density_j24_methods}
 \end{figure}
@@
 \caption[UMi Density profiles]{Similar to
-fig.~\ref{fig:sculptor_observed_profiles} except for Ursa
-Minor}\label{fig:umi_observed_profiles}
-\end{figure}
-
-Note that a full rigorous statistical analysis would require a
-simulation study of injecting dwarfs into Gaia and assessing the
-reliability of various methods of membership and density profiles. This
-is beyond the scope of this thesis.
-
-\begin{verbatim}
-SELECT TOP 1000
-       *
-FROM delve_dr2.objects
-WHERE 11 < ra
-and ra < 19
-and -37.7 < dec
-and dec < -29.7
-\end{verbatim}
-
-\section{Velocity modelling and
-comparisons}\label{velocity-modelling-and-comparisons}
-
-Here, we describe in additional detail, our methods and comparisons for
-RV modelling between studies.
-
-Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
-smoothed samples from posterior/prior.
-
-\begin{table*}[t]
-\centering
-\caption{Summary of velocity measurements and derived properties.}
-\begin{tabular}{lllllllll}
-\toprule
- & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
-\midrule
-Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
- & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
- & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
- & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
- & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
-UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
- & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
- & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
- & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
- & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{table*}[t]
-\centering
-\caption[Sculptor RV fits]{MCMC fits for different RV datasets for Sculptor among 3 different models. }
-\label{tbl:scl_rv_mcmc}
-\begin{tabular}{lllllll}
-\toprule
-study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log B_2/B_1$\\
-\midrule
-all &  &  &  &  &  & \\
- & $111.3\pm0.2$ & $9.64\pm0.16$ & - & - & - & 0\\
- & $111.3 \pm 0.2$ & $9.61 \pm0.16$ & - & $4.3\pm1.3$ & $-147_{-13}^{+17}$ & -1.9\\
- & $111.2\pm0.2$ & $9.66\pm0.16$ & $0.07\pm0.02$ & - & - & -3.4\\
-tolstoy+23 &  &  &  &  &  & \\
- & $111.3 \pm 0.3$ & $9.79 \pm 0.18$ & - & - & - & 0\\
- & $111.3\pm0.3$ & $9.77\pm0.19$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ & -1.3\\
- & $111.2 \pm 0.3$ & $9.73\pm0.19$ & $0.085 \pm 0.023$ & – & – & -4.6\\
-walker+09 &  &  &  &  &  & \\
- & $111.1\pm0.3$ & $9.5\pm0.2$ & – & – & – & 0\\
- & $111.1\pm0.3$ & $9.5\pm0.2$ & - & $5.2_{-1.6}^{+1.7}$ & $-134_{-16}^{+22}$ & -1.9\\
- & $111.1\pm0.3$ & $9.6\pm0.2$ & $0.06\pm0.03$ & – & – & +0.3\\
-apogee &  &  &  &  &  & \\
- & $111.2\pm0.7$ & $8.6\pm0.5$ & – & – & – & –\\
- & $111.2\pm0.7$ & $8.5\pm0.5$ & – & $6\pm3$ & $-126_{-33}^{+45}$ & +0.1\\
- & $111.1\pm0.7$ & $8.5\pm0.5$ & $0.07\pm0.06$ & – & – & +0.9\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{table*}[t]
-\centering
-\caption[Ursa Minor RV fits]{MCMC fits for UMi velocity dispersion. }
-\label{tbl:umi_rv_mcmc}
-\begin{tabular}{lllll}
-\toprule
-study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
-\midrule
-all & $-245.8\pm0.3$ & $8.8\pm0.2$ & +1.3 & +0.9\\
-pace & $-244.6\pm0.4$ & $9.0\pm0.3$ & +0.3 & +0.5\\
-spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.06\\
-apogee & $-245.6\pm1.2$ & $10.0_{-0.8}^{+1.0}$ & +1.0 & +0.5\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
-\caption[Scl velocity sample]{RV members of Sculptor plotted in the
-tangent plane coloured by corrected velocity difference from mean
-\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
-fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
-motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
-scale).}
+Fig.~\ref{fig:scl_observed_profiles} except for Ursa
+Minor}\label{fig:umi_density_extras}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
-\caption[Scl velocity gradient]{The corrected LOS velocity along the
-best fit rotational axis. RV members are black points, the systematic
-\(v_z\) is the horizontal grey line, blue lines represent the
-(projected) gradient from MCMC samples, and the orange line is a rolling
-median (with a window size of 50).}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_j24.pdf}}
+\caption[UMi density methods]{Similar to
+Fig.~\ref{fig:scl_density_j24_methods} except for Ursa
+Minor.}\label{fig:umi_density_j24_methods}
 \end{figure}

### 2025-06-26 thesis/introduction.tex
thesis w full ch2 draft

@@
   \citet{IH1995} structural parameters
+\item
+  \citet{mateo1998}, \citet{simon2019} ARAA on ultrafaints.
+\item
+  \citet{walker+penarrubia2011}: measuring cores and cusps
+\item
+  \citet{battaglia+nipoti2022}: dynamics and dwarf galaxies review
+\item
+  \citet{dsouza+bell2022}, \citet{santistevan+2024}. challenges to
+  backwards time integration
 \end{itemize}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/mw_satellites_ongaia.pdf}}
+\includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_onsky_scl_umi_for_1.pdf}
 \caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
@@
 the relevance to this work (classical dwarfs, Magellanic clouds REF, and
-the systems in question or comparison). \textbf{TODO: use
-abbreviations}}
+the systems in question or comparison). The background image is from
+ESA/Gaia/DPAC
+(https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).}\label{fig:mw_satellite_system}
 \end{figure}
 
-\subsection{Cosmological context}\label{cosmological-context}
+\subsection{Motivation}\label{motivation}
+
+\begin{itemize}
+\tightlist
+\item
+  Sculptor and Ursa Minor are possibly unusual.
+\end{itemize}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_vs_penarrubia.pdf}}
+\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
+profiles are well-matched to \citet{PNM2008}.}\label{fig:toy_profiles}
+\end{figure}
+
+\section{Cosmological context}\label{cosmological-context}
 
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_vs_penarrubia.pdf}}
-\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
-profiles are well-matched to \citet{PNM2008}.}\label{fig:toy_profiles}
-\end{figure}
+Origin of NFW profiles
 
-\subsection{Theoretical Background}\label{theoretical-background}
+\section{Theoretical Background}\label{theoretical-background}
 
@@
 unbound is, as stated in \citet{PNM2008}, the break radius
-\begin{equation}
+\begin{equation}{
 R_b = C\,\sigma_{v}\,\Delta t
-\end{equation} where \(\sigma_v\) is the present line of sight velocity
+}\end{equation} where \(\sigma_v\) is the present line of sight velocity
 dispersion , \(\delta t\) is the time since pericentre, and
@@
 
-As illustrated in fig.~\ref{fig:toy_profiles}, the density profile
+As illustrated in Fig.~\ref{fig:toy_profiles}, the density profile
 initially stars off exponential. At increasing times since the first
@@
   Methods
+\end{itemize}
+
+\subsubsection{Background}\label{background}
+
+\subsubsection{The N-body method}\label{the-n-body-method}
+
+\begin{itemize}
+\tightlist
+\item
+  Nbody as a solution to the collisionless boltzman equation
+\item
+  limitations and interpretation (MC sampling of phase space)
+\item
+  Tree approximations and the BH method
+\end{itemize}
+
+\subsection{Tidal effects on dark matter
+halos}\label{tidal-effects-on-dark-matter-halos}
+
+\begin{itemize}
+\tightlist
+\item
+  EN2020 tidal tracks
+\item
+  peñarrubia2009/etc.
+\item
+  assumtions and limitations
 \end{itemize}
@@
 
-From the \citet{nfw1996} paper, eqns. 3 \& 4 \begin{equation}
+\subsubsection{NFW}\label{nfw}
+
+From \citet{nfw1996}, eqns. 3 \& 4 \begin{equation}{
 \frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
-\end{equation} where \begin{equation}
+}\end{equation} where \begin{equation}{
 \delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
-\end{equation} \(c\) is concentration parameter, \(\rho_c\) is the
+}\end{equation} \(c\) is concentration parameter, \(\rho_c\) is the
 critical density of the universe, and \(r_s\) is the characteristic
@@
 200 times the critical density of the universe, and \(M_{200}\) is the
-mass contained inside \(r_{200}\). As equations: \begin{equation}
+mass contained inside \(r_{200}\). As equations: \begin{equation}{
 \rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
-\end{equation}
+}\end{equation}
 
-\begin{equation}
+\begin{equation}{
 r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
 }}}
-\end{equation}
+}\end{equation}
 
-\begin{equation}
+\begin{equation}{
 M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
-\end{equation}
+}\end{equation}
 
 \(M_{200}\) is also sometimes called the virial mass of the halo.
-\(r_{200}\) is directly related to \(r_s\) by \begin{equation}
+\(r_{200}\) is directly related to \(r_s\) by \begin{equation}{
 r_{200} = c\,r_s
-\end{equation}
-
-Another useful definition is \begin{equation}
-A(x) \equiv \log (1+x) - \frac{x}{1+x}.
-\end{equation}
-
-We will also define a dimensionless radius \begin{equation}
-x \equiv r/r_s.
-\end{equation}
-
-A simple substitution to the definition gives
-
-\begin{equation}
-\rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
-\end{equation} where \begin{equation}
-\rho_s = \frac{c^3}{A(c)} \rho_{200}
-\end{equation} and \(A(c)\) is as above. The characteristic density can
-also be written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see
-below), giving \begin{equation}
-\rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
-\end{equation} Note that the NFW density profile is the same as an
-alpha-beta-gamma profile where \(\alpha=\gamma=1\) and \(\beta =3\).
-
-\subsubsection{Circular velocity}\label{circular-velocity}
+}\end{equation}
 
 The circular velocity in terms of
-\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \begin{equation}
+\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \begin{equation}{
 \left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
-\end{equation}
+}\end{equation}
 
-or in terms of \(M_s\) and \(r_s\), \begin{equation}
+or in terms of \(M_s\) and \(r_s\), \begin{equation}{
 v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
-\end{equation} Another parameterization of the NFW profile is in terms
+}\end{equation} Another parameterization of the NFW profile is in terms
 of the maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the
 radius at which it is reached \(r_{\rm circ}^{\rm max}\). Given the
-scale radius, \begin{equation}
+scale radius, \begin{equation}{
 r_{\rm circ}^{\rm max} = \alpha\ r_s
-\end{equation}
+}\end{equation}
 

### 2025-06-26 thesis/methods.tex
thesis w full ch2 draft

@@
 
-fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
+Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
 each component of our fiducially potential. We adopt the potential
@@
 
-\begin{equation}
+\begin{equation}{
 \Phi(r) = - \frac{GM}{r + a}
-\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
+}\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
 \(M=2.1 \times 10^{10}\,\Mo\) is the total mass.
@@
 The thin and thick disks are represented with the
-\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}
+\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}{
 \Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
-\end{equation}
+}\end{equation}
 
@@
 
-Finally, in chapter REF, we consider the influence of the large Milky
-Cloud on the orbits and evolution of Scl. We adopt the
-\citet{vasiliev2024} multipole approximation of an N-body simulation of
-the LMC and MW. Their initial conditions are
+Finally, we also consider the influence of the large Milky Cloud on the
+orbits and evolution of Scl. We adopt the \citet{vasiliev2024} multipole
+approximation of an N-body simulation of the LMC and MW. Their initial
+conditions are
 
@@
 the stars are merely collisionless tracers embedded in this potential
-(added on in post-processing). The density is exponentially truncated
-with a profile
+(added on in post-processing). The density is a cubic-exponentially
+truncated with a profile
 \begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
-\rho_{\rm tNFW} = e^{-r/r_t}\ \rho_{\rm NFW}(t)
-}\end{equation} where we adopt \(r_t = 100 r_s\) or approximately 10
-times \(r_{200}\).
+\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t)
+}\end{equation} where we adopt \(r_t = 20 r_s\) or approximately
+\(r_{200}\) for our Sculptor-like fiducial halo. (Using \(r_{200}\)
+depends on the chosen scale of the halo, we chose to be scale free
+here.) This truncation causes the density to drop quickly past
+\(20 r_s\) but retails \textgreater{} 90 \(\%\) agreement with an NFW
+within \(18r_s\). This choice is purely for numerical convenience,
+however it is unlikely that the outer density profile of loosely bound
+particles past \(20\) kpc affects the evolution of the halo. A more
+detailed cosmological consideration would be ideal.
 
@@
 relaxed and well-converged, we run the simulation in isolation (no
-external potential) for 5 Gyr (or approximately the crossing timescale).
-Our fiducial isolation halo uses \(r_s=2.76\) kpc and
-\(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for any
-length or mass scale.
+external potential) for 5 Gyr (or about 3 times the crossing timescale
+at the virial radius). Our fiducial isolation halo uses \(r_s=2.76\) kpc
+and \(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for
+any length or mass scale.
 
-We adopt a softening length of \begin{equation}
+We adopt a softening length of \begin{equation}{
 h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-\end{equation} See appendix REF for a discussion of this choice, which
+}\end{equation} See appendix REF for a discussion of this choice, which
 is similar to the \citet{power+2003} suggested softening.
@@
 by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise,
-\begin{equation}
+\begin{equation}{
 t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
 = {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
-\end{equation} This works out to be about 6-10 times (increasing with
+}\end{equation} This works out to be about 6-10 times (increasing with
 particle number) our adopted softening length for NFW halos given our
@@
 
-\begin{equation}
+\begin{equation}{
 v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
-\end{equation}
+}\end{equation}
 
@@
 energy), then we assign the stellar weight for a given particle with
-energy \({\cal E}\) is \begin{equation}
+energy \({\cal E}\) is \begin{equation}{
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-\end{equation} While \(f({\cal E})\) is a phase-space density, the
+}\end{equation} While \(f({\cal E})\) is a phase-space density, the
 differential energy distribution includes an additional \(g({\cal E})\)
@@
 We use Eddington inversion to find the distribution function, (eq.
-4-140b in BT87) \begin{equation}
+4-140b in BT87) \begin{equation}{
 f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-\end{equation}
+}\end{equation}
 
@@
 describe dwarf galaxy's density profiles as well as stellar disks. The
-stellar profile is given by \begin{equation}
+stellar profile is given by \begin{equation}{
 \Sigma_\star(R) = A e^{-R / R_s}
-\end{equation} where \(A\) is the normalization and \(R_s\) is the
+}\end{equation} where \(A\) is the normalization and \(R_s\) is the
 exponential scale radius. The 3D deprojected profile is found through
-Abel inversion \citep[e.g.][]{errani+2024}: \begin{equation}
+Abel inversion \citep[e.g.][]{errani+2024}: \begin{equation}{
 \rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
-\end{equation} where \(K_0\) is the 0th order modified Bessel function
+}\end{equation} where \(K_0\) is the 0th order modified Bessel function
 of the second type.
@@
 dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
-profile is defined by \begin{equation}
+profile is defined by \begin{equation}{
 \Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
-\end{equation} where \(M\) is the total mass and \(R_s\) is the
-characteristic scale radius. The 3D density is \begin{equation}
+}\end{equation} where \(M\) is the total mass and \(R_s\) is the
+characteristic scale radius. The 3D density is \begin{equation}{
 \rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
-\end{equation}
+}\end{equation}
 
@@
 (balancing integration time and only compromising resolution in
-collisional regime) is \begin{equation}
+collisional regime) is \begin{equation}{
 h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
-\end{equation}
+}\end{equation}
 

### 2025-06-26 thesis/sculptor.tex
thesis w full ch2 draft

@@
 
+\subsection{Derived properties}\label{derived-properties}
+
+\begin{table*}[t]
+\centering
+\begin{tabular}{lll}
+\toprule
+parameter & value & reference\\
+\midrule
+$L_\star$ & $1.82_{-0.22}^{+0.25}\times10^6\ L_\odot$ & \\
+$M_\star$ & $2.7_{-0.6}^{+0.7} \times10^6\ {\rm M}_\odot$ & \\
+$M_{200}$ & $0.48_{-0.25}^{+0.52}\ M_0$ & \\
+$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
+$v_{\rm circ, max}$ & $31\pm8\,\kms$ & \\
+$r_{\rm circ, max}$ & $5.9 \pm 2.9$ kpc & \\
+$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & \\
+$t_{\rm break, obs}$ & $110\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
 \section{Sculptor (MW only)}\label{sculptor-mw-only}
@@
 
-fig.~\ref{fig:unref} contains example orbits
+Fig.~\ref{fig:unref} contains example orbits
 
@@
 influenced by the presence of an LMC. Figure
-fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
+Fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
 Scl's pericentre with the MW and implies that Scl may be on its first

### 2025-06-26 presentations/pittsburgh25/arroyo-polonio_scale.jl

@@ -1,0 +1,2056 @@
+### A Pluto.jl notebook ###
+# v0.20.6
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ f115979c-3f07-11f0-2ff6-7ddc61640d3e
+using Arya, CairoMakie
+
+# ╔═╡ bdca1255-292a-434e-9be9-914cdc0fcc95
+R_h = 9.46 / sqrt(1 - 0.36) / 60
+
+# ╔═╡ 428559de-3172-438d-91ed-18a62f0212c5
+range_deg = (0.01786759961324086,  1.27)
+
+# ╔═╡ a0b04e51-fbb7-48a5-9ccd-06cca6241752
+range_R_h = range_deg ./ R_h
+
+# ╔═╡ a0c3397d-5b38-4489-9510-c57e5764429e
+let
+	fig = Figure(figure_padding=20)
+	ax1 = Axis(fig[1,1], limits=(range_deg, (0.0,1e-5)), 
+			   xticksmirrored=false, aspect=1e4,xaxisposition=:top, 
+			   xscale=log10, xticks=[0.1, 1.0], xtickalign=0.0, xminortickalign=0.0)
+
+	
+	ax2 = Axis(fig[1,1], limits=(log10.(range_R_h), (0,1e-5)), xticksmirrored=false, aspect=1e4, xlabel=L"R / R_h", xtickalign=0.0, xminortickalign=0.0)
+
+	
+	hidespines!(ax2)
+	hideydecorations!(ax1)
+	hideydecorations!(ax2)
+
+	resize_to_layout!(fig)
+	fig
+end
+
+# ╔═╡ 164f32b6-e6e6-4112-8cf5-0ccca1d49861
+let
+	fig = Figure(figure_padding=20)
+	ax1 = Axis(fig[1,1], limits=(log10.(range_deg .* 60), (0.0,1e-5)), 
+			   xticksmirrored=false, aspect=1e4,xaxisposition=:top, 
+			   xlabel="Radius / arcmin", xtickalign=0.0, xminortickalign=0.0)
+
+	
+	ax2 = Axis(fig[1,1], limits=(log10.(range_R_h), (0,1e-5)), xticksmirrored=false, aspect=1e4, xlabel=L"R / R_h", xtickalign=0.0, xminortickalign=0.0)
+
+	
+	hidespines!(ax2)
+	hideydecorations!(ax1)
+	hideydecorations!(ax2)
+
+	resize_to_layout!(fig)
[…]

### 2025-06-26 thesis/compile.sh

@@ -1,17 +1,17 @@
 set -e 
 rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
-pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections --filter equation.py"
+# pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections --filter equation.py"
+pandocoptions="--defaults=pandoc_params.yaml"
 
 pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
 pandoc $pandocoptions -o sculptor.tex sculptor.md
+
 # pandoc $pandocoptions -o ursa_minor.tex ursa_minor.md
 # pandoc $pandocoptions -o conclusion.tex conclusion.md
 
-
-pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=chapter --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections"
-pandoc $pandocoptions -o appendix.tex appendix.md
+# pandoc $pandocoptions -o appendix.tex appendix.md --top-level-division=chapter
 

### 2025-07-15 thesis/abstract.tex
major thesis restructure

Dwarf galaxies are unique cosmological probes, testing our understanding
of the smallest scales of structure formation in the Universe. Among the
classical dwarf galaxies, Sculptor and Ursa Minor appear unusual, with
more stars than expected in the outer regions compared with the typical
Exponential density profile, rememnescent of tidal interaction. I verify
the validity of this excess, checking \citet{jensen+2024}'s bayesian
membership catalogue for completeness, simple cuts, and other surveys.
To evaluate if tides indeed can explain these density features, I
conduct idealized N-body simulation of both galaxies in the tidal field
of the Milky Way. In both cases, the orbits are well constrained,
producing pericenters larger than expected to produce substantial tidal
impacts. In addition, the time since the last pericenter does not agree
with a tidal interpretation. However, Sculptor's orbit is strongly
influenced by the presence of an LMC. With an LMC, Sculptor's long term
orbital history is highly uncertain, however recent tidal effects due to
the Milky Way and LMC only minimally affect this galaxy. Because tides
likely do not explain the presence of these density features, I review
other theories about the formation of dwarf galaxy ``halos,'' including
mergers / accretion events and complex multi-component star formation
histories. The density profiles of Sculptor and Ursa Minor, with their
deviations from the classically expected exponential, present an
interesting puzzle, likely with a solution involving their formation and
history in the early universe.

### 2025-07-15 thesis/appendix.tex
major thesis restructure

--- 2025-04-15 thesis/numerical_convergence.tex
@@
-Here, we describe some convergence tests to ensure our methods and
-results are minimally impacted by numerical limitations and assumptions.
-See \citet{power+2003} for a detailed discussion of various assumptions
-and parameters used in N-body simulations.
+\chapter{Data selection}\label{data-selection}
 
-\subsubsection{Softening}\label{softening}
+\section{Describution of algorithm}\label{describution-of-algorithm}
 
-\citet{power+2003} suggest the empirical rule that the ideal softening
-(balancing integration time and only compromising resolution in
-collisional regime) is \[
-h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
-\]
-
-For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
-particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
-slightly smaller softening which was reduced by a factor of
-\(\sqrt{10}\) which appears to improve agreement slightly in the
-innermost regions. However, this choice likely unnecessarily increases
-computation time for the relative gain in accuracy.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_softening.png}}
-\caption{Softening convergence}\label{fig:softening_convergence}
-\end{figure}
-
-\subsubsection{Timestepping and force
-accuracy}\label{timestepping-and-force-accuracy}
-
-In general, we use adaptive timestepping and relative opening criteria
-for gravitational force computations. To verify that these choices and
-associated accuracy parameters minimally impact convergence or speed, we
-show a few more isolation runs (using only 1e5 particles)
+To create a high-quality sample, J+24 select stars initially from Gaia
+within a 2--4 degree circular region centred on the dwarf satisfying:
 
@@
 \item
-  constant timestep (\ldots), approximantly half of minimum timestep
-  with adaptive timestepping
+  Solved astrometry, magnitude, and colour.
 \item
-  geometric opening, with \(\theta = 0.5\).
+  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
+  quality astrometry. \texttt{ruwe} is a measure of the excess
+  astrometric noise on fitting a consistent parallax-proper motion
+  solution \citep[see][]{lindegren+2021}.\\
 \item
-  strict integration accuracy, (facc = \ldots.)
+  3\(\sigma\) consistency of measured parallax with dwarf's distance
+  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
+  correction).
+\item
+  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
+  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
+  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
+\item
+  Corrected colour excess is within 3\(\sigma\) of the expected
+  distribution from \citet{riello+2021}. Removes stars with unreliable
+  photometry.
+\item
+  De-reddened \(G\) magnitude is between
+  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
+  and stars significantly brighter than the tip of the red giant branch
+  (TRGB) magnitude plus the distance modulus uncertainty
+  \(\sigma_{\rm DM}\).
+\item
+  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
+  Removes stars substantially outside the expected CMD.
 \end{itemize}
 
-\subsubsection{Alternative methods}\label{alternative-methods}
+Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
+extinction maps.
 
-\begin{itemize}
-\tightlist
-\item
-  FMM
-\item
-  PMM-tree
-\item
-  Gadget2
-\item
-  etc.
-\end{itemize}
+J+24 define likelihoods \({\cal L}\) representing the probability
+density that a star is consistent with either the MW stellar background
+(\({\cal L}_{\rm bg}\)) or the satellite galaxy
+(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
+product of a spatial, PM, and CMD term: \begin{equation}{
+{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
+}\end{equation}
+
+Each likelihood is normalized over their respective 2D parameter space
+for both the satellite. To control the relative frequency of member and
+background stars, \(f_{\rm sat}\) representing the fraction of member
+stars in the field. The total likelihood for any star in this model is
+the sum of the satellite and background likelihoods, weighted by their
+relative frequencies, \begin{equation}{
+{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
+}\end{equation} The probability that any star belongs to the satellite
+is then given by \begin{equation}{
+P_{\rm sat} = 
+\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
+= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
+}\end{equation}
+
+For the satellite's spatial likelihood, J+24 consider both one-component
+and a two-component density models. The one component model is
+constructed as a single exponential profile ( surface density
+\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\)
+fixed to the value in table 1 of \citet{MV2020a} from \citet{munoz+2018}
+(for a Sérsic fit). Additionally, structural uncertainties (for position
+angle, ellipticity, and scale radius) are sampled over to construct the
+final likelihood map. The two-component model instead adds a second
+exponential,
+\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
+scale radius is fixed, and the outer scale radius and magnitude of the
+second component \(R_{\rm outer}\), \(B\) are free parameters.
+Structural property uncertainties are not included in the two-component
+model.
+
+The PM likelihood is a bivariate gaussian with variance and covariance
+equal to each star's proper motions. J+24 assume the stellar PM errors
+are the main source of uncertainty.
+
+The satellite's CMD likelihood is based on a Padova isochrone
+\citep{girardi+2002}. The isochrone has a matching metallicity and 12
+Gyr age (except 2 Gyr is used for Fornax). The (gaussian) colour width
+is assumed to be 0.1 mag plus the Gaia colour uncertainty at each
+magnitude. The horizontal branch is modelled as a constant magnitude
+extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB stars and a
+0.1 mag width plus the mean colour error). A likelihood map is
+constructed by sampling the distance modulus in addition to the CMD
+width, taking the maximum of RGB and HB likelihoods.
+
+The background likelihoods are instead empirically constructed. Stars
+stars outside of 5\(R_h\) passing the quality cuts estimate the
+background density in PM and CMD space. The density is a sum of
+bivariate gaussians with variances based on Gaia uncertainties (and
+covariance for proper motions). The spatial background likelihood is
+assumed to be constant.
+
+J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
+\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
+with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
+informative. The proper motion single component prior is same as
+\citet{MV2020a}: a normal distribution with mean 0 and standard
+deviation \(100\ \kms\). If 2-component spatial, instead is a uniform
+distribution spanning 5\(\sigma\) of single component case w/ systematic
+uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
+\(R_{\rm outer}\) has a uniform prior only restricting
+\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
+then reported and used to calculate the final \(P_{\rm sat}\) values.
+
+\section{Additional density tests}\label{additional-density-tests}
+
+In this section, we discuss additional tests and verification of the
+derived density profiles. In particular, we check that methodology
+(simpler cuts, circularized radii, algorithm version) do not
+substantially affect the density profile. We also compile density
+profiles presented in the literature as reference. In all cases, the
+density profiles appear to have excellent convergence out to
+\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
+where the background dominates.
+
+Discuss selection criteria for DELVE and UNIONS samples, literature
+comparison, simple selection criteria, MCMC density profiles and when
+\citet{jensen+2024} becomes background-limited.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_methods.png}}
-\caption{Isolation method convergence}\label{fig:methods_convergence}
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/scl_density_methods_extra.pdf}}
+\caption[Scl density comparison]{Density profiles for various
+assumptions for Sculptor. PSAT is our fiducial 2-component J+24 sample,
+circ is a 2-component bayesian model assuming circular radii, simple is
+the series of simple cuts described, bright is the sample of the
+brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
+(background subtracted and rescaled to
+match).}\label{fig:scl_density_extras}
 \end{figure}
 
-\subsubsection{Fiducial Parameters}\label{fiducial-parameters}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/scl_density_methods_j24.pdf}}
+\caption[Scl density methods]{Comparison of density profiles for each
+J+24 method. The fiducial is a 2-component elliptical model. However,
+the 1-component is still elliptical but only contains 1 component and
+the circular model assumes a circular outer density profile and bins in
+circular bins instead of elliptical
+bins.}\label{fig:scl_density_j24_methods}
+\end{figure}
 
-Note that we use code units which assume that \(G=1\) for convenience
-and numerical stability. The conversion between code units to physical
-units is (for our convention):
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/umi_density_methods_extra.pdf}}
+\caption[UMi density comparison]{Similar to
+Fig.~\ref{fig:scl_observed_profiles} except for Ursa
+Minor}\label{fig:umi_density_extras}
+\end{figure}
 
-\begin{itemize}
-\tightlist
-\item
-  1 length = 1 kpc
-\item
-  1 mass unit = \(10^{10}\) Msun
-\item
-  1 velocity unit = 207.4 km/s
-\item
-  1 time unit = 4.715 Myr
-\end{itemize}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/umi_density_methods_j24.pdf}}
+\caption[UMi density methods]{Similar to
+Fig.~\ref{fig:scl_density_j24_methods} except for Ursa
+Minor.}\label{fig:umi_density_j24_methods}
+\end{figure}
 
-Most parameters below are not too relevant or have been discussed or are
-merely dealing with cpu and IO details. The changes between simulation
-runs primarily affect the integration time, output frequency, and
-softening. Otherwise, we leave all other parameters fixed.
+\section{Comparison to Literature}\label{comparison-to-literature}
 
-\begin{verbatim}
-#======IO parameters======
-
-#---Filenames
-InitCondFile                initial
-OutputDir                   ./out
-SnapshotFileBase            snapshot
-OutputListFilename          outputs.txt
-
-#---File formats 
-ICFormat                    3       # use HDF5
-SnapFormat                  3
-
-#---Mem & CPU limits
-TimeLimitCPU                86400
-CpuTimeBetRestartFile       7200
-MaxMemSize                  2400
-
-#---Time
-TimeBegin                   0
-TimeMax                    2120      # 10 Gyr
-
-#---Output frequency
-OutputListOn                0
-TimeBetSnapshot             10
-TimeOfFirstSnapshot         0 
-TimeBetStatistics           10
-NumFilesPerSnapshot         1
-MaxFilesWithConcurrentIO    1
-
-#=======Gravity======
-
-#---Timestep accuracy
-ErrTolIntAccuracy           0.01
-CourantFac                  0.1     # ignored; for SPH
-MaxSizeTimestep             0.5
-MinSizeTimestep             0.0
-
-#---Tree algorithm
-TypeOfOpeningCriterion      1       # Relative
-ErrTolTheta                 0.5     # mostly used for Barnes-Hut
-ErrTolThetaMax              1.0     # (used only for relative)
-ErrTolForceAcc              0.005   # (used only for relative)
-
-#---Domain decomposition: should only affect performance
-TopNodeFactor                       3.0
-ActivePartFracForNewDomainDecomp    0.02
-
-#---Gravitational Softening
-SofteningComovingClass0      0.044  # HALO dependent
-SofteningMaxPhysClass0       0      # ignored; for cosmological
-SofteningClassOfPartType0    0
-SofteningClassOfPartType1    0
-
-#=======Miscellanius=======
-
-# probably do not need to change the options below
-
-#---Unit System
-UnitLength_in_cm            1 
-UnitMass_in_g               1
-UnitVelocity_in_cm_per_s    1 
-GravityConstantInternal     1
-
-#---Cosmological Parameters 
-ComovingIntegrationOn      0 # no cosmology
-Omega0                     0
-OmegaLambda                 0 
-OmegaBaryon                 0
-HubbleParam                 1
-Hubble                      100
-BoxSize                     0
-
-#---SPH
-ArtBulkViscConst             0.8
-MinEgySpec                   0
-InitGasTemp                  100
-
-#---Initial density estimate (SPH)
-DesNumNgb                   64
-MaxNumNgbDeviation          1
-\end{verbatim}
+Here, we compare our density profiles against past derivations of
+density profiles for Sculptor and Ursa Minor

### 2025-07-15 thesis/data.tex
major thesis restructure

@@
-\section{Abstract}\label{abstract}
-
 To provide an empirical benchmark for simulations in the next chapters,
@@
 
-\section{Introduction}\label{introduction}
-
-Since the discovery of dwarf galaxies around the Milky Way,
-observational work has attempted to measure and refine their basic
-properties. While the Milky Way's satellites are close (by extragalactic
-standards), their low numbers of stars and large apparent sizes present
-challenges for observational work. In particular, the properties of the
-faintest dwarf galaxy systems and especially \emph{ambiguous} systems
-remains a pressing challenge as a unique probe into the nature of the
-early universe and dark matter.
-
-Among local dwarf galaxies, classical systems are perhaps the systems
-best studied and with the best derived properties.
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} present our
-adopted observed properties for Sculptor and Ursa Minor. While extending
-1-2 degrees across the sky (e.g.
-Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
-classical dwarfs larger number of bright (giant) stars allow for
-thousands of stars to be observed with deep photometry and spectroscopy
-(e.g. \citet{tolstoy+2023}, \citet{pace+2020}). As a result, the
-determination of fundamental properties such as the position, size,
-orientation, distance (from 100s of RRL stars, e.g. \citet{tran+2022};
-\citet{garofalo+2025}), proper motions (from Gaia, described in next
-section), line-of-sight (LOS) velocity, and velocity dispersion
-\(\sigma_v\) are relatively well constrained today and show excellent
-convergence among different works. However, ongoing research still is
-refining our understanding of the detailed structure of Milky Way
-satellites, hinting that these objects may be more complex than at first
-glance.
-
-A number of works have previously speculated that dwarf galaxies such as
-Sculptor and Ursa Minor may have signatures of tidal perturbations,
-however not without contention. While \citet{hodge1961} and
-\citet{demers+krautter+kunkel1980} were some of the earliest work to
-derive the density profile of Sculptor, \citet{innanen+papp1979} was
-perhaps the first to speculate that Sculptor may harbour a substantial
-population of ``extratidal stars'' (stars beyond the tidal radius),
-finding candidate members out to 180' in an elongated distribution from
-\citet{vanagt1978}`s catalogue of variable stars.\footnote{Interestingly,
-  \citet{innanen+papp1979} also speculate about Uranus's satellite
-  distribution, covering gravitational dynamics at two very different
-  scales.} Additionally, \citet{eskridge1988} showed a possible excess
-of stars in Sculptor relative to a \citet{king1962} and Exponential
-density profile beginning around 50', but suggest that this excess may
-not be unusual. Later work by \citet{IH1995}, \citet{walcher+2003}, and
-\citet{westfall+2006} also showed evidence of an extended component of
-Sculptor's density profile, interpreting these stars as either evidence
-of tidal debris or a dwarf galaxy halo. On the other hand,
-\citet{coleman+dacosta+bland-hawthorn2005} show that Sculptor is well
-described by a two component density profile, they additionally mention
-that it is unlikely many of Sculptor's stars (less than 2\%) are
-extratidal. While \citet{munoz+2018} is the most recent and deepest
-photometric study of Sculptor, they only cover an area of out to radii
-of about 30', so they are unable to study the possible extended
-components of Sculptor detected in past works. In addition to the
-presence of ``extratidal stars'', many studies note that Sculptor's
-stars appear to become more elliptical with radius, consistent with
-tidal effects (\citet{IH1995}; \citet{westfall+2006}).
-
-Ursa Minor was often noted not for a density excess but for unusual
-features. Starting with the first density profile from
-\citet{hodge1964}, Ursa Minor is known to be highly elliptical. Studies
-by \citet{olszewski+aaronson1985}, \citet{demers+1995}, \citet{IH1995},
-\citet{kleyna+1998}, and \citet{bellazzini+2002} note that Ursa Minor
-appears to contain substructure along its major axis, but without strong
-interpretation for the causes. As one interpretation \citet{kleyna+2003}
-suggest that this feature is a long-lived star cluster residing in a
-cored dark matter halo to be dynamically stable. One of the first direct
-suggestions of tidal features came from \citet{martinez-delgado+2001},
-who find that stars extend far beyond the nominal tidal radius for Ursa
-Minor in the direction of the galaxies elongation, interpreting this as
-evidence for tidal effects. \citet{palma+2003} corroborate many of these
-earlier findings, showing further evidence for Ursa Minor's peculiar
-morphology including S-shaped contours, a possible extended extratidal
-population of stars, and a failure for a King density profile to
-adequately capture the actual stellar distribution. Each of these
-characteristics appear to indicate strong tidal disruption. Using
-velocity observations, \citet{pace+2014} additionally show that there
-are multiple components in spatial-velocity information. However,
-\citet{munoz+2018}'s more modern photometry show a more regular
-distribution of stars. Irregardless, Ursa Minor has had enough work
-suggesting peculiarities that a deeper investigation into the
-possibility of tidal effects is worthwhile.
-
-However, everything discussed so far is before \emph{Gaia}. Thus, the
-knowledge of the orbits of these systems is largely unknown and most
-surveys could only filter out stars using photometry. As a result, the
-tidal radius could only be guessed based on either the density profile
-or current distance of the dwarf galaxy. With \emph{Gaia} (discussed
-more below), recent work has used Bayesian frameworks to derive
-systematic proper motions of many dwarf galaxies (e.g. \citet{MV2020a})
-and filter away foreground stars using proper motions and parallax to
-detect distant members, and study the 6D internal kinematics of these
-galaxies (e.g. \citet{tolstoy+2023}). Most relevant here,
-\citet{jensen+2024} present a bayesian algorithm to determine likely
-members in \emph{Gaia} (described below), and detect extended secondary
-components for 9 dwarf galaxies, including Sculptor and Ursa Minor.
-\citet{sestito+2023a} and \citet{sestito+2023b} followed up a few of the
-most distant stars detected in \citet{jensen+2024} confirming that
-members exist in each galaxy as far ass 9-12 \(R_h\) from the centres.
-In this chapter, we discuss the origin and reliability of these
-detections, confirming that the density profiles are indeed robust.
-These density profiles then provide the scientific motivation to explore
-the tidal interpretation in the next chapters.
-
-\emph{Maybe define \(R_{\rm ell}\)} \emph{and} \(\xi, \eta\)
-\emph{here}? \emph{We will discuss multiple populations in future
-chapters}
-
-\begin{table*}[t]
-\centering
-\caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
-References are: 1.
-Ricardo R.
-Muñoz et al.
-(2018) Sérsic fits, 2.
-Tran et al.
-(2022), 3.
-McConnachie and Venn (2020b), 4.
-Arroyo-Polonio et al.
-(2024), 5.
-McConnachie and Venn (2020a).
-}
-\label{tbl:scl_obs_props}
-\begin{tabular}{lll}
-\toprule
-parameter & value & Source\\
-\midrule
-$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
-$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
-distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
-distance & $83.2 \pm 2$ kpc & 2\\
-$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
-$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & 4\\
-$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
-ellipticity & $0.37 \pm 0.01$ & 1\\
-position angle & $94\pm1^\circ$ & 1\\
-$M_V$ & $-10.82\pm0.14$ & 1\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{table*}[t]
-\centering
-\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor.
-References are: (1) Ricardo R.
-Muñoz et al.
-(2018) Sérsic fits, (2) Garofalo et al.
-(2025), (3) McConnachie and Venn (2020a), (4) Pace et al.
-(2020), average of MMT and Keck results.
-}
-\label{tbl:umi_obs_props}
-\begin{tabular}{lll}
-\toprule
-parameter & value & Source\\
-\midrule
-$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
-$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
-distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
-distance & $70.1 \pm 3.6$ kpc & 2\\
-$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
-LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
-$\sigma_v$ & $8.6 \pm 0.3$ & 4\\
-$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
-ellipticity & $0.55 \pm 0.01$ & 1\\
-position angle & $50 \pm 1^\circ$ & 1\\
-$M_V$ & $-9.03 \pm 0.05$ & 1\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\subsection{\texorpdfstring{The \emph{Gaia}
-mission}{The Gaia mission}}\label{the-gaia-mission}
-
-Some of the most fundamental properties of astronomical objects are
-their positions and velocities. Unfortunately, determining distances to
-stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
-are easily determined from spectroscopy, the tangental velocities,
-perpendicular to the LOS, are best measured through proper motions,
-typically requiring accurate determinations of small changes in a star's
-position. \emph{Gaia}'s mission is to produce extremely precise
-astrometry enabling measurements of unprecedented accuracy and scale for
-proper motions, parallaxes, and magnitudes.
-
-\emph{Gaia} was designed to revolutionize proper motion and parallax
-measurements. \emph{Gaia} is a space-based, all-sky survey telescope
-with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
-point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
-completing its mission in 2025 (but with two more data releases
-planned). By imagining two patches of sky on the same focal plane,
-separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
-measure absolute proper motions. Stars in different regions of the sky
-are affected by parallax differently, so by observing the relative
-positions of stars separated by large angles during multiple epochs over
-a year, an absolute all-sky reference frame can be derived
-\citep{gaiacollaboration+2016}. In addition to precise astrometric
-information, \emph{Gaia} measures the magnitude of stars in the very
-wide \emph{G} band (330-1050nm), blue and red colours using the blue and
-red photometers (BP and RP, 330-680, 640-1050 respectively), and takes
-low resolution BP-RP spectra and radial velocity measurements of bright
-stars (\emph{Gaia} radial velocity magnitudes \textless16)
-\citep{gaiacollaboration+2016}. For our purposes, most useful are
-\emph{Gaia}'s measurements of \(G\) magnitude,
-\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
-the proper motions in RA and declination \(\mu_{\alpha*}\) and
-\(\mu_\delta\) (where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\)
-corrects for projection effects).
-
-\emph{Gaia} has revolutionized many astronomical disciplines, the least
-of which is local group and Milky Way science. While proper motions of a
-dwarf galaxies has been measured in a case by case basis by the Hubble
-Space Telescope, a full systematic determination of proper motions for
-most dwarf galaxies was unavailable until \emph{Gaia}
-\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
-the detection and improved measurements of halo substructures
-(CITATIONS), streams (CITAITONS), and Milky Way structure and
-kinematics, among many other applications.
-
-\section{Gaia Membership Selection}\label{gaia-membership-selection}
-
-\subsection{Quality cuts and membership
-model}\label{quality-cuts-and-membership-model}
+\section{\texorpdfstring{\emph{Gaia} data}{Gaia data}}\label{gaia-data}
 
@@
 
-To create a high-quality sample, J+24 select stars initially from Gaia
-within a 2--4 degree circular region centred on the dwarf satisfying:
-
-\begin{itemize}
-\tightlist
-\item
-  Solved astrometry, magnitude, and colour.
-\item
-  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
-  quality astrometry. \texttt{ruwe} is a measure of the excess
-  astrometric noise on fitting a consistent parallax-proper motion
-  solution \citep[see][]{lindegren+2021}.\\
-\item
-  3\(\sigma\) consistency of measured parallax with dwarf's distance
-  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
-  correction).
-\item
-  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
-  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
-  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
-\item
-  Corrected colour excess is within 3\(\sigma\) of the expected
-  distribution from \citet{riello+2021}. Removes stars with unreliable
-  photometry.
-\item
-  De-reddened \(G\) magnitude is between
-  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
-  and stars significantly brighter than the tip of the red giant branch
-  (TRGB) magnitude plus the distance modulus uncertainty
-  \(\sigma_{\rm DM}\).
-\item
-  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
-  Removes stars substantially outside the expected CMD.
-\end{itemize}
-
-Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
-extinction maps.
-
 J+24 define likelihoods \({\cal L}\) representing the probability
@@
 the sum of the satellite and background likelihoods, weighted by their
-relative frequencies,
-\begin{equation}\protect\phantomsection\label{eq:Ltot}{
+relative frequencies, \begin{equation}{
 {\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
@@
 
-For the satellite's spatial likelihood, J+24 consider both one-component
-and a two-component density models. The one component model is
-constructed as a single exponential profile ( surface density
-\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\)
-fixed to the value in table 1 of \citet{MV2020a} from \citet{munoz+2018}
-(for a Sérsic fit). Additionally, structural uncertainties (for position
-angle, ellipticity, and scale radius) are sampled over to construct the
-final likelihood map. The two-component model instead adds a second
-exponential,
-\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
-scale radius is fixed, and the outer scale radius and magnitude of the
-second component \(R_{\rm outer}\), \(B\) are free parameters.
-Structural property uncertainties are not included in the two-component
-model.
-
-The PM likelihood is a bivariate gaussian with variance and covariance
-equal to each star's proper motions. J+24 assume the stellar PM errors
-are the main source of uncertainty.
-
-The satellite's CMD likelihood is based on a Padova isochrone
-\citep{girardi+2002}. The isochrone has a matching metallicity and 12
-Gyr age (except 2 Gyr is used for Fornax). The (gaussian) colour width
-is assumed to be 0.1 mag plus the Gaia colour uncertainty at each
-magnitude. The horizontal branch is modelled as a constant magnitude
-extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB stars and a
-0.1 mag width plus the mean colour error). A likelihood map is
-constructed by sampling the distance modulus in addition to the CMD
-width, taking the maximum of RGB and HB likelihoods.
-
-The background likelihoods are instead empirically constructed. Stars
-stars outside of 5\(R_h\) passing the quality cuts estimate the
-background density in PM and CMD space. The density is a sum of
-bivariate gaussians with variances based on Gaia uncertainties (and
-covariance for proper motions). The spatial background likelihood is
-assumed to be constant.
-
-J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
-\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
-with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
-informative. The proper motion single component prior is same as
-\citet{MV2020a}: a normal distribution with mean 0 and standard
-deviation \(100\ \kms\). If 2-component spatial, instead is a uniform
-distribution spanning 5\(\sigma\) of single component case w/ systematic
-uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
-\(R_{\rm outer}\) has a uniform prior only restricting
-\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
-then reported and used to calculate the final \(P_{\rm sat}\) values.
-
-\subsection{Selected samples}\label{selected-samples}
+\section{Selected stars}\label{selected-stars}
 
@@
 \(\xi\) and \(\eta\) are the distances in RA and declination as measured
-on a plane tangent to the dwarf galaxies centre.
+on a plane tangent to the dwarf galaxies centre. \(R_{\rm ell}\) is
+defined as the circularized elliptical radius, so
+\(R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2)\) where
+\(\xi'\) and \(\eta'\) are rotated to align with the position angle
+direction \(\theta\).
 
@@
 
-\subsection{Density Profiles}\label{density-profiles}
+\section{Density profiles}\label{density-profiles}
 
@@
 
-\subsection{Caveats}\label{caveats}
-
-The J+24 method was designed in particular to detect the presence of a
-density excess and find individual stars at large radii to be followed
-up. We are more interested in accurately quantifying the density profile
-and size of any perturbations. One potential problem with using J+24's
-candidate members is that the algorithm assumes the density is either
-described by a single or double exponential. If this model does not
-accurately match the actual density profile of the dwarf galaxy, we want
-to understand the impact of this assumption.
-
-In particular, in Fig.~\ref{fig:scl_observed_profiles}, notice that the
-\(P_{\rm sat}\) selection method produces small errorbars, even when the
-density is more than 1 dex below the local background. These stars are
-likely selecting stars from the statistical MW background consistent
-with UMi PM / CMD, recovering the assumed density profile. As a result,
-the reliability of these density profiles below the CMD+PM background
-may be questionable. A more robust analysis, removing this particular
-density assumption, would be required to more appropriately represent
-the knowledge of the density profile as the background begins to
-dominate.
-
-J+24 do not account for structural uncertainties in dwarfs for the two
-component case. We assume constant ellipticity and position angle. Dwarf
-galaxies, in reality, are not necessarily smooth or have constant
-ellipticity. J+24 test an alternative method using circular radii for
-the extended density component, and we find these density profiles are
-very similar to the fully elliptical case, even when assuming circular
-bins for the circular outer component. As such, even reducing the
-assumed ellipticity from \(0.37-0.55\) to 0 does not substantially
-impact the density profiles.
-
-Finally, our uncertainties are likely underrepresented. A more self
-consistent model would fit the density profile to the entire field at
-once, eliminating possible misrepresentation of the uncertainties. In
-the Appendix to this section, we \emph{will} also discuss an alternative
-method which runs a MCMC model using the likelihoods above to solve for
-the density in each elliptical bin. From these tests, we note that the
-density profile and uncertainties derived from the J+24 sample are
-reliable insofar as the dwarf density is above the background from MW
-CMD+PM-consistent interlopers. We estimate that this effect comprimises
-the density profiles past \(\log R / {\rm arcmin} = 1.8\) for Sculptor
-and Ursa Minor, but agreement is good before then.
-
-While \emph{Gaia} has shown excellent performance, some notable
-limitations may introduce problems in our interpretation and reliability
-of density profiles. Gaia systematics in proper motions and parallaxes
-are typically smaller than the values for sources of magnitudes
-\(G\in[18,20]\). Since we use proper motions and parallaxes as general
-consistency with the dwarf, and factor in systematic uncertainties in
-each case, these effects should not be too significant. However, the
-systematic proper motion uncertainties becomes the dominant source of
-uncertainty in the derived systemic proper motions of each galaxy (see
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
-
-\emph{Gaia} shows high but imperfect completeness, particularly showing
-limitations in crowded fields and for faint sources (\(G\gtrapprox20\)).
-As discussed in \citet{fabricius+2021}, for the high stellar densities
-in globular clusters, the completeness relative to HST varies
-significantly with the stellar density. However, the typical stellar
-densities of dwarf galaxies are much lower, at about 20 stars/arcmin =
-90,000 stars / degree, lower than the lowest globular cluster densities
-and safely below the crowding limit of 750,000 objects/degree for BP/RP
-photometry. In \citet{fabricius+2021}, for the lowest density globular
-clusters, the completeness down to \(G\approx 20\) is \(\sim 80\%\).
-Closely separated stars pose problems for Gaia's on-board processing, as
-the pixel size is 59x177 mas on the sky. This results in a reduction of
-stars separated by less than 1.5'' and especially for stars separated by
-less than 0.6 arcseconds. The astrometric parameters of closely
-separated stars furthermore tends to be of lower quality
-\citep{fabricius+2021}. However, even for the denser field of Fornax,
-only about 3\% of stars have a neighbour within 2 arc seconds, so
-multiplicity should not affect completeness too much (except for
-unresolved binaries). One potential issue is that the previous analyses
-do not account for our cuts on quality and number of astrometric
-parameters. These could worsen completeness, particularly since the
-BP-RP spectra are more sensitive to dense fields. In the appendix to
-this section, we test if magnitude cuts impact the resulting density
-profiles, finding that this is likely not an issue.
-
-Finally, an excellent test of systematics and methods in \emph{Gaia} is
-to compare against another survey. In the appendix, we show that
-profiles derived from J+24 agree with DELVE or UNIONS data within
-uncertainties. Systematics, completion effects, and selection methods
-are unlikely to substantially change the density profiles presented
-here.
-
-\section{Discussion and Conclusions}\label{discussion-and-conclusions}
+\section{Summary}\label{summary}
 
@@
 the observed properties of Sculptor and Ursa Minor.
-
-Are dwarf galaxies indeed expected to be one-component exponential-like
-density profiles? The suggestion of exponential density profiles dates
-back to \citet{faber+lin1983}. Expand\ldots{}
-
-\section{Additional density profile
-tests}\label{additional-density-profile-tests}
-
-In this section, we discuss additional tests and verification of the
-derived density profiles. In particular, we check that methodology
-(simpler cuts, circularized radii, algorithm version) do not
-substantially affect the density profile. We also compile density
-profiles presented in the literature as reference. In all cases, the
-density profiles appear to have excellent convergence out to
-\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
-where the background dominates.
-
-Discuss selection criteria for DELVE and UNIONS samples, literature
-comparison, simple selection criteria, MCMC density profiles and when
-\citet{jensen+2024} becomes background-limited.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
-\caption[Density profiles]{Density profiles for various assumptions for
-Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
-2-component bayesian model assuming circular radii, simple is the series
-of simple cuts described, bright is the sample of the brightest half of
-stars (scaled by 2), DELVE is a sample of RGB stars (background
-subtracted and rescaled to match).}\label{fig:scl_density_extras}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_j24.pdf}}
-\caption[Density profiles]{Comparison of density profiles for each J+24
-method. The fiducial is a 2-component elliptical model. However, the
-1-component is still elliptical but only contains 1 component and the
-circular model assumes a circular outer density profile and bins in
-circular bins instead of elliptical
-bins.}\label{fig:scl_density_j24_methods}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
-\caption[UMi Density profiles]{Similar to
-Fig.~\ref{fig:scl_observed_profiles} except for Ursa
-Minor}\label{fig:umi_density_extras}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_j24.pdf}}
-\caption[UMi density methods]{Similar to
-Fig.~\ref{fig:scl_density_j24_methods} except for Ursa
-Minor.}\label{fig:umi_density_j24_methods}
-\end{figure}

### 2025-07-15 thesis/discussion.tex
major thesis restructure

\section{Reliability and comparison of observed density
profiles}\label{reliability-and-comparison-of-observed-density-profiles}

\subsection{Caveats}\label{caveats}

While the J+24 method is excellent and has been verified, there are
several possible limitations we discuss here. In general, while these
limitations are likely real, they do not substantially affect our
conclusions up to where we suggest the density profiles are unreliable.

\textbf{Spatial likelihood.} J+24 method was designed in particular to
detect the presence of a density excess and find individual stars at
large radii to be followed up. We are more interested in accurately
quantifying the density profile and size of any perturbations. One
potential problem with using J+24's candidate members is that the
algorithm assumes the density is either described by a single or double
exponential. If this model does not accurately match the actual density
profile of the dwarf galaxy, we want to understand the impact of this
assumption.In particular, in Fig.~\ref{fig:scl_observed_profiles},
notice that the \(P_{\rm sat}\) selection method produces small
errorbars, even when the density is more than 1 dex below the local
background. These stars are likely selecting stars from the statistical
MW background consistent with UMi PM / CMD, recovering the assumed
density profile. As a result, the reliability of these density profiles
below the CMD+PM background may be questionable. A more robust analysis,
removing this particular density assumption, would be required to more
appropriately represent the knowledge of the density profile as the
background begins to dominate.

\textbf{Uncertainty misrepresentation.} A more self consistent model
would fit the density profile to the entire field at once, eliminating
possible misrepresentation of the uncertainties. In the Appendix to this
section, we \emph{will} also discuss an alternative method which runs a
MCMC model using the likelihoods above to solve for the density in each
elliptical bin. From these tests, we note that the density profile and
uncertainties derived from the J+24 sample are reliable insofar as the
dwarf density is above the background from MW CMD+PM-consistent
interlopers. We estimate that this effect comprimises the density
profiles past \(\log R / {\rm arcmin} = 1.8\) for Sculptor and Ursa
Minor, but agreement is good before then.

\textbf{\emph{Gaia} systematics}. While \emph{Gaia} has shown excellent
performance, some notable limitations may introduce problems in our
interpretation and reliability of density profiles. Gaia systematics in
proper motions and parallaxes are typically smaller than the values for
sources of magnitudes \(G\in[18,20]\). Since we use proper motions and
parallaxes as general consistency with the dwarf, and factor in
systematic uncertainties in each case, these effects should not be too
significant. However, the systematic proper motion uncertainties becomes
the dominant source of uncertainty in the derived systemic proper
motions of each galaxy (see
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).

\textbf{Completeness}. \emph{Gaia} shows high but imperfect
completeness, particularly showing limitations in crowded fields and for
faint sources (\(G\gtrapprox20\)). As discussed in
\citet{fabricius+2021}, for the high stellar densities in globular
clusters, the completeness relative to HST varies significantly with the
stellar density. However, the typical stellar densities of dwarf
galaxies are much lower, at about 20 stars/arcmin = 90,000 stars /
degree, lower than the lowest globular cluster densities and safely
below the crowding limit of 750,000 objects/degree for BP/RP photometry.
In \citet{fabricius+2021}, for the lowest density globular clusters, the
completeness down to \(G\approx 20\) is \(\sim 80\%\). Closely separated
stars pose problems for Gaia's on-board processing, as the pixel size is
59x177 mas on the sky. This results in a reduction of stars separated by
less than 1.5'' and especially for stars separated by less than 0.6
arcseconds. The astrometric parameters of closely separated stars
furthermore tends to be of lower quality \citep{fabricius+2021}.
However, even for the denser field of Fornax, only about 3\% of stars
have a neighbour within 2 arc seconds, so multiplicity should not affect
completeness too much (except for unresolved binaries). One potential
issue is that the previous analyses do not account for our cuts on
quality and number of astrometric parameters. These could worsen
completeness, particularly since the BP-RP spectra are more sensitive to
dense fields. In the appendix to this section, we test if magnitude cuts
impact the resulting density profiles, finding that this is likely not
an issue.

\textbf{Structural uncertainties}. J+24 do not account for structural
uncertainties in dwarfs for the two component case. We assume constant
ellipticity and position angle. Dwarf galaxies, in reality, are not
necessarily smooth or have constant ellipticity. J+24 test an
alternative method using circular radii for the extended density
component, and we find these density profiles are very similar to the
fully elliptical case, even when assuming circular bins for the circular
outer component. As such, even reducing the assumed ellipticity from
\(0.37-0.55\) to 0 does not substantially impact the density profiles.

\textbf{Outlook.} Finally, an excellent test of systematics and methods
in \emph{Gaia} is to compare against another survey. In the appendix, we
show that profiles derived from J+24 agree with DELVE or UNIONS data
within uncertainties. Systematics, completion effects, and selection
methods are unlikely to substantially change the density profiles
presented here.

\subsection{Comparison to literature}\label{comparison-to-literature}

A number of works have previously speculated that dwarf galaxies such as
Sculptor and Ursa Minor may have signatures of tidal perturbations,
however not without contention. While \citet{hodge1961} and
\citet{demers+krautter+kunkel1980} were some of the earliest work to
derive the density profile of Sculptor, \citet{innanen+papp1979} was
perhaps the first to speculate that Sculptor may harbour a substantial
population of ``extratidal stars'' (stars beyond the tidal radius),
finding candidate members out to 180' in an elongated distribution from
\citet{vanagt1978}`s catalogue of variable stars.\footnote{Interestingly,
  \citet{innanen+papp1979} also speculate about Uranus's satellite
  distribution, covering gravitational dynamics at two very different
  scales.} Additionally, \citet{eskridge1988} showed a possible excess
of stars in Sculptor relative to a \citet{king1962} and Exponential
density profile beginning around 50', but suggest that this excess may
not be unusual. Later work by \citet{IH1995}, \citet{walcher+2003}, and
\citet{westfall+2006} also showed evidence of an extended component of
Sculptor's density profile, interpreting these stars as either evidence
of tidal debris or a dwarf galaxy halo. On the other hand,
\citet{coleman+dacosta+bland-hawthorn2005} show that Sculptor is well
described by a two component density profile, they additionally mention
that it is unlikely many of Sculptor's stars (less than 2\%) are
extratidal. While \citet{munoz+2018} is the most recent and deepest
photometric study of Sculptor, they only cover an area of out to radii
of about 30', so they are unable to study the possible extended
components of Sculptor detected in past works. In addition to the
presence of ``extratidal stars'', many studies note that Sculptor's
stars appear to become more elliptical with radius, consistent with
tidal effects (\citet{IH1995}; \citet{westfall+2006}).

Ursa Minor was often noted not for a density excess but for unusual
features. Starting with the first density profile from
\citet{hodge1964}, Ursa Minor is known to be highly elliptical. Studies
by \citet{olszewski+aaronson1985}, \citet{demers+1995}, \citet{IH1995},
\citet{kleyna+1998}, and \citet{bellazzini+2002} note that Ursa Minor
appears to contain substructure along its major axis, but without strong
interpretation for the causes. As one interpretation \citet{kleyna+2003}
suggest that this feature is a long-lived star cluster residing in a
cored dark matter halo to be dynamically stable. One of the first direct
suggestions of tidal features came from \citet{martinez-delgado+2001},
who find that stars extend far beyond the nominal tidal radius for Ursa
Minor in the direction of the galaxies elongation, interpreting this as
evidence for tidal effects. \citet{palma+2003} corroborate many of these
earlier findings, showing further evidence for Ursa Minor's peculiar
morphology including S-shaped contours, a possible extended extratidal
population of stars, and a failure for a King density profile to
adequately capture the actual stellar distribution. Each of these
characteristics appear to indicate strong tidal disruption. Using
velocity observations, \citet{pace+2014} additionally show that there
are multiple components in spatial-velocity information. However,
\citet{munoz+2018}'s more modern photometry show a more regular
distribution of stars. Irregardless, Ursa Minor has had enough work
suggesting peculiarities that a deeper investigation into the
possibility of tidal effects is worthwhile.

However, everything discussed so far is before \emph{Gaia}. Thus, the
knowledge of the orbits of these systems is largely unknown and most
surveys could only filter out stars using photometry. As a result, the
tidal radius could only be guessed based on either the density profile
or current distance of the dwarf galaxy. With \emph{Gaia} (discussed
more below), recent work has used Bayesian frameworks to derive
systematic proper motions of many dwarf galaxies (e.g. \citet{MV2020a})
and filter away foreground stars using proper motions and parallax to
detect distant members, and study the 6D internal kinematics of these
galaxies (e.g. \citet{tolstoy+2023}). Most relevant here,
\citet{jensen+2024} present a bayesian algorithm to determine likely
members in \emph{Gaia} (described below), and detect extended secondary
components for 9 dwarf galaxies, including Sculptor and Ursa Minor.
\citet{sestito+2023a} and \citet{sestito+2023b} followed up a few of the
most distant stars detected in \citet{jensen+2024} confirming that
members exist in each galaxy as far ass 9-12 \(R_h\) from the centres.
In this chapter, we discuss the origin and reliability of these
detections, confirming that the density profiles are indeed robust.
These density profiles then provide the scientific motivation to explore
the tidal interpretation in the next chapters.

\section{Limitations of simulations}\label{limitations-of-simulations}

Variations to the potential of the inner disk (exclusion of a bar,
spiral arms) should minimally affect our results as no orbit we consider
reaches less than \textasciitilde15 kpc of the MW centre. We exclude the
mass evolution of the halo from this analysis. Over \(10\,\)Gyr, this
would be fairly significant (factor of \(\sim 2\)in MW mass, REF) but
since we want to determine the upper limit of tidal effects, it is safe
to neglect this.

\subsection{Long term orbital history}\label{long-term-orbital-history}

\citet{dsouza+bell2022}, \citet{santistevan+2024}. challenges to
backwards time integration

\subsection{Deviations from the NfW}\label{deviations-from-the-nfw}

\begin{itemize}
\tightlist
\item
  \citet{dicintio+2013}
\end{itemize}

\section{Comparison to other work}\label{comparison-to-other-work}

\subsection{Sculptor}\label{sculptor}

Theoretical work on Sculptor

\begin{itemize}
\tightlist
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\item
  \citet{amorisco+zavala+deboer2014}
\item
  \citet{battaglia+2008}
\item
  \citet{breddels+2013}
\item
  \citet{breddels+helmi2013}
\item
  \citet{richardson+fairbairn2014}
\item
  \citet{SFW2017}
\item
  \citet{innanen+papp1979}
\item
  \citet{wilkinson+2002}
\item
  \citet{yang+2025}: chemical evolution in Scl.
\item
  \citet{skuladottir+2024}; \citet{skuladottir+2021};
  \citet{lee+jeon+bromm2024}, pop III high res spectro?
\item
  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
\item
  \citet{tang+2023}: apogee modeling of Scl.
\item
  \citet{pandey+west2022} chem evo OMEGA and isotopes.
\item
  \citet{an+koposov2022}: distance gradients?
\item
  \citet{kawata+2006}, 2pop origins?
\item
  \citet{grcevich+putman2009}, \citet{carignan+1998} ambiguous HI clouds
  near Scl ?
\item
  \citet{agnello+evans2012}, claim that Sculptor cannot hold two
  populations in NFW halo?
\end{itemize}

Observational work on Scl

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023a}
\item
  \citet{westfall+2006} wide degree survey for extended structure.
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
\item
  \citet{coleman+dacosta+bland-hawthorn2005}
\item
  \citet{DQ1994}
\item
  \citet{WMO2009}
\item
  \citet{IH1995},
\item
  \citet{munoz+2018}: Using Megacam to derive density profiles and
  structural properties of many dwarf spheroidal galaxies.
\item
  \citet{kirby+2009}
\item
  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
\item
  \citet{grebel1996}
\item
  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
  metallicity gradient and search for metal poor stars in Scl.
\end{itemize}

Future ideas:

\begin{itemize}
\tightlist
\item
  \citet{evslin2016}: measuring ellipticities of halos w TMT.
\end{itemize}

\subsection{Ursa Minor}\label{ursa-minor}

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023b}
\item
  \citet{pace+2020}
\item
  \citet{pace+2014} substructure
\item
  \citet{bellazzini+2002}
\item
  \citet{hargreaves+1994}
\item
  \citet{martinez-delgado+2001}
\item
  \citet{munoz+2005}: velocity dispersion profile
\item
  \citet{palma+2003}
\item
  \citet{spencer+2018}
\item
  \citet{vitral+2023}
\item
  \citet{piatek+2005}: old HST Pm meas.
\item
  \citet{gallagher+2003}: no ionizaed gas.
\item
  \citet{shetrone+cote+stetson2001}: RGBs
\item
  \citet{wilkinson+2004}
\item
  \citet{kleyna+2003}: dynamical fossil.
\item
  \citet{pryor+kormendy1990}, \citet{lake1990} MLR / velocity dispersion
\item
  \citet{armandroff+olszewski+pryor1995};
  \citet{olszewski+pryor+armandroff1996} MLR
\item
  \citet{aaronson+olszewski+hodge1983}; \citet{aaronson1983} carbon
  stars and velocities
\item
  \citet{stetson1984} early spectroscopy.
\item
  \citet{tsujimoto+shigeyama2002} chem evo.
\end{itemize}

Theoretical work

\begin{itemize}
\tightlist
\item
  robles+bullock2021
\item
  \citet{caproni+lanfranchi2021}, \citet{caproni+2015}: simulation of
  gass loss.
\item
  \citet{bajkova+bobylev2017}, local orbits
\item
  \citet{gomez-flechoso+martinez-delgado2003}: MLR of UMi
\item
  \citet{lynden-bell1976} early discussion of orbits and LMC plane.
\end{itemize}

\section{Are Sculptor and Ursa Minor
typical?}\label{are-sculptor-and-ursa-minor-typical}

\subsection{The formation of Exponential
profiles}\label{the-formation-of-exponential-profiles}

As mentioned earlier, many dwarf galaxies appear to have exponential
stellar density profiles.

Disk galaxies have long been known to be exponential-like across a wide
range of scales. As such, several theoretical works have aimed to
undertsand the formation of exponential stellar disks. One prevailing
theory posits that scattering of stars in a disk naturally forms an
exponential \citep[\citet{wu+2022}]{elmegreen+struck2013}, or that the
disk is due to conservation of angular momentum during spherical
collapse.

Given that dwarf galaxies were noticed to be exponential like above,
\citet{faber+lin1983} proposed that dwarf spheroidals formed from disky
galaxies maintaining the typical exponential density profile

\citet{reed+gilmore2005} propsed that exponential dSph form through
impusive mass loss, redistributing the stellar component into an
exponential like profile.

\begin{itemize}
\tightlist
\item
  disk galaxy formation \citet{fall+efstathiou1980}, \citet{mestel1963}
\end{itemize}

One possible explanation is that \citet{mayer+2001a}

Other explanations range from the effects of mass loss, feedback,
angular momentum,

\citet{klimentowski2007}, \citet{klimentowski2009}.

In summary, while there is some theoretical reasons why exponential
spheroidal stellar profiles may form, the emperical sucess of an
exponential law still requires explanations. Deviations from this
exponential thus test wether this exponential law is indeed
near-universal or if other ingredients in the formation and history of
dwarf galaxies may be responsible for changes from the empirical
expectation.

\section{Understanding the extended density profiles of dwarf
galaxies}\label{understanding-the-extended-density-profiles-of-dwarf-galaxies}

Are dwarf galaxies indeed expected to be one-component exponential-like
density profiles? The suggestion of exponential density profiles dates
back to \citet{faber+lin1983}. Expand\ldots{}

A number of recent work has also confirmed the presence of extended,
likely non-tidal, density profiles. For example, \citet{chiti+2021};
\citet{chiti+2023} spectroscopically confirm members out to \(~9 R_h\)
in Tucana II.

\begin{itemize}
\tightlist
\item
  \citet{revaz+jablonka2018}
\item
  \citet{tau+vivas+martinez-vazquez2024} (RRL stars )
\item
  \citet{roderick+2016} Boötes I
\item
  \citet{mcconnachie+penarrubia+navarro2007} (velocity space \& multiple
  populations).
\end{itemize}

\subsection{Multi-epoch star formation and
feedback}\label{multi-epoch-star-formation-and-feedback}

The star formation history of dwarf galaxies is typically thought to be
\emph{bursty}: a series of discrete episodes of intense star formation
separated by periods of quiescence.

For example, in the simulations of \citet{wheeler+2019}, etc., dwarf
galaxies exhibit strongly bursty SFHs.

\begin{itemize}
\tightlist
\item
  \citet{maxwell+2012}
\item
  \citet{wright+2019}
\item
  \citet{azartash-namin+2024}
\end{itemize}

\subsubsection{Multi-component density
profiles}\label{multi-component-density-profiles}

A related note is that several dwarf galaxies show evidence for multiple
chemodynamical components in their stars. While spectra tend to focus on
the inner regions (high probability members), the existence of
multiple-component stellar populations in the inner regions of dwarf
galaxies hints at a complex star formation history capable of creating
the observed profiles.

\begin{itemize}
\tightlist
\item
  \citet{benitez-llambay+2016}
\item
  \citet{mercado+2021}: formation of metallicity gradients from SFH
\item
  \citet{revaz+jablonka2018} natural formation of multi-component and
  gradients from cosmological simulations of dwarf galaxies. Originate
  from dynamical heating of metal rich population and outside in star
  formation?
\item
  \citet{el-badry+2016}. Feedback drives fluctuations in size + radial
  migration =\textgreater{} population gradients.
\end{itemize}

Observational evidence:

\begin{itemize}
\tightlist
\item
  \citet{arronyo-polonio+2024}
\item
  \citet{pace+2020}
\item
  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
  Carina)
\end{itemize}

\subsection{Past mergers and
accretion}\label{past-mergers-and-accretion}

\begin{itemize}
\tightlist
\item
  \citet{deason+2014}
\item
  \citet{deason+2022}
\item
  \citet{ricotti+2022}
\item
  \citet{querci+2025}
\item
  \citet{amorisco+evans+vandeven2014}
\item
  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
  context of Tuc II, merger of two galaxies in the early universe
  produces a more extended density profile and metallicity gradient.
\item
  \citet{lokas+2014}: And II from merger,
\item
  rotation from merger: \citet{cardona-barrero+2021}
\end{itemize}

\subsection{Preprocessing}\label{preprocessing}

Another alternative is that the dwarf galaxies have been
``preprocessed'' by another dwarf/LMC like galaxy earlier on and this
has resulted in their unique properties

\subsection{Tidal dwarfs}\label{tidal-dwarfs}

On e possibility is that Sculptor and Ursa Minor are not formed from
dark matter halos but instead due to the collision of gas. This
possibility is not a convincing explanation because we know the star
formations histories are extended and relatively old, and the galaxies
are close to apocentre, making achieving the observed velocity
dispersion challenging in this framework.

\subsection{\texorpdfstring{Alternatives to
\(\Lambda\)CDM}{Alternatives to \textbackslash LambdaCDM}}\label{alternatives-to-lambdacdm}

Alternatives to the prevailing cosmology we assume may help explain the
nature of these dwarf galaxies. t

\begin{itemize}
\tightlist
\item
  WDM: forms cores, doesn't help
\item
  SIDM: can cause faster disruption, doesn't fix orbital path or change
  tidal effects?
\item
  MOND: \citet{sanchez-salcedo+hernandez2007}: mond in dsph
\end{itemize}

### 2025-07-15 thesis/introduction.tex
major thesis restructure

--- 2025-05-02 thesis/introduction.tex
@@
-\section{Background \& Past Work}\label{background-past-work}
+Dwarf galaxies host, in many ways, the most extreme galactic
+environments in the universe. These little galaxies are typically
+defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\),
+e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
+galaxies, dwarfs are the most numerous galaxies in the universe. Dwarf
+galaxies are highly \emph{dark-matter dominated}, with mass to light
+ratios sometimes exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of
+their small total masses, many dwarf galaxies are \emph{quenched}, with
+little to no recent star formation. Their stellar populations are
+\emph{relics} from the early universe, consisting of many of the oldest
+and most metal poor stars. Understanding the properties of dwarf
+galaxies thus has relevance across astronomy, from cosmological
+structure formation on the smallest scale to extremely metal-poor
+stellar populations to chemical evolution to galactic dynamics. Of the
+classical dwarfs, the Sculptor and Ursa Minor dwarf spheroidal galaxies
+stand out with a more extended density profile relative to an
+exponential, hinting at tidal effects. As a case study in tidal effects
+and our understanding of dwarf galaxy formation and evolution, we aim to
+understand the dynamical history of these galaxies.
+
+In this section, we first describe the observational history of dwarf
+galaxies, what \emph{Gaia} is and how this telescope has changed our
+understanding of the local group, the present state of dwarf galaxy
+knowledge, and why Sculptor and Ursa Minor stand out. We then describe
+the role of dwarf galaxies in cosmology, their formation and our current
+theory of hierarchichal structure formation. We later discuss idealized
+simulations, tidal effects, break radii, and recent developments in
+tidal simulations. Finally, we provide a roadmap to the thesis as a
+whole at the end of this section.
+
+\section{Observational context}\label{observational-context}
+
+\subsection{Early observations}\label{early-observations}
+
+The Magellanic clouds are the most prominent of the Milky Way satellites
+and were likely always known to humans in the southern hemisphere.
+However, the first discovered dwarf galaxies in 1938, Sculptor and
+Fornax, heralds i n an era of discovery and investigation into a novel
+astronomical system. \citet{shapley1938} present Sculptor and Fornax as
+some kind of \emph{stellar system} resembling the Magellanic clouds and
+globular clusters. The first direct hints that these stellar systems are
+galaxies originates from early spectroscopic work
+\citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
+\citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
+\citet{olszewski1998}.]{aaronson1983}. Based on their velocity
+dispersions given their physical sizes, the mass to light ratios of
+these dwarf galaxies is far larger than expected for star clusters
+without dark matter. In general, these objects have reached a consensus
+that they have high velocity dispersion, old spheroidal stellar
+populations, and low stellar masses. While we now know the basic
+properties of dwarf galaxies, tremendous progress has been made in the
+past 10-20 years with large area deep photometric surveys, multi-object
+spectrographs, high resolution chemistry, and theoretical work,
+unveiling that dwarf galaxies are each unique systems with different
+formation pathways \citep[e.g.][]{simon2019}.
+
+Dwarf galaxies span a large range of size and morphology.
+Fig.~\ref{fig:galaxy_images} displays several images of dwarf galaxies.
+The largest in the local group but not always considered a dwarf galaxy,
+the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
+Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. From
+this figure, we notice that while large LMC-like dwarf galaxies exhibit
+complex structure such as a bar, young blue stars, and irregular shapes,
+the classical and faint dwarf galaxies are composed mostly of very old,
+faint, red stars, very diffuse and often large on the sky if local.
+These galaxies are typically classified as dwarf spheroidals (dSph),
+however we typically omit the dSph designation and refer to these
+galaxies by their constellation hereafter.
+
+Since the discovery of dwarf galaxies around the Milky Way,
+observational work has attempted to measure and refine their basic
+properties \citep[e.g.][]{mateo1998}. While the Milky Way's satellites
+are close (by extragalactic standards), their low numbers of stars and
+large apparent sizes present challenges for observational work. In
+particular, the properties of the faintest dwarf galaxy systems and
+especially \emph{ambiguous} systems remains a pressing challenge as a
+unique probe into the nature of the early universe and dark matter.
+
+\textbf{Classical systems}, the brightest of the local group
+(non-Magellanic) dwarf galaxies, are perhaps the systems best studied
+and with the most stringingly derived properties. Defined\footnote{Note
+  that some work like \citet{simon2019} instead define classical dwarfs
+  based on luminosity as compared to ultra faint dwarfs. This may
+  include galaxies such as Antlia I, Crater II, Canes Venatici I, and
+  Sagittarius. We stick with the historical definition for simplicity
+  and since the Gaia observations are most numerous for the historical
+  classicals.} as the first six discovered dwarf spheroidal galaxies,
+(Sculptor, Ursa Minor, Fornax, Draco, Leo I, Leo II, and Carina),
+classical systems are typically old, spheroidal, and gas-poor. While
+extending 1-2 degrees across the sky (e.g.
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
+these dwarfs contain large numbers of bright (giant) stars allow for
+thousands of stars to be observed with deep photometry and spectroscopy
+\citep[e.g.][\citet{pace+2020}]{tolstoy+2023}. As a result, the
+determination of fundamental properties such as the position, size,
+orientation, distance \citep[from 100s of RRL stars,
+e.g.][]{tran+2022, garofalo+2025}, proper motions (from \emph{Gaia},
+e.g. \citet{battaglia+2022}, \citet{MV2020a}), line-of-sight (LOS)
+velocity and dispersion \(\sigma_v\), are all relatively well
+constrained today and show excellent convergence among different works.
+However, ongoing research still is refining our understanding of the
+detailed structure of Milky Way satellites, hinting that these objects
+may be more complex than at first glance.
+
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
+\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Sculptor (DES
+2), Ursa Minor (UNWISE with Gaia point sources over-plotted), and Bootes
+V (SDSS). While the LMC is very prominent in the sky, even the classical
+dwarf galaxies are not obvious except in terms of star counts (e.g.
+Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the central
+third to sixth of the image.}\label{fig:galaxy_images}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
+\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
+on the sky. We label the classical dwarf galaxies (green diamonds),
+fainter dwarfs (blue squares), globular clusters (orange circles), and
+ambiguous systems (pink open hexagons). Globular clusters are more
+centrally concentrated, but dwarf galaxies are preferentially found away
+from the MW disk. Sculptor and Ursa Minor are highlighted as two dwarfs
+we study later. The background image is from ESA/Gaia/DPAC
+(https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).
+Dwarf galaxies (confirmed), globular clusters, and ambiguous systems are
+from the \citet{pace2024} catalogue (version
+1.0.3).}\label{fig:mw_satellite_system}
+\end{figure}
+
+\subsection{\texorpdfstring{The era of
+\emph{Gaia}}{The era of Gaia}}\label{the-era-of-gaia}
+
+Some of the most fundamental properties of astronomical objects are
+their positions and velocities. Unfortunately, determining distances to
+stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
+are easily determined from spectroscopy, the tangental velocities,
+perpendicular to the LOS, are best measured through proper motions,
+typically requiring accurate determinations of small changes in a star's
+position. \emph{Gaia}'s mission addresses these problems by making
+extremely precise measurements of stellar positions over time.
+
+\emph{Gaia} was designed to revolutionize proper motion and parallax
+measurements. \emph{Gaia} is a space-based, all-sky survey telescope
+with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
+point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
+completing its space-based mission in 2025 (with two more data releases
+planned). By imagining two patches of sky on the same focal plane,
+separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
+measure absolute proper motions. Stars in different regions of the sky
+are affected by parallax differently, so by precisely observing the
+relative positions of stars separated by large angles during multiple
+epochs over a year, an absolute all-sky reference frame can be derived
+\citep{gaiacollaboration+2016}. In addition to precise astrometric
+information, \emph{Gaia} measures the magnitude of stars in the very
+wide \emph{G} band (330-1050nm), blue and red colours using the blue and
+red photometers (BP and RP, 330-680 nm and 640-1050 nm respectively),
+and takes low resolution BP-RP spectra and radial velocity measurements
+of bright stars (\emph{Gaia} radial velocity magnitudes \textless16)
+\citep{gaiacollaboration+2016}. For our purposes, most useful are
+\emph{Gaia}'s measurements of \(G\) magnitude,
+\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
+the proper motions in RA and declination \(\mu_{\alpha*}\) and
+\(\mu_\delta\).\footnote{\(\mu_{\alpha*}\) allows where
+  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
+  effects{]}.}
+
+\emph{Gaia} has revolutionized many astronomical disciplines, the least
+of which is local group and Milky Way science. While proper motions of a
+dwarf galaxies has been measured in a case by case basis by the Hubble
+Space Telescope, a full systematic determination of proper motions for
+most dwarf galaxies was unavailable until \emph{Gaia}
+\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
+the detection and improved measurements of halo substructures and
+streams \citep{bonaca+price-whelan2025}, Milky Way structure and
+kinematics \citep{hunt+vasiliev2025}.
+
+For Local group dwarf spheroidals, \emph{Gaia} has allowed for many of
+the first proper motion measurements. While the proper motion
+uncertainty on a typical dwarf member star is often with large
+uncertainties, by combining the proper motions for 100s to 1000s of
+stars in \emph{Gaia}, incredibly precise proper motion measurements can
+be determined \citep[e.g.][]{MV2020a}, sometimes only limited by
+\emph{Gaia}'s systematic error floor. Proper motions have ushered in a
+new dynamical era for MW satellites studies, where we often have precise
+determinations of the dwarf galaxy's (short term) orbit under a given MW
+potential.
+
+\subsection{Dwarf galaxies today}\label{dwarf-galaxies-today}
+
+\subsubsection{Exponential stellar density
+profiles}\label{exponential-stellar-density-profiles}
+
+For a long time, exponential surface density profiles have been used as
+an description for spiral galaxies {[}REFS{]}.
+
+Since \citet{faber+lin1983} demonstrated that an Exponential density
+profile is a reasonable empirical fit to dwarf galaxies, many works have
+fitted an Exponential profile (\citet{Eskridge} 1988a,c; Hodge et al
+1991a,b; \citet{IH1995}, ) in addition to Sérsic, King and various other
+profiles. \citet{lelli+2014} (example of using exponential.)
+
+\citet{wang+2019} for fornax, \citet{mcconnachie+irwin2006} for Exp,
+Plummer, King in other galaxies.
+
+However, even starting from \citet{aparicio1997}, some people have noted
+deviations from this simple empirical rule. Additionally,
+\citet{herrmann+hunter+elmegreen2013},
+\citet{herrmann+hunter+elmegreen2016} note that at least in a
+photometric sample of more distant dwarf disky/Irr/blue compact dwarfs,
+that many dwarfs show deviations from exponential profiles.
 
@@
 \item
-  What is dark matter? Why do we look at dwarfs?
+  \citet{makarov+2012} isolated local volume dwarf galaxy with central
+  deprivation compared to exponential, spheroidal
 \item
-  Forms of dark matter, lambda-CDM, and dwarf galaxies
+  \citet{martin+2016}, exponential fits to many pandas dwarf galaxies
 \item
-  How does gravity affect dwarfs, theory of tidal perturbations
+  \citet{moskowitz+walker2020}, generalized plummer fits.
+\end{itemize}
+
+\subsubsection{Hints of tides and
+disruptions}\label{hints-of-tides-and-disruptions}
+
+\subsubsection{Kinematics and structure}\label{kinematics-and-structure}
+
+\subsection{Sculptor and Ursa Minor: Hints of tidal
+signatures?}\label{sculptor-and-ursa-minor-hints-of-tidal-signatures}
+
+Sculptor and Ursa Minor may appear to be typical dwarf galaxies at first
+glance. Table~\ref{tbl:scl_obs_props}. describe the present-day
+properties of each galaxy. Indeed, Sculptor as the first discovered
+dwarf galaxy, is often described as a ``prototypical'' dSph. However,
+both galaxies have a long history of speculation that they may be
+influenced by the Milky Way's gravity (see discussion). Only recently,
+with \emph{Gaia} data and spectroscopic followup, has the detection of a
+density excess become more unmistakable. Using \citet{jensen+2024}'s
+algorithm (described briefly below), \citet{sestito+2023a},
+\citet{sestito+2023b} report that a ``kink'' in the density profile,
+beginning around 30 arcmin for both Sculptor and Ursa Minor. They
+spectroscopically followup some of the most distant stars, finding
+multiple members between 6-12 half-light radii from the centre of each
+dwarf (REF fig.~XXX). If dwarfs initially begin with exponential
+profiles like Fornax, then these stars should be much rarer.
+
+Perhaps the most straighforward explanation for the density profiles of
+Sculptor and Ursa Minor are tidal interactions. \citet{sestito+2023a},
+\citet{sestito+2023b} conclude that tides are likely the explanation.
+Fig.~\ref{fig:scl_umi_fnx_vs_penarrubia} reproduces their comparison,
+showing that Sculptor and Ursa Minor's density profiles deviate
+substantially from fornax's in the outskirts, and are well-described by
+the tidal model from \citet{PNM2008}. Our goal is to determine, assuming
+\(\Lambda\)CMD, if the effects of the Milky Way (or other satellites)
+may indeed create observable tidal signatures in these galaxies. If
+tides cannot explain these features, Sculptor and Ursa Minor may instead
+contain a extended stellar ``halo'' or second component of the
+galaxy---illustrating evidence of complex history or formation in each
+galaxy and forcing formation models to confront a diversity of density
+profiles in the local group.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_vs_penarrubia.pdf}}
+\caption[Sculptor and Ursa Minor match tidal models]{A plot of the
+surface density profiles of Sculptor, Ursa Minor, and Fornax scaled to
+their half-light radius and the density at half-light radius (data
+described in Section~\ref{sec:data}). The lines represent the initial
+and final models from \citet{PNM2008}'s model (Fig. 4 top left, stellar
+segregation of 0.20 moving on a highly eccentric orbit peri:apo = 1 :
+100)).}\label{fig:scl_umi_vs_penarrubia}
+\end{figure}
+
+\begin{table*}[t]
+\centering
+\caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
+References are: 1.
+Muñoz et al.
+(2018) Sérsic fits, 2.
+Tran et al.
+(2022) RR lyrae distance, 3.
+Alan W.
+McConnachie and Venn (2020b), 4.
+Arroyo-Polonio et al.
+(2024).
+}
+\label{tbl:scl_obs_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
+$\delta$ & $-33.7186 \pm 0.0007^\circ$ & –\\
+distance modulus & $19.60 \pm 0.05$ & 2\\
+distance & $83.2 \pm 2$ kpc & –\\
+$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
+LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
+$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & –\\
+$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
+ellipticity & $0.37 \pm 0.01$ & –\\
+position angle & $94\pm1^\circ$ & –\\
+$M_V$ & $-10.82\pm0.14$ & –\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor.
+References are: (1) Muñoz et al.
+(2018) Sérsic fits, (2) Garofalo et al.
+(2025) RR lyrae distance, (3) Alan W.
+McConnachie and Venn (2020a), (4) Pace et al.
+(2020), average of MMT and Keck results.
+}
+\label{tbl:umi_obs_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & Source\\
+\midrule
+$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
+$\delta$ & $67.2221 \pm 0.0016$˚ & –\\
+distance modulus & $19.23 \pm 0.11$ & 2\\
+distance & $70.1 \pm 3.6$ kpc & –\\
+$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
+$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
+LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
+$\sigma_v$ & $8.6 \pm 0.3$ & –\\
+$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
+ellipticity & $0.55 \pm 0.01$ & –\\
+position angle & $50 \pm 1^\circ$ & –\\
+$M_V$ & $-9.03 \pm 0.05$ & –\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\section{Cosmological and theoretical
+context}\label{cosmological-and-theoretical-context}
+
+The leading theory of cosmology, \(\Lambda\)CMD (cold dark matter) holds
+that the universe is composed of about 68\% dark energy (\(\Lambda\)),
+27\% dark matter (DM), and 5\% regular baryons
+\citep{planckcollaboration+2020}. While baryonic matter is the only
+matter we can directly understand or observe, we still know the general
+properties of dark matter and dark energy. Dark energy causes the
+acceleration of the expansion of the universe on large scales. We do not
+discuss dark energy here as it does not substantially affect the local
+group today. Dark matter makes up the vast majority of mass in galaxies.
+Typically, galaxies have baryonic to dark matter ratios of between 1:5
+to beyond 1:100. The faintest dwarf galaxies harbouring the largest
+proportion of dark matter. In \(\Lambda\)CMD, dark matter is assumed to
+only interact gravitationally with itself and other matter (transparent
+to light, or \emph{dark}). Dark matter is also created \emph{cold},
+i.e.~typical velocities much smaller than the speed of light in the
+early universe. Dark matter is a key ingredient in cosmological
+structural formation, dynamical galaxy structure, and galaxy-galaxy
+interactions.
+
+\subsection{Formation of dwarf
+galaxies}\label{formation-of-dwarf-galaxies}
+
+The very early universe was almost featureless. As observed from the
+cosmic microwave background created at redshift \(z\sim 1100\), the
+temperature is a near perfect blackbody with fluctuations only at 1 part
+in 10,000. These tiny perturbations began to grow as the universe ages
+and expands. Initially hot and dense, baryonic matter was prevented from
+collapsing by thermal / radiation pressure. However, dark matter was
+uninhibited, freely passing through the universe and collapsing into
+denser regions. Each self-gravitating overdensity of dark matter is
+known as a \emph{halo}. As the universe continued to expand and cool
+down, baryonic matter began to collapse into dark matter halos.
+Eventually, the densest regions of baryons situated near the centres of
+halos would become the first stars of the first galaxies.
+
+Halos rarely evolve as isolated systems. Instead, structure formation is
+\emph{hierarchical}, where smaller \emph{subhalos} are accreted into
+larger halos which may accrete into yet larger halos. The most massive
+halos are made up of thousands of subhalos (verify), with each subhalo
+becoming entire galaxies making up a galaxy cluster. But, even the
+smallest galaxies reside in a \emph{halo}, with the dwarf galaxies
+representing entities with simpler \emph{assembly histories}.
+
+\subsection{Cosmological dark matter density
+profiles}\label{cosmological-dark-matter-density-profiles}
+
+From cosmological simulations, we know that across many orders of
+magnitude, dark matter halos appear self-similar. In
+\citet{NFW1996};\citet{NFW1997}, hereafter NFW, the authors observe that
+the radial density profiles \(\rho(r)\) follow a two-parameter law:
+\begin{equation}{
+\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2}
+}\end{equation} where \(r_s\) and \(\rho_s\) are the scale radius and
+density (see below). This density profile has shown remarkable success
+as a description of \(\Lambda\)CMD halos \citep[REFS, but see
+also][darkEXP, ect.]{navarro2010}. This density profile is \emph{cuspy},
+where the density continuously rises like \(\rho \sim 1/r\) at the
+centre. Also note that the total mass of the NFW profile logarithmically
+diverges. Typically, the virial mass of NFW halos is discussed instead,
+defined as \(M_{200}\), the mass within a region \(r_{200}\) containing
+a mean density 200 times the critical density of the universe
+\(\rho_{\rm crit}\), so
+\(M_{200} = M(r<r_{200}) = 200\ (4\pi/3) \ r_{200}^3\ \rho_{\rm crit}\).
+
+There are several parameterizations of the NFW profile. \citet{NFW1996}
+initially write the halo in terms of \(r_s\) and a concentration
+\(c= r_{200} / r_s\) relating the virial radius \(r_{200}\) to \(r_s\).
+. In this case,
+\(\rho_s = \rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\). However, we
+chose to discuss halos in terms of velocity space. The circular velocity
+of the NFW halo, \(v_{\rm circ}(r) = \sqrt{G M(r) / r}\) related to the
+enclosed mass \(M(r)\), reaches a maximum of \(v_{\rm max}\) at
+\(r_{\rm max} = \alpha\,r_s\) where \(\alpha\approx2.16258\). Thus
+\(r_{\rm max}\) is directly related to the scale radius and
+\(v_{\rm max}\) is related to the total mass.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095831443.png}}
+\caption[The NFW density profile]{showing the fits to the NFW halos for
+several different halos from \citet{NFW1996}.}
+\end{figure}
+
+\subsection{Typical properties}\label{typical-properties}
+
+A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
+relation, describing the amount of stellar mass forms in a (sub)halo of
+a given size. In particular, the SMHM relation grows especially steep in
+the dwarf galaxy regime---many dwarf galaxies are formed in halos of
+similar masses. \citet{fattahi+2013}, \citet{fattahi+2018}
 
@@
 \item
-    \citet{EN2021}, \citet{PNM2008}, etc.
-  \end{itemize}
-\item
-  Instances of dwarfs undergoing weird processess
-\item
-  Alternative processes and uncertainties in the evolution of dwarfs
+  \citet{battaglia+nipoti2022}
 \end{itemize}
 
-The classical dwarfs are some of the earliest discovered systems,
-begining with \citet{shapley1938}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}}
+\caption[stellar mass halo mass relation]{The stellar mass halo mass
+relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
+\citet{fattahi2018}.}
+\end{figure}
+
+\subsection{Challenges to cosmology or modelling
+limitations?}\label{challenges-to-cosmology-or-modelling-limitations}
+
+One of the open questions in dwarf galaxy evolution is the formation of
+dark matter \emph{cores} where the density becomes constant instead
+\(\rho \sim 1\). A leading theory is cores form through of baryonic
+effects and feedback. If dark matter deviates from \(\Lambda\)CDM, then
+the NFW density profile may not describe halos anymore. \emph{Warm} dark
+matter, relativistic in the early universe but cooler now, smooths out
+small-scale features and softens the cusps of dark matter halos.
+\emph{Self-interacting} dark matter instead can form cores but also the
+cores can collapse into a density peak.
 
@@
 \item
-  \citet{fattahi+2013}, \citet{fattahi+2018}
+  \citet{walker+penarrubia2011}: measuring cores and cusps
 \item
-  \citet{sanchez-salcedo+hernandez2007}: mond in dsph
-\item
-  \citet{mayer+2001} theory of tidal stripping
-\item
-  \citet{IH1995} structural parameters
+  \citet{dicintio+2013}, \citet{navarro+2010}, NFW profiles may actually
+  be more Einasto like
 \end{itemize}
 
-\subsection{Cosmological context}\label{cosmological-context}
+\section{Simulating tidal effects}\label{simulating-tidal-effects}
 
-\begin{figure}
-\centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/power_spectrum.png}
-\caption[Cosmological Power Spectrum]{The matter power spectrum under
-different assumptions for dark matter. Dwarf galaxies occupy the middle
-and low end of the blue region (10\^{}10 - 10\^{}8 solar masses),
-enabling a unique window into properties of dark matter on small scales.
-The smaller scales we can understand dark matter, the better we are able
-to test different models of dark matter. figure 1 from
-\citet{bechtol+2022}.}\label{fig:cosmological_power_spectrum}
-\end{figure}
+Since the discovery of dwarf galaxies, a large body of work has
+speculated, considered, or simulated tidal effects. However, before
+\emph{Gaia}, the orbits of most systems were essentially unconstrained.
+The theory developed still relates to today's work.
 
-Figure: Density profiles of comological simulated halos, matching
-approximantly the NFW formula.
+Because we only have finite compute resources, major assumptions or
+limitations are common in simulations. Cosmological simulations,
+starting from the very early universe, create a realistic environment,
+structure formation, and merger history. But, cosmological simulations
+typically barely resolve dwarf galaxies {[}REFS{]}, leading to numerical
+artifacts and artificial disruption. On the other hand, idealized
+simulations are able to reach high resolution and numerical convergence
+for a single dwarf galaxy. But, idealization may not be the most
+realistic environment, neglecting mergers, assembly history, and often
+baryonic physics. We use idealized simulations here which are more
+powerful in accurately assessing tidal effects, given that the
+idealizations do not impact these galaxies's recent history too much.
 
-\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}Figure:
-Image of the Sculptor dwarf spheroidal galaxy from Dark Energy Survey
-Data Release 2 {[}\citet{abbott+2021}; image created with HiPS2FITS{]}.
-Sculptor appears as a fairly prominent, extended over density of
-predominantly faint, red stars. 0.5 degree field of view centred on
-Sculptor.
+\subsection{Simulating large gravitational systems: The N-body
+method}\label{simulating-large-gravitational-systems-the-n-body-method}
 
-\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}Figure:
-Image of Ursa Minor dwarf galaxy from the Digitized Sky Survey 2 (0.75
-deg field of view tangent plane). UMi appears as a diagonal/elliptical
-haze of faint, reddish stars from the top left to the bottom right. Even
-as classical dwarf, Ursa minor is fairly diffuse and does not stand
-obviously out from the background.
+While the dominant force on large scales in the universe, calculating
+the gravitational interaction between a large number of objects becomes
+computationally prohibitive. So, how do we manage to create simulations
+with billions of particles across vast volumes of the universe?
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_vs_penarrubia.pdf}}
-\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
-profiles are well-matched to \citet{PNM2008}.}\label{fig:toy_profiles}
-\end{figure}
+Perhaps the simplest method to compute the evolution of dark matter is
+through \emph{N-body simulations}. A dark matter halo is represented as
+a large number of dark matter particles (bodies).
 
-\subsection{Theoretical Background}\label{theoretical-background}
+Naively, the newtonian gravitational force requires adding together the
+forces from each particle at each particle, causing a computational cost
+that scales quadratically with the number of particles, or \(O(N^2)\).
+With this method, simulating a large number of particles, such as
+10\^{}6, would require about 10\^{}12 force evaluations at each time
+step, making cosmological and high-resolution studies unfeasible.
+However, only long-range gravitational interactions tend to be important
+for CMD, so we can utilize the \emph{tree method} to compute the
+gravitational force vastly more efficiently.
+
+The first gravitational tree code was introduced in
+\citet{barnes+hut1986}, and is still in use today. We utilize the
+massively parallel code \emph{Gadget 4} \citep{gadget4}. Particles are
+spatially split into an \emph{octotree}. The tree construction stars
+with one large node, a box containing all of the particles. If there is
+more than one particle in a box/node , the box is then divided into 8
+more nodes (halving the side length in each dimension) and this step is
+repeated until each node only contains 1 particle. With this
+heirarchichal organization, if a particle is sufficiently far away from
+a node, then the force is well approximated by the force from the centre
+of mass of the node. As such, each force calculation only requires a
+walk through the tree, only descending farther into the tree as
+necessary to retain accuracy. The total force calculations reduce from
+\(N^2\) to \(O(N\,\log N)\), representing orders of magnitude speedup.
+Modern codes such as \emph{Gadget} utilize many other performance
+tricks, such as splitting particles across many supercomputer nodes,
+efficient memory storage, adaptive time stepping, and parallel file
+writing to retain fast performance for large scale simulations, forming
+the foundation for many cosmological simulation codes.
+
+\subsection{Galactic tides and idealized
+simulations}\label{galactic-tides-and-idealized-simulations}
+
+Despite the power of cosmological simulations for understanding large
+scale structure, a fundamental limitation is their ability to resolve
+small scales such as dwarf galaxies. Dwarf galaxies, without sufficient
+resolution, are likely to artificially disrupt as a result of numerical
+noise. For understanding the tidal evolution of a dwarf galaxy, this
+tool has limited accuracy then.
+
+Idealized simulations take a different approach. Instead of starting
+from the universe as a whole, just a single subhalo is simulating,
+allowing resolution several orders of magnitude than cosmological
+simulations. For instance, the highest resolution dark matter only
+simulation, the \emph{Aquarious Project} only reached a mass resolution
+of XXX. In constrast, our dark matter mass resolution in later chapters
+is of order 1000s of solar masses.
+
+As such, a long history of work has resolved to understand the tidal
+evolution of dwarf galaxies using these \emph{idealized} simulations.
+
+Some of the earliest theory work on tidal stripping of dwarf galaxies
+originate from \citep[\citet{moore+davis1994},
+\citet{johnston+spergel+hernquist1995}, \citet{oh+lin+aarseth1995},
+\citet{piatek+pryor1995}, \citet{velazquez+white1995},
+\citet{kroupa1997}]{allen+richstone1988}. Many of these works used
+similar techniques as we continue to use today.
+
+More recently, \citet{PNM2008} run idealized simulations of a local
+group dwarf galaxy in a milky way halo.
 
@@
 \item
-  Cosmology foundations
-
-  \begin{itemize}
-  \tightlist
-  \item
-    power spectrum plot
-  \end{itemize}
-\item
-  NFW plot (density or energy), maybe borrow from paper?
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Explain origin of NFW
-  \end{itemize}
-\item
-  Dwarf galaxy formation, halos
+  \citet{mayer+2001} theory of tidal stripping
 \end{itemize}
 
-N-body DM simulations
+\subsubsection{Break and tidal radii}\label{break-and-tidal-radii}
 
-\begin{itemize}
-\tightlist
-\item
-  Collisionless Boltzmann equation and meaning of such simulations
-\item
-  Assumptions \& context \& past work
-\item
-  Evolution under tidal field
-\end{itemize}
+Analytic approximations are powerful tools to generally determine the
+influence of tides given a dwarf galaxy and a host.
 
-To motivate why a tidal interaction may give rise to the observed
-density profiles, we create a toy simulation following \citet{PNM2008}.
+The Jacobi radius represents the approximate radius where stars become
+unbound for a galaxy in a circular orbit around a host galaxy.
+Calculated from an approximation of the location of the L1 and L2
+lagrange points, the Jacobi radius is where the mean density of the
+dwarf galaxy is three times the mean interior density of the host galaxy
+at pericentre, or \begin{equation}{
+3\bar \rho_{\rm MW}(r_{\rm peri}) = \bar \rho_{\rm dwarf}(r_J).
+}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
+we should expect to find unbound, \emph{extratidal} stars. While most
+valid for circular orbits, assuming \(r_{\rm peri}\) for the host-dwarf
+distance works as most stars are lost near pericentre.
 
-\begin{itemize}
-\item
-  NFW initial conditions (sculptor like, vcm, rcm)
-\item
-  Evolved in x-y plane using \citet{EP2020} potential for
-  \textasciitilde{} 5Gyr with pericentre of 15 kpc and apocentre of 100
-  kpc.
-\item
-  Exponential initial stellar profile.
-\end{itemize}
-
-As a dark matter halo is perturbed on a pericentric passage with the
-milky way,
-
-\begin{itemize}
-\tightlist
-\item
-  Tidal stress heats halo slightly
-\item
-  Mass loss, particularly of loosely bound particles
-\end{itemize}
-
-The stellar component tracers will similarly follow the behaviour of the
-dark matter.
-
-An emperical estimate of where the simulation's stars are becoming
-unbound is, as stated in \citet{PNM2008}, the break radius
-\begin{equation}
-R_b = C\,\sigma_{v}\,\Delta t
-\end{equation} where \(\sigma_v\) is the present line of sight velocity
-dispersion , \(\delta t\) is the time since pericentre, and
-\(C \approx 0.55\) is a fit. The idea motivating this equation is stars
-in the inner regions will have dynamically equilibriated to the new
-potential (phase mixed), however the outer regions are no longer in
-steady state, so we have to wait until the crossing time reaches them as
-well.
-
-As illustrated in fig.~\ref{fig:toy_profiles}, the density profile
-initially stars off exponential. At increasing times since the first
-pericentric passage, the break radius, appearing as an apparent
-separation between the slopes of the inner and outer profile, increases.
-
-\subsection{Introduction to Dark Matter
-simulations}\label{introduction-to-dark-matter-simulations}
-
-In this section, we will cover
-
-\begin{itemize}
-\tightlist
-\item
-  How are dark matter simulations conducted
-\item
-  Interpretations and uncertainties
-\item
-  Methods
-\end{itemize}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
-\caption[Break radius validation]{The break radius of the simulations is
-set by the time since pericentre.}\label{fig:idealized_break_radius}
-\end{figure}
+We use the break radius as defined in \citet{penarrubia+2009}, marking
+where the galaxy is still in disequilibrium. The break radius
+\(r_{\rm break}\) is proportional to the velocity dispersion
+\(\sigma_v\) and time since pericentre \(\Delta t\), \begin{equation}{
+r_{\rm break} = C\,\sigma_{v}\,\Delta t
+}\end{equation} where the scaling constant \(C \approx 0.55\) is a fit.
+\(r_{\rm break}\) describes where the dynamical timescale is longer than
+the time since the perturbation, i.e.~the radius within which the galaxy
+should have dynamically relaxed. As illustrated in
+Fig.~\ref{fig:toy_profiles}, for an idealized model with exponential
+stars in a NFW halo, after some time past pericentre, the stellar
+component is smooth but contains a change in slope around
+\(r_{\rm break}\), where the radial velocities of the stars becomes
+positive as they still readjust to a new equilibrium.
 
@@
 
-\subsection{Potentials}\label{potentials}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
+\caption[Break radius validation]{The break radius of the simulations is
+set by the time since pericentre. The initial conditions are
+Sculptor-like, exponential stars embedded in NFW, evolved in
+\citet{EP2020} potential with a pericentre of 15 kpc and apocentre of
+100 kpc. See Section~\ref{sec:methods} for a description of our
+simulation setup. \textbf{TODO:} mark \(r_J\), plot initial, make sure
+1st pericentre, simplify models.}\label{fig:idealized_break_radius}
+\end{figure}
 
-From the \citet{nfw1996} paper, eqns. 3 \& 4 \begin{equation}
-\frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
-\end{equation} where \begin{equation}
-\delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
-\end{equation} \(c\) is concentration parameter, \(\rho_c\) is the
-critical density of the universe, and \(r_s\) is the characteristic
-scale length of the halo.
+\subsubsection{Tidal evolution}\label{tidal-evolution}
 
-The NFW halo is sometimes described by \(M_{200}\). \(r_{200}\) is the
-radius at which the mean density of the halo interior to \(r_{200}\) is
-200 times the critical density of the universe, and \(M_{200}\) is the
-mass contained inside \(r_{200}\). As equations: \begin{equation}
-\rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
-\end{equation}
+While the break and jacobi radii help with understanding when tidal
+effects matter, we still need to know what these effects do and how a
+galaxy evolves.
 
-\begin{equation}
-r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
-}}}
-\end{equation}
+As a galaxy evolves in a tidal field, several changes happen.
 
-\begin{equation}
-M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
-\end{equation}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \emph{Mass is lost}. In particular, particles and stars on weakly
+  bound orbits are most likely to be removed by tides.
+\item
+  \emph{Steams form}. If tides are strong enough, mass lost from the
+  galaxy becomes part of tidal tails, evolving along a similar orbit but
+  leading or trailing the galaxy.
+\item
+  \emph{The galaxy reshapes}. Bound mass of the galaxy redistributes. As
+  mentioned in the last section, this is visible as a wave of outward
+  moving material, with the outermost material reaching equilibrium
+  last.
+\item
+  \emph{A new equilibrium}. With mass loss, the gravitational potential
+  decreases, resulting in a more compact dark matter halo and stars
+  which adiobadically expand to a larger scale radius.
+\end{enumerate}
 
-\(M_{200}\) is also sometimes called the virial mass of the halo.
-\(r_{200}\) is directly related to \(r_s\) by \begin{equation}
-r_{200} = c\,r_s
-\end{equation}
+Assuming that galaxies are spherical, isotropic, and evolve in a
+constant tidal field, most dwarf galaxies should evolve along similar
+tidal tracks.
 
-Another useful definition is \begin{equation}
-A(x) \equiv \log (1+x) - \frac{x}{1+x}.
-\end{equation}
+\citet{EN2021} derive tidal tracks for dwarf galaxies, showing that NFW
+halos all evolve along a similar trajectory in terms of \(r_{\rm max}\)
+and \(v_{\rm max}\). \begin{equation}{
+\frac{v_{\rm max}}{v_{\rm max, 0}} = 
+2^\alpha 
+\left(\frac{r_{\rm max}}{r_{\rm max, 0}}\right)^{\beta}\left[1 + \left(r_{\rm max} / r_{\rm max, 0}\right)^2\right]^{-\beta}
+}\end{equation} \{eq:tidal\_track\} where \(\alpha=0.4\), \(\beta=0.65\)
+are empirical fits. As illustrated in Fig.~\ref{fig:tidal_tracks}, this
+formula works for both circular and elliptical orbits and is independent
+of the initial subhalo size or distance to the host.
 
-We will also define a dimensionless radius \begin{equation}
-x \equiv r/r_s.
-\end{equation}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}}
+\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies.
+Fig. 6 from \citet{EN2021}.}\label{fig:tidal_tracks}
+\end{figure}
 
-A simple substitution to the definition gives
+\subsection{Notable related works}\label{notable-related-works}
 
-\begin{equation}
-\rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
-\end{equation} where \begin{equation}
-\rho_s = \frac{c^3}{A(c)} \rho_{200}
-\end{equation} and \(A(c)\) is as above. The characteristic density can
-also be written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see
-below), giving \begin{equation}
-\rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
-\end{equation} Note that the NFW density profile is the same as an
-alpha-beta-gamma profile where \(\alpha=\gamma=1\) and \(\beta =3\).
+Case studies:
 
-\subsubsection{Circular velocity}\label{circular-velocity}
+\begin{itemize}
+\tightlist
+\item
+  \citet{dicintio+2024}
+\item
+  \citet{iorio+2019}
+\end{itemize}
 
-The circular velocity in terms of
-\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \begin{equation}
-\left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
-\end{equation}
+Dynamical
 
-or in terms of \(M_s\) and \(r_s\), \begin{equation}
-v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
-\end{equation} Another parameterization of the NFW profile is in terms
-of the maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the
-radius at which it is reached \(r_{\rm circ}^{\rm max}\). Given the
-scale radius, \begin{equation}
-r_{\rm circ}^{\rm max} = \alpha\ r_s
-\end{equation}
+\begin{itemize}
+\tightlist
+\item
+  \citet{read+2006}
+\item
+  \citet{bullock+johnston+2005}
+\item
+  \citet{PNM2008}
+\item
+  \citet{borukhovetskaya+2022a}, \citet{borukhovetskaya+2022}
+\item
+  \citet{errani+2023a}, \citet{fattahi+2018}
+\item
+  \citet{wang+2017},
+\item
+  \citet{pryor1996}
+\item
+  \citet{klimentowski+2009}
+\end{itemize}
 
-where \(\alpha\approx2.16258\), and \(v_{\rm circ}^{\rm max}\) can be
-found from either of the equations above.
+\section{Thesis Outline}\label{thesis-outline}
+
+In chapter 2, we describe how we compute observational density profiles
+from \citet{jensen+202}. In chapter 3, we review our simulation methods.
+Next, we present our results for the tidal stripping of Sculptor and
+Ursa Minor in Chapter 4. We discuss our results, limitations, and
+implications in Chapter 5. Finally, in Chapter 6, we summarize this work
+and discuss future directions for similar work and the field of dwarf
+galaxies.

### 2025-07-15 thesis/methods.tex
major thesis restructure

--- 2025-05-02 thesis/methods.tex
@@
-In this section, we discuss our simulation methods from orbital
-estimation and initial conditions to post-processing and software.
+In this section, we discuss our methods used for the following two
+chapters. We utilize a typical analytic, idealized Milky Way potential
+based on \citet{mcmillan2011}. Our simulations are collisionless cold
+dark matter N-body simulations integrated with Gadget-4. We check for
+numerical convergence and stability of initial conditions in isolation.
 
-We use Agama \citep{agama} for potential specification and initial
-conditions, and Gadget-4 \citep{gadget4} for orbit integration and dark
-matter simulations (with a custom patch to integrate with Agama).
+\section{Milky Way potential}\label{milky-way-potential}
 
-\section{Milky Way Potential}\label{milky-way-potential}
+We adopt the Milky Way potential described in \citet{EP2020}, an
+analytic approximation of \citet{mcmillan2011}.
+Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
+each component of our fiducially potential. The potential consisting of
+a stellar Bulge, a thin and thick disk, and a dark matter NFW halo.
+Especially in the regime where Sculptor and Ursa Minor orbit, the dark
+matter halo is the dominant component of the potential. The details of
+the stellar disk and bulge are less important.
 
-fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
-each component of our fiducially potential. We adopt the potential
-described in \citet{EP2020}, an analytic approximation of
-\citet{mcmillan2011}. The potential consisting of a stellar Bulge, a
-thin and thick disk, and a dark matter NFW halo. Especially in the
-regime where Sculptor and Ursa Minor orbit, the dark matter halo is the
-dominant component of the potential. The details of the stellar disk and
-bulge are less important.
+The potential is specified as follows. The galactic bulge is described
+by a \citet{hernquist1990} potential. The potential is
 
-The galactic bulge is described by a \citet{hernquist1990} potential.
-The potential is
+\begin{equation}\protect\phantomsection\label{eq:hernquist}{
+\Phi(r) = - \frac{GM}{r + a}
+}\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
+\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
+disks are represented with the \citet{miyamoto+nagai1975} cylindrical
+potential:
 
-\begin{equation}
-\Phi(r) = - \frac{GM}{r + a}
-\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
-\(M=2.1 \times 10^{10}\,\Mo\) is the total mass.
-
-The thin and thick disks are represented with the
-\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}
+\begin{equation}\protect\phantomsection\label{eq:mn75}{
 \Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
-\end{equation}
+}\end{equation}
 
@@
 \(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
-
-The halo is a NFW dark matter halo (REF) with
+The halo is a NFW dark matter halo (Eq.~\ref{eq:nfw}) with
 \(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.
-
-Variations to the potential of the inner disk (exclusion of a bar)
-should minimally affect our results as no orbit we consider reaches less
-than \textasciitilde15 kpc of the MW centre. We exclude the mass
-evolution of the halo from this analysis. Over \(10\,\)Gyr, this would
-be fairly significant (factor of \(\sim 2\)in MW mass, REF) but since we
-want to determine the upper limit of tidal effects, it is safe to
-neglect this.
-
-Finally, in chapter REF, we consider the influence of the large Milky
-Cloud on the orbits and evolution of Scl. We adopt the
-\citet{vasiliev2024} multipole approximation of an N-body simulation of
-the LMC and MW. Their initial conditions are
-
-\begin{itemize}
-\tightlist
-\item
-  MW halo:
-\item
-  MW bulge (static):
-\item
-  MW disk (static):
-\item
-  LMC halo:
-\end{itemize}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.png}}
 \caption[Circular velocity of potential]{Circular velocity profile of
-\citet{EP2020} potential. We also show the MW and LMC potential from
-\citet{vasiliev2024} for their L3M11 model.}\label{fig:v_circ_potential}
+\citet{EP2020} potential.}\label{fig:v_circ_potential}
 \end{figure}
 
-\section{Orbital Estimation}\label{orbital-estimation}
+\section{Orbital estimation}\label{orbital-estimation}
 
-To estimate the orbits of a dwarf galaxy, we perform a Monte Carlo
-sampling of the present-day observables. We integrate each sampled
-observable back in time for 10 Gyr in Gadget as massless point particles
-outputting positions every 5Myr (with otherwise similar parameters to
-n-body runs below.)
+To estimate the possible orbits of a dwarf galaxy, we perform a Monte
+Carlo sampling of the present-day observables. The present-day position,
+distance modulus, LOS velocity, and proper motions are each sampled from
+normal distributions given the reported uncertainties in
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
+each sampled observable back in time for 10 Gyr in Gadget as massless
+point particles outputting positions every 5Myr (with otherwise similar
+parameters to n-body runs below.) When selecting the initial position
+and velocity of an N-body model, we select the position and velocity of
+the first apocentre occurring after 10 Gyr ago.
 
-To convert from Gaia to galactocentric coordinates, we use the
-\citet{astropy} v4 galactocentric frame. This frame assumes the galactic
-centre (Sagittarius A*) appears:
+We assume the galaxy is represented as a point particle for this
+analysis. In detail, (self) dynamical friction likely influences the
+orbit, but a point particle is an adequate description.
 
-\begin{itemize}
-\tightlist
-\item
-  J2000 `ra' = 17h45m37.224s, `dec' = -28°56'10.23'',
-  \(\mu_{\alpha*}=-3.151\pm0.018\) mas/yr, \(\mu_\delta=-5.547\pm0.026\)
-  from the appendix and Table 2 of \citet{reid+brunthaler2004}.
-\item
-  distance = \(8.122\pm0.033\,\)kpc, solar radial velocity =
-  \(11 + 1.9 \pm 3\) km/s. from \citet{gravitycollaboration+2018}.
-\end{itemize}
+To convert from Gaia to galactocentric coordinates, we use the Astropy
+v4 galactocentric frame \citep{astropycollaboration+2022}. This frame
+assumes the galactic centre appears at position
+\(\alpha = {\rm 17h45m37.224s}\), \(\delta = -28°56'10.23''\) with
+proper motions \(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
+\(\mu_\delta=-5.547\pm0.026 \masyr\) (from the appendix and Table 2 of
+\citet{reid+brunthaler2004}). The galactic centre distance is
+\(8.122\pm0.033\,\)kpc with a radial velocity = \(11 + 1.9 \pm 3\) km/s.
+from \citet{gravitycollaboration+2018}. Finally, adding that the sun is
+\(20.8\pm0.3\,\)pc above the disk from \citet{bennett+bovy2019}, and
+using the procedure outlined in \citet{drimmel+poggio2018}, the solar
+velocity relative to the galactic rest frame is `v\_sun' =
+\([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
+uncertainties in the reference frame are typically smaller than the
+uncertainties on the dwarf galaxy's position and velocity.
 
-Finally, adding that the sun is \(20.8\pm0.3\,\)pc above the disk from
-bennett+bovy2019, and using the procedure described in
-\citet{drimmel+poggio2018}, the solar velocity relative to the galactic
-rest frame is
-
-\begin{itemize}
-\tightlist
-\item
-  `v\_sun' = \([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s
-\end{itemize}
-
-We select the initial conditions from the location of the first
-apocentre of a selected orbit.
-
-While we assume that the galaxy can be described as a point particle,
-not subject to (internal) dynamical friction, this approximation works
-well if the galaxy is not strongly affected by tides (like Scl).
-Analytic approximations of dynamical friction tend to be inadequate to
-more accurately model the N-body trajectory.
-
-\section{N-Body Modelling}\label{n-body-modelling}
+\section{N-Body modelling}\label{n-body-modelling}
 
@@
 We use Agama \citep{agama} to generate initial conditions. We initially
-assume galaxies are described by an NFW dark matter potential (REF) and
-the stars are merely collisionless tracers embedded in this potential
-(added on in post-processing). The density is exponentially truncated
-with a profile
+assume galaxies are described by an NFW dark matter potential
+Eq.~\ref{eq:nfw} and the stars are merely collisionless tracers embedded
+in this potential (added on in post-processing). The density is a
+cubic-exponentially truncated with a profile
 \begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
-\rho_{\rm tNFW} = e^{-r/r_t}\ \rho_{\rm NFW}(t)
-}\end{equation} where we adopt \(r_t = 100 r_s\) or approximately 10
-times \(r_{200}\).
+\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t)
+}\end{equation} where we adopt \(r_t = 20 r_s\) or approximately
+\(r_{200}\) for our Sculptor-like fiducial halo. Using \(r_{200}\) for
+\(r_t\) would depend on the chosen scale of the halo. So our adopted
+\(r_t\) is an approximate upper limit of \(r_{200}\) for typical dwarf
+galaxy halos \citep{ludlow+2016}. It is unlikely that the outer density
+profile of loosely bound particles past \(20\) kpc affects the tidal
+evolution of a subhalo.
 
@@
 relaxed and well-converged, we run the simulation in isolation (no
-external potential) for 5 Gyr (or approximately the crossing timescale).
-Our fiducial isolation halo uses \(r_s=2.76\) kpc and
-\(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for any
-length or mass scale.
+external potential) for 5 Gyr (or about 3 times the crossing timescale
+at the virial radius). Our fiducial isolation halo uses \(r_s=2.76\) kpc
+and \(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for
+any length or mass scale.
 
-We adopt a softening length of \begin{equation}
+For our simulation parameters, we adopt a softening length of
+\begin{equation}{
 h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-\end{equation} See appendix REF for a discussion of this choice, which
-is similar to the \citet{power+2003} suggested softening.
-
-We use the relative tree opening criterion with the accuracy parameter
-set to 0.005, and adaptive time stepping with integration accuracy set
-to 0.01.
+}\end{equation} See appendix Section~\ref{sec:extra_convergence} for a
+discussion of this choice, which is similar to the \citet{power+2003}
+suggested softening. We use the relative tree opening criterion with the
+accuracy parameter set to 0.005, and adaptive time stepping with
+integration accuracy set to 0.01.
 
@@
 collisionless assumption breaks down). The relaxation timescale is given
-by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise,
-\begin{equation}
+by \begin{equation}\protect\phantomsection\label{eq:t_relax}{
 t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
 = {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
-\end{equation} This works out to be about 6-10 times (increasing with
-particle number) our adopted softening length for NFW halos given our
-assumptions. As such, at full resolution, we can only trust density
-profiles down to \(\sim10\epsilon\), just enough to resolve stellar
-density profiles.
-
-Note that by decreasing the truncation radius, the effective particle
-number is improved, so future experiments could be less generous with
-this parameter to improve computational performance (order 30\% better
-resolution).
+}\end{equation} where \(N(r)\) is the number of particles within \(r\),
+\(t_{\rm circ}\) is the circular velocity timescale, and \(\bar \rho\)
+is the mean interior density \citep{power+2003}. The converged radius
+\(r_{\rm relax}(t)\) is the radius above which \(t_{\rm relax} > t\).
+Typically, \(r_{\rm relax}(10\Gyr)\) is about 6-10 times our adopted
+softening length, increasing with particle number. As such, at full
+resolution, we can only trust density profiles down to \(\sim10\) times
+our softening length, sufficient to resolve stellar density profiles.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.png}}
 \caption[Numerical halo convergence]{Numerical convergence test for
 circular velocity as a function of log radius for simulations with
-different total numbers of particles in isolation. Residuals in lower
+different total numbers of particles in isolation. Residuals in bottom
 panel are relative to NFW. The initial conditions are dotted and the
-converged radius is marked by arrows
-(REF).}\label{fig:numerical_convergance}
+converged radius is marked by arrows (Eq.~\ref{eq:t_relax}). Note that a
+slight reduction in density starting around \$r = 30 \$kpc is expected
+given our truncation choice.}\label{fig:numerical_convergance}
 \end{figure}
@@
 To perform the simulations of a given galaxy in a given potential, we
-centre the isolation run's final snapshot and place the dwarf galaxy in
-the specified orbit in the given potential. We typically run the
-simulation for 10 Gyr, which allows us to orbit slightly past the
-expected initial conditions.
+centre the isolation run's final snapshot
+(Section~\ref{sec:shrinking_spheres}) and place the dwarf galaxy in the
+specified orbit in the given potential. We typically run the simulation
+for 10 Gyr, which allows us to orbit slightly past the expected initial
+conditions.
 
-\section{Post Processing}\label{post-processing}
+\section{Post processing}\label{post-processing}
 
-\subsection{Centring}\label{centring}
+\subsection{Centring}\label{sec:shrinking_spheres}
 
@@
   shown for every 200th particle (ranked from the centre outwards)
-\end{itemize}
-
-\begin{equation}
-v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
-\end{equation}
-
-\begin{itemize}
-\tightlist
 \item
@@
 
-\subsection{Stellar Probabilities}\label{stellar-probabilities}
+\subsection{Stellar probabilities}\label{stellar-probabilities}
 
-We assign stellar probabilities (assuming spherical, isotropic symmetry)
-via Eddington inversion.
+We ``paint'' stars onto dark matter particles using the particle tagging
+method \citep[e.g.][]{bullock+johnston2005}, assuming spherical
+symmetry. Let \(\Psi\) be the potential (normalized to vanish at
+infinity) and \({\cal E}\) is the binding energy
+\({\cal E} = \Psi - 1/2 v^2\). If we know \(f({\cal E})\), the
+distribution function (phase-space density in energy), then we assign
+the stellar weight for a given particle with energy \({\cal E}\) is
 
-Let \(\Psi\) be the potential (normalized to vanish at infinity) and
-\({\cal E}\) is the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
-know \(f({\cal E})\), the distribution function (phase-space density in
-energy), then we assign the stellar weight for a given particle with
-energy \({\cal E}\) is \begin{equation}
+\begin{equation}{
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-\end{equation} While \(f({\cal E})\) is a phase-space density, the
+}\end{equation} While \(f({\cal E})\) is a phase-space density, the
 differential energy distribution includes an additional \(g({\cal E})\)
-occupation term (BTXXX).
+occupation term (BT87). We use Eddington inversion to find the
+distribution function, (eq. 4-140b in BT87)
 
-We use Eddington inversion to find the distribution function, (eq.
-4-140b in BT87) \begin{equation}
+\begin{equation}{
 f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-\end{equation}
+}\end{equation}
 
@@
 \(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\) which goes to
-zero provided that \(n > 1\).
-
-\(\Psi\) is taken as known from the underlying assumed analytic dark
-matter potential. \(\rho_\star\) is pre-specified or calculated via Abel
-integration / deprojection
+zero provided that \(n > 1\). We take \(\Psi\) from the underlying
+assumed analytic dark matter potential. \(\rho_\star\) is pre-specified
+or calculated via Abel integration / deprojection.
 
@@
 
-The 2D exponential stellar profile is given by \begin{equation}
+Perhaps the simplest in form, the 2D exponential is commonly used to
+describe dwarf galaxy's density profiles as well as stellar disks. The
+stellar profile is given by \begin{equation}{
 \Sigma_\star(R) = A e^{-R / R_s}
-\end{equation} where \(A\) is some normalization and \(R_s\) is the
-exponential scale radius. The 3D deprojected profile is found through
-abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
-\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
-\end{equation} where \(K_0\) is the 0th order modified Bessel function
-of the second type.
+}\end{equation} where \(A\) is the normalization and \(R_s\) is the
+exponential scale radius. \#\#\# Plummer
 
-\subsubsection{Plummer}\label{plummer}
-
-A Plummer profile is defined by \begin{equation}
+The Plummer density profile is generally considered to be a reasonable
+empirical fit to Globular Clusters, where it was first proposed
+\citep{sadfjk}. Plummer profiles are also sometimes used to describe
+dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
+profile is defined by \begin{equation}{
 \Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
-\end{equation} where \(M\) is the total mass and \(R_s\) is the
-characteristic scale radius. The density is \begin{equation}
-\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
-\end{equation}
+}\end{equation} where \(M\) is the total mass and \(R_s\) is the
+characteristic scale radius.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Stellar velocity dispersion is calculated for all stars within 1kpc
-  (3D) of the centre by assuming
+Stellar velocity dispersion is calculated for all stars within a 1kpc
+sphere of the centre. We assume
 \(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
-  i.e.~isotropic velocity dispersion
-\item
-  We calculate density profiles similar to stars and assume a constant
-  bin width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\)
-  (Freedman-Diaconis prescription). \(R\) may be either the cylendrical
-  radius in \(x-y\) or the on-sky project \(\xi, \eta\) tangent plane
-  coordinates.
-\item
-  Because our dwarfs are assumed to be spherical/isotropic, we retain
-  this assumption when calculating predicted density profiles.
-\end{itemize}
+i.e.~isotropic velocity dispersion.
+
+We calculate density profiles similar to stars and assume a constant bin
+width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\) (Freedman-Diaconis
+prescription). \(R\) may be either the cylendrical radius in \(x-y\) or
+the on-sky project \(\xi, \eta\) tangent plane coordinates.
+
+Because our dwarfs are assumed to be spherical/isotropic, we retain this
+assumption when calculating predicted density profiles.

### 2025-07-15 thesis/numerical_convergence.tex
major thesis restructure

@@
+\chapter{Numerical convergence and
+parameters}\label{sec:extra_convergence}
+
 Here, we describe some convergence tests to ensure our methods and
@@
 
-\subsubsection{Softening}\label{softening}
+\section{Softening}\label{softening}
 
-\citet{power+2003} suggest the empirical rule that the ideal softening
-(balancing integration time and only compromising resolution in
-collisional regime) is \[
-h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
-\]
+One challenge of N-body integration is close, collisional encounters,
+assuming point particles, cause divergences in the local force. However,
+this should not occur in a \emph{collisionless} simulation. As a result,
+most collisionless N-body codes adopt a gravitational softening, a
+length scale below which the force of gravity begins to decrease between
+point particles.
+
+\citet{power+2003} empirically suggest that the ideal softening is
+\begin{equation}{
+h_{\rm grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}},
+}\end{equation}
+
+where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
+number of particles within \(R_{200}\). This choice balances integration
+time and only compromises resolution in collisional regime.
 
@@
 \(\sqrt{10}\) which appears to improve agreement slightly in the
-innermost regions. However, this choice likely unnecessarily increases
-computation time for the relative gain in accuracy.
+innermost regions.
 
@@
 
-\subsubsection{Timestepping and force
-accuracy}\label{timestepping-and-force-accuracy}
+\section{Time stepping and force
+accuracy}\label{time-stepping-and-force-accuracy}
 
@@
 
-\subsubsection{Alternative methods}\label{alternative-methods}
-
-\begin{itemize}
-\tightlist
-\item
-  FMM
-\item
-  PMM-tree
-\item
-  Gadget2
-\item
-  etc.
-\end{itemize}
-
 \begin{figure}
@@
 
-\subsubsection{Fiducial Parameters}\label{fiducial-parameters}
+\section{Fiducial parameters}\label{fiducial-parameters}
 

### 2025-07-15 thesis/rv_models.tex
major thesis restructure

\chapter{Radial velocity modeling}\label{sec:rv_obs}

\section{Data selection}\label{data-selection}

For both Sculptor and Ursa Minor, we construct literature samples of
radial velocity measurements. We combine these samples with J+24's
members to produce RV consistent stars and to compute velocity
dispersion, systematic velocities, and test for the appearance of
velocity gradients.

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1-3
arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
measurement from each study using the inverse-variance weighted mean
\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\). We remove stars with significant velocity dispersions
as measured between observations in a study or between studies. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically).

Next, we need to correct the coordinate frames for the solar motion and
on-sky size of the galaxy. We transform the frame into the galactic
standard of rest (GSR). The next step is to account for the slight
differences in the direction of each radial velocity. Let the \(\hat z\)
be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
the angular distance between the centre of the galaxy and the individual
star, the corrected radial velocity is then \begin{equation}{
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
velocities in RA and Dec, and \(\theta\) is the position angle of the
star with respect to the centre of the dwarf. The correction from both
effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We then use the \(v_z\) values for the
following modelling, however repeating with uncorrected, heliocentric
velocities does not significantly affect the results.

The combined likelihood, including RV information, becomes
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
}\end{equation} where we assume that the satellite and background
distributions are Gaussian. Specifically, \begin{equation}{
\begin{split}
{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
will dominate the uncertainty budget here. We assume a halo/background
velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

Similar to above, we retain stars with the resulting membership
probability of greater than 0.2. Because of the additional information
from radial velocities, most stars have probabilities close to 1 or 0 so
the probability cut is not too significant.

We assume priors on systematic velocity and velocity dispersion of
\begin{equation}{
\begin{split}
\mu_{v} &= N(0\,\kms, \sigma_{\rm halo}^2) \\ 
\sigma_{v} &= U(0, 20\,\kms)
\end{split}
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above, a reasonable
assumption for dwarfs in orbit around the MW.

\section{Results}\label{sec:rv_results}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
terms of \(v_z\) (REF). Orange points are from our crossmatched RV
membership sample.}
\end{figure}

For Sculptor, we combine radial velocity measurements from APOGEE,
\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
measurements. We find that there is no significant velocity shift in
crossmatched stars between catalogues. After crossmatching to high
quality Gaia stars and excluding significant stellar velocity
dispersions, we have a sample of 1918 members.

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
detailed comparison between different samples and additional tests.

We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
Several past work has attempted to detect a gradient in Sculptor, but no
consensus has been reached. \citet{arroyo-polonio+2024} detect a
velocity gradient of \(4\pm1.5\,\kmsdeg\) in a similar direction using
the \citet{tolstoy+2023} sample, finding inconclusive statistical
evidence. They additionally suggest a third chemodynamical component of
the galaxy which may bias rotation measurements. \citet{battaglia+2008}
also detect a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along
the major axis, approximately the same direction. Instead,
\citet{strigari2010}; \citet{martinez-garcia+2023} detect no significant
gradient in Sculptor using \citet{WMO2009} sample. Note that
pre-\emph{Gaia} work did not have as strong of a constraint on the
proper motion of Scl, which limits conclusions of the intrinsic velocity
gradient in Scl.

For UMi, we collect radial velocities from, APOGEE,
\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
the median difference in velocities in this crossmatch is about 2.2
km/s, we adopt 1 km/s as the approximate systematic error here. Our
final sample includes 831 members.

We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
do not find evidence for a velocity gradient, consistent with past work
\citep{pace+2020, martinez-garcia+2023}.

\section{Discussion and limitations}\label{discussion-and-limitations}

Our model here is relatively simple. Some things which we note as
systematics or limitations:

\begin{itemize}
\tightlist
\item
  Inter-study systematics and biases. While basic crossmatches and a
  simple velocity shift, combining data from multiple instruments is
  challenging. This appears to be a minor issue (Sculptor) or is
  corrected for (Ursa Minor).
\item
  Misrepresentative uncertainties. Inspection of the variances compared
  to the standard deviations within a study seems to imply that errors
  are accurately reported. APOGEE notes that their RV uncertainties are
  known to be underestimates but are a small proportion of our sample.
\item
  Binarity. While not too large of a change for classical dwarfs, this
  could inflate velocity dispersions of about \(9\,\kms\) by about
  \(1\,\kms\)\citep{spencer+2017}. Thus, our measurement is likely
  slightly inflated given the high binarity fractions measured in these
  systems \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
\item
  Multiple populations. Both Sculptor and Ursa Minor likely contain
  multiple populations \citep[\citet{pace+2020},
  \citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a
  single population, and each population may have a different extent and
  velocity dispersion, this could result in biased velocity dispersions.
  However, it is unclear how to uniquely determine an overall velocity
  dispersion in a multi-population system.
\item
  Selection effects. RV studies each have their own selection effects,
  which may affect the resulting dispersion, especially if different
  populations or regions of the galaxy have different velocities or
  velocity dispersions. We do not attempt to correct for this.
\end{itemize}

For both Ursa Minor and Sculptor, we also fit models to only data from
individual surveys (see REF). Since the resulting parameters are very
similar, we conclude that many of the systematic uncertainties are
likely smaller than the present errors or that each large survey has
similar biases.

\section{Velocity modelling and
comparisons}\label{velocity-modelling-and-comparisons}

Here, we describe in additional detail, our methods and comparisons for
RV modelling between studies.

Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
smoothed samples from posterior/prior.

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties.}
\begin{tabular}{lllllllll}
\toprule
 & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
\midrule
Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
 & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
 & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
 & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
 & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
 & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
 & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
 & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
 & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Sculptor RV fits]{MCMC fits for different RV datasets for Sculptor among 3 different models. }
\label{tbl:scl_rv_mcmc}
\begin{tabular}{lllllll}
\toprule
study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log B_2/B_1$\\
\midrule
all &  &  &  &  &  & \\
 & $111.3\pm0.2$ & $9.64\pm0.16$ & - & - & - & 0\\
 & $111.3 \pm 0.2$ & $9.61 \pm0.16$ & - & $4.3\pm1.3$ & $-147_{-13}^{+17}$ & -1.9\\
 & $111.2\pm0.2$ & $9.66\pm0.16$ & $0.07\pm0.02$ & - & - & -3.4\\
tolstoy+23 &  &  &  &  &  & \\
 & $111.3 \pm 0.3$ & $9.79 \pm 0.18$ & - & - & - & 0\\
 & $111.3\pm0.3$ & $9.77\pm0.19$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ & -1.3\\
 & $111.2 \pm 0.3$ & $9.73\pm0.19$ & $0.085 \pm 0.023$ & – & – & -4.6\\
walker+09 &  &  &  &  &  & \\
 & $111.1\pm0.3$ & $9.5\pm0.2$ & – & – & – & 0\\
 & $111.1\pm0.3$ & $9.5\pm0.2$ & - & $5.2_{-1.6}^{+1.7}$ & $-134_{-16}^{+22}$ & -1.9\\
 & $111.1\pm0.3$ & $9.6\pm0.2$ & $0.06\pm0.03$ & – & – & +0.3\\
apogee &  &  &  &  &  & \\
 & $111.2\pm0.7$ & $8.6\pm0.5$ & – & – & – & –\\
 & $111.2\pm0.7$ & $8.5\pm0.5$ & – & $6\pm3$ & $-126_{-33}^{+45}$ & +0.1\\
 & $111.1\pm0.7$ & $8.5\pm0.5$ & $0.07\pm0.06$ & – & – & +0.9\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Ursa Minor RV fits]{MCMC fits for UMi velocity dispersion. }
\label{tbl:umi_rv_mcmc}
\begin{tabular}{lllll}
\toprule
study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
\midrule
all & $-245.8\pm0.3$ & $8.8\pm0.2$ & +1.3 & +0.9\\
pace & $-244.6\pm0.4$ & $9.0\pm0.3$ & +0.3 & +0.5\\
spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.06\\
apogee & $-245.6\pm1.2$ & $10.0_{-0.8}^{+1.0}$ & +1.0 & +0.5\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_rv_scatter.pdf}}
\caption[Scl velocity sample]{RV members of Sculptor plotted in the
tangent plane coloured by corrected velocity difference from mean
\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
scale).}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
\caption[Scl velocity gradient]{The corrected LOS velocity along the
best fit rotational axis. RV members are black points, the systematic
\(v_z\) is the horizontal grey line, blue lines represent the
(projected) gradient from MCMC samples, and the orange line is a rolling
median (with a window size of 50).}
\end{figure}

### 2025-07-15 thesis/sculptor.tex
major thesis restructure

--- 2025-04-25 thesis/sculptor.tex
@@
-\section{Sculptor (MW only)}\label{sculptor-mw-only}
+\section{Sculptor}\label{sculptor}
 
-\subsection{Orbital estimation}\label{orbital-estimation}
+\subsection{Derived properties}\label{derived-properties}
 
-fig.~\ref{fig:unref} contains example orbits
+\begin{table*}[t]
+\centering
+\begin{tabular}{lll}
+\toprule
+parameter & value & reference\\
+\midrule
+$L_\star$ & $1.82_{-0.22}^{+0.25}\times10^6\ L_\odot$ & \\
+$M_\star$ & $2.7_{-0.6}^{+0.7} \times10^6\ {\rm M}_\odot$ & \\
+$M_{200}$ & $0.48_{-0.25}^{+0.52}\ M_0$ & \\
+$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
+$v_{\rm circ, max}$ & $31\pm8\,\kms$ & \\
+$r_{\rm circ, max}$ & $5.9 \pm 2.9$ kpc & \\
+$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & \\
+$t_{\rm break, obs}$ & $110\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Milky Way tides}\label{milky-way-tides}
+
+\subsubsection{Orbital properties}\label{orbital-properties}
+
+Fig.~\ref{fig:unref} contains example orbits
 
@@
 
-\subsection{Simulations}\label{simulations}
+\subsubsection{Tidal effects}\label{tidal-effects}
 
@@
 
-\section{Sculptor (LMC orbits)}\label{sculptor-lmc-orbits}
+\subsection{Effects of the LMC}\label{effects-of-the-lmc}
+
+\subsubsection{Orbital effects}\label{orbital-effects}
 
@@
 rapidly but relatively close to the LMC in the past 100 Myr.
+
+Finally, we also consider the influence of the large Milky Cloud on the
+orbits and evolution of Scl. We adopt the \citet{vasiliev2024} multipole
+approximation of an N-body simulation of the LMC and MW. Their initial
+conditions are
+
+\begin{itemize}
+\tightlist
+\item
+  MW halo:
+\item
+  MW bulge (static):
+\item
+  MW disk (static):
+\item
+  LMC halo:
+\end{itemize}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}}
-\caption{image-20250415113122385}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}
+\caption[Selected orbits of Sculptor with an LMC]{Selected orbits of
+Sculptor in the MW frame. \textbf{TODO}: include LMC distance or frame,
+maybe xyz instead?}
 \end{figure}
 
-\begin{figure}
+\subsubsection{Tidal effects}\label{tidal-effects-1}
+
+\section{Ursa Minor}\label{ursa-minor}
+
+\subsection{Derived properties}\label{derived-properties-1}
+
+\begin{table*}[t]
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113133353.png}}
-\caption{image-20250415113133353}
-\end{figure}
+\begin{tabular}{lll}
+\toprule
+parameter & value & reference\\
+\midrule
+$L_\star$ &  & \\
+$M_\star$ &  & \\
+$M_{200}$ &  & \\
+$c_{\rm NFW}$ &  & \\
+$v_{\rm circ, max}$ & $\,\kms$ & \\
+$r_{\rm circ, max}$ & kpc & \\
+$r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
+$t_{\rm break, obs}$ & $120\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Orbital properties}\label{orbital-properties-1}
+
+\begin{table*}[t]
+\centering
+\begin{tabular}{lllllllll}
+\toprule
+Orbit & Mean &  & smallperi &  &  &  &  & \\
+\midrule
+ra &  &  &  &  &  &  &  & \\
+dec &  &  &  &  &  &  &  & \\
+distance &  &  &  &  &  &  &  & \\
+pmra &  &  &  &  &  &  &  & \\
+pmdec &  &  &  &  &  &  &  & \\
+rv &  &  &  &  &  &  &  & \\
+x &  &  &  &  &  &  &  & \\
+y &  &  &  &  &  &  &  & \\
+z &  &  &  &  &  &  &  & \\
+vx &  &  &  &  &  &  &  & \\
+vy &  &  &  &  &  &  &  & \\
+vz &  &  &  &  &  &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Tidal effects}\label{tidal-effects-2}

### 2025-07-15 thesis/compile.sh

@@ -9,9 +9,12 @@ pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
 pandoc $pandocoptions -o sculptor.tex sculptor.md
+pandoc $pandocoptions -o discussion.tex discussion.md
 
-# pandoc $pandocoptions -o ursa_minor.tex ursa_minor.md
-# pandoc $pandocoptions -o conclusion.tex conclusion.md
+pandoc $pandocoptions -o ursa_minor.tex ursa_minor.md
+pandoc $pandocoptions -o conclusion.tex conclusion.md
 
-# pandoc $pandocoptions -o appendix.tex appendix.md --top-level-division=chapter
+pandoc $pandocoptions -o appendix.tex appendix.md --top-level-division=chapter
+pandoc $pandocoptions -o rv_models.tex rv_models.md --top-level-division=chapter
+pandoc $pandocoptions -o numerical_convergence.tex numerical_convergence.md --top-level-division=chapter
 

### 2025-07-15 thesis/md_figure.py

@@ -10,11 +10,11 @@ def rename_url(blocks, i):
                 if isinstance(con2, Image):
                     url = con2.url
                     print("Found image at ", url, file=sys.stderr)
-                    if url.startswith("figures"):
-                        url_new = path.splitext(url)[0] + ".pdf"
-                        if path.isfile(url_new):
-                            con2.url = url_new
-                            print("Renaming ", url,"=>", con2.url, file=sys.stderr)
+                    # if url.startswith("figures"):
+                    #     url_new = path.splitext(url)[0] + ".pdf"
+                    #     if path.isfile(url_new):
+                    #         con2.url = url_new
+                    #         print("Renaming ", url,"=>", con2.url, file=sys.stderr)
 
 
 def process_figures(doc):

### 2025-07-15 thesis/plots/galaxy_pictures.jl

@@ -1,0 +1,119 @@
+### A Pluto.jl notebook ###
+# v0.20.6
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 0125bdd2-f9db-11ef-3d22-63d25909a69a
+begin
+	using Pkg; Pkg.activate()
+
+	FIGDIR = "figures"
+
+	using LilGuys
+	using CairoMakie
+	using Arya
+
+end
+
+# ╔═╡ f5c22abc-2634-4774-8516-fbd07aa690aa
+include("./paper_style.jl")
+
+# ╔═╡ 0e6f0654-dee0-4fee-b9fa-e7fc68de7b91
+import FileIO
+
+# ╔═╡ 5eaf3b50-886e-47ac-9a7c-80d693bc3c17
+CairoMakie.activate!(type=:png)
+
+# ╔═╡ 4b115ec9-6848-4d91-a492-f282d1fb1795
+lmc = FileIO.load("LMC_4deg_DSS2_color.jpg")
+
+# ╔═╡ 7d9745e3-c413-4bac-b877-21d8ca9adf78
+scl = FileIO.load("./Scl_1deg_DES-DR2_ColorIRG.jpg")
+
+# ╔═╡ 228c0e35-73bc-4eb7-b006-ed71e0b8d7bd
+umi = FileIO.load("./umi_2deg_unwise_plus_gaia.png")
+
+# ╔═╡ 67f24990-55fd-412d-aa99-aa69b9c774a2
+boo5 = FileIO.load("./Boo_V_0.15deg_P_SDSS9_color.jpg")
+
+# ╔═╡ 00c8ab2a-08e8-467e-a9ba-1793d9af8ea5
+function plot_scalebar!(width_degrees, bar_degrees, bar_label)
+	lines!([0.1, 0.1+bar_degrees / width_degrees], [0.1, 0.1], space=:relative, color=:white, linewidth = theme(:linewidth)[]/2)
+	text!(0.1, 0.1, text=bar_label, color=:white, align=(:left, :bottom), space=:relative, fontsize=10, font="TeX Gyre Heros Makie")
+	text!
+end
+
+# ╔═╡ 2188831f-d8ed-4a37-ad92-bc1f0d90ebbc
+function plot_label!(galaxy, fov)
[…]

### 2025-07-15 thesis/sync.sh

@@ -1,2 +1,2 @@
-rsync -av ~/thesis/ .
-rsync -av ~/dwarfs_figures/thesis/plots/figures ~/thesis/
+rsync -av ~/thesis_2/ .
+rsync -av ~/dwarfs_figures/thesis/plots/figures ~/thesis_2/

### 2025-07-21 thesis/appendix.tex
introduction updates

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/scl_density_methods_extra.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
 \caption[Scl density comparison]{Density profiles for various
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/scl_density_methods_j24.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_j24.pdf}}
 \caption[Scl density methods]{Comparison of density profiles for each
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/umi_density_methods_extra.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
 \caption[UMi density comparison]{Similar to
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/umi_density_methods_j24.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_j24.pdf}}
 \caption[UMi density methods]{Similar to

### 2025-07-21 thesis/introduction.tex
introduction updates

@@
 e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
-galaxies, dwarfs are the most numerous galaxies in the universe. Dwarf
-galaxies are highly \emph{dark-matter dominated}, with mass to light
-ratios sometimes exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of
-their small total masses, many dwarf galaxies are \emph{quenched}, with
-little to no recent star formation. Their stellar populations are
-\emph{relics} from the early universe, consisting of many of the oldest
-and most metal poor stars. Understanding the properties of dwarf
-galaxies thus has relevance across astronomy, from cosmological
-structure formation on the smallest scale to extremely metal-poor
-stellar populations to chemical evolution to galactic dynamics. Of the
-classical dwarfs, the Sculptor and Ursa Minor dwarf spheroidal galaxies
-stand out with a more extended density profile relative to an
-exponential, hinting at tidal effects. As a case study in tidal effects
-and our understanding of dwarf galaxy formation and evolution, we aim to
-understand the dynamical history of these galaxies.
+galaxies, dwarfs are the most numerous galaxies. Dwarf galaxies are
+highly \emph{dark-matter dominated}, with mass to light ratios sometimes
+exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
+masses, many dwarf galaxies are \emph{quenched}, with little to no
+recent star formation. Their stellar populations are \emph{relics} from
+the early universe, consisting of many of the oldest and most metal poor
+stars. Understanding the properties of dwarf galaxies thus has
+implications across astronomy, from cosmological structure formation on
+the smallest scale to extremely metal-poor stellar populations to
+chemical evolution to galactic dynamics. Of the classical dwarfs, the
+Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
+extended density profile relative to an exponential, hinting at tidal
+effects. As a case study in tidal effects and our understanding of dwarf
+galaxy formation and evolution, we aim to understand the dynamical
+history of these galaxies.
 
-In this section, we first describe the observational history of dwarf
-galaxies, what \emph{Gaia} is and how this telescope has changed our
-understanding of the local group, the present state of dwarf galaxy
-knowledge, and why Sculptor and Ursa Minor stand out. We then describe
-the role of dwarf galaxies in cosmology, their formation and our current
-theory of hierarchichal structure formation. We later discuss idealized
-simulations, tidal effects, break radii, and recent developments in
-tidal simulations. Finally, we provide a roadmap to the thesis as a
-whole at the end of this section.
+In this section, we first describe cosmological structure formation and
+the role of dwarf galaxies. Then, we explore the observational history
+of dwarf galaxies, the \emph{Gaia} telescope, open questions, and why
+Sculptor and Ursa Minor stand out. We discuss idealized simulations,
+tidal effects, break radii, and recent developments in tidal
+simulations. Finally, we provide a roadmap to the thesis as a whole at
+the end of this section.
+
+\section{Cosmological context}\label{cosmological-context}
+
+We only understand a tiny fraction of the composition of the universe.
+The leading theory of cosmology, \(\Lambda\)CDM (cold dark matter),
+posits that the universe is composed of about 68\% dark energy
+(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
+  like to change definitions of words. \emph{Baryons} here means
+  baryons+leptons, i.e.~any standard model massive fermion. The photon
+  energy density is negligible today} \citep{planckcollaboration+2020}.
+While the composition of dark matter and dark energy remains elusive, we
+know their general properties. Dark energy causes the acceleration of
+the expansion of the universe on large scales. We do not discuss dark
+energy here---it does not substantially affect the local group today.
+Dark matter, instead, makes up the vast majority of mass in galaxies.
+Typically, galaxies have baryonic to dark matter ratios of between 1:5
+to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
+matter is assumed to only interact gravitationally, passing through
+matter without effect (transparent to light, or \emph{dark}). Dark
+matter is also \emph{cold}, i.e.~typical velocities much smaller than
+the speed of light in the early universe. Implications of dark matter
+properties range from cosmological structural formation, galaxy
+structure, and galaxy interactions.
+
+\subsection{Structure formation and dwarf
+galaxies}\label{structure-formation-and-dwarf-galaxies}
+
+The very early universe was almost featureless. Our earliest
+observations of the universe stem from the cosmic microwave background
+(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
+But tiny perturbations in the CMB, temperature fluctuations of 1 part in
+10,000, became the seeds of large-scale cosmological structure. Governed
+by cosmological expansion, gravitational collapse, and baryonic physics,
+each perturbation grows into larger structures. Initially, baryonic
+matter was coupled to radiation and resisted collapse. Dark matter, only
+influenced by gravity instead, freely collapsed into the first
+structures. Each self-gravitating overdensity of dark matter is known as
+a \emph{halo}. After recombination, where electrons combined with atomic
+nuclei to form atoms, baryons decoupled from radiation and fell into the
+dark matter halos. The densest pockets of baryons later formed the first
+stars and galaxies.
+
+Dark matter halos, and their associated galaxies, rarely evolve in
+isolation. Instead, structure formation is \emph{hierarchical}. Small
+dark matter halos collapse first and hierarchically merge into
+progressively larger halos \citep[e.g.][\citet{blumenthal+1986},
+\citet{white+rees1978}, \citet{white+frenk1991},
+\citet{somerville+dave2015}]{blumenthal+1984}. Structure formation
+happens on a wide range of scales. The largest structures become the
+cosmic web---composed of voids, filaments, and clusters---and the
+smallest structures directly observable host dwarf galaxies.
+Hierarchical assembly is evident through the large scale structure of
+the universe, remnants of past mergers within the milky way, and tidal
+disruption of dwarf galaxies and their streams around nearby galaxies.
+
+Small-scale structure formation is sensitive to deviations from
+\(\Lambda\)CDM cosmology \citep[e.g.][]{bechtol+2022}. Many alternative
+models, such as warm dark matter or self-interacting dark matter, may
+smooth out small-scale features and reduce the abundance of small halos
+or change their structure \citep[e.g.][]{lovell+2014}. Dwarf galaxies,
+occupying the smallest dark matter halos, are promising windows into
+small-scale cosmological features. Alternative dark matter properties
+can influence dwarf galaxy structure, formation, and tidal evolution.
+The nearby dwarf galaxies of the local group, with detailed
+observations, present a promising opportunity to understand the
+evolution of these objects and test our understanding of cosmology. To
+understand the evolution of these objects, we need to understand the
+general predictions from \(\Lambda\)CDM for the properties of dwarf
+galaxies.
+
+\subsection{Cosmological dark matter density
+profiles}\label{cosmological-dark-matter-density-profiles}
+
+In \(\Lambda\)CDM cosmological simulations, dark matter halos are
+remarkably self-similar. In \citet{NFW1996};\citet{NFW1997}, hereafter
+NFW, the authors observe that the spherical radial density profiles
+\(\rho(r)\) are well described by a two-parameter law: \begin{equation}{
+\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
+}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
+density . This profile has shown remarkable success in describing
+\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
+profiles are \emph{cuspy}, where the density continuously rises like
+\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
+density profile increases at \(r \sim r_s\) and at large radii, the
+density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
+shows an example of a NFW halo and a cored dark matter halo.
+
+The total mass of an NFW profile diverges, so halos are commonly
+characterized by an overdensity criterion. The virial mass, \(M_{200}\),
+is defined as the mass within a region \(r_{200}\) containing a mean
+enclosed density 200 times\footnote{200 smells like an arbitrary value.
+  From approximate analytic arguments about the mean density of a
+  self-gravitating spherical collapse halo in an expanding universe, the
+  expected virial overdensity is XXX, rounding up to 200 (REFS).} the
+critical density of the universe: \begin{equation}{
+M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
+}\end{equation} A standard second parameter is the halo concentration,
+\(c=r_{200} / r_s\) describing how the characteristic size of the halo
+compares to its virial radius. In this case, the scale density is a
+function of \(c\) alone,
+\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
+\citep{NFW1996}. However, we characterize halos by their circular
+velocity profiles. The circular velocity,
+\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
+\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
+\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
+halo mass and the observed line of sight velocity disperion, and
+\(r_{\rm max}\) relates to the scale radius.
+
+While two free parameters characterize a NFW halo, \(M_{200}\) and
+\(c\), these parameters are not independent. Smaller dark matter halos
+often collapse earlier, when the universe was denser. As a result, small
+subhalos tend to be more concentrated \citep[e.g.][]{NFW1997}. The
+relationship between \(M_{200}\) and c, or the mass-concentration
+relation, how more massive halos become less concentrated in a
+predictable way \citep[e.g.][\citet{ludlow+2016}]{bullock+2001}. While
+there is a general trend in concentration with mass, the concentration
+values still have substantial scatter. Other parameters such as the halo
+spin or shape or slight changes in the halo density profile add
+additional variability to the present-day distribution of halos
+\citep[see e.g.][\citet{dutton+maccio2014}, darkEXP,
+ect.]{navarro+2010}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
+\caption[Example density profiles]{Example stellar and dark matter
+density profiles for a Sculptor-like galaxy. The dark matter is more
+extended and massive than the star across the entire
+galaxy.}\label{fig:nfw_density}
+\end{figure}
+
+\subsection{Connecting stars and dark
+matter}\label{connecting-stars-and-dark-matter}
+
+Both cosmology and observations find fundamental relationships between
+properties of galaxies, particularly for the total mass and luminosity.
+A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
+relation, describing the amount of stellar mass forms in a (sub)halo of
+a given size. In particular, the SMHM relation grows especially steep in
+the dwarf galaxy regime---many dwarf galaxies are formed in halos of
+similar masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
+\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
+apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
+the range of possible \(v_{\rm max}\) is fairly narrow across \(\sim 5\)
+decades in stellar mass. Thus, if we know the initial stellar mass of a
+dwarf galaxy, we also know, at some level, its dark matter mass.
+
+Several challenges complicate a simple SMHM trend including environment,
+assembly history, and tidal effects. By being closer to a massive host,
+most dwarfs quench earlier, resulting in lower stellar masses
+\citep[e.g.][]{christensen+2024}. Additionally, the time of formation
+(relative to reionization) can influence the resulting stellar content
+\citep{kim+2024}. Finally, tides influence both the dark matter and
+stellar mass but in different amounts \citep[e.g.][]{PNM2008}.
+Consequently, tides may reduce the halo mass more than the stellar mass,
+adding additional scatter to the SMHM trend, particularly for Milky Way
+satellites {[}@.e.g fattahi+2018{]}. Understanding the effects of tides
+on local group dwarf galaxies is essential to determining where and how
+these galaxies formed in a cosmological context.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
+\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
+relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
+\citet{fattahi+2018}. Right: the mass-concentration relation for NFW
+halos, but parameterized in terms of \(r_{\rm max}\) and
+\(v_{\rm max}\). Together, these plots represent the cosmologically
+expected properties of underlying dark matter halos in any dwarf galaxy.
+The velocity dispersion directly constrains the mass contained within a
+half-light radius, so the underlying halo in \(\Lambda\)CDM is
+well-constrained.}\label{fig:smhm}
+\end{figure}
 
@@
 
-The Magellanic clouds are the most prominent of the Milky Way satellites
-and were likely always known to humans in the southern hemisphere.
-However, the first discovered dwarf galaxies in 1938, Sculptor and
-Fornax, heralds i n an era of discovery and investigation into a novel
-astronomical system. \citet{shapley1938} present Sculptor and Fornax as
-some kind of \emph{stellar system} resembling the Magellanic clouds and
-globular clusters. The first direct hints that these stellar systems are
-galaxies originates from early spectroscopic work
+Dwarf galaxies have often puzzled astronomers and our understanding of
+the universe. While we now generally understand how dwarf galaxies fit
+into the universe, this was not always the case. The first discovered
+dwarf galaxies in 1938\footnote{technically, the LMC and SMC may be
+  classified as dwarf galaxies, but these were likely always known to
+  humans at southern latitudes.}, Sculptor and Fornax, heralds an era of
+discovery and investigation into these astronomical system.
+\citet{shapley1938} present Sculptor and Fornax as a new type of
+\emph{stellar system} resembling the Magellanic clouds and globular
+clusters, but do not attempt to speculate on the exact nature. The first
+direct hints that these systems are galaxies originates from early
+spectroscopic work deriving the first velocity dispersions
 \citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
 \citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
-\citet{olszewski1998}.]{aaronson1983}. Based on their velocity
-dispersions given their physical sizes, the mass to light ratios of
-these dwarf galaxies is far larger than expected for star clusters
-without dark matter. In general, these objects have reached a consensus
-that they have high velocity dispersion, old spheroidal stellar
-populations, and low stellar masses. While we now know the basic
-properties of dwarf galaxies, tremendous progress has been made in the
-past 10-20 years with large area deep photometric surveys, multi-object
-spectrographs, high resolution chemistry, and theoretical work,
-unveiling that dwarf galaxies are each unique systems with different
-formation pathways \citep[e.g.][]{simon2019}.
+\citet{olszewski1998}.]{aaronson1983}. Based on the physical size and
+velocity dispersions, the mass to light ratios of these dwarf galaxies
+is far larger than expected for star clusters without dark matter. In
+general, these objects have reached a consensus that they have high
+velocity dispersion, old spheroidal stellar populations, and low stellar
+masses.
 
@@
 the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
-Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. From
-this figure, we notice that while large LMC-like dwarf galaxies exhibit
-complex structure such as a bar, young blue stars, and irregular shapes,
-the classical and faint dwarf galaxies are composed mostly of very old,
-faint, red stars, very diffuse and often large on the sky if local.
-These galaxies are typically classified as dwarf spheroidals (dSph),
-however we typically omit the dSph designation and refer to these
-galaxies by their constellation hereafter.
+Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. Large
+LMC-like dwarf galaxies exhibit complex structure such as a bar, young
+blue stars, and irregular shapes. On the other hand, the classical and
+faint dwarf galaxies are composed mostly of old, faint, and red stars.
+These galaxies are typically classified as dwarf spheroidals (dSph)
+(although we typically omit the dSph designation and refer to these
+galaxies by their constellation alone).
 
@@
 
-\textbf{Classical systems}, the brightest of the local group
+\emph{Classical systems}, the brightest of the local group
 (non-Magellanic) dwarf galaxies, are perhaps the systems best studied
-and with the most stringingly derived properties. Defined\footnote{Note
+and with the most stringently derived properties. Defined\footnote{Note
   that some work like \citet{simon2019} instead define classical dwarfs
@@
 Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
-these dwarfs contain large numbers of bright (giant) stars allow for
+these dwarfs contain large numbers of bright (giant) stars, allowing
 thousands of stars to be observed with deep photometry and spectroscopy
@@
 constrained today and show excellent convergence among different works.
-However, ongoing research still is refining our understanding of the
+However, ongoing research continues to redefine our understanding of the
 detailed structure of Milky Way satellites, hinting that these objects
@@
 
-For Local group dwarf spheroidals, \emph{Gaia} has allowed for many of
+For Local group dwarf spheroidal, \emph{Gaia} has allowed for many of
 the first proper motion measurements. While the proper motion
@@
 
-\subsubsection{Exponential stellar density
-profiles}\label{exponential-stellar-density-profiles}
+With the help of \emph{Gaia}, high resolution spectroscopy, high-quality
+photometry, and large aperture multi-object spectroscopic instruments,
+we can observe dwarf galaxies in unprecedented detail. Today, dwarf
+galaxies are used to search for dark matter cores (REFS), test our
+understanding of early nucleosynthesis (REFS), the details of galaxy
+formation (REFS), and \ldots.
+
+While we now know the basic properties of dwarf galaxies, tremendous
+progress has been made in the past 10-20 years with large area deep
+photometric surveys, multi-object spectrographs, high resolution
+chemistry, and theoretical work, unveiling that dwarf galaxies are each
+unique systems with different formation pathways
+\citep[e.g.][]{simon2019}.
+
+\citet{battaglia+nipoti2022}
+
+One of the open questions in dwarf galaxy evolution is the formation of
+dark matter \emph{cores} where the density becomes constant instead
+\(\rho \sim 1\). A leading theory is cores form through of baryonic
+effects and feedback. If dark matter deviates from \(\Lambda\)CDM, then
+the NFW density profile may not describe halos anymore. \emph{Warm} dark
+matter, relativistic in the early universe but cooler now, smooths out
+small-scale features and softens the cusps of dark matter halos.
+\emph{Self-interacting} dark matter instead can form cores but also the
+cores can collapse into a density peak.
+
+\begin{itemize}
+\tightlist
+\item
+  \citet{walker+penarrubia2011}: measuring cores and cusps
+\item
+  \citet{dicintio+2013}, \citet{navarro+2010}, NFW profiles may actually
+  be more Einasto like
+\end{itemize}
+
+\subsection{Stellar density profiles}\label{stellar-density-profiles}
+
+Given star counts for a local group dwarf, one of the main quantities of
+interest is the surface density profile. Long used to characterize and
+understand larger galaxies, these profiles can provide hints as to the
+formation of dwarf galaxies, determine the physical extent of the
+stellar component, and the total stellar mass. Previous works tend to
+focus on fitting 3 different surface brightness profiles: an
+Exponential, a Plummer, or a King profile.
+
+The exponential profile is perhaps the simplest, as a 1-parameter
+profile \begin{equation}{
+\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
+}\end{equation}
 
@@
 
-Since \citet{faber+lin1983} demonstrated that an Exponential density
-profile is a reasonable empirical fit to dwarf galaxies, many works have
-fitted an Exponential profile (\citet{Eskridge} 1988a,c; Hodge et al
-1991a,b; \citet{IH1995}, ) in addition to Sérsic, King and various other
-profiles. \citet{lelli+2014} (example of using exponential.)
+\citet{plummer1911} proposed a 1-parameter solution for a polytropic
+potential to fit globular cluster density profiles:
 
-\citet{wang+2019} for fornax, \citet{mcconnachie+irwin2006} for Exp,
-Plummer, King in other galaxies.
+\begin{equation}{
+\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2}
+}\end{equation}
 
-However, even starting from \citet{aparicio1997}, some people have noted
+The \citet{king1962} profile, originating as an empirical fit to
+globular clusters, also sometimes reasonably fits dwarf galaxies. This
+density profile includes an additional parameter \(R_t\), a truncation
+radius where the density sharply drops to 0. \begin{equation}{
+\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
+}\end{equation}
+
+Finally, the Sérsic profile represents an extension of the exponential.
+Typically parameterized in terms of a half-light radius \(R_h\), the
+density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\), the
+profile's equation is \begin{equation}{
+\Sigma_{\rm Sérsic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
+}\end{equation} where \(b_n\) is a constant depending on \(n\). Less
+commonly used in studies of dwarf galaxies, \citet{munoz+2018} advocate
+for using the Sérsic profile since the added flexibility allows for more
+profiles to be fit.
+
+While there are not clear theoretical explanations why any profile is
+best, the Exponential stellar density profile is commonly assumed as a
+single parameter density profile which fits many dwarf galaxies well
+\citep[e.g.][\citet{eskridge1988a}, \citet{hodge1991a},
+\citet{hodge1991b}, \citet{IH1995}, etc.]{MV2020a} \citet{faber+lin1983}
+demonstrated that an Exponential density profile is a reasonable
+empirical fit to dwarf galaxies, theorizing that dwarf spheroidal evolve
+from exponential disks, maintaining a similar light profile.
+
+in addition to Sérsic, King and other profiles. \citet{lelli+2014}
+(example of using exponential.) \citet{wang+2019} for fornax,
+\citet{mcconnachie+irwin2006} for Exp, Plummer, King in other galaxies.
+
+However, an exponential profile was often noted to not fit every dwarf
+galaxy. Even starting from \citet{aparicio1997}, some people have noted
 deviations from this simple empirical rule. Additionally,
@@
 \end{itemize}
-
-\subsubsection{Hints of tides and
-disruptions}\label{hints-of-tides-and-disruptions}
-
-\subsubsection{Kinematics and structure}\label{kinematics-and-structure}
 
@@
 $\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
-$\delta$ & $-33.7186 \pm 0.0007^\circ$ & –\\
+$\delta$ & $-33.7186 \pm 0.0007^\circ$ & ”\\
 distance modulus & $19.60 \pm 0.05$ & 2\\
-distance & $83.2 \pm 2$ kpc & –\\
+distance & $83.2 \pm 2$ kpc & ”\\
 $\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
+$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & ”\\
 LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
-$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & –\\
+$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & ”\\
 $R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
-ellipticity & $0.37 \pm 0.01$ & –\\
-position angle & $94\pm1^\circ$ & –\\
-$M_V$ & $-10.82\pm0.14$ & –\\
+ellipticity & $0.37 \pm 0.01$ & ”\\
+position angle & $94\pm1^\circ$ & ”\\
+$M_V$ & $-10.82\pm0.14$ & ”\\
 \bottomrule
@@
 $\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
-$\delta$ & $67.2221 \pm 0.0016$˚ & –\\
+$\delta$ & $67.2221 \pm 0.0016$˚ & ”\\
 distance modulus & $19.23 \pm 0.11$ & 2\\
-distance & $70.1 \pm 3.6$ kpc & –\\
+distance & $70.1 \pm 3.6$ kpc & ”\\
 $\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
-$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
+$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & ”\\
 LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
-$\sigma_v$ & $8.6 \pm 0.3$ & –\\
+$\sigma_v$ & $8.6 \pm 0.3$ & ”\\
 $R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
-ellipticity & $0.55 \pm 0.01$ & –\\
-position angle & $50 \pm 1^\circ$ & –\\
-$M_V$ & $-9.03 \pm 0.05$ & –\\
+ellipticity & $0.55 \pm 0.01$ & ”\\
+position angle & $50 \pm 1^\circ$ & ”\\
+$M_V$ & $-9.03 \pm 0.05$ & ”\\
 \bottomrule
@@
 
-\section{Cosmological and theoretical
-context}\label{cosmological-and-theoretical-context}
+\section{Simulating tidal effects}\label{simulating-tidal-effects}
 
-The leading theory of cosmology, \(\Lambda\)CMD (cold dark matter) holds
-that the universe is composed of about 68\% dark energy (\(\Lambda\)),
-27\% dark matter (DM), and 5\% regular baryons
-\citep{planckcollaboration+2020}. While baryonic matter is the only
-matter we can directly understand or observe, we still know the general
-properties of dark matter and dark energy. Dark energy causes the
-acceleration of the expansion of the universe on large scales. We do not
-discuss dark energy here as it does not substantially affect the local
-group today. Dark matter makes up the vast majority of mass in galaxies.
-Typically, galaxies have baryonic to dark matter ratios of between 1:5
-to beyond 1:100. The faintest dwarf galaxies harbouring the largest
-proportion of dark matter. In \(\Lambda\)CMD, dark matter is assumed to
-only interact gravitationally with itself and other matter (transparent
-to light, or \emph{dark}). Dark matter is also created \emph{cold},
-i.e.~typical velocities much smaller than the speed of light in the
-early universe. Dark matter is a key ingredient in cosmological
-structural formation, dynamical galaxy structure, and galaxy-galaxy
-interactions.
+Since the discovery of dwarf galaxies, a large body of work has
+speculated, considered, or simulated tidal effects. While
+pre-\emph{Gaia} work often did not know the specific orbits of dwarf
+galaxies, the theory for satellites serves as an excellent framework for
+understanding specific systems.
 
-\subsection{Formation of dwarf
-galaxies}\label{formation-of-dwarf-galaxies}
+While cosmological simulations predict that mergers and satellite-galaxy
+interactions are common \citep[e.g.,][]{riley+2024}, they struggle to
+understand the evolution of individual dwarf galaxies. In particular,
+dwarf galaxies are often barely resolved. The highest resolution
+simulation of a Milky Way dark matter halo, the Aquarius project
+\citep{springel+2008}, acchieved a DM resolution of
+\(1.712 \times10^3 \Mo\), enough to barely resolve Sculptor like halos
+(see methods). On the other hand, idealized simulations are able to
+reach high resolution and numerical convergence for a single dwarf
+galaxy. But, idealization may not be the most realistic environment,
+neglecting mergers, assembly history, and often baryonic physics. We use
+idealized simulations here which are more powerful in accurately
+assessing tidal effects, given that the idealizations do not impact
+these galaxies's recent history too much. As such, a long history of
+work has resolved to understand the tidal evolution of dwarf galaxies
+using these \emph{idealized} simulations.
 
-The very early universe was almost featureless. As observed from the
-cosmic microwave background created at redshift \(z\sim 1100\), the
-temperature is a near perfect blackbody with fluctuations only at 1 part
-in 10,000. These tiny perturbations began to grow as the universe ages
-and expands. Initially hot and dense, baryonic matter was prevented from
-collapsing by thermal / radiation pressure. However, dark matter was
-uninhibited, freely passing through the universe and collapsing into
-denser regions. Each self-gravitating overdensity of dark matter is
-known as a \emph{halo}. As the universe continued to expand and cool
-down, baryonic matter began to collapse into dark matter halos.
-Eventually, the densest regions of baryons situated near the centres of
-halos would become the first stars of the first galaxies.
+Some of the earliest theory work on tidal stripping of dwarf galaxies
+originate from \citep[\citet{moore+davis1994},
+\citet{johnston+spergel+hernquist1995}, \citet{oh+lin+aarseth1995},
+\citet{piatek+pryor1995}, \citet{velazquez+white1995},
+\citet{kroupa1997}]{allen+richstone1988}. Many of these works used
+similar techniques as we continue to use today, setting up a dark matter
+halo in a static Milky Way potential and predicting the evolution of
+quantities of interest.
 
-Halos rarely evolve as isolated systems. Instead, structure formation is
-\emph{hierarchical}, where smaller \emph{subhalos} are accreted into
-larger halos which may accrete into yet larger halos. The most massive
-halos are made up of thousands of subhalos (verify), with each subhalo
-becoming entire galaxies making up a galaxy cluster. But, even the
-smallest galaxies reside in a \emph{halo}, with the dwarf galaxies
-representing entities with simpler \emph{assembly histories}.
-
-\subsection{Cosmological dark matter density
-profiles}\label{cosmological-dark-matter-density-profiles}
-
-From cosmological simulations, we know that across many orders of
-magnitude, dark matter halos appear self-similar. In
-\citet{NFW1996};\citet{NFW1997}, hereafter NFW, the authors observe that
-the radial density profiles \(\rho(r)\) follow a two-parameter law:
-\begin{equation}{
-\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2}
-}\end{equation} where \(r_s\) and \(\rho_s\) are the scale radius and
-density (see below). This density profile has shown remarkable success
-as a description of \(\Lambda\)CMD halos \citep[REFS, but see
-also][darkEXP, ect.]{navarro2010}. This density profile is \emph{cuspy},
-where the density continuously rises like \(\rho \sim 1/r\) at the
-centre. Also note that the total mass of the NFW profile logarithmically
-diverges. Typically, the virial mass of NFW halos is discussed instead,
-defined as \(M_{200}\), the mass within a region \(r_{200}\) containing
-a mean density 200 times the critical density of the universe
-\(\rho_{\rm crit}\), so
-\(M_{200} = M(r<r_{200}) = 200\ (4\pi/3) \ r_{200}^3\ \rho_{\rm crit}\).
-
-There are several parameterizations of the NFW profile. \citet{NFW1996}
-initially write the halo in terms of \(r_s\) and a concentration
-\(c= r_{200} / r_s\) relating the virial radius \(r_{200}\) to \(r_s\).
-. In this case,
-\(\rho_s = \rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\). However, we
-chose to discuss halos in terms of velocity space. The circular velocity
-of the NFW halo, \(v_{\rm circ}(r) = \sqrt{G M(r) / r}\) related to the
-enclosed mass \(M(r)\), reaches a maximum of \(v_{\rm max}\) at
-\(r_{\rm max} = \alpha\,r_s\) where \(\alpha\approx2.16258\). Thus
-\(r_{\rm max}\) is directly related to the scale radius and
-\(v_{\rm max}\) is related to the total mass.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095831443.png}}
-\caption[The NFW density profile]{showing the fits to the NFW halos for
-several different halos from \citet{NFW1996}.}
-\end{figure}
-
-\subsection{Typical properties}\label{typical-properties}
-
-A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
-relation, describing the amount of stellar mass forms in a (sub)halo of
-a given size. In particular, the SMHM relation grows especially steep in
-the dwarf galaxy regime---many dwarf galaxies are formed in halos of
-similar masses. \citet{fattahi+2013}, \citet{fattahi+2018}
+More recently, \citet{PNM2008} run idealized simulations of a local
+group dwarf galaxy in a milky way halo.
 
@@
 \item
-  \citet{battaglia+nipoti2022}
+  \citet{mayer+2001} theory of tidal stripping
 \end{itemize}
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}}
-\caption[stellar mass halo mass relation]{The stellar mass halo mass
-relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
-\citet{fattahi2018}.}
-\end{figure}
+With precise orbits and a better understanding of the Milky Way
+potential and system, more recent work began to directly probe the
+dynamical histories of individual dwarf galaxies. Examples include
+\citet{iorio+2019} for Sculptor, \citet{borukhovetskaya+2022};
+\citet{dicintio+2024} for Fornax, \citet{borukhovetskaya+2022a} for
+Antlia II,
 
-\subsection{Challenges to cosmology or modelling
-limitations?}\label{challenges-to-cosmology-or-modelling-limitations}
-
-One of the open questions in dwarf galaxy evolution is the formation of
-dark matter \emph{cores} where the density becomes constant instead
-\(\rho \sim 1\). A leading theory is cores form through of baryonic
-effects and feedback. If dark matter deviates from \(\Lambda\)CDM, then
-the NFW density profile may not describe halos anymore. \emph{Warm} dark
-matter, relativistic in the early universe but cooler now, smooths out
-small-scale features and softens the cusps of dark matter halos.
-\emph{Self-interacting} dark matter instead can form cores but also the
-cores can collapse into a density peak.
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{walker+penarrubia2011}: measuring cores and cusps
-\item
-  \citet{dicintio+2013}, \citet{navarro+2010}, NFW profiles may actually
-  be more Einasto like
-\end{itemize}
-
-\section{Simulating tidal effects}\label{simulating-tidal-effects}
-
-Since the discovery of dwarf galaxies, a large body of work has
-speculated, considered, or simulated tidal effects. However, before
-\emph{Gaia}, the orbits of most systems were essentially unconstrained.
-The theory developed still relates to today's work.
-
-Because we only have finite compute resources, major assumptions or
-limitations are common in simulations. Cosmological simulations,
-starting from the very early universe, create a realistic environment,
-structure formation, and merger history. But, cosmological simulations
-typically barely resolve dwarf galaxies {[}REFS{]}, leading to numerical
-artifacts and artificial disruption. On the other hand, idealized
-simulations are able to reach high resolution and numerical convergence
-for a single dwarf galaxy. But, idealization may not be the most
-realistic environment, neglecting mergers, assembly history, and often
-baryonic physics. We use idealized simulations here which are more
-powerful in accurately assessing tidal effects, given that the
-idealizations do not impact these galaxies's recent history too much.
+Our goal is to apply a similar framework to Sculptor and Ursa Minor.
 
@@
 
-While the dominant force on large scales in the universe, calculating
-the gravitational interaction between a large number of objects becomes
-computationally prohibitive. So, how do we manage to create simulations
-with billions of particles across vast volumes of the universe?
-
-Perhaps the simplest method to compute the evolution of dark matter is
-through \emph{N-body simulations}. A dark matter halo is represented as
-a large number of dark matter particles (bodies).
+Modelling gravitational evolution is essential for understanding
+properties of galaxies. Perhaps the simplest method to compute the
+evolution of dark matter is through \emph{N-body simulations}. A dark
+matter halo is represented as a large number of dark matter particles
+(bodies). Each body represents a monte-carlo sample from the underlying
+matter distribution. However, galaxies are often assumed to be
+\emph{collisionless}---particles are not strongly affected by
+small-scale gravitational encounters. In contrast, star clusters evolve
+differently as star-star gravitational collisions changes the dynamics
+of the system. While we use individual gravitating bodies in N-body
+simulations, the Newtonian gravitational force is softened to be a
+Plummer sphere, so encounters closer than a softening length do not
+substantially impact the dynamical evolution.
 
@@
 necessary to retain accuracy. The total force calculations reduce from
-\(N^2\) to \(O(N\,\log N)\), representing orders of magnitude speedup.
-Modern codes such as \emph{Gadget} utilize many other performance
-tricks, such as splitting particles across many supercomputer nodes,
-efficient memory storage, adaptive time stepping, and parallel file
-writing to retain fast performance for large scale simulations, forming
-the foundation for many cosmological simulation codes.
+\(O(N^2)\) to \(O(N\,\log N)\), representing orders of magnitude
+speedup. Modern codes such as \emph{Gadget} utilize many other
+performance tricks, such as splitting particles across many
+supercomputer nodes, efficient memory storage, adaptive time stepping,
+and parallel file writing to retain fast performance for large scale
+simulations, forming the foundation for many cosmological simulation
+codes.
 
-\subsection{Galactic tides and idealized
-simulations}\label{galactic-tides-and-idealized-simulations}
-
-Despite the power of cosmological simulations for understanding large
-scale structure, a fundamental limitation is their ability to resolve
-small scales such as dwarf galaxies. Dwarf galaxies, without sufficient
-resolution, are likely to artificially disrupt as a result of numerical
-noise. For understanding the tidal evolution of a dwarf galaxy, this
-tool has limited accuracy then.
-
-Idealized simulations take a different approach. Instead of starting
-from the universe as a whole, just a single subhalo is simulating,
-allowing resolution several orders of magnitude than cosmological
-simulations. For instance, the highest resolution dark matter only
-simulation, the \emph{Aquarious Project} only reached a mass resolution
-of XXX. In constrast, our dark matter mass resolution in later chapters
-is of order 1000s of solar masses.
-
-As such, a long history of work has resolved to understand the tidal
-evolution of dwarf galaxies using these \emph{idealized} simulations.
-
-Some of the earliest theory work on tidal stripping of dwarf galaxies
-originate from \citep[\citet{moore+davis1994},
-\citet{johnston+spergel+hernquist1995}, \citet{oh+lin+aarseth1995},
-\citet{piatek+pryor1995}, \citet{velazquez+white1995},
-\citet{kroupa1997}]{allen+richstone1988}. Many of these works used
-similar techniques as we continue to use today.
-
-More recently, \citet{PNM2008} run idealized simulations of a local
-group dwarf galaxy in a milky way halo.
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{mayer+2001} theory of tidal stripping
-\end{itemize}
-
-\subsubsection{Break and tidal radii}\label{break-and-tidal-radii}
+\subsection{Break and tidal radii}\label{break-and-tidal-radii}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
 \caption[Break radius validation]{The break radius of the simulations is
@@
 \citet{EP2020} potential with a pericentre of 15 kpc and apocentre of
-100 kpc. See Section~\ref{sec:methods} for a description of our
-simulation setup. \textbf{TODO:} mark \(r_J\), plot initial, make sure
-1st pericentre, simplify models.}\label{fig:idealized_break_radius}
+100 kpc. In this model, the jacobi radius is close to the break radius.
+See Section~\ref{sec:methods} for a description of our simulation
+setup.}\label{fig:idealized_break_radius}
 \end{figure}
 
-\subsubsection{Tidal evolution}\label{tidal-evolution}
+\subsection{Tidal evolution}\label{tidal-evolution}
 
@@
 \left(\frac{r_{\rm max}}{r_{\rm max, 0}}\right)^{\beta}\left[1 + \left(r_{\rm max} / r_{\rm max, 0}\right)^2\right]^{-\beta}
-}\end{equation} \{eq:tidal\_track\} where \(\alpha=0.4\), \(\beta=0.65\)
-are empirical fits. As illustrated in Fig.~\ref{fig:tidal_tracks}, this
-formula works for both circular and elliptical orbits and is independent
-of the initial subhalo size or distance to the host.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}}
-\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies.
-Fig. 6 from \citet{EN2021}.}\label{fig:tidal_tracks}
-\end{figure}
-
-\subsection{Notable related works}\label{notable-related-works}
+}\end{equation} where \(\alpha=0.4\), \(\beta=0.65\) are empirical fits.
+As illustrated in Fig.~\ref{fig:tidal_tracks}, this formula works for
+both circular and elliptical orbits and is independent of the initial
+subhalo size or distance to the host.
 
 Case studies:
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{dicintio+2024}
-\item
-  \citet{iorio+2019}
-\end{itemize}
 
@@
 \item
-  \citet{bullock+johnston+2005}
+  \citet{bullock+johnston2005}
 \item
   \citet{PNM2008}
-\item
-  \citet{borukhovetskaya+2022a}, \citet{borukhovetskaya+2022}
 \item
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}
+\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies.
+Fig. 6 from \citet{EN2021}.}\label{fig:tidal_tracks}
+\end{figure}
+
 \section{Thesis Outline}\label{thesis-outline}
 
+In this thesis, our goal is to review the evidence for an extended
+density profile in Ursa Minor and Sculptor, to assess the impact of
+tidal effects on each galaxy, and to discuss possible interpretations
+for the structure of these galaxies.
+
 In chapter 2, we describe how we compute observational density profiles
-from \citet{jensen+202}. In chapter 3, we review our simulation methods.
-Next, we present our results for the tidal stripping of Sculptor and
-Ursa Minor in Chapter 4. We discuss our results, limitations, and
-implications in Chapter 5. Finally, in Chapter 6, we summarize this work
-and discuss future directions for similar work and the field of dwarf
-galaxies.
+from \citet{jensen+2024}. In chapter 3, we review our simulation
+methods. Next, we present our results for the tidal stripping of
+Sculptor and Ursa Minor in Chapter 4. We discuss our results,
+limitations, and implications in Chapter 5. Finally, in Chapter 6, we
+summarize this work and discuss future directions for similar work and
+the field of dwarf galaxies.

### 2025-07-21 thesis/numerical_convergence.tex
introduction updates

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_softening.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
 \caption{Softening convergence}\label{fig:softening_convergence}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_methods.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
 \caption{Isolation method convergence}\label{fig:methods_convergence}

### 2025-07-21 thesis/rv_models.tex
introduction updates

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_rv_fits.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
 \caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_rv_scatter.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
 \caption[Scl velocity sample]{RV members of Sculptor plotted in the
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
 \caption[Scl velocity gradient]{The corrected LOS velocity along the

### 2025-07-21 thesis/sculptor.tex
introduction updates

@@
 \midrule
-$L_\star$ & $1.82_{-0.22}^{+0.25}\times10^6\ L_\odot$ & \\
-$M_\star$ & $2.7_{-0.6}^{+0.7} \times10^6\ {\rm M}_\odot$ & \\
-$M_{200}$ & $0.48_{-0.25}^{+0.52}\ M_0$ & \\
+$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
+$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
+$M_\star / L_\star$ & 1.7, 0.17 dex uncertainties. & (woo+2009?)\\
+$M_{200}$ & $0.5_{-0.3}^{+0.6} \times10^10\ M_0$ & \\
 $c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
-$v_{\rm circ, max}$ & $31\pm8\,\kms$ & \\
-$r_{\rm circ, max}$ & $5.9 \pm 2.9$ kpc & \\
+$v_{\rm circ, max}$ & $31_{-6}^{+8}\,\kms$ & \\
+$r_{\rm circ, max}$ & $6_{-2}^{+3}$ kpc & \\
 $r_{\rm break, obs}$ & $25 \pm 5$ arcmin & \\
 $t_{\rm break, obs}$ & $110\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+For the mass-to-light ratio, we use the values from \citet{woo+2009}.
+Note that we assume a scatter of 0.17 dex for the mass-to-light ratios
+as reported for mass uncertainties. \citet{delosreyes+2024} also
+indicate that other methods for stellar mass-to-light ratios result in
+similar scatters and systematic biases. However these values are just
+approximate.
+
+\begin{table*}[t]
+\centering
+\begin{tabular}{lllll}
+\toprule
+Halo name & Rmax & Vmax & M200 & c\\
+\midrule
+fiducial & 3.2 & 31 & 0.33 & 21\\
+extended & 4.2 & 31 & 0.39 & 17\\
 \bottomrule
@@
 \midrule
-$L_\star$ &  & \\
-$M_\star$ &  & \\
-$M_{200}$ &  & \\
-$c_{\rm NFW}$ &  & \\
-$v_{\rm circ, max}$ & $\,\kms$ & \\
-$r_{\rm circ, max}$ & kpc & \\
+$L_\star$ & $3.5 \pm 0.1 \times 10^5\,L_\odot$ & \\
+$M_\star$ & $7_{-2}^{+3} \times 10^5\,\Mo$ & \\
+$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & (woo+2009?)\\
+$M_{200}$ & $3_{-2}^{+4} \times 10^9\,\Mo$ & \\
+$c$ & 14? & \\
+$v_{\rm circ, max}$ & $27_{-6}^{+7}\,\kms$ & \\
+$r_{\rm circ, max}$ & $5_{-2}^{+1}$ kpc & \\
 $r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
 $t_{\rm break, obs}$ & $120\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\begin{tabular}{lllll}
+\toprule
+Halo name & Rmax & Vmax & M200 & c\\
+\midrule
+fiducial & 5 & 37 & 0.67 & 17\\
+compact & 4 & 38 & 0.62 & 21\\
 \bottomrule

### 2025-07-23 thesis/discussion.tex
refining intro

@@
   disk galaxy formation \citet{fall+efstathiou1980}, \citet{mestel1963}
+\item
+  valcke+derijcke+dejonghe2008 with early idealized hydro simulations
+  show Sérsic values from simulations between 0.8 and 1 for dSph,
+  indicating a natural formation of Exponential or slightly steeper
+  density profiles.
 \end{itemize}
@@
 expectation.
+
+\subsection{Alternatives to exponential
+profiles}\label{alternatives-to-exponential-profiles}
+
+\begin{itemize}
+\tightlist
+\item
+  elmegreen+hunter2006: Discuss the formation of double exponentials in
+  disk galaxies. While a simpler idealized / Semi-analytic model of star
+  formation, show that assuming exponentail gas and KS-relation with SF
+  threshhold, that reduction of turbulance (driven by a number of
+  processes) in outer disk leads to a steeper decline in SFR as crossing
+  the critical SF threshold becomes less likely.
+\end{itemize}
 
@@
   migration =\textgreater{} population gradients.
+\item
 \end{itemize}
@@
   Carina)
+\item
+  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
 \end{itemize}

### 2025-07-23 thesis/introduction.tex
refining intro

@@
 environments in the universe. These little galaxies are typically
-defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\),
-e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
-galaxies, dwarfs are the most numerous galaxies. Dwarf galaxies are
-highly \emph{dark-matter dominated}, with mass to light ratios sometimes
+defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\) ,
+e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
+dwarfs are the most numerous galaxies. Dwarf galaxies are highly
+\emph{dark-matter dominated}, with mass to light ratios sometimes
 exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
@@
 dark matter halos collapse first and hierarchically merge into
-progressively larger halos \citep[e.g.][\citet{blumenthal+1986},
-\citet{white+rees1978}, \citet{white+frenk1991},
-\citet{somerville+dave2015}]{blumenthal+1984}. Structure formation
-happens on a wide range of scales. The largest structures become the
-cosmic web---composed of voids, filaments, and clusters---and the
-smallest structures directly observable host dwarf galaxies.
-Hierarchical assembly is evident through the large scale structure of
-the universe, remnants of past mergers within the milky way, and tidal
-disruption of dwarf galaxies and their streams around nearby galaxies.
+progressively larger halos
+\citep[e.g.,][]{blumenthal+1984, blumenthal+1986, white+rees1978, white+frenk1991, somerville+dave2015}.
+Structure formation happens on a wide range of scales. The largest
+structures become the cosmic web---composed of voids, filaments, and
+clusters---and the smallest structures directly observable host dwarf
+galaxies. Hierarchical assembly is evident through the large scale
+structure of the universe, remnants of past mergers within the milky
+way, and tidal disruption of dwarf galaxies and their streams around
+nearby galaxies.
 
 Small-scale structure formation is sensitive to deviations from
-\(\Lambda\)CDM cosmology \citep[e.g.][]{bechtol+2022}. Many alternative
+\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
 models, such as warm dark matter or self-interacting dark matter, may
 smooth out small-scale features and reduce the abundance of small halos
-or change their structure \citep[e.g.][]{lovell+2014}. Dwarf galaxies,
+or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
 occupying the smallest dark matter halos, are promising windows into
@@
 
-\subsection{Cosmological dark matter density
-profiles}\label{cosmological-dark-matter-density-profiles}
+\subsection{Typical structure of dark matter
+halos}\label{typical-structure-of-dark-matter-halos}
 
 In \(\Lambda\)CDM cosmological simulations, dark matter halos are
-remarkably self-similar. In \citet{NFW1996};\citet{NFW1997}, hereafter
-NFW, the authors observe that the spherical radial density profiles
-\(\rho(r)\) are well described by a two-parameter law: \begin{equation}{
+remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
+authors observe that the spherical radial density profiles \(\rho(r)\)
+are well described by a two-parameter law: \begin{equation}{
 \rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
@@
 
-While two free parameters characterize a NFW halo, \(M_{200}\) and
-\(c\), these parameters are not independent. Smaller dark matter halos
-often collapse earlier, when the universe was denser. As a result, small
-subhalos tend to be more concentrated \citep[e.g.][]{NFW1997}. The
-relationship between \(M_{200}\) and c, or the mass-concentration
-relation, how more massive halos become less concentrated in a
-predictable way \citep[e.g.][\citet{ludlow+2016}]{bullock+2001}. While
-there is a general trend in concentration with mass, the concentration
-values still have substantial scatter. Other parameters such as the halo
-spin or shape or slight changes in the halo density profile add
-additional variability to the present-day distribution of halos
-\citep[see e.g.][\citet{dutton+maccio2014}, darkEXP,
-ect.]{navarro+2010}.
+While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
+these are not independent. Smaller dark matter halos often collapse
+earlier, when the universe was denser. As a result, small subhalos tend
+to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
+between \(M_{200}\) and c, or the mass-concentration relation, how more
+massive halos become less concentrated in a predictable way
+\citep[e.g.,][]{bullock+2001, ludlow+2016}. While there is a general
+trend in concentration with mass, the concentration values still have
+substantial scatter. Other parameters such as the halo spin or shape or
+slight changes in the halo density profile add additional variability to
+the present-day distribution of halos \citep[see
+e.g.,][darkEXP]{navarro+2010, dicintio+2013, dutton+maccio2014}.
 
@@
 
-\subsection{Connecting stars and dark
-matter}\label{connecting-stars-and-dark-matter}
+\subsection{Connecting dark matter to
+stars}\label{connecting-dark-matter-to-stars}
 
@@
 most dwarfs quench earlier, resulting in lower stellar masses
-\citep[e.g.][]{christensen+2024}. Additionally, the time of formation
+\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
 (relative to reionization) can influence the resulting stellar content
 \citep{kim+2024}. Finally, tides influence both the dark matter and
-stellar mass but in different amounts \citep[e.g.][]{PNM2008}.
+stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
 Consequently, tides may reduce the halo mass more than the stellar mass,
 adding additional scatter to the SMHM trend, particularly for Milky Way
-satellites {[}@.e.g fattahi+2018{]}. Understanding the effects of tides
-on local group dwarf galaxies is essential to determining where and how
-these galaxies formed in a cosmological context.
+satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
+tides on local group dwarf galaxies is essential to determining where
+and how these galaxies formed in a cosmological context.
 
@@
 \caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
-relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
-\citet{fattahi+2018}. Right: the mass-concentration relation for NFW
-halos, but parameterized in terms of \(r_{\rm max}\) and
-\(v_{\rm max}\). Together, these plots represent the cosmologically
-expected properties of underlying dark matter halos in any dwarf galaxy.
-The velocity dispersion directly constrains the mass contained within a
-half-light radius, so the underlying halo in \(\Lambda\)CDM is
-well-constrained.}\label{fig:smhm}
+relation, where \(v_{\rm max}\) is related to the halo mass. Right: the
+mass-concentration relation for NFW halos, but parameterized in terms of
+\(r_{\rm max}\) and \(v_{\rm max}\). Together, these plots represent the
+cosmologically expected properties of underlying dark matter halos in
+any dwarf galaxy. The velocity dispersion directly constrains the mass
+contained within a half-light radius, so the underlying halo in
+\(\Lambda\)CDM is well-constrained. Adapted from figure 1 of
+\citet{fattahi+2018}, stellar mass and \(v_{\rm max}\) of APOSTLE
+satellites at peak \(v_{\rm max}\).}\label{fig:smhm}
 \end{figure}
@@
 
-\subsection{Early observations}\label{early-observations}
-
-Dwarf galaxies have often puzzled astronomers and our understanding of
-the universe. While we now generally understand how dwarf galaxies fit
-into the universe, this was not always the case. The first discovered
-dwarf galaxies in 1938\footnote{technically, the LMC and SMC may be
-  classified as dwarf galaxies, but these were likely always known to
-  humans at southern latitudes.}, Sculptor and Fornax, heralds an era of
-discovery and investigation into these astronomical system.
+Dwarf galaxies have long challenged questioned theories of galaxy
+formation. The first discovered dwarf galaxies in 1938\footnote{technically,
+  the LMC and SMC may be classified as dwarf galaxies, but these were
+  likely always known to humans at southern latitudes.}, Sculptor and
+Fornax, already presented such a puzzle. Having no close analogues,
 \citet{shapley1938} present Sculptor and Fornax as a new type of
 \emph{stellar system} resembling the Magellanic clouds and globular
-clusters, but do not attempt to speculate on the exact nature. The first
-direct hints that these systems are galaxies originates from early
-spectroscopic work deriving the first velocity dispersions
-\citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
-\citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
-\citet{olszewski1998}.]{aaronson1983}. Based on the physical size and
-velocity dispersions, the mass to light ratios of these dwarf galaxies
-is far larger than expected for star clusters without dark matter. In
-general, these objects have reached a consensus that they have high
-velocity dispersion, old spheroidal stellar populations, and low stellar
-masses.
+clusters, but not attempting to speculate on the exact nature. While
+generally understood to be galaxies based on the inferred luminosities
+and sizes, the nature and formation remained unclear
+\citep[e.g.,][]{hodge1971}.
 
-Dwarf galaxies span a large range of size and morphology.
-Fig.~\ref{fig:galaxy_images} displays several images of dwarf galaxies.
-The largest in the local group but not always considered a dwarf galaxy,
-the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
-Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. Large
-LMC-like dwarf galaxies exhibit complex structure such as a bar, young
-blue stars, and irregular shapes. On the other hand, the classical and
-faint dwarf galaxies are composed mostly of old, faint, and red stars.
-These galaxies are typically classified as dwarf spheroidals (dSph)
-(although we typically omit the dSph designation and refer to these
-galaxies by their constellation alone).
+The earliest spectroscopic work hinted that dwarf galaxies may contain
+substantial dark matter. From the first measurements of velocity
+dispersions
+\citep[e.g.,][]{aaronson1983, gallagher+wyse1994, pryor1996}, the
+inferred mass-to-light ratios were about 10 times the expectation for
+globular clusters \citep{matteo1998}. Subsequently, many theories
+attempting to understand the formation and observed properties of these
+objects were proposed. Examples included that dwarf galaxies were formed
+tidally in the gas clouds of satellites; dwarf galaxies are undergoing
+tidal dissolution resulting in extreme mass-to-light ratios
+\citep[e.g.,][]{pryor1996}, or Since then, we have known that dwarf
+galaxies are among the darkest objects in the universe, and
+understanding their properties is critical to understanding the
+universe.
 
-Since the discovery of dwarf galaxies around the Milky Way,
-observational work has attempted to measure and refine their basic
-properties \citep[e.g.][]{mateo1998}. While the Milky Way's satellites
-are close (by extragalactic standards), their low numbers of stars and
-large apparent sizes present challenges for observational work. In
-particular, the properties of the faintest dwarf galaxy systems and
-especially \emph{ambiguous} systems remains a pressing challenge as a
-unique probe into the nature of the early universe and dark matter.
-
-\emph{Classical systems}, the brightest of the local group
-(non-Magellanic) dwarf galaxies, are perhaps the systems best studied
-and with the most stringently derived properties. Defined\footnote{Note
-  that some work like \citet{simon2019} instead define classical dwarfs
-  based on luminosity as compared to ultra faint dwarfs. This may
-  include galaxies such as Antlia I, Crater II, Canes Venatici I, and
-  Sagittarius. We stick with the historical definition for simplicity
-  and since the Gaia observations are most numerous for the historical
-  classicals.} as the first six discovered dwarf spheroidal galaxies,
-(Sculptor, Ursa Minor, Fornax, Draco, Leo I, Leo II, and Carina),
-classical systems are typically old, spheroidal, and gas-poor. While
-extending 1-2 degrees across the sky (e.g.
-Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
-these dwarfs contain large numbers of bright (giant) stars, allowing
-thousands of stars to be observed with deep photometry and spectroscopy
-\citep[e.g.][\citet{pace+2020}]{tolstoy+2023}. As a result, the
-determination of fundamental properties such as the position, size,
-orientation, distance \citep[from 100s of RRL stars,
-e.g.][]{tran+2022, garofalo+2025}, proper motions (from \emph{Gaia},
-e.g. \citet{battaglia+2022}, \citet{MV2020a}), line-of-sight (LOS)
-velocity and dispersion \(\sigma_v\), are all relatively well
-constrained today and show excellent convergence among different works.
-However, ongoing research continues to redefine our understanding of the
-detailed structure of Milky Way satellites, hinting that these objects
-may be more complex than at first glance.
+Dwarf galaxies span a large range of physical sizes, luminosities, and
+morphologies. Broadly, there are three classes of dwarf galaxies based
+on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
+\textbf{bright dwarfs galaxies} with magnitudes
+\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
+and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
+an example of an irregular (and slightly bright) dwarf galaxy displaying
+a bar. \textbf{Classical dwarf} galaxies occupy intermediate
+luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
+systems are old, gas-poor, and spheroidal. All Milky Way satellites
+discovered before digital sky surveys are classicals. The 12 classical
+dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
+Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
+I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
+  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
+  dSph, we omit this suffix for brevity.} The ultra-faints occupy the
+very faintest magnitudes (\(-7.7 \lesssim M_V\)). These galaxies have
+miniscule stellar masses, tend to be more compact, and are the darkest
+known galaxies. An example is Boötes V as shown in
+Fig.~\ref{fig:galaxy_images}. Altogether, dwarf galaxies span about 7
+orders of magnitude in stellar mass.
 
@@
 V (SDSS). While the LMC is very prominent in the sky, even the classical
-dwarf galaxies are not obvious except in terms of star counts (e.g.
-Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the central
-third to sixth of the image.}\label{fig:galaxy_images}
+dwarf galaxies are not obvious except in terms of star counts
+(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
+central third to sixth of the image.}\label{fig:galaxy_images}
 \end{figure}
@@
 
-Some of the most fundamental properties of astronomical objects are
-their positions and velocities. Unfortunately, determining distances to
-stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
-are easily determined from spectroscopy, the tangental velocities,
-perpendicular to the LOS, are best measured through proper motions,
-typically requiring accurate determinations of small changes in a star's
-position. \emph{Gaia}'s mission addresses these problems by making
-extremely precise measurements of stellar positions over time.
+A star's or galaxy's position and velocity are fundamental quantities
+for understanding its orbit and origin. Unfortunately, determining
+distances and full 3D velocities is challenging. The most direct
+measurement of distance, parallax, requires precise tracking of a star's
+sky position across a year. And while line-of-sight (LOS) velocities are
+easily determined from spectroscopy, tangental velocities, derived from
+proper motions, are much more challenging. Typically, measuring proper
+motions requires accurate (much less than arcsecond) determinations of
+small changes in a star's position over baselines of several years to
+decades. The full 6D position and velocity information for stars was
+historically only known for a handful of stars.
 
-\emph{Gaia} was designed to revolutionize proper motion and parallax
-measurements. \emph{Gaia} is a space-based, all-sky survey telescope
-with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
-point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
-completing its space-based mission in 2025 (with two more data releases
-planned). By imagining two patches of sky on the same focal plane,
-separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
-measure absolute proper motions. Stars in different regions of the sky
-are affected by parallax differently, so by precisely observing the
-relative positions of stars separated by large angles during multiple
-epochs over a year, an absolute all-sky reference frame can be derived
-\citep{gaiacollaboration+2016}. In addition to precise astrometric
-information, \emph{Gaia} measures the magnitude of stars in the very
-wide \emph{G} band (330-1050nm), blue and red colours using the blue and
-red photometers (BP and RP, 330-680 nm and 640-1050 nm respectively),
-and takes low resolution BP-RP spectra and radial velocity measurements
-of bright stars (\emph{Gaia} radial velocity magnitudes \textless16)
-\citep{gaiacollaboration+2016}. For our purposes, most useful are
-\emph{Gaia}'s measurements of \(G\) magnitude,
-\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
-the proper motions in RA and declination \(\mu_{\alpha*}\) and
-\(\mu_\delta\).\footnote{\(\mu_{\alpha*}\) allows where
+\emph{Gaia} has redefined astrometry, providing photometry, positions,
+proper motions, and parallaxes for over 1 billion stars
+\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
+space-based, all-sky survey telescope situated at the Sun-Earth L2
+Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
+completed its space-based mission in 2025, two forthcoming data releases
+remain. Determining absolute parallax measurement is facilitated by the
+observation that stars in different regions of the sky are affected by
+parallax motion with different phases. By imagining two regions
+separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
+tiny changes in relative positions of stars across small and large
+angles. Combining measurements from multiple epochs across several
+years, an absolute all-sky reference frame is derived from which
+parallax and proper motions are derived \citep{gaiacollaboration+2016}.
+In addition to astrometry, \emph{Gaia} measures photometry in the wide
+\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
+330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
+additionally provides low resolution BP-RP spectra and radial velocity
+measurements of bright stars (\emph{Gaia} radial velocity magnitudes
+\textless16) \citep{gaiacollaboration+2016}. For this work,
+\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
+\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
+\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{\(\mu_\alpha\)
+  and \(\mu_\delta\) are the apparent changes in \(\alpha\) and
+  \(\delta\). \(\mu_{\alpha*}\) allows where
   \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
-  effects{]}.}
+  effects in \(\alpha\).}
 
-\emph{Gaia} has revolutionized many astronomical disciplines, the least
-of which is local group and Milky Way science. While proper motions of a
-dwarf galaxies has been measured in a case by case basis by the Hubble
-Space Telescope, a full systematic determination of proper motions for
-most dwarf galaxies was unavailable until \emph{Gaia}
-\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
-the detection and improved measurements of halo substructures and
-streams \citep{bonaca+price-whelan2025}, Milky Way structure and
-kinematics \citep{hunt+vasiliev2025}.
+\emph{Gaia} has revolutionized many fields of astronomy including the
+Milky Way and the Local Group. For example, 6D kinematic classification
+of Milky Way stars has lead to the discovery of past mergers or Milky
+Way building blocks like \emph{Gaia}-sausage Enceladus
+\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
+\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and effects of the
+Milky Way's bulge and bar \citep{hunt+vasiliev2025}. Moving to the Milky
+Way halo, numerous streams \citep{bonaca+price-whelan2025}. \emph{Gaia}
+has revealed the hierarchical formation and messy structure of our own
+galaxy.
 
-For Local group dwarf spheroidal, \emph{Gaia} has allowed for many of
-the first proper motion measurements. While the proper motion
-uncertainty on a typical dwarf member star is often with large
+For Milky Way satellites, \emph{Gaia} has enabled well-constrained
+orbital analysis and facilitated precise stellar membership
+determination. Before, proper motions of dwarf galaxies were spase or
+undetermined and measured by the Hubble Space Telescope or other methods
+for only a few systems. \emph{Gaia} allowed for the first systematic
+determinations of Milky Way satellite proper motions, and of
+unprecedented precision \citep{MV2020a, pace+li2019}. While the proper
+motion uncertainty on a typical dwarf member star is often with large
 uncertainties, by combining the proper motions for 100s to 1000s of
 stars in \emph{Gaia}, incredibly precise proper motion measurements can
-be determined \citep[e.g.][]{MV2020a}, sometimes only limited by
-\emph{Gaia}'s systematic error floor. Proper motions have ushered in a
-new dynamical era for MW satellites studies, where we often have precise
-determinations of the dwarf galaxy's (short term) orbit under a given MW
-potential.
+be determined, sometimes only limited by \emph{Gaia}'s systematic error
+floor \citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new
+dynamical era for MW satellites studies, where we often have precise
+determinations of the dwarf galaxy's recent orbit under a given MW
+potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
+foreground stars. By measuring parallaxes or proper motions
+significantly different from the dwarf galaxy, many stars can be
+immediately removed as non-members
+\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.
 
@@
 
-With the help of \emph{Gaia}, high resolution spectroscopy, high-quality
-photometry, and large aperture multi-object spectroscopic instruments,
-we can observe dwarf galaxies in unprecedented detail. Today, dwarf
-galaxies are used to search for dark matter cores (REFS), test our
-understanding of early nucleosynthesis (REFS), the details of galaxy
-formation (REFS), and \ldots.
+Today, we know the Milky Way system is teeming with satellites.
+Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
+including dwarf galaxies, ambiguous systems, and globular clusters.
+Advances in telescopes and observations has accelerated progress on the
+nature of dwarf galaxy and introduced new questions. Deep digital
+photometric surveys have more than quadrupled the number of known Milky
+Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing surveys, like
+the Vera Rubin Observatory's Legacy Survey for Space and Time, will
+continue to probe fainter and fainter dwarf galaxies. In addition, large
+aperture multi-object spectrographs have revealed the detailed and
+complex inner chemodynamical structure of dwarf galaxies. Beyond precise
+structural and kinematic properties of dwarf galaxies, modern
+observations allow for the separation of multiple stellar populations,
+detailed constraints on the dark matter density profiles, and hints of
+tidal disruption or stellar halos.
 
-While we now know the basic properties of dwarf galaxies, tremendous
-progress has been made in the past 10-20 years with large area deep
-photometric surveys, multi-object spectrographs, high resolution
-chemistry, and theoretical work, unveiling that dwarf galaxies are each
-unique systems with different formation pathways
-\citep[e.g.][]{simon2019}.
+Of local dwarfs, the classical systems still remain the best studied and
+with most strongly constrained parameters. While extending 1-2 degrees
+across the sky, making deriving properties trickier
+\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
+(giant) stars, allowing thousands of stars to be observed with deep
+photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
+a result, the determination of fundamental properties such as the
+position, size, orientation, distances, proper motions, line-of-sight
+(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
+relatively well constrained today and show excellent convergence among
+different works. However, ongoing research continues to redefine our
+understanding of the detailed structure of Milky Way satellites, hinting
+that these objects may be more complex than at first glance.
 
-\citet{battaglia+nipoti2022}
+Observations of dwarf galaxies have been the origin of several
+\emph{small-scale} problems in cosmology
+\citep[e.g.,][]{bullock+boylan-kolchin2017}. For example, the mismatch
+between the expected number of dwarf galaxies from simulations and the
+observed abundance was known as the \emph{missing satellites problem}.
 
-One of the open questions in dwarf galaxy evolution is the formation of
-dark matter \emph{cores} where the density becomes constant instead
-\(\rho \sim 1\). A leading theory is cores form through of baryonic
-effects and feedback. If dark matter deviates from \(\Lambda\)CDM, then
-the NFW density profile may not describe halos anymore. \emph{Warm} dark
-matter, relativistic in the early universe but cooler now, smooths out
-small-scale features and softens the cusps of dark matter halos.
-\emph{Self-interacting} dark matter instead can form cores but also the
-cores can collapse into a density peak.
+\emph{Core Cusp Problem}. One of the open questions in dwarf galaxy
+evolution is the formation of dark matter \emph{cores} where the density
+becomes constant instead \(\rho \sim 1\). A leading theory is cores form
+through of baryonic effects and feedback. If dark matter deviates from
+\(\Lambda\)CDM, then the NFW density profile may not describe halos
+anymore. \emph{Warm} dark matter, relativistic in the early universe but
+cooler now, smooths out small-scale features and softens the cusps of
+dark matter halos. \emph{Self-interacting} dark matter instead can form
+cores but also the cores can collapse into a density peak.
 
@@
 \item
-  \citet{dicintio+2013}, \citet{navarro+2010}, NFW profiles may actually
-  be more Einasto like
+  \citet{oh+2011} survey of cusyp / cored dwarf galaxies.
 \end{itemize}
+
+Understanding the p
 
@@
 
-Given star counts for a local group dwarf, one of the main quantities of
-interest is the surface density profile. Long used to characterize and
-understand larger galaxies, these profiles can provide hints as to the
-formation of dwarf galaxies, determine the physical extent of the
-stellar component, and the total stellar mass. Previous works tend to
-focus on fitting 3 different surface brightness profiles: an
-Exponential, a Plummer, or a King profile.
+Surface density profiles efficiently characterize the shape of a galaxy.
+At the most basic, fitting stellar densities provide properties such as
+the shape, location, size, and orientation of a dwarf galaxy. However,
+the details of a stellar density profile are essential for interpreting
+the total mass of dwarf galaxies, their assembly and dynamical history,
+and understanding correlations between dwarf galaxies (REFS).
 
-The exponential profile is perhaps the simplest, as a 1-parameter
-profile \begin{equation}{
+Typically, three different surface density profiles are applied to dwarf
+galaxies: Exponential, Plummer, or a King profile
+\citep[e.g.,][]{munoz+18, cicuendez+2018}. The exponential profile is
+perhaps the simplest, as a 1-parameter profile \begin{equation}{
 \Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
@@
 For a long time, exponential surface density profiles have been used as
-an description for spiral galaxies {[}REFS{]}.
+an description for spiral galaxies (see discussion before).
 
-\citet{plummer1911} proposed a 1-parameter solution for a polytropic
-potential to fit globular cluster density profiles:
-
-\begin{equation}{
-\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2}
+To fit globular cluster density profiles, \citet{plummer1911} proposed a
+1-parameter solution for a density polytrope\footnote{where density and
+  pressure are assumed to be correlated}, \begin{equation}{
+\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2},
 }\end{equation}
 
-The \citet{king1962} profile, originating as an empirical fit to
-globular clusters, also sometimes reasonably fits dwarf galaxies. This
-density profile includes an additional parameter \(R_t\), a truncation
-radius where the density sharply drops to 0. \begin{equation}{
+where \(M\) is the total mass and \(R_h\) is the 2D half-light radius.
+Now superceded by the King profile for globular clusters, the Plummer
+model is still a good fit to dwarf spheroidals and XXX.
+
+The \citet{king1962} profile, also an empirical fit to globular
+clusters, is also commonly applied to . This density profile includes an
+additional parameter \(R_t\), a truncation radius where the density
+sharply drops to 0. \begin{equation}{
 \Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
@@
 
-Finally, the Sérsic profile represents an extension of the exponential.
+Where \(R_c\) is the characteristic ``core'' radius and \(\Sigma_0\) is
+a characteristic density related to the total mass.
+
+Finally, the Sérsic profile represents a generalization of an
+exponential profile, and describes galaxies of many kinds well.
 Typically parameterized in terms of a half-light radius \(R_h\), the
@@
 single parameter density profile which fits many dwarf galaxies well
-\citep[e.g.][\citet{eskridge1988a}, \citet{hodge1991a},
-\citet{hodge1991b}, \citet{IH1995}, etc.]{MV2020a} \citet{faber+lin1983}
-demonstrated that an Exponential density profile is a reasonable
-empirical fit to dwarf galaxies, theorizing that dwarf spheroidal evolve
-from exponential disks, maintaining a similar light profile.
+\citetext{\citealp[e.g.,][]{eskridge1988a}; \citealp{hodge1971a}; \citealp{hodge1991b}; \citealp{IH1995}; \citealp[
+but with nuclear regions and only for Mb \textless{} -16
+virgo]{binggeli+cameron1991}; \citealp{caldwell+bothun1987}; \citealp{simon+2005}}.
 
-in addition to Sérsic, King and other profiles. \citet{lelli+2014}
-(example of using exponential.) \citet{wang+2019} for fornax,
-\citet{mcconnachie+irwin2006} for Exp, Plummer, King in other galaxies.
+\citet{faber+lin1983} demonstrated that an Exponential density profile
+is a reasonable empirical fit to dwarf galaxies, theorizing that dwarf
+spheroidal evolve from exponential disks, maintaining a similar light
+profile.
+
+As a result, many studies assume an exponential density profile in
+theoretical or observational modelling
+\citep[e.g.,][]{MV2020a, kowalczyk+2013}.
+
+\citet{lelli+2014} (example of using exponential.)
 
 However, an exponential profile was often noted to not fit every dwarf
-galaxy. Even starting from \citet{aparicio1997}, some people have noted
-deviations from this simple empirical rule. Additionally,
-\citet{herrmann+hunter+elmegreen2013},
+galaxy. Even starting from \citet{aparicio1997}, and \citet[for coma
+cluster dE]{graham+guzman2003}, some people have noted deviations from
+this simple empirical rule. Additionally, \citet{hunter+elmegreen};
+\citet{herrmann+hunter+elmegreen2013};
 \citet{herrmann+hunter+elmegreen2016} note that at least in a
 photometric sample of more distant dwarf disky/Irr/blue compact dwarfs,
-that many dwarfs show deviations from exponential profiles.
+that many or most dwarfs show deviations from exponential profiles. Irr
+do represent a substantially different class though so it is unclear if
+these conclusions apply.
+
+Interestingly, many works also show that differences between the
+different density profiles remain barely significant for some galaxies
+\citep[e.g.,][]{cicuendez+2018}. \citet{wang+2019} for Fornax showing
+similar performance but favour Sérsic, \citet{mcconnachie+irwin2006} for
+Exp, Plummer, King in other galaxies.
 
 \begin{itemize}
-\tightlist
 \item
@@
 
+Irregardless of if all dwarf galaxies are indeed exponential, there is
+nevertheless diversity requiring interpretation in the density profiles
+of dwarf galaxies.
+
 \subsection{Sculptor and Ursa Minor: Hints of tidal
@@
 Sculptor and Ursa Minor may appear to be typical dwarf galaxies at first
-glance. Table~\ref{tbl:scl_obs_props}. describe the present-day
-properties of each galaxy. Indeed, Sculptor as the first discovered
-dwarf galaxy, is often described as a ``prototypical'' dSph. However,
-both galaxies have a long history of speculation that they may be
-influenced by the Milky Way's gravity (see discussion). Only recently,
-with \emph{Gaia} data and spectroscopic followup, has the detection of a
-density excess become more unmistakable. Using \citet{jensen+2024}'s
-algorithm (described briefly below), \citet{sestito+2023a},
+glance. Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} describe
+the present-day properties of each galaxy. Indeed, Sculptor as the first
+discovered dwarf galaxy, is often described as a ``prototypical'' dSph.
+However, both galaxies have a long history of speculation that they may
+be influenced by the Milky Way's gravity (see discussion). With
+\emph{Gaia} data and spectroscopic followup, the detection of a density
+excess has become more unmistakable. Using \citet{jensen+2024}'s
+algorithm (described briefly below), \citet{sestito+2023a};
 \citet{sestito+2023b} report that a ``kink'' in the density profile,
@@
 Perhaps the most straighforward explanation for the density profiles of
-Sculptor and Ursa Minor are tidal interactions. \citet{sestito+2023a},
+Sculptor and Ursa Minor are tidal interactions. \citet{sestito+2023a};
 \citet{sestito+2023b} conclude that tides are likely the explanation.
@@
 Some of the earliest theory work on tidal stripping of dwarf galaxies
-originate from \citep[\citet{moore+davis1994},
-\citet{johnston+spergel+hernquist1995}, \citet{oh+lin+aarseth1995},
-\citet{piatek+pryor1995}, \citet{velazquez+white1995},
-\citet{kroupa1997}]{allen+richstone1988}. Many of these works used
-similar techniques as we continue to use today, setting up a dark matter
-halo in a static Milky Way potential and predicting the evolution of
-quantities of interest.
+originate from
+\citep{allen+richstone1988, moore+davis1994, johnston+spergel+hernquist1995, oh+lin+aarseth1995, piatek+pryor1995, velazquez+white1995, kroupa1997, pryor1996}.
+Many of these works used similar techniques as we continue to use today,
+setting up a dark matter halo in a static Milky Way potential and
+predicting the evolution of quantities of interest.
 
-More recently, \citet{PNM2008} run idealized simulations of a local
-group dwarf galaxy in a milky way halo.
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{mayer+2001} theory of tidal stripping
-\end{itemize}
+More recent works include \citet{read+2006};
+\citet{bullock+johnston2005}; \citet{PNM2008}; \citet{penarrubia+2009};
+\citet{klimentowski+2009}; \citet{errani+2023a}; \citet{fattahi+2018};
+\citet{stucker+2023}; \citet{wang+2017}.
 
@@
 \citet{iorio+2019} for Sculptor, \citet{borukhovetskaya+2022};
-\citet{dicintio+2024} for Fornax, \citet{borukhovetskaya+2022a} for
-Antlia II,
-
-Our goal is to apply a similar framework to Sculptor and Ursa Minor.
+\citet{dicintio+2024} for Fornax; \citet{borukhovetskaya+2022a} for
+Antlia II. Our goal is to apply a similar framework to Sculptor and Ursa
+Minor.
 
@@
 \(O(N^2)\) to \(O(N\,\log N)\), representing orders of magnitude
-speedup. Modern codes such as \emph{Gadget} utilize many other
-performance tricks, such as splitting particles across many
-supercomputer nodes, efficient memory storage, adaptive time stepping,
-and parallel file writing to retain fast performance for large scale
-simulations, forming the foundation for many cosmological simulation
-codes.
+speedup. Modern codes such as \emph{Gadget} utilize other performance
+tricks, such as splitting particles across many supercomputer nodes,
+efficient memory storage, adaptive time stepping, and parallel file
+writing to retain fast performance for large scale simulations, forming
+the foundation for many cosmological simulation codes.
 
@@
 Analytic approximations are powerful tools to generally determine the
-influence of tides given a dwarf galaxy and a host.
+influence of tides given a dwarf galaxy and a host. A variety of ``tidal
+radii'' exist in the literature, we focus on two simple forms, the
+Jacobi radius and the break radius.
 
-The Jacobi radius represents the approximate radius where stars become
-unbound for a galaxy in a circular orbit around a host galaxy.
+The \textbf{Jacobi radius} represents the approximate radius where stars
+become unbound for a galaxy in a circular orbit around a host galaxy.
 Calculated from an approximation of the location of the L1 and L2
@@
 
-We use the break radius as defined in \citet{penarrubia+2009}, marking
-where the galaxy is still in disequilibrium. The break radius
+We use the \textbf{break radius} as defined in \citet{penarrubia+2009},
+marking where the galaxy is still in disequilibrium. The break radius
 \(r_{\rm break}\) is proportional to the velocity dispersion
@@
 From this argument, we note that the following properties must be
-approximately true for tides to occur:
+approximately true to interpret a break is
 
@@
 \item
-  Close enough pericentre. The other break radius \(r_J\) implies that
-  if the host density is 3x the satellite, stars will be lost
+  \emph{Small pericentre}. If the pericentre is too large and
+  \(r_J \gg R_h\), then few stars will be lost.
 \item
@@
   smaller, compacter DM halo with a more extended stellar component.
-
-  \begin{itemize}
-  \tightlist
 \item
   This information is mostly related to the statistical initial
-    distribution of satellites from cosmology {[}ludlow+2016;
-    \citet{fattahi+2018}{]}
-  \end{itemize}
+  distribution of satellites from cosmology
+  \citep{ludlow+2016, fattahi+2018}
 \end{itemize}
@@
 
-As a galaxy evolves in a tidal field, several changes happen.
+As a galaxy evolves in a tidal field, several changes happen:
 
@@
 \item
-  \emph{Steams form}. If tides are strong enough, mass lost from the
-  galaxy becomes part of tidal tails, evolving along a similar orbit but
-  leading or trailing the galaxy.
+  \emph{Steams form}. If tides are strong enough, unbound mass becomes
+  part of tidal tails, evolving along a similar orbit but leading or
+  trailing the galaxy.
 \item
-  \emph{The galaxy reshapes}. Bound mass of the galaxy redistributes. As
+  \emph{Mass redistributes}. Bound mass of the galaxy redistributes. As
   mentioned in the last section, this is visible as a wave of outward
@@
 constant tidal field, most dwarf galaxies should evolve along similar
-tidal tracks.
+tidal tracks. \citet{EN2021} derive tidal tracks for dwarf galaxies,
+showing that NFW halos all evolve along a similar trajectory in terms of
+\(r_{\rm max}\) and \(v_{\rm max}\).
 
-\citet{EN2021} derive tidal tracks for dwarf galaxies, showing that NFW
-halos all evolve along a similar trajectory in terms of \(r_{\rm max}\)
-and \(v_{\rm max}\). \begin{equation}{
+\begin{equation}{
 \frac{v_{\rm max}}{v_{\rm max, 0}} = 
@@
 
-Case studies:
-
-Dynamical
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{read+2006}
-\item
-  \citet{bullock+johnston2005}
-\item
-  \citet{PNM2008}
-\item
-  \citet{errani+2023a}, \citet{fattahi+2018}
-\item
-  \citet{wang+2017},
-\item
-  \citet{pryor1996}
-\item
-  \citet{klimentowski+2009}
-\end{itemize}
-
 \begin{figure}
 \centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}
-\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies.
-Fig. 6 from \citet{EN2021}.}\label{fig:tidal_tracks}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}
+\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies,
+the logarithm of maximum circular velocity and radius of relative to the
+initial conditions for satellites on a variety of orbits. Almost all
+satellites follow the tidal track suggested by \citet{EN2021}. Adapted
+from fig.~6 of \citet{EN2021}.}\label{fig:tidal_tracks}
 \end{figure}

### 2025-07-25 thesis/introduction.tex
more introduction updates

@@
   names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
-  dSph, we omit this suffix for brevity.} The ultra-faints occupy the
-very faintest magnitudes (\(-7.7 \lesssim M_V\)). These galaxies have
-miniscule stellar masses, tend to be more compact, and are the darkest
-known galaxies. An example is Boötes V as shown in
+  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
+occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
+galaxies have minuscule stellar masses, tend to be more compact, and are
+the darkest known galaxies. An example is Boötes V as shown in
 Fig.~\ref{fig:galaxy_images}. Altogether, dwarf galaxies span about 7
@@
 Way halo, numerous streams \citep{bonaca+price-whelan2025}. \emph{Gaia}
-has revealed the hierarchical formation and messy structure of our own
-galaxy.
+has revealed the hierarchical formation and complex, evolving structure
+of our own galaxy.
 

### 2025-07-25 notebooks/analytic_profile_comparison.jl

@@ -1,0 +1,153 @@
+### A Pluto.jl notebook ###
+# v0.20.6
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 28ed6182-698c-11f0-3368-f5b5af3ea05c
+begin
+	import Pkg; Pkg.activate()
+
+
+	using LilGuys
+	using CairoMakie, Arya
+
+end
+
+# ╔═╡ 4671142c-e341-4927-bfe7-809b41daf251
+CairoMakie.activate!(type=:png)
+
+# ╔═╡ 903cd96e-0768-4edb-9126-4b6eafeb4ca9
+prof_exp = LilGuys.Exp2D()
+
+# ╔═╡ cb7dc629-fbc9-4e08-ba56-2074009ee874
+prof_sersic_n2 = LilGuys.Sersic(n=0.8)
+
+# ╔═╡ 5fb7529d-0b1a-4fa1-bbaf-ed13f2f01027
+prof_plummer = LilGuys.Plummer()
+
+# ╔═╡ 645b8702-c0df-4152-b6c3-7e21f11ab305
+import SpecialFunctions: gamma
+
+# ╔═╡ 929dfb76-3580-49e2-bf92-098e131f4c01
+Base.@kwdef struct GeneralizedPlummer <: LilGuys.SphericalProfile
+	r_s::Float64
+	M::Float64
+	n::Float64
+
+	_k::Float64 = gamma(n / 2) / (gamma((n-3)/2) * π^(3/2))
+	
+end
+
+# ╔═╡ 6e612d6d-4605-47de-af5f-c38d022650d8
+LilGuys.density(prof::GeneralizedPlummer, r::Real) = prof._k * prof.M * prof.r_s^(prof.n - 3) / (prof.r_s^2 + r^2)^(prof.n/2)
+
+# ╔═╡ f2725829-5ca7-43a5-b43e-1dfddbac1f7e
+LilGuys.R_h(prof::LilGuys.Sersic) = prof.R_h
+
+# ╔═╡ 15ed2abd-a8b4-4bc4-94ab-48276475e9a8
+steep_plummer = GeneralizedPlummer(M=1, r_s=1, n=9)
+
+# ╔═╡ c0e9018f-5ac6-4f1c-b800-bb7348a4f5f7
+profiles = [
+	"exponential" => prof_exp,
+	"sersic" => prof_sersic_n2,
[…]

### 2025-08-13 thesis/appendix.tex
thesis updates and ludlow notebook

@@
 density profiles for Sculptor and Ursa Minor
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/analytic_profile_comparison.pdf}}
+\caption[Comparison of analytic density profiles]{A comparison of
+different parameterizations for dwarf galaxy density profiles. Note that
+deviations between profiles only become apparent past 3 R\_h, and only
+the Plummer profile, in contrast to more commonly assumed profiles,
+deviates by \textasciitilde1 dex positive before 6 Rh. Since this
+profile is a far minority in the literature, deviations from exponential
+and close relatives are interesting and worth further consideration.}
+\end{figure}

### 2025-08-13 thesis/discussion.tex
thesis updates and ludlow notebook

@@
 \citet{westfall+2006} also showed evidence of an extended component of
-Sculptor's density profile, interpreting these stars as either evidence
-of tidal debris or a dwarf galaxy halo. On the other hand,
-\citet{coleman+dacosta+bland-hawthorn2005} show that Sculptor is well
-described by a two component density profile, they additionally mention
-that it is unlikely many of Sculptor's stars (less than 2\%) are
-extratidal. While \citet{munoz+2018} is the most recent and deepest
-photometric study of Sculptor, they only cover an area of out to radii
-of about 30', so they are unable to study the possible extended
-components of Sculptor detected in past works. In addition to the
-presence of ``extratidal stars'', many studies note that Sculptor's
+Sculptor's density profile (among other dwarfs), interpreting these
+stars as either evidence of tidal debris or a dwarf galaxy halo. On the
+other hand, \citet{coleman+dacosta+bland-hawthorn2005} show that
+Sculptor is well described by a two component density profile, they
+additionally mention that it is unlikely many of Sculptor's stars (less
+than 2\%) are extratidal. While \citet{munoz+2018} is the most recent
+and deepest photometric study of Sculptor, they only cover an area of
+out to radii of about 30', so they are unable to study the possible
+extended components of Sculptor detected in past works. In addition to
+the presence of ``extratidal stars'', many studies note that Sculptor's
 stars appear to become more elliptical with radius, consistent with
@@
   populations).
+\item
+  \citet{cicuendez+battaglia2018}
 \end{itemize}

### 2025-08-13 thesis/introduction.tex
thesis updates and ludlow notebook

@@
 environments in the universe. These little galaxies are typically
-defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\) ,
+defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
+similarly \(L \lesssim 10^9 L_\odot\),
 e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
-dwarfs are the most numerous galaxies. Dwarf galaxies are highly
-\emph{dark-matter dominated}, with mass to light ratios sometimes
-exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
-masses, many dwarf galaxies are \emph{quenched}, with little to no
-recent star formation. Their stellar populations are \emph{relics} from
-the early universe, consisting of many of the oldest and most metal poor
-stars. Understanding the properties of dwarf galaxies thus has
-implications across astronomy, from cosmological structure formation on
-the smallest scale to extremely metal-poor stellar populations to
-chemical evolution to galactic dynamics. Of the classical dwarfs, the
-Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
-extended density profile relative to an exponential, hinting at tidal
-effects. As a case study in tidal effects and our understanding of dwarf
-galaxy formation and evolution, we aim to understand the dynamical
-history of these galaxies.
+dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
+highly \emph{dark-matter dominated}, with mass to light ratios exceeding
+100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
+many dwarf galaxies are \emph{quenched}, with little to no recent star
+formation. Their stellar populations are \emph{relics} from the early
+universe, consisting of many of the oldest and most metal poor stars.
+Understanding the properties of dwarf galaxies thus has implications
+across astronomy, from cosmological structure formation on the smallest
+scale to extremely metal-poor stellar populations to chemical evolution
+to galactic dynamics. Of the classical dwarfs, the Sculptor and Ursa
+Minor dwarf spheroidal galaxies stand out with a more extended density
+profile relative to an exponential, hinting at tidal effects. As a case
+study in tidal effects and our understanding of dwarf galaxy formation
+and evolution, we aim to understand the dynamical history of these
+galaxies.
 
@@
 
-We only understand a tiny fraction of the composition of the universe.
-The leading theory of cosmology, \(\Lambda\)CDM (cold dark matter),
-posits that the universe is composed of about 68\% dark energy
-(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
-  like to change definitions of words. \emph{Baryons} here means
+We only understand a tiny fraction of the universe's composition. The
+leading theory of cosmology, \(\Lambda\)CDM (cold dark matter), posits
+that the universe is composed of about 68\% dark energy (\(\Lambda\)),
+27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
+  to change definitions of words. \emph{Baryons} here means
   baryons+leptons, i.e.~any standard model massive fermion. The photon
@@
 progressively larger halos
-\citep[e.g.,][]{blumenthal+1984, blumenthal+1986, white+rees1978, white+frenk1991, somerville+dave2015}.
+\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
 Structure formation happens on a wide range of scales. The largest
@@
 small-scale cosmological features. Alternative dark matter properties
-can influence dwarf galaxy structure, formation, and tidal evolution.
-The nearby dwarf galaxies of the local group, with detailed
+can influence dwarf galaxy abundance, formation, structure,, and tidal
+evolution. The nearby dwarf galaxies of the local group, with detailed
 observations, present a promising opportunity to understand the
@@
 is defined as the mass within a region \(r_{200}\) containing a mean
-enclosed density 200 times\footnote{200 smells like an arbitrary value.
-  From approximate analytic arguments about the mean density of a
-  self-gravitating spherical collapse halo in an expanding universe, the
-  expected virial overdensity is XXX, rounding up to 200 (REFS).} the
-critical density of the universe: \begin{equation}{
+enclosed density 200 times\footnote{From approximate analytic arguments
+  about the mean density of a self-gravitating spherical collapse halo
+  in an expanding universe, the expected virial overdensity is \ldots,
+  commonly rounded to 200 (REFS).} the critical density of the universe:
+\begin{equation}{
 M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
@@
 massive halos become less concentrated in a predictable way
-\citep[e.g.,][]{bullock+2001, ludlow+2016}. While there is a general
-trend in concentration with mass, the concentration values still have
-substantial scatter. Other parameters such as the halo spin or shape or
-slight changes in the halo density profile add additional variability to
-the present-day distribution of halos \citep[see
-e.g.,][darkEXP]{navarro+2010, dicintio+2013, dutton+maccio2014}.
+\citep[e.g.,][]{bullock+2001, ludlow+2016}. The lower panel of
+Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
+relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
+a general trend in concentration with mass, the concentration values
+still have substantial scatter. Other parameters such as the halo spin
+or shape may add additional variability to present-day distribution of
+halos, but this variation is typically expected to be small
+\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
-\caption[Example density profiles]{Example stellar and dark matter
-density profiles for a Sculptor-like galaxy. The dark matter is more
-extended and massive than the star across the entire
-galaxy.}\label{fig:nfw_density}
+\caption[Example density profiles]{Density profiles in log density
+versus log radius for the stars and dark matter of a Sculptor-like
+galaxy. The dark matter is more extended and massive than the star
+across the entire galaxy---dynamical evolution is driven by the dark
+matter mass distribution.}\label{fig:nfw_density}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
+\caption[Stellar-mass halo-mass relation]{\textbf{Top} Stellar mass
+(top) as a function of maximum circular velocity. The solid line with
+the 1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
+with scatter points simulated central galaxies from APOSTLE in
+\citet{fattahi+2018}. \textbf{Bottom:} Radius of maximum circular
+velocity \(r_{\rm max}\) as a function of maximum circular velocity. The
+solid line with 1-\(\sigma\) shaded region and dashed line are the
+relations from \citet{ludlow+2016} for \(z=0\) and \(z=2\) respectively.
+\emph{TODO}: add legend / annotations.}\label{fig:smhm}
 \end{figure}
@@
 Both cosmology and observations find fundamental relationships between
-properties of galaxies, particularly for the total mass and luminosity.
-A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
-relation, describing the amount of stellar mass forms in a (sub)halo of
-a given size. In particular, the SMHM relation grows especially steep in
-the dwarf galaxy regime---many dwarf galaxies are formed in halos of
-similar masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
+the dynamical or total mass and stellar masses of galaxies. A key
+prediction of \(\Lambda\)CDM is the stellar mass halo mass relation,
+describing the amount of stellar mass forms in a (sub)halo of a given
+size. In particular, the SMHM relation grows especially steep in the
+dwarf galaxy regime---many dwarf galaxies are formed in halos of similar
+masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
 \(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
 apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
-the range of possible \(v_{\rm max}\) is fairly narrow across \(\sim 5\)
-decades in stellar mass. Thus, if we know the initial stellar mass of a
-dwarf galaxy, we also know, at some level, its dark matter mass.
+the range of predicted \(v_{\rm max}\) is fairly narrow across
+\(\sim 4\) decades in stellar mass. Because lower mass galaxies have
+increasingly shallow potential wells, feedback becomes more effective
+removing gas more quickly. In addition, re-ionization suppresses late
+star formation in the faintest galaxies. Thus, if we know the initial
+stellar mass of a dwarf galaxy, we also know, at some level, the
+properties of its dark matter halo.
 
 Several challenges complicate a simple SMHM trend including environment,
-assembly history, and tidal effects. By being closer to a massive host,
-most dwarfs quench earlier, resulting in lower stellar masses
-\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
-(relative to reionization) can influence the resulting stellar content
-\citep{kim+2024}. Finally, tides influence both the dark matter and
-stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
+assembly history, and tidal effects. Additionally, the detailed physics
+of galaxy formation can impact the SMHM relationship. By being closer to
+a massive host, many dwarfs quench earlier, resulting in lower stellar
+masses \citep[e.g.,][]{christensen+2024}. Additionally, the time of
+formation (relative to reionization) can influence the resulting stellar
+content \citep{kim+2024}. Finally, tides influence both the dark matter
+and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
 Consequently, tides may reduce the halo mass more than the stellar mass,
-adding additional scatter to the SMHM trend, particularly for Milky Way
+adding additional scatter to the SMHM trend, particularly for low-mass
 satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
-\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
-relation, where \(v_{\rm max}\) is related to the halo mass. Right: the
-mass-concentration relation for NFW halos, but parameterized in terms of
-\(r_{\rm max}\) and \(v_{\rm max}\). Together, these plots represent the
-cosmologically expected properties of underlying dark matter halos in
-any dwarf galaxy. The velocity dispersion directly constrains the mass
-contained within a half-light radius, so the underlying halo in
-\(\Lambda\)CDM is well-constrained. Adapted from figure 1 of
-\citet{fattahi+2018}, stellar mass and \(v_{\rm max}\) of APOSTLE
-satellites at peak \(v_{\rm max}\).}\label{fig:smhm}
-\end{figure}
-
 \section{Observational context}\label{observational-context}
 
-Dwarf galaxies have long challenged questioned theories of galaxy
-formation. The first discovered dwarf galaxies in 1938\footnote{technically,
-  the LMC and SMC may be classified as dwarf galaxies, but these were
-  likely always known to humans at southern latitudes.}, Sculptor and
-Fornax, already presented such a puzzle. Having no close analogues,
-\citet{shapley1938} present Sculptor and Fornax as a new type of
-\emph{stellar system} resembling the Magellanic clouds and globular
-clusters, but not attempting to speculate on the exact nature. While
-generally understood to be galaxies based on the inferred luminosities
-and sizes, the nature and formation remained unclear
-\citep[e.g.,][]{hodge1971}.
+Dwarf galaxies have long raised conundrums in theories of galaxy
+formation. The discovery of Fornax and Sculptor in 1938
+\citep{shapley1938}\footnote{technically, the LMC and SMC may be
+  classified as dwarf galaxies, but these were likely always known to
+  humans at southern latitudes.}, with no known analogues, already
+presented such an enigma. Shapley presented these dwarfs as a new type
+of \emph{stellar system} resembling the Magellanic clouds and globular
+clusters but did not attempt to speculate on the exact nature. While
+dwarf galaxies were quickly understood to be galaxies based on the
+inferred luminosities and sizes, their nature and formation remained
+unclear for decades \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
 
 The earliest spectroscopic work hinted that dwarf galaxies may contain
-substantial dark matter. From the first measurements of velocity
-dispersions
-\citep[e.g.,][]{aaronson1983, gallagher+wyse1994, pryor1996}, the
-inferred mass-to-light ratios were about 10 times the expectation for
-globular clusters \citep{matteo1998}. Subsequently, many theories
+substantial dark matter masses. From early determinations of the
+velocity dispersion for Sculptor and Ursa Minor dSph
+\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
+mass-to-light ratios were at least 10 times the expectation for globular
+clusters. While only determined by a few stars initially, these values
+have solidified with time with larger and more precise samples
+\citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
 attempting to understand the formation and observed properties of these
-objects were proposed. Examples included that dwarf galaxies were formed
-tidally in the gas clouds of satellites; dwarf galaxies are undergoing
+objects were proposed. Examples include: dwarf galaxies are undergoing
 tidal dissolution resulting in extreme mass-to-light ratios
-\citep[e.g.,][]{pryor1996}, or Since then, we have known that dwarf
+\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
+holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark matter
+free out-of-equilibrium ``tidal dwarfs'' from past mergers
+\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
+of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
+formation was also not out of the question
+\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
 galaxies are among the darkest objects in the universe, and
@@
 galaxies have minuscule stellar masses, tend to be more compact, and are
-the darkest known galaxies. An example is Boötes V as shown in
-Fig.~\ref{fig:galaxy_images}. Altogether, dwarf galaxies span about 7
+the darkest known galaxies. Altogether, dwarf galaxies span about 7
 orders of magnitude in stellar mass.
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
-\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Sculptor (DES
-2), Ursa Minor (UNWISE with Gaia point sources over-plotted), and Bootes
-V (SDSS). While the LMC is very prominent in the sky, even the classical
+\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
+DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
+over-plotted). Each image includes the galaxies luminosity and a 200 pc
+scalebar. While the LMC is very prominent in the sky, the classical
 dwarf galaxies are not obvious except in terms of star counts
 (e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
-central third to sixth of the image.}\label{fig:galaxy_images}
+central \(\sim\)third of the image.}\label{fig:galaxy_images}
 \end{figure}
@@
 \(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
-\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{\(\mu_\alpha\)
-  and \(\mu_\delta\) are the apparent changes in \(\alpha\) and
-  \(\delta\). \(\mu_{\alpha*}\) allows where
-  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
-  effects in \(\alpha\).}
+\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
+  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
+  change in right ascension \(\alpha\) and declination \(\delta\),
+  typically in units of mili-arcsecond (mas) per year. \(\mu_{\alpha*}\)
+  allows where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for
+  projection effects in \(\alpha\).}
 
@@
 \citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
-\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and effects of the
-Milky Way's bulge and bar \citep{hunt+vasiliev2025}. Moving to the Milky
-Way halo, numerous streams \citep{bonaca+price-whelan2025}. \emph{Gaia}
-has revealed the hierarchical formation and complex, evolving structure
-of our own galaxy.
+\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and dynamical effects
+of the Milky Way's spiral arms and the bar in the solar neighbourhood
+\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
+helped find and constrain numerous stellar streams
+\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
+the hierarchical formation and complex, evolving structure of our own
+galaxy.
 
@@
 determination. Before, proper motions of dwarf galaxies were spase or
-undetermined and measured by the Hubble Space Telescope or other methods
-for only a few systems. \emph{Gaia} allowed for the first systematic
-determinations of Milky Way satellite proper motions, and of
-unprecedented precision \citep{MV2020a, pace+li2019}. While the proper
-motion uncertainty on a typical dwarf member star is often with large
-uncertainties, by combining the proper motions for 100s to 1000s of
-stars in \emph{Gaia}, incredibly precise proper motion measurements can
-be determined, sometimes only limited by \emph{Gaia}'s systematic error
-floor \citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new
-dynamical era for MW satellites studies, where we often have precise
+undetermined. Few galaxies had precisely measured proper motions, often
+from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
+\emph{Gaia} allowed for the first systematic determinations of Milky Way
+satellite proper motions, and of unprecedented precision
+\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
+typical dwarf member star is often with large uncertainties, by
+combining the proper motions for 100s to 1000s of stars in \emph{Gaia},
+incredibly precise proper motion measurements can be determined,
+sometimes only limited by \emph{Gaia}'s systematic error floor
+\citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new dynamical
+era for MW satellites studies, where we often have precise
 determinations of the dwarf galaxy's recent orbit under a given MW
@@
 (LOS) velocities, and velocity dispersions \(\sigma_v\), are all
-relatively well constrained today and show excellent convergence among
-different works. However, ongoing research continues to redefine our
-understanding of the detailed structure of Milky Way satellites, hinting
-that these objects may be more complex than at first glance.
+relatively well constrained and converged among different studies today.
+However, ongoing research continues to redefine our understanding of the
+detailed structure of Milky Way satellites, hinting that these objects
+may be more complex than at first glance.
 
-Observations of dwarf galaxies have been the origin of several
-\emph{small-scale} problems in cosmology
-\citep[e.g.,][]{bullock+boylan-kolchin2017}. For example, the mismatch
-between the expected number of dwarf galaxies from simulations and the
-observed abundance was known as the \emph{missing satellites problem}.
+Observations of dwarf galaxies have been the origin of several disputes
+or \emph{small-scale} problems in cosmology \citep[see
+review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
+the expected number of dwarf galaxies from simulations and the observed
+abundance was known as the \emph{missing satellites problem}.
+Additionally, a number of observations showed that many dwarf galaxies,
+but not all, possess dark matter ``cores'' as measured from circular
+velocity curves \citep[e.g.,][]{moore1994, adams+2014, oh+2015} or dwarf
+spheroidal \citep[albeit less constraining,
+e.g.,][]{walker+penarrubia2011, read+walker+steger2019}. As a result,
+numerous alternative forms of dark matter have been advocated as
+solutions. \emph{Warm} dark matter, relativistic in the early universe
+but cooler now, smooths out small-scale features and softens the cusps
+of dark matter halos. \emph{Self-interacting} dark matter instead can
+form cores but also the cores can collapse into a density peak.
 
-\emph{Core Cusp Problem}. One of the open questions in dwarf galaxy
-evolution is the formation of dark matter \emph{cores} where the density
-becomes constant instead \(\rho \sim 1\). A leading theory is cores form
-through of baryonic effects and feedback. If dark matter deviates from
-\(\Lambda\)CDM, then the NFW density profile may not describe halos
-anymore. \emph{Warm} dark matter, relativistic in the early universe but
-cooler now, smooths out small-scale features and softens the cusps of
-dark matter halos. \emph{Self-interacting} dark matter instead can form
-cores but also the cores can collapse into a density peak.
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{walker+penarrubia2011}: measuring cores and cusps
-\item
-  \citet{oh+2011} survey of cusyp / cored dwarf galaxies.
-\end{itemize}
-
-Understanding the p
+However, these tensions have eased as a result of improved consideration
+of baryonic physics. For instance, understanding observational
+limitations and completeness has eased tension with the abundance of
+dwarf galaxies \citep[e.g.,][]{kim+peter+hargis2018}. Also, recent
+hydrodynamic simulations in particular have shown that strong feedback
+can produce dark matter cores
+\citep[e.g.,][]{tollet+2016, fitts+2017, orkney+2021}. Altogether, the
+numerous past challenges for \(\Lambda\)CDM in the dwarf galaxy regime
+illustrates the opportunity for dwarf galaxies to test the understanding
+of galaxy formation and dark matter physics.
 
@@
 the total mass of dwarf galaxies, their assembly and dynamical history,
-and understanding correlations between dwarf galaxies (REFS).
+and understanding correlations between dwarf galaxies
+\citep[e.g.,][\citet{penarrubia+2009}, \citet{querci+2025},
+\citet{lee+2018}]{herrmann+hunter+elmegreen2013}.
 
-Typically, three different surface density profiles are applied to dwarf
-galaxies: Exponential, Plummer, or a King profile
-\citep[e.g.,][]{munoz+18, cicuendez+2018}. The exponential profile is
-perhaps the simplest, as a 1-parameter profile \begin{equation}{
+Four different surface density profiles are frequently applied to dwarf
+galaxies: Exponential, Plummer, King, or Sérsic profiles
+\citep[e.g.,][]{munoz+18}. The exponential profile is perhaps the
+simplest, as a 1-parameter profile \begin{equation}{
 \Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
@@
 For a long time, exponential surface density profiles have been used as
-an description for spiral galaxies (see discussion before).
+a description for spiral galaxies {[}\citet{mateo1998}; discussion
+below{]}.
 
@@
 where \(M\) is the total mass and \(R_h\) is the 2D half-light radius.
-Now superceded by the King profile for globular clusters, the Plummer
-model is still a good fit to dwarf spheroidals and XXX.
+Now mostly superseded by the King profile for globular clusters, the
+Plummer model is still a good fit to many dwarf spheroidals {[}e.g.,
+moskowitz+walker2020{]}.
 
 The \citet{king1962} profile, also an empirical fit to globular
-clusters, is also commonly applied to . This density profile includes an
-additional parameter \(R_t\), a truncation radius where the density
-sharply drops to 0. \begin{equation}{
-\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
+clusters, is also often applied to dwarf galaxies, especially in older
+literature. Using three parameters, a core radius \(R_c\), a truncation
+radius \(R_t\), and a characteristic density, \(\Sigma_0\), the King
+profile is \begin{equation}{
+\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
 }\end{equation}
 
-Where \(R_c\) is the characteristic ``core'' radius and \(\Sigma_0\) is
-a characteristic density related to the total mass.
+In much of the older literature, \(R_t\) was often called and
+interpreted as ``tidal radius'' as well, after the similar
+interpretation for globular clusters
+\citep[e.g.,][\citet{hodge1961}]{IH1995}.
 
-Finally, the Sérsic profile represents a generalization of an
-exponential profile, and describes galaxies of many kinds well.
+Finally, the \citet{sersic1963} profile represents a generalization of
+an exponential profile, and describes most galaxy light profiles well.
 Typically parameterized in terms of a half-light radius \(R_h\), the
@@
 profile's equation is \begin{equation}{
-\Sigma_{\rm Sérsic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
-}\end{equation} where \(b_n\) is a constant depending on \(n\). Less
-commonly used in studies of dwarf galaxies, \citet{munoz+2018} advocate
-for using the Sérsic profile since the added flexibility allows for more
-profiles to be fit.
+\Sigma_{\rm S\acute ersic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
+}\end{equation} where \(b_n\) is a constant depending on \(n\). \(n=1\)
+provides an exponential profile and \(n=4\) recovers
+\citet{devaucouleurs1948}'s profile for elliptical galaxies. While a
+Sérsic profile is not always used for dwarf galaxies, \citet{munoz+2018}
+advocate for using the Sérsic profile since the added flexibility allows
+for more profiles to be fit.
 
 While there are not clear theoretical explanations why any profile is
-best, the Exponential stellar density profile is commonly assumed as a
-single parameter density profile which fits many dwarf galaxies well
-\citetext{\citealp[e.g.,][]{eskridge1988a}; \citealp{hodge1971a}; \citealp{hodge1991b}; \citealp{IH1995}; \citealp[
-but with nuclear regions and only for Mb \textless{} -16
-virgo]{binggeli+cameron1991}; \citealp{caldwell+bothun1987}; \citealp{simon+2005}}.
+best, the Exponential density profile is commonly assumed for dwarf
+spheroidal galaxies. \citet{faber+lin1983} was one of the first to
+demonstrate that an Exponential density profile is a reasonable
+empirical fit to dwarf galaxies, theorizing that dwarf spheroidal evolve
+from exponential disk galaxies and maintained a similar light profile.
+Later, \citet{read+gilmore2005} showed that exponential profiles may
+originate from general mass loss during the evolution of dwarf galaxies.
+Many later photometric works for dwarf spheroidal galaxies apply
+exponential fits, finding that exponential and king profiles both
+provide good descriptions
+\citetext{\citealp[\citet{mateo1998}]{binggeli+sandage+tarenghi1984}; \citealp{mcconnachie+irwin2006}; \citealp{cicuendez+2018}}.
+As a result, many studies assume an exponential density profile for
+dwarf galaxies in theoretical or observational modelling
+\citetext{\citealp[e.g.,][
+\citet{MV2020a}]{martin+2016}; \citealp{battaglia+2022}; \citealp[ but
+is for disk]{kowalczyk+2013}}.
 
-\citet{faber+lin1983} demonstrated that an Exponential density profile
-is a reasonable empirical fit to dwarf galaxies, theorizing that dwarf
-spheroidal evolve from exponential disks, maintaining a similar light
-profile.
+An exponential is far from the only density profile advocated for. Some
+authors fitting Sérsic or King profiles (often in addition to
+Exponential), find that the additional parameter somewhat improves fits
+relative to an exponential
+\citep{IH1995, vanzee+barton+skillman2004, munoz+2018, wang+2019}. Note
+that the Sérsic indices are typically \(n \lesssim 1\), implying that
+most galaxies tend to be an exponential or slightly steeper.
 
-As a result, many studies assume an exponential density profile in
-theoretical or observational modelling
-\citep[e.g.,][]{MV2020a, kowalczyk+2013}.
-
-\citet{lelli+2014} (example of using exponential.)
+Beyond local group dwarf spheroidals, exponential densities appear to be
+common, but sometimes with significant modifications. In particular,
+many extragalactic dwarf Elliptical and blue compact / Irr galaxies
+galaxies are fit well with an exponential + central cusp / nuclear
+region \citep[\citet{noeske+2003}]{caldwell+bothun1987}. Contrastingly,
+some studies find an inner deprivation of stellar material as compared
+to exponentials \citep[e.g.,][]{caldwell+1992, makarov+2012}.
 
 However, an exponential profile was often noted to not fit every dwarf
-galaxy. Even starting from \citet{aparicio1997}, and \citet[for coma
-cluster dE]{graham+guzman2003}, some people have noted deviations from
-this simple empirical rule. Additionally, \citet{hunter+elmegreen};
+galaxy, particular for more distant dwarf galaxies. Even starting from
+\citet{aparicio1997}, and \citet[for coma cluster
+dE]{graham+guzman2003}, some people have noted deviations from this
+simple empirical rule. Additionally, \citet{hunter+elmegreen};
 \citet{herrmann+hunter+elmegreen2013};
-\citet{herrmann+hunter+elmegreen2016} note that at least in a
-photometric sample of more distant dwarf disky/Irr/blue compact dwarfs,
-that many or most dwarfs show deviations from exponential profiles. Irr
-do represent a substantially different class though so it is unclear if
-these conclusions apply.
+\citet{herrmann+hunter+elmegreen2016}; \citet{lee+2018} note that at
+least in a photometric sample of more distant dwarf disky/Irr/blue
+compact dwarfs, that many or most dwarfs show two-part exponential
+profiles. Irr do represent a substantially different class though so it
+is unclear if these conclusions apply. \citet{caldwell+1992} mostly well
+fit outer exponential, inner deviations photometric M31. Note that while
+diversity is also commented on here, dwarf galaxies with outer
+flattening profiles are often a minority in these studies.
 
-Interestingly, many works also show that differences between the
-different density profiles remain barely significant for some galaxies
-\citep[e.g.,][]{cicuendez+2018}. \citet{wang+2019} for Fornax showing
-similar performance but favour Sérsic, \citet{mcconnachie+irwin2006} for
-Exp, Plummer, King in other galaxies.
+Alternatively, \citet{moskowitz+walker2020}, use a plummer profile and
+modified plummer profiles with steeper outer slopes (gamma=-8 vs -4).
+For most galaxies, they do not have convincing statistical evidence to
+prefer one model over another, but show that for 8 / 15 of dwarfs with
+sufficient data, a plummer or steeper profile is preferred, with only
+the stepper one ever strongly preferred.
 
-\begin{itemize}
-\item
-  \citet{makarov+2012} isolated local volume dwarf galaxy with central
-  deprivation compared to exponential, spheroidal
-\item
-  \citet{martin+2016}, exponential fits to many pandas dwarf galaxies
-\item
-  \citet{moskowitz+walker2020}, generalized plummer fits.
-\end{itemize}
-
-Irregardless of if all dwarf galaxies are indeed exponential, there is
-nevertheless diversity requiring interpretation in the density profiles
-of dwarf galaxies.
+Altogether, while there is some natural variation in the density
+profiles of dwarf galaxies, an exponential is an excellent first-order
+approximation. Typically, deviations from exponentials are in a
+direction of steeper outer cutoff or changes to the inner slope of a
+dwarf galaxy (due to a nuclear region/etc.). As such, deviations in
+directions of density profiles becoming steeper in the very outer
+regions are unusual. Explaining the origin, similarity, and diversity of
+dwarf galaxy density profiles is a pressing question in our theories of
+dwarf galaxy formation and evolution. Irregardless of if all dwarf
+galaxies are indeed exponential, there is nevertheless diversity
+requiring interpretation in the density profiles of dwarf galaxies.
 
@@
 their half-light radius and the density at half-light radius (data
-described in Section~\ref{sec:data}). The lines represent the initial
-and final models from \citet{PNM2008}'s model (Fig. 4 top left, stellar
-segregation of 0.20 moving on a highly eccentric orbit peri:apo = 1 :
-100)).}\label{fig:scl_umi_vs_penarrubia}
+described in Section~\ref{sec:data}). The right panel adds an idealized
+model with initial, final, and break radius as a dotted line, solid line
+and an arrow respectively see
+\ref{sec:break_radii}.}\label{fig:scl_umi_vs_penarrubia}
 \end{figure}
@@
 
-Since the discovery of dwarf galaxies, a large body of work has
-speculated, considered, or simulated tidal effects. While
-pre-\emph{Gaia} work often did not know the specific orbits of dwarf
-galaxies, the theory for satellites serves as an excellent framework for
-understanding specific systems.
+Since the discovery of dwarf galaxies, a large body of work has either
+considered or simulated tidal effects. While pre-\emph{Gaia} work often
+did not know the specific orbits of dwarf galaxies, the theory of tidal
+mass loss remains largely the same.
 
-While cosmological simulations predict that mergers and satellite-galaxy
-interactions are common \citep[e.g.,][]{riley+2024}, they struggle to
-understand the evolution of individual dwarf galaxies. In particular,
-dwarf galaxies are often barely resolved. The highest resolution
-simulation of a Milky Way dark matter halo, the Aquarius project
-\citep{springel+2008}, acchieved a DM resolution of
-\(1.712 \times10^3 \Mo\), enough to barely resolve Sculptor like halos
-(see methods). On the other hand, idealized simulations are able to
-reach high resolution and numerical convergence for a single dwarf
-galaxy. But, idealization may not be the most realistic environment,
-neglecting mergers, assembly history, and often baryonic physics. We use
-idealized simulations here which are more powerful in accurately
-assessing tidal effects, given that the idealizations do not impact
-these galaxies's recent history too much. As such, a long history of
-work has resolved to understand the tidal evolution of dwarf galaxies
-using these \emph{idealized} simulations.
+Simulating dwarf galaxies accurately in a cosmological context remains a
+substantial challenge. Currently, Cosmological simulations can predict
+the overall abundance of larger mass dwarf galaxies {[}e.g., {]} and
+broadly predict effects of tides \citep[e.g.,][]{riley+2024}. But, dwarf
+galaxies are often near the resolution limit. Insufficient resolution
+can lead to artificial disruption of dwarf galaxies and overestimation
+of tidal effects \citep[e.g.,][]{santos-santos+2024}. For example, the
+highest resolution cosmological simulation of a Milky Way-size dark
+matter halo, the Aquarius project \citep{springel+2008}, achieved a DM
+resolution of \(1.712 \times10^3 \Mo\), enough to barely resolve
+Sculptor like halos (see Section~\ref{sec:Methods}). To address this
+challenge, idealized simulations only simulate a single subhalo in an
+approximate host potential. As a result, idealized simulations can reach
+excellent numerical convergence at the cost of realism. For instance,
+our simulations later are about 3x higher resolution than Aquarius but
+with a fraction of the computational cost (400x less particles).
+Idealized simulations make numerous simplifications: neglecting mergers,
+cosmological context and evolution, mass assembly, and often baryonic
+physics. We use idealized simulations here which are more powerful in
+accurately assessing tidal effects after infall, given that the
+idealizations do not impact these galaxies's recent history too much. As
+such, a long history of work has resolved to understand the tidal
+evolution of dwarf galaxies using these \emph{idealized} simulations.
 
-Some of the earliest theory work on tidal stripping of dwarf galaxies
-originate from
-\citep{allen+richstone1988, moore+davis1994, johnston+spergel+hernquist1995, oh+lin+aarseth1995, piatek+pryor1995, velazquez+white1995, kroupa1997, pryor1996}.
-Many of these works used similar techniques as we continue to use today,
-setting up a dark matter halo in a static Milky Way potential and
-predicting the evolution of quantities of interest.
-
-More recent works include \citet{read+2006};
-\citet{bullock+johnston2005}; \citet{PNM2008}; \citet{penarrubia+2009};
-\citet{klimentowski+2009}; \citet{errani+2023a}; \citet{fattahi+2018};
-\citet{stucker+2023}; \citet{wang+2017}.
+Some of the earliest theory work on tidal mass loss of dwarf galaxies
+originate from \citet{oh+lin+aarseth1995}; \citet{piatek+pryor1995};
+\citet{moore+davis1994}; \citet{johnston+spergel+hernquist1995}.
+Already, these works used similar techniques as we continue to use today
+and laid the foundation for understanding tidal effects. While the
+detailed assumptions have evolved (e.g.~assumed total mass of dwarf
+galaxies and evolving beyond \citet{king1966} stellar), many of their
+conclusions still hold true. More recent works expanding this theory
+include \citet{read+2006}; \citet{bullock+johnston2005};
+\citet{PNM2008}; \citet{penarrubia+2009}; \citet{klimentowski+2009};
+\citet{errani+2023a}; \citet{fattahi+2018}; \citet{stucker+2023};
+\citet{wang+2017}.
 
@@
 potential and system, more recent work began to directly probe the
-dynamical histories of individual dwarf galaxies. Examples include
-\citet{iorio+2019} for Sculptor, \citet{borukhovetskaya+2022};
-\citet{dicintio+2024} for Fornax; \citet{borukhovetskaya+2022a} for
-Antlia II. Our goal is to apply a similar framework to Sculptor and Ursa
-Minor.
+dynamical histories of individual dwarf galaxies (although early work
+began this for Sagittarius / etc). Examples include \citet{iorio+2019}
+for Sculptor, \citet{borukhovetskaya+2022}; \citet{dicintio+2024} for
+Fornax; \citet{borukhovetskaya+2022a} for Antlia II. Our goal is to
+apply a similar framework to Sculptor and Ursa Minor.
 
@@
 
-\subsection{Break and tidal radii}\label{break-and-tidal-radii}
+\subsection{Break and tidal radii}\label{sec:break_radii}
 
@@
 positive as they still readjust to a new equilibrium.
-
-From this argument, we note that the following properties must be
-approximately true to interpret a break is
-
-\begin{itemize}
-\tightlist
-\item
-  \emph{Small pericentre}. If the pericentre is too large and
-  \(r_J \gg R_h\), then few stars will be lost.
-\item
-  Corresponding time since last pericentre: If the time since last
-  pericentre is not \(\sim\) consistent with an observed break in the
-  density profile, then tides
-\item
-  Halo evolution. As found in \citet{EN2021}, galaxies evolve along well
-  defined tidal tracks (assuming spherical, isotropic, NFW halo, which
-  may not be true, see \ldots). These tracks tend to ``puff up'' the
-  stellar component while also removing dark matter mass, leaving a
-  smaller, compacter DM halo with a more extended stellar component.
-\item
-  This information is mostly related to the statistical initial
-  distribution of satellites from cosmology
-  \citep{ludlow+2016, fattahi+2018}
-\end{itemize}
 
@@
 
-\section{Thesis Outline}\label{thesis-outline}
+\section{Thesis outline}\label{thesis-outline}
 
@@
 from \citet{jensen+2024}. In chapter 3, we review our simulation
-methods. Next, we present our results for the tidal stripping of
-Sculptor and Ursa Minor in Chapter 4. We discuss our results,
-limitations, and implications in Chapter 5. Finally, in Chapter 6, we
-summarize this work and discuss future directions for similar work and
-the field of dwarf galaxies.
+methods. Next, we present our results for the tidal effects on Sculptor
+and Ursa Minor in Chapter 4. We discuss our results, limitations, and
+implications in Chapter 5. Finally, Chapter 6 summarizes this work and
+discuss future directions for similar work and the field of dwarf
+galaxies.

### 2025-08-13 notebooks/ludlow.jl

@@ -1,0 +1,655 @@
+### A Pluto.jl notebook ###
+# v0.20.13
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 9991bd2a-77c4-11f0-2bf0-2fbe36425e63
+begin 
+	import Pkg; Pkg.activate()
+
+	using LilGuys
+	using Arya
+	using CairoMakie
+end
+
+# ╔═╡ 4f471846-ae8b-419e-8cd5-920032f19c82
+import LilGuys: R200, M200, find_zero
+
+# ╔═╡ b88a3b23-6f3e-487b-848c-24fab0e8a286
+a1, a2, a3 = [2.9, 0.614, 0.995] #[2.521, 0.729, 0.988]
+
+# ╔═╡ c8f78a5b-fbf5-4b0f-849a-eec18e8a4639
+function c_mah(c)
+	return ((c / a1)^(1/a3)  - 1) / a2
+end
+
+# ╔═╡ 5ff465a3-846e-4fac-b2c3-bbe92a5a068e
+function c_enclosed_density(c)
+	return a1 * (1 + a2 * c)^a3
+end
+
+# ╔═╡ cc36e1fc-cd68-45c8-a1dd-4a75e19e3057
+ρ_c(z) = LilGuys.ρ_crit * (1+z)^3 
+
+# ╔═╡ d690b0c4-3f9a-4c46-b356-f439982adbba
+ρ_0 = ρ_c(0.)
+
+# ╔═╡ 74437045-5ee0-4c80-b8b9-90adf1a06f61
+function scale_density(halo)
+	return LilGuys.mean_density(halo, halo.r_s)
+end
+
+# ╔═╡ f433807e-620a-4191-b94e-ebe9f68d3a21
+function scale_point_density(halo)
+	return LilGuys.density(halo, halo.r_s)
+end
+
+# ╔═╡ 8325d421-a7a7-41ee-a0e3-3ff1b0df0c22
+function mass_given_density(halo, ρ)
+	log_r = LilGuys.find_zero(lr -> LilGuys.mean_density(halo, 10 ^ lr) - ρ, 
+			log10(halo.r_s))
+	return LilGuys.mass(halo, 10 ^ log_r)
+end
+
+# ╔═╡ cd58027b-ff2d-45d6-9b37-08d95eaf40ab
+function MAH(nfw; c_scale=1)
[…]

### 2025-08-13 thesis/plots/analytic_profile_comparison.jl

@@ -14,6 +14,12 @@ begin
 
 end
 
+# ╔═╡ ec95bcc8-2c24-40d7-80f1-32e83c4c17aa
+include("./paper_style.jl")
+
+# ╔═╡ 7efef2ac-d002-4a86-a8c3-e9e7036de668
+FIGDIR = "./figures"
+
 # ╔═╡ 4671142c-e341-4927-bfe7-809b41daf251
 CairoMakie.activate!(type=:png)
 
@@ -21,7 +27,10 @@ CairoMakie.activate!(type=:png)
 prof_exp = LilGuys.Exp2D()
 
 # ╔═╡ cb7dc629-fbc9-4e08-ba56-2074009ee874
-prof_sersic_n2 = LilGuys.Sersic(n=0.8)
+prof_sersic_n0_8 = LilGuys.Sersic(n=0.8)
+
+# ╔═╡ d8694866-324a-4bac-8856-77a130559a8f
+prof_sersic_n1_2 = LilGuys.Sersic(n=1.2)
 
 # ╔═╡ 5fb7529d-0b1a-4fa1-bbaf-ed13f2f01027
 prof_plummer = LilGuys.Plummer()
@@ -42,6 +51,9 @@ end
 # ╔═╡ 6e612d6d-4605-47de-af5f-c38d022650d8
 LilGuys.density(prof::GeneralizedPlummer, r::Real) = prof._k * prof.M * prof.r_s^(prof.n - 3) / (prof.r_s^2 + r^2)^(prof.n/2)
 
+# ╔═╡ cd2f1f91-2f9d-476d-8f84-6e17abf6df66
+outer_slope(n) = n - 1
+
 # ╔═╡ f2725829-5ca7-43a5-b43e-1dfddbac1f7e
 LilGuys.R_h(prof::LilGuys.Sersic) = prof.R_h
 
@@ -50,11 +62,12 @@ steep_plummer = GeneralizedPlummer(M=1, r_s=1, n=9)
 
 # ╔═╡ c0e9018f-5ac6-4f1c-b800-bb7348a4f5f7
 profiles = [
-	"exponential" => prof_exp,
-	"sersic" => prof_sersic_n2,
-	"plummer" => prof_plummer,
-	"steep plummer" => steep_plummer,
-	"king c=10" => LilGuys.KingProfile(R_c=1, R_t=10, k=1)
+	"Exponential" => prof_exp,
[…]

### 2025-08-18 thesis/abstract.tex
thesis julio ch1 edits

@@
-Dwarf galaxies are unique cosmological probes, testing our understanding
-of the smallest scales of structure formation in the Universe. Among the
-classical dwarf galaxies, Sculptor and Ursa Minor appear unusual, with
-more stars than expected in the outer regions compared with the typical
-Exponential density profile, rememnescent of tidal interaction. I verify
-the validity of this excess, checking \citet{jensen+2024}'s bayesian
-membership catalogue for completeness, simple cuts, and other surveys.
-To evaluate if tides indeed can explain these density features, I
-conduct idealized N-body simulation of both galaxies in the tidal field
-of the Milky Way. In both cases, the orbits are well constrained,
-producing pericenters larger than expected to produce substantial tidal
-impacts. In addition, the time since the last pericenter does not agree
-with a tidal interpretation. However, Sculptor's orbit is strongly
-influenced by the presence of an LMC. With an LMC, Sculptor's long term
-orbital history is highly uncertain, however recent tidal effects due to
-the Milky Way and LMC only minimally affect this galaxy. Because tides
-likely do not explain the presence of these density features, I review
-other theories about the formation of dwarf galaxy ``halos,'' including
-mergers / accretion events and complex multi-component star formation
-histories. The density profiles of Sculptor and Ursa Minor, with their
-deviations from the classically expected exponential, present an
-interesting puzzle, likely with a solution involving their formation and
-history in the early universe.
+Dwarf galaxies are unique cosmological probes, which test our
+understanding of the smallest scales of structure formation in the
+Universe. \emph{Among the classical dwarf galaxies, we show that
+Sculptor and Ursa Minor appear unusual, with more stars than expected in
+the outer regions compared with the typical Exponential density
+profile.}\footnote{need to refine this sentence} I verify the validity
+of this excess, checking \citet{jensen+2024}'s bayesian membership
+catalogue for completeness, simple cuts, and other surveys. To evaluate
+if tides indeed can explain these density features, I conduct idealized
+N-body simulation of both galaxies in the tidal field of the Milky Way.
+In both cases, the orbits are well constrained, producing pericenters
+larger than expected to produce substantial tidal impacts. In addition,
+the time since the last pericenter does not agree with a tidal
+interpretation. However, Sculptor's orbit is strongly influenced by the
+presence of an LMC. With an LMC, Sculptor's long term orbital history is
+highly uncertain, however recent tidal effects due to the Milky Way and
+LMC only minimally affect this galaxy. Because tides likely do not
+explain the presence of these density features, I review other theories
+about the formation of dwarf galaxy ``halos,'' including mergers /
+accretion events and complex multi-component star formation histories.
+The density profiles of Sculptor and Ursa Minor, with their deviations
+from the classically expected exponential, present an interesting
+puzzle, likely with a solution involving their formation and history in
+the early universe.

### 2025-08-18 thesis/data.tex
thesis julio ch1 edits

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
 \caption[Sculptor sample selection]{The selection criteria for Scl
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
 \caption[Ursa Minor sample selection]{Similar to
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.pdf}}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
 \caption[Fornax sample selection]{Similar to

### 2025-08-18 thesis/discussion.tex
thesis julio ch1 edits

@@
 
-\citet{reed+gilmore2005} propsed that exponential dSph form through
+\citet{read+gilmore2005} propsed that exponential dSph form through
 impusive mass loss, redistributing the stellar component into an
@@
 
-\citet{klimentowski2007}, \citet{klimentowski2009}.
+\citet{klimentowski+2007}, \citet{klimentowski+2009}.
 
@@
 \item
-  \citet{arronyo-polonio+2024}
+  \citet{arroyo-polonio+2024}
 \item
@@
 \item
-  \citet{ricotti+2022}
+  \citet{ricotti+polisensky+cleland2022}
 \item

### 2025-08-18 thesis/introduction.tex
thesis julio ch1 edits

@@
 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
-many dwarf galaxies are \emph{quenched}, with little to no recent star
-formation. Their stellar populations are \emph{relics} from the early
-universe, consisting of many of the oldest and most metal poor stars.
-Understanding the properties of dwarf galaxies thus has implications
-across astronomy, from cosmological structure formation on the smallest
-scale to extremely metal-poor stellar populations to chemical evolution
-to galactic dynamics. Of the classical dwarfs, the Sculptor and Ursa
+many dwarf galaxy satellites of the Milky Way are \emph{quenched}, with
+little to no recent star formation. Milky Way dwarfs contain stellar
+populations which are \emph{relics} from the early universe, consisting
+of many of the oldest and most metal poor stars. Understanding the
+properties of dwarf galaxies thus has implications across astronomy,
+from cosmological structure formation on the smallest scale to extremely
+metal-poor stellar populations to chemical evolution to galactic
+dynamics. Of the classical dwarfs, we find that the Sculptor and Ursa
 Minor dwarf spheroidal galaxies stand out with a more extended density
@@
 
-\section{Cosmological context}\label{cosmological-context}
-
-We only understand a tiny fraction of the universe's composition. The
-leading theory of cosmology, \(\Lambda\)CDM (cold dark matter), posits
-that the universe is composed of about 68\% dark energy (\(\Lambda\)),
-27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
-  to change definitions of words. \emph{Baryons} here means
-  baryons+leptons, i.e.~any standard model massive fermion. The photon
-  energy density is negligible today} \citep{planckcollaboration+2020}.
-While the composition of dark matter and dark energy remains elusive, we
-know their general properties. Dark energy causes the acceleration of
-the expansion of the universe on large scales. We do not discuss dark
-energy here---it does not substantially affect the local group today.
-Dark matter, instead, makes up the vast majority of mass in galaxies.
-Typically, galaxies have baryonic to dark matter ratios of between 1:5
-to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
-matter is assumed to only interact gravitationally, passing through
-matter without effect (transparent to light, or \emph{dark}). Dark
-matter is also \emph{cold}, i.e.~typical velocities much smaller than
-the speed of light in the early universe. Implications of dark matter
-properties range from cosmological structural formation, galaxy
-structure, and galaxy interactions.
-
-\subsection{Structure formation and dwarf
-galaxies}\label{structure-formation-and-dwarf-galaxies}
-
-The very early universe was almost featureless. Our earliest
-observations of the universe stem from the cosmic microwave background
-(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
-But tiny perturbations in the CMB, temperature fluctuations of 1 part in
-10,000, became the seeds of large-scale cosmological structure. Governed
-by cosmological expansion, gravitational collapse, and baryonic physics,
-each perturbation grows into larger structures. Initially, baryonic
-matter was coupled to radiation and resisted collapse. Dark matter, only
-influenced by gravity instead, freely collapsed into the first
-structures. Each self-gravitating overdensity of dark matter is known as
-a \emph{halo}. After recombination, where electrons combined with atomic
-nuclei to form atoms, baryons decoupled from radiation and fell into the
-dark matter halos. The densest pockets of baryons later formed the first
-stars and galaxies.
-
-Dark matter halos, and their associated galaxies, rarely evolve in
-isolation. Instead, structure formation is \emph{hierarchical}. Small
-dark matter halos collapse first and hierarchically merge into
-progressively larger halos
-\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
-Structure formation happens on a wide range of scales. The largest
-structures become the cosmic web---composed of voids, filaments, and
-clusters---and the smallest structures directly observable host dwarf
-galaxies. Hierarchical assembly is evident through the large scale
-structure of the universe, remnants of past mergers within the milky
-way, and tidal disruption of dwarf galaxies and their streams around
-nearby galaxies.
-
-Small-scale structure formation is sensitive to deviations from
-\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
-models, such as warm dark matter or self-interacting dark matter, may
-smooth out small-scale features and reduce the abundance of small halos
-or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
-occupying the smallest dark matter halos, are promising windows into
-small-scale cosmological features. Alternative dark matter properties
-can influence dwarf galaxy abundance, formation, structure,, and tidal
-evolution. The nearby dwarf galaxies of the local group, with detailed
-observations, present a promising opportunity to understand the
-evolution of these objects and test our understanding of cosmology. To
-understand the evolution of these objects, we need to understand the
-general predictions from \(\Lambda\)CDM for the properties of dwarf
-galaxies.
-
-\subsection{Typical structure of dark matter
-halos}\label{typical-structure-of-dark-matter-halos}
-
-In \(\Lambda\)CDM cosmological simulations, dark matter halos are
-remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
-authors observe that the spherical radial density profiles \(\rho(r)\)
-are well described by a two-parameter law: \begin{equation}{
-\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
-}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
-density . This profile has shown remarkable success in describing
-\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
-profiles are \emph{cuspy}, where the density continuously rises like
-\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
-density profile increases at \(r \sim r_s\) and at large radii, the
-density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
-shows an example of a NFW halo and a cored dark matter halo.
-
-The total mass of an NFW profile diverges, so halos are commonly
-characterized by an overdensity criterion. The virial mass, \(M_{200}\),
-is defined as the mass within a region \(r_{200}\) containing a mean
-enclosed density 200 times\footnote{From approximate analytic arguments
-  about the mean density of a self-gravitating spherical collapse halo
-  in an expanding universe, the expected virial overdensity is \ldots,
-  commonly rounded to 200 (REFS).} the critical density of the universe:
-\begin{equation}{
-M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
-}\end{equation} A standard second parameter is the halo concentration,
-\(c=r_{200} / r_s\) describing how the characteristic size of the halo
-compares to its virial radius. In this case, the scale density is a
-function of \(c\) alone,
-\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
-\citep{NFW1996}. However, we characterize halos by their circular
-velocity profiles. The circular velocity,
-\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
-\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
-\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
-halo mass and the observed line of sight velocity disperion, and
-\(r_{\rm max}\) relates to the scale radius.
-
-While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
-these are not independent. Smaller dark matter halos often collapse
-earlier, when the universe was denser. As a result, small subhalos tend
-to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
-between \(M_{200}\) and c, or the mass-concentration relation, how more
-massive halos become less concentrated in a predictable way
-\citep[e.g.,][]{bullock+2001, ludlow+2016}. The lower panel of
-Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
-relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
-a general trend in concentration with mass, the concentration values
-still have substantial scatter. Other parameters such as the halo spin
-or shape may add additional variability to present-day distribution of
-halos, but this variation is typically expected to be small
-\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
-\caption[Example density profiles]{Density profiles in log density
-versus log radius for the stars and dark matter of a Sculptor-like
-galaxy. The dark matter is more extended and massive than the star
-across the entire galaxy---dynamical evolution is driven by the dark
-matter mass distribution.}\label{fig:nfw_density}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
-\caption[Stellar-mass halo-mass relation]{\textbf{Top} Stellar mass
-(top) as a function of maximum circular velocity. The solid line with
-the 1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
-with scatter points simulated central galaxies from APOSTLE in
-\citet{fattahi+2018}. \textbf{Bottom:} Radius of maximum circular
-velocity \(r_{\rm max}\) as a function of maximum circular velocity. The
-solid line with 1-\(\sigma\) shaded region and dashed line are the
-relations from \citet{ludlow+2016} for \(z=0\) and \(z=2\) respectively.
-\emph{TODO}: add legend / annotations.}\label{fig:smhm}
-\end{figure}
-
-\subsection{Connecting dark matter to
-stars}\label{connecting-dark-matter-to-stars}
-
-Both cosmology and observations find fundamental relationships between
-the dynamical or total mass and stellar masses of galaxies. A key
-prediction of \(\Lambda\)CDM is the stellar mass halo mass relation,
-describing the amount of stellar mass forms in a (sub)halo of a given
-size. In particular, the SMHM relation grows especially steep in the
-dwarf galaxy regime---many dwarf galaxies are formed in halos of similar
-masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
-\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
-apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
-the range of predicted \(v_{\rm max}\) is fairly narrow across
-\(\sim 4\) decades in stellar mass. Because lower mass galaxies have
-increasingly shallow potential wells, feedback becomes more effective
-removing gas more quickly. In addition, re-ionization suppresses late
-star formation in the faintest galaxies. Thus, if we know the initial
-stellar mass of a dwarf galaxy, we also know, at some level, the
-properties of its dark matter halo.
-
-Several challenges complicate a simple SMHM trend including environment,
-assembly history, and tidal effects. Additionally, the detailed physics
-of galaxy formation can impact the SMHM relationship. By being closer to
-a massive host, many dwarfs quench earlier, resulting in lower stellar
-masses \citep[e.g.,][]{christensen+2024}. Additionally, the time of
-formation (relative to reionization) can influence the resulting stellar
-content \citep{kim+2024}. Finally, tides influence both the dark matter
-and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
-Consequently, tides may reduce the halo mass more than the stellar mass,
-adding additional scatter to the SMHM trend, particularly for low-mass
-satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
-tides on local group dwarf galaxies is essential to determining where
-and how these galaxies formed in a cosmological context.
-
 \section{Observational context}\label{observational-context}
@@
 formation. The discovery of Fornax and Sculptor in 1938
-\citep{shapley1938}\footnote{technically, the LMC and SMC may be
-  classified as dwarf galaxies, but these were likely always known to
-  humans at southern latitudes.}, with no known analogues, already
-presented such an enigma. Shapley presented these dwarfs as a new type
-of \emph{stellar system} resembling the Magellanic clouds and globular
-clusters but did not attempt to speculate on the exact nature. While
-dwarf galaxies were quickly understood to be galaxies based on the
-inferred luminosities and sizes, their nature and formation remained
-unclear for decades \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
+\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
+  Clouds (LMC, SMC) may be classified as dwarf galaxies, but these were
+  likely always known to humans at southern latitudes.}, with no known
+analogues, already presented such an enigma. Shapley presented these
+dwarfs as a new type of \emph{stellar system} resembling the Magellanic
+Clouds and globular clusters but did not attempt to speculate on the
+exact nature. While dwarf galaxies were quickly understood to be
+galaxies based on the inferred luminosities and sizes, their nature and
+formation remained unclear for decades
+\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
 
 The earliest spectroscopic work hinted that dwarf galaxies may contain
-substantial dark matter masses. From early determinations of the
-velocity dispersion for Sculptor and Ursa Minor dSph
+substantial amounts of dark matter. From early determinations of the
+velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
+(dSph) galaxies
 \citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
-mass-to-light ratios were at least 10 times the expectation for globular
-clusters. While only determined by a few stars initially, these values
-have solidified with time with larger and more precise samples
-\citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
+mass-to-light ratios were at least 10 times larger than the values from
+globular clusters scaled to the same sizes. While rather uncertain
+initially, these values have corroborated with larger and more precise
+samples \citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
 attempting to understand the formation and observed properties of these
@@
 \citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
-holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark matter
-free out-of-equilibrium ``tidal dwarfs'' from past mergers
+holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark
+matter-free ``tidal dwarfs'' from past mergers
 \citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
@@
 galaxies are among the darkest objects in the universe, and
-understanding their properties is critical to understanding the
-universe.
+understanding their properties is critical to understanding dark matter.
 
@@
 on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
-\textbf{bright dwarfs galaxies} with magnitudes
+\textbf{bright dwarfs} with magnitudes
 \(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
-and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
-an example of an irregular (and slightly bright) dwarf galaxy displaying
-a bar. \textbf{Classical dwarf} galaxies occupy intermediate
+and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
+Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
+galaxy displaying a bar. \textbf{Classical dwarfs} occupy intermediate
 luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
@@
 discovered before digital sky surveys are classicals. The 12 classical
-dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
+dwarfs satellites of our Galaxy are Sagittarius, Fornax, Leo I,
 Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
 I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
-  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
+  names we discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor
   dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
@@
 galaxies have minuscule stellar masses, tend to be more compact, and are
-the darkest known galaxies. Altogether, dwarf galaxies span about 7
-orders of magnitude in stellar mass.
+the darkest known galaxies. Altogether, dwarf galaxies span more than 15
+orders in absolute magnitude.
+
+Since Local Group dwarfs are nearby, we can study these galaxies on a
+star-by-star basis. As a result, it is possible to measure the 3D
+velocity and position of a star if we can measure the stars position,
+line-of-sight (LOS) velocity, and proper motions. Unfortunately,
+determining distances and full 3D velocities is challenging. The most
+direct measurement of distance, parallax, requires precise tracking of a
+star's sky position across a year. And while line-of-sight (LOS)
+velocities are easily determined from spectroscopy, tangential
+velocities, derived from proper motions and distances, are much more
+challenging. Typically, measuring proper motions requires accurate (much
+less than arcsecond) determinations of small changes in a star's
+position over baselines of several years to decades. The full 6D
+position and velocity information for stars has historically been known
+for only a handful of stars.
 
@@
 \centering
-\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
+\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
 \caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
 DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
-over-plotted). Each image includes the galaxies luminosity and a 200 pc
-scalebar. While the LMC is very prominent in the sky, the classical
-dwarf galaxies are not obvious except in terms of star counts
-(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
-central \(\sim\)third of the image.}\label{fig:galaxy_images}
+over-plotted). Each image includes the galaxy's luminosity and a 200 pc
+scale bar. \emph{TODO: Mark half-light radius with
+circle?}}\label{fig:galaxy_images}
 \end{figure}
@@
 
-A star's or galaxy's position and velocity are fundamental quantities
-for understanding its orbit and origin. Unfortunately, determining
-distances and full 3D velocities is challenging. The most direct
-measurement of distance, parallax, requires precise tracking of a star's
-sky position across a year. And while line-of-sight (LOS) velocities are
-easily determined from spectroscopy, tangental velocities, derived from
-proper motions, are much more challenging. Typically, measuring proper
-motions requires accurate (much less than arcsecond) determinations of
-small changes in a star's position over baselines of several years to
-decades. The full 6D position and velocity information for stars was
-historically only known for a handful of stars.
+\emph{(db: these paragraphs start the same\ldots)}
 
@@
 proper motions, and parallaxes for over 1 billion stars
-\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
+\citep{gaiacollaboration+2021}. Launched in 2013, \emph{Gaia} is a
 space-based, all-sky survey telescope situated at the Sun-Earth L2
@@
 years, an absolute all-sky reference frame is derived from which
-parallax and proper motions are derived \citep{gaiacollaboration+2016}.
-In addition to astrometry, \emph{Gaia} measures photometry in the wide
-\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
-330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
-additionally provides low resolution BP-RP spectra and radial velocity
-measurements of bright stars (\emph{Gaia} radial velocity magnitudes
-\textless16) \citep{gaiacollaboration+2016}. For this work,
-\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
-\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
+parallax and proper motions are derived. In addition to astrometry,
+\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
+and colours from the blue photometer (BP, 330-680 nm) and red photometer
+(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
+BP-RP spectra and radial velocity measurements of bright stars
+(\emph{Gaia} radial velocity magnitudes \textless16)
+\citep{gaiacollaboration+2016}. For this work, \emph{Gaia}'s most
+relevant measurement are \(G\) magnitude, \(G_{\rm BP} - G_{\rm RP}\)
+colour, \((\alpha, \delta)\) position, and
 \((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
   motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
-  change in right ascension \(\alpha\) and declination \(\delta\),
-  typically in units of mili-arcsecond (mas) per year. \(\mu_{\alpha*}\)
-  allows where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for
-  projection effects in \(\alpha\).}
+  change in right ascension, \(\alpha\), and declination, \(\delta\),
+  typically in units of mili-arcsecond (mas) per year.
+  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
+  effects in \(\alpha\).}
 
-\emph{Gaia} has revolutionized many fields of astronomy including the
-Milky Way and the Local Group. For example, 6D kinematic classification
-of Milky Way stars has lead to the discovery of past mergers or Milky
-Way building blocks like \emph{Gaia}-sausage Enceladus
+\emph{Gaia} has revolutionized many fields of astronomy including
+studies of the Milky Way and the Local Group. The Local Group is defined
+as galaxies which are approximately bound to the Milky Way-Andromeda
+system, or within about 1 Mpc from the Milky Way \citep[e.g.,][ and
+references therein]{mcconnachie2012}. For example, 6D kinematic
+classification of Milky Way stars led to the discovery of past mergers
+or Milky Way building blocks like \emph{Gaia}-Sausage Enceladus
 \citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
-\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and dynamical effects
-of the Milky Way's spiral arms and the bar in the solar neighbourhood
+\emph{Gaia} snail {[}e.g., \citet{antoja+2018}, and dynamical effects of
+the Milky Way's spiral arms and the bar in the solar neighbourhood
 \citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
@@
 the hierarchical formation and complex, evolving structure of our own
-galaxy.
+Galaxy.
 
@@
 orbital analysis and facilitated precise stellar membership
-determination. Before, proper motions of dwarf galaxies were spase or
+determination. Before, proper motions of dwarf galaxies were sparse or
 undetermined. Few galaxies had precisely measured proper motions, often
@@
 \emph{Gaia} allowed for the first systematic determinations of Milky Way
-satellite proper motions, and of unprecedented precision
+satellite proper motions of unprecedented precision
 \citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
-typical dwarf member star is often with large uncertainties, by
-combining the proper motions for 100s to 1000s of stars in \emph{Gaia},
-incredibly precise proper motion measurements can be determined,
-sometimes only limited by \emph{Gaia}'s systematic error floor
-\citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new dynamical
-era for MW satellites studies, where we often have precise
-determinations of the dwarf galaxy's recent orbit under a given MW
-potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
-foreground stars. By measuring parallaxes or proper motions
-significantly different from the dwarf galaxy, many stars can be
-immediately removed as non-members
-\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.
+typical dwarf member star is often large, by combining the proper
+motions for 100s to 1000s of stars in \emph{Gaia}, precise proper motion
+measurements can be determined, sometimes only limited by \emph{Gaia}'s
+systematic error floor \citep[e.g.,][]{MV2020a}. Proper motions have
+ushered in a new dynamical era for MW satellite studies, where we can
+derive precise orbits assuming a given MW potential. In addition,
+\emph{Gaia} helps separate out contaminating MW foreground stars. By
+measuring parallaxes or proper motions, many background and foreground
+stars can be removed as non-members
+\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.
 
@@
 
-Today, we know the Milky Way system is teeming with satellites.
+Today, we know that the Milky Way system is teeming with satellites.
 Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
-including dwarf galaxies, ambiguous systems, and globular clusters.
-Advances in telescopes and observations has accelerated progress on the
-nature of dwarf galaxy and introduced new questions. Deep digital
-photometric surveys have more than quadrupled the number of known Milky
-Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing surveys, like
-the Vera Rubin Observatory's Legacy Survey for Space and Time, will
-continue to probe fainter and fainter dwarf galaxies. In addition, large
-aperture multi-object spectrographs have revealed the detailed and
-complex inner chemodynamical structure of dwarf galaxies. Beyond precise
-structural and kinematic properties of dwarf galaxies, modern
-observations allow for the separation of multiple stellar populations,
-detailed constraints on the dark matter density profiles, and hints of
-tidal disruption or stellar halos.
+including dwarf galaxies, ambiguous systems (whose nature as a
+dwarf-like or cluster-like system remains uncertain), and globular
+clusters. Advances in observational techniques has accelerated progress
+in our understanding of dwarf galaxies and introduced new questions.
+Deep digital photometric surveys have more than quadrupled the number of
+known Milky Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing
+surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
+Time \citep{ivezic+2019}, will likely identify even fainter dwarf
+galaxies. In addition, large aperture multi-object spectrographs have
+revealed the detailed and complex inner chemodynamical structure of
+dwarf galaxies. Beyond precise structural and kinematic properties of
+dwarf galaxies, modern observations allow for the separation of multiple
+stellar populations, detailed constraints on the dark matter density
+profiles, and hints of tidal disruption or stellar halos.
 
-Of local dwarfs, the classical systems still remain the best studied and
-with most strongly constrained parameters. While extending 1-2 degrees
-across the sky, making deriving properties trickier
-\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
-(giant) stars, allowing thousands of stars to be observed with deep
-photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
-a result, the determination of fundamental properties such as the
+Of local dwarfs, the classical systems still remain the most studied
+galaxies with the best constrained parameters. Because many Milky Way
+dwarfs extend 1-2 degrees across the sky, measurements of some of their
+properties can be challenging \citep[e.g.,][]{mateo1998}. These dwarfs
+also contain large numbers of bright (giant) stars which can be followed
+up spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a
+result, the determination of fundamental properties such as the
 position, size, orientation, distances, proper motions, line-of-sight
 (LOS) velocities, and velocity dispersions \(\sigma_v\), are all
-relatively well constrained and converged among different studies today.
-However, ongoing research continues to redefine our understanding of the
-detailed structure of Milky Way satellites, hinting that these objects
-may be more complex than at first glance.
+relatively well constrained today. However, ongoing research continues
+to redefine our understanding of the detailed structure of Milky Way
+satellites, hinting that these objects may be more complex than expected
+at first glance.
 
 Observations of dwarf galaxies have been the origin of several disputes
-or \emph{small-scale} problems in cosmology \citep[see
+or \emph{small-scale} problems for \(\Lambda\)CDM \citep[see
 review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
@@
 abundance was known as the \emph{missing satellites problem}.
-Additionally, a number of observations showed that many dwarf galaxies,
-but not all, possess dark matter ``cores'' as measured from circular
-velocity curves \citep[e.g.,][]{moore1994, adams+2014, oh+2015} or dwarf
-spheroidal \citep[albeit less constraining,
-e.g.,][]{walker+penarrubia2011, read+walker+steger2019}. As a result,
-numerous alternative forms of dark matter have been advocated as
-solutions. \emph{Warm} dark matter, relativistic in the early universe
-but cooler now, smooths out small-scale features and softens the cusps
-of dark matter halos. \emph{Self-interacting} dark matter instead can
-form cores but also the cores can collapse into a density peak.
+Additionally, a number of observations suggested that some dwarf
+galaxies, although not all, possess dark matter ``cores''
+\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019}.\footnote{In
+  detail, gas-phase rotation curves are better able to differentiate
+  between cores and cusps, whereas stellar kinematics is less
+  constraining.} As a result, alternative forms of dark matter have been
+advocated as solutions, such as Warm and Self-Interacting Dark Matter.
 
-However, these tensions have eased as a result of improved consideration
-of baryonic physics. For instance, understanding observational
-limitations and completeness has eased tension with the abundance of
-dwarf galaxies \citep[e.g.,][]{kim+peter+hargis2018}. Also, recent
-hydrodynamic simulations in particular have shown that strong feedback
-can produce dark matter cores
-\citep[e.g.,][]{tollet+2016, fitts+2017, orkney+2021}. Altogether, the
-numerous past challenges for \(\Lambda\)CDM in the dwarf galaxy regime
-illustrates the opportunity for dwarf galaxies to test the understanding
-of galaxy formation and dark matter physics.
+However, these tensions have eased as a result of improved understanding
+of baryonic physics. For example, recent hydrodynamic simulations in
+particular have shown that strong feedback can produce dark matter cores
+\citep[e.g.,][]{tollet+2016, fitts+2017, benitez-llambay+2019, orkney+2021}.
+Altogether, the numerous past challenges for \(\Lambda\)CDM in the dwarf
+galaxy regime illustrates the opportunity for dwarf galaxies to test the
+understanding of galaxy formation and dark matter physics.
 
-\subsection{Stellar density profiles}\label{stellar-density-profiles}
+\subsection{Typical stellar density
+profiles}\label{typical-stellar-density-profiles}
 
-Surface density profiles efficiently characterize the shape of a galaxy.
-At the most basic, fitting stellar densities provide properties such as
-the shape, location, size, and orientation of a dwarf galaxy. However,
-the details of a stellar density profile are essential for interpreting
+Projected density profiles efficiently characterize the shape of a
+galaxy. At its most basic, density profiles provide properties such as
+the shape, location, size, and orientation of a dwarf galaxy. In
+addition, the details of a stellar density profile can help interpret
 the total mass of dwarf galaxies, their assembly and dynamical history,
-and understanding correlations between dwarf galaxies
+and to understand scaling laws between dwarf galaxy properties
 \citep[e.g.,][\citet{penarrubia+2009}, \citet{querci+2025},
-\citet{lee+2018}]{herrmann+hunter+elmegreen2013}.
+\citet{lee+2018}]{herrmann+hunter+elmegreen2013}. We note that for
+resolved galaxies, these profiles are in terms of stellar count density
+instead of surface magnitudes.
 
-Four different surface density profiles are frequently applied to dwarf
-galaxies: Exponential, Plummer, King, or Sérsic profiles
-\citep[e.g.,][]{munoz+18}. The exponential profile is perhaps the
-simplest, as a 1-parameter profile \begin{equation}{
-\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
+Four different surface density laws are frequently used to parameterize
+dwarf galaxy profiles: Exponential, Plummer, King, or Sérsic profiles
+\citep[e.g.,][]{munoz+2018}. The exponential profile is perhaps the
+simplest, defined in terms of the central surface density, \(\Sigma_0\),
+and scale radius, \(R_s\): \begin{equation}{
+\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s).
 }\end{equation}
 
-For a long time, exponential surface density profiles have been used as
-a description for spiral galaxies {[}\citet{mateo1998}; discussion
-below{]}.
+This profile is often applied to the radial light distribution of galaxy
+disks {[}\citet{freeman1970}; other classic paper?{]}.
 
 To fit globular cluster density profiles, \citet{plummer1911} proposed a
-1-parameter solution for a density polytrope\footnote{where density and
-  pressure are assumed to be correlated}, \begin{equation}{
-\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2},
+profile based on a polytropic solution\footnote{where density and
+  pressure are assumed to be related by a power law}, \begin{equation}{
+\Sigma_{\rm Plummer} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
 }\end{equation}
 
-where \(M\) is the total mass and \(R_h\) is the 2D half-light radius.
-Now mostly superseded by the King profile for globular clusters, the
-Plummer model is still a good fit to many dwarf spheroidals {[}e.g.,
-moskowitz+walker2020{]}.
+where \(\Sigma_0\) is the central surface density and \(R_h\) is the 2D
+half-light radius. Now mostly superseded by the King profile for
+globular clusters, the Plummer model is still a good fit to many dwarf
+spheroidals \citep[e.g.,][]{moskowitz+walker2020}.
 
@@
 In much of the older literature, \(R_t\) was often called and
-interpreted as ``tidal radius'' as well, after the similar
-interpretation for globular clusters
-\citep[e.g.,][\citet{hodge1961}]{IH1995}.
+interpreted as a ``tidal radius'', after the similar interpretation for
+globular clusters \citep[e.g.,][\citet{hodge1961}]{IH1995}.
 
@@
 profile's equation is \begin{equation}{
-\Sigma_{\rm S\acute ersic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
+\Sigma_{\rm S\acute ersic} = \Sigma_h \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
 }\end{equation} where \(b_n\) is a constant depending on \(n\). \(n=1\)
@@
 Sérsic profile is not always used for dwarf galaxies, \citet{munoz+2018}
-advocate for using the Sérsic profile since the added flexibility allows
-for more profiles to be fit.
+advocate for the Sérsic profile since the added flexibility allows for
+more profiles to be fit.
 
-While there are not clear theoretical explanations why any profile is
-best, the Exponential density profile is commonly assumed for dwarf
-spheroidal galaxies. \citet{faber+lin1983} was one of the first to
-demonstrate that an Exponential density profile is a reasonable
-empirical fit to dwarf galaxies, theorizing that dwarf spheroidal evolve
-from exponential disk galaxies and maintained a similar light profile.
-Later, \citet{read+gilmore2005} showed that exponential profiles may
-originate from general mass loss during the evolution of dwarf galaxies.
-Many later photometric works for dwarf spheroidal galaxies apply
-exponential fits, finding that exponential and king profiles both
-provide good descriptions
+While there are no clear theoretical preferences for any of these
+profiles, exponential density profiles are commonly used for dwarf
+spheroidal galaxies. \citet{faber+lin1983} were among the first to
+demonstrate that an exponential law is a reasonable empirical fit to
+dwarf galaxies, theorizing that dwarf spheroidal may have evolved from
+exponential disk galaxies and maintained a similar light profile. Later,
+\citet{read+gilmore2005} showed that exponential profiles may originate
+from general mass loss during the evolution of dwarf galaxies. Many
+later photometric works for dwarf spheroidal galaxies have used
+exponential fits, finding that exponential and King profiles both
+provide good descriptions in many cases
 \citetext{\citealp[\citet{mateo1998}]{binggeli+sandage+tarenghi1984}; \citealp{mcconnachie+irwin2006}; \citealp{cicuendez+2018}}.
@@
 that the Sérsic indices are typically \(n \lesssim 1\), implying that
-most galaxies tend to be an exponential or slightly steeper.
+most galaxies tend to be an exponential or slightly steeper in the outer
+regions.
 
-Beyond local group dwarf spheroidals, exponential densities appear to be
+Beyond Local Group dwarf spheroidals, exponential profiles appear to be
 common, but sometimes with significant modifications. In particular,
-many extragalactic dwarf Elliptical and blue compact / Irr galaxies
-galaxies are fit well with an exponential + central cusp / nuclear
-region \citep[\citet{noeske+2003}]{caldwell+bothun1987}. Contrastingly,
-some studies find an inner deprivation of stellar material as compared
-to exponentials \citep[e.g.,][]{caldwell+1992, makarov+2012}.
+many extragalactic dwarf elliptical and blue compact / Irr galaxies are
+fit well with an exponential + central cusp / nuclear region
+\citep[\citet{noeske+2003}]{caldwell+bothun1987}. In constrast, some
+studies find an inner decrement in density compared to exponentials
+\citep[e.g.,][]{caldwell+1992, makarov+2012}.
 
-However, an exponential profile was often noted to not fit every dwarf
-galaxy, particular for more distant dwarf galaxies. Even starting from
-\citet{aparicio1997}, and \citet[for coma cluster
+However, exponential profiles do not fit every dwarf galaxy. Even
+starting from \citet{aparicio+1997}, and \citet[for coma cluster
 dE]{graham+guzman2003}, some people have noted deviations from this
-simple empirical rule. Additionally, \citet{hunter+elmegreen};
+simple empirical rule. Additionally, \citet{hunter+elmegreen2006};
 \citet{herrmann+hunter+elmegreen2013};
@@
 profiles. Irr do represent a substantially different class though so it
-is unclear if these conclusions apply. \citet{caldwell+1992} mostly well
-fit outer exponential, inner deviations photometric M31. Note that while
-diversity is also commented on here, dwarf galaxies with outer
-flattening profiles are often a minority in these studies.
+is unclear if these conclusions apply. \citet{caldwell+1992}'s
+photometric study of M31 dwarfs show that the outer density profile
+typically fits well to an exponential but with inner deviations. Note
+that while diversity is also commented on here, dwarf galaxies with
+outer flattening profiles are often a minority in these studies.
 
-Alternatively, \citet{moskowitz+walker2020}, use a plummer profile and
-modified plummer profiles with steeper outer slopes (gamma=-8 vs -4).
+Alternatively, \citet{moskowitz+walker2020}, use a Plummer profile and
+modified Plummer profiles with steeper outer slopes
+(\(\Gamma = d\log \Sigma / d \log R = -8\) compared to \(\Gamma = -4\)).
 For most galaxies, they do not have convincing statistical evidence to
@@
 direction of steeper outer cutoff or changes to the inner slope of a
-dwarf galaxy (due to a nuclear region/etc.). As such, deviations in
-directions of density profiles becoming steeper in the very outer
-regions are unusual. Explaining the origin, similarity, and diversity of
-dwarf galaxy density profiles is a pressing question in our theories of
-dwarf galaxy formation and evolution. Irregardless of if all dwarf
-galaxies are indeed exponential, there is nevertheless diversity
-requiring interpretation in the density profiles of dwarf galaxies.
+dwarf galaxy (due to a nuclear region/etc.). Flattened density profiles
+in the outer regions are more unusual. Explaining the origin,
+similarity, and diversity of dwarf galaxy density profiles is a pressing
+question for theories of dwarf galaxy formation and evolution. Whether
+dwarf galaxies are indeed exponential or not, the diversity of density
+profile shapes of dwarf galaxies needs to be explained.
 
@@
 glance. Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} describe
-the present-day properties of each galaxy. Indeed, Sculptor as the first
-discovered dwarf galaxy, is often described as a ``prototypical'' dSph.
-However, both galaxies have a long history of speculation that they may
-be influenced by the Milky Way's gravity (see discussion). With
-\emph{Gaia} data and spectroscopic followup, the detection of a density
-excess has become more unmistakable. Using \citet{jensen+2024}'s
-algorithm (described briefly below), \citet{sestito+2023a};
+the present-day properties of each galaxy. Sculptor, as the first
+discovered classical dSph, is often described as a ``prototypical''
+dSph. There has long been speculation that both Sculptor and Ursa Minor
+may have been influenced by the Milky Way's tidal field (see
+Section~\ref{sec:discussion}). \citet{sestito+2023a};
 \citet{sestito+2023b} report that a ``kink'' in the density profile,
 beginning around 30 arcmin for both Sculptor and Ursa Minor. They
-spectroscopically followup some of the most distant stars, finding
+spectroscopically follow up some of the most distant stars, finding
 multiple members between 6-12 half-light radii from the centre of each
-dwarf (REF fig.~XXX). If dwarfs initially begin with exponential
-profiles like Fornax, then these stars should be much rarer.
+dwarf. If dwarfs had exponential profiles like Fornax, then these stars
+should be much rarer.
 
-Perhaps the most straighforward explanation for the density profiles of
-Sculptor and Ursa Minor are tidal interactions. \citet{sestito+2023a};
-\citet{sestito+2023b} conclude that tides are likely the explanation.
-Fig.~\ref{fig:scl_umi_fnx_vs_penarrubia} reproduces their comparison,
-showing that Sculptor and Ursa Minor's density profiles deviate
-substantially from fornax's in the outskirts, and are well-described by
-the tidal model from \citet{PNM2008}. Our goal is to determine, assuming
-\(\Lambda\)CMD, if the effects of the Milky Way (or other satellites)
-may indeed create observable tidal signatures in these galaxies. If
-tides cannot explain these features, Sculptor and Ursa Minor may instead
-contain a extended stellar ``halo'' or second component of the
-galaxy---illustrating evidence of complex history or formation in each
-galaxy and forcing formation models to confront a diversity of density
-profiles in the local group.
+Sculptor and Ursa Minor are not well-described by an exponential
+profile. The left panel of Fig.~\ref{fig:scl_umi_vs_penarrubia} shows
+our density profiles of Sculptor, Ursa Minor, and Fornax (see
+Section~\ref{sec:observations} for our methods). Compared to Fornax,
+both Sculptor and Ursa Minor show an excess of stars beginning around
+\(\log R/R_h\approx 0.4\). The right panel of
+Fig.~\ref{fig:scl_umi_vs_penarrubia} compares the same density profiles
+to an initial and final density profile from a simple tidal model of an
+exponential dwarf spheroidal galaxy embedded in a NFW halo in the Milky
+Way field, described in Section~\ref{sec:tidal_theory}. Fornax agrees
+well with the exponential initial conditions. However, Sculptor and Ursa
+Minor are better described by the more extended, tidally evolved density
+profile.
+
+A goal of this work is to determine, assuming \(\Lambda\)CMD, if the
+effects of the Milky Way (or other satellites) may indeed create
+observable tidal signatures in Sculptor and Ursa Minor. If tides cannot
+explain these features, these features may instead be a extended stellar
+``halo'' or second component of the galaxy---suggestive of a complex
+history or formation in each galaxy, which needs to be explained by
+galaxy formation models of Local Group dwarf galaxies.
 
@@
 their half-light radius and the density at half-light radius (data
-described in Section~\ref{sec:data}). The right panel adds an idealized
-model with initial, final, and break radius as a dotted line, solid line
-and an arrow respectively see
+described in Section~\ref{sec:observations}). The right panel adds an
+idealized model with initial, final, and break radius as a dotted line,
+solid line and an arrow respectively see
 \ref{sec:break_radii}.}\label{fig:scl_umi_vs_penarrubia}
@@
 
-\section{Simulating tidal effects}\label{simulating-tidal-effects}
+\section{Cosmological context}\label{cosmological-context}
+
+We only understand a tiny fraction of the universe's composition. The
+leading theory of cosmology, \(\Lambda\)CDM (Lambda Cold Dark Matter),
+posits that the universe is composed of about 68\% dark energy
+(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
+  like to change definitions of words. \emph{Baryons} here means
+  baryons+leptons, i.e.~any standard model massive fermion.}
+\citep{planckcollaboration+2020}. While the composition of dark matter
+and dark energy remains elusive, we know their general properties. Dark
+energy causes the acceleration of the expansion of the universe on large
+scales. We do not discuss dark energy here---it does not substantially
+affect the Local Group today. Dark matter, instead, makes up the vast
+majority of mass in galaxies. Typically, galaxies have baryonic to dark
+matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
+In \(\Lambda\)CMD, dark matter is assumed to interact only
+gravitationally. Light passes through dark matter unimpeded---in this
+sense dark matter is transparent. Dark matter is also commonly assumed
+to be \emph{cold}, i.e.~typical velocities much smaller than the speed
+of light in the early universe. Implications of dark matter properties
+range from cosmological structural formation, galaxy structure, and
+galaxy interactions.
+
+\subsection{Structure formation and dwarf
+galaxies}\label{structure-formation-and-dwarf-galaxies}
+
+The very early universe was almost featureless. Our earliest
+observations of the universe stem from the cosmic microwave background
+(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
+But tiny perturbations in the CMB, temperature fluctuations of 1 part in
+10,000, reveal the underlying seeds of large-scale cosmological
+structure.\footnote{The power spectrum from the CMB is complicated by
+  baryons, which lead to oscillations in power based on the
+  sound-crossing timescale.} Governed by cosmological expansion,
+gravitational collapse, and baryonic physics, each overdensity grows
+into larger structures. Initially, baryonic matter was coupled to
+radiation and resisted collapse. Dark matter, only influenced by gravity
+instead, freely collapsed into the first structures. For mass
+perturbations both smaller than the horizon and overdense enough to
+collapse, these overdensities of dark matter become self-gravitating,
+known as \emph{halos}. After recombination, where electrons combined
+with atomic nuclei to form atoms, baryons decoupled from radiation and
+fell into the dark matter halos. The densest pockets of baryons later
+formed the first stars and galaxies.
+
+Dark matter halos, and their associated galaxies, rarely evolve in
+isolation. Instead, structure formation is \emph{hierarchical}. Small
+dark matter halos collapse first and hierarchically merge into
+progressively larger halos
+\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
+Structure formation happens on a wide range of scales. The largest
+structures become the cosmic web---composed of voids, filaments, and
+clusters---and the smallest structures directly observable host dwarf
+galaxies. Hierarchical assembly is evident through the large scale
+structure of the universe, remnants of past mergers within the Milky
+Way, and tidal disruption of dwarf galaxies and their streams around
+nearby galaxies.
+
+Small-scale structure formation is sensitive to deviations from
+\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. One key
+prediction of \(\Lambda\)CDM is that mass perturbations are expected to
+exist on all scales, and are largest on the smallest scales, so we would
+expect the formation of halos on all scales. Many alternative models,
+such as warm dark matter or self-interacting dark matter, may smooth out
+small-scale features and reduce the abundance of small halos or change
+their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies, which
+occupy the smallest dark matter halos, are promising windows into
+small-scale cosmological features. Alternative dark matter properties
+can influence dwarf galaxy abundance, formation, structure, and tidal
+evolution. The nearby dwarf galaxies of the Local Group, with detailed
+observations, present a promising opportunity to understand the
+evolution of these objects and test our understanding of cosmology. To
+understand the evolution of these objects, we need to understand the
+general predictions of \(\Lambda\)CDM for the properties of dwarf
+galaxies.
+
+\emph{Warm} dark matter, relativistic in the early universe but cooler
+now, smooths out small-scale features and softens the cusps of dark
+matter halos. \emph{Self-interacting} dark matter instead can form cores
+but also the cores can collapse into a density peak. \textbf{TODO:
+incorporate this here.}
+
+\subsection{Typical structure of dark matter
+halos}\label{typical-structure-of-dark-matter-halos}
+
+In \(\Lambda\)CDM cosmological simulations, dark matter halos are
+remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
+authors observe that the spherically-averaged density profiles
+\(\rho(r)\) are universally well described by a two-parameter law:
+\begin{equation}\protect\phantomsection\label{eq:nfw}{
+\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
+}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
+density . This profile has shown remarkable success in describing
+\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
+profiles are \emph{cuspy}, where the density continuously rises like
+\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
+density profile increases gradually with radius, and at large radii the
+density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
+shows an example of a NFW halo.
+
+The total mass of an NFW profile diverges, so halos are commonly defined
+using an overdensity criterion. The virial mass, \(M_{200}\), is defined
+as the mass within a radius \(r_{200}\) containing a mean enclosed
+density 200 times\footnote{For the collapse of a uniform spherical
+  density, the virialized overdensity would be
+  \(\Delta = 18\pi^2\approx 178\) for a critical universe
+  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
+  this parameter may be closer to \(\Delta \approx 100\) for our
+  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
+  from][]{bryan+norman1998}.} the critical density of the universe:
+\begin{equation}{
+M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
+}\end{equation} A standard second parameter is the halo concentration,
+\(c=r_{200} / r_s\) describing how the characteristic size of the halo
+compares to its virial radius. In this case, the scale density is a
+function of \(c\) alone,
+\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
+\citep{NFW1996}. However, we characterize halos by their circular
+velocity profiles. The circular velocity,
+\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
+\(v_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
+\(v_{\rm max}\) is related to both the total halo mass and the observed
+line of sight velocity dispersion, and \(r_{\rm max}\) relates to the
+scale radius.
+
+While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
+these are not independent. Lower mass dark matter halos often collapse
+earlier, when the universe was denser. As a result, low mass subhalos
+tend to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
+between \(M_{200}\) and c, or the mass-concentration relation, describes
+the mean trend of concentration with mass
+\citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
+Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
+relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
+a general trend for concentration to increase with mass, the
+concentration values still have substantial scatter. Other parameters
+such as the halo spin or shape may affect the scatter of the
+mass-concentration relation, but their effect is typically expected to
+be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
+\caption[Example density profiles]{Density profiles in log density
+versus log radius for the stars and dark matter of a Sculptor-like
+galaxy. The dark matter is more extended and massive than the star
+across the entire galaxy---dynamical evolution is driven by the dark
+matter mass distribution.}\label{fig:nfw_density}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
+\caption[Stellar-mass halo-mass relation]{\textbf{Left} Radius of
+maximum circular velocity \(r_{\rm max}\) as a function of maximum
+circular velocity. The solid line with 1-\(\sigma\) shaded region and
+dashed line are the relations from \citet{ludlow+2016} for \(z=0\) and
+\(z=2\) respectively. \textbf{Right} Stellar mass (top) as a function of
+maximum circular velocity. The solid line with the 1-\(\sigma\) shaded
+region is the relation from \citet{fattahi+2018} with scatter points
+simulated central galaxies from APOSTLE in
+\citet{fattahi+2018}.}\label{fig:smhm}
+\end{figure}
+
+\subsection{Connecting dark matter to
+stars}\label{connecting-dark-matter-to-stars}
+
+Unfortunately, directly observing dark matter remains infeasible (lest
+dark matter wouldn't be dark). Fortunately, dwarf spheroidal galaxies
+come with a convenient population of visible stars. Stars typically
+represent a small fraction of the total mass of a dwarf galaxy. For
+example, Fig.~\ref{fig:nfw_density} shows the typical density profile
+for the stars and dark matter for a Sculptor-like galaxy. At all radii,
+dark matter are at least \textasciitilde10 times more dense, and the
+total virial mass is often \textasciitilde1000 times higher than the
+stellar mass alone. As a result, the dynamics and formation of dwarf
+galaxies is governed by the underlying dark matter halo.
+
+Both cosmology and observations find fundamental relationships between
+the dynamical or total mass and stellar masses of
+galaxies.\textbf{Improve last sentence.} A prediction of cosmological
+simulations is the Stellar mass Halo Mass relation (SMHM), describing
+the mass of stars which form in a halo of a given size. In particular,
+the SMHM relation grows especially steep in the dwarf galaxy
+regime---many dwarf galaxies are formed in halos of similar mass.
+Fig.~\ref{fig:smhm} shows the stellar mass versus \(v_{\rm max}\)
+maximum circular velocity (proxy for halo mass) for APOSTLE dwarfs from
+\citet{fattahi+2018}. The APOSTLE project \citep{sawala+2016} uses the
+hydrodynamical setup from the EAGLE project
+\citep{crain+2015, schaye+2015} to simulate Local Group analogues in a
+\(\Lambda\)CDM cosmological context (db: reorder these sentences). While
+there is some scatter, the range of predicted \(v_{\rm max}\) is fairly
+narrow across \(\sim 4\) decades in stellar mass. Because lower mass
+galaxies have increasingly shallow potential wells, feedback becomes
+more effective at removing gas. Re-ionization additionally suppresses
+late star formation in the faintest galaxies. As a result, the resulting
+stellar mass of a galaxy is highly sensitive to the dark matter mass,
+especially for faint dwarf galaxies. If we know the initial stellar mass
+of a dwarf galaxy, we also know, at some level, the properties of its
+dark matter halo.
+
+Several challenges complicate a simple SMHM trend including environment,
+assembly history, and tidal effects. Additionally, the detailed physics
+of galaxy formation can impact the SMHM relationship. By being closer to
+a massive host, many dwarfs quench earlier, resulting in lower stellar
+masses \citep[e.g.,][]{christensen+2024}. In particular, effects like
+ram-pressure stripping (removal of gas in the dwarf galaxy due to
+pressure from host's circumgalactic medium) and tidal removal of gas
+cause star formation to quench {[}REFS{]}. Additionally, the time of
+formation (relative to reionization) can influence the resulting stellar
+content \citep{kim+2024}. Finally, tides influence both the dark matter
+and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
+Consequently, tides may reduce the halo mass more than the stellar mass,
+adding additional scatter to the SMHM trend, particularly for low-mass
+satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
+tides on Local Group dwarf galaxies is essential to determine where and
+how these galaxies formed in a cosmological context.
+
+\section{Simulating tidal effects}\label{sec:tidal_theory}
 
@@
 considered or simulated tidal effects. While pre-\emph{Gaia} work often
-did not know the specific orbits of dwarf galaxies, the theory of tidal
+had few constraints on orbits of dwarf galaxies, the theory of tidal
 mass loss remains largely the same.
@@
 Simulating dwarf galaxies accurately in a cosmological context remains a
-substantial challenge. Currently, Cosmological simulations can predict
-the overall abundance of larger mass dwarf galaxies {[}e.g., {]} and
+substantial challenge. Currently, cosmological simulations can predict
+the overall abundance of larger mass? dwarf galaxies {[}e.g., {]} and
 broadly predict effects of tides \citep[e.g.,][]{riley+2024}. But, dwarf
@@
 can lead to artificial disruption of dwarf galaxies and overestimation
-of tidal effects \citep[e.g.,][]{santos-santos+2024}. For example, the
+of tidal effects \citep[e.g.,][]{santos-santos+2025}. For example, the
 highest resolution cosmological simulation of a Milky Way-size dark
 matter halo, the Aquarius project \citep{springel+2008}, achieved a DM
-resolution of \(1.712 \times10^3 \Mo\), enough to barely resolve
-Sculptor like halos (see Section~\ref{sec:Methods}). To address this
-challenge, idealized simulations only simulate a single subhalo in an
-approximate host potential. As a result, idealized simulations can reach
-excellent numerical convergence at the cost of realism. For instance,
-our simulations later are about 3x higher resolution than Aquarius but
-with a fraction of the computational cost (400x less particles).
-Idealized simulations make numerous simplifications: neglecting mergers,
-cosmological context and evolution, mass assembly, and often baryonic
-physics. We use idealized simulations here which are more powerful in
-accurately assessing tidal effects after infall, given that the
-idealizations do not impact these galaxies's recent history too much. As
-such, a long history of work has resolved to understand the tidal
-evolution of dwarf galaxies using these \emph{idealized} simulations.
+resolution of \(1.712 \times10^3 \Mo\), enough to barely resolve the
+stellar components of Sculptor like halos (see chapter
+{[}@-sec:methods{]}). To address this challenge, idealized simulations
+only simulate a single subhalo in an approximate host potential. As a
+result, idealized simulations can reach excellent numerical convergence
+at the cost of realism. For instance, our simulations later are about 3x
+higher resolution than Aquarius at a fraction of the computational cost
+(400x less particles). Idealized simulations make numerous
+simplifications: neglecting mergers, cosmological context and evolution,
+mass assembly, and often baryonic physics. We use idealized simulations
+here which are more powerful in accurately assessing tidal effects after
+infall, given that the idealizations do not impact these galaxies's
+recent history too much.
 
@@
 \citet{moore+davis1994}; \citet{johnston+spergel+hernquist1995}.
-Already, these works used similar techniques as we continue to use today
-and laid the foundation for understanding tidal effects. While the
-detailed assumptions have evolved (e.g.~assumed total mass of dwarf
-galaxies and evolving beyond \citet{king1966} stellar), many of their
-conclusions still hold true. More recent works expanding this theory
-include \citet{read+2006}; \citet{bullock+johnston2005};
-\citet{PNM2008}; \citet{penarrubia+2009}; \citet{klimentowski+2009};
-\citet{errani+2023a}; \citet{fattahi+2018}; \citet{stucker+2023};
-\citet{wang+2017}.
+Already, these works used techniques similar to what we continue to use
+today and laid the foundation for understanding tidal effects. While the
+detailed assumptions have evolved (e.g.~no longer assuming
+mass-follows-light), many of their conclusions still hold true. More
+recent work expanding this theory include \citet{read+2006};
+\citet{bullock+johnston2005}; \citet{PNM2008}; \citet{penarrubia+2009};
+\citet{klimentowski+2009}; \citet{errani+2023a}; \citet{fattahi+2018};
+\citet{stucker+2023}; \citet{wang+2017}.
 
@@
 
+\textbf{\emph{These past three paragraphs need revisions\ldots{}}}
+
 \subsection{Simulating large gravitational systems: The N-body
@@
 
-Modelling gravitational evolution is essential for understanding
-properties of galaxies. Perhaps the simplest method to compute the
-evolution of dark matter is through \emph{N-body simulations}. A dark
-matter halo is represented as a large number of dark matter particles
-(bodies). Each body represents a monte-carlo sample from the underlying
-matter distribution. However, galaxies are often assumed to be
-\emph{collisionless}---particles are not strongly affected by
-small-scale gravitational encounters. In contrast, star clusters evolve
-differently as star-star gravitational collisions changes the dynamics
-of the system. While we use individual gravitating bodies in N-body
+Modelling gravitational evolution for large systems requires special
+methods. Perhaps the simplest method to compute the evolution of dark
+matter is through \emph{N-body simulations}. A dark matter halo is
+represented as a large number of dark matter particles (bodies). Each
+body is essentially a Monte Carlo sample of the underlying phase-space
+distribution. Note that dark matter (and galaxies) are often assumed to
+be \emph{collisionless}---particles are not strongly affected by close,
+\emph{collisional} gravitational encounters which substantially change
+the momenta of involved bodies. In contrast, star clusters are often
+collisional so neglecting these encounters may not be a reasonable
+approximation. While we use individual gravitating bodies in N-body
 simulations, the Newtonian gravitational force is softened to be a
-Plummer sphere, so encounters closer than a softening length do not
-substantially impact the dynamical evolution.
+Plummer sphere as to limit strongly collisional encounters.
 
-Naively, the newtonian gravitational force requires adding together the
-forces from each particle at each particle, causing a computational cost
+Naively, the Newtonian gravitational force requires adding together the
+forces from each particle on each particle, with a computational cost
 that scales quadratically with the number of particles, or \(O(N^2)\).
 With this method, simulating a large number of particles, such as
-10\^{}6, would require about 10\^{}12 force evaluations at each time
-step, making cosmological and high-resolution studies unfeasible.
-However, only long-range gravitational interactions tend to be important
-for CMD, so we can utilize the \emph{tree method} to compute the
-gravitational force vastly more efficiently.
+\(10^6\), would require \(10^{12}\) force evaluations at each time step,
+making cosmological and high-resolution studies unfeasible. However,
+only long-range gravitational interactions tend to be important for CMD,
+so we can utilize the \emph{tree method} to compute the gravitational
+force vastly more efficiently.
 
@@
 
-\subsection{Break and tidal radii}\label{sec:break_radii}
-
-Analytic approximations are powerful tools to generally determine the
-influence of tides given a dwarf galaxy and a host. A variety of ``tidal
-radii'' exist in the literature, we focus on two simple forms, the
-Jacobi radius and the break radius.
-
-The \textbf{Jacobi radius} represents the approximate radius where stars
-become unbound for a galaxy in a circular orbit around a host galaxy.
-Calculated from an approximation of the location of the L1 and L2
-lagrange points, the Jacobi radius is where the mean density of the
-dwarf galaxy is three times the mean interior density of the host galaxy
-at pericentre, or \begin{equation}{
-3\bar \rho_{\rm MW}(r_{\rm peri}) = \bar \rho_{\rm dwarf}(r_J).
-}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
-we should expect to find unbound, \emph{extratidal} stars. While most
-valid for circular orbits, assuming \(r_{\rm peri}\) for the host-dwarf
-distance works as most stars are lost near pericentre.
-
-We use the \textbf{break radius} as defined in \citet{penarrubia+2009},
-marking where the galaxy is still in disequilibrium. The break radius
-\(r_{\rm break}\) is proportional to the velocity dispersion
-\(\sigma_v\) and time since pericentre \(\Delta t\), \begin{equation}{
-r_{\rm break} = C\,\sigma_{v}\,\Delta t
-}\end{equation} where the scaling constant \(C \approx 0.55\) is a fit.
-\(r_{\rm break}\) describes where the dynamical timescale is longer than
-the time since the perturbation, i.e.~the radius within which the galaxy
-should have dynamically relaxed. As illustrated in
-Fig.~\ref{fig:toy_profiles}, for an idealized model with exponential
-stars in a NFW halo, after some time past pericentre, the stellar
-component is smooth but contains a change in slope around
-\(r_{\rm break}\), where the radial velocities of the stars becomes
-positive as they still readjust to a new equilibrium.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
-\caption[Break radius validation]{The break radius of the simulations is
-set by the time since pericentre. The initial conditions are
-Sculptor-like, exponential stars embedded in NFW, evolved in
-\citet{EP2020} potential with a pericentre of 15 kpc and apocentre of
-100 kpc. In this model, the jacobi radius is close to the break radius.
-See Section~\ref{sec:methods} for a description of our simulation
-setup.}\label{fig:idealized_break_radius}
-\end{figure}
-
 \subsection{Tidal evolution}\label{tidal-evolution}
 
-While the break and jacobi radii help with understanding when tidal
-effects matter, we still need to know what these effects do and how a
-galaxy evolves.
+A galaxy in equilibrium will remain in equilibrium, unless acted upon by
+an external force. As an example, Fig.~\ref{fig:lagrange_points}
+illustrates the effective potential around a Sculptor-like NFW galaxy in
+a circular orbit around a Milky Way-like galaxy. The effective potential
+accounts for the centrifugal force in the rotating frame. There are two
+key saddle points in Fig.~\ref{fig:lagrange_points}, labeled \(L_1\) and
+\(L_2\) (as Lagrange points, there are another 3 not as relevant). Since
+a particles trajectory is constrained to be within the equipotential
+surface equal to the particle's total energy, the easiest (requiring the
+lowest energy) path for a particle to escape the satellites
+gravitational influence is through the windows around \(L_1\) and
+\(L_2\). Otherwise, the potential steeply increases. \(L_1\) and \(L_2\)
+are co-linear with the satellite and host-origin. In general, particles
+which have higher energy than the energy at \(L_1\) and \(L_2\) are most
+commonly unbound from the galaxy.
 
-As a galaxy evolves in a tidal field, several changes happen:
+The full evolution of a dwarf galaxy in a tidal follows an approximate
+sequence (although somewhat concurrent as well.)
 
@@
   \emph{Mass is lost}. In particular, particles and stars on weakly
-  bound orbits are most likely to be removed by tides.
+  bound orbits are most likely to be removed by tides. Particles escape
+  through \(L_1\) and \(L_2\)
 \item
-  \emph{Steams form}. If tides are strong enough, unbound mass becomes
-  part of tidal tails, evolving along a similar orbit but leading or
-  trailing the galaxy.
+  \emph{Steams form}. Unbound mass becomes part of a stream or tidal
+  tails. These particles follow similar orbits but are slightly higher
+  or lower energy (depending on which side the particle escaped from).
 \item
@@
   decreases, resulting in a more compact dark matter halo and stars
-  which adiobadically expand to a larger scale radius.
+  which adiabadically expand to a larger scale radius.
 \end{enumerate}
 
-Assuming that galaxies are spherical, isotropic, and evolve in a
-constant tidal field, most dwarf galaxies should evolve along similar
-tidal tracks. \citet{EN2021} derive tidal tracks for dwarf galaxies,
-showing that NFW halos all evolve along a similar trajectory in terms of
-\(r_{\rm max}\) and \(v_{\rm max}\).
+A tidally affected galaxy contains predictable observational clues to
+this process. Nearby to the galaxy, the density profile is flattened as
+a result of newly unbound material. Further away from the galaxy, this
+would be visible as a stream. Additionally, a satellite stream contains
+a velocity gradient along the path. Finally, due to mass loss, a
+satellite will have a reduced velocity dispersion, larger stellar size,
+and
+
+Dwarf galaxies evolve along \emph{tidal tracks}. From N-body simulations
+across a numerous initial conditions and orbits, NFW halos evolve along
+a narrow track in terms of maximum circular velocity and radius
+\(v_{\rm max}\) and \(r_{\rm max}\) . \citet{EN2021} derive an empirical
+fit to these tidal tracks, finding that
 
@@
 2^\alpha 
-\left(\frac{r_{\rm max}}{r_{\rm max, 0}}\right)^{\beta}\left[1 + \left(r_{\rm max} / r_{\rm max, 0}\right)^2\right]^{-\beta}
-}\end{equation} where \(\alpha=0.4\), \(\beta=0.65\) are empirical fits.
-As illustrated in Fig.~\ref{fig:tidal_tracks}, this formula works for
-both circular and elliptical orbits and is independent of the initial
-subhalo size or distance to the host.
+\left(\frac{r_{\rm max}}{r_{\rm max, 0}}\right)^{\beta}\left[1 + \left(r_{\rm max} / r_{\rm max, 0}\right)^2\right]^{-\beta},
+}\end{equation} where \(\alpha=0.4\) and \(\beta=0.65\). As illustrated
+in Fig.~\ref{fig:tidal_tracks}, this formula works for both circular and
+elliptical orbits and is independent of the initial subhalo size or
+distance to the host. \textbf{\emph{Rewrite this paragrah}}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{./figures/lagrange_points.pdf}}
+\caption[Lagrange points]{The potential contours for a 2-body galaxy
+system in the rotating frame. \(L_1\) and and \(L_2\) is most efficient
+as the required energy (the effective potential) is the
+smallest.}\label{fig:lagrange_points}
+\end{figure}
 
@@
 
+\subsection{Break and tidal radii}\label{sec:break_radii}
+
+Following the detailed discussion of tidal evolution, we can interpret
+two simple analytic approximations of where tidal effects should be
+visible.
+
+The \textbf{Jacobi radius} represents the approximate radius where stars
+become unbound for a galaxy in a circular orbit around a host galaxy.
+Calculated from an approximation of the location of the \(L_1\) and
+\(L_2\) Lagrange points, the Jacobi radius is where the mean density of
+the dwarf galaxy is roughly three times the mean interior density of the
+host galaxy at pericentre, or \begin{equation}{
+3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J).
+}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
+we should expect to find relatively clear signs of tidal disturbance.
+While strictly valid for circular orbits, assuming \(r_{\rm peri}\) for
+the host-dwarf distance works as most stars are lost near pericentre.
+
+We also use the \textbf{break radius} as defined in
+\citet{penarrubia+2009}, marking where the galaxy is still in
+disequilibrium. The break radius \(r_{\rm break}\) is proportional to
+the velocity dispersion ,\(\sigma_v\), and time elapsed since pericentre
+,\(\Delta t\), \begin{equation}{
+r_{\rm break} = C\,\sigma_{v}\,\Delta t
+}\end{equation} where the scaling constant \(C \approx 0.55\) is derived
+empirically. \(r_{\rm break}\) describes where the dynamical timescale
+is longer than the time since the perturbation, i.e.~the radius within
+which the galaxy should have dynamically relaxed. As illustrated in
+Fig.~\ref{fig:idealized_break_radius}, for an idealized model with
+exponential stars in a NFW halo, shortly after pericentre, the stellar
+component is smooth but contains a change in slope around
+\(r_{\rm break}\). This radius is visible in the stellar distribution as
+non-spherical S-shaped overdensities of stars. Also, \(r_{\rm break}\)
+is where where the mean radial velocities of the stars becomes positive,
+i.e.~the system is out of equilibrium and adjusting to the new density
+profile.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
+\caption[Break radius validation]{Example density and velocity
+distributions of an idealized simulation shortly after pericentre.
+\textbf{Top left}: The 2D density profile for the initial and final
+simulation with the break radius marked. The break radius of the
+simulations is set by the time since pericentre. \textbf{Top right}: The
+projected 2D stellar density in the \(x\)-\(y\) plane. The green circle
+represents the break radius and the grey arrow points towards the host
+centre. \textbf{Bottom left}: the mean radial velocity (dot product of
+relative position and velocity relative to dwarf centre) as a function
+of 2D radius. \textbf{Bottom right}: The mean radial velocity as
+projected into 2D bins. The initial conditions are Sculptor-like,
+exponential stars embedded in NFW, evolved in \citet{EP2020} potential
+with a pericentre of 10 kpc and apocentre of 100 kpc shortly after the
+3rd pericentre. See Section~\ref{sec:methods} for a description of our
+simulation setup. \emph{Note to self:} This figure may be too complex,
+not sure the best way to present everything
+yet.}\label{fig:idealized_break_radius}
+\end{figure}
+
 \section{Thesis outline}\label{thesis-outline}
@@
 
-In chapter 2, we describe how we compute observational density profiles
-from \citet{jensen+2024}. In chapter 3, we review our simulation
-methods. Next, we present our results for the tidal effects on Sculptor
-and Ursa Minor in Chapter 4. We discuss our results, limitations, and
-implications in Chapter 5. Finally, Chapter 6 summarizes this work and
-discuss future directions for similar work and the field of dwarf
-galaxies.
+In chapter {[}@-sec:observations{]}, we describe how we compute
+observational density profiles from \citet{jensen+2024}. In chapter
+{[}@-sec:methods{]}, we review our simulation methods. Next, we present
+our results for the tidal effects on Sculptor and Ursa Minor in chapter
+{[}@-sec:results{]}, We discuss our results, limitations, and
+implications in chapter {[}@-sec:discussion{]}. Finally, chapter
+{[}@-sec:summary{]} summarizes this work and discuss future directions
+for similar work and the field of dwarf galaxies.

### 2025-08-18 thesis/methods.tex
thesis julio ch1 edits

@@
 In this section, we discuss our methods used for the following two
-chapters. We utilize a typical analytic, idealized Milky Way potential
-based on \citet{mcmillan2011}. Our simulations are collisionless cold
-dark matter N-body simulations integrated with Gadget-4. We check for
-numerical convergence and stability of initial conditions in isolation.
+chapters. We utilize an analytic, static, axisymmetric Milky Way
+potential based on \citet{mcmillan2011}. Our N-body simulation assume
+collisionless, cold dark matter and are integrated with Gadget-4. We
+check for numerical convergence and stability of initial conditions in
+isolation.
 
@@
 Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
-each component of our fiducially potential. The potential consisting of
-a stellar Bulge, a thin and thick disk, and a dark matter NFW halo.
-Especially in the regime where Sculptor and Ursa Minor orbit, the dark
-matter halo is the dominant component of the potential. The details of
-the stellar disk and bulge are less important.
+each component and the total circular velocity profile for our fiducial
+profile. The potential consisting of a stellar bulge, a thin and thick
+disk, and a dark matter NFW halo. Among typical orbits for Sculptor and
+Ursa Minor, the dark matter halo dominates the potential.
 
 The potential is specified as follows. The galactic bulge is described
-by a \citet{hernquist1990} potential. The potential is
+by a \citet{hernquist1990} potential,
 
 \begin{equation}\protect\phantomsection\label{eq:hernquist}{
-\Phi(r) = - \frac{GM}{r + a}
+\Phi(r) = - \frac{GM}{r + a},
 }\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
@@
 by \begin{equation}\protect\phantomsection\label{eq:t_relax}{
-t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
+t_{\rm relax}(r) = t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
 = {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
@@
 zero provided that \(n > 1\). We take \(\Psi\) from the underlying
-assumed analytic dark matter potential. \(\rho_\star\) is pre-specified
-or calculated via Abel integration / deprojection.
-
-We consider both a Plummer and 2-dimensional exponential (Exp2D) stellar
-models:
-
-\subsubsection{Exp2D}\label{exp2d}
-
-Perhaps the simplest in form, the 2D exponential is commonly used to
-describe dwarf galaxy's density profiles as well as stellar disks. The
-stellar profile is given by \begin{equation}{
-\Sigma_\star(R) = A e^{-R / R_s}
-}\end{equation} where \(A\) is the normalization and \(R_s\) is the
-exponential scale radius. \#\#\# Plummer
-
-The Plummer density profile is generally considered to be a reasonable
-empirical fit to Globular Clusters, where it was first proposed
-\citep{sadfjk}. Plummer profiles are also sometimes used to describe
-dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
-profile is defined by \begin{equation}{
-\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
-}\end{equation} where \(M\) is the total mass and \(R_s\) is the
-characteristic scale radius.
+assumed analytic dark matter potential. \(\rho_\star\) can be calculated
+from the surface density, \(\Sigma_\star\), via the inverse Abel
+transform. In this work, we consider both Plummer and exponential
+profiles, as described in
 

### 2025-08-18 thesis/sculptor.tex
thesis julio ch1 edits

@@
+In this chapter, we present our results from our simulations of Sculptor
+and Ursa Minor. In each case, we first discuss the range of reasonable
+orbits and initial dark matter halos for each galaxy. Then, we describe
+how tides affect the evolution of each galaxy. We furthermore consider
+how the effects of the LMC changes the results, including using
+additional N-body simulations in evolving potentials. Finally, we review
+the predictions and characteristics of the present-day stellar
+populations and properties from our models.
+
 \section{Sculptor}\label{sculptor}
 
-\subsection{Derived properties}\label{derived-properties}
+Before running N-body simulations, we need to understand both the
+typical halos which could host a sculptor like galaxy and the possible
+orbits Sculptor.
+
+Fig.~\ref{fig:scl_halos} presents approximate constraints based on
+
+\subsubsection{Derived properties and
+setup}\label{derived-properties-and-setup}
 
@@
 \centering
+\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al.
+(2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008).
+Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$.
+From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al.
+(2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$.
+}
+\label{tbl:scl_derived_props}
 \begin{tabular}{lll}
@@
 $M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
-$M_\star / L_\star$ & 1.7, 0.17 dex uncertainties. & (woo+2009?)\\
-$M_{200}$ & $0.5_{-0.3}^{+0.6} \times10^10\ M_0$ & \\
+$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
 $c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
-$v_{\rm circ, max}$ & $31_{-6}^{+8}\,\kms$ & \\
-$r_{\rm circ, max}$ & $6_{-2}^{+3}$ kpc & \\
-$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & \\
-$t_{\rm break, obs}$ & $110\pm30$ Myr & \\
+$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
+$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
+$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
+$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
 \bottomrule
@@
 
-For the mass-to-light ratio, we use the values from \citet{woo+2009}.
-Note that we assume a scatter of 0.17 dex for the mass-to-light ratios
-as reported for mass uncertainties. \citet{delosreyes+2024} also
-indicate that other methods for stellar mass-to-light ratios result in
-similar scatters and systematic biases. However these values are just
-approximate.
+For the mass-to-light ratio, we use the values from
+\citet{woo+courteau+dekel2008}. Note that we assume a scatter of 0.17
+dex for the mass-to-light ratios as reported for mass uncertainties.
+\citet{delosreyes+2024} also indicate that other methods for stellar
+mass-to-light ratios result in similar scatters and systematic biases.
+However these values are just approximate.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_initial_halos.pdf}}
+\caption[Sculptor initial halos]{The suggested halos of Sculptor
+compared to cosmological predictions.}\label{fig:scl_halos}
+\end{figure}
 
@@
 \centering
+\caption[Initial halos of Sculptor]{The parameters for our initial Sculptor halos. }
+\label{tbl:scl_ini_halos}
 \begin{tabular}{lllll}
@@
 \midrule
-fiducial & 3.2 & 31 & 0.33 & 21\\
-extended & 4.2 & 31 & 0.39 & 17\\
+compact & 3.2 & 31 & 0.33 & 21\\
+lmc & 4.2 & 31 & 0.39 & 17\\
+small & 2.5 & 25 &  & \\
 \bottomrule
@@
 
-Fig.~\ref{fig:unref} contains example orbits
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_xyzr_orbits.pdf}}
+\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
+potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
+Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
+plane. Our selected \texttt{smallperi} orbit is plotted in black and
+light blue transparent orbits represent the past 5Gyr orbits sampled
+over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
+sculptor is well-constrained in this potential and it is unlikely to
+achieve a smaller pericentre than our selected orbit.
+\the\textwidth}\label{fig:scl_orbits}
+\end{figure}
 
-Given the parameters in table@REF, we
+To select an orbit with \textasciitilde{} maximum possible
+observationally-consistent tidal forces, we take the median parameters
+for orbits with a pericentre less than 2x the 3\(\sigma\) minimum
+pericentre and derive the following orbit. The initial conditions are
+from the
 
-\subsubsection{Tidal effects}\label{tidal-effects}
+\begin{table*}[t]
+\centering
+\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
+The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
+The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+}
+\label{tbl:scl_obrits}
+\begin{tabular}{llll}
+\toprule
+Property & Mean & SmallPeri & LMC\\
+\midrule
+distance & 83.2 & 82.6 & \\
+pmra & 0.099 & 0.134 & \\
+pmdec & -0.160 & -0.198 & \\
+Vlos & 111.2 & 111.2 & \\
+$t_i$ & -8.74 & -9.43 & \\
+$\hat{x}_{i}$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10] & \\
+$\vec{v}_i$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29] & \\
+pericentre & 53 & 43 & \\
+apocentre & 102 & 96 & \\
+$t_{\rm last\ peri}$ & -0.45 & -0.46 & \\
+Numer of peris &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Tidal effects}\label{tidal-effects}
 
@@
 \end{itemize}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
+\caption[Sculptor simulation snapshots]{Images of the dark matter
+evolution over a selection of past apocentres and the present day
+position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
+(approximately orbital) plane and the colourscale is logarithmic
+spanning 5 orders of magnitude between the maximum and minimum values.
+In this image, stars occupy only ever a few pixels so are not plotted.}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
+\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
+orbit. Todo: add velocity dispersion plot to RHS}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
+\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
+velocity dispersion within 1 kpc for different Scl models. In all cases,
+the evolution is mild. Note that binarity may reduce the inflate the
+observed velocity dispersion by \textasciitilde{} 1 km/s, so the
+conservative lower limit is around 8 km/s.}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
+\caption[Sculptor initial and final density profiles]{Effects on
+exponential initial stars. TODO: plot 2D sky proj. stars}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
+\caption[Sculptor Plummer initial and final density profiles]{effects on
+Plummer initial stars.}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}
-\caption[Effect of LMC on Sculptor's Orbit]{Orbits of Sculptor with and
-without the LMC. The potential is that of \citet{vasiliev+2021} in each
-case.}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.pdf}}
+\caption[Sculptor Orbits with LMC]{This figure is similar to
+Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
+without an LMC. In the bottom row, the distance from Sculptor (or the
+LMC) to the MW is plotted (left), and the Sculptor - LMC distance
+(right.)}\label{fig:scl_lmc_orbit_effect}
+\end{figure}
+
+\textbf{tinyperilmc}
+
+\begin{itemize}
+\item
+  ra = 15.0183 dec = -33.7186 distance = 73.1 pmra = 0.137 pmdec =
+  -0.156 radial\_velocity = 111.2
+
+  t\_i = -2.00 pericentre = 38.82 apocentre = 187.50 t last peri = -0.33
+  x\_i = {[}4.30 138.89 125.88{]} v\_i = {[}6.89 -56.83 52.09{]}
+\end{itemize}
+
+Smallperi
+
+\begin{itemize}
+\item
+  ra = 15.0183 dec = -33.7186 distance = 84.0 pmra = 0.166 pmdec =
+  -0.237 radial\_velocity = 111.2
+
+  t\_i = -2.00 pericentre = 28.00 apocentre = 190.45 t last peri = -0.39
+  x\_i = {[}12.79 168.76 87.33{]} v\_i = {[}0.37 -57.69 61.40{]}
+\end{itemize}
+
+\subsubsection{Tidal effects}\label{tidal-effects-1}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.pdf}}
+\caption{Sculptor Simulation Snapshots with LMC}
 \end{figure}
@@
 \centering
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}
-\caption[Selected orbits of Sculptor with an LMC]{Selected orbits of
-Sculptor in the MW frame. \textbf{TODO}: include LMC distance or frame,
-maybe xyz instead?}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
+\caption[Sculptor initial and final density with LMC]{The tidal effects
+on the stellar surface density due to the LMC today.}
 \end{figure}
-
-\subsubsection{Tidal effects}\label{tidal-effects-1}
 
@@
 
-\subsection{Derived properties}\label{derived-properties-1}
+\subsection{Derived properties}\label{derived-properties}
 
@@
 \centering
+\caption[Derived Properties of Ursa Minor]{Derived properties of Ursa Minor. }
+\label{tbl:umi_derived_props}
 \begin{tabular}{lll}
@@
 $M_\star$ & $7_{-2}^{+3} \times 10^5\,\Mo$ & \\
-$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & (woo+2009?)\\
+$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & Woo, Courteau, and Dekel (2008)\\
 $M_{200}$ & $3_{-2}^{+4} \times 10^9\,\Mo$ & \\
@@
 \centering
+\caption[Ursa Minor Initial Halos]{Initial halos for Ursa Minor. }
+\label{tbl:umi_ini_halos}
 \begin{tabular}{lllll}
@@
 
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_initial_halos.pdf}}
+\caption[Ursa Minor initial halos]{The suggested halos of Ursa Minor
+compared to cosmological predictions.}
+\end{figure}
+
 \subsection{Orbital properties}\label{orbital-properties-1}
 
-\begin{table*}[t]
+\begin{figure}
 \centering
-\begin{tabular}{lllllllll}
-\toprule
-Orbit & Mean &  & smallperi &  &  &  &  & \\
-\midrule
-ra &  &  &  &  &  &  &  & \\
-dec &  &  &  &  &  &  &  & \\
-distance &  &  &  &  &  &  &  & \\
-pmra &  &  &  &  &  &  &  & \\
-pmdec &  &  &  &  &  &  &  & \\
-rv &  &  &  &  &  &  &  & \\
-x &  &  &  &  &  &  &  & \\
-y &  &  &  &  &  &  &  & \\
-z &  &  &  &  &  &  &  & \\
-vx &  &  &  &  &  &  &  & \\
-vy &  &  &  &  &  &  &  & \\
-vz &  &  &  &  &  &  &  & \\
-\bottomrule
-\end{tabular}
-\end{table*}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_xyzr_orbits.pdf}}
+\caption{Ursa Minor Orbits}
+\end{figure}
 
-\subsection{Tidal effects}\label{tidal-effects-2}
+\subsection{Tidal Effects}\label{tidal-effects-2}
+
+ra = 227.242 dec = 67.2221 distance = 64.6 pmra = -0.158 pmdec = 0.05
+radial\_velocity = -245.75
+
+t\_i = -9.53 pericentre = 29.64 apocentre = 74.88 t last peri = -0.80
+x\_i = {[}-16.48 69.92 21.05{]} v\_i = {[}16.32 39.86 -116.99{]}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
+\caption[Ursa Minor simulation snapshots]{Ursa Minor simulation images.}
+\end{figure}
+
+Figure: Velocity dispersion evolution of Ursa Minor
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_smallperi_i_f.pdf}}
+\caption[Ursa Minor simulated density profiles]{The tidal effects on the
+stellar surface density.}
+\end{figure}
+
+\subsection{Effects of the LMC}\label{effects-of-the-lmc-1}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_lmc_xyzr_orbits.pdf}}
+\caption[Ursa Minor orbits with LMC]{Orbits of Ursa Minor with (orange)
+and without (green) an LMC. The final positions of Ursa Minor and the
+LMC are plotted as scatter points and the solid blue line represents the
+LMC trajectory. Note that the LMC only increases Ursa Minor's peri and
+apo-centres, weakening any tidal effect. Interestingly, there is a
+change that Ursa Minor may have once been bound to the LMC (diverging
+orange lines at top left of middle panel.)}
+\end{figure}

### 2025-08-18 notebooks/ludlow.jl

@@ -84,7 +84,7 @@ md"""
 """
 
 # ╔═╡ 423f83f7-3530-4689-9aa9-0fbc50df0e54
-nfw = NFW(M200 = 1e5, c=17)
+nfw = NFW(M200 = 1e5, c=5)
 
 # ╔═╡ 6fa3783a-14b3-4943-a5d8-a04a48effd43
 scale_density(nfw), scale_point_density(nfw)
@@ -460,7 +460,7 @@ end
 let
 	fig = Figure()
 	ax = Axis(fig[1,1], 
-			  xlabel = "log radius / pkpc",
+			  xlabel = "log radius / kpc",
 			  ylabel = "log density"
 			 )
 
@@ -481,6 +481,42 @@ let
 		lines!(x, y, color=z, colorrange=extrema(zs))
 		#hlines!(log10(200ρ_c(z)), color=z, colorrange=(extrema(zs)))
 		scatter!(log10(h.r_s), log10(LilGuys.density(h, h.r_s)), color=z, colorrange=extrema(zs))
+
+	end
+
+	Colorbar(fig[1,2], colorrange=(extrema(zs)))
+
+	fig
+end
+
+# ╔═╡ 990c126b-83b4-4f3f-a2d6-d775d2174f1c
+let
+	fig = Figure()
+	ax = Axis(fig[1,1], 
+			  xlabel = "log radius / R200",
+			  ylabel = "log density"
+			 )
+
+
+
+
+	zs = 0:7
+
+	for z in zs
+		h = halo_at_z(z)
+
+		x = LinRange(-2, log10(R200(h, z)), 1000) 
+		r = 10 .^ x
+		
+		v = LilGuys.density.(h, r)
+
+		y = log10.(v)
+
+		scale = R200(h)
+		
+		lines!(x .- log10(scale), y .+ 3log10(scale), color=z, colorrange=extrema(zs))
+		#hlines!(log10(200ρ_c(z)), color=z, colorrange=(extrema(zs)))
+		#
 
 	end
 
@@ -649,6 +685,7 @@ end
 # ╠═0963eb03-2c6d-45b9-b34d-344a0bc5b545
 # ╠═e1dc5fee-600a-4a81-87cd-9db46ca223e0
 # ╠═faa9c7d6-2d5e-45d7-8cb0-3e47a025f080
+# ╠═990c126b-83b4-4f3f-a2d6-d775d2174f1c
[…]

### 2025-08-18 thesis/compile.sh

@@ -1,5 +1,4 @@
 set -e 
-rsync ../dwarfs_figures/thesis/plots/figures/ figures -av
 
 # pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections --filter equation.py"
 pandocoptions="--defaults=pandoc_params.yaml"

### 2025-08-18 thesis/figures/sync.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+rsync -av cosma:~/thesis/plots/figures/* .

### 2025-08-18 thesis/sync.sh

@@ -1,2 +1,1 @@
-rsync -av ~/thesis_2/ .
-rsync -av ~/dwarfs_figures/thesis/plots/figures ~/thesis_2/
+rsync -av ~/thesis/ .

### 2025-08-23 thesis/appendix.tex
thesis julio ch1 edits
thesis refining data section

@@
 the sum of the satellite and background likelihoods, weighted by their
-relative frequencies, \begin{equation}{
+relative frequencies,
+\begin{equation}\protect\phantomsection\label{eq:Ltot}{
 {\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis_2/figures/analytic_profile_comparison.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/analytic_profile_comparison.pdf}}
 \caption[Comparison of analytic density profiles]{A comparison of

### 2025-08-23 thesis/introduction.tex
thesis refining data section

@@
 stellar components of Sculptor like halos (see chapter
-{[}@-sec:methods{]}). To address this challenge, idealized simulations
+\ref{sec:methods}). To address this challenge, idealized simulations
 only simulate a single subhalo in an approximate host potential. As a
@@
 
-In chapter {[}@-sec:observations{]}, we describe how we compute
+In chapter \ref{sec:observations}, we describe how we compute
 observational density profiles from \citet{jensen+2024}. In chapter
-{[}@-sec:methods{]}, we review our simulation methods. Next, we present
+\ref{sec:methods}, we review our simulation methods. Next, we present
 our results for the tidal effects on Sculptor and Ursa Minor in chapter
-{[}@-sec:results{]}, We discuss our results, limitations, and
-implications in chapter {[}@-sec:discussion{]}. Finally, chapter
-{[}@-sec:summary{]} summarizes this work and discuss future directions
-for similar work and the field of dwarf galaxies.
+\ref{sec:results}, We discuss our results, limitations, and implications
+in chapter \ref{sec:discussion}. Finally, chapter \ref{sec:summary}
+summarizes this work and discuss future directions for similar work and
+the field of dwarf galaxies.

### 2025-08-28 thesis/data.tex

@@
-To provide an empirical benchmark for simulations in the next chapters,
-we aim to understand the observed properties and density profiles of
-Sculptor and Ursa Minor. In this section, we first compile past
-measurements of each galaxy. We then describe the \citet{jensen+2024}
-Bayesian methodology to select high probability member stars from
-\emph{Gaia}. We show the derived density profiles are robust to changes
-in assumptions. While Fornax is well represented by a 2D exponential
-profile, Sculptor and Ursa Minor show an excess of stars relative to an
-2D exponential in the outer regions, better described by a Plummer
-profile. Future chapters assess possible explanations for these excess.
+An ongoing research question is whether local dwarf spheroidal galaxies
+contain substructure such as stellar halos. Some systems appear to be
+simple in structure, yet recent work has reported detection of stellar
+halos in several classical dwarfs. To provide an empirical benchmark for
+simulations in the next chapters, we examine the observed stellar
+structure of Sculptor and Ursa Minor. We describe the
+\citet{jensen+2024} Bayesian methodology to select high-probability
+member stars from \emph{Gaia}. The derived density profiles are robust
+to alternative selection criteria. While Fornax is well represented by
+an exponential density profile, we find that Sculptor and Ursa Minor
+exhibit an excess of stars in their outer regions relative to an
+exponential, with a Plummer profile providing a better fit. In following
+chapters, we consider possible physical explanations for these extended
+profiles.
 
-\section{\texorpdfstring{\emph{Gaia} data}{Gaia data}}\label{gaia-data}
+\section{\texorpdfstring{\emph{Gaia} sample
+selection}{Gaia sample selection}}\label{gaia-sample-selection}
 
-Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
-membership estimation method. J+24 use a Bayesian framework
-incorporating proper motion (PM), colour-magnitude diagram (CMD), and
-spatial information to determine the probability that a given star
-belongs to the satellite. By accounting for PM in particular, J+24
-produces low contamination samples of candidate member stars out to
-large distances from a dwarf galaxy. J+24 extends the algorithm
-presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
-an optional secondary, extended spatial component to find possible
-members as far as \textasciitilde10 half-light radii \(R_h\) from
-several dwarf galaxies. See also similar work by \citet{pace+li2019};
+The detection of faint features in resolved stellar systems requires
+careful consideration of whether a given star belongs to the system.
+Without removing contamination from foreground/background features,
+faint features may be lost in the noise or be of dubious association
+with the system. Historically, background discrimination was performed
+by the colour-magnitude diagram alone. Now with \emph{Gaia} data,
+stellar parallax and proper motion areavailable as additional membership
+indicators. Here, we use the \citet{jensen+2024}'s (hereafter J+24)
+membership probabilities for \emph{Gaia} observations. J+24 use a
+Bayesian framework incorporating proper motion (PM), colour-magnitude
+diagram (CMD), and spatial information to determine the probability that
+a given star belongs to the satellite or background. By accounting for
+PM in particular, J+24 produces low contamination samples of candidate
+member stars out to large distances from a dwarf galaxy. J+24 extends
+the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
+additionally includes a secondary, extended spatial component. J+24 are
+able to find candidate members out to \textasciitilde10 half-light radii
+(\(R_h\)). Similar recent work has also included \citet{pace+li2019};
 \citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.
 
-J+24 define likelihoods \({\cal L}\) representing the probability
-density that a star is consistent with either the MW stellar background
-(\({\cal L}_{\rm bg}\)) or the satellite galaxy
-(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
-product of a spatial, PM, and CMD term: \begin{equation}{
+J+24 construct likelihoods, \({\cal L}\), representing the probability
+density that a star is consistent with either the stellar
+foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
+\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
+of a spatial, PM, and CMD components, \begin{equation}{
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
@@
 
-Each likelihood is normalized over their respective 2D parameter space
-for both the satellite. To control the relative frequency of member and
-background stars, \(f_{\rm sat}\) representing the fraction of member
-stars in the field. The total likelihood for any star in this model is
-the sum of the satellite and background likelihoods, weighted by their
-relative frequencies, \begin{equation}{
+The spatial likelihood is specified either as a 1 or 2 component
+elliptical exponential density profile (with the 2 component used when
+shown to be detected). The proper motion likelihood specifies if a
+star's proper motion is consistent with the dwarf galaxy's systemic
+motion, accounting for observational uncertainties. Finally, the CMD
+likelihood is based on the consistency of a star with theoretical
+isochrones in \emph{Gaia} \(G\) versus \(G_{\rm BP}- G_{\rm RP}\) for
+the galaxy. The background likelihood is a uniform distribution over the
+field, and the background CMD and PM likelihoods are constructed
+empirically from the stars in an annulus surrounding the dwarf galaxy.
+Each likelihood is normalized over the respective parameter space.
+
+The total likelihood is the sum of the background and satellite
+likelihoods, weighted by the fraction of stars in the field belonging to
+the satellite \(f_{\rm sat}\): \begin{equation}{
 {\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
@@
 
-\section{Selected stars}\label{selected-stars}
-
-In
-Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
-we illustrate the resulting samples from the algorithm in the tangent
-plane, CMD, and proper motion space. The tangent plane coordinates
-\(\xi\) and \(\eta\) are the distances in RA and declination as measured
-on a plane tangent to the dwarf galaxies centre. \(R_{\rm ell}\) is
-defined as the circularized elliptical radius, so
-\(R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2)\) where
-\(\xi'\) and \(\eta'\) are rotated to align with the position angle
-direction \(\theta\).
+J+24 run MCMC simulations to solve for the proper motions,
+\(f_{\rm sat}\), and structural properties of the second exponential
+density profile (if included). The median parameter are then used to
+calculate \(P_{\rm sat}\) which we use to select our samples of likely
+member stars.
 
@@
 magnitude.
+
+\section{Observed stellar
+distributions}\label{observed-stellar-distributions}
+
+In
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
+we illustrate the resulting samples from the algorithm in the tangent
+plane, CMD, and proper motion space. The tangent plane coordinates
+\(\xi\) and \(\eta\) are the distances in RA and declination as measured
+on a plane tangent to the dwarf galaxies centre. \(R_{\rm ell}\) is
+defined as the circularized elliptical radius, so
+\(R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2)\) where
+\(\xi'\) and \(\eta'\) are rotated to align with the position angle
+direction \(\theta\).
 
@@
 unlikely that there are additional faint, tidal features detectable with
-\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
-cuts and thresholds, finding no evidence of other structure in both
-\emph{Gaia} and DELVE/UNIONS data.
+\emph{Gaia}.
 
-Finally, we illustrate the location of RV-confirmed members from
-Section~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
-brighter than a typical \emph{Gaia} candidate, RV members typically have
-more precise PMs. However, the RV members fill about the same area on
-the CMD down to the magnitude limit. Additionally, while the number of
-stars with RV measurements decreases at large radii, there are still
-confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
-stars illustrate the extended profiles of each galaxy. Note that for an
-exponential profile, 99.95\% of stars should be within \(6R_h\), so the
-discovery of any stars beyond \(6R_h\) hints at a deviation from an
-exponential profile. These \emph{extratidal} stars hint at something
-tidal in nature, and our goal is to explore possible interpretations.
+Finally, we illustrate the location of distant, RV-confirmed members
+from Section~\ref{sec:rv_obs} in each panel. These stars illustrate the
+extended profiles of each galaxy. Note that for an exponential profile,
+99.95\% of stars should be within \(6R_h\), so the discovery of any
+stars beyond \(6R_h\) hints at a deviation from an exponential profile.
+Distant member stars beg for an explanation as to their origin.
 

### 2025-08-28 thesis/sculptor.tex

@@
 
-Before running N-body simulations, we need to understand both the
-typical halos which could host a sculptor like galaxy and the possible
-orbits Sculptor.
-
-Fig.~\ref{fig:scl_halos} presents approximate constraints based on
-
 \subsubsection{Derived properties and
 setup}\label{derived-properties-and-setup}
+
+From the observed properties of Sculptor, we can infer reasonable
+\(\Lambda\)CMD dark matter halo hosts. Table~\ref{tbl:scl_derived_props}
+reports our inferred halo and kinematic properties of Sculptor. First,
+taking the absolute magnitude from \citet{munoz+2018} with the
+mass-to-light ratio from \citet{woo+courteau+dekel2008} (1.7 with
+\(\sim\) 0.17 dex uncertainty), the total current stellar mass of
+Sculptor is \(M_\star \sim 3.1 \times 10^6 \Mo\). Based on the stellar
+mass-\(v_{\rm max}\) relation \citep[from][]{fattahi+2018}, see also
+\ref{fig:SMHM}, the halo should have
+\(v_{\rm max} \sim 31 \pm 3 \,\kms\). Finally, using the
+\citet{ludlow+2016} \(z=0\) mass-concentration relation, this constraint
+translates into a \(r_{\rm max} \sim 6 \pm 2\,{\rm kpc}\), or
+alternatively \(M_{200} = 5\times10^9\) with \(c=13\).
+
+Fig.~\ref{fig:scl_halos} illustrates these estimates visually. The
+stellar-mass \(v_{\rm max}\) constraint translates to an estimate of
+\(v_{\rm max}\) only (horizontal band). The mass-concentration relation
+describes the relationship between \(v_{\rm max}\) and \(r_{\rm max}\),
+an approximately diagonal linear band. And finally, we include a curved
+line which illustrates where the initial velocity dispersion is
+approximantly larger than the present-day observed dispersion
+\(v_{\rm circ}(R_h) / \sqrt{3} \sim \sigma_v \gtrsim 9\,\kms\).
+
+The intersection of the constraints on the halo leads us to select an
+initial halo, \texttt{compact}, which is more compact than the mean but
+with a observationally consistent initial velocity dispersion. We also
+have the \texttt{light} halo which occupies the intersection between
+3\(\sigma\) lowest mass and the velocity dispersion constraints, which
+we come back to in Section~\ref{sec:scl_lmc}.
+
+While there is some range in the choice of initial halo, reasonable
+changes to the initial halo do not substantially affect the tidal
+evolution for Sculptor. The velocity dispersion, in particular,
+translates to an estimate of the mass within \(r_h\). As a result, halos
+with the same velocity dispersion only differ in total mass, but the
+density slope and circular velocity near the stellar component is
+similar. These halos will have a similarly affected stellar component.
 
@@
 $M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
+$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
+$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
 $M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
 $c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
-$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
-$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
 $r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
@@
 
-For the mass-to-light ratio, we use the values from
-\citet{woo+courteau+dekel2008}. Note that we assume a scatter of 0.17
-dex for the mass-to-light ratios as reported for mass uncertainties.
-\citet{delosreyes+2024} also indicate that other methods for stellar
-mass-to-light ratios result in similar scatters and systematic biases.
-However these values are just approximate.
-
 \begin{figure}
@@
 \caption[Sculptor initial halos]{The suggested halos of Sculptor
-compared to cosmological predictions.}\label{fig:scl_halos}
+compared to cosmological predictions. \textbf{TODO: simplify
+\(\sigma_v\) calculation}.}\label{fig:scl_halos}
 \end{figure}
@@
 
+Next, we calculate the range of possible orbits of Sculptor in our
+fiducial Milky Way potential. Fig.~\ref{fig:scl_orbits} illustrates
+point particle orbits for 100 samples of Sculptor's observed kinematics
+integrated backwards 5 Gyr in both galactocentric coordinate slices (x,
+y, z) and in galactocentric radius with time. In terms of the x, y, z
+planes, note that all sampled orbits of Sculptor have nearly the same
+morphology---they each orbit mostly in the y-z plane, passing through
+the same number of pericentres and maintaining similar orbits across
+5Gyr. Sculptor's orbital history is relatively well-constrained.
+
+We select an orbit with the \(\sim 3\sigma\) smallest pericentre among
+all possible orbits. We achieve this by taking the median parameters of
+all orbits with a pericentre less than the 0.0027th quantile pericentre,
+yielding a pericentre of 43 kpc.\footnote{only slightly smaller than the
+  0.00135th quantile pericentre} It is unlikely given the current
+observations that Sculptor has a smaller pericentre. Since tidal effects
+depend most strongly on the pericentre of an orbit, this orbit should
+maximize the tidal force from the Milky Way.
+
+We take the first apocentre after a lookback time of 10 Gyr, or at 9.43
+Gyr, as the initial conditions for our model of Sculptor.
+Table~\ref{tbl:scl_orbits} notes the initial conditions for our
+simulation.
+
 \begin{figure}
@@
 sculptor is well-constrained in this potential and it is unlikely to
-achieve a smaller pericentre than our selected orbit.
-\the\textwidth}\label{fig:scl_orbits}
+achieve a smaller pericentre than our selected
+orbit.}\label{fig:scl_orbits}
 \end{figure}
-
-To select an orbit with \textasciitilde{} maximum possible
-observationally-consistent tidal forces, we take the median parameters
-for orbits with a pericentre less than 2x the 3\(\sigma\) minimum
-pericentre and derive the following orbit. The initial conditions are
-from the
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  dark matter loss
-\item
-  little influence to stars
-\item
-  smallest pericentre does not help
-\end{itemize}
+The Milky Way tides indeed affect Sculptor's dark matter halo.
+Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
+of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
+previous pericentres. This simulation uses the \texttt{compact} initial
+halo on the \texttt{smallperi} orbit. The dark matter forms large
+streams which orbit the Milky Way several times. However, note that the
+central dark matter cusp is similar in appearance across all simulation
+snapshots.
+
+More quantitatively, the halo loses about 99\% of its initial dark
+matter mass. This corresponds in a reduction from the initial
+\(v_{\rm max} = 31\,\kms\) to \(v_{\rm max} = 22\,\kms\). However, the
+stellar component of Sculptor is relatively unaffected.
+
+While mass loss leads to adiabatic expansion of the stellar component,
+no features of tidal disruption or disequilibrium are currently
+observable. Fig.~\ref{fig:scl_smallperi_i_f} shows the final projected
+density of stars on the sky, and the initial and final radially averaged
+density profiles. No non-spherical density features are apparent, even
+at 5 decades fainter in surface density than the central density. In
+addition, the initial and final density profiles look identical up to
+some scale, with a tidally-induced excess of stars only appearing
+\(\sim 3\) orders of magnitude fainter than the faintest our density
+profile measures.
+
+The analytically motivated break and tidal radii corroborate a weak
+tidal effect on stars. In both cases, these radii work out to be
+\(\gtrsim 100\) arcminutes, outside where we can measure the density
+profile. In particular, because we have chosen the most extreme
+observationally permissible orbit, it is unlikely, for similar
+\(\Lambda\)CDM motivated initial conditions, that these radii would ever
+approach where the break in Sculptor's density profile appears.
 
@@
 spanning 5 orders of magnitude between the maximum and minimum values.
-In this image, stars occupy only ever a few pixels so are not plotted.}
+In this image, stars occupy only ever a few pixels so are not
+plotted.}\label{fig:scl_sim_images}
 \end{figure}
@@
 \caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
-orbit. Todo: add velocity dispersion plot to RHS}
+orbit. Todo: add velocity dispersion plot to
+RHS}\label{fig:scl_tidal_track}
 \end{figure}
@@
 observed velocity dispersion by \textasciitilde{} 1 km/s, so the
-conservative lower limit is around 8 km/s.}
+conservative lower limit is around 8 km/s. \textbf{TODO}, add bound mass
+with time. Maybe combine with tidal tracks and radius / time orbit.}
 \end{figure}
@@
 \caption[Sculptor initial and final density profiles]{Effects on
-exponential initial stars. TODO: plot 2D sky proj. stars}
+exponential initial stars. TODO: plot 2D sky proj.
+stars}\label{fig:scl_smallperi_i_f}
 \end{figure}
@@
 \caption[Sculptor Plummer initial and final density profiles]{effects on
-Plummer initial stars.}
+Plummer initial stars.}\label{fig:scl_smallperi_plummer_i_f}
 \end{figure}
@@
 
+The Milky Way isn't the only galaxy in town. Becoming more apparent, the
+infall of the LMC substantially affects the Milky Way system. The mass
+of the LMC may be as high as 1/5 of the Milky Way mass, so including an
+LMC in the Milky Way potential may change conclusions about properties
+and orbits of the Milky Way satellite system.
+
 \subsubsection{Orbital effects}\label{orbital-effects}
@@
 As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
-influenced by the presence of an LMC. Figure
-Fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
-Scl's pericentre with the MW and implies that Scl may be on its first
-infall with the MW, in contrast to the discussion above. Thus, the LMC
-has a critical impact on the evolution of Scl. Additionally, Scl passes
-rapidly but relatively close to the LMC in the past 100 Myr.
+influenced by the presence of an LMC.
 
-Finally, we also consider the influence of the large Milky Cloud on the
-orbits and evolution of Scl. We adopt the \citet{vasiliev2024} multipole
-approximation of an N-body simulation of the LMC and MW. Their initial
-conditions are
+To explore the effects of the LMC on Sculptor, we use the publically
+available L3M11 model from \citet{vasiliev2024}. This model stars with a
+lighter Milky Way than our fiducial \citet{EP2020}, \citet{mcmillan2011}
+like model. The L3M11 potential is an evolving multipole approximation
+of an N-body simulation with both a live MW and LMC dark matter halo.
+The MW intital conditions were an NFW with \(r_s=16.5\,\)kpc and mass
+\(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
+\(r_s=11.7\) and \(M_{100} = 2.76 \times 10^{11} \Mo\). This model had a
+previous LMC pericentre at about 6 Gyr ago.
 
-\begin{itemize}
-\tightlist
-\item
-  MW halo:
-\item
-  MW bulge (static):
-\item
-  MW disk (static):
-\item
-  LMC halo:
-\end{itemize}
+Fig.~\ref{fig:scl_lmc_orbit_effect} displays the effect of including an
+LMC in the potential. The green samples are in the initial MW-only
+potential in the \texttt{L3M11} model, and the orange samples are
+integrated in the evolving MW and LMC \texttt{L3M11} model. The past 1
+Gyr is similar in both cases, but the orbits diverge significantly
+afterwards. The recent passage of Sculptor with the LMC around 0.1 Gyr
+ago allows for Sculptor to begin as far as 300 kpc from the Milky Way
+centre. The evolving potential also adds significant long term
+variability in the possible orbits of Sculptor. Sculptor, however, is
+orbiting in the opposite direction of the LMC so is likely not
+associated with the LMC system.
 
-We now consider 3 orbits under the L3M11 potential model from
-\citet{vasiliev2024}. We focus on this LMC potential as a lighter LMC or
-MW should only reduce tidal impacts and the recent orbit of Scl is
-minimally affected by the LMC potential.
+Given the large uncertainties of the LMC model, we conservatively double
+all of the observational parameters of Sculptor. This has a similar
+effect to including LMC mass and orbital uncertainties but is
+considerably simpler. From this larger range of orbits, we once again
+select the orbit with the median final observables of all orbits with
+pericentres less than the 0.0027th quantile. This orbit, the
+\texttt{smallperilmc} orbit is plotted in black and is only integrated
+up to 2 Gyr ago, to isolate recent tidal effects.
 
@@
 
-Smallperi
+\subsubsection{Tidal effects}\label{tidal-effects-1}
 
-\begin{itemize}
-\item
-  ra = 15.0183 dec = -33.7186 distance = 84.0 pmra = 0.166 pmdec =
-  -0.237 radial\_velocity = 111.2
+Unlike the previous mdoel, this model only has one pericentre with the
+Milky Way. Fig.~\ref{fig:scl_lmc_sim_images} shows snapshots of Sculptor
+over the past 2 Gyr while marking the position of the LMC. With only one
+pericentre, and a larger one than the \texttt{smallperi} orbit,
+Sculptor's dark matter is substantially less disrupted. And while, based
+on the tidal tensor values, the LMC induces a greater instantaneous
+tidal effect than the Milky Way, Sculptor's dark matter component does
+not show strong effects due to the LMC.
 
-  t\_i = -2.00 pericentre = 28.00 apocentre = 190.45 t last peri = -0.39
-  x\_i = {[}12.79 168.76 87.33{]} v\_i = {[}0.37 -57.69 61.40{]}
-\end{itemize}
+Finally, Fig.~\ref{fig:scl_lmc_i_f} shows the projected on-sky final
+stellar distribution and the initial and final stellar density profile
+for this model with exponential stars. The stellar component does not
+change at all. While the break radius set by the time since the last LMC
+pericentre would agree with the location of the break in the observed
+density profile, no stellar effect would be observable.
 
-\subsubsection{Tidal effects}\label{tidal-effects-1}
+The LMC flyby encounter is an approximately impulsive encounter, in
+contrast with more adiabadic mass loss due to the Milky Way. Impulsive
+encounters tend to inject energy into the stellar and dark matter
+distribution, and can initially cause the galaxy to contract. In
+addition, the tidal force is required to be far larger than for slower,
+adiabadic mass loss, because the galaxy experiences this tidal field for
+far less time. Even in models where Sculptor passes through the the LMC,
+tidal tails do not form immediately after this encounter.
+
+A final note is that this model only considers the recent tidal effects
+.As illustrated in the range of possible orbits in
+Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
+experienced an extremely small pericentre with the Milky Way. This
+pericentre has the potential to substantially rearrange the stellar
+component, drive large tidal mass loss, and create a Plummer-like
+stellar density profile. However, this encounter is highly dependent on
+the choice of the MW-LMC potential model, and may not occur at all. Long
+time integration in dynamic potentials amplifies uncertainties in the
+inputs and may not be reliable. \textbf{TODO: I would love to get this
+model to work\ldots{}}
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.pdf}}
-\caption{Sculptor Simulation Snapshots with LMC}
+\caption{Sculptor Simulation Snapshots with
+LMC}\label{fig:scl_lmc_sim_images}
 \end{figure}

### 2025-08-29 analysis/sculptor/1e6_new_v31_r3.2/L3M11_9Gyr_smallperi.3b/paths.sh
durham 2025 seminar presentation

@@ -1,0 +1,3 @@
+source ../paths.sh
+
+export out_path="$DWARFS_SIMS/sculptor/1e6_new_v31_r3.2/L3M11_9Gyr_smallperi.3b/"

### 2025-08-29 notebooks/sim_analy/analyze_orbit.jl

@@ -708,7 +708,9 @@ let
 	
 	orbital_properties = OrderedDict(
 		"pericentre" => r_peri,
-		"apocentre" => r[idx_apos[idx_apos .< idx_f]],
+		"apocentre" => r[last(idx_apos[idx_apos .< idx_f])],
+		"pericentres" => r[idx_peris],
+		"apocentres" => r[idx_apos],
 		"period" => period,
 		"idx_f" => idx_f,
 		"idx_peri" => idx_peri,

### 2025-08-29 presentations/durham25/classical_dwarf_densities.jl

@@ -16,11 +16,17 @@ begin
 
 end
 
+# ╔═╡ 8aab441c-3268-449d-b401-94f8e15eb5f3
+using CSV, DataFrames
+
 # ╔═╡ cf655228-1528-4e77-9629-07e99984951f
 using OrderedCollections
 
 # ╔═╡ 6d2bff6d-b347-49c4-87df-ad58d8a27ff3
 include("./style.jl")
+
+# ╔═╡ b47ea4ad-9f71-44a5-abc0-df6a748d5247
+CairoMakie.activate!(type=:png)
 
 # ╔═╡ 67dd3488-79e1-4ba2-b9ac-f417765d55de
 import TOML
@@ -113,6 +119,9 @@ galaxynames = [
 	"crater2",
 ]
 
+# ╔═╡ 4db3ed1d-ef02-4f28-8c9b-b6c6d491199f
+[print(galaxy, ", ") for galaxy in galaxynames]
+
 # ╔═╡ f4e8b66c-5f18-45fd-8859-32479d7227bc
 begin 
 	profiles = OrderedDict(name => load_profile(name) for name in galaxynames)
@@ -123,14 +132,48 @@ end
 # ╔═╡ d34dfbe4-c80a-46e8-9082-30148cbcbd0b
 profiles["sculptor"]
 
+# ╔═╡ ae186cf2-6fd5-4958-acc0-8a87eab45358
+p08_notides = CSV.read(joinpath(ENV["DWARFS_ROOT"], "observations/all/data", "penarrubia08_notides.csv"), DataFrame)
+
+# ╔═╡ 180d22b5-21f0-4559-9696-03887675d9f9
+p08_r_scale = 0.7
+
+# ╔═╡ fe81daf4-d2a9-484f-8ba4-9d4f7348d3fd
+p08_m_scale = 5 * p08_r_scale
+
+# ╔═╡ eb85be07-d72e-4d22-8cef-a8521d465a4c
+p08_tides = CSV.read(joinpath(ENV["DWARFS_ROOT"], "observations/all/data", "penarrubia08_apo.csv"), DataFrame)
+
+# ╔═╡ 5eb61a4e-063f-4c11-8dea-07be62cee089
+begin 
+	prof_i = LilGuys.SurfaceDensityProfile(
+		log_R=p08_notides.x,
+		log_Sigma=p08_notides." y",
+		R_units="", 
+		log_R_bins=[]
+	) 
+	
[…]

### 2025-08-29 presentations/durham25/initial_conditions.jl

@@ -29,7 +29,7 @@ stars_dir = joinpath(ENV["DWARFS_ROOT"], "analysis", "sculptor/1e7_new_v31_r3.2/
 snap = Snapshot(joinpath(stars_dir, "iso_paint.hdf5"))
 
 # ╔═╡ a3a3a875-d0cc-4594-86c2-b242e859efef
-df_probs = LilGuys.read_hdf5_table(joinpath(stars_dir, "exp2d_rs0.08", "probabilities_stars.hdf5"))
+df_probs = LilGuys.read_hdf5_table(joinpath(stars_dir, "exp2d_rs0.10", "probabilities_stars.hdf5"))
 
 # ╔═╡ 01e34341-e806-44ef-9591-d226b345dec6
 snap.weights = df_probs.probability[snap.index]
@@ -38,16 +38,22 @@ snap.weights = df_probs.probability[snap.index]
 prof_dm = DensityProfile(snap, bins=-2:0.1:2)
 
 # ╔═╡ 3edb5998-a200-4e94-9598-5b73c29def33
-sum(densities(prof_dm) .* diff(π*LilGuys.radius_bins(prof_dm) .^3)) 
+sum(densities(prof_dm) .* diff(4π/3*LilGuys.radius_bins(prof_dm) .^3)) 
+
+# ╔═╡ d76e4d60-f37a-4f27-8183-d5b2b3bda78d
+sum(snap.masses)
 
 # ╔═╡ 5743880e-567f-4af5-acb5-e2ec420110aa
-Mstar = 3e6 / M2MSUN
+Mstar =  3e6 / M2MSUN
+
+# ╔═╡ c635c9e5-582d-48f1-ae15-9e842e21cbac
+sum(snap.weights * Mstar)
 
 # ╔═╡ 098e33d5-22a5-496b-943b-b8c194cdcc62
 prof_stars = DensityProfile(snap, snap.weights * Mstar)
 
 # ╔═╡ f6d51dd1-91df-4a6b-85a8-cd9b26d2c9c9
-sum(densities(prof_stars) .* diff(π*LilGuys.radius_bins(prof_stars) .^3)) 
+sum(densities(prof_stars) .* diff(4π/3*LilGuys.radius_bins(prof_stars) .^3)) 
 
 # ╔═╡ 546e88aa-c3c5-41b6-84a7-ce733e8f309f
 COLORS
@@ -59,13 +65,15 @@ COLORS
[…]

### 2025-08-29 presentations/durham25/scl_density_i_f.jl

@@ -185,7 +185,7 @@ end
 @savefig "scl_smallperi_i_f" compare_profiles("1e7_new_v31_r3.2", "orbit_smallperi", "exp2d_rs0.10", norm_shift=0.15, break_height=-1.0, r_j=true)
 
 # ╔═╡ f053b7f9-f934-4cd4-ae46-32a0a373ff7e
-# @savefig "scl_mean_i_f" compare_profiles("1e7_V31_r3.2", "orbit_mean", "exp2d_rs0.10", norm_shift=0.15, break_height=-5, r_j=get_r_j("1e7_V31_r3.2", "orbit_mean"))
+
 
 # ╔═╡ 4f02015a-dfc0-4c7c-935d-9992b0690576
 @savefig "scl_plummer_i" compare_profiles("1e7_new_v31_r3.2", "orbit_smallperi", "plummer_rs0.20", norm_shift=0.0, plot_final=false, break_height=nothing)
@@ -197,8 +197,13 @@ end
 @savefig "scl_lmc_i_f" compare_profiles("1e7_new_v25_r2.5", "smallperilmc", "exp2d_rs0.13", 
 	norm_shift=0.1, break_height=-6, lmc=true, r_j=true)
 
+# ╔═╡ c3ae7b58-68b5-49ed-a874-611debd890bc
+@savefig "scl_lmc_smallperi_i_f" compare_profiles("1e6_new_v31_r3.2", "L3M11_9Gyr_smallperi.3b", "exp2d_rs0.13", 
+	norm_shift=-0.1, break_height=-6)
+
 # ╔═╡ fab0c970-7d9a-4ed6-9632-72a7b0d52c5b
-LilGuys.Plummer(r_s=0.3)
+ compare_profiles("1e6_new_v31_r3.2", "L3M11_9Gyr_smallperi", "exp2d_rs0.13", 
+	norm_shift=0.2, break_height=-6)
 
 # ╔═╡ 084e8e1d-daec-4859-bbe2-888d484d6123
 @savefig "scl_lmc_i_f_mw_break_radii" compare_profiles("1e7_new_v25_r2.5", "smallperilmc", "exp2d_rs0.13", 
@@ -228,5 +233,6 @@ LilGuys.Plummer(r_s=0.3)
 # ╠═4f02015a-dfc0-4c7c-935d-9992b0690576
[…]

### 2025-08-29 presentations/durham25/scl_lmc_orbits.jl

@@ -155,8 +155,8 @@ lw_thick = @lift 2*$(theme(:linewidth))
 # ╔═╡ ff89d7db-c954-48b8-b87a-446ccfb2d79b
 trajectories = OrderedDict(
 
-	"no lmc" => orbits_no_lmc,
-	 "lmc" => orbits_w_lmc,
+	"Scl: MW only" => orbits_no_lmc,
+	 "Scl: MW & LMC" => orbits_w_lmc,
 )
 
 # ╔═╡ 92fb5b1f-5dc6-4766-b69d-f8644861d4cd
@@ -168,12 +168,7 @@ md"""
 orbits_m_lmc = orbits_w_lmc .- [orbit_lmc]
 
 # ╔═╡ 081e6a08-92e3-4f9d-bc8e-7ff80ca74693
-trajectories_radii = OrderedDict(
-	"no lmc" => orbits_no_lmc,
-	 "lmc" => orbits_w_lmc,
-	"scl - lmc" => orbits_m_lmc
 
-)
 
 # ╔═╡ 6f1a1958-708a-4aac-935d-0e5a87091a98
 function plot_traj_lmc!(; x_direction = 2, t_min=-5/T2GYR)
@@ -261,10 +256,10 @@ function plot_orbits_w_lmc(trajectories; lmc_visible=false, act_visible=false)
 end
 
 # ╔═╡ ed9ab01f-1906-4a2a-8f5d-6559f30a1c69
-@savefig "scl_nonolmc_mc_orbits_xy" plot_orbits_w_lmc(Dict("MW only" => orbits_no_lmc), lmc_visible=false)
+@savefig "scl_nonolmc_mc_orbits_xy" plot_orbits_w_lmc(Dict("Scl: MW only" => orbits_no_lmc), lmc_visible=false)
 
 # ╔═╡ 05af926b-9af5-4fc1-a2b0-0fa1a0502ed9
-@savefig "scl_nolmc_mc_orbits_xy" plot_orbits_w_lmc(Dict("MW only" => orbits_no_lmc), lmc_visible=true)
+@savefig "scl_nolmc_mc_orbits_xy" plot_orbits_w_lmc(Dict("Scl: MW only" => orbits_no_lmc), lmc_visible=true)
 
 # ╔═╡ 2683e73a-e9d9-4f79-a101-8f306d36e428
 @savefig "scl_lmc_mc_orbits_xy" plot_orbits_w_lmc(trajectories, lmc_visible=true)

### 2025-08-29 presentations/durham25/umi_lmc_orbits.jl

@@ -45,10 +45,10 @@ end
 galaxyname = "ursa_minor"
 
 # ╔═╡ 432dc7d3-c390-48cb-9e51-d4d0ff21620c
-Nmax = 1000
+Nmax = 100
 
 # ╔═╡ 5eaf3b50-886e-47ac-9a7c-80d693bc3c17
-CairoMakie.activate!(type=:png, px_per_unit=4)
+CairoMakie.activate!(type=:png)
 
 # ╔═╡ b9b99de2-1c01-4562-8295-75b770e232dc
 md"""
@@ -69,7 +69,7 @@ function read_traj(modelname)
 end
 
 # ╔═╡ 70d81f4a-8e01-45b1-a695-234d0249cca9
-function plot_x_y_traj!(orbits; thin=1, x_direction=2, y_direction=3, alpha=0.01, color=:black, t_min = -2/T2GYR, kwargs...)
+function plot_x_y_traj!(orbits; thin=1, x_direction=2, y_direction=3, alpha=0.1, color=:black, t_min = -2/T2GYR, kwargs...)
 
     for i in 1:thin:length(orbits)
 		filt = orbits[i].times .> t_min 

### 2025-08-29 presentations/durham25/v_circ_potential.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.8
+# v0.20.15
 
 using Markdown
 using InteractiveUtils

### 2025-08-29 thesis/plots/initial_conditions.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.13
+# v0.20.15
 
 using Markdown
 using InteractiveUtils
@@ -23,13 +23,13 @@ include("./paper_style.jl")
 CairoMakie.activate!(type=:png)
 
 # ╔═╡ caf046d9-e475-4810-844e-b65e9c08c832
-stars_dir = joinpath(ENV["DWARFS_ROOT"], "analysis", "sculptor/1e7_V31_r3.2/", "stars")
+stars_dir = joinpath(ENV["DWARFS_ROOT"], "analysis", "sculptor/1e7_new_v31_r3.2/", "stars")
 
 # ╔═╡ 7a6d6345-fefb-4017-8230-3ff5bea55c83
 snap = Snapshot(joinpath(stars_dir, "iso_paint.hdf5"))
 
 # ╔═╡ a3a3a875-d0cc-4594-86c2-b242e859efef
-df_probs = LilGuys.read_hdf5_table(joinpath(stars_dir, "exp2d_rs0.08", "probabilities_stars.hdf5"))
+df_probs = LilGuys.read_hdf5_table(joinpath(stars_dir, "exp2d_rs0.10", "probabilities_stars.hdf5"))
 
 # ╔═╡ 01e34341-e806-44ef-9591-d226b345dec6
 snap.weights = df_probs.probability[snap.index]
@@ -38,10 +38,10 @@ snap.weights = df_probs.probability[snap.index]
 prof_dm = DensityProfile(snap, bins=-2:0.1:2.1)
 
 # ╔═╡ 932473fc-535f-4dab-bb03-53c8a71aff16
-prof_dm_ana = LilGuys.load_profile(joinpath(ENV["DWARFS_ROOT"], "simulations/sculptor/1e7_V31_r3.2/halo.toml"))
+prof_dm_ana = NFW(r_circ_max=6, v_circ_max=31/V2KMS)
 
 # ╔═╡ a6fd5c3f-c9ae-40a0-a8b7-c09b9dd9d2af
-prof_stars_ana = LilGuys.load_profile(joinpath(ENV["DWARFS_ROOT"], "analysis/sculptor/1e7_V31_r3.2/stars/exp2d_rs0.08/profile.toml"))
+prof_stars_ana = LilGuys.load_profile(joinpath(ENV["DWARFS_ROOT"], "analysis/sculptor/1e7_new_v31_r3.2/stars/exp2d_rs0.10/profile.toml"))
[…]

### 2025-08-29 thesis/abstract.tex
julio edits at durham

--- 2025-07-15 thesis/abstract.tex
@@
-Dwarf galaxies are unique cosmological probes, testing our understanding
-of the smallest scales of structure formation in the Universe. Among the
-classical dwarf galaxies, Sculptor and Ursa Minor appear unusual, with
-more stars than expected in the outer regions compared with the typical
-Exponential density profile, rememnescent of tidal interaction. I verify
-the validity of this excess, checking \citet{jensen+2024}'s bayesian
-membership catalogue for completeness, simple cuts, and other surveys.
-To evaluate if tides indeed can explain these density features, I
-conduct idealized N-body simulation of both galaxies in the tidal field
-of the Milky Way. In both cases, the orbits are well constrained,
-producing pericenters larger than expected to produce substantial tidal
-impacts. In addition, the time since the last pericenter does not agree
-with a tidal interpretation. However, Sculptor's orbit is strongly
-influenced by the presence of an LMC. With an LMC, Sculptor's long term
-orbital history is highly uncertain, however recent tidal effects due to
-the Milky Way and LMC only minimally affect this galaxy. Because tides
-likely do not explain the presence of these density features, I review
-other theories about the formation of dwarf galaxy ``halos,'' including
-mergers / accretion events and complex multi-component star formation
-histories. The density profiles of Sculptor and Ursa Minor, with their
-deviations from the classically expected exponential, present an
-interesting puzzle, likely with a solution involving their formation and
-history in the early universe.
+The satellite galaxies of the Milky Way (MW), with the exception of the
+Magellanic clouds, are all dwarf spheroidals (dSphs)---gas-free,
+non-rotating stellar systems who's gravity is dominated by dark matter.
+The light profiles of dSphs are usually well-approximated by an
+exponential law, with a sharp outer cutoff. Yet systems like the
+Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
+stars far from the main body, out to \textasciitilde10 effective radii.
+The origin of these extended profiles is not clear, and may reflect
+either the influence of Galactic tides or instead an innate feature. In
+this thesis, I review the excestence of an extended light profiles in
+Scl and UMi, validating the Bayesian membership catalogue of
+\citet{jensen+2024}. To evaluate if tides indeed can explain these
+density features, I conduct idealized N-body simulation of both galaxies
+in the tidal field of the Milky Way. I find that neigher dwarf has been
+subjected to tidal forces strong enough to create the observed extended
+light profiles from initially exponential profiles. One complication is
+that Sculptor's orbit is strongly influenced by the presence of the
+Large Magellanic Cloud (LMC). Because the LMC's mass is poorly
+constrained, Sculptor's long term orbital history is highly uncertain.
+Despite this uncertainty, our models still suggest that the combined
+tides of the LMC and MW are still unable to explain Scl's outer profile
+over the past 5 Gyr. I conclude that the extended light profiles of Scl
+and UMi are innate, likely reflecting their formation histories.
+Mergers, accretion events, or complex, multi-component star formation
+episodes are possible explanations for such an extended outer profile.

### 2025-08-29 thesis/data.tex
julio edits at durham

--- 2025-08-18 thesis/data.tex
@@
-To provide an empirical benchmark for simulations in the next chapters,
-we aim to understand the observed properties and density profiles of
-Sculptor and Ursa Minor. In this section, we first compile past
-measurements of each galaxy. We then describe the \citet{jensen+2024}
-Bayesian methodology to select high probability member stars from
-\emph{Gaia}. We show the derived density profiles are robust to changes
-in assumptions. While Fornax is well represented by a 2D exponential
-profile, Sculptor and Ursa Minor show an excess of stars relative to an
-2D exponential in the outer regions, better described by a Plummer
-profile. Future chapters assess possible explanations for these excess.
+As discussed in Section~\ref{sec:exponential_profiles}, the projected
+luminosity/stellar mass density profile of dwarf galaxies is generally
+well-described by an exponential law Eq.~\ref{eq:exponential_profiles}.
+A prototypical example is the Fornax dSph --- well-fit by an exponential
+over 4.5 decades in surface density. On the other hand, dSph like
+Sculptor or Ursa Minor have profiles with deviate significantly from an
+exponential profile fitted to the inner regions, with a clear excess of
+stars/light in the outer regions. In this chapter, we critically review
+the density profile of MW classical dwarfs. This thesis concerns the
+origin of this excess, with a tidal interpretation explored in
+subsequent chapters.
 
-\section{\texorpdfstring{\emph{Gaia} data}{Gaia data}}\label{gaia-data}
+\textbf{question: how much about density profiles in this
+paragraph\ldots{}}
 
-Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
-membership estimation method. J+24 use a Bayesian framework
-incorporating proper motion (PM), colour-magnitude diagram (CMD), and
-spatial information to determine the probability that a given star
-belongs to the satellite. By accounting for PM in particular, J+24
-produces low contamination samples of candidate member stars out to
-large distances from a dwarf galaxy. J+24 extends the algorithm
-presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
-an optional secondary, extended spatial component to find possible
-members as far as \textasciitilde10 half-light radii \(R_h\) from
-several dwarf galaxies. See also similar work by \citet{pace+li2019};
-\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.
+\section{\texorpdfstring{Satellite stellar membership with
+\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{satellite-stellar-membership-with-gaia}
 
-J+24 define likelihoods \({\cal L}\) representing the probability
-density that a star is consistent with either the MW stellar background
-(\({\cal L}_{\rm bg}\)) or the satellite galaxy
-(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
-product of a spatial, PM, and CMD term: \begin{equation}{
+Measuring the light profile of a resolved galaxy requires careful
+consideration of whether a given star belongs to the system. Without
+removing contamination from foreground/background features, faint
+features may be lost in the noise or be of uncertain association. When
+only photometric data was available, the galaxy membership of stars were
+ascertained using the colour-magnitude diagram alone \citep[e.g.,
+matched filter methods like][]{rockosi+2002}. Now that \emph{Gaia} data
+are available, stellar parallax and proper motion are also available to
+improve membership assignment.
+
+Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
+probabilities for \emph{Gaia} observations. J+24 used a Bayesian
+framework incorporating proper motion (PM), colour-magnitude diagram
+(CMD), and spatial information to determine the probability that a given
+star belongs to the satellite or foreground/background. By accounting
+for PM in particular, J+24 produced low contamination samples of
+candidate member stars out to large distances from a dwarf galaxy. J+24
+extended the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
+additionally includes a secondary, extended spatial component. J+24
+detected candidate members out to \textasciitilde10 half-light radii
+from the centres of some galaxies (\(R_h\)). Similar recent work has
+also included \citet{pace+li2019}; \citet{battaglia+2022};
+\citet{pace+erkal+li2022}; \citet{qi+2022}.
+
+J+24 construct likelihoods, \({\cal L}\), representing the probability
+density that a star is consistent with either the foreground/background,
+\({\cal L}_{\rm bg}\), or the satellite galaxy, \({\cal L}_{\rm sat}\).
+In either case, the likelihoods are the product of a spatial, PM, and
+CMD component, \begin{equation}{
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
@@
 
-Each likelihood is normalized over their respective 2D parameter space
-for both the satellite. To control the relative frequency of member and
-background stars, \(f_{\rm sat}\) representing the fraction of member
-stars in the field. The total likelihood for any star in this model is
-the sum of the satellite and background likelihoods, weighted by their
-relative frequencies, \begin{equation}{
+For the satellite, the \emph{spatial likelihood} is specified either as
+a one- or two-component elliptical exponential profile. A second
+component is included only when the preferred amplitude of the second
+component is non-zero. The \emph{proper-motion likelihood} quantifies
+the agreement of a star's motion with the dwarf galaxy's systemic
+motion, accounting for observational uncertainties. The \emph{CMD
+likelihood} measures the consistency of a star's \emph{Gaia} \(G\)
+magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
+isochrones for the galaxy. For the background likelihoods, the spatial
+likelihood is a uniform distribution over the field, and the background
+CMD and PM likelihoods are constructed empirically from the stars in an
+annulus far from the satellite. Each likelihood is normalized as a
+probability density over the respective parameter space.
+
+The total likelihood is a mixture model of the satellite and background,
+weighted by the fraction of stars in the field belonging to the
+satellite, \(f_{\rm sat}\): \begin{equation}{
 {\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
-}\end{equation} The probability that any star belongs to the satellite
-is then given by \begin{equation}{
+}\end{equation} The probability that a given star belongs to the
+satellite is then \begin{equation}{
 P_{\rm sat} = 
@@
 
-\section{Selected stars}\label{selected-stars}
+J+24 fit this model using Monte Carlo Markov chain simulations, solving
+for the proper motions, satellite membership fraction (\(f_{\rm sat}\)),
+and structural properties of the second exponential density profile (if
+included). The median parameters from the samples are then used to
+calculate the \(P_{\rm sat}\) we use for sample selection.
 
-In
+For our fiducial sample, we adopt a minimum probability of
+\(P_{\rm sat} = 0.2\). Most stars have \(P_{\rm  sat}\) values which are
+nearly 0 or 1, so the exact choice of probability threshold has little
+effect on the resulting sample. Even at our relatively generous
+probability threshold of 0.2, the purity remains high when validated
+against spectroscopic line-of-sight (LOS) velocities
+(\textasciitilde90\%, J+24). We note, however, that these purity
+estimates may be biased. Stars with LOS velocities are typically
+brighter and, as a result, have more precise \emph{Gaia} measurements.
+However, we find that our conclusions are unchanged when limiting
+samples to only the brightest stars. Altogether, the J+24 method
+provides a high-quality, low-contamination sample of dwarf galaxy
+candidate member stars, which we will now investigate.
+
+\section{The effects of membership
+criteria}\label{the-effects-of-membership-criteria}
+
+We analyze stellar distributions in the tangent plane, considering as
+well the projected shape of a galaxy. The tangent plane coordinates
+\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
+the plane tangent to the galaxy centre. To account for the elliptical
+shape of the galaxy, we use \(R_{\rm ell}\), which we define as the
+circularized elliptical radius, \begin{equation}{
+R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2),
+}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
+coordinate rotated to align with a dwarf galaxy's major and minor axis,
+and \(a\) and \(b\) are the semi-major and semi-minor axis of the
+galaxy.
+
+We illustrate the likelihood-based membership selection criteria through
+a progressive tightening of criteria. First, we consider a
+minimally-refined sample, \textbf{all}, only excluding stars with poor
+astrometry, unreliable photometry, or inconsistent parallaxes. Next,
+incorporating CMD and PM information, we use the \textbf{CMD+PM} to test
+a selection method agnostic to the spatial position. This sample
+includes stars where the CMD and PM combined likelihood favours
+satellite membership:
+\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
+{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
+}\end{equation} Next, our \textbf{fiducial sample},
+\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
+J+24. Finally, the \textbf{RV members} sample adds line-of-sight
+velocity information (see Appendix for details). While the latter is the
+best motivated membership sample, we do not use this sample for stellar
+density analysis due to its incompleteness and complex selection
+function.
+
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
+shows our fiducial sample for each galaxy, and the impact of alternate
+samples, in tangent-plane coordinates, the CMD, and in proper-motion
+space. The \emph{all} sample extends uniformly across the tangent plane,
+but includes a substantial background population. The \emph{CMD+PM}
+sample has a much lower background density in the tangent plane,
+revealing the satellite more clearly. The fiducial sample appears
+similar to the \emph{CMD+PM} sample in the CMD and PM planes, but
+excludes improbably distant stars. The \emph{RV member} sample also
+traces out similar distributions in CMD and PM space as the fiducial
+sample, with less dispersion, likely reflecting the brighter magnitudes
+of stars with spectroscopic follow-up. Each component plays an important
+role in ensuring a high-quality membership criterion.
+
+Based on the fiducial sample's distribution in
 Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
-we illustrate the resulting samples from the algorithm in the tangent
-plane, CMD, and proper motion space. The tangent plane coordinates
-\(\xi\) and \(\eta\) are the distances in RA and declination as measured
-on a plane tangent to the dwarf galaxies centre. \(R_{\rm ell}\) is
-defined as the circularized elliptical radius, so
-\(R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2)\) where
-\(\xi'\) and \(\eta'\) are rotated to align with the position angle
-direction \(\theta\).
+Ursa Minor stands out because of its higher ellipticity. Sculptor and
+Ursa Minor show a population of members past \(6R_h\) in radius, but
+Fornax does not, even though Fornax has many more stars. Fornax also has
+a bluer CMD, indicative of more recent star formation. Otherwise, there
+are no standout morphological features between these galaxies.
 
-For our fiducial sample, we adopt a probability cut of
-\(P_{\rm sat} = 0.2\). Most stars are assigned probabilities close to
-either 0 or 1, so the choice of probability threshold is only marginally
-significant. Additionally, even for a probability cut of 0.2, the purity
-of the resulting sample with RV measurements is very high
-(\textasciitilde90\%, J+24). Note there is likely a systematic bias in
-using stars with RV measurements to measure purity. Fainter stars are
-less likely to have been targeted and have poorer astrometry and colour
-information. We find no difference in the resulting density
-distributions when restricting stars to be brighter than a specific
-magnitude.
+One limitation of the J+24 method is the assumption of a specific
+density profile for the dwarf galaxy (a one- or two-component
+exponential), which may impact the membership probability of distant
+stars. However, there are no clear extensions or overdensities in the
+\emph{CMD+PM} selected sample, which does not include a spatial
+likelihood. If even fainter, more extended features exist, they are
+apparently not detectable with \emph{Gaia}.
 
-For all galaxies, each criteria is commensurate in sifting out
-nonmembers. The CMD is well defined, and probable members only extend a
-few times the colour uncertainty from the CMD. In proper motion space,
-selected stars are within \(\masyr\) from the systematic proper motion.
-The stars furthest away in proper motion space typically have large
-uncertainties \(\sim 1\masyr\), so most members are reasonably
-consistent. In any case, the algorithm removes stars with inconsistent
-PMs. Finally, the spatial likelihood reduces the probabilities of stars
-distant from the dwarfs centres. Member stars are rare outside of
-\(\sim5R_h\), resulting from both the likelihood specification and the
-lower density of background stars consistent with the CMD and PM of the
-satellite.
-
-One limitation of the J+24 method is that it assumes a spatial form of
-the dwarf galaxy, possibly limiting the detection of very distant stars
-or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
-illustrates the distribution of stars selected without a spatial
-criterion. We define the \emph{CMD+PM} selection as
-\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
-{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
-}\end{equation} with the likelihoods from J+24 as described above. These
-stars are distributed similar to the fiducial (probable members) sample
-but with a background uniform distribution across the entire field. This
-illustrates the approximate background of stars which may be confused as
-members. Additionally, since there is no clear spatial structure in the
-\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
-unlikely that there are additional faint, tidal features detectable with
-\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
-cuts and thresholds, finding no evidence of other structure in both
-\emph{Gaia} and DELVE/UNIONS data.
-
-Finally, we illustrate the location of RV-confirmed members from
-Section~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
-brighter than a typical \emph{Gaia} candidate, RV members typically have
-more precise PMs. However, the RV members fill about the same area on
-the CMD down to the magnitude limit. Additionally, while the number of
-stars with RV measurements decreases at large radii, there are still
-confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
-stars illustrate the extended profiles of each galaxy. Note that for an
-exponential profile, 99.95\% of stars should be within \(6R_h\), so the
-discovery of any stars beyond \(6R_h\) hints at a deviation from an
-exponential profile. These \emph{extratidal} stars hint at something
-tidal in nature, and our goal is to explore possible interpretations.
+To illustrate the extent of Sculptor and Ursa Minor, we highlight the
+most distant, spectroscopically confirmed members with red-outlined
+stars in Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In
+particular, \citet{sestito+2023a}; \citet{sestito+2023b} targeted the
+most distant stars using the J+24 candidate membership list. These stars
+illustrate an extended stellar distribution for each galaxy. For an
+exponential profile, 99.95\% of stars fall within a radius of \(6R_h\),
+so the mere presence of very distant member stars provides direct
+evidence for deviations from a simple, exponential profile.
 
@@
 \includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
-\caption[Sculptor sample selection]{The selection criteria for Scl
-members. Light grey points represent field stars (satisfying quality
-criteria), turquoise points CMD+PM selected stars
-(Eq.~\ref{eq:sel_cmd_pm}), blue xs probable members (2-component), and
-RV confirmed members indigo diamonds. We mark the two stars from
-\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
-Tangent plane \(\xi, \eta\). The orange ellipse represents 3 half-light
-radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\)
-magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot the Padova
-12Gyr {[}Fe/H{]}=-1.68 isochone in orange. The black bar in the top left
-represents the median colour error. \textbf{Bottom right:} Proper motion
-in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\) (corrected) the
-orange circle represents the \citet{MV2020b} proper motion. The black
-cross represents the median proper motion
-error.}\label{fig:scl_selection}
+\caption[Sculptor sample selection]{The distributions of various samples
+of \emph{Gaia} stars for Sculptor. We plot light grey points for field
+stars (with consistent parallaxes and reliable photometry and
+astrometry), turquoise points for \emph{CMD+PM} selected stars
+(Eq.~\ref{eq:sel_cmd_pm}), blue \texttt{x}s for the fiducial sample
+(\(P_{\rm sat} > 0.2\)), and indigo diamonds for the RV confirmed
+members. We mark the two far-outlier stars from \citet{sestito+2023a}
+with rust-outlined indigo stars. \textbf{Top:} Tangent plane
+\(\xi, \eta\). The orange ellipse represents 3 half-light radii.
+\textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\) magnitude
+versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a Padova \(12\,\)Gyr,
+\({\rm [Fe/H] }=-1.68\) isochone in orange. The black bar in the top
+left represents the median colour error. \textbf{Bottom right:} Proper
+motion in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
+(corrected). The orange point marks the systemic \citet{MV2020b} proper
+motion. The black cross represents the median proper motion error.
+\textbf{TODO: reduce png sizes}}\label{fig:scl_selection}
 \end{figure}
@@
 Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
-members outside of \(6R_h\) in black stars (from \citet{sestito+2023b},
-\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
+members outside a radius of \(6R_h\) with red stars (from
+\citet{sestito+2023b}, \citet{pace+2020} and \citet{spencer+2018}).
+\textbf{TODO: plot 6Rh ellipse\ldots{}}}\label{fig:umi_selection}
 \end{figure}
@@
 Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
-\citet{WMO2009}. While possibly a limitation of RV sample selection,
-Fornax does not show the same extended outer halo of probable members as
-Sculptor or Ursa Minor despite having many more
+\citet{WMO2009}. Fornax does not show the same extended outer halo of
+probable members as Sculptor or Ursa Minor despite having many more
 stars.}\label{fig:fornax_selection}
@@
 
-Density profiles are an essential observational constraint for our later
-simulations. To derive density profiles, we use 0.05 dex bins in log
-radius. We remove bins at smaller (larger) radii than the first empty
-bin working outwards. We use symmetric poisson uncertainties. As
-discussed below, these uncertainties are straightforward but are likely
-under-representative. We use the values from \citet{munoz+2018}'s Sérsic
-maximum likelihood fits for \(R_h\), which are more precisely derived
-given their deeper photometry.
+We derive density profiles by binning member stars in constant width
+bins in \(\log R_{\rm ell}\) of 0.05 dex. We ignore bins interior or
+exterior to the first empty bin in either direction. We use symmetric
+Poisson uncertainties as ``error bars'' in the density estimate at each
+bin.
 
@@
 for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
-different selections of stars from above: all quality stars, CMD+PM
-only, and the probable member samples. In each case, all samples are the
-same towards the inner regions of the satellite. Classical dwarfs
-dominate the stellar density in their cores. However, the density
-profile \emph{all stars} plateaus at the total background in the field
-at radii of 30-60 arcminutes. By restricting stars to being most likely
-satellite members by CMD + PM, the background is reduced by 1-2 dex. The
-CMD+PM background plateau likely represents the density of background
-stars which could be mistaken as members. Finally, the background
-(BG)-subtracted profile results from subtracting the apparent background
-in the \emph{all} profile.
+different selections of stars from above: \emph{all} stars, CMD+PM only,
+and the fiducial sample. In each case, all samples are the same towards
+the inner regions of the satellite. However, the density profile
+\emph{quality stars} plateaus at the total background in the field at
+radii of 30-60 arcminutes, depending on the galaxy. By restricting
+stellar membership with CMD + PM, the background is reduced by 1-2 dex.
+The CMD+PM background plateau represents the density of background stars
+which could be mistaken as members because of their coincident colours
+and proper motions. Finally, the background (BG)-subtracted profile
+results from subtracting the apparent background in the \emph{all}
+profile.
 
-Note that the probable members (fiducial) density profile continues to
-confidently estimate the density profile below the CMD+PM background.
-These points are likely unreliable (see discussion below). However,
-before this point, both the BG subtracted and probable members density
-profiles are strikingly similar. Assumptions about the details of the
-likelihood and spatial dependence have marginal influence on the
-resulting density profile when the satellite is higher density than the
-background. Thus the detection of excesses of stars in Sculptor (past 20
-arcmin) and UMi (past 15-20 arcmin) are robust.
+\textbf{Describe R\_break, etc. in Fig.2.4} Note that, in all three
+galaxies, the data does not extend out to the break radius.
 
-Nearby to UMi, there is a small, likely unassociated, ultrafaint star
+While the density profiles of all samples agrees in the inner regions,
+they begin to deviate outside \(6 R_h\). Outside this radius, the
+fiducial profile extends 1--2 magnitudes below the CMD+PM selection
+background. Because the fiducial profile in this region depends
+critically on the spatial likelihood, its shape is vulnerable to the
+assumed density profile (an exponential in J+24). We explore this in
+more detail in Appendix {[}@-sec:density\_extra{]}.
+
+Nearto UMi, there is a small, likely unassociated, ultrafaint star
 cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
 position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
-to an elliptical radius of 36 arcminutes. However this cluster does not
-have a bright RGB, so has few stars brighter than a \(G\) mag of 22.
-While Muñoz 1 may contribute \(\sim\) 5 stars to the density profile of
-UMi, this would possibly only affect 1-2 bins due to its compact size.
+to an elliptical radius of 36 arcminutes. This cluster does not have a
+bright RGB, so has few stars brighter than a magnitude \(G=22\). The
+cluster has little effect on the elliptically-averaged density profile.
+\textbf{Mark Muñoz 1}
 
-To illustrate the differences between each dwarf galaxy, in
-Fig.~\ref{fig:classical_dwarfs_densities}, we compare Sculptor, Ursa
-Minor, and other classical dwarfs against 2D-exponential and Plummer
-density profiles (REF). All dwarf galaxies appear to be well described
-by an exponential profile in the inner regions. However, while an
-exponential continues to be a good description for other dwarf galaxies,
-Sculptor and Ursa Minor diverge, showing a substantial excess over an
-exponential, better fit by a Plummer instead. In fact, Sculptor and Ursa
-Minor show excesses of between 2-3 orders of magnitude beyond what is
-expected of an exponential at the very outer regions!
+Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
+profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
+galaxies. Of the classicals, we exclude Antlia II due to the extremely
+high background, and Sagittarius which was not included in J+24. The
+density profiles are scaled to match at the half-light radius from
+\citet{munoz+2018}. All of the classical dwarfs appear to be well
+described by an exponential profile in the inner regions. In the outer
+regions, however, Sculptor and Ursa Minor deviate, and show a clear
+outer excess over an exponential law (solid black line). These galaxies
+are better fit by a Plummer law (dashed black line). The deviation from
+an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
+of magnitude. This excess is remarkable because the spatial likelihood
+used to derive the fiducial sample probabilities is the same (a single
+component exponential) in all cases. The remainder of this thesis will
+be devoted to assessing whether the outer excess shown by Scl and UMi
+are due to Galactic tides.
 
@@
 half-light radius (vertical dashed line) and the break radius (black
-arrow, REF).}\label{fig:scl_observed_profiles}
+arrow, Eq.~\ref{eq:r_break}). \textbf{rename
+bg-subtracted}}\label{fig:scl_observed_profiles}
 \end{figure}
@@
 \end{figure}
-
-\section{Summary}\label{summary}
-
-In summary, we have used J+24 data to derive the density profiles for
-the classical dwarf galaxies. In each case, the density profile is
-robust against different selection criteria. Both Sculptor and Ursa
-Minor show strong evidence for deviations from an exponential profile.
-We find no evidence of additional (velocity or stellar) substructure in
-either galaxy. Our density profiles are consistent with many previous
-works, all indicating divergence from a King or Exponential density
-profile with a break between 30 and 60 arcminutes. Our goal in the
-following chapters is to test if tides provide a viable explanation for
-the observed properties of Sculptor and Ursa Minor.

### 2025-08-29 thesis/introduction.tex
julio edits at durham

@@
 Dwarf galaxies host, in many ways, the most extreme galactic
-environments in the universe. These little galaxies are typically
-defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
-similarly \(L \lesssim 10^9 L_\odot\),
-e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
-dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
-highly \emph{dark-matter dominated}, with mass to light ratios exceeding
-100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
-many dwarf galaxy satellites of the Milky Way are \emph{quenched}, with
-little to no recent star formation. Milky Way dwarfs contain stellar
-populations which are \emph{relics} from the early universe, consisting
-of many of the oldest and most metal poor stars. Understanding the
-properties of dwarf galaxies thus has implications across astronomy,
-from cosmological structure formation on the smallest scale to extremely
-metal-poor stellar populations to chemical evolution to galactic
-dynamics. Of the classical dwarfs, we find that the Sculptor and Ursa
-Minor dwarf spheroidal galaxies stand out with a more extended density
-profile relative to an exponential, hinting at tidal effects. As a case
-study in tidal effects and our understanding of dwarf galaxy formation
-and evolution, we aim to understand the dynamical history of these
-galaxies.
+environments in the universe. These galaxies are typically defined to be
+fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
+or similarly \(M_\star \lesssim 10^9 M_\odot\) {[}e.g.,
+\citet{hodge1971}; \citet{mcconnachie2012}). Because the galaxy
+luminosity function increases towards fainter objects, dwarfs are the
+most numerous galaxies in the Universe. Dwarf galaxies are also highly
+\emph{dark-matter dominated}, with mass to light ratios which may exceed
+1000 \(M_\odot/ L_\odot\). With the exception of the Magellanic Cloud,
+all dwarf galaxy satellites of the Milky Way (MW) are \emph{quenched},
+with little to no recent star formation. Indeed, most faint MW
+satellites contain stellar populations which are \emph{relics} from the
+early universe, consisting of many of the oldest and most metal poor
+stars.
 
-In this section, we first describe cosmological structure formation and
-the role of dwarf galaxies. Then, we explore the observational history
-of dwarf galaxies, the \emph{Gaia} telescope, open questions, and why
-Sculptor and Ursa Minor stand out. We discuss idealized simulations,
-tidal effects, break radii, and recent developments in tidal
-simulations. Finally, we provide a roadmap to the thesis as a whole at
-the end of this section.
+Understanding the properties of dwarf galaxies thus has implications
+across astronomy, from cosmological structure formation on small scales
+to the formation of metal-poor stellar populations.
 
-\section{Observational context}\label{observational-context}
+\emph{Remove this paragraph?} Of the classical dwarfs, we find that the
+Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
+extended density profile relative to an exponential, hinting at tidal
+effects. As a case study in tidal effects and our understanding of dwarf
+galaxy formation and evolution, we aim to understand the dynamical
+history of these galaxies.
 
-Dwarf galaxies have long raised conundrums in theories of galaxy
+In this section, we first describe the general observed properties of
+local dwarf galaxies. Next, we summarize the cosmological origin and
+role of dwarf galaxies. We then review recent updates to the observed
+properties of dwarf galaxies made on new facilities. We then discuss how
+tides influence satellite galaxies in general. Finally, we provide a
+roadmap to the thesis. \textbf{Look at this paragraph again}
+
+\section{Observational history of dwarf
+galaxies}\label{observational-history-of-dwarf-galaxies}
+
+Dwarf galaxies have long raised conundrums for theories of galaxy
 formation. The discovery of Fornax and Sculptor in 1938
 \citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
-  Clouds (LMC, SMC) may be classified as dwarf galaxies, but these were
-  likely always known to humans at southern latitudes.}, with no known
-analogues, already presented such an enigma. Shapley presented these
-dwarfs as a new type of \emph{stellar system} resembling the Magellanic
-Clouds and globular clusters but did not attempt to speculate on the
-exact nature. While dwarf galaxies were quickly understood to be
-galaxies based on the inferred luminosities and sizes, their nature and
-formation remained unclear for decades
+  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
+  were likely always known to humans at southern latitudes.}, with no
+known analogues, already presented such an enigma. Shapley presented
+these dwarfs as a new type of \emph{stellar system} resembling the
+Magellanic Clouds and globular clusters but did not attempt to speculate
+on their exact nature. While dwarf galaxies were quickly understood to
+be galaxies based on the inferred luminosities and sizes, their nature
+and formation remained unclear for decades
 \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
@@
 velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
-(dSph) galaxies
-\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
-mass-to-light ratios were at least 10 times larger than the values from
-globular clusters scaled to the same sizes. While rather uncertain
-initially, these values have corroborated with larger and more precise
-samples \citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
-attempting to understand the formation and observed properties of these
-objects were proposed. Examples include: dwarf galaxies are undergoing
-tidal dissolution resulting in extreme mass-to-light ratios
-\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
-holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark
-matter-free ``tidal dwarfs'' from past mergers
-\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
-of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
-formation was also not out of the question
-\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
-galaxies are among the darkest objects in the universe, and
-understanding their properties is critical to understanding dark matter.
+(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
+inferred mass-to-light ratios were at least 10 times larger than the
+values of globular clusters (GCs). While rather uncertain initially,
+these values have been corroborated with larger and more precise samples
+\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
+proposed to explain these unusually high mass-to-light ratios. Examples
+include: ongoing tidal disruption artificially inflating the
+line-of-sight velocity dispersions \citep[e.g.,][]{kuhn+miller1989}, the
+presence of massive central black holes
+\citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
+\citep{milgrom1995}. Over time, however, a consensus developed where the
+high mass-to-light ratios of dwarf galaxies was due to the presence of a
+dark matter halo, which was lacking in GCs
+\citep[e.g.,][\textbf{REFS}]{dekel+silk1986}. Since then, the properties
+of dwarf galaxies have played an increasingly important role in our
+understanding of the clustering of dark matter on small scales {[}e.g.,
+sales+2022, bullock+boylan-kolchyn2017{]}.
 
@@
 Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
-galaxy displaying a bar. \textbf{Classical dwarfs} occupy intermediate
-luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
-systems are old, gas-poor, and spheroidal. All Milky Way satellites
+galaxy, where most stars are in a rotationally-supported thin disk (seen
+nearly face-on) with a prominent bar. \textbf{Classical dwarf
+spheroidals} occupy intermediate luminosities (
+\(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these systems are old,
+non-star forming, gas-poor, and spheroidal. All Milky Way satellites
 discovered before digital sky surveys are classicals. The 12 classical
-dwarfs satellites of our Galaxy are Sagittarius, Fornax, Leo I,
-Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
-I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
-  names we discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor
-  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
-occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
-galaxies have minuscule stellar masses, tend to be more compact, and are
-the darkest known galaxies. Altogether, dwarf galaxies span more than 15
-orders in absolute magnitude.
+dwarf satellites of our Galaxy are (in order of decreasing luminosity)
+Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II, Carina, Draco,
+Ursa Minor, Canes Venatici I, Sextans I, and Crater II.\footnote{While
+  formally the dwarf galaxy names we discuss contain ``dwarf
+  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
+  brevity.} The \textbf{ultra-faint}s dwarfs (UFDs) occupy the very
+faintest magnitude range (\(-7.7 \lesssim M_V\)). These galaxies have
+minuscule stellar masses, tend to be more compact, and include the
+faintest known galaxies (\textbf{ref}). Altogether, known dwarf galaxies
+span more than 15 orders in magnitude, or over 7 decades in stellar
+mass. \textbf{Rewrite to include stellar mass ranges as well}
 
-Since Local Group dwarfs are nearby, we can study these galaxies on a
-star-by-star basis. As a result, it is possible to measure the 3D
-velocity and position of a star if we can measure the stars position,
-line-of-sight (LOS) velocity, and proper motions. Unfortunately,
-determining distances and full 3D velocities is challenging. The most
-direct measurement of distance, parallax, requires precise tracking of a
-star's sky position across a year. And while line-of-sight (LOS)
-velocities are easily determined from spectroscopy, tangential
-velocities, derived from proper motions and distances, are much more
-challenging. Typically, measuring proper motions requires accurate (much
-less than arcsecond) determinations of small changes in a star's
-position over baselines of several years to decades. The full 6D
-position and velocity information for stars has historically been known
-for only a handful of stars.
+A common definition today for a dwarf galaxy is a stellar system with
+evidence for dark matter (high mass-to-light ratios, e.g.~\textbf{REF}).
+In contrast, star clusters (like GCs) have no clear evidence for dark
+matter.
 
@@
 \end{figure}
+
+\section{Cosmological context}\label{cosmological-context}
+
+We only understand a tiny fraction of the universe's composition. The
+leading theory of cosmology, \LCDM (Lambda Cold Dark Matter), posits
+that the universe is composed of about 68\% dark energy (\(\Lambda\)),
+27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
+  to change definitions of words. \emph{Baryons} here means
+  baryons+leptons, i.e.~any standard model massive fermion.}
+\citep{planckcollaboration+2020}. While the composition of dark matter
+and dark energy remains elusive, we know their general properties. Dark
+energy causes the acceleration of the expansion of the universe on large
+scales. We do not discuss dark energy here---it does not substantially
+affect the Local Group today. Dark matter, instead, makes up the vast
+majority of mass in galaxies. Typically, galaxies have baryonic to dark
+matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
+In \LCDM, dark matter is assumed to interact only gravitationally. Light
+passes through dark matter unimpeded---in this sense dark matter is
+transparent. Dark matter is also commonly assumed to be \emph{cold},
+i.e.~typical velocities much smaller than the speed of light in the
+early universe. If dark matter is cold, then it should condense on all
+scales, forming non-linear structures (or \emph{halos}) from the size of
+galaxy clusters to smaller than the faintest dwarf galaxies.
+Implications of dark matter properties range from cosmological
+structural formation, galaxy structure, and galaxy interactions.
+
+\textbf{is this ordered logically?}
+
+\subsection{\texorpdfstring{Dwarf galaxy formation in
+\LCDM}{Dwarf galaxy formation in }}\label{dwarf-galaxy-formation-in}
+
+\textbf{In the next section, we consider galaxy formatin in LCDM}
+
+The very early universe was almost featureless. Our earliest
+observations of the universe stem from the cosmic microwave background
+(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
+But tiny perturbations in the CMB, temperature fluctuations of 1 part in
+10,000, reveal the underlying seeds of large-scale cosmological
+structure. In an expanding universe, gravitational instability makes CDM
+overdensities grow and collapse hierarchically onto larger structures.
+Initially, baryonic matter was coupled to radiation and resisted
+collapse. Dark matter, only influenced by gravity instead, freely
+collapsed into the first structures. For mass perturbations both smaller
+than the horizon and overdense enough to collapse, these overdensities
+of dark matter become self-gravitating, known as \emph{halos}. After
+recombination, where electrons combined with atomic nuclei to form
+atoms, baryons decoupled from radiation and fell into the dark matter
+halos. The densest pockets of baryons later formed the first stars and
+galaxies.
+
+Dark matter halos, and their associated galaxies, rarely evolve in
+isolation. Instead, structure formation is \emph{hierarchical}. Small
+dark matter halos collapse first and hierarchically merge into
+progressively larger halos
+\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
+Structure formation happens on a wide range of scales. The largest
+structures become the cosmic web---composed of voids, filaments, and
+sheets---and the smallest non-linear structures (halos, \textbf{define
+non-linear}) host dwarf galaxies. Hierarchical assembly is evident
+through the large scale structure of the universe, remnants of past
+mergers within the Milky Way, and tidal disruption of dwarf galaxies and
+their streams around nearby galaxies.
+
+Small-scale structure formation is sensitive to deviations from
+\LCDM cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of
+\LCDM is that mass perturbations are expected to exist on all scales,
+and are largest on the smallest scales, so we would expect the formation
+of halos on all scales. Many alternative models, such as warm dark
+matter, may smooth out small-scale features and reduce the abundance of
+small halos or change their structure \citep[e.g.,][]{lovell+2014}.
+Dwarf galaxies, which occupy the smallest dark matter halos, are
+promising windows into small-scale structures. To understand the
+evolution of these objects, we need to understand the general
+predictions of \LCDM for the properties of dwarf galaxies.
+
+\subsection{Typical structure of dark matter
+halos}\label{typical-structure-of-dark-matter-halos}
+
+In \LCDM cosmological simulations, dark matter halos are remarkably
+self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
+observe that the spherically-averaged density profiles \(\rho(r)\) are
+universally well described by a two-parameter law:
+\begin{equation}\protect\phantomsection\label{eq:nfw}{
+\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
+}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
+density . This profile has shown remarkable success in describing
+\LCDM halos across several orders of magnitude in mass. NFW profiles are
+\emph{cuspy}, where the density continuously rises like
+\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
+density profile increases gradually with radius, and at large radii the
+density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
+shows an example NFW halo.
+
+The total mass of an NFW profile formally diverges, so halo masses are
+commonly defined using an overdensity criterion. The virial mass,
+\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
+containing a mean enclosed density 200 times\footnote{For the collapse
+  of a uniform spherical density, the virialized overdensity would be
+  \(\Delta = 18\pi^2\approx 178\) for a critical universe
+  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
+  this parameter may be closer to \(\Delta \approx 100\) for our
+  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
+  from][]{bryan+norman1998}.} the critical density of the universe:
+\begin{equation}{
+M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
+}\end{equation} A standard second parameter of NFW halos is the halo
+concentration, \(c=r_{200} / r_s\) which describes how the
+characteristic radius scale of the halo compares to the virial radius.
+Using this parameter, the scale density is a function of \(c\) alone,
+\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
+\citep{NFW1996}.
+
+An equivalent, alternative characterization of NFW halos uses their
+circular velocity profiles. The circular velocity,
+\(\V_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum
+\(\V_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
+\(\V_{\rm max}\) and \(r_{\rm max}\), like \(M_{200}\) and \(c\), fully
+specify an NFW halo,
+
+The two parameters of an NFW profile are not independent. Lower mass
+dark matter halos often collapse earlier, when the universe was denser.
+As a result, low mass subhalos tend to be more concentrated
+\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
+the mass-concentration relation, describes the mean trend of
+concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
+left panel of Fig.~\ref{fig:smhm} illustrates the present-day
+mass-concentration relation in terms of \(r_{\rm max}\) and
+\(\V_{\rm max}\). While there is a general trend for concentration to
+increase with mass, the concentration values still have substantial
+scatter. Other parameters such as the halo spin or shape may affect the
+scatter of the mass-concentration relation, but their effect is
+typically expected to be small
+\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
+\caption[Example density profiles]{Density profiles in log density
+versus log radius for stars and dark matter in a Sculptor-like galaxy.
+The dark matter is more extended and massive than the star across the
+entire galaxy.}\label{fig:nfw_density}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
+\caption[Stellar-mass halo-mass relation]{\textbf{Left} Radius of
+maximum circular velocity \(r_{\rm max}\) as a function of maximum
+circular velocity. The solid line with 1-\(\sigma\) shaded region and
+dashed line are the relations from \citet{ludlow+2016} for \(z=0\) and
+\(z=2\) respectively. \textbf{Right} Stellar mass (top) as a function of
+maximum circular velocity. The solid line with the 1-\(\sigma\) shaded
+region is the relation from \citet{fattahi+2018} with scatter points
+simulated central galaxies from APOSTLE in
+\citet{fattahi+2018}.}\label{fig:smhm}
+\end{figure}
+
+\subsection{\texorpdfstring{Formation of stars within
+\LCDM halos}{Formation of stars within halos}}\label{formation-of-stars-within-halos}
+
+The observed abundance of galaxies may be compared with the abundance of
+\LCDM halos to derive constraints regarding which galaxies inhabit which
+halos. This technique, dubbed ``abundance matching,'' implies a tight
+relation between the stellar mass of a galaxy and the mass of the halo
+it inhabits \citep{li+white, moster+naab+white2013}.
+
+Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
+(SMHM, with halo mass represented by \(\V_{\rm max}\)) predicted by the
+\LCDM cosmological hydrodynamical simulations from the the APOSTLE
+project \citep{sawala+2016}, which uses the hydrodynamical setup from
+the EAGLE project \citep{crain+2015, schaye+2015} to simulate Local
+Group analogues in a \LCDM cosmological context \textbf{(db: reorder
+these sentences)}. While there is some scatter, the range of predicted
+\({\rm v}_{\rm max}\) is fairly narrow across \(\sim 4\) decades in
+stellar mass. This figure indicates that the SMHM relation becomes
+increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
+formed in relatively massive halos. Because lower mass galaxies have
+increasingly shallow potential wells, feedback becomes more effective at
+removing gas. Re-ionization additionally suppresses late star formation
+in the faintest galaxies. As a result, the resulting stellar mass of a
+galaxy is highly sensitive to the dark matter mass, especially for faint
+dwarf galaxies. If we know the initial stellar mass of a dwarf galaxy,
+we also know, at some level, the properties of its dark matter halo.
+\textbf{fix this paragraph}
+
+In \LCDM galaxy formation, the majority of mass in dwarf galaxies comes
+from Dark Matter. Given the SMHM, Fig.~\ref{fig:nfw_density} shows an
+example stellar component for a classical dwarf galaxy with the expected
+underlying dark matter halo. Even when the stars are most dense in the
+inner regions, the dark matter remains nearly an order of magnitude
+higher in density. Stars have a small contribution to the gravitational
+structure of dwarf galaxies--instead stars are reasonably approximated
+as tracer particles of the underlying dark matter halo. \textbf{add
+something about stellar compactness}
+
+Several challenges complicate the SMHM trend, including environment,
+assembly history, tidal effects, and the details of galaxy formation. In
+particular, effects like ram-pressure stripping (removal of gas in the
+dwarf galaxy due to pressure from host's circumgalactic medium) and
+tidal removal of gas cause star formation to quench {[}\textbf{REFS}{]}.
+Additionally, the time of formation (relative to reionization) can
+influence the resulting stellar content \citep{kim+2024}. Finally, tides
+influence both the dark matter and stellar mass but in different amounts
+\citep[e.g.,][]{PNM2008}. Consequently, tides may reduce the halo mass
+more than the stellar mass, adding additional scatter to the SMHM trend,
+particularly for low-mass satellites \citep[e.g.,][]{fattahi+2018}.
+Understanding the effects of tides on Local Group dwarf galaxies is
+essential to determine where and how these galaxies formed in a
+cosmological context.
+
+\section{Observing faint features of dwarf
+galaxies}\label{observing-faint-features-of-dwarf-galaxies}
+
+\subsection{\texorpdfstring{The era of
+\emph{Gaia}}{The era of Gaia}}\label{the-era-of-gaia}
+
+Since Local Group dwarfs are nearby, they are resolved into individual
+stars, and we can study these galaxies on a star-by-star basis. As a
+result, it is possible to measure the 3D velocity and position of a star
+if we can measure the its position, distance, line-of-sight (LOS)
+velocity, and proper motions. Unfortunately, determining distances and
+full 3D velocities is challenging. The most direct measurement of
+distance, parallax, requires precise tracking of a star's sky position
+across a year. And while line-of-sight (LOS) velocities are relatively
+easily determined from spectroscopy, tangential velocities, derived from
+proper motions and distances, are much more challenging. Typically,
+measuring proper motions requires accurate (much less than arcsecond)
+determinations of small changes in a star's position over baselines of
+several years to decades. The full 6D position and velocity information
+for stars has, until recently, been known for only a handful of stars.
+
+Launched in 2013, \emph{Gaia} is a space-based, all-sky survey telescope
+situated at the Sun-Earth L2 Lagrange point
+\citep{gaiacollaboration+2016}. \emph{Gaia} has redefined astrometry,
+providing photometry, positions, proper motions, and parallaxes for over
+1 billion stars \citep{gaiacollaboration+2021}. While \emph{Gaia}
+completed its space-based mission in 2025, two forthcoming data releases
+are expected.
+
+Determining absolute parallax measurement is facilitated by the
+observation that stars in different regions of the sky are affected by
+parallax motion with different phases. By imagining two regions
+separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
+tiny changes in relative positions of stars across small and large
+angles. Combining measurements from multiple epochs across several
+years, an absolute all-sky reference frame is derived from which
+parallax and proper motions are derived. In addition to astrometry,
+\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
+and colours from the blue photometer (BP, 330-680 nm) and red photometer
+(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
+BP-RP spectra and radial velocity measurements of bright stars (of
+magnitudes \(G_{\rm RVS} < 16\)) \citep{gaiacollaboration+2016}. For our
+work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
+\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
+\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
+  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
+  change in right ascension, \(\alpha\), and declination, \(\delta\),
+  typically in units of mili-arcsecond (mas) per year.
+  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
+  effects in \(\alpha\).}
+
+\subsection{The Local Group of
+Galaxies}\label{the-local-group-of-galaxies}
+
+The Local Group (LG) is defined as galaxies which are within \(\sim 1\)
+Mpc from the Milky Way-Andromeda system \citep[e.g.,][ and references
+therein]{mcconnachie2012}. \emph{Gaia} has revolutionized LG studies.
+For example, the 6D dynamical measurements and metallicities of MW stars
+led to the (re)discovery of past mergers or Milky Way building blocks
+like \emph{Gaia}-Sausage Enceladus
+\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
+see also][]{mesa+2005}}, out-of-equilibrium structures like the
+\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
+the Milky Way's spiral arms and the bar in the solar neighbourhood
+\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
+helped find and constrain numerous stellar streams
+\citep[\citet{bonaca+price-whelan2025}]{ibata+malhan+martin2019}.
+Altogether, \emph{Gaia} has revealed the hierarchical formation and
+complex, evolving structure of our own Galaxy.
+
+For Milky Way satellites, \emph{Gaia} has enabled well-constrained
+orbital analysis and facilitated precise stellar membership
+determination. Before, proper motions of dwarf galaxies were ill-known
+or undetermined. Few galaxies had precisely measured proper motions,
+more recently from Hubble Space Telescope observations
+\citep[e.g.,][]{sohn+2017}. \emph{Gaia} allowed the first systematic
+determinations of Milky Way satellite proper motions of unprecedented
+precision \citep{pace+li2019, MV2020a}. While the proper motion
+uncertainty of a typical dwarf member star is often large, by combining
+the proper motions of 100s or 1000s of stars from \emph{Gaia}, average
+precise proper motion measurements can be determined, sometimes only
+limited by \emph{Gaia}'s systematic error floor
+\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new
+dynamical era for MW satellite studies, where we can derive precise
+orbits for any satellite, assuming a given MW potential. In addition,
+\emph{Gaia} helps establish membership by separate out contaminating MW
+foreground stars. By measuring parallaxes and/or proper motions, many
+background and foreground stars can be classified as non-members
+\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.
 
@@
 
-\subsection{\texorpdfstring{The era of
-\emph{Gaia}}{The era of Gaia}}\label{the-era-of-gaia}
-
-\emph{(db: these paragraphs start the same\ldots)}
-
-\emph{Gaia} has redefined astrometry, providing photometry, positions,
-proper motions, and parallaxes for over 1 billion stars
-\citep{gaiacollaboration+2021}. Launched in 2013, \emph{Gaia} is a
-space-based, all-sky survey telescope situated at the Sun-Earth L2
-Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
-completed its space-based mission in 2025, two forthcoming data releases
-remain. Determining absolute parallax measurement is facilitated by the
-observation that stars in different regions of the sky are affected by
-parallax motion with different phases. By imagining two regions
-separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
-tiny changes in relative positions of stars across small and large
-angles. Combining measurements from multiple epochs across several
-years, an absolute all-sky reference frame is derived from which
-parallax and proper motions are derived. In addition to astrometry,
-\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
-and colours from the blue photometer (BP, 330-680 nm) and red photometer
-(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
-BP-RP spectra and radial velocity measurements of bright stars
-(\emph{Gaia} radial velocity magnitudes \textless16)
-\citep{gaiacollaboration+2016}. For this work, \emph{Gaia}'s most
-relevant measurement are \(G\) magnitude, \(G_{\rm BP} - G_{\rm RP}\)
-colour, \((\alpha, \delta)\) position, and
-\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
-  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
-  change in right ascension, \(\alpha\), and declination, \(\delta\),
-  typically in units of mili-arcsecond (mas) per year.
-  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
-  effects in \(\alpha\).}
-
-\emph{Gaia} has revolutionized many fields of astronomy including
-studies of the Milky Way and the Local Group. The Local Group is defined
-as galaxies which are approximately bound to the Milky Way-Andromeda
-system, or within about 1 Mpc from the Milky Way \citep[e.g.,][ and
-references therein]{mcconnachie2012}. For example, 6D kinematic
-classification of Milky Way stars led to the discovery of past mergers
-or Milky Way building blocks like \emph{Gaia}-Sausage Enceladus
-\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
-\emph{Gaia} snail {[}e.g., \citet{antoja+2018}, and dynamical effects of
-the Milky Way's spiral arms and the bar in the solar neighbourhood
-\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
-helped find and constrain numerous stellar streams
-\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
-the hierarchical formation and complex, evolving structure of our own
-Galaxy.
-
-For Milky Way satellites, \emph{Gaia} has enabled well-constrained
-orbital analysis and facilitated precise stellar membership
-determination. Before, proper motions of dwarf galaxies were sparse or
-undetermined. Few galaxies had precisely measured proper motions, often
-from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
-\emph{Gaia} allowed for the first systematic determinations of Milky Way
-satellite proper motions of unprecedented precision
-\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
-typical dwarf member star is often large, by combining the proper
-motions for 100s to 1000s of stars in \emph{Gaia}, precise proper motion
-measurements can be determined, sometimes only limited by \emph{Gaia}'s
-systematic error floor \citep[e.g.,][]{MV2020a}. Proper motions have
-ushered in a new dynamical era for MW satellite studies, where we can
-derive precise orbits assuming a given MW potential. In addition,
-\emph{Gaia} helps separate out contaminating MW foreground stars. By
-measuring parallaxes or proper motions, many background and foreground
-stars can be removed as non-members
-\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.
-
 \subsection{Dwarf galaxies today}\label{dwarf-galaxies-today}
@@
 Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
-including dwarf galaxies, ambiguous systems (whose nature as a
-dwarf-like or cluster-like system remains uncertain), and globular
-clusters. Advances in observational techniques has accelerated progress
-in our understanding of dwarf galaxies and introduced new questions.
-Deep digital photometric surveys have more than quadrupled the number of
-known Milky Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing
-surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
-Time \citep{ivezic+2019}, will likely identify even fainter dwarf
+including dwarf galaxies, globular clusters, and ambiguous systems,
+whose nature as a galaxy or star cluster remains uncertain. Advances in
+observational techniques has accelerated progress in our understanding
+of dwarf galaxies and introduced new questions. Deep digital photometric
+surveys have more than quadrupled the number of known Milky Way dwarf
+galaxies since 2005 \citep{simon2019}. Upcoming surveys, like the Vera
+Rubin Observatory's Legacy Survey for Space and Time
+\citep[LSST,][]{ivezic+2019}, will likely identify even fainter dwarf
 galaxies. In addition, large aperture multi-object spectrographs have
 revealed the detailed and complex inner chemodynamical structure of
-dwarf galaxies. Beyond precise structural and kinematic properties of
-dwarf galaxies, modern observations allow for the separation of multiple
-stellar populations, detailed constraints on the dark matter density
-profiles, and hints of tidal disruption or stellar halos.
+dwarf galaxies \textbf{REFS}. Beyond precise structural and kinematic
+properties of dwarf galaxies, modern observations allow for the
+separation of multiple stellar populations, detailed constraints on the
+dark matter density profiles, and hints of tidal disruption or stellar
+halos.
 
 Of local dwarfs, the classical systems still remain the most studied
-galaxies with the best constrained parameters. Because many Milky Way
-dwarfs extend 1-2 degrees across the sky, measurements of some of their
-properties can be challenging \citep[e.g.,][]{mateo1998}. These dwarfs
-also contain large numbers of bright (giant) stars which can be followed
-up spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a
-result, the determination of fundamental properties such as the
-position, size, orientation, distances, proper motions, line-of-sight
-(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
-relatively well constrained today. However, ongoing research continues
-to redefine our understanding of the detailed structure of Milky Way
-satellites, hinting that these objects may be more complex than expected
-at first glance.
+galaxies with the best constrained parameters. These dwarfs also contain
+large numbers of bright (giant) stars which can be followed up
+spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
+the determination of fundamental properties such as the position, size,
+orientation, distances, proper motions, line-of-sight (LOS) velocities,
+and velocity dispersions \(\sigma_\V\), are all relatively well
+constrained today.
 
 Observations of dwarf galaxies have been the origin of several disputes
-or \emph{small-scale} problems for \(\Lambda\)CDM \citep[see
-review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
-the expected number of dwarf galaxies from simulations and the observed
-abundance was known as the \emph{missing satellites problem}.
-Additionally, a number of observations suggested that some dwarf
-galaxies, although not all, possess dark matter ``cores''
-\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019}.\footnote{In
+or \emph{small-scale} problems for
+\LCDM [see reviews by @bullock+boylan-kolchin2017; sales+2022]. For
+example, the mismatch between the expected number of dwarf galaxies from
+simulations and the observed abundance has been known as the
+\emph{missing satellites problem}. Additionally, a number of
+observations suggested that some dwarf galaxies, although not all,
+possess dark matter ``cores,''
+\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019},\footnote{In
   detail, gas-phase rotation curves are better able to differentiate
   between cores and cusps, whereas stellar kinematics is less
-  constraining.} As a result, alternative forms of dark matter have been
-advocated as solutions, such as Warm and Self-Interacting Dark Matter.
+  constraining.} contrary to the expectation from \LCDM of ``cuspy''
+inter dark matter profiles \citep{NFW1996, NFW1997}. As a result,
+alternative forms of dark matter have been advocated as solutions, such
+as Warm or Self-Interacting Dark Matter.
 
@@
 particular have shown that strong feedback can produce dark matter cores
-\citep[e.g.,][]{tollet+2016, fitts+2017, benitez-llambay+2019, orkney+2021}.
-Altogether, the numerous past challenges for \(\Lambda\)CDM in the dwarf
-galaxy regime illustrates the opportunity for dwarf galaxies to test the
+\citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
+Altogether, the numerous past challenges for \LCDM in the dwarf galaxy
+regime illustrate the opportunity for dwarf galaxies to test the
 understanding of galaxy formation and dark matter physics.
 
-\subsection{Typical stellar density
-profiles}\label{typical-stellar-density-profiles}
+\subsection{Dwarf galaxy light profiles}\label{sec:exponential_profiles}
 
-Projected density profiles efficiently characterize the shape of a
-galaxy. At its most basic, density profiles provide properties such as
-the shape, location, size, and orientation of a dwarf galaxy. In
-addition, the details of a stellar density profile can help interpret
-the total mass of dwarf galaxies, their assembly and dynamical history,
-and to understand scaling laws between dwarf galaxy properties
-\citep[e.g.,][\citet{penarrubia+2009}, \citet{querci+2025},
-\citet{lee+2018}]{herrmann+hunter+elmegreen2013}. We note that for
-resolved galaxies, these profiles are in terms of stellar count density
-instead of surface magnitudes.
+Projected luminosity / stellar density profiles efficiently characterize
+the radial structure of a galaxy. At its most basic, light profiles
+synthesize properties such as the shape, location, size, and orientation
+of a dwarf galaxy. In addition, the details of a stellar density profile
+can help interpret a galaxy's assembly and dynamical history
+\citep[e.g.,][, \citet{lee+2018}, \citet{querci+2025}]{penarrubia+2009}.
+We note that for resolved galaxies, these profiles are in terms of
+stellar count density instead of surface brightness.
 
@@
 simplest, defined in terms of the central surface density, \(\Sigma_0\),
-and scale radius, \(R_s\): \begin{equation}{
+and scale radius, \(R_s\):
+\begin{equation}\protect\phantomsection\label{eq:exponential_law}{
 \Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s).
@@
 To fit globular cluster density profiles, \citet{plummer1911} proposed a
-profile based on a polytropic solution\footnote{where density and
-  pressure are assumed to be related by a power law}, \begin{equation}{
-\Sigma_{\rm Plummer} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
+profile based on a self-gravitating polytrope,\footnote{where density
+  and pressure are assumed to be related by a power law}
+\begin{equation}{
+\Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
 }\end{equation}
 
-where \(\Sigma_0\) is the central surface density and \(R_h\) is the 2D
-half-light radius. Now mostly superseded by the King profile for
-globular clusters, the Plummer model is still a good fit to many dwarf
-spheroidals \citep[e.g.,][]{moskowitz+walker2020}.
+where \(\Sigma_0\) is the central surface density and \(R_h\) is the
+projected half-light radius. Now mostly superseded by the King profile
+for globular clusters, the Plummer model is still a good fit to many
+dwarf spheroidals \citep[e.g.,][]{moskowitz+walker2020}.
 
 The \citet{king1962} profile, also an empirical fit to globular
-clusters, is also often applied to dwarf galaxies, especially in older
+clusters, is also used to describe dwarf galaxies, especially in older
 literature. Using three parameters, a core radius \(R_c\), a truncation
 radius \(R_t\), and a characteristic density, \(\Sigma_0\), the King
-profile is \begin{equation}{
-\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
+profile may be written as \begin{equation}{
+\Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
 }\end{equation}
 
-In much of the older literature, \(R_t\) was often called and
-interpreted as a ``tidal radius'', after the similar interpretation for
-globular clusters \citep[e.g.,][\citet{hodge1961}]{IH1995}.
+In much of the older literature, \(R_t\) was often interpreted as a
+``tidal radius'', after an analogous interpretation for globular
+clusters \citep[e.g.,][\citet{hodge1961}]{IH1995}.
 
 Finally, the \citet{sersic1963} profile represents a generalization of
-an exponential profile, and describes most galaxy light profiles well.
-Typically parameterized in terms of a half-light radius \(R_h\), the
-density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\), the
-profile's equation is \begin{equation}{
-\Sigma_{\rm S\acute ersic} = \Sigma_h \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
-}\end{equation} where \(b_n\) is a constant depending on \(n\). \(n=1\)
-provides an exponential profile and \(n=4\) recovers
-\citet{devaucouleurs1948}'s profile for elliptical galaxies. While a
-Sérsic profile is not always used for dwarf galaxies, \citet{munoz+2018}
-advocate for the Sérsic profile since the added flexibility allows for
-more profiles to be fit.
+an exponential profile, and describes most dwarf galaxy light profiles
+well. Typically parameterized in terms of a half-light radius \(R_h\),
+the density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\),
+the profile's equation is \begin{equation}{
+\Sigma_{\rm S} = \Sigma_h \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
+}\end{equation} where \(b_n\) is a constant that depends on \(n\). A
+Sérsic profile with \(n=1\) is equivalent to an exponential profile,
+while \(n=4\) recovers \citet{devaucouleurs1948}'s profile for
+elliptical galaxies. Although a Sérsic profile is not always used for
+dwarf galaxies, \citet{munoz+2018} advocate for the Sérsic profile since
+the added flexibility allows for more profiles to be fit with a single
+law.
 
@@
 demonstrate that an exponential law is a reasonable empirical fit to
-dwarf galaxies, theorizing that dwarf spheroidal may have evolved from
+dwarf galaxies, theorizing that dwarf spheroidals may have evolved from
 exponential disk galaxies and maintained a similar light profile. Later,
 \citet{read+gilmore2005} showed that exponential profiles may originate
-from general mass loss during the evolution of dwarf galaxies. Many
-later photometric works for dwarf spheroidal galaxies have used
-exponential fits, finding that exponential and King profiles both
-provide good descriptions in many cases
+from mass loss during the evolution of dwarf galaxies. Many subsequent
+photometric works for dwarf spheroidal galaxies have used exponential
+fits, finding that exponential and King profiles both provide good
+descriptions in many cases
 \citetext{\citealp[\citet{mateo1998}]{binggeli+sandage+tarenghi1984}; \citealp{mcconnachie+irwin2006}; \citealp{cicuendez+2018}}.
-As a result, many studies assume an exponential density profile for
-dwarf galaxies in theoretical or observational modelling
-\citetext{\citealp[e.g.,][
+As a result, it has become conventional to assume an exponential density
+profile to describe dwarf galaxies in theoretical or observational
+modelling \citetext{\citealp[e.g.,][
 \citet{MV2020a}]{martin+2016}; \citealp{battaglia+2022}; \citealp[ but
@@
 
-An exponential is far from the only density profile advocated for. Some
-authors fitting Sérsic or King profiles (often in addition to
-Exponential), find that the additional parameter somewhat improves fits
-relative to an exponential
-\citep{IH1995, vanzee+barton+skillman2004, munoz+2018, wang+2019}. Note
-that the Sérsic indices are typically \(n \lesssim 1\), implying that
-most galaxies tend to be an exponential or slightly steeper in the outer
-regions.
-
-Beyond Local Group dwarf spheroidals, exponential profiles appear to be
-common, but sometimes with significant modifications. In particular,
-many extragalactic dwarf elliptical and blue compact / Irr galaxies are
-fit well with an exponential + central cusp / nuclear region
+Outside Local Group dwarf spheroidals, exponential profiles are fairly
+common, but sometimes with significant modifications. For example, many
+extragalactic dwarf elliptical and blue compact / Irr galaxies are fit
+well with an exponential + central cusp / nuclear region
 \citep[\citet{noeske+2003}]{caldwell+bothun1987}. In constrast, some
@@
 However, exponential profiles do not fit every dwarf galaxy. Even
-starting from \citet{aparicio+1997}, and \citet[for coma cluster
-dE]{graham+guzman2003}, some people have noted deviations from this
-simple empirical rule. Additionally, \citet{hunter+elmegreen2006};
+starting from \citet{aparicio+1997}, and \citet[for Coma cluster dwarf
+ellipticals]{graham+guzman2003}, some people have noted deviations from
+this simple empirical rule. Additionally, \citet{hunter+elmegreen2006};
 \citet{herrmann+hunter+elmegreen2013};
 \citet{herrmann+hunter+elmegreen2016}; \citet{lee+2018} note that at
-least in a photometric sample of more distant dwarf disky/Irr/blue
-compact dwarfs, that many or most dwarfs show two-part exponential
-profiles. Irr do represent a substantially different class though so it
-is unclear if these conclusions apply. \citet{caldwell+1992}'s
-photometric study of M31 dwarfs show that the outer density profile
-typically fits well to an exponential but with inner deviations. Note
-that while diversity is also commented on here, dwarf galaxies with
-outer flattening profiles are often a minority in these studies.
+least in a photometric sample of more distant dwarf disky/irregular/blue
+compact dwarfs, many or most dwarfs are best fit with two nested
+exponential profiles. Dwarf irregular galaxies do represent a
+substantially different class though so it is unclear how these
+conclusions apply to dwarf spheroidals. \citet{caldwell+1992}'s
+photometric study of M31 dwarfs show that the outer density profile is
+typically well fit by an exponential, although deviations are evident in
+the inner regions.
 
-Alternatively, \citet{moskowitz+walker2020}, use a Plummer profile and
-modified Plummer profiles with steeper outer slopes
-(\(\Gamma = d\log \Sigma / d \log R = -8\) compared to \(\Gamma = -4\)).
+More recently, \citet{moskowitz+walker2020}, have used a Plummer profile
+and modified Plummer profiles with steeper outer slopes
+(\(\Gamma = d\log \Sigma / d \log R = -8\) rather than \(\Gamma = -4\)).
 For most galaxies, they do not have convincing statistical evidence to
 prefer one model over another, but show that for 8 / 15 of dwarfs with
-sufficient data, a plummer or steeper profile is preferred, with only
-the stepper one ever strongly preferred.
+sufficient data, a Plummer or steeper profile is (slightly) preferred.
+\textbf{This sentence is problematic}
 
-Altogether, while there is some natural variation in the density
-profiles of dwarf galaxies, an exponential is an excellent first-order
+Altogether, while there is some variation in the density profiles of
+dwarf galaxies, an exponential is an excellent first-order
 approximation. Typically, deviations from exponentials are in a
-direction of steeper outer cutoff or changes to the inner slope of a
-dwarf galaxy (due to a nuclear region/etc.). Flattened density profiles
-in the outer regions are more unusual. Explaining the origin,
-similarity, and diversity of dwarf galaxy density profiles is a pressing
-question for theories of dwarf galaxy formation and evolution. Whether
-dwarf galaxies are indeed exponential or not, the diversity of density
-profile shapes of dwarf galaxies needs to be explained.
+direction of a steeper outer cutoff or changes to the inner slope of a
+dwarf galaxy (due, for example, to a nuclear star cluster.). Flattened
+density profiles in the outer regions are more unusual. Explaining the
+origin, similarity, and diversity of dwarf galaxy density profiles is a
+pressing question for theories of dwarf galaxy formation and evolution.
+Whether dwarf galaxies are indeed exponential or not, the diversity of
+density profile shapes of dwarf galaxies needs to be explained.
 
-\subsection{Sculptor and Ursa Minor: Hints of tidal
-signatures?}\label{sculptor-and-ursa-minor-hints-of-tidal-signatures}
+\subsection{The extended light profiles of Sculptor and Ursa Minor:
+Hints of tidal
+signatures?}\label{the-extended-light-profiles-of-sculptor-and-ursa-minor-hints-of-tidal-signatures}
 
-Sculptor and Ursa Minor may appear to be typical dwarf galaxies at first
-glance. Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} describe
-the present-day properties of each galaxy. Sculptor, as the first
-discovered classical dSph, is often described as a ``prototypical''
-dSph. There has long been speculation that both Sculptor and Ursa Minor
-may have been influenced by the Milky Way's tidal field (see
-Section~\ref{sec:discussion}). \citet{sestito+2023a};
-\citet{sestito+2023b} report that a ``kink'' in the density profile,
-beginning around 30 arcmin for both Sculptor and Ursa Minor. They
+Sculptor and Ursa Minor appear to be typical dwarf spheroidal galaxies
+at first glance. Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}
+describe the present-day properties of each galaxy. Sculptor, as the
+first discovered classical dSph, is often described as a
+``prototypical'' dSph. There has long been speculation that both
+Sculptor and Ursa Minor may have been influenced by the Milky Way's
+tidal field (see Section~\ref{sec:discussion}). \citet{sestito+2023a};
+\citet{sestito+2023b} report a ``kink'' in the density profile,
+beginning around 30 arcmin for both Sculptor and Ursa Minor, that they
+interpret as caused by the effects of Galactic tides. They
 spectroscopically follow up some of the most distant stars, finding
 multiple members between 6-12 half-light radii from the centre of each
-dwarf. If dwarfs had exponential profiles like Fornax, then these stars
+dwarf. If dwarfs had exponential profiles, like Fornax, then these stars
 should be much rarer.
@@
 profile. The left panel of Fig.~\ref{fig:scl_umi_vs_penarrubia} shows
-our density profiles of Sculptor, Ursa Minor, and Fornax (see
-Section~\ref{sec:observations} for our methods). Compared to Fornax,
-both Sculptor and Ursa Minor show an excess of stars beginning around
-\(\log R/R_h\approx 0.4\). The right panel of
-Fig.~\ref{fig:scl_umi_vs_penarrubia} compares the same density profiles
-to an initial and final density profile from a simple tidal model of an
-exponential dwarf spheroidal galaxy embedded in a NFW halo in the Milky
-Way field, described in Section~\ref{sec:tidal_theory}. Fornax agrees
-well with the exponential initial conditions. However, Sculptor and Ursa
-Minor are better described by the more extended, tidally evolved density
-profile.
+the density profiles of Sculptor, Ursa Minor, and Fornax (see
+Section~\ref{sec:observations} for details on how these profiles are
+measured). Compared to Fornax, both Sculptor and Ursa Minor show an
+excess of stars beginning around \(\log R/R_h\approx 0.4\). The right
+panel of Fig.~\ref{fig:scl_umi_vs_penarrubia} compares the same density
+profiles to an initial and final density profile from a simple tidal
+model of an exponential dwarf spheroidal galaxy embedded in a NFW halo
+in the Milky Way field, described in Section~\ref{sec:tidal_theory}
+(\textbf{Move to later or describe?}). Fornax agrees well with the
+exponential initial conditions. However, Sculptor and Ursa Minor are
+better described by the more extended, tidally evolved density profile.
 
-A goal of this work is to determine, assuming \(\Lambda\)CMD, if the
-effects of the Milky Way (or other satellites) may indeed create
-observable tidal signatures in Sculptor and Ursa Minor. If tides cannot
-explain these features, these features may instead be a extended stellar
-``halo'' or second component of the galaxy---suggestive of a complex
-history or formation in each galaxy, which needs to be explained by
-galaxy formation models of Local Group dwarf galaxies.
+A goal of this work is to determine, assuming \LCDM, if the effects of
+the Milky Way (or other satellites) are indeed responsible for the
+extended outer light profiles of Sculptor and Ursa Minor. If tides
+cannot explain these features, these features may instead be due to an
+extended stellar ``halo'' or second component of the galaxy---suggestive
+of a complex star formation or accretion history.
 
@@
 idealized model with initial, final, and break radius as a dotted line,
-solid line and an arrow respectively see
-\ref{sec:break_radii}.}\label{fig:scl_umi_vs_penarrubia}
+solid line and an arrow respectively see \ref{sec:break_radii}.
+\textbf{Move RHS to later\ldots{}}}\label{fig:scl_umi_vs_penarrubia}
 \end{figure}
@@
 
-\section{Cosmological context}\label{cosmological-context}
-
-We only understand a tiny fraction of the universe's composition. The
-leading theory of cosmology, \(\Lambda\)CDM (Lambda Cold Dark Matter),
-posits that the universe is composed of about 68\% dark energy
-(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
-  like to change definitions of words. \emph{Baryons} here means
-  baryons+leptons, i.e.~any standard model massive fermion.}
-\citep{planckcollaboration+2020}. While the composition of dark matter
-and dark energy remains elusive, we know their general properties. Dark
-energy causes the acceleration of the expansion of the universe on large
-scales. We do not discuss dark energy here---it does not substantially
-affect the Local Group today. Dark matter, instead, makes up the vast
-majority of mass in galaxies. Typically, galaxies have baryonic to dark
-matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
-In \(\Lambda\)CMD, dark matter is assumed to interact only
-gravitationally. Light passes through dark matter unimpeded---in this
-sense dark matter is transparent. Dark matter is also commonly assumed
-to be \emph{cold}, i.e.~typical velocities much smaller than the speed
-of light in the early universe. Implications of dark matter properties
-range from cosmological structural formation, galaxy structure, and
-galaxy interactions.
-
-\subsection{Structure formation and dwarf
-galaxies}\label{structure-formation-and-dwarf-galaxies}
-
-The very early universe was almost featureless. Our earliest
-observations of the universe stem from the cosmic microwave background
-(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
-But tiny perturbations in the CMB, temperature fluctuations of 1 part in
-10,000, reveal the underlying seeds of large-scale cosmological
-structure.\footnote{The power spectrum from the CMB is complicated by
-  baryons, which lead to oscillations in power based on the
-  sound-crossing timescale.} Governed by cosmological expansion,
-gravitational collapse, and baryonic physics, each overdensity grows
-into larger structures. Initially, baryonic matter was coupled to
-radiation and resisted collapse. Dark matter, only influenced by gravity
-instead, freely collapsed into the first structures. For mass
-perturbations both smaller than the horizon and overdense enough to
-collapse, these overdensities of dark matter become self-gravitating,
-known as \emph{halos}. After recombination, where electrons combined
-with atomic nuclei to form atoms, baryons decoupled from radiation and
-fell into the dark matter halos. The densest pockets of baryons later
-formed the first stars and galaxies.
-
-Dark matter halos, and their associated galaxies, rarely evolve in
-isolation. Instead, structure formation is \emph{hierarchical}. Small
-dark matter halos collapse first and hierarchically merge into
-progressively larger halos
-\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
-Structure formation happens on a wide range of scales. The largest
-structures become the cosmic web---composed of voids, filaments, and
-clusters---and the smallest structures directly observable host dwarf
-galaxies. Hierarchical assembly is evident through the large scale
-structure of the universe, remnants of past mergers within the Milky
-Way, and tidal disruption of dwarf galaxies and their streams around
-nearby galaxies.
-
-Small-scale structure formation is sensitive to deviations from
-\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. One key
-prediction of \(\Lambda\)CDM is that mass perturbations are expected to
-exist on all scales, and are largest on the smallest scales, so we would
-expect the formation of halos on all scales. Many alternative models,
-such as warm dark matter or self-interacting dark matter, may smooth out
-small-scale features and reduce the abundance of small halos or change
-their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies, which
-occupy the smallest dark matter halos, are promising windows into
-small-scale cosmological features. Alternative dark matter properties
-can influence dwarf galaxy abundance, formation, structure, and tidal
-evolution. The nearby dwarf galaxies of the Local Group, with detailed
-observations, present a promising opportunity to understand the
-evolution of these objects and test our understanding of cosmology. To
-understand the evolution of these objects, we need to understand the
-general predictions of \(\Lambda\)CDM for the properties of dwarf
-galaxies.
-
-\emph{Warm} dark matter, relativistic in the early universe but cooler
-now, smooths out small-scale features and softens the cusps of dark
-matter halos. \emph{Self-interacting} dark matter instead can form cores
-but also the cores can collapse into a density peak. \textbf{TODO:
-incorporate this here.}
-
-\subsection{Typical structure of dark matter
-halos}\label{typical-structure-of-dark-matter-halos}
-
-In \(\Lambda\)CDM cosmological simulations, dark matter halos are
-remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
-authors observe that the spherically-averaged density profiles
-\(\rho(r)\) are universally well described by a two-parameter law:
-\begin{equation}\protect\phantomsection\label{eq:nfw}{
-\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
-}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
-density . This profile has shown remarkable success in describing
-\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
-profiles are \emph{cuspy}, where the density continuously rises like
-\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
-density profile increases gradually with radius, and at large radii the
-density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
-shows an example of a NFW halo.
-
-The total mass of an NFW profile diverges, so halos are commonly defined
-using an overdensity criterion. The virial mass, \(M_{200}\), is defined
-as the mass within a radius \(r_{200}\) containing a mean enclosed
-density 200 times\footnote{For the collapse of a uniform spherical
-  density, the virialized overdensity would be
-  \(\Delta = 18\pi^2\approx 178\) for a critical universe
-  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
-  this parameter may be closer to \(\Delta \approx 100\) for our
-  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
-  from][]{bryan+norman1998}.} the critical density of the universe:
-\begin{equation}{
-M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
-}\end{equation} A standard second parameter is the halo concentration,
-\(c=r_{200} / r_s\) describing how the characteristic size of the halo
-compares to its virial radius. In this case, the scale density is a
-function of \(c\) alone,
-\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
-\citep{NFW1996}. However, we characterize halos by their circular
-velocity profiles. The circular velocity,
-\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
-\(v_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
-\(v_{\rm max}\) is related to both the total halo mass and the observed
-line of sight velocity dispersion, and \(r_{\rm max}\) relates to the
-scale radius.
-
-While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
-these are not independent. Lower mass dark matter halos often collapse
-earlier, when the universe was denser. As a result, low mass subhalos
-tend to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
-between \(M_{200}\) and c, or the mass-concentration relation, describes
-the mean trend of concentration with mass
-\citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
-Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
-relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
-a general trend for concentration to increase with mass, the
-concentration values still have substantial scatter. Other parameters
-such as the halo spin or shape may affect the scatter of the
-mass-concentration relation, but their effect is typically expected to
-be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
-\caption[Example density profiles]{Density profiles in log density
-versus log radius for the stars and dark matter of a Sculptor-like
-galaxy. The dark matter is more extended and massive than the star
-across the entire galaxy---dynamical evolution is driven by the dark
-matter mass distribution.}\label{fig:nfw_density}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
-\caption[Stellar-mass halo-mass relation]{\textbf{Left} Radius of
-maximum circular velocity \(r_{\rm max}\) as a function of maximum
-circular velocity. The solid line with 1-\(\sigma\) shaded region and
-dashed line are the relations from \citet{ludlow+2016} for \(z=0\) and
-\(z=2\) respectively. \textbf{Right} Stellar mass (top) as a function of
-maximum circular velocity. The solid line with the 1-\(\sigma\) shaded
-region is the relation from \citet{fattahi+2018} with scatter points
-simulated central galaxies from APOSTLE in
-\citet{fattahi+2018}.}\label{fig:smhm}
-\end{figure}
-
-\subsection{Connecting dark matter to
-stars}\label{connecting-dark-matter-to-stars}
-
-Unfortunately, directly observing dark matter remains infeasible (lest
-dark matter wouldn't be dark). Fortunately, dwarf spheroidal galaxies
-come with a convenient population of visible stars. Stars typically
-represent a small fraction of the total mass of a dwarf galaxy. For
-example, Fig.~\ref{fig:nfw_density} shows the typical density profile
-for the stars and dark matter for a Sculptor-like galaxy. At all radii,
-dark matter are at least \textasciitilde10 times more dense, and the
-total virial mass is often \textasciitilde1000 times higher than the
-stellar mass alone. As a result, the dynamics and formation of dwarf
-galaxies is governed by the underlying dark matter halo.
-
-Both cosmology and observations find fundamental relationships between
-the dynamical or total mass and stellar masses of
-galaxies.\textbf{Improve last sentence.} A prediction of cosmological
-simulations is the Stellar mass Halo Mass relation (SMHM), describing
-the mass of stars which form in a halo of a given size. In particular,
-the SMHM relation grows especially steep in the dwarf galaxy
-regime---many dwarf galaxies are formed in halos of similar mass.
-Fig.~\ref{fig:smhm} shows the stellar mass versus \(v_{\rm max}\)
-maximum circular velocity (proxy for halo mass) for APOSTLE dwarfs from
-\citet{fattahi+2018}. The APOSTLE project \citep{sawala+2016} uses the
-hydrodynamical setup from the EAGLE project
-\citep{crain+2015, schaye+2015} to simulate Local Group analogues in a
-\(\Lambda\)CDM cosmological context (db: reorder these sentences). While
-there is some scatter, the range of predicted \(v_{\rm max}\) is fairly
-narrow across \(\sim 4\) decades in stellar mass. Because lower mass
-galaxies have increasingly shallow potential wells, feedback becomes
-more effective at removing gas. Re-ionization additionally suppresses
-late star formation in the faintest galaxies. As a result, the resulting
-stellar mass of a galaxy is highly sensitive to the dark matter mass,
-especially for faint dwarf galaxies. If we know the initial stellar mass
-of a dwarf galaxy, we also know, at some level, the properties of its
-dark matter halo.
-
-Several challenges complicate a simple SMHM trend including environment,
-assembly history, and tidal effects. Additionally, the detailed physics
-of galaxy formation can impact the SMHM relationship. By being closer to
-a massive host, many dwarfs quench earlier, resulting in lower stellar
-masses \citep[e.g.,][]{christensen+2024}. In particular, effects like
-ram-pressure stripping (removal of gas in the dwarf galaxy due to
-pressure from host's circumgalactic medium) and tidal removal of gas
-cause star formation to quench {[}REFS{]}. Additionally, the time of
-formation (relative to reionization) can influence the resulting stellar
-content \citep{kim+2024}. Finally, tides influence both the dark matter
-and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
-Consequently, tides may reduce the halo mass more than the stellar mass,
-adding additional scatter to the SMHM trend, particularly for low-mass
-satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
-tides on Local Group dwarf galaxies is essential to determine where and
-how these galaxies formed in a cosmological context.
-
-\section{Simulating tidal effects}\label{sec:tidal_theory}
-
-Since the discovery of dwarf galaxies, a large body of work has either
-considered or simulated tidal effects. While pre-\emph{Gaia} work often
-had few constraints on orbits of dwarf galaxies, the theory of tidal
-mass loss remains largely the same.
+\section{Interpreting tidal signatures}\label{sec:tidal_theory}
 
 Simulating dwarf galaxies accurately in a cosmological context remains a
-substantial challenge. Currently, cosmological simulations can predict
-the overall abundance of larger mass? dwarf galaxies {[}e.g., {]} and
-broadly predict effects of tides \citep[e.g.,][]{riley+2024}. But, dwarf
-galaxies are often near the resolution limit. Insufficient resolution
-can lead to artificial disruption of dwarf galaxies and overestimation
-of tidal effects \citep[e.g.,][]{santos-santos+2025}. For example, the
-highest resolution cosmological simulation of a Milky Way-size dark
-matter halo, the Aquarius project \citep{springel+2008}, achieved a DM
-resolution of \(1.712 \times10^3 \Mo\), enough to barely resolve the
-stellar components of Sculptor like halos (see chapter
-\ref{sec:methods}). To address this challenge, idealized simulations
-only simulate a single subhalo in an approximate host potential. As a
-result, idealized simulations can reach excellent numerical convergence
-at the cost of realism. For instance, our simulations later are about 3x
-higher resolution than Aquarius at a fraction of the computational cost
-(400x less particles). Idealized simulations make numerous
-simplifications: neglecting mergers, cosmological context and evolution,
-mass assembly, and often baryonic physics. We use idealized simulations
-here which are more powerful in accurately assessing tidal effects after
-infall, given that the idealizations do not impact these galaxies's
-recent history too much.
+substantial challenge. As discussed above, cosmological simulations can
+predict the overall abundance of the most massive dwarf galaxies
+{[}e.g., \textbf{references}{]} and broadly examine the effects of tides
+\citep[e.g.,][]{riley+2024}. However, dwarf galaxies are often near the
+resolution limit. Insufficient resolution can lead to artificial
+disruption of dwarf galaxies and overestimation of tidal effects
+\citep[e.g.,][]{santos-santos+2025}. To address this challenge,
+idealized simulations are often used to simulate a single subhalo in an
+approximate host potential. As a result, idealized simulations can reach
+excellent numerical convergence For instance, our simulations we
+describe later reach 3x higher resolution than Aquarius
+\citep{springel+2008} at a fraction of the computational cost (400x
+fewer particles). Idealized simulations make numerous simplifications:
+neglecting mergers, cosmological context and evolution, mass assembly,
+and often baryonic physics. We use idealized simulations here to assess
+tidal effects after infall.
 
-Some of the earliest theory work on tidal mass loss of dwarf galaxies
-originate from \citet{oh+lin+aarseth1995}; \citet{piatek+pryor1995};
-\citet{moore+davis1994}; \citet{johnston+spergel+hernquist1995}.
-Already, these works used techniques similar to what we continue to use
-today and laid the foundation for understanding tidal effects. While the
-detailed assumptions have evolved (e.g.~no longer assuming
-mass-follows-light), many of their conclusions still hold true. More
-recent work expanding this theory include \citet{read+2006};
-\citet{bullock+johnston2005}; \citet{PNM2008}; \citet{penarrubia+2009};
-\citet{klimentowski+2009}; \citet{errani+2023a}; \citet{fattahi+2018};
-\citet{stucker+2023}; \citet{wang+2017}.
+Some of the earliest simulation work on tidal mass loss of dwarf
+galaxies originate from \citet{oh+lin+aarseth1995};
+\citet{piatek+pryor1995}; \citet{moore+davis1994};
+\citet{johnston+spergel+hernquist1995}. Already, these works used
+techniques similar to what we continue to use today and laid the
+foundation for understanding tidal effects. While the detailed
+assumptions have evolved (e.g.~no longer assuming mass-follows-light),
+many of their conclusions still hold true. More recent work expanding
+this theory include \citet{read+2006}; \citet{bullock+johnston2005};
+\citet{PNM2008}; \citet{penarrubia+2009}; \citet{klimentowski+2009};
+\citet{errani+2023a}; \citet{fattahi+2018}; \citet{stucker+2023};
+\citet{wang+2017}.
 
@@
 
-\textbf{\emph{These past three paragraphs need revisions\ldots{}}}
+*\textbf{Finish last two paragraphs}
 
-\subsection{Simulating large gravitational systems: The N-body
-method}\label{simulating-large-gravitational-systems-the-n-body-method}
+\subsection{A simple tidal simulation}\label{a-simple-tidal-simulation}
 
-Modelling gravitational evolution for large systems requires special
-methods. Perhaps the simplest method to compute the evolution of dark
-matter is through \emph{N-body simulations}. A dark matter halo is
-represented as a large number of dark matter particles (bodies). Each
-body is essentially a Monte Carlo sample of the underlying phase-space
-distribution. Note that dark matter (and galaxies) are often assumed to
-be \emph{collisionless}---particles are not strongly affected by close,
-\emph{collisional} gravitational encounters which substantially change
-the momenta of involved bodies. In contrast, star clusters are often
-collisional so neglecting these encounters may not be a reasonable
-approximation. While we use individual gravitating bodies in N-body
-simulations, the Newtonian gravitational force is softened to be a
-Plummer sphere as to limit strongly collisional encounters.
+To illustrate the effects of tides on an NFW halo in a Milky-Way like
+potential, we consider a toy model. See also \citet{PNM2008}?.
 
-Naively, the Newtonian gravitational force requires adding together the
-forces from each particle on each particle, with a computational cost
-that scales quadratically with the number of particles, or \(O(N^2)\).
-With this method, simulating a large number of particles, such as
-\(10^6\), would require \(10^{12}\) force evaluations at each time step,
-making cosmological and high-resolution studies unfeasible. However,
-only long-range gravitational interactions tend to be important for CMD,
-so we can utilize the \emph{tree method} to compute the gravitational
-force vastly more efficiently.
+We evolve an NFW profile in a logarithmic host potential along an orbit
+where the pericentre to apocentre is 1:50.
 
-The first gravitational tree code was introduced in
-\citet{barnes+hut1986}, and is still in use today. We utilize the
-massively parallel code \emph{Gadget 4} \citep{gadget4}. Particles are
-spatially split into an \emph{octotree}. The tree construction stars
-with one large node, a box containing all of the particles. If there is
-more than one particle in a box/node , the box is then divided into 8
-more nodes (halving the side length in each dimension) and this step is
-repeated until each node only contains 1 particle. With this
-heirarchichal organization, if a particle is sufficiently far away from
-a node, then the force is well approximated by the force from the centre
-of mass of the node. As such, each force calculation only requires a
-walk through the tree, only descending farther into the tree as
-necessary to retain accuracy. The total force calculations reduce from
-\(O(N^2)\) to \(O(N\,\log N)\), representing orders of magnitude
-speedup. Modern codes such as \emph{Gadget} utilize other performance
-tricks, such as splitting particles across many supercomputer nodes,
-efficient memory storage, adaptive time stepping, and parallel file
-writing to retain fast performance for large scale simulations, forming
-the foundation for many cosmological simulation codes.
-
-\subsection{Tidal evolution}\label{tidal-evolution}
-
-A galaxy in equilibrium will remain in equilibrium, unless acted upon by
-an external force. As an example, Fig.~\ref{fig:lagrange_points}
-illustrates the effective potential around a Sculptor-like NFW galaxy in
-a circular orbit around a Milky Way-like galaxy. The effective potential
-accounts for the centrifugal force in the rotating frame. There are two
-key saddle points in Fig.~\ref{fig:lagrange_points}, labeled \(L_1\) and
-\(L_2\) (as Lagrange points, there are another 3 not as relevant). Since
-a particles trajectory is constrained to be within the equipotential
-surface equal to the particle's total energy, the easiest (requiring the
-lowest energy) path for a particle to escape the satellites
-gravitational influence is through the windows around \(L_1\) and
-\(L_2\). Otherwise, the potential steeply increases. \(L_1\) and \(L_2\)
-are co-linear with the satellite and host-origin. In general, particles
-which have higher energy than the energy at \(L_1\) and \(L_2\) are most
-commonly unbound from the galaxy.
-
-The full evolution of a dwarf galaxy in a tidal follows an approximate
-sequence (although somewhat concurrent as well.)
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Mass is lost}. In particular, particles and stars on weakly
-  bound orbits are most likely to be removed by tides. Particles escape
-  through \(L_1\) and \(L_2\)
-\item
-  \emph{Steams form}. Unbound mass becomes part of a stream or tidal
-  tails. These particles follow similar orbits but are slightly higher
-  or lower energy (depending on which side the particle escaped from).
-\item
-  \emph{Mass redistributes}. Bound mass of the galaxy redistributes. As
-  mentioned in the last section, this is visible as a wave of outward
-  moving material, with the outermost material reaching equilibrium
-  last.
-\item
-  \emph{A new equilibrium}. With mass loss, the gravitational potential
-  decreases, resulting in a more compact dark matter halo and stars
-  which adiabadically expand to a larger scale radius.
-\end{enumerate}
-
-A tidally affected galaxy contains predictable observational clues to
-this process. Nearby to the galaxy, the density profile is flattened as
-a result of newly unbound material. Further away from the galaxy, this
-would be visible as a stream. Additionally, a satellite stream contains
-a velocity gradient along the path. Finally, due to mass loss, a
-satellite will have a reduced velocity dispersion, larger stellar size,
-and
-
-Dwarf galaxies evolve along \emph{tidal tracks}. From N-body simulations
-across a numerous initial conditions and orbits, NFW halos evolve along
-a narrow track in terms of maximum circular velocity and radius
-\(v_{\rm max}\) and \(r_{\rm max}\) . \citet{EN2021} derive an empirical
-fit to these tidal tracks, finding that
-
-\begin{equation}{
-\frac{v_{\rm max}}{v_{\rm max, 0}} = 
-2^\alpha 
-\left(\frac{r_{\rm max}}{r_{\rm max, 0}}\right)^{\beta}\left[1 + \left(r_{\rm max} / r_{\rm max, 0}\right)^2\right]^{-\beta},
-}\end{equation} where \(\alpha=0.4\) and \(\beta=0.65\). As illustrated
-in Fig.~\ref{fig:tidal_tracks}, this formula works for both circular and
-elliptical orbits and is independent of the initial subhalo size or
-distance to the host. \textbf{\emph{Rewrite this paragrah}}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{./figures/lagrange_points.pdf}}
-\caption[Lagrange points]{The potential contours for a 2-body galaxy
-system in the rotating frame. \(L_1\) and and \(L_2\) is most efficient
-as the required energy (the effective potential) is the
-smallest.}\label{fig:lagrange_points}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715095615423.png}
-\caption[Tidal tracks of dwarf galaxies]{Tidal tracks of dwarf galaxies,
-the logarithm of maximum circular velocity and radius of relative to the
-initial conditions for satellites on a variety of orbits. Almost all
-satellites follow the tidal track suggested by \citet{EN2021}. Adapted
-from fig.~6 of \citet{EN2021}.}\label{fig:tidal_tracks}
-\end{figure}
-
-\subsection{Break and tidal radii}\label{sec:break_radii}
-
-Following the detailed discussion of tidal evolution, we can interpret
-two simple analytic approximations of where tidal effects should be
-visible.
-
-The \textbf{Jacobi radius} represents the approximate radius where stars
-become unbound for a galaxy in a circular orbit around a host galaxy.
-Calculated from an approximation of the location of the \(L_1\) and
-\(L_2\) Lagrange points, the Jacobi radius is where the mean density of
-the dwarf galaxy is roughly three times the mean interior density of the
-host galaxy at pericentre, or \begin{equation}{
-3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J).
-}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
-we should expect to find relatively clear signs of tidal disturbance.
-While strictly valid for circular orbits, assuming \(r_{\rm peri}\) for
-the host-dwarf distance works as most stars are lost near pericentre.
-
-We also use the \textbf{break radius} as defined in
-\citet{penarrubia+2009}, marking where the galaxy is still in
-disequilibrium. The break radius \(r_{\rm break}\) is proportional to
-the velocity dispersion ,\(\sigma_v\), and time elapsed since pericentre
-,\(\Delta t\), \begin{equation}{
-r_{\rm break} = C\,\sigma_{v}\,\Delta t
-}\end{equation} where the scaling constant \(C \approx 0.55\) is derived
-empirically. \(r_{\rm break}\) describes where the dynamical timescale
-is longer than the time since the perturbation, i.e.~the radius within
-which the galaxy should have dynamically relaxed. As illustrated in
-Fig.~\ref{fig:idealized_break_radius}, for an idealized model with
-exponential stars in a NFW halo, shortly after pericentre, the stellar
-component is smooth but contains a change in slope around
-\(r_{\rm break}\). This radius is visible in the stellar distribution as
-non-spherical S-shaped overdensities of stars. Also, \(r_{\rm break}\)
-is where where the mean radial velocities of the stars becomes positive,
-i.e.~the system is out of equilibrium and adjusting to the new density
-profile.
+As illustrated in Fig.~\ref{fig:idealized_break_radius}, for an
+idealized model with exponential stars in a NFW halo, shortly after
+pericentre, the stellar component is smooth but contains a change in
+slope around \(r_{\rm break}\). This radius is visible in the stellar
+distribution as non-spherical S-shaped overdensities of stars. Also,
+\(r_{\rm break}\) is where where the mean radial velocities of the stars
+becomes positive, i.e.~the system is out of equilibrium and adjusting to
+the new density profile.
 
@@
 3rd pericentre. See Section~\ref{sec:methods} for a description of our
-simulation setup. \emph{Note to self:} This figure may be too complex,
-not sure the best way to present everything
-yet.}\label{fig:idealized_break_radius}
+simulation setup.}\label{fig:idealized_break_radius}
 \end{figure}
+
+\subsection{Tidal and ``break'' radii}\label{sec:break_radii}
+
+For a given orbit in a given potential, there are characteristic radii
+which help gauge the effects of tides on a dwarf galaxy model.
+
+The \textbf{Jacobi radius} represents the approximate radius where stars
+become unbound for a galaxy in a circular orbit around a host galaxy.
+Calculated from an approximation of the location of the \(L_1\) and
+\(L_2\) Lagrange points, the Jacobi radius is where the mean density of
+the dwarf galaxy is roughly three times the mean interior density of the
+host galaxy at pericentre, or \begin{equation}{
+3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J).
+}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
+we should expect to find relatively clear signs of tidal disturbance.
+While strictly valid for circular orbits, assuming \(r_{\rm peri}\) for
+the host-dwarf distance works as most stars are lost near pericentre.
+
+We also use the \textbf{break radius} as defined in
+\citet{penarrubia+2009}, marking where the galaxy is still in
+disequilibrium after pericentre in a highly-eccentric orbit. The break
+radius \(r_{\rm break}\) is proportional to the velocity dispersion
+,\(\sigma_v\), and time elapsed since pericentre ,\(\Delta t\),
+\begin{equation}\protect\phantomsection\label{eq:r_break}{
+r_{\rm break} = C\,\sigma_{v}\,\Delta t
+}\end{equation} where the scaling constant \(C \approx 0.55\) was
+derived empirically. \(r_{\rm break}\) describes where the dynamical
+timescale is longer than the time since the perturbation, i.e.~the
+radius within which the galaxy has had enough time to dynamically relax.
 

### 2025-08-29 thesis/main.tex
thesis updates and completed test fbest notebook
updated structure and table shortcaptions
thesis w full ch2 draft
major thesis restructure
thesis julio ch1 edits
julio edits at durham

@@
 
-\title{Timid tides or tall tails? \\
-    Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies
+\title{
+    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?
 }
@@
 
-\noindent Dr. Kim Venn, Co-Supervisor\\
+\noindent Dr. Kim Venn\\
 Department of Astronomy\\
@@
 
-\chapter{Introduction}
+\chapter{Introduction}\label{sec:introduction}
 \input{introduction}
 
-\chapter{Data Selection}
+\chapter{The Light Profiles of the Classical Dwarfs}\label{sec:observations}
 \input{data}
 
-\chapter{Methods}
+\chapter{Simulation Methods}\label{sec:methods}
 \input{methods}
+
+\chapter{The Effects of Tides}\label{sec:results}
+\input{sculptor}
+
+%\chapter{The effects of tides on the Ursa Minor Dwarf Spheriodal Galaxy}
+%\input{ursa_minor}
+
+\chapter{Discussion}\label{sec:discussion}
+\input{discussion}
+
+\chapter{Summary and Outlook}\label{sec:summary}
 
@@
 \appendix
-% \chapter*{Appendix}
-% \addcontentsline{toc}{chapter}{Appendix}
-% \renewcommand{\thesection}{A.\arabic{section}}
-% \renewcommand\thefigure{A.\arabic{figure}}    
-% \renewcommand\theequation{A.\arabic{equation}}    
-% \setcounter{figure}{0}
-% \setcounter{equation}{0}
-% 
-% \renewcommand*{\theHsection}{A.\arabic{section}}%
-% \renewcommand*{\theHsubsection}{A.\arabic{subsection}}%
-% \renewcommand*{\theHfigure}{A.\arabic{figure}}%
-% \renewcommand*{\theHequation}{A.\arabic{equation}}%
 
-\chapter{Model parameters}
-
+\input{appendix}
+\input{rv_models}
 \input{numerical_convergence}
 
-⁄\end{document}
+\end{document}

### 2025-08-29 thesis/methods.tex
julio edits at durham

@@
-In this section, we discuss our methods used for the following two
-chapters. We utilize an analytic, static, axisymmetric Milky Way
-potential based on \citet{mcmillan2011}. Our N-body simulation assume
-collisionless, cold dark matter and are integrated with Gadget-4. We
-check for numerical convergence and stability of initial conditions in
-isolation.
+In the previous chapters, we have seen that the Scl and UMi classical
+dSphs have outer density profiles that appear to deviate from the
+exponential law that approximates well all other classical dSphs. Our
+main intension is to assess whether such excesses result from the
+effects of Galactic tides. To this purpose, we intend to use N-body
+simulations of the evolution of CDM halos in a Galactic potential,
+constrained to have the orbital parameters consistent with a dwarf's
+present-day position and velocity. We shall assume that the Galactic
+potential is the static, analytic potential inferred by
+\citet{mcmillan+2011} from observations of kinematic tracers. We also
+assume the potential of each dwarf may be initially approximated by a
+cuspy NFW profile. Since the dwarfs in question are heavily dark matter
+dominated, we shall use a carefully selected sample of dark matter
+particles to mimic and track the evolution of an embedded tracer stellar
+component. In this Chapter, I describe the Galactic potential used, the
+orbital estimation method, the initial conditions setup, and the N-body
+method used.
 
-\section{Milky Way potential}\label{milky-way-potential}
+\section{Orbital estimation}\label{orbital-estimation}
 
-We adopt the Milky Way potential described in \citet{EP2020}, an
-analytic approximation of \citet{mcmillan2011}.
+To choose the possible orbits of a dwarf galaxy, we perform a Monte
+Carlo sampling of the present-day observables. The present-day position,
+distance modulus, LOS velocity, and proper motions are each sampled from
+normal distributions given the reported uncertainties in
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
+each sampled observable back in time for 10 Gyr using \agama [@agama].
+Dynamical is not expected to impact orbits substantially because of the
+low masses of the dwarfs, and because the dwarfs orbit mainly in the
+periphery of the Galaxy. When selecting the initial position and
+velocity of an N-body model, we select the position and velocity of the
+first apocentre occurring 10 Gyr ago.
+
+To convert from Gaia to Galactocentric coordinates, we use the Astropy
+v4 Galactocentric frame \citep{astropycollaboration+2022}. This frame
+assumes the Galactic centre is at position
+\(\alpha = {\rm 17h\,45m\,37.224s}\),
+\(\delta = -28^\circ\,56'\,10.23''\) with proper motions
+\(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
+\(\mu_\delta=-5.547\pm0.026 \masyr\) (from the appendix and Table 2 of
+\citet{reid+brunthaler2004}). The Galactic centre distance from the Sun
+is \(8.122\pm0.033\,\)kpc with a radial velocity =
+\(11 + 1.9 \pm 3\,\kms\), from \citet{gravitycollaboration+2018}.
+Finally, adding that the Sun is \(20.8\pm0.3\,\)pc above the disk from
+\citet{bennett+bovy2019}, and using the procedure outlined in
+\citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
+rest frame is
+\(\v_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
+uncertainties in the reference frame are typically smaller than the
+uncertainties on a dwarf galaxy's position and velocity.
+
+\subsection{Milky Way Potential}\label{milky-way-potential}
+
+We adopt the Milky Way potential described in \citet{EP2020}, which is
+an analytic approximation to that proposed by \citet{mcmillan2011}.
 Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
 each component and the total circular velocity profile for our fiducial
-profile. The potential consisting of a stellar bulge, a thin and thick
-disk, and a dark matter NFW halo. Among typical orbits for Sculptor and
-Ursa Minor, the dark matter halo dominates the potential.
+profile. The potential includes a stellar bulge, a thin and thick disk,
+and a dark matter NFW halo.
 
-The potential is specified as follows. The galactic bulge is described
-by a \citet{hernquist1990} potential,
+The galactic bulge is described by a \citet{hernquist1990} potential,
 
-\begin{equation}\protect\phantomsection\label{eq:hernquist}{
+\begin{equation}{
 \Phi(r) = - \frac{GM}{r + a},
@@
 
-\begin{equation}\protect\phantomsection\label{eq:mn75}{
+\begin{equation}{
 \Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
@@
 The halo is a NFW dark matter halo (Eq.~\ref{eq:nfw}) with
-\(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.
+\(r_{\rm max} = 43.7\,\)kpc and \(\V_{\rm max} = 191\,\kms\).
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
 \caption[Circular velocity of potential]{Circular velocity profile of
@@
 
-\section{Orbital estimation}\label{orbital-estimation}
+\subsection{Orbits of Sculptor}\label{orbits-of-sculptor}
 
-To estimate the possible orbits of a dwarf galaxy, we perform a Monte
-Carlo sampling of the present-day observables. The present-day position,
-distance modulus, LOS velocity, and proper motions are each sampled from
-normal distributions given the reported uncertainties in
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
-each sampled observable back in time for 10 Gyr in Gadget as massless
-point particles outputting positions every 5Myr (with otherwise similar
-parameters to n-body runs below.) When selecting the initial position
-and velocity of an N-body model, we select the position and velocity of
-the first apocentre occurring after 10 Gyr ago.
+Next, we calculate the range of possible orbits of Sculptor in our
+fiducial Milky Way potential. Fig.~\ref{fig:scl_orbits} illustrates
+point particle orbits for 100 samples of Sculptor's observed kinematics
+integrated backwards 5 Gyr in both galactocentric coordinate slices (x,
+y, z) and in galactocentric radius with time. In terms of the x, y, z
+planes, note that all sampled orbits of Sculptor have nearly the same
+morphology---they each orbit mostly in the y-z plane, passing through
+the same number of pericentres and maintaining similar orbits across
+5Gyr. Sculptor's orbital history is relatively well-constrained.
 
-We assume the galaxy is represented as a point particle for this
-analysis. In detail, (self) dynamical friction likely influences the
-orbit, but a point particle is an adequate description.
+We select an orbit with the \(\sim 3\sigma\) smallest pericentre among
+all possible orbits. We achieve this by taking the median parameters of
+all orbits with a pericentre less than the 0.0027th quantile pericentre,
+yielding a pericentre of 43 kpc.\footnote{only slightly smaller than the
+  0.00135th quantile pericentre} It is unlikely given the current
+observations that Sculptor has a smaller pericentre. Since tidal effects
+depend most strongly on the pericentre of an orbit, this orbit should
+maximize the tidal force from the Milky Way.
 
-To convert from Gaia to galactocentric coordinates, we use the Astropy
-v4 galactocentric frame \citep{astropycollaboration+2022}. This frame
-assumes the galactic centre appears at position
-\(\alpha = {\rm 17h45m37.224s}\), \(\delta = -28°56'10.23''\) with
-proper motions \(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
-\(\mu_\delta=-5.547\pm0.026 \masyr\) (from the appendix and Table 2 of
-\citet{reid+brunthaler2004}). The galactic centre distance is
-\(8.122\pm0.033\,\)kpc with a radial velocity = \(11 + 1.9 \pm 3\) km/s.
-from \citet{gravitycollaboration+2018}. Finally, adding that the sun is
-\(20.8\pm0.3\,\)pc above the disk from \citet{bennett+bovy2019}, and
-using the procedure outlined in \citet{drimmel+poggio2018}, the solar
-velocity relative to the galactic rest frame is `v\_sun' =
-\([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
-uncertainties in the reference frame are typically smaller than the
-uncertainties on the dwarf galaxy's position and velocity.
+We take the first apocentre after a lookback time of 10 Gyr, or at 9.43
+Gyr, as the initial conditions for our model of Sculptor.
+Table~\ref{tbl:scl_orbits} notes the initial conditions for our
+simulation.
 
-\section{N-Body modelling}\label{n-body-modelling}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
+\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
+potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
+Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
+plane. Our selected \texttt{smallperi} orbit is plotted in black and
+light blue transparent orbits represent the past 5Gyr orbits sampled
+over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
+sculptor is well-constrained in this potential and it is unlikely to
+achieve a smaller pericentre than our selected
+orbit.}\label{fig:scl_orbits}
+\end{figure}
 
-\subsection{Initial conditions}\label{initial-conditions}
+\begin{table*}[t]
+\centering
+\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
+The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
+The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+}
+\label{tbl:scl_orbits}
+\begin{tabular}{llll}
+\toprule
+Property & Mean & SmallPeri & LMC\\
+\midrule
+distance & 83.2 & 82.6 & \\
+pmra & 0.099 & 0.134 & \\
+pmdec & -0.160 & -0.198 & \\
+Vlos & 111.2 & 111.2 & \\
+$t_i$ & -8.74 & -9.43 & \\
+$\hat{x}_{i}$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10] & \\
+$\vec{v}_i$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29] & \\
+pericentre & 53 & 43 & \\
+apocentre & 102 & 96 & \\
+$t_{\rm last\ peri}$ & -0.45 & -0.46 & \\
+Numer of peris &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
 
-We use Agama \citep{agama} to generate initial conditions. We initially
+\subsection{Orbits of Ursa Minor}\label{orbits-of-ursa-minor}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_xyzr_orbits.pdf}}
+\caption{Ursa Minor Orbits}
+\end{figure}
+
+\subsection{}\label{section}
+
+ra = 227.242 dec = 67.2221 distance = 64.6 pmra = -0.158 pmdec = 0.05
+radial\_velocity = -245.75
+
+t\_i = -9.53 pericentre = 29.64 apocentre = 74.88 t last peri = -0.80
+x\_i = {[}-16.48 69.92 21.05{]} v\_i = {[}16.32 39.86 -116.99{]}
+
+\section{Initial conditions}\label{initial-conditions}
+
+We use \agama [@agama] to generate initial conditions. We initially
 assume galaxies are described by an NFW dark matter potential
@@
 in this potential (added on in post-processing). The density is a
-cubic-exponentially truncated with a profile
-\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
+cubic-exponentially truncated with a profile \begin{equation}{
 \rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t)
@@
 
+\subsection{Halos for Sculptor}\label{halos-for-sculptor}
+
+From the observed properties of Sculptor, we can infer reasonable
+\(\Lambda\)CMD dark matter halo hosts. Table~\ref{tbl:scl_derived_props}
+reports our inferred halo and kinematic properties of Sculptor. First,
+taking the absolute magnitude from \citet{munoz+2018} with the
+mass-to-light ratio from \citet{woo+courteau+dekel2008} (1.7 with
+\(\sim\) 0.17 dex uncertainty), the total current stellar mass of
+Sculptor is \(M_\star \sim 3.1 \times 10^6 \Mo\). Based on the stellar
+mass-\(v_{\rm max}\) relation \citep[from][]{fattahi+2018}, see also
+\ref{fig:smhm}, the halo should have
+\(v_{\rm max} \sim 31 \pm 3 \,\kms\). Finally, using the
+\citet{ludlow+2016} \(z=0\) mass-concentration relation, this constraint
+translates into a \(r_{\rm max} \sim 6 \pm 2\,{\rm kpc}\), or
+alternatively \(M_{200} = 5\times10^9\) with \(c=13\).
+
+Fig.~\ref{fig:scl_halos} illustrates these estimates visually. The
+stellar-mass \(v_{\rm max}\) constraint translates to an estimate of
+\(v_{\rm max}\) only (horizontal band). The mass-concentration relation
+describes the relationship between \(v_{\rm max}\) and \(r_{\rm max}\),
+an approximately diagonal linear band. And finally, we include a curved
+line which illustrates where the initial velocity dispersion is
+approximantly larger than the present-day observed dispersion
+\(v_{\rm circ}(R_h) / \sqrt{3} \sim \sigma_v \gtrsim 9\,\kms\).
+
+The intersection of the constraints on the halo leads us to select an
+initial halo, \texttt{compact}, which is more compact than the mean but
+with a observationally consistent initial velocity dispersion. We also
+have the \texttt{light} halo which occupies the intersection between
+3\(\sigma\) lowest mass and the velocity dispersion constraints, which
+we come back to in Section~\ref{sec:scl_lmc}.
+
+While there is some range in the choice of initial halo, reasonable
+changes to the initial halo do not substantially affect the tidal
+evolution for Sculptor. The velocity dispersion, in particular,
+translates to an estimate of the mass within \(r_h\). As a result, halos
+with the same velocity dispersion only differ in total mass, but the
+density slope and circular velocity near the stellar component is
+similar. These halos will have a similarly affected stellar component.
+
+\begin{table*}[t]
+\centering
+\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al.
+(2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008).
+Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$.
+From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al.
+(2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$.
+}
+\label{tbl:scl_derived_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & reference\\
+\midrule
+$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
+$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
+$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
+$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
+$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
+$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
+$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
+$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_initial_halos.pdf}}
+\caption[Sculptor initial halos]{The suggested halos of Sculptor
+compared to cosmological predictions. \textbf{TODO: simplify
+\(\sigma_v\) calculation}.}\label{fig:scl_halos}
+\end{figure}
+
+\begin{table*}[t]
+\centering
+\caption[Initial halos of Sculptor]{The parameters for our initial Sculptor halos. }
+\label{tbl:scl_ini_halos}
+\begin{tabular}{lllll}
+\toprule
+Halo name & Rmax & Vmax & M200 & c\\
+\midrule
+compact & 3.2 & 31 & 0.33 & 21\\
+lmc & 4.2 & 31 & 0.39 & 17\\
+small & 2.5 & 25 &  & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Halos for Ursa Minor}\label{halos-for-ursa-minor}
+
+\begin{table*}[t]
+\centering
+\caption[Derived Properties of Ursa Minor]{Derived properties of Ursa Minor. }
+\label{tbl:umi_derived_props}
+\begin{tabular}{lll}
+\toprule
+parameter & value & reference\\
+\midrule
+$L_\star$ & $3.5 \pm 0.1 \times 10^5\,L_\odot$ & \\
+$M_\star$ & $7_{-2}^{+3} \times 10^5\,\Mo$ & \\
+$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & Woo, Courteau, and Dekel (2008)\\
+$M_{200}$ & $3_{-2}^{+4} \times 10^9\,\Mo$ & \\
+$c$ & 14? & \\
+$v_{\rm circ, max}$ & $27_{-6}^{+7}\,\kms$ & \\
+$r_{\rm circ, max}$ & $5_{-2}^{+1}$ kpc & \\
+$r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
+$t_{\rm break, obs}$ & $120\pm30$ Myr & \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Ursa Minor Initial Halos]{Initial halos for Ursa Minor. }
+\label{tbl:umi_ini_halos}
+\begin{tabular}{lllll}
+\toprule
+Halo name & Rmax & Vmax & M200 & c\\
+\midrule
+fiducial & 5 & 37 & 0.67 & 17\\
+compact & 4 & 38 & 0.62 & 21\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_initial_halos.pdf}}
+\caption[Ursa Minor initial halos]{The suggested halos of Ursa Minor
+compared to cosmological predictions.}
+\end{figure}
+
+\subsection{}\label{section-1}
+
+\section{N-Body modelling}\label{n-body-modelling}
+
+Modelling gravitational evolution for large systems requires special
+methods. Perhaps the simplest method to compute the evolution of dark
+matter is through \emph{N-body simulations}. A dark matter halo is
+represented as a large number of dark matter particles (bodies). Each
+body is essentially a Monte Carlo sample of the underlying phase-space
+distribution. Note that dark matter (and galaxies) are often assumed to
+be \emph{collisionless}---particles are not strongly affected by close,
+\emph{collisional} gravitational encounters which substantially change
+the momenta of involved bodies. In contrast, star clusters are often
+collisional so neglecting these encounters may not be a reasonable
+approximation. While we use individual gravitating bodies in N-body
+simulations, the Newtonian gravitational force is softened to be a
+Plummer sphere as to limit strongly collisional encounters.
+
+Naively, the Newtonian gravitational force requires adding together the
+forces from each particle on each particle, with a computational cost
+that scales quadratically with the number of particles, or \(O(N^2)\).
+With this method, simulating a large number of particles, such as
+\(10^6\), would require \(10^{12}\) force evaluations at each time step,
+making cosmological and high-resolution studies unfeasible. However,
+only long-range gravitational interactions tend to be important for CDM,
+so we can utilize the \emph{tree method} to compute the gravitational
+force vastly more efficiently.
+
+The first gravitational tree code was introduced in
+\citet{barnes+hut1986}, and is still in use today. We utilize the
+massively parallel code \emph{Gadget 4} \citep{gadget4}. Particles are
+spatially split into an \emph{octotree}. The tree construction stars
+with one large node, a box containing all of the particles. If there is
+more than one particle in a box/node , the box is then divided into 8
+more nodes (halving the side length in each dimension) and this step is
+repeated until each node only contains 1 particle. With this
+heirarchichal organization, if a particle is sufficiently far away from
+a node, then the force is well approximated by the force from the centre
+of mass of the node. As such, each force calculation only requires a
+walk through the tree, only descending farther into the tree as
+necessary to retain accuracy. The total force calculations reduce from
+\(O(N^2)\) to \(O(N\,\log N)\), representing orders of magnitude
+speedup. Modern codes such as \emph{Gadget} utilize other performance
+tricks, such as splitting particles across many supercomputer nodes,
+efficient memory storage, adaptive time stepping, and parallel file
+writing to retain fast performance for large scale simulations, forming
+the foundation for many cosmological simulation codes.
+
+\textbf{could be 1 paragraph}
+
 \subsection{Isolation runs and simulation
@@
 at the virial radius). Our fiducial isolation halo uses \(r_s=2.76\) kpc
-and \(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for
+and \(M_s = 0.29 \times 10^{10}\,\Mo\) , but can be easily rescaled for
 any length or mass scale.
@@
 h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-}\end{equation} See appendix Section~\ref{sec:extra_convergence} for a
+}\end{equation} See Appendix Section~\ref{sec:extra_convergence} for a
 discussion of this choice, which is similar to the \citet{power+2003}
@@
 
-\subsubsection{Numerical convergence}\label{numerical-convergence}
+\subsection{Numerical fidelity}\label{numerical-fidelity}
 
-As discussed in \citet{power+2003}, the region where density is
-converged is related to the region which becomes collisionally relaxed
-over the time of the universe (so about 5-10 Gyr). (i.e.~where the
-collisionless assumption breaks down). The relaxation timescale is given
-by \begin{equation}\protect\phantomsection\label{eq:t_relax}{
-t_{\rm relax}(r) = t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
-= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
-}\end{equation} where \(N(r)\) is the number of particles within \(r\),
-\(t_{\rm circ}\) is the circular velocity timescale, and \(\bar \rho\)
-is the mean interior density \citep{power+2003}. The converged radius
-\(r_{\rm relax}(t)\) is the radius above which \(t_{\rm relax} > t\).
-Typically, \(r_{\rm relax}(10\Gyr)\) is about 6-10 times our adopted
-softening length, increasing with particle number. As such, at full
-resolution, we can only trust density profiles down to \(\sim10\) times
-our softening length, sufficient to resolve stellar density profiles.
+Fig.~\ref{fig:numerical_convergence} illustrates how well our numerical
+setup is able to reproduce the desired initial conditions, before and
+after running the model in isolation. Circular velocity is computed
+assuming spherical symmetry and only shown for every 200th particle
+(ranked from the centre outwards). With increasing particle number, the
+circular velocity profile maintains close agreement with the expected
+NFW velocity profile until \(r_{\rm relax}\), as marked by arrows. The
+reduction of mass (and consequently circular velocity) interior to
+\(r_{\rm relax}\) is likely due to collisional effects which would
+continue to reduce with higher resolution. Typically,
+\(r_{\rm relax}(10\Gyr)\) is about 6-10 times our adopted softening
+length, increasing with particle number. As such, at full resolution, we
+can only trust density profiles down to \(\sim10\) times the softening
+length, sufficient to resolve stellar density profiles.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
 \caption[Numerical halo convergence]{Numerical convergence test for
@@
 slight reduction in density starting around \$r = 30 \$kpc is expected
-given our truncation choice.}\label{fig:numerical_convergance}
+given our truncation choice. \textbf{Add
+softening}}\label{fig:numerical_convergence}
 \end{figure}
 
-\subsection{Orbital runs}\label{orbital-runs}
+\subsection{Orbital evolution}\label{orbital-evolution}
 
@@
 
-\section{Post processing}\label{post-processing}
-
 \subsection{Centring}\label{sec:shrinking_spheres}
 
-Shrinking spheres centres inspired by \citet{power+2003}
+Accurately determining the centre of the subhalo at each timestep is
+essential to most analysis. We use a shrinking-spheres centre method
+inspired by \citet{power+2003}. First, we start with an initial centre
+estimate from the last timestep. Then, we calculate the radius of all
+particles from the centre, remove particles with a radius beyond the
+0.975 quantile of the centre, and recalculating centroid until radius is
+less than \textasciitilde1kpc or fewer than 0.1\% of particles remain.
+Finally, we remove all unbound particles based on the instantaneous
+N-body potential with all particles. For all future timesteps, we use
+the snapshot centre's position, velocity, and acceleration to predict
+the location of the next centre. We also consider only particles
+retained from the previous iteration.
 
-\begin{itemize}
-\tightlist
-\item
-  Recursively shrink radius by 0.975 quantile and recalculating centroid
-  until radius is less than \textasciitilde1kpc or fewer than 0.1\% of
-  particles remain.
-\item
-  Remove bound particles (using instantanious potential).
-\item
-  Use previous snapshot centre and acceleration to predict new centre
-  for next snapshot and only include particles included in previous
-  centring timestep.
-\end{itemize}
-
-The statistical centring uncertainty for the 1e7 particle isolation run
-is of order 0.003 kpc, however oscillations in the centre are of order
-0.03 kpc. This is about three times the softening length but is less
-than the numerically converged radius scale.
-
-\subsection{Velocity profiles}\label{velocity-profiles}
-
-\begin{itemize}
-\tightlist
-\item
-  Circular velocity is computed assuming spherical symmetry and only
-  shown for every 200th particle (ranked from the centre outwards)
-\item
-  \(v_{\rm circ\ max}\) and \(r_{\rm circ,\ max}\) is found by
-  least-squares fitting the NFW functional velocity form to the points
-  of the velocity profile that have the 10\% highest velocities. This is
-  not a necessarily good fit, especially as the halo becomes stripped,
-  but accurate enough to find a reliable maximum.
-\end{itemize}
+The statistical centring uncertainty for the full resolution (\(10^7\)
+particle) isolation run is of order 0.003 kpc, but oscillations in the
+centre are of order 0.03 kpc. This is about three times the softening
+length but is less than the numerically converged radius scale.
 
@@
 from the surface density, \(\Sigma_\star\), via the inverse Abel
-transform. In this work, we consider both Plummer and exponential
-profiles, as described in
+transform.
 
-\subsection{Stellar density profiles and velocity
-dispersion}\label{stellar-density-profiles-and-velocity-dispersion}
-
-Stellar velocity dispersion is calculated for all stars within a 1kpc
-sphere of the centre. We assume
-\(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
-i.e.~isotropic velocity dispersion.
-
-We calculate density profiles similar to stars and assume a constant bin
-width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\) (Freedman-Diaconis
-prescription). \(R\) may be either the cylendrical radius in \(x-y\) or
-the on-sky project \(\xi, \eta\) tangent plane coordinates.
-
-Because our dwarfs are assumed to be spherical/isotropic, we retain this
-assumption when calculating predicted density profiles.
+We find the stellar profiles created in this manner are stable in the
+isolated systems and retail excellent agreement with the assumed stellar
+density profile.

### 2025-08-29 thesis/rv_models.tex
thesis refining data section
julio edits at durham

@@
 not report GaiaDR3 source ID's, we match to the nearest star within 1-3
-arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
-measurement from each study using the inverse-variance weighted mean
-\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
+arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
+mean RV measurement from each study using the inverse-variance weighted
+mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
 variance \(s^2\). We remove stars with significant velocity dispersions
@@
 \centering
-\caption{Summary of velocity measurements and derived properties.}
+\caption[Spectroscopic LOS velocity measurements]{Summary of velocity measurements and derived properties. }
+\label{tbl:rv_measurements}
 \begin{tabular}{lllllllll}
@@
 \end{table*}
+
+measurement
 

### 2025-08-29 thesis/sculptor.tex
julio edits at durham

@@
 
-\subsubsection{Derived properties and
-setup}\label{derived-properties-and-setup}
-
-From the observed properties of Sculptor, we can infer reasonable
-\(\Lambda\)CMD dark matter halo hosts. Table~\ref{tbl:scl_derived_props}
-reports our inferred halo and kinematic properties of Sculptor. First,
-taking the absolute magnitude from \citet{munoz+2018} with the
-mass-to-light ratio from \citet{woo+courteau+dekel2008} (1.7 with
-\(\sim\) 0.17 dex uncertainty), the total current stellar mass of
-Sculptor is \(M_\star \sim 3.1 \times 10^6 \Mo\). Based on the stellar
-mass-\(v_{\rm max}\) relation \citep[from][]{fattahi+2018}, see also
-\ref{fig:SMHM}, the halo should have
-\(v_{\rm max} \sim 31 \pm 3 \,\kms\). Finally, using the
-\citet{ludlow+2016} \(z=0\) mass-concentration relation, this constraint
-translates into a \(r_{\rm max} \sim 6 \pm 2\,{\rm kpc}\), or
-alternatively \(M_{200} = 5\times10^9\) with \(c=13\).
-
-Fig.~\ref{fig:scl_halos} illustrates these estimates visually. The
-stellar-mass \(v_{\rm max}\) constraint translates to an estimate of
-\(v_{\rm max}\) only (horizontal band). The mass-concentration relation
-describes the relationship between \(v_{\rm max}\) and \(r_{\rm max}\),
-an approximately diagonal linear band. And finally, we include a curved
-line which illustrates where the initial velocity dispersion is
-approximantly larger than the present-day observed dispersion
-\(v_{\rm circ}(R_h) / \sqrt{3} \sim \sigma_v \gtrsim 9\,\kms\).
-
-The intersection of the constraints on the halo leads us to select an
-initial halo, \texttt{compact}, which is more compact than the mean but
-with a observationally consistent initial velocity dispersion. We also
-have the \texttt{light} halo which occupies the intersection between
-3\(\sigma\) lowest mass and the velocity dispersion constraints, which
-we come back to in Section~\ref{sec:scl_lmc}.
-
-While there is some range in the choice of initial halo, reasonable
-changes to the initial halo do not substantially affect the tidal
-evolution for Sculptor. The velocity dispersion, in particular,
-translates to an estimate of the mass within \(r_h\). As a result, halos
-with the same velocity dispersion only differ in total mass, but the
-density slope and circular velocity near the stellar component is
-similar. These halos will have a similarly affected stellar component.
-
-\begin{table*}[t]
-\centering
-\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al.
-(2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008).
-Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$.
-From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al.
-(2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$.
-}
-\label{tbl:scl_derived_props}
-\begin{tabular}{lll}
-\toprule
-parameter & value & reference\\
-\midrule
-$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
-$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
-$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
-$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
-$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
-$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
-$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
-$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
-$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_initial_halos.pdf}}
-\caption[Sculptor initial halos]{The suggested halos of Sculptor
-compared to cosmological predictions. \textbf{TODO: simplify
-\(\sigma_v\) calculation}.}\label{fig:scl_halos}
-\end{figure}
-
-\begin{table*}[t]
-\centering
-\caption[Initial halos of Sculptor]{The parameters for our initial Sculptor halos. }
-\label{tbl:scl_ini_halos}
-\begin{tabular}{lllll}
-\toprule
-Halo name & Rmax & Vmax & M200 & c\\
-\midrule
-compact & 3.2 & 31 & 0.33 & 21\\
-lmc & 4.2 & 31 & 0.39 & 17\\
-small & 2.5 & 25 &  & \\
-\bottomrule
-\end{tabular}
-\end{table*}
-
 \subsection{Milky Way tides}\label{milky-way-tides}
-
-\subsubsection{Orbital properties}\label{orbital-properties}
-
-Next, we calculate the range of possible orbits of Sculptor in our
-fiducial Milky Way potential. Fig.~\ref{fig:scl_orbits} illustrates
-point particle orbits for 100 samples of Sculptor's observed kinematics
-integrated backwards 5 Gyr in both galactocentric coordinate slices (x,
-y, z) and in galactocentric radius with time. In terms of the x, y, z
-planes, note that all sampled orbits of Sculptor have nearly the same
-morphology---they each orbit mostly in the y-z plane, passing through
-the same number of pericentres and maintaining similar orbits across
-5Gyr. Sculptor's orbital history is relatively well-constrained.
-
-We select an orbit with the \(\sim 3\sigma\) smallest pericentre among
-all possible orbits. We achieve this by taking the median parameters of
-all orbits with a pericentre less than the 0.0027th quantile pericentre,
-yielding a pericentre of 43 kpc.\footnote{only slightly smaller than the
-  0.00135th quantile pericentre} It is unlikely given the current
-observations that Sculptor has a smaller pericentre. Since tidal effects
-depend most strongly on the pericentre of an orbit, this orbit should
-maximize the tidal force from the Milky Way.
-
-We take the first apocentre after a lookback time of 10 Gyr, or at 9.43
-Gyr, as the initial conditions for our model of Sculptor.
-Table~\ref{tbl:scl_orbits} notes the initial conditions for our
-simulation.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_xyzr_orbits.pdf}}
-\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
-potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
-Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
-plane. Our selected \texttt{smallperi} orbit is plotted in black and
-light blue transparent orbits represent the past 5Gyr orbits sampled
-over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
-sculptor is well-constrained in this potential and it is unlikely to
-achieve a smaller pericentre than our selected
-orbit.}\label{fig:scl_orbits}
-\end{figure}
-
-\begin{table*}[t]
-\centering
-\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
-The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
-The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
-}
-\label{tbl:scl_obrits}
-\begin{tabular}{llll}
-\toprule
-Property & Mean & SmallPeri & LMC\\
-\midrule
-distance & 83.2 & 82.6 & \\
-pmra & 0.099 & 0.134 & \\
-pmdec & -0.160 & -0.198 & \\
-Vlos & 111.2 & 111.2 & \\
-$t_i$ & -8.74 & -9.43 & \\
-$\hat{x}_{i}$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10] & \\
-$\vec{v}_i$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29] & \\
-pericentre & 53 & 43 & \\
-apocentre & 102 & 96 & \\
-$t_{\rm last\ peri}$ & -0.45 & -0.46 & \\
-Numer of peris &  &  & \\
-\bottomrule
-\end{tabular}
-\end{table*}
 
@@
 
-\subsection{Effects of the LMC}\label{effects-of-the-lmc}
+\subsection{Effects of the LMC}\label{sec:scl_lmc}
 
@@
 
-Fig.~\ref{fig:scl_lmc_orbit_effect} displays the effect of including an
+Fig.~\ref{fig:scl_lmc_orbits_effect} displays the effect of including an
 LMC in the potential. The green samples are in the initial MW-only
@@
 LMC) to the MW is plotted (left), and the Sculptor - LMC distance
-(right.)}\label{fig:scl_lmc_orbit_effect}
+(right.)}\label{fig:scl_lmc_orbits_effect}
 \end{figure}
@@
 \caption[Sculptor initial and final density with LMC]{The tidal effects
-on the stellar surface density due to the LMC today.}
+on the stellar surface density due to the LMC
+today.}\label{fig:scl_lmc_i_f}
 \end{figure}
@@
 \section{Ursa Minor}\label{ursa-minor}
-
-\subsection{Derived properties}\label{derived-properties}
-
-\begin{table*}[t]
-\centering
-\caption[Derived Properties of Ursa Minor]{Derived properties of Ursa Minor. }
-\label{tbl:umi_derived_props}
-\begin{tabular}{lll}
-\toprule
-parameter & value & reference\\
-\midrule
-$L_\star$ & $3.5 \pm 0.1 \times 10^5\,L_\odot$ & \\
-$M_\star$ & $7_{-2}^{+3} \times 10^5\,\Mo$ & \\
-$M_\star / L_\star$ & 1.9 (pm 0.17 dex) & Woo, Courteau, and Dekel (2008)\\
-$M_{200}$ & $3_{-2}^{+4} \times 10^9\,\Mo$ & \\
-$c$ & 14? & \\
-$v_{\rm circ, max}$ & $27_{-6}^{+7}\,\kms$ & \\
-$r_{\rm circ, max}$ & $5_{-2}^{+1}$ kpc & \\
-$r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
-$t_{\rm break, obs}$ & $120\pm30$ Myr & \\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{table*}[t]
-\centering
-\caption[Ursa Minor Initial Halos]{Initial halos for Ursa Minor. }
-\label{tbl:umi_ini_halos}
-\begin{tabular}{lllll}
-\toprule
-Halo name & Rmax & Vmax & M200 & c\\
-\midrule
-fiducial & 5 & 37 & 0.67 & 17\\
-compact & 4 & 38 & 0.62 & 21\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_initial_halos.pdf}}
-\caption[Ursa Minor initial halos]{The suggested halos of Ursa Minor
-compared to cosmological predictions.}
-\end{figure}
-
-\subsection{Orbital properties}\label{orbital-properties-1}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_xyzr_orbits.pdf}}
-\caption{Ursa Minor Orbits}
-\end{figure}
-
-\subsection{Tidal Effects}\label{tidal-effects-2}
-
-ra = 227.242 dec = 67.2221 distance = 64.6 pmra = -0.158 pmdec = 0.05
-radial\_velocity = -245.75
-
-t\_i = -9.53 pericentre = 29.64 apocentre = 74.88 t last peri = -0.80
-x\_i = {[}-16.48 69.92 21.05{]} v\_i = {[}16.32 39.86 -116.99{]}
 
@@
 
-\subsection{Effects of the LMC}\label{effects-of-the-lmc-1}
+\subsection{Effects of the LMC}\label{effects-of-the-lmc}
 

### 2025-08-31 thesis/abstract.tex
thesis julio ch1-3 edits

@@
 Magellanic clouds, are all dwarf spheroidals (dSphs)---gas-free,
-non-rotating stellar systems who's gravity is dominated by dark matter.
+non-rotating stellar systems whos gravity is dominated by dark matter.
 The light profiles of dSphs are usually well-approximated by an
@@
 Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
-stars far from the main body, out to \textasciitilde10 effective radii.
-The origin of these extended profiles is not clear, and may reflect
-either the influence of Galactic tides or instead an innate feature. In
-this thesis, I review the excestence of an extended light profiles in
-Scl and UMi, validating the Bayesian membership catalogue of
+stars far from the main body, out to \(\sim10\) effective radii. The
+origin of these extended profiles is not clear, and may reflect either
+the influence of Galactic tides or instead an innate feature. In this
+thesis, I review the excestence of an extended light profiles in Scl and
+UMi, validating the Bayesian membership catalogue of
 \citet{jensen+2024}. To evaluate if tides indeed can explain these

### 2025-08-31 thesis/data.tex
thesis julio ch1-3 edits

@@
 luminosity/stellar mass density profile of dwarf galaxies is generally
-well-described by an exponential law Eq.~\ref{eq:exponential_profiles}.
-A prototypical example is the Fornax dSph --- well-fit by an exponential
+well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
+prototypical example is the Fornax dSph --- well-fit by an exponential
 over 4.5 decades in surface density. On the other hand, dSph like
@@
 subsequent chapters.
-
-\textbf{question: how much about density profiles in this
-paragraph\ldots{}}
 
@@
 For our fiducial sample, we adopt a minimum probability of
-\(P_{\rm sat} = 0.2\). Most stars have \(P_{\rm  sat}\) values which are
-nearly 0 or 1, so the exact choice of probability threshold has little
-effect on the resulting sample. Even at our relatively generous
-probability threshold of 0.2, the purity remains high when validated
-against spectroscopic line-of-sight (LOS) velocities
+\(P_{\rm sat} = 0.2\). We use the \(P_{\rm sat}\) values from the
+elliptical 2-component runs if a galaxy showed evidence for an outer
+component, the 1-component run otherwise. Most stars have
+\(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact choice of
+probability threshold has little effect on the resulting sample. Even at
+our relatively generous probability threshold of 0.2, the purity remains
+high when validated against spectroscopic line-of-sight (LOS) velocities
 (\textasciitilde90\%, J+24). We note, however, that these purity
@@
 circularized elliptical radius, \begin{equation}{
-R_{\rm ell}^2 = a\,b\,({\xi'}^2 / a^2 + {\eta'}^2 / b^2),
+R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
 }\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
@@
 {\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
-}\end{equation} Next, our \textbf{fiducial sample},
+}\end{equation} Next, our \textbf{fiducial} sample,
 \(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
-J+24. Finally, the \textbf{RV members} sample adds line-of-sight
-velocity information (see Appendix for details). While the latter is the
-best motivated membership sample, we do not use this sample for stellar
-density analysis due to its incompleteness and complex selection
-function.
+J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
+line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
+details). While the latter is the best motivated membership sample, we
+do not use this sample for stellar density analysis due to its
+incompleteness and complex selection function.
 
@@
 shows our fiducial sample for each galaxy, and the impact of alternate
-samples, in tangent-plane coordinates, the CMD, and in proper-motion
-space. The \emph{all} sample extends uniformly across the tangent plane,
-but includes a substantial background population. The \emph{CMD+PM}
-sample has a much lower background density in the tangent plane,
-revealing the satellite more clearly. The fiducial sample appears
-similar to the \emph{CMD+PM} sample in the CMD and PM planes, but
-excludes improbably distant stars. The \emph{RV member} sample also
-traces out similar distributions in CMD and PM space as the fiducial
-sample, with less dispersion, likely reflecting the brighter magnitudes
-of stars with spectroscopic follow-up. Each component plays an important
-role in ensuring a high-quality membership criterion.
+samples, in tangent-plane coordinates, the CMD, and proper-motion space.
+The ``all'' sample extends uniformly across the tangent plane, but
+includes a substantial background population. The ``CMD+PM'' sample has
+a much lower background density in the tangent plane, revealing the
+satellite more clearly. The fiducial sample appears similar to the
+``CMD+PM'' sample in the CMD and PM planes, but excludes improbably
+distant stars. The ``RV member'' sample also traces out similar
+distributions in CMD and PM space as the fiducial sample, with less
+dispersion, likely reflecting the brighter magnitudes of stars with
+spectroscopic follow-up. Each selection criteria---spatial, CMD, and
+PM---contributes towards a high-quality membership criterion.
 
@@
 a bluer CMD, indicative of more recent star formation. Otherwise, there
-are no standout morphological features between these galaxies.
+are no major morphological differences between these galaxies.
 
@@
 stars. However, there are no clear extensions or overdensities in the
-\emph{CMD+PM} selected sample, which does not include a spatial
-likelihood. If even fainter, more extended features exist, they are
-apparently not detectable with \emph{Gaia}.
+``CMD+PM'' selected sample, which does not include a spatial likelihood.
+If even fainter, more extended features exist, they are apparently not
+detectable with \emph{Gaia}.
 
-To illustrate the extent of Sculptor and Ursa Minor, we highlight the
-most distant, spectroscopically confirmed members with red-outlined
-stars in Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In
-particular, \citet{sestito+2023a}; \citet{sestito+2023b} targeted the
-most distant stars using the J+24 candidate membership list. These stars
-illustrate an extended stellar distribution for each galaxy. For an
-exponential profile, 99.95\% of stars fall within a radius of \(6R_h\),
-so the mere presence of very distant member stars provides direct
-evidence for deviations from a simple, exponential profile.
+To illustrate the extent of Sculptor and Ursa Minor, we highlight
+distant, spectroscopically confirmed members with red-outlined stars in
+Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
+\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
+stars from the J+24 candidate membership list. The most distant stars
+confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
+of each dwarf. For an exponential profile, 99.95\% of stars fall within
+a radius of \(6R_h\), so the presence of very distant member stars
+provides evidence for deviations from an exponential profile.
 
@@
 \caption[Sculptor sample selection]{The distributions of various samples
-of \emph{Gaia} stars for Sculptor. We plot light grey points for field
-stars (with consistent parallaxes and reliable photometry and
-astrometry), turquoise points for \emph{CMD+PM} selected stars
-(Eq.~\ref{eq:sel_cmd_pm}), blue \texttt{x}s for the fiducial sample
-(\(P_{\rm sat} > 0.2\)), and indigo diamonds for the RV confirmed
-members. We mark the two far-outlier stars from \citet{sestito+2023a}
-with rust-outlined indigo stars. \textbf{Top:} Tangent plane
-\(\xi, \eta\). The orange ellipse represents 3 half-light radii.
-\textbf{Bottom left:} Colour magnitude diagram in Gaia \(G\) magnitude
-versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a Padova \(12\,\)Gyr,
-\({\rm [Fe/H] }=-1.68\) isochone in orange. The black bar in the top
-left represents the median colour error. \textbf{Bottom right:} Proper
-motion in declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
-(corrected). The orange point marks the systemic \citet{MV2020b} proper
-motion. The black cross represents the median proper motion error.
-\textbf{TODO: reduce png sizes}}\label{fig:scl_selection}
+of \emph{Gaia} stars for Sculptor. We plot light grey points for ``all''
+field stars (with consistent parallaxes and reliable photometry and
+astrometry), turquoise points for ``CMD+PM'' selected stars
+(Eq.~\ref{eq:sel_cmd_pm}), blue squares for the ``fiducial'' sample
+(with membership probability \(P_{\rm sat} > 0.2\)), and indigo diamonds
+for the ``RV members'' sample. We mark the two far-outlier stars from
+\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
+Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
+half-light radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia
+\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
+Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
+black bar in the top left represents the median colour error.
+\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
+\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
+\citet{MV2020b} proper motion. The black cross represents the median
+proper motion error.}\label{fig:scl_selection}
 \end{figure}
@@
 \caption[Ursa Minor sample selection]{Similar to
-Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline RV
-members outside a radius of \(6R_h\) with red stars (from
-\citet{sestito+2023b}, \citet{pace+2020} and \citet{spencer+2018}).
-\textbf{TODO: plot 6Rh ellipse\ldots{}}}\label{fig:umi_selection}
+Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
+``velocity confirmed'' members outside a radius of \(6R_h\) with red
+stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
+mark the location of Muñoz 1 with a pink
+circle.}\label{fig:umi_selection}
 \end{figure}
@@
 \caption[Fornax sample selection]{Similar to
-Fig.~\ref{fig:scl_selection} except for Fornax. RV members are from
+Fig.~\ref{fig:scl_selection} except for Fornax. RV measurements are from
 \citet{WMO2009}. Fornax does not show the same extended outer halo of
@@
 for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
-different selections of stars from above: \emph{all} stars, CMD+PM only,
-and the fiducial sample. In each case, all samples are the same towards
-the inner regions of the satellite. However, the density profile
-\emph{quality stars} plateaus at the total background in the field at
-radii of 30-60 arcminutes, depending on the galaxy. By restricting
-stellar membership with CMD + PM, the background is reduced by 1-2 dex.
-The CMD+PM background plateau represents the density of background stars
-which could be mistaken as members because of their coincident colours
-and proper motions. Finally, the background (BG)-subtracted profile
-results from subtracting the apparent background in the \emph{all}
-profile.
-
-\textbf{Describe R\_break, etc. in Fig.2.4} Note that, in all three
-galaxies, the data does not extend out to the break radius.
+three different samples from above: ``all'', ``CMD+PM'', and the
+``fiducial'' sample. In each case, all samples are the same towards the
+inner regions of the satellite. However, the density profile from
+``all'' plateaus at the total background in the field at radii of 30-60
+arcminutes, depending on the galaxy. By restricting stellar membership
+with CMD + PM, the background is reduced by 1-2 dex. The ``CMD+PM''
+background plateau represents the density of background stars which
+could be mistaken as members because of their coincident colours and
+proper motions. Finally, the ``all -- background'' profile results from
+subtracting the apparent background in the ``all'' profile.
 
@@
 assumed density profile (an exponential in J+24). We explore this in
-more detail in Appendix {[}@-sec:density\_extra{]}.
+more detail in Appendix \ref{sec:density_extra}.
 
-Nearto UMi, there is a small, likely unassociated, ultrafaint star
-cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
-position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
-to an elliptical radius of 36 arcminutes. This cluster does not have a
-bright RGB, so has few stars brighter than a magnitude \(G=22\). The
-cluster has little effect on the elliptically-averaged density profile.
-\textbf{Mark Muñoz 1}
+Near to UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
+ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
+relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
+corresponding to an elliptical radius of 37 arcminutes. This cluster
+does not have a bright RGB, so has few stars brighter than a magnitude
+\(G=22\). The cluster has little effect on the elliptically-averaged
+density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).
 
@@
 profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
-galaxies. Of the classicals, we exclude Antlia II due to the extremely
-high background, and Sagittarius which was not included in J+24. The
+galaxies. Of the classicals, we exclude Antlia II, due to the extremely
+high background, and Sagittarius, which was not included in J+24. The
 density profiles are scaled to match at the half-light radius from
@@
 an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
-of magnitude. This excess is remarkable because the spatial likelihood
-used to derive the fiducial sample probabilities is the same (a single
-component exponential) in all cases. The remainder of this thesis will
-be devoted to assessing whether the outer excess shown by Scl and UMi
-are due to Galactic tides.
+of magnitude. The remainder of this thesis will be devoted to assessing
+whether the outer excess shown by Scl and UMi are due to Galactic tides.
 
@@
 Ursa Minor, and Fornax for different selection criteria, ploted as log
-surface dencity versus log elliptical radius. \emph{CMD+PM} select stars
-more likely to be members according to CMD and PM only, \emph{all}
-selects any high quality star, and \emph{BG subtracted} is the
-background-subtracted density derived from \emph{all} stars. We mark the
-half-light radius (vertical dashed line) and the break radius (black
-arrow, Eq.~\ref{eq:r_break}). \textbf{rename
-bg-subtracted}}\label{fig:scl_observed_profiles}
+surface dencity versus log elliptical radius. The samples are: ``all''
+selects any high quality and parallax-consistent star (grey
+circles),``all--background'' subtracts a uniform background density from
+the ``all'' profile (orange pentagons), ``CMD+PM'' select stars
+according to CMD and PM only (teal triangles), and ``fiducial'' also
+includes spatial information (blue squares). We mark the half-light
+radius with a vertical dashed line and the background density with the
+horizontal grey line. For Ursa Minor, we show the expected location of
+Muñoz 1 stars as a horizontal bar (ranging from plus/minus 3 half-light
+radii).}\label{fig:scl_observed_profiles}
 \end{figure}

### 2025-08-31 thesis/introduction.tex
thesis julio ch1-3 edits

@@
 fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
-or similarly \(M_\star \lesssim 10^9 M_\odot\) {[}e.g.,
-\citet{hodge1971}; \citet{mcconnachie2012}). Because the galaxy
+or similarly \(M_\star \lesssim 10^9 M_\odot\)
+\citep[e.g.,][]{hodge1971, mcconnachie2012}. Because the galaxy
 luminosity function increases towards fainter objects, dwarfs are the
-most numerous galaxies in the Universe. Dwarf galaxies are also highly
+most numerous galaxies in the Universe
+\citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
 \emph{dark-matter dominated}, with mass to light ratios which may exceed
-1000 \(M_\odot/ L_\odot\). With the exception of the Magellanic Cloud,
-all dwarf galaxy satellites of the Milky Way (MW) are \emph{quenched},
-with little to no recent star formation. Indeed, most faint MW
+1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.
+With the exception of the Magellanic Clouds, most dwarf galaxy
+satellites of the Milky Way (MW) are \emph{quenched}, with little to no
+recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
 satellites contain stellar populations which are \emph{relics} from the
 early universe, consisting of many of the oldest and most metal poor
-stars.
-
-Understanding the properties of dwarf galaxies thus has implications
-across astronomy, from cosmological structure formation on small scales
-to the formation of metal-poor stellar populations.
-
-\emph{Remove this paragraph?} Of the classical dwarfs, we find that the
-Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
-extended density profile relative to an exponential, hinting at tidal
-effects. As a case study in tidal effects and our understanding of dwarf
-galaxy formation and evolution, we aim to understand the dynamical
-history of these galaxies.
+stars. Understanding the properties of dwarf galaxies thus has
+implications across astronomy, from cosmological structure formation on
+small scales to the formation of metal-poor stellar populations.
 
@@
 local dwarf galaxies. Next, we summarize the cosmological origin and
-role of dwarf galaxies. We then review recent updates to the observed
-properties of dwarf galaxies made on new facilities. We then discuss how
-tides influence satellite galaxies in general. Finally, we provide a
-roadmap to the thesis. \textbf{Look at this paragraph again}
+role of dwarf galaxies. We later review recent advancements and problems
+concerning dwarf galaxies. Next, we contextualize the extended stellar
+density profiles of Sculptor and Ursa Minor. We then discuss how tides
+influence satellite galaxies in general. Finally, we provide a roadmap
+to the thesis.
 
@@
 proposed to explain these unusually high mass-to-light ratios. Examples
-include: ongoing tidal disruption artificially inflating the
-line-of-sight velocity dispersions \citep[e.g.,][]{kuhn+miller1989}, the
-presence of massive central black holes
-\citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
-\citep{milgrom1995}. Over time, however, a consensus developed where the
-high mass-to-light ratios of dwarf galaxies was due to the presence of a
-dark matter halo, which was lacking in GCs
-\citep[e.g.,][\textbf{REFS}]{dekel+silk1986}. Since then, the properties
-of dwarf galaxies have played an increasingly important role in our
-understanding of the clustering of dark matter on small scales {[}e.g.,
-sales+2022, bullock+boylan-kolchyn2017{]}.
+include: ongoing tidal disruption inflating inferred velocity
+dispersions \citep[e.g.,][]{kuhn+miller1989}, the presence of massive
+central black holes \citep[e.g.,][]{strobel+lake1994}, or modified
+theories of gravity \citep{milgrom1995}. Over time, however, a consensus
+developed where the high mass-to-light ratios of dwarf galaxies was due
+to the presence of a dark matter halo, which was lacking in GCs
+\citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}. Since then, the
+properties of dwarf galaxies have played an increasingly important role
+in our understanding of the clustering of dark matter on small scales
+\citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.
+
+Today, a common definition for a dwarf galaxy is a gravitationally bound
+stellar system with dark matter \citep[or, more specifically, not
+consistent with newtonian dynamics of visible matter
+alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
+have no clear evidence for dark matter. The boundary between these two
+classes becomes uncertain for faint, compact systems, known as
+``ambiguous'' systems.
+
+The Local Group (LG) is defined as the group consisting of galaxies
+within \(\sim 1\) Mpc from the Milky Way-Andromeda centre \citep[e.g.,][
+and references therein]{mcconnachie2012}. Today, we know that the Milky
+Way system is teeming with satellites.
+Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
+including dwarf galaxies, globular clusters, and ambiguous systems. Such
+a nearby population of dwarf galaxies are useful for resolved, detailed
+studies aiming to understand the nature of these systems.
 
@@
 \textbf{bright dwarfs} with magnitudes
-\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
-and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
-Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
-galaxy, where most stars are in a rotationally-supported thin disk (seen
-nearly face-on) with a prominent bar. \textbf{Classical dwarf
-spheroidals} occupy intermediate luminosities (
-\(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these systems are old,
-non-star forming, gas-poor, and spheroidal. All Milky Way satellites
-discovered before digital sky surveys are classicals. The 12 classical
-dwarf satellites of our Galaxy are (in order of decreasing luminosity)
-Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II, Carina, Draco,
-Ursa Minor, Canes Venatici I, Sextans I, and Crater II.\footnote{While
-  formally the dwarf galaxy names we discuss contain ``dwarf
-  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
-  brevity.} The \textbf{ultra-faint}s dwarfs (UFDs) occupy the very
-faintest magnitude range (\(-7.7 \lesssim M_V\)). These galaxies have
-minuscule stellar masses, tend to be more compact, and include the
-faintest known galaxies (\textbf{ref}). Altogether, known dwarf galaxies
-span more than 15 orders in magnitude, or over 7 decades in stellar
-mass. \textbf{Rewrite to include stellar mass ranges as well}
+\(-18 \lesssim M_V \lesssim  -14\) or stellar masses
+\(10^9\,\Mo \gtrsim M_\star \gtrsim 3\times10^7\,\Mo\) \footnote{assuming
+  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
+  populations}, often exhibit irregular morphologies and recent star
+formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
+(LMC) as an example of an irregular, bright dwarf galaxy, where most
+stars are in a rotationally-supported thin disk (seen nearly face-on)
+with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
+intermediate luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\) or
+\(3\times10^7\,\Mo \gtrsim M_\star \gtrsim 10^5\,\Mo\)). Typically,
+these systems are old, non-star forming, gas-poor, and spheroidal. All
+Milky Way satellites discovered before digital sky surveys are
+classicals. The 12 classical dwarf satellites of our Galaxy are (in
+order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
+Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
+I, and Crater II.\footnote{While formally the dwarf galaxy names we
+  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
+  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
+occupy the very faintest magnitude range (\(-7.7 \lesssim M_V\), or
+\(M_\star < 10^5\,\Mo\)). These galaxies have minuscule stellar masses,
+typically compact sizes, and very metal-poor stellar populations
+\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
+more than 15 orders in magnitude, or over 7 decades in stellar mass.
 
-A common definition today for a dwarf galaxy is a stellar system with
-evidence for dark matter (high mass-to-light ratios, e.g.~\textbf{REF}).
-In contrast, star clusters (like GCs) have no clear evidence for dark
-matter.
+Here, we focus on classical dwarf galaxies, which are ideal for detailed
+dynamical study. These galaxies are dark matter dominated but also have
+large numbers of stars with radial velocities and proper motions, and
+have generally well-constrained properties.
 
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
-\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
-DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
-over-plotted). Each image includes the galaxy's luminosity and a 200 pc
-scale bar. \emph{TODO: Mark half-light radius with
-circle?}}\label{fig:galaxy_images}
+\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax
+\citep[DES DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
+\citep[UNWISE,][with \textit{Gaia} point sources
+over-plotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
+The grey ellipse represents the half-light radius for the three dwarf
+spheroidals, and the luminosity is derived from the absolute V-band
+magnitude of each galaxy.}\label{fig:galaxy_images}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
+\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
+on the sky. We label the classical dwarf galaxies (green diamonds),
+fainter dwarfs (blue squares), globular clusters (orange circles), and
+ambiguous systems (pink open hexagons). Globular clusters are more
+centrally concentrated, but dwarf galaxies are preferentially found away
+from the MW disk. Sculptor and Ursa Minor are highlighted as two dwarfs
+we study later. The background image is from ESA/Gaia/DPAC
+(https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).
+Dwarf galaxies (confirmed), globular clusters, and ambiguous systems are
+from the \citet{pace2024} catalogue (version
+1.0.3).}\label{fig:mw_satellite_system}
 \end{figure}
@@
 
-We only understand a tiny fraction of the universe's composition. The
-leading theory of cosmology, \LCDM (Lambda Cold Dark Matter), posits
-that the universe is composed of about 68\% dark energy (\(\Lambda\)),
-27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
-  to change definitions of words. \emph{Baryons} here means
+We only understand a fraction of the universe's composition. The leading
+theory of cosmology, \LCDM{} (Lambda Cold Dark Matter), posits that the
+universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
+matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
   baryons+leptons, i.e.~any standard model massive fermion.}
@@
 and dark energy remains elusive, we know their general properties. Dark
-energy causes the acceleration of the expansion of the universe on large
+energy drives the acceleration of the expansion of the universe on large
 scales. We do not discuss dark energy here---it does not substantially
@@
 majority of mass in galaxies. Typically, galaxies have baryonic to dark
-matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
-In \LCDM, dark matter is assumed to interact only gravitationally. Light
-passes through dark matter unimpeded---in this sense dark matter is
-transparent. Dark matter is also commonly assumed to be \emph{cold},
+matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
+\citep[e.g.,][]{hayashi+2023}.
+
+In \LCDM{}, dark matter is assumed to interact only gravitationally.
+Light passes through dark matter unimpeded---in this sense dark matter
+is transparent. Dark matter is also commonly assumed to be \emph{cold},
 i.e.~typical velocities much smaller than the speed of light in the
@@
 galaxy clusters to smaller than the faintest dwarf galaxies.
-Implications of dark matter properties range from cosmological
-structural formation, galaxy structure, and galaxy interactions.
+Implications of dark matter properties include cosmological structural
+formation, galaxy structure, and galaxy interactions.
 
-\textbf{is this ordered logically?}
-
-\subsection{\texorpdfstring{Dwarf galaxy formation in
-\LCDM}{Dwarf galaxy formation in }}\label{dwarf-galaxy-formation-in}
-
-\textbf{In the next section, we consider galaxy formatin in LCDM}
+\subsection{\texorpdfstring{Structure formation in
+\LCDM{}}{Structure formation in }}\label{structure-formation-in}
 
@@
 \citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
-Structure formation happens on a wide range of scales. The largest
-structures become the cosmic web---composed of voids, filaments, and
-sheets---and the smallest non-linear structures (halos, \textbf{define
-non-linear}) host dwarf galaxies. Hierarchical assembly is evident
-through the large scale structure of the universe, remnants of past
-mergers within the Milky Way, and tidal disruption of dwarf galaxies and
-their streams around nearby galaxies.
+Hierarchical assembly is evident through the large scale structure of
+the universe, remnants of past mergers within the Milky Way, and tidal
+disruption of dwarf galaxies and their streams around nearby galaxies.
 
-Small-scale structure formation is sensitive to deviations from
-\LCDM cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of
-\LCDM is that mass perturbations are expected to exist on all scales,
-and are largest on the smallest scales, so we would expect the formation
-of halos on all scales. Many alternative models, such as warm dark
-matter, may smooth out small-scale features and reduce the abundance of
-small halos or change their structure \citep[e.g.,][]{lovell+2014}.
-Dwarf galaxies, which occupy the smallest dark matter halos, are
-promising windows into small-scale structures. To understand the
-evolution of these objects, we need to understand the general
-predictions of \LCDM for the properties of dwarf galaxies.
+Small-scale structure formation is sensitive to deviations from \LCDM{}
+cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of \LCDM{}
+is that mass perturbations are expected to exist on all scales, and are
+largest on the smallest scales, so we would expect the formation of
+halos on all scales. Many alternative models, such as warm dark matter,
+may smooth out small-scale features and reduce the abundance of small
+halos or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf
+galaxies, which occupy the smallest dark matter halos, are promising
+windows into the behaviour of dark matter on small scales.
 
@@
 
-In \LCDM cosmological simulations, dark matter halos are remarkably
+In \LCDM{} cosmological simulations, dark matter halos are remarkably
 self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
 observe that the spherically-averaged density profiles \(\rho(r)\) are
-universally well described by a two-parameter law:
+universally well described by a two-parameter law,
 \begin{equation}\protect\phantomsection\label{eq:nfw}{
@@
 }\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
-density . This profile has shown remarkable success in describing
-\LCDM halos across several orders of magnitude in mass. NFW profiles are
-\emph{cuspy}, where the density continuously rises like
-\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
-density profile increases gradually with radius, and at large radii the
-density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
-shows an example NFW halo.
+density. This profile has shown remarkable success in describing \LCDM{}
+halos across several orders of magnitude in mass. NFW profiles are
+\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
+radii \(r \ll r_s\). The steepness of the density profile increases
+gradually with radius, and at large radii the density falls off like
+\(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density} shows an example NFW
+halo.
 
@@
 \begin{equation}{
-M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
+M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G.
 }\end{equation} A standard second parameter of NFW halos is the halo
-concentration, \(c=r_{200} / r_s\) which describes how the
+concentration, \(c=r_{200} / r_s\), which describes how the
 characteristic radius scale of the halo compares to the virial radius.
@@
 circular velocity profiles. The circular velocity,
-\(\V_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum
-\(\V_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
-\(\V_{\rm max}\) and \(r_{\rm max}\), like \(M_{200}\) and \(c\), fully
-specify an NFW halo,
+\(\vcirc(r) = \sqrt{G M(r) / r}\), reaches a maximum \(\vmax\) at radius
+\(\rmax \approx 2.16258\,r_s\). \(\vmax\) and \(r_{\rm max}\), like
+\(M_{200}\) and \(c\), fully specify an NFW halo.
 
-The two parameters of an NFW profile are not independent. Lower mass
+The two parameters of an NFW profile are not independent. Lower-mass
 dark matter halos often collapse earlier, when the universe was denser.
@@
 left panel of Fig.~\ref{fig:smhm} illustrates the present-day
-mass-concentration relation in terms of \(r_{\rm max}\) and
-\(\V_{\rm max}\). While there is a general trend for concentration to
-increase with mass, the concentration values still have substantial
-scatter. Other parameters such as the halo spin or shape may affect the
-scatter of the mass-concentration relation, but their effect is
-typically expected to be small
+mass-concentration from \citet{ludlow+2016}. While concentration tends
+to decrease with increasing mass, the relation has substantial scatter.
+Other parameters such as the halo spin or shape may affect the scatter
+of the mass-concentration relation, but their effect is typically
+expected to be small
 \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
 \caption[Example density profiles]{Density profiles in log density
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
-\caption[Stellar-mass halo-mass relation]{\textbf{Left} Radius of
-maximum circular velocity \(r_{\rm max}\) as a function of maximum
-circular velocity. The solid line with 1-\(\sigma\) shaded region and
-dashed line are the relations from \citet{ludlow+2016} for \(z=0\) and
-\(z=2\) respectively. \textbf{Right} Stellar mass (top) as a function of
-maximum circular velocity. The solid line with the 1-\(\sigma\) shaded
-region is the relation from \citet{fattahi+2018} with scatter points
-simulated central galaxies from APOSTLE in
-\citet{fattahi+2018}.}\label{fig:smhm}
+\caption[Stellar-mass halo-mass relation]{\textbf{Left} The NFW halo
+concentration \(c=r_{200} / r_s\) as a function of virial mass
+\(M_{200}\). The solid line with 1-\(\sigma\) shaded region is the
+mass-concentration relation from \citet{ludlow+2016} for \(z=0\).
+\textbf{Right} Stellar mass (top) as a function of maximum circular
+velocity. The solid line with the 1-\(\sigma\) shaded region is the
+relation from \citet{fattahi+2018} with scatter points simulated central
+galaxies from \apostle{} in \citet{fattahi+2018}.}\label{fig:smhm}
 \end{figure}
 
-\subsection{\texorpdfstring{Formation of stars within
-\LCDM halos}{Formation of stars within halos}}\label{formation-of-stars-within-halos}
+\subsection{\texorpdfstring{Formation of galaxies within \LCDM{}
+halos}{Formation of galaxies within  halos}}\label{formation-of-galaxies-within-halos}
 
 The observed abundance of galaxies may be compared with the abundance of
-\LCDM halos to derive constraints regarding which galaxies inhabit which
-halos. This technique, dubbed ``abundance matching,'' implies a tight
-relation between the stellar mass of a galaxy and the mass of the halo
-it inhabits \citep{li+white, moster+naab+white2013}.
+\LCDM{} halos to derive constraints regarding which galaxies inhabit
+which halos. This technique, dubbed ``abundance matching,'' implies a
+tight relation between the stellar mass of a galaxy and the mass of the
+halo it inhabits \citep{li+white2009, moster+naab+white2013}.
 
 Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
-(SMHM, with halo mass represented by \(\V_{\rm max}\)) predicted by the
-\LCDM cosmological hydrodynamical simulations from the the APOSTLE
-project \citep{sawala+2016}, which uses the hydrodynamical setup from
-the EAGLE project \citep{crain+2015, schaye+2015} to simulate Local
-Group analogues in a \LCDM cosmological context \textbf{(db: reorder
-these sentences)}. While there is some scatter, the range of predicted
-\({\rm v}_{\rm max}\) is fairly narrow across \(\sim 4\) decades in
-stellar mass. This figure indicates that the SMHM relation becomes
+(SMHM, with halo mass represented by \(\vmax\)) predicted by the \LCDM{}
+cosmological hydrodynamical simulations of Local Group analogues from
+the \apostle{} project \citep{sawala+2016}.\footnote{\apostle{}
+  simulated Local Group analogues in a \LCDM{} cosmological context with
+  the hydrodynamical setup from the \eagle{} simulations
+  \citep{crain+2015, schaye+2015}}. While there is some scatter, the
+range of predicted \(\vmax\) is fairly narrow across \(\sim 4\) decades
+in stellar mass. This figure indicates that the SMHM relation becomes
 increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
@@
 in the faintest galaxies. As a result, the resulting stellar mass of a
-galaxy is highly sensitive to the dark matter mass, especially for faint
-dwarf galaxies. If we know the initial stellar mass of a dwarf galaxy,
-we also know, at some level, the properties of its dark matter halo.
-\textbf{fix this paragraph}
+dwarf galaxy is highly sensitive to the dark matter halo mass.
 
-In \LCDM galaxy formation, the majority of mass in dwarf galaxies comes
-from Dark Matter. Given the SMHM, Fig.~\ref{fig:nfw_density} shows an
-example stellar component for a classical dwarf galaxy with the expected
-underlying dark matter halo. Even when the stars are most dense in the
-inner regions, the dark matter remains nearly an order of magnitude
-higher in density. Stars have a small contribution to the gravitational
-structure of dwarf galaxies--instead stars are reasonably approximated
-as tracer particles of the underlying dark matter halo. \textbf{add
-something about stellar compactness}
+In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxies
+comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
+shows an example exponential stellar component for a classical dwarf
+galaxy with the expected underlying dark matter halo. Even when the
+stars are most dense in the inner regions, the dark matter remains
+nearly an order of magnitude higher in density. Stars have a small
+contribution to the gravitational structure of dwarf galaxies--instead
+stars are reasonably approximated as tracer particles of the underlying
+dark matter halo. In addition, the stellar component is much more
+compact than the dark matter halo.
 
@@
 dwarf galaxy due to pressure from host's circumgalactic medium) and
-tidal removal of gas cause star formation to quench {[}\textbf{REFS}{]}.
-Additionally, the time of formation (relative to reionization) can
-influence the resulting stellar content \citep{kim+2024}. Finally, tides
-influence both the dark matter and stellar mass but in different amounts
-\citep[e.g.,][]{PNM2008}. Consequently, tides may reduce the halo mass
-more than the stellar mass, adding additional scatter to the SMHM trend,
-particularly for low-mass satellites \citep[e.g.,][]{fattahi+2018}.
-Understanding the effects of tides on Local Group dwarf galaxies is
-essential to determine where and how these galaxies formed in a
-cosmological context.
+tidal removal of gas cause star formation to quench
+\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
+(relative to reionization) can influence the resulting stellar content
+\citep{kim+2024}. Finally, tides influence both the dark matter and
+stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
+Consequently, tides may reduce the halo mass more than the stellar mass,
+adding additional scatter to the SMHM trend, particularly for low-mass
+satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
+tides on Local Group dwarf galaxies is essential for determining where
+and how these galaxies formed in a cosmological context.
 
@@
 determinations of small changes in a star's position over baselines of
-several years to decades. The full 6D position and velocity information
-for stars has, until recently, been known for only a handful of stars.
+years to decades. The full 6D position and velocity information for
+stars has, until recently, been known for only a handful of stars.
 
@@
 (RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
-BP-RP spectra and radial velocity measurements of bright stars (of
-magnitudes \(G_{\rm RVS} < 16\)) \citep{gaiacollaboration+2016}. For our
+BP-RP spectra and radial velocity measurements of bright stars \citep[of
+magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
 work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
@@
 
-The Local Group (LG) is defined as galaxies which are within \(\sim 1\)
-Mpc from the Milky Way-Andromeda system \citep[e.g.,][ and references
-therein]{mcconnachie2012}. \emph{Gaia} has revolutionized LG studies.
-For example, the 6D dynamical measurements and metallicities of MW stars
-led to the (re)discovery of past mergers or Milky Way building blocks
-like \emph{Gaia}-Sausage Enceladus
+\emph{Gaia} has revolutionized LG studies. For example, the 6D dynamical
+measurements and metallicities of MW stars led to the (re)discovery of
+past mergers or Milky Way building blocks like \emph{Gaia}-Sausage
+Enceladus
 \citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
-see also][]{mesa+2005}}, out-of-equilibrium structures like the
+see also][]{meza+2005}}, out-of-equilibrium structures like the
 \emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
 the Milky Way's spiral arms and the bar in the solar neighbourhood
-\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
-helped find and constrain numerous stellar streams
-\citep[\citet{bonaca+price-whelan2025}]{ibata+malhan+martin2019}.
+\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
+Way halo, \emph{Gaia} has helped find and constrain numerous stellar
+streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
 Altogether, \emph{Gaia} has revealed the hierarchical formation and
@@
 
-For Milky Way satellites, \emph{Gaia} has enabled well-constrained
-orbital analysis and facilitated precise stellar membership
-determination. Before, proper motions of dwarf galaxies were ill-known
-or undetermined. Few galaxies had precisely measured proper motions,
-more recently from Hubble Space Telescope observations
-\citep[e.g.,][]{sohn+2017}. \emph{Gaia} allowed the first systematic
-determinations of Milky Way satellite proper motions of unprecedented
-precision \citep{pace+li2019, MV2020a}. While the proper motion
-uncertainty of a typical dwarf member star is often large, by combining
-the proper motions of 100s or 1000s of stars from \emph{Gaia}, average
-precise proper motion measurements can be determined, sometimes only
-limited by \emph{Gaia}'s systematic error floor
-\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new
-dynamical era for MW satellite studies, where we can derive precise
+For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
+facilitated stellar membership determination. Before \emph{Gaia}, few
+galaxies had precisely measured proper motions, more recently from
+Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
+\emph{Gaia} allowed the first systematic and precise determinations of
+Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
+the proper motion uncertainty of a typical dwarf member star is often
+large, by combining the proper motions of 100s or 1000s of stars from
+\emph{Gaia}, precise average proper motion measurements can be
+determined, sometimes only limited by \emph{Gaia}'s systematic error
+floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
+new dynamical era for MW satellite studies, where we can derive precise
 orbits for any satellite, assuming a given MW potential. In addition,
@@
 background and foreground stars can be classified as non-members
-\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.
-
-\begin{figure}
-\centering
-\includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
-\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
-on the sky. We label the classical dwarf galaxies (green diamonds),
-fainter dwarfs (blue squares), globular clusters (orange circles), and
-ambiguous systems (pink open hexagons). Globular clusters are more
-centrally concentrated, but dwarf galaxies are preferentially found away
-from the MW disk. Sculptor and Ursa Minor are highlighted as two dwarfs
-we study later. The background image is from ESA/Gaia/DPAC
-(https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).
-Dwarf galaxies (confirmed), globular clusters, and ambiguous systems are
-from the \citet{pace2024} catalogue (version
-1.0.3).}\label{fig:mw_satellite_system}
-\end{figure}
+\citep[e.g.,][]{battaglia+2022, jensen+2024}.
 
@@
 
-Today, we know that the Milky Way system is teeming with satellites.
-Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
-including dwarf galaxies, globular clusters, and ambiguous systems,
-whose nature as a galaxy or star cluster remains uncertain. Advances in
-observational techniques has accelerated progress in our understanding
-of dwarf galaxies and introduced new questions. Deep digital photometric
-surveys have more than quadrupled the number of known Milky Way dwarf
-galaxies since 2005 \citep{simon2019}. Upcoming surveys, like the Vera
-Rubin Observatory's Legacy Survey for Space and Time
-\citep[LSST,][]{ivezic+2019}, will likely identify even fainter dwarf
-galaxies. In addition, large aperture multi-object spectrographs have
-revealed the detailed and complex inner chemodynamical structure of
-dwarf galaxies \textbf{REFS}. Beyond precise structural and kinematic
-properties of dwarf galaxies, modern observations allow for the
-separation of multiple stellar populations, detailed constraints on the
-dark matter density profiles, and hints of tidal disruption or stellar
-halos.
+Advances in observational techniques has accelerated progress in our
+understanding of dwarf galaxies and introduced new questions. Deep
+digital photometric surveys have more than quadrupled the number of
+known Milky Way dwarf galaxies since 2005 \citep{simon2019}. Upcoming
+surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
+Time \citep[LSST,][]{ivezic+2019}, will likely identify even fainter
+dwarf galaxies. In addition, large aperture multi-object spectrographs
+have revealed the detailed and complex inner chemodynamical structure of
+dwarf galaxies \citep[e.g.,][]{battaglia+2006, fabrizio+2016}. Beyond
+precise structural and kinematic properties of dwarf galaxies, modern
+observations allow for the separation of multiple stellar populations,
+detailed constraints on the dark matter density profiles, and hints of
+tidal disruption or stellar halos.
 
@@
 Observations of dwarf galaxies have been the origin of several disputes
-or \emph{small-scale} problems for
-\LCDM [see reviews by @bullock+boylan-kolchin2017; sales+2022]. For
-example, the mismatch between the expected number of dwarf galaxies from
-simulations and the observed abundance has been known as the
-\emph{missing satellites problem}. Additionally, a number of
-observations suggested that some dwarf galaxies, although not all,
-possess dark matter ``cores,''
+or \emph{small-scale} problems for \LCDM{} \citep[see reviews
+by][]{bullock+boylan-kolchin2017, sales+2022}. For example, the mismatch
+between the expected number of dwarf galaxies from simulations and the
+observed abundance has been known as the \emph{missing satellites
+problem}. Additionally, a number of observations suggested that some
+dwarf galaxies, although not all, possess dark matter ``cores,''
 \citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019},\footnote{In
@@
   between cores and cusps, whereas stellar kinematics is less
-  constraining.} contrary to the expectation from \LCDM of ``cuspy''
+  constraining.} contrary to the expectation from \LCDM{} of ``cuspy''
 inter dark matter profiles \citep{NFW1996, NFW1997}. As a result,
@@
 \citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
-Altogether, the numerous past challenges for \LCDM in the dwarf galaxy
+Altogether, the numerous past challenges for \LCDM{} in the dwarf galaxy
 regime illustrate the opportunity for dwarf galaxies to test the
@@
 can help interpret a galaxy's assembly and dynamical history
-\citep[e.g.,][, \citet{lee+2018}, \citet{querci+2025}]{penarrubia+2009}.
-We note that for resolved galaxies, these profiles are in terms of
-stellar count density instead of surface brightness.
+\citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. We note that
+for resolved galaxies, these profiles are in terms of stellar count
+density instead of surface brightness.
 
@@
 \Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s).
-}\end{equation}
-
-This profile is often applied to the radial light distribution of galaxy
-disks {[}\citet{freeman1970}; other classic paper?{]}.
+}\end{equation} This profile is often applied to the radial light
+distribution of galaxy disks
+\citep{devaucouleurs1959a, freeman1970, kent1985}.
 
@@
 \Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
-}\end{equation}
-
-where \(\Sigma_0\) is the central surface density and \(R_h\) is the
-projected half-light radius. Now mostly superseded by the King profile
-for globular clusters, the Plummer model is still a good fit to many
-dwarf spheroidals \citep[e.g.,][]{moskowitz+walker2020}.
+}\end{equation} where \(\Sigma_0\) is the central surface density and
+\(R_h\) is the projected half-light radius. Now mostly superseded by the
+King profile for globular clusters, the Plummer model is still a good
+fit to many dwarf spheroidals \citep[e.g.,][]{moskowitz+walker2020}.
 
@@
 \Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
-}\end{equation}
-
-In much of the older literature, \(R_t\) was often interpreted as a
-``tidal radius'', after an analogous interpretation for globular
-clusters \citep[e.g.,][\citet{hodge1961}]{IH1995}.
+}\end{equation} In much of the older literature, \(R_t\) was often
+interpreted as a ``tidal radius'', after an analogous interpretation for
+globular clusters \citep[e.g.,][]{IH1995, hodge1961}.
 
@@
 descriptions in many cases
-\citetext{\citealp[\citet{mateo1998}]{binggeli+sandage+tarenghi1984}; \citealp{mcconnachie+irwin2006}; \citealp{cicuendez+2018}}.
+\citep{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018}.
 As a result, it has become conventional to assume an exponential density
 profile to describe dwarf galaxies in theoretical or observational
-modelling \citetext{\citealp[e.g.,][
-\citet{MV2020a}]{martin+2016}; \citealp{battaglia+2022}; \citealp[ but
-is for disk]{kowalczyk+2013}}.
+modelling
+\citep[e.g.,][]{martin+2016, MV2020a, battaglia+2022, kowalczyk+2013}.
 
 Outside Local Group dwarf spheroidals, exponential profiles are fairly
-common, but sometimes with significant modifications. For example, many
-extragalactic dwarf elliptical and blue compact / Irr galaxies are fit
-well with an exponential + central cusp / nuclear region
-\citep[\citet{noeske+2003}]{caldwell+bothun1987}. In constrast, some
-studies find an inner decrement in density compared to exponentials
-\citep[e.g.,][]{caldwell+1992, makarov+2012}.
+common, but sometimes with modifications. For example, many
+extragalactic dwarf elliptical, blue compact, and irregular dwarf
+galaxies are better described with an exponential profile including a
+central cusp or nuclear region \citep{caldwell+bothun1987, noeske+2003}.
+Instead, some studies find an inner density decrement compared to
+exponentials \citep[e.g.,][]{caldwell+1992, makarov+2012}.
 
@@
 \citet{herrmann+hunter+elmegreen2013};
-\citet{herrmann+hunter+elmegreen2016}; \citet{lee+2018} note that at
-least in a photometric sample of more distant dwarf disky/irregular/blue
+\citet{herrmann+hunter+elmegreen2016}; and \citet{lee+2018} note that at
+least in a photometric sample of more distant disky, irregular, and blue
 compact dwarfs, many or most dwarfs are best fit with two nested
-exponential profiles. Dwarf irregular galaxies do represent a
-substantially different class though so it is unclear how these
-conclusions apply to dwarf spheroidals. \citet{caldwell+1992}'s
-photometric study of M31 dwarfs show that the outer density profile is
-typically well fit by an exponential, although deviations are evident in
-the inner regions.
+exponential profiles. It is unclear how these conclusions apply to dwarf
+spheroidals.
 
@@
 For most galaxies, they do not have convincing statistical evidence to
-prefer one model over another, but show that for 8 / 15 of dwarfs with
-sufficient data, a Plummer or steeper profile is (slightly) preferred.
-\textbf{This sentence is problematic}
+prefer one model over another, except for a few galaxies which are
+better described by a steep outer slope.
 
@@
 direction of a steeper outer cutoff or changes to the inner slope of a
-dwarf galaxy (due, for example, to a nuclear star cluster.). Flattened
+dwarf galaxy (due, for example, to a nuclear star cluster). Flattened
 density profiles in the outer regions are more unusual. Explaining the
-origin, similarity, and diversity of dwarf galaxy density profiles is a
-pressing question for theories of dwarf galaxy formation and evolution.
-Whether dwarf galaxies are indeed exponential or not, the diversity of
-density profile shapes of dwarf galaxies needs to be explained.
+origin, similarity, and diversity of dwarf galaxy density profiles is an
+open question for theories of dwarf galaxy formation and evolution.
 
@@
 Sculptor and Ursa Minor may have been influenced by the Milky Way's
-tidal field (see Section~\ref{sec:discussion}). \citet{sestito+2023a};
-\citet{sestito+2023b} report a ``kink'' in the density profile,
-beginning around 30 arcmin for both Sculptor and Ursa Minor, that they
-interpret as caused by the effects of Galactic tides. They
+tidal field (see Section~\ref{sec:discussion}).
+\citet{sestito+2023a, sestito+2023b} report a ``kink'' in the density
+profile, beginning around 30 arcmin for both Sculptor and Ursa Minor,
+that they interpret as caused by the effects of Galactic tides. They
 spectroscopically follow up some of the most distant stars, finding
-multiple members between 6-12 half-light radii from the centre of each
-dwarf. If dwarfs had exponential profiles, like Fornax, then these stars
-should be much rarer.
+multiple members between 6 and 12 half-light radii from the centre of
+each dwarf. If these dwarfs had exponential profiles, like Fornax, then
+these stars should be much rarer.
 
 Sculptor and Ursa Minor are not well-described by an exponential
-profile. The left panel of Fig.~\ref{fig:scl_umi_vs_penarrubia} shows
-the density profiles of Sculptor, Ursa Minor, and Fornax (see
+profile. The left panel of Fig.~\ref{fig:scl_umi_vs_fornax} shows the
+density profiles of Sculptor, Ursa Minor, and Fornax (see
 Section~\ref{sec:observations} for details on how these profiles are
 measured). Compared to Fornax, both Sculptor and Ursa Minor show an
-excess of stars beginning around \(\log R/R_h\approx 0.4\). The right
-panel of Fig.~\ref{fig:scl_umi_vs_penarrubia} compares the same density
-profiles to an initial and final density profile from a simple tidal
-model of an exponential dwarf spheroidal galaxy embedded in a NFW halo
-in the Milky Way field, described in Section~\ref{sec:tidal_theory}
-(\textbf{Move to later or describe?}). Fornax agrees well with the
-exponential initial conditions. However, Sculptor and Ursa Minor are
-better described by the more extended, tidally evolved density profile.
+excess of stars beginning after \(\log R/R_h\approx 0.4\) and possibly
+exceeding 100 times the density of the exponential fit at large radii.
 
-A goal of this work is to determine, assuming \LCDM, if the effects of
+A goal of this work is to determine, assuming \LCDM{}, if the effects of
 the Milky Way (or other satellites) are indeed responsible for the
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_vs_penarrubia.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_vs_fornax.pdf}}
 \caption[Sculptor and Ursa Minor match tidal models]{A plot of the
-surface density profiles of Sculptor, Ursa Minor, and Fornax scaled to
-their half-light radius and the density at half-light radius (data
-described in Section~\ref{sec:observations}). The right panel adds an
-idealized model with initial, final, and break radius as a dotted line,
-solid line and an arrow respectively see \ref{sec:break_radii}.
-\textbf{Move RHS to later\ldots{}}}\label{fig:scl_umi_vs_penarrubia}
+surface density profiles of Sculptor (orange squares), Ursa Minor (red
+triangles), and Fornax (green circles) scaled to their half-light radius
+and the density at half-light radius (data described in
+Section~\ref{sec:observations}). The solid black line is an exponential
+profile (Eq.~\ref{eq:exponential_law}).}\label{fig:scl_umi_vs_fornax}
 \end{figure}
@@
 Simulating dwarf galaxies accurately in a cosmological context remains a
-substantial challenge. As discussed above, cosmological simulations can
-predict the overall abundance of the most massive dwarf galaxies
-{[}e.g., \textbf{references}{]} and broadly examine the effects of tides
-\citep[e.g.,][]{riley+2024}. However, dwarf galaxies are often near the
-resolution limit. Insufficient resolution can lead to artificial
-disruption of dwarf galaxies and overestimation of tidal effects
-\citep[e.g.,][]{santos-santos+2025}. To address this challenge,
-idealized simulations are often used to simulate a single subhalo in an
-approximate host potential. As a result, idealized simulations can reach
-excellent numerical convergence For instance, our simulations we
-describe later reach 3x higher resolution than Aquarius
-\citep{springel+2008} at a fraction of the computational cost (400x
-fewer particles). Idealized simulations make numerous simplifications:
-neglecting mergers, cosmological context and evolution, mass assembly,
-and often baryonic physics. We use idealized simulations here to assess
-tidal effects after infall.
+challenge. Cosmological simulations can predict the overall abundance of
+more massive dwarf galaxies \citep[e.g.,][]{sawala+2016} and broadly
+examine the effects of tides \citep[e.g.,][]{riley+2024}. However, dwarf
+galaxies are often near the resolution limit. Insufficient resolution
+can lead to artificial disruption of dwarf galaxies and overprediction
+of tidal effects \citep[e.g.,][]{santos-santos+2025}. To address this
+challenge, idealized simulations are often used to simulate a single
+subhalo in an approximate host potential, acchieving excellent numerical
+convergence. For instance, our simulations we describe later reach 3
+times higher resolution than Aquarius \citep{springel+2008} at a
+fraction of the computational cost (400 times fewer particles).
+Idealized simulations make numerous simplifications, neglecting mergers,
+cosmological context, mass assembly, and often baryonic physics. We use
+idealized simulations here to assess tidal effects after infall.
 
@@
 \citet{piatek+pryor1995}; \citet{moore+davis1994};
-\citet{johnston+spergel+hernquist1995}. Already, these works used
-techniques similar to what we continue to use today and laid the
-foundation for understanding tidal effects. While the detailed
-assumptions have evolved (e.g.~no longer assuming mass-follows-light),
-many of their conclusions still hold true. More recent work expanding
-this theory include \citet{read+2006}; \citet{bullock+johnston2005};
-\citet{PNM2008}; \citet{penarrubia+2009}; \citet{klimentowski+2009};
-\citet{errani+2023a}; \citet{fattahi+2018}; \citet{stucker+2023};
-\citet{wang+2017}.
+\citet{johnston+spergel+hernquist1995}. Already, using similar numerical
+techniques as we do later on, these works described the evolution of
+stellar and dark matter mass loss, the formation of symmetric tidal
+streams, and the minor changes to the central structure.
 
 With precise orbits and a better understanding of the Milky Way
-potential and system, more recent work began to directly probe the
-dynamical histories of individual dwarf galaxies (although early work
-began this for Sagittarius / etc). Examples include \citet{iorio+2019}
-for Sculptor, \citet{borukhovetskaya+2022}; \citet{dicintio+2024} for
-Fornax; \citet{borukhovetskaya+2022a} for Antlia II. Our goal is to
-apply a similar framework to Sculptor and Ursa Minor.
-
-*\textbf{Finish last two paragraphs}
-
-\subsection{A simple tidal simulation}\label{a-simple-tidal-simulation}
-
-To illustrate the effects of tides on an NFW halo in a Milky-Way like
-potential, we consider a toy model. See also \citet{PNM2008}?.
-
-We evolve an NFW profile in a logarithmic host potential along an orbit
-where the pericentre to apocentre is 1:50.
-
-As illustrated in Fig.~\ref{fig:idealized_break_radius}, for an
-idealized model with exponential stars in a NFW halo, shortly after
-pericentre, the stellar component is smooth but contains a change in
-slope around \(r_{\rm break}\). This radius is visible in the stellar
-distribution as non-spherical S-shaped overdensities of stars. Also,
-\(r_{\rm break}\) is where where the mean radial velocities of the stars
-becomes positive, i.e.~the system is out of equilibrium and adjusting to
-the new density profile.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
-\caption[Break radius validation]{Example density and velocity
-distributions of an idealized simulation shortly after pericentre.
-\textbf{Top left}: The 2D density profile for the initial and final
-simulation with the break radius marked. The break radius of the
-simulations is set by the time since pericentre. \textbf{Top right}: The
-projected 2D stellar density in the \(x\)-\(y\) plane. The green circle
-represents the break radius and the grey arrow points towards the host
-centre. \textbf{Bottom left}: the mean radial velocity (dot product of
-relative position and velocity relative to dwarf centre) as a function
-of 2D radius. \textbf{Bottom right}: The mean radial velocity as
-projected into 2D bins. The initial conditions are Sculptor-like,
-exponential stars embedded in NFW, evolved in \citet{EP2020} potential
-with a pericentre of 10 kpc and apocentre of 100 kpc shortly after the
-3rd pericentre. See Section~\ref{sec:methods} for a description of our
-simulation setup.}\label{fig:idealized_break_radius}
-\end{figure}
+potential and system, recent work continues to directly probe the
+dynamical histories of individual dwarf galaxies.
+\citet{borukhovetskaya+2022, dicintio+2024} both ran simulations tuned
+to Fornax, showing this galaxy's stellar component or globular clusters
+are likely not affected by tides. \citet{borukhovetskaya+2022a} used
+N-body simulations to analyze Crater II, showing that the present-day
+properties of the galaxy are challenging to reconcile with \LCDM{}
+initial conditions with tidal evolution. Most relevantly,
+\citet{iorio+2019} conducted simulations of Sculptor, concluding tides
+likely do not affect this galaxy. Our goal is to apply a similar
+framework to Sculptor and Ursa Minor.
 
@@
 the dwarf galaxy is roughly three times the mean interior density of the
-host galaxy at pericentre, or \begin{equation}{
+host galaxy at pericentre, or
+\begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
 3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J).
@@
 
+\subsection{A simple tidal simulation}\label{a-simple-tidal-simulation}
+
+To illustrate the effects of tides on an NFW halo due to a larger
+galaxy, we consider a toy model. We evolve an NFW subhalo with
+\(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) in an NFW static host halo with
+\(\rmax=25\,\kpc\) and \(\vmax = 207.4\,\kms\). The stars are initially
+an exponential profile with a scale radius \(R_s=0.25\,\kpc\). We then
+evolve the orbit through a pericentre of 10 kpc starting from an
+apocentre of 100 kpc. We show the results 170 Myr after the first
+pericentre. See Section~\ref{sec:methods} for a more detailed
+description of our general simulation setup.
+
+Fig.~\ref{fig:idealized_break_radius} illustrates the properties of this
+idealized simulation shortly after the first pericentre. The projected
+density of stars (right) is relatively undisturbed and spherical in the
+centre, but becomes non-isotropic outside the break radius and shows
+nascent tidal tails. As a surface density profile (top left), these
+tidally disturbed stars appear as an extended, outer density excess
+relative to the initial conditions. This excess appears shortly after
+the break radius. The break radius also marks where the mean radial
+velocity of the stars transitions from 0 to positive --- the galaxy is
+out of equilibrium outside \(r_{\rm break}\).
+
+To first order, the final density profile of this toy simulation indeed
+resembles Sculptor and Ursa Minor outer density excess. This thesis aims
+to investigate if a tidal explanation is permissible given the current
+properties of each galaxy.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
+\caption[Break radius validation]{Example density and velocity
+distributions of an idealized simulation shortly after pericentre.
+\textbf{Top left}: The 2D density profile for the initial and final
+simulation with the break radius marked. The break radius of the
+simulations is set by the time since pericentre. \textbf{Bottom left}:
+the mean radial velocity (dot product of relative position and velocity
+relative to dwarf centre) as a function of 2D radius. \textbf{Right}:
+The projected 2D stellar density in the \(x\)-\(y\) plane. The green
+circle represents the break radius and the grey arrow points towards the
+host centre.}\label{fig:idealized_break_radius}
+\end{figure}
+
 \section{Thesis outline}\label{thesis-outline}

### 2025-08-31 thesis/main.tex
thesis julio ch1-3 edits

@@
 
+The Digitized Sky Surveys were produced at the Space Telescope Science Institute under U.S.
+Government grant NAG W-2166.
+The images of these surveys are based on photographic data obtained using the Oschin Schmidt Telescope on Palomar Mountain and the UK Schmidt Telescope.
+The plates were processed into the present compressed digital form with the permission of these institutions.
+
+The National Geographic Society --- Palomar Observatory Sky Atlas (POSS-I) was made by the California Institute of Technology with grants from the National Geographic Society.
+
+The Second Palomar Observatory Sky Survey (POSS-II) was made by the California Institute of Technology with funds from the National Science Foundation, the National Geographic Society, the Sloan Foundation, the Samuel Oschin Foundation, and the Eastman Kodak Corporation.
+
+The Oschin Schmidt Telescope is operated by the California Institute of Technology and Palomar Observatory.
+
+The UK Schmidt Telescope was operated by the Royal Observatory Edinburgh, with funding from the UK Science and Engineering Research Council (later the UK Particle Physics and Astronomy Research Council), until 1988 June, and thereafter by the Anglo-Australian Observatory.
+The blue plates of the southern Sky Atlas and its Equatorial Extension (together known as the SERC-J), as well as the Equatorial Red (ER), and the Second Epoch [red] Survey (SES) were all taken with the UK Schmidt.
+
+All data are subject to the copyright given in the copyright summary. Copyright information specific to individual plates is provided in the downloaded FITS headers.
+
+Supplemental funding for sky-survey work at the ST ScI is provided by the European Southern Observatory.
+
 %% Lists
@@
 \chapter{The Effects of Tides}\label{sec:results}
-\input{sculptor}
+\input{results}
 

### 2025-08-31 thesis/methods.tex
thesis julio ch1-3 edits

--- 2025-08-18 thesis/methods.tex
@@
-In this section, we discuss our methods used for the following two
-chapters. We utilize an analytic, static, axisymmetric Milky Way
-potential based on \citet{mcmillan2011}. Our N-body simulation assume
-collisionless, cold dark matter and are integrated with Gadget-4. We
-check for numerical convergence and stability of initial conditions in
-isolation.
+In the previous chapters, we have seen that the Scl and UMi classical
+dSphs have outer density profiles that appear to deviate from the
+exponential law that approximates well all other classical dSphs. Our
+main intention is to assess whether such excesses result from the
+effects of Galactic tides. To this purpose, we intend to use N-body
+simulations of the evolution of CDM halos in a Galactic potential,
+constrained to have the orbital parameters consistent with a dwarf's
+present-day position and velocity. We shall assume that the Galactic
+potential is the static, analytic potential inferred by
+\citet{mcmillan2011} from observations of kinematic tracers. We also
+assume the potential of each dwarf may be initially approximated by a
+cuspy NFW profile. Since the dwarfs in question are heavily dark matter
+dominated, we shall use a carefully selected sample of dark matter
+particles to mimic and track the evolution of an embedded tracer stellar
+component. In this Chapter, I describe the Galactic potential used, the
+orbital estimation method, the initial conditions setup, and the N-body
+method used.
 
-\section{Milky Way potential}\label{milky-way-potential}
+\section{Orbital estimation}\label{orbital-estimation}
 
-We adopt the Milky Way potential described in \citet{EP2020}, an
-analytic approximation of \citet{mcmillan2011}.
+To explore the possible orbits of a dwarf galaxy, we perform a Monte
+Carlo sampling of the present-day observables. The present-day position,
+distance modulus, LOS velocity, and proper motions are each sampled from
+normal distributions given the reported uncertainties in
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
+each sampled position/velocity back in time for 10 Gyr using \agama{}
+\citep{agama}. Dynamical friction is not expected to impact orbits
+substantially because of the low masses and large pericentres of the
+dwarfs.
+
+\subsection{Galactocentric frame}\label{galactocentric-frame}
+
+To convert observed positions and velocities to Galactocentric
+coordinates, we use the Astropy v4 Galactocentric frame
+\citep{astropycollaboration+2022}. This frame assumes the Galactic
+centre is at position \(\alpha = {\rm 17h\,45m\,37.224s}\),
+\(\delta = -28^\circ\,56'\,10.23''\) with proper motions
+\(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
+\(\mu_\delta=-5.547\pm0.026 \masyr\) \citep[from the appendix and Table
+2 of][]{reid+brunthaler2004}. The Galactic centre is at a distance from
+the Sun of \(8.122\pm0.033\,\)kpc with a radial velocity =
+\(11 + 1.9 \pm 3\,\kms\) \citep{gravitycollaboration+2018}. The Sun is
+assumed to be \(20.8\pm0.3\,\)pc above the disk
+\citep{bennett+bovy2019}. Using the procedure outlined in
+\citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
+rest frame is then
+\(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
+uncertainties in the reference frame are typically smaller than the
+uncertainties on a dwarf galaxy's position and velocity.
+
+\subsection{Milky Way Potential}\label{milky-way-potential}
+
+We adopt the Milky Way potential described in \citet{EP2020}, which is
+an analytic approximation to that proposed by \citet{mcmillan2011}.
 Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
 each component and the total circular velocity profile for our fiducial
-profile. The potential consisting of a stellar bulge, a thin and thick
-disk, and a dark matter NFW halo. Among typical orbits for Sculptor and
-Ursa Minor, the dark matter halo dominates the potential.
+profile. The potential includes a stellar bulge, a thin and thick disk,
+and a dark matter NFW halo.
 
-The potential is specified as follows. The galactic bulge is described
-by a \citet{hernquist1990} potential,
+The Galactic bulge is described by a \citet{hernquist1990} potential,
 
-\begin{equation}\protect\phantomsection\label{eq:hernquist}{
+\begin{equation}{
 \Phi(r) = - \frac{GM}{r + a},
@@
 disks are represented with the \citet{miyamoto+nagai1975} cylindrical
-potential:
-
-\begin{equation}\protect\phantomsection\label{eq:mn75}{
-\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
-}\end{equation}
-
-where \(a\) is the disc radial scale length, \(b\) is the scale height,
-and \(M\) is the total mass of the disk. For the thin disk,
-\(a=3.944\,\)kpc, \(b=0.311\,\)kpc,
+potential, \begin{equation}{
+\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}},
+}\end{equation} where \(a\) is the disc radial scale length, \(b\) is
+the scale height, and \(M\) is the total mass of the disk. For the thin
+disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
 \(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
 \(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
-The halo is a NFW dark matter halo (Eq.~\ref{eq:nfw}) with
-\(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.
+The halo is an NFW dark matter halo (Eq.~\ref{eq:nfw}) with
+\(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\).
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
 \caption[Circular velocity of potential]{Circular velocity profile of
-\citet{EP2020} potential.}\label{fig:v_circ_potential}
+\citet{EP2020} potential. The total circular velocity (thick black line)
+is composed of an NFW halo (green dashed line), a think and thick
+\citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
+\citet{hernquist1990} bulge (blue dotted
+line).}\label{fig:v_circ_potential}
 \end{figure}
 
-\section{Orbital estimation}\label{orbital-estimation}
+\subsection{Orbits of Sculptor}\label{orbits-of-sculptor}
 
-To estimate the possible orbits of a dwarf galaxy, we perform a Monte
-Carlo sampling of the present-day observables. The present-day position,
-distance modulus, LOS velocity, and proper motions are each sampled from
-normal distributions given the reported uncertainties in
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
-each sampled observable back in time for 10 Gyr in Gadget as massless
-point particles outputting positions every 5Myr (with otherwise similar
-parameters to n-body runs below.) When selecting the initial position
-and velocity of an N-body model, we select the position and velocity of
-the first apocentre occurring after 10 Gyr ago.
+Sculptor's orbital history is relatively well-constrained.
+Fig.~\ref{fig:scl_orbits} illustrates point particle orbits for 100
+samples of Sculptor's observed kinematics integrated backwards 5 Gyr in
+both Galactocentric coordinate slices (\(x\), \(y\), \(z\)) and in
+Galactocentric radius with time. All sampled orbits of Sculptor have
+nearly the same morphology---the orbit primarily resides in the
+\(y\)--\(z\) plane and has a similar number of periods and
+pericentre/apocentre.
 
-We assume the galaxy is represented as a point particle for this
-analysis. In detail, (self) dynamical friction likely influences the
-orbit, but a point particle is an adequate description.
+To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
+smallest pericentre among all possible orbits. We achieve this by taking
+the median parameters of all orbits with a pericentre less than the
+0.0027th quantile pericentre, yielding a pericentre of 43 kpc. Given the
+current observations, it is unlikely that Sculptor has a significantly
+smaller pericentre than our selected orbit.
 
-To convert from Gaia to galactocentric coordinates, we use the Astropy
-v4 galactocentric frame \citep{astropycollaboration+2022}. This frame
-assumes the galactic centre appears at position
-\(\alpha = {\rm 17h45m37.224s}\), \(\delta = -28°56'10.23''\) with
-proper motions \(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
-\(\mu_\delta=-5.547\pm0.026 \masyr\) (from the appendix and Table 2 of
-\citet{reid+brunthaler2004}). The galactic centre distance is
-\(8.122\pm0.033\,\)kpc with a radial velocity = \(11 + 1.9 \pm 3\) km/s.
-from \citet{gravitycollaboration+2018}. Finally, adding that the sun is
-\(20.8\pm0.3\,\)pc above the disk from \citet{bennett+bovy2019}, and
-using the procedure outlined in \citet{drimmel+poggio2018}, the solar
-velocity relative to the galactic rest frame is `v\_sun' =
-\([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
-uncertainties in the reference frame are typically smaller than the
-uncertainties on the dwarf galaxy's position and velocity.
+We take the first apocentre after a look-back time of 10 Gyr, or at 9.43
+Gyr, as the initial conditions for our model of Sculptor, noted in
+Table~\ref{tbl:scl_orbits}.
 
-\section{N-Body modelling}\label{n-body-modelling}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
+\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
+potential in Galactocentric \(x\), \(y\), and \(z\) coordinates (top)
+and in Galactocentric radius \(r\) versus time (bottom). The Milky Way
+is at the centre with the disk lying in the \(x\)--\(y\) plane. Our
+selected \texttt{smallperi} orbit is plotted in black and light blue
+transparent orbits represent the past 5Gyr orbits sampled over Sculptor
+observables in Table~\ref{tbl:scl_obs_props}. The orbit of sculptor is
+well-constrained in this potential and it is unlikely to achieve a
+smaller pericentre than our selected orbit.}\label{fig:scl_orbits}
+\end{figure}
 
-\subsection{Initial conditions}\label{initial-conditions}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_xyzr_orbits.pdf}}
+\caption[Ursa Minor Orbits]{Similar to Fig.~\ref{fig:scl_orbits}, the
+orbits of Ursa Minor in a static Milky Way potential in Galactocentric
+\(x\), \(y\), and \(z\) coordinates. In the lower panel, we show the
+radius versus time for only three orbits of Ursa
+Minor.}\label{fig:umi_orbits}
+\end{figure}
 
-We use Agama \citep{agama} to generate initial conditions. We initially
-assume galaxies are described by an NFW dark matter potential
-Eq.~\ref{eq:nfw} and the stars are merely collisionless tracers embedded
-in this potential (added on in post-processing). The density is a
-cubic-exponentially truncated with a profile
-\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
-\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t)
-}\end{equation} where we adopt \(r_t = 20 r_s\) or approximately
-\(r_{200}\) for our Sculptor-like fiducial halo. Using \(r_{200}\) for
-\(r_t\) would depend on the chosen scale of the halo. So our adopted
-\(r_t\) is an approximate upper limit of \(r_{200}\) for typical dwarf
-galaxy halos \citep{ludlow+2016}. It is unlikely that the outer density
-profile of loosely bound particles past \(20\) kpc affects the tidal
-evolution of a subhalo.
+\begin{table*}[t]
+\centering
+\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
+The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
+The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+}
+\label{tbl:scl_orbits}
+\begin{tabular}{lll}
+\toprule
+Property & Mean & SmallPeri\\
+\midrule
+distance / kpc & 83.2 & 82.6\\
+$\pmra / \masyr$ & 0.099 & 0.134\\
+$\pmdec / \masyr$ & -0.160 & -0.198\\
+LOS velocity / $\kms$ & 111.2 & 111.2\\
+$t_i / \Gyr$ & -8.74 & -9.43\\
+$\vec{x}_{i} / \kpc$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10]\\
+$\vec{v}_i / \kms$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29]\\
+pericentre / kpc & 53 & 43\\
+apocentre / kpc & 102 & 96\\
+$t_{\rm last\ peri} / {\rm Gyr}$ & -0.45 & -0.46\\
+Number of pericentres & 6 & 6\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Orbits of Ursa Minor}\label{orbits-of-ursa-minor}
+
+Similar to Sculptor, Ursa Minor has a well-constrained orbit.
+Fig.~\ref{fig:umi_orbits} shows 100 random point-orbits of Ursa Minor.
+Initially, we select an orbit with approximately the \(3\sigma\)
+smallest pericentre as in Sculptor.
+
+Ursa Minor's N-body orbit diverges from the point particle orbit. To
+ensure the final conditions of the N-body simulation are close to the
+intended final position, we iteratively adjust Ursa Minor's initial
+conditions. Initially, starting with low-resolution runs, we adjust the
+cylindrical actions of the initial orbit by the final difference in
+actions at the end of orbital evolution. After the initial actions have
+converged (2 iterations), we change the initial action angles by the
+final difference in action angles. This method converges within 4
+iterations to an orbit agreeing with the observed kinematics of Ursa
+Minor.
+
+\begin{table*}[t]
+\centering
+\caption[Ursa Minor Selected Orbits]{Properties of selected orbits for Ursa Minor.
+The "smallperi" orbit is the initial point orbit and the "smallperi.5" is the initial orbit for the N-body simulation.
+}
+\label{tbl:umi_orbits}
+\begin{tabular}{llll}
+\toprule
+Property & Mean & SmallPeri.1 & SmallPeri.5\\
+\midrule
+distance / kpc & $70.1$ & 64.6 & \\
+$\pmra / \masyr$ & $-0.124 \pm 0.17$ & -0.158 & \\
+$\pmdec / \masyr$ & $0.078\pm0.17$ & 0.05 & \\
+$v_{\rm LOS} / \kms$ & $-245.9\pm 1$ & -245.75 & \\
+$t_i / \Gyr$ & -8.74 & -9.53 & -9.53\\
+$\vec{x}_{i} / \kpc$ & [4.88, -65.11, 50.78] & [-16.48 69.92 21.05] & [17.40, 74.51, 21.34]\\
+$\vec{v}_i / \kms$ & [-34.28, 77.11, 101.38] & [16.32, 39.86, -116.99] & [14.27, 48.62, -114.08]\\
+pericentre / kpc & 37 & 29 & 28\\
+apocentre / kpc & 83 & 75 & 72\\
+$t_{\rm last\ peri} / \Gyr$ & $-0.96$ & $-0.80$ & $-0.81$\\
+Number of pericentres & 9 & 6 & 6\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\section{Initial conditions}\label{initial-conditions}
+
+We use \agama{} \citep{agama} to generate the initial N-body dark matter
+halo. We assume galaxies are described by an NFW dark matter potential
+(Eq.~\ref{eq:nfw}). We also assume the stars do not contribute to the
+potential. The dark matter density is truncated in the outer regions by
+\begin{equation}{
+\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t),
+}\end{equation} where we adopt \(r_t = 20 r_s\).
+
+\subsection{Initial dark matter halos for Sculptor and Ursa
+Minor}\label{initial-dark-matter-halos-for-sculptor-and-ursa-minor}
+
+From the observed properties of Sculptor and Ursa Minor, we infer
+reasonable \LCDM{} initial halo conditions.
+
+Table~\ref{tbl:derived_props} reports our inferred halo and kinematic
+properties of Sculptor and Ursa Minor. First, taking the absolute
+magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
+\citet{woo+courteau+dekel2008} (with \(\sim\) 0.17 dex uncertainty), the
+total current stellar mass of Sculptor and Ursa Minor are
+\(M_\star \sim 3.1 \times 10^6 \Mo\) and
+\(M_\star \sim 7 \times 10^5 \Mo\). Based on the stellar mass-\(\vmax\)
+relation \citep[from][]{fattahi+2018}, see also \ref{fig:smhm}, Sculptor
+and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
+\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
+\(z=0\) mass-concentration relation, this constraint translates into a
+\(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
+galaxy.
+
+Fig.~\ref{fig:scl_halos} illustrates these estimates visually for both
+galaxies. The stellar-mass \(\vmax\) constraint translates to an
+estimate of \(\vmax\) only (horizontal band). The mass-concentration
+relation describes the relationship between \(\vmax\) and \(\rmax\)
+(diagonal linear band). And finally, we include a curved line which
+illustrates the halo \(\vmax\) which has a specified initial LOS
+velocity dispersion given \(\rmax\)
+\(, \vcirc(R_h) / \sqrt{3} \approx \sigma_v\).
+\(R_h \approx 0.24\,\kpc\) for both galaxies.
+
+While there is some range in the choice of initial halo, reasonable
+changes to the initial halo do not substantially affect the tidal
+evolution for either galaxy. The observed velocity dispersion is
+directly related to the mass within a half-light radius
+\citep[e.g.,][]{wolf+2010}. As a result, halos with the same velocity
+dispersion may differ in total mass but should have similar mass within
+\(R_h\). So, the tidal effects on stars should be similar for halos with
+similar velocity dispersions.
+
+Our selected halos (in Table~\ref{tbl:initial_halos}) for each
+simulation run are based on the cosmological constraints and the need to
+match the present-day velocity dispersion at the end of the simulation.
+
+\begin{table*}[t]
+\centering
+\caption[Derived Properties of Sculptor and Ursa Minor]{Inferred properties of the stellar component and halo for Sculptor and Ursa Minor.
+We record the total luminosity, stellar mass, mass-to-light ratio, dark matter halo $\vmax$ and $\rmax$, and dark matter halo virial mass $M_{200}$ and concentration $c_{\rm NFW}$.
+}
+\label{tbl:derived_props}
+\begin{tabular}{lll}
+\toprule
+parameter & Sculptor & Ursa Minor\\
+\midrule
+$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & $3.5 \pm 0.1 \times 10^5\,L_\odot$\\
+$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & $7_{-2}^{+3} \times 10^5\,\Mo$\\
+$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & $1.9 \times 10^{\pm 0.17}$\\
+$\vmax$ & $31\pm 3\,\kms$ & $27_{-6}^{+7}\,\kms$\\
+$\rmax$ & $6 \pm 2$ kpc & $5_{-2}^{+1}$ kpc\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $3_{-2}^{+4} \times 10^9\,\Mo$\\
+$c_{\rm NFW}$ & $13_{-3}^{+4}$ & 14?\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_halo_selection.pdf}}
+\caption[Sculptor initial halos]{Selection of initial halos for Sculptor
+and Ursa Minor. The grey line and pink line with shaded regions
+represent the \citet{ludlow+2016} mass-concentration relation and
+\citet{fattahi+2018} SMHM relation respectively. The curved lines
+represent the velocity dispersion of the initial halo given the
+present-day half-light radius \citep[via the][ mass
+estimator]{wolf+2010}.}\label{fig:scl_halos}
+\end{figure}
+
+\begin{table*}[t]
+\centering
+\caption[Initial halos]{The initial conditions for our initial dark matter halos. }
+\label{tbl:initial_halos}
+\begin{tabular}{lllll}
+\toprule
+Halo name & $\rmax$ & $\vmax$ & $M_{200}$ & $c_{\rm NFW}$\\
+\midrule
+Scl: fiducial & 3.2 & 31 & 0.33 & 21\\
+Scl: small & 2.5 & 25 &  & \\
+UMi: fiducial & 4 & 38 & 0.62 & 21\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\section{Numerical methods}\label{numerical-methods}
+
+To simulate the tidal evolution of galaxies, we use ``N-body''
+simulations integrated with the parallel, gravitational-tree program
+\gadget{} \citep{gadget4}. The N-body method calculates and evolves the
+gravitational accelerations between a large number of collisionless
+particles to approximate the dynamical evolution of matter. To
+approximate a collisionless system (i.e., without strong
+particle-particle gravitational deflections), the force is softened
+below a ``softening length.'' Regions which are smaller than the
+softening length, or have few particles, are not well-resolved. Tree
+codes organize particles into a spatial tree, enabling grouping of the
+gravitational forces from nearby particles. A gravitational tree code
+substantially reduces the required number of force computations, as
+compared to the exact, direct-summation method.
 
@@
 
-To ensure that the initial conditionss of the simulation are dynamically
-relaxed and well-converged, we run the simulation in isolation (no
-external potential) for 5 Gyr (or about 3 times the crossing timescale
-at the virial radius). Our fiducial isolation halo uses \(r_s=2.76\) kpc
-and \(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for
-any length or mass scale.
+To ensure that the initial conditions of the simulation are dynamically
+relaxed and well-converged, we run a halo first in isolation using
+\gadget{}. Since gravity is scale-free, we use the same isolation run
+for all halos. We adopt \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)
+for the isolation halo based on Sculptor's mean properties. We run this
+model for 5 Gyr (about three times the free fall timescale
+\(t_{\rm ff} = \vcirc / r \approx 1.5\,\Gyr\) at \(r_{200}=36\,\)kpc).
 
 For our simulation parameters, we adopt a softening length of
-\begin{equation}{
-h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-}\end{equation} See appendix Section~\ref{sec:extra_convergence} for a
+\begin{equation}\protect\phantomsection\label{eq:softening_length}{
+h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
+}\end{equation} See Appendix Section~\ref{sec:extra_convergence} for a
 discussion of this choice, which is similar to the \citet{power+2003}
@@
 
-\subsubsection{Numerical convergence}\label{numerical-convergence}
+\subsection{Numerical fidelity}\label{numerical-fidelity}
 
-As discussed in \citet{power+2003}, the region where density is
-converged is related to the region which becomes collisionally relaxed
-over the time of the universe (so about 5-10 Gyr). (i.e.~where the
-collisionless assumption breaks down). The relaxation timescale is given
-by \begin{equation}\protect\phantomsection\label{eq:t_relax}{
-t_{\rm relax}(r) = t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
-= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
-}\end{equation} where \(N(r)\) is the number of particles within \(r\),
-\(t_{\rm circ}\) is the circular velocity timescale, and \(\bar \rho\)
-is the mean interior density \citep{power+2003}. The converged radius
-\(r_{\rm relax}(t)\) is the radius above which \(t_{\rm relax} > t\).
-Typically, \(r_{\rm relax}(10\Gyr)\) is about 6-10 times our adopted
-softening length, increasing with particle number. As such, at full
-resolution, we can only trust density profiles down to \(\sim10\) times
-our softening length, sufficient to resolve stellar density profiles.
+Fig.~\ref{fig:numerical_convergence} illustrates how well our numerical
+setup is able to reproduce the desired initial conditions, before and
+after running the model in isolation. This figure shows that our
+numerical methods are able to approximate well an NFW halo down to an
+innermost radius that strongly depends on resolution. The larger the
+number of particles, the smaller the radius that is effectively
+``resolved'' in a given simulation. For the Sculptor halo shown in this
+figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\) ), a
+simulation with \(10^7\) particles is needed to resolve the innermost
+100 pc. For reference, the half-light radius of Sculptor is roughly 100
+pc, which means that at least 10 million particles would be needed to
+follow faithfully its tidal evolution. Vertical arrows in
+Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
+defined by \citet[eq.\textasciitilde13]{power+2003} for NFW halos formed
+in cosmological N-body simulations. This radius marks the region where
+collisional effects driven by the finite number of particles used to
+describe the innermost regions of a halo become important. The softening
+length (from Eq.~\ref{eq:softening_length}) is typically a few times
+smaller than the converged length.
 
@@
 different total numbers of particles in isolation. Residuals in bottom
-panel are relative to NFW. The initial conditions are dotted and the
-converged radius is marked by arrows (Eq.~\ref{eq:t_relax}). Note that a
-slight reduction in density starting around \$r = 30 \$kpc is expected
-given our truncation choice.}\label{fig:numerical_convergance}
+panel are relative to NFW. The initial conditions are dotted, the
+converged radius is marked by arrows \citep[from][ eq. 13]{power+2003},
+and the softening length is marked by a vertical bar. Note that a slight
+reduction in density starting around \(r = 30\,\kpc\) is expected given
+our truncation choice.}\label{fig:numerical_convergence}
 \end{figure}
 
-\subsection{Orbital runs}\label{orbital-runs}
+\subsection{Orbital evolution}\label{orbital-evolution}
 
-To perform the simulations of a given galaxy in a given potential, we
-centre the isolation run's final snapshot
-(Section~\ref{sec:shrinking_spheres}) and place the dwarf galaxy in the
-specified orbit in the given potential. We typically run the simulation
-for 10 Gyr, which allows us to orbit slightly past the expected initial
-conditions.
+Next, we evolve the halo in the Galactic potential. We scale the relaxed
+snapshot and softening length to match the initial halo in
+\ref{tbl:initial_halos}, and shift the snapshot to the initial
+conditions from the orbital analysis (see
+Tables~\ref{tbl:scl_orbits}, \ref{tbl:umi_orbits}). We then evolve the
+full N-body NFW model forward in time in the Galactic potential and
+follow it in time until the present time, when the halo is closest to
+the observed position of the galaxy.
 
-\section{Post processing}\label{post-processing}
+\subsection{Tidal mass losses}\label{sec:shrinking_spheres}
 
-\subsection{Centring}\label{sec:shrinking_spheres}
+To accurately follow the evolution of a halo, it is important to
+determine the centre of the halo at each time chosen for analysis. We
+use a shrinking-spheres centre method inspired by \citet{power+2003}.
+First, we start with an initial centre estimate from the last timestep.
+Then, we calculate the radius of all particles from the centre, remove
+particles with a radius beyond the 0.975 quantile of the centre, and
+recalculating the centre of mass. The procedure is repeated until the
+selection radius is less than \textasciitilde1kpc or fewer than 0.1\% of
+particles remain. After a centre has been chosen, we remove all unbound
+particles based on the \gadget{} calculated potential of the halo. For
+all future timesteps, we consider only particles retained from the
+previous iteration.
 
-Shrinking spheres centres inspired by \citet{power+2003}
+The statistical centring uncertainty for the full resolution (\(10^7\)
+particle) isolation run is of order 0.003 kpc, but fluctuations are
+observed of \(\sim 0.03\,\kpc\). This is about three times the softening
+length but is less than the numerically converged radius scale.
 
-\begin{itemize}
-\tightlist
-\item
-  Recursively shrink radius by 0.975 quantile and recalculating centroid
-  until radius is less than \textasciitilde1kpc or fewer than 0.1\% of
-  particles remain.
-\item
-  Remove bound particles (using instantanious potential).
-\item
-  Use previous snapshot centre and acceleration to predict new centre
-  for next snapshot and only include particles included in previous
-  centring timestep.
-\end{itemize}
+\subsection{The stellar component}\label{sec:painting_stars}
 
-The statistical centring uncertainty for the 1e7 particle isolation run
-is of order 0.003 kpc, however oscillations in the centre are of order
-0.03 kpc. This is about three times the softening length but is less
-than the numerically converged radius scale.
+We ``paint'' stars onto dark matter particles using the particle-tagging
+method \citep[e.g.][]{bullock+johnston2005}, assuming spherical
+symmetry. We initially assume exponential stars with
+\(R_s = 0.10\,,\kpc\) for both galaxies.
 
-\subsection{Velocity profiles}\label{velocity-profiles}
-
-\begin{itemize}
-\tightlist
-\item
-  Circular velocity is computed assuming spherical symmetry and only
-  shown for every 200th particle (ranked from the centre outwards)
-\item
-  \(v_{\rm circ\ max}\) and \(r_{\rm circ,\ max}\) is found by
-  least-squares fitting the NFW functional velocity form to the points
-  of the velocity profile that have the 10\% highest velocities. This is
-  not a necessarily good fit, especially as the halo becomes stripped,
-  but accurate enough to find a reliable maximum.
-\end{itemize}
-
-\subsection{Stellar probabilities}\label{stellar-probabilities}
-
-We ``paint'' stars onto dark matter particles using the particle tagging
-method \citep[e.g.][]{bullock+johnston2005}, assuming spherical
-symmetry. Let \(\Psi\) be the potential (normalized to vanish at
-infinity) and \({\cal E}\) is the binding energy
-\({\cal E} = \Psi - 1/2 v^2\). If we know \(f({\cal E})\), the
-distribution function (phase-space density in energy), then we assign
-the stellar weight for a given particle with energy \({\cal E}\) is
-
-\begin{equation}{
+Let \(\Psi\) be the potential (normalized to vanish at infinity) and
+\({\cal E}\) the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
+know \(f({\cal E})\), the distribution function (phase-space density in
+energy), then we assign a stellar weight to a given particle with energy
+\({\cal E}\) using \begin{equation}{
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
@@
 occupation term (BT87). We use Eddington inversion to find the
-distribution function, (eq. 4-140b in BT87)
+distribution function, (eq. 4-140b in BT87) \begin{equation}{
+f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
+}\end{equation} In practice the right, boundary term is zero as
+\(\Psi \to 0\) as \(r\to\infty\), and if \(\rho \propto r^{-n}\) at
+large \(r\) and \(\Psi \sim r^{-1}\) then
+\(d\rho / d\Psi \sim r^{-n+1}\) which goes to zero provided that
+\(n > 1\). We take \(\Psi\) from the underlying assumed analytic dark
+matter potential. \(\rho_\star\) can be calculated from the surface
+density, \(\Sigma_\star\), via the inverse Abel transform.
 
-\begin{equation}{
-f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-}\end{equation}
-
-In practice the right, boundary term is zero as \(\Psi \to 0\) as
-\(r\to\infty\), and if \(\rho \propto r^{-n}\) at large \(r\) and
-\(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\) which goes to
-zero provided that \(n > 1\). We take \(\Psi\) from the underlying
-assumed analytic dark matter potential. \(\rho_\star\) can be calculated
-from the surface density, \(\Sigma_\star\), via the inverse Abel
-transform. In this work, we consider both Plummer and exponential
-profiles, as described in
-
-\subsection{Stellar density profiles and velocity
-dispersion}\label{stellar-density-profiles-and-velocity-dispersion}
-
-Stellar velocity dispersion is calculated for all stars within a 1kpc
-sphere of the centre. We assume
-\(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
-i.e.~isotropic velocity dispersion.
-
-We calculate density profiles similar to stars and assume a constant bin
-width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\) (Freedman-Diaconis
-prescription). \(R\) may be either the cylendrical radius in \(x-y\) or
-the on-sky project \(\xi, \eta\) tangent plane coordinates.
-
-Because our dwarfs are assumed to be spherical/isotropic, we retain this
-assumption when calculating predicted density profiles.
+We find the stellar profiles created in this manner are stable in the
+isolated systems and show excellent agreement with the assumed stellar
+density profile.

### 2025-08-31 thesis/results.tex
thesis julio ch1-3 edits

--- 2025-08-29 thesis/sculptor.tex
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}}
 \caption[Sculptor Orbits with LMC]{This figure is similar to
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.png}}
 \caption{Sculptor Simulation Snapshots with
@@
 \section{Ursa Minor}\label{ursa-minor}
+
+\begin{table*}[t]
+\centering
+\caption{Given the observed break radii, the required time of last pericentre and pericentre to produce the observed tidal effect.}
+\begin{tabular}{llll}
+\toprule
+parameter & Scl & UMi & \\
+\midrule
+$r_{\rm break}$ observed & $25 \pm 5$ arcmin & $30 \pm 5$ arcmin & (sestito+2024a?); (sestito+2024b?)\\
+$t_{\rm last\ peri}$ required & $110\pm30$ Myr & $120\pm30$ Myr & Eq. \ref{eq:r_break}\\
+$r_{\rm peri}$ required & 16kpc & 14kpc & Eq. \ref{eq:r_jacobi}\\
+\bottomrule
+\end{tabular}
+\end{table*}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_lmc_xyzr_orbits.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_lmc_xyzr_orbits.png}}
 \caption[Ursa Minor orbits with LMC]{Orbits of Ursa Minor with (orange)

### 2025-08-31 observations/munoz1/relative_location.jl

@@ -1,0 +1,15 @@
+using LilGuys
+using TOML
+
+
+props_munoz1 = TOML.parsefile("observed_properties.toml")
+props_umi = TOML.parsefile("../ursa_minor/observed_properties.toml")
+
+xi, eta = 60 .* LilGuys.to_tangent(props_munoz1["ra"], props_munoz1["dec"], props_umi["ra"], props_umi["dec"])
+
+@info "units = arcmin"
+@info "xi, eta = $xi, $eta"
+
+R_ell = LilGuys.calc_R_ell(xi, eta, props_umi["ellipticity"], props_umi["position_angle"])
+
+@info "R_ell = $R_ell"

### 2025-08-31 thesis/compile.sh

@@ -7,10 +7,8 @@ pandoc $pandocoptions -o abstract.tex abstract.md
 pandoc $pandocoptions -o introduction.tex introduction.md
 pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
-pandoc $pandocoptions -o sculptor.tex sculptor.md
+pandoc $pandocoptions -o results.tex results.md
 pandoc $pandocoptions -o discussion.tex discussion.md
-
-pandoc $pandocoptions -o ursa_minor.tex ursa_minor.md
 pandoc $pandocoptions -o conclusion.tex conclusion.md
 
 pandoc $pandocoptions -o appendix.tex appendix.md --top-level-division=chapter

### 2025-09-02 thesis/abstract.tex
julio sep 1 edits to §1, 2

@@
 The satellite galaxies of the Milky Way (MW), with the exception of the
-Magellanic clouds, are all dwarf spheroidals (dSphs)---gas-free,
-non-rotating stellar systems whos gravity is dominated by dark matter.
+Magellanic Clouds, are all dwarf spheroidals (dSphs)---gas-free,
+non-rotating stellar systems whose gravity is dominated by dark matter.
 The light profiles of dSphs are usually well-approximated by an
@@
 Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
-stars far from the main body, out to \(\sim10\) effective radii. The
-origin of these extended profiles is not clear, and may reflect either
-the influence of Galactic tides or instead an innate feature. In this
-thesis, I review the excestence of an extended light profiles in Scl and
+stars far from the main body, out to \(\sim10\) effective radii,
+indicative of outer deviations from an exponential law. The origin of
+these extended profiles is not clear, and may reflect either the
+influence of Galactic tides or, instead, an innate feature. In this
+thesis, we review the existence of extended light profiles in Scl and
 UMi, validating the Bayesian membership catalogue of
 \citet{jensen+2024}. To evaluate if tides indeed can explain these
-density features, I conduct idealized N-body simulation of both galaxies
-in the tidal field of the Milky Way. I find that neigher dwarf has been
-subjected to tidal forces strong enough to create the observed extended
-light profiles from initially exponential profiles. One complication is
-that Sculptor's orbit is strongly influenced by the presence of the
-Large Magellanic Cloud (LMC). Because the LMC's mass is poorly
-constrained, Sculptor's long term orbital history is highly uncertain.
-Despite this uncertainty, our models still suggest that the combined
-tides of the LMC and MW are still unable to explain Scl's outer profile
-over the past 5 Gyr. I conclude that the extended light profiles of Scl
-and UMi are innate, likely reflecting their formation histories.
-Mergers, accretion events, or complex, multi-component star formation
-episodes are possible explanations for such an extended outer profile.
+density features, we conduct idealized N-body simulation of both
+galaxies in the tidal field of the Milky Way. We find that neither dwarf
+has been subjected to tidal forces strong enough to create the observed
+extended light profiles from initially exponential profiles. One
+complication is that Sculptor's orbit is strongly influenced by the
+presence of the Large Magellanic Cloud (LMC). Because the LMC's mass is
+poorly constrained, Sculptor's long term orbital history is highly
+uncertain. Despite this uncertainty, our models still suggest that the
+combined tides of the LMC and MW are unable to explain Scl's outer
+profile. We conclude that the extended light profiles of Scl and UMi are
+innate, likely reflecting their assembly histories. Mergers, accretion
+events, or complex, multi-component star formation episodes are possible
+explanations for the observations ed extended light profiles.

### 2025-09-02 thesis/data.tex
julio sep 1 edits to §1, 2

@@
 prototypical example is the Fornax dSph --- well-fit by an exponential
-over 4.5 decades in surface density. On the other hand, dSph like
-Sculptor or Ursa Minor have profiles with deviate significantly from an
+over 4.5 decades in surface density. On the other hand, dSphs like
+Sculptor or Ursa Minor have profiles which deviate significantly from an
 exponential profile fitted to the inner regions, with a clear excess of
 stars/light in the outer regions. In this chapter, we critically review
-the density profile of MW classical dwarfs. This thesis concerns the
+the density profile of the MW classical dwarfs. This thesis concerns the
 origin of this excess, with a tidal interpretation explored in
@@
 Measuring the light profile of a resolved galaxy requires careful
-consideration of whether a given star belongs to the system. Without
-removing contamination from foreground/background features, faint
+consideration of whether any given star belongs to the system or not.
+Without removing contamination from foreground/background sources, faint
 features may be lost in the noise or be of uncertain association. When
-only photometric data was available, the galaxy membership of stars were
+only photometric data was available, the membership of stars was
 ascertained using the colour-magnitude diagram alone \citep[e.g.,
-matched filter methods like][]{rockosi+2002}. Now that \emph{Gaia} data
-are available, stellar parallax and proper motion are also available to
-improve membership assignment.
+matched filter methods like those used by][]{rockosi+2002}. Now that
+\emph{Gaia} data are available, stellar parallax and proper motion are
+also available to improve membership assignment.
 
 Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
-probabilities for \emph{Gaia} observations. J+24 used a Bayesian
-framework incorporating proper motion (PM), colour-magnitude diagram
-(CMD), and spatial information to determine the probability that a given
-star belongs to the satellite or foreground/background. By accounting
-for PM in particular, J+24 produced low contamination samples of
-candidate member stars out to large distances from a dwarf galaxy. J+24
-extended the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
-additionally includes a secondary, extended spatial component. J+24
+probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
+incorporating proper motion (PM), colour-magnitude diagram (CMD), and
+spatial information to determine the probability that a given star
+belongs to the satellite or foreground/background. By accounting for PM
+in particular, J+24 produced low contamination samples of candidate
+member stars out to large distances from a dwarf galaxy. J+24 extended
+the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
+additionally including a secondary, extended spatial component. J+24
 detected candidate members out to \textasciitilde10 half-light radii
@@
 
-J+24 construct likelihoods, \({\cal L}\), representing the probability
-density that a star is consistent with either the foreground/background,
-\({\cal L}_{\rm bg}\), or the satellite galaxy, \({\cal L}_{\rm sat}\).
-In either case, the likelihoods are the product of a spatial, PM, and
-CMD component, \begin{equation}{
+J+24's method uses likelihoods, \({\cal L}\), representing the
+probability density that a star is consistent with either the
+foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
+\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
+of a spatial, PM, and CMD component, \begin{equation}{
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
@@
 
-For the satellite, the \emph{spatial likelihood} is specified either as
-a one- or two-component elliptical exponential profile. A second
-component is included only when the preferred amplitude of the second
-component is non-zero. The \emph{proper-motion likelihood} quantifies
-the agreement of a star's motion with the dwarf galaxy's systemic
-motion, accounting for observational uncertainties. The \emph{CMD
-likelihood} measures the consistency of a star's \emph{Gaia} \(G\)
-magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
-isochrones for the galaxy. For the background likelihoods, the spatial
-likelihood is a uniform distribution over the field, and the background
-CMD and PM likelihoods are constructed empirically from the stars in an
-annulus far from the satellite. Each likelihood is normalized as a
-probability density over the respective parameter space.
+For a satellite, the \emph{spatial likelihood} is specified either as a
+one- or two-component elliptical exponential profile. A second component
+is included only when the preferred amplitude of the second component is
+non-zero. The \emph{proper-motion likelihood} quantifies the agreement
+of a star's motion with the dwarf galaxy's systemic motion, accounting
+for observational uncertainties. The \emph{CMD likelihood} measures the
+consistency of a star's \emph{Gaia} \(G\) magnitude and
+\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
+galaxy. For the background likelihoods, the spatial likelihood is a
+uniform distribution over the field, and the background CMD and PM
+likelihoods are constructed empirically from the stars in an annulus far
+from the satellite. Each likelihood is normalized as a probability
+density over the respective parameter space.
 
@@
 For our fiducial sample, we adopt a minimum probability of
-\(P_{\rm sat} = 0.2\). We use the \(P_{\rm sat}\) values from the
-elliptical 2-component runs if a galaxy showed evidence for an outer
-component, the 1-component run otherwise. Most stars have
-\(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact choice of
-probability threshold has little effect on the resulting sample. Even at
-our relatively generous probability threshold of 0.2, the purity remains
-high when validated against spectroscopic line-of-sight (LOS) velocities
-(\textasciitilde90\%, J+24). We note, however, that these purity
-estimates may be biased. Stars with LOS velocities are typically
-brighter and, as a result, have more precise \emph{Gaia} measurements.
-However, we find that our conclusions are unchanged when limiting
-samples to only the brightest stars. Altogether, the J+24 method
-provides a high-quality, low-contamination sample of dwarf galaxy
-candidate member stars, which we will now investigate.
+\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
+J+24's quality cuts typically only include stars with \(G < 21\). We use
+the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
+galaxy shows evidence for an outer component, the 1-component run
+otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
+1, so the exact choice of probability threshold has little effect on the
+resulting sample. Even at our relatively generous probability threshold
+of 0.2, the purity remains high when validated against spectroscopic
+line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
+however, that these purity estimates may be biased. Stars with LOS
+velocities are typically brighter and, as a result, have more precise
+\emph{Gaia} measurements. However, we find that our conclusions are
+unchanged when limiting samples to only the brightest stars. Altogether,
+the J+24 method provides a high-quality, low-contamination sample of
+dwarf galaxy candidate member stars, which we will now investigate.
 
@@
 \(\xi\) and \(\eta\) are offsets in RA and declination as measured on
-the plane tangent to the galaxy centre. To account for the elliptical
-shape of the galaxy, we use \(R_{\rm ell}\), which we define as the
-circularized elliptical radius, \begin{equation}{
+the plane tangent relative to the galaxy centre. To account for the
+elliptical shape of the galaxy, we use \(R_{\rm ell}\), which we define
+as the circularized elliptical radius, \begin{equation}{
 R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
@@
 a progressive tightening of criteria. First, we consider a
-minimally-refined sample, \textbf{all}, only excluding stars with poor
-astrometry, unreliable photometry, or inconsistent parallaxes. Next,
-incorporating CMD and PM information, we use the \textbf{CMD+PM} to test
-a selection method agnostic to the spatial position. This sample
+minimally-refined sample, \textbf{all}, which only excludes stars with
+poor astrometry, unreliable photometry, or inconsistent parallaxes.
+Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
+to test a selection method agnostic to the spatial position. This sample
 includes stars where the CMD and PM combined likelihood favours
@@
 Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
-shows our fiducial sample for each galaxy, and the impact of alternate
-samples, in tangent-plane coordinates, the CMD, and proper-motion space.
-The ``all'' sample extends uniformly across the tangent plane, but
-includes a substantial background population. The ``CMD+PM'' sample has
-a much lower background density in the tangent plane, revealing the
-satellite more clearly. The fiducial sample appears similar to the
-``CMD+PM'' sample in the CMD and PM planes, but excludes improbably
-distant stars. The ``RV member'' sample also traces out similar
-distributions in CMD and PM space as the fiducial sample, with less
-dispersion, likely reflecting the brighter magnitudes of stars with
-spectroscopic follow-up. Each selection criteria---spatial, CMD, and
-PM---contributes towards a high-quality membership criterion.
+show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
+Fornax), as well as impact of different samples, in tangent-plane
+coordinates, the CMD, and proper-motion space. The ``all'' sample
+extends uniformly across the tangent plane, but includes a substantial
+background population. The ``CMD+PM'' sample has a much lower background
+density in the tangent plane, revealing the satellite more clearly. The
+fiducial sample appears similar to the ``CMD+PM'' sample in the CMD and
+PM planes, but excludes improbably distant stars. The ``RV member''
+sample also traces out similar distributions in CMD and PM space as the
+fiducial sample, with less dispersion, likely reflecting the brighter
+magnitudes of stars with spectroscopic follow-up. Each selection
+criteria---spatial, CMD, and PM---contributes towards a high-quality
+membership assignment.
 
@@
 ``CMD+PM'' selected sample, which does not include a spatial likelihood.
-If even fainter, more extended features exist, they are apparently not
+If even fainter, more extended features exist, they are not clearly
 detectable with \emph{Gaia}.
@@
 of each dwarf. For an exponential profile, 99.95\% of stars fall within
-a radius of \(6R_h\), so the presence of very distant member stars
+a radius of \(6R_h\), so the mere presence of very distant member stars
 provides evidence for deviations from an exponential profile.
@@
 three different samples from above: ``all'', ``CMD+PM'', and the
-``fiducial'' sample. In each case, all samples are the same towards the
+``fiducial'' sample. In each case, all samples coincide towards the
 inner regions of the satellite. However, the density profile from
@@
 
-While the density profiles of all samples agrees in the inner regions,
+While the density profiles of all samples agree in the inner regions,
 they begin to deviate outside \(6 R_h\). Outside this radius, the
@@
 
-Near to UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
+Near UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
 ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
@@
 high background, and Sagittarius, which was not included in J+24. The
-density profiles are scaled to match at the half-light radius from
-\citet{munoz+2018}. All of the classical dwarfs appear to be well
+density profiles are scaled to match at the half-light radius, taken
+from \citet{munoz+2018}. All of the classical dwarfs appear to be well
 described by an exponential profile in the inner regions. In the outer

### 2025-09-02 thesis/introduction.tex
julio sep 1 edits to §1, 2

@@
 1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.
+
 With the exception of the Magellanic Clouds, most dwarf galaxy
@@
 recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
-satellites contain stellar populations which are \emph{relics} from the
-early universe, consisting of many of the oldest and most metal poor
-stars. Understanding the properties of dwarf galaxies thus has
+satellites contain old stellar populations which are \emph{relics} from
+the early universe, consisting of many of the oldest and most metal poor
+stars known. Understanding the properties of dwarf galaxies thus has
 implications across astronomy, from cosmological structure formation on
@@
 
-In this section, we first describe the general observed properties of
-local dwarf galaxies. Next, we summarize the cosmological origin and
-role of dwarf galaxies. We later review recent advancements and problems
-concerning dwarf galaxies. Next, we contextualize the extended stellar
-density profiles of Sculptor and Ursa Minor. We then discuss how tides
-influence satellite galaxies in general. Finally, we provide a roadmap
-to the thesis.
+In this Chapter, we first describe the general observed properties of
+local dwarf galaxies. Next, we summarize our understanding of the
+cosmological origin of dwarf galaxies. We later review recent
+advancements and pending questions concerning dwarf galaxies, and
+introduce the puzzle posed by the extended stellar density profiles of
+Sculptor and Ursa Minor. We end with a brief roadmap to the remainder of
+this dissertation.
 
-\section{Observational history of dwarf
-galaxies}\label{observational-history-of-dwarf-galaxies}
+\section{Observations of dwarf
+galaxies}\label{observations-of-dwarf-galaxies}
 
@@
 be galaxies based on the inferred luminosities and sizes, their nature
-and formation remained unclear for decades
+remained unclear for decades
 \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
@@
 substantial amounts of dark matter. From early determinations of the
-velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
-(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
-inferred mass-to-light ratios were at least 10 times larger than the
-values of globular clusters (GCs). While rather uncertain initially,
-these values have been corroborated with larger and more precise samples
-\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
-proposed to explain these unusually high mass-to-light ratios. Examples
-include: ongoing tidal disruption inflating inferred velocity
-dispersions \citep[e.g.,][]{kuhn+miller1989}, the presence of massive
-central black holes \citep[e.g.,][]{strobel+lake1994}, or modified
-theories of gravity \citep{milgrom1995}. Over time, however, a consensus
-developed where the high mass-to-light ratios of dwarf galaxies was due
-to the presence of a dark matter halo, which was lacking in GCs
-\citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}. Since then, the
-properties of dwarf galaxies have played an increasingly important role
-in our understanding of the clustering of dark matter on small scales
-\citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.
+velocity dispersion for the Sculptor (Scl) and Ursa Minor (UMi) dwarf
+spheroidal (dSph) galaxies, inferred mass-to-light ratios were at least
+10 times larger than the values of globular clusters {[}GCs; e.g.,
+\citet{aaronson1983}; \citet{aaronson+olszewski1987}{]}. While rather
+uncertain initially, these values were later corroborated with larger
+and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the time,
+several theories were proposed to explain these unusually high
+mass-to-light ratios. Examples include: ongoing tidal disruption
+inflating inferred velocity dispersions
+\citep[e.g.,][]{kuhn+miller1989}, the presence of massive central black
+holes \citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
+\citep{milgrom1995}. Over time, however, a consensus developed where the
+high mass-to-light ratios of dwarf galaxies was due to the presence of a
+dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
+Since then, the properties of dwarf galaxies have played an increasingly
+important role in our understanding of the clustering of dark matter on
+small scales \citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.
 
@@
 stellar system with dark matter \citep[or, more specifically, not
-consistent with newtonian dynamics of visible matter
+consistent with Newtonian dynamics of visible matter
 alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
 have no clear evidence for dark matter. The boundary between these two
-classes becomes uncertain for faint, compact systems, known as
-``ambiguous'' systems.
-
-The Local Group (LG) is defined as the group consisting of galaxies
-within \(\sim 1\) Mpc from the Milky Way-Andromeda centre \citep[e.g.,][
-and references therein]{mcconnachie2012}. Today, we know that the Milky
-Way system is teeming with satellites.
-Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
-including dwarf galaxies, globular clusters, and ambiguous systems. Such
-a nearby population of dwarf galaxies are useful for resolved, detailed
-studies aiming to understand the nature of these systems.
+classes becomes uncertain for faint, compact stellar associations, known
+as ``ambiguous'' systems.
 
@@
 on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
-\textbf{bright dwarfs} with magnitudes
-\(-18 \lesssim M_V \lesssim  -14\) or stellar masses
-\(10^9\,\Mo \gtrsim M_\star \gtrsim 3\times10^7\,\Mo\) \footnote{assuming
+\textbf{bright dwarfs} with magnitudes \(-14 \gtrsim M_V \gtrsim  -18\)
+or stellar masses
+\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
   stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
@@
 with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
-intermediate luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\) or
-\(3\times10^7\,\Mo \gtrsim M_\star \gtrsim 10^5\,\Mo\)). Typically,
+intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
+\(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
 these systems are old, non-star forming, gas-poor, and spheroidal. All
@@
   omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
-occupy the very faintest magnitude range (\(-7.7 \lesssim M_V\), or
-\(M_\star < 10^5\,\Mo\)). These galaxies have minuscule stellar masses,
-typically compact sizes, and very metal-poor stellar populations
+occupy the very faintest magnitude range (\(M_V \gtrsim -7.7\) or
+\(M_\star \lesssim 10^5\,\Mo\)). These galaxies have minuscule stellar
+masses, typically compact sizes, and very metal-poor stellar populations
 \citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
@@
 
-Here, we focus on classical dwarf galaxies, which are ideal for detailed
-dynamical study. These galaxies are dark matter dominated but also have
-large numbers of stars with radial velocities and proper motions, and
-have generally well-constrained properties.
+Of local dwarfs, the classical systems still remain among the most
+studied galaxies with the best constrained parameters. These dwarfs also
+contain large numbers of bright (giant) stars which can be followed up
+spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
+the determination of fundamental properties such as the position, size,
+orientation, distances, proper motions, line-of-sight (LOS) velocities,
+and velocity dispersions \(\sigma_\V\), are all relatively well
+constrained today. \textbf{revisit this\ldots JFN cuts a similar
+version}
+
+Most well-studied dwarf galaxies lie near the Milky Way or in its
+vicinity, the Local Group (LG) of galaxies. TheLG is defined as the
+group consisting of galaxies within \(\sim 1\) Mpc from the Milky
+Way-Andromeda centre \citep[e.g.,][ and references
+therein]{mcconnachie2012}. Today, we know that the Milky Way system is
+teeming with dwarfs, many of which are satellites of either the Milky
+Way or(MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
+the MW satellite system, including dwarf galaxies, globular clusters,
+and ambiguous systems. Such a nearby population of dwarf galaxies are
+useful for resolved, detailed studies aiming to understand the nature of
+these systems.
 
@@
 
-\section{Cosmological context}\label{cosmological-context}
+\section{Dwarf galaxies in a cosmological
+context}\label{dwarf-galaxies-in-a-cosmological-context}
 
 We only understand a fraction of the universe's composition. The leading
-theory of cosmology, \LCDM{} (Lambda Cold Dark Matter), posits that the
+theory of cosmology, Lambda Cold Dark Matter (\LCDM{}), posits that the
 universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
@@
 In \LCDM{}, dark matter is assumed to interact only gravitationally.
-Light passes through dark matter unimpeded---in this sense dark matter
+Light passes through dark matter unimpeded---in this sense, dark matter
 is transparent. Dark matter is also commonly assumed to be \emph{cold},
@@
 
-\subsection{Typical structure of dark matter
-halos}\label{typical-structure-of-dark-matter-halos}
+\subsection{The structure of cold dark matter
+halos}\label{the-structure-of-cold-dark-matter-halos}
 
@@
 gradually with radius, and at large radii the density falls off like
-\(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density} shows an example NFW
-halo.
+\(\rho \sim 1/r^3\).
 
 The total mass of an NFW profile formally diverges, so halo masses are
-commonly defined using an overdensity criterion. The virial mass,
+conventionally defined using an overdensity criterion. The virial mass,
 \(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
@@
 \begin{equation}{
-M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G.
-}\end{equation} A standard second parameter of NFW halos is the halo
+M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G,
+}\end{equation} and \(H(z)\) is the Hubble constant as a function of
+redshift. Another way of characterizing NFW halos is through the halo
 concentration, \(c=r_{200} / r_s\), which describes how the
@@
 
+The blue solid curve in Fig.~\ref{fig:nfw_density} shows an example NFW
+halo for a galaxy with \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
+\(M_{200} = 1\times10^10\,\Mo\) and \(c=12.5\), the cosmological mean
+for a Fornax stellar mass galaxy.
+
 The two parameters of an NFW profile are not independent. Lower-mass
@@
 the mass-concentration relation, describes the mean trend of
-concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
-left panel of Fig.~\ref{fig:smhm} illustrates the present-day
-mass-concentration from \citet{ludlow+2016}. While concentration tends
-to decrease with increasing mass, the relation has substantial scatter.
-Other parameters such as the halo spin or shape may affect the scatter
-of the mass-concentration relation, but their effect is typically
-expected to be small
-\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
+concentration with mass or, equivalently, the dependence of \(\vmax\) on
+\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
+Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration from
+\citet{ludlow+2016}. While concentration tends to decrease with
+increasing mass, the relation has substantial scatter. Other parameters
+such as the halo spin or shape may affect the scatter of the
+mass-concentration relation, but their effect is typically expected to
+be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
 
@@
 \centering
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
 \caption[Example density profiles]{Density profiles in log density
-versus log radius for stars and dark matter in a Sculptor-like galaxy.
-The dark matter is more extended and massive than the star across the
-entire galaxy.}\label{fig:nfw_density}
+versus log radius for stars and dark matter in a Fornax-like galaxy. The
+dark matter is more extended and massive than the star across the entire
+galaxy.}\label{fig:nfw_density}
 \end{figure}
@@
 relation from \citet{fattahi+2018} with scatter points simulated central
-galaxies from \apostle{} in \citet{fattahi+2018}.}\label{fig:smhm}
+galaxies from \apostle{} in \citet{fattahi+2018}. The pink star
+illustrates the location of the Fornax galaxy, whose density profiles
+are shown in Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
 \end{figure}
 
-\subsection{\texorpdfstring{Formation of galaxies within \LCDM{}
-halos}{Formation of galaxies within  halos}}\label{formation-of-galaxies-within-halos}
+\subsection{\texorpdfstring{Galaxies formation in
+\LCDM{}}{Galaxies formation in }}\label{galaxies-formation-in}
 
@@
 
-In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxies
+In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxy
 comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
-shows an example exponential stellar component for a classical dwarf
-galaxy with the expected underlying dark matter halo. Even when the
-stars are most dense in the inner regions, the dark matter remains
-nearly an order of magnitude higher in density. Stars have a small
-contribution to the gravitational structure of dwarf galaxies--instead
-stars are reasonably approximated as tracer particles of the underlying
-dark matter halo. In addition, the stellar component is much more
-compact than the dark matter halo.
+shows an example exponential stellar component for the Fornax dwarf
+galaxy with the expected underlying dark matter halo. This galaxy has a
+stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
+radius \(0.65\,\kpc\) \citep[ with \citet{woo+courteau+dekel2008}
+mass-to-light ratio]{munoz+2018}, and a dark matter halo with
+\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\). Even when the stars are
+densest (the inner regions), the dark matter remains nearly an order of
+magnitude higher in density. Stars make a small contribution to the
+gravitational structure of dwarf galaxies--indeed, stars are reasonably
+approximated as tracer particles of the underlying dark matter halo. In
+addition, the stellar component is confined to the central regions of
+the dark matter halo.
 
-Several challenges complicate the SMHM trend, including environment,
-assembly history, tidal effects, and the details of galaxy formation. In
+Several factors affect the SMHM trend, including environment, assembly
+history, tidal effects, and the details of galaxy formation. In
 particular, effects like ram-pressure stripping (removal of gas in the
-dwarf galaxy due to pressure from host's circumgalactic medium) and
+dwarf galaxy due to pressure from the host's circumgalactic medium) and
 tidal removal of gas cause star formation to quench
@@
 
-\section{Observing faint features of dwarf
-galaxies}\label{observing-faint-features-of-dwarf-galaxies}
+\subsection{Open questions and ongoing challenges in dwarf
+galaxies}\label{open-questions-and-ongoing-challenges-in-dwarf-galaxies}
 
-\subsection{\texorpdfstring{The era of
-\emph{Gaia}}{The era of Gaia}}\label{the-era-of-gaia}
-
-Since Local Group dwarfs are nearby, they are resolved into individual
-stars, and we can study these galaxies on a star-by-star basis. As a
-result, it is possible to measure the 3D velocity and position of a star
-if we can measure the its position, distance, line-of-sight (LOS)
-velocity, and proper motions. Unfortunately, determining distances and
-full 3D velocities is challenging. The most direct measurement of
-distance, parallax, requires precise tracking of a star's sky position
-across a year. And while line-of-sight (LOS) velocities are relatively
-easily determined from spectroscopy, tangential velocities, derived from
-proper motions and distances, are much more challenging. Typically,
-measuring proper motions requires accurate (much less than arcsecond)
-determinations of small changes in a star's position over baselines of
-years to decades. The full 6D position and velocity information for
-stars has, until recently, been known for only a handful of stars.
-
-Launched in 2013, \emph{Gaia} is a space-based, all-sky survey telescope
-situated at the Sun-Earth L2 Lagrange point
-\citep{gaiacollaboration+2016}. \emph{Gaia} has redefined astrometry,
-providing photometry, positions, proper motions, and parallaxes for over
-1 billion stars \citep{gaiacollaboration+2021}. While \emph{Gaia}
-completed its space-based mission in 2025, two forthcoming data releases
-are expected.
-
-Determining absolute parallax measurement is facilitated by the
-observation that stars in different regions of the sky are affected by
-parallax motion with different phases. By imagining two regions
-separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
-tiny changes in relative positions of stars across small and large
-angles. Combining measurements from multiple epochs across several
-years, an absolute all-sky reference frame is derived from which
-parallax and proper motions are derived. In addition to astrometry,
-\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
-and colours from the blue photometer (BP, 330-680 nm) and red photometer
-(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
-BP-RP spectra and radial velocity measurements of bright stars \citep[of
-magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
-work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
-\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
-\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
-  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
-  change in right ascension, \(\alpha\), and declination, \(\delta\),
-  typically in units of mili-arcsecond (mas) per year.
-  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
-  effects in \(\alpha\).}
-
-\subsection{The Local Group of
-Galaxies}\label{the-local-group-of-galaxies}
-
-\emph{Gaia} has revolutionized LG studies. For example, the 6D dynamical
-measurements and metallicities of MW stars led to the (re)discovery of
-past mergers or Milky Way building blocks like \emph{Gaia}-Sausage
-Enceladus
-\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
-see also][]{meza+2005}}, out-of-equilibrium structures like the
-\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
-the Milky Way's spiral arms and the bar in the solar neighbourhood
-\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
-Way halo, \emph{Gaia} has helped find and constrain numerous stellar
-streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
-Altogether, \emph{Gaia} has revealed the hierarchical formation and
-complex, evolving structure of our own Galaxy.
-
-For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
-facilitated stellar membership determination. Before \emph{Gaia}, few
-galaxies had precisely measured proper motions, more recently from
-Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
-\emph{Gaia} allowed the first systematic and precise determinations of
-Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
-the proper motion uncertainty of a typical dwarf member star is often
-large, by combining the proper motions of 100s or 1000s of stars from
-\emph{Gaia}, precise average proper motion measurements can be
-determined, sometimes only limited by \emph{Gaia}'s systematic error
-floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
-new dynamical era for MW satellite studies, where we can derive precise
-orbits for any satellite, assuming a given MW potential. In addition,
-\emph{Gaia} helps establish membership by separate out contaminating MW
-foreground stars. By measuring parallaxes and/or proper motions, many
-background and foreground stars can be classified as non-members
-\citep[e.g.,][]{battaglia+2022, jensen+2024}.
-
-\subsection{Dwarf galaxies today}\label{dwarf-galaxies-today}
-
-Advances in observational techniques has accelerated progress in our
+Advances in observational techniques have accelerated progress in our
 understanding of dwarf galaxies and introduced new questions. Deep
@@
 detailed constraints on the dark matter density profiles, and hints of
-tidal disruption or stellar halos.
-
-Of local dwarfs, the classical systems still remain the most studied
-galaxies with the best constrained parameters. These dwarfs also contain
-large numbers of bright (giant) stars which can be followed up
-spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
-the determination of fundamental properties such as the position, size,
-orientation, distances, proper motions, line-of-sight (LOS) velocities,
-and velocity dispersions \(\sigma_\V\), are all relatively well
-constrained today.
+tidal disruption or stellar halos. \textbf{does this ¶ (or section)
+belong here?}
 
@@
 
+\section{The structure of nearby dwarf
+galaxies}\label{the-structure-of-nearby-dwarf-galaxies}
+
+\subsection{\texorpdfstring{The \emph{Gaia} space
+telescope}{The Gaia space telescope}}\label{the-gaia-space-telescope}
+
+Since Local Group dwarfs are nearby, they are resolved into individual
+stars, and therefore we can study these galaxies on a star-by-star
+basis. As a result, it is possible to measure the 3D velocity and
+position of a star if we can measure its position, distance,
+line-of-sight (LOS) velocity, and proper motion. Unfortunately,
+determining distances and full 3D velocities is challenging. The most
+direct measurement of distance, parallax, requires precise tracking of a
+star's sky position across a year. And while line-of-sight (LOS)
+velocities are relatively easily determined from spectroscopy,
+tangential velocities, derived from proper motions and distances, are
+much more challenging. Typically, measuring proper motions requires
+accurate (much less than arcsecond) determinations of small changes in a
+star's position over baselines of years to decades. The full 6D position
+and velocity information for stars has, until recently, been known for
+only a handful of stars.
+
+Launched in 2013, \emph{Gaia} is a space-based, all-sky survey telescope
+situated at the Sun-Earth L2 Lagrange point
+\citep{gaiacollaboration+2016}. \emph{Gaia} has redefined astrometry,
+providing photometry, positions, proper motions, and parallaxes for over
+1 billion stars \citep{gaiacollaboration+2021}. While \emph{Gaia}
+completed its space-based mission in 2025, two further data releases are
+expected.
+
+Determining absolute parallax measurement is facilitated by the
+observation that stars in different regions of the sky are affected by
+parallax motion with different phases. By imaging two regions separated
+by 106.5 degrees on the same focal plane, \emph{Gaia} measures tiny
+changes in relative positions of stars across small and large angles.
+Combining measurements from multiple epochs across several years, an
+absolute all-sky reference frame is derived from which parallax and
+proper motions are derived. In addition to astrometry, \emph{Gaia}
+measures photometry in the wide \emph{G} band (330-1050nm) and colours
+from the blue photometer (BP, 330-680 nm) and red photometer (RP,
+640-1050 nm). \emph{Gaia} additionally provides low resolution BP-RP
+spectra and radial velocity measurements of bright stars \citep[of
+magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
+work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
+\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
+\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
+  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
+  change in right ascension, \(\alpha\), and declination, \(\delta\),
+  typically in units of mili-arcsecond (mas) per year.
+  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
+  effects in \(\alpha\).}
+
+\subsection{\texorpdfstring{\emph{Gaia}'s impact on Milky Way
+studies}{Gaia's impact on Milky Way studies}}\label{gaias-impact-on-milky-way-studies}
+
+\emph{Gaia} has revolutionized our understanding of Milky Way structure.
+For example, the 6D dynamical measurements and metallicities of MW stars
+led to the (re)discovery of past mergers or Milky Way building blocks
+like \emph{Gaia}-Sausage Enceladus
+\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
+see also][]{meza+2005}}, out-of-equilibrium structures like the
+\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
+the Milky Way's spiral arms and the bar in the solar neighbourhood
+\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
+Way halo, \emph{Gaia} has helped find and constrain numerous stellar
+streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
+Altogether, \emph{Gaia} has revealed the hierarchical formation and
+complex, evolving structure of our own Galaxy.
+
+For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
+facilitated stellar membership determination. Before \emph{Gaia}, few
+galaxies had precisely measured proper motions, more recently from
+Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
+\emph{Gaia} allowed the first systematic and precise determinations of
+Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
+the proper motion uncertainty of a typical dwarf member star is often
+large, by combining the proper motions of 100s or 1000s of stars from
+\emph{Gaia}, precise average proper motion measurements can be
+determined, sometimes only limited by \emph{Gaia}'s systematic error
+floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
+new era for MW satellite dynamical studies, where we can derive precise
+orbits for any satellite, assuming a given MW potential. In addition,
+\emph{Gaia} helps establish membership by separating out contaminating
+MW foreground stars. By measuring parallaxes and/or proper motions, many
+background and foreground stars can be classified as non-members
+\citep[e.g.,][]{battaglia+2022, jensen+2024}.
+
 \subsection{Dwarf galaxy light profiles}\label{sec:exponential_profiles}
@@
 \citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. We note that
-for resolved galaxies, these profiles are in terms of stellar count
-density instead of surface brightness.
+for resolved galaxies, these profiles are in terms of stellar counts
+instead of surface brightness.
 
@@
 \Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
-}\end{equation} In much of the older literature, \(R_t\) was often
-interpreted as a ``tidal radius'', after an analogous interpretation for
-globular clusters \citep[e.g.,][]{IH1995, hodge1961}.
+}\end{equation} In much of the older literature, \(R_t\) was interpreted
+as a ``tidal radius'', after an analogous interpretation for globular
+clusters \citep[e.g.,][]{hodge1961, IH1995}.
 
@@
 dwarf galaxies, \citet{munoz+2018} advocate for the Sérsic profile since
-the added flexibility allows for more profiles to be fit with a single
-law.
+the added flexibility allows more profiles to be fit with a single law.
 
 While there are no clear theoretical preferences for any of these
-profiles, exponential density profiles are commonly used for dwarf
+profiles, exponential density profiles have been commonly used for dwarf
 spheroidal galaxies. \citet{faber+lin1983} were among the first to
-demonstrate that an exponential law is a reasonable empirical fit to
-dwarf galaxies, theorizing that dwarf spheroidals may have evolved from
-exponential disk galaxies and maintained a similar light profile. Later,
+demonstrate that an exponential law is a reasonable empirical fit,
+theorizing that dwarf spheroidals may have evolved from exponential disk
+galaxies and maintained a similar light profile. Later,
 \citet{read+gilmore2005} showed that exponential profiles may originate
 from mass loss during the evolution of dwarf galaxies. Many subsequent
-photometric works for dwarf spheroidal galaxies have used exponential
+photometric studies of dwarf spheroidal galaxies have used exponential
 fits, finding that exponential and King profiles both provide good
@@
 extragalactic dwarf elliptical, blue compact, and irregular dwarf
-galaxies are better described with an exponential profile including a
-central cusp or nuclear region \citep{caldwell+bothun1987, noeske+2003}.
-Instead, some studies find an inner density decrement compared to
-exponentials \citep[e.g.,][]{caldwell+1992, makarov+2012}.
+galaxies are better described with an exponential profile to which is
+added a central cusp or nuclear region
+\citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
+studies find an inner density decrement compared to exponentials
+\citep[e.g.,][]{caldwell+1992, makarov+2012}.
 
-However, exponential profiles do not fit every dwarf galaxy. Even
-starting from \citet{aparicio+1997}, and \citet[for Coma cluster dwarf
+However, exponential profiles do not fit every dwarf galaxy. For example
+\citet{aparicio+1997}, and \citet[for Coma cluster dwarf
 ellipticals]{graham+guzman2003}, some people have noted deviations from
@@
 \citet{herrmann+hunter+elmegreen2013};
-\citet{herrmann+hunter+elmegreen2016}; and \citet{lee+2018} note that at
-least in a photometric sample of more distant disky, irregular, and blue
-compact dwarfs, many or most dwarfs are best fit with two nested
+\citet{herrmann+hunter+elmegreen2016}; and \citet{lee+2018} noted that
+at least in a photometric sample of more distant disky, irregular, and
+blue compact dwarfs, many or most dwarfs are best fit with two nested
 exponential profiles. It is unclear how these conclusions apply to dwarf
@@
 dwarf galaxies, an exponential is an excellent first-order
-approximation. Typically, deviations from exponentials are in a
+approximation. Typically, deviations from exponentials are in the
 direction of a steeper outer cutoff or changes to the inner slope of a
 dwarf galaxy (due, for example, to a nuclear star cluster). Flattened
-density profiles in the outer regions are more unusual. Explaining the
-origin, similarity, and diversity of dwarf galaxy density profiles is an
-open question for theories of dwarf galaxy formation and evolution.
+density profiles in the outer regions are more unusual. Explaining in
+detail the origin, similarity, and diversity of dwarf galaxy density
+profiles is an open question for theories of dwarf galaxy formation and
+evolution.
 
@@
 tidal field (see Section~\ref{sec:discussion}).
-\citet{sestito+2023a, sestito+2023b} report a ``kink'' in the density
-profile, beginning around 30 arcmin for both Sculptor and Ursa Minor,
-that they interpret as caused by the effects of Galactic tides. They
+\citet{sestito+2023a, sestito+2023b}, for example, have recently
+reported a ``kink'' in the density profile, beginning around 30 arcmin
+for from the centre of each of these systems, that they interpret as
+potentially caused by the effects of Galactic tides. They
 spectroscopically follow up some of the most distant stars, finding
-multiple members between 6 and 12 half-light radii from the centre of
-each dwarf. If these dwarfs had exponential profiles, like Fornax, then
-these stars should be much rarer.
+members as far as 6 and 12 half-light radii from the centre of each
+dwarf. If these dwarfs had exponential profiles, like Fornax, then these
+far-outlying stars should be much rarer.
 
@@
 measured). Compared to Fornax, both Sculptor and Ursa Minor show an
-excess of stars beginning after \(\log R/R_h\approx 0.4\) and possibly
-exceeding 100 times the density of the exponential fit at large radii.
+excess of stars outside \(\log R/R_h\approx 0.4\) and which exceeds 100
+times the density of the exponential fit at large radii.
 
-A goal of this work is to determine, assuming \LCDM{}, if the effects of
-the Milky Way (or other satellites) are indeed responsible for the
-extended outer light profiles of Sculptor and Ursa Minor. If tides
-cannot explain these features, these features may instead be due to an
-extended stellar ``halo'' or second component of the galaxy---suggestive
-of a complex star formation or accretion history.
+A goal of this work is to determine, in the context of \LCDM{}, if the
+effects of the Milky Way (or other satellites) are indeed responsible
+for the extended outer light profiles of Sculptor and Ursa Minor. If
+tides cannot explain these features, these features may instead be due
+to an extended stellar ``halo'' or second component of the
+galaxy---suggestive of a complex star formation or assembly history.
 
@@
 challenge. Cosmological simulations can predict the overall abundance of
-more massive dwarf galaxies \citep[e.g.,][]{sawala+2016} and broadly
-examine the effects of tides \citep[e.g.,][]{riley+2024}. However, dwarf
-galaxies are often near the resolution limit. Insufficient resolution
-can lead to artificial disruption of dwarf galaxies and overprediction
-of tidal effects \citep[e.g.,][]{santos-santos+2025}. To address this
+the most massive dwarf galaxies \citep[e.g.,][]{sawala+2016} and broadly
+examine the effects of tides \citep[e.g.,][]{riley+2024}. However,
+simulated dwarf galaxies are often near the resolution limit.
+Insufficient resolution can lead to artificial disruption of dwarf
+galaxies and over-prediction of tidal effects
+\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To address this
 challenge, idealized simulations are often used to simulate a single
-subhalo in an approximate host potential, acchieving excellent numerical
-convergence. For instance, our simulations we describe later reach 3
+subhalo in an approximate host potential, achieving excellent numerical
+convergence. For instance, the simulations we describe later reach 3
 times higher resolution than Aquarius \citep{springel+2008} at a
@@
 Idealized simulations make numerous simplifications, neglecting mergers,
-cosmological context, mass assembly, and often baryonic physics. We use
-idealized simulations here to assess tidal effects after infall.
+cosmological context, mass assembly, and often baryonic physics. We
+shall use idealized simulations here to assess tidal effects after
+infall.
 
-Some of the earliest simulation work on tidal mass loss of dwarf
-galaxies originate from \citet{oh+lin+aarseth1995};
-\citet{piatek+pryor1995}; \citet{moore+davis1994};
-\citet{johnston+spergel+hernquist1995}. Already, using similar numerical
-techniques as we do later on, these works described the evolution of
-stellar and dark matter mass loss, the formation of symmetric tidal
-streams, and the minor changes to the central structure.
+Some early simulation work on tidal mass loss of dwarf galaxies
+originate from \citet{oh+lin+aarseth1995}; \citet{piatek+pryor1995};
+\citet{moore+davis1994}; \citet{johnston+spergel+hernquist1995}.
+Already, these works described the relative evolution of stellar and
+dark matter mass loss, that most mass loss happens near pericentre, the
+formation of symmetric tidal streams, and the minor changes to the
+central structure.
+
+Later, \citet{read+2006} showed that tidal tails would likely cause
+increasing velocity dispersion profiles, and interpreted the lack of
+similar observational features as indicating tides likely did not affect
+the stellar component of dwarf galaxies \citep[see
+also][]{klimentowski+2009a}.
+
+\citet{hayashi+2003}
+
+\citet{klimentowski+2009} shows that tides can cause a disky dwarf
+galaxy to transition into a spheroidal galaxy, and the velocity
+dispersion and maximum circular velocity are reasonable tracers of total
+mass.
+
+CDM halos also are found to be highly resilient to full disruption
+{[}\citet{EP2020}; {]}. In addition, NFW halos evolve along ``tidal
+tracks'' {[}e.g, \citet{PNM2008}, \citet{EN2021}; {]}. On the other
+hand, cored dark matter halos likely disrupt faster
+\citep[e.g.,][]{penarrubia+2010, errani+2023a}.
+
+\citet{green+vandenbosch2019}
+
+Recent, work by
+\citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
+showed that tidal effects are nearly entirely described as the removal
+of particles below a certain energy \citep[see
+also][]{choi+weinberg+katz2009}. \citet{strucker+2023} expanded on this
+idea, creating an analytic model for adiabatic tidal mass loss,
+confirming that an NFW halo is likely unable to be disrupted and
+explaining the evolution of NFW halos along similar tidal tracks.
 
 With precise orbits and a better understanding of the Milky Way
-potential and system, recent work continues to directly probe the
-dynamical histories of individual dwarf galaxies.
+potential, recent work continues to directly probe the dynamical
+histories of individual dwarf galaxies.
 \citet{borukhovetskaya+2022, dicintio+2024} both ran simulations tuned
-to Fornax, showing this galaxy's stellar component or globular clusters
-are likely not affected by tides. \citet{borukhovetskaya+2022a} used
-N-body simulations to analyze Crater II, showing that the present-day
-properties of the galaxy are challenging to reconcile with \LCDM{}
-initial conditions with tidal evolution. Most relevantly,
+to Fornax, showing that this galaxy's stellar component or globular
+clusters are likely not affected by tides. \citet{borukhovetskaya+2022a}
+also used N-body simulations to analyze Crater II, showing that the
+present-day properties of the galaxy are challenging to reconcile with
+\LCDM{} initial conditions with tidal evolution. Most relevantly,
 \citet{iorio+2019} conducted simulations of Sculptor, concluding tides
@@
 \begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
-3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J).
-}\end{equation} If \(r_J\) occurs within the visible extent of a galaxy,
-we should expect to find relatively clear signs of tidal disturbance.
-While strictly valid for circular orbits, assuming \(r_{\rm peri}\) for
-the host-dwarf distance works as most stars are lost near pericentre.
+3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
+}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) occurs within the
+visible extent of a galaxy, we should expect to find relatively clear
+signs of tidal disturbance. While strictly valid for circular orbits,
+assuming \(r_{\rm peri}\) for the host-dwarf distance works as most
+stars are lost near pericentre.
 
@@
 \citet{penarrubia+2009}, marking where the galaxy is still in
-disequilibrium after pericentre in a highly-eccentric orbit. The break
-radius \(r_{\rm break}\) is proportional to the velocity dispersion
-,\(\sigma_v\), and time elapsed since pericentre ,\(\Delta t\),
-\begin{equation}\protect\phantomsection\label{eq:r_break}{
+disequilibrium after pericentric passage in a highly-eccentric orbit.
+The break radius \(r_{\rm break}\) is proportional to the velocity
+dispersion, \(\sigma_v\), and time elapsed since pericentre,
+\(\Delta t\), \begin{equation}\protect\phantomsection\label{eq:r_break}{
 r_{\rm break} = C\,\sigma_{v}\,\Delta t
@@
 galaxy, we consider a toy model. We evolve an NFW subhalo with
-\(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) in an NFW static host halo with
-\(\rmax=25\,\kpc\) and \(\vmax = 207.4\,\kms\). The stars are initially
-an exponential profile with a scale radius \(R_s=0.25\,\kpc\). We then
-evolve the orbit through a pericentre of 10 kpc starting from an
-apocentre of 100 kpc. We show the results 170 Myr after the first
-pericentre. See Section~\ref{sec:methods} for a more detailed
-description of our general simulation setup.
+\(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another NFW static
+host halo with \(\rmax=25\,\kpc\) and \(\vmax = 207.4\,\kms\). These
+choices are motivated by the inferred structure and masses of dSphs and
+the Milky Way, respectively. The stars follow initially an exponential
+profile with a scale radius \(R_s=0.25\,\kpc\). We then evolve the orbit
+through a pericentre of 10 kpc starting from an apocentre of 100 kpc. We
+show the results 170 Myr after the first pericentre. See
+Section~\ref{sec:methods} for a more detailed description of our general
+simulation setup.
 
@@
 centre, but becomes non-isotropic outside the break radius and shows
-nascent tidal tails. As a surface density profile (top left), these
-tidally disturbed stars appear as an extended, outer density excess
-relative to the initial conditions. This excess appears shortly after
-the break radius. The break radius also marks where the mean radial
-velocity of the stars transitions from 0 to positive --- the galaxy is
-out of equilibrium outside \(r_{\rm break}\).
+nascent tidal tails.These tidally disturbed stars appear as an extended,
+outer density excess relative to the initial conditions. This excess
+appears just outside the break radius. The break radius also marks where
+the mean radial velocity of the stars transitions from 0 to positive ---
+the galaxy is out of equilibrium outside \(r_{\rm break}\).
 
 To first order, the final density profile of this toy simulation indeed
-resembles Sculptor and Ursa Minor outer density excess. This thesis aims
-to investigate if a tidal explanation is permissible given the current
+resembles the outer excess in Sculptor and Ursa Minor. This thesis aims
+to investigate if a tidal explanation is viable given the current
 properties of each galaxy.
@@
 
-In this thesis, our goal is to review the evidence for an extended
+The goal of this thesis is to review the evidence for an extended
 density profile in Ursa Minor and Sculptor, to assess the impact of
@@
 
-In chapter \ref{sec:observations}, we describe how we compute
-observational density profiles from \citet{jensen+2024}. In chapter
+In Chapter \ref{sec:observations}, we describe how we compute
+observational density profiles following \citet{jensen+2024}. In Chapter
 \ref{sec:methods}, we review our simulation methods. Next, we present
-our results for the tidal effects on Sculptor and Ursa Minor in chapter
+our results for the tidal effects on Sculptor and Ursa Minor in Chapter
 \ref{sec:results}, We discuss our results, limitations, and implications
-in chapter \ref{sec:discussion}. Finally, chapter \ref{sec:summary}
-summarizes this work and discuss future directions for similar work and
+in Chapter \ref{sec:discussion}. Finally, Chapter \ref{sec:summary}
+summarizes this work and discuss future directions for similar work in
 the field of dwarf galaxies.

### 2025-09-02 thesis/methods.tex
julio sep 1 edits to §1, 2

@@
 main intention is to assess whether such excesses result from the
-effects of Galactic tides. To this purpose, we intend to use N-body
-simulations of the evolution of CDM halos in a Galactic potential,
-constrained to have the orbital parameters consistent with a dwarf's
-present-day position and velocity. We shall assume that the Galactic
-potential is the static, analytic potential inferred by
-\citet{mcmillan2011} from observations of kinematic tracers. We also
-assume the potential of each dwarf may be initially approximated by a
-cuspy NFW profile. Since the dwarfs in question are heavily dark matter
-dominated, we shall use a carefully selected sample of dark matter
-particles to mimic and track the evolution of an embedded tracer stellar
-component. In this Chapter, I describe the Galactic potential used, the
-orbital estimation method, the initial conditions setup, and the N-body
-method used.
+effects of Galactic tides. To this purpose, we use N-body simulations of
+the evolution of CDM halos in a Galactic potential, constrained to have
+the orbital parameters consistent with a dwarf's present-day position
+and velocity. We shall assume that the Galactic potential is the static,
+analytic potential inferred by \citet{mcmillan2011} from observations of
+kinematic tracers. We also assume that the potential of each dwarf may
+be initially approximated by a cuspy NFW profile. Since the dwarfs in
+question are heavily dark matter dominated, we shall use a carefully
+selected sample of dark matter particles to mimic and track the
+evolution of an embedded tracer stellar component. In this Chapter, I
+describe the Galactic potential used, the orbital estimation method, the
+initial conditions setup, and the N-body method used.
 
@@
 We ``paint'' stars onto dark matter particles using the particle-tagging
-method \citep[e.g.][]{bullock+johnston2005}, assuming spherical
+method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
 symmetry. We initially assume exponential stars with
@@
 differential energy distribution includes an additional \(g({\cal E})\)
-occupation term (BT87). We use Eddington inversion to find the
-distribution function, (eq. 4-140b in BT87) \begin{equation}{
+occupation term \citep{BT1987}. We use Eddington inversion to find the
+distribution function, \citep[eq. 4-140b in][]{BT1987} \begin{equation}{
 f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).

### 2025-09-09 thesis/introduction.tex
thesis updates

@@
 Idealized simulations make numerous simplifications, neglecting mergers,
-cosmological context, mass assembly, and often baryonic physics. We
-shall use idealized simulations here to assess tidal effects after
-infall.
+cosmological context, mass assembly, and often baryonic physics {[}e.g.,
+\citet{hayashi+2003}; bullock+johnston2005; \citet{klimentowski+2009};
+\citet{ogiya+2019}{]}. We shall use idealized simulations here to assess
+tidal effects after infall.
 
@@
 
-\citet{hayashi+2003}
-
-\citet{klimentowski+2009} shows that tides can cause a disky dwarf
-galaxy to transition into a spheroidal galaxy, and the velocity
-dispersion and maximum circular velocity are reasonable tracers of total
-mass.
-
 CDM halos also are found to be highly resilient to full disruption
@@
 
-\citet{green+vandenbosch2019}
+\citet{hayashi+2003}, \citet{green+vandenbosch2019}
 
@@
 framework to Sculptor and Ursa Minor.
+
+\textbf{todo: finish this synthesis}
 

### 2025-09-09 thesis/methods.tex
thesis updates

@@
 substantially because of the low masses and large pericentres of the
-dwarfs.
+dwarfs, so we assume a single point-mass particle for the backwards
+integration.
 
@@
 \citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
-rest frame is then
+rest frame is then in Cartesian coordinates
 \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
-uncertainties in the reference frame are typically smaller than the
+uncertainties in this reference frame are typically smaller than the
 uncertainties on a dwarf galaxy's position and velocity.
@@
 Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
-each component and the total circular velocity profile for our fiducial
-profile. The potential includes a stellar bulge, a thin and thick disk,
-and a dark matter NFW halo.
+each component and the total circular velocity profile for this
+potential. The potential includes a stellar bulge, a thin and thick
+disk, and a dark matter NFW halo.
 
@@
 \citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
-\citet{hernquist1990} bulge (blue dotted
-line).}\label{fig:v_circ_potential}
+\citet{hernquist1990} bulge (blue dotted line). \textbf{make radius
+lowercase}}\label{fig:v_circ_potential}
 \end{figure}
 
-\subsection{Orbits of Sculptor}\label{orbits-of-sculptor}
+\subsection{Sculptor's Orbit}\label{sculptors-orbit}
 
-Sculptor's orbital history is relatively well-constrained.
-Fig.~\ref{fig:scl_orbits} illustrates point particle orbits for 100
-samples of Sculptor's observed kinematics integrated backwards 5 Gyr in
-both Galactocentric coordinate slices (\(x\), \(y\), \(z\)) and in
-Galactocentric radius with time. All sampled orbits of Sculptor have
+Sculptor's orbital in the assumed potential is relatively
+well-constrained. Fig.~\ref{fig:scl_orbits} illustrates point particle
+orbits for 100 samples of Sculptor's observed kinematics integrated
+backwards 5 Gyr in both Galactocentric coordinate slices (\(x\), \(y\),
+\(z\)) and in Galactocentric radius with time. These orbits sample the
+uncertainties in distance, proper motion, and radial velocity, as given
+in Table~\ref{tbl:scl_obs_props}. All sampled orbits of Sculptor have
 nearly the same morphology---the orbit primarily resides in the
@@
 0.0027th quantile pericentre, yielding a pericentre of 43 kpc. Given the
-current observations, it is unlikely that Sculptor has a significantly
-smaller pericentre than our selected orbit.
+current observations, it is unlikely that Sculptor has had a
+significantly smaller pericentre than our selected orbit, which we refer
+to as the \smallperi{} orbit.
 
@@
 is at the centre with the disk lying in the \(x\)--\(y\) plane. Our
-selected \texttt{smallperi} orbit is plotted in black and light blue
+selected \smallperi{} orbit is plotted in black and light green
 transparent orbits represent the past 5Gyr orbits sampled over Sculptor
-observables in Table~\ref{tbl:scl_obs_props}. The orbit of sculptor is
+observables in Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is
 well-constrained in this potential and it is unlikely to achieve a
-smaller pericentre than our selected orbit.}\label{fig:scl_orbits}
+smaller pericentre than the \smallperi{} orbit.}\label{fig:scl_orbits}
 \end{figure}
@@
 The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
-The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
 }
@@
 \toprule
-Property & Mean & SmallPeri\\
+Property & mean & \smallperi{}\\
 \midrule
@@
 $t_i / \Gyr$ & -8.74 & -9.43\\
-$\vec{x}_{i} / \kpc$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10]\\
-$\vec{v}_i / \kms$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29]\\
+${x}_{i} / \kpc$ & 16.13 & -2.49\\
+${y}_{i} / \kpc$ & 92.47 & -42.78\\
+${z}_{i} / \kpc$ & 39.63 & 86.10\\
+$\V_{x\,i} / \kms$ & -2.37, & -20.56\\
+$\V_{y\,i} / \kms$ & -54.70 & -114.83\\
+$\V_{z\,i} / \kms$ & 128.96 & -57.29\\
 pericentre / kpc & 53 & 43\\
@@
 
-\subsection{Orbits of Ursa Minor}\label{orbits-of-ursa-minor}
+\subsection{Ursa Minor's Orbit}\label{ursa-minors-orbit}
 
-Similar to Sculptor, Ursa Minor has a well-constrained orbit.
-Fig.~\ref{fig:umi_orbits} shows 100 random point-orbits of Ursa Minor.
-Initially, we select an orbit with approximately the \(3\sigma\)
-smallest pericentre as in Sculptor.
+Similar to Sculptor, Ursa Minor has a well-constrained orbit in the
+assumed MW potential. Fig.~\ref{fig:umi_orbits} shows 100 random
+point-orbits of Ursa Minor. As for Sculptor, we select an orbit with
+approximately the \(3\sigma\) smallest pericentre, the Ursa Minor
+\(\smallperi{}\) orbit.
 
-Ursa Minor's N-body orbit diverges from the point particle orbit. To
-ensure the final conditions of the N-body simulation are close to the
+We shall see later that the orbit of Ursa Minor's N-body model differs
+from the point particle orbit because of the effects of tidal mass loss.
+To ensure the final conditions of the N-body simulation are close to the
 intended final position, we iteratively adjust Ursa Minor's initial
@@
 iterations to an orbit agreeing with the observed kinematics of Ursa
-Minor.
+Minor. Since Sculptor's orbit is less strongly affected by tides, we do
+not carry out this correction for Sculptor.
 
@@
 \caption[Ursa Minor Selected Orbits]{Properties of selected orbits for Ursa Minor.
-The "smallperi" orbit is the initial point orbit and the "smallperi.5" is the initial orbit for the N-body simulation.
+The \smallperi{} orbit is the initial point orbit and the \smallperi{}.5 is the initial orbit for the N-body simulation.
 }
@@
 \toprule
-Property & Mean & SmallPeri.1 & SmallPeri.5\\
+Property & Mean & \smallperi{}.1 & \smallperi{}.5\\
 \midrule
@@
 $t_i / \Gyr$ & -8.74 & -9.53 & -9.53\\
-$\vec{x}_{i} / \kpc$ & [4.88, -65.11, 50.78] & [-16.48 69.92 21.05] & [17.40, 74.51, 21.34]\\
-$\vec{v}_i / \kms$ & [-34.28, 77.11, 101.38] & [16.32, 39.86, -116.99] & [14.27, 48.62, -114.08]\\
+${x}_{i} / \kpc$ & 4.88 & -16.48 & 17.40\\
+${y}_{i} / \kpc$ & -65.11 & 69.92 & 74.51\\
+${z}_{i} / \kpc$ & 50.78 & 21.05 & 21.34\\
+$\V_{x\,i} / \kms$ & -34.28 & 16.32 & 14.27\\
+$\V_{y\,i} / \kms$ & 77.11 & 39.86 & 48.62,\\
+$\V_{z\,i} / \kms$ & 101.38 & -116.99 & -114.08\\
 pericentre / kpc & 37 & 29 & 28\\
@@
 We use \agama{} \citep{agama} to generate the initial N-body dark matter
-halo. We assume galaxies are described by an NFW dark matter potential
-(Eq.~\ref{eq:nfw}). We also assume the stars do not contribute to the
-potential. The dark matter density is truncated in the outer regions by
-\begin{equation}{
+halo. We assume galaxies are described by a spherical, isotropic NFW
+dark matter potential (Eq.~\ref{eq:nfw}). We also assume the stars do
+not contribute to the potential. The dark matter density is truncated in
+the outer regions by \begin{equation}{
 \rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t),
@@
 
-From the observed properties of Sculptor and Ursa Minor, we infer
-reasonable \LCDM{} initial halo conditions.
+The observed half-light radius and velocity dispersion of Sculptor and
+Ursa Minor, together with the mass-concentration relation of \LCDM{}
+halos, determines the properties of the N-body models adopted for each
+dwarf.
 
-Table~\ref{tbl:derived_props} reports our inferred halo and kinematic
+Table~\ref{tbl:derived_props} lists the our inferred halo and kinematic
 properties of Sculptor and Ursa Minor. First, taking the absolute
@@
 \(M_\star \sim 3.1 \times 10^6 \Mo\) and
-\(M_\star \sim 7 \times 10^5 \Mo\). Based on the stellar mass-\(\vmax\)
-relation \citep[from][]{fattahi+2018}, see also \ref{fig:smhm}, Sculptor
-and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
-\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
+\(M_\star \sim 7 \times 10^5 \Mo\), respectively. Based on the stellar
+mass-\(\vmax\) relation {[}fig.~1 from \citet{fattahi+2018};{]},
+Sculptor and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\)
+and \(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
 \(z=0\) mass-concentration relation, this constraint translates into a
 \(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
-galaxy.
+galaxy. These choices set the velocity dispersion of a (massless)
+stellar system embedded in such a halo, depending on its half-light
+radius.
 
-Fig.~\ref{fig:scl_halos} illustrates these estimates visually for both
-galaxies. The stellar-mass \(\vmax\) constraint translates to an
-estimate of \(\vmax\) only (horizontal band). The mass-concentration
-relation describes the relationship between \(\vmax\) and \(\rmax\)
-(diagonal linear band). And finally, we include a curved line which
-illustrates the halo \(\vmax\) which has a specified initial LOS
-velocity dispersion given \(\rmax\)
-\(, \vcirc(R_h) / \sqrt{3} \approx \sigma_v\).
-\(R_h \approx 0.24\,\kpc\) for both galaxies.
+The situation is illustrated in Fig.~\ref{fig:scl_halos}. Here we show,
+in the NFW \(\rmax\)--\(\vmax\) plane, the velocity dispersion of
+stellar systems with half-light radius matching that of Sculptor and
+Ursa Minor (\(R_h=240\,\)pc for both galaxies). For Scl, assuming
+\(\vmax = 31\,\kms\) and \(\rmax = 6\,\kpc\) gives
+\(\sigma_v \approx = XXX\), in contrast with the observed dispersion of
+\textbf{add}. To match the observed value, we choose a somewhat smaller
+value for r\_max=?? kpc (shown by the blue circle in the left panel of
+Fig. 3.4). We have also experimented with a smaller rmax and Vmax for
+Scl (yellow square in the same panel, to account for the fact that the
+half-light radius and velocity dispersion may be affected by tidal mass
+losses.
+
+For Ursa Minor, the cosmological mean predicts \ldots. We chose a halo
+with a higher \(\vmax\) and \(\rmax\) so the initial dispersion is
+\textbf{add}. While above the observed value, we will show in the next
+section this results in a final velocity dispersion near to the observed
+values. \textbf{add}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_velocity.pdf}}
+\caption{Initial halo velocity profiles}
+\end{figure}
 
@@
 \(R_h\). So, the tidal effects on stars should be similar for halos with
-similar velocity dispersions.
+similar velocity dispersions. \textbf{add curve for density at
+pericentre}
 
@@
 match the present-day velocity dispersion at the end of the simulation.
+\textbf{Make figure of initial halos, mark observed half-light radius
+and velocity dispersion, pericentric MW density?}
 
@@
 
+\subsection{\texorpdfstring{The N-body code:
+\gadget{}}{The N-body code: }}\label{the-n-body-code}
+
 To simulate the tidal evolution of galaxies, we use ``N-body''
@@
 \gadget{}. Since gravity is scale-free, we use the same isolation run
-for all halos. We adopt \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)
-for the isolation halo based on Sculptor's mean properties. We run this
-model for 5 Gyr (about three times the free fall timescale
-\(t_{\rm ff} = \vcirc / r \approx 1.5\,\Gyr\) at \(r_{200}=36\,\)kpc).
+for all halos and rescale the results to the desired values of size and
+mass. We adopt a fiducial value of \(\rmax = 6.0\,\)kpc and
+\(\vmax = 31\,\kms\) for the isolation halo based on Sculptor's mean
+properties. We run this model for 5 Gyr (about three times the crossing
+time \(t_{\rm cross} = \vcirc / r \approx 1.5\,\Gyr\) at
+\(r_{200}=36\,\)kpc).
 
@@
 discussion of this choice, which is similar to the \citet{power+2003}
-suggested softening. We use the relative tree opening criterion with the
-accuracy parameter set to 0.005, and adaptive time stepping with
-integration accuracy set to 0.01.
+suggested softening. We use in \gadget{} the relative tree opening
+criterion with the accuracy parameter set to \(\alpha =0.005\) (so nodes
+are opened when \(M\,l/r^3 < \alpha |a|\) for a distance from the node
+of mass \(M\) of \(r\), \(l\) is the side length of the node, and \(a\)
+is the particle's total acceleration), and adaptive time stepping with
+integration accuracy set to \(\eta=0.01\) (particles must take
+time-steps smaller than \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for
+acceleration \(a\)).
 
@@
 after running the model in isolation. This figure shows that our
-numerical methods are able to approximate well an NFW halo down to an
-innermost radius that strongly depends on resolution. The larger the
-number of particles, the smaller the radius that is effectively
+numerical methods are able to approximate well an NFW halo down to a
+given innermost radius that strongly depends on resolution. The larger
+the number of particles, the smaller the radius that is effectively
 ``resolved'' in a given simulation. For the Sculptor halo shown in this
@@
 100 pc. For reference, the half-light radius of Sculptor is roughly 100
-pc, which means that at least 10 million particles would be needed to
-follow faithfully its tidal evolution. Vertical arrows in
+pc, which means that at least 10 million particles are needed to follow
+faithfully its tidal evolution. Vertical arrows in
 Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
@@
 reduction in density starting around \(r = 30\,\kpc\) is expected given
-our truncation choice.}\label{fig:numerical_convergence}
+our outer truncation choice.}\label{fig:numerical_convergence}
 \end{figure}
@@
 Next, we evolve the halo in the Galactic potential. We scale the relaxed
-snapshot and softening length to match the initial halo in
-\ref{tbl:initial_halos}, and shift the snapshot to the initial
-conditions from the orbital analysis (see
+snapshot and softening length to match the initial halo (see
+Table~\ref{tbl:initial_halos}), and shift the snapshot to the initial
+conditions inferred from the orbital analysis (see
 Tables~\ref{tbl:scl_orbits}, \ref{tbl:umi_orbits}). We then evolve the
-full N-body NFW model forward in time in the Galactic potential and
-follow it in time until the present time, when the halo is closest to
-the observed position of the galaxy.
+full N-body NFW model forward in time in the Galactic potential until
+the present time, when the halo is closest to its present-day observed
+position in the MW halo.
 
-\subsection{Tidal mass losses}\label{sec:shrinking_spheres}
+\subsection{Halo centring}\label{sec:shrinking_spheres}
 
 To accurately follow the evolution of a halo, it is important to
-determine the centre of the halo at each time chosen for analysis. We
-use a shrinking-spheres centre method inspired by \citet{power+2003}.
-First, we start with an initial centre estimate from the last timestep.
-Then, we calculate the radius of all particles from the centre, remove
-particles with a radius beyond the 0.975 quantile of the centre, and
-recalculating the centre of mass. The procedure is repeated until the
-selection radius is less than \textasciitilde1kpc or fewer than 0.1\% of
-particles remain. After a centre has been chosen, we remove all unbound
-particles based on the \gadget{} calculated potential of the halo. For
-all future timesteps, we consider only particles retained from the
-previous iteration.
+determine the centre of the self-bound halo remnant at each time chosen
+for analysis. We use a shrinking-spheres centre method inspired by
+\citet{power+2003}. First, we start with an initial centre estimate from
+the last timestep. Then, we calculate the radius of all particles from
+the centre, remove particles with a radius beyond the 0.975 quantile of
+the centre, and recalculate the centre of mass. The procedure is
+repeated until the selection radius is less than \textasciitilde1kpc or
+fewer than 0.1\% of particles remain. After a centre has been chosen, we
+remove all unbound particles based on the \gadget{} calculated potential
+of the halo. For all future timesteps, we consider only particles
+retained from the previous iteration.
 
@@
 
-\subsection{The stellar component}\label{sec:painting_stars}
+\subsection{Sculptor and Ursa Minor's initial stellar
+components}\label{sec:painting_stars}
 
@@
 method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
-symmetry. We initially assume exponential stars with
-\(R_s = 0.10\,,\kpc\) for both galaxies.
+symmetry. We initially assume stars follow a projected exponential law
+(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,,\kpc\) for both
+galaxies. The tagging method assigns a probability to each dark matter
+particle, which is proportional to the ``light-to-mass'' ratio required
+to match the assumed stellar light profile. We briefly describe the
+procedure next. We refer interested readers to \textbf{REF (rapha?)} for
+further details. \emph{code may be found at?}
 
@@
 \({\cal E}\) the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
-know \(f({\cal E})\), the distribution function (phase-space density in
-energy), then we assign a stellar weight to a given particle with energy
-\({\cal E}\) using \begin{equation}{
+know \(f({\cal E})\), the distribution function (phase-space density as
+a function of energy), then we assign a stellar weight to a given
+particle with energy \({\cal E}\) using \begin{equation}{
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
@@
 isolated systems and show excellent agreement with the assumed stellar
-density profile.
+density profile. \textbf{Add figure showing stellar and dark matter
+density profiles\ldots{}}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_conditions.pdf}}
+\caption[Initial stellar systems]{The initial dark matter and stellar
+profiles for Sculptor and Ursa Minor. The initial profiles are dotted
+and final solid.}
+\end{figure}

### 2025-09-09 thesis/results.tex
thesis updates

@@
-In this chapter, we present our results from our simulations of Sculptor
-and Ursa Minor. In each case, we first discuss the range of reasonable
-orbits and initial dark matter halos for each galaxy. Then, we describe
-how tides affect the evolution of each galaxy. We furthermore consider
-how the effects of the LMC changes the results, including using
-additional N-body simulations in evolving potentials. Finally, we review
-the predictions and characteristics of the present-day stellar
-populations and properties from our models.
+As discussed in Chapters \ref{sec:introduction}, \ref{sec:data},
+Sculptor and Ursa Minor have unusually extended stellar light profiles.
+Using our simulation setup introduced in Chapter \ref{sec:methods}, we
+present our simulation results for each galaxy in this Chapter. We find
+that tides may drive significant evolution of each galaxies total mass.
+However, the compact stellar distribution for each galaxy is nearly
+unaffected. We also show that the LMC strongly perturbs Scl's orbit.
+But, the tidal effects when considering an LMC are yet weaker than in
+the Milky Way only case. We conclude that tides are unlikely to affect
+the stellar component of either galaxy.
 
-\section{Sculptor}\label{sculptor}
+\subsection{Tidal evolution in the Milky
+Way}\label{tidal-evolution-in-the-milky-way}
 
-\subsection{Milky Way tides}\label{milky-way-tides}
-
-\subsection{Tidal effects}\label{tidal-effects}
-
-The Milky Way tides indeed affect Sculptor's dark matter halo.
+The Milky Way tides drive significant mass loss for Sculptor.
 Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
 of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
-previous pericentres. This simulation uses the \texttt{compact} initial
-halo on the \texttt{smallperi} orbit. The dark matter forms large
-streams which orbit the Milky Way several times. However, note that the
-central dark matter cusp is similar in appearance across all simulation
-snapshots.
+previous pericentres. This simulation uses the fiducial initial halo on
+the \smallperi{} orbit. The dark matter forms large streams which orbit
+the Milky Way several times. Note that, in contrast to the outer
+regions, the central dark matter cusp is similar in appearance across
+all simulation snapshots.
 
-More quantitatively, the halo loses about 99\% of its initial dark
-matter mass. This corresponds in a reduction from the initial
-\(v_{\rm max} = 31\,\kms\) to \(v_{\rm max} = 22\,\kms\). However, the
-stellar component of Sculptor is relatively unaffected.
+Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
+velocity profile of the dark matter, and the evolution of the maximum
+circular velocity. Sculptor's final dark matter halo evolves from
+\(\vmax=31\,\kms\) at \(\rmax=3.2\,\kms\) to \(\vmax = 22\,\kms\) at
+\(\rmax = 2\,\kpc\). In terms of bound mass, the mass evolves from
+\(0.40 \times 10^{10}\,\Mo\) to \(0.026\times10^{10}\,\Mo\). The
+circular velocity maximum evolves along the tidal track predicted by
+\citet{EN2021}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
+\caption[Sculptor simulation snapshots]{Images of the dark matter
+evolution over a selection of past apocentres and the present day
+position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
+(approximately orbital) plane and the colourscale is logarithmic
+spanning 5 orders of magnitude between the maximum and minimum values.
+The green dot represents the final (observed) position of the galaxy and
+the grey curve represents the orbit over one radial
+oscillation.}\label{fig:scl_sim_images}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
+\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
+orbit. Todo: add velocity dispersion plot to RHS, label initial and
+final with times}\label{fig:scl_tidal_track}
+\end{figure}
+
+\subsection{}\label{section}
 
@@
 
-The analytically motivated break and tidal radii corroborate a weak
+The analytically motivated break and tidal radii also support a weak
 tidal effect on stars. In both cases, these radii work out to be
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
-\caption[Sculptor simulation snapshots]{Images of the dark matter
-evolution over a selection of past apocentres and the present day
-position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
-(approximately orbital) plane and the colourscale is logarithmic
-spanning 5 orders of magnitude between the maximum and minimum values.
-In this image, stars occupy only ever a few pixels so are not
-plotted.}\label{fig:scl_sim_images}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
-\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
-orbit. Todo: add velocity dispersion plot to
-RHS}\label{fig:scl_tidal_track}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
-\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
-velocity dispersion within 1 kpc for different Scl models. In all cases,
-the evolution is mild. Note that binarity may reduce the inflate the
-observed velocity dispersion by \textasciitilde{} 1 km/s, so the
-conservative lower limit is around 8 km/s. \textbf{TODO}, add bound mass
-with time. Maybe combine with tidal tracks and radius / time orbit.}
-\end{figure}
-
-\begin{figure}
-\centering
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
-\caption[Sculptor initial and final density profiles]{Effects on
-exponential initial stars. TODO: plot 2D sky proj.
-stars}\label{fig:scl_smallperi_i_f}
+\caption[Sculptor initial and final density profiles]{The tidal effects
+on Scl's stellar component, for the \smallperi{} orbit with the fiducial
+halo and exponential stars with \(R_s=0.10\,\kpc\). \textbf{Left:} the
+final 2D projected density of stars on the sky, with colours
+representing logarithmically increasing density. The circle marks the
+break radius. \textbf{Right:} The initial (dotted) and final (solid)
+stellar density profiles as compared to the observed stellar density
+profile. Arrows mark the break and Jacobi (Eq.~\ref{eq:jacobi_radius})
+radii. \textbf{add initial sky-projected density, t=xxx in legend,
+sculptor in figure title}}\label{fig:scl_smallperi_i_f}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
-\caption[Sculptor Plummer initial and final density profiles]{effects on
-Plummer initial stars.}\label{fig:scl_smallperi_plummer_i_f}
+\caption[Sculptor Plummer initial and final density profiles]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f} except for Plummer initial stars with
+\(R_h = 0.20\,\kpc\). While a faint stream may be visible with deeper
+observations, effects on the stellar profile are minimal. \emph{could
+overlay orbit on left}}\label{fig:scl_smallperi_plummer_i_f}
 \end{figure}
@@
 
-The Milky Way isn't the only galaxy in town. Becoming more apparent, the
-infall of the LMC substantially affects the Milky Way system. The mass
-of the LMC may be as high as 1/5 of the Milky Way mass, so including an
-LMC in the Milky Way potential may change conclusions about properties
-and orbits of the Milky Way satellite system.
+The Milky Way isn't the only galaxy in town. Recently, work has shown
+that the infall of the LMC may substantially affects the Milky Way
+system
+\citep[e.g.,][]{erkal+2019, cautun+2019, garavito-camargo+2021, vasiliev2023}.
+The mass of the LMC may be as high as 1/5 of the Milky Way mass.
+Including an LMC in the Milky Way potential may change conclusions about
+properties and orbits of the Milky Way satellite system, including Scl
+\citep[e.g.,][]{patel+2020, battaglia+2022}.
 
@@
 
-As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
-influenced by the presence of an LMC.
-
-To explore the effects of the LMC on Sculptor, we use the publically
-available L3M11 model from \citet{vasiliev2024}. This model stars with a
-lighter Milky Way than our fiducial \citet{EP2020}, \citet{mcmillan2011}
-like model. The L3M11 potential is an evolving multipole approximation
-of an N-body simulation with both a live MW and LMC dark matter halo.
-The MW intital conditions were an NFW with \(r_s=16.5\,\)kpc and mass
+To explore the effects of the LMC on Sculptor, we use the publicly
+available L3M11 model from \citet{vasiliev2024}. While
+\citet{vasiliev2024} present other models, the L3M11 model has the
+highest MW and LMC masses, and the (recent) orbital history appears to
+be independent of the potential details. This model stars with a lighter
+Milky Way than our fiducial, \citet{mcmillan2011-like} model. The L3M11
+potential is an evolving multipole approximation of an N-body simulation
+with both a live MW and LMC dark matter halo. The potential additionally
+includes the reflex motion of the MW due to the LMC. The MW was
+initialized as an NFW dark matter halo with \(r_s=16.5\,\)kpc and mass
 \(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
@@
 previous LMC pericentre at about 6 Gyr ago.
+
+For our orbit, we select the orbit with the \(3\sigma\)-smallest LMC
+pericentre. Results are similar when selecting for a small MW pericentre
+instead. We also use, instead, the \texttt{small} halo for Sculptor, as
+the tidal effects do not reduce the velocity dispersion, unlike for the
+MW-only model.
 
@@
 LMC) to the MW is plotted (left), and the Sculptor - LMC distance
-(right.)}\label{fig:scl_lmc_orbits_effect}
+(right.) \textbf{cut scale on buttom
+panel}}\label{fig:scl_lmc_orbits_effect}
 \end{figure}
 
-\textbf{tinyperilmc}
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_orbits_mass_effect.png}}
+\caption[Sculptor Orbits with LMC]{The long-term orbital history of
+Sculptor is uncertain. In light, transparant lines, the orbits of
+Sculptor in three different LMC/MW mass models from \citet{vasiliev2024}
+are shown for random samples of the observed properties. The LMC orbits
+are in solid, thick lines of the corresponding colour. The L2M11 has a
+lighter LMC mass, and the L3M10 model has a lighter MW mass than our
+fiducial L3M11 LMC model.}\label{fig:scl_lmc_orbits_mass}
+\end{figure}
 
-\begin{itemize}
-\item
-  ra = 15.0183 dec = -33.7186 distance = 73.1 pmra = 0.137 pmdec =
-  -0.156 radial\_velocity = 111.2
+\begin{table*}[t]
+\centering
+\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
+The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
+The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+}
+\label{tbl:scl_orbits}
+\begin{tabular}{ll}
+\toprule
+Property & smallperi-LMC\\
+\midrule
+distance / kpc & 73.1\\
+$\pmra / \masyr$ & 0.137\\
+$\pmdec / \masyr$ & –0.156\\
+LOS velocity / $\kms$ & 111.2\\
+$t_i / \Gyr$ & -2\\
+${x}_{i} / \kpc$ & 4.30\\
+${y}_{i} / \kpc$ & 138.89\\
+${z}_{i} / \kpc$ & 125.88\\
+$\V_{x\,i} / \kms$ & 6.89\\
+$\V_{y\,i} / \kms$ & -56.83\\
+$\V_{z\,i} / \kms$ & 52.09\\
+pericentre / kpc & 38.82\\
+peri-LMC / kpc & ?\\
+apocentre / kpc & 187.50\\
+$t_{\rm last\ peri} / {\rm Gyr}$ & -0.46\\
+$t_{\rm last\ peri-LMC} / \Gyr$ & -0.1?\\
+Number of pericentres & 1-2\\
+\bottomrule
+\end{tabular}
+\end{table*}
 
-  t\_i = -2.00 pericentre = 38.82 apocentre = 187.50 t last peri = -0.33
-  x\_i = {[}4.30 138.89 125.88{]} v\_i = {[}6.89 -56.83 52.09{]}
-\end{itemize}
-
-\subsubsection{Tidal effects}\label{tidal-effects-1}
+\subsubsection{Tidal effects}\label{tidal-effects}
 
@@
 
-A final note is that this model only considers the recent tidal effects
-.As illustrated in the range of possible orbits in
-Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
-experienced an extremely small pericentre with the Milky Way. This
-pericentre has the potential to substantially rearrange the stellar
-component, drive large tidal mass loss, and create a Plummer-like
-stellar density profile. However, this encounter is highly dependent on
-the choice of the MW-LMC potential model, and may not occur at all. Long
-time integration in dynamic potentials amplifies uncertainties in the
-inputs and may not be reliable. \textbf{TODO: I would love to get this
-model to work\ldots{}}
-
 \begin{figure}
@@
 \end{figure}
+
+\subsubsection{A possible second MW
+pericentre}\label{a-possible-second-mw-pericentre}
+
+A final note is that this model only considers the recent tidal effects
+.As illustrated in the range of possible orbits in
+Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
+experienced an extremely small pericentre with the Milky Way. This
+pericentre has the potential to substantially rearrange the stellar
+component, drive large tidal mass loss, and create a Plummer-like
+stellar density profile. However, this encounter is highly dependent on
+the choice of the MW-LMC potential model, and may not occur at all. Long
+time integration in dynamic potentials amplifies uncertainties in the
+inputs and may not be reliable.
+
+To illustrate the possible effects of this previous pericentre, we
+create a model with a goal of matching the \(3\sigma\) smallest
+pericentre with the MW. Because of strong tidal shocking, the final
+conditions depend strongly on the initial conditions. In addition,
+actions are not conserved in this evolving potential, so we use the
+\citet{vasiliev2024} method of iteratively updating the initial
+conditions by empirically estimating the Jacobian. Our final position
+reaches adequate, but not perfect, agreement with the intended final
+position.
+
+As a result of this proceedure, while the point particle reaches a
+pericentre of \(\sim 4\,\kpc\), the adjusted N-body orbit instead only
+reaches a pericentre of \(\sim 12\,\kpc\).
 

### 2025-09-24 thesis/abstract.tex
half-done discussion

@@
-The satellite galaxies of the Milky Way (MW), with the exception of the
-Magellanic Clouds, are all dwarf spheroidals (dSphs)---gas-free,
-non-rotating stellar systems whose gravity is dominated by dark matter.
-The light profiles of dSphs are usually well-approximated by an
-exponential law, with a sharp outer cutoff. Yet systems like the
-Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
-stars far from the main body, out to \(\sim10\) effective radii,
-indicative of outer deviations from an exponential law. The origin of
-these extended profiles is not clear, and may reflect either the
-influence of Galactic tides or, instead, an innate feature. In this
-thesis, we review the existence of extended light profiles in Scl and
-UMi, validating the Bayesian membership catalogue of
-\citet{jensen+2024}. To evaluate if tides indeed can explain these
-density features, we conduct idealized N-body simulation of both
-galaxies in the tidal field of the Milky Way. We find that neither dwarf
-has been subjected to tidal forces strong enough to create the observed
-extended light profiles from initially exponential profiles. One
-complication is that Sculptor's orbit is strongly influenced by the
-presence of the Large Magellanic Cloud (LMC). Because the LMC's mass is
-poorly constrained, Sculptor's long term orbital history is highly
-uncertain. Despite this uncertainty, our models still suggest that the
-combined tides of the LMC and MW are unable to explain Scl's outer
-profile. We conclude that the extended light profiles of Scl and UMi are
-innate, likely reflecting their assembly histories. Mergers, accretion
-events, or complex, multi-component star formation episodes are possible
-explanations for the observations ed extended light profiles.
+The satellite galaxies of the Milky Way (MW), except the Magellanic
+Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating
+stellar systems dynamically dominated by dark matter. The light profiles
+of dSphs are usually well-approximated by an exponential law, with a
+steep outer cutoff. Yet systems like the Sculptor (Scl) and Ursa Minor
+(UMi) dSphs have distant member stars, out to \(\sim10\) effective radii
+from the centre, indicative of outer deviations from an exponential law.
+The origin of these extended profiles is unclear, and may reflect either
+the influence of Galactic tides or an innate feature. In this thesis, we
+review the existence of extended light profiles in Scl and UMi,
+validating the Bayesian membership catalogue of \citet{jensen+2024}. To
+evaluate if tides indeed can explain these density features, we conduct
+idealized N-body simulation of both galaxies in the tidal field of the
+Milky Way. We find that neither dwarf has been subjected to tidal forces
+strong enough to create the observed extended light profiles from
+initially exponential profiles. One complication is that Sculptor's
+orbit is strongly influenced by the presence of the Large Magellanic
+Cloud (LMC). Our models still suggest that the combined tides of the LMC
+and MW are unable to explain Scl's outer profile. We conclude that the
+extended light profiles of Scl and UMi are innate, likely reflecting
+their assembly histories. Mergers, accretion events, or episodic star
+formation are possible explanations for the observations of extended
+light profiles.

### 2025-09-24 thesis/data.tex
half-done discussion

@@
 from the satellite. Each likelihood is normalized as a probability
-density over the respective parameter space.
+density over the respective parameter space. The appendix discusses the
+details of quality cuts and the likelihood calculation.
 
@@
 
-We analyze stellar distributions in the tangent plane, considering as
-well the projected shape of a galaxy. The tangent plane coordinates
-\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
-the plane tangent relative to the galaxy centre. To account for the
+We analyze stellar distributions in the tangent plane and consider the
+projected shape of a galaxy. The tangent plane coordinates \(\xi\) and
+\(\eta\) are offsets in right ascension (RA) and declination as measured
+on the plane tangent relative to the galaxy centre. To account for the
 elliptical shape of the galaxy, we use \(R_{\rm ell}\), which we define
@@
 Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
-Ursa Minor stands out because of its higher ellipticity. Sculptor and
-Ursa Minor show a population of members past \(6R_h\) in radius, but
-Fornax does not, even though Fornax has many more stars. Fornax also has
-a bluer CMD, indicative of more recent star formation. Otherwise, there
+Ursa Minor stands out because of its higher ellipticity. While Sculptor
+and Ursa Minor show an extended population of members past \(6R_h\) in
+radius Fornax does not despite having many more stars. Fornax also has a
+bluer CMD, indicative of more recent star formation. Otherwise, there
 are no major morphological differences between these galaxies.
@@
 ``CMD+PM'' selected sample, which does not include a spatial likelihood.
-If even fainter, more extended features exist, they are not clearly
+If even fainter, more extended features exist, they are not obviously
 detectable with \emph{Gaia}.
@@
 exterior to the first empty bin in either direction. We use symmetric
-Poisson uncertainties as ``error bars'' in the density estimate at each
-bin.
+Poisson uncertainties as error bars in the density estimate at each bin.
 
@@
 they begin to deviate outside \(6 R_h\). Outside this radius, the
-fiducial profile extends 1--2 magnitudes below the CMD+PM selection
+fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
 background. Because the fiducial profile in this region depends
@@
 relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
-corresponding to an elliptical radius of 37 arcminutes. This cluster
-does not have a bright RGB, so has few stars brighter than a magnitude
-\(G=22\). The cluster has little effect on the elliptically-averaged
+corresponding to an elliptical radius of 37 arcminutes. Muñoz 1 does not
+have a bright RGB, so has few stars above the \emph{Gaia} magnitude
+limit. The cluster has little effect on the elliptically-averaged
 density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).

### 2025-09-24 thesis/discussion.tex
julio edits at durham
half-done discussion

@@
 
-\subsection{Long term orbital history}\label{long-term-orbital-history}
+\subsection{Challenges to backwards-time
+integration}\label{sec:scl_orbit_variability}
 
@@
 backwards time integration
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
+\caption[Sculptor Orbits with LMC]{The long-term orbital history of
+Sculptor is uncertain. In light, transparent lines, the orbits of
+Sculptor in three different LMC/MW mass models from \citet{vasiliev2024}
+are shown for random samples of the observed properties. The LMC orbits
+are in solid, thick lines of the corresponding colour. The L2M11 has a
+lighter LMC mass, and the L3M10 model has a lighter MW mass than our
+fiducial L3M11 LMC model.}\label{fig:scl_lmc_orbits_mass}
+\end{figure}
+
+Our LMC model only simulates recent tidal effects---the long-term orbit
+of Scl is essentially unknown. Fig.~\ref{fig:scl_lmc_orbits_effect}
+illustrates the orbits of Scl for three different LMC-MW models in
+\citet{vasiliev2024}: L2M11 has a lighter LMC and L3M10 has a lighter MW
+than the L3M11 potential we used. All orbits diverge significantly after
+2 Gyr ago. In addition, other affects such as the lumpiness of the MW
+halo, the growth of the MW halo, dynamical friction, and limitations of
+this potential make long-term backwards time integration yet more
+unreliable \citep[e.g.,][]{dsouza+bell2022}.
+
+In Fig.~\ref{fig:scl_lmc_orbits_effect}, some orbits have a small
+\(\lesssim 10\,\kpc\) pericentre with the Milky Way. In the Appendix, we
+consider a model with a previous MW pericentre of 4kpc. While this
+encounter should affect the inner stellar distribution of Sculptor based
+on the Jacobi radius, the observed light profile remains unaffected.
+Because this model experiences just one tidal shock, stronger tidal
+streams form, but the tidal impact is not strong enough to reshape the
+observed stellar distribution. We conclude that
+
+As discussed in \citet{vasiliev2024}, if the LMC has had a past, second
+pericentre with the Milky Way, there is a chance that UMi was previously
+bound to the LMC. We find that this occurs about 15\% of the time for
+this choice of potential. However, backwards integration may not
+appropriately represent the true probability of a Magellanic origin,
+since tidal stripping is not time reversible. Because of the possibility
+of this past encounter, and the high uncertainties associated with
+backwards integration, we note this as a possibility but defer to the
+discussion for our interpretations.
 
@@
 
-\section{Comparison to other work}\label{comparison-to-other-work}
+\section{Comparison to previous work}\label{comparison-to-previous-work}
+
+For both Sculptor and Ursa minor, these galaxies have been studied
+extensively in both a theoretical and observational context.
 
@@
   \citet{amorisco+zavala+deboer2014}
-\item
-  \citet{battaglia+2008}
 \item
@@
 \item
+  \citet{tolstoy+2004}: detection of two populations.
+\item
+  \citet{battaglia+2008a}
+\item
   \citet{sestito+2023a}
@@
   metallicity gradient and search for metal poor stars in Scl.
-\end{itemize}
-
-Future ideas:
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{evslin2016}: measuring ellipticities of halos w TMT.
 \end{itemize}
@@
 
-\section{Are Sculptor and Ursa Minor
-typical?}\label{are-sculptor-and-ursa-minor-typical}
+\section{Should we expect exponential
+profiles?}\label{should-we-expect-exponential-profiles}
 
-\subsection{The formation of Exponential
-profiles}\label{the-formation-of-exponential-profiles}
-
-As mentioned earlier, many dwarf galaxies appear to have exponential
-stellar density profiles.
+As discussed in Section~\ref{sec:exponential_profiles}, the exponential
+profile has long been used for dwarf galaxies.
 
@@
 collapse.
-
-Given that dwarf galaxies were noticed to be exponential like above,
-\citet{faber+lin1983} proposed that dwarf spheroidals formed from disky
-galaxies maintaining the typical exponential density profile
 
@@
 
-In summary, while there is some theoretical reasons why exponential
-spheroidal stellar profiles may form, the emperical sucess of an
-exponential law still requires explanations. Deviations from this
-exponential thus test wether this exponential law is indeed
-near-universal or if other ingredients in the formation and history of
-dwarf galaxies may be responsible for changes from the empirical
-expectation.
-
-\subsection{Alternatives to exponential
-profiles}\label{alternatives-to-exponential-profiles}
-
-\begin{itemize}
-\tightlist
-\item
 elmegreen+hunter2006: Discuss the formation of double exponentials in
@@
 the critical SF threshold becomes less likely.
-\end{itemize}
 
@@
 galaxies}\label{understanding-the-extended-density-profiles-of-dwarf-galaxies}
-
-Are dwarf galaxies indeed expected to be one-component exponential-like
-density profiles? The suggestion of exponential density profiles dates
-back to \citet{faber+lin1983}. Expand\ldots{}
 
@@
   \citet{cicuendez+battaglia2018}
+\item
+  \citet{delpino+aparicio+hidalgo2015}: Fornax
+\item
+  \citet{delpino+2017}, \citet{delpino+2017a}
 \end{itemize}
 
-\subsection{Multi-epoch star formation and
-feedback}\label{multi-epoch-star-formation-and-feedback}
-
-The star formation history of dwarf galaxies is typically thought to be
-\emph{bursty}: a series of discrete episodes of intense star formation
-separated by periods of quiescence.
-
-For example, in the simulations of \citet{wheeler+2019}, etc., dwarf
-galaxies exhibit strongly bursty SFHs.
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{maxwell+2012}
-\item
-  \citet{wright+2019}
-\item
-  \citet{azartash-namin+2024}
-\end{itemize}
-
-\subsubsection{Multi-component density
-profiles}\label{multi-component-density-profiles}
+\subsection{Multi-component density profiles: clues to the
+past?}\label{multi-component-density-profiles-clues-to-the-past}
 
@@
 \begin{itemize}
-\tightlist
 \item
-  \citet{benitez-llambay+2016}
+  \citet{benitez-llambay+2016}, \citet{benitez-llambay+2015}
 \item
@@
   migration =\textgreater{} population gradients.
-\item
 \end{itemize}
@@
 \item
-  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
-  Carina)
+  \citet{battaglia+2012}, \citet{fabrizio+2016}, \citet{kordopatis+2016}
+  (multi-component in Carina)
 \item
   \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
+\item
+  \citet{battaglia+2011} sextans
+\item
+  \citet{ho+2012}: And II
 \end{itemize}
+
+\subsection{Multi-epoch star
+formation}\label{multi-epoch-star-formation}
+
+The star formation history of dwarf galaxies is typically thought to be
+\emph{bursty}: a series of discrete episodes of intense star formation
+separated by periods of quiescence.
+
+For example, in the simulations of \citet{wheeler+2019}, etc., dwarf
+galaxies exhibit strongly bursty SFHs.
+
+\begin{itemize}
+\item
+  \citet{maxwell+2012}
+\item
+  \citet{wright+2019}
+\item
+  \citet{azartash-namin+2024}
+\item
+  \citet{dong+lin+murray2003}
+\end{itemize}
+
+\subsection{Dynamical heating of old
+stars}\label{dynamical-heating-of-old-stars}
+
+For a variety of reasons, older stars in a galaxy may be preferentially
+``hotter'' (i.e., higher typical random velocities) than younger stars.
+
+In the case of dwarf galaxies, a few mechanisms have been proposed which
+may heat the stellar components: heating by strong feedback, heating by
+dark sub-subhalos, and heating
+
+Because dwarf galaxies have extended star formation histories, like the
+stellar-feedback scattering method,
+
+One compl
+
+\emph{cluster dynamics} Could processess which happen in globular
+clusters also occur in dsarf galaxies, such as mass segragation and the
+evaporation of the smallest mass stars? This seems unlikely at first
+glance because dwarf galaxies contain such
+
+\citet{baumgardt+2022}.
 
@@
 accretion}\label{past-mergers-and-accretion}
+
+When galaxies merge, traces of the effects of the merger can remain for
+long afterwards. For dwarf galaxies, mergers may appear as multiple
+populations which may have different sizes, chemistry, and kinematics.
+
+A number of studies has investigated mergers in dwarf galaxies, showing
+that this is a feasible scenario for halo formation
 
@@
   rotation from merger: \citet{cardona-barrero+2021}
+\item
+  \citet{starkenburg+helmi+sales2016}
+\item
+  \citet{fouquet+2017}: And II via merger
+\item
+  \citet{genina+2019}: \apostle{} galaxies: Mergers with gas can heat
+  old population and form new population subsequently. Additionally,
+  ram-pressure stripping and interactions with cosmic filimants can
+  remove outer gas and trigger inner star formation.
 \end{itemize}
@@
 ``preprocessed'' by another dwarf/LMC like galaxy earlier on and this
-has resulted in their unique properties
+has resulted in their unique properties.
+
+This could involve a strong tidal interaction very early on in the dwarf
+galaxy.
 
@@
 \end{itemize}
+
+\section{Summary and outlook}\label{summary-and-outlook}
+
+In this thesis, we have investigated tidal effects on the dwarf galaxies
+Sculptor and Ursa Minor.
+
+We first showed, using data from J+24, that each galaxy contains a
+substantial, extended population when compared to an exponential density
+profile. We show that this population is reasonably independent of the
+details of the data selection or methodology, implying that this
+``density excess'' is likely a real feature of each galaxy. Sculptor and
+Ursa Minor are more extended in the outer regions than other nearby
+dwarf classical, given current observations.
+
+We then investigated if tides were a permissable explanation. By
+modelling each galaxy based on cosmological initial conditions, we
+showed that tides do not strongly affect either galaxy. However, the
+presence of an LMC in the MW possibly complicates our conclusions, but
+we find that the effect of the LMC means Scl could be on first infal and
+has experienced a significantlly weaker tidal history, and Ursa Minor
+may only have a more extended orbit. In all of our investigation, we
+show that tides are very unlikely to affect the currently-observed
+density profiles. A stream is possibly detectable, depending on the
+initial stellar extent of each galaxy and how much deeper future
+observations are able to reach.
+
+As a result, we conclude that the observed extended density of each
+galaxy is inconsistent with a tidal origin, and requires a different
+explanation. In this (final) chapter, we have summarized possible other
+explanations in the literature, including the effects of mergrs,
+episodic star formation histories, stellar halos,
+
+While we have discussed a number of different avenues through which
+stellar halos form, these need not be exclusive.
+
+As an example, consider +frenk+carlos+++: when two gas-rich dwarf
+galaxies collide, the merger can introduce new star formation, possibly
+forming a new, distint population of stars. The formation of stars may
+drive a burst which heats up old stars. The merger may scatter old stars
+into a stellar halo. And the merger may distribute the two populations
+of the old galaxies into separate components. As a result, many
+different processes can contribute to the formation of chemically and
+dynamically distinct populations.
+
+Is there hope for determining the nature of a stellar halo? Perhaps the
+most promising tool will be chemistry. In particular, if we reach a
+point where we can chemically distinguish the populations of different
+galaxies, then chemistry provides an opportunity to test the merger
+history of galaxies.
+
+We propose a categorization of the (\LCDM{}) scenarios above
+
+\begin{itemize}
+\tightlist
+\item
+  Gas-poor mergers or accretion: more than one \emph{chemodynamically
+  independent} population (e.g.~GSE in the MW halo)
+\item
+  Gas-rich merger: more than one \emph{chemodynamically independent}
+  population plus an additional, younger population with possibly shared
+  heritage.
+\item
+  Continuous dynamical heating (e.g.~subsubhalos) : chemically
+  contiguous population with an old to young population gradient.
+\item
+  Strong stellar feedback: Associated bursty star formation history in
+  chemical abundance space, but reasonably chemically-contiguous
+  population.
+\end{itemize}
+
+In addition, the first two scenarios could show significant variation
+among dwarfs (especially for (near) major-mergers), whereas the last two
+should be relatively similar among similar dwarfs.
+
+Measure halo ellipticity via TMT, \citet{evslin2016}

### 2025-09-24 thesis/introduction.tex
half-done discussion

@@
 the early universe, consisting of many of the oldest and most metal poor
-stars known. Understanding the properties of dwarf galaxies thus has
-implications across astronomy, from cosmological structure formation on
-small scales to the formation of metal-poor stellar populations.
+stars known \citep{simon2019}. Understanding the properties of dwarf
+galaxies thus has implications across astronomy, from cosmological
+structure formation on small scales to the formation of metal-poor
+stellar populations.
 
@@
   were likely always known to humans at southern latitudes.}, with no
-known analogues, already presented such an enigma. Shapley presented
+known analogues, already presented such an enigma. H. Shapley presented
 these dwarfs as a new type of \emph{stellar system} resembling the
 Magellanic Clouds and globular clusters but did not attempt to speculate
-on their exact nature. While dwarf galaxies were quickly understood to
-be galaxies based on the inferred luminosities and sizes, their nature
+on their exact nature. While dwarf galaxies were soon understood to be
+galaxies based on the inferred luminosities and sizes, their nature
 remained unclear for decades
@@
 The earliest spectroscopic work hinted that dwarf galaxies may contain
-substantial amounts of dark matter. From early determinations of the
-velocity dispersion for the Sculptor (Scl) and Ursa Minor (UMi) dwarf
-spheroidal (dSph) galaxies, inferred mass-to-light ratios were at least
-10 times larger than the values of globular clusters {[}GCs; e.g.,
-\citet{aaronson1983}; \citet{aaronson+olszewski1987}{]}. While rather
-uncertain initially, these values were later corroborated with larger
-and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the time,
-several theories were proposed to explain these unusually high
+substantial amounts of dark matter. From early velocity dispersion
+measurements for dwarf spheroidal (dSph) galaxies, inferred
+mass-to-light ratios were at least 10 times larger than for globular
+clusters \citep[GCs, e.g.,][]{aaronson1983, aaronson+olszewski1987}.
+While uncertain initially, these values were later corroborated with
+larger and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the
+time, several theories were proposed to explain these unusually high
 mass-to-light ratios. Examples include: ongoing tidal disruption
@@
 Today, a common definition for a dwarf galaxy is a gravitationally bound
-stellar system with dark matter \citep[or, more specifically, not
-consistent with Newtonian dynamics of visible matter
-alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
+stellar system with dark matter.\footnote{Or, more generally, systems
+  inconsistent with Newtonian dynamics of visible matter alone,
+  \citet{willman+strader2012}} In contrast, star clusters (like GCs)
 have no clear evidence for dark matter. The boundary between these two
-classes becomes uncertain for faint, compact stellar associations, known
-as ``ambiguous'' systems.
+classes blurs for faint, compact stellar associations, known as
+``ambiguous'' systems.
 
-Dwarf galaxies span a large range of physical sizes, luminosities, and
+Dwarf galaxies span a large range of sizes, luminosities, and
 morphologies. Broadly, there are three classes of dwarf galaxies based
-on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
-\textbf{bright dwarfs} with magnitudes \(-14 \gtrsim M_V \gtrsim  -18\)
-or stellar masses
+on luminosity. Local \textbf{bright dwarfs} with magnitudes
+\(-14 \gtrsim M_V \gtrsim  -18\) or stellar masses
 \(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
@@
 stars are in a rotationally-supported thin disk (seen nearly face-on)
-with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
+with a prominent bar. \textbf{Classical dwarf spheroidals}\footnote{While
+  formally the dwarf galaxy names we discuss contain ``dwarf
+  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
+  brevity. Additionally, the 12 classical dwarf satellites of our Galaxy
+  are (in order of decreasing luminosity) Sagittarius, Fornax, Leo I,
+  Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
+  I, Sextans I, and Crater II. Antlia II, Crater II, and Canes Venatici
+  I are the only post digital sky surveys additions.} occupy
 intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
@@
 Milky Way satellites discovered before digital sky surveys are
-classicals. The 12 classical dwarf satellites of our Galaxy are (in
-order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
-Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
-I, and Crater II.\footnote{While formally the dwarf galaxy names we
-  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
-  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
-occupy the very faintest magnitude range (\(M_V \gtrsim -7.7\) or
-\(M_\star \lesssim 10^5\,\Mo\)). These galaxies have minuscule stellar
-masses, typically compact sizes, and very metal-poor stellar populations
-\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
-more than 15 orders in magnitude, or over 7 decades in stellar mass.
-
-Of local dwarfs, the classical systems still remain among the most
-studied galaxies with the best constrained parameters. These dwarfs also
-contain large numbers of bright (giant) stars which can be followed up
-spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
-the determination of fundamental properties such as the position, size,
-orientation, distances, proper motions, line-of-sight (LOS) velocities,
-and velocity dispersions \(\sigma_\V\), are all relatively well
-constrained today. \textbf{revisit this\ldots JFN cuts a similar
-version}
+classicals, and these systems remain among the best studied. The
+\textbf{ultra-faint}s dwarfs (UFDs) occupy the very faintest magnitude
+range (\(M_V \gtrsim -7.7\) or \(M_\star \lesssim 10^5\,\Mo\)). These
+galaxies have minuscule stellar masses, typically compact sizes, and
+very metal-poor stellar populations \citep[see review][]{simon2019}.
+Altogether, known dwarf galaxies span more than 15 orders in magnitude,
+or over 7 decades in stellar mass.
 
 Most well-studied dwarf galaxies lie near the Milky Way or in its
-vicinity, the Local Group (LG) of galaxies. TheLG is defined as the
+vicinity, the Local Group (LG) of galaxies. The LG is defined as the
 group consisting of galaxies within \(\sim 1\) Mpc from the Milky
@@
 universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
-matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
-  baryons+leptons, i.e.~any standard model massive fermion.}
+matter (DM), and 5\% regular old baryons\footnote{\emph{Baryons} here
+  means baryons+leptons, i.e.~any standard model massive fermion.}
 \citep{planckcollaboration+2020}. While the composition of dark matter
@@
 collapse. Dark matter, only influenced by gravity instead, freely
-collapsed into the first structures. For mass perturbations both smaller
-than the horizon and overdense enough to collapse, these overdensities
-of dark matter become self-gravitating, known as \emph{halos}. After
-recombination, where electrons combined with atomic nuclei to form
-atoms, baryons decoupled from radiation and fell into the dark matter
-halos. The densest pockets of baryons later formed the first stars and
-galaxies.
+collapsed into the first structures. Mass perturbations sufficiently
+small and overdense become self-gravitating structures, known as
+\emph{halos}. After recombination, where electrons combined with atomic
+nuclei to form atoms, baryons decoupled from radiation and fell into the
+dark matter halos. The densest pockets of baryons later formed the first
+stars and galaxies.
 
 Dark matter halos, and their associated galaxies, rarely evolve in
-isolation. Instead, structure formation is \emph{hierarchical}. Small
-dark matter halos collapse first and hierarchically merge into
+isolation. Instead, \LCDM{} structure formation is \emph{hierarchical}.
+Small dark matter halos collapse first and hierarchically merge into
 progressively larger halos
@@
 gradually with radius, and at large radii the density falls off like
-\(\rho \sim 1/r^3\).
+\(\rho \sim 1/r^3\). The blue solid curve in Fig.~\ref{fig:nfw_density}
+shows an example NFW halo.
 
@@
 
-The blue solid curve in Fig.~\ref{fig:nfw_density} shows an example NFW
-halo for a galaxy with \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
-\(M_{200} = 1\times10^10\,\Mo\) and \(c=12.5\), the cosmological mean
-for a Fornax stellar mass galaxy.
-
 The two parameters of an NFW profile are not independent. Lower-mass
@@
 mass-concentration relation, but their effect is typically expected to
-be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
+be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}. The
+mass-concentration relation informs our assumed structure of Scl and UMi
+later.
 
@@
 \centering
-\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
 \caption[Example density profiles]{Density profiles in log density
@@
 dark matter is more extended and massive than the star across the entire
-galaxy.}\label{fig:nfw_density}
+galaxy. The halo has \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
+\(M_{200} = 1\times10^{10}\,\Mo\) and \(c=12.5\), the cosmological mean
+for a Fornax stellar mass galaxy. The galaxy has a stellar mass
+\(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
+\(0.65\,\kpc\)
+\citep{munoz+2018, woo+courteau+dekel2008}.}\label{fig:nfw_density}
 \end{figure}
@@
 shows an example exponential stellar component for the Fornax dwarf
-galaxy with the expected underlying dark matter halo. This galaxy has a
-stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
-radius \(0.65\,\kpc\) \citep[ with \citet{woo+courteau+dekel2008}
-mass-to-light ratio]{munoz+2018}, and a dark matter halo with
-\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\). Even when the stars are
-densest (the inner regions), the dark matter remains nearly an order of
-magnitude higher in density. Stars make a small contribution to the
-gravitational structure of dwarf galaxies--indeed, stars are reasonably
-approximated as tracer particles of the underlying dark matter halo. In
-addition, the stellar component is confined to the central regions of
-the dark matter halo.
+galaxy with the expected underlying dark matter halo. Even when the
+stars are densest in the inner regions, the dark matter remains nearly
+an order of magnitude higher in density. Stars make a small contribution
+to the gravitational structure of dwarf galaxies---indeed, stars are
+reasonably approximated as tracer particles of the underlying dark
+matter halo. In addition, the stellar component is confined to the
+central regions of the dark matter halo.
 
 Several factors affect the SMHM trend, including environment, assembly
-history, tidal effects, and the details of galaxy formation. In
-particular, effects like ram-pressure stripping (removal of gas in the
+history, tidal effects, and the details of galaxy formation. For
+example, effects like ram-pressure stripping (removal of gas in the
 dwarf galaxy due to pressure from the host's circumgalactic medium) and
@@
 \citep{kim+2024}. Finally, tides influence both the dark matter and
-stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
-Consequently, tides may reduce the halo mass more than the stellar mass,
-adding additional scatter to the SMHM trend, particularly for low-mass
-satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
-tides on Local Group dwarf galaxies is essential for determining where
-and how these galaxies formed in a cosmological context.
+stellar mass but in different amounts, adding additional scatter to the
+SMHM trend for satellites \citep[e.g.,][]{PNM2008, fattahi+2018}.
+Understanding the effects of tides on Local Group dwarf galaxies aids in
+revealing where and how these galaxies formed in a cosmological context.
 
-\subsection{Open questions and ongoing challenges in dwarf
-galaxies}\label{open-questions-and-ongoing-challenges-in-dwarf-galaxies}
-
-Advances in observational techniques have accelerated progress in our
-understanding of dwarf galaxies and introduced new questions. Deep
-digital photometric surveys have more than quadrupled the number of
-known Milky Way dwarf galaxies since 2005 \citep{simon2019}. Upcoming
-surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
-Time \citep[LSST,][]{ivezic+2019}, will likely identify even fainter
-dwarf galaxies. In addition, large aperture multi-object spectrographs
-have revealed the detailed and complex inner chemodynamical structure of
-dwarf galaxies \citep[e.g.,][]{battaglia+2006, fabrizio+2016}. Beyond
-precise structural and kinematic properties of dwarf galaxies, modern
-observations allow for the separation of multiple stellar populations,
-detailed constraints on the dark matter density profiles, and hints of
-tidal disruption or stellar halos. \textbf{does this ¶ (or section)
-belong here?}
+\subsection{Challenges and questions concerning dwarf
+galaxies}\label{challenges-and-questions-concerning-dwarf-galaxies}
 
@@
 
-However, these tensions have eased as a result of improved understanding
-of baryonic physics. For example, recent hydrodynamic simulations in
-particular have shown that strong feedback can produce dark matter cores
+However, some of these tensions have eased as a result of improved
+understanding of baryonic physics. For example, recent hydrodynamic
+simulations in particular have shown that strong feedback can produce
+dark matter cores
 \citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
-Altogether, the numerous past challenges for \LCDM{} in the dwarf galaxy
-regime illustrate the opportunity for dwarf galaxies to test the
-understanding of galaxy formation and dark matter physics.
+Several open questions remain, concerning e.g.~the plane of satellites,
+and the details of the sizes and rotation curves of dwarf galaxies, the
+existence and nature of stellar halos and accretion in dwarf galaxies
+\citep[e.g.,][]{sales+2022}. Altogether, the numerous past and ongoing
+challenges for \LCDM{} in the dwarf galaxy regime illustrate the
+opportunity for dwarf galaxies to test the understanding of galaxy
+formation and dark matter physics.
 
@@
 parallax motion with different phases. By imaging two regions separated
-by 106.5 degrees on the same focal plane, \emph{Gaia} measures tiny
-changes in relative positions of stars across small and large angles.
-Combining measurements from multiple epochs across several years, an
-absolute all-sky reference frame is derived from which parallax and
-proper motions are derived. In addition to astrometry, \emph{Gaia}
-measures photometry in the wide \emph{G} band (330-1050nm) and colours
-from the blue photometer (BP, 330-680 nm) and red photometer (RP,
-640-1050 nm). \emph{Gaia} additionally provides low resolution BP-RP
-spectra and radial velocity measurements of bright stars \citep[of
-magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
-work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
+by 106.5 degrees on the same focal plane, \emph{Gaia} measures changes
+in relative positions of stars across small and large angles. Combining
+measurements from multiple epochs across several years, an absolute
+all-sky reference frame is derived from which parallax and proper
+motions are derived. In addition to astrometry, \emph{Gaia} measures
+photometry in the wide \emph{G} band (330--1050nm) and colours from the
+blue photometer (BP, 330--680 nm) and red photometer (RP, 640--1050 nm).
+\emph{Gaia} additionally provides low resolution BP-RP spectra and
+radial velocity measurements of bright stars \citep[of magnitudes
+\(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our work,
+\emph{Gaia}'s most relevant measurements are \(G\) magnitude,
 \(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
@@
 the Milky Way's spiral arms and the bar in the solar neighbourhood
-\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
-Way halo, \emph{Gaia} has helped find and constrain numerous stellar
-streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
-Altogether, \emph{Gaia} has revealed the hierarchical formation and
-complex, evolving structure of our own Galaxy.
+\citep[ and references therein]{hunt+vasiliev2025}. In the Milky Way
+halo, \emph{Gaia} has helped find and constrain numerous stellar streams
+\citep{ibata+malhan+martin2019, bonaca+price-whelan2025}. Altogether,
+\emph{Gaia} has revealed the hierarchical formation and complex,
+evolving structure of our own Galaxy.
 
 For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
-facilitated stellar membership determination. Before \emph{Gaia}, few
-galaxies had precisely measured proper motions, more recently from
-Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
-\emph{Gaia} allowed the first systematic and precise determinations of
-Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
-the proper motion uncertainty of a typical dwarf member star is often
-large, by combining the proper motions of 100s or 1000s of stars from
-\emph{Gaia}, precise average proper motion measurements can be
-determined, sometimes only limited by \emph{Gaia}'s systematic error
-floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
-new era for MW satellite dynamical studies, where we can derive precise
-orbits for any satellite, assuming a given MW potential. In addition,
+facilitated robust stellar membership determination. Before \emph{Gaia},
+few galaxies had precisely measured proper motions \citep[e.g.~using
+Hubble Space Telescope,][]{sohn+2017}. \emph{Gaia} allowed the first
+systematic and precise determinations of Milky Way satellite proper
+motions \citep{pace+li2019, MV2020a}. While the proper motion
+uncertainty of a typical dwarf member star is often large, by combining
+the proper motions of 100s or 1000s of stars from \emph{Gaia}, precise
+average proper motion measurements can be determined, sometimes only
+limited by \emph{Gaia}'s systematic error floor
+\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new era
+for MW satellite dynamical studies, where we can derive precise orbits
+for any satellite, assuming a given MW potential. In addition,
 \emph{Gaia} helps establish membership by separating out contaminating
 MW foreground stars. By measuring parallaxes and/or proper motions, many
-background and foreground stars can be classified as non-members
+more background and foreground stars can be classified as non-members
 \citep[e.g.,][]{battaglia+2022, jensen+2024}.
@@
 the radial structure of a galaxy. At its most basic, light profiles
-synthesize properties such as the shape, location, size, and orientation
-of a dwarf galaxy. In addition, the details of a stellar density profile
-can help interpret a galaxy's assembly and dynamical history
-\citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. We note that
-for resolved galaxies, these profiles are in terms of stellar counts
-instead of surface brightness.
+synthesize properties such as the shape, size, and orientation of a
+dwarf galaxy. In addition, the details of a stellar density profile can
+help interpret a galaxy's assembly and dynamical history
+\citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. Note that for
+resolved galaxies, these profiles are expressed in stellar count
+densities instead of surface brightness.
 
@@
 
-The \citet{king1962} profile, also an empirical fit to globular
-clusters, is also used to describe dwarf galaxies, especially in older
-literature. Using three parameters, a core radius \(R_c\), a truncation
-radius \(R_t\), and a characteristic density, \(\Sigma_0\), the King
-profile may be written as \begin{equation}{
+The \citet{king1962} profile, also a fit to globular clusters, is also
+used to describe dwarf galaxies, more so in older literature. Using
+three parameters, a core radius \(R_c\), a truncation radius \(R_t\),
+and a characteristic density, \(\Sigma_0\), the King profile may be
+written as \begin{equation}{
 \Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
@@
 while \(n=4\) recovers \citet{devaucouleurs1948}'s profile for
-elliptical galaxies. Although a Sérsic profile is not always used for
-dwarf galaxies, \citet{munoz+2018} advocate for the Sérsic profile since
-the added flexibility allows more profiles to be fit with a single law.
+elliptical galaxies. Although a Sérsic profile is less commonly applied
+to dwarf galaxies, \citet{munoz+2018} advocate for the Sérsic profile
+since the added flexibility allows more profiles to be fit with a single
+law.
 
@@
 \citep{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018}.
-As a result, it has become conventional to assume an exponential density
-profile to describe dwarf galaxies in theoretical or observational
-modelling
-\citep[e.g.,][]{martin+2016, MV2020a, battaglia+2022, kowalczyk+2013}.
+More recently, \citet{moskowitz+walker2020} fit instead generalized
+Plummer profiles, but most of their fits would be consistent with a
+single-component exponential. As a result, it has become conventional to
+assume an exponential density profile to describe dwarf galaxies in
+theoretical or observational modelling
+\citep[e.g.,][]{kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
 
-Outside Local Group dwarf spheroidals, exponential profiles are fairly
-common, but sometimes with modifications. For example, many
-extragalactic dwarf elliptical, blue compact, and irregular dwarf
-galaxies are better described with an exponential profile to which is
-added a central cusp or nuclear region
+Outside the MW system, exponential profiles are common, but sometimes
+with modifications. For example, many extragalactic dwarf elliptical,
+blue compact, and irregular dwarf galaxies are better described with an
+exponential profile to which is added a central cusp or nuclear region
 \citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
 studies find an inner density decrement compared to exponentials
-\citep[e.g.,][]{caldwell+1992, makarov+2012}.
-
-However, exponential profiles do not fit every dwarf galaxy. For example
-\citet{aparicio+1997}, and \citet[for Coma cluster dwarf
-ellipticals]{graham+guzman2003}, some people have noted deviations from
-this simple empirical rule. Additionally, \citet{hunter+elmegreen2006};
-\citet{herrmann+hunter+elmegreen2013};
-\citet{herrmann+hunter+elmegreen2016}; and \citet{lee+2018} noted that
-at least in a photometric sample of more distant disky, irregular, and
-blue compact dwarfs, many or most dwarfs are best fit with two nested
-exponential profiles. It is unclear how these conclusions apply to dwarf
-spheroidals.
-
-More recently, \citet{moskowitz+walker2020}, have used a Plummer profile
-and modified Plummer profiles with steeper outer slopes
-(\(\Gamma = d\log \Sigma / d \log R = -8\) rather than \(\Gamma = -4\)).
-For most galaxies, they do not have convincing statistical evidence to
-prefer one model over another, except for a few galaxies which are
-better described by a steep outer slope.
+\citep[e.g.,][]{caldwell+1992, makarov+2012}, or that dwarfs are better
+fit by two nested exponentials
+\citep[e.g.,][]{aparicio+1997, graham+guzman2003, hunter+elmegreen2006, lee+2018}.
+It is unclear how these conclusions apply to dwarf spheroidals.
 
@@
 
-Sculptor and Ursa Minor are not well-described by an exponential
-profile. The left panel of Fig.~\ref{fig:scl_umi_vs_fornax} shows the
-density profiles of Sculptor, Ursa Minor, and Fornax (see
+Sculptor and Ursa Minor are poorly described by an exponential profile.
+The left panel of Fig.~\ref{fig:scl_umi_vs_fornax} shows the density
+profiles of Sculptor, Ursa Minor, and Fornax (see
 Section~\ref{sec:observations} for details on how these profiles are
 measured). Compared to Fornax, both Sculptor and Ursa Minor show an
-excess of stars outside \(\log R/R_h\approx 0.4\) and which exceeds 100
+excess of stars outside \(\log R/R_h\approx 0.4\), which exceeds 100
 times the density of the exponential fit at large radii.
 
-A goal of this work is to determine, in the context of \LCDM{}, if the
-effects of the Milky Way (or other satellites) are indeed responsible
-for the extended outer light profiles of Sculptor and Ursa Minor. If
-tides cannot explain these features, these features may instead be due
-to an extended stellar ``halo'' or second component of the
-galaxy---suggestive of a complex star formation or assembly history.
+A goal of this work is to determine if tidal effects are indeed
+responsible for the extended outer light profiles of Sculptor and Ursa
+Minor. If tides cannot explain these features, these features may
+instead be due to an extended stellar ``halo'' or second component of
+the galaxy---suggestive of a complex star formation or assembly history.
 
@@
 Idealized simulations make numerous simplifications, neglecting mergers,
-cosmological context, mass assembly, and often baryonic physics {[}e.g.,
-\citet{hayashi+2003}; bullock+johnston2005; \citet{klimentowski+2009};
-\citet{ogiya+2019}{]}. We shall use idealized simulations here to assess
-tidal effects after infall.
+cosmological context, mass assembly, and often baryonic physics
+\citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
+We shall use idealized simulations here to assess tidal effects after
+infall.
 
-Some early simulation work on tidal mass loss of dwarf galaxies
-originate from \citet{oh+lin+aarseth1995}; \citet{piatek+pryor1995};
-\citet{moore+davis1994}; \citet{johnston+spergel+hernquist1995}.
-Already, these works described the relative evolution of stellar and
-dark matter mass loss, that most mass loss happens near pericentre, the
-formation of symmetric tidal streams, and the minor changes to the
-central structure.
-
-Later, \citet{read+2006} showed that tidal tails would likely cause
-increasing velocity dispersion profiles, and interpreted the lack of
-similar observational features as indicating tides likely did not affect
-the stellar component of dwarf galaxies \citep[see
-also][]{klimentowski+2009a}.
+Conclusions from early work on tidal evolution remain mostly relevant
+today. A galaxy undergoing tidal effects loses most mass near the
+Lagrange points at pericentre, forms symmetric tidal streams with a
+velocity gradient, and retains its central structure
+\citep{oh+lin+aarseth1995, piatek+pryor1995, moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
 
 CDM halos also are found to be highly resilient to full disruption
-{[}\citet{EP2020}; {]}. In addition, NFW halos evolve along ``tidal
-tracks'' {[}e.g, \citet{PNM2008}, \citet{EN2021}; {]}. On the other
+\citep{EP2020}. In addition, NFW halos evolve along ``tidal tracks''
+\citep[e.g,][]{PNM2008, green+vandenbosch2019, EN2021}. On the other
 hand, cored dark matter halos likely disrupt faster
@@
 
-\citet{hayashi+2003}, \citet{green+vandenbosch2019}
-
-Recent, work by
 \citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
@@
 of particles below a certain energy \citep[see
-also][]{choi+weinberg+katz2009}. \citet{strucker+2023} expanded on this
-idea, creating an analytic model for adiabatic tidal mass loss,
-confirming that an NFW halo is likely unable to be disrupted and
-explaining the evolution of NFW halos along similar tidal tracks.
+also][]{choi+weinberg+katz2009}. \citet{stucker+2023} generalized this
+idea, creating a model for adiabatic tidal mass loss in an isotropic
+tidal field, confirming that an NFW halo is likely unable to be
+disrupted and explaining the evolution of NFW halos along similar tidal
+tracks.
 
@@
 framework to Sculptor and Ursa Minor.
-
-\textbf{todo: finish this synthesis}
 
@@
 show the results 170 Myr after the first pericentre. See
-Section~\ref{sec:methods} for a more detailed description of our general
-simulation setup.
+Section~\ref{sec:methods} for a more detailed description of our
+numerical setup.
 
@@
 in Chapter \ref{sec:discussion}. Finally, Chapter \ref{sec:summary}
-summarizes this work and discuss future directions for similar work in
-the field of dwarf galaxies.
+summarizes this thesis and discusses future directions for similar work.

### 2025-09-24 thesis/main.tex
julio sep 1 edits to §1, 2
thesis updates
half-done discussion

@@
 \title{
-    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?
+    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?
 }
@@
 
-\chapter{The Light Profiles of the Classical Dwarfs}\label{sec:observations}
+\chapter{The Light Profiles of the Classical Dwarf Spheroidals}\label{sec:observations}
 \input{data}
@@
 
-\chapter{The Effects of Tides}\label{sec:results}
+\chapter{Galactic tidal effects on Sculptor and Ursa Minor}\label{sec:results}
 \input{results}

### 2025-09-24 thesis/methods.tex
half-done discussion

@@
 question are heavily dark matter dominated, we shall use a carefully
-selected sample of dark matter particles to mimic and track the
-evolution of an embedded tracer stellar component. In this Chapter, I
-describe the Galactic potential used, the orbital estimation method, the
-initial conditions setup, and the N-body method used.
+selected sample of dark matter particles to emulate the evolution of an
+embedded tracer stellar component. In this Chapter, we describe our
+choice of Galactic potential, orbital estimation, initial conditions
+setup, and N-body methods.
 
-\section{Orbital estimation}\label{orbital-estimation}
+\section{Orbital estimation}\label{sec:orbital_estimation}
 
@@
 coordinates, we use the Astropy v4 Galactocentric frame
-\citep{astropycollaboration+2022}. This frame assumes the Galactic
-centre is at position \(\alpha = {\rm 17h\,45m\,37.224s}\),
-\(\delta = -28^\circ\,56'\,10.23''\) with proper motions
-\(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
-\(\mu_\delta=-5.547\pm0.026 \masyr\) \citep[from the appendix and Table
-2 of][]{reid+brunthaler2004}. The Galactic centre is at a distance from
-the Sun of \(8.122\pm0.033\,\)kpc with a radial velocity =
-\(11 + 1.9 \pm 3\,\kms\) \citep{gravitycollaboration+2018}. The Sun is
-assumed to be \(20.8\pm0.3\,\)pc above the disk
-\citep{bennett+bovy2019}. Using the procedure outlined in
-\citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
-rest frame is then in Cartesian coordinates
-\(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
-uncertainties in this reference frame are typically smaller than the
-uncertainties on a dwarf galaxy's position and velocity.
+\citep{astropycollaboration+2022}. In this frame, the solar position is
+\([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
+\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
+velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
+km/s
+\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
+The uncertainties in this reference frame are typically smaller than the
+uncertainties on a dwarf galaxy's position and (tangental) velocity.
 
@@
 The halo is an NFW dark matter halo (Eq.~\ref{eq:nfw}) with
-\(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\).
+\(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\), or
+\(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/v_circ_potential.png}
 \caption[Circular velocity of potential]{Circular velocity profile of
@@
 \citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
-\citet{hernquist1990} bulge (blue dotted line). \textbf{make radius
-lowercase}}\label{fig:v_circ_potential}
+\citet{hernquist1990} bulge (blue dotted
+line).}\label{fig:v_circ_potential}
 \end{figure}
@@
 significantly smaller pericentre than our selected orbit, which we refer
-to as the \smallperi{} orbit.
-
-We take the first apocentre after a look-back time of 10 Gyr, or at 9.43
-Gyr, as the initial conditions for our model of Sculptor, noted in
-Table~\ref{tbl:scl_orbits}.
+to as the \smallperi{} orbit. We take the first apocentre after a
+look-back time of 10 Gyr, or at \(\sim9.1\,\Gyr\), as the initial
+conditions for our model of Sculptor, noted in
+Table~\ref{tbl:orbit_ics}.
 
@@
 \centering
-\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor.
-The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}.
+\caption[Orbit initial conditions]{The orbital initial conditions for models presented in this work.
+The observables represent the medians from orbital integration used to derived the orbits.
+Instead, the initial position and velocity represents the initialization of the actual N-body model.
 The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+We describe the \texttt{LMC-flyby} orbit in Section \ref{sec:scl_lmc}.
 }
-\label{tbl:scl_orbits}
-\begin{tabular}{lll}
+\label{tbl:orbit_ics}
+\begin{tabular}{llll}
 \toprule
-Property & mean & \smallperi{}\\
+Property & Scl: \smallperi{} & Scl: \verb|LMC-flyby| & Umi: \smallperi{}\\
 \midrule
-distance / kpc & 83.2 & 82.6\\
-$\pmra / \masyr$ & 0.099 & 0.134\\
-$\pmdec / \masyr$ & -0.160 & -0.198\\
-LOS velocity / $\kms$ & 111.2 & 111.2\\
-$t_i / \Gyr$ & -8.74 & -9.43\\
-${x}_{i} / \kpc$ & 16.13 & -2.49\\
-${y}_{i} / \kpc$ & 92.47 & -42.78\\
-${z}_{i} / \kpc$ & 39.63 & 86.10\\
-$\V_{x\,i} / \kms$ & -2.37, & -20.56\\
-$\V_{y\,i} / \kms$ & -54.70 & -114.83\\
-$\V_{z\,i} / \kms$ & 128.96 & -57.29\\
-pericentre / kpc & 53 & 43\\
-apocentre / kpc & 102 & 96\\
-$t_{\rm last\ peri} / {\rm Gyr}$ & -0.45 & -0.46\\
-Number of pericentres & 6 & 6\\
+distance / kpc & 82.6 & 72.9 & 64.6\\
+$\pmra / \masyr$ & 0.134 & 0.137 & -0.158\\
+$\pmdec / \masyr$ & -0.198 & -0.157 & 0.050\\
+LOS velocity / $\kms$ & 111.2 & 111.2 & -245.75\\
+$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
+${x}_{i} / \kpc$ & -2.49 & 4.30 & 17.40\\
+${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
+${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
+$\V_{x\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
+$\V_{y\,i} / \kms$ & -114.83 & -56.83 & 48.62,\\
+$\V_{z\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
 \bottomrule
@@
 approximately the \(3\sigma\) smallest pericentre, the Ursa Minor
-\(\smallperi{}\) orbit.
+\(\smallperi{}\) orbit (see Fig.~\ref{fig:umi_orbits} and
+Table~\ref{tbl:orbit_ics}).
 
@@
 not carry out this correction for Sculptor.
-
-\begin{table*}[t]
-\centering
-\caption[Ursa Minor Selected Orbits]{Properties of selected orbits for Ursa Minor.
-The \smallperi{} orbit is the initial point orbit and the \smallperi{}.5 is the initial orbit for the N-body simulation.
-}
-\label{tbl:umi_orbits}
-\begin{tabular}{llll}
-\toprule
-Property & Mean & \smallperi{}.1 & \smallperi{}.5\\
-\midrule
-distance / kpc & $70.1$ & 64.6 & \\
-$\pmra / \masyr$ & $-0.124 \pm 0.17$ & -0.158 & \\
-$\pmdec / \masyr$ & $0.078\pm0.17$ & 0.05 & \\
-$v_{\rm LOS} / \kms$ & $-245.9\pm 1$ & -245.75 & \\
-$t_i / \Gyr$ & -8.74 & -9.53 & -9.53\\
-${x}_{i} / \kpc$ & 4.88 & -16.48 & 17.40\\
-${y}_{i} / \kpc$ & -65.11 & 69.92 & 74.51\\
-${z}_{i} / \kpc$ & 50.78 & 21.05 & 21.34\\
-$\V_{x\,i} / \kms$ & -34.28 & 16.32 & 14.27\\
-$\V_{y\,i} / \kms$ & 77.11 & 39.86 & 48.62,\\
-$\V_{z\,i} / \kms$ & 101.38 & -116.99 & -114.08\\
-pericentre / kpc & 37 & 29 & 28\\
-apocentre / kpc & 83 & 75 & 72\\
-$t_{\rm last\ peri} / \Gyr$ & $-0.96$ & $-0.80$ & $-0.81$\\
-Number of pericentres & 9 & 6 & 6\\
-\bottomrule
-\end{tabular}
-\end{table*}
 
@@
 total current stellar mass of Sculptor and Ursa Minor are
-\(M_\star \sim 3.1 \times 10^6 \Mo\) and
-\(M_\star \sim 7 \times 10^5 \Mo\), respectively. Based on the stellar
-mass-\(\vmax\) relation {[}fig.~1 from \citet{fattahi+2018};{]},
-Sculptor and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\)
-and \(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
+\(M_\star \approx 3.1 \times 10^6 \Mo\) and
+\(M_\star \approx 7 \times 10^5 \Mo\), respectively. Based on the
+stellar mass-\(\vmax\) relation \citep[from][]{fattahi+2018}, Sculptor
+and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
+\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
 \(z=0\) mass-concentration relation, this constraint translates into a
 \(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
-galaxy. These choices set the velocity dispersion of a (massless)
-stellar system embedded in such a halo, depending on its half-light
-radius.
+galaxy.
 
-The situation is illustrated in Fig.~\ref{fig:scl_halos}. Here we show,
-in the NFW \(\rmax\)--\(\vmax\) plane, the velocity dispersion of
-stellar systems with half-light radius matching that of Sculptor and
-Ursa Minor (\(R_h=240\,\)pc for both galaxies). For Scl, assuming
-\(\vmax = 31\,\kms\) and \(\rmax = 6\,\kpc\) gives
-\(\sigma_v \approx = XXX\), in contrast with the observed dispersion of
-\textbf{add}. To match the observed value, we choose a somewhat smaller
-value for r\_max=?? kpc (shown by the blue circle in the left panel of
-Fig. 3.4). We have also experimented with a smaller rmax and Vmax for
-Scl (yellow square in the same panel, to account for the fact that the
-half-light radius and velocity dispersion may be affected by tidal mass
-losses.
-
-For Ursa Minor, the cosmological mean predicts \ldots. We chose a halo
-with a higher \(\vmax\) and \(\rmax\) so the initial dispersion is
-\textbf{add}. While above the observed value, we will show in the next
-section this results in a final velocity dispersion near to the observed
-values. \textbf{add}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_velocity.pdf}}
-\caption{Initial halo velocity profiles}
-\end{figure}
+The observed velocity dispersion constrains the halo mass. From the
+\citet{wolf+2010} mass estimator, the mass within the 3D half-light
+radius \(r_h\) is
+\begin{equation}\protect\phantomsection\label{eq:wolf_mass}{
+M(r_h) \approx \frac{3}{G} \sigma_\V^2\,r_h
+}\end{equation} for a velocity dispersion \(\sigma_\V\). For Scl and
+UMi, the cosmological mean parameters above would predict
+\(\sigma_{\V, i}  \approx 8.5\,\kms\) and \(8.0\,\kms\) assuming
+\(r_h=240\,\)pc. This is below the observed values of \(9.7\,\kms\) and
+\(8.7\,\kms\), and tidal evolution will further reduce the dispersion.
+To match the observed \(\sigma_\V\) at the end of the simulation, we
+choose \(\vmax = 31\,\kms\) and \(\rmax=3.2\,\kpc\) for Scl and
+\(\vmax=38\,\kms\) and \(\rmax=4\,\kpc\) for UMi.
 
@@
 changes to the initial halo do not substantially affect the tidal
-evolution for either galaxy. The observed velocity dispersion is
-directly related to the mass within a half-light radius
-\citep[e.g.,][]{wolf+2010}. As a result, halos with the same velocity
-dispersion may differ in total mass but should have similar mass within
-\(R_h\). So, the tidal effects on stars should be similar for halos with
-similar velocity dispersions. \textbf{add curve for density at
-pericentre}
-
-Our selected halos (in Table~\ref{tbl:initial_halos}) for each
-simulation run are based on the cosmological constraints and the need to
-match the present-day velocity dispersion at the end of the simulation.
-\textbf{Make figure of initial halos, mark observed half-light radius
-and velocity dispersion, pericentric MW density?}
+evolution for either galaxy. So, the tidal effects on stars should be
+similar for halos with similar velocity dispersions.
 
@@
 $\rmax$ & $6 \pm 2$ kpc & $5_{-2}^{+1}$ kpc\\
-$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $3_{-2}^{+4} \times 10^9\,\Mo$\\
-$c_{\rm NFW}$ & $13_{-3}^{+4}$ & 14?\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $0.3_{-0.2}^{+0.4} \times 10^{10}\,\Mo$\\
+$c_{\rm NFW}$ & $13_{-3}^{+4}$ & $13.5_{-3}^{+4}$\\
 \bottomrule
@@
 represent the velocity dispersion of the initial halo given the
-present-day half-light radius \citep[via the][ mass
-estimator]{wolf+2010}.}\label{fig:scl_halos}
+present-day half-light radius via
+Eq.~\ref{eq:wolf_mass}.}\label{fig:scl_halos}
 \end{figure}
-
-\begin{table*}[t]
-\centering
-\caption[Initial halos]{The initial conditions for our initial dark matter halos. }
-\label{tbl:initial_halos}
-\begin{tabular}{lllll}
-\toprule
-Halo name & $\rmax$ & $\vmax$ & $M_{200}$ & $c_{\rm NFW}$\\
-\midrule
-Scl: fiducial & 3.2 & 31 & 0.33 & 21\\
-Scl: small & 2.5 & 25 &  & \\
-UMi: fiducial & 4 & 38 & 0.62 & 21\\
-\bottomrule
-\end{tabular}
-\end{table*}
 
@@
 approximate a collisionless system (i.e., without strong
-particle-particle gravitational deflections), the force is softened
-below a ``softening length.'' Regions which are smaller than the
-softening length, or have few particles, are not well-resolved. Tree
-codes organize particles into a spatial tree, enabling grouping of the
-gravitational forces from nearby particles. A gravitational tree code
-substantially reduces the required number of force computations, as
-compared to the exact, direct-summation method.
+particle-particle gravitational deflections), the force is tapered below
+a ``softening length.'' Resolution is limited by the number of particles
+and the softening length. Tree codes organize particles into a spatial
+tree, enabling grouping of the gravitational forces from nearby
+particles. A gravitational tree code substantially reduces the required
+number of force computations, as compared to the exact, direct-summation
+method.
 
@@
 h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-}\end{equation} See Appendix Section~\ref{sec:extra_convergence} for a
+}\end{equation} See Appendix \ref{sec:extra_convergence} for a
 discussion of this choice, which is similar to the \citet{power+2003}
@@
 the number of particles, the smaller the radius that is effectively
-``resolved'' in a given simulation. For the Sculptor halo shown in this
-figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\) ), a
+``resolved'' in a given simulation. For the Sculptor-like halo shown in
+this figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)), a
 simulation with \(10^7\) particles is needed to resolve the innermost
@@
 panel are relative to NFW. The initial conditions are dotted, the
-converged radius is marked by arrows \citep[from][ eq. 13]{power+2003},
-and the softening length is marked by a vertical bar. Note that a slight
+converged radius is marked by arrows \citep[eq. 13,][]{power+2003}, and
+the softening length is marked by a vertical bar. Note that a slight
 reduction in density starting around \(r = 30\,\kpc\) is expected given
@@
 Next, we evolve the halo in the Galactic potential. We scale the relaxed
-snapshot and softening length to match the initial halo (see
-Table~\ref{tbl:initial_halos}), and shift the snapshot to the initial
-conditions inferred from the orbital analysis (see
-Tables~\ref{tbl:scl_orbits}, \ref{tbl:umi_orbits}). We then evolve the
-full N-body NFW model forward in time in the Galactic potential until
-the present time, when the halo is closest to its present-day observed
-position in the MW halo.
+snapshot and softening length to match the initial halo, and shift the
+snapshot to the initial conditions inferred from the orbital analysis
+(see Table~\ref{tbl:orbit_ics}). We then evolve the full N-body NFW
+model forward in time in the Galactic potential until the present time,
+when the halo is closest to its present-day observed position in the MW
+halo.
 
@@
 symmetry. We initially assume stars follow a projected exponential law
-(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,,\kpc\) for both
+(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,\kpc\) for both
 galaxies. The tagging method assigns a probability to each dark matter
@@
 to match the assumed stellar light profile. We briefly describe the
-procedure next. We refer interested readers to \textbf{REF (rapha?)} for
-further details. \emph{code may be found at?}
+procedure next, but refer interested readers to \citet{EP2020}.
 
 Let \(\Psi\) be the potential (normalized to vanish at infinity) and
-\({\cal E}\) the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
-know \(f({\cal E})\), the distribution function (phase-space density as
-a function of energy), then we assign a stellar weight to a given
+\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 v^2\). If we
+know the distribution function \(f({\cal E})\) (the phase-space density
+as a function of energy), then we assign a stellar weight to a given
 particle with energy \({\cal E}\) using \begin{equation}{
@@
 f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-}\end{equation} In practice the right, boundary term is zero as
-\(\Psi \to 0\) as \(r\to\infty\), and if \(\rho \propto r^{-n}\) at
-large \(r\) and \(\Psi \sim r^{-1}\) then
-\(d\rho / d\Psi \sim r^{-n+1}\) which goes to zero provided that
-\(n > 1\). We take \(\Psi\) from the underlying assumed analytic dark
-matter potential. \(\rho_\star\) can be calculated from the surface
-density, \(\Sigma_\star\), via the inverse Abel transform.
+}\end{equation} In practice the right, boundary term is zero.\footnote{If,
+  at large \(r\), \(\rho \propto r^{-n}\) with \(n>1\), so
+  \(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\).} We take
+\(\Psi\) from the underlying assumed analytic dark matter potential.
+\(\rho_\star\) can be calculated from the surface density,
+\(\Sigma_\star\), via the inverse Abel transform.
 
-We find the stellar profiles created in this manner are stable in the
-isolated systems and show excellent agreement with the assumed stellar
-density profile. \textbf{Add figure showing stellar and dark matter
-density profiles\ldots{}}
+Fig.~\ref{fig:scl_umi_initial_isolation} shows the initial circular
+velocity profiles of stars and dark matter, demonstrating that these
+galaxies are stable in isolation and the stellar component would
+contribute negligibly to the total mass (\(M \sim \V_{\rm circ}^2 R\)).
+Our cosmologically-motivated, observationally-based, and
+numerically-converged initial conditions enable us to accurately
+consider tidal effects in the next Chapter.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_conditions.pdf}}
-\caption[Initial stellar systems]{The initial dark matter and stellar
-profiles for Sculptor and Ursa Minor. The initial profiles are dotted
-and final solid.}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_velocity.pdf}}
+\caption[Initial halo velocity profiles]{The initial circular velocity
+profiles of dark matter (blue) and stars (orange) for Sculptor and Ursa
+Minor. The initial conditions are dotted and the isolation evolved
+profiles are solid. The green cross marks the present day half-light
+radius and velocity dispersion, and the black band represents the
+1-sigma mean density of the MW at pericentre across orbits. Initial
+conditions are stable in isolation and mass is dominated by dark
+matter.}\label{fig:scl_umi_initial_isolation}
 \end{figure}

### 2025-09-24 thesis/results.tex
half-done discussion

--- 2025-08-31 thesis/results.tex
@@
-In this chapter, we present our results from our simulations of Sculptor
-and Ursa Minor. In each case, we first discuss the range of reasonable
-orbits and initial dark matter halos for each galaxy. Then, we describe
-how tides affect the evolution of each galaxy. We furthermore consider
-how the effects of the LMC changes the results, including using
-additional N-body simulations in evolving potentials. Finally, we review
-the predictions and characteristics of the present-day stellar
-populations and properties from our models.
+As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
+we test whether Galactic tides can form the extended density profiles of
+Scl and UMi. In this Chapter, we analyze tailored N-body simulations,
+using the methods described in Chapter \ref{sec:methods}, to assess
+tidal impacts on each galaxy. We find that tides drive dark matter loss
+in both systems but leave their compact stellar components largely
+unaffected. The Large Magellanic Cloud (LMC) perturbs the orbit of Scl,
+and somewhat UMi, yet the resulting tidal effects are weaker than in a
+Milky Way-only potential. Our simulations demonstrate that recent tides
+are unlikely to alter the stellar structure of Scl or UMi.
 
-\section{Sculptor}\label{sculptor}
+In this Chapter, we consider Scl first, describing tidal effects from
+the MW on dark matter and stellar components, and then consider the
+effects from the LMC. We then analyze UMi in a similar manner,
+considering in turn the dark matter evolution, stellar evolution, and
+orbital effects of the LMC.
 
-\subsection{Milky Way tides}\label{milky-way-tides}
+\section{Tidal effects on Sculptor}\label{tidal-effects-on-sculptor}
 
-\subsection{Tidal effects}\label{tidal-effects}
+\subsection{Dark matter evolution}\label{dark-matter-evolution}
 
-The Milky Way tides indeed affect Sculptor's dark matter halo.
-Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
-of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
-previous pericentres. This simulation uses the \texttt{compact} initial
-halo on the \texttt{smallperi} orbit. The dark matter forms large
-streams which orbit the Milky Way several times. However, note that the
-central dark matter cusp is similar in appearance across all simulation
-snapshots.
+As a representation of the most extreme tidal history, we initially
+investigate the \smallperi{} Scl model.
 
-More quantitatively, the halo loses about 99\% of its initial dark
-matter mass. This corresponds in a reduction from the initial
-\(v_{\rm max} = 31\,\kms\) to \(v_{\rm max} = 22\,\kms\). However, the
-stellar component of Sculptor is relatively unaffected.
+Scl experiences moderate tidal mass loss. Fig.~\ref{fig:scl_sim_images}
+shows the stripping of dark matter and formation of diffuse streams
+trailing and leading Sculptor's orbit. Because tidal stripping can be
+described as removal of the lowest energy particles, most mass loss
+occurs in the outer halo.
 
-While mass loss leads to adiabatic expansion of the stellar component,
-no features of tidal disruption or disequilibrium are currently
-observable. Fig.~\ref{fig:scl_smallperi_i_f} shows the final projected
-density of stars on the sky, and the initial and final radially averaged
-density profiles. No non-spherical density features are apparent, even
-at 5 decades fainter in surface density than the central density. In
-addition, the initial and final density profiles look identical up to
-some scale, with a tidally-induced excess of stars only appearing
-\(\sim 3\) orders of magnitude fainter than the faintest our density
-profile measures.
+N-body models may deviate from a point-particle trajectory due to
+dynamical self-friction \citep[e.g.,][]{miller+2020}. However, this
+effect is slight for Scl, which ends near the observed position, without
+adjusting the initial conditions (the green point in
+Fig.~\ref{fig:scl_sim_images}).
 
-The analytically motivated break and tidal radii corroborate a weak
-tidal effect on stars. In both cases, these radii work out to be
-\(\gtrsim 100\) arcminutes, outside where we can measure the density
-profile. In particular, because we have chosen the most extreme
-observationally permissible orbit, it is unlikely, for similar
-\(\Lambda\)CDM motivated initial conditions, that these radii would ever
-approach where the break in Sculptor's density profile appears.
+The inner density cusp is tidally resilient.
+Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
+velocity profiles, and the evolution of the maximum circular velocity.
+The maximum velocity drops from \(31\,\kms\) to \(22\,\kms\), evolving
+along the tidal track from \citet{EN2021}. The final circular velocity
+profile resembles the initial with an inner cusp, but has a sharper
+outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
+mass, as listed in Table~\ref{tbl:scl_sim_results}. However, the inner
+structure is not expected to be affected, as the Jacobi radius is over
+3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
+significant mass, but mostly from the outer halo.
 
@@
 \caption[Sculptor simulation snapshots]{Images of the dark matter
-evolution over a selection of past apocentres and the present day
-position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
-(approximately orbital) plane and the colourscale is logarithmic
-spanning 5 orders of magnitude between the maximum and minimum values.
-In this image, stars occupy only ever a few pixels so are not
-plotted.}\label{fig:scl_sim_images}
+evolution over a selection of past apocentres and the present day.
+Limits range from -150 to 150 kpc in the \(y\)--\(z\) (approximately
+orbital) plane and the colourscale is logarithmic spanning 5 orders of
+magnitude between the maximum and minimum values. The green dot
+represents the final expected position of the galaxy and the solid and
+dotted grey curves represents the orbit over one previous or future
+radial oscillation respectively.}\label{fig:scl_sim_images}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
-\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
-orbit. Todo: add velocity dispersion plot to
-RHS}\label{fig:scl_tidal_track}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
+\caption[Sculptor tidal tracks]{Dark matter evolution for the
+\smallperi{} model of Sculptor. Blue solid and orange dashed lines show
+the initial and final circular velocity profiles. The points represent
+the evolution of the maximum circular velocity, and the dotted black
+line shows the tidal track from \citet{EN2021}. To calculate the
+velocity profiles, unbound particles are iteratively removed,
+recalculating the potential at each step assuming spherical
+symmetry.}\label{fig:scl_tidal_track}
 \end{figure}
 
-\begin{figure}
+\begin{table*}[t]
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
-\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
-velocity dispersion within 1 kpc for different Scl models. In all cases,
-the evolution is mild. Note that binarity may reduce the inflate the
-observed velocity dispersion by \textasciitilde{} 1 km/s, so the
-conservative lower limit is around 8 km/s. \textbf{TODO}, add bound mass
-with time. Maybe combine with tidal tracks and radius / time orbit.}
-\end{figure}
+\caption[Simulation results for Sculptor’s dark matter]{The orbital and dark matter properties for the simulation of Sculptor.
+The random samples column shows the distributions from point orbits, and the \smallperi{} column contains the results from the N-body simulation.
+}
+\label{tbl:scl_sim_results}
+\begin{tabular}{lll}
+\toprule
+Property & random samples & \smallperi{}\\
+\midrule
+pericentre & $53\pm3$ & 29\\
+apocentre & $102\pm3$ & 94.4\\
+time of last pericentre / Gyr & $-0.45 \pm 0.2$ & -0.47\\
+number of pericentres & 6 & 6\\
+Jacobi radius / kpc & $4.5 \pm 0.3$ & 3.5\\
+Jacobi radius / arcmin & $186\pm12$ & 101\\
+final heliocentric distance / kpc & $83.2\pm2$ & 81.6\\
+$\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.695\\
+$r_\textrm{max, f} / r_\textrm{max, i}$ &  & 0.406\\
+fractional final bound mass &  & 0.0893\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Stellar evolution}\label{stellar-evolution}
+
+Tides minimally affect the stellar component of Sculptor in the
+\smallperi{} orbit. In Fig.~\ref{fig:scl_smallperi_i_f}, the projected
+stellar distribution displays no prominent distortions, and the radial
+density profile is nearly unchanged. Only at a surface density
+\(\sim10^8\) times fainter than the centre may faint tidal features
+emerge. The total stellar mass lost corresponds to \(\sim 10\) stars in
+total (see Table~\ref{tbl:scl_sim_results})---a formidable challenge
+with the best of observations.
+
+More extended stellar profiles are more sensitive to tides. With Plummer
+initial conditions, as shown in
+Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the model loses more stellar
+mass and forms more luminous tidal tails. Observations reaching surface
+densities \(\sim10\) times fainter than our data could reveal a stream
+in this case. Nevertheless, the radial extent probed by our data remains
+nearly unchanged by tidal evolution.
+
+Table~\ref{tbl:scl_sim_stars_results} quantifies the evolution of
+stellar properties. The stellar velocity dispersion decreases by only
+\(\sim1\,\kms\) and the half-light radius expands by \(\sim 10\%\). This
+is consistent with adiabatic expansion due to the reduction of the total
+mass \citep[e.g.,][]{stucker+2023}. In addition, the break and Jacobi
+radii are \(\gtrsim 100\) arcminutes on the sky---tidal signatures would
+be beyond the reach of our data. Altogether, Galactic tides negligibly
+impact Scl's stellar component.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
-\caption[Sculptor initial and final density profiles]{Effects on
-exponential initial stars. TODO: plot 2D sky proj.
-stars}\label{fig:scl_smallperi_i_f}
+\caption[Sculptor initial and final density profiles]{The tidal effects
+on Scl's stellar component, for the \smallperi{} orbit with the fiducial
+halo and exponential stars with \(R_s=0.10\,\kpc\). \textbf{Top:} the
+initial (left) and final (right) 2D projected density of stars on the
+sky. The solid circle marks \(6R_h\), the dotted circle the break
+radius, and the (dotted) line the past (future) orbit. \textbf{Bottom:}
+The initial (dotted) and final (solid) stellar density profiles as
+compared to the observed stellar density profile. Arrows mark the break
+and Jacobi radii (Eqs.~\ref{eq:r_break}, \ref{eq:r_jacobi})
+.}\label{fig:scl_smallperi_i_f}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
-\caption[Sculptor Plummer initial and final density profiles]{effects on
-Plummer initial stars.}\label{fig:scl_smallperi_plummer_i_f}
+\caption[Sculptor Plummer initial and final density profiles]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f} except for Plummer initial stars with
+\(R_h = 0.20\,\kpc\). While a faint stream may be visible with deeper
+observations, effects on the stellar profile are
+minimal.}\label{fig:scl_smallperi_plummer_i_f}
 \end{figure}
 
-\subsection{Effects of the LMC}\label{sec:scl_lmc}
+\begin{table*}[t]
+\centering
+\caption[Simulation results for Sculptor’s stars]{The present-day stellar properties for the simulations of Sculptor.
+In each row, we have the initial stellar velocity dispersion (within 1kpc), the final velocity dispersion, the fraction of stellar mass unbound, the initial half-light radius, the final half-light radius, and the break radius in arcmin and kpc (Eq. \ref{eq:r_break}).
+}
+\label{tbl:scl_sim_stars_results}
+\begin{tabular}{lll}
+\toprule
+Property & Exponential & Plummer\\
+\midrule
+$\sigma_{\V, i}\,/\,\kms$ & 9.8 & 10.7\\
+$\sigma_{\V, f} \,/\,\kms$ & 8.8 & 9.4\\
+fractional stellar mass loss & $2.1\times 10^{-6}$ & $0.024$\\
+$R_{h, i}\,/\,\kpc$ & 0.169 & 0.202\\
+$R_{h, f}\,/\,\kpc$ & 0.189 & 0.227\\
+break radius / arcmin & $98$ & $105$\\
+break radius / kpc & 2.3 & 2.5\\
+\bottomrule
+\end{tabular}
+\end{table*}
 
-The Milky Way isn't the only galaxy in town. Becoming more apparent, the
-infall of the LMC substantially affects the Milky Way system. The mass
-of the LMC may be as high as 1/5 of the Milky Way mass, so including an
-LMC in the Milky Way potential may change conclusions about properties
-and orbits of the Milky Way satellite system.
+\subsection{Orbital effects of the LMC}\label{sec:scl_lmc}
 
-\subsubsection{Orbital effects}\label{orbital-effects}
+The Milky Way isn't the only galaxy in town. Recently, work has shown
+that the infall of the LMC may substantially affect the Milky Way system
+\citep[e.g.,][]{erkal+2019, cautun+2019, garavito-camargo+2021, vasiliev2023}.
+With a mass up to one fifth of the MW \citep[e.g.,][]{penarrubia+2015},
+the LMC infall affects conclusions about the MW properties and the
+orbits of satellites \citep[see
+e.g.,][]{patel+2020, battaglia+2022, correamagnus+vasiliev2022}. In this
+section, we examine how the LMC affects the orbital history of Sculptor.
 
-As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
-influenced by the presence of an LMC.
+We use the \texttt{L3M11} model of the MW and LMC potential from
+\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
+multipole approximation of an N-body simulation including a live MW and
+LMC dark matter halo. The potential includes a static MW bulge and disk,
+evolving MW and LMC halos, and the MW reflex motion. In their
+simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
+\(M_{\rm 200}= 98.4\times10^{10}\Mo\) , and the LMC a NFW halo with
+\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
+\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
+potential. Notably, this model has a previous LMC pericentre
+\(\sim6\,\Gyr\) ago.
 
-To explore the effects of the LMC on Sculptor, we use the publically
-available L3M11 model from \citet{vasiliev2024}. This model stars with a
-lighter Milky Way than our fiducial \citet{EP2020}, \citet{mcmillan2011}
-like model. The L3M11 potential is an evolving multipole approximation
-of an N-body simulation with both a live MW and LMC dark matter halo.
-The MW intital conditions were an NFW with \(r_s=16.5\,\)kpc and mass
-\(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
-\(r_s=11.7\) and \(M_{100} = 2.76 \times 10^{11} \Mo\). This model had a
-previous LMC pericentre at about 6 Gyr ago.
+The inclusion of the LMC reshapes Scl's orbital history, as shown in
+Fig.~\ref{fig:scl_lmc_orbits_effect}. In the MW-only potential, Scl's
+orbit is typical of a long-term MW satellite. However, Scl's passage by
+the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term orbit---Scl instead
+reaches an apocentre of nearly \(300\,\kpc\). Scl may even be on first
+infall, depending on the MW and LMC mass. Scl's orbit is
+counter-rotating to the LMC, so Scl is unlikely to be a LMC satellite.
 
-Fig.~\ref{fig:scl_lmc_orbits_effect} displays the effect of including an
-LMC in the potential. The green samples are in the initial MW-only
-potential in the \texttt{L3M11} model, and the orange samples are
-integrated in the evolving MW and LMC \texttt{L3M11} model. The past 1
-Gyr is similar in both cases, but the orbits diverge significantly
-afterwards. The recent passage of Sculptor with the LMC around 0.1 Gyr
-ago allows for Sculptor to begin as far as 300 kpc from the Milky Way
-centre. The evolving potential also adds significant long term
-variability in the possible orbits of Sculptor. Sculptor, however, is
-orbiting in the opposite direction of the LMC so is likely not
-associated with the LMC system.
+The timing of the LMC encounter implies a break radius consistent with
+the observed density kink. We select the orbit with the
+\(6\sigma\)-smallest LMC pericentre. The orbit is selected following the
+procedure in Section~\ref{sec:orbital_estimation} except with
+observational uncertainties doubled. Results are similar when selecting
+for a small MW pericentre instead. This \texttt{LMC-flyby} orbit, is
+integrated back in time \(2\,\Gyr\) ago to isolate recent tidal effects.
+We modify Scl's initial halo to have \(\rmax = 2.5\,\kpc\) and
+\(\vmax = 25\,\kms\), slightly reducing the initial stellar velocity
+dispersion. Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected
+orbit in black and Table~\ref{tbl:orbit_ics} records the initial
+conditions.
 
-Given the large uncertainties of the LMC model, we conservatively double
-all of the observational parameters of Sculptor. This has a similar
-effect to including LMC mass and orbital uncertainties but is
-considerably simpler. From this larger range of orbits, we once again
-select the orbit with the median final observables of all orbits with
-pericentres less than the 0.0027th quantile. This orbit, the
-\texttt{smallperilmc} orbit is plotted in black and is only integrated
-up to 2 Gyr ago, to isolate recent tidal effects.
+\begin{table*}[t]
+\centering
+\caption[Orbits and results for Scl in the LMC potential.]{The orbital properties and dark matter evolution for the models including an LMC.
+Similar to Table \ref{tbl:scl_sim_results} except quantities with respect to the LMC are in parentheses.
+}
+\label{tbl:scl_lmc_sims}
+\begin{tabular}{lll}
+\toprule
+Property & random samples & \texttt{LMC-flyby}\\
+\midrule
+pericentre & $44\pm 3$ ($29 \pm 2$) & 39 (20)\\
+apocentre & $218 \pm 8$ & –\\
+time of last pericentre / Gyr & $-0.38\pm0.01$ (-0.11) & -0.33 (-0.10)\\
+number of pericentres & 1 (2) & 1 (1)\\
+Jacobi radius / kpc & $4.1\pm0.3$ ($4.5\pm0.2$) & 2.8 (2.6)\\
+Jacobi radius / arcmin & $168 \pm 11$ ($186\pm6$) & 132 (121)\\
+final heliocentric distance / kpc & $83.2\pm2$ & 72.9\\
+$\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.928\\
+$r_\textrm{max, f} / r_\textrm{max, i}$ &  & 0.763\\
+fractional final bound mass &  & 0.5402\\
+\bottomrule
+\end{tabular}
+\end{table*}
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}}
-\caption[Sculptor Orbits with LMC]{This figure is similar to
-Fig.~\ref{fig:scl_orbits} except that we are showing the orbits with and
-without an LMC. In the bottom row, the distance from Sculptor (or the
-LMC) to the MW is plotted (left), and the Sculptor - LMC distance
-(right.)}\label{fig:scl_lmc_orbits_effect}
+\caption[Sculptor orbits with LMC]{Similar to Fig.~\ref{fig:scl_orbits}
+except for orbits with (orange) and without (green lines) the inclusion
+of an LMC (blue line) in the potential. The bottom row additionally
+shows the distance between Scl and the LMC over
+time.}\label{fig:scl_lmc_orbits_effect}
 \end{figure}
 
-\textbf{tinyperilmc}
+\subsection{Tidal effects from the
+LMC}\label{tidal-effects-from-the-lmc}
 
-\begin{itemize}
-\item
-  ra = 15.0183 dec = -33.7186 distance = 73.1 pmra = 0.137 pmdec =
-  -0.156 radial\_velocity = 111.2
+The combined tidal force of the MW and LMC are even weaker for Scl than
+in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
+matter evolution of Scl and the passage of the LMC. With only one MW
+pericentre, Scl's dark matter is less disrupted than the previous
+MW-only model. The subsequent LMC passage has little effect. As a
+result, the dark matter structure evolves mildly and \(\sim 50\%\) of
+mass remains bound (Table~\ref{tbl:scl_lmc_sims}).
 
-  t\_i = -2.00 pericentre = 38.82 apocentre = 187.50 t last peri = -0.33
-  x\_i = {[}4.30 138.89 125.88{]} v\_i = {[}6.89 -56.83 52.09{]}
-\end{itemize}
-
-\subsubsection{Tidal effects}\label{tidal-effects-1}
-
-Unlike the previous mdoel, this model only has one pericentre with the
-Milky Way. Fig.~\ref{fig:scl_lmc_sim_images} shows snapshots of Sculptor
-over the past 2 Gyr while marking the position of the LMC. With only one
-pericentre, and a larger one than the \texttt{smallperi} orbit,
-Sculptor's dark matter is substantially less disrupted. And while, based
-on the tidal tensor values, the LMC induces a greater instantaneous
-tidal effect than the Milky Way, Sculptor's dark matter component does
-not show strong effects due to the LMC.
-
-Finally, Fig.~\ref{fig:scl_lmc_i_f} shows the projected on-sky final
-stellar distribution and the initial and final stellar density profile
-for this model with exponential stars. The stellar component does not
-change at all. While the break radius set by the time since the last LMC
-pericentre would agree with the location of the break in the observed
-density profile, no stellar effect would be observable.
-
-The LMC flyby encounter is an approximately impulsive encounter, in
-contrast with more adiabadic mass loss due to the Milky Way. Impulsive
-encounters tend to inject energy into the stellar and dark matter
-distribution, and can initially cause the galaxy to contract. In
-addition, the tidal force is required to be far larger than for slower,
-adiabadic mass loss, because the galaxy experiences this tidal field for
-far less time. Even in models where Sculptor passes through the the LMC,
-tidal tails do not form immediately after this encounter.
-
-A final note is that this model only considers the recent tidal effects
-.As illustrated in the range of possible orbits in
-Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
-experienced an extremely small pericentre with the Milky Way. This
-pericentre has the potential to substantially rearrange the stellar
-component, drive large tidal mass loss, and create a Plummer-like
-stellar density profile. However, this encounter is highly dependent on
-the choice of the MW-LMC potential model, and may not occur at all. Long
-time integration in dynamic potentials amplifies uncertainties in the
-inputs and may not be reliable. \textbf{TODO: I would love to get this
-model to work\ldots{}}
+Correspondingly, the stellar component is nearly unchanged.
+Fig.~\ref{fig:scl_lmc_i_f} shows the projected stellar distributions and
+density profiles of this model. While the break radius is within the
+observed density profile, tidal effects are too weak to be detectible.
+Structural properties of the stars similarly evolve little
+(Table~\ref{tbl:scl_lmc_sim_stars}). While the instantaneous tidal force
+from the LMC is larger than the MW, Scl does not experience the LMC
+tidal field long enough to display disturbances.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.png}}
-\caption{Sculptor Simulation Snapshots with
-LMC}\label{fig:scl_lmc_sim_images}
+\caption[Sculptor simulation snapshots with LMC]{Similar to
+Fig.~\ref{fig:scl_sim_images} except for the case where the potential
+includes an LMC. The current position and path of the LMC are
+represented by the green dot and line respectively. We also plot the
+full orbit (over the past 2Gyr) for both Scl and the LMC, as less than
+one radial period happens over this time
+frame.}\label{fig:scl_lmc_sim_images}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_i_f.pdf}}
-\caption[Sculptor initial and final density with LMC]{The tidal effects
-on the stellar surface density due to the LMC
-today.}\label{fig:scl_lmc_i_f}
+\caption[Sculptor initial and final density with LMC]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f} except for the \texttt{LMC-flyby}
+model. With only one MW pericentre and a recent, rapid LMC encounter,
+tidal forces do not appear to affect the stellar
+distribution.}\label{fig:scl_lmc_i_f}
 \end{figure}
-
-\section{Ursa Minor}\label{ursa-minor}
 
@@
 \centering
-\caption{Given the observed break radii, the required time of last pericentre and pericentre to produce the observed tidal effect.}
-\begin{tabular}{llll}
+\caption[Simulation results for Sculptor’s stars in the LMC+MW potential]{Similar to Table \ref{tbl:scl_sim_stars_results}, but for the properties of the stellar components of the \texttt{LMC-flyby} model of Sculptor.
+}
+\label{tbl:scl_lmc_sim_stars}
+\begin{tabular}{lll}
 \toprule
-parameter & Scl & UMi & \\
+Property & Scl: LMC-exponential & LMC-Plummer\\
 \midrule
-$r_{\rm break}$ observed & $25 \pm 5$ arcmin & $30 \pm 5$ arcmin & (sestito+2024a?); (sestito+2024b?)\\
-$t_{\rm last\ peri}$ required & $110\pm30$ Myr & $120\pm30$ Myr & Eq. \ref{eq:r_break}\\
-$r_{\rm peri}$ required & 16kpc & 14kpc & Eq. \ref{eq:r_jacobi}\\
+$\sigma_{\V, i}\,/\,\kms$ & 9.0 & 9.4\\
+$\sigma_{\V, f} \,/\,\kms$ & 8.8 & 9.2\\
+fractional stellar mass loss & $<10^{-12}$ & 0.0013\\
+$R_{h, i}$ / kpc & 0.186 & 0.201\\
+$R_{h, f}$ / kpc & 0.189 & 0.205\\
+break radius & $77'$, 1.6 kpc & $80'$, 1.7 kpc\\
+LMC break radius & $23'$, 0.49 kpc & $24'$, 0.52 kpc\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Robustness to initial
+conditions}\label{robustness-to-initial-conditions}
+
+Some initial conditions may be slightly more susceptible to tides. Dark
+matter halos which are cored or less concentrated disrupt or lose mass
+faster \citep[e.g.,][]{stucker+2023}. We consider alternative dark
+matter halo initial conditions in the Appendix. We find that while the
+details of the strength of tidal impacts depend on the halo, our models
+here remain representative of the tidal effects.
+
+\subsection{Summary}\label{summary}
+
+We find, including only the MW, that tides only remove dark matter from
+the outskirts of Scl. The central cusp and compact stellar distribution
+are resilient to tides and any tidal effects are well-outside the reach
+of current observations. We then find that the LMC strongly perturbs
+Scl's orbit---in this case, Scl may be on first infall. However, with
+only 2 pericentres, the combined tides of the LMC and MW tides are
+weaker than for our initial model. In either case, tides are unlikely to
+affect Sculptor's stellar component.
+
+\section{Tidal effects on Ursa Minor}\label{tidal-effects-on-ursa-minor}
+
+The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
+case. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM evolution.
+UMi loses more DM mass, with only about 3\% of the total mass remaining
+({[}Table~\ref{tbl:umi_sim_results}; {]}). However, tidal features in
+the observed stellar components are still extremely faint, appearing
+outside 100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
+moderately affect the observed size and velocity dispersion of Ursa
+Minor (Table~\ref{tbl:umi_sim_stars_results}).
+
+A stream may just be beyond observational limits if Ursa Minor's stars
+were initially more extended. As seen in Fig.~\ref{fig:umi_plummer_i_f},
+the tidal tails are more luminous than for Sculptor and would begin near
+the outermost density measurement. The properties of the stream are
+shown in Fig.~\ref{fig:umi_tidal_stream}. The stream stars become
+apparent outside of about 2 degrees, or the Jacobi radius. As expected
+for tidal tails, the stream shows gradients in each observable, with a
+dispersion in each observable similar to that of the dwarf progenitor.
+
+\begin{table*}[t]
+\centering
+\caption[Simulation results for Ursa Minor’s dark matter]{The present-day properties for Ursa Minor’s final dark matter halo. See Table \ref{tbl:scl_sim_results} for details. }
+\label{tbl:umi_sim_results}
+\begin{tabular}{lll}
+\toprule
+Property & Random orbits & \smallperi{}\\
+\midrule
+pericentre & $37\pm3$ & 30\\
+apocentre & $83 \pm 4$ & 75\\
+time of last pericentre & $-0.97 \pm 0.07$ & -0.80\\
+number of pericentres & $\sim 8$ & 8\\
+Jacobi radius / kpc & $3.7 \pm 0.3$ & 2.9\\
+Jacobi radius / arcmin & $184 \pm 12$ & 156\\
+final heliocentric distance & $70.1 \pm 3.6$ & 64.7\\
+${\vmax}_f / {\vmax}_i$ &  & 0.511\\
+${\rmax}_f / {\rmax}_i$ &  & 0.249\\
+fractional dm final mass &  & 0.035\\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption[Simulation results for Ursa Minor’s stars]{Similar to Table \ref{tbl:scl_sim_stars_results}, the present-day stellar properties for the simulation of Ursa Minor for exponential and Plummer stars.
+}
+\label{tbl:umi_sim_stars_results}
+\begin{tabular}{lll}
+\toprule
+Property & smallperi-exp & smallperi-Plummer\\
+\midrule
+$\sigma_{\V, i}\,/\,\kms$ & 10.0 & 10.9\\
+$\sigma_{\V, f}\,/\,\kms$ & 8.2 & 8.5\\
+fractional stellar mass loss & $0.00015$ & 0.039\\
+$R_{h, i}\,/\,\kpc$ & 0.135 & 0.151\\
+$R_{h, f}\,/\,\kpc$ & 0.169 & 0.191\\
+break radius & 197 arcmin, 3.7 kpc & 204 arcmin, 3.8 kpc\\
 \bottomrule
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
-\caption[Ursa Minor simulation snapshots]{Ursa Minor simulation images.}
+\caption[Ursa Minor simulation snapshots]{Similar to
+Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
+orbit.}\label{fig:umi_sim_images}
 \end{figure}
-
-Figure: Velocity dispersion evolution of Ursa Minor
 
@@
 \caption[Ursa Minor simulated density profiles]{The tidal effects on the
-stellar surface density.}
+stellar surface density of Ursa Minor for exponential stars on the
+\smallperi{} orbit.}\label{fig:umi_smallperi_i_f}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_plummer_i_f.pdf}}
+\caption[Ursa Minor simulated density profiles]{The tidal effects on the
+stellar surface density of Ursa Minor for Plummer stars on the
+\smallperi{} orbit.}\label{fig:umi_plummer_i_f}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
+\caption[Ursa Minor Predicted stream]{The properties of the stream
+around the UMi \smallperi{} orbit with Plummer stars. The panels are all
+as a function of \(\xi'\), the distance along the stream as defined by
+the current GSR proper motion vector. The top panels show the GSR proper
+motions in RA and Dec, and the bottom two show the distance and GSR
+radial velocities. To sample the stream, we randomly draw 100,000
+samples from the snapshot based on the stellar weights. A detectible
+gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
+the stream is tracked across several
+degrees.}\label{fig:umi_tidal_stream}
 \end{figure}
@@
 \subsection{Effects of the LMC}\label{effects-of-the-lmc}
+
+Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
+the orbit of Ursa Minor. Predominantly, the effect is to increase the
+orbital period, apocentre and pericentre. However, the orbit remains in
+a similar plane and with similar morphology. As UMi is on the opposite
+side of the Galaxy of the LMC, this is not unexpected. The deviation
+from the MW-only orbit is mostly due to the LMC-induced reflex motion of
+the Milky Way.
 
@@
 LMC are plotted as scatter points and the solid blue line represents the
-LMC trajectory. Note that the LMC only increases Ursa Minor's peri and
-apo-centres, weakening any tidal effect. Interestingly, there is a
-change that Ursa Minor may have once been bound to the LMC (diverging
-orange lines at top left of middle panel.)}
+LMC trajectory. Note that the LMC mostly increases Ursa Minor's
+pericentres and apocentres.}\label{fig:umi_orbits_lmc}
 \end{figure}
+
+\subsection{Summary}\label{summary-1}
+
+While tides affect UMi more strongly than Scl, the tidal effects are
+insufficient to reshape the observed stellar density profile. Faint
+tidal tails may be observable with deeper data. Finally, including the
+LMC in the potential further weakens the tides experience by UMi.

### 2025-09-27 thesis/abstract.tex
a compltet thesis draft

@@
 The satellite galaxies of the Milky Way (MW), except the Magellanic
-Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating
-stellar systems dynamically dominated by dark matter. The light profiles
-of dSphs are usually well-approximated by an exponential law, with a
-steep outer cutoff. Yet systems like the Sculptor (Scl) and Ursa Minor
-(UMi) dSphs have distant member stars, out to \(\sim10\) effective radii
-from the centre, indicative of outer deviations from an exponential law.
-The origin of these extended profiles is unclear, and may reflect either
-the influence of Galactic tides or an innate feature. In this thesis, we
-review the existence of extended light profiles in Scl and UMi,
-validating the Bayesian membership catalogue of \citet{jensen+2024}. To
-evaluate if tides indeed can explain these density features, we conduct
-idealized N-body simulation of both galaxies in the tidal field of the
-Milky Way. We find that neither dwarf has been subjected to tidal forces
-strong enough to create the observed extended light profiles from
-initially exponential profiles. One complication is that Sculptor's
-orbit is strongly influenced by the presence of the Large Magellanic
-Cloud (LMC). Our models still suggest that the combined tides of the LMC
-and MW are unable to explain Scl's outer profile. We conclude that the
-extended light profiles of Scl and UMi are innate, likely reflecting
-their assembly histories. Mergers, accretion events, or episodic star
-formation are possible explanations for the observations of extended
-light profiles.
+Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating,
+dark-matter dominated stellar systems. The light profiles of dSphs
+typically follow an exponential law, with the outer density sharply
+declining . Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
+dSphs host member stars out to \(\sim10\) effective radii, indicating
+outer deviations from an exponential law. The origin of these extended
+profiles is unclear, possibly arising from Galactic tides or intrinsic
+properties. In this thesis, we review the evidence for extended light
+profiles in Scl and UMi, validating the Bayesian membership catalogue of
+\citet{jensen+2024}. To evaluate if tides can produce these density
+features, we conduct idealized N-body simulation of both galaxies in the
+tidal field of the Milky Way. We find that neither dwarf experiences
+tides sufficient to affect their stellar distribution. One complication
+is that Sculptor's orbit is strongly influenced by the presence of the
+Large Magellanic Cloud (LMC). Our models still suggest that the combined
+tides of the LMC and MW are unable to explain Scl's outer profile. We
+conclude that the extended light profiles of Scl and UMi not of tidal
+origin. They are instead likely innate, possibly explained by past
+mergers, accretion events, or episodic star formation.

### 2025-09-27 thesis/data.tex
a compltet thesis draft

@@
 from the satellite. Each likelihood is normalized as a probability
-density over the respective parameter space. The appendix discusses the
-details of quality cuts and the likelihood calculation.
+density over the respective parameter space. Appendix
+\ref{sec:density_extra} discusses the details of quality cuts and the
+likelihood calculation.
 
@@
 confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
-of each dwarf. For an exponential profile, 99.95\% of stars fall within
-a radius of \(6R_h\), so the mere presence of very distant member stars
-provides evidence for deviations from an exponential profile.
+of each dwarf. For an exponential profile, 0.05\% of stars fall outside
+a radius of \(6R_h\) and less than \(1/10^6\) fall outside
+\(10R_h\)---the mere presence of very distant member stars provides
+evidence for deviations from an exponential profile.
 
@@
 \centering
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.png}}
 \caption[Sculptor sample selection]{The distributions of various samples
-of \emph{Gaia} stars for Sculptor. We plot light grey points for ``all''
-field stars (with consistent parallaxes and reliable photometry and
-astrometry), turquoise points for ``CMD+PM'' selected stars
-(Eq.~\ref{eq:sel_cmd_pm}), blue squares for the ``fiducial'' sample
-(with membership probability \(P_{\rm sat} > 0.2\)), and indigo diamonds
-for the ``RV members'' sample. We mark the two far-outlier stars from
-\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Top:}
+of \emph{Gaia} stars for Sculptor. In each panel, we plot light grey
+points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
+selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
+``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
+``RV members'' sample. We mark the two far-outlier stars from
+\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
 Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
-half-light radii. \textbf{Bottom left:} Colour magnitude diagram in Gaia
+half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
 \(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
@@
 \centering
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.png}}
 \caption[Ursa Minor sample selection]{Similar to
@@
 \centering
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.png}}
 \caption[Fornax sample selection]{Similar to
-Fig.~\ref{fig:scl_selection} except for Fornax. RV measurements are from
-\citet{WMO2009}. Fornax does not show the same extended outer halo of
-probable members as Sculptor or Ursa Minor despite having many more
+Fig.~\ref{fig:scl_selection} except for Fornax. The isocrone is instead
+for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
+Fornax does not show the same extended outer halo of probable members as
+Sculptor or Ursa Minor despite having many more
 stars.}\label{fig:fornax_selection}
@@
 While the density profiles of all samples agree in the inner regions,
-they begin to deviate outside \(6 R_h\). Outside this radius, the
+they begin to deviate outside \(6R_h\). Outside this radius, the
 fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
@@
 
+Our density profiles are robust to alternative methodologies and
+magnitude cuts. Density profiles based on J+24 candidates may be biased
+by their assumed spatial likelihoods. To address this, we consider a
+Bayesian model with a non-parametric spatial likelihood. We also explore
+selections using absolute cuts, and the effect of structural assumptions
+of the density profile (Appendix \ref{sec:extra_density}). These
+methodologies are consistent except when the derived density drops below
+the background of satellite-like stars (from the CMD+PM sample). We also
+do not find evidence of magnitude biases in the density profiles,
+indicating inhomogeneous magnitude-completeness is small. Our density
+profiles furthermore agree with photometric surveys and similar
+literature (see Section~\ref{sec:extra_density}). We conclude that
+selection criteria do not influence our results.
+
 Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
@@
 from \citet{munoz+2018}. All of the classical dwarfs appear to be well
-described by an exponential profile in the inner regions. In the outer
-regions, however, Sculptor and Ursa Minor deviate, and show a clear
-outer excess over an exponential law (solid black line). These galaxies
-are better fit by a Plummer law (dashed black line). The deviation from
-an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
-of magnitude. The remainder of this thesis will be devoted to assessing
-whether the outer excess shown by Scl and UMi are due to Galactic tides.
+described by an exponential profile in the inner regions.\footnote{Comparing
+  density profiles of dwarf galaxies is complicated by variations in the
+  effective depth between galaxies. However, the deviations between Scl,
+  UMi and other galaxies is apparent even when data covers all dwarfs.}
+In the outer regions, however, Sculptor and Ursa Minor deviate, and show
+a clear outer excess over an exponential law (solid black line). These
+galaxies are better fit by a Plummer law (dashed black line). The
+deviation from an exponential grows outwards, and at \(\sim 8 R_h\), may
+reach 2 orders of magnitude. The remainder of this thesis will be
+devoted to assessing whether the outer excess shown by Scl and UMi are
+due to Galactic tides.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
-\caption[Sculptor density profiles]{The density profile of Sculptor,
-Ursa Minor, and Fornax for different selection criteria, ploted as log
-surface dencity versus log elliptical radius. The samples are: ``all''
-selects any high quality and parallax-consistent star (grey
-circles),``all--background'' subtracts a uniform background density from
-the ``all'' profile (orange pentagons), ``CMD+PM'' select stars
-according to CMD and PM only (teal triangles), and ``fiducial'' also
-includes spatial information (blue squares). We mark the half-light
-radius with a vertical dashed line and the background density with the
-horizontal grey line. For Ursa Minor, we show the expected location of
-Muñoz 1 stars as a horizontal bar (ranging from plus/minus 3 half-light
+\caption[Density profiles for different Gaia samples]{The density
+profile of Sculptor, Ursa Minor, and Fornax for different selection
+criteria, ploted as log surface dencity versus log elliptical radius.
+The samples are: ``all'' selects any high quality and
+parallax-consistent star (grey circles),``all--background'' subtracts a
+uniform background density from the ``all'' profile (orange pentagons),
+``CMD+PM'' select stars according to CMD and PM only (turquoise
+triangles), and ``fiducial'' also includes spatial information (blue
+squares). We mark the half-light radius with a vertical dashed line and
+the background density with the horizontal grey line. For Ursa Minor, we
+show the expected location of Muñoz 1 stars as a horizontal bar (ranging
+from \(\pm3\) times the cluster's half-light
 radii).}\label{fig:scl_observed_profiles}

### 2025-09-27 thesis/discussion.tex
a compltet thesis draft

--- 2025-08-18 thesis/discussion.tex
@@
-\section{Reliability and comparison of observed density
-profiles}\label{reliability-and-comparison-of-observed-density-profiles}
+In the previous chapters, we have shown that (1) Sculptor and Ursa Minor
+have unusually extended density profiles for MW classical satellites and
+(2) recent tides are an unlikely explanation. Here, we first consider
+the reliability and assumptions supporting our conclusions. We then
+consider possible formation scenarios of extended density profiles,
+including multi-epoch star formation, dwarf mergers and subhalo
+encounters, and alternative theories of dwarf galaxy structure. We
+finish with an outlook on prospects for disentangling various stellar
+halo formation scenarios.
 
-\subsection{Caveats}\label{caveats}
+While tides have been a popular explanation for some features of dwarf
+galaxies \citep[discussion above, and
+e.g.,][]{tsujimoto+shigeyama2002, mayer+2001a}, our work, and many
+others, have disfavoured this for Scl, UMi, and many other MW dwarfs.
+For example, \citet{read+2006} suggest that the lack of observational
+evidence for a rising velocity dispersion profile with radii indicates
+most dwarfs are not tidally affected. Later, \citet{penarrubia+2009}
+define and use the break radius, Eq.~\ref{eq:r_break}, to show that the
+(then understood) orbits of most satellites are inconsistent with a
+tidal feature. In addition, \citet{pace+erkal+li2022} use a criterion
+based on the observed mean density and orbital pericentre (like the
+Jacobi radius) to show most dwarfs are unlikely to be undergoing
+significant tidal disruption. Most recently,
+\citet{tchiorniy+genina2025} use a similar idealized framework to ours,
+with a focus on the inner density, showing tides do not strongly affect
+equilibrium assumptions for several classical dwarfs. The general tidal
+evolution of our simulations is furthermore consistent with much of the
+literature \citep[e.g.,][]{robles+bullock2021, EN2021}. Tides seem to
+now play a subdominant role in the evolution of most Milky Way
+satellites.
 
-While the J+24 method is excellent and has been verified, there are
-several possible limitations we discuss here. In general, while these
-limitations are likely real, they do not substantially affect our
-conclusions up to where we suggest the density profiles are unreliable.
+\section{Comparison with prior work}\label{comparison-with-prior-work}
 
-\textbf{Spatial likelihood.} J+24 method was designed in particular to
-detect the presence of a density excess and find individual stars at
-large radii to be followed up. We are more interested in accurately
-quantifying the density profile and size of any perturbations. One
-potential problem with using J+24's candidate members is that the
-algorithm assumes the density is either described by a single or double
-exponential. If this model does not accurately match the actual density
-profile of the dwarf galaxy, we want to understand the impact of this
-assumption.In particular, in Fig.~\ref{fig:scl_observed_profiles},
-notice that the \(P_{\rm sat}\) selection method produces small
-errorbars, even when the density is more than 1 dex below the local
-background. These stars are likely selecting stars from the statistical
-MW background consistent with UMi PM / CMD, recovering the assumed
-density profile. As a result, the reliability of these density profiles
-below the CMD+PM background may be questionable. A more robust analysis,
-removing this particular density assumption, would be required to more
-appropriately represent the knowledge of the density profile as the
-background begins to dominate.
+For both Scl and UMi, these galaxies have been studied extensively in
+both a theoretical and observational context. While many works
+considered mass modelling, star formation histories, chemistry, and
+other observational properties, we focus on the observations and models
+concerning tides and dynamical evolution here.
 
-\textbf{Uncertainty misrepresentation.} A more self consistent model
-would fit the density profile to the entire field at once, eliminating
-possible misrepresentation of the uncertainties. In the Appendix to this
-section, we \emph{will} also discuss an alternative method which runs a
-MCMC model using the likelihoods above to solve for the density in each
-elliptical bin. From these tests, we note that the density profile and
-uncertainties derived from the J+24 sample are reliable insofar as the
-dwarf density is above the background from MW CMD+PM-consistent
-interlopers. We estimate that this effect comprimises the density
-profiles past \(\log R / {\rm arcmin} = 1.8\) for Sculptor and Ursa
-Minor, but agreement is good before then.
+However, some cosmological simulations have suggested that tidal effects
+may be ubiquitous. \citep{riley+2024, shipp+2023} find that a majority
+of satellites form substantial streams, and not due to resolution
+effects \citep[see also][]{panithanpaisal+2021}. This result appears to
+be in tension with ours and many previous works. Some reasons
+cosmological and idealized simulations may give opposite results could
+include resolution effects \citep[e.g.,][]{santos-santos+2025},
+additional perturbations from dark matter substructure, tidal
+``pre-processing'' by other infalling satellites, or the streams may be
+below detection limits \citep[e.g.,][]{shipp+2023}. More work is needed
+to understand why streams appear to form more readily in cosmological
+simulations than is observed in the Milky Way or through idealized
+simulations.
 
-\textbf{\emph{Gaia} systematics}. While \emph{Gaia} has shown excellent
-performance, some notable limitations may introduce problems in our
-interpretation and reliability of density profiles. Gaia systematics in
-proper motions and parallaxes are typically smaller than the values for
-sources of magnitudes \(G\in[18,20]\). Since we use proper motions and
-parallaxes as general consistency with the dwarf, and factor in
-systematic uncertainties in each case, these effects should not be too
-significant. However, the systematic proper motion uncertainties becomes
-the dominant source of uncertainty in the derived systemic proper
-motions of each galaxy (see
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
-
-\textbf{Completeness}. \emph{Gaia} shows high but imperfect
-completeness, particularly showing limitations in crowded fields and for
-faint sources (\(G\gtrapprox20\)). As discussed in
-\citet{fabricius+2021}, for the high stellar densities in globular
-clusters, the completeness relative to HST varies significantly with the
-stellar density. However, the typical stellar densities of dwarf
-galaxies are much lower, at about 20 stars/arcmin = 90,000 stars /
-degree, lower than the lowest globular cluster densities and safely
-below the crowding limit of 750,000 objects/degree for BP/RP photometry.
-In \citet{fabricius+2021}, for the lowest density globular clusters, the
-completeness down to \(G\approx 20\) is \(\sim 80\%\). Closely separated
-stars pose problems for Gaia's on-board processing, as the pixel size is
-59x177 mas on the sky. This results in a reduction of stars separated by
-less than 1.5'' and especially for stars separated by less than 0.6
-arcseconds. The astrometric parameters of closely separated stars
-furthermore tends to be of lower quality \citep{fabricius+2021}.
-However, even for the denser field of Fornax, only about 3\% of stars
-have a neighbour within 2 arc seconds, so multiplicity should not affect
-completeness too much (except for unresolved binaries). One potential
-issue is that the previous analyses do not account for our cuts on
-quality and number of astrometric parameters. These could worsen
-completeness, particularly since the BP-RP spectra are more sensitive to
-dense fields. In the appendix to this section, we test if magnitude cuts
-impact the resulting density profiles, finding that this is likely not
-an issue.
-
-\textbf{Structural uncertainties}. J+24 do not account for structural
-uncertainties in dwarfs for the two component case. We assume constant
-ellipticity and position angle. Dwarf galaxies, in reality, are not
-necessarily smooth or have constant ellipticity. J+24 test an
-alternative method using circular radii for the extended density
-component, and we find these density profiles are very similar to the
-fully elliptical case, even when assuming circular bins for the circular
-outer component. As such, even reducing the assumed ellipticity from
-\(0.37-0.55\) to 0 does not substantially impact the density profiles.
-
-\textbf{Outlook.} Finally, an excellent test of systematics and methods
-in \emph{Gaia} is to compare against another survey. In the appendix, we
-show that profiles derived from J+24 agree with DELVE or UNIONS data
-within uncertainties. Systematics, completion effects, and selection
-methods are unlikely to substantially change the density profiles
-presented here.
-
-\subsection{Comparison to literature}\label{comparison-to-literature}
-
-A number of works have previously speculated that dwarf galaxies such as
-Sculptor and Ursa Minor may have signatures of tidal perturbations,
-however not without contention. While \citet{hodge1961} and
-\citet{demers+krautter+kunkel1980} were some of the earliest work to
-derive the density profile of Sculptor, \citet{innanen+papp1979} was
-perhaps the first to speculate that Sculptor may harbour a substantial
-population of ``extratidal stars'' (stars beyond the tidal radius),
-finding candidate members out to 180' in an elongated distribution from
-\citet{vanagt1978}`s catalogue of variable stars.\footnote{Interestingly,
-  \citet{innanen+papp1979} also speculate about Uranus's satellite
-  distribution, covering gravitational dynamics at two very different
-  scales.} Additionally, \citet{eskridge1988} showed a possible excess
-of stars in Sculptor relative to a \citet{king1962} and Exponential
-density profile beginning around 50', but suggest that this excess may
-not be unusual. Later work by \citet{IH1995}, \citet{walcher+2003}, and
-\citet{westfall+2006} also showed evidence of an extended component of
-Sculptor's density profile (among other dwarfs), interpreting these
-stars as either evidence of tidal debris or a dwarf galaxy halo. On the
-other hand, \citet{coleman+dacosta+bland-hawthorn2005} show that
-Sculptor is well described by a two component density profile, they
-additionally mention that it is unlikely many of Sculptor's stars (less
-than 2\%) are extratidal. While \citet{munoz+2018} is the most recent
-and deepest photometric study of Sculptor, they only cover an area of
-out to radii of about 30', so they are unable to study the possible
-extended components of Sculptor detected in past works. In addition to
-the presence of ``extratidal stars'', many studies note that Sculptor's
-stars appear to become more elliptical with radius, consistent with
-tidal effects (\citet{IH1995}; \citet{westfall+2006}).
-
-Ursa Minor was often noted not for a density excess but for unusual
-features. Starting with the first density profile from
-\citet{hodge1964}, Ursa Minor is known to be highly elliptical. Studies
-by \citet{olszewski+aaronson1985}, \citet{demers+1995}, \citet{IH1995},
-\citet{kleyna+1998}, and \citet{bellazzini+2002} note that Ursa Minor
-appears to contain substructure along its major axis, but without strong
-interpretation for the causes. As one interpretation \citet{kleyna+2003}
-suggest that this feature is a long-lived star cluster residing in a
-cored dark matter halo to be dynamically stable. One of the first direct
-suggestions of tidal features came from \citet{martinez-delgado+2001},
-who find that stars extend far beyond the nominal tidal radius for Ursa
-Minor in the direction of the galaxies elongation, interpreting this as
-evidence for tidal effects. \citet{palma+2003} corroborate many of these
-earlier findings, showing further evidence for Ursa Minor's peculiar
-morphology including S-shaped contours, a possible extended extratidal
-population of stars, and a failure for a King density profile to
-adequately capture the actual stellar distribution. Each of these
-characteristics appear to indicate strong tidal disruption. Using
-velocity observations, \citet{pace+2014} additionally show that there
-are multiple components in spatial-velocity information. However,
-\citet{munoz+2018}'s more modern photometry show a more regular
-distribution of stars. Irregardless, Ursa Minor has had enough work
-suggesting peculiarities that a deeper investigation into the
-possibility of tidal effects is worthwhile.
-
-However, everything discussed so far is before \emph{Gaia}. Thus, the
-knowledge of the orbits of these systems is largely unknown and most
-surveys could only filter out stars using photometry. As a result, the
-tidal radius could only be guessed based on either the density profile
-or current distance of the dwarf galaxy. With \emph{Gaia} (discussed
-more below), recent work has used Bayesian frameworks to derive
-systematic proper motions of many dwarf galaxies (e.g. \citet{MV2020a})
-and filter away foreground stars using proper motions and parallax to
-detect distant members, and study the 6D internal kinematics of these
-galaxies (e.g. \citet{tolstoy+2023}). Most relevant here,
-\citet{jensen+2024} present a bayesian algorithm to determine likely
-members in \emph{Gaia} (described below), and detect extended secondary
-components for 9 dwarf galaxies, including Sculptor and Ursa Minor.
-\citet{sestito+2023a} and \citet{sestito+2023b} followed up a few of the
-most distant stars detected in \citet{jensen+2024} confirming that
-members exist in each galaxy as far ass 9-12 \(R_h\) from the centres.
-In this chapter, we discuss the origin and reliability of these
-detections, confirming that the density profiles are indeed robust.
-These density profiles then provide the scientific motivation to explore
-the tidal interpretation in the next chapters.
-
-\section{Limitations of simulations}\label{limitations-of-simulations}
-
-Variations to the potential of the inner disk (exclusion of a bar,
-spiral arms) should minimally affect our results as no orbit we consider
-reaches less than \textasciitilde15 kpc of the MW centre. We exclude the
-mass evolution of the halo from this analysis. Over \(10\,\)Gyr, this
-would be fairly significant (factor of \(\sim 2\)in MW mass, REF) but
-since we want to determine the upper limit of tidal effects, it is safe
-to neglect this.
-
-\subsection{Long term orbital history}\label{long-term-orbital-history}
-
-\citet{dsouza+bell2022}, \citet{santistevan+2024}. challenges to
-backwards time integration
-
-\subsection{Deviations from the NfW}\label{deviations-from-the-nfw}
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{dicintio+2013}
-\end{itemize}
-
-\section{Comparison to other work}\label{comparison-to-other-work}
+Closely related to our work, \citet{iorio+2019} applied idealized N-body
+simulations to study tidal effects on Scl. They similarly found weak
+tidal effects, even for a dark-matter-free model. However, their orbits
+are less well-constrained and they did not consider the LMC.
 
@@
 
-Theoretical work on Sculptor
+Scl has long been speculated to be disturbed. \citet{innanen+papp1979}
+found RR Lyrae candidate Scl members \citep[from][]{vanagt1978} out to
+180' in an elongated distribution, speculating this to be tidal
+disruption. Many later density profile determinations noted Scl's
+elongation and apparent outer density excess similar to J+24's detection
+\citep[but see also
+\citet{coleman+dacosta+bland-hawthorn2005}]{eskridge1988, IH1995, walcher+2003, westfall+2006}.
+Scl's ellipticity and extended density profile were often interpreted as
+evidence of tidal debris or sometimes a dwarf galaxy halo.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{battaglia+2008}
-\item
-  \citet{iorio+2019}
-\item
-  \citet{amorisco+zavala+deboer2014}
-\item
-  \citet{battaglia+2008}
-\item
-  \citet{breddels+2013}
-\item
-  \citet{breddels+helmi2013}
-\item
-  \citet{richardson+fairbairn2014}
-\item
-  \citet{SFW2017}
-\item
-  \citet{innanen+papp1979}
-\item
-  \citet{wilkinson+2002}
-\item
-  \citet{yang+2025}: chemical evolution in Scl.
-\item
-  \citet{skuladottir+2024}; \citet{skuladottir+2021};
-  \citet{lee+jeon+bromm2024}, pop III high res spectro?
-\item
-  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
-\item
-  \citet{tang+2023}: apogee modeling of Scl.
-\item
-  \citet{pandey+west2022} chem evo OMEGA and isotopes.
-\item
-  \citet{an+koposov2022}: distance gradients?
-\item
-  \citet{kawata+2006}, 2pop origins?
-\item
-  \citet{grcevich+putman2009}, \citet{carignan+1998} ambiguous HI clouds
-  near Scl ?
-\item
-  \citet{agnello+evans2012}, claim that Sculptor cannot hold two
-  populations in NFW halo?
-\end{itemize}
-
-Observational work on Scl
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{sestito+2023a}
-\item
-  \citet{westfall+2006} wide degree survey for extended structure.
-\item
-  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
-  \citet{arroyo-polonio+2024}
-\item
-  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
-\item
-  \citet{coleman+dacosta+bland-hawthorn2005}
-\item
-  \citet{DQ1994}
-\item
-  \citet{WMO2009}
-\item
-  \citet{IH1995},
-\item
-  \citet{munoz+2018}: Using Megacam to derive density profiles and
-  structural properties of many dwarf spheroidal galaxies.
-\item
-  \citet{kirby+2009}
-\item
-  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
-\item
-  \citet{grebel1996}
-\item
-  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
-  metallicity gradient and search for metal poor stars in Scl.
-\end{itemize}
-
-Future ideas:
-
-\begin{itemize}
-\tightlist
-\item
-  \citet{evslin2016}: measuring ellipticities of halos w TMT.
-\end{itemize}
+In addition, Scl hosts at least two populations, as revealed through
+photometry \citep{tolstoy+2004}, kinematics
+\citep{battaglia+2008, tolstoy+2023, arroyo-polonio+2024}, dynamical
+structure \citep{breddels+helmi2014}, age gradients \citep{deboer+2011},
+and chemistry \citep{kirby+2009}. These two populations both have radii
+smaller than the outer density excess. Scl challenges the idea of a
+simple, single-component dwarf spheroidal population and hints at
+episodic star formation or hierarchical assembly.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{sestito+2023b}
-\item
-  \citet{pace+2020}
-\item
-  \citet{pace+2014} substructure
-\item
-  \citet{bellazzini+2002}
-\item
-  \citet{hargreaves+1994}
-\item
-  \citet{martinez-delgado+2001}
-\item
-  \citet{munoz+2005}: velocity dispersion profile
-\item
-  \citet{palma+2003}
-\item
-  \citet{spencer+2018}
-\item
-  \citet{vitral+2023}
-\item
-  \citet{piatek+2005}: old HST Pm meas.
-\item
-  \citet{gallagher+2003}: no ionizaed gas.
-\item
-  \citet{shetrone+cote+stetson2001}: RGBs
-\item
-  \citet{wilkinson+2004}
-\item
-  \citet{kleyna+2003}: dynamical fossil.
-\item
-  \citet{pryor+kormendy1990}, \citet{lake1990} MLR / velocity dispersion
-\item
-  \citet{armandroff+olszewski+pryor1995};
-  \citet{olszewski+pryor+armandroff1996} MLR
-\item
-  \citet{aaronson+olszewski+hodge1983}; \citet{aaronson1983} carbon
-  stars and velocities
-\item
-  \citet{stetson1984} early spectroscopy.
-\item
-  \citet{tsujimoto+shigeyama2002} chem evo.
-\end{itemize}
+UMi has garnered claims of inner substructure such as stellar or
+kinematic ``clumps''. Studies by \citet{olszewski+aaronson1985},
+\citet{demers+1995}, \citet{IH1995}, \citet{kleyna+1998},
+\citet{battinelli+demers1999}, and \citet{bellazzini+2002} note that
+Ursa Minor appears to contain ``clumps'' along its major axis. One clump
+has furthermore been shown to be kinematically distinct and colder
+\citep[e.g.,][]{pace+2014}. If a star cluster, than the cluster should
+disolve in \(\sim 3\) Gyrs provided the UMi's halo is not cored and did
+not interact with dark (sub)subhalos---the survival of such substructure
+depends on the nature of dark matter \citep{kleyna+2003, lora+2012}.
+\citet{wilkinson+2004} additionally find a puzzling drop in the velocity
+dispersion in the outskirts. The nature of any clumps or substructure
+remains unclear, but \citet{munoz+2018}'s are unable to find signs of
+any substructure with modern, deeper photometry.
 
-Theoretical work
+In addition, several works have found supposed tidal features in UMi.
+\citet{hargreaves+1994} first detected a velocity gradient in UMi,
+suggestive of tidal disruption. Later, \citet{martinez-delgado+2001}
+find that stars extend far beyond the ``tidal radius'' (from a King
+profile fit) for Ursa Minor, in the direction of the galaxy's elongation
+\citep[see corresponding simulations
+by][]{gomez-flechoso+martinez-delgado2003}. \citet{palma+2003} further
+showed evidence for S-shaped contours, and an extended population of
+``extratidal'' stars. Our density profiles are consistent with these
+works. However, strong evidence of tidal disruption has not yet been
+found.
 
-\begin{itemize}
-\tightlist
-\item
-  robles+bullock2021
-\item
-  \citet{caproni+lanfranchi2021}, \citet{caproni+2015}: simulation of
-  gass loss.
-\item
-  \citet{bajkova+bobylev2017}, local orbits
-\item
-  \citet{gomez-flechoso+martinez-delgado2003}: MLR of UMi
-\item
-  \citet{lynden-bell1976} early discussion of orbits and LMC plane.
-\end{itemize}
+\section{Forming a stellar halo}\label{sec:stellar_halos}
 
-\section{Are Sculptor and Ursa Minor
-typical?}\label{are-sculptor-and-ursa-minor-typical}
+As a tidal origin of Scl and UMi's extended density profiles is
+disfavoured, we consider alternative processes which may form a dwarf
+galaxy ``stellar halo''. Some recent works have also confirmed the
+presence of extended, likely non-tidal, density profiles. For example,
+\citet{chiti+2021}; \citet{chiti+2023} spectroscopically confirm members
+out to \(~9 R_h\) in Tucana II. Given these stars are misaligned with
+the orbit and the lack of a velocity gradient, tides seem to be an
+unlikely explanation for Tucana II as well. The prevalence of stellar
+halos around dwarf galaxies remains an active research topic.
 
-\subsection{The formation of Exponential
-profiles}\label{the-formation-of-exponential-profiles}
+Many dwarf galaxies also host multiple chemodynamical stellar
+populations. Typically, older populations are extended, kinematically
+hotter, and metal-poo , whereas the younger populations are more
+compact, colder, and metal-rich. Examples include Carina
+\citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
+Fornax \citep[\citet{amorisco+evans2012},
+\citet{delpino+aparicio+hidalgo2015}]{battaglia+2006}, Sextans
+\citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
+Andromeda II
+\citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}. Evidence
+of multiple-populations in dwarf galaxies suggests that effects like
+galaxy mergers or episodic star formation shape the stellar structure of
+dwarf galaxies.
 
-As mentioned earlier, many dwarf galaxies appear to have exponential
-stellar density profiles.
+We now explore possible origins of a ``stellar halo'' which may resemble
+the extended density profiles in Scl and UMi.
 
-Disk galaxies have long been known to be exponential-like across a wide
-range of scales. As such, several theoretical works have aimed to
-undertsand the formation of exponential stellar disks. One prevailing
-theory posits that scattering of stars in a disk naturally forms an
-exponential \citep[\citet{wu+2022}]{elmegreen+struck2013}, or that the
-disk is due to conservation of angular momentum during spherical
-collapse.
+\subsection{Internal processess}\label{internal-processess}
 
-Given that dwarf galaxies were noticed to be exponential like above,
-\citet{faber+lin1983} proposed that dwarf spheroidals formed from disky
-galaxies maintaining the typical exponential density profile
+\textbf{Dynamical heating of old stars}. In a galaxy, older stars
+generally have higher random velocities (i.e., kinematically ``hotter'')
+than younger stars. In dwarf galaxies, several mechanism may heat
+stellar components, including supernovae kicks in forming stars,
+feedback-driven potential fluctuations
+\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021}, or
+heating by dark sub-subhalos \citep{penarrubia+2025}. Over time, these
+processes naturally produce a more extended, older stellar population.
 
-\citet{read+gilmore2005} propsed that exponential dSph form through
-impusive mass loss, redistributing the stellar component into an
-exponential like profile.
+\textbf{Episodic star formation and feedback.} Dwarf galaxies are
+thought to experience bursty star formation--i.e., consisting of several
+episodes of intense star formation separated by periods of quiescence
+\citep[e.g.,][]{salvadori+ferrara+schneider2008, valcke+derijcke+dejonghe2008, wheeler+2019, azartash-namin+2024}.
+Stellar feedback in a dwarf galaxy's shallow potential well drives
+oscillations in the star formation rate. Alternatively, re-ionization
+may temporarily suspending star formation \citep{benitez-llambay+2015}.
+In many simulations, shrinking gas reservoirs form
+centrally-concentrated later generations of stars, naturally spawning
+multiple populations \citep{kawata+2006, revaz+jablonka2018}. However,
+star formation is also sensitive to a dwarf's environment.
 
-\begin{itemize}
-\tightlist
-\item
-  disk galaxy formation \citet{fall+efstathiou1980}, \citet{mestel1963}
-\item
-  valcke+derijcke+dejonghe2008 with early idealized hydro simulations
-  show Sérsic values from simulations between 0.8 and 1 for dSph,
-  indicating a natural formation of Exponential or slightly steeper
-  density profiles.
-\end{itemize}
+\subsection{External processes}\label{external-processes}
 
-One possible explanation is that \citet{mayer+2001a}
+\textbf{Induced star formation.} Star formation may quench and reignite
+due to an external perturbation. Examples include tidal compression
+\citep{mayer+2001a, dong+lin+murray2003}, gaseous filaments
+\citep{genina+2019}, dark halos \citep{starkenburg+helmi+sales2016}, or
+shocks with the MW corona \citep{wright+2019}. Induced burst would be
+more stochastic than internal bursts, possibly explaining diversity in
+dwarf galaxy structure.
 
-Other explanations range from the effects of mass loss, feedback,
-angular momentum,
+\textbf{Major mergers.} When galaxies merge, they may leave signatures
+such as population gradients and stellar halos. Classical dwarfs have a
+\(\sim 10\%\) chance of undergoing a major merger since redshift \(z=1\)
+\citep{deason+wetzel+garrison-kimmel2014}. Mergers may preferentially
+disperse the lower-mass galaxy's stars, forming a halo and a metallicity
+gradient \citep{benitez-llambay+2016}. Intermediate-mass mergers (with
+mass-ratios of \(\sim\) 1:5) most effectively populate halos, balancing
+stellar mass from larger galaxies with better dispersion of lower-mass
+galaxies \citep{deason+2022}. Tuc II's properties are suggested to
+originate from a similar merger
+\citep{tarumi+yoshida+frebel2021, querci+2025}.
 
-\citet{klimentowski+2007}, \citet{klimentowski+2009}.
+\textbf{Gas-rich mergers.} If a merger occurs between gas-rich galaxies,
+triggered star formation may occur in the aftermath
+\citep[e.g.,][]{genina+2019}. And II and Phoenix have steep metallicity
+gradients and unusual prolate rotation, theorized to result from mergers
+of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.
 
-In summary, while there is some theoretical reasons why exponential
-spheroidal stellar profiles may form, the emperical sucess of an
-exponential law still requires explanations. Deviations from this
-exponential thus test wether this exponential law is indeed
-near-universal or if other ingredients in the formation and history of
-dwarf galaxies may be responsible for changes from the empirical
-expectation.
+\textbf{Minor mergers / accretion}. Like how the Milky Way's halo is
+believed to be built from many minor mergers, dwarf galaxies may form
+halos through accretion of yet fainter dwarfs.
+\citet{ricotti+polisensky+cleland2022} demonstrated that accretion of
+ultra-faint dwarfs may create dwarf stellar halos. A stream detected
+around And II further supports the occurrence of mergers among dwarfs
+\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
+the halo chemistry would resemble a population of ultra-faint dwarfs.
+The occurance of dwarf mergers would further constrain small-scale
+galaxy formation.
 
-\subsection{Alternatives to exponential
-profiles}\label{alternatives-to-exponential-profiles}
+\textbf{Tidal preprocessing}. Dwarf galaxies may have been
+``preprocessed'' by a larger satellite like the LMC
+\citep[e.g.,][]{santistevan+2023, riley+2024}. From the orbit
+integrations above, it is possible that UMi was once an LMC satellite.
+However, the prevalence of preprocessing remains uncertain. Like stellar
+heating, preprocessing redistributes already-present stellar
+populations.
 
-\begin{itemize}
-\tightlist
-\item
-  elmegreen+hunter2006: Discuss the formation of double exponentials in
-  disk galaxies. While a simpler idealized / Semi-analytic model of star
-  formation, show that assuming exponentail gas and KS-relation with SF
-  threshhold, that reduction of turbulance (driven by a number of
-  processes) in outer disk leads to a steeper decline in SFR as crossing
-  the critical SF threshold becomes less likely.
-\end{itemize}
+\textbf{Tidal dwarf galaxies} are cluster-like objects which form in
+gas-rich tidal streams created during the merger of two galaxies
+\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Tidal
+dwarfs may be more susceptible to tides, forming extended density
+profiles and appearing to have dark matter
+\citep{casas+2012, yang+2014, wang+2024a}. If Scl and UMi are tidal
+dwarfs, a stronger velocity (dispersion) gradient and tidal tails should
+be detectible.
 
-\section{Understanding the extended density profiles of dwarf
-galaxies}\label{understanding-the-extended-density-profiles-of-dwarf-galaxies}
+\subsection{\texorpdfstring{Beyond \LCDM{}}{Beyond }}\label{beyond}
 
-Are dwarf galaxies indeed expected to be one-component exponential-like
-density profiles? The suggestion of exponential density profiles dates
-back to \citet{faber+lin1983}. Expand\ldots{}
+\textbf{Modified Newtonian Dynamics (MOND)}. MOND modifies gravity
+instead of using dark matter to explain the rotation curves of galaxies.
+In MOND, tides may more strongly affect dwarf galaxies owing to the lack
+of dark matter mass and the stronger MW tidal field
+\citep{mcgaugh+wolf2010, brada+milgrom2000}. A tidal origin of the
+density excess is more likely in this case. If MOND is to recover the
+observed velocity dispersions of satellites, then many more dwarfs may
+be actively tidally disrupting as well
+\citep{casas+2012, yang+2014, wang+2024a}.
 
-A number of recent work has also confirmed the presence of extended,
-likely non-tidal, density profiles. For example, \citet{chiti+2021};
-\citet{chiti+2023} spectroscopically confirm members out to \(~9 R_h\)
-in Tucana II.
+\textbf{Self-interacting dark matter (SIDM).} SIDM significantly
+complicates tidal evolution. SIDM halos in isolation are not static---by
+transferring heat through collisions, these halos first undergo core
+formation and then core collapse. Besides structural changes, SIDM adds
+that pressure from the host DM halo can change the structure
+(e.g.~aiding core collapse) or remove material from the inner galaxy
+(analogous to ram pressure stripping) \citep[e.g.,][]{cartonzeng+2024}.
+An SIDM halo may be more strongly affected by tides, but the velocity
+dispersion makes a tidal disruption in SIDM still unlikely.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{revaz+jablonka2018}
-\item
-  \citet{tau+vivas+martinez-vazquez2024} (RRL stars )
-\item
-  \citet{roderick+2016} Boötes I
-\item
-  \citet{mcconnachie+penarrubia+navarro2007} (velocity space \& multiple
-  populations).
-\item
-  \citet{cicuendez+battaglia2018}
-\end{itemize}
+\textbf{Fuzzy dark matter} can also heat up stars owing to density
+perturbations in the form of interference fringes
+\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. Similar to other
+intrinsic heating methods, this model would likely affect all dwarfs
+similarly, so we should detect similar halos around similar dwarfs.
 
-\subsection{Multi-epoch star formation and
-feedback}\label{multi-epoch-star-formation-and-feedback}
+\subsection{Disentangling the origin of a stellar
+halo.}\label{disentangling-the-origin-of-a-stellar-halo.}
 
-The star formation history of dwarf galaxies is typically thought to be
-\emph{bursty}: a series of discrete episodes of intense star formation
-separated by periods of quiescence.
+We just reviewed a number of different, possibly-concurrent explanations
+for the formation of a stellar halo. While we leave the nature of Scl
+and UMi's extended stellar densities an open question, we can discuss
+possible clues to different formation scenarios.
 
-For example, in the simulations of \citet{wheeler+2019}, etc., dwarf
-galaxies exhibit strongly bursty SFHs.
+\textbf{Chemistry}. Large samples of detailed chemical abundances have
+been invaluable for understanding MW substructure (e.g.~for
+\emph{Gaia}-Sausage Enceladus). Chemistry, particularly comparing the
+inner and outer regions could test if the halo appears to originate from
+a distinct system than the dwarf, differentiating many \emph{internal}
+versus \emph{external} scenarios.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{maxwell+2012}
-\item
-  \citet{wright+2019}
-\item
-  \citet{azartash-namin+2024}
-\end{itemize}
+\textbf{Star formation histories}. Evidence of significant star
+formation episodes or lack thereof may differentiate scenarios which
+rely on strong star formation bursts (e.g.~\emph{episodic star formation
+history, induced star formation, and gas-rich mergers.})
 
-\subsubsection{Multi-component density
-profiles}\label{multi-component-density-profiles}
+\textbf{Kinematics.} Particularly, for models relying on recent tidal
+disruption, kinematic disequilibrium features should be visible. These
+would appear as velocity gradients, increasing velocity dispersions,
+outward-biased moving stars, or non-phase mixed structures
+\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.
 
-A related note is that several dwarf galaxies show evidence for multiple
-chemodynamical components in their stars. While spectra tend to focus on
-the inner regions (high probability members), the existence of
-multiple-component stellar populations in the inner regions of dwarf
-galaxies hints at a complex star formation history capable of creating
-the observed profiles.
+\textbf{Deep photometry} may find or rule out signs of dynamical
+disequilibrium and tidal tails for tidally susceptible models
+\emph{(e.g., MOND and tidal dwarfs.)}
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{benitez-llambay+2016}
-\item
-  \citet{mercado+2021}: formation of metallicity gradients from SFH
-\item
-  \citet{revaz+jablonka2018} natural formation of multi-component and
-  gradients from cosmological simulations of dwarf galaxies. Originate
-  from dynamical heating of metal rich population and outside in star
-  formation?
-\item
-  \citet{el-badry+2016}. Feedback drives fluctuations in size + radial
-  migration =\textgreater{} population gradients.
-\item
-\end{itemize}
+Ongoing, upcoming, and future facilities will be essential for testing
+these theories. For example, given the typical faint magnitudes of dSph
+stars, chemistry is best accessible through large field-of-view
+multi-object spectroscopic instruments on large telescopes (e.g., 4MOST
+and extremely large telescopes \citet{skuladottir+2023}). 3D internal
+kinematics of dwarfs may require a successor to \emph{Gaia}. However,
+using JWST and HST enables precise proper motions for small regions
+within dwarf galaxies \citep[e.g.,][]{vitral+2025}, which may be
+sufficient to constrain vastly different internal dynamic structures and
+ongoing tidal disruption. Finaly, Photometric surveys by Rubin
+Observatory and Euclid will most-emminently probe yet fainter magnitudes
+around dwarf galaxies and possibly find or constrain stellar halos and
+their star formation histories. Altogether, new surveys will likely
+uncover novel aspects about the inner workings and outer halos of dwarf
+galaxies.
 
-Observational evidence:
+In addition to new observing facilities, the next generation of dwarf
+simulations are beginning to answer questions and unravel processes in
+the formation of dwarf galaxies. With improved resolution, realism, and
+physics, the frequency and effects of various mechanisms discussed can
+be better constrained.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{arroyo-polonio+2024}
-\item
-  \citet{pace+2020}
-\item
-  \citet{fabrizio+2016}, \citet{kordopatis+2016} (multi-component in
-  Carina)
-\item
-  \citet{battaglia+2006}, \citet{amorisco+evans2012} for fornax
-\end{itemize}
+\section{Conclusion and outlook}\label{conclusion-and-outlook}
 
-\subsection{Past mergers and
-accretion}\label{past-mergers-and-accretion}
+In this thesis, we have investigated the extended, outer density excess
+of Sculptor and Ursa Minor and its possible tidal origin. We first
+verified that Scl and UMi have unusually extended density profiles,
+compared to other dwarf spheroidals. We show that the density profiles
+are robust to alternate data criteria, implying that this ``density
+excess'' is likely a real feature of each galaxy.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{deason+2014}
-\item
-  \citet{deason+2022}
-\item
-  \citet{ricotti+polisensky+cleland2022}
-\item
-  \citet{querci+2025}
-\item
-  \citet{amorisco+evans+vandeven2014}
-\item
-  \citet{tarumi+yoshida+frebel2021}: Shows cosmological simulations in
-  context of Tuc II, merger of two galaxies in the early universe
-  produces a more extended density profile and metallicity gradient.
-\item
-  \citet{lokas+2014}: And II from merger,
-\item
-  rotation from merger: \citet{cardona-barrero+2021}
-\end{itemize}
+We then investigated if tides were a permissible explanation. By
+modelling each galaxy based on cosmological initial conditions, we
+showed that tides do not strongly affect either galaxy. The LMC changes
+the orbital history of Scl and UMi, and tides become even weaker in a
+combined LMC and MW potential. While UMi may form a stellar stream, the
+stream is far fainter than is presently detectible. We conclude that
+recent tides are unlikely to shape the observed stellar distributions of
+Scl and UMi.
 
-\subsection{Preprocessing}\label{preprocessing}
-
-Another alternative is that the dwarf galaxies have been
-``preprocessed'' by another dwarf/LMC like galaxy earlier on and this
-has resulted in their unique properties
-
-\subsection{Tidal dwarfs}\label{tidal-dwarfs}
-
-On e possibility is that Sculptor and Ursa Minor are not formed from
-dark matter halos but instead due to the collision of gas. This
-possibility is not a convincing explanation because we know the star
-formations histories are extended and relatively old, and the galaxies
-are close to apocentre, making achieving the observed velocity
-dispersion challenging in this framework.
-
-\subsection{\texorpdfstring{Alternatives to
-\(\Lambda\)CDM}{Alternatives to \textbackslash LambdaCDM}}\label{alternatives-to-lambdacdm}
-
-Alternatives to the prevailing cosmology we assume may help explain the
-nature of these dwarf galaxies. t
-
-\begin{itemize}
-\tightlist
-\item
-  WDM: forms cores, doesn't help
-\item
-  SIDM: can cause faster disruption, doesn't fix orbital path or change
-  tidal effects?
-\item
-  MOND: \citet{sanchez-salcedo+hernandez2007}: mond in dsph
-\end{itemize}
+Finally, we consider alternative scenarios forming extended density
+distribution. We review mechanisms ranging from episodic star formation
+histories, to accretion, to departures from \LCDM{}. While a more
+precise explanation awaits, upcoming and ongoing surveys will uncover
+the stellar populations within dwarf galaxies in unprecedented detail.
+Future work will the Milky Way, uncovering the cosmic origins of the
+smallest structures forming building blocks of galaxies like ours.

### 2025-09-27 thesis/extra_density.tex
a compltet thesis draft

\chapter{The reliability of derived density
profiles}\label{sec:extra_density}

In this Appendix, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles of the dwarfs,
comparing methodologies, samples, and to literature. In conclusion, the
density profiles from J+24 converge among methods out to a
background-limited radius, derived here.

\section{Bayesian membership
probabilities}\label{bayesian-membership-probabilities}

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree circular region centred on the dwarf satisfying:

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegren+2021}.\\
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance,
  with \citet{lindegren+2021}'s zero-point correction. Note that the
  typical satellite parallax is very small.
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). This corresponds to tangental velocities
  of \(\gtrsim 500\) km/s at distances larger than 10 kpc.
\item
  Corrected colour excess is within 3\(\sigma\) of the expected
  distribution from \citet{riello+2021}. Removes stars with unreliable
  photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\sigma_{\rm DM}\).
\item
  De-reddened colour is between \(-0.5 < {\rm BP - RP} <  2.5\). Removes
  stars substantially outside the expected CMD.
\end{itemize}

Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
extinction maps.

J+24 define likelihoods \({\cal L}\) measuring consistency with the MW
stellar background (\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). The total likelihoods are the product of a
spatial, PM, and CMD term: \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

Each likelihood is normalized over their respective 2D parameter space
for both the satellite. To control the relative frequency of member and
background stars, \(f_{\rm sat}\) representing the fraction of member
stars in the field. The total likelihood for any star in this model is
the \(f_{\rm sat}\)-weighted sum of the foreground and background,
\begin{equation}\protect\phantomsection\label{eq:Ltot}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} Finally, the probability that any star belongs to the
satellite is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

For the satellite's spatial likelihood, J+24 consider both one-component
and a two-component density models. The one component model is
constructed as a single exponential profile ( surface density
\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\).
\(R_s\) is fixed to the equivalent \(R_h\) value from
\citet{munoz+2018}'s Sérsic fit. Additionally, structural uncertainties
(for position angle, ellipticity, and scale radius) are sampled over to
construct the final likelihood map. The two-component model instead adds
a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are free parameters.
Structural property uncertainties are not included in the two-component
model.

The PM likelihood is a bivariate gaussian with variance and covariance
equal to each star's proper motions. J+24 assume the stellar PM errors
are the main source of uncertainty.

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age (except 2 Gyr is used for Fornax). The isochrone is given a
Gaussian colour width of 0.1 mag plus the \emph{Gaia} colour uncertainty
at a given magnitude. The horizontal branch is modelled as a constant
magnitude extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB
stars and a 0.1 mag width plus the mean colour error). A likelihood map
is constructed by sampling the distance modulus in addition to the CMD
width, taking the maximum of RGB and HB likelihoods.

The background likelihoods are instead empirically constructed. Stars
stars outside of 5\(R_h\) passing the quality cuts estimate the
background density in PM and CMD space. The density is a sum of
bivariate gaussians with variances based on Gaia uncertainties (and
covariance for proper motions). The spatial background likelihood is
assumed to be constant.

J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
informative. The proper motion single component prior is same as
\citet{MV2020a}: a normal distribution with mean 0 and standard
deviation \(100\ \kms\). If using the 2-component model, the prior is
instead a uniform distribution spanning 5\(\sigma\) of single component
case w/ systematic uncertainties. \(f_{\rm sat}\) (and \(B\)) has a
uniform prior from 0 to 1. \(R_{\rm outer}\) has a uniform prior only
restricting \(R_{\rm outer} > R_s\). The mode of each parameter from the
MCMC are then reported and used to calculate the final \(P_{\rm sat}\)
values.

\section{Alternative methodologies}\label{alternative-methodologies}

\subsection{Simplistic cuts}\label{simplistic-cuts}

Besides a global background subtraction (as used in
Section~\ref{sec:observations}), absolute cuts in CMD and PM represent
perhaps the next simplest method to construct density profiles. We
create cuts using only the following

\begin{itemize}
\tightlist
\item
  (BP-RP, G) within the vertices defined in Table~\ref{tbl:colour_cuts}
\item
  PM within a \(1\masyr\) radius of the galaxies measured value.
\item
  Parallax within \(3\sigma\) of the dwarf distance
\item
  \texttt{ruwe} \textless{} 1.3
\end{itemize}

\begin{table*}[t]
\centering
\caption[Colour cuts for density profiles.]{Colour cuts used in samples in this section. Each pair of columns contains the vertices of a polygon in the CMD within which we select stars to derive density profiles. }
\label{tbl:colour_cuts}
\begin{tabular}{llllllll}
\toprule
Scl &  &  &  & UMi &  &  & \\
\midrule
Gaia &  & DELVE &  & Gaia &  & Unions & \\
BP-RP & G & g-r & g & BP-RP & G & r-i & r\\
-0.21 & 20.58 & 0.32 & 22.58 & 0.49 & 20.83 & -0.485 & 24.13\\
-0.1 & 20.1 & 0.51 & 23.07 & 1.24 & 20.9 & 0.145 & 23.11\\
0.28 & 19.77 & 0.56 & 21.49 & 1.18 & 19.65 & 0.201 & 21.41\\
0.7 & 19.64 & 0.71 & 19.7 & 1.23 & 18.12 & 0.21 & 20.05\\
1.02 & 18.5 & 0.96 & 18.49 & 1.39 & 17.09 & 0.288 & 17.23\\
1.14 & 17.59 & 0.84 & 18.35 & 1.32 & 16.94 & 0.226 & 17.21\\
1.41 & 16.73 & 0.6 & 19.65 & 0.96 & 18.58 & 0.151 & 19.3\\
1.68 & 15.94 & 0.42 & 21.42 & 0.88 & 19.22 & 0.145 & 19.75\\
1.94 & 15.94 &  &  & 0.4 & 19.52 & 0.026 & 19.86\\
1.91 & 16.8 &  &  & -0.03 & 19.78 & 0.011 & 19.52\\
1.68 & 17.01 &  &  & 0.06 & 20.16 & -0.155 & 19.95\\
1.45 & 17.87 &  &  & 0.83 & 19.72 & -0.22 & 20.34\\
1.32 & 19.16 &  &  & 0.21 & 20.83 & -0.176 & 20.47\\
1.26 & 20.1 &  &  &  &  & -0.098 & 20.33\\
1.35 & 21 &  &  &  &  & 0.008 & 19.84\\
0.46 & 21 &  &  &  &  & 0.142 & 19.75\\
0.82 & 20.27 &  &  &  &  & 0.098 & 21.45\\
0.39 & 20.33 &  &  &  &  & -0.052 & 22.38\\
0.16 & 20.76 &  &  &  &  & -0.388 & 23.42\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Alternative samples}\label{alternative-samples}

For the DELVE sample \citep{drlica-wagner+2022}, we select everything in
data release 2 between RA of 11 and 19, and declination of -37.7 and
-29.7.

We also consider samples from the Ultraviolet Near-Infrared Optical
Northern Survey \citep[UNIONS][]{gwyn+2025} for Ursa Minor. We select
stars within a 5 degree tangent square of UMi with no
\texttt{FLAGS\_CFIS} set and \texttt{s21} and \texttt{s31} both \(<3\)
(not extended sources).

\subsection{A Bayesian histogram
model}\label{a-bayesian-histogram-model}

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin. This model extends
the J+24 framework with two notable differences. (1) The systemic proper
motions are held fixed. This allows for a much more efficient
calculation of the likelihoods. (2) The spatial likelihood for the
satellite is a piecewise constant density function.While this model
theoretically has many more parameters to fit (one for each bin), each
parameter is independently estimated based on only the stars in that
specific bin.

We segregate the data into radial bins and independently fit the model
to only stars in each bin. This fits many small models instead of a
global large model. In detail, this means that, in each bin, the
normalized density of the background and satellite are \emph{equal}. The
only fit parameter is the \(f_{\rm sat}\) in the particular bin.
Therefore, the likelihood in each bin is the same as above except
\({\cal L}_{\rm space} = 1\) for the foreground and background.

In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
satellite density. The model construction is then \begin{equation}{
\theta_i \equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}})
}\end{equation}

\begin{equation}{
f_{\rm sat} = \frac{10^{\theta_i}}{1 + 10^{\theta_i}}
}\end{equation}

\begin{equation}{
\theta_i \sim {\rm Uniform}(-12, 6)
}\end{equation}

\begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM}
}\end{equation}

For each satellite, we hold the structural parameters fixed to those in
J+24, create bins which are the larger of the bin containing 20 stars or
0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
with 48 realizations using Turing.jl and the No-U Turns Sampler. The
chains appear to converge excellently.

Based on the \(f_{\rm sat}\) values in each bin, the number of members
is just the number of stars in the bin times \(f_{\rm sat}\) for the
bin, allowing the direct derivation of the density profiles from this
model.

We have also briefly explored a model which solves for \(f_{\rm sat}\)
for random realizations of the global structural uncertainties. These do
not appear to influence the results substantially. Another possible
inconsistency is that the half-light radii are estimated from a
different sample. We do briefly rederive structural parameters for the
sample using Sérsic fits (although with a fixed origin).

\section{Comparison of density profiles}\label{sec:density_extra}

In this section, we discuss additional tests and verification of the
derived density profiles. In particular, we check that methodology
(simpler cuts, circularized radii, algorithm version) do not
substantially affect the density profile. We also compile density
profiles presented in the literature as reference. In all cases, the
density profiles appear to have excellent convergence out to
\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
where the background dominates.

\textbf{Spatial likelihood.} J+24 method was designed in particular to
detect the presence of a density excess and find individual stars at
large radii to be followed up. We are more interested in accurately
quantifying the density profile and size of any perturbations. One
potential problem with using J+24's candidate members is that the
algorithm assumes the density is either described by a single or double
exponential. If this model does not accurately match the actual density
profile of the dwarf galaxy, we want to understand the impact of this
assumption.In particular, in Fig.~\ref{fig:scl_observed_profiles},
notice that the \(P_{\rm sat}\) selection method produces small
errorbars, even when the density is more than 1 dex below the local
background. These stars are likely selecting stars from the statistical
MW background consistent with UMi PM+CMD, recovering the assumed density
profile. As a result, the reliability of these density profiles below
the CMD+PM background may be questionable. A more robust analysis,
removing this particular density assumption, would be required to more
appropriately represent the knowledge of the density profile as the
background begins to dominate.

\textbf{\emph{Gaia} systematics}. While \emph{Gaia} has shown excellent
performance, some notable limitations may introduce problems in our
interpretation and reliability of density profiles. Gaia systematics in
proper motions and parallaxes are typically smaller than the values for
sources of magnitudes \(G\in[18,20]\). Since we use proper motions and
parallaxes as general consistency with the dwarf, and factor in
systematic uncertainties in each case, these effects should not be too
significant. However, the systematic proper motion uncertainties becomes
the dominant source of uncertainty in the derived systemic proper
motions of each galaxy (see
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).

\textbf{Completeness}. \emph{Gaia} shows high but imperfect
completeness, particularly showing limitations in crowded fields and for
faint sources (\(G\gtrapprox20\)). As discussed in
\citet{fabricius+2021}, for the high stellar densities in globular
clusters, the completeness relative to HST varies significantly with the
stellar density. However, the typical stellar densities of dwarf
galaxies are much lower, at about 20 stars/arcmin = 90,000 stars /
degree, lower than the lowest globular cluster densities and safely
below the crowding limit of 750,000 objects/degree for BP/RP photometry.
In \citet{fabricius+2021}, for the lowest density globular clusters, the
completeness down to \(G\approx 20\) is \(\sim 80\%\). Closely separated
stars pose problems for Gaia's on-board processing, as the pixel size is
59x177 mas on the sky. This results in a reduction of stars separated by
less than 1.5'' and especially for stars separated by less than 0.6
arcseconds. The astrometric parameters of closely separated stars
furthermore tends to be of lower quality \citep{fabricius+2021}.
However, even for the denser field of Fornax, only about 3\% of stars
have a neighbour within 2 arc seconds, so multiplicity should not affect
completeness too much (except for unresolved binaries). One potential
issue is that the previous analyses do not account for our cuts on
quality and number of astrometric parameters. These could worsen
completeness, particularly since the BP-RP spectra are more sensitive to
dense fields.

In Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we
show the results of limiting the density profile to the brightest half
of stars. We find no substantive differences. If \emph{Gaia}'s
incompleteness is inhomogeneous, then the variations in \emph{Gaia}'s
completeness do not strongly affect the density profiles.

\textbf{Structural uncertainties}. J+24 do not account for structural
uncertainties in dwarfs for the two component case. We assume constant
ellipticity and position angle. Dwarf galaxies, in reality, are not
necessarily smooth or have constant ellipticity. J+24 test an
alternative method using circular radii for the extended density
component, and we find these density profiles are very similar to the
fully elliptical case, even when assuming circular bins for the circular
outer component. As such, even reducing the assumed ellipticity from
\(0.37-0.55\) to 0 does not substantially impact the density profiles.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
\caption[Scl density comparison]{Density profiles for various
assumptions for Sculptor. PSAT is our fiducial 2-component J+24 sample,
circ is a 2-component bayesian model assuming circular radii, simple is
the series of simple cuts described, bright is the sample of the
brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
(background subtracted and rescaled to
match).}\label{fig:scl_density_extras}
\end{figure}

Comparison of density profiles for each J+24 method. The fiducial is a
2-component elliptical model. However, the 1-component is still
elliptical but only contains 1 component and the circular model assumes
a circular outer density profile and bins in circular bins instead of
elliptical bins.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Scl literature density profiles]{A comparison of the Scl
derived density profile and historical works.}
\end{figure}

\subsection{Ursa Minor}\label{ursa-minor}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
\caption[UMi density comparison]{Similar to
Fig.~\ref{fig:scl_observed_profiles} except for Ursa
Minor.}\label{fig:umi_density_extras}
\end{figure}

\begin{table*}[t]
\centering
\caption[Properties of probabilistically-derived density profiles]{For each classical dwarf, we have: the BG-limited radius $R_{\rm limit}$, where the density of stars is no longer reliably derived, the approximate number of member stars in Gaia, and the derived Sérsic indices from the density profiles. }
\label{tbl:mcmc_props}
\begin{tabular}{lllll}
\toprule
Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$ & num stars & $n$\\
\midrule
Fornax & 5.25 & 79.1 & 23154 & $0.794^{+0.034}_{-0.018}$\\
Sculptor & 6.39 & 64.1 & 7024 & $1.281^{+0.04}_{-0.04}$\\
Leo I & 4.23 & 13.5 & 1242 & $0.719^{+0.059}_{-0.053}$\\
Ursa Minor & 6.42 & 86.4 & 2314 & $1.425^{+0.10}_{-0.093}$\\
Leo II & 3.63 & 8.76 & 347 & $0.707^{+0.11}_{-0.092}$\\
Carina & 4.16 & 33.3 & 2389 & $1.025^{+0.08}_{-0.076}$\\
Draco & 3.59 & 27.3 & 1781 & $1.106^{+0.084}_{-0.078}$\\
Canes Venatici I & 1.95 & 12.5 & 156 & $1.18^{+0.36}_{-0.28}$\\
Sextans I & 3.42 & 67.9 & 1830 & $1.136^{+0.10}_{-0.09}$\\
Crater II & 1.93 & 39.0 & 507 & $0.535^{+0.12}_{-0.09}$\\
\bottomrule
\end{tabular}
\end{table*}

\section{A comparison of density
profiles}\label{a-comparison-of-density-profiles}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
histogram density profiles s compared to J+24. In general, both
methodologies agree well. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails. Because J+24 use only one or two components across the
entire dwarf galaxy, the faint regions become sensitive to the spatial
likelihood. Antlia II is an extreme case, where the satellite is hidden
behind a large number of foreground stars. We derive a much more
poorly-constrained and fainter total density profile than J+24, and the
divergence between the MCMC method here and J+24's profile likely shows
that \emph{Gaia} observations are insufficient to properly constrain
Antia II's density profile. This is likely due to uncertainties in the
background density of satellite-like stars in Antlia II.

. Once the true density of stars drops below the background of
satellite-like stars, J+24's method will occasionally select consistent
background stars with a density directly dependent on the spatial
likelihood. The uncertainties in the outer regions are likely
underestimated.

To properly compare density profiles before background-limiting effects
become important, we only calculate our profiles in the main text out to
the radii in Table~\ref{tbl:mcmc_props}. We derive these
``background-limited'' radii based on the outermost derived density in
the MCMC histogram with an uncertainty lower than 1 dex. This typically
corresponds to the empirical background from the CMD+PM samples in the
main text.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms.png}}
\caption[Probabilistic density profiles]{A comparison between the MCMC
histogram method and J+24. The MCMC samples in each bin are black
transparent dots, and the J+24 derived density profiles (with their
\(P_{\rm sat} < 0.2\)) are orange solid dots.}\label{fig:mcmc_hists}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms2.png}}
\caption[Probabilistic density profiles continued]{A continuation of
Fig.~\ref{fig:mcmc_hists}.}\label{fig:mcmc_hists2}
\end{figure}

### 2025-09-27 thesis/extra_numerical_convergence.tex
a compltet thesis draft

--- 2025-07-21 thesis/numerical_convergence.tex
@@
 \end{figure}
-
-\section{Fiducial parameters}\label{fiducial-parameters}
-
-Note that we use code units which assume that \(G=1\) for convenience
-and numerical stability. The conversion between code units to physical
-units is (for our convention):
-
-\begin{itemize}
-\tightlist
-\item
-  1 length = 1 kpc
-\item
-  1 mass unit = \(10^{10}\) Msun
-\item
-  1 velocity unit = 207.4 km/s
-\item
-  1 time unit = 4.715 Myr
-\end{itemize}
-
-Most parameters below are not too relevant or have been discussed or are
-merely dealing with cpu and IO details. The changes between simulation
-runs primarily affect the integration time, output frequency, and
-softening. Otherwise, we leave all other parameters fixed.
-
-\begin{verbatim}
-#======IO parameters======
-
-#---Filenames
-InitCondFile                initial
-OutputDir                   ./out
-SnapshotFileBase            snapshot
-OutputListFilename          outputs.txt
-
-#---File formats 
-ICFormat                    3       # use HDF5
-SnapFormat                  3
-
-#---Mem & CPU limits
-TimeLimitCPU                86400
-CpuTimeBetRestartFile       7200
-MaxMemSize                  2400
-
-#---Time
-TimeBegin                   0
-TimeMax                     2120     # 10 Gyr
-
-#---Output frequency
-OutputListOn                0
-TimeBetSnapshot             10
-TimeOfFirstSnapshot         0 
-TimeBetStatistics           10
-NumFilesPerSnapshot         1
-MaxFilesWithConcurrentIO    1
-
-#=======Gravity======
-
-#---Timestep accuracy
-ErrTolIntAccuracy           0.01
-CourantFac                  0.1     # ignored; for SPH
-MaxSizeTimestep             0.5
-MinSizeTimestep             0.0
-
-#---Tree algorithm
-TypeOfOpeningCriterion      1       # Relative
-ErrTolTheta                 0.5     # mostly used for Barnes-Hut
-ErrTolThetaMax              1.0     # (used only for relative)
-ErrTolForceAcc              0.005   # (used only for relative)
-
-#---Domain decomposition: should only affect performance
-TopNodeFactor                       3.0
-ActivePartFracForNewDomainDecomp    0.02
-
-#---Gravitational Softening
-SofteningComovingClass0     0.044  # HALO dependent
-SofteningMaxPhysClass0      0      # ignored; for cosmological
-SofteningClassOfPartType0   0
-SofteningClassOfPartType1   0
-
-#=======Miscellanius=======
-
-# probably do not need to change the options below
-
-#---Unit System
-UnitLength_in_cm            1 
-UnitMass_in_g               1
-UnitVelocity_in_cm_per_s    1 
-GravityConstantInternal     1
-
-#---Cosmological Parameters 
-ComovingIntegrationOn         0 # no cosmology
-Omega0                      0
-OmegaLambda                 0 
-OmegaBaryon                 0
-HubbleParam                 1
-Hubble                      100
-BoxSize                     0
-
-#---SPH
-ArtBulkViscConst            0.8
-MinEgySpec                  0
-InitGasTemp                 100
-
-#---Initial density estimate (SPH)
-DesNumNgb                   64
-MaxNumNgbDeviation          1
-\end{verbatim}

### 2025-09-27 thesis/extra_results.tex
a compltet thesis draft

\chapter{Additional results an tests}\label{sec:extra_results}

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
oblateness.

\section{Alternative initial
conditions}\label{alternative-initial-conditions}

The initial concentration (i.e.~choice of \(\vmax\) and \(\rmax\))
affects total dark matter evolution.
Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
\smallperi{} orbit with different initial halo parameters. The heavier
halo has \(\vmax=43\) and \(\rmax=7\), whereas the lighter halo has
\(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
more mass, the halo evolves to similar final structural parameters, as
expected in \citet{EN2021}. We note these choices were constrained to
match Scl's present-day velocity dispersion.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_halo_boundmass.pdf}}
\caption[Tidal dependence on halo concentration]{A comparison of the
evolution of different N-body models for different halo concentrations
(heavier and lighter halo), and the mean orbit instead of the
\smallperi{} orbit. In the left, the maximum circular velocity \(\vmax\)
of the dark matter halo is plotted as a function of simulation time. In
the right, \(\vmax\) is instead plotted as a function of
\(\rmax\).}\label{fig:tidal_tracks_concentration}
\end{figure}

We have good observational evidence for some dwarf galaxies to be cored.
Even Sculptor has debated claims for having a core
\citep[e.g.,][]{amorisco+zavala+deboer2014, breddels+helmi2013, battaglia+2008, walker+2009a, richardson+fairbairn2014, agnello+evans2012}.
To test if a core matter, we adopt a ``cored-NFW'' model,
\begin{equation}{
\rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
}\end{equation} where \(r_c\) is the core radius. We adopt
\(r_c = 0.1r_s\) and adopt an extreme initial mass of \(\vmax=49\,\kms\)
and \(\rmax=7.1\,\kpc\).

Sculptor and Ursa Minor may also be highly elliptical in 3D
\citep[e.g.,][]{an+koposov2022}, especially given their projected
shapes. We consider

We find the evolution of the oblate halo to be nearly identical to the
spherical halo, similar to \citet{battaglia+sollima+nipoti2015}.

While we have compared the evolution of orbits in the MW potential for
simplicity, because the tidal evolution of Scl is very weak in the LMC
model, the differences between models become even more slight.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
\caption[Tidal dependence on halo structure]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
including a core, velocity anisotropy, and evolving an oblate halo.
While the normalization may differ, the tidal evolution is
similar.}\label{fig:tidal_tracks_structure}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_orbits_boundmass.pdf}}
\caption[Ursa Minor tidal dependence on orbit]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
orbits in the LMC. Most LMC models evolve more weakly than the MW
models, however these do not reach full agreement with the present-day
position so misrepresent recent tidal
evolution.}\label{fig:tidal_tracks_umi}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_impact_i_f.pdf}}
\caption[Scl MW impact stellar densities]{The initial and final density
profiles for the Sculptor model which passes through the MW previously.
The impact does not substantially affect the inner density region. The
hints of a tidal stream in this model are misaligned to the proper
motion, likely a result of the LMC.}
\end{figure}

### 2025-09-27 thesis/extra_rv_models.tex
a compltet thesis draft

--- 2025-08-29 thesis/rv_models.tex
@@
 
+\citet{martinez-garcia+2023}: rotation gradients.
+
 \section{Velocity modelling and
@@
 \label{tbl:rv_measurements}
-\begin{tabular}{lllllllll}
+\begin{tabular}{llllllll}
 \toprule
- & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
+ & Study & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$\\
 \midrule
-Scl & combined &  & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
- & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
- & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
- & walker+09 & MMFS & 1818 & 1522 & 1417 & 1328 & 1.8 & 3\\
- & APOGEE & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
-UMi & combined &  & 4714 & 1225 & 1148 & 863 & 2.1 & \\
- & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
- & pace+20 & DEIMOS & 1716 & 1538 & 829 & 678 & 2.5 & 1\\
- & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & 2\\
- & APOGEE & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
+Scl & combined & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
+ & tolstoy+23 & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
+ & sestito+23a & 2 & 2 & 2 & 2 & 13 & –\\
+ & walker+09 & 1818 & 1522 & 1417 & 1328 & 1.8 & 3”\\
+ & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
+UMi & combined & 4714 & 1225 & 1148 & 863 & 2.1 & \\
+ & sestito+23b & 5 & 5 & 5 & 5 & 1.8 & –\\
+ & pace+20 & 1716 & 1538 & 829 & 678 & 2.5 & 1”\\
+ & spencer+18 & 1407 & 970 & 596 & 406 & 0.9 & 2”\\
+ & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
 \bottomrule
@@
 
-measurement
-
-\begin{table*}[t]
-\centering
-\caption[Sculptor RV fits]{MCMC fits for different RV datasets for Sculptor among 3 different models. }
-\label{tbl:scl_rv_mcmc}
-\begin{tabular}{lllllll}
-\toprule
-study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\log B_2/B_1$\\
-\midrule
-all &  &  &  &  &  & \\
- & $111.3\pm0.2$ & $9.64\pm0.16$ & - & - & - & 0\\
- & $111.3 \pm 0.2$ & $9.61 \pm0.16$ & - & $4.3\pm1.3$ & $-147_{-13}^{+17}$ & -1.9\\
- & $111.2\pm0.2$ & $9.66\pm0.16$ & $0.07\pm0.02$ & - & - & -3.4\\
-tolstoy+23 &  &  &  &  &  & \\
- & $111.3 \pm 0.3$ & $9.79 \pm 0.18$ & - & - & - & 0\\
- & $111.3\pm0.3$ & $9.77\pm0.19$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ & -1.3\\
- & $111.2 \pm 0.3$ & $9.73\pm0.19$ & $0.085 \pm 0.023$ & – & – & -4.6\\
-walker+09 &  &  &  &  &  & \\
- & $111.1\pm0.3$ & $9.5\pm0.2$ & – & – & – & 0\\
- & $111.1\pm0.3$ & $9.5\pm0.2$ & - & $5.2_{-1.6}^{+1.7}$ & $-134_{-16}^{+22}$ & -1.9\\
- & $111.1\pm0.3$ & $9.6\pm0.2$ & $0.06\pm0.03$ & – & – & +0.3\\
-apogee &  &  &  &  &  & \\
- & $111.2\pm0.7$ & $8.6\pm0.5$ & – & – & – & –\\
- & $111.2\pm0.7$ & $8.5\pm0.5$ & – & $6\pm3$ & $-126_{-33}^{+45}$ & +0.1\\
- & $111.1\pm0.7$ & $8.5\pm0.5$ & $0.07\pm0.06$ & – & – & +0.9\\
-\bottomrule
-\end{tabular}
-\end{table*}
+\input{rv_table.tex}
 
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/sigma_v_gradient.pdf}}
+\caption[Possible gradients in the velocity dispersion]{The observed
+velocity dispersion gradient (black) in 10 equal number bins and the
+derived slopes from the model fitting. Scl shows moderate evidence for
+an increasing velocity dispersion with radius. UMi's evidence for a
+radial gradient is weaker.}
+\end{figure}

### 2025-09-27 thesis/introduction.tex
a compltet thesis draft

@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
-\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax
+\caption[Images of dwarf galaxies]{Images of the LMC (DSS2), Fornax
 \citep[DES DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
@@
 \includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
-\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
-on the sky. We label the classical dwarf galaxies (green diamonds),
-fainter dwarfs (blue squares), globular clusters (orange circles), and
-ambiguous systems (pink open hexagons). Globular clusters are more
-centrally concentrated, but dwarf galaxies are preferentially found away
-from the MW disk. Sculptor and Ursa Minor are highlighted as two dwarfs
-we study later. The background image is from ESA/Gaia/DPAC
+\caption[The on-sky distribution of Milky Way satellites]{The location
+of MW dwarf galaxies on the sky. We label the classical dwarf galaxies
+(green diamonds), fainter dwarfs (blue squares), globular clusters
+(orange circles), and ambiguous systems (pink open hexagons). Globular
+clusters are more centrally concentrated, but dwarf galaxies are
+preferentially found away from the MW disk. Sculptor and Ursa Minor are
+highlighted as two dwarfs we study later. The background image is from
+ESA/Gaia/DPAC
 (https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).
@@
 \includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
-\caption[Example density profiles]{Density profiles in log density
-versus log radius for stars and dark matter in a Fornax-like galaxy. The
-dark matter is more extended and massive than the star across the entire
-galaxy. The halo has \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
-\(M_{200} = 1\times10^{10}\,\Mo\) and \(c=12.5\), the cosmological mean
-for a Fornax stellar mass galaxy. The galaxy has a stellar mass
+\caption[Example dark matter and stellar density profiles]{Density
+profiles in log density versus log radius for stars and dark matter in a
+Fornax-like galaxy. The dark matter is more extended and massive than
+the star across the entire galaxy. The galaxy has a stellar mass
 \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
-\(0.65\,\kpc\)
-\citep{munoz+2018, woo+courteau+dekel2008}.}\label{fig:nfw_density}
+\(0.65\,\kpc\) \citep{munoz+2018, woo+courteau+dekel2008}. The halo
+represents the cosmological mean given the stellar mass, with
+\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
+\(M_{200} = 1\times10^{10}\,\Mo\) and
+\(c=12.5\).}\label{fig:nfw_density}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
-\caption[Stellar-mass halo-mass relation]{\textbf{Left} The NFW halo
-concentration \(c=r_{200} / r_s\) as a function of virial mass
-\(M_{200}\). The solid line with 1-\(\sigma\) shaded region is the
-mass-concentration relation from \citet{ludlow+2016} for \(z=0\).
-\textbf{Right} Stellar mass (top) as a function of maximum circular
-velocity. The solid line with the 1-\(\sigma\) shaded region is the
-relation from \citet{fattahi+2018} with scatter points simulated central
-galaxies from \apostle{} in \citet{fattahi+2018}. The pink star
-illustrates the location of the Fornax galaxy, whose density profiles
-are shown in Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
+\caption[Cosmological halo mass and stellar mass
+relations]{\textbf{Left} The NFW halo concentration \(c=r_{200} / r_s\)
+as a function of virial mass \(M_{200}\). The solid line with
+1-\(\sigma\) shaded region is the mass-concentration relation from
+\citet{ludlow+2016} for \(z=0\). \textbf{Right} Stellar mass (top) as a
+function of maximum circular velocity. The solid line with the
+1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
+with scatter points simulated central galaxies from \apostle{} in
+\citet{fattahi+2018}. The pink star illustrates the location of the
+Fornax galaxy, whose density profiles are shown in
+Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
 \end{figure}
@@
 few galaxies had precisely measured proper motions \citep[e.g.~using
-Hubble Space Telescope,][]{sohn+2017}. \emph{Gaia} allowed the first
-systematic and precise determinations of Milky Way satellite proper
-motions \citep{pace+li2019, MV2020a}. While the proper motion
+Hubble Space Telescope,][]{piatek+2005, sohn+2017}. \emph{Gaia} allowed
+the first systematic and precise determinations of Milky Way satellite
+proper motions \citep{pace+li2019, MV2020a}. While the proper motion
 uncertainty of a typical dwarf member star is often large, by combining
@@
 
-While there are no clear theoretical preferences for any of these
-profiles, exponential density profiles have been commonly used for dwarf
-spheroidal galaxies. \citet{faber+lin1983} were among the first to
-demonstrate that an exponential law is a reasonable empirical fit,
-theorizing that dwarf spheroidals may have evolved from exponential disk
-galaxies and maintained a similar light profile. Later,
+\citet{faber+lin1983} were among the first to demonstrate that an
+exponential law is a reasonable empirical fit, theorizing that dwarf
+spheroidals may have evolved from exponential disk galaxies\footnote{While
+  studied for far longer, the exponential origin of disk galaxies is no
+  better understood. A sample of ideas range from scattering of stars
+  \citep{elmegreen+struck2013, wu+2022} to angular momentum transport,
+  to disk viscosity-driven radial gas flows
+  \citep{lin+pringle1987, wang+lilly2022} or spherical collapse
+  \citep{freeman1970}.} and maintained a similar light profile. Later,
 \citet{read+gilmore2005} showed that exponential profiles may originate
-from mass loss during the evolution of dwarf galaxies. Many subsequent
-photometric studies of dwarf spheroidal galaxies have used exponential
-fits, finding that exponential and King profiles both provide good
-descriptions in many cases
+from mass loss during the evolution of dwarf galaxies. Tides are a
+possible mechanism for this transformation---the ``tidal stirring
+hypothesis'' \citep{mayer+2001a, klimentowski+2009}. However, the
+theoretical origin of exponential disks is unclear.
+
+Nevertheless, exponential density profiles have been commonly used for
+dwarf spheroidal galaxies. Many subsequent photometric studies of dwarf
+spheroidal galaxies have used exponential fits, finding that exponential
+and King profiles both provide good descriptions in many cases
 \citep{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018}.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_vs_fornax.pdf}}
-\caption[Sculptor and Ursa Minor match tidal models]{A plot of the
+\pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_fornax_exp.pdf}}
+\caption[Scl and UMi have an extended density profile]{A plot of the
 surface density profiles of Sculptor (orange squares), Ursa Minor (red
@@
 Section~\ref{sec:observations}). The solid black line is an exponential
-profile (Eq.~\ref{eq:exponential_law}).}\label{fig:scl_umi_vs_fornax}
+profile (Eq.~\ref{eq:exponential_law}). Scl and UMi show a clear
+overdensity at large radii.}\label{fig:scl_umi_vs_fornax}
 \end{figure}
@@
 
-Simulating dwarf galaxies accurately in a cosmological context remains a
-challenge. Cosmological simulations can predict the overall abundance of
-the most massive dwarf galaxies \citep[e.g.,][]{sawala+2016} and broadly
-examine the effects of tides \citep[e.g.,][]{riley+2024}. However,
-simulated dwarf galaxies are often near the resolution limit.
-Insufficient resolution can lead to artificial disruption of dwarf
-galaxies and over-prediction of tidal effects
-\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To address this
-challenge, idealized simulations are often used to simulate a single
-subhalo in an approximate host potential, achieving excellent numerical
-convergence. For instance, the simulations we describe later reach 3
-times higher resolution than Aquarius \citep{springel+2008} at a
-fraction of the computational cost (400 times fewer particles).
-Idealized simulations make numerous simplifications, neglecting mergers,
-cosmological context, mass assembly, and often baryonic physics
+Tidal disruption of dwarfs in the Local Group continues to this day. The
+Magellanic stream, a massive, gas-rich feature emanating from the
+Magellanic clouds, is believed to arise partially from the MW's tides
+\citep{putman+1998, diaz+bekki2012, donghia+fox2016}. Other clear
+examples of tidal streams include the Sagittarius stream, the Andromeda
+Giant Southern stream, and the Tucana III stream
+\citep[e.g.,][]{ibata+gilmore+irwin1994, ibata+2001, li+2018}. These
+examples illustrate that hierarchical accretion remains an active
+process. Interpreting such observations relies on simulations of tidal
+disruption.
+
+Cosmological simulations struggle to resolve tidal effects on dwarfs.
+Since many dwarfs are near the resolution limit, they are vulnerable to
+artifical disruption
+\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
+numerical challenges, idealized simulations model a single subhalo in an
+analytic host potential, achieving excellent numerical convergence. For
+instance, the simulations we describe later reach the times higher
+resolution than Aquarius \citep{springel+2008} with 400 times fewer
+particles. Idealized simulations make numerous simplifications,
+neglecting mergers, cosmological context, mass assembly, and often
+baryonic physics
 \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
@@
 
-Conclusions from early work on tidal evolution remain mostly relevant
-today. A galaxy undergoing tidal effects loses most mass near the
-Lagrange points at pericentre, forms symmetric tidal streams with a
-velocity gradient, and retains its central structure
-\citep{oh+lin+aarseth1995, piatek+pryor1995, moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
-
-CDM halos also are found to be highly resilient to full disruption
-\citep{EP2020}. In addition, NFW halos evolve along ``tidal tracks''
-\citep[e.g,][]{PNM2008, green+vandenbosch2019, EN2021}. On the other
-hand, cored dark matter halos likely disrupt faster
+Idealized simulations predict clear properties of tidally disrupting
+dwarf galaxies. Tidally stripped stars form tidal
+streams---approximately uniform-density stellar tails with a bulk
+velocity gradient
+\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
+Most mass loss happens near pericentre, where tides are strongest.
+However, the central structure of a dwarf galaxy often remains
+undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
+cuspy CDM halos also are found to be nearly immune to full disruption
+\citep{EP2020}, but cored dark matter halos may disrupt more quickly
 \citep[e.g.,][]{penarrubia+2010, errani+2023a}.
 
+To first order, tidal mass loss peels away the outer energy-layers of a
+dwarf galaxy.
 \citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
 showed that tidal effects are nearly entirely described as the removal
-of particles below a certain energy \citep[see
+of particles above a truncation energy \citep[see
 also][]{choi+weinberg+katz2009}. \citet{stucker+2023} generalized this
 idea, creating a model for adiabatic tidal mass loss in an isotropic
-tidal field, confirming that an NFW halo is likely unable to be
-disrupted and explaining the evolution of NFW halos along similar tidal
-tracks.
+tidal field. They explain the resilience of NFW halos and origin of
+characteristic ``tidal tracks'' \citep[as observed
+in][]{PNM2008, green+vandenbosch2019, EN2021}.
 
-With precise orbits and a better understanding of the Milky Way
-potential, recent work continues to directly probe the dynamical
+With precise orbital constraints and improved models of the Milky Way
+potential, recent studies continue to directly probe the dynamical
 histories of individual dwarf galaxies.
@@
 to Fornax, showing that this galaxy's stellar component or globular
-clusters are likely not affected by tides. \citet{borukhovetskaya+2022a}
-also used N-body simulations to analyze Crater II, showing that the
-present-day properties of the galaxy are challenging to reconcile with
-\LCDM{} initial conditions with tidal evolution. Most relevantly,
-\citet{iorio+2019} conducted simulations of Sculptor, concluding tides
-likely do not affect this galaxy. Our goal is to apply a similar
-framework to Sculptor and Ursa Minor.
+clusters are likely not affected by tides. Similarly,
+\citet{borukhovetskaya+2022a} analyzed Crater II, showing that the
+present-day structure challenging to reconcile with \LCDM{} initial
+conditions and/or tidal evolution. Most relevantly, \citet{iorio+2019}
+modelled Sculptor, concluding tides likely do not affect the stellar
+component.
+
+Building on this body of work, we will use idealized simulations to
+understand the severity of tidal effects on Sculptor and Ursa Minor.
 
@@
 The \textbf{Jacobi radius} represents the approximate radius where stars
-become unbound for a galaxy in a circular orbit around a host galaxy.
-Calculated from an approximation of the location of the \(L_1\) and
-\(L_2\) Lagrange points, the Jacobi radius is where the mean density of
-the dwarf galaxy is roughly three times the mean interior density of the
-host galaxy at pericentre, or
+become unbound for a galaxy in a circular orbit around a host
+galaxy.\footnote{Possibly first calculated before \citet{laplace1798},
+  this radius also bears other names such as the Hill radius
+  \citep[from][]{hill1878}. Likely only names after Jacobi for the
+  Jacobi integral \citep{jacobi1836}.} Calculated from an approximation
+of the location of the \(L_1\) and \(L_2\) Lagrange points, the Jacobi
+radius is where the mean density of the dwarf galaxy is roughly three
+times the mean interior density of the host galaxy at pericentre, or
 \begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
@@
 r_{\rm break} = C\,\sigma_{v}\,\Delta t
-}\end{equation} where the scaling constant \(C \approx 0.55\) was
-derived empirically. \(r_{\rm break}\) describes where the dynamical
-timescale is longer than the time since the perturbation, i.e.~the
-radius within which the galaxy has had enough time to dynamically relax.
+}\end{equation} where the empirical constant \(C \approx 0.55\).
+\(r_{\rm break}\) describes where the dynamical timescale is longer than
+the time since the perturbation, i.e.~the radius within which the galaxy
+is dynamically relaxed.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/idealized_break_radius.pdf}}
-\caption[Break radius validation]{Example density and velocity
+\caption[Example tidal simulation]{Example density and velocity
 distributions of an idealized simulation shortly after pericentre.
-\textbf{Top left}: The 2D density profile for the initial and final
-simulation with the break radius marked. The break radius of the
-simulations is set by the time since pericentre. \textbf{Bottom left}:
-the mean radial velocity (dot product of relative position and velocity
-relative to dwarf centre) as a function of 2D radius. \textbf{Right}:
-The projected 2D stellar density in the \(x\)-\(y\) plane. The green
-circle represents the break radius and the grey arrow points towards the
-host centre.}\label{fig:idealized_break_radius}
+\textbf{Top}: The projected 2D stellar density in the \(x\)--\(y\) plane
+for the initial (left) and final (right) simulation. The dashed green
+circle represents the break radius (Eq.~\ref{eq:r_break}) and the solid
+(dotted) blue line represents the past (future) orbit. \textbf{Bottom
+left}: The projected stellar density profile for the initial (dotted)
+and final (solid) simulation snapshot. The green arrow marks the break
+radius. \textbf{Bottom left}: the mean radial velocity (dot product of
+relative position and velocity relative to dwarf centre) as a function
+of projected radius. The green arrow marks the break
+radius.}\label{fig:idealized_break_radius}
 \end{figure}
@@
 our results for the tidal effects on Sculptor and Ursa Minor in Chapter
-\ref{sec:results}, We discuss our results, limitations, and implications
-in Chapter \ref{sec:discussion}. Finally, Chapter \ref{sec:summary}
-summarizes this thesis and discusses future directions for similar work.
+\ref{sec:results}. We discuss the implications of our results and ending
+with a summary and outlook in Chapter \ref{sec:discussion}.

### 2025-09-27 thesis/methods.tex
a compltet thesis draft

@@
 \includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/v_circ_potential.png}
-\caption[Circular velocity of potential]{Circular velocity profile of
-\citet{EP2020} potential. The total circular velocity (thick black line)
-is composed of an NFW halo (green dashed line), a think and thick
-\citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
+\caption[Circular velocity of the Milky Way potential]{Circular velocity
+profile of \citet{EP2020} potential. The total circular velocity (thick
+black line) is composed of an NFW halo (green dashed line), a think and
+thick \citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
 \citet{hernquist1990} bulge (blue dotted
@@
 \pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
-\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
-potential in Galactocentric \(x\), \(y\), and \(z\) coordinates (top)
-and in Galactocentric radius \(r\) versus time (bottom). The Milky Way
-is at the centre with the disk lying in the \(x\)--\(y\) plane. Our
-selected \smallperi{} orbit is plotted in black and light green
-transparent orbits represent the past 5Gyr orbits sampled over Sculptor
-observables in Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is
-well-constrained in this potential and it is unlikely to achieve a
-smaller pericentre than the \smallperi{} orbit.}\label{fig:scl_orbits}
+\caption[Sculptor's possible orbits]{The orbits of Sculptor in a static
+Milky Way potential in Galactocentric \(x\), \(y\), and \(z\)
+coordinates (top) and in Galactocentric radius \(r\) versus time
+(bottom). The Milky Way is at the centre with the disk lying in the
+\(x\)--\(y\) plane. Our selected \smallperi{} orbit is plotted in black
+and light green transparent orbits represent the past 5Gyr orbits
+sampled over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The
+orbit of Sculptor is well-constrained in this potential and it is
+unlikely to achieve a smaller pericentre than the \smallperi{}
+orbit.}\label{fig:scl_orbits}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_xyzr_orbits.pdf}}
-\caption[Ursa Minor Orbits]{Similar to Fig.~\ref{fig:scl_orbits}, the
-orbits of Ursa Minor in a static Milky Way potential in Galactocentric
-\(x\), \(y\), and \(z\) coordinates. In the lower panel, we show the
-radius versus time for only three orbits of Ursa
-Minor.}\label{fig:umi_orbits}
+\caption[Ursa Minor's possible orbits]{Similar to
+Fig.~\ref{fig:scl_orbits}, the orbits of Ursa Minor in a static Milky
+Way potential in Galactocentric \(x\), \(y\), and \(z\) coordinates. In
+the lower panel, we show the radius versus time for only three orbits of
+Ursa Minor.}\label{fig:umi_orbits}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/initial_halo_selection.pdf}}
-\caption[Sculptor initial halos]{Selection of initial halos for Sculptor
+\caption[Initial halo selection]{Selection of initial halos for Sculptor
 and Ursa Minor. The grey line and pink line with shaded regions
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
-\caption[Numerical halo convergence]{Numerical convergence test for
-circular velocity as a function of log radius for simulations with
-different total numbers of particles in isolation. Residuals in bottom
-panel are relative to NFW. The initial conditions are dotted, the
-converged radius is marked by arrows \citep[eq. 13,][]{power+2003}, and
-the softening length is marked by a vertical bar. Note that a slight
-reduction in density starting around \(r = 30\,\kpc\) is expected given
-our outer truncation choice.}\label{fig:numerical_convergence}
+\caption[Numerical convergence of the N-body simulation]{Numerical
+convergence test for circular velocity as a function of log radius for
+simulations with different total numbers of particles in isolation.
+Residuals in bottom panel are relative to NFW. The initial conditions
+are dotted, the converged radius is marked by arrows \citep[eq.
+13,][]{power+2003}, and the softening length is marked by a vertical
+bar. Note that a slight reduction in density starting around
+\(r = 30\,\kpc\) is expected given our outer truncation
+choice.}\label{fig:numerical_convergence}
 \end{figure}

### 2025-09-27 thesis/results.tex
a compltet thesis draft

@@
 evolution over a selection of past apocentres and the present day.
-Limits range from -150 to 150 kpc in the \(y\)--\(z\) (approximately
-orbital) plane and the colourscale is logarithmic spanning 5 orders of
-magnitude between the maximum and minimum values. The green dot
-represents the final expected position of the galaxy and the solid and
-dotted grey curves represents the orbit over one previous or future
-radial oscillation respectively.}\label{fig:scl_sim_images}
+Limits range from -150 to 150 kpc in the \(y\)--\(z\) (\(\sim\)orbital)
+plane, and the colourscale is logarithmic spanning 5 orders of magnitude
+between the maximum and minimum values. The green dot represents the
+final expected position of the galaxy and the solid and dotted grey
+curves represents the orbit over one previous or future radial
+oscillation respectively.}\label{fig:scl_sim_images}
 \end{figure}
@@
 
-\subsection{Robustness to initial
-conditions}\label{robustness-to-initial-conditions}
-
-Some initial conditions may be slightly more susceptible to tides. Dark
-matter halos which are cored or less concentrated disrupt or lose mass
-faster \citep[e.g.,][]{stucker+2023}. We consider alternative dark
-matter halo initial conditions in the Appendix. We find that while the
-details of the strength of tidal impacts depend on the halo, our models
-here remain representative of the tidal effects.
-
 \subsection{Summary}\label{summary}
@@
 UMi loses more DM mass, with only about 3\% of the total mass remaining
-({[}Table~\ref{tbl:umi_sim_results}; {]}). However, tidal features in
-the observed stellar components are still extremely faint, appearing
-outside 100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
+(see Fig.~\ref{fig:umi_tidal_track} and
+Table~\ref{tbl:umi_sim_results}). However, tidal features in the
+observed stellar components are still extremely faint, appearing outside
+100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
 moderately affect the observed size and velocity dispersion of Ursa
@@
 \centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
+\caption[Ursa Minor tidal tracks]{Similar to
+Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
+substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
+\end{figure}
+
+\begin{figure}
+\centering
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_smallperi_i_f.pdf}}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_plummer_i_f.pdf}}
-\caption[Ursa Minor simulated density profiles]{The tidal effects on the
+\caption[Ursa Minor Plummer model density]{The tidal effects on the
 stellar surface density of Ursa Minor for Plummer stars on the
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
-\caption[Ursa Minor Predicted stream]{The properties of the stream
+\caption[Ursa Minor predicted stream]{The properties of the stream
 around the UMi \smallperi{} orbit with Plummer stars. The panels are all
@@
 LMC in the potential further weakens the tides experience by UMi.
+
+\section{Modeling uncertainties}\label{modeling-uncertainties}
+
+\subsection{Modelling assumptions}\label{modelling-assumptions}
+
+As Section~\ref{sec:results} shows, tides marginally affected the
+stellar components of Scl and UMi, even under extreme orbits. While
+Section~\ref{sec:results} only presents select models, alternative
+initial conditions do not affect our qualitative conclusions on tidal
+effects.
+
+We neglect baryonic physics. Since Scl and UMi have predominantly stars
+older than \(\sim 9\) Gyr
+\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025},
+gas dynamics are unlikely to affect recent evolution. A collisionless
+dark-matter only simulation should be an excellent approximation.
+
+Cored or less concentrated dark matter halos disrupt quicker
+\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo in particular is
+among the least concentrated halos consistent with UMi's velocity
+dispersion. Although Scl's fiducial halo is more concentrated, less
+concentrated and cored halos evolve similarly (see
+Section~\ref{sec:extra_results}).
+
+Galaxies are rarely perfect isotropic spheres. Sculptor and Ursa Minor
+are elliptical, and halos are expected to be radially anisotropic
+\citep[e.g.,][]{navarro+2010}. We test example non-spherical and
+anisotropic models in Section~\ref{sec:extra_results}, finding that
+these assumptions likely do not alter our conclusions.
+
+A system's observed velocity dispersion directly constrains the mean
+density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
+required to disrupt the stellar component does not strongly depend on
+the inner halo structure (via the Jacobi radius). Alternative initial
+conditions may influence the total mass evolution but would produce a
+similar final stellar structure.
+
+\subsection{Orbital uncertainties}\label{orbital-uncertainties}
+
+The long-term orbits of satellites are uncertain. Analytic Milky Way
+potentials neglect many unknown details, including triaxiality, halo
+twisting, mass evolution, and substructure. As a result, calculated
+orbits may diverge significantly from the true orbits of satellites
+\citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the Milky Way and
+dynamical friction implies that orbits are typically less bound in the
+past (less affected by tides). However, orbital energy and angular
+momentum are not conserved in cosmological simulations. Orbits in
+analytic potentials may overestimate the pericentre and underestimate
+the maximum tidal stress \citep[although typically not by enough to
+change our conclusions,][]{santistevan+2023, santistevan+2024}.
+
+Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates the uncertainties of the
+long-term orbits of Scl and UMi. Scl distance could range from
+4--\(300\,\kpc\) any time more than 4Gyr ago. And, if the LMC had a
+previous pericentre, UMi may have been an LMC satellite
+\citep[see][]{vasiliev2024}. Notably, these potentials are still
+idealized in many ways.
+
+Motivated by Fig.~\ref{fig:scl_orbit_lmc_uncert}, we examine an extreme
+pericentre of \(4\,\kpc\) of Scl with the MW in Appendix
+\ref{sec:extra_results}, finding it still insufficient to produce the
+observed density profile. We conclude our simulated orbits represent
+reasonable extremes for \emph{recent} tidal effects. Past encounters
+with the LMC are revisited below as a form of ``pre-processing'' in
+Section~\ref{sec:stellar_halos}.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
+\caption[Sculptor Orbits with LMC]{The long-term orbital history of
+Sculptor (\textbf{top}) and UMi (\textbf{bottom}) are uncertain. In both
+panels, light, transparent lines represent randomly-sampled orbit of the
+satellites (alla ref) in three different LMC/MW mass models from
+\citet{vasiliev2024}. The LMC orbits are in solid, thick lines of the
+corresponding colour. The L2M11 has a lighter LMC mass, and the L3M10
+model has a lighter MW mass than our fiducial L3M11 LMC
+model.}\label{fig:scl_orbit_lmc_uncert}
+\end{figure}
+
+\subsection{Summary}\label{summary-2}
+
+While the long-term tidal evolution is unconstrained, we conclude that
+our models are reasonable representations of recent tidal effects. As a
+result, recent tides are unlikely to affect the stellar distributions of
+Sculptor or Ursa Minor.

### 2025-09-27 agama/halos_elliptical/run_oblate.sh

@@ -1,1 +1,3 @@
-python example_doublepowerlaw.py density="potential_oblate_0.5.ini" out="1e6_oblate_0.5.dat" nbody=1000000
+#python example_doublepowerlaw.py density="potential_spherical.ini" > spherical.log
+python example_doublepowerlaw.py density="potential_oblate_0.5.ini" > oblate.log
+

### 2025-09-27 observations/mcmc/toy_model.jl

@@ -1,0 +1,409 @@
+### A Pluto.jl notebook ###
+# v0.20.18
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 1368181a-97d7-11f0-3b4f-b90907cb3133
+begin
+	import Pkg; Pkg.activate()
+
+	using LilGuys
+	using Arya
+	using DataFrames, CSV
+	import TOML
+	using CairoMakie
+end
+
+# ╔═╡ 19112874-f462-4cfb-bbef-ca2a43e72fd1
+using StatsBase, Distributions
+
+# ╔═╡ 7288a3ad-375a-4930-a98b-2d225d8ed709
+using PyFITS
+
+# ╔═╡ f1b3eef7-7aae-4622-ba01-9470a2867b39
+f_mix = 1
+
+# ╔═╡ 446bb406-8ec0-4dba-822b-187c0b97d73b
+N_stars = 100_000
+
+# ╔═╡ cc93ac36-795d-443e-998e-968a2d01f858
+R_max = 30
+
+# ╔═╡ a235a388-a42c-494f-8ef9-26554c5238b8
+f_sat_true = 0.03
+
+# ╔═╡ 82ff3270-fbe2-45df-943c-946f1292075a
+P_cut = 0.5
+
+# ╔═╡ 688e1152-81db-4a93-a314-1d902c87f054
+x_dist_memb = Beta(0.1, 8)
+
+# ╔═╡ 0feed8ce-18c4-4213-8519-5749a4b16552
+prof_memb_assumed = LilGuys.Plummer(r_s=3.0)
+
+# ╔═╡ ce09dc9f-0cea-4121-b0da-60a808399b2f
+prof_memb = LilGuys.Exp2D(R_s=0.8)
+
+# ╔═╡ 86671511-ac67-4837-bc29-1f4c8597519f
+N_memb = ceil(Int, N_stars*f_sat_true)
+
+# ╔═╡ a6af0647-c18e-4157-bf71-ed63186fe546
+N_bkd = ceil(Int, N_stars*(1-f_sat_true))
+
+# ╔═╡ 2ee203cd-1370-441f-96de-bf086ca48bc6
+f_sat = f_sat_true
+
+# ╔═╡ b719c6db-ec7c-45c3-a1d1-da9778c47477
+md"""
+# Toy Model
+"""
+
+# ╔═╡ 7d78cfc8-18b9-416b-9676-1c86c9f84841
+df_scl = let 
[…]

### 2025-09-27 thesis/compile.sh

@@ -1,6 +1,5 @@
 set -e 
 
-# pandocoptions="--filter ./table_shortcaption.py --filter pandoc-crossref --citeproc -M --top-level-division=section --bibliography=main.bib --natbib -M --autoSectionLabels=false -M --autoEqnLabels=true --filter ./table.py --filter ./md_figure.py --number-sections --filter equation.py"
 pandocoptions="--defaults=pandoc_params.yaml"
 
 pandoc $pandocoptions -o abstract.tex abstract.md
@@ -9,9 +8,9 @@ pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
 pandoc $pandocoptions -o results.tex results.md
 pandoc $pandocoptions -o discussion.tex discussion.md
-pandoc $pandocoptions -o conclusion.tex conclusion.md
 
-pandoc $pandocoptions -o appendix.tex appendix.md --top-level-division=chapter
-pandoc $pandocoptions -o rv_models.tex rv_models.md --top-level-division=chapter
-pandoc $pandocoptions -o numerical_convergence.tex numerical_convergence.md --top-level-division=chapter
+pandoc $pandocoptions -o extra_density.tex extra_density.md --top-level-division=chapter
+pandoc $pandocoptions -o extra_rv_models.tex extra_rv_models.md --top-level-division=chapter
+pandoc $pandocoptions -o extra_numerical_convergence.tex extra_numerical_convergence.md --top-level-division=chapter
+pandoc $pandocoptions -o extra_results.tex extra_results.md --top-level-division=chapter
 

### 2025-09-29 thesis/discussion.tex
thesis edits

@@
 halo formation scenarios.
+
+\section{Comparison with prior work}\label{comparison-with-prior-work}
+
+For both Scl and UMi, these galaxies have been studied extensively in
+both a theoretical and observational context. While many works
+considered mass modelling, star formation histories, chemistry, and
+other observational properties, we focus on the observations and models
+concerning tides and dynamical evolution here.
 
@@
 
-\section{Comparison with prior work}\label{comparison-with-prior-work}
-
-For both Scl and UMi, these galaxies have been studied extensively in
-both a theoretical and observational context. While many works
-considered mass modelling, star formation histories, chemistry, and
-other observational properties, we focus on the observations and models
-concerning tides and dynamical evolution here.
-
 However, some cosmological simulations have suggested that tidal effects
@@
 simulations.
-
-Closely related to our work, \citet{iorio+2019} applied idealized N-body
-simulations to study tidal effects on Scl. They similarly found weak
-tidal effects, even for a dark-matter-free model. However, their orbits
-are less well-constrained and they did not consider the LMC.
 
@@
 episodic star formation or hierarchical assembly.
+
+Our work is inconsistent with claims of observational tidal disruption.
+Closely related to our work, \citet{iorio+2019} applied idealized N-body
+simulations to study tidal effects on Scl. They similarly found weak
+tidal effects, even for a dark-matter-free model. However, their orbits
+are less well-constrained and they did not consider the LMC. If the
+extended density profile of Scl is formed through another mechanism
+which also creates multiple populations (see discussion below), then
+both observational features would become consistent with such an
+explanation.
 
@@
 
+As our models disagree with a tidal interpretation for the density
+profile of Ursa Minor, the observed density excess likely originate from
+another process. However, also find no evidence for a velocity gradient
+with current spectroscopic samples Section~\ref{sec:extra_rv_models}.
+While the origin of inner substructure in UMi is beyond the scope of
+this work, this structure, if real, points towards a complex origin of
+Ursa Minor.
+
 \section{Forming a stellar halo}\label{sec:stellar_halos}
 
-As a tidal origin of Scl and UMi's extended density profiles is
-disfavoured, we consider alternative processes which may form a dwarf
+As we disfavour a tidal origin of Scl and UMi's extended density
+profiles, we consider alternative processes which may form a dwarf
 galaxy ``stellar halo''. Some recent works have also confirmed the
-presence of extended, likely non-tidal, density profiles. For example,
-\citet{chiti+2021}; \citet{chiti+2023} spectroscopically confirm members
-out to \(~9 R_h\) in Tucana II. Given these stars are misaligned with
-the orbit and the lack of a velocity gradient, tides seem to be an
-unlikely explanation for Tucana II as well. The prevalence of stellar
-halos around dwarf galaxies remains an active research topic.
+presence of extended, likely non-tidal, density profiles. Tucana II is
+another dwarf with spectroscopic confirmation of outer ``halo'' stars
+out to \(9R_h\) and inconsistent with a tidal origin
+\citep{chiti+2021, chiti+2023}. The prevalence of stellar halos around
+dwarf galaxies remains an active research topic.
 
@@
 populations. Typically, older populations are extended, kinematically
-hotter, and metal-poo , whereas the younger populations are more
-compact, colder, and metal-rich. Examples include Carina
+hotter, and metal-poo, whereas the younger populations are more compact,
+colder, and metal-rich. Examples include Carina
 \citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
@@
 processes naturally produce a more extended, older stellar population.
+If internal dynamical heating explains stellar halos, then similar
+stellar halos should exist around similar dwarfs.
 
@@
 centrally-concentrated later generations of stars, naturally spawning
-multiple populations \citep{kawata+2006, revaz+jablonka2018}. However,
-star formation is also sensitive to a dwarf's environment.
+multiple populations \citep{kawata+2006, revaz+jablonka2018}. If
+episodic or vigorous star formation helps create stellar halos, then
+corresponding evidence of bursty star formation histories should be
+apparent.
 
@@
 originate from a similar merger
-\citep{tarumi+yoshida+frebel2021, querci+2025}.
+\citep{tarumi+yoshida+frebel2021, querci+2025}. Major mergers should
+produce multiple chemodynamical populations consistent with two
+distinct, but similar mass dwarf galaxies.
 
@@
 of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.
+Gas-rich mergers should have at least 3-distinct chemodynamical
+populations, one from each galaxy and one from the induced starburst.
 
@@
 ultra-faint dwarfs may create dwarf stellar halos. A stream detected
-around And II further supports the occurrence of mergers among dwarfs
-\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
-the halo chemistry would resemble a population of ultra-faint dwarfs.
-The occurance of dwarf mergers would further constrain small-scale
-galaxy formation.
+around And II further supports the existence of mergers among dwarfs
+\citep{amorisco+evans+vandeven2014, roederer+2023}. The occurrence of
+dwarf mergers would further constrain small-scale galaxy formation. In
+this scenario, the halo chemistry would resemble a population of
+ultra-faint dwarfs.
 
@@
 However, the prevalence of preprocessing remains uncertain. Like stellar
-heating, preprocessing redistributes already-present stellar
-populations.
+heating, preprocessing redistributes already-present stellar populations
+without creating new populations, but would occur more stochastically.
+Distant, unbound stream stars may be found around a preprocessed dwarf.
 
@@
 observed velocity dispersions of satellites, then many more dwarfs may
-be actively tidally disrupting as well
-\citep{casas+2012, yang+2014, wang+2024a}.
+be actively tidally disrupting
+\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
+also][]{sanchez-salcedo+hernandez2007}}.
 
@@
 
-\textbf{Chemistry}. Large samples of detailed chemical abundances have
+\emph{Chemistry}. Large samples of detailed chemical abundances have
 been invaluable for understanding MW substructure (e.g.~for
@@
 
-\textbf{Star formation histories}. Evidence of significant star
-formation episodes or lack thereof may differentiate scenarios which
-rely on strong star formation bursts (e.g.~\emph{episodic star formation
+\emph{Star formation histories}. Evidence of significant star formation
+episodes or lack thereof may differentiate scenarios which rely on
+strong star formation bursts (e.g.~\emph{episodic star formation
 history, induced star formation, and gas-rich mergers.})
 
-\textbf{Kinematics.} Particularly, for models relying on recent tidal
+\emph{Kinematics}. Particularly, for models relying on recent tidal
 disruption, kinematic disequilibrium features should be visible. These
@@
 
-\textbf{Deep photometry} may find or rule out signs of dynamical
+\emph{Deep photometry} may find or rule out signs of dynamical
 disequilibrium and tidal tails for tidally susceptible models
-\emph{(e.g., MOND and tidal dwarfs.)}
+\emph{(e.g., MOND and tidal dwarfs.)} In addition, photometry will help
+constrain the prevalence and nature of extended features in dwarf
+galaxies.
 

### 2025-09-29 thesis/extra_results.tex
thesis edits

@@
 concentration, cored halos, alternative orbits, anisotropy, and
-oblateness.
+ellipsoidal models.
 
@@
 \end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_sim_stream.pdf}}
+\caption[Ursa Minor predicted stream]{The properties of the stream
+around the UMi \smallperi{} orbit with Plummer stars. The panels are all
+as a function of \(\xi'\), the distance along the stream as defined by
+the current GSR proper motion vector. The top panels show the GSR proper
+motions in RA and Dec, and the bottom two show the distance and GSR
+radial velocities. To sample the stream, we randomly draw 100,000
+samples from the snapshot based on the stellar weights. A detectible
+gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
+the stream is tracked across several
+degrees.}\label{fig:umi_tidal_stream}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
+\caption[Scl predicted stream]{The predicted properties of a tidal tail
+in the Scl model.}\label{fig:scl_tidal_stream}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_stream.pdf}}
+\caption[Scl predicted stream]{the predicted properties of the stream of
+stars around Scl in the MW impact model. Note that, compared to
+Fig.~\ref{fig:scl_sim_stream}, this model produces a much more scattered
+and misaligned stream, due to the long time since stream formation and
+nonlinear effects of the LMC encounter. \textbf{todo check 2d velocity
+gradients in this case}}\label{fig:scl_mw_impact_tidal_stream}
+\end{figure}

### 2025-09-29 thesis/extra_rv_models.tex
thesis edits

@@
-\chapter{Radial velocity modeling}\label{sec:rv_obs}
+\chapter{LOS velocity modeling}\label{sec:rv_obs}
+
+In this section, we analyze the observed line-of-sight (LOS) velocity
+distributions for Scl and UMi. Our goal is to derive the systemic
+velocity and velocity disperions, and understand if the galaxies show
+evidence of tidal features, e.g.~a velocity gradient or a rising outer
+velocity dispersion. Our derived velocity distributions are consistent
+with literature. We find weak evidence for a velocity gradient in Scl
+and a rising outer velocity dispersion. However, Scl's velocity gradient
+is misaligned with its proper motion, likely inconsistent with a tidal
+interpretation. We find no evidence of a gradient in mean velocity or
+velocity dispersion in UMi. We conclude that Scl and UMi do not show
+observable features of tidal disruption given current velocity
+observations.
 
@@
 radial velocity measurements. We combine these samples with J+24's
-members to produce RV consistent stars and to compute velocity
-dispersion, systematic velocities, and test for the appearance of
-velocity gradients.
+Bayesian likelihoods to produce likely members using additional RV
+information.
 
 First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
-not report GaiaDR3 source ID's, we match to the nearest star within 1-3
+not report GaiaDR3 source ID's, we match to the nearest star within 1--3
 arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
@@
 mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
-variance \(s^2\). We remove stars with significant velocity dispersions
-as measured between observations in a study or between studies. By using
+variance \(s^2\).
+
+We remove stars with significant velocity dispersions as measured
+between observations in a study or between studies, to filter out
+unreliable measurements or possible spectroscopic binaries. By using
 that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
@@
 \(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
-number of measurements is 1-3 typically).
+number of measurements is 1-3 typically)..
 
@@
 star, the corrected radial velocity is then \begin{equation}{
-v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
-}\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
-in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
-velocities in RA and Dec, and \(\theta\) is the position angle of the
-star with respect to the centre of the dwarf. The correction from both
-effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
+v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
+}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the line of
+sight velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
+tangental velocities in RA and Dec, and \(\theta\) is the position angle
+of the star with respect to the centre of the dwarf. The correction from
+both effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
 Sculptor and less for Ursa Minor \citep[see
@@
 individual measurement uncertainty. Typically, the velocity dispersion
-will dominate the uncertainty budget here. We assume a halo/background
-velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
+dominates the uncertainty budget. We assume a halo/background velocity
+dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
 \citep[e.g.][]{brown+2010}.
@@
 Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2. Because of the additional information
-from radial velocities, most stars have probabilities close to 1 or 0 so
-the probability cut is not too significant.
+probability of greater than 0.2. Like in J+24, most stars have
+probabilities of nearly 1 or 0, so the precise choice of cut marginally
+affects the resulting sample.
 
-We assume priors on systematic velocity and velocity dispersion of
-\begin{equation}{
-\begin{split}
-\mu_{v} &= N(0\,\kms, \sigma_{\rm halo}^2) \\ 
-\sigma_{v} &= U(0, 20\,\kms)
-\end{split}
+\section{MCMC modeling}\label{mcmc-modeling}
+
+We use Monte-Carlo Markov chains (MCMC) to calculate the posterior
+distributions of our models of the intrinsic velocity dispersion and
+possible velocity gradients in Scl and UMi.
+
+The likelihood is based on the distribution of stars with respect to the
+model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
+in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
+likelihood is \begin{equation}{
+\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
+}\end{equation} where \(f\) is the probability distribution of a
+standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
+star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
+velocity dispersion at the \(i\)th star's position \(\xi_i\) and
+\(\eta_i\).
+
+For all models, we ado \begin{equation}{
+\mu_{0} \sim N(0\,\kms, \sigma^2_{\rm halo})\\
+\sigma_{0} \sim U(0\,\kms, 20\,\kms),
 }\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
@@
 assumption for dwarfs in orbit around the MW.
+
+For the simplest model, \(\mu\) and \(\sigma\) are held
+fixed---\(\mu(\xi, \eta) = \mu_0\) and \(\sigma(\xi, \eta) = \sigma_0\).
+There are no additional parameters besides \(\mu_0\) and \(\sigma_0\).
+
+To model a velocity gradient, we add two parameters, the gradient
+\begin{equation}{
+\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta\\
+\sigma = \sigma_0
+}\end{equation}
+
+To model an increasing velocity dispersion, we add a third parameter,
+\(c\) which applies as \begin{equation}{
+\log \sigma(R_{\rm ell}) = \log \sigma_0 + c\,\log\left(R_{\rm ell} / R_h\right) \\
+\mu = \mu_0 + a\,\xi + b\,\eta\\
+}\end{equation} with \(c \sim U(0, 1)\) where \(R_{\rm ell}\) is the
+elliptical radius. We only consider this model including the velocity
+gradient as a velocity gradient is the simplest way to increase the
+dispersion gradient.
+
+Bayes factors quantify the relative evidence between Bayesian model. If
+the Bayes factor
+
+Our models are \emph{nested}, where the fiducial model (\(\sigma_0\) and
+\(\mu_0\) only) always arises from setting parameters of the more
+complex models to zero (\(a=0, b=0\) and/or \(c=0\) ). As a result, we
+can use the Savage-Dickey method to calculate the Bayes factors
+\citep{dickey+lientz1970}, where the Bayes factor is then the relative
+density of posterior versus prior samples where \(a=b=c=0\). To
+calculate densities, we use a Silverman-bandwidth KDE to smooth the
+samples.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption[LOS velocity fit to Scl.]{Velocity histogram of Scl and UMi in
-terms of \(v_z\) (REF). Orange points are from our crossmatched RV
-membership sample.}
+\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
+terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
+points with error bars are from the crossmatched observed sample, and
+green lines represent MCMC fits to the velocity dispersion.}
 \end{figure}
-
-For Sculptor, we combine radial velocity measurements from APOGEE,
-\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
-\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
-measurements. We find that there is no significant velocity shift in
-crossmatched stars between catalogues. After crossmatching to high
-quality Gaia stars and excluding significant stellar velocity
-dispersions, we have a sample of 1918 members.
-
-We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
-velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
-consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{battaglia+2008}]{walker+2009}. See appendix REF for a more
-detailed comparison between different samples and additional tests.
-
-We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
-a position angle of \(-149_{-13}^{+17}\) degrees (see appendix REF).
-Several past work has attempted to detect a gradient in Sculptor, but no
-consensus has been reached. \citet{arroyo-polonio+2024} detect a
-velocity gradient of \(4\pm1.5\,\kmsdeg\) in a similar direction using
-the \citet{tolstoy+2023} sample, finding inconclusive statistical
-evidence. They additionally suggest a third chemodynamical component of
-the galaxy which may bias rotation measurements. \citet{battaglia+2008}
-also detect a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along
-the major axis, approximately the same direction. Instead,
-\citet{strigari2010}; \citet{martinez-garcia+2023} detect no significant
-gradient in Sculptor using \citet{WMO2009} sample. Note that
-pre-\emph{Gaia} work did not have as strong of a constraint on the
-proper motion of Scl, which limits conclusions of the intrinsic velocity
-gradient in Scl.
-
-For UMi, we collect radial velocities from, APOGEE,
-\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
-shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
-\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
-183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
-dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
-the median difference in velocities in this crossmatch is about 2.2
-km/s, we adopt 1 km/s as the approximate systematic error here. Our
-final sample includes 831 members.
-
-We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
-dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
-\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
-do not find evidence for a velocity gradient, consistent with past work
-\citep{pace+2020, martinez-garcia+2023}.
-
-\section{Discussion and limitations}\label{discussion-and-limitations}
-
-Our model here is relatively simple. Some things which we note as
-systematics or limitations:
-
-\begin{itemize}
-\tightlist
-\item
-  Inter-study systematics and biases. While basic crossmatches and a
-  simple velocity shift, combining data from multiple instruments is
-  challenging. This appears to be a minor issue (Sculptor) or is
-  corrected for (Ursa Minor).
-\item
-  Misrepresentative uncertainties. Inspection of the variances compared
-  to the standard deviations within a study seems to imply that errors
-  are accurately reported. APOGEE notes that their RV uncertainties are
-  known to be underestimates but are a small proportion of our sample.
-\item
-  Binarity. While not too large of a change for classical dwarfs, this
-  could inflate velocity dispersions of about \(9\,\kms\) by about
-  \(1\,\kms\)\citep{spencer+2017}. Thus, our measurement is likely
-  slightly inflated given the high binarity fractions measured in these
-  systems \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
-\item
-  Multiple populations. Both Sculptor and Ursa Minor likely contain
-  multiple populations \citep[\citet{pace+2020},
-  \citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a
-  single population, and each population may have a different extent and
-  velocity dispersion, this could result in biased velocity dispersions.
-  However, it is unclear how to uniquely determine an overall velocity
-  dispersion in a multi-population system.
-\item
-  Selection effects. RV studies each have their own selection effects,
-  which may affect the resulting dispersion, especially if different
-  populations or regions of the galaxy have different velocities or
-  velocity dispersions. We do not attempt to correct for this.
-\end{itemize}
-
-For both Ursa Minor and Sculptor, we also fit models to only data from
-individual surveys (see REF). Since the resulting parameters are very
-similar, we conclude that many of the systematic uncertainties are
-likely smaller than the present errors or that each large survey has
-similar biases.
-
-\citet{martinez-garcia+2023}: rotation gradients.
-
-\section{Velocity modelling and
-comparisons}\label{velocity-modelling-and-comparisons}
-
-Here, we describe in additional detail, our methods and comparisons for
-RV modelling between studies.
-
-Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
-smoothed samples from posterior/prior.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
-\caption[Scl velocity sample]{RV members of Sculptor plotted in the
-tangent plane coloured by corrected velocity difference from mean
-\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
-Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
-motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
-scale).}
-\end{figure}
-
-\begin{figure}
-\centering
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
@@
 (projected) gradient from MCMC samples, and the orange line is a rolling
-median (with a window size of 50).}
+median (with a window size of
+50).}\label{fig:scl_velocity_gradient_scatter}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
+\caption[Sculptor's velocity gradient orientation]{RV members of
+Sculptor plotted in the tangent plane coloured by corrected velocity
+difference from mean \(v_z - \bar v_z\) . The black ellipse marks the
+half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
+arrows mark the proper motion (PM, GSR frame) and derived velocity
+gradient (rot) vectors (to scale).}
 \end{figure}
@@
 \end{figure}
+
+For Sculptor, we combine radial velocity measurements from APOGEE,
+\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
+\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
+measurements. We find that there is no significant velocity shift in
+crossmatched stars between catalogues. After crossmatching to high
+quality Gaia stars and excluding significant stellar velocity
+dispersions, we have a sample of 1918 members.
+
+We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
+velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
+consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
+\citet{battaglia+2008}]{walker+2009}.
+
+We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
+a position angle of \(-149_{-13}^{+17}\) degrees. Several past work has
+attempted to detect a gradient in Sculptor, but no consensus has been
+reached. \citet{arroyo-polonio+2024} detect a velocity gradient of
+\(4\pm1.5\,\kmsdeg\) in a similar direction using the
+\citet{tolstoy+2023} sample, finding inconclusive statistical evidence.
+They additionally suggest a third chemodynamical component of the galaxy
+which may bias rotation measurements. \citet{battaglia+2008} also detect
+a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along the major
+axis, approximately the same direction. Instead, \citet{strigari2010};
+\citet{martinez-garcia+2023} detect no significant gradient in Sculptor
+using \citet{WMO2009} sample. Note that pre-\emph{Gaia} work did not
+have as strong of a constraint on the proper motion of Scl, which limits
+conclusions of the intrinsic velocity gradient in Scl.
+
+Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
+sample and the MCMC samples for the velocity gradient. While some
+samples have a consistent gradient with 0, most samples have a positive
+velocity gradient, consistent with the rolling median trend.
+
+The velocity gradient in Scl is misaligned with the proper motion
+
+For UMi, we collect radial velocities from, APOGEE,
+\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
+shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
+\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
+183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
+dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
+the median difference in velocities in this crossmatch is about 2.2
+km/s, we adopt 1 km/s as the approximate systematic error here. Our
+final sample includes 831 members.
+
+We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
+dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
+\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
+do not find evidence for a velocity gradient, consistent with past work
+\citep{pace+2020, martinez-garcia+2023}.
+
+For both Ursa Minor and Sculptor, we also fit models to only data from
+individual surveys. Since the resulting parameters are very similar, we
+conclude that many of the systematic uncertainties are likely smaller
+than the present errors or that each large survey has similar biases.
+
+\section{Discussion and limitations}\label{discussion-and-limitations}
+
+Our model here is relatively simple. Some things which we note as
+systematics or limitations:
+
+\emph{Inter-study systematics and biases}. While basic crossmatches and
+a simple velocity shift, combining data from multiple instruments is
+challenging. This appears to be a minor issue (Sculptor) or is corrected
+for (Ursa Minor).
+
+\emph{Misrepresentative uncertainties}. Inspection of the variances
+compared to the standard deviations within a study seems to imply that
+errors are accurately reported. APOGEE notes that their RV uncertainties
+are known to be underestimates but are a small proportion of our sample.
+We find that most samples report internally-consistent uncertainties for
+multiple observations of the same star.
+
+\emph{Binarity}. Stars in binary systems can inflate the inferred
+system's velocity dispersion. For example, a system with an apparent
+velocity dispersion of \(9\,\kms\) may have a true velocity dispersion
+of \(8\,\kms\) \citep{spencer+2017}. Thus, our measurement is likely
+slightly inflated given the high binary fractions measured in these
+systems \citep{spencer+2018, arroyo-polonio+2023, gration+2025}.
+
+\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
+multiple populations \citep[\citet{pace+2020},
+\citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a single
+population, and each population may have a different extent and velocity
+dispersion, this could result in biased velocity dispersions. However,
+it is unclear how to uniquely determine an overall velocity dispersion
+in a multi-population system.
+
+\emph{Selection effects}. RV studies each have their own selection
+effects, which may affect the resulting dispersion, especially if
+different populations or regions of the galaxy have different velocities
+or velocity dispersions. We do not attempt to correct for this. Given
+that samples extend through the half-light radius of each system, we do
+not expect current samples to miss large features except in the very
+outskirts.
+
+Could we observe velocity gradients instead in \emph{Gaia}.
+Unfortunately, given Scl and UMi's distances (\(\gtrsim 70\,\kpc\))
+\emph{Gaia's} systemic proper motion uncertainty of \(0.017\,\masyr\)
+corresponds to a velocity of \(5\,\kms\), making the detection of
+gradients of amplitude less than \(5\,\kms\) challenging. In addition,
+the typical uncertainties of member stars of Scl and UMi are even
+larger. While in future data releases, \emph{Gaia} uncertainties may
+permit study in the internal motion of dwarfs, the current \emph{Gaia}
+proper motions are not yet precise enough.
+
+\section{Summary}\label{summary}
+
+In this section, we analyze literature samples of LOS velocity
+measurements for both Scl and UMi. In each case, we find systemic LOS
+velocities and velocity dispersions consistent with past work. We detect
+weak evidence for a velocity gradient in Scl, however the gradient is
+mis-aligned with the proper motion of Scl so is unlikely to be of tidal
+origin. In both galaxies, the velocity dispersion profile is consistent
+with a constant velocity dispersion with radius. The lack of
+observational evidence for ongoing tidal disruption in the velocity
+distribution of stars in Scl and UMi further supports our
+interpretations in the main text.

### 2025-09-29 thesis/introduction.tex
thesis edits

@@
   were likely always known to humans at southern latitudes.}, with no
-known analogues, already presented such an enigma. H. Shapley presented
-these dwarfs as a new type of \emph{stellar system} resembling the
-Magellanic Clouds and globular clusters but did not attempt to speculate
-on their exact nature. While dwarf galaxies were soon understood to be
-galaxies based on the inferred luminosities and sizes, their nature
-remained unclear for decades
+known analogues at the time, already presented such an enigma. H.
+Shapley presented these dwarfs as a new type of \emph{stellar system}
+resembling the Magellanic Clouds and globular clusters but did not
+attempt to speculate on their exact nature. While dwarf galaxies were
+soon understood to be galaxies based on the inferred luminosities and
+sizes, their nature remained unclear for decades
 \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
@@
 have no clear evidence for dark matter. The boundary between these two
-classes blurs for faint, compact stellar associations, known as
-``ambiguous'' systems.
+classes blurs for faint, compact stellar associations. Systems with
+shared characteristics of globular clusters and dwarf galaxies, and
+without evidence as to their nature, are known as ``ambiguous'' systems
+(\textbf{ref}).
 
@@
 \(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
-  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
-  populations}, often exhibit irregular morphologies and recent star
-formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
-(LMC) as an example of an irregular, bright dwarf galaxy, where most
-stars are in a rotationally-supported thin disk (seen nearly face-on)
-with a prominent bar. \textbf{Classical dwarf spheroidals}\footnote{While
+  stellar mass-to-light ratio of \(1 \Mo / L_{\odot}\), may be
+  \(\sim 2 \Mo/L_\odot\) for older populations.}, often exhibit
+irregular morphologies and recent star formation.
+Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud (LMC) as
+an example of an irregular, bright dwarf galaxy, where most stars are in
+a rotationally-supported thin disk (seen nearly face-on) with a
+prominent bar. \textbf{Classical dwarf spheroidals}\footnote{While
   formally the dwarf galaxy names we discuss contain ``dwarf
@@
 galaxies have minuscule stellar masses, typically compact sizes, and
-very metal-poor stellar populations \citep[see review][]{simon2019}.
-Altogether, known dwarf galaxies span more than 15 orders in magnitude,
-or over 7 decades in stellar mass.
+very metal-poor stellar populations \citep[see the review
+by][]{simon2019}. Altogether, known dwarf galaxies span more than 15
+magnitudes, or over 7 decades in stellar mass.
 
@@
 teeming with dwarfs, many of which are satellites of either the Milky
-Way or(MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
+Way (MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
 the MW satellite system, including dwarf galaxies, globular clusters,
-and ambiguous systems. Such a nearby population of dwarf galaxies are
-useful for resolved, detailed studies aiming to understand the nature of
-these systems.
+and ambiguous systems. This nearby population of dwarf galaxies are
+opportune for resolved, detailed studies aimed to understand the nature
+of these systems.
 
@@
 observations of the universe stem from the cosmic microwave background
-(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
-But tiny perturbations in the CMB, temperature fluctuations of 1 part in
-10,000, reveal the underlying seeds of large-scale cosmological
+(CMB)---revealing a nearly uniform, isotropic blackbody emission. But
+tiny perturbations in the CMB, temperature fluctuations of 1 part in
+100,000, reveal the underlying seeds of large-scale cosmological
 structure. In an expanding universe, gravitational instability makes CDM
@@
 \emph{halos}. After recombination, where electrons combined with atomic
-nuclei to form atoms, baryons decoupled from radiation and fell into the
-dark matter halos. The densest pockets of baryons later formed the first
+nuclei to form atoms, baryons decoupled from radiation and fell into
+dark matter halos, where the condensed at the centre through radiative
+energy losses. The densest pockets of baryons later formed the first
 stars and galaxies.
@@
 progressively larger halos
-\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
+\citep[e.g.,][]{white+rees1978, blumenthal+1984, white+frenk1991}.
 Hierarchical assembly is evident through the large scale structure of
@@
 galaxies, which occupy the smallest dark matter halos, are promising
-windows into the behaviour of dark matter on small scales.
+probes into the behaviour of dark matter on small scales.
 
@@
 }\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
-density. This profile has shown remarkable success in describing \LCDM{}
+density. This profile has shown remarkable success at describing \LCDM{}
 halos across several orders of magnitude in mass. NFW profiles are
@@
 M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G,
-}\end{equation} and \(H(z)\) is the Hubble constant as a function of
-redshift. Another way of characterizing NFW halos is through the halo
-concentration, \(c=r_{200} / r_s\), which describes how the
-characteristic radius scale of the halo compares to the virial radius.
+}\end{equation} and \(H(z)\) is the Hubble constant, which depends on
+redshift. Another way of characterizing NFW halos is through the
+concentration parameter, \(c=r_{200} / r_s\), which describes how the
+characteristic radial scale of the halo compares to the virial radius.
 Using this parameter, the scale density is a function of \(c\) alone,
@@
 The two parameters of an NFW profile are not independent. Lower-mass
-dark matter halos often collapse earlier, when the universe was denser.
-As a result, low mass subhalos tend to be more concentrated
+dark matter halos typically collapse earlier, when the universe was
+denser. As a result, low mass subhalos tend to be more concentrated
 \citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
@@
 mass-concentration relation, but their effect is typically expected to
-be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}. The
-mass-concentration relation informs our assumed structure of Scl and UMi
-later.
+be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.
 
@@
 \caption[Example dark matter and stellar density profiles]{Density
-profiles in log density versus log radius for stars and dark matter in a
-Fornax-like galaxy. The dark matter is more extended and massive than
-the star across the entire galaxy. The galaxy has a stellar mass
-\(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
-\(0.65\,\kpc\) \citep{munoz+2018, woo+courteau+dekel2008}. The halo
-represents the cosmological mean given the stellar mass, with
-\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
-\(M_{200} = 1\times10^{10}\,\Mo\) and
+profiles in log 3D density versus log 3D radius for stars and dark
+matter in a Fornax-like galaxy. The dark matter is more extended and
+massive than the star across the entire galaxy. Fornax has a stellar
+mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
+\(0.65\,\kpc\) \citep{munoz+2018, woo+courteau+dekel2008}. The
+corresponding cosmological-mean halo for Fornax has \(\vmax=40\,\kpc\)
+and \(\rmax=8\,\kpc\), or \(M_{200} = 1\times10^{10}\,\Mo\) and
 \(c=12.5\).}\label{fig:nfw_density}
@@
 Fornax galaxy, whose density profiles are shown in
-Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
+Fig.~\ref{fig:nfw_density}. \textbf{add rmax-vmax plot
+too}}\label{fig:smhm}
 \end{figure}
 
-\subsection{\texorpdfstring{Galaxies formation in
-\LCDM{}}{Galaxies formation in }}\label{galaxies-formation-in}
+\subsection{\texorpdfstring{Galaxy formation in
+\LCDM{}}{Galaxy formation in }}\label{galaxy-formation-in}
 
@@
 \LCDM{} halos to derive constraints regarding which galaxies inhabit
-which halos. This technique, dubbed ``abundance matching,'' implies a
-tight relation between the stellar mass of a galaxy and the mass of the
-halo it inhabits \citep{li+white2009, moster+naab+white2013}.
+which halos. One simple technique, dubbed ``abundance matching,''
+implies a tight relation between the stellar mass of a galaxy and the
+mass of the halo it inhabits
+\citep{li+white2009, moster+naab+white2013}.
 

### 2025-09-29 thesis/methods.tex
thesis edits

@@
 \label{tbl:orbit_ics}
-\begin{tabular}{llll}
+\begin{tabular}{lllll}
 \toprule
-Property & Scl: \smallperi{} & Scl: \verb|LMC-flyby| & Umi: \smallperi{}\\
+Property & Scl: \smallperi{} & Scl: \verb|LMC-flyby| & UMI: \smallperi{} & Umi: \smallperi{}.5\\
 \midrule
-distance / kpc & 82.6 & 72.9 & 64.6\\
-$\pmra / \masyr$ & 0.134 & 0.137 & -0.158\\
-$\pmdec / \masyr$ & -0.198 & -0.157 & 0.050\\
-LOS velocity / $\kms$ & 111.2 & 111.2 & -245.75\\
-$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
-${x}_{i} / \kpc$ & -2.49 & 4.30 & 17.40\\
-${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
-${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
-$\V_{x\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
-$\V_{y\,i} / \kms$ & -114.83 & -56.83 & 48.62,\\
-$\V_{z\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
+distance / kpc & 82.6 & 72.9 & - & 64.6\\
+$\pmra / \masyr$ & 0.134 & 0.137 & - & -0.158\\
+$\pmdec / \masyr$ & -0.198 & -0.157 & - & 0.050\\
+LOS velocity / $\kms$ & 111.2 & 111.2 & - & -245.75\\
+$t_i / \Gyr$ & -9.17 & -2.0 & - & -9.67\\
+${x}_{i} / \kpc$ & -2.49 & 4.30 &  & 17.40\\
+${y}_{i} / \kpc$ & -42.78 & 138.89 &  & 74.51\\
+${z}_{i} / \kpc$ & 86.10 & 125.88 &  & 21.34\\
+$\V_{x\,i} / \kms$ & -20.56 & 6.89 &  & 14.27\\
+$\V_{y\,i} / \kms$ & -114.83 & -56.83 &  & 48.62,\\
+$\V_{z\,i} / \kms$ & -57.29 & 52.09 &  & -114.08\\
 \bottomrule
@@
 
-\subsection{Ursa Minor's Orbit}\label{ursa-minors-orbit}
+\subsection{Ursa Minor's Orbit}\label{sec:orbit_corrections}
 

### 2025-09-29 thesis/results.tex
thesis edits

@@
 As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
-we test whether Galactic tides can form the extended density profiles of
-Scl and UMi. In this Chapter, we analyze tailored N-body simulations,
-using the methods described in Chapter \ref{sec:methods}, to assess
-tidal impacts on each galaxy. We find that tides drive dark matter loss
-in both systems but leave their compact stellar components largely
-unaffected. The Large Magellanic Cloud (LMC) perturbs the orbit of Scl,
-and somewhat UMi, yet the resulting tidal effects are weaker than in a
-Milky Way-only potential. Our simulations demonstrate that recent tides
-are unlikely to alter the stellar structure of Scl or UMi.
+we test whether Galactic tides are responsible for the extended density
+profiles of Scl and UMi. In this Chapter, we analyze tailored N-body
+simulations, using the methods described in Chapter \ref{sec:methods},
+to assess the tidal impact of the Galactic potential. To anticipate our
+main conclusion, we find that tides drive dark matter loss in both
+systems but leave their compact stellar components largely unaffected.
+The Large Magellanic Cloud (LMC) perturbs the orbit of Scl, and somewhat
+UMi, yet the resulting tidal effects are weaker than in a Milky Way-only
+potential. Our simulations demonstrate that recent tides are unlikely to
+have altered the stellar structure of Scl or UMi.
 
@@
 
-\subsection{Dark matter evolution}\label{dark-matter-evolution}
+\subsection{Evolution of Sculptor's dark matter
+halo}\label{evolution-of-sculptors-dark-matter-halo}
 
-As a representation of the most extreme tidal history, we initially
+As a representation of an extreme tidal history, we initially
 investigate the \smallperi{} Scl model.
 
-Scl experiences moderate tidal mass loss. Fig.~\ref{fig:scl_sim_images}
-shows the stripping of dark matter and formation of diffuse streams
-trailing and leading Sculptor's orbit. Because tidal stripping can be
-described as removal of the lowest energy particles, most mass loss
-occurs in the outer halo.
+Scl experiences moderate tidal mass loss after 10 Gyrs of evolution.
+Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and
+formation of diffuse streams trailing and leading Sculptor's orbit.
+Because tidal stripping can be described as a gradual removal of the
+least bound particles, most mass loss occurs in the outer halo.
 
 N-body models may deviate from a point-particle trajectory due to
-dynamical self-friction \citep[e.g.,][]{miller+2020}. However, this
-effect is slight for Scl, which ends near the observed position, without
-adjusting the initial conditions (the green point in
+dynamical self-friction \citep[e.g.,][]{white1983, miller+2020}.
+However, this effect is slight for Scl, which ends near the observed
+position, without adjusting the initial conditions (the green point in
 Fig.~\ref{fig:scl_sim_images}).
@@
 3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
-significant mass, but mostly from the outer halo.
+significant amounts of mass, but mostly from the outer halo.
 
@@
 velocity profiles, unbound particles are iteratively removed,
-recalculating the potential at each step assuming spherical
-symmetry.}\label{fig:scl_tidal_track}
+recalculating the potential at each step assuming spherical symmetry.
+\textbf{mark jacobi radius and stellar contributions\ldots, maybe sigmav
+/ r\_h}}\label{fig:scl_tidal_track}
 \end{figure}
@@
 
-\subsection{Stellar evolution}\label{stellar-evolution}
+\subsection{Evolution of Sculptor's
+stars}\label{evolution-of-sculptors-stars}
 
@@
 density profile is nearly unchanged. Only at a surface density
-\(\sim10^8\) times fainter than the centre may faint tidal features
+\(\sim10^8\) times fainter than the centre do some faint tidal features
 emerge. The total stellar mass lost corresponds to \(\sim 10\) stars in
@@
 
-More extended stellar profiles are more sensitive to tides. With Plummer
-initial conditions, as shown in
-Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the model loses more stellar
-mass and forms more luminous tidal tails. Observations reaching surface
-densities \(\sim10\) times fainter than our data could reveal a stream
-in this case. Nevertheless, the radial extent probed by our data remains
-nearly unchanged by tidal evolution.
+Sculptor may not have been born as an exponential galaxy, and more
+extended stellar profiles would be more sensitive to tides. As an
+approximation of the present-day extended density profile, we initialize
+a model with Plummer stars instead. As shown in
+Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the Plummer model loses more
+stellar mass and forms more luminous tidal tails. Observations reaching
+surface densities \(\sim10\) times fainter than our data could reveal a
+stream in this case. Nevertheless, the radial extent probed by our data
+remains nearly unchanged by tidal evolution.
+
+The Jacobi and break radii further support that tidal effects should not
+be apparent in the observed stellar component. As calculated for this
+model, (see
+Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
+break and Jacobi radii both fall outside of \(\sim 100\) arcminutes for
+either stellar component. Indeed, the stellar component only begins to
+deviate from an exponential profile around this break radius
+(Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
+Given that no orbits of Scl in this potential produce significantly
+smaller break or Jacobi radii, it is unlikely any other orbit would
+produce a significant, observable density excess.
 
@@
 \texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
-potential. Notably, this model has a previous LMC pericentre
-\(\sim6\,\Gyr\) ago.
+potential.
 
@@
 The timing of the LMC encounter implies a break radius consistent with
-the observed density kink. We select the orbit with the
-\(6\sigma\)-smallest LMC pericentre. The orbit is selected following the
-procedure in Section~\ref{sec:orbital_estimation} except with
+the beginning of the observed density excess. We select the orbit with
+the \(6\sigma\)-smallest LMC pericentre. The orbit is selected following
+the procedure in Section~\ref{sec:orbital_estimation} except with
 observational uncertainties doubled. Results are similar when selecting
@@
 
-The combined tidal force of the MW and LMC are even weaker for Scl than
+The combined tidal effect of the MW and LMC are even weaker for Scl than
 in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
@@
 pericentre, Scl's dark matter is less disrupted than the previous
-MW-only model. The subsequent LMC passage has little effect. As a
-result, the dark matter structure evolves mildly and \(\sim 50\%\) of
-mass remains bound (Table~\ref{tbl:scl_lmc_sims}).
+MW-only model. The subsequent LMC passage modifies Scl's orbit but has
+otherwise little effect. As a result, the dark matter structure evolves
+mildly and \(\sim 50\%\) of mass remains bound
+(Table~\ref{tbl:scl_lmc_sims}).
 
@@
 density profiles of this model. While the break radius is within the
-observed density profile, tidal effects are too weak to be detectible.
+observed density profile, tidal effects are too weak to be detectable.
 Structural properties of the stars similarly evolve little
-(Table~\ref{tbl:scl_lmc_sim_stars}). While the instantaneous tidal force
-from the LMC is larger than the MW, Scl does not experience the LMC
-tidal field long enough to display disturbances.
+(Table~\ref{tbl:scl_lmc_sim_stars}).
+
+While the instantaneous tidal force from the LMC is larger than the MW,
+Scl does not experience the LMC tidal field long enough to display
+disturbances. Scl and the LMC have a similar orbital plane, but orbit in
+opposite directions (Fig.~\ref{fig:scl_lmc_orbits_effect}). As a result,
+the interaction between Scl and the LMC is brief as their relative
+velocity is \(\sim 400\,\kms\) at closest approach.
 
@@
 Fig.~\ref{fig:scl_smallperi_i_f} except for the \texttt{LMC-flyby}
-model. With only one MW pericentre and a recent, rapid LMC encounter,
-tidal forces do not appear to affect the stellar
-distribution.}\label{fig:scl_lmc_i_f}
+model. The Jacobi and break radii here are calculated with respect to
+the LMC, the corresponding radii with respect to the MW are larger. With
+only one MW pericentre and a recent, rapid LMC encounter, tidal forces
+do not appear to affect the stellar distribution. \textbf{todo fix break
+radius position, orbits as arrows\ldots{}}}\label{fig:scl_lmc_i_f}
 \end{figure}
@@
 
-We find, including only the MW, that tides only remove dark matter from
-the outskirts of Scl. The central cusp and compact stellar distribution
-are resilient to tides and any tidal effects are well-outside the reach
-of current observations. We then find that the LMC strongly perturbs
-Scl's orbit---in this case, Scl may be on first infall. However, with
-only 2 pericentres, the combined tides of the LMC and MW tides are
-weaker than for our initial model. In either case, tides are unlikely to
-affect Sculptor's stellar component.
+We find, including only the MW potential, that tides only remove dark
+matter from the outskirts of Scl. The central cusp and compact stellar
+distribution are resilient to tides and any tidal effects are
+well-outside the reach of current observations. We then find that the
+LMC strongly perturbs Scl's orbit---in this case, Scl may be on first
+infall. However, with only 1 pericentre each for the LMC and MW, the
+combined tides are weaker than for our initial model. In either case, we
+conclude that tides are unlikely to affect Sculptor's stellar component.
 
@@
 
-The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
-case. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM evolution.
-UMi loses more DM mass, with only about 3\% of the total mass remaining
-(see Fig.~\ref{fig:umi_tidal_track} and
-Table~\ref{tbl:umi_sim_results}). However, tidal features in the
-observed stellar components are still extremely faint, appearing outside
-100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
-moderately affect the observed size and velocity dispersion of Ursa
-Minor (Table~\ref{tbl:umi_sim_stars_results}).
+\subsection{Evolution of Ursa Minor's dark matter
+halo}\label{evolution-of-ursa-minors-dark-matter-halo}
 
-A stream may just be beyond observational limits if Ursa Minor's stars
-were initially more extended. As seen in Fig.~\ref{fig:umi_plummer_i_f},
-the tidal tails are more luminous than for Sculptor and would begin near
-the outermost density measurement. The properties of the stream are
-shown in Fig.~\ref{fig:umi_tidal_stream}. The stream stars become
-apparent outside of about 2 degrees, or the Jacobi radius. As expected
-for tidal tails, the stream shows gradients in each observable, with a
-dispersion in each observable similar to that of the dwarf progenitor.
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
+\caption[Ursa Minor simulation snapshots]{Similar to
+Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
+orbit. Dark matter evolution is more dramatic than for
+Scl.}\label{fig:umi_sim_images}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
+\caption[Ursa Minor tidal tracks]{Similar to
+Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
+substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
+\end{figure}
+
+The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
+potential. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM
+evolution. UMi loses significantly more DM mass than Scl, forming
+substantial dark matter streams encircling the MW several times.
+
+UMi only retains 3\% of its total mass after 9 Gyr
+(Table~\ref{tbl:umi_sim_results}). As a result, the final dark matter
+component is much smaller than the initial, but still evolves along the
+predicted tidal track Fig.~\ref{fig:umi_tidal_track}. Despite the more
+substantial tidal evolution, the Jacobi radius is still large, lying at
+\(\sim 4\,\kpc\), well beyond the final \(\rmax\).
+
+Because of UMi's mass loss, the orbit deviates substantially from a
+point orbit. Through our orbit-adjustment procedure in
+Section~\ref{sec:orbit_corrections}, we recover nearly exactly the
+present-day position of Ursa Minor by changing the initial positions by
+\(20\,\kpc\) and \(\sim 9\,\kms\). These adjustments do not
+significantly affect the qualitative structure or pericentre of the
+orbit.
+
+\subsection{Evolution of Ursa Minor's
+stars}\label{evolution-of-ursa-minors-stars}
+
 \begin{table*}[t]
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
-\caption[Ursa Minor simulation snapshots]{Similar to
-Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
-orbit.}\label{fig:umi_sim_images}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
-\caption[Ursa Minor tidal tracks]{Similar to
-Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
-substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
-\end{figure}
-
-\begin{figure}
-\centering
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_smallperi_i_f.pdf}}
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
-\caption[Ursa Minor predicted stream]{The properties of the stream
-around the UMi \smallperi{} orbit with Plummer stars. The panels are all
-as a function of \(\xi'\), the distance along the stream as defined by
-the current GSR proper motion vector. The top panels show the GSR proper
-motions in RA and Dec, and the bottom two show the distance and GSR
-radial velocities. To sample the stream, we randomly draw 100,000
-samples from the snapshot based on the stellar weights. A detectible
-gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
-the stream is tracked across several
-degrees.}\label{fig:umi_tidal_stream}
-\end{figure}
+Tidal features in UMi's stellar component are still extremely faint,
+becoming apparent only outside 100 arcminutes in
+Fig.~\ref{fig:umi_smallperi_i_f}. Tides also only moderately affect the
+observed size and velocity dispersion of Ursa Minor
+(Table~\ref{tbl:umi_sim_stars_results}). For exponential initial
+conditions, tidal effects are unlikely to ever be observable in the near
+future.
+
+Similar to Scl, we consider a model where UMi's stars are initially a
+more extended Plummer profile, resembling the present-day density
+profile. The stellar evolution of the Plummer stellar component is
+similar to the exponential stellar profile (\citet{ig:umi_plummer_i_f}).
+However, because there are more loosely-bound stars, the Plummer model
+looses nearly 4\% of its initial stellar mass to tides
+(Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
+detectible if we measure densities 2 orders of magnitude fainter than
+our present data. We show the properties of a stream in the Appendix
+(Fig.~\ref{fig:umi_tidal_stream}), but such a stream is unlikely to be
+observable in the near future.
 
 \subsection{Effects of the LMC}\label{effects-of-the-lmc}
-
-Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
-the orbit of Ursa Minor. Predominantly, the effect is to increase the
-orbital period, apocentre and pericentre. However, the orbit remains in
-a similar plane and with similar morphology. As UMi is on the opposite
-side of the Galaxy of the LMC, this is not unexpected. The deviation
-from the MW-only orbit is mostly due to the LMC-induced reflex motion of
-the Milky Way.
 
@@
 
+Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
+the orbit of Ursa Minor. Predominantly, the effect is to increase the
+orbital period, apocentre and pericentre. However, the orbit remains in
+a similar plane and with similar morphology. As UMi is on the opposite
+side as the Galaxy of the LMC, this is not unexpected.
+
+The deviation from the MW-only orbit is mostly due to the LMC-induced
+reflex motion of the Milky Way. Because the MW centre, and our origin,
+is accelerated towards the LMC, UMi experiences a fictitious
+acceleration opposite to the acceleration experienced by the MW centre,
+causing the orbit to increase in characteristic radius.
+
 \subsection{Summary}\label{summary-1}
@@
 tidal tails may be observable with deeper data. Finally, including the
-LMC in the potential further weakens the tides experience by UMi.
+LMC in the potential further weakens the tides experienced by UMi.
 
@@
 initial conditions do not affect our qualitative conclusions on tidal
-effects.
+effects expected for Scl and UMi in the Galactic (and LMC) potential.
 

### 2025-10-02 thesis/data.tex
discussion revisions

@@
 prototypical example is the Fornax dSph --- well-fit by an exponential
-over 4.5 decades in surface density. On the other hand, dSphs like
-Sculptor or Ursa Minor have profiles which deviate significantly from an
-exponential profile fitted to the inner regions, with a clear excess of
-stars/light in the outer regions. In this chapter, we critically review
-the density profile of the MW classical dwarfs. This thesis concerns the
-origin of this excess, with a tidal interpretation explored in
-subsequent chapters.
+over 4.5 decades in surface density. On the other hand, Sculptor or Ursa
+Minor have profiles that deviate significantly from an exponential
+profile fitted to the inner regions, with a clear excess of stars/light
+in the outer regions. In this chapter, we critically review the density
+profile of the MW classical dwarfs. This thesis concerns the origin of
+the outer deviations from exponential profiles, with a tidal
+interpretation explored in subsequent chapters.
 
@@
 For a satellite, the \emph{spatial likelihood} is specified either as a
-one- or two-component elliptical exponential profile. A second component
-is included only when the preferred amplitude of the second component is
-non-zero. The \emph{proper-motion likelihood} quantifies the agreement
-of a star's motion with the dwarf galaxy's systemic motion, accounting
-for observational uncertainties. The \emph{CMD likelihood} measures the
-consistency of a star's \emph{Gaia} \(G\) magnitude and
-\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
-galaxy. For the background likelihoods, the spatial likelihood is a
-uniform distribution over the field, and the background CMD and PM
-likelihoods are constructed empirically from the stars in an annulus far
-from the satellite. Each likelihood is normalized as a probability
-density over the respective parameter space. Appendix
+one- or two-component elliptical exponential profile. A second
+exponential component is considered only when the preferred amplitude of
+the second component is non-zero. The \emph{proper-motion likelihood}
+quantifies the agreement of a star's motion with the dwarf galaxy's
+systemic motion, accounting for observational uncertainties. The
+\emph{CMD likelihood} measures the consistency of a star's \emph{Gaia}
+\(G\) magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
+isochrones for the dwarf. For the background likelihoods, the spatial
+likelihood is a uniform distribution over the field, and the background
+CMD and PM likelihoods are constructed empirically from the stars in an
+annulus far from the satellite. Each likelihood is normalized as a
+probability density over the respective parameter space. Appendix
 \ref{sec:density_extra} discusses the details of quality cuts and the
@@
 the J+24 method provides a high-quality, low-contamination sample of
-dwarf galaxy candidate member stars, which we will now investigate.
+dwarf galaxy candidate member stars, which we will now investigate in
+further detail.
 
@@
 line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
-details). While the latter is the best motivated membership sample, we
+details). While the latter is the highest-purity membership sample, we
 do not use this sample for stellar density analysis due to its

### 2025-10-02 thesis/discussion.tex
discussion revisions

--- 2025-09-27 thesis/discussion.tex
@@
-In the previous chapters, we have shown that (1) Sculptor and Ursa Minor
-have unusually extended density profiles for MW classical satellites and
-(2) recent tides are an unlikely explanation. Here, we first consider
-the reliability and assumptions supporting our conclusions. We then
-consider possible formation scenarios of extended density profiles,
-including multi-epoch star formation, dwarf mergers and subhalo
-encounters, and alternative theories of dwarf galaxy structure. We
-finish with an outlook on prospects for disentangling various stellar
-halo formation scenarios.
+In the previous chapters, we have shown that Sculptor and Ursa Minor
+have more extended density profiles compared to other MW classical
+satellites, and that recent tides are an unlikely explanation for such
+extended density profiles. We therefore suggest that Scl and UMi host a
+``stellar halo''---an additional, extended population of stars. With our
+data, other classical dwarfs lack evidence for similar stellar halos.
+The absence of similar halos in other classical dwarfs suggests Scl and
+UMi have followed a divergent evolutionary path.
 
-While tides have been a popular explanation for some features of dwarf
-galaxies \citep[discussion above, and
-e.g.,][]{tsujimoto+shigeyama2002, mayer+2001a}, our work, and many
-others, have disfavoured this for Scl, UMi, and many other MW dwarfs.
-For example, \citet{read+2006} suggest that the lack of observational
-evidence for a rising velocity dispersion profile with radii indicates
-most dwarfs are not tidally affected. Later, \citet{penarrubia+2009}
-define and use the break radius, Eq.~\ref{eq:r_break}, to show that the
-(then understood) orbits of most satellites are inconsistent with a
-tidal feature. In addition, \citet{pace+erkal+li2022} use a criterion
-based on the observed mean density and orbital pericentre (like the
-Jacobi radius) to show most dwarfs are unlikely to be undergoing
-significant tidal disruption. Most recently,
-\citet{tchiorniy+genina2025} use a similar idealized framework to ours,
-with a focus on the inner density, showing tides do not strongly affect
-equilibrium assumptions for several classical dwarfs. The general tidal
-evolution of our simulations is furthermore consistent with much of the
-literature \citep[e.g.,][]{robles+bullock2021, EN2021}. Tides seem to
-now play a subdominant role in the evolution of most Milky Way
-satellites.
+In this chapter, we place our results in context with past work. We
+review both work considering tidal effects on Scl and UMi and detections
+of peculiarities. We then consider possible origins of stellar halos for
+these galaxies. Scenarios include induced star formation forming
+multiple populations, past mergers depositing stars into the outskirts,
+and even departures from the standard dwarf galaxy structure. We later
+discuss how future observations and simulations may distinguish halo
+formation scenarios. Finally, we summarize the main conclusions of this
+thesis.
 
@@
 
-For both Scl and UMi, these galaxies have been studied extensively in
-both a theoretical and observational context. While many works
-considered mass modelling, star formation histories, chemistry, and
-other observational properties, we focus on the observations and models
-concerning tides and dynamical evolution here.
+\subsection{The role of tides}\label{the-role-of-tides}
 
-However, some cosmological simulations have suggested that tidal effects
-may be ubiquitous. \citep{riley+2024, shipp+2023} find that a majority
-of satellites form substantial streams, and not due to resolution
-effects \citep[see also][]{panithanpaisal+2021}. This result appears to
-be in tension with ours and many previous works. Some reasons
-cosmological and idealized simulations may give opposite results could
-include resolution effects \citep[e.g.,][]{santos-santos+2025},
-additional perturbations from dark matter substructure, tidal
-``pre-processing'' by other infalling satellites, or the streams may be
-below detection limits \citep[e.g.,][]{shipp+2023}. More work is needed
-to understand why streams appear to form more readily in cosmological
-simulations than is observed in the Milky Way or through idealized
-simulations.
+Tidal effects have been long invoked to explain features of dwarf
+galaxies e.g. \ref{sec:scl_umi_obs_tides},
+\citep{tsujimoto+shigeyama2002, mayer+2001a}. Yet, growing work suggests
+tides may not be as commonplace as suspected. For example,
+\citet{read+2006} suggest that the absence of rising radial velocity
+dispersion profiles in dwarfs implies a lack of tidal disruption.
+\citet{penarrubia+2009}, introducing the break radius as a diagnostic
+(Eq.~\ref{eq:r_break}), showing that most satellite orbits were
+inconsistent with observable tidal effects. Recently,
+\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
+(Eq.~\ref{eq:r_jacobi}), reaching a similar conclusion. With the
+exception of the Sagittarius dSph, these results suggest tides have mild
+influence on the stellar components of classical dwarfs.
 
-Closely related to our work, \citet{iorio+2019} applied idealized N-body
-simulations to study tidal effects on Scl. They similarly found weak
-tidal effects, even for a dark-matter-free model. However, their orbits
-are less well-constrained and they did not consider the LMC.
+A few works have modelled Scl or UMi specifically, finding similar
+conclusions. \citet{iorio+2019} applied idealized N-body simulations to
+study tidal effects on Scl. They similarly found weak tidal effects,
+even for a dark-matter-free model. Most recently,
+\citet{tchiorniy+genina2025} also use idealized simulations tuned to
+five classical dwarfs with a focus on the inner density, concluding that
+tides do not strongly affect equilibrium assumptions or the stellar
+component.
 
-\subsection{Sculptor}\label{sculptor}
+Our work extends these models, adding updated structural properties, a
+broader range of orbital histories, and the influence of the LMC.
+Despite these considerations, we still reach similar conclusions: tidal
+effects do not shape Scl or UMi's stellar component. A non-tidal
+explanation likely underlies the origin of these galaxies' density
+excess.
 
-Scl has long been speculated to be disturbed. \citet{innanen+papp1979}
-found RR Lyrae candidate Scl members \citep[from][]{vanagt1978} out to
-180' in an elongated distribution, speculating this to be tidal
-disruption. Many later density profile determinations noted Scl's
-elongation and apparent outer density excess similar to J+24's detection
-\citep[but see also
-\citet{coleman+dacosta+bland-hawthorn2005}]{eskridge1988, IH1995, walcher+2003, westfall+2006}.
-Scl's ellipticity and extended density profile were often interpreted as
-evidence of tidal debris or sometimes a dwarf galaxy halo.
+\subsection{Peculiarities in the
+galaxies}\label{peculiarities-in-the-galaxies}
 
-In addition, Scl hosts at least two populations, as revealed through
-photometry \citep{tolstoy+2004}, kinematics
-\citep{battaglia+2008, tolstoy+2023, arroyo-polonio+2024}, dynamical
-structure \citep{breddels+helmi2014}, age gradients \citep{deboer+2011},
-and chemistry \citep{kirby+2009}. These two populations both have radii
-smaller than the outer density excess. Scl challenges the idea of a
-simple, single-component dwarf spheroidal population and hints at
-episodic star formation or hierarchical assembly.
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_fe_h_gradient.pdf}}
+\caption[Metallicity gradients in Scl and UMi]{Density profiles and
+metallicity gradients in Scl (left) and UMi (right). In the top panel,
+we plot the J+24 density profiles as function of elliptical radius in
+black, and the two components of a double-exponential density fit as
+curved, transparent lines. The vertical lines represent the transition
+from the inner to the outer exponential profile. \textbf{Bottom}:
+Metallicies as a function of elliptical radius for stars from
+\citet{tolstoy+2023}; \citet{apogee}; and \citet{sestito+2023} (left)
+and \citet{pace+2020}; \citet{apogee}; and \citet{sestito+2023b}
+(right)}\label{fig:metallicity_gradients}
+\end{figure}
 
-\subsection{Ursa Minor}\label{ursa-minor}
+Besides extended stellar density profiles, Scl and UMi display other
+peculiarities which may hold clues to their formation. Scl and UMi both
+hosts at least two chemodynamic populations, as revealed through their
+photometric or metallicity-velocity structure
+\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner, compact
+population is younger, more metal-rich, and dynamically colder, whereas
+the outer, extended population older, more metal-poor, and dynamically
+hotter.
 
-UMi has garnered claims of inner substructure such as stellar or
-kinematic ``clumps''. Studies by \citet{olszewski+aaronson1985},
-\citet{demers+1995}, \citet{IH1995}, \citet{kleyna+1998},
-\citet{battinelli+demers1999}, and \citet{bellazzini+2002} note that
-Ursa Minor appears to contain ``clumps'' along its major axis. One clump
-has furthermore been shown to be kinematically distinct and colder
-\citep[e.g.,][]{pace+2014}. If a star cluster, than the cluster should
-disolve in \(\sim 3\) Gyrs provided the UMi's halo is not cored and did
-not interact with dark (sub)subhalos---the survival of such substructure
-depends on the nature of dark matter \citep{kleyna+2003, lora+2012}.
-\citet{wilkinson+2004} additionally find a puzzling drop in the velocity
-dispersion in the outskirts. The nature of any clumps or substructure
-remains unclear, but \citet{munoz+2018}'s are unable to find signs of
-any substructure with modern, deeper photometry.
+Could these internal stellar populations map to the observed density
+excess? In Scl, both populations have scale radii likely smaller than
+the half-light radius. But, the third, more tentatively detected
+population in \citet{arroyo-polonio+2024} would reside within the
+stellar halo. In Ursa Minor, the transition between the metal-rich and
+metal-poor stellar components occurs at \(\sim 30\,\)arcmin
+\citep{pace2020}, consummate with the beginning of the density excess.
+This metal-poor population may thus be related to the stellar halo of
+Ursa Minor. Fig:metallicity\_gradients shows the radial metallicity
+distribution in both galaxies. Scl and UMi both show evidence of a
+metallicity gradient, but the outskirts have few stars. Further
+observations would be necessary to understand if the stellar halo
+represents a chemically distinct population.
 
-In addition, several works have found supposed tidal features in UMi.
-\citet{hargreaves+1994} first detected a velocity gradient in UMi,
-suggestive of tidal disruption. Later, \citet{martinez-delgado+2001}
-find that stars extend far beyond the ``tidal radius'' (from a King
-profile fit) for Ursa Minor, in the direction of the galaxy's elongation
-\citep[see corresponding simulations
-by][]{gomez-flechoso+martinez-delgado2003}. \citet{palma+2003} further
-showed evidence for S-shaped contours, and an extended population of
-``extratidal'' stars. Our density profiles are consistent with these
-works. However, strong evidence of tidal disruption has not yet been
-found.
+Ursa Minor has also shown evidence for possible inner substructure such
+as stellar or kinematic ``clumps''
+\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
+While one clump was re-detected kinematically in \citet{pace+2014},
+\citet{munoz+2018} find no evidence of substructure with modern
+photometry. In any case, the survival of a cold clump depends on the
+nature of dark matter, disrupting quickly in a \LCDM{} halo
+\citep{kleyna+2003, lora+2012}.
+
+Given their multiple stellar populations and possible inner
+substructure, this hints that Scl and UMi's distinct structure today
+arises from an unusual past. A mechanism forming a stellar halo could
+also explain multiple populations.
 
@@
 
-As a tidal origin of Scl and UMi's extended density profiles is
-disfavoured, we consider alternative processes which may form a dwarf
-galaxy ``stellar halo''. Some recent works have also confirmed the
-presence of extended, likely non-tidal, density profiles. For example,
-\citet{chiti+2021}; \citet{chiti+2023} spectroscopically confirm members
-out to \(~9 R_h\) in Tucana II. Given these stars are misaligned with
-the orbit and the lack of a velocity gradient, tides seem to be an
-unlikely explanation for Tucana II as well. The prevalence of stellar
-halos around dwarf galaxies remains an active research topic.
+As we disfavour a tidal origin for Scl and UMi's extended density
+profiles, we now consider alternative processes which may form a stellar
+halo in these dwarfs.
 
-Many dwarf galaxies also host multiple chemodynamical stellar
-populations. Typically, older populations are extended, kinematically
-hotter, and metal-poo , whereas the younger populations are more
-compact, colder, and metal-rich. Examples include Carina
-\citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
-Fornax \citep[\citet{amorisco+evans2012},
-\citet{delpino+aparicio+hidalgo2015}]{battaglia+2006}, Sextans
-\citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
-Andromeda II
-\citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}. Evidence
-of multiple-populations in dwarf galaxies suggests that effects like
-galaxy mergers or episodic star formation shape the stellar structure of
-dwarf galaxies.
+\textbf{Episodic star formation.} Star formation may quench and
+reignite, creating successive stellar generations with differing
+distributions. External star formation triggers include tidal
+compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
+gaseous filaments \citep{genina+2019}, perturbations from dark halos
+\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
+\citep{wright+2019}. More common mechanisms, like feedback or
+reionization-driven quenching, may also form multiple stellar
+generations
+\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
+such processes would not explain why stellar halo formation may be
+non-universal. If a stellar halo was formed in an older, induced burst
+of star formation, then the star formation history should contain
+evidence of a corresponding burst.
 
-We now explore possible origins of a ``stellar halo'' which may resemble
-the extended density profiles in Scl and UMi.
+\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
+\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
+the lower-mass galaxy are preferentially dispersed, forming a stellar
+halo and population gradient \citep{benitez-llambay+2016, deason+2022}.
+If the galaxies contains gas, the merger can trigger new star formation,
+forming a younger population of stars \citep[e.g.,][]{genina+2019}. A
+few local dwarfs are suspected to have undergone a major merger,
+including Tucana II, Andromeda II, and Phoenix
+\citep{lokas+2014, fouquet+2017, tarumi+2021, cardona-barrero+2021, querci+2025}.
+A galaxy having undergone a major merger should show evidence of at
+least two or three populations from distinct origins.
 
-\subsection{Internal processess}\label{internal-processess}
+\textbf{Minor mergers / accretion}. Just as the Milky Way's halo is
+believed to be built from many minor mergers, dwarf galaxies may build
+stellar halos composed of even fainter satellites
+\citep{ricotti+polisensky+cleland2022}. Few examples are known, but a
+stream around And II may be an instance of dwarf accretion
+\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
+the stellar halo's would contain chemical signatures from several
+distinct ultra-faint dwarf galaxies.
 
-\textbf{Dynamical heating of old stars}. In a galaxy, older stars
-generally have higher random velocities (i.e., kinematically ``hotter'')
-than younger stars. In dwarf galaxies, several mechanism may heat
-stellar components, including supernovae kicks in forming stars,
-feedback-driven potential fluctuations
-\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021}, or
-heating by dark sub-subhalos \citep{penarrubia+2025}. Over time, these
-processes naturally produce a more extended, older stellar population.
+\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
+``preprocessed'' by a massive satellite like the LMC
+\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
+redistributes already-present stellar populations without creating new
+populations, and may mimic a stellar halo. Key evidence suggesting
+preprocessing may include distant dwarf stars or stellar streams.
 
-\textbf{Episodic star formation and feedback.} Dwarf galaxies are
-thought to experience bursty star formation--i.e., consisting of several
-episodes of intense star formation separated by periods of quiescence
-\citep[e.g.,][]{salvadori+ferrara+schneider2008, valcke+derijcke+dejonghe2008, wheeler+2019, azartash-namin+2024}.
-Stellar feedback in a dwarf galaxy's shallow potential well drives
-oscillations in the star formation rate. Alternatively, re-ionization
-may temporarily suspending star formation \citep{benitez-llambay+2015}.
-In many simulations, shrinking gas reservoirs form
-centrally-concentrated later generations of stars, naturally spawning
-multiple populations \citep{kawata+2006, revaz+jablonka2018}. However,
-star formation is also sensitive to a dwarf's environment.
-
-\subsection{External processes}\label{external-processes}
-
-\textbf{Induced star formation.} Star formation may quench and reignite
-due to an external perturbation. Examples include tidal compression
-\citep{mayer+2001a, dong+lin+murray2003}, gaseous filaments
-\citep{genina+2019}, dark halos \citep{starkenburg+helmi+sales2016}, or
-shocks with the MW corona \citep{wright+2019}. Induced burst would be
-more stochastic than internal bursts, possibly explaining diversity in
-dwarf galaxy structure.
-
-\textbf{Major mergers.} When galaxies merge, they may leave signatures
-such as population gradients and stellar halos. Classical dwarfs have a
-\(\sim 10\%\) chance of undergoing a major merger since redshift \(z=1\)
-\citep{deason+wetzel+garrison-kimmel2014}. Mergers may preferentially
-disperse the lower-mass galaxy's stars, forming a halo and a metallicity
-gradient \citep{benitez-llambay+2016}. Intermediate-mass mergers (with
-mass-ratios of \(\sim\) 1:5) most effectively populate halos, balancing
-stellar mass from larger galaxies with better dispersion of lower-mass
-galaxies \citep{deason+2022}. Tuc II's properties are suggested to
-originate from a similar merger
-\citep{tarumi+yoshida+frebel2021, querci+2025}.
-
-\textbf{Gas-rich mergers.} If a merger occurs between gas-rich galaxies,
-triggered star formation may occur in the aftermath
-\citep[e.g.,][]{genina+2019}. And II and Phoenix have steep metallicity
-gradients and unusual prolate rotation, theorized to result from mergers
-of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.
-
-\textbf{Minor mergers / accretion}. Like how the Milky Way's halo is
-believed to be built from many minor mergers, dwarf galaxies may form
-halos through accretion of yet fainter dwarfs.
-\citet{ricotti+polisensky+cleland2022} demonstrated that accretion of
-ultra-faint dwarfs may create dwarf stellar halos. A stream detected
-around And II further supports the occurrence of mergers among dwarfs
-\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
-the halo chemistry would resemble a population of ultra-faint dwarfs.
-The occurance of dwarf mergers would further constrain small-scale
-galaxy formation.
-
-\textbf{Tidal preprocessing}. Dwarf galaxies may have been
-``preprocessed'' by a larger satellite like the LMC
-\citep[e.g.,][]{santistevan+2023, riley+2024}. From the orbit
-integrations above, it is possible that UMi was once an LMC satellite.
-However, the prevalence of preprocessing remains uncertain. Like stellar
-heating, preprocessing redistributes already-present stellar
-populations.
-
-\textbf{Tidal dwarf galaxies} are cluster-like objects which form in
-gas-rich tidal streams created during the merger of two galaxies
-\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Tidal
-dwarfs may be more susceptible to tides, forming extended density
-profiles and appearing to have dark matter
-\citep{casas+2012, yang+2014, wang+2024a}. If Scl and UMi are tidal
-dwarfs, a stronger velocity (dispersion) gradient and tidal tails should
-be detectible.
-
-\subsection{\texorpdfstring{Beyond \LCDM{}}{Beyond }}\label{beyond}
-
-\textbf{Modified Newtonian Dynamics (MOND)}. MOND modifies gravity
-instead of using dark matter to explain the rotation curves of galaxies.
-In MOND, tides may more strongly affect dwarf galaxies owing to the lack
-of dark matter mass and the stronger MW tidal field
-\citep{mcgaugh+wolf2010, brada+milgrom2000}. A tidal origin of the
-density excess is more likely in this case. If MOND is to recover the
-observed velocity dispersions of satellites, then many more dwarfs may
-be actively tidally disrupting as well
+\textbf{Dark-matter free dwarfs.} Tidal dwarfs form in gas-rich tidal
+streams created during the merger of two galaxies
+\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
+dark matter halo, tidal dwarfs may be more susceptible to tides, forming
+extended density profiles and inflating the velocity dispersion
 \citep{casas+2012, yang+2014, wang+2024a}.
 
-\textbf{Self-interacting dark matter (SIDM).} SIDM significantly
-complicates tidal evolution. SIDM halos in isolation are not static---by
-transferring heat through collisions, these halos first undergo core
-formation and then core collapse. Besides structural changes, SIDM adds
-that pressure from the host DM halo can change the structure
-(e.g.~aiding core collapse) or remove material from the inner galaxy
-(analogous to ram pressure stripping) \citep[e.g.,][]{cartonzeng+2024}.
-An SIDM halo may be more strongly affected by tides, but the velocity
-dispersion makes a tidal disruption in SIDM still unlikely.
+Similarly, in Modified Newtonian Dynamics (MOND), galaxies do not
+contain dark matter. Instead the gravitational force law is altered to
+explain rotation curves. Dwarfs would similarly experience stronger
+tidal effects in MOND \citep{mcgaugh+wolf2010, brada+milgrom2000}.
 
-\textbf{Fuzzy dark matter} can also heat up stars owing to density
-perturbations in the form of interference fringes
-\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. Similar to other
-intrinsic heating methods, this model would likely affect all dwarfs
-similarly, so we should detect similar halos around similar dwarfs.
+In both cases, tides more plausibly produce extended density profiles.
+However, recovering the observed velocity dispersions often requires
+ongoing tidal disruption
+\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
+also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
+evidence of such features (see Section~\ref{sec:extra_rv_models}), so
+future data would need to uncover signs of disruption to support this
+hypothesis.
 
-\subsection{Disentangling the origin of a stellar
-halo.}\label{disentangling-the-origin-of-a-stellar-halo.}
+\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
+than younger stars due to processes including stellar feedback
+\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
+sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
+matter interference fringes
+\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
+these processes should operate similarly across dwarf galaxies---stellar
+halos should be more common if these processes form substantial halos.
+
+\section{Disentangling the origin of a stellar
+halo}\label{disentangling-the-origin-of-a-stellar-halo}
 
@@
 
-\textbf{Chemistry}. Large samples of detailed chemical abundances have
-been invaluable for understanding MW substructure (e.g.~for
-\emph{Gaia}-Sausage Enceladus). Chemistry, particularly comparing the
-inner and outer regions could test if the halo appears to originate from
-a distinct system than the dwarf, differentiating many \emph{internal}
-versus \emph{external} scenarios.
+\emph{Precise chemistry}. Chemistry, particularly comparing the inner
+and outer regions, could test if the halo appears to originate from a
+distinct system than the dwarf. This would test if the halo is a
+separate chemical population, may reveal the properties of any merged
+galaxies, or show if the halo is consistent with originating from the
+same galaxy.
 
-\textbf{Star formation histories}. Evidence of significant star
-formation episodes or lack thereof may differentiate scenarios which
-rely on strong star formation bursts (e.g.~\emph{episodic star formation
-history, induced star formation, and gas-rich mergers.})
-
-\textbf{Kinematics.} Particularly, for models relying on recent tidal
+\emph{Detailed kinematics}. For models relying on recent tidal
 disruption, kinematic disequilibrium features should be visible. These
@@
 outward-biased moving stars, or non-phase mixed structures
-\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.
+\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}. In
+addition, measurements of the ellipticity and anisotropy would help
+verify or alter our understanding of the mass structure and if DM-free
+models may be permissible.
 
-\textbf{Deep photometry} may find or rule out signs of dynamical
+\emph{Deep photometry} may find or rule out signs of dynamical
 disequilibrium and tidal tails for tidally susceptible models
-\emph{(e.g., MOND and tidal dwarfs.)}
+\emph{(e.g., pre-processing and DM-free dwarfs.)} In addition,
+photometry will help constrain the prevalence and nature of extended
+features in dwarf galaxies.
 
-Ongoing, upcoming, and future facilities will be essential for testing
-these theories. For example, given the typical faint magnitudes of dSph
-stars, chemistry is best accessible through large field-of-view
-multi-object spectroscopic instruments on large telescopes (e.g., 4MOST
-and extremely large telescopes \citet{skuladottir+2023}). 3D internal
-kinematics of dwarfs may require a successor to \emph{Gaia}. However,
-using JWST and HST enables precise proper motions for small regions
-within dwarf galaxies \citep[e.g.,][]{vitral+2025}, which may be
-sufficient to constrain vastly different internal dynamic structures and
-ongoing tidal disruption. Finaly, Photometric surveys by Rubin
-Observatory and Euclid will most-emminently probe yet fainter magnitudes
-around dwarf galaxies and possibly find or constrain stellar halos and
-their star formation histories. Altogether, new surveys will likely
-uncover novel aspects about the inner workings and outer halos of dwarf
-galaxies.
+\emph{Star formation histories} can be derived through photometry or
+chemistry. Evidence of significant star formation episodes or lack
+thereof may differentiate scenarios which rely on strong star formation
+bursts (e.g.~\emph{episodic star formation history, induced star
+formation, and gas-rich mergers.})
 
-In addition to new observing facilities, the next generation of dwarf
-simulations are beginning to answer questions and unravel processes in
-the formation of dwarf galaxies. With improved resolution, realism, and
-physics, the frequency and effects of various mechanisms discussed can
-be better constrained.
+Ongoing and future facilities will be essential for testing dwarf galaxy
+halo formation. Given the typical faint magnitudes of dSph stars,
+chemistry is best accessible through multi-object spectroscopic
+instruments on large telescopes (e.g., 4MOST and extremely large
+telescopes \citet{skuladottir+2023}). 3D internal kinematics of dwarfs
+may require a successor to \emph{Gaia}. However, using JWST and HST
+enables precise proper motions for small regions within dwarf galaxies
+\citep[e.g.,][]{vitral+2025}, which may be sufficient to constrain
+vastly different internal dynamic structures and ongoing tidal
+disruption. Finally, ongoing photometric surveys by Rubin Observatory
+\citep{ivezic+2019} and the Euclid mission
+\citep{euclidcollaboration+2025} will probe even fainter features around
+dwarf galaxies, possibly finding more halos, tidal streams, and
+improving our understanding of their stellar populations. Altogether,
+new surveys will likely uncover novel aspects about the inner workings
+and outer halos of dwarf galaxies.
+
+The next generation of dwarf simulations are beginning to answer
+questions and unravel processes in the formation of dwarf galaxies.
+Notable recent simulation projects include EDGE \citep{agertz+2020} and
+LYRA \citep{gutcke+2021}, both of which can simulate the cosmological
+formation of isolated dwarf galaxies while resolving individual
+(massive) stars. By including more realism (e.g., black holes, magnetic
+fields), refining models of poorly understood processes (feedback and
+star formation), and improving numerical resolution, simulations will
+continue to explore these processes and refine our understanding of
+dwarf galaxy formation.
 
@@
 In this thesis, we have investigated the extended, outer density excess
-of Sculptor and Ursa Minor and its possible tidal origin. We first
-verified that Scl and UMi have unusually extended density profiles,
-compared to other dwarf spheroidals. We show that the density profiles
-are robust to alternate data criteria, implying that this ``density
-excess'' is likely a real feature of each galaxy.
+of Sculptor and Ursa Minor and its possible tidal origin. We show that
+the density profiles are robust, implying that Scl and UMi contain a
+true density excess relative to other classical dwarfs.
 
-We then investigated if tides were a permissible explanation. By
-modelling each galaxy based on cosmological initial conditions, we
-showed that tides do not strongly affect either galaxy. The LMC changes
-the orbital history of Scl and UMi, and tides become even weaker in a
-combined LMC and MW potential. While UMi may form a stellar stream, the
-stream is far fainter than is presently detectible. We conclude that
-recent tides are unlikely to shape the observed stellar distributions of
-Scl and UMi.
+We then investigated if tides are a possible explanation. By modelling
+each galaxy based on cosmological initial conditions, we showed that
+tides do not strongly affect either galaxy. The LMC changes the orbital
+history of Scl and UMi, but tides become even weaker in a combined LMC
+and MW potential. We conclude that recent tides are unlikely to shape
+the outer radial distribution of stars in Scl and UMi.
 
-Finally, we consider alternative scenarios forming extended density
-distribution. We review mechanisms ranging from episodic star formation
+Finally, we discuss the implications of a non-tidal density excess in
+Scl and UMi. We review mechanisms ranging from episodic star formation
 histories, to accretion, to departures from \LCDM{}. While a more
@@
 the stellar populations within dwarf galaxies in unprecedented detail.
-Future work will the Milky Way, uncovering the cosmic origins of the
-smallest structures forming building blocks of galaxies like ours.
+Future work will continue to uncover the cosmic origins of the smallest
+structures forming building blocks of galaxies like ours.

### 2025-10-02 thesis/extra_density.tex
discussion revisions

@@
 
-In this Appendix, we describe the details of J+24's likelihoods. We then
-test the reliability of the derived density profiles of the dwarfs,
-comparing methodologies, samples, and to literature. In conclusion, the
-density profiles from J+24 converge among methods out to a
-background-limited radius, derived here.
+In this section, we describe the details of J+24's likelihoods. We then
+test the reliability of the derived density profiles for Scl and UMi,
+comparing methodologies, samples, and to literature. Finally, we present
+a comparison of J+24's samples to a non-parametric Bayesian-derived
+density profile, finding similar results except in the very outer
+regions. Past a ``limiting radius'', the density profiles become
+susceptible to the likelihood specification and may select mostly
+background stars. In conclusion, the density profiles from J+24 converge
+among methods out to our limiting radii derived here.
 
@@
 \caption[Scl literature density profiles]{A comparison of the Scl
-derived density profile and historical works.}
+derived density profile and historical works. Unlike most profiles in
+this thesis, this density profile is plotted with respect to the
+semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
+The solid black line is a 2D exponential with corresponding scale radius
+to Scl's half-light radius, and residuals in the lower panel are
+relative to this exponential.}
 \end{figure}
@@
 Minor.}\label{fig:umi_density_extras}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
+\caption[UMi density profiles in literatures]{Similar to}
 \end{figure}

### 2025-10-02 thesis/extra_numerical_convergence.tex
discussion revisions

@@
 
-Here, we describe some convergence tests to ensure our methods and
-results are minimally impacted by numerical limitations and assumptions.
-See \citet{power+2003} for a detailed discussion of various assumptions
-and parameters used in N-body simulations.
+Here, we describe the numerical specifics of our simulation and
+analysis. We present convergence tests supporting our choices of
+softening and additional parameters. We find alternate choices of
+numerical parameters are unlikely to improve our convergence or result
+in significantly different evolution.
 
@@
 number of particles within \(R_{200}\). This choice balances integration
-time and only compromises resolution in collisional regime.
-
-For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
-particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
-slightly smaller softening which was reduced by a factor of
-\(\sqrt{10}\) which appears to improve agreement slightly in the
-innermost regions.
+time and only compromises resolution in collisional regime. For our
+isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
+particles, this works out to be \(0.044\,{\rm kpc}\), about
+\(\sqrt{10}\) times larger than our adopted softening
+(Eq.~\ref{eq:softening_length}).
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
-\caption{Softening convergence}\label{fig:softening_convergence}
+\caption[Softening convergence]{Similar to
+Fig.~\ref{fig:numerical_convergence} except for simulations with
+different softening lengths, \(h_{\rm grav}\). \(0.044\,\kpc\) is our
+fiducial softening length for this halo. All simulations here were ran
+in isolation of 5 Gyr with \(10^{6}\)
+particles.}\label{fig:softening_convergence}
 \end{figure}
 
-\section{Time stepping and force
+Fig.~\ref{fig:softening_convergence} illustrates the influence of
+softening length on isolation evolution. The simulation with
+\(h=0.14\,\kpc\) deviates the most from the expected profile. Instead,
+the final profiles for \(h=0.044\,\kpc\) and \(h=0.014\,\kpc\) are
+nearly identical. Smaller values of softening than \(h=0.044\,\kpc\)
+likely will not affect the evolution in the inner regions. As
+computation time increases moderately with decreasing softening length,
+\(h=0.044\,\kpc\) is an ideal choice balancing efficiency and accuracy.
+Our simulations therefore adopt this choice (as in
+Eq.~\ref{eq:softening_length}).
+
+\section{Time-stepping and force
 accuracy}\label{time-stepping-and-force-accuracy}
 
-In general, we use adaptive timestepping and relative opening criteria
+In general, we use adaptive time-stepping and relative opening criteria
 for gravitational force computations. To verify that these choices and
 associated accuracy parameters minimally impact convergence or speed, we
-show a few more isolation runs (using only 1e5 particles)
+show a few more isolation runs (using only \(10^5\) particles).
 
-\begin{itemize}
-\tightlist
-\item
-  constant timestep (\ldots), approximantly half of minimum timestep
-  with adaptive timestepping
-\item
-  geometric opening, with \(\theta = 0.5\).
-\item
-  strict integration accuracy, (facc = \ldots.)
-\end{itemize}
+We use in \gadget{} the relative tree opening criterion with the
+accuracy parameter set to \(\alpha =0.005\) (so nodes are opened when
+\(M\,l/r^3 < \alpha |a|\) for a distance from the node of mass \(M\) of
+\(r\), \(l\) is the side length of the node, and \(a\) is the particle's
+total acceleration), and adaptive time stepping with integration
+accuracy set to \(\eta=0.01\) (particles must take time-steps smaller
+than \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).
+
+Fig.~\ref{fig:methods_convergence} tests various additional simulation
+parameters in isolation. We compare against an equivalent model ran
+instead with \texttt{GADGET-2}, a model ran with constant timestep of
+0.5Myr (\texttt{dt0.1} named after our dimensionless units), a
+simulation with the time-step accuracy set to \(\eta=0.003\) (high
+accuracy), and a simulation using a geometric opening criterian instead
+of the force (geometric). In all cases, the final isolation velocity
+profile is nearly the same after 5Gyr. The numerical details likely do
+not limit our convergence (as marked by the arrow).
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
-\caption{Isolation method convergence}\label{fig:methods_convergence}
+\caption[Isolation method convergence]{A comparison against different
+integration methods. See text for details on each
+model.}\label{fig:methods_convergence}
 \end{figure}

### 2025-10-02 thesis/extra_rv_models.tex
discussion revisions

@@
 In this section, we analyze the observed line-of-sight (LOS) velocity
-distributions for Scl and UMi. Our goal is to derive the systemic
-velocity and velocity disperions, and understand if the galaxies show
-evidence of tidal features, e.g.~a velocity gradient or a rising outer
-velocity dispersion. Our derived velocity distributions are consistent
-with literature. We find weak evidence for a velocity gradient in Scl
-and a rising outer velocity dispersion. However, Scl's velocity gradient
-is misaligned with its proper motion, likely inconsistent with a tidal
-interpretation. We find no evidence of a gradient in mean velocity or
+distributions for Sculptor (Scl) and Ursa Minor (UMi). Our goal is to
+understand if the galaxies show evidence of tidal features---for
+example, a velocity gradient or rising outer velocity dispersion. Our
+derived systemic velocities and dispersions are consistent with
+literature. We find weak evidence for a velocity gradient in Scl and a
+rising outer velocity dispersion. However, Scl's velocity gradient is
+misaligned with its proper motion, likely inconsistent with a tidal
+interpretation. Scl's velocity dispersion also mostly rises within the
+effective radius, more likely a feature of inner structure than tidal
+disruption. We find no evidence of a gradient in mean velocity or
 velocity dispersion in UMi. We conclude that Scl and UMi do not show
-observable features of tidal disruption given current velocity
-observations.
+clear features of tidal disruption given current velocity observations.
 
@@
 
-For both Sculptor and Ursa Minor, we construct literature samples of
-radial velocity measurements. We combine these samples with J+24's
-Bayesian likelihoods to produce likely members using additional RV
-information.
-
-First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
-not report GaiaDR3 source ID's, we match to the nearest star within 1--3
-arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
-mean RV measurement from each study using the inverse-variance weighted
-mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
-variance \(s^2\).
-
-We remove stars with significant velocity dispersions as measured
-between observations in a study or between studies, to filter out
-unreliable measurements or possible spectroscopic binaries. By using
-that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
-\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
-distribution with \(N-1\) measurements. This cut typically removes stars
-with reduced chi-squared values
-\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
-number of measurements is 1-3 typically)..
-
-Next, we need to correct the coordinate frames for the solar motion and
-on-sky size of the galaxy. We transform the frame into the galactic
-standard of rest (GSR). The next step is to account for the slight
-differences in the direction of each radial velocity. Let the \(\hat z\)
-be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
-the angular distance between the centre of the galaxy and the individual
-star, the corrected radial velocity is then \begin{equation}{
-v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
-}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the line of
-sight velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
-tangental velocities in RA and Dec, and \(\theta\) is the position angle
-of the star with respect to the centre of the dwarf. The correction from
-both effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
-Sculptor and less for Ursa Minor \citep[see
-also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
-the distance uncertainty and velocity dispersion in quadrature to the RV
-uncertainties for each star. We then use the \(v_z\) values for the
-following modelling, however repeating with uncorrected, heliocentric
-velocities does not significantly affect the results.
-
-The combined likelihood, including RV information, becomes
-\begin{equation}{
-{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
-}\end{equation} where we assume that the satellite and background
-distributions are Gaussian. Specifically, \begin{equation}{
-\begin{split}
-{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
-{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
-\end{split}
-}\end{equation} where \(f\) is the probability density of a standard
-normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
-velocity and dispersion of the satellite, and \(\delta v_i\) is the
-individual measurement uncertainty. Typically, the velocity dispersion
-dominates the uncertainty budget. We assume a halo/background velocity
-dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
-\citep[e.g.][]{brown+2010}.
-
-Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2. Like in J+24, most stars have
-probabilities of nearly 1 or 0, so the precise choice of cut marginally
-affects the resulting sample.
-
-\section{MCMC modeling}\label{mcmc-modeling}
-
-We use Monte-Carlo Markov chains (MCMC) to calculate the posterior
-distributions of our models of the intrinsic velocity dispersion and
-possible velocity gradients in Scl and UMi.
-
-The likelihood is based on the distribution of stars with respect to the
-model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
-in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
-likelihood is \begin{equation}{
-\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
-}\end{equation} where \(f\) is the probability distribution of a
-standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
-star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
-velocity dispersion at the \(i\)th star's position \(\xi_i\) and
-\(\eta_i\).
-
-For all models, we ado \begin{equation}{
-\mu_{0} \sim N(0\,\kms, \sigma^2_{\rm halo})\\
-\sigma_{0} \sim U(0\,\kms, 20\,\kms),
-}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
-the velocity dispersion of the MW halo adopted above, a reasonable
-assumption for dwarfs in orbit around the MW.
-
-For the simplest model, \(\mu\) and \(\sigma\) are held
-fixed---\(\mu(\xi, \eta) = \mu_0\) and \(\sigma(\xi, \eta) = \sigma_0\).
-There are no additional parameters besides \(\mu_0\) and \(\sigma_0\).
-
-To model a velocity gradient, we add two parameters, the gradient
-\begin{equation}{
-\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta\\
-\sigma = \sigma_0
-}\end{equation}
-
-To model an increasing velocity dispersion, we add a third parameter,
-\(c\) which applies as \begin{equation}{
-\log \sigma(R_{\rm ell}) = \log \sigma_0 + c\,\log\left(R_{\rm ell} / R_h\right) \\
-\mu = \mu_0 + a\,\xi + b\,\eta\\
-}\end{equation} with \(c \sim U(0, 1)\) where \(R_{\rm ell}\) is the
-elliptical radius. We only consider this model including the velocity
-gradient as a velocity gradient is the simplest way to increase the
-dispersion gradient.
-
-Bayes factors quantify the relative evidence between Bayesian model. If
-the Bayes factor
-
-Our models are \emph{nested}, where the fiducial model (\(\sigma_0\) and
-\(\mu_0\) only) always arises from setting parameters of the more
-complex models to zero (\(a=0, b=0\) and/or \(c=0\) ). As a result, we
-can use the Savage-Dickey method to calculate the Bayes factors
-\citep{dickey+lientz1970}, where the Bayes factor is then the relative
-density of posterior versus prior samples where \(a=b=c=0\). To
-calculate densities, we use a Silverman-bandwidth KDE to smooth the
-samples.
-
-\section{Results}\label{sec:rv_results}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
-terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
-points with error bars are from the crossmatched observed sample, and
-green lines represent MCMC fits to the velocity dispersion.}
-\end{figure}
+For both Sculptor and Ursa Minor, we construct literature samples of LOS
+velocity measurements. We combine these samples with J+24's Bayesian
+likelihoods to produce LOS velocity-informed membership probabilities.
+We select stars with moderate to large membership probabilities for
+future analysis.
 
@@
 
-\input{rv_table.tex}
+First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
+not report GaiaDR3 source ID's, we match to the nearest star within 1--3
+arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
+mean RV measurement from each study using the inverse-variance weighted
+mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
+variance \(s^2\).
 
-\begin{table*}[t]
-\centering
-\caption[Ursa Minor RV fits]{MCMC fits for UMi velocity dispersion. }
-\label{tbl:umi_rv_mcmc}
-\begin{tabular}{lllll}
-\toprule
-study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
-\midrule
-all & $-245.8\pm0.3$ & $8.8\pm0.2$ & +1.3 & +0.9\\
-pace & $-244.6\pm0.4$ & $9.0\pm0.3$ & +0.3 & +0.5\\
-spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.06\\
-apogee & $-245.6\pm1.2$ & $10.0_{-0.8}^{+1.0}$ & +1.0 & +0.5\\
-\bottomrule
-\end{tabular}
-\end{table*}
+We next remove stars with significant velocity dispersions as measured
+between observations in a study or between studies, to filter out
+unreliable measurements or possible spectroscopic binaries. By using
+that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
+\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
+distribution with \(N-1\) measurements. This cut typically removes stars
+with reduced chi-squared values
+\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (as a star
+typically fewer than four velocity measurements).
+
+We correct the velocities for the solar motion and on-sky size of the
+galaxy. First, we transform the velocities into the galactic standard of
+rest (GSR), a reference frame centred at the Sun's position but at rest
+with respect to the Galactic centre. Next, we transform velocities to
+correct for the apparent gradient induced by the dwarf's proper motion.
+Let the \(\hat z\) be the direction from the sun to the dwarf galaxy.
+Then if \(\phi\) is the angular distance between the centre of the
+galaxy and the individual star, the corrected radial velocity is then
+\begin{equation}{
+v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
+}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the LOS
+velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
+tangental velocities in RA and Dec in the GSR frame, and \(\theta\) is
+the position angle of the star with respect to the centre of the dwarf.
+The correction from both effects induces an apparent gradient of about
+\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor \citep[see
+also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
+the distance uncertainty and velocity dispersion in quadrature to the RV
+uncertainties for each star. We use the \(v_z\) values for the following
+modelling.
+
+We build on J+24's likelihood framework to add velocity information (see
+Section~\ref{sec:extra_density}). The total likelihood of a star
+belonging to the satellite or foreground/background population is
+\begin{equation}{
+{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm los},
+}\end{equation} where we have added \({\cal L}_{\rm los}\), LOS-velocity
+likelihood term. We assume that the satellite and background
+\(v_{\rm los, gsr}\) distributions are Gaussian. Specifically,
+\begin{equation}{
+\begin{split}
+{\cal L}_{\rm los, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
+{\cal L}_{\rm los, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
+\end{split}
+}\end{equation} where \(f\) is the probability density of a standard
+normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
+velocity and dispersion of the satellite, and \(\delta v_i\) is the
+individual measurement uncertainty. Typically, the velocity dispersion
+dominates the uncertainty budget. We assume a halo/background velocity
+dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
+\citep[e.g.][]{brown+2010}.
+
+Similar to above, we retain stars with the resulting membership
+probability of greater than 0.2. Like in J+24, most stars have
+probabilities of nearly 1 or 0, so the precise choice of cut marginally
+affects the resulting sample.
+
+\section{MCMC modeling}\label{mcmc-modeling}
+
+To model for possible gradients. We use Monte-Carlo Markov chains (MCMC)
+to calculate the posterior distributions of our models of the intrinsic
+velocity dispersion and possible velocity gradients in Scl and UMi.
+
+The likelihood is based on the distribution of stars with respect to the
+model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
+in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
+likelihood is \begin{equation}{
+\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
+}\end{equation} where \(f\) is the probability distribution of a
+standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
+star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
+velocity dispersion at the \(i\)th star's position \(\xi_i\) and
+\(\eta_i\).
+
+We consider a single model designed to derive systemic velocity and the
+velocity dispersion, detect velocity gradients, and detect a rising
+velocity dispersion. The mean velocity at a given point on the sky,
+\(\mu\), is assumed to be, \begin{equation}{
+\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
+}\end{equation}
+
+for tangent-plane coordinates \(\xi\) and \(\eta\), systemtic velocity
+\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
+dispersion at a given position, \(\sigma\), is assumed to depend as a
+power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
+\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
+}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
+at \(R_h\), and \(c\) is the velocity dispersion gradient slope.
+
+Our model has parameters with priors \begin{equation}{
+\mu_{0} \sim N(0, \sigma^2_{\rm halo}) \,\kms\\
+\sigma_{0} \sim U(0, 20)\ \kms \\
+a \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
+b \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
+c\sim N(0, 0.3^2)\ {\rm dex} \\
+}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
+the velocity dispersion of the MW halo adopted above,
+\(N(\mu, \sigma^2)\) is a standard normal distribution with mean \(\mu\)
+and variance \(\sigma^2\), and \(U(l, h)\) is a uniform distribution
+between \(l\) and \(h\).
+
+Bayes factors quantify the relative evidence between Bayesian models. We
+can use Bayes factors to compare the above model to simpler models
+without gradients in \(\sigma\) or \(\mu\). Assuming that a
+gradient-free model represents the case \(a=b=0\) (gradient free) or
+\(c=0\) (fixed dispersion) models, we can quickly calculate the relative
+Bayes factors without running another model. We use the Savage-Dickey
+method \citep{dickey+lientz1970}, where the Bayes factor is then the
+relative density of posterior versus prior samples when \(a=b=0\)
+(gradient free) or \(c=0\) (fixed dispersion). In order to calculate
+these densities, we use a kernel density estimator on the posterior
+samples with a Silvermann bandwidth in each dimension.
+
+While this is a more complex model, we find each parameter agrees with
+models of reduced complexity.
+
+\section{Results}\label{sec:rv_results}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
-\caption[Scl velocity gradient]{The corrected LOS velocity along the
-best fit rotational axis. RV members are black points, the systematic
-\(v_z\) is the horizontal grey line, blue lines represent the
-(projected) gradient from MCMC samples, and the orange line is a rolling
-median (with a window size of
-50).}\label{fig:scl_velocity_gradient_scatter}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
+\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
+terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
+points with error bars are from the crossmatched observed sample, and
+green lines represent MCMC fits to the velocity dispersion.}
 \end{figure}
+
+\input{rv_table.tex}
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
-\caption[Sculptor's velocity gradient orientation]{RV members of
-Sculptor plotted in the tangent plane coloured by corrected velocity
-difference from mean \(v_z - \bar v_z\) . The black ellipse marks the
-half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
-arrows mark the proper motion (PM, GSR frame) and derived velocity
-gradient (rot) vectors (to scale).}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
+\caption[Scl velocity gradient]{\textbf{Top} members of Sculptor plotted
+in the tangent plane coloured by corrected velocity difference from mean
+\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
+Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
+motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
+scale). \textbf{Bottom}: The corrected LOS velocity along the best fit
+rotational axis. RV members are black points, the systematic \(v_z\) is
+the horizontal grey line, blue lines represent the (projected) gradient
+from MCMC samples, and the orange line is a rolling median (with a
+window size of 50).}\label{fig:scl_velocity_gradient_scatter}
 \end{figure}
@@
 an increasing velocity dispersion with radius. UMi's evidence for a
-radial gradient is weaker.}
+radial gradient is weaker, and Scl's velocity dispersion appears to be
+flat outside of \(R_h\).}
 \end{figure}

### 2025-10-02 thesis/introduction.tex
discussion revisions

@@
 mass-to-light ratios were at least 10 times larger than for globular
-clusters \citep[GCs, e.g.,][]{aaronson1983, aaronson+olszewski1987}.
-While uncertain initially, these values were later corroborated with
-larger and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the
-time, several theories were proposed to explain these unusually high
+clusters \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987}. While
+uncertain initially, these values were later corroborated with larger
+and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the time,
+several theories were proposed to explain these unusually high
 mass-to-light ratios. Examples include: ongoing tidal disruption
@@
   inconsistent with Newtonian dynamics of visible matter alone,
-  \citet{willman+strader2012}} In contrast, star clusters (like GCs)
-have no clear evidence for dark matter. The boundary between these two
-classes blurs for faint, compact stellar associations. Systems with
-shared characteristics of globular clusters and dwarf galaxies, and
+  \citet{willman+strader2012}} In contrast, star clusters (like globular
+clusters) have no clear evidence for dark matter. The boundary between
+these two classes blurs for faint, compact stellar associations. Systems
+with shared characteristics of globular clusters and dwarf galaxies, and
 without evidence as to their nature, are known as ``ambiguous'' systems
-(\textbf{ref}).
+\citep[e.g.,][]{smith+2024}.
 
@@
 classicals, and these systems remain among the best studied. The
-\textbf{ultra-faint}s dwarfs (UFDs) occupy the very faintest magnitude
-range (\(M_V \gtrsim -7.7\) or \(M_\star \lesssim 10^5\,\Mo\)). These
-galaxies have minuscule stellar masses, typically compact sizes, and
-very metal-poor stellar populations \citep[see the review
-by][]{simon2019}. Altogether, known dwarf galaxies span more than 15
-magnitudes, or over 7 decades in stellar mass.
+\textbf{ultra-faint dwarfs} occupy the very faintest magnitude range
+(\(M_V \gtrsim -7.7\) or \(M_\star \lesssim 10^5\,\Mo\)). These galaxies
+have minuscule stellar masses, typically compact sizes, and very
+metal-poor stellar populations \citep[see the review by][]{simon2019}.
+Altogether, known dwarf galaxies span more than 15 magnitudes, or over 7
+decades in stellar mass.
 
 Most well-studied dwarf galaxies lie near the Milky Way or in its
-vicinity, the Local Group (LG) of galaxies. The LG is defined as the
+vicinity, the Local Group of galaxies. The Local Group is defined as the
 group consisting of galaxies within \(\sim 1\) Mpc from the Milky
@@
 
-\subsection{The structure of cold dark matter
-halos}\label{the-structure-of-cold-dark-matter-halos}
+\subsection{The structure of cold dark matter halos}\label{sec:NFW}
 
@@
 \subsection{\texorpdfstring{Galaxy formation in
-\LCDM{}}{Galaxy formation in }}\label{galaxy-formation-in}
+\LCDM{}}{Galaxy formation in }}\label{sec:galaxy_formation}
 
@@
 Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
-(SMHM, with halo mass represented by \(\vmax\)) predicted by the \LCDM{}
+(SMHM, with halo mass represented by \(\vmax\)) predicted by \LCDM{}
 cosmological hydrodynamical simulations of Local Group analogues from
@@
 increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
-formed in relatively massive halos. Because lower mass galaxies have
-increasingly shallow potential wells, feedback becomes more effective at
-removing gas. Re-ionization additionally suppresses late star formation
-in the faintest galaxies. As a result, the resulting stellar mass of a
-dwarf galaxy is highly sensitive to the dark matter halo mass.
+formed in halos of similar mass. Because lower mass galaxies have
+shallower potential wells, feedback becomes more effective at removing
+gas. Re-ionization additionally suppresses late star formation in the
+faintest galaxies. As a result, the resulting stellar mass of a dwarf
+galaxy is highly sensitive to dark matter halo mass.
 
@@
 shows an example exponential stellar component for the Fornax dwarf
-galaxy with the expected underlying dark matter halo. Even when the
-stars are densest in the inner regions, the dark matter remains nearly
-an order of magnitude higher in density. Stars make a small contribution
-to the gravitational structure of dwarf galaxies---indeed, stars are
-reasonably approximated as tracer particles of the underlying dark
-matter halo. In addition, the stellar component is confined to the
-central regions of the dark matter halo.
+galaxy with its surrounding dark matter NFW halo \citep[with parameters
+matching the][ and \citet{fattahi+2018} relations]{ludlow+2016}. Where
+the stars are densest, the dark matter remains nearly an order of
+magnitude higher in density. Stars make a small contribution to the
+gravitational structure of dwarf galaxies---indeed, stars are reasonably
+approximated as tracer particles of the underlying dark matter halo. In
+addition, the stellar component is typically confined to the central
+regions of the dark matter halo.
 
@@
 (relative to reionization) can influence the resulting stellar content
-\citep{kim+2024}. Finally, tides influence both the dark matter and
-stellar mass but in different amounts, adding additional scatter to the
-SMHM trend for satellites \citep[e.g.,][]{PNM2008, fattahi+2018}.
-Understanding the effects of tides on Local Group dwarf galaxies aids in
-revealing where and how these galaxies formed in a cosmological context.
+\citep{kim+2024}. Finally, Galactic tides influence both the dark matter
+and stellar mass but in different amounts, adding additional scatter to
+the SMHM trend for satellites \citep[e.g.,][]{PNM2008, fattahi+2018}.
+Understanding the effects of tides on Local Group dwarf galaxies may
+help us understand where and how these galaxies formed in a cosmological
+context.
 
@@
 observed abundance has been known as the \emph{missing satellites
-problem}. Additionally, a number of observations suggested that some
-dwarf galaxies, although not all, possess dark matter ``cores,''
+problem}. Additionally, a number of observations suggest that some dwarf
+galaxies, although not all, possess dark matter ``cores,''
 \citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019},\footnote{In
@@
   constraining.} contrary to the expectation from \LCDM{} of ``cuspy''
-inter dark matter profiles \citep{NFW1996, NFW1997}. As a result,
+inner dark matter profiles \citep{NFW1996, NFW1997}. As a result,
 alternative forms of dark matter have been advocated as solutions, such
@@
 Several open questions remain, concerning e.g.~the plane of satellites,
-and the details of the sizes and rotation curves of dwarf galaxies, the
+the details of the sizes and rotation curves of dwarf galaxies, and the
 existence and nature of stellar halos and accretion in dwarf galaxies
@@
 determining distances and full 3D velocities is challenging. The most
-direct measurement of distance, parallax, requires precise tracking of a
-star's sky position across a year. And while line-of-sight (LOS)
+direct measurement of distance, the parallax, requires precise tracking
+of a star's sky position across a year. And while line-of-sight (LOS)
 velocities are relatively easily determined from spectroscopy,
@@
 much more challenging. Typically, measuring proper motions requires
-accurate (much less than arcsecond) determinations of small changes in a
-star's position over baselines of years to decades. The full 6D position
-and velocity information for stars has, until recently, been known for
-only a handful of stars.
+precise (\(\ll\) arcsecond) determinations of small changes in a star's
+position over baselines of years to decades. The full 6D position and
+velocity information for stars has, until recently, been known for only
+a handful of stars.
 
@@
 completed its space-based mission in 2025, two further data releases are
-expected.
+still expected.
 
@@
 For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
-facilitated robust stellar membership determination. Before \emph{Gaia},
-few galaxies had precisely measured proper motions \citep[e.g.~using
-Hubble Space Telescope,][]{piatek+2005, sohn+2017}. \emph{Gaia} allowed
-the first systematic and precise determinations of Milky Way satellite
-proper motions \citep{pace+li2019, MV2020a}. While the proper motion
-uncertainty of a typical dwarf member star is often large, by combining
-the proper motions of 100s or 1000s of stars from \emph{Gaia}, precise
-average proper motion measurements can be determined, sometimes only
-limited by \emph{Gaia}'s systematic error floor
-\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new era
-for MW satellite dynamical studies, where we can derive precise orbits
-for any satellite, assuming a given MW potential. In addition,
-\emph{Gaia} helps establish membership by separating out contaminating
-MW foreground stars. By measuring parallaxes and/or proper motions, many
-more background and foreground stars can be classified as non-members
+facilitated robust stellar membership determinations. Before
+\emph{Gaia}, few galaxies had precisely measured proper motions
+\citep[e.g.~using Hubble Space Telescope,][]{piatek+2005, sohn+2017}.
+\emph{Gaia} allowed for some of the first systematic and precise
+determinations of Milky Way satellite proper motions
+\citep{pace+li2019, MV2020a}. While the proper motion uncertainty of a
+typical dwarf member star is often large, by combining the proper
+motions of 100s or 1000s of stars from \emph{Gaia}, precise average
+proper motion measurements can be determined, sometimes only limited by
+\emph{Gaia}'s systematic error floor \citep[e.g.,][]{MV2020a}. Proper
+motions have thus ushered in a new era for MW satellite dynamical
+studies, where we can derive precise orbits for any satellite, assuming
+a given MW potential. In addition, \emph{Gaia} helps establish
+membership by separating out contaminating MW foreground stars. By
+measuring parallaxes and/or proper motions, many more background and
+foreground stars can be classified as non-members
 \citep[e.g.,][]{battaglia+2022, jensen+2024}.
@@
 simplest, defined in terms of the central surface density, \(\Sigma_0\),
-and scale radius, \(R_s\):
+and projected scale radius, \(R_s\):
 \begin{equation}\protect\phantomsection\label{eq:exponential_law}{
 \Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s).
-}\end{equation} This profile is often applied to the radial light
+}\end{equation} This profile is also often applied to the radial light
 distribution of galaxy disks
@@
 written as \begin{equation}{
-\Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
-}\end{equation} In much of the older literature, \(R_t\) was interpreted
-as a ``tidal radius'', after an analogous interpretation for globular
-clusters \citep[e.g.,][]{hodge1961, IH1995}.
+\Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right); \qquad \text{for } R<R_t
+}\end{equation} and \(\Sigma_{\rm K}=0\) for \(R \geq R_t\). In much of
+the older literature, \(R_t\) was interpreted as a ``tidal radius'',
+after an analogous interpretation for globular clusters
+\citep[e.g.,][]{hodge1961, IH1995}.
 
@@
 \Sigma_{\rm S} = \Sigma_h \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
-}\end{equation} where \(b_n\) is a constant that depends on \(n\). A
-Sérsic profile with \(n=1\) is equivalent to an exponential profile,
-while \(n=4\) recovers \citet{devaucouleurs1948}'s profile for
-elliptical galaxies. Although a Sérsic profile is less commonly applied
-to dwarf galaxies, \citet{munoz+2018} advocate for the Sérsic profile
-since the added flexibility allows more profiles to be fit with a single
-law.
+}\end{equation} where \(b_n\) the coefficient which solves
+\(\Gamma(2n) = 2\gamma(2n, b_n)\) with \(\Gamma\) the Gamma function and
+\(\gamma\) the lower incomplete gamma function
+\citep{graham+driver2005}. A Sérsic profile with \(n=1\) is equivalent
+to an exponential profile, while \(n=4\) recovers
+\citet{devaucouleurs1948}'s profile for elliptical galaxies. Although a
+Sérsic profile is less commonly applied to dwarf galaxies,
+\citet{munoz+2018} advocate for the Sérsic profile since the added
+flexibility allows more profiles to be fit with a single law.
 
-\citet{faber+lin1983} were among the first to demonstrate that an
-exponential law is a reasonable empirical fit, theorizing that dwarf
-spheroidals may have evolved from exponential disk galaxies\footnote{While
-  studied for far longer, the exponential origin of disk galaxies is no
-  better understood. A sample of ideas range from scattering of stars
-  \citep{elmegreen+struck2013, wu+2022} to angular momentum transport,
-  to disk viscosity-driven radial gas flows
+While there are no clear theoretical preferences for any of these
+profiles, exponential density profiles have been commonly used for dwarf
+spheroidal galaxies. \citet{faber+lin1983} were among the first to
+demonstrate that an exponential law is a reasonable empirical fit,
+theorizing that dwarf spheroidals may have evolved from exponential disk
+galaxies\footnote{While studied for far longer, the exponential origin
+  of disk galaxies is no better understood. A sample of ideas range from
+  scattering of stars \citep{elmegreen+struck2013, wu+2022} to angular
+  momentum transport, to disk viscosity-driven radial gas flows
   \citep{lin+pringle1987, wang+lilly2022} or spherical collapse
@@
 
-Nevertheless, exponential density profiles have been commonly used for
-dwarf spheroidal galaxies. Many subsequent photometric studies of dwarf
-spheroidal galaxies have used exponential fits, finding that exponential
-and King profiles both provide good descriptions in many cases
+Many subsequent photometric studies of dwarf spheroidal galaxies have
+used exponential fits, finding that exponential and King profiles both
+provide good descriptions in many cases
 \citep{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018}.
@@
 blue compact, and irregular dwarf galaxies are better described with an
-exponential profile to which is added a central cusp or nuclear region
+exponential profile to which a central cusp or nuclear region is added
 \citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
-studies find an inner density decrement compared to exponentials
+studies find an inner density decrement relative to exponentials
 \citep[e.g.,][]{caldwell+1992, makarov+2012}, or that dwarfs are better
@@
 \citep[e.g.,][]{aparicio+1997, graham+guzman2003, hunter+elmegreen2006, lee+2018}.
-It is unclear how these conclusions apply to dwarf spheroidals.
+It is unclear how these conclusions apply to the dwarf spheroidals of
+the Local Group.
 
@@
 \subsection{The extended light profiles of Sculptor and Ursa Minor:
-Hints of tidal
-signatures?}\label{the-extended-light-profiles-of-sculptor-and-ursa-minor-hints-of-tidal-signatures}
+Hints of tidal signatures?}\label{sec:scl_umi_obs_tides}
 
-Sculptor and Ursa Minor appear to be typical dwarf spheroidal galaxies
-at first glance. Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}
-describe the present-day properties of each galaxy. Sculptor, as the
-first discovered classical dSph, is often described as a
-``prototypical'' dSph. There has long been speculation that both
-Sculptor and Ursa Minor may have been influenced by the Milky Way's
-tidal field (see Section~\ref{sec:discussion}).
-\citet{sestito+2023a, sestito+2023b}, for example, have recently
-reported a ``kink'' in the density profile, beginning around 30 arcmin
-for from the centre of each of these systems, that they interpret as
-potentially caused by the effects of Galactic tides. They
-spectroscopically follow up some of the most distant stars, finding
-members as far as 6 and 12 half-light radii from the centre of each
-dwarf. If these dwarfs had exponential profiles, like Fornax, then these
-far-outlying stars should be much rarer.
+Sculptor (Scl) and Ursa Minor (UMi) appear to be typical dwarf
+spheroidal galaxies at first glance (see Fig.~\ref{fig:galaxy_images}).
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} describe the
+structural parameters of each galaxy. Sculptor, as the first discovered
+classical dSph, is even described as a ``prototypical'' dSph
+\citep[e.g.,][]{mcconnachie2012}.
+
+However, there has long been speculation that both Scl and UMi may have
+been influenced by the Milky Way's tidal field. Already,
+\citet{innanen+papp1979} found RR Lyrae candidate Scl members
+\citep[from][]{vanagt1978} out to 180' in an elongated distribution,
+speculating this to be tidal disruption. Many later density profile
+determinations noted Scl's elongation and apparent outer density excess
+\citetext{\citealp{eskridge1988}; \citealp{IH1995}; \citealp{walcher+2003}; \citealp{westfall+2006}; \citealp[but
+see also][]{coleman+dacosta+bland-hawthorn2005}}. These works often
+interpreted these features as evidence of tidal effects
+\citep[e.g.,][]{walcher+2003} or sometimes a dwarf galaxy halo
+\citep{westfall+2006}.
+
+UMi also attracted similar suspicions. \citet{hargreaves+1994} first
+detected a velocity gradient in UMi, suggestive of tidal disruption.
+Later, \citet{martinez-delgado+2001} find that stars extend far beyond
+the King-profile ``tidal radius'', aligned with UMi's elongation,
+consistent with tidally-stripping simulations in
+\citet{gomez-flechoso+martinez-delgado2003}. \citet{palma+2003}
+furthermore detected S-shaped isophotes and an extended population of
+``extratidal'' stars.
+
+Adding to the evidence, \citet{sestito+2023a, sestito+2023b} report a
+``kink'' in the density profile of each galaxy. They spectroscopically
+follow up some of the most distant stars, finding members as far as 6
+and 12 half-light radii from the centre of each dwarf. If these dwarfs
+had exponential profiles, like Fornax, then these far-outlying stars
+should be much rarer.
 
@@
 Muñoz et al.
-(2018) Sérsic fits, 2.
+(2018, Sérsic fit), 2.
 Tran et al.
-(2022) RR lyrae distance, 3.
+(2022, RR lyrae distance), 3.
 Alan W.
@@
 References are: (1) Muñoz et al.
-(2018) Sérsic fits, (2) Garofalo et al.
-(2025) RR lyrae distance, (3) Alan W.
+(2018, Sérsic fit), (2) Garofalo et al.
+(2025) pRR lyrae distance], (3) Alan W.
 McConnachie and Venn (2020a), (4) Pace et al.
-(2020), average of MMT and Keck results.
+(2020), average of MMT and Keck results with systematic uncertainty from Section \ref{sec:extra_rv_models} discussion.
 }
@@
 
-Tidal disruption of dwarfs in the Local Group continues to this day. The
+The Local Group hosts several examples of ongoing tidal disruption. The
 Magellanic stream, a massive, gas-rich feature emanating from the
@@
 analytic host potential, achieving excellent numerical convergence. For
-instance, the simulations we describe later reach the times higher
-resolution than Aquarius \citep{springel+2008} with 400 times fewer
-particles. Idealized simulations make numerous simplifications,
-neglecting mergers, cosmological context, mass assembly, and often
-baryonic physics
+instance, the simulations we describe later in this thesis reach the
+times higher resolution than Aquarius \citep{springel+2008} with 400
+times fewer particles. Idealized simulations make numerous
+simplifications, neglecting mergers, cosmological context, mass
+assembly, and often baryonic physics
 \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
+Cosmological simulations appear to predict that tidal disruption is more
+common than idealized work or the Milky Way system suggests, but the
+origin and role of numerics and assumptions in this discrepancy is
+unclear \citep[ \citet{riley+2024}]{panithanpaisal+2021, shipp+2023l}.
 We shall use idealized simulations here to assess tidal effects after
-infall.
+the satellite's infall into the MW halo.
 
 Idealized simulations predict clear properties of tidally disrupting
-dwarf galaxies. Tidally stripped stars form tidal
+dwarf spheroidal galaxies. Tidally stripped stars form tidal
 streams---approximately uniform-density stellar tails with a bulk
@@
 undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
-cuspy CDM halos also are found to be nearly immune to full disruption
-\citep{EP2020}, but cored dark matter halos may disrupt more quickly
+NFW halos also are found to be nearly immune to tidal disruption
+\citep{EP2020}, but cored dark matter halos may disrupt fully and faster
 \citep[e.g.,][]{penarrubia+2010, errani+2023a}.
 
-To first order, tidal mass loss peels away the outer energy-layers of a
-dwarf galaxy.
+To first order, tidal mass loss peels away the outer layers of a dwarf
+galaxy in energy space.
 \citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
@@
 idea, creating a model for adiabatic tidal mass loss in an isotropic
-tidal field. They explain the resilience of NFW halos and origin of
-characteristic ``tidal tracks'' \citep[as observed
-in][]{PNM2008, green+vandenbosch2019, EN2021}.
+tidal field. They explain the resilience of NFW halos against full tidal
+disruption and origin of well-defined ``tidal tracks'' \citep[as
+observed in][]{PNM2008, green+vandenbosch2019, EN2021}.
 
 With precise orbital constraints and improved models of the Milky Way
-potential, recent studies continue to directly probe the dynamical
-histories of individual dwarf galaxies.
+potential, recent studies has continued to probe the dynamical histories
+of individual dwarf galaxies.
 \citet{borukhovetskaya+2022, dicintio+2024} both ran simulations tuned
@@
 \citet{borukhovetskaya+2022a} analyzed Crater II, showing that the
-present-day structure challenging to reconcile with \LCDM{} initial
-conditions and/or tidal evolution. Most relevantly, \citet{iorio+2019}
+present-day structure is challenging to reconcile with NFW initial
+conditions and Galactic tides. Most relevantly, \citet{iorio+2019}
 modelled Sculptor, concluding tides likely do not affect the stellar
@@
 For a given orbit in a given potential, there are characteristic radii
-which help gauge the effects of tides on a dwarf galaxy model.
+which help gauge the effects of tides on a dwarf galaxy.
 
@@
 3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
-}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) occurs within the
-visible extent of a galaxy, we should expect to find relatively clear
-signs of tidal disturbance. While strictly valid for circular orbits,
-assuming \(r_{\rm peri}\) for the host-dwarf distance works as most
-stars are lost near pericentre.
+}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) is comparable to
+the visible extent of a galaxy, we should expect to find clear signs of
+tidal disturbance. While strictly valid for circular orbits, assuming
+\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
+lost during pericentric passages.
 
 We also use the \textbf{break radius} as defined in
-\citet{penarrubia+2009}, marking where the galaxy is still in
-disequilibrium after pericentric passage in a highly-eccentric orbit.
-The break radius \(r_{\rm break}\) is proportional to the velocity
-dispersion, \(\sigma_v\), and time elapsed since pericentre,
-\(\Delta t\), \begin{equation}\protect\phantomsection\label{eq:r_break}{
+\citet{penarrubia+2009}, marking the outermost radius within which the
+dwarf has been able to achieve equilibrium after pericentric passage in
+a highly-eccentric orbit. The break radius \(r_{\rm break}\) is
+proportional to the velocity dispersion, \(\sigma_v\), and time elapsed
+since pericentre, \(\Delta t\),
+\begin{equation}\protect\phantomsection\label{eq:r_break}{
 r_{\rm break} = C\,\sigma_{v}\,\Delta t
@@
 
-To illustrate the effects of tides on an NFW halo due to a larger
-galaxy, we consider a toy model. We evolve an NFW subhalo with
-\(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another NFW static
-host halo with \(\rmax=25\,\kpc\) and \(\vmax = 207.4\,\kms\). These
-choices are motivated by the inferred structure and masses of dSphs and
-the Milky Way, respectively. The stars follow initially an exponential
-profile with a scale radius \(R_s=0.25\,\kpc\). We then evolve the orbit
-through a pericentre of 10 kpc starting from an apocentre of 100 kpc. We
-show the results 170 Myr after the first pericentre. See
-Section~\ref{sec:methods} for a more detailed description of our
-numerical setup.
+To illustrate the effects of tides on an NFW halo due to a massive host,
+we consider a toy model. We evolve an N-body realization of an NFW
+subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
+NFW static host halo with \(\rmax=25\,\kpc\) and
+\(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
+structure and masses of dSphs and the Milky Way, respectively. We
+simulate the stellar component by assigning stellar weights to each
+particle based on the relative densities of the distribution functions
+in energy space (see \citet{ref:sec_painting_stars}). The stars follow
+initially a 2D exponential profile with a scale radius
+\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We then
+evolve the orbit through a pericentre of 20 kpc starting from an
+apocentre of 100 kpc. See Section~\ref{sec:methods} for a more detailed
+description of our numerical setup.
 
 Fig.~\ref{fig:idealized_break_radius} illustrates the properties of this
-idealized simulation shortly after the first pericentre. The projected
-density of stars (right) is relatively undisturbed and spherical in the
-centre, but becomes non-isotropic outside the break radius and shows
-nascent tidal tails.These tidally disturbed stars appear as an extended,
-outer density excess relative to the initial conditions. This excess
-appears just outside the break radius. The break radius also marks where
-the mean radial velocity of the stars transitions from 0 to positive ---
-the galaxy is out of equilibrium outside \(r_{\rm break}\).
+idealized simulation at the second apocentre, as seen from the centre of
+the host. The projected density of stars, is relatively undisturbed and
+spherical in the centre, but becomes non-isotropic outside the break
+radius and shows nascent tidal tails. These tidally disturbed stars
+appear as an extended, outer density excess relative to the initial
+conditions. This excess appears just outside the break radius. The break
+radius also marks where the mean 3D radial velocity of the stars (with
+respect to the galaxy's centre) transitions from 0 to positive --- the
+galaxy is out of equilibrium outside \(r_{\rm break}\).
 
@@
 resembles the outer excess in Sculptor and Ursa Minor. This thesis aims
-to investigate if a tidal explanation is viable given the current
-properties of each galaxy.
+to investigate if a tidal explanation like this is viable given the
+current properties of each galaxy.
 
@@
 \caption[Example tidal simulation]{Example density and velocity
-distributions of an idealized simulation shortly after pericentre.
-\textbf{Top}: The projected 2D stellar density in the \(x\)--\(y\) plane
-for the initial (left) and final (right) simulation. The dashed green
-circle represents the break radius (Eq.~\ref{eq:r_break}) and the solid
-(dotted) blue line represents the past (future) orbit. \textbf{Bottom
-left}: The projected stellar density profile for the initial (dotted)
-and final (solid) simulation snapshot. The green arrow marks the break
-radius. \textbf{Bottom left}: the mean radial velocity (dot product of
-relative position and velocity relative to dwarf centre) as a function
-of projected radius. The green arrow marks the break
-radius.}\label{fig:idealized_break_radius}
+distributions of an idealized simulation at the second apocentre
+(initialized on the first apocentre). \textbf{Top}: The projected 2D
+stellar density in the \(y'\)--\(z'\) plane for the initial (left) and
+final (right) simulation. \(y'\) is the direction of tangental orbital
+motion and \(z'\) is the direction of angular momentum, as seen from the
+centre of the host. The dashed green circle represents the break radius
+(Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
+direction. \textbf{Bottom left}: The projected stellar density profile
+for the initial (dotted) and final (solid) simulation snapshot.
+\textbf{Bottom left}: the mean radial velocity (dot product of relative
+position and velocity relative to dwarf centre) as a function of
+projected radius. The green arrow marks the break radius in both lower
+panels.}\label{fig:idealized_break_radius}
 \end{figure}

### 2025-10-02 thesis/methods.tex
discussion revisions

--- 2025-09-27 thesis/methods.tex
@@
 exponential law that approximates well all other classical dSphs. Our
-main intention is to assess whether such excesses result from the
+main intention is to assess whether such deviations result from the
 effects of Galactic tides. To this purpose, we use N-body simulations of
@@
 the orbital parameters consistent with a dwarf's present-day position
-and velocity. We shall assume that the Galactic potential is the static,
-analytic potential inferred by \citet{mcmillan2011} from observations of
-kinematic tracers. We also assume that the potential of each dwarf may
-be initially approximated by a cuspy NFW profile. Since the dwarfs in
-question are heavily dark matter dominated, we shall use a carefully
-selected sample of dark matter particles to emulate the evolution of an
-embedded tracer stellar component. In this Chapter, we describe our
-choice of Galactic potential, orbital estimation, initial conditions
-setup, and N-body methods.
+and velocity. We shall assume that, over the past 10 Gyr, the Galactic
+potential is the static, analytic potential inferred by
+\citet{mcmillan2011} from observations of kinematic tracers. We also
+assume that the potential of each dwarf may be initially approximated by
+a cuspy NFW profile. Since the dwarfs in question are heavily dark
+matter dominated, we shall use a carefully selected sample of dark
+matter particles to emulate the evolution of an embedded tracer stellar
+component. In this Chapter, we describe our choice of Galactic
+potential, orbital estimation, initial conditions setup, and N-body
+methods.
 
@@
 coordinates, we use the Astropy v4 Galactocentric frame
-\citep{astropycollaboration+2022}. In this frame, the solar position is
+\citep{astropycollaboration+2022}. Our Cartesian Galactocentric
+coordinates here assume the Galactic centre is at
+\([x, y, z] = [0,0,0]\), where \(x\) is the direction from the sun to
+the Galactic centre, \(y\) is the direction of the motion of the Local
+Standard of Rest, and \(z\) is the direction perpendicular to the
+Galactic plane. The coordinate frame is also right-handed, such that the
+\(z\)-angular momentum of the sun is negative (since the sun is at
+\(x<0\)). In this frame, the solar position is
 \([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
@@
 The uncertainties in this reference frame are typically smaller than the
-uncertainties on a dwarf galaxy's position and (tangental) velocity.
+uncertainties on a dwarf galaxy's distance and tangental velocity.
 
@@
 profile of \citet{EP2020} potential. The total circular velocity (thick
-black line) is composed of an NFW halo (green dashed line), a think and
-thick \citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
+black line) is composed of an NFW halo (green dashed line), thin and
+thick \citet{miyamoto+nagai1975} disks (orange dash-dotted line), and a
 \citet{hernquist1990} bulge (blue dotted
@@
 
-Sculptor's orbital in the assumed potential is relatively
+Sculptor's orbit in the assumed potential is relatively
 well-constrained. Fig.~\ref{fig:scl_orbits} illustrates point particle
 orbits for 100 samples of Sculptor's observed kinematics integrated
-backwards 5 Gyr in both Galactocentric coordinate slices (\(x\), \(y\),
-\(z\)) and in Galactocentric radius with time. These orbits sample the
-uncertainties in distance, proper motion, and radial velocity, as given
-in Table~\ref{tbl:scl_obs_props}. All sampled orbits of Sculptor have
-nearly the same morphology---the orbit primarily resides in the
-\(y\)--\(z\) plane and has a similar number of periods and
-pericentre/apocentre.
+backwards for \(10\,\Gyr\) in both Galactocentric Cartesian coordinates
+(\(x\), \(y\), \(z\)) and in Galactocentric radius with time. These
+orbits sample the uncertainties in distance, proper motion, and radial
+velocity, as given in Table~\ref{tbl:scl_obs_props}. All sampled orbits
+of Sculptor have nearly the same morphology---the orbit primarily
+resides in the \(y\)--\(z\) plane and completes a similar number of
+periods and pericentric and apocentric passages.
 
 To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
-smallest pericentre among all possible orbits. We achieve this by taking
-the median parameters of all orbits with a pericentre less than the
-0.0027th quantile pericentre, yielding a pericentre of 43 kpc. Given the
-current observations, it is unlikely that Sculptor has had a
-significantly smaller pericentre than our selected orbit, which we refer
-to as the \smallperi{} orbit. We take the first apocentre after a
-look-back time of 10 Gyr, or at \(\sim9.1\,\Gyr\), as the initial
-conditions for our model of Sculptor, noted in
-Table~\ref{tbl:orbit_ics}.
+smallest pericentre among all possible orbits integrated backwards for
+\(10\,\Gyr\). We achieve this by taking the median parameters of all
+orbits with a pericentre less than the 0.0027th quantile pericentre,
+yielding a pericentre of 43 kpc. Given the current observations, it is
+unlikely that Sculptor has had a significantly smaller pericentre than
+our selected orbit, which we refer to as the \smallperi{} orbit. We take
+the first apocentre after a look-back time of 10 Gyr, or at
+\(\sim9.1\,\Gyr\), as the initial conditions for our model of Sculptor,
+noted in Table~\ref{tbl:orbit_ics}.
 
@@
 \(x\)--\(y\) plane. Our selected \smallperi{} orbit is plotted in black
-and light green transparent orbits represent the past 5Gyr orbits
+and light green transparent orbits represent the past 10Gyr orbits
 sampled over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The
@@
 the lower panel, we show the radius versus time for only three orbits of
-Ursa Minor.}\label{fig:umi_orbits}
+Ursa Minor, representing the \smallperi{} point-mass orbit (black), the
+mean orbit, and the orbit with the \(3\sigma\)-largest
+pericentre.}\label{fig:umi_orbits}
 \end{figure}
@@
 $t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
-${x}_{i} / \kpc$ & -2.49 & 4.30 & 17.40\\
+${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
 ${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
@@
 $\V_{x\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
-$\V_{y\,i} / \kms$ & -114.83 & -56.83 & 48.62,\\
+$\V_{y\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
 $\V_{z\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
@@
 assumed MW potential. Fig.~\ref{fig:umi_orbits} shows 100 random
-point-orbits of Ursa Minor. As for Sculptor, we select an orbit with
-approximately the \(3\sigma\) smallest pericentre, the Ursa Minor
+point-mass orbits of Ursa Minor. As for Sculptor, we select an orbit
+with approximately the \(3\sigma\) smallest pericentre, the Ursa Minor
 \(\smallperi{}\) orbit (see Fig.~\ref{fig:umi_orbits} and
@@
 the outer regions by \begin{equation}{
-\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t),
+\rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
 }\end{equation} where we adopt \(r_t = 20 r_s\).
@@
 magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
-\citet{woo+courteau+dekel2008} (with \(\sim\) 0.17 dex uncertainty), the
-total current stellar mass of Sculptor and Ursa Minor are
-\(M_\star \approx 3.1 \times 10^6 \Mo\) and
+\citet{woo+courteau+dekel2008}, the total current stellar mass of
+Sculptor and Ursa Minor are \(M_\star \approx 3.1 \times 10^6 \Mo\) and
 \(M_\star \approx 7 \times 10^5 \Mo\), respectively. Based on the
@@
 
-The observed velocity dispersion constrains the halo mass. From the
-\citet{wolf+2010} mass estimator, the mass within the 3D half-light
-radius \(r_h\) is
+The observed velocity dispersion constrains the total mass within the
+stellar half-light radius. From the \citet{wolf+2010} mass estimator,
+the mass contained within the 3D half-light radius \(r_h\) is
 \begin{equation}\protect\phantomsection\label{eq:wolf_mass}{
@@
 We record the total luminosity, stellar mass, mass-to-light ratio, dark matter halo $\vmax$ and $\rmax$, and dark matter halo virial mass $M_{200}$ and concentration $c_{\rm NFW}$.
+Uncertainties are the 16-84th percentile derived using Monte-Carlo sampling, assuming 0.035 and 0.1 dex uncertainties in the $\vmax(M_\star)$ and $c(M_{200})$ relations.
 }
@@
 $M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & $1.9 \times 10^{\pm 0.17}$\\
-$\vmax$ & $31\pm 3\,\kms$ & $27_{-6}^{+7}\,\kms$\\
-$\rmax$ & $6 \pm 2$ kpc & $5_{-2}^{+1}$ kpc\\
-$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $0.3_{-0.2}^{+0.4} \times 10^{10}\,\Mo$\\
+$\vmax$ & $31\pm 3\,\kms$ & $27_{-2}^{+3}\,\kms$\\
+$\rmax$ & $6 \pm 2$ kpc & $5_{-1}^{+2}$ kpc\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $0.3_{-0.1}^{+0.2} \times 10^{10}\,\Mo$\\
 $c_{\rm NFW}$ & $13_{-3}^{+4}$ & $13.5_{-3}^{+4}$\\
@@
 
-To simulate the tidal evolution of galaxies, we use ``N-body''
-simulations integrated with the parallel, gravitational-tree program
-\gadget{} \citep{gadget4}. The N-body method calculates and evolves the
+To simulate the tidal evolution of galaxies, we use N-body simulations
+integrated with the parallel, gravitational-tree program \gadget{}
+\citep{gadget4}. The N-body method calculates and evolves the
 gravitational accelerations between a large number of collisionless
@@
 \(\vmax = 31\,\kms\) for the isolation halo based on Sculptor's mean
-properties. We run this model for 5 Gyr (about three times the crossing
-time \(t_{\rm cross} = \vcirc / r \approx 1.5\,\Gyr\) at
+properties. We run this model for 5 Gyr (about one-half crossing time
+\(t_{\rm cross} = 2\pi\,r /\vcirc  \approx 9\,\Gyr\) at
 \(r_{200}=36\,\)kpc).
@@
 \begin{equation}\protect\phantomsection\label{eq:softening_length}{
-h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
-}\end{equation} See Appendix \ref{sec:extra_convergence} for a
-discussion of this choice, which is similar to the \citet{power+2003}
-suggested softening. We use in \gadget{} the relative tree opening
-criterion with the accuracy parameter set to \(\alpha =0.005\) (so nodes
-are opened when \(M\,l/r^3 < \alpha |a|\) for a distance from the node
-of mass \(M\) of \(r\), \(l\) is the side length of the node, and \(a\)
-is the particle's total acceleration), and adaptive time stepping with
-integration accuracy set to \(\eta=0.01\) (particles must take
-time-steps smaller than \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for
-acceleration \(a\)).
+h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
+}\end{equation} where the force is softened by a spline kernel
+\citetext{\citealp[eqs. 70--71
+in][]{springel+yoshida+white2001}; \citealp[but with the spline
+characteristic radius \(2.8h_{\rm grav}\), see][]{gadget4}}. See
+Appendix \ref{sec:extra_convergence} for a discussion of this choice and
+our simulation parameters, which is similar to the \citet{power+2003}
+suggested softening.
 
@@
 collisional effects driven by the finite number of particles used to
-describe the innermost regions of a halo become important. The softening
-length (from Eq.~\ref{eq:softening_length}) is typically a few times
-smaller than the converged length.
+describe the innermost regions of a halo become important over a Hubble
+time in a cosmological simulation. The softening length (from
+Eq.~\ref{eq:softening_length}) is typically a few times smaller than the
+converged length.
 
@@
 for analysis. We use a shrinking-spheres centre method inspired by
-\citet{power+2003}. First, we start with an initial centre estimate from
-the last timestep. Then, we calculate the radius of all particles from
-the centre, remove particles with a radius beyond the 0.975 quantile of
-the centre, and recalculate the centre of mass. The procedure is
-repeated until the selection radius is less than \textasciitilde1kpc or
-fewer than 0.1\% of particles remain. After a centre has been chosen, we
-remove all unbound particles based on the \gadget{} calculated potential
-of the halo. For all future timesteps, we consider only particles
-retained from the previous iteration.
+\citet{power+2003}. First, we start with an initial centre estimate
+using all bound particles from the last timestep. Then, we calculate the
+radius of all particles from the centre, remove particles with a radius
+beyond the 0.975 quantile of the centre, and recalculate the centre of
+mass. The procedure is repeated until the selection radius is less than
+\textasciitilde1kpc or fewer than 0.1\% of particles remain. After a
+centre has been chosen, we remove all unbound particles based on the
+\gadget{} calculated potential of the halo. For all future timesteps, we
+consider only particles retained from the previous iteration.
 
@@
 observed of \(\sim 0.03\,\kpc\). This is about three times the softening
-length but is less than the numerically converged radius scale.
+length but is less than the numerically converged radial scale.
 

### 2025-10-02 thesis/results.tex
discussion revisions

@@
 In this Chapter, we consider Scl first, describing tidal effects from
-the MW on dark matter and stellar components, and then consider the
-effects from the LMC. We then analyze UMi in a similar manner,
-considering in turn the dark matter evolution, stellar evolution, and
-orbital effects of the LMC.
+the MW on dark matter and stellar components. Next, we consider how
+accounting for the LMC amends our conclusions. We then analyze UMi in a
+similar manner, considering in turn the dark matter evolution, stellar
+evolution, and orbital effects of the LMC.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
-\caption[Sculptor tidal tracks]{Dark matter evolution for the
-\smallperi{} model of Sculptor. Blue solid and orange dashed lines show
-the initial and final circular velocity profiles. The points represent
-the evolution of the maximum circular velocity, and the dotted black
-line shows the tidal track from \citet{EN2021}. To calculate the
-velocity profiles, unbound particles are iteratively removed,
-recalculating the potential at each step assuming spherical symmetry.
-\textbf{mark jacobi radius and stellar contributions\ldots, maybe sigmav
-/ r\_h}}\label{fig:scl_tidal_track}
+\caption[Sculptor tidal tracks]{Dynamical evolution for the \smallperi{}
+model of Sculptor. Dotted and solid lines show the initial and final
+circular velocity profiles, and blue and orange lines show the dark
+matter and stellar (2D exponential) profiles. The points represent the
+evolution of the maximum circular velocity, and the dotted black line
+shows the tidal track from \citet{EN2021}. To calculate the velocity
+profiles, unbound particles are iteratively removed, recalculating the
+potential at each step assuming spherical
+symmetry.}\label{fig:scl_tidal_track}
 \end{figure}
@@
 \midrule
-pericentre & $53\pm3$ & 29\\
+pericentre & $53\pm3$ & 42\\
 apocentre & $102\pm3$ & 94.4\\
@@
 sky. The solid circle marks \(6R_h\), the dotted circle the break
-radius, and the (dotted) line the past (future) orbit. \textbf{Bottom:}
-The initial (dotted) and final (solid) stellar density profiles as
-compared to the observed stellar density profile. Arrows mark the break
-and Jacobi radii (Eqs.~\ref{eq:r_break}, \ref{eq:r_jacobi})
+radius, and the blue arrow the orbital direction. \textbf{Bottom:} The
+initial (dotted) and final (solid) stellar density profiles as compared
+to the observed stellar density profile. Arrows mark the half-light
+(\(R_h\)), break, and Jacobi radii
+(Eqs.~\ref{eq:r_break}, \ref{eq:r_jacobi})
 .}\label{fig:scl_smallperi_i_f}
@@
 only one MW pericentre and a recent, rapid LMC encounter, tidal forces
-do not appear to affect the stellar distribution. \textbf{todo fix break
-radius position, orbits as arrows\ldots{}}}\label{fig:scl_lmc_i_f}
+do not appear to affect the stellar
+distribution.}\label{fig:scl_lmc_i_f}
 \end{figure}
@@
 $R_{h, f}$ / kpc & 0.189 & 0.205\\
-break radius & $77'$, 1.6 kpc & $80'$, 1.7 kpc\\
+break radius & $78'$, 1.6 kpc & $81'$, 1.7 kpc\\
 LMC break radius & $23'$, 0.49 kpc & $24'$, 0.52 kpc\\
@@
 becoming apparent only outside 100 arcminutes in
-Fig.~\ref{fig:umi_smallperi_i_f}. Tides also only moderately affect the
-observed size and velocity dispersion of Ursa Minor
+Fig.~\ref{fig:umi_smallperi_i_f}. The observed size and velocity
+dispersion of Ursa Minor evolve little
 (Table~\ref{tbl:umi_sim_stars_results}). For exponential initial
@@
 profile. The stellar evolution of the Plummer stellar component is
-similar to the exponential stellar profile (\citet{ig:umi_plummer_i_f}).
-However, because there are more loosely-bound stars, the Plummer model
-looses nearly 4\% of its initial stellar mass to tides
-(Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
-detectible if we measure densities 2 orders of magnitude fainter than
-our present data. We show the properties of a stream in the Appendix
-(Fig.~\ref{fig:umi_tidal_stream}), but such a stream is unlikely to be
-observable in the near future.
+similar (Fig.~\ref{fig:umi_plummer_i_f}). However, because there are
+more loosely-bound stars, the Plummer model looses nearly 4\% of its
+initial stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}),
+and tidal features may be detectible if we measure densities 2 orders of
+magnitude fainter than our present data. We show the properties of a
+stream in the Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a
+stream is unlikely to be observable in the near future. We conclude that
+tides do not strongly affect the stellar component of this model.
 

### 2025-10-02 orbits/bootes3/mean_reflex_motion.jl

@@ -1,0 +1,266 @@
+### A Pluto.jl notebook ###
+# v0.20.18
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ b088fc88-9e6a-11f0-3095-3d9c1fe82a36
+begin
+	import Pkg; Pkg.activate()
+
+	import Agama
+	using LilGuys
+	
+	using CairoMakie
+	using Arya
+end
+
+# ╔═╡ 91ab3185-a7a3-453d-ac72-9fd026d30e77
+import TOML
+
+# ╔═╡ ac195c2d-e80d-49d8-9524-9d440844f59d
+module OrbitUtils
+	include("../orbit_utils.jl")
+end
+
+# ╔═╡ 76bc084b-6714-47f8-9032-90baa407267f
+galaxyname = "bootes3"
+
+# ╔═╡ 309abb2f-9c6a-4562-a7da-caebbed9bc03
+modelname = "vasiliev+21"
+
+# ╔═╡ 203f2d03-15f5-4553-9eeb-2699635c212c
+pot = Agama.Potential(file=joinpath(modelname, "agama_potential.ini"))
+
+# ╔═╡ cea8a018-e315-4968-93b6-f2a59a1bcc44
+pot_noreflex = Agama.Potential(file=joinpath("vasiliev+21_noreflex", "agama_potential.ini"))
+
+# ╔═╡ 5678931c-6d87-42d8-8332-bf282c46dce8
+pot_mw =  Agama.Potential(file=joinpath("vasiliev+21", "potential_mw.ini"))
+
+# ╔═╡ ad14e47b-c5d4-4aea-aff4-5a466ece1157
+pot_mw_static =  Agama.Potential(file=joinpath("vasiliev+21", "potential_mw_init.ini"))
+
+# ╔═╡ 59946955-acdc-41d8-ad67-d7843087f309
+pot_lmc =  Agama.Potential(file=joinpath("vasiliev+21", "potential_lmc.ini"))
+
+# ╔═╡ dad0a7e3-db11-4bb3-b30f-3b68b8c1ce29
+pot_reflex =  Agama.Potential(file=joinpath("vasiliev+21", "potential_reflex.ini"))
+
+# ╔═╡ 6195c401-2dc7-425b-a082-8bbda16d0ba3
[…]

### 2025-10-06 thesis/acronyms.tex
thesis updates / reorganization of RV modeling
major thesis restructure
julio edits at durham
thesis julio ch1-3 edits
thesis updates
thesis; julio oct 4 comments

@@
 % Ordinary abbreviations
+% units
 \newcommand{\kms}{{\rm km\,s^{-1}}}
 \newcommand{\kmsdeg}{{\rm km\,s^{-1}\,deg^{-1}}}
+\newcommand{\masyr}{{\rm mas^{-1}\,yr^{-1}}}
+\newcommand{\Gyr}{{\rm Gyr}}
+\newcommand{\kpc}{{\rm kpc}}
+\newcommand{\Msun}{{\rm M_\odot}}
+
+% match shortcuts
+\newcommand{\pmra}{\mu_{\alpha*}}
+\newcommand{\pmdec}{\mu_{\delta}}
+\newcommand{\V}{{\rm v}}
+\newcommand{\vmax}{\V_{\rm max}}
+\newcommand{\vcirc}{\V_{\rm circ}}
+\newcommand{\rmax}{r_{\rm max}}
+
+\newcommand{\LCDM}{$\Lambda$CDM}
+\newcommand{\agama}{{\sc Agama}}
+\newcommand{\gadget}{{\sc gadget-4}}
+\newcommand{\apostle}{{\sc apostle}}
+\newcommand{\eagle}{{\sc eagle}}
+\newcommand{\smallperi}{{\tt smallperi}}
 
@@
 
-\newacronym
-[description={
-Apache Point Observatory Galactic Evolution Experiment. 
-A large near-infrared spectroscopic survey of stars in the Milky Way. \citep{apogee17.}}]
 {apogee}{apogee}{Apache Point Observatory Galactic Evolution Experiment}

### 2025-10-06 thesis/data.tex
thesis; julio oct 4 comments

@@
 
-\section{Density profiles}\label{density-profiles}
+\section{Density profiles}\label{sec:data_density_profiles}
 
@@
 profiles furthermore agree with photometric surveys and similar
-literature (see Section~\ref{sec:extra_density}). We conclude that
+literature (see Appendix \ref{sec:extra_density}). We conclude that
 selection criteria do not influence our results.

### 2025-10-06 thesis/discussion.tex
thesis; julio oct 4 comments

@@
 satellites, and that recent tides are an unlikely explanation for such
-extended density profiles. We therefore suggest that Scl and UMi host a
-``stellar halo''---an additional, extended population of stars. With our
-data, other classical dwarfs lack evidence for similar stellar halos.
-The absence of similar halos in other classical dwarfs suggests Scl and
-UMi have followed a divergent evolutionary path.
+extended density profiles. We therefore suggest that Scl and UMi may
+host an additional, extended population of stars. With our data, other
+classical dwarfs lack evidence for similar extended populations. The
+absence of similar extended populations in other classical dwarfs
+suggests Scl and UMi have followed a divergent evolutionary path.
 
@@
 review both work considering tidal effects on Scl and UMi and detections
-of peculiarities. We then consider possible origins of stellar halos for
-these galaxies. Scenarios include induced star formation forming
-multiple populations, past mergers depositing stars into the outskirts,
-and even departures from the standard dwarf galaxy structure. We later
+of peculiarities. We then consider possible origins of extended stellar
+populations for these galaxies. Scenarios include multi-epoch star
+formation, past mergers depositing stars into the outskirts, and even
+departures from the typical exponential dwarf galaxy structure. We later
 discuss how future observations and simulations may distinguish halo
@@
 Tidal effects have been long invoked to explain features of dwarf
-galaxies e.g. \ref{sec:scl_umi_obs_tides},
-\citep{tsujimoto+shigeyama2002, mayer+2001a}. Yet, growing work suggests
-tides may not be as commonplace as suspected. For example,
-\citet{read+2006} suggest that the absence of rising radial velocity
-dispersion profiles in dwarfs implies a lack of tidal disruption.
-\citet{penarrubia+2009}, introducing the break radius as a diagnostic
-(Eq.~\ref{eq:r_break}), showing that most satellite orbits were
-inconsistent with observable tidal effects. Recently,
+galaxies e.g.~Section \ref{sec:scl_umi_obs_tides},
+\citep{mayer+2001a, tsujimoto+shigeyama2002} \textbf{Check this
+citation}. Yet, growing work suggests tides may not be as commonplace as
+suspected. For example, \citet{read+2006} suggest that the absence of
+rising radial velocity dispersion profiles in dwarfs implies a lack of
+tidal disruption. \citet{penarrubia+2009}, introducing the break radius
+as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
+orbits were inconsistent with observable tidal effects. Recently,
 \citet{pace+erkal+li2022} use a Jacobi radius-based criterion
-(Eq.~\ref{eq:r_jacobi}), reaching a similar conclusion. With the
+(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the
 exception of the Sagittarius dSph, these results suggest tides have mild
@@
 effects do not shape Scl or UMi's stellar component. A non-tidal
-explanation likely underlies the origin of these galaxies' density
+explanation likely underlies the origin of these galaxies' outer density
 excess.
@@
 from the inner to the outer exponential profile. \textbf{Bottom}:
-Metallicies as a function of elliptical radius for stars from
-\citet{tolstoy+2023}; \citet{apogee}; and \citet{sestito+2023} (left)
-and \citet{pace+2020}; \citet{apogee}; and \citet{sestito+2023b}
-(right)}\label{fig:metallicity_gradients}
+Metallicities as a function of elliptical radius for stars from
+\citet{tolstoy+2023}; \citet{abdurrouf+2022}; and \citet{sestito+2023a}
+(left) and \citet{pace+2020}; \citet{abdurrouf+2022}; and
+\citet{sestito+2023b} (right), crossmatched against J+24 as described in
+Appendix \ref{sec:extra_rv_models}.}\label{fig:metallicity_gradients}
 \end{figure}
@@
 the half-light radius. But, the third, more tentatively detected
-population in \citet{arroyo-polonio+2024} would reside within the
-stellar halo. In Ursa Minor, the transition between the metal-rich and
+population in \citet{arroyo-polonio+2024} would reside within the outer
+density excess. In Ursa Minor, the transition between the metal-rich and
 metal-poor stellar components occurs at \(\sim 30\,\)arcmin
-\citep{pace2020}, consummate with the beginning of the density excess.
-This metal-poor population may thus be related to the stellar halo of
-Ursa Minor. Fig:metallicity\_gradients shows the radial metallicity
-distribution in both galaxies. Scl and UMi both show evidence of a
-metallicity gradient, but the outskirts have few stars. Further
-observations would be necessary to understand if the stellar halo
-represents a chemically distinct population.
+\citep{pace+2020}, which coincides with the density excess. This
+metal-poor population may thus be related to the extended stellar
+density of Ursa Minor.
+
+Fig.~\ref{fig:metallicity_gradients} shows the radial metallicity
+distribution in both galaxies. To aid in comparison, we fit a
+2-component exponential to the density profiles. We also mark where the
+densities of each component become equal, the ``transition'' radius. Scl
+and UMi both show evidence of a metallicity gradient. But, the relation
+between the metallicity gradient and the transition from the inner to
+the outer density profile is unclear. UMi also has few measurements past
+the transition radius. Further observations would be necessary to
+understand if the extended stellar component represents a chemically
+distinct population.
 
@@
 photometry. In any case, the survival of a cold clump depends on the
-nature of dark matter, disrupting quickly in a \LCDM{} halo
-\citep{kleyna+2003, lora+2012}.
+inner structure of the halo, disrupting quickly in halos with
+substructure or inner density cusps but survining longer in smooth,
+cored halos \citep{kleyna+2003, lora+2012}.
 
@@
 substructure, this hints that Scl and UMi's distinct structure today
-arises from an unusual past. A mechanism forming a stellar halo could
-also explain multiple populations.
+arises from an unusual past. For example, a merger may naturally explain
+both extended and multiple stellar populations.
 
-\section{Forming a stellar halo}\label{sec:stellar_halos}
+\section{Forming an extended stellar
+population}\label{sec:stellar_halos}
 
 As we disfavour a tidal origin for Scl and UMi's extended density
-profiles, we now consider alternative processes which may form a stellar
-halo in these dwarfs.
+profiles, we now consider six alternative processes which may form an
+extended stellar population in these dwarfs.
 
@@
 \citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
-such processes would not explain why stellar halo formation may be
-non-universal. If a stellar halo was formed in an older, induced burst
-of star formation, then the star formation history should contain
-evidence of a corresponding burst.
+such processes would not explain why extended stellar populations appear
+to be non-universal. If an extended population was formed in an older,
+induced burst of star formation, then the star formation history should
+contain evidence of a corresponding burst.
 
@@
 \citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
-the lower-mass galaxy are preferentially dispersed, forming a stellar
-halo and population gradient \citep{benitez-llambay+2016, deason+2022}.
-If the galaxies contains gas, the merger can trigger new star formation,
-forming a younger population of stars \citep[e.g.,][]{genina+2019}. A
-few local dwarfs are suspected to have undergone a major merger,
-including Tucana II, Andromeda II, and Phoenix
-\citep{lokas+2014, fouquet+2017, tarumi+2021, cardona-barrero+2021, querci+2025}.
+the lower-mass galaxy are preferentially dispersed, forming an extended
+stellar component and population gradient
+\citep{benitez-llambay+2016, deason+2022}. If the galaxies contains gas,
+the merger can trigger new star formation, forming a younger population
+of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
+to have undergone a major merger, including Tucana II, Andromeda II, and
+Phoenix
+\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
 A galaxy having undergone a major merger should show evidence of at
@@
 
-\textbf{Minor mergers / accretion}. Just as the Milky Way's halo is
-believed to be built from many minor mergers, dwarf galaxies may build
-stellar halos composed of even fainter satellites
-\citep{ricotti+polisensky+cleland2022}. Few examples are known, but a
-stream around And II may be an instance of dwarf accretion
-\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
-the stellar halo's would contain chemical signatures from several
-distinct ultra-faint dwarf galaxies.
+Just as the Milky Way's halo is believed to be built from many minor
+mergers, dwarf galaxies may build accreted ``stellar halos'' composed of
+even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
+examples are known, but a stream around And II may be an instance of
+dwarf accretion \citep{amorisco+evans+vandeven2014, roederer+2023}. In
+this scenario, the extended stellar populations would contain chemical
+signatures from several distinct ultra-faint dwarf galaxies.
 
@@
 also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
-evidence of such features (see Section~\ref{sec:extra_rv_models}), so
+evidence of such features (see Appendix \ref{sec:extra_rv_models}), so
 future data would need to uncover signs of disruption to support this
@@
 \citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
-these processes should operate similarly across dwarf galaxies---stellar
-halos should be more common if these processes form substantial halos.
+these processes should operate similarly across dwarf
+galaxies---extended stellar populations should be more common if these
+processes form substantial halos.
 
-\section{Disentangling the origin of a stellar
-halo}\label{disentangling-the-origin-of-a-stellar-halo}
+\section{Disentangling the origin of extended stellar
+populations}\label{disentangling-the-origin-of-extended-stellar-populations}
 
-We just reviewed a number of different, possibly-concurrent explanations
-for the formation of a stellar halo. While we leave the nature of Scl
+We reviewed a number of different, possibly-concurrent explanations for
+the formation of extended populations. While we leave the nature of Scl
 and UMi's extended stellar densities an open question, we can discuss
@@
 
-Ongoing and future facilities will be essential for testing dwarf galaxy
-halo formation. Given the typical faint magnitudes of dSph stars,
-chemistry is best accessible through multi-object spectroscopic
-instruments on large telescopes (e.g., 4MOST and extremely large
-telescopes \citet{skuladottir+2023}). 3D internal kinematics of dwarfs
-may require a successor to \emph{Gaia}. However, using JWST and HST
-enables precise proper motions for small regions within dwarf galaxies
-\citep[e.g.,][]{vitral+2025}, which may be sufficient to constrain
-vastly different internal dynamic structures and ongoing tidal
-disruption. Finally, ongoing photometric surveys by Rubin Observatory
-\citep{ivezic+2019} and the Euclid mission
-\citep{euclidcollaboration+2025} will probe even fainter features around
-dwarf galaxies, possibly finding more halos, tidal streams, and
-improving our understanding of their stellar populations. Altogether,
-new surveys will likely uncover novel aspects about the inner workings
-and outer halos of dwarf galaxies.
+Ongoing and future observational surveys will be instrumental for
+testing dwarf galaxy halo formation. Because dSph stars are usually
+faint, detailed chemical analysis will rely on medium--high resolution
+multi-object spectroscopy on large telescopes, such as 4MOST
+\citep{skuladottir+2023} or extremely large telescopes. Mapping the 3D
+kinematics across a dwarf may require a \emph{Gaia} successor, but
+ongoing surveys with JWST and HST can derive precise proper motions
+within select regions of dwarf galaxies, able to reconstruct internal
+structure and anisotropies \citep[e.g.,][]{vitral+2025}. Current
+wide-field photometric surveys by Rubin Observatory \citep{ivezic+2019}
+and the Euclid mission \citep{euclidcollaboration+2025} will trace even
+fainter features around dwarf galaxies, possibly finding halos, tidal
+streams, and unusual stellar populations. Together, these observational
+programs promise a new window into the detailed structure, evolution,
+and outskirts of dwarf galaxies.
 
-The next generation of dwarf simulations are beginning to answer
-questions and unravel processes in the formation of dwarf galaxies.
-Notable recent simulation projects include EDGE \citep{agertz+2020} and
-LYRA \citep{gutcke+2021}, both of which can simulate the cosmological
+The next generation of dwarf simulations will likewise continue to
+refine our understanding of dwarf galaxy formation. Notable recent
+simulation projects include EDGE \citep{agertz+2020} and LYRA
+\citep{gutcke+2021}, both of which can simulate the cosmological
 formation of isolated dwarf galaxies while resolving individual
-(massive) stars. By including more realism (e.g., black holes, magnetic
-fields), refining models of poorly understood processes (feedback and
-star formation), and improving numerical resolution, simulations will
-continue to explore these processes and refine our understanding of
-dwarf galaxy formation.
+(massive) stars. Next steps for simulations include improved realism
+(e.g., magnetic fields), refining physical models (including feedback
+and star formation), and improving numerical resolution. With these
+continuing advances, future simulations may shed light on viability of
+halo formation scenarios and the details of dwarf galaxy assembly.
 
@@
 Scl and UMi. We review mechanisms ranging from episodic star formation
-histories, to accretion, to departures from \LCDM{}. While a more
-precise explanation awaits, upcoming and ongoing surveys will uncover
-the stellar populations within dwarf galaxies in unprecedented detail.
-Future work will continue to uncover the cosmic origins of the smallest
-structures forming building blocks of galaxies like ours.
+histories, to accretion, to departures from \LCDM{}. Future simulations
+and observational programs will be able to investigate the nature of
+dwarfs with ever increasing detail. Abundant opportunities await to
+uncover the cosmic origins and history of dwarf galaxies, the building
+blocks of galaxies like ours.

### 2025-10-06 thesis/extra_density.tex
thesis; julio oct 4 comments

@@
 comparing methodologies, samples, and to literature. Finally, we present
-a comparison of J+24's samples to a non-parametric Bayesian-derived
-density profile, finding similar results except in the very outer
-regions. Past a ``limiting radius'', the density profiles become
-susceptible to the likelihood specification and may select mostly
-background stars. In conclusion, the density profiles from J+24 converge
-among methods out to our limiting radii derived here.
+a comparison of J+24's samples to a non-parametric Bayesian density
+profile, finding similar results in the inner regions. Past a ``limiting
+radius'', J+24's density profiles become susceptible to the likelihood
+specification. However, all methods agree up to the limiting radii. In
+conclusion, the density profiles from J+24 are robust except in the very
+outer regions.
 
@@
 \item
-  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
+  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\). Ensures high
   quality astrometry. \texttt{ruwe} is a measure of the excess
   astrometric noise on fitting a consistent parallax-proper motion
-  solution \citep[see][]{lindegren+2021}.\\
+  solution \citep[see][]{lindegren+2021}.
 \item
@@
 covariance for proper motions). The spatial background likelihood is
-assumed to be constant.
+assumed to be uniform over the field.
 
@@
 
-\section{Alternative methodologies}\label{alternative-methodologies}
+\section{Alternative sample selection}\label{sec:simple_selection}
 
-\subsection{Simplistic cuts}\label{simplistic-cuts}
+As J+24's probabilistic model may systematically bias the density
+profile (see discussion below), we consider samples selected both from
+\emph{Gaia} and external surveys using simple, absolute selection cuts.
 
-Besides a global background subtraction (as used in
-Section~\ref{sec:observations}), absolute cuts in CMD and PM represent
-perhaps the next simplest method to construct density profiles. We
-create cuts using only the following
+For \emph{Gaia} data, we filter stars by four simple selection criteria.
+We require \texttt{ruwe} \textless{} 1.3 as above, selecting only stars
+with high-quality astrometry. We then require the star's parallax to be
+\(3\sigma\)-consistent with the dwarf's distance. Next, we select stars
+with proper motions with \(1\masyr\)\hspace{0pt} radius of the galaxy's
+measured value. Finally, we filter stars to lie on the dwarf's CMD in
+(\(G_{\rm BP} - G_{\rm RP}\),\(G\)), within the vertices defined in
+Table~\ref{tbl:colour_cuts}. This polygon is derived based on the
+colour-magnitude diagram of stars in the centre of the dwarf galaxy,
+where the dwarf dominates the stellar density.
 
-\begin{itemize}
-\tightlist
-\item
-  (BP-RP, G) within the vertices defined in Table~\ref{tbl:colour_cuts}
-\item
-  PM within a \(1\masyr\) radius of the galaxies measured value.
-\item
-  Parallax within \(3\sigma\) of the dwarf distance
-\item
-  \texttt{ruwe} \textless{} 1.3
-\end{itemize}
+In addition to \emph{Gaia}, we select stars in the Scl field from the
+deeper, photometric DELVE survey \citep{drlica-wagner+2022}. We select
+sources DELVE data release 2 within an ellipse of radius 150 arcminutes.
+We then only include sources which are categorized as likely stars and
+sources with reliable (flags \textless=4) \(g\) and \(r\) magnitudes. We
+then filter by the \(g\) and \(r\) magnitudes, selecting the empirical
+CMD cut using the vertices in Table~\ref{tbl:colour_cuts}.
 
-\begin{table*}[t]
-\centering
-\caption[Colour cuts for density profiles.]{Colour cuts used in samples in this section.
-Each pair of columns contains the vertices of a polygon in the CMD within which we select stars to derive density profiles.
-}
-\label{tbl:colour_cuts}
-\begin{tabular}{llllllll}
-\toprule
-Scl &  &  &  & UMi &  &  & \\
-\midrule
-Gaia &  & DELVE &  & Gaia &  & Unions & \\
-BP-RP & G & g-r & g & BP-RP & G & r-i & r\\
--0.21 & 20.58 & 0.32 & 22.58 & 0.49 & 20.83 & -0.485 & 24.13\\
--0.1 & 20.1 & 0.51 & 23.07 & 1.24 & 20.9 & 0.145 & 23.11\\
-0.28 & 19.77 & 0.56 & 21.49 & 1.18 & 19.65 & 0.201 & 21.41\\
-0.7 & 19.64 & 0.71 & 19.7 & 1.23 & 18.12 & 0.21 & 20.05\\
-1.02 & 18.5 & 0.96 & 18.49 & 1.39 & 17.09 & 0.288 & 17.23\\
-1.14 & 17.59 & 0.84 & 18.35 & 1.32 & 16.94 & 0.226 & 17.21\\
-1.41 & 16.73 & 0.6 & 19.65 & 0.96 & 18.58 & 0.151 & 19.3\\
-1.68 & 15.94 & 0.42 & 21.42 & 0.88 & 19.22 & 0.145 & 19.75\\
-1.94 & 15.94 &  &  & 0.4 & 19.52 & 0.026 & 19.86\\
-1.91 & 16.8 &  &  & -0.03 & 19.78 & 0.011 & 19.52\\
-1.68 & 17.01 &  &  & 0.06 & 20.16 & -0.155 & 19.95\\
-1.45 & 17.87 &  &  & 0.83 & 19.72 & -0.22 & 20.34\\
-1.32 & 19.16 &  &  & 0.21 & 20.83 & -0.176 & 20.47\\
-1.26 & 20.1 &  &  &  &  & -0.098 & 20.33\\
-1.35 & 21 &  &  &  &  & 0.008 & 19.84\\
-0.46 & 21 &  &  &  &  & 0.142 & 19.75\\
-0.82 & 20.27 &  &  &  &  & 0.098 & 21.45\\
-0.39 & 20.33 &  &  &  &  & -0.052 & 22.38\\
-0.16 & 20.76 &  &  &  &  & -0.388 & 23.42\\
-\bottomrule
-\end{tabular}
-\end{table*}
-
-\subsection{Alternative samples}\label{alternative-samples}
-
-For the DELVE sample \citep{drlica-wagner+2022}, we select everything in
-data release 2 between RA of 11 and 19, and declination of -37.7 and
--29.7.
-
-We also consider samples from the Ultraviolet Near-Infrared Optical
-Northern Survey \citep[UNIONS][]{gwyn+2025} for Ursa Minor. We select
-stars within a 5 degree tangent square of UMi with no
-\texttt{FLAGS\_CFIS} set and \texttt{s21} and \texttt{s31} both \(<3\)
-(not extended sources).
-
-\subsection{A Bayesian histogram
-model}\label{a-bayesian-histogram-model}
-
-To address the concerns discussed above, we consider here a
-non-parametric model to fit the density in each bin. This model extends
-the J+24 framework with two notable differences. (1) The systemic proper
-motions are held fixed. This allows for a much more efficient
-calculation of the likelihoods. (2) The spatial likelihood for the
-satellite is a piecewise constant density function.While this model
-theoretically has many more parameters to fit (one for each bin), each
-parameter is independently estimated based on only the stars in that
-specific bin.
-
-We segregate the data into radial bins and independently fit the model
-to only stars in each bin. This fits many small models instead of a
-global large model. In detail, this means that, in each bin, the
-normalized density of the background and satellite are \emph{equal}. The
-only fit parameter is the \(f_{\rm sat}\) in the particular bin.
-Therefore, the likelihood in each bin is the same as above except
-\({\cal L}_{\rm space} = 1\) for the foreground and background.
-
-In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
-satellite density. The model construction is then \begin{equation}{
-\theta_i \equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}})
-}\end{equation}
-
-\begin{equation}{
-f_{\rm sat} = \frac{10^{\theta_i}}{1 + 10^{\theta_i}}
-}\end{equation}
-
-\begin{equation}{
-\theta_i \sim {\rm Uniform}(-12, 6)
-}\end{equation}
-
-\begin{equation}{
-{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM}
-}\end{equation}
-
-For each satellite, we hold the structural parameters fixed to those in
-J+24, create bins which are the larger of the bin containing 20 stars or
-0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
-with 48 realizations using Turing.jl and the No-U Turns Sampler. The
-chains appear to converge excellently.
-
-Based on the \(f_{\rm sat}\) values in each bin, the number of members
-is just the number of stars in the bin times \(f_{\rm sat}\) for the
-bin, allowing the direct derivation of the density profiles from this
-model.
-
-We have also briefly explored a model which solves for \(f_{\rm sat}\)
-for random realizations of the global structural uncertainties. These do
-not appear to influence the results substantially. Another possible
-inconsistency is that the half-light radii are estimated from a
-different sample. We do briefly rederive structural parameters for the
-sample using Sérsic fits (although with a fixed origin).
-
-\section{Comparison of density profiles}\label{sec:density_extra}
-
-In this section, we discuss additional tests and verification of the
-derived density profiles. In particular, we check that methodology
-(simpler cuts, circularized radii, algorithm version) do not
-substantially affect the density profile. We also compile density
-profiles presented in the literature as reference. In all cases, the
-density profiles appear to have excellent convergence out to
-\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
-where the background dominates.
-
-\textbf{Spatial likelihood.} J+24 method was designed in particular to
-detect the presence of a density excess and find individual stars at
-large radii to be followed up. We are more interested in accurately
-quantifying the density profile and size of any perturbations. One
-potential problem with using J+24's candidate members is that the
-algorithm assumes the density is either described by a single or double
-exponential. If this model does not accurately match the actual density
-profile of the dwarf galaxy, we want to understand the impact of this
-assumption.In particular, in Fig.~\ref{fig:scl_observed_profiles},
-notice that the \(P_{\rm sat}\) selection method produces small
-errorbars, even when the density is more than 1 dex below the local
-background. These stars are likely selecting stars from the statistical
-MW background consistent with UMi PM+CMD, recovering the assumed density
-profile. As a result, the reliability of these density profiles below
-the CMD+PM background may be questionable. A more robust analysis,
-removing this particular density assumption, would be required to more
-appropriately represent the knowledge of the density profile as the
-background begins to dominate.
-
-\textbf{\emph{Gaia} systematics}. While \emph{Gaia} has shown excellent
-performance, some notable limitations may introduce problems in our
-interpretation and reliability of density profiles. Gaia systematics in
-proper motions and parallaxes are typically smaller than the values for
-sources of magnitudes \(G\in[18,20]\). Since we use proper motions and
-parallaxes as general consistency with the dwarf, and factor in
-systematic uncertainties in each case, these effects should not be too
-significant. However, the systematic proper motion uncertainties becomes
-the dominant source of uncertainty in the derived systemic proper
-motions of each galaxy (see
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}).
-
-\textbf{Completeness}. \emph{Gaia} shows high but imperfect
-completeness, particularly showing limitations in crowded fields and for
-faint sources (\(G\gtrapprox20\)). As discussed in
-\citet{fabricius+2021}, for the high stellar densities in globular
-clusters, the completeness relative to HST varies significantly with the
-stellar density. However, the typical stellar densities of dwarf
-galaxies are much lower, at about 20 stars/arcmin = 90,000 stars /
-degree, lower than the lowest globular cluster densities and safely
-below the crowding limit of 750,000 objects/degree for BP/RP photometry.
-In \citet{fabricius+2021}, for the lowest density globular clusters, the
-completeness down to \(G\approx 20\) is \(\sim 80\%\). Closely separated
-stars pose problems for Gaia's on-board processing, as the pixel size is
-59x177 mas on the sky. This results in a reduction of stars separated by
-less than 1.5'' and especially for stars separated by less than 0.6
-arcseconds. The astrometric parameters of closely separated stars
-furthermore tends to be of lower quality \citep{fabricius+2021}.
-However, even for the denser field of Fornax, only about 3\% of stars
-have a neighbour within 2 arc seconds, so multiplicity should not affect
-completeness too much (except for unresolved binaries). One potential
-issue is that the previous analyses do not account for our cuts on
-quality and number of astrometric parameters. These could worsen
-completeness, particularly since the BP-RP spectra are more sensitive to
-dense fields.
-
-In Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we
-show the results of limiting the density profile to the brightest half
-of stars. We find no substantive differences. If \emph{Gaia}'s
-incompleteness is inhomogeneous, then the variations in \emph{Gaia}'s
-completeness do not strongly affect the density profiles.
-
-\textbf{Structural uncertainties}. J+24 do not account for structural
-uncertainties in dwarfs for the two component case. We assume constant
-ellipticity and position angle. Dwarf galaxies, in reality, are not
-necessarily smooth or have constant ellipticity. J+24 test an
-alternative method using circular radii for the extended density
-component, and we find these density profiles are very similar to the
-fully elliptical case, even when assuming circular bins for the circular
-outer component. As such, even reducing the assumed ellipticity from
-\(0.37-0.55\) to 0 does not substantially impact the density profiles.
+Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
+Survey \citep[UNIONS][]{gwyn+2025}. We select sources within an ellipse
+of radius 230 arminutes. We keep sources with no \texttt{FLAGS\_CFIS}
+set and the \texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended
+sources). We then use the polygon bounded by the vertices in
+Table~\ref{tbl:colour_cuts} to select the observed red giant branch,
+horizontal branch, and main sequence of the dwarf.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
-\caption[Scl density comparison]{Density profiles for various
-assumptions for Sculptor. PSAT is our fiducial 2-component J+24 sample,
-circ is a 2-component bayesian model assuming circular radii, simple is
-the series of simple cuts described, bright is the sample of the
-brightest half of stars (scaled by 2), DELVE is a sample of RGB stars
-(background subtracted and rescaled to
-match).}\label{fig:scl_density_extras}
-\end{figure}
-
-Comparison of density profiles for each J+24 method. The fiducial is a
-2-component elliptical model. However, the 1-component is still
-elliptical but only contains 1 component and the circular model assumes
-a circular outer density profile and bins in circular bins instead of
-elliptical bins.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
-\caption[Scl literature density profiles]{A comparison of the Scl
-derived density profile and historical works. Unlike most profiles in
-this thesis, this density profile is plotted with respect to the
-semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
-The solid black line is a 2D exponential with corresponding scale radius
-to Scl's half-light radius, and residuals in the lower panel are
-relative to this exponential.}
-\end{figure}
-
-\subsection{Ursa Minor}\label{ursa-minor}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_density_methods_extra.pdf}}
-\caption[UMi density comparison]{Similar to
-Fig.~\ref{fig:scl_observed_profiles} except for Ursa
-Minor.}\label{fig:umi_density_extras}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/extra_cmd_selection.png}}
+\caption[CMD cuts]{Colour-magnitude diagram cuts used in samples in this
+section. Orange points are selected, black are non-selected stars within
+\(1R_h\), and grey stars are for all stars. \textbf{todo, only plot
+black/grey and cmd outline\ldots{}}}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
-\caption[UMi density profiles in literatures]{Similar to}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
+\caption[DELVE and UNIONS spatial distribution of stars]{The
+distribution of DELVE and UNIONS selected stars in Scl and UMI}
 \end{figure}
+
+\section{\texorpdfstring{Uncertainties and possible biases in
+\emph{Gaia}-derived density
+profiles}{Uncertainties and possible biases in Gaia-derived density profiles}}\label{sec:density_extra}
+
+In this section, we discuss additional tests and verification of the
+derived density profiles. In particular, we check that alternative
+methodologies do not substantially affect the density profile. In all
+cases, the density profiles appear to have excellent convergence out to
+\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
+where the background dominates.
+
+\textbf{Spatial likelihood.} J+24's algorithm was designed specifically
+to detect the presence of a density excess in satellites and identify
+individual stars for spectroscopic follow-up. We are instead interested
+in constraining a dwarf galaxy's stellar structure. As such, one
+potential problem with using J+24's candidate members is that the
+stellar density model is either a single or double exponential. If the
+(double) exponential model does not adequately describe a dwarf galaxy,
+then J+24's derived density profiles may be systematically biased
+towards the assumed (double) exponential profile.
+
+As an example, in the lower panels of Fig.~\ref{fig:scl_density_extras};
+Fig.~\ref{fig:umi_density_extras}, the density profile derived from the
+fiducial (J+24) sample appears to be well-constrained at densities far
+below the CMD+PM or \texttt{simple} model's background density. In
+addition, the fiducial and 1-component density profiles deviate, despite
+being derived identically besides the spatial likelihood assumption.
+
+These stars are likely selecting stars from the statistical MW
+background consistent either galaxy's CMD and PMs, recovering the
+assumed density profile. As a result, the reliability of these density
+profiles below the CMD+PM background may be questionable. A more robust
+analysis, removing this particular density assumption, would be required
+to more appropriately represent the knowledge of the density profile as
+the background begins to dominate.
+
+We revisit possible biases in spatial likelihood in
+Section~\ref{sec:mcmc_hists}.
+
+\textbf{Completeness}. \emph{Gaia} shows high but imperfect
+completeness, particularly showing limitations in crowded fields and for
+faint sources (\(G\gtrapprox20\)). In \citet{fabricius+2021}, for the
+lowest density globular clusters, the completeness down to
+\(G\approx 20\) is \(\sim 80\%\). As the lowest density globular
+clusters are still far denser than dwarf galaxies, so the completeness
+is likely higher for classical dwarfs. \emph{Gaia} is likely more
+incomplete in terms of BP-RP photometry (as we use here) and for stars
+with small pairwise separations (\(\lesssim 1.4''\)). The influence of
+these details is unclear. Nevertheless, \emph{Gaia} should be more
+complete for brighter stars.
+
+We find little evidence for magnitude-dependent biases in the density
+profiles. In
+Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we show
+density profiles derived using only stars brighter than the median
+magnitude. We find no substantive differences compared to our fiducial
+density profile. If \emph{Gaia}'s incompleteness is inhomogeneous, then
+these variations do not appear to affect the density profiles.
+
+If \emph{Gaia} has systematic density biases, the density profile should
+vary among different facilities. We show the density profiles derived
+from deep, ground-based UNIONS and DEVLE photometry (lower panels of
+Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). For
+both Scl and UMi, these alternative surveys result in near-identical
+density profiles. Variations between density profiles derived from these
+surveys appear to also be with uncertainties.
+
+\textbf{Structural uncertainties}. The assumed structural parameters of
+a dwarf may be misrepresentative or vary with radius. J+24 derive
+probabilities where the outer stellar component is circular instead of
+elliptical. Using the circular 2-component probabilities and binning in
+circular radius, we find little difference in the derived density
+profiles (the \texttt{circ} points in
+Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). Thus,
+our conclusions are robust to extreme variations in the ellipticity.
+Reasonable changes in other structural parameters (centre, position
+angle) are unlikely to have a more substantial effect.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
+\caption[Sculptor density methodology comparison]{Density profiles for
+various assumptions for Sculptor. \texttt{2-exp} is the fiducial
+double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
+one-component exponential spatial likelihood, and \texttt{simple} uses
+the simple position-independent selection criteria \texttt{circ} is a
+2-component bayesian model assuming circular radii, simple is the series
+of simple cuts described, bright is the sample of the brightest half of
+stars (scaled by 2).}\label{fig:scl_density_extras}
+\end{figure}
+
+\section{A Bayesian density profile}\label{sec:mcmc_hists}
+
+To address the concerns discussed above, we consider here a
+non-parametric model to fit the density in each bin.
 
@@
 
-\section{A comparison of density
-profiles}\label{a-comparison-of-density-profiles}
+\subsection{Methodology}\label{methodology}
+
+This model extends the J+24 framework with two notable differences. (1)
+The systemic proper motions are held fixed. This allows for a much more
+efficient calculation of the likelihoods. (2) The spatial likelihood for
+the satellite is a piecewise constant density function.While this model
+theoretically has many more parameters to fit (one for each bin), each
+parameter is independently estimated based on only the stars in that
+specific bin.
+
+We segregate the data into radial bins and independently fit the model
+to only stars in each bin. This fits many small models instead of a
+global large model. In detail, this means that, in each bin, the
+normalized density of the background and satellite are \emph{equal}. The
+only fit parameter is the \(f_{\rm sat}\) in the particular bin.
+Therefore, the likelihood in each bin is the same as above except
+\({\cal L}_{\rm space} = 1\) for the foreground and background.
+
+In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
+satellite density. The model construction is then \begin{equation}{
+\begin{align}
+\theta_i &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
+f_{\rm sat} &= \frac{10^{\theta_i}}{1 + 10^{\theta_i}} \\
+\theta_i &\sim {\rm Uniform}(-12, 6) \\
+{\cal L}_{\rm tot} &= f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM} \\
+\end{align}
+}\end{equation}
+
+For each satellite, we hold the structural parameters fixed to those in
+J+24, create bins which are the larger of the bin containing 20 stars or
+0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
+for 48 walkers using Turing.jl and the No-U Turns Sampler.
+
+Based on the \(f_{\rm sat}\) values in each bin, the number of members
+is just the number of stars in the bin times \(f_{\rm sat}\) for the
+bin, allowing the direct derivation of the density profiles from this
+model.
+
+We have also briefly explored a model which solves for \(f_{\rm sat}\)
+for random realizations of the global structural uncertainties. These do
+not appear to influence the results substantially. Another possible
+inconsistency is that the half-light radii are estimated from a
+different sample. We do briefly rederive structural parameters for the
+sample using Sérsic fits (although with a fixed origin).
+
+\subsection{Results}\label{results}
 
 Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
-histogram density profiles s compared to J+24. In general, both
+histogram density profiles compared to J+24. In general, both
 methodologies agree well. However, J+24 tend to systematically
@@
 
-. Once the true density of stars drops below the background of
+Once the true density of stars drops below the background of
 satellite-like stars, J+24's method will occasionally select consistent
@@
 \end{figure}
+
+\section{Comparison to literature}\label{comparison-to-literature}
+
+We finally compare our density profiles against a sample of
+literature-derived density profiles in
+Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. The
+J+24-derived density profiles are representative of state-of-the-art
+density profiles. A few photometric surveys may extend to slightly
+fainter densities, but typically with large uncertanties. Deviations
+between sources are slight despite the range of methods across time. All
+density profiles extending into the outskirts of Scl and UMi show a
+similar overdensity to the ones we find. The extended stellar profiles
+of Scl and UMi appear to be a robust result across the literature.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
+\caption[Sculptor literature density profiles]{A comparison of the Scl
+derived density profile and historical works. Unlike most profiles in
+this thesis, this density profile is plotted with respect to the
+semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
+The solid black line is a 2D exponential with corresponding scale radius
+to Scl's half-light radius, and the residuals in the bottom panel are
+with respect to this profile. \textbf{remove delve, umi title\ldots,,
+combine??}}\label{fig:scl_lit_profiles}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
+\caption[Ursa Minor literature density profiles]{Similar to
+Fig.~\ref{fig:scl_lit_profiles} except for Ursa Minor. The references
+are (in order) \citet{sato+2025}; \citet{palma+2003};
+\citet{martinez-delgado+2001}; \citet{kleyna+1998}; \citet{IH1995}; and
+\citet{Hodge1964}.}
+\end{figure}
+
+\section{Summary}\label{summary}
+
+In this Appendix, we discussed the methodological details of the J+24
+sample selection algorithm. We consider possible biases due to the
+assumed spatial likelihood, \emph{Gaia}'s completeness, and structural
+parameters. In all cases, we find these assumptions likely do not cause
+major biases in the derived density profiles. However, we show that
+J+24's density profiles become unreliable when dropping below the
+background of satellite-like stars. We address this concern through an
+algorithm rederiving the density in elliptical bins. We find excellent
+agreement with J+24 for most galaxies out to a background-limited
+radius. Finally, comparing our density profiles against the literature,
+we find our results to be consistent. We conclude that the detection of
+an extended density profile in Scl and UMi is robust to incompleteness,
+methodology, alternative surveys, and across the literature.

### 2025-10-06 thesis/extra_numerical_convergence.tex
thesis; julio oct 4 comments

--- 2025-09-27 thesis/extra_numerical_convergence.tex
@@
 
-Here, we describe some convergence tests to ensure our methods and
-results are minimally impacted by numerical limitations and assumptions.
-See \citet{power+2003} for a detailed discussion of various assumptions
-and parameters used in N-body simulations.
+Here, we describe the numerical specifics of our simulation and
+analysis. We present convergence tests supporting our choices of
+softening and additional parameters. We find that alternate choices of
+numerical parameters neither improve convergence nor the resulting
+evolution.
 
-\section{Softening}\label{softening}
+Due to the finite resolution of N-body simulations, collisional (close)
+encounters between particles are inevitable. Such collisional encounters
+are both unphysical (with possible arbitrary acceleration, violating the
+``collisionless'' assumption) and computationally expensive (requiring
+many small time-steps). As a remedy, many N-body codes use a
+``softened'' gravitational force law, where the force weakens when
+closer to a particle than a ``softening length''. The choice of
+softening length ideally balances resolution and computational speed.
 
-One challenge of N-body integration is close, collisional encounters,
-assuming point particles, cause divergences in the local force. However,
-this should not occur in a \emph{collisionless} simulation. As a result,
-most collisionless N-body codes adopt a gravitational softening, a
-length scale below which the force of gravity begins to decrease between
-point particles.
-
-\citet{power+2003} empirically suggest that the ideal softening is
-\begin{equation}{
+Empirically, \citet{power+2003} suggest that the ideal softening is
+\begin{equation}\protect\phantomsection\label{eq:power_softening}{
 h_{\rm grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}},
@@
 number of particles within \(R_{200}\). This choice balances integration
-time and only compromises resolution in collisional regime.
+time and only compromises resolution in collisional regime. For our
+isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
+particles, this works out to be \(0.044\,{\rm kpc}\). Next we consider
+the effect of softening as applied to our isolation halo.
 
-For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
-particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
-slightly smaller softening which was reduced by a factor of
-\(\sqrt{10}\) which appears to improve agreement slightly in the
-innermost regions.
+The two other key numerical parameters in \gadget{} are the tree-force
+and time-step accuracy parameters. Specifically, \gadget{} opens a node
+if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
+the particle is \(r\), side-length is \(l\), and the particle's total
+acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
+adaptive time stepping with integration accuracy set to \(\eta=0.01\).
+(particles must take time-steps smaller than
+\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).
+
+The top panel of Fig.~\ref{fig:methods_convergence} illustrates the
+influence of softening length for simulations of Scl's \smallperi{}
+orbit with \(10^5\) particles. We show a model with the fiducial
+softening length (\(h=0.14\,\kpc\)), and larger and smaller softenings
+by a factor \(\sqrt{10}\). The larger softening length is consistent
+with Eq.~\ref{eq:power_softening}'s prediction. With a larger softening
+length, the halo diverges from the expectation (using \(100\times\) more
+particles) across most radii. On the other hand, the fiducial and
+smaller softening lengths predict similar final results, although the
+smaller softening length simulation deviates slightly more than the
+fiducial, likely a result of more collisional encounters. As computation
+time increases moderately with decreasing softening length, the fiducial
+softening length balances efficiency and accuracy for this simulation.
+While smaller than \citet{power+2003}'s suggestion, our choice is
+designed for more compact, dwarf satellite halos instead of large,
+galaxy-mass isolated halos.
+
+The lower panel of Fig.~\ref{fig:methods_convergence} tests changes to
+these parameters. We include a model with smaller integration accuracy
+(``small timestep'') and a stricter tree-force tolerance (``high acc.
+force''). More precise tolerances on these parameters does not affect
+the evolution within the uncertainties of the final profile.
+
+As demonstrated in this subsection, stricter numerical accuracy and
+modified softening lengths do not affect our results. We thus conclude
+that our simulations are numerically well-converged (up to the
+``convergence'' radius).
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
-\caption{Softening convergence}\label{fig:softening_convergence}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
+\caption[Isolation method convergence]{A comparison of the final
+profiles using different simulation methods for Sculptor's \smallperi{}
+model. The benchmark model is our fiducial \(10^7\) particle run, and
+all other models use \(10^5\) paritcles, with their ``converged radius''
+marked by the black arrow and softening by the vertical bar.
+\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
+lengths. \textbf{Bottom:} Models with more precise timestep accuracy
+(\(\eta=0.001\) compared to fiducial \(\eta=0.005\)) and gravitational
+force accuracy (\(\alpha=0.003\) versus \(\alpha=0.01\)).
+\textbf{Summary:} Except for the model with a larger softening length,
+all simulations agree within uncertainties. \textbf{Todo: softening
+along lower axis always\ldots, changing
+linewidth?}}\label{fig:methods_convergence}
 \end{figure}
-
-\section{Time stepping and force
-accuracy}\label{time-stepping-and-force-accuracy}
-
-In general, we use adaptive timestepping and relative opening criteria
-for gravitational force computations. To verify that these choices and
-associated accuracy parameters minimally impact convergence or speed, we
-show a few more isolation runs (using only 1e5 particles)
-
-\begin{itemize}
-\tightlist
-\item
-  constant timestep (\ldots), approximantly half of minimum timestep
-  with adaptive timestepping
-\item
-  geometric opening, with \(\theta = 0.5\).
-\item
-  strict integration accuracy, (facc = \ldots.)
-\end{itemize}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
-\caption{Isolation method convergence}\label{fig:methods_convergence}
-\end{figure}

### 2025-10-06 thesis/extra_results.tex
thesis; julio oct 4 comments

@@
-\chapter{Additional results an tests}\label{sec:extra_results}
+\chapter{Additional simulation tests}\label{sec:extra_results}
 
@@
 concentration, cored halos, alternative orbits, anisotropy, and
-ellipsoidal models.
+ellipsoidal models. While each of these variables influences resulting
+dark matter evolution, the final stellar component and dark matter
+remnant remains largely across alternate initial conditions. We conclude
+that consideration of these effects likely would not change our
+conclusions.
+
+The following models aim to reproduce the velocity dispersion
+\(\lesssim 1\,\kms\) with a similar present-day half-light radius.
 
@@
 
-The initial concentration (i.e.~choice of \(\vmax\) and \(\rmax\))
-affects total dark matter evolution.
+\subsection{Halo concentration}\label{halo-concentration}
+
+Changing the initial concentration (or \(\vmax\) and \(\rmax\))
+primarily affects total dark matter evolution.
 Fig.~\ref{fig:tidal_tracks_concentration} compares models on the
@@
 \(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
-more mass, the halo evolves to similar final structural parameters, as
-expected in \citet{EN2021}. We note these choices were constrained to
-match Scl's present-day velocity dispersion.
+more mass, the halo evolves to similar final structural parameters.
 
@@
 
-We have good observational evidence for some dwarf galaxies to be cored.
-Even Sculptor has debated claims for having a core
-\citep[e.g.,][]{amorisco+zavala+deboer2014, breddels+helmi2013, battaglia+2008, walker+2009a, richardson+fairbairn2014, agnello+evans2012}.
-To test if a core matter, we adopt a ``cored-NFW'' model,
-\begin{equation}{
+\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
+
+Figure: The evolution of velocity anisotropy.
+
+\subsection{Dark matter cores}\label{dark-matter-cores}
+
+Many dwarf galaxies appear to have dark matter cores. In Scl, the
+presence or absence of a core has been debated
+\citep[e.g.,][]{battaglia+2008, walker+2009a, agnello+evans2012, breddels+helmi2013, amorisco+zavala+deboer2014, richardson+fairbairn2014}.
+To simulate the evolution of a cored dark matter halo, we adopt a
+``cored-NFW'' model, \begin{equation}{
 \rho/\rho   _s = \frac{1}{(1+r/r_s)^2 (r_c/r_s + r/r_s)},
-}\end{equation} where \(r_c\) is the core radius. We adopt
-\(r_c = 0.1r_s\) and adopt an extreme initial mass of \(\vmax=49\,\kms\)
-and \(\rmax=7.1\,\kpc\).
+}\end{equation} where \(r_c\) is the core radius. We set
+\(M_s=0.79\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
+\(r_s=2.76\,\kpc\), \(r_c=0.1r_s\), and use the same truncation as our
+fiducial halo (Eq.~\ref{eq:trunc_nfw}). While this is not the largest
+possible core, the core is similar in size to the stellar component.
+Larger cores would require a more massive halo in order to match the
+velocity dispersion at a constant scale radius.
 
-Sculptor and Ursa Minor may also be highly elliptical in 3D
-\citep[e.g.,][]{an+koposov2022}, especially given their projected
-shapes. We consider
+In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
+cored model as compared to a heavier halo. The cored halo loses more
+total dark matter mass, evolving further along the tidal track as
+expected. However, in order to reproduce the observed velocity
+dispersion, the final halo reaches a \(\rmax\) and \(\vmax\) as the
+fiducial halo. A cored dark matter halo likely cannot be completely
+tidally disrupted and still recover the observed velocity dispersion.
+
+\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
+
+Figure: Cored versus cuspy initial and final density (todo, fix)..
+
+\subsection{An ellipsoidal halo}\label{an-ellipsoidal-halo}
+
+Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
+violating our assumption of spherical symmetry
+\citep[e.g.,][]{an+koposov2022}.
+
+We create an oblate initial dark matter halo using \agama{}. The initial
+snapshot is sampled from the distribution function \begin{equation}{
+\begin{split}
+f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right]
+\end{split},
+}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
+\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
+\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
+\texttt{example\_doublepowerlaw.py} script in \Agama{} solves for the
+best parameters matching a given density profile. We chose to create a
+model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
+For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
+are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
+\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
+\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
+are stable in isolation and produce a halo with near the desired density
+profile and ellipticity. We scale the halo to have a major axis density
+profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
+\(\vmax=48\,\kpc\).
 
 We find the evolution of the oblate halo to be nearly identical to the
-spherical halo, similar to \citet{battaglia+sollima+nipoti2015}.
+spherical halo, similar to \citet{battaglia+sollima+nipoti2015}. The
+oblate halo however becomes spherical as time progresses. If Scl's dark
+matter halo is indeed elliptical, this model may not be an adequate
+description. Possibly, internal rotation, alternative density structure,
+or velocity anisotropy play a role in the stability of oblate initial
+conditions to tidal stripping. A full investigation of these effects is
+left to future work.
 
-While we have compared the evolution of orbits in the MW potential for
-simplicity, because the tidal evolution of Scl is very weak in the LMC
-model, the differences between models become even more slight.
+\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_density_i_f.pdf}}
+
+Slices of the oblate density before and after tidal evolution.
+\textbf{TODO: isodensity contours on RHS projected major/minor axis}.
 
@@
 
+\subsection{Orbital variation}\label{orbital-variation}
+
+In the main text, we focus on recent tidal impacts on Scl and UMi. But,
+due to the long-term orbital uncertainties of Scl and UMi with an LMC
+(see Fig.~\ref{fig:scl_umi_orbit_uncert}), tidally extreme orbits are
+entirely possible. Here, we consider a few examples of extreme orbital
+evolution in the long term case. While the actual orbit of either galaxy
+almost certainly deviates from these examples (discussion in
+Section~\ref{sec:orbital_uncertainties}), these orbits still represent
+extreme past tidal histories.
+
+For a more massive LMC (e.g., the L3M11 model), Scl may have undergone
+an extreme pericentric passage with the MW \(\sim 6\,\Gyr\) ago. We find
+that the 3\(\sigma\) smallest pericentre is \(4\,\kpc\). To test if this
+model forms a different density profile, we simulate this model.
+
+Because of the strong tidal interaction with the MW, the trajectory is
+substantially perturbed from a point orbit. We adjust the initial
+conditions by comparing the change in actions (as calculated in the
+static MW-only potential) before and after the pericentre.
+Table~\ref{tbl:orbit_adjustments} records the iterations we take. Our
+final model is able to approximately reproduce the observed position of
+Scl, but more refinement is required for a closer match.
+
 \begin{figure}
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_orbits_boundmass.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_orbits_boundmass.pdf}}
 \caption[Ursa Minor tidal dependence on orbit]{Similar to
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_impact_i_f.pdf}}
-\caption[Scl MW impact stellar densities]{The initial and final density
-profiles for the Sculptor model which passes through the MW previously.
-The impact does not substantially affect the inner density region. The
-hints of a tidal stream in this model are misaligned to the proper
-motion, likely a result of the LMC.}
-\end{figure}
+\section{Present-day properties}\label{present-day-properties}
+
+We find that this model also does not display strong signs of tidal
+signatures. Fig.~\ref{fig:scl_mw_impact_i_f} shows the initial and final
+stellar profiles for this model (integrated with a lower-resolution
+\(10^6\) particles). The model evolves little more than the
+\smallperi{}. Given that the Jacobi radius would be within \(R_h\), why
+does this model not show strong density perturbations?
+
+We propose that th case Scl undergoes a small MW pericentre is an
+impulsive encounter. Because Scl's apocentre is still \(\sim 300\,\kpc\)
+in this model, the galaxy passes through the MW with velocity
+\(600\,\kms?\). Since the crossing time at the half-light radius is
+about 90 Myr, in one crossing-time, the model moves from pericentre to a
+Galactocentric distance of \(\sim20\,\kpc\), where tides forced have
+dropped substantially. As a result, after \(6\Gyr\) of relatively
+isolated evolution, the break radius falls well outside the observed
+density profile, but the injection of energy uniformly throughout the
+halo only results in moderate DM loss predominantly in the outskirts,
+not a reshaping of the inner density profile. Note that the Jacobi
+radius is derived on circular orbits, not highly elliptical orbits like
+this case.
+
+However, we only note that the stars appear to be consistent but are
+well below the \citet{power+2003}'s numerically converged radius.
+
+\section{The formation of tidal
+tails}\label{the-formation-of-tidal-tails}
+
+As we predict that any tidal tails formed in Scl and UMi would be
+extremely faint, we deferred discussion of the properties of any such
+tails to this section. We note that in extreme cases, \(\sim 2\%\) of
+stars may become unbound in either galaxy, forming tidal tails which may
+be detectible with substantially better observations.
 
@@
 \caption[Scl predicted stream]{The predicted properties of a tidal tail
-in the Scl model.}\label{fig:scl_tidal_stream}
+in the Scl model. \textbf{TODO: titles}}\label{fig:scl_tidal_stream}
 \end{figure}
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_stream.pdf}}
-\caption[Scl predicted stream]{the predicted properties of the stream of
-stars around Scl in the MW impact model. Note that, compared to
-Fig.~\ref{fig:scl_sim_stream}, this model produces a much more scattered
-and misaligned stream, due to the long time since stream formation and
-nonlinear effects of the LMC encounter. \textbf{todo check 2d velocity
-gradients in this case}}\label{fig:scl_mw_impact_tidal_stream}
-\end{figure}
+\section{Summary}\label{summary}
+
+While we have compared the evolution of orbits in the MW potential for
+simplicity, because the tidal evolution of Scl is very weak in the LMC
+model, the differences between models become even more slight. And while
+UMi has a lower pericentre than Scl, the differences between halo
+structure should apply similarly to UMi.

### 2025-10-06 thesis/extra_rv_models.tex
thesis; julio oct 4 comments

@@
-\chapter{LOS velocity modeling}\label{sec:rv_obs}
+\chapter{Line-of-sight velocities: Sample selection and
+modelling}\label{sec:extra_rv_models}
 
 In this section, we analyze the observed line-of-sight (LOS) velocity
-distributions for Sculptor (Scl) and Ursa Minor (UMi). Our goal is to
-understand if the galaxies show evidence of tidal features---for
-example, a velocity gradient or rising outer velocity dispersion. Our
-derived systemic velocities and dispersions are consistent with
-literature. We find weak evidence for a velocity gradient in Scl and a
-rising outer velocity dispersion. However, Scl's velocity gradient is
-misaligned with its proper motion, likely inconsistent with a tidal
-interpretation. Scl's velocity dispersion also mostly rises within the
-effective radius, more likely a feature of inner structure than tidal
-disruption. We find no evidence of a gradient in mean velocity or
+distributions for Sculptor (Scl) and Ursa Minor (UMi). Our aim is to
+test for kinematic tidal signatures by combining literature LOS
+velocities with J+24's membership framework. Our derived systemic
+velocities and dispersions are consistent with literature. We find weak
+evidence for a velocity gradient in Scl. As the gradient is misaligned
+with the orbit, the gradient more likely reflects intrinsic rotation
+than tidal disruption. Scl's velocity dispersion also rises within the
+effective radius, but predominantly in the inner regions, contrary to
+tidal disruption. We find no evidence of a gradient in mean velocity or
 velocity dispersion in UMi. We conclude that Scl and UMi do not show
@@
 
-\section{Data selection}\label{data-selection}
+\section{Data processing and
+selection}\label{data-processing-and-selection}
 
-For both Sculptor and Ursa Minor, we construct literature samples of LOS
-velocity measurements. We combine these samples with J+24's Bayesian
-likelihoods to produce LOS velocity-informed membership probabilities.
-We select stars with moderate to large membership probabilities for
-future analysis.
+We compile LOS velocities from several spectroscopic surveys. For
+Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
+\citet{sestito+2023a}; and \apogee{} \citep[DR17,][]{abdurrouf+2022}.
+For Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
+\citet{sestito+2023b}; and \apogee{}. We then crossmatch all catalogues
+to J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we
+match to the nearest star within 1--3 arcseconds. We combine
+measurements of the same star using inverse-variance weighting.
 
-\begin{table*}[t]
-\centering
-\caption[Spectroscopic LOS velocity measurements]{Summary of velocity measurements and derived properties. }
-\label{tbl:rv_measurements}
-\begin{tabular}{llllllll}
-\toprule
- & Study & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$\\
-\midrule
-Scl & combined & 8945 & 2280 & 2096 & 1981 & 0.9 & \\
- & tolstoy+23 & 3311 & 1701 & 1522 & 1482 & 0.65 & –\\
- & sestito+23a & 2 & 2 & 2 & 2 & 13 & –\\
- & walker+09 & 1818 & 1522 & 1417 & 1328 & 1.8 & 3”\\
- & APOGEE & 5082 & 253 & 170 & 164 & 0.5 & –\\
-UMi & combined & 4714 & 1225 & 1148 & 863 & 2.1 & \\
- & sestito+23b & 5 & 5 & 5 & 5 & 1.8 & –\\
- & pace+20 & 1716 & 1538 & 829 & 678 & 2.5 & 1”\\
- & spencer+18 & 1407 & 970 & 596 & 406 & 0.9 & 2”\\
- & APOGEE & 9500 & 279 & 37 & 67 & 0.6 & –\\
-\bottomrule
-\end{tabular}
-\end{table*}
+To reduce likely binaries, we remove stars with significant velocity
+dispersions.\footnote{Specifically, using that
+  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
+  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
+  distribution with \(N-1\) measurements.}
 
-First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
-not report GaiaDR3 source ID's, we match to the nearest star within 1--3
-arcseconds (see REF Table~\ref{tbl:rv_measurements}). We combine the
-mean RV measurement from each study using the inverse-variance weighted
-mean \(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
-variance \(s^2\).
+We build on J+24's likelihood by adding multiplicative terms in the
+total likelihood for the velocity consistency (see
+Section~\ref{sec:extra_density}). We assume that the satellite and
+background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
+halo/background velocity dispersion of a constant
+\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
+\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
+satellite membership probabilities of greater than 0.2. For Scl, we find
+1918 unique members and UMi, 831.
 
-We next remove stars with significant velocity dispersions as measured
-between observations in a study or between studies, to filter out
-unreliable measurements or possible spectroscopic binaries. By using
-that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
-\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
-distribution with \(N-1\) measurements. This cut typically removes stars
-with reduced chi-squared values
-\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (as a star
-typically fewer than four velocity measurements).
+For UMi, we shifted the velocities of \citet{spencer+2018}
+(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
+a systematic velocity offset. Otherwise all studies appear to agree.
 
 We correct the velocities for the solar motion and on-sky size of the
-galaxy. First, we transform the velocities into the galactic standard of
-rest (GSR), a reference frame centred at the Sun's position but at rest
-with respect to the Galactic centre. Next, we transform velocities to
-correct for the apparent gradient induced by the dwarf's proper motion.
-Let the \(\hat z\) be the direction from the sun to the dwarf galaxy.
-Then if \(\phi\) is the angular distance between the centre of the
-galaxy and the individual star, the corrected radial velocity is then
-\begin{equation}{
-v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
-}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the LOS
-velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
-tangental velocities in RA and Dec in the GSR frame, and \(\theta\) is
-the position angle of the star with respect to the centre of the dwarf.
-The correction from both effects induces an apparent gradient of about
-\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor \citep[see
-also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
-the distance uncertainty and velocity dispersion in quadrature to the RV
-uncertainties for each star. We use the \(v_z\) values for the following
-modelling.
-
-We build on J+24's likelihood framework to add velocity information (see
-Section~\ref{sec:extra_density}). The total likelihood of a star
-belonging to the satellite or foreground/background population is
-\begin{equation}{
-{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm los},
-}\end{equation} where we have added \({\cal L}_{\rm los}\), LOS-velocity
-likelihood term. We assume that the satellite and background
-\(v_{\rm los, gsr}\) distributions are Gaussian. Specifically,
-\begin{equation}{
-\begin{split}
-{\cal L}_{\rm los, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
-{\cal L}_{\rm los, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
-\end{split}
-}\end{equation} where \(f\) is the probability density of a standard
-normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
-velocity and dispersion of the satellite, and \(\delta v_i\) is the
-individual measurement uncertainty. Typically, the velocity dispersion
-dominates the uncertainty budget. We assume a halo/background velocity
-dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
-\citep[e.g.][]{brown+2010}.
-
-Similar to above, we retain stars with the resulting membership
-probability of greater than 0.2. Like in J+24, most stars have
-probabilities of nearly 1 or 0, so the precise choice of cut marginally
-affects the resulting sample.
+galaxy. We transform the velocities into the galactic standard of rest
+(GSR, i.e.~same location as ICRS but velocities relative to the galactic
+centre), and then transform velocities to correct for the apparent
+gradient induced by the dwarf's proper motion.\footnote{Specifically,
+  sub} We define \(v_{\rm gsr}'\) to be velocities in the The correction
+from both effects induces an apparent gradient of about \(1.3\,\kmsdeg\)
+for Sculptor and less for Ursa Minor \citep[see
+also][]{WMO2008, strigari2010}. The uncertainty on this velocity
+correction are less than the individual star uncertainties and our
+derived velocity gradient.
 
@@
 
-To model for possible gradients. We use Monte-Carlo Markov chains (MCMC)
-to calculate the posterior distributions of our models of the intrinsic
-velocity dispersion and possible velocity gradients in Scl and UMi.
+We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
+velocity, velocity dispersion, and possible gradient. We assume that the
+galaxy follows a planar velocity gradient. The mean velocity at a given
+point on the sky, \(\mu\), is assumed to be,
 
-The likelihood is based on the distribution of stars with respect to the
-model's predicted \(\mu_\V\) and \(\sigma_\V\) at each star's position
-in the tangent plane \(\xi\) and \(\eta\). Specifically, the log
-likelihood is \begin{equation}{
-\log {\cal L} = \sum_i \frac{1}{\sigma_i} \log f\left(\frac{\V_i - \mu_i}{\sigma_i}\right),
-}\end{equation} where \(f\) is the probability distribution of a
-standard normal distribution, \(\V_i\) is the velocity of the \(i\)th
-star, and \(\mu_i\) and \(\sigma_i\) are the model's predicted mean and
-velocity dispersion at the \(i\)th star's position \(\xi_i\) and
-\(\eta_i\).
-
-We consider a single model designed to derive systemic velocity and the
-velocity dispersion, detect velocity gradients, and detect a rising
-velocity dispersion. The mean velocity at a given point on the sky,
-\(\mu\), is assumed to be, \begin{equation}{
+\begin{equation}{
 \mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
@@
 
-for tangent-plane coordinates \(\xi\) and \(\eta\), systemtic velocity
+for tangent-plane coordinates \(\xi\) and \(\eta\), systemic velocity
 \(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
@@
 }\end{equation} where \(\sigma_0\) is the system's velocity dispersion
-at \(R_h\), and \(c\) is the velocity dispersion gradient slope.
-
-Our model has parameters with priors \begin{equation}{
-\mu_{0} \sim N(0, \sigma^2_{\rm halo}) \,\kms\\
-\sigma_{0} \sim U(0, 20)\ \kms \\
-a \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
-b \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
-c\sim N(0, 0.3^2)\ {\rm dex} \\
-}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
-the velocity dispersion of the MW halo adopted above,
-\(N(\mu, \sigma^2)\) is a standard normal distribution with mean \(\mu\)
-and variance \(\sigma^2\), and \(U(l, h)\) is a uniform distribution
-between \(l\) and \(h\).
-
-Bayes factors quantify the relative evidence between Bayesian models. We
-can use Bayes factors to compare the above model to simpler models
-without gradients in \(\sigma\) or \(\mu\). Assuming that a
-gradient-free model represents the case \(a=b=0\) (gradient free) or
-\(c=0\) (fixed dispersion) models, we can quickly calculate the relative
-Bayes factors without running another model. We use the Savage-Dickey
-method \citep{dickey+lientz1970}, where the Bayes factor is then the
-relative density of posterior versus prior samples when \(a=b=0\)
-(gradient free) or \(c=0\) (fixed dispersion). In order to calculate
-these densities, we use a kernel density estimator on the posterior
-samples with a Silvermann bandwidth in each dimension.
-
-While this is a more complex model, we find each parameter agrees with
-models of reduced complexity.
+at \(R_h\), and \(c\) is the velocity dispersion gradient slope. We use
+weakly-informative priors, as described in Table~\ref{tbl:scl_rv_mcmc},
+and with \(a, b \sim N(0, 6^2)\,\kmsdeg\).
 
@@
 \end{figure}
+
+Figure: The observed velocity dispersion gradient (black) in 10 equal
+number bins and the derived slopes from the model fitting. Scl shows
+moderate evidence for an increasing velocity dispersion with radius.
+UMi's evidence for a radial gradient is weaker, and Scl's velocity
+dispersion appears to be flat outside of \(R_h\).
 
@@
 
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/sigma_v_gradient.pdf}}
-\caption[Possible gradients in the velocity dispersion]{The observed
-velocity dispersion gradient (black) in 10 equal number bins and the
-derived slopes from the model fitting. Scl shows moderate evidence for
-an increasing velocity dispersion with radius. UMi's evidence for a
-radial gradient is weaker, and Scl's velocity dispersion appears to be
-flat outside of \(R_h\).}
-\end{figure}
-
-For Sculptor, we combine radial velocity measurements from APOGEE,
-\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
-\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
-measurements. We find that there is no significant velocity shift in
-crossmatched stars between catalogues. After crossmatching to high
-quality Gaia stars and excluding significant stellar velocity
-dispersions, we have a sample of 1918 members.
-
 We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
@@
 consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{battaglia+2008}]{walker+2009}.
-
-We detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at
-a position angle of \(-149_{-13}^{+17}\) degrees. Several past work has
-attempted to detect a gradient in Sculptor, but no consensus has been
-reached. \citet{arroyo-polonio+2024} detect a velocity gradient of
-\(4\pm1.5\,\kmsdeg\) in a similar direction using the
-\citet{tolstoy+2023} sample, finding inconclusive statistical evidence.
-They additionally suggest a third chemodynamical component of the galaxy
-which may bias rotation measurements. \citet{battaglia+2008} also detect
-a \(-7.6_{-2.2}^{+3.0}\,\kmsdeg\) velocity gradient along the major
-axis, approximately the same direction. Instead, \citet{strigari2010};
-\citet{martinez-garcia+2023} detect no significant gradient in Sculptor
-using \citet{WMO2009} sample. Note that pre-\emph{Gaia} work did not
-have as strong of a constraint on the proper motion of Scl, which limits
-conclusions of the intrinsic velocity gradient in Scl.
+\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
+gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
+\(-149_{-13}^{+17}\) degrees, similar to past detections
+\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
+see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.
 
@@
 samples have a consistent gradient with 0, most samples have a positive
-velocity gradient, consistent with the rolling median trend.
-
-The velocity gradient in Scl is misaligned with the proper motion
-
-For UMi, we collect radial velocities from, APOGEE,
-\citet{sestito+2023b}, \citet{pace+2020}, and \citet{spencer+2018}. We
-shifted the velocities of \citet{spencer+2018} (\(-1.1\,\kms\)) and
-\citet{pace+2020} (\(+1.1\,\kms\) ) to reach the same scale. We found
-183 crossmatched common stars (passing 3\(\sigma\) RV cut, velocity
-dispersion cut, and PSAT J+24 \textgreater{} 0.2 w/o velocities). Since
-the median difference in velocities in this crossmatch is about 2.2
-km/s, we adopt 1 km/s as the approximate systematic error here. Our
-final sample includes 831 members.
+velocity gradient, consistent with the rolling median trend. The
+velocity gradient in Scl is misaligned with the proper motion
 
 We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
-dispersion of \(8.8\pm0.2\,\kms\) for UMi. This is consistent with
-\citet{pace+2020} and to a lesser extent with \citet{spencer+2018}. We
-do not find evidence for a velocity gradient, consistent with past work
-\citep{pace+2020, martinez-garcia+2023}.
+dispersion of \(8.8\pm0.2\,\kms\) for UMi. We do not find evidence for a
+velocity gradient. This is consistent with previous work
+\citetext{\citealp{pace+2020}; \citealp[somewhat
+with][]{spencer+2018}; \citealp{martinez-garcia+2023}}.
 
-For both Ursa Minor and Sculptor, we also fit models to only data from
-individual surveys. Since the resulting parameters are very similar, we
-conclude that many of the systematic uncertainties are likely smaller
-than the present errors or that each large survey has similar biases.
-
-\section{Discussion and limitations}\label{discussion-and-limitations}
-
-Our model here is relatively simple. Some things which we note as
-systematics or limitations:
-
-\emph{Inter-study systematics and biases}. While basic crossmatches and
-a simple velocity shift, combining data from multiple instruments is
-challenging. This appears to be a minor issue (Sculptor) or is corrected
-for (Ursa Minor).
-
-\emph{Misrepresentative uncertainties}. Inspection of the variances
-compared to the standard deviations within a study seems to imply that
-errors are accurately reported. APOGEE notes that their RV uncertainties
-are known to be underestimates but are a small proportion of our sample.
-We find that most samples report internally-consistent uncertainties for
-multiple observations of the same star.
+\section{Discussion and caveats}\label{discussion-and-caveats}
 
 \emph{Binarity}. Stars in binary systems can inflate the inferred
-system's velocity dispersion. For example, a system with an apparent
-velocity dispersion of \(9\,\kms\) may have a true velocity dispersion
-of \(8\,\kms\) \citep{spencer+2017}. Thus, our measurement is likely
-slightly inflated given the high binary fractions measured in these
-systems \citep{spencer+2018, arroyo-polonio+2023, gration+2025}.
+system's velocity dispersion. For Scl and UMi with high measured
+velocity dispersion, binaries likely add \(\sim 1\,\kms\) to the
+velocity dispersion \citep{spencer+2018, gration+2025}.
 
 \emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
-multiple populations \citep[\citet{pace+2020},
-\citet{tolstoy+2004}]{arroyo-polonio+2024}. Since we only model a single
-population, and each population may have a different extent and velocity
-dispersion, this could result in biased velocity dispersions. However,
-it is unclear how to uniquely determine an overall velocity dispersion
-in a multi-population system.
+multiple populations with different kinematics\citep[\citet{pace+2020},
+\citet{tolstoy+2004}]{arroyo-polonio+2024}. We do not model these
+separately---it is unclear how to define a velocity dispersion in such a
+case.
 
-\emph{Selection effects}. RV studies each have their own selection
-effects, which may affect the resulting dispersion, especially if
-different populations or regions of the galaxy have different velocities
-or velocity dispersions. We do not attempt to correct for this. Given
-that samples extend through the half-light radius of each system, we do
-not expect current samples to miss large features except in the very
-outskirts.
+\emph{Inter-study biases}. While basic crossmatches and a simple
+velocity shift, combining data from multiple instruments is challenging.
+Studies may also have biased selection effects or misrepresentative
+uncertainties, biasing our results. However, besides the \(\sim1\,\kms\)
+velocity offset in Ursa Minor, our results are consistent across
+samples.
 
-Could we observe velocity gradients instead in \emph{Gaia}.
-Unfortunately, given Scl and UMi's distances (\(\gtrsim 70\,\kpc\))
-\emph{Gaia's} systemic proper motion uncertainty of \(0.017\,\masyr\)
-corresponds to a velocity of \(5\,\kms\), making the detection of
-gradients of amplitude less than \(5\,\kms\) challenging. In addition,
-the typical uncertainties of member stars of Scl and UMi are even
-larger. While in future data releases, \emph{Gaia} uncertainties may
-permit study in the internal motion of dwarfs, the current \emph{Gaia}
-proper motions are not yet precise enough.
+\emph{Tangental motions.} Precise star-by-star proper motions would
+further test for kinematic disequilibrium. Presently, \emph{Gaia}
+uncertainties are too large to permit such studies for Sculptor and Ursa
+Minor.\footnote{Specifically, typical proper motion uncertntanties for
+  faint, member stars of Scl and UMi are around \(0.5\,\masyr\),
+  corresponding to velocities \(\sim 100\,\kms\).}
 

### 2025-10-06 thesis/introduction.tex
thesis; julio oct 4 comments

@@
 
-However, there has long been speculation that both Scl and UMi may have
-been influenced by the Milky Way's tidal field. Already,
-\citet{innanen+papp1979} found RR Lyrae candidate Scl members
-\citep[from][]{vanagt1978} out to 180' in an elongated distribution,
-speculating this to be tidal disruption. Many later density profile
-determinations noted Scl's elongation and apparent outer density excess
+However, many works speculated that Scl and UMi were influenced by the
+Milky Way's tidal field. Already, \citet{innanen+papp1979} found RR
+Lyrae candidate Scl members \citep[from][]{vanagt1978} out to 180' in an
+elongated distribution, speculating this to be tidal disruption. Later
+density profile determinations noted Scl's elongation and apparent outer
+density excess
 \citetext{\citealp{eskridge1988}; \citealp{IH1995}; \citealp{walcher+2003}; \citealp{westfall+2006}; \citealp[but
@@
 origin and role of numerics and assumptions in this discrepancy is
-unclear \citep[ \citet{riley+2024}]{panithanpaisal+2021, shipp+2023l}.
-We shall use idealized simulations here to assess tidal effects after
-the satellite's infall into the MW halo.
+unclear \citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall
+use idealized simulations here to assess tidal effects after the
+satellite's infall into the MW halo.
 
@@
 particle based on the relative densities of the distribution functions
-in energy space (see \citet{ref:sec_painting_stars}). The stars follow
+in energy space (see Section~\ref{sec:painting_stars}). The stars follow
 initially a 2D exponential profile with a scale radius

### 2025-10-06 thesis/results.tex
thesis; julio oct 4 comments

@@
-As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
-we test whether Galactic tides are responsible for the extended density
-profiles of Scl and UMi. In this Chapter, we analyze tailored N-body
-simulations, using the methods described in Chapter \ref{sec:methods},
-to assess the tidal impact of the Galactic potential. To anticipate our
-main conclusion, we find that tides drive dark matter loss in both
-systems but leave their compact stellar components largely unaffected.
-The Large Magellanic Cloud (LMC) perturbs the orbit of Scl, and somewhat
-UMi, yet the resulting tidal effects are weaker than in a Milky Way-only
-potential. Our simulations demonstrate that recent tides are unlikely to
-have altered the stellar structure of Scl or UMi.
+As discussed in Chapters \ref{sec:introduction} and
+\ref{sec:observations}, this thesis aims to test whether Galactic tides
+are responsible for the extended density profiles of Scl and UMi. In
+this Chapter, we analyze tailored N-body simulations, using the methods
+described in Chapter \ref{sec:methods}, to assess the tidal impact of
+the Galactic potential. To anticipate our main conclusion, we find that
+tides drive dark matter loss in both systems but leave their compact
+stellar components largely unaffected. The Large Magellanic Cloud (LMC)
+has the potential of substantially perturbing Scl's and even UMi's
+orbit, yet the resulting tidal effects are still too weak to account for
+the extended outer profiles. Our simulations demonstrate that recent
+tides are unlikely to have altered the stellar structure of Scl or UMi.
 
 In this Chapter, we consider Scl first, describing tidal effects from
-the MW on dark matter and stellar components. Next, we consider how
-accounting for the LMC amends our conclusions. We then analyze UMi in a
-similar manner, considering in turn the dark matter evolution, stellar
-evolution, and orbital effects of the LMC.
+the MW on its dark matter and stellar components. Next, we consider how
+accounting for the LMC may affect our conclusions. We then analyze UMi
+in a similar manner, considering in turn the dark matter evolution,
+stellar evolution, and orbital effects of the LMC.
 
@@
 As a representation of an extreme tidal history, we initially
-investigate the \smallperi{} Scl model.
+investigate the \smallperi{} orbit, described in
+Section~\ref{sec:scl_smallperi}, chosen to maximize possible effects of
+Galactic tides.
 
 Scl experiences moderate tidal mass loss after 10 Gyrs of evolution.
-Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and
+Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and the
 formation of diffuse streams trailing and leading Sculptor's orbit.
 Because tidal stripping can be described as a gradual removal of the
-least bound particles, most mass loss occurs in the outer halo.
+least bound particles, most mass loss occurs in the outer halo. Instead,
+the inner regions of the galaxy may be relatively unaffected.
 
@@
 outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
-mass, as listed in Table~\ref{tbl:scl_sim_results}. However, the inner
-structure is not expected to be affected, as the Jacobi radius is over
-3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
-significant amounts of mass, but mostly from the outer halo.
+mass (see Table~\ref{tbl:scl_sim_results}). However, the inner structure
+is not expected to be affected, as the Jacobi radius is over 3kpc,
+outside of the initial and final \(\rmax\) (see
+Table~\ref{tbl:scl_sim_results} and Fig.~\ref{fig:scl_tidal_track}).
+Thus, tides may remove significant amounts of mass, but mostly from the
+outer halo.
 
@@
 
-\subsection{Evolution of Sculptor's
-stars}\label{evolution-of-sculptors-stars}
+\subsection{Evolution of Sculptor's stars}\label{sec:scl_sim_stars}
 
@@
 
-Sculptor may not have been born as an exponential galaxy, and more
-extended stellar profiles would be more sensitive to tides. As an
-approximation of the present-day extended density profile, we initialize
-a model with Plummer stars instead. As shown in
-Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the Plummer model loses more
-stellar mass and forms more luminous tidal tails. Observations reaching
-surface densities \(\sim10\) times fainter than our data could reveal a
-stream in this case. Nevertheless, the radial extent probed by our data
-remains nearly unchanged by tidal evolution.
+This result implies that Scl's extended profile cannot be reproduced by
+Galactic tides operating on an initially exponential profile. The weak
+effect of tides suggests that the outer profile of Sculptor is innate,
+and not the result of tidal evolution. We check this assertion by
+choosing a different initial stellar profile which matches the observed
+profile and assessing how it evolves on the smallperi orbit. We show in
+Fig.~\ref{fig:scl_smallperi_plummer_i_f} that a Plummer profile (instead
+of an exponential) provides an adequate fit to Scl's observed profile.
+The Plummer model loses more stellar mass and forms more luminous tidal
+tails. Observations reaching surface densities \(\sim10\) times fainter
+than our data could reveal a stream in this case. Nevertheless, over the
+radial extent probed by our data, the stellar profile remains nearly
+unchanged by tidal evolution.
 
@@
 (Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
-Given that no orbits of Scl in this potential produce significantly
-smaller break or Jacobi radii, it is unlikely any other orbit would
-produce a significant, observable density excess.
+Since no orbits of Scl produce significantly smaller break or Jacobi
+radii, it is unlikely any orbit would produce an observable density
+excess.
 
@@
 e.g.,][]{patel+2020, battaglia+2022, correamagnus+vasiliev2022}. In this
-section, we examine how the LMC affects the orbital history of Sculptor.
+section, we examine how the LMC may affect the orbital history of
+Sculptor.
 
@@
 Fig.~\ref{fig:scl_lmc_orbits_effect}. In the MW-only potential, Scl's
-orbit is typical of a long-term MW satellite. However, Scl's passage by
-the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term orbit---Scl instead
-reaches an apocentre of nearly \(300\,\kpc\). Scl may even be on first
-infall, depending on the MW and LMC mass. Scl's orbit is
-counter-rotating to the LMC, so Scl is unlikely to be a LMC satellite.
+orbit is typical of a long-term MW satellite. However, Scl's closest
+approach to the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term
+orbit---Scl is inferred to reach an apocentre of nearly \(300\,\kpc\).
+Scl may even be on first infall, depending on the MW and LMC mass. Scl's
+is orbiting the Milky Way on a similar plane to the LMC, but in the
+opposite direction---Scl is unlikely to be a LMC satellite.
 
-The timing of the LMC encounter implies a break radius consistent with
-the beginning of the observed density excess. We select the orbit with
-the \(6\sigma\)-smallest LMC pericentre. The orbit is selected following
-the procedure in Section~\ref{sec:orbital_estimation} except with
-observational uncertainties doubled. Results are similar when selecting
-for a small MW pericentre instead. This \texttt{LMC-flyby} orbit, is
-integrated back in time \(2\,\Gyr\) ago to isolate recent tidal effects.
-We modify Scl's initial halo to have \(\rmax = 2.5\,\kpc\) and
-\(\vmax = 25\,\kms\), slightly reducing the initial stellar velocity
-dispersion. Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected
-orbit in black and Table~\ref{tbl:orbit_ics} records the initial
-conditions.
+Promisingly, the timing of the LMC encounter implies a break radius
+(\(\sim 25'\), from Table~\ref{tbl:scl_lmc_sim_stars}) consistent with
+the beginning of Scl's observed density excess (see
+Fig.~\ref{fig:classical_dwarfs_densities}, and
+Section~\ref{sec:data_density_profiles}). To probe this further, we
+select an orbit with the smallest LMC-Scl pericentre (20,kpc) in the
+\texttt{L3M11} model, consistent with Scl's present-day positions and
+velocities. The orbit is selected following the procedure in
+Section~\ref{sec:orbital_estimation} (with uncertainties doubled). This
+\texttt{LMC-flyby} orbit, is integrated back in time \(2\,\Gyr\) ago to
+isolate recent tidal effects. We modify Scl's initial halo to have
+\(\rmax = 2.5\,\kpc\) and \(\vmax = 25\,\kms\), slightly reducing the
+initial stellar velocity dispersion.
+Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected orbit in black
+and Table~\ref{tbl:orbit_ics} records the initial conditions.
 
@@
 
-The combined tidal effect of the MW and LMC are even weaker for Scl than
-in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
-matter evolution of Scl and the passage of the LMC. With only one MW
-pericentre, Scl's dark matter is less disrupted than the previous
-MW-only model. The subsequent LMC passage modifies Scl's orbit but has
-otherwise little effect. As a result, the dark matter structure evolves
-mildly and \(\sim 50\%\) of mass remains bound
-(Table~\ref{tbl:scl_lmc_sims}).
+Perhaps surprisingly, the combined tidal effect of the MW and LMC are
+weaker for Scl than in the MW-only case.
+Fig.~\ref{fig:scl_lmc_sim_images} shows the dark matter evolution of Scl
+and the passage of the LMC. With only one MW pericentre, Scl's dark
+matter is less disrupted than the previous MW-only model. The subsequent
+LMC passage modifies Scl's orbit but has otherwise little effect. As a
+result, the dark matter structure evolves mildly and \(\sim 50\%\) of
+mass remains bound (Table~\ref{tbl:scl_lmc_sims}).
 
-Correspondingly, the stellar component is nearly unchanged.
-Fig.~\ref{fig:scl_lmc_i_f} shows the projected stellar distributions and
-density profiles of this model. While the break radius is within the
-observed density profile, tidal effects are too weak to be detectable.
-Structural properties of the stars similarly evolve little
-(Table~\ref{tbl:scl_lmc_sim_stars}).
+JFN: This is mainly because of the altered orbit of Sculptor, which is
+now inferred to come from a much larger apocentre, and, thus, to have
+completed fewer orbits than in the previous MW-only ``smallperi'' orbit.
+This reduces the number of pericentric passages, and therefore the net
+tidal effect.
+
+Correspondingly, the stellar component is nearly unchanged by the
+combined MW and LMC tides. Fig.~\ref{fig:scl_lmc_i_f} shows the
+projected stellar distributions and density profiles of this model.
+While the break radius is within the observed density profile, tidal
+effects are too weak to be detectable. Structural properties of the
+stars similarly evolve little (Table~\ref{tbl:scl_lmc_sim_stars}).
 
@@
 the interaction between Scl and the LMC is brief as their relative
-velocity is \(\sim 400\,\kms\) at closest approach.
+velocity is \(\sim 400\,\kms\) at closest approach. Furthermore, the
+Jacobi radius due to the LMC still falls outside the observed density
+profile (Fig.~\ref{fig:scl_lmc_i_f}) and the MW Jacobi radius is even
+larger. As a result, tides in a MW and LMC potential are even weaker
+overall than for the MW-only orbit.
 
@@
 distribution are resilient to tides and any tidal effects are
-well-outside the reach of current observations. We then find that the
-LMC strongly perturbs Scl's orbit---in this case, Scl may be on first
-infall. However, with only 1 pericentre each for the LMC and MW, the
-combined tides are weaker than for our initial model. In either case, we
-conclude that tides are unlikely to affect Sculptor's stellar component.
+well-outside the reach of current observations. We have also found that
+the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
+first infall. However, with only 1 pericentre each for the LMC and MW,
+the combined tides are weaker than for our initial model. In either
+case, we conclude that tides are unlikely to affect Sculptor's stellar
+component.
 
@@
 Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
-substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
+substantially more mass than Sculptor. \textbf{todo: thinner and
+truncated tidal track, thinner stellar
+component}}\label{fig:umi_tidal_track}
 \end{figure}
@@
 component is much smaller than the initial, but still evolves along the
-predicted tidal track Fig.~\ref{fig:umi_tidal_track}. Despite the more
+predicted tidal track (Fig.~\ref{fig:umi_tidal_track}). Despite the more
 substantial tidal evolution, the Jacobi radius is still large, lying at
@@
 
-Similar to Scl, we consider a model where UMi's stars are initially a
-more extended Plummer profile, resembling the present-day density
-profile. The stellar evolution of the Plummer stellar component is
-similar (Fig.~\ref{fig:umi_plummer_i_f}). However, because there are
-more loosely-bound stars, the Plummer model looses nearly 4\% of its
-initial stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}),
-and tidal features may be detectible if we measure densities 2 orders of
-magnitude fainter than our present data. We show the properties of a
-stream in the Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a
-stream is unlikely to be observable in the near future. We conclude that
-tides do not strongly affect the stellar component of this model.
+The break and Jacobi radii fall well outside the observed stellar
+profile. Tides would have to be far stronger to affect the observed
+stellar component. As a result, the minimal tidal evolution of this
+model is not unexpected.
+
+As for Scl (Section~\ref{sec:scl_sim_stars}), we also consider a model
+where UMi's stars are initially a more extended Plummer profile,
+resembling the present-day density profile. The stellar evolution of
+this Plummer stellar component is similar
+(Fig.~\ref{fig:umi_plummer_i_f}). However, because there are more
+loosely-bound stars, the Plummer model looses nearly 4\% of its initial
+stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}), and tidal
+features may be detectable if we measure densities 2 orders of magnitude
+fainter than our present data. We show the properties of a stream in the
+Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a stream is
+unlikely to be observable in the near future. We conclude that tides do
+not strongly affect the stellar component of this model.
 
@@
 the orbit of Ursa Minor. Predominantly, the effect is to increase the
-orbital period, apocentre and pericentre. However, the orbit remains in
-a similar plane and with similar morphology. As UMi is on the opposite
-side as the Galaxy of the LMC, this is not unexpected.
+orbital period, apocentre and pericentre. Yet, the orbit remains in a
+similar plane and with similar morphology. As UMi is on the opposite
+side as the Galaxy of the LMC, and has a closest LMC approach of
+\(\gtrsim 100\,\kpc\), this is not surprising.
 
 The deviation from the MW-only orbit is mostly due to the LMC-induced
-reflex motion of the Milky Way. Because the MW centre, and our origin,
-is accelerated towards the LMC, UMi experiences a fictitious
-acceleration opposite to the acceleration experienced by the MW centre,
-causing the orbit to increase in characteristic radius.
+reflex motion of the Milky Way. Because the MW centre is accelerated
+towards the LMC and away from UMi, UMi's orbit to increase in
+characteristic radius.
 
@@
 
-\section{Modeling uncertainties}\label{modeling-uncertainties}
+\section{Modelling uncertainties}\label{modelling-uncertainties}
 
@@
 
-As Section~\ref{sec:results} shows, tides marginally affected the
-stellar components of Scl and UMi, even under extreme orbits. While
-Section~\ref{sec:results} only presents select models, alternative
-initial conditions do not affect our qualitative conclusions on tidal
-effects expected for Scl and UMi in the Galactic (and LMC) potential.
+As the above results show, tides only marginally affected the stellar
+components of Scl and UMi, even orbit chosen to have the smallest
+observationally-consistent pericentres. While we have only presented
+select models, alternative initial conditions do not affect our
+qualitative conclusions on tidal effects expected for Scl and UMi in the
+Galactic (and LMC) potential.
 
-We neglect baryonic physics. Since Scl and UMi have predominantly stars
-older than \(\sim 9\) Gyr
-\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025},
-gas dynamics are unlikely to affect recent evolution. A collisionless
-dark-matter only simulation should be an excellent approximation.
+Although our analysis neglects baryonic physics, Scl and UMi have
+predominantly stars older than \(\sim 9\) Gyr
+\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025}.
+So, gas dynamics are unlikely to affect recent evolution. A
+collisionless dark-matter only simulation should therefore be an
+excellent approximation.
 
@@
 dispersion. Although Scl's fiducial halo is more concentrated, less
-concentrated and cored halos evolve similarly (see
-Section~\ref{sec:extra_results}).
+concentrated and cored halos evolve similarly (see Appendix
+\ref{sec:extra_results}).
 
@@
 are elliptical, and halos are expected to be radially anisotropic
-\citep[e.g.,][]{navarro+2010}. We test example non-spherical and
-anisotropic models in Section~\ref{sec:extra_results}, finding that
-these assumptions likely do not alter our conclusions.
+\citep[e.g.,][]{navarro+2010}. We test non-spherical and anisotropic
+models in Appendix \ref{sec:extra_results}, finding that these
+assumptions likely do not alter our conclusions.
 
-A system's observed velocity dispersion directly constrains the mean
-density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
-required to disrupt the stellar component does not strongly depend on
-the inner halo structure (via the Jacobi radius). Alternative initial
-conditions may influence the total mass evolution but would produce a
-similar final stellar structure.
+Alternative initial conditions may influence the total mass evolution
+but should produce a similar final stellar structure. A system's
+observed velocity dispersion directly constrains the mean density within
+\(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force required to
+disrupt the stellar component does not strongly depend on the inner halo
+structure (via the Jacobi radius).
 
@@
 potentials neglect many unknown details, including triaxiality, halo
-twisting, mass evolution, and substructure. As a result, calculated
-orbits may diverge significantly from the true orbits of satellites
-\citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the Milky Way and
-dynamical friction implies that orbits are typically less bound in the
-past (less affected by tides). However, orbital energy and angular
-momentum are not conserved in cosmological simulations. Orbits in
-analytic potentials may overestimate the pericentre and underestimate
-the maximum tidal stress \citep[although typically not by enough to
-change our conclusions,][]{santistevan+2023, santistevan+2024}.
+twisting, mass evolution, and substructure. Due to these inadequacies,
+calculated orbits may diverge significantly from the true orbits of
+satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the
+Milky Way and dynamical friction imply that orbits are typically less
+bound in the past (less affected by tides). Orbital energy and angular
+momentum of subhalos are not conserved in cosmological N-body
+simulations. Consequently, orbits in analytic potentials may
+overestimate the pericentre and underestimate the maximum tidal stress
+\citep[although typically not by enough to change our
+conclusions,][]{santistevan+2023, santistevan+2024}.
 
-Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates the uncertainties of the
-long-term orbits of Scl and UMi. Scl distance could range from
-4--\(300\,\kpc\) any time more than 4Gyr ago. And, if the LMC had a
-previous pericentre, UMi may have been an LMC satellite
-\citep[see][]{vasiliev2024}. Notably, these potentials are still
-idealized in many ways.
+As an example, Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates how
+changes to the LMC potential modifies the long-term orbital trajectories
+of Scl and UMi. More than 4 Gyr ago, orbits of Scl diverge
+substantially. Some orbits are near apocentres of \(\sim 300\,\kpc\)
+when others approach pericentres as small as \(\sim 10\,\kpc\). Ursa
+Minor's orbit is more stable until the possible previous LMC pericentre.
+In some cases, Ursa Minor may have been bound to the LMC.
 
@@
 \ref{sec:extra_results}, finding it still insufficient to produce the
-observed density profile. We conclude our simulated orbits represent
-reasonable extremes for \emph{recent} tidal effects. Past encounters
-with the LMC are revisited below as a form of ``pre-processing'' in
-Section~\ref{sec:stellar_halos}.
+observed density profile. Regardless, we conclude our simulated orbits
+represent reasonable extremes for \emph{recent} tidal effects. Past
+encounters with the LMC are revisited below as a form of
+``pre-processing'' in Section~\ref{sec:stellar_halos}.
 
@@
 result, recent tides are unlikely to affect the stellar distributions of
-Sculptor or Ursa Minor.
+Sculptor and Ursa Minor.

### 2025-10-06 thesis/data.tex
full draft of appendicies, improved formatting

@@
 \section{\texorpdfstring{Satellite stellar membership with
-\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{satellite-stellar-membership-with-gaia}
+\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{sec:the_algorithm}
 
@@
 \ref{sec:density_extra} discusses the details of quality cuts and the
-likelihood calculation.
+likelihood calculation \citep[see also,][]{MV2020a, jensen+2024}.
 
@@
 J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
-line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
-details). While the latter is the highest-purity membership sample, we
-do not use this sample for stellar density analysis due to its
-incompleteness and complex selection function.
+line-of-sight velocity information (see Appendix
+\ref{sec:extra_rv_models} for details). While the latter is the
+highest-purity membership sample, we do not use this sample for stellar
+density analysis due to its incompleteness and complex selection
+function.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
 \caption[Sculptor sample selection]{The distributions of various samples
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
 \caption[Ursa Minor sample selection]{Similar to
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
 \caption[Fornax sample selection]{Similar to

### 2025-10-06 thesis/discussion.tex
full draft of appendicies, improved formatting

@@
 Tidal effects have been long invoked to explain features of dwarf
-galaxies e.g.~Section \ref{sec:scl_umi_obs_tides},
-\citep{mayer+2001a, tsujimoto+shigeyama2002} \textbf{Check this
-citation}. Yet, growing work suggests tides may not be as commonplace as
-suspected. For example, \citet{read+2006} suggest that the absence of
-rising radial velocity dispersion profiles in dwarfs implies a lack of
-tidal disruption. \citet{penarrubia+2009}, introducing the break radius
-as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
-orbits were inconsistent with observable tidal effects. Recently,
-\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
+galaxies \citep[e.g.,
+Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}
+\textbf{Check this citation}. Yet, growing work suggests tides may not
+be as commonplace as suspected. For example, \citet{read+2006} suggest
+that the absence of rising radial velocity dispersion profiles in dwarfs
+implies a lack of tidal disruption. \citet{penarrubia+2009}, introducing
+the break radius as a diagnostic (Eq.~\ref{eq:r_break}), showed that
+most satellite orbits were inconsistent with observable tidal effects.
+Recently, \citet{pace+erkal+li2022} use a Jacobi radius-based criterion
 (Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the

### 2025-10-06 thesis/extra_density.tex
full draft of appendicies, improved formatting

@@
 
-In this section, we describe the details of J+24's likelihoods. We then
-test the reliability of the derived density profiles for Scl and UMi,
-comparing methodologies, samples, and to literature. Finally, we present
-a comparison of J+24's samples to a non-parametric Bayesian density
-profile, finding similar results in the inner regions. Past a ``limiting
-radius'', J+24's density profiles become susceptible to the likelihood
-specification. However, all methods agree up to the limiting radii. In
-conclusion, the density profiles from J+24 are robust except in the very
-outer regions.
+In this section, we detail the J+24 membership algorithm and test the
+resulting density profiles for Scl and UMi. We compare profiles derived
+with alternative methodologies and literature results. We also present a
+non-parametric Bayesian density profile, which reproduces J+24's results
+in the inner regions but diverges when the background dominates. In
+summary, we find the density profiles are robust up to the
+background-limited radius derived here.
 
@@
 To create a high-quality sample, J+24 select stars initially from Gaia
-within a 2--4 degree circular region centred on the dwarf satisfying:
+within a 2--4 degree region around each satellite with high-quality
+astrometry, reliable photometry, and consistent parallaxes and broadly
+consistent proper motions and colours. Stars are removed if they have
+excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
+see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
+expectations \citep[from][]{riello+2021}, proper motions magnitudes
+\(>10\,\masyr\) in \(\mu_{\alpha*}\) or \(\mu_\delta\), magnitudes
+brighter than the tip of the red giant branch or fainter than \(G=22\),
+or colours outside \(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry
+is dereddened with \citet{schlegel+finkbeiner+davis1998} extinction
+maps.
 
-\begin{itemize}
-\tightlist
-\item
-  Solved astrometry, magnitude, and colour.
-\item
-  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\). Ensures high
-  quality astrometry. \texttt{ruwe} is a measure of the excess
-  astrometric noise on fitting a consistent parallax-proper motion
-  solution \citep[see][]{lindegren+2021}.
-\item
-  3\(\sigma\) consistency of measured parallax with dwarf's distance,
-  with \citet{lindegren+2021}'s zero-point correction. Note that the
-  typical satellite parallax is very small.
-\item
-  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
-  10\(\,{\rm mas\ yr^{-1}}\). This corresponds to tangental velocities
-  of \(\gtrsim 500\) km/s at distances larger than 10 kpc.
-\item
-  Corrected colour excess is within 3\(\sigma\) of the expected
-  distribution from \citet{riello+2021}. Removes stars with unreliable
-  photometry.
-\item
-  De-reddened \(G\) magnitude is between
-  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
-  and stars significantly brighter than the tip of the red giant branch
-  (TRGB) magnitude plus the distance modulus uncertainty
-  \(\sigma_{\rm DM}\).
-\item
-  De-reddened colour is between \(-0.5 < {\rm BP - RP} <  2.5\). Removes
-  stars substantially outside the expected CMD.
-\end{itemize}
-
-Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
-extinction maps.
-
-J+24 define likelihoods \({\cal L}\) measuring consistency with the MW
-stellar background (\({\cal L}_{\rm bg}\)) or the satellite galaxy
-(\({\cal L}_{\rm sat}\)). The total likelihoods are the product of a
-spatial, PM, and CMD term: \begin{equation}{
-{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
-}\end{equation}
-
-Each likelihood is normalized over their respective 2D parameter space
-for both the satellite. To control the relative frequency of member and
-background stars, \(f_{\rm sat}\) representing the fraction of member
-stars in the field. The total likelihood for any star in this model is
-the \(f_{\rm sat}\)-weighted sum of the foreground and background,
-\begin{equation}\protect\phantomsection\label{eq:Ltot}{
-{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
-}\end{equation} Finally, the probability that any star belongs to the
-satellite is then given by \begin{equation}{
-P_{\rm sat} = 
-\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
-= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
-}\end{equation}
-
-For the satellite's spatial likelihood, J+24 consider both one-component
-and a two-component density models. The one component model is
-constructed as a single exponential profile ( surface density
-\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\).
-\(R_s\) is fixed to the equivalent \(R_h\) value from
-\citet{munoz+2018}'s Sérsic fit. Additionally, structural uncertainties
-(for position angle, ellipticity, and scale radius) are sampled over to
-construct the final likelihood map. The two-component model instead adds
-a second exponential,
-\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
-scale radius is fixed, and the outer scale radius and magnitude of the
-second component \(R_{\rm outer}\), \(B\) are free parameters.
-Structural property uncertainties are not included in the two-component
-model.
+J+24 construct satellite likelihoods in spatial, proper motion, and CMD
+space following expected satellite properties. J+24 model the spatial
+likelihood as either a one or two-component exponential
+(Eq.~\ref{eq:exponential_law}). The structural parameters of the inner
+component are fixed, and marginalized over if one-component. The outer
+profile scale radius and normalization are free parameters.
 
 The PM likelihood is a bivariate gaussian with variance and covariance
-equal to each star's proper motions. J+24 assume the stellar PM errors
-are the main source of uncertainty.
+equal to each star's proper motions.
 
-The satellite's CMD likelihood is based on a Padova isochrone
-\citep{girardi+2002}. The isochrone has a matching metallicity and 12
-Gyr age (except 2 Gyr is used for Fornax). The isochrone is given a
-Gaussian colour width of 0.1 mag plus the \emph{Gaia} colour uncertainty
-at a given magnitude. The horizontal branch is modelled as a constant
-magnitude extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB
-stars and a 0.1 mag width plus the mean colour error). A likelihood map
-is constructed by sampling the distance modulus in addition to the CMD
-width, taking the maximum of RGB and HB likelihoods.
+J+24 model the CMD as a Padova isochrone \citep{girardi+2002}, with
+matching metallicity, 12 Gyr age (or 2 Gyr for Fornax), and Gaussian
+width of 0.1 mag plus the \emph{Gaia} colour uncertainty at each
+magnitude. The horizontal branch is modelled as a
+constant-magnitude\footnote{Specifically, the mean magnitude of the 12
+  Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
+sequence extending blue of the CMD with the same width as the RGB. The
+CMD likelihoods are marginalized over the distance modulus and take the
+maximum of the RGB and horizontal branch likelihoods.
 
-The background likelihoods are instead empirically constructed. Stars
-stars outside of 5\(R_h\) passing the quality cuts estimate the
-background density in PM and CMD space. The density is a sum of
-bivariate gaussians with variances based on Gaia uncertainties (and
-covariance for proper motions). The spatial background likelihood is
-assumed to be uniform over the field.
+The background likelihoods are determined empirically as a kernel
+density estimates from stars outside 5\(R_h\) in PM and CMD space. The
+spatial background likelihood is uniform.
 
-J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
-\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
-with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
-informative. The proper motion single component prior is same as
-\citet{MV2020a}: a normal distribution with mean 0 and standard
-deviation \(100\ \kms\). If using the 2-component model, the prior is
-instead a uniform distribution spanning 5\(\sigma\) of single component
-case w/ systematic uncertainties. \(f_{\rm sat}\) (and \(B\)) has a
-uniform prior from 0 to 1. \(R_{\rm outer}\) has a uniform prior only
-restricting \(R_{\rm outer} > R_s\). The mode of each parameter from the
-MCMC are then reported and used to calculate the final \(P_{\rm sat}\)
-values.
+J+24 derive the distributions of parameters (proper motions, satellite
+fraction \(f_{\rm sat}\), and second spatial component if included)
+through Monte Carlo Markov chain sampling with broad or
+weakly-informative priors. The posterior modes are used to calculate the
+final \(P_{\rm sat}\) values.
 
-\section{Alternative sample selection}\label{sec:simple_selection}
+\section{\texorpdfstring{Possible biases in \emph{Gaia}-derived density
+profiles}{Possible biases in Gaia-derived density profiles}}\label{sec:density_extra}
 
-As J+24's probabilistic model may systematically bias the density
-profile (see discussion below), we consider samples selected both from
-\emph{Gaia} and external surveys using simple, absolute selection cuts.
+In this section, we test how assumptions in J+24's algorithm or biases
+in \emph{Gaia} may affect density profiles. In all cases, density
+profiles converge until \(R_{\rm ell} \approx 60\,{\rm arcmin}\), where
+background contamination likely dominates.
 
-For \emph{Gaia} data, we filter stars by four simple selection criteria.
-We require \texttt{ruwe} \textless{} 1.3 as above, selecting only stars
-with high-quality astrometry. We then require the star's parallax to be
-\(3\sigma\)-consistent with the dwarf's distance. Next, we select stars
-with proper motions with \(1\masyr\)\hspace{0pt} radius of the galaxy's
-measured value. Finally, we filter stars to lie on the dwarf's CMD in
-(\(G_{\rm BP} - G_{\rm RP}\),\(G\)), within the vertices defined in
-Table~\ref{tbl:colour_cuts}. This polygon is derived based on the
-colour-magnitude diagram of stars in the centre of the dwarf galaxy,
-where the dwarf dominates the stellar density.
-
-In addition to \emph{Gaia}, we select stars in the Scl field from the
-deeper, photometric DELVE survey \citep{drlica-wagner+2022}. We select
-sources DELVE data release 2 within an ellipse of radius 150 arcminutes.
-We then only include sources which are categorized as likely stars and
-sources with reliable (flags \textless=4) \(g\) and \(r\) magnitudes. We
-then filter by the \(g\) and \(r\) magnitudes, selecting the empirical
-CMD cut using the vertices in Table~\ref{tbl:colour_cuts}.
-
-Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
-Survey \citep[UNIONS][]{gwyn+2025}. We select sources within an ellipse
-of radius 230 arminutes. We keep sources with no \texttt{FLAGS\_CFIS}
-set and the \texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended
-sources). We then use the polygon bounded by the vertices in
-Table~\ref{tbl:colour_cuts} to select the observed red giant branch,
-horizontal branch, and main sequence of the dwarf.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/extra_cmd_selection.png}}
-\caption[CMD cuts]{Colour-magnitude diagram cuts used in samples in this
-section. Orange points are selected, black are non-selected stars within
-\(1R_h\), and grey stars are for all stars. \textbf{todo, only plot
-black/grey and cmd outline\ldots{}}}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
-\caption[DELVE and UNIONS spatial distribution of stars]{The
-distribution of DELVE and UNIONS selected stars in Scl and UMI}
-\end{figure}
-
-\section{\texorpdfstring{Uncertainties and possible biases in
-\emph{Gaia}-derived density
-profiles}{Uncertainties and possible biases in Gaia-derived density profiles}}\label{sec:density_extra}
-
-In this section, we discuss additional tests and verification of the
-derived density profiles. In particular, we check that alternative
-methodologies do not substantially affect the density profile. In all
-cases, the density profiles appear to have excellent convergence out to
-\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
-where the background dominates.
-
-\textbf{Spatial likelihood.} J+24's algorithm was designed specifically
-to detect the presence of a density excess in satellites and identify
-individual stars for spectroscopic follow-up. We are instead interested
-in constraining a dwarf galaxy's stellar structure. As such, one
-potential problem with using J+24's candidate members is that the
-stellar density model is either a single or double exponential. If the
-(double) exponential model does not adequately describe a dwarf galaxy,
-then J+24's derived density profiles may be systematically biased
-towards the assumed (double) exponential profile.
-
-As an example, in the lower panels of Fig.~\ref{fig:scl_density_extras};
-Fig.~\ref{fig:umi_density_extras}, the density profile derived from the
-fiducial (J+24) sample appears to be well-constrained at densities far
-below the CMD+PM or \texttt{simple} model's background density. In
-addition, the fiducial and 1-component density profiles deviate, despite
-being derived identically besides the spatial likelihood assumption.
-
-These stars are likely selecting stars from the statistical MW
-background consistent either galaxy's CMD and PMs, recovering the
-assumed density profile. As a result, the reliability of these density
-profiles below the CMD+PM background may be questionable. A more robust
-analysis, removing this particular density assumption, would be required
-to more appropriately represent the knowledge of the density profile as
-the background begins to dominate.
-
-We revisit possible biases in spatial likelihood in
-Section~\ref{sec:mcmc_hists}.
-
-\textbf{Completeness}. \emph{Gaia} shows high but imperfect
-completeness, particularly showing limitations in crowded fields and for
-faint sources (\(G\gtrapprox20\)). In \citet{fabricius+2021}, for the
-lowest density globular clusters, the completeness down to
-\(G\approx 20\) is \(\sim 80\%\). As the lowest density globular
-clusters are still far denser than dwarf galaxies, so the completeness
-is likely higher for classical dwarfs. \emph{Gaia} is likely more
-incomplete in terms of BP-RP photometry (as we use here) and for stars
-with small pairwise separations (\(\lesssim 1.4''\)). The influence of
-these details is unclear. Nevertheless, \emph{Gaia} should be more
-complete for brighter stars.
-
-We find little evidence for magnitude-dependent biases in the density
-profiles. In
-Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we show
-density profiles derived using only stars brighter than the median
-magnitude. We find no substantive differences compared to our fiducial
-density profile. If \emph{Gaia}'s incompleteness is inhomogeneous, then
-these variations do not appear to affect the density profiles.
-
-If \emph{Gaia} has systematic density biases, the density profile should
-vary among different facilities. We show the density profiles derived
-from deep, ground-based UNIONS and DEVLE photometry (lower panels of
-Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). For
-both Scl and UMi, these alternative surveys result in near-identical
-density profiles. Variations between density profiles derived from these
-surveys appear to also be with uncertainties.
+\textbf{Spatial likelihood.} As J+24 include a spatial likelihood, our
+density profiles risks ``double fitting''---the spatial distribution of
+stars informs the membership catalogue from which density profiles are
+derived. If the assumed spatial likelihood is misrepresentative, the
+derived density profiles may bias towards the assumed likelihood.
+Fig.~\ref{fig:scl_umi_density_extras} illustrates that selecting the
+one- or two-component spatial models affects the outer extent of the
+dwarf galaxy. Promisingly, the density excess persists when assuming a
+one-component exponential profile. We revisit this concern using
+spatially independent methods below.
 
 \textbf{Structural uncertainties}. The assumed structural parameters of
-a dwarf may be misrepresentative or vary with radius. J+24 derive
-probabilities where the outer stellar component is circular instead of
-elliptical. Using the circular 2-component probabilities and binning in
-circular radius, we find little difference in the derived density
-profiles (the \texttt{circ} points in
-Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). Thus,
-our conclusions are robust to extreme variations in the ellipticity.
-Reasonable changes in other structural parameters (centre, position
-angle) are unlikely to have a more substantial effect.
+a dwarf (centre, position angle, ellipticity, scale radius) may be
+biased or vary radially. As a test, we use a membership sample J+24
+derive assuming the outer component is circular. Even when binning in
+circular radius, the density profiles remain unchanged (the
+\texttt{circ} points in Fig.~\ref{fig:scl_umi_density_extras}). Thus,
+our conclusions appear to be robust to an extreme shift in ellipticity.
+Reasonable changes in other structural parameters are unlikely to have a
+stronger effects.
+
+\textbf{Completeness}. \emph{Gaia} appears to have high but imperfect
+completeness, more limited for crowded fields, faint sources
+(\(G\gtrsim20\)), and BP-RP magnitudes. In \citet{fabricius+2021}, the
+completeness down to \(G\approx 20\) is \(\sim 80\%\) for low-density
+globular clusters. As dwarfs are even less dense, their completeness is
+likely higher. To test for magnitude-dependent biases, we show density
+profiles derived for only the brighter half of member stars in
+Fig.~\ref{fig:scl_umi_density_extras}. We find no substantive change
+compared to our fiducial density profile. If \emph{Gaia}'s
+incompleteness is inhomogeneous, then these variations likely do not
+affect our density profiles.
 
@@
 double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
-one-component exponential spatial likelihood, and \texttt{simple} uses
-the simple position-independent selection criteria \texttt{circ} is a
-2-component bayesian model assuming circular radii, simple is the series
-of simple cuts described, bright is the sample of the brightest half of
-stars (scaled by 2).}\label{fig:scl_density_extras}
+one-component exponential spatial likelihood, \texttt{simple} uses the
+simple position-independent selection criteria, \texttt{circ} is a
+2-component bayesian model assuming circular radii, \texttt{bright} only
+includes stars brighter than the median magnitude, and \texttt{DELVE} or
+\texttt{UNIONS} use photometry from external surveys with background
+subtraction. In all cases, the density profiles are nearly identical
+until the background-limited regime (grey shaded
+region).}\label{fig:scl_umi_density_extras}
+\end{figure}
+
+\section{Comparison against alternate
+samples}\label{sec:simple_selection}
+
+Next, to compare against simpler methods, we create samples using
+absolute filters in the CMD, PM space. We also compare against similarly
+selected samples from deep photometry-only data.
+
+For the \texttt{simple} \emph{Gaia} samples, we require high-quality
+astrometry (\texttt{ruwe} \textless{} 1.3), parallaxes
+\(3\sigma\)-consistent with the dwarf's distance, proper motions with
+\(1\masyr\) radius of the galaxy's value, and within the dwarf's
+empirical CMD (Fig.~\ref{fig:extra_cmd}).
+
+We also determine density profiles from the deeper photometric surveys.
+For Scl, we use DELVE DR2 survey \citep{drlica-wagner+2022}. We select
+sources within an ellipse of radius 150 arcminutes, categorized as
+likely stars, with reliable \(g\) and \(r\) magnitudes (associated flags
+\(\leq 4\)), and within Scl's observed CMD (Fig.~\ref{fig:extra_cmd}).
+Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
+Survey \citep[UNIONS,][]{gwyn+2025}. We select sources within an ellipse
+of radius 230 arminutes, with no \texttt{FLAGS\_CFIS} set, the
+\texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended sources),
+and within the CMD cut in Fig.~\ref{fig:extra_cmd}.
+Fig.~\ref{fig:delve_unions_tangent} shows the resulting spatial
+distribution for both samples. The only remarkable structure is Muñoz 1
+nearby to Ursa Minor.
+
+Finally, Fig.~\ref{fig:scl_umi_density_extras} compares the density
+profiles from each sample. All samples agree until the ``limiting
+radius'', where the \texttt{simple} and DELVE/UNIONS samples reach their
+respective backgrounds. The very outskirts of J+24's density profiles
+may also extend below the background of apparent member stars,
+complicating the reliability of the outer density profiles.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
+\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
+cuts used in samples in this section. The likely-member stars in the
+inner \(1R_h\) of each dwarf (black points) are used as a guide to
+determine the CMD cuts (orange polygons). Light grey points instead show
+the full distribution of stars across the field.}\label{fig:extra_cmd}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
+\caption[DELVE and UNIONS spatial distribution of stars]{The
+distribution of DELVE and UNIONS selected stars in Scl and UMi as grey
+points overdrawn with isophotes. The clump to the lower right of UMi is
+Muñoz 1.}\label{fig:delve_unions_tangent}
 \end{figure}
@@
 To address the concerns discussed above, we consider here a
-non-parametric model to fit the density in each bin.
+non-parametric model to fit the density in each bin. We demonstrate that
+the non-parametric model fails to find evidence for stellar density
+where J+24 still detects members. Notably, our density profiles diverge
+substantially from J+24 for Antlia II. We suggest these differences
+arise from background contamination, motivating our ``limiting radii''
+representing where densities may become unreliable.
+
+\subsection{Methodology}\label{methodology}
+
+As a non-parametric but similar model to J+24, we consider a piecewise
+constant spatial likelihood. In this model, the stars are divided into
+radial bins which are then considered independently. Each bin has only
+one free parameter, the fraction of satellite stars in that bin,
+\(f_{\rm sat}\). This model thus directly derives the density profile
+from the data in a single step. If a bin contains insufficient
+information to estimate a precise satellite density, than the posterior
+\(f_{\rm sat}\) should reflect this uncertainty.
+
+We parameterize \(f_{\rm sat}\) in terms of a log-relative density,
+\(\theta\) \begin{equation}{
+\begin{split}
+\theta &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
+f_{\rm sat} &= \frac{10^{\theta}}{1 + 10^{\theta}}.
+\end{split}
+}\end{equation} We then adopt a broad uniform prior on \(\theta\) from
+-12 to 6. The CMD and PM likelihoods are unchanged from J+24, except the
+systemic PM is fixed for efficiency.
+
+To bin stars, we hold fixed J+24's structural parameters and create bins
+of the wider of the width \(\Delta \log R=0.05\) or containing the next
+20 stars. We then derive posterior \(f_{\rm sat}\) distributions with
+MCMC (48 walkers of 1000 steps each, and using the No-U-Turn sampler as
+implemented in Turing.jl). The final density profile is directly derived
+from \(f_{\rm sat}\) and the number of stars in each bin.
+
+\subsection{Results}\label{results}
 
@@
 \centering
-\caption[Properties of probabilistically-derived density profiles]{For each classical dwarf, we have: the BG-limited radius $R_{\rm limit}$, where the density of stars is no longer reliably derived, the approximate number of member stars in Gaia, and the derived Sérsic indices from the density profiles.
+\caption[The limiting radii of Gaia-derived density profiles]{For each classical dwarf, the limiting radius $R_{\rm limit}$ in units of $R_h$ and arcminutes.
+$R_{\rm limit}$ represents where there no longer appears to be evidence of stars in Emph(Str(Gaia)) using the nonparametric MCMC density profiles.
 }
 \label{tbl:mcmc_props}
-\begin{tabular}{lllll}
+\begin{tabular}{lll}
 \toprule
-Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$ & num stars & $n$\\
+Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$\\
 \midrule
-Fornax & 5.25 & 79.1 & 23154 & $0.794^{+0.034}_{-0.018}$\\
-Sculptor & 6.39 & 64.1 & 7024 & $1.281^{+0.04}_{-0.04}$\\
-Leo I & 4.23 & 13.5 & 1242 & $0.719^{+0.059}_{-0.053}$\\
-Ursa Minor & 6.42 & 86.4 & 2314 & $1.425^{+0.10}_{-0.093}$\\
-Leo II & 3.63 & 8.76 & 347 & $0.707^{+0.11}_{-0.092}$\\
-Carina & 4.16 & 33.3 & 2389 & $1.025^{+0.08}_{-0.076}$\\
-Draco & 3.59 & 27.3 & 1781 & $1.106^{+0.084}_{-0.078}$\\
-Canes Venatici I & 1.95 & 12.5 & 156 & $1.18^{+0.36}_{-0.28}$\\
-Sextans I & 3.42 & 67.9 & 1830 & $1.136^{+0.10}_{-0.09}$\\
-Crater II & 1.93 & 39.0 & 507 & $0.535^{+0.12}_{-0.09}$\\
+Fornax & 5.25 & 79.1\\
+Sculptor & 6.39 & 64.1\\
+Leo I & 4.23 & 13.5\\
+Ursa Minor & 6.42 & 86.4\\
+Leo II & 3.63 & 8.76\\
+Carina & 4.16 & 33.3\\
+Draco & 3.59 & 27.3\\
+Canes Venatici I & 1.95 & 12.5\\
+Sextans I & 3.42 & 67.9\\
+Crater II & 1.93 & 39.0\\
 \bottomrule
@@
 
-\subsection{Methodology}\label{methodology}
+Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
+density profiles as compared to J+24. In general, both methodologies are
+consistent. However, J+24 tend to systematically overestimate faint
+densities and confidently derive densities where the MCMC model fails to
+derive a density estimate.
 
-This model extends the J+24 framework with two notable differences. (1)
-The systemic proper motions are held fixed. This allows for a much more
-efficient calculation of the likelihoods. (2) The spatial likelihood for
-the satellite is a piecewise constant density function.While this model
-theoretically has many more parameters to fit (one for each bin), each
-parameter is independently estimated based on only the stars in that
-specific bin.
+In the outskirts of satellites, more background / foreground stars may
+have consistent PM and CMD properties than satellite members. Improperly
+estimating the satellite's density in this ``background-limited'' regime
+likely affects the inferred density of members. If the satellite's
+density is severely overestimated, then a J+24-like sample may select
+many additional background stars. The density of candidates would then
+be biased towards the assumed density. As J+24 only use a one or
+two-component density profile across the entire dwarf, the density
+profile is likely biased where the CMD+PM background dominates. The
+nonparametric MCMC model instead does not assume a local density, so
+should better represent the underlying satellite density.
 
-We segregate the data into radial bins and independently fit the model
-to only stars in each bin. This fits many small models instead of a
-global large model. In detail, this means that, in each bin, the
-normalized density of the background and satellite are \emph{equal}. The
-only fit parameter is the \(f_{\rm sat}\) in the particular bin.
-Therefore, the likelihood in each bin is the same as above except
-\({\cal L}_{\rm space} = 1\) for the foreground and background.
-
-In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
-satellite density. The model construction is then \begin{equation}{
-\begin{align}
-\theta_i &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
-f_{\rm sat} &= \frac{10^{\theta_i}}{1 + 10^{\theta_i}} \\
-\theta_i &\sim {\rm Uniform}(-12, 6) \\
-{\cal L}_{\rm tot} &= f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM} \\
-\end{align}
-}\end{equation}
-
-For each satellite, we hold the structural parameters fixed to those in
-J+24, create bins which are the larger of the bin containing 20 stars or
-0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
-for 48 walkers using Turing.jl and the No-U Turns Sampler.
-
-Based on the \(f_{\rm sat}\) values in each bin, the number of members
-is just the number of stars in the bin times \(f_{\rm sat}\) for the
-bin, allowing the direct derivation of the density profiles from this
-model.
-
-We have also briefly explored a model which solves for \(f_{\rm sat}\)
-for random realizations of the global structural uncertainties. These do
-not appear to influence the results substantially. Another possible
-inconsistency is that the half-light radii are estimated from a
-different sample. We do briefly rederive structural parameters for the
-sample using Sérsic fits (although with a fixed origin).
-
-\subsection{Results}\label{results}
-
-Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
-histogram density profiles compared to J+24. In general, both
-methodologies agree well. However, J+24 tend to systematically
-overestimate faint densities and confidently derive densities where the
-MCMC model fails. Because J+24 use only one or two components across the
-entire dwarf galaxy, the faint regions become sensitive to the spatial
-likelihood. Antlia II is an extreme case, where the satellite is hidden
-behind a large number of foreground stars. We derive a much more
-poorly-constrained and fainter total density profile than J+24, and the
-divergence between the MCMC method here and J+24's profile likely shows
-that \emph{Gaia} observations are insufficient to properly constrain
-Antia II's density profile. This is likely due to uncertainties in the
-background density of satellite-like stars in Antlia II.
-
-Once the true density of stars drops below the background of
-satellite-like stars, J+24's method will occasionally select consistent
-background stars with a density directly dependent on the spatial
-likelihood. The uncertainties in the outer regions are likely
-underestimated.
+Antlia II represents an extreme example---J+24 and the non-parametric
+density model systematically disagree across the entire galaxy.
+Additionally, the piecewise model derives densities with larger
+uncertainties and over a smaller range than J+24. The methodological
+divergence here likely arises due to an extraordinary
+background/foreground of MW stars. \emph{Gaia} data alone may be unable
+to properly constrain the density of such background-contaminated
+objects.
 
@@
 ``background-limited'' radii based on the outermost derived density in
-the MCMC histogram with an uncertainty lower than 1 dex. This typically
-corresponds to the empirical background from the CMD+PM samples in the
-main text.
+the MCMC non-parametric model with an uncertainty lower than 1 dex. This
+closely corresponds to the background density from the ``CMD+PM''
+samples in the main text (see Fig.~\ref{fig:scl_observed_profiles}).
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/mcmc_histograms.png}
 \caption[Probabilistic density profiles]{A comparison between the MCMC
 histogram method and J+24. The MCMC samples in each bin are black
-transparent dots, and the J+24 derived density profiles (with their
-\(P_{\rm sat} < 0.2\)) are orange solid dots.}\label{fig:mcmc_hists}
+transparent dots (with added jitter), and the J+24 derived density
+profiles (i.e., \(P_{\rm sat} < 0.2\)) are orange solid
+dots.}\label{fig:mcmc_hists}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms2.png}}
-\caption[Probabilistic density profiles continued]{A continuation of
-Fig.~\ref{fig:mcmc_hists}.}\label{fig:mcmc_hists2}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/mcmc_histograms2.png}
+\caption[Probabilistic density profiles
+continued]{Fig.~\ref{fig:mcmc_hists} continued.}\label{fig:mcmc_hists2}
 \end{figure}
@@
 
-We finally compare our density profiles against a sample of
+Finally, we compare our density profiles against a sample of
 literature-derived density profiles in
-Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. The
-J+24-derived density profiles are representative of state-of-the-art
-density profiles. A few photometric surveys may extend to slightly
-fainter densities, but typically with large uncertanties. Deviations
-between sources are slight despite the range of methods across time. All
-density profiles extending into the outskirts of Scl and UMi show a
-similar overdensity to the ones we find. The extended stellar profiles
-of Scl and UMi appear to be a robust result across the literature.
+Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. Deviations
+between different profiles are slight, despite the range of methods
+across time, and all density profiles extending into the outskirts of
+Scl and UMi show a similar overdensity to the ones we find. The extended
+stellar profiles of Scl and UMi appear to be a robust result across the
+literature.
 
@@
 to Scl's half-light radius, and the residuals in the bottom panel are
-with respect to this profile. \textbf{remove delve, umi title\ldots,,
-combine??}}\label{fig:scl_lit_profiles}
+with respect to this profile. References are (in order),
+\citet{munoz+2018}; \citet{westfall+2006}; \citet{walcher+2003};
+\citet{eskridge1988}; \citet{demers+krautter+kunkel1980}; and
+\citet{hodge1961}.}\label{fig:scl_lit_profiles}
 \end{figure}
@@
 Fig.~\ref{fig:scl_lit_profiles} except for Ursa Minor. The references
-are (in order) \citet{sato+2025}; \citet{palma+2003};
-\citet{martinez-delgado+2001}; \citet{kleyna+1998}; \citet{IH1995}; and
-\citet{Hodge1964}.}
+are (in order) \citet[derived using their minor axis
+profile]{sato+2025}; \citet{palma+2003}; \citet{martinez-delgado+2001};
+\citet{kleyna+1998}; \citet{IH1995}; and
+\citet{Hodge1964}.}\label{fig:umi_lit_profiles}
 \end{figure}
@@
 major biases in the derived density profiles. However, we show that
-J+24's density profiles become unreliable when dropping below the
-background of satellite-like stars. We address this concern through an
-algorithm rederiving the density in elliptical bins. We find excellent
-agreement with J+24 for most galaxies out to a background-limited
-radius. Finally, comparing our density profiles against the literature,
-we find our results to be consistent. We conclude that the detection of
-an extended density profile in Scl and UMi is robust to incompleteness,
-methodology, alternative surveys, and across the literature.
+J+24's density profiles may become unreliable when dropping below the
+background of satellite-like stars. Finally, comparing our density
+profiles against the literature, we find our results to be consistent.
+We conclude that the detection of an extended density profile in Scl and
+UMi is robust to incompleteness, methodology, alternative surveys, and
+across the literature.

### 2025-10-06 thesis/extra_numerical_convergence.tex
full draft of appendicies, improved formatting

@@
 
-The two other key numerical parameters in \gadget{} are the tree-force
-and time-step accuracy parameters. Specifically, \gadget{} opens a node
-if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
-the particle is \(r\), side-length is \(l\), and the particle's total
-acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
-adaptive time stepping with integration accuracy set to \(\eta=0.01\).
-(particles must take time-steps smaller than
-\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).
-
 The top panel of Fig.~\ref{fig:methods_convergence} illustrates the
@@
 particles) across most radii. On the other hand, the fiducial and
-smaller softening lengths predict similar final results, although the
-smaller softening length simulation deviates slightly more than the
-fiducial, likely a result of more collisional encounters. As computation
-time increases moderately with decreasing softening length, the fiducial
-softening length balances efficiency and accuracy for this simulation.
-While smaller than \citet{power+2003}'s suggestion, our choice is
-designed for more compact, dwarf satellite halos instead of large,
-galaxy-mass isolated halos.
+smaller softening lengths predict similar final results. As computation
+time increases with decreasing softening length, the fiducial softening
+length balances efficiency and accuracy for this simulation.
 
-The lower panel of Fig.~\ref{fig:methods_convergence} tests changes to
-these parameters. We include a model with smaller integration accuracy
-(``small timestep'') and a stricter tree-force tolerance (``high acc.
-force''). More precise tolerances on these parameters does not affect
-the evolution within the uncertainties of the final profile.
+The two other key numerical parameters in \gadget{} are the tree-force
+and time-step accuracy parameters. Specifically, \gadget{} opens a node
+if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
+the particle is \(r\), side-length is \(l\), and the particle's total
+acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
+adaptive time stepping with integration accuracy set to \(\eta=0.01\).
+(particles must take time-steps smaller than
+\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)). The
+lower panel of Fig.~\ref{fig:methods_convergence} tests changes to these
+parameters. We include a model with smaller integration accuracy
+(\(\eta=0.003\) in ``small timestep'') and a stricter tree-force
+tolerance (\(\alpha=0.001\), ``high acc. force''). More precise
+tolerances on these parameters does not affect the evolution within the
+uncertainties of the final profile.
 
@@
 all other models use \(10^5\) paritcles, with their ``converged radius''
-marked by the black arrow and softening by the vertical bar.
+marked by the black arrow and softening by the vertical bar(s).
 \textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
-lengths. \textbf{Bottom:} Models with more precise timestep accuracy
-(\(\eta=0.001\) compared to fiducial \(\eta=0.005\)) and gravitational
-force accuracy (\(\alpha=0.003\) versus \(\alpha=0.01\)).
-\textbf{Summary:} Except for the model with a larger softening length,
-all simulations agree within uncertainties. \textbf{Todo: softening
-along lower axis always\ldots, changing
-linewidth?}}\label{fig:methods_convergence}
+lengths than the fiducial. \textbf{Bottom:} Models with more precise
+timestep accuracy and gravitational force accuracy. \textbf{Summary:}
+Except for the model with a larger softening length, all simulations
+agree within uncertainties.}\label{fig:methods_convergence}
 \end{figure}

### 2025-10-06 thesis/extra_results.tex
full draft of appendicies, improved formatting

@@
 concentration, cored halos, alternative orbits, anisotropy, and
-ellipsoidal models. While each of these variables influences resulting
-dark matter evolution, the final stellar component and dark matter
-remnant remains largely across alternate initial conditions. We conclude
-that consideration of these effects likely would not change our
-conclusions.
-
-The following models aim to reproduce the velocity dispersion
-\(\lesssim 1\,\kms\) with a similar present-day half-light radius.
+ellipsoidal models. While each of these variables influences total dark
+matter evolution, the resulting inner structure is similar across
+models. We conclude that consideration of these effects likely would not
+change our conclusions.
 
@@
 conditions}\label{alternative-initial-conditions}
+
+The following models aim to reproduce the stellar velocity dispersion of
+Scl to within \(\lesssim 1\,\kms\) with a similar present-day half-light
+radius.
 
@@
 
+The heavier halo diverges from the initial point orbit more
+substantially. We show the results of correcting the orbit (akin to Ursa
+Minor's correction, see Section~\ref{sec:orbit_corrections}) in
+Fig.~\ref{fig:tidal_tracks_concentration} as the ``heavier, new orbit''
+halo. Since correcting the orbit has a relatively small effect on the
+tidal evolution, we neglect these corrections in further comparisons for
+simplicity.
+
 \begin{figure}
@@
 evolution of different N-body models for different halo concentrations
-(heavier and lighter halo), and the mean orbit instead of the
-\smallperi{} orbit. In the left, the maximum circular velocity \(\vmax\)
-of the dark matter halo is plotted as a function of simulation time. In
-the right, \(\vmax\) is instead plotted as a function of
+(heavier and lighter halo), and the heavy halo with the action-angle
+corrected orbit. In the left, the maximum circular velocity \(\vmax\) of
+the dark matter halo is plotted as a function of simulation time. In the
+right, \(\vmax\) is instead plotted as a function of
 \(\rmax\).}\label{fig:tidal_tracks_concentration}
 \end{figure}
-
-\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
-
-Figure: The evolution of velocity anisotropy.
 
@@
 }\end{equation} where \(r_c\) is the core radius. We set
-\(M_s=0.79\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
-\(r_s=2.76\,\kpc\), \(r_c=0.1r_s\), and use the same truncation as our
-fiducial halo (Eq.~\ref{eq:trunc_nfw}). While this is not the largest
-possible core, the core is similar in size to the stellar component.
-Larger cores would require a more massive halo in order to match the
-velocity dispersion at a constant scale radius.
+\(M_s=0.54\times10^{10}\,\Mo\) where \(M_s = 4\pi/3\ r_s^3 \rho_s\),
+\(r_s=1.08\,\kpc\), \(r_c=r_s\), and use the same truncation as our
+fiducial halo (Eq.~\ref{eq:trunc_nfw}). We plot the initial and final
+density, as compared to a cuspy NFW, in Fig.~\ref{fig:cored_i_f}.
 
@@
 expected. However, in order to reproduce the observed velocity
-dispersion, the final halo reaches a \(\rmax\) and \(\vmax\) as the
-fiducial halo. A cored dark matter halo likely cannot be completely
-tidally disrupted and still recover the observed velocity dispersion.
+dispersion, the final halo maintains a higher \(\vmax\) than the
+fiducial halo.
 
+\begin{figure}
+\centering
 \pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
+\caption[Cored initial and final density profile]{The initial (dotted
+orange) and final (solid orange) 3D density profiles for the cored model
+of Scl on the \smallperi{} orbit. The blue thin line represents an NFW
+halo with the same \(M_s\) and \(r_s\).}\label{fig:cored_i_f}
+\end{figure}
 
-Figure: Cored versus cuspy initial and final density (todo, fix)..
+\subsection{Velocity Anisotropy}\label{velocity-anisotropy}
+
+Radial velocity anisotropy may cause halos to disrupt faster
+\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
+velocity anisotropy, we initialize a model with a velocity anistropy
+with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
+to \(\beta=1\) and infinity, with scale length \(4r_s\).
+
+Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
+profiles. The dwarf galaxy becomes more isotropic with tidal evolution.
+Particles on more radially-anisotropic orbits are more easily stripped
+as they have larger apocentres than more circular orbits of the
+equivalent energy. Regardless, the overall tidal evolution is very
+similar to our fiducial, isotropic case (see
+Fig.~\ref{fig:tidal_tracks_structure}).
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
+\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
+profiles for the initially-anisotropic model of Scl on the \smallperi{}
+orbit. The final profile is far more isotropic (\(\beta=0\)) than the
+initial.}\label{fig:anisotropy_i_f}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
+\caption[Tidal dependence on halo structure]{Similar to
+Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
+including a core, velocity anisotropy, and evolving an oblate halo.
+While the normalization may differ, the tidal evolution is
+similar.}\label{fig:tidal_tracks_structure}
+\end{figure}
 
@@
 Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
-violating our assumption of spherical symmetry
-\citep[e.g.,][]{an+koposov2022}.
+violating the assumption of spherical symmetry
+\citep[e.g.,][]{an+koposov2022}. The shape of the underlying dark matter
+halo is unknown, however.
 
-We create an oblate initial dark matter halo using \agama{}. The initial
-snapshot is sampled from the distribution function \begin{equation}{
+We create an oblate initial dark matter halo using \agama{}.
+Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
+isolation) isodensity contours of our model. The initial snapshot is
+sampled from the distribution function \begin{equation}{
 \begin{split}
@@
 \(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
-\texttt{example\_doublepowerlaw.py} script in \Agama{} solves for the
+\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
 best parameters matching a given density profile. We chose to create a
@@
 We find the evolution of the oblate halo to be nearly identical to the
-spherical halo, similar to \citet{battaglia+sollima+nipoti2015}. The
-oblate halo however becomes spherical as time progresses. If Scl's dark
-matter halo is indeed elliptical, this model may not be an adequate
-description. Possibly, internal rotation, alternative density structure,
-or velocity anisotropy play a role in the stability of oblate initial
-conditions to tidal stripping. A full investigation of these effects is
-left to future work.
-
-\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_density_i_f.pdf}}
-
-Slices of the oblate density before and after tidal evolution.
-\textbf{TODO: isodensity contours on RHS projected major/minor axis}.
+spherical halo (see Fig.~\ref{fig:tidal_tracks_structure}), similar to
+\citet{battaglia+sollima+nipoti2015}. The oblate halo however becomes
+spherical after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's
+dark matter halo is indeed elliptical today, this model may not be an
+adequate description.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
-\caption[Tidal dependence on halo structure]{Similar to
-Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
-including a core, velocity anisotropy, and evolving an oblate halo.
-While the normalization may differ, the tidal evolution is
-similar.}\label{fig:tidal_tracks_structure}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
+\caption[Oblate halo before and after tidal evolution]{The initial and
+final (after 9Gyr of tidal evolution) projected density profiles for the
+oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
+\(z'\) the major and minor axes. The contours are drawn assuming normal
+smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
+dex.}\label{fig:oblate_i_f}
 \end{figure}
@@
 
-In the main text, we focus on recent tidal impacts on Scl and UMi. But,
-due to the long-term orbital uncertainties of Scl and UMi with an LMC
-(see Fig.~\ref{fig:scl_umi_orbit_uncert}), tidally extreme orbits are
-entirely possible. Here, we consider a few examples of extreme orbital
-evolution in the long term case. While the actual orbit of either galaxy
-almost certainly deviates from these examples (discussion in
-Section~\ref{sec:orbital_uncertainties}), these orbits still represent
-extreme past tidal histories.
-
-For a more massive LMC (e.g., the L3M11 model), Scl may have undergone
-an extreme pericentric passage with the MW \(\sim 6\,\Gyr\) ago. We find
-that the 3\(\sigma\) smallest pericentre is \(4\,\kpc\). To test if this
-model forms a different density profile, we simulate this model.
+As discussed in Section~\ref{sec:scl_umi_orbit_uncert}, the long-term
+orbit of Scl is uncertain. For a more massive LMC (e.g., the L3M11
+model), Scl may have undergone an extreme pericentric passage with the
+MW \(\sim 6\,\Gyr\) ago. We find that the 3\(\sigma\) smallest
+pericentre is \(4\,\kpc\), and simulate this model to test if such a
+pericentre may be sufficient.
 
@@
 conditions by comparing the change in actions (as calculated in the
-static MW-only potential) before and after the pericentre.
-Table~\ref{tbl:orbit_adjustments} records the iterations we take. Our
-final model is able to approximately reproduce the observed position of
-Scl, but more refinement is required for a closer match.
+static MW-only potential) before and after the pericentre. Our final
+model, the \texttt{MW\ impact} model, is able to approximately reproduce
+the observed position of Scl.
+
+Fig.~\ref{fig:tidal_tracks_umi} compares the tidal evolution of Scl on
+the mean, \texttt{MW\ impact} , and \smallperi{} orbit. The mean orbit
+evloves less down the tidal track than the \smallperi{} model. Instead,
+the \texttt{MW\ impact} orbit experiences most tidal evolution during
+its first MW-pericentre. While evolving further along the tidal track,
+the stars of this model nevertheless remain exponential
+(Fig.~\ref{fig:scl_mw_impact_i_f}). We suggest that the impulsive
+pericentric passage does not occur for long enough in this model to
+produce the expected extended density profile. A yet more extreme
+orbital history would be necessary to tidally transform Scl's stars.
 
@@
 
-\section{Present-day properties}\label{present-day-properties}
-
-We find that this model also does not display strong signs of tidal
-signatures. Fig.~\ref{fig:scl_mw_impact_i_f} shows the initial and final
-stellar profiles for this model (integrated with a lower-resolution
-\(10^6\) particles). The model evolves little more than the
-\smallperi{}. Given that the Jacobi radius would be within \(R_h\), why
-does this model not show strong density perturbations?
-
-We propose that th case Scl undergoes a small MW pericentre is an
-impulsive encounter. Because Scl's apocentre is still \(\sim 300\,\kpc\)
-in this model, the galaxy passes through the MW with velocity
-\(600\,\kms?\). Since the crossing time at the half-light radius is
-about 90 Myr, in one crossing-time, the model moves from pericentre to a
-Galactocentric distance of \(\sim20\,\kpc\), where tides forced have
-dropped substantially. As a result, after \(6\Gyr\) of relatively
-isolated evolution, the break radius falls well outside the observed
-density profile, but the injection of energy uniformly throughout the
-halo only results in moderate DM loss predominantly in the outskirts,
-not a reshaping of the inner density profile. Note that the Jacobi
-radius is derived on circular orbits, not highly elliptical orbits like
-this case.
-
-However, we only note that the stars appear to be consistent but are
-well below the \citet{power+2003}'s numerically converged radius.
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_impact_i_f.pdf}}
+\caption[Sculptor MW-impact density profiles]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f} except for the orbit of Scl passing
+through the MW.}\label{fig:scl_mw_impact_i_f}
+\end{figure}
 
@@
 
-As we predict that any tidal tails formed in Scl and UMi would be
-extremely faint, we deferred discussion of the properties of any such
-tails to this section. We note that in extreme cases, \(\sim 2\%\) of
-stars may become unbound in either galaxy, forming tidal tails which may
-be detectible with substantially better observations.
+As discussed in Section~\ref{sec:results}, any hints of a possible tidal
+stream around Scl and UMi are beyond the reach of current observational
+facilities. However, we can still predict the properties of such a
+stream.
+
+Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream}
+shows the resulting distributions of velocities and distance along the
+stream orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
+detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
+motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
+such a gradient would require tracing stars across several degrees on
+the sky.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
+\caption[Sculptor predicted stream]{The predicted properties of a tidal
+tail in the Scl model. The panels are all as a function of \(\xi'\), the
+distance along the stream as defined by the current GSR proper motion
+vector. The top panels show the GSR proper motions in RA and Dec, and
+the bottom two show the distance and GSR radial velocities. To sample
+the stream, we randomly draw 100,000 samples from the snapshot based on
+the stellar weights. A detectible gradient in \(\mu_{\alpha*}\) and LOS
+velocity should be detectible if the stream is tracked across several
+degrees.}\label{fig:scl_tidal_stream}
+\end{figure}
 
@@
 \caption[Ursa Minor predicted stream]{The properties of the stream
-around the UMi \smallperi{} orbit with Plummer stars. The panels are all
-as a function of \(\xi'\), the distance along the stream as defined by
-the current GSR proper motion vector. The top panels show the GSR proper
-motions in RA and Dec, and the bottom two show the distance and GSR
-radial velocities. To sample the stream, we randomly draw 100,000
-samples from the snapshot based on the stellar weights. A detectible
-gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
-the stream is tracked across several
-degrees.}\label{fig:umi_tidal_stream}
-\end{figure}
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_stream.pdf}}
-\caption[Scl predicted stream]{The predicted properties of a tidal tail
-in the Scl model. \textbf{TODO: titles}}\label{fig:scl_tidal_stream}
+around the UMi \smallperi{} orbit with Plummer
+stars.}\label{fig:umi_tidal_stream}
 \end{figure}

### 2025-10-06 thesis/extra_rv_models.tex
full draft of appendicies, improved formatting

@@
 Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
-\citet{sestito+2023a}; and \apogee{} \citep[DR17,][]{abdurrouf+2022}.
-For Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
-\citet{sestito+2023b}; and \apogee{}. We then crossmatch all catalogues
-to J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we
-match to the nearest star within 1--3 arcseconds. We combine
-measurements of the same star using inverse-variance weighting.
-
-To reduce likely binaries, we remove stars with significant velocity
+\citet{sestito+2023a}; and APOGEE \citep[DR17,][]{abdurrouf+2022}. For
+Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
+\citet{sestito+2023b}; and APOGEE. We then crossmatch all catalogues to
+J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we match
+to the nearest star within 1--3 arcseconds. We combine measurements of
+the same star using inverse-variance weighting. To reduce likely
+binaries, we remove stars with significant velocity
 dispersions.\footnote{Specifically, using that
@@
 total likelihood for the velocity consistency (see
-Section~\ref{sec:extra_density}). We assume that the satellite and
+Section~\ref{sec:the_algorithm}). We assume that the satellite and
 background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
@@
 (\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
-a systematic velocity offset. Otherwise all studies appear to agree.
+a systematic velocity offset. Otherwise, all studies appear to be on a
+similar velocity scale.
 
@@
 centre), and then transform velocities to correct for the apparent
-gradient induced by the dwarf's proper motion.\footnote{Specifically,
-  sub} We define \(v_{\rm gsr}'\) to be velocities in the The correction
-from both effects induces an apparent gradient of about \(1.3\,\kmsdeg\)
-for Sculptor and less for Ursa Minor \citep[see
-also][]{WMO2008, strigari2010}. The uncertainty on this velocity
-correction are less than the individual star uncertainties and our
-derived velocity gradient.
+gradient induced by the dwarf's proper motion \citep[see
+also][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
+velocities in the GSR frame subtracting the PM-induced gradient. The
+correction from both effects induces an apparent gradient of about
+\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor. The uncertainty
+on this velocity correction are less than the individual star
+uncertainties and any derived velocity gradients.
 
@@
 We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
-velocity, velocity dispersion, and possible gradient. We assume that the
-galaxy follows a planar velocity gradient. The mean velocity at a given
-point on the sky, \(\mu\), is assumed to be,
+velocity, velocity dispersion, and possible gradients in velocities or
+velocity dispersions. We assume that the galaxy follows a planar
+velocity gradient. The mean velocity at a given point on the sky,
+\(\mu\), is assumed to be,
 
 \begin{equation}{
-\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
+\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta,
 }\end{equation}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption[LOS velocity fit to Scl]{Velocity histogram of Scl and UMi in
-terms of projected-correction GSR LOS velocity (Eq.~\ref{eq:v_z}). Black
+\caption[LOS velocity fit to Scl]{\textbf{Left}: Velocity histogram of
+Scl and UMi in terms of projected-correction GSR LOS velocity. Black
 points with error bars are from the crossmatched observed sample, and
-green lines represent MCMC fits to the velocity dispersion.}
+green lines represent MCMC fits to the velocity dispersion.
+\textbf{Right} The observed velocity dispersion gradient (black points)
+in 10 equal number bins and the derived slopes for MCMC samples (green
+lines). Scl shows moderate evidence for an increasing velocity
+dispersion with radius, but Scl's velocity dispersion appears to flatten
+outside of \(R_h\). UMi's evidence for a radial gradient is
+weaker.}\label{fig:rv_hists}
 \end{figure}
-
-Figure: The observed velocity dispersion gradient (black) in 10 equal
-number bins and the derived slopes from the model fitting. Scl shows
-moderate evidence for an increasing velocity dispersion with radius.
-UMi's evidence for a radial gradient is weaker, and Scl's velocity
-dispersion appears to be flat outside of \(R_h\).
 
@@
 We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
-velocity dispersion \(9.64\pm0.16\,\kms\). Our values are very
-consistent with previous work \citep[e.g.][\citet{arroyo-polonio+2024},
+velocity dispersion \(9.64\pm0.16\,\kms\) (see Fig.~\ref{fig:rv_hists}).
+Our values are very consistent with previous work
+\citep[e.g.][\citet{arroyo-polonio+2024},
 \citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
@@
 We derive a mean \(-245.8\pm0.3_{\rm stat}\,\kms\) and velocity
-dispersion of \(8.8\pm0.2\,\kms\) for UMi. We do not find evidence for a
-velocity gradient. This is consistent with previous work
-\citetext{\citealp{pace+2020}; \citealp[somewhat
+dispersion of \(8.8\pm0.2\,\kms\) for UMi (see Fig.~\ref{fig:rv_hists}).
+We do not find evidence for a velocity gradient. This is consistent with
+previous work \citetext{\citealp{pace+2020}; \citealp[somewhat
 with][]{spencer+2018}; \citealp{martinez-garcia+2023}}.

### 2025-10-06 thesis/introduction.tex
full draft of appendicies, improved formatting

@@
 or similarly \(M_\star \lesssim 10^9 M_\odot\)
-\citep[e.g.,][]{hodge1971, mcconnachie2012}. Because the galaxy
-luminosity function increases towards fainter objects, dwarfs are the
-most numerous galaxies in the Universe
+\citep[e.g.,][]{mcconnachie2012, bullock+boylan-kolchin2017}. Because
+the galaxy luminosity function increases towards fainter objects, dwarfs
+are the most numerous galaxies in the Universe
 \citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
@@
 important role in our understanding of the clustering of dark matter on
-small scales \citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.
+small scales \citep[e.g.,][]{bullock+boylan-kolchin2017, sales+2022}.
 
@@
   inconsistent with Newtonian dynamics of visible matter alone,
-  \citet{willman+strader2012}} In contrast, star clusters (like globular
-clusters) have no clear evidence for dark matter. The boundary between
-these two classes blurs for faint, compact stellar associations. Systems
-with shared characteristics of globular clusters and dwarf galaxies, and
-without evidence as to their nature, are known as ``ambiguous'' systems
-\citep[e.g.,][]{smith+2024}.
+  \citet{willman+strader2012}.} In contrast, star clusters (like
+globular clusters) have no clear evidence for dark matter. The boundary
+between these two classes blurs for faint, compact stellar associations.
+Systems with shared characteristics of globular clusters and dwarf
+galaxies, and without evidence as to their nature, are known as
+``ambiguous'' systems \citep[e.g.,][]{smith+2024}.
 
@@
 an example of an irregular, bright dwarf galaxy, where most stars are in
-a rotationally-supported thin disk (seen nearly face-on) with a
-prominent bar. \textbf{Classical dwarf spheroidals}\footnote{While
-  formally the dwarf galaxy names we discuss contain ``dwarf
-  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
-  brevity. Additionally, the 12 classical dwarf satellites of our Galaxy
-  are (in order of decreasing luminosity) Sagittarius, Fornax, Leo I,
-  Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
-  I, Sextans I, and Crater II. Antlia II, Crater II, and Canes Venatici
-  I are the only post digital sky surveys additions.} occupy
-intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
+a rotationally-supported disk (seen nearly face-on) with a prominent
+bar. \textbf{Classical dwarf spheroidals}\footnote{While formally the
+  dwarf galaxy names we discuss contain ``dwarf spheroidal'' (dSph),
+  e.g.~Sculptor dSph, we omit this suffix for brevity. Additionally, the
+  12 classical dwarf satellites of our Galaxy are (in order of
+  decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor, Antlia
+  II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I,
+  and Crater II. Antlia II, Crater II, and Canes Venatici I are the only
+  post digital sky surveys additions.} occupy intermediate luminosities
+( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
 \(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
@@
 \centering
-\includegraphics[width=5.41667in,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/mw_satellites_1.jpg}
 \caption[The on-sky distribution of Milky Way satellites]{The location
@@
 universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
-matter (DM), and 5\% regular old baryons\footnote{\emph{Baryons} here
-  means baryons+leptons, i.e.~any standard model massive fermion.}
+matter (DM), and 5\% regular old baryons\footnote{In a classic,
+  astronomer's corruption of jargon, \emph{Baryons} here means
+  baryons+leptons, i.e.~any standard model massive fermion.}
 \citep{planckcollaboration+2020}. While the composition of dark matter
@@
 \centering
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
+\includegraphics[width=3.5in,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
 \caption[Example dark matter and stellar density profiles]{Density
@@
 \citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
-Several open questions remain, concerning e.g.~the plane of satellites,
-the details of the sizes and rotation curves of dwarf galaxies, and the
-existence and nature of stellar halos and accretion in dwarf galaxies
-\citep[e.g.,][]{sales+2022}. Altogether, the numerous past and ongoing
-challenges for \LCDM{} in the dwarf galaxy regime illustrate the
-opportunity for dwarf galaxies to test the understanding of galaxy
-formation and dark matter physics.
+Several open questions remain, concerning, e.g., the plane of
+satellites, the details of the sizes and rotation curves of dwarf
+galaxies, and the existence and nature of stellar halos and accretion in
+dwarf galaxies \citep[e.g.,][]{sales+2022}. Altogether, the numerous
+past and ongoing challenges for \LCDM{} in the dwarf galaxy regime
+illustrate the opportunity for dwarf galaxies to test the understanding
+of galaxy formation and dark matter physics.
 
@@
 McConnachie and Venn (2020a), (4) Pace et al.
-(2020), average of MMT and Keck results with systematic uncertainty from Section \ref{sec:extra_rv_models} discussion.
+(2020), average of MMT and Keck results with systematic uncertainty from Appendix \ref{sec:extra_rv_models} discussion.
 }
@@
 become unbound for a galaxy in a circular orbit around a host
-galaxy.\footnote{Possibly first calculated before \citet{laplace1798},
-  this radius also bears other names such as the Hill radius
-  \citep[from][]{hill1878}. Likely only names after Jacobi for the
-  Jacobi integral \citep{jacobi1836}.} Calculated from an approximation
-of the location of the \(L_1\) and \(L_2\) Lagrange points, the Jacobi
-radius is where the mean density of the dwarf galaxy is roughly three
-times the mean interior density of the host galaxy at pericentre, or
+galaxy.\footnote{The Jacobi radius was derived at least as early as
+  \citet{laplace1798}. This radius also bears other names, such as the
+  Hill radius \citep[from][]{hill1878}. Likely only named after Jacobi
+  for the Jacobi integral \citep{jacobi1836}.} Calculated from an
+approximation of the location of the \(L_1\) and \(L_2\) Lagrange
+points, the Jacobi radius is where the mean density of the dwarf galaxy
+is roughly three times the mean interior density of the host galaxy at
+pericentre, or
 \begin{equation}\protect\phantomsection\label{eq:r_jacobi}{

### 2025-10-06 thesis/main.tex
a compltet thesis draft
thesis edits
full draft of appendicies, improved formatting

@@
 \title{
-    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?
+    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?
 }
@@
 \addcontentsline{toc}{chapter}{Acknowledgements}
+I promise I will fill this out thoroughly.
 
@@
 %Friends and family and collegues (fill in later).
-
-This work has made use of data from the European Space Agency (ESA) mission
-{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
-Data Processing and Analysis Consortium (DPAC,
-\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
-has been provided by national institutions, in particular the institutions
-participating in the {\it Gaia} Multilateral Agreement. 
-\citep{gaiacollaboration+2016, gaiacollaboration+2023}.
-
-This research made use of hips2fits,\footnote{https://alasky.cds.unistra.fr/hips-image-services/hips2fits} a service provided by CDS.
-
-The Digitized Sky Surveys were produced at the Space Telescope Science Institute under U.S.
-Government grant NAG W-2166.
-The images of these surveys are based on photographic data obtained using the Oschin Schmidt Telescope on Palomar Mountain and the UK Schmidt Telescope.
-The plates were processed into the present compressed digital form with the permission of these institutions.
-
-The National Geographic Society --- Palomar Observatory Sky Atlas (POSS-I) was made by the California Institute of Technology with grants from the National Geographic Society.
-
-The Second Palomar Observatory Sky Survey (POSS-II) was made by the California Institute of Technology with funds from the National Science Foundation, the National Geographic Society, the Sloan Foundation, the Samuel Oschin Foundation, and the Eastman Kodak Corporation.
-
-The Oschin Schmidt Telescope is operated by the California Institute of Technology and Palomar Observatory.
-
-The UK Schmidt Telescope was operated by the Royal Observatory Edinburgh, with funding from the UK Science and Engineering Research Council (later the UK Particle Physics and Astronomy Research Council), until 1988 June, and thereafter by the Anglo-Australian Observatory.
-The blue plates of the southern Sky Atlas and its Equatorial Extension (together known as the SERC-J), as well as the Equatorial Red (ER), and the Second Epoch [red] Survey (SES) were all taken with the UK Schmidt.
-
-All data are subject to the copyright given in the copyright summary. Copyright information specific to individual plates is provided in the downloaded FITS headers.
-
-Supplemental funding for sky-survey work at the ST ScI is provided by the European Southern Observatory.
 
@@
 
-%\chapter{The effects of tides on the Ursa Minor Dwarf Spheriodal Galaxy}
-%\input{ursa_minor}
-
-\chapter{Discussion}\label{sec:discussion}
+\chapter{Discussion \& Conclusions}\label{sec:discussion}
 \input{discussion}
 
-\chapter{Summary and Outlook}\label{sec:summary}
+\chapter*{Research Acknowledgements}
+\addcontentsline{toc}{chapter}{Research Acknowledgements}
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement. 
+\citep{gaiacollaboration+2016, gaiacollaboration+2023}.
+
+This research made use of hips2fits,\footnote{https://alasky.cds.unistra.fr/hips-image-services/hips2fits} a service provided by CDS.
+
+The Digitized Sky Surveys were produced at the Space Telescope Science Institute under U.S.
+Government grant NAG W-2166.
+The images of these surveys are based on photographic data obtained using the Oschin Schmidt Telescope on Palomar Mountain and the UK Schmidt Telescope.
+The plates were processed into the present compressed digital form with the permission of these institutions.
+
+The National Geographic Society --- Palomar Observatory Sky Atlas (POSS-I) was made by the California Institute of Technology with grants from the National Geographic Society.
+
+The Second Palomar Observatory Sky Survey (POSS-II) was made by the California Institute of Technology with funds from the National Science Foundation, the National Geographic Society, the Sloan Foundation, the Samuel Oschin Foundation, and the Eastman Kodak Corporation.
+
+The Oschin Schmidt Telescope is operated by the California Institute of Technology and Palomar Observatory.
+
+The UK Schmidt Telescope was operated by the Royal Observatory Edinburgh, with funding from the UK Science and Engineering Research Council (later the UK Particle Physics and Astronomy Research Council), until 1988 June, and thereafter by the Anglo-Australian Observatory.
+The blue plates of the southern Sky Atlas and its Equatorial Extension (together known as the SERC-J), as well as the Equatorial Red (ER), and the Second Epoch [red] Survey (SES) were all taken with the UK Schmidt.
+
+All data are subject to the copyright given in the copyright summary. Copyright information specific to individual plates is provided in the downloaded FITS headers.
+
+Supplemental funding for sky-survey work at the ST ScI is provided by the European Southern Observatory.
 
@@
 
-\input{appendix}
-\input{rv_models}
-\input{numerical_convergence}
+\input{extra_density}
+\input{extra_rv_models}
+\input{extra_numerical_convergence}
+\input{extra_results}
 

### 2025-10-08 thesis/abstract.tex
grammar/spelling edits

@@
 Clouds, are all dwarf spheroidals (dSphs)---gas-free, non-rotating,
-dark-matter dominated stellar systems. The light profiles of dSphs
+dark-matter-dominated stellar systems. The light profiles of dSphs
 typically follow an exponential law, with the outer density sharply
-declining . Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
+declining. Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
 dSphs host member stars out to \(\sim10\) effective radii, indicating
@@
 \citet{jensen+2024}. To evaluate if tides can produce these density
-features, we conduct idealized N-body simulation of both galaxies in the
-tidal field of the Milky Way. We find that neither dwarf experiences
+features, we conduct idealized N-body simulations of both galaxies in
+the tidal field of the Milky Way. We find that neither dwarf experiences
 tides sufficient to affect their stellar distribution. One complication
@@
 tides of the LMC and MW are unable to explain Scl's outer profile. We
-conclude that the extended light profiles of Scl and UMi not of tidal
-origin. They are instead likely innate, possibly explained by past
+conclude that the extended light profiles of Scl and UMi are not of
+tidal origin. They are instead likely innate, possibly explained by past
 mergers, accretion events, or episodic star formation.

### 2025-10-08 thesis/acknowledgments.tex
grammar/spelling edits

While written by my hand, I cannot pretend that I alone created this
work. I am ever thankful and greatly indebted to the multitude of people
who have supported, inspired, and nurtured me as a human being.

To my supervisor, Julio Navarro, for being the critical eye and guiding
reason to this work. I will always approach research differently now.

To Raphaël, who has helped inspire and direct this project. To Isabel
and Ryan.

To Jax, who has been a support along the way, guiding my research
methods and project. Akshara and Ewould.

To all of my new friends I have made along the way. Ruxin, Aviv, Anya,
Pranav, Ted, Adrianne, Will, Maeve, and everyone else I have met. We
have build a wonderful community of students in this department. And to
my friends across UVic, Wyatt, Rosemary, Maggie,

To my past advisors, James who has continued to support me long after
our paths diverged, David and Jennifer for guiding my first research
project.

To Wayne for showing us the joy of astronomy and outreach. To my mentor,
Ebtihal.

To my past roommates, Matthew, Josh, and Bryn, for putting up with my
chaos and our shared time together.

To Fierce Brosnen and our trivia team. Melody, Lily, Gillian, Stephen,
Jillian, and so many more.

To my long-time friends. Nikola and Eric. Mikey and Anthony. Anya, Kaia,
Alyssa, Aahliyah, Cameron. May we continue to find each other even as we
drift apart.

To my musical friends, Adrienne, Alex for our shared joy in music
making. And Steve, for showing me the joy in a simple life and beautiful
music. To my fellow knitters. Anne, Anya, Leise, Wes. To the
international community.

To my parents for their continuous support throghout this project. To my
grandparents for always being there. To my cousins and my extended
family. And to my siblings for always being a source of joy.

To the stars and the moon. To the land of Victoria, and its historical
caretakers. To the waters and beautiful trees, the ocean waves, rocks
and hills and jagged coastlines. To the rich, enveloping forests and
rust-red arbutus tress, scraggly Garry oaks and Camus flowers. I could
not imagine a more beautiful place to study and to always remind my
place in the world.

### 2025-10-08 thesis/data.tex
grammar/spelling edits

@@
 
-Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
+Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
 probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
@@
 member stars out to large distances from a dwarf galaxy. J+24 extended
-the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
-additionally including a secondary, extended spatial component. J+24
-detected candidate members out to \textasciitilde10 half-light radii
-from the centres of some galaxies (\(R_h\)). Similar recent work has
-also included \citet{pace+li2019}; \citet{battaglia+2022};
+the algorithm presented in \citet{MV2020a, MV2020b} by additionally
+including a secondary, extended spatial component. J+24 detected
+candidate members out to \textasciitilde10 half-light radii from the
+centres of some galaxies (\(R_h\)). Similar recent work has also
+included \citet{pace+li2019}; \citet{battaglia+2022};
 \citet{pace+erkal+li2022}; \citet{qi+2022}.
@@
 }\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
-coordinate rotated to align with a dwarf galaxy's major and minor axis,
-and \(a\) and \(b\) are the semi-major and semi-minor axis of the
+coordinates rotated to align with a dwarf galaxy's major and minor axes,
+and \(a\) and \(b\) are the semi-major and semi-minor axes of the
 galaxy.
@@
 show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
-Fornax), as well as impact of different samples, in tangent-plane
+Fornax), as well as the impact of different samples, in tangent-plane
 coordinates, the CMD, and proper-motion space. The ``all'' sample
@@
 fiducial sample appears similar to the ``CMD+PM'' sample in the CMD and
-PM planes, but excludes improbably distant stars. The ``RV member''
+PM planes, but excludes improbably distant stars. The ``RV members''
 sample also traces out similar distributions in CMD and PM space as the
@@
 magnitudes of stars with spectroscopic follow-up. Each selection
-criteria---spatial, CMD, and PM---contributes towards a high-quality
+criterion---spatial, CMD, and PM---contributes towards a high-quality
 membership assignment.
@@
 and Ursa Minor show an extended population of members past \(6R_h\) in
-radius Fornax does not despite having many more stars. Fornax also has a
-bluer CMD, indicative of more recent star formation. Otherwise, there
+radius, Fornax does not, despite having many more stars. Fornax also has
+a bluer CMD, indicative of more recent star formation. Otherwise, there
 are no major morphological differences between these galaxies.
@@
 \caption[Sculptor sample selection]{The distributions of various samples
-of \emph{Gaia} stars for Sculptor. In each panel, we plot light grey
+of \emph{Gaia} stars for Sculptor. In each panel, we use light grey
 points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
@@
 \citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
-Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
+Tangent plane \(\xi, \eta\). The orange ellipses represent 3 and 6
 half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
@@
 \caption[Fornax sample selection]{Similar to
-Fig.~\ref{fig:scl_selection} except for Fornax. The isocrone is instead
+Fig.~\ref{fig:scl_selection} except for Fornax. The isochrone is instead
 for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
 Fornax does not show the same extended outer halo of probable members as
-Sculptor or Ursa Minor despite having many more
-stars.}\label{fig:fornax_selection}
+Sculptor or Ursa Minor.}\label{fig:fornax_selection}
 \end{figure}
@@
 
-We derive density profiles by binning member stars in constant width
+We derive density profiles by binning member stars in constant-width
 bins in \(\log R_{\rm ell}\) of 0.05 dex. We ignore bins interior or
@@
 
-Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
+Fig.~\ref{fig:scl_observed_profiles} shows the derived density profiles
 for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
@@
 arcminutes, depending on the galaxy. By restricting stellar membership
-with CMD + PM, the background is reduced by 1-2 dex. The ``CMD+PM''
-background plateau represents the density of background stars which
-could be mistaken as members because of their coincident colours and
-proper motions. Finally, the ``all -- background'' profile results from
-subtracting the apparent background in the ``all'' profile.
+with CMD and PM information, the background is reduced by 1--2 dex. The
+``CMD+PM'' background plateau represents the density of background stars
+that could be mistaken as members because of their coincident colours
+and proper motions. Finally, the ``all -- background'' profile results
+from subtracting the apparent background in the ``all'' profile.
 
@@
 corresponding to an elliptical radius of 37 arcminutes. Muñoz 1 does not
-have a bright RGB, so has few stars above the \emph{Gaia} magnitude
-limit. The cluster has little effect on the elliptically-averaged
-density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).
+have a bright RGB, so the cluster has few stars above the \emph{Gaia}
+magnitude limit. The cluster has little effect on the
+elliptically-averaged density profile (see location on
+Fig.~\ref{fig:scl_observed_profiles}).
 
@@
 selections using absolute cuts, and the effect of structural assumptions
-of the density profile (Appendix \ref{sec:extra_density}). These
+of the density profile in Appendix \ref{sec:extra_density}. These
 methodologies are consistent except when the derived density drops below
@@
 do not find evidence of magnitude biases in the density profiles,
-indicating inhomogeneous magnitude-completeness is small. Our density
-profiles furthermore agree with photometric surveys and similar
+indicating that inhomogeneous magnitude-completeness is small. Our
+density profiles furthermore agree with photometric surveys and similar
 literature (see Appendix \ref{sec:extra_density}). We conclude that
@@
   effective depth between galaxies. However, the deviations between Scl,
-  UMi and other galaxies is apparent even when data covers all dwarfs.}
-In the outer regions, however, Sculptor and Ursa Minor deviate, and show
-a clear outer excess over an exponential law (solid black line). These
-galaxies are better fit by a Plummer law (dashed black line). The
-deviation from an exponential grows outwards, and at \(\sim 8 R_h\), may
-reach 2 orders of magnitude. The remainder of this thesis will be
-devoted to assessing whether the outer excess shown by Scl and UMi are
-due to Galactic tides.
+  UMi, and other galaxies are apparent even where the data is complete
+  across all dwarfs.} In the outer regions, however, Sculptor and Ursa
+Minor deviate and show a clear outer excess over an exponential law
+(solid black line). These galaxies are better fit by a Plummer law
+(dashed black line). The deviation from an exponential grows outwards,
+and at \(\sim 8 R_h\), may reach 2 orders of magnitude. The remainder of
+this thesis will be devoted to assessing whether the outer excesses
+shown by Scl and UMi are due to Galactic tides.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_fnx_density_methods.pdf}}
-\caption[Density profiles for different Gaia samples]{The density
-profile of Sculptor, Ursa Minor, and Fornax for different selection
-criteria, ploted as log surface dencity versus log elliptical radius.
-The samples are: ``all'' selects any high quality and
-parallax-consistent star (grey circles),``all--background'' subtracts a
+\caption[Density profiles for different \textit{Gaia} samples]{The
+density profile of Sculptor, Ursa Minor, and Fornax for different
+selection criteria, plotted as log surface density versus log elliptical
+radius. The samples are: ``all'' selects any high-quality and
+parallax-consistent star (grey circles), ``all--background'' subtracts a
 uniform background density from the ``all'' profile (orange pentagons),
-``CMD+PM'' select stars according to CMD and PM only (turquoise
+``CMD+PM'' selects stars according to CMD and PM only (turquoise
 triangles), and ``fiducial'' also includes spatial information (blue
@@
 Sculptor, Ursa Minor, and other classical dwarfs compared to Exp2D and
-Plummer density profiles. Dwarf galaxies are scaled to the same
-half-light radius and density at half-light radius. Sculptor and Ursa
-Minor have an excess of stars in the outer regions (past
-\(\log R/R_h \sim 0.3\)) compared with other classical
+Plummer density profiles. Dwarf galaxies are scaled by the half-light
+radius and density at half-light radius. Sculptor and Ursa Minor have an
+excess of stars in the outer regions (past \(\log R/R_h \sim 0.3\))
+compared with other classical
 dwarfs.}\label{fig:classical_dwarfs_densities}

### 2025-10-08 thesis/discussion.tex
grammar/spelling edits

--- 2025-10-06 thesis/discussion.tex
@@
 have more extended density profiles compared to other MW classical
-satellites, and that recent tides are an unlikely explanation for such
-extended density profiles. We therefore suggest that Scl and UMi may
-host an additional, extended population of stars. With our data, other
-classical dwarfs lack evidence for similar extended populations. The
-absence of similar extended populations in other classical dwarfs
-suggests Scl and UMi have followed a divergent evolutionary path.
+satellites, and that recent tides are an unlikely explanation for their
+extended outer structure. We therefore suggest that Scl and UMi may host
+an additional, extended population of stars, such as a ``stellar halo''.
+With our data, other classical dwarfs lack evidence for similar extended
+populations. Thus, Scl and UMi have followed a divergent evolutionary
+path, motivating our discussion of extended density formation pathways
+here.
 
 In this chapter, we place our results in context with past work. We
-review both work considering tidal effects on Scl and UMi and detections
-of peculiarities. We then consider possible origins of extended stellar
-populations for these galaxies. Scenarios include multi-epoch star
-formation, past mergers depositing stars into the outskirts, and even
-departures from the typical exponential dwarf galaxy structure. We later
-discuss how future observations and simulations may distinguish halo
-formation scenarios. Finally, we summarize the main conclusions of this
-thesis.
+review both works, considering tidal effects on Scl and UMi and
+detections of peculiarities. We then consider possible origins of
+extended stellar populations for these galaxies. Scenarios include
+multi-epoch star formation, past mergers depositing stars into the
+outskirts, and even departures from \LCDM{}. We later discuss how future
+observations and simulations may distinguish among formation scenarios.
+Finally, we summarize the main conclusions of this thesis.
 
@@
 
-Tidal effects have been long invoked to explain features of dwarf
-galaxies e.g.~Section \ref{sec:scl_umi_obs_tides},
-\citep{mayer+2001a, tsujimoto+shigeyama2002} \textbf{Check this
-citation}. Yet, growing work suggests tides may not be as commonplace as
-suspected. For example, \citet{read+2006} suggest that the absence of
-rising radial velocity dispersion profiles in dwarfs implies a lack of
-tidal disruption. \citet{penarrubia+2009}, introducing the break radius
-as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
-orbits were inconsistent with observable tidal effects. Recently,
-\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
-(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the
-exception of the Sagittarius dSph, these results suggest tides have mild
-influence on the stellar components of classical dwarfs.
+Tidal effects have long been invoked to explain features of dwarf
+galaxies \citep[e.g.,
+Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}.
+Yet, growing work suggests tides may not be as commonplace as suspected.
+For example, \citet{read+2006} suggested that the absence of rising
+radial velocity dispersion profiles in dwarfs implies a lack of tidal
+disruption. \citet{penarrubia+2009}, introducing the break radius as a
+diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite orbits
+were inconsistent with observable tidal effects. Recently,
+\citet{pace+erkal+li2022} used a Jacobi radius-based criterion
+(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. Except for the
+Sagittarius dSph, these results suggest tides have a mild influence on
+the stellar components of classical dwarfs.
 
@@
 even for a dark-matter-free model. Most recently,
-\citet{tchiorniy+genina2025} also use idealized simulations tuned to
+\citet{tchiorniy+genina2025} also used idealized simulations tuned to
 five classical dwarfs with a focus on the inner density, concluding that
@@
 
-\subsection{Peculiarities in the
-galaxies}\label{peculiarities-in-the-galaxies}
+\subsection{Peculiarities in the galaxies}\label{sec:peculiarities}
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_fe_h_gradient.pdf}}
-\caption[Metallicity gradients in Scl and UMi]{Density profiles and
-metallicity gradients in Scl (left) and UMi (right). In the top panel,
-we plot the J+24 density profiles as function of elliptical radius in
-black, and the two components of a double-exponential density fit as
-curved, transparent lines. The vertical lines represent the transition
-from the inner to the outer exponential profile. \textbf{Bottom}:
-Metallicities as a function of elliptical radius for stars from
-\citet{tolstoy+2023}; \citet{abdurrouf+2022}; and \citet{sestito+2023a}
-(left) and \citet{pace+2020}; \citet{abdurrouf+2022}; and
-\citet{sestito+2023b} (right), crossmatched against J+24 as described in
-Appendix \ref{sec:extra_rv_models}.}\label{fig:metallicity_gradients}
+\caption[Metallicity gradients in Sculptor and Ursa Minor]{Density
+profiles and metallicity gradients in Scl (left) and UMi (right).
+\textbf{Top}: Black points show the J+24 density profiles as a function
+of elliptical radius, and transparent lines represent the two components
+of a double-exponential density fit. The vertical lines represent the
+transition from the inner to the outer exponential profile.
+\textbf{Bottom}: Metallicities as a function of elliptical radius for
+stars crossmatched against J+24 as described in Appendix
+\ref{sec:extra_rv_models}. Larger, red stars show the members from
+\citet{sestito+2023a, sestito+2023b}.}\label{fig:metallicity_gradients}
 \end{figure}
@@
 Besides extended stellar density profiles, Scl and UMi display other
-peculiarities which may hold clues to their formation. Scl and UMi both
-hosts at least two chemodynamic populations, as revealed through their
-photometric or metallicity-velocity structure
-\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner, compact
-population is younger, more metal-rich, and dynamically colder, whereas
-the outer, extended population older, more metal-poor, and dynamically
-hotter.
+peculiarities that may hold clues to their formation. Scl and UMi both
+host at least two chemodynamic populations, as revealed through their
+photometric or metallicity-velocity structure\footnote{Other examples of
+  galaxies with multiple populations include Carina
+  \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
+  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
+  Sextans
+  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
+  Andromeda II
+  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
+\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
+younger, higher metallicity, and dynamically colder, whereas the outer
+population is older, lower metallicity, and dynamically hotter.
 
@@
 
-Ursa Minor has also shown evidence for possible inner substructure such
+Ursa Minor has also shown evidence for possible inner substructure, such
 as stellar or kinematic ``clumps''
@@
 inner structure of the halo, disrupting quickly in halos with
-substructure or inner density cusps but survining longer in smooth,
+substructure or inner density cusps but surviving longer in smooth,
 cored halos \citep{kleyna+2003, lora+2012}.
@@
 arises from an unusual past. For example, a merger may naturally explain
-both extended and multiple stellar populations.
+both extended outer and multiple stellar populations.
 
@@
 As we disfavour a tidal origin for Scl and UMi's extended density
-profiles, we now consider six alternative processes which may form an
-extended stellar population in these dwarfs.
+profiles, we now consider six other processes that may form an extended
+stellar population in these dwarfs.
 
@@
 stellar component and population gradient
-\citep{benitez-llambay+2016, deason+2022}. If the galaxies contains gas,
+\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
 the merger can trigger new star formation, forming a younger population
@@
 \citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
-A galaxy having undergone a major merger should show evidence of at
-least two or three populations from distinct origins.
+A galaxy having undergone a major merger should harbour at least two or
+three populations from distinct origins.
 
@@
 even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
-examples are known, but a stream around And II may be an instance of
-dwarf accretion \citep{amorisco+evans+vandeven2014, roederer+2023}. In
-this scenario, the extended stellar populations would contain chemical
+examples of dwarf-dwarf accretion are known, but a possible stream
+around And II or chemical peculiarities around Sextans I may be such
+instances \citep{amorisco+evans+vandeven2014, roederer+2023}. In this
+scenario, the extended stellar populations would contain chemical
 signatures from several distinct ultra-faint dwarf galaxies.
@@
 \citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
-redistributes already-present stellar populations without creating new
-populations, and may mimic a stellar halo. Key evidence suggesting
-preprocessing may include distant dwarf stars or stellar streams.
+redistributes already-present stellar populations and may mimic a
+stellar halo. Key evidence suggestive of preprocessing may include
+distant dwarf stars or stellar streams.
 
-\textbf{Dark-matter free dwarfs.} Tidal dwarfs form in gas-rich tidal
+\textbf{Dark-matter-free dwarfs.} Tidal dwarfs form in gas-rich tidal
 streams created during the merger of two galaxies
@@
 Similarly, in Modified Newtonian Dynamics (MOND), galaxies do not
-contain dark matter. Instead the gravitational force law is altered to
+contain dark matter. Instead, the gravitational force law is altered to
 explain rotation curves. Dwarfs would similarly experience stronger
@@
 also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
-evidence of such features (see Appendix \ref{sec:extra_rv_models}), so
-future data would need to uncover signs of disruption to support this
-hypothesis.
+evidence of such features (see Appendix
+\ref{sec:extra_rv_models})---future data would need to uncover signs of
+disruption to support this hypothesis.
 
@@
 galaxies---extended stellar populations should be more common if these
-processes form substantial halos.
+processes are important.
 
@@
 
-We reviewed a number of different, possibly-concurrent explanations for
+We reviewed a number of different, possibly concurrent explanations for
 the formation of extended populations. While we leave the nature of Scl
@@
 addition, measurements of the ellipticity and anisotropy would help
-verify or alter our understanding of the mass structure and if DM-free
-models may be permissible.
+verify or alter our understanding of the mass structure and whether
+DM-free models may be permissible.
 
 \emph{Deep photometry} may find or rule out signs of dynamical
-disequilibrium and tidal tails for tidally susceptible models
-\emph{(e.g., pre-processing and DM-free dwarfs.)} In addition,
+disequilibrium and tidal tails for tidally susceptible models (e.g.,
+\emph{pre-processing} and \emph{DM-free dwarfs}). In addition,
 photometry will help constrain the prevalence and nature of extended
@@
 chemistry. Evidence of significant star formation episodes or lack
-thereof may differentiate scenarios which rely on strong star formation
-bursts (e.g.~\emph{episodic star formation history, induced star
-formation, and gas-rich mergers.})
+thereof may differentiate scenarios that rely on strong star formation
+bursts (e.g., \emph{episodic star formation history, induced star
+formation,} and \emph{gas-rich mergers}).
 
@@
 multi-object spectroscopy on large telescopes, such as 4MOST
-\citep{skuladottir+2023} or extremely large telescopes. Mapping the 3D
-kinematics across a dwarf may require a \emph{Gaia} successor, but
-ongoing surveys with JWST and HST can derive precise proper motions
-within select regions of dwarf galaxies, able to reconstruct internal
-structure and anisotropies \citep[e.g.,][]{vitral+2025}. Current
-wide-field photometric surveys by Rubin Observatory \citep{ivezic+2019}
-and the Euclid mission \citep{euclidcollaboration+2025} will trace even
-fainter features around dwarf galaxies, possibly finding halos, tidal
-streams, and unusual stellar populations. Together, these observational
-programs promise a new window into the detailed structure, evolution,
-and outskirts of dwarf galaxies.
+\citep{skuladottir+2023} or with the extremely large telescope
+\citep{jagourel+2018}. Mapping the 3D kinematics across a dwarf may
+require a \emph{Gaia} successor, but ongoing surveys with JWST and HST
+can derive precise proper motions within select regions of dwarf
+galaxies, able to reconstruct internal structure and anisotropies
+\citep[e.g.,][]{vitral+2025}. Current wide-field photometric surveys by
+Rubin Observatory \citep{ivezic+2019} and the Euclid mission
+\citep{euclidcollaboration+2025} will trace even fainter features around
+dwarf galaxies, possibly finding halos, tidal streams, and unusual
+stellar populations. Together, these observational programs promise a
+new window into the detailed structure, evolution, and outskirts of
+dwarf galaxies.
 
@@
 and star formation), and improving numerical resolution. With these
-continuing advances, future simulations may shed light on viability of
-halo formation scenarios and the details of dwarf galaxy assembly.
+continuing advances, future simulations may shed light on the viability
+of extended stellar population formation scenarios and the details of
+dwarf galaxy assembly.
 
@@
 
-We then investigated if tides are a possible explanation. By modelling
-each galaxy based on cosmological initial conditions, we showed that
-tides do not strongly affect either galaxy. The LMC changes the orbital
-history of Scl and UMi, but tides become even weaker in a combined LMC
-and MW potential. We conclude that recent tides are unlikely to shape
-the outer radial distribution of stars in Scl and UMi.
+We then investigated whether tides are a possible explanation. By
+modelling each galaxy based on cosmological initial conditions, we
+showed that tides do not strongly affect either galaxy. The LMC changes
+the orbital history of Scl and UMi, but tides become even weaker in a
+combined LMC and MW potential. We conclude that recent tides are
+unlikely to shape the outer radial distribution of stars in Scl and UMi.
 
@@
 and observational programs will be able to investigate the nature of
-dwarfs with ever increasing detail. Abundant opportunities await to
+dwarfs with ever-increasing detail. Abundant opportunities await to
 uncover the cosmic origins and history of dwarf galaxies, the building

### 2025-10-08 thesis/extra_density.tex
grammar/spelling edits

@@
-\chapter{The reliability of derived density
-profiles}\label{sec:extra_density}
+\chapter{\texorpdfstring{The Reliability of Projected Number Density
+Profiles Derived from
+\emph{Gaia}}{The Reliability of Projected Number Density Profiles Derived from Gaia}}\label{sec:extra_density}
 
@@
 non-parametric Bayesian density profile, which reproduces J+24's results
-in the inner regions but diverges when the background dominates. In
-summary, we find the density profiles are robust up to the
-background-limited radius derived here.
+in the inner regions but differs when the background dominates. We find
+the density profiles are robust up to the background-limited radius
+derived here.
 
@@
 within a 2--4 degree region around each satellite with high-quality
-astrometry, reliable photometry, and consistent parallaxes and broadly
+astrometry, reliable photometry, consistent parallaxes, and broadly
 consistent proper motions and colours. Stars are removed if they have
@@
 see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
-expectations \citep[from][]{riello+2021}, proper motions magnitudes
-\(>10\,\masyr\) in \(\mu_{\alpha*}\) or \(\mu_\delta\), magnitudes
-brighter than the tip of the red giant branch or fainter than \(G=22\),
-or colours outside \(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry
-is dereddened with \citet{schlegel+finkbeiner+davis1998} extinction
-maps.
+expectations \citep[from][]{riello+2021}, proper motions \(>10\,\masyr\)
+in \(|\mu_{\alpha*}|\) or \(|\mu_\delta|\), magnitudes brighter than the
+tip of the red giant branch\footnote{Derived from the maximum RGB
+  magnitude from the associated Padova isochrone plus
+  \(5\sigma\)\hspace{0pt} times the distance modulus uncertainty.} or
+fainter than \(G=22\), or colours outside
+\(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry is dereddened with
+\citet{schlegel+finkbeiner+davis1998} extinction maps.
 
@@
 
-The PM likelihood is a bivariate gaussian with variance and covariance
+The PM likelihood is a bivariate Gaussian with variance and covariance
 equal to each star's proper motions.
@@
 J+24 model the CMD as a Padova isochrone \citep{girardi+2002}, with
-matching metallicity, 12 Gyr age (or 2 Gyr for Fornax), and Gaussian
-width of 0.1 mag plus the \emph{Gaia} colour uncertainty at each
-magnitude. The horizontal branch is modelled as a
-constant-magnitude\footnote{Specifically, the mean magnitude of the 12
-  Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
+metallicity matching that obtained spectroscopically, 12 Gyr age (or 2
+Gyr for Fornax), and a Gaussian width of 0.1 mag plus the \emph{Gaia}
+colour uncertainty at each magnitude. The horizontal branch is modelled
+as a constant-magnitude\footnote{Specifically, the mean magnitude of the
+  12 Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
 sequence extending blue of the CMD with the same width as the RGB. The
@@
 
-The background likelihoods are determined empirically as a kernel
-density estimates from stars outside 5\(R_h\) in PM and CMD space. The
+The background likelihoods are determined empirically from kernel
+density estimates of stars outside 5\(R_h\) in PM and CMD space. The
 spatial background likelihood is uniform.
@@
 
-\section{\texorpdfstring{Possible biases in \emph{Gaia}-derived density
-profiles}{Possible biases in Gaia-derived density profiles}}\label{sec:density_extra}
+\section{Independent sample
+selection}\label{independent-sample-selection}
 
-In this section, we test how assumptions in J+24's algorithm or biases
-in \emph{Gaia} may affect density profiles. In all cases, density
-profiles converge until \(R_{\rm ell} \approx 60\,{\rm arcmin}\), where
-background contamination likely dominates.
+To test for possible limitations in the J+24 methodology (discussed
+below), we compare their density profiles against samples derived from
+absolute cuts in \emph{Gaia} and external photometric surveys.
 
-\textbf{Spatial likelihood.} As J+24 include a spatial likelihood, our
-density profiles risks ``double fitting''---the spatial distribution of
-stars informs the membership catalogue from which density profiles are
-derived. If the assumed spatial likelihood is misrepresentative, the
-derived density profiles may bias towards the assumed likelihood.
-Fig.~\ref{fig:scl_umi_density_extras} illustrates that selecting the
-one- or two-component spatial models affects the outer extent of the
-dwarf galaxy. Promisingly, the density excess persists when assuming a
-one-component exponential profile. We revisit this concern using
-spatially independent methods below.
-
-\textbf{Structural uncertainties}. The assumed structural parameters of
-a dwarf (centre, position angle, ellipticity, scale radius) may be
-biased or vary radially. As a test, we use a membership sample J+24
-derive assuming the outer component is circular. Even when binning in
-circular radius, the density profiles remain unchanged (the
-\texttt{circ} points in Fig.~\ref{fig:scl_umi_density_extras}). Thus,
-our conclusions appear to be robust to an extreme shift in ellipticity.
-Reasonable changes in other structural parameters are unlikely to have a
-stronger effects.
-
-\textbf{Completeness}. \emph{Gaia} appears to have high but imperfect
-completeness, more limited for crowded fields, faint sources
-(\(G\gtrsim20\)), and BP-RP magnitudes. In \citet{fabricius+2021}, the
-completeness down to \(G\approx 20\) is \(\sim 80\%\) for low-density
-globular clusters. As dwarfs are even less dense, their completeness is
-likely higher. To test for magnitude-dependent biases, we show density
-profiles derived for only the brighter half of member stars in
-Fig.~\ref{fig:scl_umi_density_extras}. We find no substantive change
-compared to our fiducial density profile. If \emph{Gaia}'s
-incompleteness is inhomogeneous, then these variations likely do not
-affect our density profiles.
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
-\caption[Sculptor density methodology comparison]{Density profiles for
-various assumptions for Sculptor. \texttt{2-exp} is the fiducial
-double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
-one-component exponential spatial likelihood, \texttt{simple} uses the
-simple position-independent selection criteria, \texttt{circ} is a
-2-component bayesian model assuming circular radii, \texttt{bright} only
-includes stars brighter than the median magnitude, and \texttt{DELVE} or
-\texttt{UNIONS} use photometry from external surveys with background
-subtraction. In all cases, the density profiles are nearly identical
-until the background-limited regime (grey shaded
-region).}\label{fig:scl_umi_density_extras}
-\end{figure}
-
-\section{Comparison against alternate
-samples}\label{sec:simple_selection}
-
-Next, to compare against simpler methods, we create samples using
-absolute filters in the CMD, PM space. We also compare against similarly
-selected samples from deep photometry-only data.
-
-For the \texttt{simple} \emph{Gaia} samples, we require high-quality
+The \texttt{simple} sample is selected from stars in \emph{Gaia} using
+absolute astrometric and quality cuts. We require high-quality
 astrometry (\texttt{ruwe} \textless{} 1.3), parallaxes
 \(3\sigma\)-consistent with the dwarf's distance, proper motions with
-\(1\masyr\) radius of the galaxy's value, and within the dwarf's
-empirical CMD (Fig.~\ref{fig:extra_cmd}).
+\(1\masyr\) of the systemic value, and colours within the dwarf's
+apparent CMD as determined from stars near the dwarf's centre (see
+Fig.~\ref{fig:extra_cmd}).
 
 We also determine density profiles from the deeper photometric surveys.
-For Scl, we use DELVE DR2 survey \citep{drlica-wagner+2022}. We select
-sources within an ellipse of radius 150 arcminutes, categorized as
-likely stars, with reliable \(g\) and \(r\) magnitudes (associated flags
-\(\leq 4\)), and within Scl's observed CMD (Fig.~\ref{fig:extra_cmd}).
-Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
+For Scl, we use data from the DECam Local Volume Exploration Survey
+(DELVE) DR2 survey \citep{drlica-wagner+2022}. We select sources within
+an ellipse of radius 150 arcminutes, categorized as likely stars, with
+reliable \(g\) and \(r\) magnitudes (associated flags \(\leq 4\)), and
+within Scl's CMD selection (orange polygon in Fig.~\ref{fig:extra_cmd}).
+Ursa Minor lies within the Ultraviolet Near-Infrared Optical Northern
 Survey \citep[UNIONS,][]{gwyn+2025}. We select sources within an ellipse
-of radius 230 arminutes, with no \texttt{FLAGS\_CFIS} set, the
-\texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended sources),
-and within the CMD cut in Fig.~\ref{fig:extra_cmd}.
-Fig.~\ref{fig:delve_unions_tangent} shows the resulting spatial
-distribution for both samples. The only remarkable structure is Muñoz 1
-nearby to Ursa Minor.
-
-Finally, Fig.~\ref{fig:scl_umi_density_extras} compares the density
-profiles from each sample. All samples agree until the ``limiting
-radius'', where the \texttt{simple} and DELVE/UNIONS samples reach their
-respective backgrounds. The very outskirts of J+24's density profiles
-may also extend below the background of apparent member stars,
-complicating the reliability of the outer density profiles.
+of radius 230 armin, with no \texttt{FLAGS\_CFIS} set, both \texttt{s21}
+and \texttt{s31} \(<3\) (i.e., not extended sources), and within the CMD
+polygon in Fig.~\ref{fig:extra_cmd}.
 
@@
 \caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
-cuts used in samples in this section. The likely-member stars in the
-inner \(1R_h\) of each dwarf (black points) are used as a guide to
-determine the CMD cuts (orange polygons). Light grey points instead show
-the full distribution of stars across the field.}\label{fig:extra_cmd}
+cuts used in samples in this section. Black points represent stars
+within \(1R_h\), light grey shows all field stars passing quality cuts,
+and the orange polygon shows the CMD cut. The central, likely-member
+stars inform the CMD selections.}\label{fig:extra_cmd}
 \end{figure}
+
+\section{\texorpdfstring{Possible biases in \emph{Gaia}-derived density
+profiles}{Possible biases in Gaia-derived density profiles}}\label{sec:density_extra}
+
+We now discuss potential biases in J+24's algorithm for \emph{Gaia}
+data. We then test how these biases may influence density profiles. In
+all cases, density profiles converge within
+\(R_{\rm ell} \approx 60\,{\rm arcmin}\), where background contamination
+begins to dominate.
+
+\textbf{Spatial likelihood.} Since J+24 include a spatial likelihood,
+the resulting density profiles risk ``double fitting''---the same
+spatial distribution of stars informs the membership catalogue and the
+subsequent density profiles. If the spatial likelihood misrepresents the
+underlying satellite structure, then the density profiles may be biased
+towards the assumed shape.
+
+\textbf{Structural uncertainties.} The assumed structural parameters of
+a dwarf (centre, position angle, ellipticity, scale radius) may be
+inaccurate or vary radially. This would change the elliptical radii of
+stars (\(R_{\rm ell}\)) and likewise the density profiles.
+
+\textbf{Completeness.} \emph{Gaia}'s completeness appears high but
+imperfect, more limited in crowded fields, for faint sources
+(\(G\gtrsim20\)), and in \(G_{\rm BP}\)-\(G_{\rm RP}\) magnitudes. In
+\citet{fabricius+2021}, the completeness down to \(G\approx 20\) is
+\(\sim 80\%\) for low-density globular clusters. As dwarfs are even less
+dense, their completeness is likely higher.
+
+The top row of Fig.~\ref{fig:scl_umi_density_extras} compares the
+density profiles assuming different spatial likelihoods: the
+\texttt{2-exp} and \texttt{1-exp} use two or one-component spatial
+likelihoods in J+24, and \texttt{simple} does not assume a spatial
+likelihood (described above). All profiles agree within the
+background-limited (BG-limited) radius, derived below. The
+\texttt{1-exp} profile predicts lower outer densities, yet the outer
+density excess in Sculptor and Ursa Minor persists. Thus, the spatial
+likelihood can influence the outer regions but has less effect on the
+inner regions.
+
+The lower panels of Fig.~\ref{fig:scl_umi_density_extras} test other
+possible biases. The \texttt{circ} profile uses circular bins and the
+J+24 sample, assuming a circular outer profile. As the profile is nearly
+identical to the fiducial model, an extreme change in ellipticity does
+not affect our conclusions. The \texttt{bright} profile is derived using
+only the brighter half of stars. We find no substantive change compared
+to our fiducial density profiles, showing completeness-related magnitude
+biases are likely small. Finally, the DELVE and UNIONS density profiles
+nearly identically trace the \texttt{2-exp} profile, but both become
+background-limited near the BG-limited radius.
+
+In all cases, there is little evidence for biases in \emph{Gaia}'s
+completeness or sensitivity to structural parameters. As the spatial
+likelihood appears to be the most severe bias, we next consider a more
+flexible model.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
-\caption[DELVE and UNIONS spatial distribution of stars]{The
-distribution of DELVE and UNIONS selected stars in Scl and UMi as grey
-points overdrawn with isophotes. The clump to the lower right of UMi is
-Muñoz 1.}\label{fig:delve_unions_tangent}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
+\caption[Density methodology comparison]{Density profiles for various
+assumptions for Sculptor (left) and Ursa Minor (right). Each profile is
+plotted as points in log surface density (scaled to the \texttt{2-exp}
+profile) versus log radius, the black line marks a 2D exponential
+profile, and the grey shaded region represents the background-limited
+region derived in
+Section~\ref{sec:mcmc_hists}.}\label{fig:scl_umi_density_extras}
 \end{figure}
@@
 non-parametric model to fit the density in each bin. We demonstrate that
-the non-parametric model fails to find evidence for stellar density
+the non-parametric model fails to find evidence for stellar density,
 where J+24 still detects members. Notably, our density profiles diverge
@@
 constant spatial likelihood. In this model, the stars are divided into
-radial bins which are then considered independently. Each bin has only
+radial bins, which are then considered independently. Each bin has only
 one free parameter, the fraction of satellite stars in that bin,
@@
 from the data in a single step. If a bin contains insufficient
-information to estimate a precise satellite density, than the posterior
+information to estimate a precise satellite density, then the posterior
 \(f_{\rm sat}\) should reflect this uncertainty.
@@
 }\end{equation} We then adopt a broad uniform prior on \(\theta\) from
--12 to 6. The CMD and PM likelihoods are unchanged from J+24, except the
-systemic PM is fixed for efficiency.
+-12 to 6. The CMD and PM likelihoods are unchanged from J+24, except
+that the systemic PM is fixed for efficiency.
 
@@
 \caption[The limiting radii of Gaia-derived density profiles]{For each classical dwarf, the limiting radius $R_{\rm limit}$ in units of $R_h$ and arcminutes.
-$R_{\rm limit}$ represents where there no longer appears to be evidence of stars in Emph(Str(Gaia)) using the nonparametric MCMC density profiles.
+$R_{\rm limit}$ represents where there is no longer evidence of Emph(Str(Gaia)) members using the nonparametric MCMC density profiles.
 }
@@
 
-Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
-density profiles as compared to J+24. In general, both methodologies are
-consistent. However, J+24 tend to systematically overestimate faint
-densities and confidently derive densities where the MCMC model fails to
-derive a density estimate.
+Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived MCMC
+density profiles as compared to J+24. In general, both methodologies
+produce similar densities. However, J+24 tend to systematically
+overestimate faint densities and confidently derive densities where the
+MCMC model fails to derive a density estimate.
 
-In the outskirts of satellites, more background / foreground stars may
+In the outskirts of satellites, more background/foreground stars may
 have consistent PM and CMD properties than satellite members. Improperly
@@
 profile is likely biased where the CMD+PM background dominates. The
-nonparametric MCMC model instead does not assume a local density, so
+nonparametric MCMC model instead does not assume a local density, so it
 should better represent the underlying satellite density.
@@
 Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. Deviations
-between different profiles are slight, despite the range of methods
+between different profiles are small, despite the range of methods
 across time, and all density profiles extending into the outskirts of
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
-\caption[Sculptor literature density profiles]{A comparison of the Scl
-derived density profile and historical works. Unlike most profiles in
-this thesis, this density profile is plotted with respect to the
+\caption[Sculptor literature density profiles]{A comparison of density
+profiles of Sculptor for works from the literature. Unlike most profiles
+in this thesis, this density profile is plotted with respect to the
 semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
-The solid black line is a 2D exponential with corresponding scale radius
-to Scl's half-light radius, and the residuals in the bottom panel are
-with respect to this profile. References are (in order),
+The solid black line is a 2D exponential with a corresponding scale
+radius to Scl's half-light radius, and the residuals in the bottom panel
+are with respect to this profile. References are (in order),
 \citet{munoz+2018}; \citet{westfall+2006}; \citet{walcher+2003};
@@
 assumed spatial likelihood, \emph{Gaia}'s completeness, and structural
-parameters. In all cases, we find these assumptions likely do not cause
-major biases in the derived density profiles. However, we show that
-J+24's density profiles may become unreliable when dropping below the
-background of satellite-like stars. Finally, comparing our density
+parameters. In all cases, we find that these assumptions likely do not
+cause major biases in the derived density profiles. However, we show
+that J+24's density profiles may become unreliable when dropping below
+the background of satellite-like stars. Finally, comparing our density
 profiles against the literature, we find our results to be consistent.

### 2025-10-08 thesis/extra_numerical_convergence.tex
grammar/spelling edits

@@
 number of particles within \(R_{200}\). This choice balances integration
-time and only compromises resolution in collisional regime. For our
+time and only compromises resolution in the collisional regime. For our
 isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
-particles, this works out to be \(0.044\,{\rm kpc}\). Next we consider
+particles, this works out to be \(0.044\,{\rm kpc}\). Next, we consider
 the effect of softening as applied to our isolation halo.
@@
 and time-step accuracy parameters. Specifically, \gadget{} opens a node
-if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
-the particle is \(r\), side-length is \(l\), and the particle's total
-acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
-adaptive time stepping with integration accuracy set to \(\eta=0.01\).
-(particles must take time-steps smaller than
+if \(M\,l/r^3 < \alpha |a|\), where the node's mass is \(M\), the
+distance from the particle is \(r\), the side-length is \(l\), and the
+particle's total acceleration is \(a\). We adopt \(\alpha =0.005\). We
+also elect to use adaptive time stepping with integration accuracy set
+to \(\eta=0.01\). (particles must take time-steps smaller than
 \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)). The
@@
 tolerance (\(\alpha=0.001\), ``high acc. force''). More precise
-tolerances on these parameters does not affect the evolution within the
+tolerances on these parameters do not affect the evolution within the
 uncertainties of the final profile.
@@
 model. The benchmark model is our fiducial \(10^7\) particle run, and
-all other models use \(10^5\) paritcles, with their ``converged radius''
+all other models use \(10^5\) particles, with their ``converged radius''
 marked by the black arrow and softening by the vertical bar(s).

### 2025-10-08 thesis/extra_results.tex
grammar/spelling edits

@@
-\chapter{Additional simulation tests}\label{sec:extra_results}
+\chapter{Additional Simulation Results}\label{sec:extra_results}
 
 In this section, we briefly explore additional simulations testing
-assumptions not varied in the main text: variations in the halo
-concentration, cored halos, alternative orbits, anisotropy, and
-ellipsoidal models. While each of these variables influences total dark
-matter evolution, the resulting inner structure is similar across
-models. We conclude that consideration of these effects likely would not
-change our conclusions.
+variations in the halo concentration, cored halos, alternative orbits,
+anisotropies, and ellipticities. While each of these variables
+influences total dark matter evolution, the resulting inner structure is
+similar across models. We conclude that consideration of these effects
+likely would not change our conclusions.
 
@@
 The following models aim to reproduce the stellar velocity dispersion of
-Scl to within \(\lesssim 1\,\kms\) with a similar present-day half-light
-radius.
+Sculptor to within \(\lesssim 1\,\kms\) with a similar present-day
+half-light radius.
 
@@
 (heavier and lighter halo), and the heavy halo with the action-angle
-corrected orbit. In the left, the maximum circular velocity \(\vmax\) of
-the dark matter halo is plotted as a function of simulation time. In the
-right, \(\vmax\) is instead plotted as a function of
+corrected orbit. \textbf{Left:} The maximum circular velocity \(\vmax\)
+of the dark matter halo is plotted as a function of simulation time.
+\textbf{Right:} \(\vmax\) is instead plotted as a function of
 \(\rmax\).}\label{fig:tidal_tracks_concentration}
@@
 In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
-cored model as compared to a heavier halo. The cored halo loses more
-total dark matter mass, evolving further along the tidal track as
-expected. However, in order to reproduce the observed velocity
-dispersion, the final halo maintains a higher \(\vmax\) than the
-fiducial halo.
+cored model as compared to a heavier halo. The cored halo appears to
+evolve more mildly than the NFW halo. This is possibly because, to match
+the velocity dispersion, the cored halo has \(\sim 50\%\) more mass
+within \(1\,\kpc\) than a similar cuspy halo.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
-\caption[Cored initial and final density profile]{The initial (dotted
+\caption[Tidal evolution of a cored density profile]{The initial (dotted
 orange) and final (solid orange) 3D density profiles for the cored model
@@
 
-\subsection{Velocity Anisotropy}\label{velocity-anisotropy}
-
-Radial velocity anisotropy may cause halos to disrupt faster
-\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
-velocity anisotropy, we initialize a model with a velocity anistropy
-with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
-to \(\beta=1\) and infinity, with scale length \(4r_s\).
-
-Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
-profiles. The dwarf galaxy becomes more isotropic with tidal evolution.
-Particles on more radially-anisotropic orbits are more easily stripped
-as they have larger apocentres than more circular orbits of the
-equivalent energy. Regardless, the overall tidal evolution is very
-similar to our fiducial, isotropic case (see
-Fig.~\ref{fig:tidal_tracks_structure}).
-
-\begin{figure}
-\centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
-\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
-profiles for the initially-anisotropic model of Scl on the \smallperi{}
-orbit. The final profile is far more isotropic (\(\beta=0\)) than the
-initial.}\label{fig:anisotropy_i_f}
-\end{figure}
-
 \begin{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
-\caption[Tidal dependence on halo structure]{Similar to
+\caption[Tidal tracks depending on halo substructure]{Similar to
 Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
@@
 
+\subsection{Velocity anisotropy}\label{velocity-anisotropy}
+
+Radial velocity anisotropy may cause halos to disrupt faster
+\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
+velocity anisotropy, we initialize a model with a velocity anisotropy
+with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
+to \(\beta=1\) and infinity, with scale length \(4r_s\).
+
+Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
+profiles after tidal evolution on Scl's \smallperi{} orbit. The dwarf
+galaxy becomes more isotropic with tidal evolution. Particles on more
+radially anisotropic orbits are more easily stripped as they have larger
+apocentres than more circular orbits of the equivalent energy.
+Regardless, the overall tidal evolution is very similar to our fiducial,
+isotropic case (see Fig.~\ref{fig:tidal_tracks_structure}).
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
+\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
+profiles for the initially anisotropic model of Scl on the \smallperi{}
+orbit. The final profile is far more isotropic (\(\beta=0\)) than the
+initial.}\label{fig:anisotropy_i_f}
+\end{figure}
+
 \subsection{An ellipsoidal halo}\label{an-ellipsoidal-halo}
@@
 violating the assumption of spherical symmetry
-\citep[e.g.,][]{an+koposov2022}. The shape of the underlying dark matter
-halo is unknown, however.
+\citep[e.g.,][]{an+koposov2022}. Although, the shape of the underlying
+dark matter halo is unknown.
 
@@
 best parameters matching a given density profile. We chose to create a
-model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
+model resembling an NFW but scaled by a factor of 0.5 in the \(z\)-axis.
 For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
@@
 spherical halo (see Fig.~\ref{fig:tidal_tracks_structure}), similar to
-\citet{battaglia+sollima+nipoti2015}. The oblate halo however becomes
-spherical after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's
-dark matter halo is indeed elliptical today, this model may not be an
-adequate description.
+\citet{battaglia+sollima+nipoti2015}. The oblate halo becomes spherical
+after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's dark matter
+halo is indeed elliptical today, this model may not be an adequate
+description.
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
-\caption[Oblate halo before and after tidal evolution]{The initial and
-final (after 9Gyr of tidal evolution) projected density profiles for the
+\caption[Oblate halo projected density snapshots]{The initial and final
+(after 9Gyr of tidal evolution) projected density profiles for the
 oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
@@
 static MW-only potential) before and after the pericentre. Our final
-model, the \texttt{MW\ impact} model, is able to approximately reproduce
-the observed position of Scl.
+model, the \texttt{MW\ impact} model, can approximately reproduce the
+observed position of Scl.
 
 Fig.~\ref{fig:tidal_tracks_umi} compares the tidal evolution of Scl on
-the mean, \texttt{MW\ impact} , and \smallperi{} orbit. The mean orbit
-evloves less down the tidal track than the \smallperi{} model. Instead,
-the \texttt{MW\ impact} orbit experiences most tidal evolution during
-its first MW-pericentre. While evolving further along the tidal track,
-the stars of this model nevertheless remain exponential
+the mean, \texttt{MW\ impact}, and \smallperi{} orbit. The mean orbit
+loses less mass than the \smallperi{} model. Instead, the
+\texttt{MW\ impact} orbit experiences most tidal evolution during its
+first MW pericentre. While evolving further along the tidal track, the
+stars of this model nevertheless remain exponential
 (Fig.~\ref{fig:scl_mw_impact_i_f}). We suggest that the impulsive
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_orbits_boundmass.pdf}}
-\caption[Ursa Minor tidal dependence on orbit]{Similar to
+\caption[Sculptor's tidal evolution for different orbits]{Similar to
 Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
 orbits in the LMC. Most LMC models evolve more weakly than the MW
-models, however these do not reach full agreement with the present-day
-position so misrepresent recent tidal
-evolution.}\label{fig:tidal_tracks_umi}
+models.}\label{fig:tidal_tracks_umi}
 \end{figure}
@@
 
-Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream}
-shows the resulting distributions of velocities and distance along the
-stream orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
+Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream} show
+the resulting distributions of velocities and distance along the stream
+orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
 detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper

### 2025-10-08 thesis/extra_rv_models.tex
grammar/spelling edits

@@
-\chapter{Line-of-sight velocities: Sample selection and
-modelling}\label{sec:extra_rv_models}
+\chapter{Line-of-Sight Velocities: Sample Selection and
+Modelling}\label{sec:extra_rv_models}
 
 In this section, we analyze the observed line-of-sight (LOS) velocity
-distributions for Sculptor (Scl) and Ursa Minor (UMi). Our aim is to
-test for kinematic tidal signatures by combining literature LOS
-velocities with J+24's membership framework. Our derived systemic
-velocities and dispersions are consistent with literature. We find weak
-evidence for a velocity gradient in Scl. As the gradient is misaligned
-with the orbit, the gradient more likely reflects intrinsic rotation
-than tidal disruption. Scl's velocity dispersion also rises within the
-effective radius, but predominantly in the inner regions, contrary to
-tidal disruption. We find no evidence of a gradient in mean velocity or
+distributions for Sculptor (Scl) and Ursa Minor (UMi). We aim to test
+for kinematic tidal signatures by combining literature LOS velocities
+with J+24's membership framework. Our derived systemic velocities and
+dispersions are consistent with the literature. We find weak evidence
+for a velocity gradient in Scl. As the gradient is misaligned with the
+orbit, the gradient more likely reflects intrinsic rotation than tidal
+disruption. Scl's velocity dispersion also rises within the effective
+radius, but predominantly in the inner regions, contrary to tidal
+disruption. We find no evidence of a gradient in mean velocity or
 velocity dispersion in UMi. We conclude that Scl and UMi do not show
@@
 Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
-\citet{sestito+2023b}; and APOGEE. We then crossmatch all catalogues to
-J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we match
-to the nearest star within 1--3 arcseconds. We combine measurements of
-the same star using inverse-variance weighting. To reduce likely
-binaries, we remove stars with significant velocity
+\citet{sestito+2023b}; and APOGEE. We then cross-match all catalogues to
+J+24 Gaia stars. If a study did not report \emph{Gaia} DR3 source IDs,
+we match to the nearest star within 1--3 arcseconds. We combine
+measurements of the same star using inverse-variance weighting. To
+reduce likely binaries, we remove stars with significant velocity
 dispersions.\footnote{Specifically, using that
@@
 Section~\ref{sec:the_algorithm}). We assume that the satellite and
-background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
-halo/background velocity dispersion of a constant
-\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
-\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
-satellite membership probabilities of greater than 0.2. For Scl, we find
-1918 unique members and UMi, 831.
+background \(v_{\rm los, gsr}\) distributions are Gaussian in the
+Galactic Standard of Rest (GSR, i.e., same location as ICRS but
+velocities relative to the galactic centre). For the satellite, we adopt
+a mean and standard deviation based on
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}, and, for the
+background, mean \(0\,\kms\) and dispersion
+\(\sigma_{\rm halo} = 100\,\kms\) \citep[e.g.][]{brown+2010}. We select
+stars with velocity-informed satellite membership probabilities of
+greater than 0.2. For Scl, we find 1918 unique members and UMi, 831.
 
@@
 
-We correct the velocities for the solar motion and on-sky size of the
-galaxy. We transform the velocities into the galactic standard of rest
-(GSR, i.e.~same location as ICRS but velocities relative to the galactic
-centre), and then transform velocities to correct for the apparent
-gradient induced by the dwarf's proper motion \citep[see
-also][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
-velocities in the GSR frame subtracting the PM-induced gradient. The
+We correct the velocities for the solar motion and the on-sky size of
+the galaxy. We transform the velocities into the GSR and correct for the
+apparent gradient induced by the dwarf's proper motion
+\citep[see,][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
+velocities in the GSR frame, subtracting the PM-induced gradient. The
 correction from both effects induces an apparent gradient of about
 \(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor. The uncertainty
-on this velocity correction are less than the individual star
+on this velocity correction is less than the individual star
 uncertainties and any derived velocity gradients.
 
-\section{MCMC modeling}\label{mcmc-modeling}
+\section{Monte Carlo Markov chain
+modelling}\label{monte-carlo-markov-chain-modelling}
 
-We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
+We fit Monte Carlo Markov chain (MCMC) models to solve for the systemic
 velocity, velocity dispersion, and possible gradients in velocities or
@@
 \mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta,
-}\end{equation}
-
-for tangent-plane coordinates \(\xi\) and \(\eta\), systemic velocity
-\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
-dispersion at a given position, \(\sigma\), is assumed to depend as a
-power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
+}\end{equation} for tangent plane coordinates \(\xi\) and \(\eta\),
+systemic velocity \(\mu_0\), and velocity gradient slopes \(a\) and
+\(b\). The velocity dispersion at a given position, \(\sigma\), is
+assumed to depend as a power-law on elliptical radius \(R_{\rm ell}\)
+alone: \begin{equation}{
 \log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
-\caption[LOS velocity fit to Scl]{\textbf{Left}: Velocity histogram of
+\caption[Velocity dispersion fits]{\textbf{Left:} Velocity histogram of
 Scl and UMi in terms of projected-correction GSR LOS velocity. Black
-points with error bars are from the crossmatched observed sample, and
+points with error bars are from the sample of LOS-velocity members, and
 green lines represent MCMC fits to the velocity dispersion.
-\textbf{Right} The observed velocity dispersion gradient (black points)
+\textbf{Right:} The observed velocity dispersion gradient (black points)
 in 10 equal number bins and the derived slopes for MCMC samples (green
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
-\caption[Scl velocity gradient]{\textbf{Top} members of Sculptor plotted
-in the tangent plane coloured by corrected velocity difference from mean
-\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
-Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
-motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
-scale). \textbf{Bottom}: The corrected LOS velocity along the best fit
-rotational axis. RV members are black points, the systematic \(v_z\) is
-the horizontal grey line, blue lines represent the (projected) gradient
-from MCMC samples, and the orange line is a rolling median (with a
-window size of 50).}\label{fig:scl_velocity_gradient_scatter}
+\caption[A possible velocity gradient in Sculptor]{\textbf{Top} members
+of Sculptor plotted in the tangent plane, coloured by corrected velocity
+difference from mean \(v_z - \bar v_z\). The black ellipse marks the
+half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
+arrows mark the proper motion (PM, GSR frame) and derived velocity
+gradient (rot) vectors (to scale). \textbf{Bottom}: The corrected LOS
+velocity along the most likely rotational axis. RV members are black
+points, the systematic \(v_z\) is the horizontal grey line, blue lines
+represent the (projected) gradient from MCMC samples, and the orange
+line is a rolling median (with a window size of
+50).}\label{fig:scl_velocity_gradient_scatter}
 \end{figure}
 
-We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
+We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\) with
 velocity dispersion \(9.64\pm0.16\,\kms\) (see Fig.~\ref{fig:rv_hists}).
 Our values are very consistent with previous work
-\citep[e.g.][\citet{arroyo-polonio+2024},
-\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
-gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
-\(-149_{-13}^{+17}\) degrees, similar to past detections
+\citep[e.g.,][]{walker+2009, arroyo-polonio+2024, battaglia+2008}. We
+detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at a
+position angle of \(-149_{-13}^{+17}\) degrees, similar to past
+detections
 \citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
@@
 sample and the MCMC samples for the velocity gradient. While some
-samples have a consistent gradient with 0, most samples have a positive
+samples have a gradient consistent with 0, most samples have a positive
 velocity gradient, consistent with the rolling median trend. The
-velocity gradient in Scl is misaligned with the proper motion
+velocity gradient in Scl is misaligned with the proper motion.
 
@@
 
-\emph{Binarity}. Stars in binary systems can inflate the inferred
-system's velocity dispersion. For Scl and UMi with high measured
-velocity dispersion, binaries likely add \(\sim 1\,\kms\) to the
-velocity dispersion \citep{spencer+2018, gration+2025}.
+\emph{Binarity}. Binary star systems may inflate the inferred system's
+velocity dispersion. For Scl and UMi with high measured velocity
+dispersion, binaries likely add \(\sim 1\,\kms\) to the velocity
+dispersion \citep{spencer+2018, gration+2025}.
 
 \emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
-multiple populations with different kinematics\citep[\citet{pace+2020},
-\citet{tolstoy+2004}]{arroyo-polonio+2024}. We do not model these
-separately---it is unclear how to define a velocity dispersion in such a
-case.
+multiple populations with different kinematics (see
+Section~\ref{sec:peculiarities}). We do not model these separately---it
+is unclear how to define a velocity dispersion in such a case.
 
@@
 
-\emph{Tangental motions.} Precise star-by-star proper motions would
+\emph{Tangential motions.} Precise star-by-star proper motions would
 further test for kinematic disequilibrium. Presently, \emph{Gaia}
 uncertainties are too large to permit such studies for Sculptor and Ursa
-Minor.\footnote{Specifically, typical proper motion uncertntanties for
+Minor.\footnote{Specifically, typical proper motion uncertainties for
   faint, member stars of Scl and UMi are around \(0.5\,\masyr\),
@@
 In this section, we analyze literature samples of LOS velocity
-measurements for both Scl and UMi. In each case, we find systemic LOS
-velocities and velocity dispersions consistent with past work. We detect
-weak evidence for a velocity gradient in Scl, however the gradient is
-mis-aligned with the proper motion of Scl so is unlikely to be of tidal
-origin. In both galaxies, the velocity dispersion profile is consistent
-with a constant velocity dispersion with radius. The lack of
-observational evidence for ongoing tidal disruption in the velocity
-distribution of stars in Scl and UMi further supports our
-interpretations in the main text.
+measurements for both Scl and UMi. In each case, we find systemic
+velocities and dispersions consistent with past work. We detect weak
+evidence for a velocity gradient in Scl, but the gradient is misaligned
+with the proper motion of Scl, so it is unlikely to be of tidal origin.
+Sculptor also shows evidence for a mildly rising velocity dispersion
+with radius; however, the outer regions appear to be flat in velocity
+dispersion. The lack of observational evidence for ongoing tidal
+disruption in the velocity distribution of stars in Scl and UMi further
+supports our interpretations in the main text.

### 2025-10-08 thesis/introduction.tex
grammar/spelling edits

@@
 the galaxy luminosity function increases towards fainter objects, dwarfs
-are the most numerous galaxies in the Universe
+are the most numerous of galaxies
 \citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
 \emph{dark-matter dominated}, with mass to light ratios which may exceed
-1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.
+1000 \(M_\odot/ L_\odot\) \citep[implying \(\sim1000\) times more dark
+matter than stellar matter, e.g.,][]{simon+geha2007, hayashi+2023}.
 
-With the exception of the Magellanic Clouds, most dwarf galaxy
-satellites of the Milky Way (MW) are \emph{quenched}, with little to no
-recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
-satellites contain old stellar populations which are \emph{relics} from
-the early universe, consisting of many of the oldest and most metal poor
-stars known \citep{simon2019}. Understanding the properties of dwarf
-galaxies thus has implications across astronomy, from cosmological
-structure formation on small scales to the formation of metal-poor
-stellar populations.
+Except for the Magellanic Clouds, most dwarf galaxy satellites of the
+Milky Way (MW) are \emph{quenched}, with little to no recent star
+formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW satellites
+contain old stellar populations which are \emph{relics} from the early
+universe, consisting of many of the oldest and most metal-poor stars
+known \citep{simon2019}. Understanding the properties of dwarf galaxies
+thus has implications across astronomy, from cosmological structure
+formation on small scales to the formation of metal-poor stellar
+populations.
 
@@
   were likely always known to humans at southern latitudes.}, with no
-known analogues at the time, already presented such an enigma. H.
-Shapley presented these dwarfs as a new type of \emph{stellar system}
-resembling the Magellanic Clouds and globular clusters but did not
-attempt to speculate on their exact nature. While dwarf galaxies were
-soon understood to be galaxies based on the inferred luminosities and
-sizes, their nature remained unclear for decades
+known analogues at the time, already presented an enigma. H. Shapley
+presented these dwarfs as a new type of \emph{stellar system} resembling
+the Magellanic Clouds and globular clusters but did not attempt to
+speculate on their nature. While dwarf galaxies were soon understood to
+be galaxies based on the inferred luminosities and sizes, their exact
+nature remained unclear for decades
 \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.
@@
 \citep{milgrom1995}. Over time, however, a consensus developed where the
-high mass-to-light ratios of dwarf galaxies was due to the presence of a
-dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
+high mass-to-light ratios of dwarf galaxies were due to the presence of
+a dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
 Since then, the properties of dwarf galaxies have played an increasingly
@@
   dwarf galaxy names we discuss contain ``dwarf spheroidal'' (dSph),
-  e.g.~Sculptor dSph, we omit this suffix for brevity. Additionally, the
-  12 classical dwarf satellites of our Galaxy are (in order of
+  e.g., Sculptor dSph, we omit this suffix for brevity. Additionally,
+  the 12 classical dwarf satellites of our Galaxy are (in order of
   decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor, Antlia
@@
   and Crater II. Antlia II, Crater II, and Canes Venatici I are the only
-  post digital sky surveys additions.} occupy intermediate luminosities
+  post-digital sky surveys additions.} occupy intermediate luminosities
 ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
 \(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
-these systems are old, non-star forming, gas-poor, and spheroidal. All
+these systems are old, non-star-forming, gas-poor, and spheroidal. All
 Milky Way satellites discovered before digital sky surveys are
@@
 
-Most well-studied dwarf galaxies lie near the Milky Way or in its
-vicinity, the Local Group of galaxies. The Local Group is defined as the
-group consisting of galaxies within \(\sim 1\) Mpc from the Milky
+Most well-studied dwarf galaxies lie in the vicinity of the Milky Way,
+the Local Group of galaxies. The Local Group is defined as the group
+consisting of galaxies within \(\sim 1\) Mpc from the Milky
 Way-Andromeda centre \citep[e.g.,][ and references
@@
 the MW satellite system, including dwarf galaxies, globular clusters,
-and ambiguous systems. This nearby population of dwarf galaxies are
-opportune for resolved, detailed studies aimed to understand the nature
-of these systems.
+and ambiguous systems. This nearby population of dwarf galaxies is
+opportune for resolved, detailed studies aimed at understanding their
+nature.
 
@@
 \includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
-\caption[Images of dwarf galaxies]{Images of the LMC (DSS2), Fornax
-\citep[DES DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
+\caption[Images of dwarf galaxies]{Images\footnotemark{} of the LMC
+\citep[Digitized Sky Survey II,][]{lasker+1996}, Fornax \citep[DES
+DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
 \citep[UNWISE,][with \textit{Gaia} point sources
-over-plotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
+overplotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
 The grey ellipse represents the half-light radius for the three dwarf
@@
 \end{figure}
+\footnotetext{Created with hips2fits
+  (https://alasky.cds.unistra.fr/hips-image-services/hips2fits), a
+  service provided by CDS.}
 
@@
 \caption[The on-sky distribution of Milky Way satellites]{The location
-of MW dwarf galaxies on the sky. We label the classical dwarf galaxies
+of MW satellites on the sky. We label the classical dwarf galaxies
 (green diamonds), fainter dwarfs (blue squares), globular clusters
@@
 highlighted as two dwarfs we study later. The background image is from
-ESA/Gaia/DPAC
-(https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2).
-Dwarf galaxies (confirmed), globular clusters, and ambiguous systems are
-from the \citet{pace2024} catalogue (version
-1.0.3).}\label{fig:mw_satellite_system}
+ESA/Gaia/DPAC.\footnotemark{} Dwarf galaxies (confirmed), globular
+clusters, and ambiguous systems are from the \citet{pace2024} catalogue
+(version 1.0.3).}\label{fig:mw_satellite_system}
 \end{figure}
+\footnotetext{https://www.esa.int/ESA\_Multimedia/Images/2018/04/Gaia\_s\_sky\_in\_colour2}
 
@@
 universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
-matter (DM), and 5\% regular old baryons\footnote{In a classic,
-  astronomer's corruption of jargon, \emph{Baryons} here means
-  baryons+leptons, i.e.~any standard model massive fermion.}
+matter (DM), and 5\% regular old baryons\footnote{In a classic
+  astronomer's corruption of jargon, \emph{Baryons} here means baryons
+  and leptons, i.e., protons, neutrons, and electrons.}
 \citep{planckcollaboration+2020}. While the composition of dark matter
@@
 In \LCDM{}, dark matter is assumed to interact only gravitationally.
-Light passes through dark matter unimpeded---in this sense, dark matter
-is transparent. Dark matter is also commonly assumed to be \emph{cold},
-i.e.~typical velocities much smaller than the speed of light in the
-early universe. If dark matter is cold, then it should condense on all
-scales, forming non-linear structures (or \emph{halos}) from the size of
-galaxy clusters to smaller than the faintest dwarf galaxies.
-Implications of dark matter properties include cosmological structural
-formation, galaxy structure, and galaxy interactions.
+Light and matter pass through dark matter unimpeded---in this sense,
+dark matter is transparent. Dark matter is also commonly assumed to be
+\emph{cold}, i.e., with typical velocities much smaller than the speed
+of light in the early universe. If dark matter is cold, then it should
+condense on all scales, forming non-linear structures (or \emph{halos})
+from the size of galaxy clusters to smaller than the faintest dwarf
+galaxies. Implications of dark matter properties include cosmological
+structural formation, galaxy structure, and galaxy interactions.
 
@@
 nuclei to form atoms, baryons decoupled from radiation and fell into
-dark matter halos, where the condensed at the centre through radiative
+dark matter halos, where they condensed at the centre through radiative
 energy losses. The densest pockets of baryons later formed the first
@@
 
-Dark matter halos, and their associated galaxies, rarely evolve in
+Dark matter halos and their associated galaxies rarely evolve in
 isolation. Instead, \LCDM{} structure formation is \emph{hierarchical}.
@@
 \citep[e.g.,][]{white+rees1978, blumenthal+1984, white+frenk1991}.
-Hierarchical assembly is evident through the large scale structure of
+Hierarchical assembly is evident through the large-scale structure of
 the universe, remnants of past mergers within the Milky Way, and tidal
@@
 \citet{ludlow+2016}. While concentration tends to decrease with
-increasing mass, the relation has substantial scatter. Other parameters
-such as the halo spin or shape may affect the scatter of the
+increasing mass, the relation has substantial scatter. Other parameters,
+such as the halo spin or shape, may affect the scatter of the
 mass-concentration relation, but their effect is typically expected to
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
-\caption[Cosmological halo mass and stellar mass
+\caption[Cosmological mass-concentration and stellar mass-halo mass
 relations]{\textbf{Left} The NFW halo concentration \(c=r_{200} / r_s\)
@@
 1-\(\sigma\) shaded region is the mass-concentration relation from
-\citet{ludlow+2016} for \(z=0\). \textbf{Right} Stellar mass (top) as a
-function of maximum circular velocity. The solid line with the
-1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
-with scatter points simulated central galaxies from \apostle{} in
-\citet{fattahi+2018}. The pink star illustrates the location of the
-Fornax galaxy, whose density profiles are shown in
-Fig.~\ref{fig:nfw_density}. \textbf{add rmax-vmax plot
-too}}\label{fig:smhm}
+\citet{ludlow+2016} for \(z=0\). \textbf{Middle}: Equivalent to the left
+except in terms of the halo maximum circular velocity, \(\vmax\), and
+radius where the velocity is maximized, \(\rmax\). \textbf{Right}
+Stellar mass (top) as a function of maximum circular velocity. The solid
+line with the 1-\(\sigma\) shaded region is the relation from
+\citet{fattahi+2018} with scatter points simulated central galaxies from
+\apostle{} in \citet{fattahi+2018}. The pink star illustrates the
+location of the Fornax galaxy, whose density profiles are shown in
+Fig.~\ref{fig:nfw_density}.}\label{fig:smhm}
 \end{figure}
@@
 shallower potential wells, feedback becomes more effective at removing
-gas. Re-ionization additionally suppresses late star formation in the
+gas. Reionization additionally suppresses late star formation in the
 faintest galaxies. As a result, the resulting stellar mass of a dwarf
@@
 observed abundance has been known as the \emph{missing satellites
-problem}. Additionally, a number of observations suggest that some dwarf
+problem}. Additionally, several observations suggest that some dwarf
 galaxies, although not all, possess dark matter ``cores,''
@@
 Since Local Group dwarfs are nearby, they are resolved into individual
-stars, and therefore we can study these galaxies on a star-by-star
+stars, and therefore, we can study these galaxies on a star-by-star
 basis. As a result, it is possible to measure the 3D velocity and
@@
 by 106.5 degrees on the same focal plane, \emph{Gaia} measures changes
-in relative positions of stars across small and large angles. Combining
-measurements from multiple epochs across several years, an absolute
-all-sky reference frame is derived from which parallax and proper
-motions are derived. In addition to astrometry, \emph{Gaia} measures
-photometry in the wide \emph{G} band (330--1050nm) and colours from the
-blue photometer (BP, 330--680 nm) and red photometer (RP, 640--1050 nm).
-\emph{Gaia} additionally provides low resolution BP-RP spectra and
-radial velocity measurements of bright stars \citep[of magnitudes
-\(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our work,
-\emph{Gaia}'s most relevant measurements are \(G\) magnitude,
+in the relative positions of stars across small and large angles.
+Combining measurements from multiple epochs across several years, an
+absolute all-sky reference frame is derived from which parallax and
+proper motions are derived. In addition to astrometry, \emph{Gaia}
+measures photometry in the wide \emph{G} band (330--1050nm) and colours
+from the blue photometer (BP, 330--680 nm) and red photometer (RP,
+640--1050 nm). \emph{Gaia} additionally provides low-resolution BP-RP
+spectra and radial velocity measurements of bright stars \citep[of
+magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
+work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
 \(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
@@
   change in right ascension, \(\alpha\), and declination, \(\delta\),
-  typically in units of mili-arcsecond (mas) per year.
+  typically in units of milli-arcsecond (mas) per year.
   \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
@@
 \emph{Gaia}, few galaxies had precisely measured proper motions
-\citep[e.g.~using Hubble Space Telescope,][]{piatek+2005, sohn+2017}.
-\emph{Gaia} allowed for some of the first systematic and precise
-determinations of Milky Way satellite proper motions
-\citep{pace+li2019, MV2020a}. While the proper motion uncertainty of a
-typical dwarf member star is often large, by combining the proper
-motions of 100s or 1000s of stars from \emph{Gaia}, precise average
-proper motion measurements can be determined, sometimes only limited by
-\emph{Gaia}'s systematic error floor \citep[e.g.,][]{MV2020a}. Proper
-motions have thus ushered in a new era for MW satellite dynamical
-studies, where we can derive precise orbits for any satellite, assuming
-a given MW potential. In addition, \emph{Gaia} helps establish
-membership by separating out contaminating MW foreground stars. By
-measuring parallaxes and/or proper motions, many more background and
-foreground stars can be classified as non-members
+\citep[e.g., using the Hubble Space
+Telescope,][]{piatek+2005, sohn+2017}. \emph{Gaia} allowed for some of
+the first systematic and precise determinations of Milky Way satellite
+proper motions \citep{pace+li2019, MV2020a}. While the proper motion
+uncertainty of a typical dwarf member star is often large, by combining
+the proper motions of 100s or 1000s of stars from \emph{Gaia}, precise
+average proper motion measurements can be determined, sometimes only
+limited by \emph{Gaia}'s systematic error floor
+\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new era
+for MW satellite dynamical studies, where we can derive precise orbits
+for any satellite, assuming a given MW potential. In addition,
+\emph{Gaia} helps establish membership by filtering contaminating MW
+foreground stars. By measuring parallaxes and/or proper motions, many
+more background and foreground stars can be classified as non-members
 \citep[e.g.,][]{battaglia+2022, jensen+2024}.
@@
 
-Projected luminosity / stellar density profiles efficiently characterize
+Projected luminosity/stellar density profiles efficiently characterize
 the radial structure of a galaxy. At its most basic, light profiles
@@
 Finally, the \citet{sersic1963} profile represents a generalization of
-an exponential profile, and describes most dwarf galaxy light profiles
+an exponential profile and describes most dwarf galaxy light profiles
 well. Typically parameterized in terms of a half-light radius \(R_h\),
-the density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\),
+the density at half-light radius \(\Sigma_h\), and a Sérsic index \(n\),
 the profile's equation is \begin{equation}{
@@
 to an exponential profile, while \(n=4\) recovers
-\citet{devaucouleurs1948}'s profile for elliptical galaxies. Although a
+\citepos{devaucouleurs1948} profile for elliptical galaxies. Although a
 Sérsic profile is less commonly applied to dwarf galaxies,
@@
 galaxies\footnote{While studied for far longer, the exponential origin
-  of disk galaxies is no better understood. A sample of ideas range from
-  scattering of stars \citep{elmegreen+struck2013, wu+2022} to angular
-  momentum transport, to disk viscosity-driven radial gas flows
+  of disk galaxies is no better understood. Ideas range from scattering
+  of stars \citep{elmegreen+struck2013, wu+2022} to angular momentum
+  transport, to disk viscosity-driven radial gas flows
   \citep{lin+pringle1987, wang+lilly2022} or spherical collapse
@@
 detected a velocity gradient in UMi, suggestive of tidal disruption.
-Later, \citet{martinez-delgado+2001} find that stars extend far beyond
+Later, \citet{martinez-delgado+2001} found stars extending far beyond
 the King-profile ``tidal radius'', aligned with UMi's elongation,
@@
 \pandocbounded{\includegraphics[keepaspectratio]{./figures/scl_umi_fornax_exp.pdf}}
-\caption[Scl and UMi have an extended density profile]{A plot of the
-surface density profiles of Sculptor (orange squares), Ursa Minor (red
-triangles), and Fornax (green circles) scaled to their half-light radius
-and the density at half-light radius (data described in
+\caption[The extended stellar profiles of Sculptor and Ursa
+Minor]{Surface density profiles of Sculptor (orange squares), Ursa Minor
+(red triangles), and Fornax (green circles) scaled to their half-light
+radius and the density at half-light radius (data described in
 Section~\ref{sec:observations}). The solid black line is an exponential
-profile (Eq.~\ref{eq:exponential_law}). Scl and UMi show a clear
-overdensity at large radii.}\label{fig:scl_umi_vs_fornax}
+profile (Eq.~\ref{eq:exponential_law}). Scl and UMi show a clear excess
+over an exponential at large radii.}\label{fig:scl_umi_vs_fornax}
 \end{figure}
@@
 \caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
-References are: 1.
+References are:(1).
 Muñoz et al.
-(2018, Sérsic fit), 2.
-Tran et al.
-(2022, RR lyrae distance), 3.
-Alan W.
-McConnachie and Venn (2020b), 4.
-Arroyo-Polonio et al.
+(2018, Sérsic fit), (2) Tran et al.
+(2022, RR lyrae distance), (3) Alan W.
+McConnachie and Venn (2020b), (4) Arroyo-Polonio et al.
 (2024).
@@
 (2018, Sérsic fit), (2) Garofalo et al.
-(2025) pRR lyrae distance], (3) Alan W.
+(2025, RR lyrae distance), (3) Alan W.
 McConnachie and Venn (2020a), (4) Pace et al.
@@
 Since many dwarfs are near the resolution limit, they are vulnerable to
-artifical disruption
+artificial disruption
 \citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
@@
 analytic host potential, achieving excellent numerical convergence. For
-instance, the simulations we describe later in this thesis reach the
+instance, the simulations we describe later in this thesis reach three
 times higher resolution than Aquarius \citep{springel+2008} with 400
@@
 common than idealized work or the Milky Way system suggests, but the
-origin and role of numerics and assumptions in this discrepancy is
+origin and role of numerics and assumptions in this discrepancy are
 unclear \citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall
@@
 undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
-NFW halos also are found to be nearly immune to tidal disruption
+NFW halos are also found to be nearly immune to tidal disruption
 \citep{EP2020}, but cored dark matter halos may disrupt fully and faster
@@
 tidal field. They explain the resilience of NFW halos against full tidal
-disruption and origin of well-defined ``tidal tracks'' \citep[as
+disruption and the origin of well-defined ``tidal tracks'' \citep[as
 observed in][]{PNM2008, green+vandenbosch2019, EN2021}.
@@
 With precise orbital constraints and improved models of the Milky Way
-potential, recent studies has continued to probe the dynamical histories
-of individual dwarf galaxies.
-\citet{borukhovetskaya+2022, dicintio+2024} both ran simulations tuned
-to Fornax, showing that this galaxy's stellar component or globular
-clusters are likely not affected by tides. Similarly,
-\citet{borukhovetskaya+2022a} analyzed Crater II, showing that the
-present-day structure is challenging to reconcile with NFW initial
-conditions and Galactic tides. Most relevantly, \citet{iorio+2019}
-modelled Sculptor, concluding tides likely do not affect the stellar
-component.
+potential, recent studies have continued to probe the dynamical
+histories of individual dwarf galaxies.
+\citet{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024}
+both ran simulations tuned to Fornax, showing that this galaxy's stellar
+component or globular clusters are likely not affected by tides.
+Similarly, \citet{borukhovetskaya+2022a} analyzed Crater II, showing
+that the present-day structure is challenging to reconcile with NFW
+initial conditions and Galactic tides. Most relevantly,
+\citet{iorio+2019} also tailored simulations to Scl, finding weak tidal
+effects.
 
@@
 dwarf has been able to achieve equilibrium after pericentric passage in
-a highly-eccentric orbit. The break radius \(r_{\rm break}\) is
+a highly eccentric orbit. The break radius \(r_{\rm break}\) is
 proportional to the velocity dispersion, \(\sigma_v\), and time elapsed
@@
 \(r_{\rm break}\) describes where the dynamical timescale is longer than
-the time since the perturbation, i.e.~the radius within which the galaxy
-is dynamically relaxed.
+the time since the perturbation, i.e., the radius within which the
+galaxy is dynamically relaxed.
 
@@
 particle based on the relative densities of the distribution functions
-in energy space (see Section~\ref{sec:painting_stars}). The stars follow
-initially a 2D exponential profile with a scale radius
+in energy space (see Section~\ref{sec:painting_stars}). The stars
+initially follow a 2D exponential profile with a scale radius
 \(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We then
@@
 idealized simulation at the second apocentre, as seen from the centre of
-the host. The projected density of stars, is relatively undisturbed and
+the host. The projected density of stars is relatively undisturbed and
 spherical in the centre, but becomes non-isotropic outside the break
@@
 resembles the outer excess in Sculptor and Ursa Minor. This thesis aims
-to investigate if a tidal explanation like this is viable given the
+to investigate whether a tidal explanation like this is viable given the
 current properties of each galaxy.
@@
 \caption[Example tidal simulation]{Example density and velocity
-distributions of an idealized simulation at the second apocentre
-(initialized on the first apocentre). \textbf{Top}: The projected 2D
-stellar density in the \(y'\)--\(z'\) plane for the initial (left) and
-final (right) simulation. \(y'\) is the direction of tangental orbital
-motion and \(z'\) is the direction of angular momentum, as seen from the
-centre of the host. The dashed green circle represents the break radius
+distributions of an idealized dwarf galaxy model undergoing tidal
+stripping. \textbf{Top}: The projected 2D stellar density in the
+\(y'\)--\(z'\) plane for the initial (left) and final (right)
+simulation. \(y'\) is the direction of tangential orbital motion, and
+\(z'\) is the direction of angular momentum, as seen from the centre of
+the host. The dashed green circle represents the break radius
 (Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
@@
 for the initial (dotted) and final (solid) simulation snapshot.
-\textbf{Bottom left}: the mean radial velocity (dot product of relative
-position and velocity relative to dwarf centre) as a function of
-projected radius. The green arrow marks the break radius in both lower
-panels.}\label{fig:idealized_break_radius}
+\textbf{Bottom right}: the mean 3D radial velocity (dot product of
+relative position and velocity relative to dwarf centre) as a function
+of projected radius. The green arrow marks the break radius in both
+lower panels.}\label{fig:idealized_break_radius}
 \end{figure}
@@
 our results for the tidal effects on Sculptor and Ursa Minor in Chapter
-\ref{sec:results}. We discuss the implications of our results and ending
+\ref{sec:results}. We discuss the implications of our results and end
 with a summary and outlook in Chapter \ref{sec:discussion}.

### 2025-10-08 thesis/main.tex
grammar/spelling edits

@@
 
-\chapter*{Acknowledgements}
-\addcontentsline{toc}{chapter}{Acknowledgements}
-I promise I will fill this out thoroughly.
-
-%Julio Navarro, my supervisor. Rapha and other collegues
-
-%Friends and family and collegues (fill in later).
-
-%% Lists
-
 \tableofcontents
@@
 
+\chapter*{Acknowledgments}
+\addcontentsline{toc}{chapter}{Acknowledgments}
+\input{acknowledgments}
+
+%% Lists
 %\chapter*{Symbols}
@@
 
+\thispagestyle{empty}
+\renewcommand{\epigraphflush}{center}
+\vfill
+\setlength{\epigraphwidth}{0.5\textwidth}
+\epigraph{The sun had fallen beneath the horizon in the east, and the long day was about to end.
+But tomorrow, the sun would rise in the west again, bathing the world in a kinder light.}{Baoshu, translation by Ken Liu}
+
+\epigraph{
+In the land of my birth, there is a hill, and upon that hill is a rock where I have sat and watched a changing world.
+This rock will not change, I would say.
+I'll pass from this Earth, but the rock will remain ever watchful of the land.
+
+But my feet kicked the dirt beneath me. Were the granules of soil not once like this rock? Is the rock not a collection of their multitudes?
+
+[We are] neither destroyed nor created, but a collection of dust meant to observe a changing world and then return to dust. 
+}{Joe Fisher, {\it Midnight Burger}}
+\vfill
+
+\newpage
 \pagenumbering{arabic}
@@
 
-\chapter{Galactic tidal effects on Sculptor and Ursa Minor}\label{sec:results}
+\chapter{Galactic Tidal Effects on Sculptor and Ursa Minor}\label{sec:results}
 \input{results}
@@
 
-\chapter*{Research Acknowledgements}
-\addcontentsline{toc}{chapter}{Research Acknowledgements}
-This work has made use of data from the European Space Agency (ESA) mission
-{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
-Data Processing and Analysis Consortium (DPAC,
-\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
-has been provided by national institutions, in particular the institutions
-participating in the {\it Gaia} Multilateral Agreement. 
-\citep{gaiacollaboration+2016, gaiacollaboration+2023}.
-
-This research made use of hips2fits,\footnote{https://alasky.cds.unistra.fr/hips-image-services/hips2fits} a service provided by CDS.
-
-The Digitized Sky Surveys were produced at the Space Telescope Science Institute under U.S.
-Government grant NAG W-2166.
-The images of these surveys are based on photographic data obtained using the Oschin Schmidt Telescope on Palomar Mountain and the UK Schmidt Telescope.
-The plates were processed into the present compressed digital form with the permission of these institutions.
-
-The National Geographic Society --- Palomar Observatory Sky Atlas (POSS-I) was made by the California Institute of Technology with grants from the National Geographic Society.
-
-The Second Palomar Observatory Sky Survey (POSS-II) was made by the California Institute of Technology with funds from the National Science Foundation, the National Geographic Society, the Sloan Foundation, the Samuel Oschin Foundation, and the Eastman Kodak Corporation.
-
-The Oschin Schmidt Telescope is operated by the California Institute of Technology and Palomar Observatory.
-
-The UK Schmidt Telescope was operated by the Royal Observatory Edinburgh, with funding from the UK Science and Engineering Research Council (later the UK Particle Physics and Astronomy Research Council), until 1988 June, and thereafter by the Anglo-Australian Observatory.
-The blue plates of the southern Sky Atlas and its Equatorial Extension (together known as the SERC-J), as well as the Equatorial Red (ER), and the Second Epoch [red] Survey (SES) were all taken with the UK Schmidt.
-
-All data are subject to the copyright given in the copyright summary. Copyright information specific to individual plates is provided in the downloaded FITS headers.
-
-Supplemental funding for sky-survey work at the ST ScI is provided by the European Southern Observatory.
+\chapter*{Research Acknowledgments}
+\addcontentsline{toc}{chapter}{Research Acknowledgments}
+\input{research_acknowledgments}
 

### 2025-10-08 thesis/methods.tex
thesis; julio oct 4 comments
full draft of appendicies, improved formatting
grammar/spelling edits

@@
 assume that the potential of each dwarf may be initially approximated by
-a cuspy NFW profile. Since the dwarfs in question are heavily dark
-matter dominated, we shall use a carefully selected sample of dark
+a cuspy NFW profile. Since the dwarfs in question are heavily
+dark-matter-dominated, we shall use a carefully selected sample of dark
 matter particles to emulate the evolution of an embedded tracer stellar
@@
 The uncertainties in this reference frame are typically smaller than the
-uncertainties on a dwarf galaxy's distance and tangental velocity.
+uncertainties on a dwarf galaxy's distance and tangential velocity.
 
-\subsection{Milky Way Potential}\label{milky-way-potential}
+\subsection{Milky Way potential}\label{milky-way-potential}
 
@@
 \centering
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/v_circ_potential.png}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
 \caption[Circular velocity of the Milky Way potential]{Circular velocity
@@
 
-\subsection{Sculptor's Orbit}\label{sculptors-orbit}
+\subsection{Sculptor's orbit}\label{sec:scl_smallperi}
 
@@
 (bottom). The Milky Way is at the centre with the disk lying in the
-\(x\)--\(y\) plane. Our selected \smallperi{} orbit is plotted in black
-and light green transparent orbits represent the past 10Gyr orbits
-sampled over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The
-orbit of Sculptor is well-constrained in this potential and it is
-unlikely to achieve a smaller pericentre than the \smallperi{}
-orbit.}\label{fig:scl_orbits}
+\(x\)--\(y\) plane. Our selected \smallperi{} orbit is plotted in black,
+and light green transparent lines represent orbits sampled over Sculptor
+observables in Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is
+well-constrained in this potential, and it is unlikely to achieve a
+smaller pericentre than the \smallperi{} orbit.}\label{fig:scl_orbits}
 \end{figure}
@@
 \caption[Orbit initial conditions]{The orbital initial conditions for models presented in this work.
-The observables represent the medians from orbital integration used to derived the orbits.
-Instead, the initial position and velocity represents the initialization of the actual N-body model.
+The observables represent the medians from orbital integration used to derive the orbits.
+Instead, the initial position and velocity represent the initialization of the actual N-body model.
 The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
@@
 
-\subsection{Ursa Minor's Orbit}\label{sec:orbit_corrections}
+\subsection{Ursa Minor's orbit}\label{sec:orbit_corrections}
 
@@
 not contribute to the potential. The dark matter density is truncated in
-the outer regions by \begin{equation}{
+the outer regions by
+\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
 \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
@@
 Ursa Minor, together with the mass-concentration relation of \LCDM{}
-halos, determines the properties of the N-body models adopted for each
+halos, determine the properties of the N-body models adopted for each
 dwarf.
 
-Table~\ref{tbl:derived_props} lists the our inferred halo and kinematic
+Table~\ref{tbl:derived_props} lists our inferred halo and kinematic
 properties of Sculptor and Ursa Minor. First, taking the absolute
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/initial_halo_selection.pdf}}
-\caption[Initial halo selection]{Selection of initial halos for Sculptor
-and Ursa Minor. The grey line and pink line with shaded regions
-represent the \citet{ludlow+2016} mass-concentration relation and
-\citet{fattahi+2018} SMHM relation respectively. The curved lines
-represent the velocity dispersion of the initial halo given the
-present-day half-light radius via
-Eq.~\ref{eq:wolf_mass}.}\label{fig:scl_halos}
+\caption[Initial halo parameter choice]{The blue circle indicates our
+choice of initial halo parameters for Sculptor and Ursa Minor. The grey
+line and green line with shaded regions represent the
+\citet{ludlow+2016} mass-concentration relation and \citet{fattahi+2018}
+SMHM relation, respectively. The curved orange lines represent the
+velocity dispersion of an exponential stellar component with the
+observed half-light radius of Sculptor and Ursa
+Minor.}\label{fig:scl_halos}
 \end{figure}
@@
 and the softening length. Tree codes organize particles into a spatial
-tree, enabling grouping of the gravitational forces from nearby
+tree, enabling the grouping of the gravitational forces from nearby
 particles. A gravitational tree code substantially reduces the required
@@
 simulations with different total numbers of particles in isolation.
-Residuals in bottom panel are relative to NFW. The initial conditions
-are dotted, the converged radius is marked by arrows \citep[eq.
-13,][]{power+2003}, and the softening length is marked by a vertical
-bar. Note that a slight reduction in density starting around
+Residuals in the bottom panel are relative to NFW. The initial
+conditions are dotted, the converged radius is marked by arrows
+\citep[eq. 13,][]{power+2003}, and the softening length is marked by a
+vertical bar. Note that a slight reduction in density starting around
 \(r = 30\,\kpc\) is expected given our outer truncation
@@
 profiles of dark matter (blue) and stars (orange) for Sculptor and Ursa
-Minor. The initial conditions are dotted and the isolation evolved
-profiles are solid. The green cross marks the present day half-light
+Minor. The initial conditions are dotted, and the isolation-evolved
+profiles are solid. The green cross marks the present-day half-light
 radius and velocity dispersion, and the black band represents the
 1-sigma mean density of the MW at pericentre across orbits. Initial
-conditions are stable in isolation and mass is dominated by dark
+conditions are stable in isolation, and mass is dominated by dark
 matter.}\label{fig:scl_umi_initial_isolation}

### 2025-10-08 thesis/research_acknowledgments.tex
grammar/spelling edits

This work used the DiRAC@Durham facility managed by the Institute for
Computational Cosmology on behalf of the STFC DiRAC HPC Facility
(www.dirac.ac.uk). The equipment was funded by BEIS capital funding via
STFC capital grants ST/K00042X/1, ST/P002293/1, ST/R002371/1 and
ST/S002502/1, Durham University and STFC operations grant ST/R000832/1.
DiRAC is part of the National e-Infrastructure.

Computations were performed on the Niagara supercomputer
\citep{ponce+2019} at the SciNet HPC Consortium \citep{loken+2010}.
SciNet is funded by Innovation, Science and Economic Development Canada;
the Digital Research Alliance of Canada; the Ontario Research Fund:
Research Excellence; and the University of Toronto.

This work has made use of data from the European Space Agency (ESA)
mission \emph{Gaia} https://www.cosmos.esa.int/gaia), processed by the
\emph{Gaia} Data Processing and Analysis Consortium (DPAC),
https://www.cosmos.esa.int/web/gaia/dpac/consortium. Funding for the
DPAC has been provided by national institutions, in particular the
institutions participating in the \emph{Gaia} Multilateral Agreement.

This work has made use of software including Jupyter, astropy, Makie,
Turing, and Pluto.

### 2025-10-08 thesis/results.tex
full draft of appendicies, improved formatting
grammar/spelling edits

@@
 the MW on its dark matter and stellar components. Next, we consider how
-accounting for the LMC may affect our conclusions. We then analyze UMi
-in a similar manner, considering in turn the dark matter evolution,
-stellar evolution, and orbital effects of the LMC.
+accounting for the LMC may affect our conclusions. We then similarly
+analyze UMi, considering in turn the dark matter evolution, stellar
+evolution, and orbital effects of the LMC.
 
@@
 profile resembles the initial with an inner cusp, but has a sharper
-outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
-mass (see Table~\ref{tbl:scl_sim_results}). However, the inner structure
-is not expected to be affected, as the Jacobi radius is over 3kpc,
-outside of the initial and final \(\rmax\) (see
+outer truncation. Quantitatively, the halo loses \(>90\%\) of its
+initial mass (see Table~\ref{tbl:scl_sim_results}). However, the inner
+structure is not expected to be affected, as the Jacobi radius is over
+3kpc, outside of the initial and final \(\rmax\) (see
 Table~\ref{tbl:scl_sim_results} and Fig.~\ref{fig:scl_tidal_track}).
@@
 Limits range from -150 to 150 kpc in the \(y\)--\(z\) (\(\sim\)orbital)
-plane, and the colourscale is logarithmic spanning 5 orders of magnitude
-between the maximum and minimum values. The green dot represents the
-final expected position of the galaxy and the solid and dotted grey
-curves represents the orbit over one previous or future radial
-oscillation respectively.}\label{fig:scl_sim_images}
+plane, and the colourscale is logarithmic, spanning 5 orders of
+magnitude between the maximum and minimum values. The green dot
+represents the final expected position of the galaxy, and the solid and
+dotted grey curves represent the orbit over one previous or future
+radial oscillation, respectively.}\label{fig:scl_sim_images}
 \end{figure}
@@
 be apparent in the observed stellar component. As calculated for this
-model, (see
+model (see
 Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
@@
 simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
-\(M_{\rm 200}= 98.4\times10^{10}\Mo\) , and the LMC a NFW halo with
+\(M_{\rm 200}= 98.4\times10^{10}\Mo\), and the LMC a NFW halo with
 \(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
@@
 is orbiting the Milky Way on a similar plane to the LMC, but in the
-opposite direction---Scl is unlikely to be a LMC satellite.
+opposite direction---Scl is unlikely to be an LMC satellite.
 
@@
 Section~\ref{sec:orbital_estimation} (with uncertainties doubled). This
-\texttt{LMC-flyby} orbit, is integrated back in time \(2\,\Gyr\) ago to
+\texttt{LMC-flyby} orbit is integrated back in time \(2\,\Gyr\) ago to
 isolate recent tidal effects. We modify Scl's initial halo to have
@@
 \centering
-\caption[Orbits and results for Scl in the LMC potential.]{The orbital properties and dark matter evolution for the models including an LMC.
+\caption[Orbits and results for Sculptor in the MW+LMC potential.]{The orbital properties and dark matter evolution for the models including an LMC.
 Similar to Table \ref{tbl:scl_sim_results} except quantities with respect to the LMC are in parentheses.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_lmc_xyzr_orbits.png}
 \caption[Sculptor orbits with LMC]{Similar to Fig.~\ref{fig:scl_orbits}
@@
 
-Perhaps surprisingly, the combined tidal effect of the MW and LMC are
+Perhaps surprisingly, the combined tidal effect of the MW and LMC is
 weaker for Scl than in the MW-only case.
@@
 matter is less disrupted than the previous MW-only model. The subsequent
-LMC passage modifies Scl's orbit but has otherwise little effect. As a
-result, the dark matter structure evolves mildly and \(\sim 50\%\) of
-mass remains bound (Table~\ref{tbl:scl_lmc_sims}).
-
-JFN: This is mainly because of the altered orbit of Sculptor, which is
-now inferred to come from a much larger apocentre, and, thus, to have
-completed fewer orbits than in the previous MW-only ``smallperi'' orbit.
-This reduces the number of pericentric passages, and therefore the net
-tidal effect.
+LMC passage modifies Scl's orbit but has otherwise little effect. The
+dark matter structure evolves mildly and \(\sim 50\%\) of mass remains
+bound (Table~\ref{tbl:scl_lmc_sims}).
 
@@
 
-While the instantaneous tidal force from the LMC is larger than the MW,
-Scl does not experience the LMC tidal field long enough to display
-disturbances. Scl and the LMC have a similar orbital plane, but orbit in
-opposite directions (Fig.~\ref{fig:scl_lmc_orbits_effect}). As a result,
-the interaction between Scl and the LMC is brief as their relative
-velocity is \(\sim 400\,\kms\) at closest approach. Furthermore, the
-Jacobi radius due to the LMC still falls outside the observed density
-profile (Fig.~\ref{fig:scl_lmc_i_f}) and the MW Jacobi radius is even
-larger. As a result, tides in a MW and LMC potential are even weaker
-overall than for the MW-only orbit.
+The reduced tides of the LMC-including model are likely a result of the
+altered orbit of Sculptor. Compared to the MW-only \smallperi{} orbit,
+the \texttt{LMC-flyby} model completes fewer orbits and therefore
+experiences a reduced net tidal effect. And while the instantaneous
+tidal force from the LMC is larger than the MW, Scl does not experience
+the LMC tidal field long enough to display disturbances. Furthermore,
+the Jacobi radius due to the LMC still falls outside the observed
+density profile (Fig.~\ref{fig:scl_lmc_i_f}), and the MW Jacobi radius
+is even larger. As a result, tides in an MW and LMC potential are even
+weaker overall than for the MW-only orbit.
 
@@
 \caption[Sculptor simulation snapshots with LMC]{Similar to
-Fig.~\ref{fig:scl_sim_images} except for the case where the potential
-includes an LMC. The current position and path of the LMC are
-represented by the green dot and line respectively. We also plot the
-full orbit (over the past 2Gyr) for both Scl and the LMC, as less than
-one radial period happens over this time
-frame.}\label{fig:scl_lmc_sim_images}
+Fig.~\ref{fig:scl_sim_images} for the case where the potential includes
+an LMC. The current position and path of the LMC are represented by the
+green dot and line, respectively. We also plot the full orbit (over the
+past 2Gyr) for both Scl and the LMC, as less than one radial period
+happens over this time frame.}\label{fig:scl_lmc_sim_images}
 \end{figure}
@@
 \caption[Sculptor initial and final density with LMC]{Similar to
-Fig.~\ref{fig:scl_smallperi_i_f} except for the \texttt{LMC-flyby}
-model. The Jacobi and break radii here are calculated with respect to
-the LMC, the corresponding radii with respect to the MW are larger. With
-only one MW pericentre and a recent, rapid LMC encounter, tidal forces
-do not appear to affect the stellar
-distribution.}\label{fig:scl_lmc_i_f}
+Fig.~\ref{fig:scl_smallperi_i_f} for the \texttt{LMC-flyby} model. The
+Jacobi and break radii here are calculated with respect to the LMC; the
+corresponding radii with respect to the MW are larger. With only one MW
+pericentre and a recent, rapid LMC encounter, tidal forces do not appear
+to affect the stellar distribution.}\label{fig:scl_lmc_i_f}
 \end{figure}
@@
 \centering
-\caption[Simulation results for Sculptor’s stars in the LMC+MW potential]{Similar to Table \ref{tbl:scl_sim_stars_results}, but for the properties of the stellar components of the \texttt{LMC-flyby} model of Sculptor.
+\caption[Simulation results for Sculptor’s stars in the MW+LMC potential]{Similar to Table \ref{tbl:scl_sim_stars_results}, but for the properties of the stellar components of the \texttt{LMC-flyby} model of Sculptor.
 }
@@
 matter from the outskirts of Scl. The central cusp and compact stellar
-distribution are resilient to tides and any tidal effects are
-well-outside the reach of current observations. We have also found that
+distribution are resilient to tides. Any tidal effects would also be
+well outside the reach of current observations. We have also found that
 the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
@@
 \caption[Ursa Minor simulation snapshots]{Similar to
-Fig.~\ref{fig:scl_sim_images} except for Ursa Minor on the \smallperi{}
+Fig.~\ref{fig:scl_sim_images} but for Ursa Minor on the \smallperi{}
 orbit. Dark matter evolution is more dramatic than for
@@
 \caption[Ursa Minor tidal tracks]{Similar to
-Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
-substantially more mass than Sculptor. \textbf{todo: thinner and
-truncated tidal track, thinner stellar
-component}}\label{fig:umi_tidal_track}
+Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor loses
+substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
 \end{figure}
 
-The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
-potential. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM
-evolution. UMi loses significantly more DM mass than Scl, forming
+The tidal evolution of Ursa Minor is similar to that of Sculptor in the
+MW-only potential. Fig.~\ref{fig:umi_sim_images} shows snapshots of the
+DM evolution. UMi loses significantly more DM mass than Scl, forming
 substantial dark matter streams encircling the MW several times.
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_smallperi_i_f.pdf}}
-\caption[Ursa Minor simulated density profiles]{The tidal effects on the
-stellar surface density of Ursa Minor for exponential stars on the
-\smallperi{} orbit.}\label{fig:umi_smallperi_i_f}
+\caption[Ursa Minor simulated density profiles]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f}: the tidal effects on the stellar
+surface density of Ursa Minor for exponential stars on the \smallperi{}
+orbit.}\label{fig:umi_smallperi_i_f}
 \end{figure}
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/umi_plummer_i_f.pdf}}
-\caption[Ursa Minor Plummer model density]{The tidal effects on the
-stellar surface density of Ursa Minor for Plummer stars on the
-\smallperi{} orbit.}\label{fig:umi_plummer_i_f}
+\caption[Ursa Minor Plummer model density]{Similar to
+Fig.~\ref{fig:scl_smallperi_i_f}: the tidal effects on the stellar
+surface density of Ursa Minor for Plummer stars on the \smallperi{}
+orbit.}\label{fig:umi_plummer_i_f}
 \end{figure}
@@
 (Fig.~\ref{fig:umi_plummer_i_f}). However, because there are more
-loosely-bound stars, the Plummer model looses nearly 4\% of its initial
+loosely-bound stars, the Plummer model loses nearly 4\% of its initial
 stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}), and tidal
@@
 and without (green) an LMC. The final positions of Ursa Minor and the
-LMC are plotted as scatter points and the solid blue line represents the
-LMC trajectory. Note that the LMC mostly increases Ursa Minor's
+LMC are plotted as scatter points, and the solid blue line represents
+the LMC trajectory. Note that the LMC mostly increases Ursa Minor's
 pericentres and apocentres.}\label{fig:umi_orbits_lmc}
@@
 the orbit of Ursa Minor. Predominantly, the effect is to increase the
-orbital period, apocentre and pericentre. Yet, the orbit remains in a
+orbital period, apocentre, and pericentre. Yet, the orbit remains in a
 similar plane and with similar morphology. As UMi is on the opposite
-side as the Galaxy of the LMC, and has a closest LMC approach of
+side of the Galaxy of the LMC, and has a closest LMC approach of
 \(\gtrsim 100\,\kpc\), this is not surprising.
@@
 reflex motion of the Milky Way. Because the MW centre is accelerated
-towards the LMC and away from UMi, UMi's orbit to increase in
+towards the LMC and away from UMi, UMi's orbit increases in
 characteristic radius.
@@
 
-\subsection{Modelling assumptions}\label{modelling-assumptions}
+\subsection{Halo structure}\label{halo-structure}
 
 As the above results show, tides only marginally affected the stellar
-components of Scl and UMi, even orbit chosen to have the smallest
+components of Scl and UMi, even with orbits chosen to have the smallest
 observationally-consistent pericentres. While we have only presented
@@
 So, gas dynamics are unlikely to affect recent evolution. A
-collisionless dark-matter only simulation should therefore be an
+collisionless dark-matter-only simulation should therefore be an
 excellent approximation.
@@
 Cored or less concentrated dark matter halos disrupt quicker
-\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo in particular is
+\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo, in particular, is
 among the least concentrated halos consistent with UMi's velocity
@@
 
-Alternative initial conditions may influence the total mass evolution
-but should produce a similar final stellar structure. A system's
-observed velocity dispersion directly constrains the mean density within
-\(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force required to
-disrupt the stellar component does not strongly depend on the inner halo
-structure (via the Jacobi radius).
+While alternative initial conditions may influence the total mass
+evolution, they should produce a similar final stellar structure. A
+system's observed velocity dispersion directly constrains the mean
+density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
+required to disrupt the stellar component does not strongly depend on
+the inner halo structure (via the Jacobi radius).
 
-\subsection{Orbital uncertainties}\label{orbital-uncertainties}
+\subsection{Orbital uncertainties}\label{sec:scl_umi_orbit_uncert}
 
@@
 satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the
-Milky Way and dynamical friction imply that orbits are typically less
+Milky Way and dynamical friction imply that orbits were typically less
 bound in the past (less affected by tides). Orbital energy and angular
@@
 As an example, Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates how
-changes to the LMC potential modifies the long-term orbital trajectories
-of Scl and UMi. More than 4 Gyr ago, orbits of Scl diverge
+changes to the LMC potential modify the long-term orbital trajectories
+of Scl and UMi. More than 4 Gyr ago, the orbits of Scl diverge
 substantially. Some orbits are near apocentres of \(\sim 300\,\kpc\)
@@
 \pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
-\caption[Sculptor Orbits with LMC]{The long-term orbital history of
-Sculptor (\textbf{top}) and UMi (\textbf{bottom}) are uncertain. In both
-panels, light, transparent lines represent randomly-sampled orbit of the
-satellites (alla ref) in three different LMC/MW mass models from
-\citet{vasiliev2024}. The LMC orbits are in solid, thick lines of the
-corresponding colour. The L2M11 has a lighter LMC mass, and the L3M10
-model has a lighter MW mass than our fiducial L3M11 LMC
-model.}\label{fig:scl_orbit_lmc_uncert}
+\caption[Long term orbital uncertainties]{The long-term orbital history
+of Sculptor (\textbf{top}) and UMi (\textbf{bottom}) is uncertain. In
+both panels, light, transparent lines represent randomly sampled orbits
+of the satellites (following Section~\ref{sec:scl_smallperi}) in three
+different LMC/MW mass models from \citet{vasiliev2024}. The LMC orbits
+are in solid, thick lines of the corresponding colour. The L2M11 has a
+lighter LMC mass, and the L3M10 model has a lighter MW mass than our
+fiducial L3M11 LMC model.}\label{fig:scl_orbit_lmc_uncert}
 \end{figure}

### 2025-10-08 thesis/compile.sh

@@ -8,6 +8,8 @@ pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
 pandoc $pandocoptions -o results.tex results.md
 pandoc $pandocoptions -o discussion.tex discussion.md
+pandoc $pandocoptions -o acknowledgments.tex acknowledgments.md
+pandoc $pandocoptions -o research_acknowledgments.tex research_acknowledgments.md
 
 pandoc $pandocoptions -o extra_density.tex extra_density.md --top-level-division=chapter
 pandoc $pandocoptions -o extra_rv_models.tex extra_rv_models.md --top-level-division=chapter

### 2025-10-14 thesis/methods.tex
julio final edits

@@
-In the previous chapters, we have seen that the Scl and UMi classical
+In the previous Chapters, we have seen that the Scl and UMi classical
 dSphs have outer density profiles that appear to deviate from the
@@
 velocity, as given in Table~\ref{tbl:scl_obs_props}. All sampled orbits
-of Sculptor have nearly the same morphology---the orbit primarily
-resides in the \(y\)--\(z\) plane and completes a similar number of
-periods and pericentric and apocentric passages.
+of Sculptor have nearly the same shape---the orbit primarily resides in
+the \(y\)--\(z\) plane and completes a similar number of periods and
+pericentric and apocentric passages.
 
@@
 Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
-defined by \citet[eq.\textasciitilde13]{power+2003} for NFW halos formed
-in cosmological N-body simulations. This radius marks the region where
+defined by \citet[their eq. 13]{power+2003} for NFW halos formed in
+cosmological N-body simulations. This radius marks the region where
 collisional effects driven by the finite number of particles used to
@@
 \citet{power+2003}. First, we start with an initial centre estimate
-using all bound particles from the last timestep. Then, we calculate the
-radius of all particles from the centre, remove particles with a radius
-beyond the 0.975 quantile of the centre, and recalculate the centre of
-mass. The procedure is repeated until the selection radius is less than
-\textasciitilde1kpc or fewer than 0.1\% of particles remain. After a
-centre has been chosen, we remove all unbound particles based on the
-\gadget{} calculated potential of the halo. For all future timesteps, we
-consider only particles retained from the previous iteration.
+using all bound particles from the previous snapshot. Then, we calculate
+the distance of all particles from the centre, remove particles with a
+distance beyond the 0.975 quantile of the centre, and recalculate the
+centre of mass. The procedure is repeated until the selection radius is
+less than \textasciitilde1kpc or fewer than 0.1\% of particles remain.
+After a centre has been chosen, we remove all unbound particles based on
+the \gadget{} calculated potential of the halo. For all future
+timesteps, we consider only particles retained from the previous
+iteration.
 
@@
 particle) isolation run is of order 0.003 kpc, but fluctuations are
-observed of \(\sim 0.03\,\kpc\). This is about three times the softening
-length but is less than the numerically converged radial scale.
+observed of order \(\sim 0.03\,\kpc\). This is about three times the
+softening length but is less than the numerically converged radial
+scale.
 
@@
 velocity profiles of stars and dark matter, demonstrating that these
-galaxies are stable in isolation and the stellar component would
-contribute negligibly to the total mass (\(M \sim \V_{\rm circ}^2 R\)).
+galaxies are stable in isolation and that the stellar component
+contributes negligibly to the total mass (\(M \sim \V_{\rm circ}^2 R\)).
 Our cosmologically-motivated, observationally-based, and

### 2025-10-14 thesis/results.tex
julio final edits

@@
 orbit, yet the resulting tidal effects are still too weak to account for
-the extended outer profiles. Our simulations demonstrate that recent
-tides are unlikely to have altered the stellar structure of Scl or UMi.
+the extended outer profiles. Our simulations thus demonstrate that
+recent tides are unlikely to have altered the stellar structure of Scl
+or UMi.
 
@@
 N-body models may deviate from a point-particle trajectory due to
-dynamical self-friction \citep[e.g.,][]{white1983, miller+2020}.
+dynamical ``self-friction'' \citep[e.g.,][]{white1983, miller+2020}.
 However, this effect is slight for Scl, which ends near the observed
@@
 initial mass (see Table~\ref{tbl:scl_sim_results}). However, the inner
-structure is not expected to be affected, as the Jacobi radius is over
-3kpc, outside of the initial and final \(\rmax\) (see
-Table~\ref{tbl:scl_sim_results} and Fig.~\ref{fig:scl_tidal_track}).
-Thus, tides may remove significant amounts of mass, but mostly from the
-outer halo.
+structure is not affected, as the Jacobi radius is over 3kpc, outside of
+the initial and final \(\rmax\) (see Table~\ref{tbl:scl_sim_results} and
+Fig.~\ref{fig:scl_tidal_track}). Thus, tides may remove significant
+amounts of mass, but mostly from the outer halo.
 
@@
 emerge. The total stellar mass lost corresponds to \(\sim 10\) stars in
-total (see Table~\ref{tbl:scl_sim_results})---a formidable challenge
-with the best of observations.
+total (see Table~\ref{tbl:scl_sim_results})---a formidable challenge to
+detect even with the best of observations.
 
@@
 choosing a different initial stellar profile which matches the observed
-profile and assessing how it evolves on the smallperi orbit. We show in
-Fig.~\ref{fig:scl_smallperi_plummer_i_f} that a Plummer profile (instead
-of an exponential) provides an adequate fit to Scl's observed profile.
-The Plummer model loses more stellar mass and forms more luminous tidal
-tails. Observations reaching surface densities \(\sim10\) times fainter
-than our data could reveal a stream in this case. Nevertheless, over the
-radial extent probed by our data, the stellar profile remains nearly
-unchanged by tidal evolution.
+profile and assessing how it evolves on the \smallperi{} orbit. We show
+in Fig.~\ref{fig:scl_smallperi_plummer_i_f} that a Plummer profile
+(instead of an exponential) provides an adequate fit to Scl's observed
+profile. The Plummer model loses more stellar mass and forms more
+luminous tidal tails. Observations reaching surface densities \(\sim10\)
+times fainter than our data could reveal a stream in this case.
+Nevertheless, over the radial extent probed by our data, the stellar
+profile remains nearly unchanged by tidal evolution.
 
-The Jacobi and break radii further support that tidal effects should not
-be apparent in the observed stellar component. As calculated for this
-model (see
+The Jacobi and break radii further support the conclusion that tidal
+effects should not be apparent in the observed stellar component. As
+calculated for this model (see
 Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
@@
 Since no orbits of Scl produce significantly smaller break or Jacobi
-radii, it is unlikely any orbit would produce an observable density
+radii, it is unlikely tha any orbit would produce an observable density
 excess.
@@
 With a mass up to one fifth of the MW \citep[e.g.,][]{penarrubia+2015},
-the LMC infall affects conclusions about the MW properties and the
-orbits of satellites \citep[see
+the LMC infall affects the MW properties and the orbits of satellites
+\citep[see
 e.g.,][]{patel+2020, battaglia+2022, correamagnus+vasiliev2022}. In this
@@
 approach to the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term
-orbit---Scl is inferred to reach an apocentre of nearly \(300\,\kpc\).
-Scl may even be on first infall, depending on the MW and LMC mass. Scl's
-is orbiting the Milky Way on a similar plane to the LMC, but in the
-opposite direction---Scl is unlikely to be an LMC satellite.
+orbit---Scl is inferred to have reach an apocentre of nearly
+\(300\,\kpc\). According to this, Scl may even be on first infall,
+depending on the MW and LMC mass. Scl's is orbiting the Milky Way on a
+similar plane to the LMC, but in the opposite direction---thus, Scl is
+unlikely to be an LMC satellite.
 
-Promisingly, the timing of the LMC encounter implies a break radius
+Interestingly, the timing of the LMC encounter implies a break radius
 (\(\sim 25'\), from Table~\ref{tbl:scl_lmc_sim_stars}) consistent with
@@
 Section~\ref{sec:data_density_profiles}). To probe this further, we
-select an orbit with the smallest LMC-Scl pericentre (20,kpc) in the
-\texttt{L3M11} model, consistent with Scl's present-day positions and
-velocities. The orbit is selected following the procedure in
+select an orbit with the smallest LMC-Scl pericentre (\(20\,\kpc\)) in
+the \texttt{L3M11} model, consistent with Scl's present-day position and
+velocity. The orbit is selected following the procedure in
 Section~\ref{sec:orbital_estimation} (with uncertainties doubled). This
@@
 and the passage of the LMC. With only one MW pericentre, Scl's dark
-matter is less disrupted than the previous MW-only model. The subsequent
-LMC passage modifies Scl's orbit but has otherwise little effect. The
-dark matter structure evolves mildly and \(\sim 50\%\) of mass remains
-bound (Table~\ref{tbl:scl_lmc_sims}).
+matter is less disrupted than in the previous MW-only model. The
+subsequent LMC passage modifies Scl's orbit but has otherwise little
+effect. The dark matter structure evolves mildly and \(\sim 50\%\) of
+mass remains bound (Table~\ref{tbl:scl_lmc_sims}).
 
@@
 matter from the outskirts of Scl. The central cusp and compact stellar
-distribution are resilient to tides. Any tidal effects would also be
-well outside the reach of current observations. We have also found that
-the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
+distribution are resilient to tides. Any tidal effects are predicted to
+be well outside the reach of current observations. We have also found
+that the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
 first infall. However, with only 1 pericentre each for the LMC and MW,
@@
 Section~\ref{sec:orbit_corrections}, we recover nearly exactly the
-present-day position of Ursa Minor by changing the initial positions by
-\(20\,\kpc\) and \(\sim 9\,\kms\). These adjustments do not
-significantly affect the qualitative structure or pericentre of the
-orbit.
+present-day position of Ursa Minor by changing the initial
+position/velocity by \(20\,\kpc\) and \(\sim 9\,\kms\). These
+adjustments do not significantly affect the qualitative structure or
+pericentre of the orbit.
 
@@
 \toprule
-Property & smallperi-exp & smallperi-Plummer\\
+Property & \smallperi{}-exp & \smallperi{}-Plummer\\
 \midrule
@@
 (Table~\ref{tbl:umi_sim_stars_results}). For exponential initial
-conditions, tidal effects are unlikely to ever be observable in the near
+conditions, tidal effects are unlikely to be observable in the near
 future.
@@
 As for Scl (Section~\ref{sec:scl_sim_stars}), we also consider a model
-where UMi's stars are initially a more extended Plummer profile,
-resembling the present-day density profile. The stellar evolution of
-this Plummer stellar component is similar
-(Fig.~\ref{fig:umi_plummer_i_f}). However, because there are more
-loosely-bound stars, the Plummer model loses nearly 4\% of its initial
-stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}), and tidal
-features may be detectable if we measure densities 2 orders of magnitude
-fainter than our present data. We show the properties of a stream in the
-Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a stream is
-unlikely to be observable in the near future. We conclude that tides do
-not strongly affect the stellar component of this model.
+where UMi's stars follow initially a Plummer profile, resembling the
+present-day density profile. The stellar evolution of this Plummer
+stellar component is similar (Fig.~\ref{fig:umi_plummer_i_f}). However,
+because there are more loosely-bound stars, the Plummer model loses
+nearly 4\% of its initial stellar mass to tides
+(Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
+detectable if we measure densities 2 orders of magnitude fainter than
+our present data. We show the properties of a stream in the Appendix
+(Fig.~\ref{fig:umi_tidal_stream}), but such a stream is unlikely to be
+observable in the near future. We conclude that tides do not strongly
+affect the stellar component of this model.
 
@@
 orbital period, apocentre, and pericentre. Yet, the orbit remains in a
-similar plane and with similar morphology. As UMi is on the opposite
-side of the Galaxy of the LMC, and has a closest LMC approach of
+similar plane and with similar shape. As UMi is on the opposite side of
+the Galaxy of the LMC, and has a closest LMC approach of
 \(\gtrsim 100\,\kpc\), this is not surprising.
@@
 
-As the above results show, tides only marginally affected the stellar
+As the above results show, tides only marginally affect the stellar
 components of Scl and UMi, even with orbits chosen to have the smallest
@@
 density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
-required to disrupt the stellar component does not strongly depend on
-the inner halo structure (via the Jacobi radius).
+required to disrupt the stellar component depends mainly on the total
+mass inside the stellar half-mass radius, and less on the actual shape
+of the density profile of the dark matter inside that radius.
 
@@
 
-The long-term orbits of satellites are uncertain. Analytic Milky Way
-potentials neglect many unknown details, including triaxiality, halo
-twisting, mass evolution, and substructure. Due to these inadequacies,
-calculated orbits may diverge significantly from the true orbits of
-satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the
-Milky Way and dynamical friction imply that orbits were typically less
-bound in the past (less affected by tides). Orbital energy and angular
-momentum of subhalos are not conserved in cosmological N-body
-simulations. Consequently, orbits in analytic potentials may
-overestimate the pericentre and underestimate the maximum tidal stress
-\citep[although typically not by enough to change our
+The long-term orbital evolution of satellites are uncertain. Analytic
+Milky Way potentials neglect many unknown details, including
+triaxiality, mass evolution, and substructure. Due to these
+inadequacies, calculated orbits may diverge significantly from the true
+orbits of satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth
+of the Milky Way and dynamical friction imply that orbits were typically
+less bound in the past. Orbital energy and angular momentum of subhalos
+are not conserved in cosmological N-body simulations. Consequently,
+orbits in analytic potentials may overestimate the pericentre and
+underestimate the maximum tidal stress \citep[although typically not by
+enough to change our
 conclusions,][]{santistevan+2023, santistevan+2024}.

### 2025-10-17 orbits/orbit_utils.jl
minor plot adjustments

@@ -52,8 +52,11 @@ function orbital_properties(pot, orbits; agama_units, actions=false)
     
     @info "calculating orbit timescales"
     periods_errs = orbit_period.(orbits)
-    periods = first.(periods_errs)
-    errs = last.(periods_errs)
+    periods = [p[1] for p in periods_errs]
+    errs = [p[2] for p in periods_errs]
+    period_apos = [p[3] for p in periods_errs]
+    n_peris = [p[4] for p in periods_errs]
+
     max_err = maximum(errs)
     @info "max pericentre err: $max_err"
 
@@ -76,12 +79,14 @@ function orbital_properties(pot, orbits; agama_units, actions=false)
     properties = DataFrame(
         "pericentre" => peris,
         "apocentre" => apos,
+        "n_peris" => n_peris,
         "time_last_peri" => t_last,
         "Lx_f" => L[1, :],
         "Ly_f" => L[2, :],
         "Lz_f" => L[3, :],
         "E_f" => E, 
         "period" => periods,
+        "period_apo" => period_apos,
         "dt" => dt,
         "peri_apo_max_err" => errs,
         "x_i" => [orbit.positions[1, end] for orbit in orbits],
@@ -109,10 +114,18 @@ end
 
 
 function orbit_period(orbit)
-    _, idxs, _, _, err = LilGuys.all_peris_apos(orbit)
+    peris, idxs, apos, idx_apos, err = LilGuys.all_peris_apos(orbit)
     period = LilGuys.mean(diff(orbit.times[idxs]))
-    return period, err
+    if length(idx_apos) > 1
+        period_apo = diff(orbit.times[idx_apos])[1] # since most recent is lower index
+    else
+        period_apo = NaN
+    end
+
+    n_peris = length(idxs)
+    return period, err, period_apo, n_peris
 end
+
 
 to_sym_mat(x) = [x[1] x[4] x[6] 
 				x[4] x[2] x[5]

### 2025-10-17 orbits/ursa_minor/EP2020/run.sh

@@ -1,0 +1,1 @@
+julia ../../calc_orbit_props.jl -N 100000 --seed 496  --galaxy ursa_minor .

### 2025-10-17 thesis/plots/cored_model.jl

@@ -45,6 +45,15 @@ function get_profiles(modelname)
 	return mass_profs[1].second, mass_profs[idx_f].second
 end
 
+# ╔═╡ 04794a51-ee18-4de9-86a7-19f350148f08
+"sculptor/1e6_Ms0.54_rs1.08_c1/orbit_smallperi"
+
+# ╔═╡ 578fa6ed-3d63-49cf-af74-5c927f38dd28
+modelnames = TOML.parsefile("model_key.toml")
+
+# ╔═╡ 0337d33f-d46f-446f-84fa-43995bd3ef3c
+galaxyname, modelname, starsname = modelnames["cored"]
+
 # ╔═╡ cb38048e-99d9-431c-ae90-de8817e02ff7
 let
 	fig = Figure(size=(4*72, 3*72))
@@ -58,7 +67,7 @@ let
 	y = log10.(LilGuys.density.(h, 10 .^ x))
 	lines!(x, y, linewidth=theme(:linewidth)[]/2, label="NFW")
 
-	profs = get_profiles("sculptor/1e6_Ms0.54_rs1.08_c1/orbit_smallperi")
+	profs = get_profiles(joinpath(galaxyname, modelname))
 
 	lines!(profs[1].log_r, LilGuys.log_densities(profs[1]), linestyle=:dot, color=COLORS[2], label="cored initial")
 	lines!(profs[2].log_r, LilGuys.log_densities(profs[2]), color=COLORS[2], label="cored final")
@@ -79,4 +88,7 @@ end
 # ╠═5eaf3b50-886e-47ac-9a7c-80d693bc3c17
 # ╠═8081e889-c1d4-4e50-b8e1-91aa8e82adf1
 # ╠═f49c9325-4319-4435-ac33-a8980e1c4f7f
+# ╠═04794a51-ee18-4de9-86a7-19f350148f08
+# ╠═578fa6ed-3d63-49cf-af74-5c927f38dd28
+# ╠═0337d33f-d46f-446f-84fa-43995bd3ef3c
 # ╠═cb38048e-99d9-431c-ae90-de8817e02ff7

### 2025-10-17 thesis/plots/cosmological_means.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.18
+# v0.20.19
 
 using Markdown
 using InteractiveUtils
@@ -64,7 +64,7 @@ smallfontsize = @lift 0.8 * $(theme(:fontsize))
 
 # ╔═╡ 3c032178-8d48-4f9c-bcec-9bf704718ea9
 @savefig "cosmological_means" let
-	fig = Figure(figure_padding=10, size=( 5.4*72, 2.0*72,))
+	fig = Figure(figure_padding=10, size=( 6*72, 2.5*72,))
 
 
 
@@ -107,7 +107,7 @@ smallfontsize = @lift 0.8 * $(theme(:fontsize))
 				yminorticks = 10:10:100,
 				xticks=[1, 10, 100],
 				ylabel = L"$\textrm{v}_\textrm{max}$ / km\,s$^{-1}$",
-			   xlabel = L"$\textrm{r}_\textrm{max}$",
+			   xlabel = L"$\textrm{r}_\textrm{max}$ / kpc",
 						 limits=(1, 50, 10, 100)
 			  )
 
@@ -158,7 +158,10 @@ smallfontsize = @lift 0.8 * $(theme(:fontsize))
 		  space=:relative, fontsize=10)
 
 	scatter!(LilGuys.v_circ_max(halo_initial) * V2KMS, M_s_initial, marker=:star5, color=COLORS[5], markersize=markersize)
-	
+
+	rowsize!(fig.layout, 1, Aspect(1, 1))
+
+	resize_to_layout!()
 	fig
 end
 

### 2025-10-17 thesis/plots/density_comparison.jl

@@ -50,8 +50,59 @@ module ModelUtils
 	include("model_utils.jl")
 end
 
+# ╔═╡ 7948bbd7-2dda-46c8-a850-d80944ba9096
+modelnames = TOML.parsefile("model_key.toml")
+
 # ╔═╡ c2eed577-ff2c-4a28-b043-3ba7f4a68d3a
 compare_both = ModelUtils.compare_both
+
+# ╔═╡ 3317cfec-be70-418c-944a-0a6741f03cf3
+@savefig "scl_smallperi_i_f" compare_both(modelnames["scl_smallperi"]...,  r_j=true, norm_shift=0.2, 	break_height=-3, title="Sculptor: smallperi-exponential")
+
+# ╔═╡ fc45829c-bf28-4d7b-9bea-e223e4b29e7d
+@savefig "scl_plummer_i_f" compare_both(modelnames["scl_smallperi_plummer"]..., norm_shift=0,  r_j=true, break_height=-2, title="Sculptor: smallperi-Plummer")
+
+# ╔═╡ 5f989e91-8870-462e-a363-767be64dac5c
+@savefig "scl_lmc_i_f" compare_both(modelnames["scl_lmc"]..., norm_shift=0.1, lmc=true, r_j=true, break_height=-7.590274001669055, title="Sculptor: LMC–exponential")
+
+# ╔═╡ e9a0ce9f-27a5-47dc-8c4d-22887f6a7fc1
+@savefig "scl_lmc_plummer_i_f" compare_both(modelnames["scl_lmc_plummer"]..., norm_shift=0.1, lmc=true, r_j=true, break_height=-7.590274001669055, title="Sculptor: LMC–Plummer")
+
+# ╔═╡ 77c057e4-78da-4201-b9f8-cb3135508a93
+md"""
+# Ursa Minor
+"""
+
+# ╔═╡ 3263677b-e0e3-4d30-8e28-309de99a8603
+@savefig "umi_smallperi_i_f" compare_both(modelnames["umi_smallperi"]...,  r_j=true, norm_shift=0.0, 	break_height=-5.5, title="Ursa Minor: smallperi-exponential")
+
+# ╔═╡ d0ce116f-aa51-43d8-9748-806eb44b8409
[…]

### 2025-10-17 thesis/plots/density_profiles_lit_comparison.jl

@@ -344,10 +344,6 @@ function compare_profiles(profiles, galaxyname; kwargs...)
 	return fig
 end
 
-# ╔═╡ 1673a2ce-0a0a-4564-b7f2-8ac9cff83b81
-let
-	prof_scl_eskridge.log_R[1:2:end] + prof_scle
-
 # ╔═╡ 4314ec9e-8359-4b2a-8eea-30f71d92b743
 compare_profiles(profiles, "sculptor")
 
@@ -468,17 +464,20 @@ all_profiles_scl = merge(profiles, profiles_old)
 
 # ╔═╡ a9194007-e598-49a3-b462-a1710ebab5b2
 @savefig "scl_umi_literature_profiles" let
-	fig = Figure(size=(5.39*72, 4*72))
+	fig = Figure(size=(5*72, 7*72))
 	ax1 = compare_profiles(fig[1,1], all_profiles_scl, "sculptor", title="Sculptor")
 	ax2 = compare_profiles_res(fig[2,1], all_profiles_scl, "sculptor")
+	# hidexdecorations!(ticks=false, minorticks=false)
 
-	ax3 = compare_profiles(fig[1, 2], profiles_umi, "ursa_minor", title="Ursa Minor")
-	hideydecorations!(ticks=false, minorticks=false)
-	ax4 = compare_profiles_res(fig[2, 2], profiles_umi, "ursa_minor",)
-	hideydecorations!(ticks=false, minorticks=false)
+	ax3 = compare_profiles(fig[3, 1], profiles_umi, "ursa_minor", title="Ursa Minor")
+	ax4 = compare_profiles_res(fig[4, 1], profiles_umi, "ursa_minor",)
+	# hideydecorations!(ticks=false, minorticks=false)
 
-	rowsize!(fig.layout, 2, Relative(0.2))
-	rowgap!(fig.layout, 0)
+	rowsize!(fig.layout, 2, Relative(0.1))
+	rowsize!(fig.layout, 4, Relative(0.1))
+
+	rowgap!(fig.layout, 1, 0)
+	rowgap!(fig.layout, 3, 0)
 	colgap!(fig.layout, 0)
 
 
@@ -535,7 +534,6 @@ end
 # ╠═6c04f96a-b14b-4c60-91de-716c9379491f
[…]

### 2025-10-17 thesis/plots/extra_density_profiles.jl

@@ -174,29 +174,38 @@ md"""
 """
 
 # ╔═╡ 5917e45e-50bb-441f-9d12-6901025a81f3
-profiles_scl_j24 = OrderedDict(
-	"2-exp" => load_profile("sculptor", "jax_2c_eqw") |> LilGuys.filter_empty_bins,
-	"1-exp" => load_profile("sculptor", "jax_eqw") |> LilGuys.filter_empty_bins,
-	"simple" => load_profile("sculptor", "simple") |> LilGuys.filter_empty_bins,
-	# "PM+CMD" => load_profile("sculptor", "jax_LLR_0_eqw") |> LilGuys.filter_empty_bins,
+profiles_scl_j24 = let
+	profs = OrderedDict(
+		"2-exp" => load_profile("sculptor", "jax_2c_eqw") |> LilGuys.filter_empty_bins,
+		"1-exp" => load_profile("sculptor", "jax_eqw") |> LilGuys.filter_empty_bins,
+		# "PM+CMD" => load_profile("sculptor", "jax_LLR_0_eqw") |> LilGuys.filter_empty_bins,
+		"circ" => load_profile("sculptor", "jax_circ_eqw") |> LilGuys.filter_empty_bins,	
+		"bright" => load_profile("sculptor", "jax_bright") |> LilGuys.filter_empty_bins,
+	
+	)
 
-)
+
+	scale!(profs, "bright", 1, 2)
+
+	profs
+end
 
 # ╔═╡ 66761f62-bfe3-4c5e-8630-17a8b63721e7
 begin 
 	profiles_scl_extra = OrderedDict(
 		"2-exp" => load_profile("sculptor", "jax_2c_eqw") |> LilGuys.filter_empty_bins,
-		"circ" => load_profile("sculptor", "jax_circ_eqw") |> LilGuys.filter_empty_bins,	
+	"simple" => load_profile("sculptor", "simple") |> LilGuys.filter_empty_bins,
 
-		"bright" => load_profile("sculptor", "jax_bright") |> LilGuys.filter_empty_bins,
 		"DELVE" => load_profile("sculptor", "delve_new_sub"),
 	)
 
-	scale!(profiles_scl_extra, "bright", 1, 2)
 	scale!(profiles_scl_extra, "DELVE", 1, 0.2)
 
 	profiles_scl_extra
 end
+
[…]

### 2025-10-17 thesis/plots/galaxy_pictures.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.18
+# v0.20.19
 
 using Markdown
 using InteractiveUtils
@@ -69,6 +69,9 @@ function plot_ellipse!(galaxy)
 
 	Utils.ellipse!(R_h_rel, obs_props["ellipticity"], obs_props["position_angle"], color=(:white, 0.5), linewidth=theme(:linewidth)[]/2)
 end
+
+# ╔═╡ 10bffed3-7a8b-43b0-874f-3086631e471d
+Utils.get_obs_props("lmc")
 
 # ╔═╡ 3c032178-8d48-4f9c-bcec-9bf704718ea9
 @savefig "galaxy_pictures" let
@@ -159,4 +162,5 @@ end
 # ╠═00c8ab2a-08e8-467e-a9ba-1793d9af8ea5
 # ╠═2188831f-d8ed-4a37-ad92-bc1f0d90ebbc
 # ╠═9f38819d-7ec1-4fe0-b3d3-b628b2bf34cb
+# ╠═10bffed3-7a8b-43b0-874f-3086631e471d
 # ╠═3c032178-8d48-4f9c-bcec-9bf704718ea9

### 2025-10-17 thesis/plots/initial_vcirc.jl

@@ -28,6 +28,9 @@ import Agama
 # ╔═╡ b74a4535-1e12-422e-a524-135db4d8a9f7
 import TOML
 
+# ╔═╡ d41f4781-5063-48c0-aade-fcd3980e19e7
+modelnames = TOML.parsefile("model_key.toml")
+
 # ╔═╡ 20c338e3-3d10-41f4-b8ee-d0eda4e755bd
 CairoMakie.activate!(type=:png)
 
@@ -35,6 +38,11 @@ CairoMakie.activate!(type=:png)
 module Utils
 	include("utils.jl")
 end
+
+# ╔═╡ 43dad299-d8d2-4146-8320-c90a62f3a3f0
+md"""
+# Data loading
+"""
 
 # ╔═╡ 0dbc4770-fd1d-4a57-9825-3576900dd7a3
 function load_vcircs(galaxyname, haloname)
@@ -49,31 +57,20 @@ function load_vcircs(galaxyname, haloname)
 	return prof_i, prof_f
 end
 
-# ╔═╡ 5a4853a9-1060-4be8-a872-c78297ed586f
-function assign_weights!(snap, df)
-	snap.weights = df.probability[snap.index]
-end
-
-# ╔═╡ b7572e12-85a0-4f80-b7aa-ab2db8a81a13
-function get_M_star(galaxyname)
-	obs_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "observations", galaxyname, "observed_properties.toml"))
-	M_star = LilGuys.mag_to_L(obs_props["Mv"]) * obs_props["M_L_s"]
-	return M_star / M2MSUN
-end
-
 # ╔═╡ 890fa950-ca5e-4bec-9ef2-a2e5b3909909
-function load_stars(galaxyname, haloname, starsname)
+function load_stars(galaxyname, modelname, starsname)
+	haloname = joinpath(modelname, "..")
 	stars_dir = joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, haloname, "stars")
 
 	df_probs = LilGuys.read_hdf5_table(joinpath(stars_dir, starsname, "probabilities_stars.hdf5"))
 
 	snap_i = Snapshot(joinpath(stars_dir, "iso_paint.hdf5"))
[…]

### 2025-10-17 thesis/plots/sim_pictures.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.18
+# v0.20.19
 
 using Markdown
 using InteractiveUtils
@@ -28,6 +28,9 @@ galaxyname = "ursa_minor"
 # ╔═╡ bf49209c-fbfc-4439-a7d8-cfad5ceba8cc
 import TOML
 
+# ╔═╡ 280e0cef-896a-4c14-a1ac-a611418be57e
+modelnames = TOML.parsefile("model_key.toml")
+
 # ╔═╡ 4e5ddbe9-e90a-42cf-a0f1-fabb3d3f1675
 xy_iso = CSV.read("resources/EP2020_iso_xy.csv", DataFrame, tasks=1)
 
@@ -42,9 +45,9 @@ CairoMakie.activate!(type=:png)
 
 # ╔═╡ b9600d93-5946-4380-a2ae-9b5f673bbaf5
 modelname = if galaxyname == "sculptor"
-	"sculptor/1e7_new_v31_r3.2/orbit_smallperi"
+	joinpath(modelnames["scl_smallperi"][1:2]...)
 elseif galaxyname == "ursa_minor"
-	"ursa_minor/1e7_new_v38_r4.0/orbit_smallperi.5"
+	joinpath(modelnames["umi_smallperi"][1:2]...)
 end
 
 # ╔═╡ 0f71807d-d698-4164-9f30-49af8dd8ba55
@@ -185,6 +188,7 @@ end
 # ╠═b0a0dddc-fb5a-4bb2-b048-a54859d0b703
 # ╠═bf49209c-fbfc-4439-a7d8-cfad5ceba8cc
 # ╠═f5c22abc-2634-4774-8516-fbd07aa690aa
+# ╠═280e0cef-896a-4c14-a1ac-a611418be57e
 # ╠═4e5ddbe9-e90a-42cf-a0f1-fabb3d3f1675
 # ╠═53cdcc20-96d4-4bd5-8028-df9a38af71ae
 # ╠═0f71807d-d698-4164-9f30-49af8dd8ba55

### 2025-10-17 thesis/plots/tidal_tails.jl

@@ -189,8 +189,11 @@ md"""
 # sculptors tidal tails
 """
 
+# ╔═╡ 934b1a11-2646-4860-afa3-593218a9964b
+modelnames = TOML.parsefile("model_key.toml")
+
 # ╔═╡ 9ae7b57f-7649-443e-a438-4e5bd4b4d048
-@savefig "scl_sim_stream" plot_sample("sculptor", "1e7_new_v31_r3.2/orbit_smallperi", "plummer_rs0.20")
+@savefig "scl_sim_stream" plot_sample(modelnames["scl_smallperi_plummer"]...)
 
 # ╔═╡ 189db37c-1e59-4e69-9487-8525d45a9c3c
 # plot_sample("sculptor", "1e6_new_v31_r3.2/orbit_smallperi", "plummer_rs0.25")
@@ -199,7 +202,7 @@ md"""
 # plot_sample(load_stars("sculptor", "1e7_new_v25_r2.5/smallperilmc", "plummer_rs0.20"))
 
 # ╔═╡ 7b274c9f-4cdd-41ef-bdef-8a649331d1d9
-@savefig "scl_mw_impact_stream" plot_sample("sculptor", "1e6_new_v31_r3.2/L3M11_9Gyr_smallperi.a4", "plummer_rs0.20")
+# @savefig "scl_mw_impact_stream" plot_sample(modelnames["scl_smallperi_plummer"]...)
 
 # ╔═╡ 5a56e59a-c846-4829-9ddf-496c8a0383fb
 # plot_sample(load_stars("sculptor", "1e6_new_v31_r3.2/L3M11_9Gyr_smallperi.a4", "exp2d_rs0.10"))
@@ -213,7 +216,7 @@ md"""
 """
 
 # ╔═╡ 52c9a0fc-6db3-4e67-8d01-2f0eb0c1663f
-@savefig "umi_sim_stream" plot_sample("ursa_minor", "1e7_new_v38_r4.0/orbit_smallperi.5", "plummer_rs0.20")
+@savefig "umi_sim_stream" plot_sample(modelnames["umi_smallperi_plummer"]...)
 
 # ╔═╡ b50e2128-aa69-4ef9-bd7a-a08f412c504e
 # plot_sample(load_stars("ursa_minor", "1e6_v37_r5.0/orbit_mean.2", "plummer_rs0.20"))
[…]

### 2025-10-17 thesis/plots/umi_trajectories.jl

@@ -39,7 +39,7 @@ The most important variable is to set the modelname to the appropriate directory
 """
 
 # ╔═╡ 6ca3fe17-3f13-43fe-967b-881078135ead
-modelname = "EP2020"
+
 
 # ╔═╡ 46348ecb-ee07-4b6a-af03-fc4f2635f57b
 FIGDIR = "./figures"
@@ -69,6 +69,9 @@ import TOML
 # ╔═╡ a7111062-b025-43a9-bdb1-aee08deb60e9
 CairoMakie.activate!(type=:png)
 
+# ╔═╡ fc3e1290-d33d-4a32-844a-d4fef24e503c
+modelnames = TOML.parsefile("model_key.toml")
+
 # ╔═╡ 16f4ac20-d8cf-4218-8c01-c15e04e567fb
 md"""
 # The example orbits
@@ -77,8 +80,11 @@ md"""
 # ╔═╡ 35ce583b-0938-429e-af5d-b17b399f6690
 Nmax = 100 # number of orbits to plot
 
+# ╔═╡ 891618e0-5c9c-443b-a874-34dce7660c66
+modelname = joinpath(modelnames["umi_orbits"])
+
 # ╔═╡ 15863916-6601-4f45-9f45-4cd303bbcc4d
-modeldir = joinpath(ENV["DWARFS_ROOT"], "orbits/ursa_minor", modelname)
+modeldir = joinpath(ENV["DWARFS_ROOT"], "orbits", modelname)
 
 # ╔═╡ 3e4f0aa7-22d4-4a19-aafd-cf86c380ade3
 scale_theme_element!(:linewidth, 0.5)
@@ -92,7 +98,7 @@ idx, orbits = let
 end
 
 # ╔═╡ 127a988e-74ab-41f4-8979-6800622f3ff4
-best_orbit = LilGuys.Orbit(modeldir * "_special_cases/orbit_smallperi.csv" ) |> reverse
+best_orbit = OrbitUtils.load_best_orbit(modelnames["umi_smallperi"][1:2]...) |> reverse
 
 # ╔═╡ 367eb7e4-8c54-4fa9-a01c-41c3c7ff7f2a
 other_orbits = LilGuys.Orbit.(modeldir * "_special_cases/" .* ["orbit_smallperi.csv", "orbit_largeperi.csv", "orbit_mean.csv"])
[…]

### 2025-10-17 thesis/plots/utils.jl

@@ -11,6 +11,11 @@ function get_obs_props(galaxyname)
     return TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "observations/$galaxyname/observed_properties.toml"))
 end
 
+function get_M_star(galaxyname)
+    obs_props = get_obs_props(galaxyname)
+	M_star = LilGuys.mag_to_L(obs_props["Mv"]) * obs_props["M_L_s"]
+	return M_star / M2MSUN
+end
 
 function ellipse!(radius, ellipticity, position_angle; x0=0, y0=0, kwargs...)
 	t = LinRange(0, 2π, 1000) 
@@ -64,4 +69,16 @@ end
 function log_derivative(f, x0; h=0.001)
 	y0 = f(x0)
 	return (log10(f(x0 * 10^h)) - log10(y0)) / h
+end
+
+
+
+function text_along_line_log!(x, y, x_0; text, h=0.03, kwargs...)
+	f = LilGuys.lerp(x, y)
+	y_0 = f(x_0)
+	dy = log_derivative(f, x_0, h=h)
+	rf = rotation_factor(Makie.current_axis(), true)
+	θ = @lift atan($rf * dy)
+
+	text!(x_0, y_0, text=text, rotation=θ; kwargs...)
 end

### 2025-10-17 thesis/plots/velocity_dispersions.jl

@@ -73,18 +73,18 @@ module ModelUtils
 end
 
 # ╔═╡ cd0ea281-28b5-46fe-8ca2-23e44c623da6
-function plot_density_f!(galaxyname, modelname, starsname; norm_shift=0, kwargs...)
+function plot_density_f!(galaxyname, modelname, starsname; R_shift=0, norm_shift=0, kwargs...)
 	prof_i, prof_f, norm = ModelUtils.load_stellar_profiles(galaxyname, modelname, starsname, norm_shift=norm_shift)
 
 	model_dir = joinpath(ENV["DWARFS_ROOT"], "analysis/$galaxyname/$modelname/stars/$starsname/")
 	dist_f =  TOML.parsefile(model_dir * "../../orbital_properties.toml")["distance_f"]
 	
-	lines!(prof_f.log_R, prof_f.log_Sigma; kwargs...)
+	lines!(prof_f.log_R .+ R_shift, prof_f.log_Sigma; kwargs...)
 
 end
 
 # ╔═╡ 0cc4c7b4-d707-4f3d-9a82-1b6eb4119143
-function plot_σv_prof!(galaxyname, args...; kwargs...)
+function plot_σv_prof!(galaxyname, args...; R_shift=0, kwargs...)
 	stars = get_stars_final(galaxyname, args...)
 
 
@@ -119,7 +119,7 @@ function plot_σv_prof!(galaxyname, args...; kwargs...)
 		Ns[i] = N
 	end
 
-	lines!(LilGuys.midpoints(r_bins), log10.(σs); kwargs...)
+	lines!(LilGuys.midpoints(r_bins), log10.(σs) .- 1/2 * R_shift; kwargs...)
 
 
 	
@@ -167,11 +167,11 @@ function compare_both(galaxyname, modelnames; density_kwargs = Dict())
 
 	ax = Axis(fig[1,1], xlabel="log radius / arcmin", ylabel = "log surface density", limits=(-0.5, 3, -3, 2))
 
-	ModelUtils.plot_expected_profile!("sculptor")
+	ModelUtils.plot_expected_profile!(galaxyname)
 
 	for (i, (label, model)) in enumerate(modelnames)
 		kwargs = get(density_kwargs, label, (;))
[…]

### 2025-10-17 thesis/tables/dm_model_evo.jl

@@ -1,34 +1,28 @@
-using LilGuys
 using PyFITS
 import TOML
 using Printf
+include("utils.jl")
 
-function print_dm_props(galaxyname, orbitname, modelname)
-    modeldir = joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, orbitname, modelname)
+function print_dm_props(galaxyname, modelname, starsname)
+    modeldir = joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, modelname)
 
     orb_props = TOML.parsefile(joinpath(modeldir, "orbital_properties.toml"))
     idx_f = orb_props["idx_f"]
 
     dm_scalars = read_fits(joinpath(modeldir, "profiles_scalars.fits"))
-    @printf "r_circ_f_i\t%8.2f\n" dm_scalars.r_circ_max[idx_f]  / dm_scalars.r_circ_max[1]
-    @printf "v_circ_f_i\t%8.2f\n" dm_scalars.v_circ_max[idx_f]  / dm_scalars.v_circ_max[1] 
+    @printf "r_circ_f_i\t%8.3f\n" dm_scalars.r_circ_max[idx_f]  / dm_scalars.r_circ_max[1]
+    @printf "v_circ_f_i\t%8.3f\n" dm_scalars.v_circ_max[idx_f]  / dm_scalars.v_circ_max[1] 
     @printf "boundmass_f_i\t%8.6f\n" dm_scalars.bound_mass[idx_f] / dm_scalars.bound_mass[1] 
-    #@printf "r_circ_i\t%8.2f\n" dm_scalars.r_circ_max[1] 
-    #@printf "r_circ_f\t%8.2f\n" dm_scalars.r_circ_max[idx_f] 
-    #@printf "v_circ_i\t%8.2f\n" dm_scalars.v_circ_max[1] * V2KMS
-    #@printf "v_circ_f\t%8.2f\n" dm_scalars.v_circ_max[idx_f] * V2KMS
-    #@printf "boundmass_i\t%8.6f\n" dm_scalars.bound_mass[1] 
-    #@printf "boundmass_f\t%8.6f\n" dm_scalars.bound_mass[idx_f] 
 end
 
 
 println("scl smallperi")
-print_dm_props("sculptor", "1e7_new_v31_r3.2", "orbit_smallperi")
+print_dm_props(modelnames["scl_smallperi"]...)
 
 println()
 println("scl lmc-flyby")
[…]

### 2025-10-17 thesis/tables/lmc_orbit_props.jl

@@ -1,0 +1,113 @@
+using LilGuys
+import TOML
+using PyFITS
+import StatsBase: quantile, median
+import DataFrames: disallowmissing
+using Printf
+import Agama
+
+include("utils.jl")
+
+units = Agama.VASILIEV_UNITS
+
+function calc_r_J(halo, rho_peri)
+    return 10 ^ LilGuys.find_zero(lr -> LilGuys.mean_density(halo, 10^lr) - 3*rho_peri, log10(halo.r_s))
+end
+
+
+function print_radius(label, r, distance)
+    r_arcmin = LilGuys.kpc2arcmin.(r, distance)
+    print_quantity(label * " / kpc", r)
+    print_quantity(label * " / arcmin", r_arcmin)
+end
+
+function LilGuys.mean_density(pot, r, units)
+    M =  Agama.enclosed_mass(pot, r, units) 
+    V = (4π/3 * r .^ 3)
+
+    return M ./ V
+end
+
+function get_orbit(galaxyname, modelname)
+    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, modelname, "orbital_properties.toml"))
+    orbit_props
+end
+
+function get_orbit_lmc(galaxyname, modelname)
+    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, modelname, "orbital_properties_lmc.toml"))
+    orbit_props
+end
+
+function get_samples(galaxyname, potname)
+    props = read_fits( joinpath(ENV["DWARFS_ROOT"], "orbits", 
+                                galaxyname, potname, "orbital_properties.fits"))
+end
+
+
+function print_orbit(galaxyname, modelname, starsname, halo)
+    orbit_props = get_orbit(galaxyname, modelname)
+    orbit_props_lmc = get_orbit_lmc(galaxyname, modelname)
+
+    for key in ["pericentre", "apocentre", "t_last_peri", "idx_peris", "distance_f"]
+        val = orbit_props[key]
+        if val isa Real
+            val = round(orbit_props[key], sigdigits=4)
+        end
+        println(key, "\t", val)
+    end
+
+    for key in ["pericentre", "t_last_peri"]
+        println(key, "\t", orbit_props_lmc[key])
+    end
+
+
[…]

### 2025-10-17 thesis/tables/orbit_initial.jl

@@ -1,17 +1,18 @@
 using LilGuys
 import TOML
+include("utils.jl")
 
 
 function get_orbit(galaxyname, orbitname)
     orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "orbits", galaxyname, orbitname * ".toml"))
 end
 
-function get_orbit(galaxyname, haloname, orbitname)
-    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, haloname, orbitname, "orbital_properties.toml"))
+function get_orbit(galaxyname, modelname, starsname)
+    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, modelname, "orbital_properties.toml"))
 end
 
-function get_orbit_i(galaxyname, haloname, orbitname)
-    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, haloname, orbitname, "simulation/orbit.toml"))
+function get_orbit_i(galaxyname, modelname, starsname)
+    orbit_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "analysis", galaxyname, modelname, "simulation/orbit.toml"))
 end
 
 
@@ -19,9 +20,18 @@ function print_orbit(args...)
     orbit_props = get_orbit(args...)
     orbit_props_i = get_orbit_i(args...)
 
-    for key in ["ra", "dec", "distance", "pmra", "pmdec", "radial_velocity", "x_i", "y_i", "z_i"]
+    for key in ["ra", "dec", "distance", "pmra", "pmdec", "radial_velocity"]
         println(key, "\t", round(orbit_props_i[key], sigdigits=4))
     end
+
+    println("t_i (nbody)", "\t", orbit_props["t_f_gyr"])
+    t_i = get(orbit_props_i, "t_i", NaN)
+    println("t_i (point)", "\t", t_i * T2GYR)
+
+    for key in ["x_i", "y_i", "z_i"]
+        println(key, "\t", round(orbit_props_i[key], sigdigits=4))
+    end
+
 
     if radii([orbit_props_i["v_x_i"], orbit_props_i["v_y_i"], orbit_props_i["v_z_i"]]) > 10
         v_scale = 1
@@ -33,9 +43,6 @@ function print_orbit(args...)
[…]

### 2025-10-17 thesis/tables/utils.jl

@@ -1,0 +1,53 @@
+import TOML
+using LilGuys
+import DataFrames: disallowmissing
+
+modelnames = TOML.parsefile("../plots/model_key.toml")
+
+
+function print_quantity(key, x::AbstractVector)
+    x = disallowmissing(replace(x, missing => NaN)) # skip missings
+    n_missing = LilGuys.mean(isnan.(x))
+    if n_missing > 0
+        @info "missing rate", n_missing
+    end
+
+    x = x[.!isnan.(x)]
+    if length(x) == 0
+        println(key, "\tNAN")
+        return
+    end
+
+    l, m, h = LilGuys.quantile(x, [0.16, 0.5, 0.84])
+    print_quantity(key, Measurement(m, m-l, h-m))
+end
+
+
+function print_quantity(key, m::Measurement)
+    println(key, "\t", latex_string(m))
+end
+
+function print_quantity(key, x::Real)
+    println(key, "\t", x)
+end
+
+
+function latex_string(m::Measurement)
+    lo = m.lower
+    hi = m.upper
+    mid = m.middle
+
+    # find order of magnitude of uncertainties
+    sig_lo = lo == 0 ? -3 : floor(Int, log10(lo))
+    sig_hi = hi == 0 ? -3 : floor(Int, log10(hi))
+    sig = min(sig_lo, sig_hi)
+
+    # we want 1 or 2 sig figs for uncertainty
+    round_digits = -sig + 1
+
+    lo_r = round(lo; sigdigits=2)
+    hi_r = round(hi; sigdigits=2)
+    mid_r = round(mid; digits=round_digits)
+
+    return "\$$(mid_r)^{+$(hi_r)}_{-$(lo_r)}\$"
+end

### 2025-10-17 thesis/acknowledgments.tex

@@
 
-To Raphaël, who has helped inspire and direct this project. To Isabel
-and Ryan.
+To Raphaël, who has helped inspire and direct this project. To everyone
+who has helped provide me opportunities to share work, and share
+wonderful ideas with me. Isabel, Jens, Eduardo, Jorge, David, Justin,
+Eugene, Azi, José, and many others.
 
 To Jax, who has been a support along the way, guiding my research
-methods and project. Akshara and Ewould.
+methods and project.
 
-To all of my new friends I have made along the way. Ruxin, Aviv, Anya,
-Pranav, Ted, Adrianne, Will, Maeve, and everyone else I have met. We
-have build a wonderful community of students in this department. And to
-my friends across UVic, Wyatt, Rosemary, Maggie,
+To all of my new friends I have made along the way. Ruxin, Aviv. To
+Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud. To Akshara for
+all your personal and professional advice. We have build a wonderful
+community of students in this department. To all the students who have
+since moved on and show what we may be: Bobby, Leo, Ryan, Katie,
+Fletcher, Ayden, Maheyer, Jess, Dhwani, Spencer, Mallory, Omar, Isaac,
+and Dasha.
+
+And to my friends across UVic, Wyatt, Rosemary, Maggie, and the many
+people I am still meeting.
+
+To the Linguistics extended family and team \emph{adequate}. Melody,
+Lily, Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are
+all wonderful human beings, cooks, and always a joy to spend time with.
+
+To my musical friends, Adrienne, Alex for our shared joy in music
+making. And Steve, for showing me the joy in a simple life and beautiful
+music.
+
+To the Victoria performing arts community and Fierce Brosnan, our local
+legend.
+
+To my fellow crafters, knitters, and crochet friends. Leise, Wes, Anne,
+Anya, Ashley, Sabrina, Sarah. To the graduate student society for
+continuing to connect us across departments.
+
+To the international community, and my international friends---Marcela,
+Bernardo, Daniel, Walla, Mehrab, and everyone else I have met.
+
+To Erica. To my students and everyone working with outreach here. I love
+being able to pass on our knowledge to the wider world.
+
+To my TEP friends---I hope you all continue to do well.
+
+To all of my grade school math nerds---Stanley, Alex, Rick, Andrew. I
+enjoyed all of our adventures together, appreciate helping myself bloom
+in curiosity and exploration, and keep on math-ing!
 
@@
 our paths diverged, David and Jennifer for guiding my first research
-project.
-
-To Wayne for showing us the joy of astronomy and outreach. To my mentor,
-Ebtihal.
+project. To Wayne for showing us the joy of astronomy and outreach. To
+my mentor, Ebtihal.
 
@@
 chaos and our shared time together.
-
-To Fierce Brosnen and our trivia team. Melody, Lily, Gillian, Stephen,
-Jillian, and so many more.
 
@@
 
-To my musical friends, Adrienne, Alex for our shared joy in music
-making. And Steve, for showing me the joy in a simple life and beautiful
-music. To my fellow knitters. Anne, Anya, Leise, Wes. To the
-international community.
+To my parents for their continuous support throughout this project. To
+my grandparents for always being there.
 
-To my parents for their continuous support throghout this project. To my
-grandparents for always being there. To my cousins and my extended
-family. And to my siblings for always being a source of joy.
+To all of my cousins, but especially Bridget and Gretchen, and my
+extended family. And to my siblings for always being a source of joy. To
+Arya. To all of my furry friends I have sat---Nano, Pluto, Hobbs,
+Penelope, Tochka, Camar, Loki, and Hermes.
 
-To the stars and the moon. To the land of Victoria, and its historical
-caretakers. To the waters and beautiful trees, the ocean waves, rocks
-and hills and jagged coastlines. To the rich, enveloping forests and
-rust-red arbutus tress, scraggly Garry oaks and Camus flowers. I could
-not imagine a more beautiful place to study and to always remind my
-place in the world.
+To the stars and the moon, for your enduring beauty, continuing to
+inspire me and posing unending questions about our natural world.
+
+To the land of Victoria, and its historical caretakers. To the waters
+and beautiful trees, the ocean waves, rocks and hills and jagged
+coastlines. To the rich, enveloping forests and rust-red arbutus tress,
+scraggly Garry oaks and Camus flowers. I could not imagine a more
+beautiful place to study and to always remind my place in the world.

### 2025-10-17 thesis/data.tex

@@
 profile fitted to the inner regions, with a clear excess of stars/light
-in the outer regions. In this chapter, we critically review the density
+in the outer regions. In this Chapter, we critically review the density
 profile of the MW classical dwarfs. This thesis concerns the origin of
 the outer deviations from exponential profiles, with a tidal
-interpretation explored in subsequent chapters.
+interpretation explored in subsequent Chapters.
 
@@
 literature (see Appendix \ref{sec:extra_density}). We conclude that
-selection criteria do not influence our results.
+selection criteria do not influence our conclusions.
 

### 2025-10-17 thesis/discussion.tex

@@
 satellites, and that recent tides are an unlikely explanation for their
-extended outer structure. We therefore suggest that Scl and UMi may host
-an additional, extended population of stars, such as a ``stellar halo''.
-With our data, other classical dwarfs lack evidence for similar extended
-populations. Thus, Scl and UMi have followed a divergent evolutionary
-path, motivating our discussion of extended density formation pathways
-here.
+extended outer structure. We therefore suggest the reason that Scl and
+UMi have extended light profiles is that they host an additional,
+extended population of stars, such as a ``stellar halo.'' With our data,
+other classical dwarfs lack evidence for similar extended populations.
+Thus, Scl and UMi have followed a divergent evolutionary path,
+motivating our discussion of extended density formation pathways here.
 
 In this chapter, we place our results in context with past work. We
-review both works, considering tidal effects on Scl and UMi and
+review that work, considering tidal effects on Scl and UMi and
 detections of peculiarities. We then consider possible origins of
-extended stellar populations for these galaxies. Scenarios include
+extended stellar populations in these galaxies. Scenarios include
 multi-epoch star formation, past mergers depositing stars into the
@@
 Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}.
-Yet, growing work suggests tides may not be as commonplace as suspected.
-For example, \citet{read+2006} suggested that the absence of rising
-radial velocity dispersion profiles in dwarfs implies a lack of tidal
-disruption. \citet{penarrubia+2009}, introducing the break radius as a
-diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite orbits
-were inconsistent with observable tidal effects. Recently,
+Yet, a growing body of work suggests tides may not be as commonplace as
+suspected. For example, \citet{read+2006} suggested that the absence of
+rising radial velocity dispersion profiles in dwarfs implies a lack of
+tidal disruption. \citet{penarrubia+2009}, introducing the break radius
+as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
+orbits were inconsistent with observable tidal effects. Recently,
 \citet{pace+erkal+li2022} used a Jacobi radius-based criterion
@@
 
-A few works have modelled Scl or UMi specifically, finding similar
+A few studies have modelled Scl or UMi specifically, finding similar
 conclusions. \citet{iorio+2019} applied idealized N-body simulations to
@@
 peculiarities that may hold clues to their formation. Scl and UMi both
-host at least two chemodynamic populations, as revealed through their
-photometric or metallicity-velocity structure\footnote{Other examples of
-  galaxies with multiple populations include Carina
+host at least two distinct chemodynamic populations, as revealed through
+their photometric or metallicity-velocity structure\footnote{Other
+  examples of galaxies with multiple populations include Carina
   \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
@@
 excess? In Scl, both populations have scale radii likely smaller than
-the half-light radius. But, the third, more tentatively detected
+the half-light radius. But, a third, more tentatively detected
 population in \citet{arroyo-polonio+2024} would reside within the outer
@@
 A galaxy having undergone a major merger should harbour at least two or
-three populations from distinct origins.
+three populations with distinct origin.
 
@@
 evidence of such features (see Appendix
-\ref{sec:extra_rv_models})---future data would need to uncover signs of
-disruption to support this hypothesis.
+\ref{sec:extra_rv_models})---future data would be needed to uncover
+signs of disruption to support this hypothesis.
 
-\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
-than younger stars due to processes including stellar feedback
+\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
+dynamically hotter than younger stars due to processes including stellar
+feedback
 \citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
@@
 
-We reviewed a number of different, possibly concurrent explanations for
-the formation of extended populations. While we leave the nature of Scl
-and UMi's extended stellar densities an open question, we can discuss
-possible clues to different formation scenarios.
+We have reviewed above a number of different, possibly concurrent
+explanations for the formation of extended populations. While we leave
+the nature of Scl and UMi's extended stellar densities an open question,
+we can discuss possible clues to different formation scenarios.
 
@@
 In this thesis, we have investigated the extended, outer density excess
-of Sculptor and Ursa Minor and its possible tidal origin. We show that
-the density profiles are robust, implying that Scl and UMi contain a
-true density excess relative to other classical dwarfs.
+of Sculptor and Ursa Minor and its possible tidal origin. We have shown
+that the density profiles are robust, implying that Scl and UMi contain
+a true density excess relative to other classical dwarfs.
 

### 2025-10-17 thesis/introduction.tex

@@
 
-Today, a common definition for a dwarf galaxy is a gravitationally bound
-stellar system with dark matter.\footnote{Or, more generally, systems
-  inconsistent with Newtonian dynamics of visible matter alone,
-  \citet{willman+strader2012}.} In contrast, star clusters (like
-globular clusters) have no clear evidence for dark matter. The boundary
-between these two classes blurs for faint, compact stellar associations.
-Systems with shared characteristics of globular clusters and dwarf
-galaxies, and without evidence as to their nature, are known as
-``ambiguous'' systems \citep[e.g.,][]{smith+2024}.
+Today, a common definition for a dwarf galaxy is that of a
+gravitationally bound stellar system with dark matter.\footnote{Or, more
+  generally, systems inconsistent with Newtonian dynamics of visible
+  matter alone, \citet{willman+strader2012}.} In contrast, star clusters
+(like globular clusters) have no clear evidence for dark matter. The
+boundary between these two classes blurs for faint, compact stellar
+associations. Systems with shared characteristics of globular clusters
+and dwarf galaxies, and without evidence as to their nature, are known
+as ``ambiguous'' systems \citep[e.g.,][]{smith+2024}.
 
@@
 metal-poor stellar populations \citep[see the review by][]{simon2019}.
-Altogether, known dwarf galaxies span more than 15 magnitudes, or over 7
+Altogether, known dwarf galaxies span more than 15 magnitudes, or over 6
 decades in stellar mass.
@@
 and ambiguous systems. This nearby population of dwarf galaxies is
-opportune for resolved, detailed studies aimed at understanding their
+amenable to resolved, detailed studies aimed at understanding their
 nature.
@@
 universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
-matter (DM), and 5\% regular old baryons\footnote{In a classic
-  astronomer's corruption of jargon, \emph{Baryons} here means baryons
-  and leptons, i.e., protons, neutrons, and electrons.}
+matter (DM), and 5\% baryons\footnote{In a classic astronomer's
+  corruption of jargon, \emph{Baryons} here means baryons and leptons,
+  i.e., protons, neutrons, and electrons.}
 \citep{planckcollaboration+2020}. While the composition of dark matter
@@
 affect the Local Group today. Dark matter, instead, makes up the vast
-majority of mass in galaxies. Typically, galaxies have baryonic to dark
+majority of mass in galaxies. Typically, galaxies have baryonic-to-dark
 matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
@@
 galaxies. Implications of dark matter properties include cosmological
-structural formation, galaxy structure, and galaxy interactions.
+structure, galaxy formation, and galaxy interactions.
 
@@
 Small-scale structure formation is sensitive to deviations from \LCDM{}
-cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of \LCDM{}
-is that mass perturbations are expected to exist on all scales, and are
-largest on the smallest scales, so we would expect the formation of
-halos on all scales. Many alternative models, such as warm dark matter,
-may smooth out small-scale features and reduce the abundance of small
-halos or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf
-galaxies, which occupy the smallest dark matter halos, are promising
-probes into the behaviour of dark matter on small scales.
+\citep[e.g.,][]{bechtol+2022}. One key prediction of \LCDM{} is that
+mass perturbations are expected to exist on all scales, and are largest
+on the smallest scales, so we would expect the formation of halos on all
+scales. Many alternative models, such as warm dark matter, may smooth
+out small-scale features and reduce the abundance of small halos or
+change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
+which occupy the smallest dark matter halos, are promising probes into
+the behaviour of dark matter on small scales.
 
@@
   this parameter may be closer to \(\Delta \approx 100\) for our
-  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
-  from][]{bryan+norman1998}.} the critical density of the universe:
-\begin{equation}{
+  universe, \(\Delta\) also increases with redshift \citep[see, e.g.,
+  eq. 6 from][]{bryan+norman1998}.} the critical density of the
+universe: \begin{equation}{
 M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G,
@@
 which halos. One simple technique, dubbed ``abundance matching,''
-implies a tight relation between the stellar mass of a galaxy and the
+assumes a tight relation between the stellar mass of a galaxy and the
 mass of the halo it inhabits
@@
 
-Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
-(SMHM, with halo mass represented by \(\vmax\)) predicted by \LCDM{}
-cosmological hydrodynamical simulations of Local Group analogues from
-the \apostle{} project \citep{sawala+2016}.\footnote{\apostle{}
-  simulated Local Group analogues in a \LCDM{} cosmological context with
-  the hydrodynamical setup from the \eagle{} simulations
-  \citep{crain+2015, schaye+2015}}. While there is some scatter, the
-range of predicted \(\vmax\) is fairly narrow across \(\sim 4\) decades
-in stellar mass. This figure indicates that the SMHM relation becomes
-increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
-formed in halos of similar mass. Because lower mass galaxies have
-shallower potential wells, feedback becomes more effective at removing
-gas. Reionization additionally suppresses late star formation in the
-faintest galaxies. As a result, the resulting stellar mass of a dwarf
-galaxy is highly sensitive to dark matter halo mass.
+The right-hand panel of Fig.~\ref{fig:smhm} shows the stellar mass
+versus halo mass relation (SMHM, with halo mass represented by
+\(\vmax\)) predicted by \LCDM{} cosmological hydrodynamical simulations
+of Local Group analogues from the \apostle{} project
+\citep{sawala+2016}.\footnote{\apostle{} simulated Local Group analogues
+  in a \LCDM{} cosmological context with the hydrodynamical setup from
+  the \eagle{} simulations \citep{crain+2015, schaye+2015}}. While there
+is some scatter, the range of predicted \(\vmax\) is fairly narrow
+across \(\sim 4\) decades in stellar mass. This figure indicates that
+the SMHM relation becomes increasingly steep in the dwarf galaxy
+regime---many dwarf galaxies are formed in halos of similar mass.
+Because lower mass galaxies have shallower potential wells, the
+energetic output of evolving stars and possibly supermassive black holes
+(i.e., ``feedback'') becomes more effective at removing gas.
+Reionization additionally suppresses late star formation in the faintest
+galaxies. As a result, the resulting stellar mass of a dwarf galaxy is
+highly sensitive to the details of halo assembly and evolution.
 
@@
 (relative to reionization) can influence the resulting stellar content
-\citep{kim+2024}. Finally, Galactic tides influence both the dark matter
+\citep{kim+2024}. Finally, Galactic tides reduce both the dark matter
 and stellar mass but in different amounts, adding additional scatter to
@@
 by][]{bullock+boylan-kolchin2017, sales+2022}. For example, the mismatch
-between the expected number of dwarf galaxies from simulations and the
-observed abundance has been known as the \emph{missing satellites
-problem}. Additionally, several observations suggest that some dwarf
-galaxies, although not all, possess dark matter ``cores,''
+between the number of dwarf galaxies and the predicted abundance of
+\LCDM{} halos has been known as the \emph{missing satellites problem}.
+Additionally, several observations suggest that some dwarf galaxies,
+although not all, possess dark matter ``cores,''
 \citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019},\footnote{In
@@
 understanding of baryonic physics. For example, recent hydrodynamic
-simulations in particular have shown that strong feedback can produce
+simulations, in particular, have shown that strong feedback can produce
 dark matter cores
 \citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
-Several open questions remain, concerning, e.g., the plane of
-satellites, the details of the sizes and rotation curves of dwarf
-galaxies, and the existence and nature of stellar halos and accretion in
-dwarf galaxies \citep[e.g.,][]{sales+2022}. Altogether, the numerous
-past and ongoing challenges for \LCDM{} in the dwarf galaxy regime
-illustrate the opportunity for dwarf galaxies to test the understanding
-of galaxy formation and dark matter physics.
+Several open questions remain, concerning, e.g., the nearly planar
+distribution of luminous Milky Way satellites, the details of the sizes
+and rotation curves of dwarf galaxies, and the existence and nature of
+stellar halos in dwarf galaxies \citep[e.g.,][]{sales+2022}. Altogether,
+the numerous past and ongoing challenges for \LCDM{} in the dwarf galaxy
+regime illustrate the opportunity for dwarf galaxies to our the
+understanding of galaxy formation and dark matter physics.
 
@@
 
-\subsection{\texorpdfstring{The \emph{Gaia} space
-telescope}{The Gaia space telescope}}\label{the-gaia-space-telescope}
+\subsection{\texorpdfstring{The \emph{Gaia}
+mission}{The Gaia mission}}\label{the-gaia-mission}
 
@@
 
-However, many works speculated that Scl and UMi were influenced by the
-Milky Way's tidal field. Already, \citet{innanen+papp1979} found RR
-Lyrae candidate Scl members \citep[from][]{vanagt1978} out to 180' in an
-elongated distribution, speculating this to be tidal disruption. Later
-density profile determinations noted Scl's elongation and apparent outer
-density excess
+However, many studies have speculated that Scl and UMi have been
+influenced by the Milky Way's tidal field. Already,
+\citet{innanen+papp1979} found RR Lyrae candidate Scl members
+\citep[from][]{vanagt1978} out to 180' in an elongated distribution,
+speculating this to be tidal disruption. Later density profile
+determinations noted Scl's elongation and apparent outer density excess
 \citetext{\citealp{eskridge1988}; \citealp{IH1995}; \citealp{walcher+2003}; \citealp{westfall+2006}; \citealp[but
-see also][]{coleman+dacosta+bland-hawthorn2005}}. These works often
+see also][]{coleman+dacosta+bland-hawthorn2005}}. These studies often
 interpreted these features as evidence of tidal effects
-\citep[e.g.,][]{walcher+2003} or sometimes a dwarf galaxy halo
-\citep{westfall+2006}.
+\citep[e.g.,][]{walcher+2003} or sometimes a ``halo''of stars
+surrounding the dwarf \citep{westfall+2006}.
 
-UMi also attracted similar suspicions. \citet{hargreaves+1994} first
+UMi has also attracted similar suspicions. \citet{hargreaves+1994} first
 detected a velocity gradient in UMi, suggestive of tidal disruption.
-Later, \citet{martinez-delgado+2001} found stars extending far beyond
-the King-profile ``tidal radius'', aligned with UMi's elongation,
-consistent with tidally-stripping simulations in
+Later, \citet{martinez-delgado+2001} found stars far beyond the
+King-profile ``tidal radius'', aligned with UMi's elongation, consistent
+with tidally-stripping simulations in
 \citet{gomez-flechoso+martinez-delgado2003}. \citet{palma+2003}
@@
 measured). Compared to Fornax, both Sculptor and Ursa Minor show an
-excess of stars outside \(\log R/R_h\approx 0.4\), which exceeds 100
-times the density of the exponential fit at large radii.
+excess of stars outside \(\log R/R_h\approx 0.4\), implying densities
+which exceed 100 times the density of the exponential fit at large
+radii.
 
@@
 Cosmological simulations appear to predict that tidal disruption is more
-common than idealized work or the Milky Way system suggests, but the
-origin and role of numerics and assumptions in this discrepancy are
-unclear \citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall
-use idealized simulations here to assess tidal effects after the
-satellite's infall into the MW halo.
+common than idealized work on the Milky Way system suggests, but the
+role of numerics and assumptions in this discrepancy are unclear
+\citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall use
+idealized simulations here to assess tidal effects after the satellite's
+infall into the MW halo.
 
@@
 dwarf spheroidal galaxies. Tidally stripped stars form tidal
-streams---approximately uniform-density stellar tails with a bulk
-velocity gradient
+streams---stellar tails with a bulk velocity gradient
 \citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
@@
 undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
-NFW halos are also found to be nearly immune to tidal disruption
+NFW halos are also found to be resilient to full tidal disruption
 \citep{EP2020}, but cored dark matter halos may disrupt fully and faster
@@
 idea, creating a model for adiabatic tidal mass loss in an isotropic
-tidal field. They explain the resilience of NFW halos against full tidal
-disruption and the origin of well-defined ``tidal tracks'' \citep[as
-observed in][]{PNM2008, green+vandenbosch2019, EN2021}.
+tidal field. Their model explains the resilience of NFW halos against
+full tidal disruption and the origin of well-defined ``tidal tracks''
+\citep[as observed in][]{PNM2008, green+vandenbosch2019, EN2021}.
 
@@
 }\end{equation} where the empirical constant \(C \approx 0.55\).
-\(r_{\rm break}\) describes where the dynamical timescale is longer than
-the time since the perturbation, i.e., the radius within which the
+\(r_{\rm break}\) describes where the dynamical timescale becomes longer
+than the time since the perturbation, i.e., the radius within which the
 galaxy is dynamically relaxed.
@@
 initially follow a 2D exponential profile with a scale radius
-\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We then
-evolve the orbit through a pericentre of 20 kpc starting from an
-apocentre of 100 kpc. See Section~\ref{sec:methods} for a more detailed
+\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We start
+the model at apocentre on an orbit with a pericentre of 20,kpc and
+apocentre of 100,kpc. See Section~\ref{sec:methods} for a more detailed
 description of our numerical setup.
@@
 simulation. \(y'\) is the direction of tangential orbital motion, and
-\(z'\) is the direction of angular momentum, as seen from the centre of
-the host. The dashed green circle represents the break radius
+\(z'\) is the direction of orbital angular momentum, as seen from the
+centre of the host. The dashed green circle represents the break radius
 (Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital

### 2025-10-20 thesis/plots/anisotropy_model.jl
final final edits

@@ -70,7 +70,7 @@ function plot_anisotropies(modelname; title="")
 	)
 
 
-	lines!(midpoints(r_bins_i), prof_i[2], label="initial", linestyle=:dot, color=COLORS[1])
+	lines!(midpoints(r_bins_i), prof_i[2], label="initial", linestyle=:dot, color=COLORS[1], )
 	lines!(midpoints(r_bins), prof_f[2], label="final", linestyle=:solid, color=COLORS[1])
 
 	axislegend(position=:lb)

### 2025-10-20 thesis/plots/cosmological_means.jl

@@ -210,7 +210,7 @@ let
 
 	fig = Figure()
 	ax = Axis(fig[1,1], 
-			 xscale=log10, yscale=log10, limits=(10, 100, 1e5, 1e9), xticks=Makie.automatic, yticks=Makie.automatic)
+			 xscale=log10, yscale=log10, limits=(10, 100, 1e4, 1e9), xticks=Makie.automatic, yticks=Makie.automatic)
 
 
 	

### 2025-10-20 thesis/plots/gaia_utils.jl

@@ -9,6 +9,7 @@ include("utils.jl")
 
 
 SELECTION_COLORS = [colorant"#cecece"; to_colormap(:YlGnBu_5)[end-2:end]]
+
 plot_labels = OrderedDict(
 	:xi => L"$\xi$\,/\,degree",
 	:eta => L"$\eta$\,/\,degree",
@@ -21,10 +22,23 @@ plot_labels = OrderedDict(
 )
 
 
-function compare_j24_samples(datasets, scatter_kwargs, observed_properties; 
+"""
+    compare_j24_samples(
+        datasets::AbstractDict, 
+        scatter_kwargs::AbstractDict, 
+        observed_properties::AbstractDict; 
+        age::Real=12, 
+        legend_position=:lt, 
+        title::AbstractString=""
+    )
+
+Compare J+24 samples by plotting tangent-plane, CMD, and PM diagrams side by side.
+"""
+function compare_j24_samples(datasets::AbstractDict, scatter_kwargs::AbstractDict, observed_properties::AbstractDict; 
         age=12, legend_position=:lt, title="") 
+
 	fig = Figure(
-        size = (5.39 * 72, 4.5*72)
+        size = (6 * 72, 5*72)
 	)
 
     if title != ""
@@ -53,8 +67,21 @@ function compare_j24_samples(datasets, scatter_kwargs, observed_properties;
 end
 
 
-function plot_tangent(gs, datasets, scatter_kwargs, observed_properties)
-    all_stars = first(datasets)[2]
+
+"""
+    plot_tangent(
+        gs::GridLayout, 
+        datasets::AbstractDict, 
+        scatter_kwargs::AbstractDict, 
+        observed_properties::AbstractDict
+    )
+
+Plots the tangent-plane projection (ξ, η) for the provided datasets, 
+including labeled half-light ellipses from observed properties.
+"""
+function plot_tangent(gs, datasets::AbstractDict, scatter_kwargs::AbstractDict, observed_properties::AbstractDict)
+    all_stars = datasets[:best]
+
     dθ = maximum(sqrt.(all_stars.xi.^2 .+ all_stars.eta .^ 2))
 
 	ax = Axis(gs,
@@ -77,8 +104,19 @@ function plot_tangent(gs, datasets, scatter_kwargs, observed_properties)
 end
 
 
-function plot_cmd(gs, datasets, scatter_kwargs, observed_properties; age)
-    df_best = datasets[(collect∘keys)(datasets)[1]]
+"""
[…]

### 2025-10-20 thesis/plots/initial_halo_selection.jl

@@ -23,7 +23,7 @@ using OrderedCollections
 include("./paper_style.jl"); scale_theme_element!(:linewidth, 0.5)
 
 # ╔═╡ eb4a5051-43b6-4afa-bf42-5a74125cf60a
-galaxy = "ursa_minor"
+
 
 # ╔═╡ ed8867fb-ca90-4577-905d-0ad93ec37097
 import TOML

### 2025-10-20 thesis/plots/iso_extra_conv.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.18
+# v0.20.19
 
 using Markdown
 using InteractiveUtils
@@ -42,7 +42,7 @@ nfw = NFW(M_s=halo.M_s, r_s=halo.r_s, c=halo.c')
 
 
 # ╔═╡ ba744355-8da6-4caa-a252-ad83ad4b8b13
-log_r_label = "log radius / kpc"
+r_label = "radius / kpc"
 
 # ╔═╡ bd023583-28b3-4c4d-8ced-7161350a5d7a
 md"""
@@ -57,7 +57,9 @@ function vcirc_axes(; xlims, ylims)
     
     fig = Figure()
     ax = Axis(fig[1, 1],
-        xlabel=log_r_label,
+        xlabel=r_label,
+		xscale=log10,
+		xticks=[0.01, 0.1, 1, 10, 100],
         ylabel=L"$\textrm{v}_\textrm{circ}$ / km\,s$^{-1}$",
         limits=(xlims[1], xlims[2], ylims[1], ylims[2]),
 		yscale=log10,
@@ -67,7 +69,9 @@ function vcirc_axes(; xlims, ylims)
 
 
     ax_res = Axis(fig[2, 1],
-        xlabel=log_r_label,
+        xlabel=r_label,
+		xscale=log10,
+		xticks=[0.01, 0.1, 1, 10, 100],
         ylabel=L"\Delta\,\textrm{v}\,/\textrm{v}_\textrm{NFW}",
         limits=(xlims[1], xlims[2], -0.2, 0.2),
     )
@@ -85,8 +89,10 @@ end
 # ╔═╡ c625ea14-662e-4dc5-851a-10dfba4df25f
 function vcirc_ax(gs, ; xlims, ylims)    
     ax = Axis(gs,
-        xlabel=log_r_label,
-        ylabel=L"$\textrm{v}_\textrm{circ}$ / km\,s$^{-1}$",
+        xlabel=r_label,
+		xscale=log10,
+		xticks=[0.01, 0.1, 1, 10, 100],
+        ylabel=L"circular velocity / km\,s$^{-1}$",
         limits=(xlims[1], xlims[2], ylims[1], ylims[2]),
 		yscale=log10,
 		yticks=[1:10; 10:10:100]
@@ -103,7 +109,7 @@ function plot_vcirc_profiles!(profiles, v_circ_0=x->0; errskip=1)
 
 		y0 = v_circ_0(profile.radii)
[…]

### 2025-10-20 thesis/plots/j+24_density_methods.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.18
+# v0.20.19
 
 using Markdown
 using InteractiveUtils
@@ -31,10 +31,18 @@ import TOML
 # ╔═╡ ad8414dd-9428-47db-b08d-65940091413e
 CairoMakie.activate!(type=:png)
 
+# ╔═╡ c51c6f13-b5ed-4671-9f08-4adb4f7b38b7
+import Statistics: median
+
 # ╔═╡ 322e01f3-32d2-4408-91ef-ab7777f0f925
 module Utils
 	include("gaia_utils.jl")
 end
+
+# ╔═╡ 476200a0-8d64-4237-8b36-490cf2172e50
+md"""
+# Plot utils
+"""
 
 # ╔═╡ 54e449b3-6bf8-45d4-98da-5eb1c6e7e6ec
 lw = theme(:linewidth)[]/2
@@ -44,6 +52,11 @@ log_r_ell_label = L"$\log\,R_\textrm{ell}$\,/\,arcmin"
 
 # ╔═╡ d42a0cd3-cc8e-4a24-8887-7100f3927961
 log_Σ_label = L"$\log\,\Sigma$\,/\,stars\ arcmin$^{-2}$"
+
+# ╔═╡ 2e1e1605-a4f9-45f7-ab1b-56417d94237e
+md"""
+# Data inputs
+"""
 
 # ╔═╡ 389fefd6-60fd-4dd8-ba77-29e87b4ed846
 obs_props_scl = TOML.parsefile(ENV["DWARFS_ROOT"] * "/observations/sculptor/observed_properties.toml")
@@ -56,22 +69,6 @@ obs_props_fornax = TOML.parsefile(ENV["DWARFS_ROOT"] * "/observations/fornax/obs
 
 # ╔═╡ 9ab4431d-e64f-44c9-a641-f16a524f3c19
 α = LilGuys.R_h(LilGuys.Exp2D())
-
-# ╔═╡ 9c594ea7-8670-4235-8bde-cb2d670fe2c3
-function get_R_h(obs_props)	
-	R_h = obs_props["R_h"]
-end
-
-# ╔═╡ 933c2e0a-db49-46fd-b3ae-a40be190022d
-R_munoz_1 = 36.58
-
-# ╔═╡ f3b219b4-7cc7-48a0-9675-91de7f40bb6d
-function get_M_s(galaxyname)
[…]

### 2025-10-20 thesis/plots/lmc_orbits_mass_effect.jl

@@ -53,22 +53,24 @@ CairoMakie.activate!(type=:png, px_per_unit=2)
 # ╔═╡ 6b0b6c8f-eef9-4c45-8227-d1906fe6b80b
 scale_theme_element!(:linewidth, 1/2)
 
+# ╔═╡ 35ce583b-0938-429e-af5d-b17b399f6690
+Nmax = 100 # number of orbits to plot
+
+# ╔═╡ 3d417fe3-a75d-476e-a4f9-8ce25914c473
+function read_orbits(modeldir)
+	structs = LilGuys.read_ordered_structs(joinpath(modeldir, "orbits.hdf5"), LilGuys.Orbit)
+
+	filt = 1:min(Nmax, length(structs))
+	last.(structs)[filt]
+end
+
 # ╔═╡ 16f4ac20-d8cf-4218-8c01-c15e04e567fb
 md"""
 # The example orbits
 """
 
-# ╔═╡ 35ce583b-0938-429e-af5d-b17b399f6690
-Nmax = 100 # number of orbits to plot
-
 # ╔═╡ 15863916-6601-4f45-9f45-4cd303bbcc4d
 modeldir = joinpath(ENV["DWARFS_ROOT"], "orbits/sculptor", "vasiliev24_L3M11_9Gyr")
-
-# ╔═╡ fd7246ec-212c-4255-a658-61bd82b9b7e0
-modeldir_best = joinpath(ENV["DWARFS_ROOT"], "orbits/sculptor", "vasiliev24_L3M11_2x_special_cases")
-
-# ╔═╡ 453d7f2e-91a4-46b5-9697-a4548f145cec
-modeldir_best_long = joinpath(ENV["DWARFS_ROOT"], "analysis/sculptor/1e6_new_v31_r3.2/", "L3M11_9Gyr_smallperi.3b")
 
 # ╔═╡ 9efc0091-ea97-439a-bbf4-5c8b5f1127fc
 modeldir_no = joinpath(ENV["DWARFS_ROOT"], "orbits/sculptor", "vasiliev24_M11")
@@ -77,146 +79,22 @@ modeldir_no = joinpath(ENV["DWARFS_ROOT"], "orbits/sculptor", "vasiliev24_M11")
 lmc_orbit = get_lmc_orbit(joinpath(ENV["DWARFS_ROOT"], "orbits/sculptor/vasiliev24_L3M11")) |> reverse
 
[…]

### 2025-10-20 thesis/plots/mcmc_hists.jl

@@ -146,7 +146,7 @@ let
 	linkaxes!(fig.content...)
 
 
-	Label(fig[end+1, :], "log Radius / arcmin", )
+	Label(fig[end+1, :], L"log $R_\textrm{ell}$ / arcmin", )
 	Label(fig[:, 0], "log surface density", rotation=π/2, )
 
 	axislegend(ax, position=:rt, unique=true, backgroundcolor=(:white, 0.8))
@@ -179,7 +179,7 @@ let
 	
 	linkaxes!(fig.content...)
 
-	Label(fig[end+1, :], "log Radius / arcmin", )
+	Label(fig[end+1, :], L"log $R_\textrm{ell}$ / arcmin", )
 	Label(fig[:, 0], "log surface density", rotation=π/2, )
 	Legend(fig[0,2], ax, tellwidth=false, unique=true)
 

### 2025-10-20 thesis/plots/model_utils.jl

@@ -9,116 +9,57 @@ import LinearAlgebra: normalize
 
 smallfontsize = @lift 0.8*$(theme(:fontsize))
 smalllinewidth = @lift 0.5 * $(theme(:linewidth))
-logdensityrange = (-8, 2)
-
-function get_R_h(galaxyname)
-	obs_props = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "observations/$galaxyname/observed_properties.toml"))
-	R_h = obs_props["R_h"]
-end
-
-function get_r_b(galaxyname, modelname, starsname; lmc=false)
-	model_dir = joinpath(ENV["DWARFS_ROOT"], "analysis/$galaxyname/$modelname/stars/$starsname/")
-
-	prof_f = SurfaceDensityProfile(model_dir * "final_profile.toml")
-
-	σv = prof_f.annotations["sigma_v"]
-	if lmc
-		props = TOML.parsefile(model_dir * "../../orbital_properties_lmc.toml")
-	else
-		props = TOML.parsefile(model_dir * "../../orbital_properties.toml")
-	end
-
-	dist_f =  TOML.parsefile(model_dir * "../../orbital_properties.toml")["distance_f"]
-
-	
-	dt = props["t_last_peri"]
-	r_b = LilGuys.break_radius(σv / V2KMS, dt / T2GYR)
-	R_h = get_R_h(galaxyname)
-
-	return LilGuys.kpc2arcmin(r_b, dist_f)
-end
 
 
-function get_r_j(galaxyname, modelname; lmc=false)
-	model_dir = joinpath(ENV["DWARFS_ROOT"], "analysis/$galaxyname/$modelname/")
+"""
+    compare_both(galaxyname, modelname, starsname, <keyword arguments)
 
-	if lmc
-		props = TOML.parsefile(model_dir * "jacobi_lmc.toml")
-	else
-		props = TOML.parsefile(model_dir * "jacobi.toml")
-	end
+Makes a comparison plot of the 2D densities of the model 
+before and after tidal evolution,
+and the initial and final density as compared with the observed density profile
 
-	return props["r_J"] 
-end
+# Arguments
+- `norm_shift`: Artificially shift the calculated normalization of the density profile in order to better agree with data
+- `lmc`: calculate break and jacobi radii with respect to the lmc
+- `title`: A plot title
+- `kwargs...`: passed to compare profiles.
 
-
[…]

### 2025-10-20 thesis/plots/orbit_utils.jl

@@ -4,8 +4,6 @@ using LilGuys
 using Arya
 import TOML
 
-import LinearAlgebra: ⋅
-
 strokewidth = @lift $(theme(:linewidth)) / 2
 
 """
@@ -165,7 +163,6 @@ function plot_rt_today!(ax_rt, orbit::Orbit, index=1; kwargs...)
     x0 = orbit.positions[:, index]
     v0 = orbit.velocities[:, index]
     r0 = radii(x0)
-    vr = (x0 ⋅ v0) / r0 / T2GYR
     t0 = orbit.times[index] * T2GYR
 
     plot_rt_today!(ax_rt, t0, r0; kwargs...)

### 2025-10-20 thesis/plots/scl_rv_gradient.jl

@@ -109,7 +109,9 @@ let
 		ylabel = L"\eta\ /\ \textrm{arcmin}",
 		xreversed=true,
 		aspect = DataAspect(), 
-			  limits=(-50, 50, -30, 30)
+			  limits=(-50, 50, -30, 30),
+		xticks = -40:20:40,
+		yticks = -40:20:40,
 	)
 
 	arrowscale = 4

### 2025-10-20 thesis/plots/scl_trajectories.jl

@@ -120,6 +120,8 @@ let
 	ylims!(0, 120)
 	plot_rt_today!(ax_rt, scl_best, length(scl_best))
 
+	Label(fig[0, :], "Sculptor", font=:bold)
+
 	rowsize!(fig.layout, 2, Aspect(1, 1.0))
 
 	resize_to_layout!(fig)

### 2025-10-20 thesis/plots/scl_umi_rv_fits.jl

@@ -174,7 +174,7 @@ end
 	v0 = median(samples_umi.μ)
 	σ = median(samples_umi.σ)
 	ax = Axis(fig[2,1], 
-		xlabel = L"$\textrm{v}_\textrm{gsr}'$ / km\,s$^{-1}$",
+		xlabel = L"$\textrm{v}_\textrm{los}'$ / km\,s$^{-1}$",
 		ylabel = "density",
 		limits=(v0-4σ, v0+4σ, 0, nothing)
 
@@ -211,8 +211,8 @@ end
 
 
 	ax2 = Axis(fig[2,2],
-		xlabel = L"$\log\ R$ / arcmin",
-		ylabel = L"$\log\ \sigma_{v, \textrm{los}}$ / km s$^{-1}$"
+		xlabel = L"$\log\ R_\textrm{ell}$ / arcmin",
+		ylabel = L"$\log\ \sigma_{\textrm{v, los}}$ / km s$^{-1}$"
 
 			  )
 

### 2025-10-20 thesis/plots/sim_pictures.jl

@@ -62,6 +62,12 @@ figname = Dict(
 	"ursa_minor" => "umi"
 )[galaxyname]
 
+# ╔═╡ 21d0e542-08ea-4f56-a07d-f1a8e7e69019
+figtitle = Dict(
+	"sculptor" => "Sculptor",
+	"ursa_minor" => "Ursa Minor"
+)[galaxyname]
+
 # ╔═╡ 3c032178-8d48-4f9c-bcec-9bf704718ea9
 out = Output(joinpath(ENV["DWARFS_ROOT"], "analysis", modelname));
 
@@ -154,7 +160,7 @@ let
 	fig = Figure()
 
 	for (i, idx) in enumerate(idxs)
-		ii, jj = (i-1) ÷ 3 , (i-1) % 3 
+		ii, jj = 1 + ((i-1) ÷ 3) , (i-1) % 3 
 		axis = Axis(fig[ii, jj])
 		plot_xy_density!(out[idx])
 		poly!(xz_iso.x, xz_iso.z, color=COLORS[9])
@@ -171,8 +177,10 @@ let
 		text!(0.05, 0.95, text= "$(round(t, digits=1)) Gyr", space=:relative, color=:white, align=(:left, :top))
 	end
 
+	Label(fig[0, :], figtitle, font=:bold)
+
 	rowsize!(fig.layout, 1, Aspect(1, 1.0))
-	rowsize!(fig.layout, 0, Aspect(1, 1.0))
+	rowsize!(fig.layout, 2, Aspect(1, 1.0))
 
 	rowgap!(fig.layout, 6)
 	colgap!(fig.layout, 6)
@@ -197,6 +205,7 @@ end
 # ╠═5eaf3b50-886e-47ac-9a7c-80d693bc3c17
 # ╠═b9600d93-5946-4380-a2ae-9b5f673bbaf5
 # ╠═32db23d9-7959-41ac-aff4-b63df5e4b94a
+# ╠═21d0e542-08ea-4f56-a07d-f1a8e7e69019
 # ╠═3c032178-8d48-4f9c-bcec-9bf704718ea9
 # ╠═a3be2d61-98eb-4037-afb4-4155ba24cc21
 # ╠═d50c120c-d72a-4fdc-b18a-9ec185c25c04

### 2025-10-20 thesis/plots/tidal_tracks.jl

@@ -133,15 +133,6 @@ function i_f_legend(ax, profs)
 	axislegend(ax, [l1, l2, l3], [L"initial ($t=%$t_i$ Gyr)", L"final ($t=0$ Gyr)", "velocity max", ], position=:rb)
 end
 
-# ╔═╡ 07fe2997-559e-4bef-b4f4-934d850cdfe2
-function plot_r_j!(r_j, y0)	
-	y_mark = y0 * 10^-0.5
-	arrows2d!([r_j], [y_mark], [0], [y_mark * (1 - 10^-0.30)], shaftwidth=theme(:linewidth)[]/2, tiplength=9, 
-			  color=:grey, minshaftlength=0)
-	text!(r_j, y_mark, text="Jacobi", rotation=-π/2, align=(:left, :center), color=:grey)
-
-end
-
 # ╔═╡ 59e9658c-7a2c-4964-8a24-fdf7b152ebcc
 theme(:Annotation)[:style][]
 
@@ -193,6 +184,15 @@ end
 
 # ╔═╡ d732b00a-acbc-4772-b7b9-103dd90a9357
 smallfontsize = 0.8 * theme(:fontsize)[]
+
+# ╔═╡ 07fe2997-559e-4bef-b4f4-934d850cdfe2
+function plot_r_j!(r_j, y0)	
+	y_mark = y0 * 10^-0.5
+	arrows2d!([r_j], [y_mark], [0], [y_mark * (1 - 10^-0.30)], shaftwidth=theme(:linewidth)[]/2, tiplength=7.5, tipwidth=7.5 * 2/sqrt(3),
+			  color=:grey, minshaftlength=0)
+	text!(r_j, y_mark, text="Jacobi", rotation=-π/2, align=(:left, :center), color=:grey, fontsize=smallfontsize)
+
+end
 
 # ╔═╡ 2d9e9ecf-32b8-4aa0-8fbb-a5bb24b7e9a1
 function plot_tidal_track_expected!(df)
@@ -266,9 +266,6 @@ end
 @savefig "scl_tidal_track"  plot_tidal_track(modelnames["scl_smallperi"]..., 
 											 title="Sculptor: smallperi orbit")
 
-# ╔═╡ f40dab86-d30b-4ba5-95b2-929b872dd887
[…]

### 2025-10-20 thesis/plots/umi_lmc_trajectories.jl

@@ -158,7 +158,8 @@ let
 	axislegend(position=:lt, backgroundcolor=(:white, 0.8))
 	#Legend(fig[4, 2], ax_rt, tellwidth=false)
 
-	
+	Label(fig[0, :], "Ursa Minor (with LMC)", font=:bold)
+
 	rowsize!(fig.layout, 2, Aspect(1, 1.0))
 
 	resize_to_layout!(fig)

### 2025-10-20 thesis/plots/v_circ_potential.jl

@@ -90,6 +90,11 @@ end
 	fig
 end
 
+# ╔═╡ a470c8bf-cfa6-4fed-be2a-5640626b80ab
+md"""
+these numbers are here to sanity check everythin
+"""
+
 # ╔═╡ e83f16b2-3d84-4a67-90f3-ffe9b439a863
 v_circ(Φ_ep._py, 100.)
 
@@ -116,6 +121,7 @@ v_circ(Φ_nfw, 100.)
 # ╠═4c0c9322-186e-4ba3-a98e-81e3c3c1b037
 # ╠═86961f65-fa3e-4899-81d5-94c6bec3b8de
 # ╠═3c032178-8d48-4f9c-bcec-9bf704718ea9
+# ╠═a470c8bf-cfa6-4fed-be2a-5640626b80ab
 # ╠═e83f16b2-3d84-4a67-90f3-ffe9b439a863
 # ╠═d01934f7-ab69-4715-b68c-e53709c57dee
 # ╠═d9699355-d786-436b-8811-ec8b8caff091

### 2025-10-20 thesis/abstract.tex
julio final edits
thesis updates

@@
 dark-matter-dominated stellar systems. The light profiles of dSphs
-typically follow an exponential law, with the outer density sharply
-declining. Yet systems like the Sculptor (Scl) and Ursa Minor (UMi)
-dSphs host member stars out to \(\sim10\) effective radii, indicating
-outer deviations from an exponential law. The origin of these extended
+typically follow an exponential law, with a sharply declining outer
+density. Yet systems like the Sculptor (Scl) and Ursa Minor (UMi) dSphs
+host member stars out to \({\sim}10\) effective radii, indicating outer
+deviations from an exponential law. The origin of these extended
 profiles is unclear, possibly arising from Galactic tides or intrinsic
@@
 the tidal field of the Milky Way. We find that neither dwarf experiences
-tides sufficient to affect their stellar distribution. One complication
-is that Sculptor's orbit is strongly influenced by the presence of the
-Large Magellanic Cloud (LMC). Our models still suggest that the combined
-tides of the LMC and MW are unable to explain Scl's outer profile. We
-conclude that the extended light profiles of Scl and UMi are not of
-tidal origin. They are instead likely innate, possibly explained by past
-mergers, accretion events, or episodic star formation.
+tides strong enough to affect their stellar distribution. One
+complication is that Sculptor's orbit is strongly influenced by the
+presence of the Large Magellanic Cloud (LMC). Our N-body models still
+suggest that the combined tides of the LMC and MW are unable to explain
+Scl's outer profile. We conclude that the extended light profiles of Scl
+and UMi are not of tidal origin. They are instead likely innate,
+possibly explained by past mergers, accretion events, or episodic star
+formation.

### 2025-10-20 thesis/acknowledgements.tex
thesis updates

--- 2025-10-17 thesis/acknowledgments.tex
@@
 To all of my new friends I have made along the way. Ruxin, Aviv. To
-Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud. To Akshara for
-all your personal and professional advice. We have build a wonderful
-community of students in this department. To all the students who have
-since moved on and show what we may be: Bobby, Leo, Ryan, Katie,
-Fletcher, Ayden, Maheyer, Jess, Dhwani, Spencer, Mallory, Omar, Isaac,
-and Dasha.
+Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud, Anne, Anya,
+Shoshanah. To Akshara and Simon for all your personal and professional
+advice. We have built a wonderful community of students in this
+department. To all the students who have since moved on and show what we
+may be: Bobby, Leo, Ryan, Katie, Fletcher, Ayden, Maheyer, Jess, Dhwani,
+Spencer, Mallory, Omar, Isaac, and Dasha.
 
-And to my friends across UVic, Wyatt, Rosemary, Maggie, and the many
-people I am still meeting.
-
-To the Linguistics extended family and team \emph{adequate}. Melody,
-Lily, Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are
-all wonderful human beings, cooks, and always a joy to spend time with.
+And to my friends across UVic, Wyatt, Rosemary, Maggie. To the
+Linguistics extended family and team \emph{adequate}. Melody, Lily,
+Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are all
+wonderful human beings, cooks, and always a joy to spend time with.
 
@@
 
-To my fellow crafters, knitters, and crochet friends. Leise, Wes, Anne,
-Anya, Ashley, Sabrina, Sarah. To the graduate student society for
-continuing to connect us across departments.
+To my fellow fibre friends and crafters. Leise, Wes, Ashley, Sabrina,
+Sarah. To the graduate student society for continuing to connect us
+across departments.
 
 To the international community, and my international friends---Marcela,
-Bernardo, Daniel, Walla, Mehrab, and everyone else I have met.
+Bernardo, Daniel, Victor, Walla, Mehrab, and everyone else I have met.
 
@@
 
-To all of my cousins, but especially Bridget and Gretchen, and my
-extended family. And to my siblings for always being a source of joy. To
-Arya. To all of my furry friends I have sat---Nano, Pluto, Hobbs,
+To Nolan and Olivia, always a joy to see each other. To Bridget and
+Gretchen. To my extended family. And to my siblings for always being a
+source of joy.
+
+To Arya. To all of my furry friends of friends---Nano, Pluto, Hobbs,
 Penelope, Tochka, Camar, Loki, and Hermes.

### 2025-10-20 thesis/data.tex
thesis updates

@@
 Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
-black bar in the top left represents the median colour error.
+black bar in the top left represents the median member colour error.
 \textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
@@
 \citet{MV2020b} proper motion. The black cross represents the median
-proper motion error.}\label{fig:scl_selection}
+member proper motion error.}\label{fig:scl_selection}
 \end{figure}
@@
 ``velocity confirmed'' members outside a radius of \(6R_h\) with red
-stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
-mark the location of Muñoz 1 with a pink
-circle.}\label{fig:umi_selection}
+stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. The
+isochrone is 12 Gyr old with \({\rm [Fe/H]} = -2.13\). We also mark the
+location of Muñoz 1 with a pink circle.}\label{fig:umi_selection}
 \end{figure}
@@
 Fig.~\ref{fig:scl_selection} except for Fornax. The isochrone is instead
-for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
-Fornax does not show the same extended outer halo of probable members as
-Sculptor or Ursa Minor.}\label{fig:fornax_selection}
+for a 2Gyr, \({\rm [Fe/H]}=-0.99\) stellar population. RV measurements
+are from \citet{WMO2009} and APOGEE. Fornax does not show the same
+extended outer halo of probable members as Sculptor or Ursa
+Minor.}\label{fig:fornax_selection}
 \end{figure}

### 2025-10-20 thesis/extra_results.tex
thesis updates

@@
 profiles for the initially anisotropic model of Scl on the \smallperi{}
-orbit. The final profile is far more isotropic (\(\beta=0\)) than the
-initial.}\label{fig:anisotropy_i_f}
+orbit. The final profile is more isotropic (closer to \(\beta=0\)) than
+the initial.}\label{fig:anisotropy_i_f}
 \end{figure}
@@
 \(z'\) the major and minor axes. The contours are drawn assuming normal
-smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
+smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
 dex.}\label{fig:oblate_i_f}
@@
 model, the \texttt{MW\ impact} model, can approximately reproduce the
-observed position of Scl.
+observed position of Scl (see Fig.~\ref{fig:scl_mw_impact_orbit}). The
+model has a Galactocentric initial position of
+\([67.83, -352.2, 110.3]\,\kpc\) and velocity of
+\([-3.68, 30.79, -22.77]\,\kpc\).
 
@@
 orbital history would be necessary to tidally transform Scl's stars.
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_orbits.pdf}}
+\caption[Sculptor MW impact orbit]{The orbit of Sculptor for the
+MW-impact model. The point (dotted) and n-body (solid) diverge by
+\(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
+\end{figure}
 

### 2025-10-20 thesis/introduction.tex
thesis updates

@@
 
-Today, a common definition for a dwarf galaxy is that of a
+Today, a common definition for a (dwarf) galaxy is that of a
 gravitationally bound stellar system with dark matter.\footnote{Or, more
@@
 the Local Group of galaxies. The Local Group is defined as the group
-consisting of galaxies within \(\sim 1\) Mpc from the Milky
-Way-Andromeda centre \citep[e.g.,][ and references
-therein]{mcconnachie2012}. Today, we know that the Milky Way system is
-teeming with dwarfs, many of which are satellites of either the Milky
-Way (MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
-the MW satellite system, including dwarf galaxies, globular clusters,
-and ambiguous systems. This nearby population of dwarf galaxies is
-amenable to resolved, detailed studies aimed at understanding their
-nature.
+consisting of galaxies within \(\sim 1\) Mpc from the MW-Andromeda
+centre \citep[e.g.,][ and references therein]{mcconnachie2012}. Today,
+we know that the Milky Way system is teeming with dwarfs, many of which
+are satellites of either the MW or Andromeda (M31).
+Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
+including dwarf galaxies, globular clusters, and ambiguous systems. This
+nearby population of dwarf galaxies is amenable to resolved, detailed
+studies aimed at understanding their nature.
 
@@
 \centering
-\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/galaxy_pictures.png}
 \caption[Images of dwarf galaxies]{Images\footnotemark{} of the LMC
@@
 observations of the universe stem from the cosmic microwave background
-(CMB)---revealing a nearly uniform, isotropic blackbody emission. But
-tiny perturbations in the CMB, temperature fluctuations of 1 part in
-100,000, reveal the underlying seeds of large-scale cosmological
-structure. In an expanding universe, gravitational instability makes CDM
-overdensities grow and collapse hierarchically onto larger structures.
-Initially, baryonic matter was coupled to radiation and resisted
-collapse. Dark matter, only influenced by gravity instead, freely
-collapsed into the first structures. Mass perturbations sufficiently
-small and overdense become self-gravitating structures, known as
-\emph{halos}. After recombination, where electrons combined with atomic
-nuclei to form atoms, baryons decoupled from radiation and fell into
-dark matter halos, where they condensed at the centre through radiative
-energy losses. The densest pockets of baryons later formed the first
-stars and galaxies.
+(CMB)---revealing a nearly uniform, isotropic blackbody emission
+\citep[e.g.,][]{ryden2016}. But tiny perturbations in the CMB,
+temperature fluctuations of 1 part in 100,000, reveal the underlying
+seeds of large-scale cosmological structure. In an expanding universe,
+gravitational instability makes CDM overdensities grow and collapse
+hierarchically onto larger structures. Initially, baryonic matter was
+coupled to radiation and resisted collapse. Dark matter, only influenced
+by gravity instead, freely collapsed into the first structures. Mass
+perturbations sufficiently small and overdense become self-gravitating
+structures, known as \emph{halos} \citep[e.g.,][]{galaxiesbook}. After
+recombination, where electrons combined with atomic nuclei to form
+atoms, baryons decoupled from radiation and fell into dark matter halos,
+where they condensed at the centre through radiative energy losses. The
+densest pockets of baryons later formed the first stars and galaxies.
 
@@
 In \LCDM{} cosmological simulations, dark matter halos are remarkably
-self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
-observe that the spherically-averaged density profiles \(\rho(r)\) are
-universally well described by a two-parameter law,
+self-similar. \citet{NFW1996, NFW1997} observe that the
+spherically-averaged density profiles \(\rho(r)\) are universally well
+described by a two-parameter law,
 \begin{equation}\protect\phantomsection\label{eq:nfw}{
@@
 }\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
-density. This profile has shown remarkable success at describing \LCDM{}
-halos across several orders of magnitude in mass. NFW profiles are
-\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
-radii \(r \ll r_s\). The steepness of the density profile increases
-gradually with radius, and at large radii the density falls off like
-\(\rho \sim 1/r^3\). The blue solid curve in Fig.~\ref{fig:nfw_density}
-shows an example NFW halo.
+density. This profile, known by the author's initials NFW, has shown
+remarkable success at describing \LCDM{} halos across several orders of
+magnitude in mass. NFW profiles are \emph{cuspy}, where the density
+rises like \(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness
+of the density profile increases gradually with radius, and at large
+radii the density falls off like \(\rho \sim 1/r^3\). The blue solid
+curve in Fig.~\ref{fig:nfw_density} shows an example NFW halo.
 
@@
 matter in a Fornax-like galaxy. The dark matter is more extended and
-massive than the star across the entire galaxy. Fornax has a stellar
-mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
-\(0.65\,\kpc\) \citep{munoz+2018, woo+courteau+dekel2008}. The
-corresponding cosmological-mean halo for Fornax has \(\vmax=40\,\kpc\)
-and \(\rmax=8\,\kpc\), or \(M_{200} = 1\times10^{10}\,\Mo\) and
+massive than the star across the entire galaxy. This galaxy has a
+stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
+radius \(0.65\,\kpc\) and 2D exponential stars \citep[based
+on,][]{munoz+2018, woo+courteau+dekel2008}. The corresponding
+cosmological-mean halo has \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
+\(M_{200} = 1\times10^{10}\,\Mo\) and
 \(c=12.5\).}\label{fig:nfw_density}
@@
   in a \LCDM{} cosmological context with the hydrodynamical setup from
-  the \eagle{} simulations \citep{crain+2015, schaye+2015}}. While there
+  the \eagle{} simulations \citep{crain+2015, schaye+2015}.} While there
 is some scatter, the range of predicted \(\vmax\) is fairly narrow
@@
 
-Outside the MW system, exponential profiles are common, but sometimes
-with modifications. For example, many extragalactic dwarf elliptical,
-blue compact, and irregular dwarf galaxies are better described with an
-exponential profile to which a central cusp or nuclear region is added
+Dwarf galaxies outside the Local Group commonly follow exponential
+profiles, but sometimes with modifications. For example, many
+extragalactic dwarf elliptical, blue compact dwarf, and irregular dwarf
+galaxies are better described with an exponential profile to which a
+central cusp or nuclear region is added
 \citep{caldwell+bothun1987, noeske+2003}. On the other hand, some
@@
 Cosmological simulations appear to predict that tidal disruption is more
-common than idealized work on the Milky Way system suggests, but the
-role of numerics and assumptions in this discrepancy are unclear
+common than idealized simulations of MW dwarfs suggest, but the role of
+numerics and assumptions in this discrepancy are unclear
 \citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall use
@@
 initially follow a 2D exponential profile with a scale radius
-\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We start
-the model at apocentre on an orbit with a pericentre of 20,kpc and
-apocentre of 100,kpc. See Section~\ref{sec:methods} for a more detailed
-description of our numerical setup.
+\(R_s=0.25\,\kpc\) and total mass \(5\times10^5\,\Mo\), embedded within
+the inner dark matter halo. We start the model at apocentre on an orbit
+with a pericentre of 20,kpc and apocentre of 100,kpc. See
+Section~\ref{sec:methods} for a more detailed description of our
+numerical setup.
 
@@
 simulation. \(y'\) is the direction of tangential orbital motion, and
-\(z'\) is the direction of orbital angular momentum, as seen from the
-centre of the host. The dashed green circle represents the break radius
-(Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
+\(z'\) is the direction of orbital angular momentum, as measured from
+the centre of the host. The dashed green circle represents the break
+radius (Eq.~\ref{eq:r_break}) and the blue arrow points in the orbital
 direction. \textbf{Bottom left}: The projected stellar density profile

### 2025-10-20 thesis/methods.tex
thesis updates

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.png}
 \caption[Sculptor's possible orbits]{The orbits of Sculptor in a static
@@
 (bottom). The Milky Way is at the centre with the disk lying in the
-\(x\)--\(y\) plane. Our selected \smallperi{} orbit is plotted in black,
-and light green transparent lines represent orbits sampled over Sculptor
-observables in Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is
-well-constrained in this potential, and it is unlikely to achieve a
-smaller pericentre than the \smallperi{} orbit.}\label{fig:scl_orbits}
+\(x\)--\(y\) plane. Our selected N-body's \smallperi{} orbit is plotted
+in black, and light green transparent lines represent orbits sampled
+over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
+Sculptor is well-constrained in this potential, and it is unlikely to
+achieve a smaller pericentre than the \smallperi{}
+orbit.}\label{fig:scl_orbits}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/umi_xyzr_orbits.pdf}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/thesis/figures/umi_xyzr_orbits.png}
 \caption[Ursa Minor's possible orbits]{Similar to
@@
 Way potential in Galactocentric \(x\), \(y\), and \(z\) coordinates. In
-the lower panel, we show the radius versus time for only three orbits of
-Ursa Minor, representing the \smallperi{} point-mass orbit (black), the
-mean orbit, and the orbit with the \(3\sigma\)-largest
-pericentre.}\label{fig:umi_orbits}
+the lower panel, we show the radius versus time for the selected
+\smallperi{} N-body orbit (black) and three point orbits of Ursa Minor,
+representing the \smallperi{}, the mean orbit, and \(3\sigma\)-largest
+pericentre orbit.}\label{fig:umi_orbits}
 \end{figure}
@@
 \midrule
-distance / kpc & 82.6 & 72.9 & 64.6\\
+distance / kpc & 82.6 & 73.1 & 64.6\\
 $\pmra / \masyr$ & 0.134 & 0.137 & -0.158\\
-$\pmdec / \masyr$ & -0.198 & -0.157 & 0.050\\
+$\pmdec / \masyr$ & -0.198 & -0.156 & 0.050\\
 LOS velocity / $\kms$ & 111.2 & 111.2 & -245.75\\
@@
 }\end{equation} In practice the right, boundary term is zero.\footnote{If,
-  at large \(r\), \(\rho \propto r^{-n}\) with \(n>1\), so
-  \(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\).} We take
-\(\Psi\) from the underlying assumed analytic dark matter potential.
-\(\rho_\star\) can be calculated from the surface density,
-\(\Sigma_\star\), via the inverse Abel transform.
+  at large \(r\), \(\rho \propto r^{-n}\) with \(n>1\), then \(\Psi\)
+  falls off no steeper than \(\Psi \sim r^{-1}\). So \(d\rho / d\Psi\)
+  decays no less rapidly than \$ r\^{}\{-n+1\}\$ which \(\rightarrow0\)
+  at large \(r\).} We take \(\Psi\) from the underlying assumed analytic
+dark matter potential. \(\rho_\star\) can be calculated from the surface
+density, \(\Sigma_\star\), via the inverse Abel transform.
 
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_velocity.pdf}}
+\includegraphics[width=6in,height=\textheight,keepaspectratio]{figures/initial_velocity.png}
 \caption[Initial halo velocity profiles]{The initial circular velocity

### 2025-10-20 thesis/results.tex
thesis updates

@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_sim_images.png}
 \caption[Sculptor simulation snapshots]{Images of the dark matter
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.png}}
+\includegraphics[width=4.5in,height=\textheight,keepaspectratio]{figures/scl_tidal_track.png}
 \caption[Sculptor tidal tracks]{Dynamical evolution for the \smallperi{}
@@
 time of last pericentre / Gyr & $-0.45 \pm 0.2$ & -0.47\\
-number of pericentres & 6 & 6\\
+number of pericentres & 5–6 & 6\\
 Jacobi radius / kpc & $4.5 \pm 0.3$ & 3.5\\
-Jacobi radius / arcmin & $186\pm12$ & 101\\
+Jacobi radius / arcmin & $186\pm12$ & 148\\
 final heliocentric distance / kpc & $83.2\pm2$ & 81.6\\
@@
 orbit---Scl is inferred to have reach an apocentre of nearly
-\(300\,\kpc\). According to this, Scl may even be on first infall,
-depending on the MW and LMC mass. Scl's is orbiting the Milky Way on a
-similar plane to the LMC, but in the opposite direction---thus, Scl is
-unlikely to be an LMC satellite.
+\(300\,\kpc\). According to this, Scl may be on first or second infall,
+depending on the MW and LMC mass.\footnote{The orbital periods may be
+  \(\gtrsim 7\,\Gyr\) for the lighter MW model, compared to
+  \(\sim 2\,\Gyr\) for the MW-only case} Scl's is orbiting the Milky Way
+on a similar plane to the LMC, but in the opposite direction---thus, Scl
+is unlikely to be an LMC satellite.
 
@@
 time of last pericentre / Gyr & $-0.38\pm0.01$ (-0.11) & -0.33 (-0.10)\\
-number of pericentres & 1 (2) & 1 (1)\\
-Jacobi radius / kpc & $4.1\pm0.3$ ($4.5\pm0.2$) & 2.8 (2.6)\\
-Jacobi radius / arcmin & $168 \pm 11$ ($186\pm6$) & 132 (121)\\
+number of pericentres & 1 (1) & 1 (1)\\
+Jacobi radius / kpc & $3.3\pm0.2$ ($3.6\pm0.2$) & 2.8 (2.6)\\
+Jacobi radius / arcmin & $136 \pm 9$ ($159\pm5$) & 132 (121)\\
 final heliocentric distance / kpc & $83.2\pm2$ & 72.9\\
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_lmc_sim_images.png}
 \caption[Sculptor simulation snapshots with LMC]{Similar to
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_images.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_sim_images.png}
 \caption[Ursa Minor simulation snapshots]{Similar to
@@
 apocentre & $83 \pm 4$ & 75\\
-time of last pericentre & $-0.97 \pm 0.07$ & -0.80\\
-number of pericentres & $\sim 8$ & 8\\
+time of last pericentre & $-0.96 \pm 0.07$ & -0.80\\
+number of pericentres & 7–8 & 8\\
 Jacobi radius / kpc & $3.7 \pm 0.3$ & 2.9\\
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
+\includegraphics[width=4.5in,height=\textheight,keepaspectratio]{figures/umi_tidal_track.png}
 \caption[Ursa Minor tidal tracks]{Similar to
@@
 \midrule
-$\sigma_{\V, i}\,/\,\kms$ & 10.0 & 10.9\\
-$\sigma_{\V, f}\,/\,\kms$ & 8.2 & 8.5\\
-fractional stellar mass loss & $0.00015$ & 0.039\\
-$R_{h, i}\,/\,\kpc$ & 0.135 & 0.151\\
-$R_{h, f}\,/\,\kpc$ & 0.169 & 0.191\\
-break radius & 197 arcmin, 3.7 kpc & 204 arcmin, 3.8 kpc\\
+$\sigma_{\V, i}\,/\,\kms$ & 11.0 & 12.1\\
+$\sigma_{\V, f}\,/\,\kms$ & 8.8 & 9.1\\
+fractional stellar mass loss & $0.00138$ & 0.068\\
+$R_{h, i}\,/\,\kpc$ & 0.201 & 0.20\\
+$R_{h, f}\,/\,\kpc$ & 0.212 & 0.257\\
+break radius & 210 arcmin, 4.0 kpc & 218 arcmin, 4.1 kpc\\
 \bottomrule
@@
 because there are more loosely-bound stars, the Plummer model loses
-nearly 4\% of its initial stellar mass to tides
+nearly 7\% of its initial stellar mass to tides
 (Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_lmc_xyzr_orbits.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_lmc_xyzr_orbits.png}
 \caption[Ursa Minor orbits with LMC]{Orbits of Ursa Minor with (orange)
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}
 \caption[Long term orbital uncertainties]{The long-term orbital history
-of Sculptor (\textbf{top}) and UMi (\textbf{bottom}) is uncertain. In
-both panels, light, transparent lines represent randomly sampled orbits
-of the satellites (following Section~\ref{sec:scl_smallperi}) in three
-different LMC/MW mass models from \citet{vasiliev2024}. The LMC orbits
-are in solid, thick lines of the corresponding colour. The L2M11 has a
-lighter LMC mass, and the L3M10 model has a lighter MW mass than our
-fiducial L3M11 LMC model.}\label{fig:scl_orbit_lmc_uncert}
+of Sculptor (\textbf{top}) and Ursa Minor (\textbf{bottom}) is
+uncertain. In both panels, light, transparent lines represent randomly
+sampled orbits of the satellites (following
+Section~\ref{sec:scl_smallperi}) in three different LMC/MW mass models
+from \citet{vasiliev2024}. The LMC orbits are in solid, thick lines of
+the corresponding colour. The L2M11 has a lighter LMC mass, and the
+L3M10 model has a lighter MW mass than our fiducial L3M11 LMC
+model.}\label{fig:scl_orbit_lmc_uncert}
 \end{figure}

### 2025-10-20 thesis/compile.sh

@@ -8,8 +8,8 @@ pandoc $pandocoptions -o data.tex data.md
 pandoc $pandocoptions -o methods.tex methods.md
 pandoc $pandocoptions -o results.tex results.md
 pandoc $pandocoptions -o discussion.tex discussion.md
-pandoc $pandocoptions -o acknowledgments.tex acknowledgments.md
-pandoc $pandocoptions -o research_acknowledgments.tex research_acknowledgments.md
+pandoc $pandocoptions -o acknowledgements.tex acknowledgements.md
+pandoc $pandocoptions -o research_acknowledgements.tex research_acknowledgements.md
 
 pandoc $pandocoptions -o extra_density.tex extra_density.md --top-level-division=chapter
 pandoc $pandocoptions -o extra_rv_models.tex extra_rv_models.md --top-level-division=chapter

### 2025-10-21 thesis/extra_numerical_convergence.tex
thesis updates
thesis text to kim

@@
 many small time-steps). As a remedy, many N-body codes use a
-``softened'' gravitational force law, where the force weakens when
-closer to a particle than a ``softening length''. The choice of
+\emph{softened} gravitational force law, where the force weakens when
+closer to a particle than a \emph{softening length.} The choice of
 softening length ideally balances resolution and computational speed.
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
-\caption[Isolation method convergence]{A comparison of the final
-profiles using different simulation methods for Sculptor's \smallperi{}
-model. The benchmark model is our fiducial \(10^7\) particle run, and
-all other models use \(10^5\) particles, with their ``converged radius''
-marked by the black arrow and softening by the vertical bar(s).
-\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
-lengths than the fiducial. \textbf{Bottom:} Models with more precise
-timestep accuracy and gravitational force accuracy. \textbf{Summary:}
-Except for the model with a larger softening length, all simulations
-agree within uncertainties.}\label{fig:methods_convergence}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/orbit_converg_methods.png}
+\caption[Numerical methods convergence]{Similar to
+Fig.~\ref{fig:numerical_convergence}, except a comparison of the final
+velocity profiles using different simulation methods for Sculptor's
+\smallperi{} model. The benchmark model is our fiducial \(10^7\)
+particle run, and all other models use \(10^5\) particles, with their
+``converged radius'' marked by the black arrow and softening by the
+vertical bar(s). \textbf{Left:} Models with \(\sqrt{10}\) larger and
+smaller softening lengths than the fiducial. \textbf{Right:} Models with
+more precise timestep accuracy and gravitational force
+accuracy.}\label{fig:methods_convergence}
 \end{figure}

### 2025-10-21 thesis/extra_results.tex
thesis text to kim

@@
 \smallperi{} orbit with different initial halo parameters. The heavier
-halo has \(\vmax=43\) and \(\rmax=7\), whereas the lighter halo has
-\(\vmax=25\) and \(\rmax=2.5\). While the less concentrated halo loses
-more mass, the halo evolves to similar final structural parameters.
+halo has \(\vmax=43\,\kms\) and \(\rmax=7\,\kpc\), whereas the lighter
+halo has \(\vmax=25\,\kms\) and \(\rmax=2.5\,\kpc\). While the less
+concentrated halo loses more mass, the halo evolves to similar final
+structural parameters.
 
@@
 
-Radial velocity anisotropy may cause halos to disrupt faster
-\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
-velocity anisotropy, we initialize a model with a velocity anisotropy
-with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
-to \(\beta=1\) and infinity, with scale length \(4r_s\).
+Velocity anisotropy quantifies the number of radial versus circular
+orbits. Specifically, the anisotropy parameter,
+\(\beta = 1 - \frac{\sigma_{\phi}^2 + \sigma_\theta^2}{2\sigma_{r}^2}\)
+where \(\sigma^2\) is the velocity dispersion in the spherical \(r\),
+\(\phi\), or \(\theta\) direction. Radial velocity anisotropy may cause
+halos to disrupt faster \citep[e.g.][]{chiang+bosch+schive2024}. To test
+the effects of moderate velocity anisotropy, we initialize a model with
+a velocity anisotropy with an Osipkov-Merritt profile rising from
+\(\beta=0.2\) at the centre to \(\beta=1\) and infinity, with scale
+length \(4r_s\).
 
@@
 We create an oblate initial dark matter halo using \agama{}.
-Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
-isolation) isodensity contours of our model. The initial snapshot is
-sampled from the distribution function \begin{equation}{
+Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after
+\(5\,\Gyr\) in isolation) isodensity contours of our model. The initial
+snapshot is sampled from the distribution function \begin{equation}{
 \begin{split}
-f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right]
+f({\bf J}) = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h({\bf J})}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g({\bf J})}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g({\bf J})}{J_{\rm cutoff}} \right)^\zeta\right]
 \end{split},
 }\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
-\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
-\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
+\(g({\bf J}) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
+\(h({\bf J}) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
 \texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
@@
 \caption[Oblate halo projected density snapshots]{The initial and final
-(after 9Gyr of tidal evolution) projected density profiles for the
-oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
-\(z'\) the major and minor axes. The contours are drawn assuming normal
-smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
+(after \(9\,\Gyr\) of tidal evolution) projected density profiles for
+the oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\)
+and \(z'\) the major and minor axes. The contours are drawn assuming
+normal smoothing of 0.4 kpc and are log-spaced with intervals of 0.1
 dex.}\label{fig:oblate_i_f}
@@
 \([67.83, -352.2, 110.3]\,\kpc\) and velocity of
-\([-3.68, 30.79, -22.77]\,\kpc\).
+\([-3.68, 30.79, -22.77]\,\kms\).
 
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_impact_orbits.pdf}}
-\caption[Sculptor MW impact orbit]{The orbit of Sculptor for the
-MW-impact model. The point (dotted) and n-body (solid) diverge by
-\(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
+\caption[Sculptor MW impact orbit]{Similar to
+Fig.~\ref{fig:scl_lmc_orbits_effect}, except for the orbit of Sculptor
+for the MW-impact model. The point (dotted) and N-body (solid) diverge
+by \(\sim 50\,\kpc\) at early times.}\label{fig:scl_mw_impact_orbit}
 \end{figure}
@@
 orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
-detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
-motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
-such a gradient would require tracing stars across several degrees on
-the sky.
+detectible gradients in radial velocities (\(\sim 10\,\kms\) over 5
+degrees) and proper motions (mostly for UMi, of \(\sim 0.1\,\masyr\)
+over 5 degrees). However, detecting such a gradient would require
+tracing stars across several degrees on the sky.
 
@@
 
-While we have compared the evolution of orbits in the MW potential for
-simplicity, because the tidal evolution of Scl is very weak in the LMC
-model, the differences between models become even more slight. And while
-UMi has a lower pericentre than Scl, the differences between halo
-structure should apply similarly to UMi.
+While we compared the evolution of orbits of Scl in the MW-only
+potential for simplicity, we expect the differences between models in
+the MW and LMC potential to be more slight due to the weaker tidal
+evolution. And while UMi has a lower pericentre than Scl, the
+differences between halo structure should apply similarly to UMi. We
+conclude that our assumptions regarding orbits, halo concentration,
+cuspyness, anisotropy, and ellipticity likely do not dramatically change
+our conclusions.

### 2025-10-21 thesis/extra_rv_models.tex
thesis updates
thesis text to kim

@@
 \caption[Velocity dispersion fits]{\textbf{Left:} Velocity histogram of
-Scl and UMi in terms of projected-correction GSR LOS velocity. Black
-points with error bars are from the sample of LOS-velocity members, and
-green lines represent MCMC fits to the velocity dispersion.
-\textbf{Right:} The observed velocity dispersion gradient (black points)
-in 10 equal number bins and the derived slopes for MCMC samples (green
-lines). Scl shows moderate evidence for an increasing velocity
-dispersion with radius, but Scl's velocity dispersion appears to flatten
-outside of \(R_h\). UMi's evidence for a radial gradient is
-weaker.}\label{fig:rv_hists}
+Scl and UMi in terms of projected-correction LOS velocity. Black points
+with error bars are a histogram of LOS-velocity members, and green lines
+represent MCMC fits to the velocity dispersion. \textbf{Right:} The
+observed velocity dispersion gradient (black points) in 10 equal number
+bins and the derived slopes for MCMC samples (green lines). Scl shows
+moderate evidence for an increasing velocity dispersion with radius, but
+Scl's velocity dispersion appears to flatten outside of \(R_h\). UMi's
+evidence for a radial gradient is weaker.}\label{fig:rv_hists}
 \end{figure}
@@
 \centering
-\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.pdf}}
 \caption[A possible velocity gradient in Sculptor]{\textbf{Top} members
 of Sculptor plotted in the tangent plane, coloured by corrected velocity
-difference from mean \(v_z - \bar v_z\). The black ellipse marks the
-half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
-arrows mark the proper motion (PM, GSR frame) and derived velocity
-gradient (rot) vectors (to scale). \textbf{Bottom}: The corrected LOS
-velocity along the most likely rotational axis. RV members are black
-points, the systematic \(v_z\) is the horizontal grey line, blue lines
-represent the (projected) gradient from MCMC samples, and the orange
-line is a rolling median (with a window size of
+difference from mean \(v_{\rm gsr}' - \bar v_{\rm gsr}'\). The black
+ellipse marks the half-light radius in Fig.~\ref{fig:scl_selection}. The
+black and green arrows mark the proper motion (PM, GSR frame) and
+derived velocity gradient (rot) vectors (to scale). \textbf{Bottom}: The
+corrected LOS velocity along the most likely rotational axis. RV members
+are black points, the systematic \(v_{\rm gsr}''\) is the horizontal
+grey line, blue lines represent the (projected) gradient from MCMC
+samples, and the orange line is a rolling median (with a window size of
 50).}\label{fig:scl_velocity_gradient_scatter}
+\end{figure}
+
+\begin{figure}
+\centering
+\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_rv_scatter_gradient.pdf}}
+\caption[velocity]{Similar to
+Fig.~\ref{fig:scl_velocity_gradient_scatter} except for Ursa Minor. Ursa
+Minor does not have clear evidence of a velocity
+gradient.}\label{fig:umi_velocity_gradient_scatter}
 \end{figure}

### 2025-10-21 thesis/introduction.tex
thesis text to kim

@@
 1000 \(M_\odot/ L_\odot\) \citep[implying \(\sim1000\) times more dark
-matter than stellar matter, e.g.,][]{simon+geha2007, hayashi+2023}.
+matter than stellar mass, e.g.,][]{simon+geha2007, hayashi+2023}.
 
@@
 formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW satellites
-contain old stellar populations which are \emph{relics} from the early
+contain stellar populations which are \emph{relics} from the early
 universe, consisting of many of the oldest and most metal-poor stars
 known \citep{simon2019}. Understanding the properties of dwarf galaxies
-thus has implications across astronomy, from cosmological structure
-formation on small scales to the formation of metal-poor stellar
-populations.
+thus has implications across astronomy, from small-scale cosmological
+structure formation to the origins of the first stars.
 
@@
 introduce the puzzle posed by the extended stellar density profiles of
-Sculptor and Ursa Minor. We end with a brief roadmap to the remainder of
-this dissertation.
+Sculptor and Ursa Minor. Then, we discuss the theory of tidal evolution.
+We end with a brief roadmap to the remainder of this dissertation.
 
@@
 The earliest spectroscopic work hinted that dwarf galaxies may contain
-substantial amounts of dark matter. From early velocity dispersion
+substantial amounts of dark matter. From velocity dispersion
 measurements for dwarf spheroidal (dSph) galaxies, inferred
@@
 holes \citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
-\citep{milgrom1995}. Over time, however, a consensus developed where the
-high mass-to-light ratios of dwarf galaxies were due to the presence of
-a dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
-Since then, the properties of dwarf galaxies have played an increasingly
-important role in our understanding of the clustering of dark matter on
-small scales \citep[e.g.,][]{bullock+boylan-kolchin2017, sales+2022}.
+\citep{milgrom1995}. Over time, a consensus developed where the high
+mass-to-light ratios of dwarf galaxies were due to the presence of a
+massive dark matter halo
+\citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}. Since then, the
+properties of dwarf galaxies have played an increasingly important role
+in our understanding of the clustering of dark matter on small scales
+\citep[e.g.,][]{bullock+boylan-kolchin2017, sales+2022}.
 
@@
   generally, systems inconsistent with Newtonian dynamics of visible
-  matter alone, \citet{willman+strader2012}.} In contrast, star clusters
+  matter alone \citep{willman+strader2012}.} In contrast, star clusters
 (like globular clusters) have no clear evidence for dark matter. The
 boundary between these two classes blurs for faint, compact stellar
-associations. Systems with shared characteristics of globular clusters
-and dwarf galaxies, and without evidence as to their nature, are known
-as ``ambiguous'' systems \citep[e.g.,][]{smith+2024}.
+associations. Systems with characteristics of both globular clusters and
+dwarf galaxies are known as ``ambiguous'' systems
+\citep[e.g.,][]{smith+2024}.
 
@@
 on luminosity. Local \textbf{bright dwarfs} with magnitudes
-\(-14 \gtrsim M_V \gtrsim  -18\) or stellar masses
-\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
+\(-14 \gtrsim M_V \gtrsim  -18\) or stellar masses\footnote{Assuming
   stellar mass-to-light ratio of \(1 \Mo / L_{\odot}\), may be
-  \(\sim 2 \Mo/L_\odot\) for older populations.}, often exhibit
+  \(\sim 2 \Mo/L_\odot\) for older populations.}
+\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\), often exhibit
 irregular morphologies and recent star formation.
@@
 \(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
-these systems are old, non-star-forming, gas-poor, and spheroidal. All
+these systems are old, non-star-forming, gas-poor, and elliptical. All
 Milky Way satellites discovered before digital sky surveys are
-classicals, and these systems remain among the best studied. The
+classicals, and these systems remain among the best studied dwarfs. The
 \textbf{ultra-faint dwarfs} occupy the very faintest magnitude range
@@
 Most well-studied dwarf galaxies lie in the vicinity of the Milky Way,
-the Local Group of galaxies. The Local Group is defined as the group
-consisting of galaxies within \(\sim 1\) Mpc from the MW-Andromeda
+the \emph{Local Group} of galaxies. The Local Group is defined as the
+group consisting of galaxies within \(\sim 1\) Mpc from the MW-Andromeda
 centre \citep[e.g.,][ and references therein]{mcconnachie2012}. Today,
@@
 including dwarf galaxies, globular clusters, and ambiguous systems. This
-nearby population of dwarf galaxies is amenable to resolved, detailed
-studies aimed at understanding their nature.
+nearby population of dwarf galaxies is amenable to resolved studies
+aimed at investigating their detailed history and structure.
 
@@
 Light and matter pass through dark matter unimpeded---in this sense,
-dark matter is transparent. Dark matter is also commonly assumed to be
+dark matter is transparent. Dark matter is also assumed to be
 \emph{cold}, i.e., with typical velocities much smaller than the speed
 of light in the early universe. If dark matter is cold, then it should
-condense on all scales, forming non-linear structures (or \emph{halos})
-from the size of galaxy clusters to smaller than the faintest dwarf
-galaxies. Implications of dark matter properties include cosmological
-structure, galaxy formation, and galaxy interactions.
+condense on all scales, from the size of galaxy clusters to smaller than
+the faintest dwarf galaxies. Implications of dark matter properties
+include cosmological structure, galaxy formation, and galaxy
+interactions.
 
@@
 observations of the universe stem from the cosmic microwave background
-(CMB)---revealing a nearly uniform, isotropic blackbody emission
+(CMB)---displaying a nearly uniform, isotropic blackbody emission
 \citep[e.g.,][]{ryden2016}. But tiny perturbations in the CMB,
@@
 coupled to radiation and resisted collapse. Dark matter, only influenced
-by gravity instead, freely collapsed into the first structures. Mass
+by gravity, freely collapsed into the first structures. Mass
 perturbations sufficiently small and overdense become self-gravitating
@@
 of the density profile increases gradually with radius, and at large
-radii the density falls off like \(\rho \sim 1/r^3\). The blue solid
+radii the density falls off like \(\rho \sim 1/r^3\). The solid blue
 curve in Fig.~\ref{fig:nfw_density} shows an example NFW halo.
@@
 concentration with mass or, equivalently, the dependence of \(\vmax\) on
-\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
+\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2014}. The left panel of
 Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration from
@@
 shows an example exponential stellar component for the Fornax dwarf
-galaxy with its surrounding dark matter NFW halo \citep[with parameters
-matching the][ and \citet{fattahi+2018} relations]{ludlow+2016}. Where
-the stars are densest, the dark matter remains nearly an order of
+galaxy with its surrounding dark matter NFW halo {[}with parameters
+matching the \citet{ludlow+2016}; \citet{fattahi+2018}; relations{]}.
+Where the stars are densest, the dark matter remains nearly an order of
 magnitude higher in density. Stars make a small contribution to the
@@
 absolute all-sky reference frame is derived from which parallax and
-proper motions are derived. In addition to astrometry, \emph{Gaia}
+proper motions are calculated. In addition to astrometry, \emph{Gaia}
 measures photometry in the wide \emph{G} band (330--1050nm) and colours
@@
 }\end{equation} and \(\Sigma_{\rm K}=0\) for \(R \geq R_t\). In much of
-the older literature, \(R_t\) was interpreted as a ``tidal radius'',
+the older literature, \(R_t\) was interpreted as a ``tidal radius,''
 after an analogous interpretation for globular clusters
@@
 theorizing that dwarf spheroidals may have evolved from exponential disk
-galaxies\footnote{While studied for far longer, the exponential origin
-  of disk galaxies is no better understood. Ideas range from scattering
-  of stars \citep{elmegreen+struck2013, wu+2022} to angular momentum
-  transport, to disk viscosity-driven radial gas flows
-  \citep{lin+pringle1987, wang+lilly2022} or spherical collapse
-  \citep{freeman1970}.} and maintained a similar light profile. Later,
+galaxies and maintained a similar light profile. Later,
 \citet{read+gilmore2005} showed that exponential profiles may originate
 from mass loss during the evolution of dwarf galaxies. Tides are a
-possible mechanism for this transformation---the ``tidal stirring
-hypothesis'' \citep{mayer+2001a, klimentowski+2009}. However, the
-theoretical origin of exponential disks is unclear.
+possible mechanism for this transformation---the \emph{tidal stirring
+hypothesis} \citep{mayer+2001a, klimentowski+2009}. However, the
+theoretical origin of exponential disks is also unknown.\footnote{While
+  studied for far longer, the exponential origin of disk galaxies is no
+  better understood. Ideas range from scattering of stars
+  \citep{elmegreen+struck2013, wu+2022} to angular momentum transport,
+  to disk viscosity-driven radial gas flows
+  \citep{lin+pringle1987, wang+lilly2022} or spherical collapse
+  \citep{freeman1970}.}
 
@@
 Plummer profiles, but most of their fits would be consistent with a
-single-component exponential. As a result, it has become conventional to
+single-component exponential.\footnote{A single-component exponential
+  profile is close to their ``Steeper'' profile in
+  \citet{moskowitz+walker2020} over the range of their data. However,
+  most of their galaxies have little evidence to prefer the Steeper or
+  Plummer density profile.} As a result, it has become conventional to
 assume an exponential density profile to describe dwarf galaxies in
@@
 
-UMi has also attracted similar suspicions. \citet{hargreaves+1994} first
+UMi has attracted similar suspicions. \citet{hargreaves+1994} first
 detected a velocity gradient in UMi, suggestive of tidal disruption.
 Later, \citet{martinez-delgado+2001} found stars far beyond the
-King-profile ``tidal radius'', aligned with UMi's elongation, consistent
+King-profile ``tidal radius,'' aligned with UMi's elongation, consistent
 with tidally-stripping simulations in
@@
 ``kink'' in the density profile of each galaxy. They spectroscopically
-follow up some of the most distant stars, finding members as far as 6
-and 12 half-light radii from the centre of each dwarf. If these dwarfs
-had exponential profiles, like Fornax, then these far-outlying stars
-should be much rarer.
+follow up distant stars, finding members as far as 6 and 12 half-light
+radii from the centre of each dwarf. If these dwarfs had exponential
+profiles, like Fornax, then these far-outlying stars should be much
+rarer.
 
@@
 \caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
-References are:(1).
-Muñoz et al.
+References are: (1) Muñoz et al.
 (2018, Sérsic fit), (2) Tran et al.
@@
 analytic host potential, achieving excellent numerical convergence. For
-instance, the simulations we describe later in this thesis reach three
-times higher resolution than Aquarius \citep{springel+2008} with 400
-times fewer particles. Idealized simulations make numerous
-simplifications, neglecting mergers, cosmological context, mass
-assembly, and often baryonic physics
+example, the simulations we describe later reach three times higher
+resolution than Aquarius \citep{springel+2008} with 400 times fewer
+particles. Idealized simulations make numerous simplifications,
+neglecting mergers, cosmological context, mass assembly, and often
+baryonic physics
 \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
@@
 Idealized simulations predict clear properties of tidally disrupting
-dwarf spheroidal galaxies. Tidally stripped stars form tidal
-streams---stellar tails with a bulk velocity gradient
+dwarf spheroidal galaxies. Tidally stripped stars form \emph{tidal
+streams}---stellar tails with a bulk velocity gradient
 \citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
@@
 initial conditions and Galactic tides. Most relevantly,
-\citet{iorio+2019} also tailored simulations to Scl, finding weak tidal
-effects.
+\citet{iorio+2019} also tailored simulations to Scl, finding weak
+Galactic tidal influence.
 
@@
 r_{\rm break} = C\,\sigma_{v}\,\Delta t
-}\end{equation} where the empirical constant \(C \approx 0.55\).
+}\end{equation} where the empirical constant is \(C \approx 0.55\).
 \(r_{\rm break}\) describes where the dynamical timescale becomes longer
@@
 we consider a toy model. We evolve an N-body realization of an NFW
-subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
-NFW static host halo with \(\rmax=25\,\kpc\) and
+subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\,\kms\) orbiting a
+static NFW host halo with \(\rmax=25\,\kpc\) and
 \(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
@@
 the inner dark matter halo. We start the model at apocentre on an orbit
-with a pericentre of 20,kpc and apocentre of 100,kpc. See
+with a pericentre of \(20\,\kpc\) and apocentre of \(100\,\kpc\). See
 Section~\ref{sec:methods} for a more detailed description of our
@@
 radius also marks where the mean 3D radial velocity of the stars (with
-respect to the galaxy's centre) transitions from 0 to positive --- the
-galaxy is out of equilibrium outside \(r_{\rm break}\).
+respect to the galaxy's centre) becomes non-zero --- the galaxy is out
+of equilibrium outside \(r_{\rm break}\).
 
 To first order, the final density profile of this toy simulation indeed
-resembles the outer excess in Sculptor and Ursa Minor. This thesis aims
-to investigate whether a tidal explanation like this is viable given the
-current properties of each galaxy.
+resembles the outer excess in Sculptor and Ursa Minor, motivating this
+thesis.
 
@@
 for the initial (dotted) and final (solid) simulation snapshot.
-\textbf{Bottom right}: the mean 3D radial velocity (dot product of
+\textbf{Bottom right}: the mean 3D radial velocity (the dot product of
 relative position and velocity relative to dwarf centre) as a function

### 2025-10-21 thesis/methods.tex
thesis text to kim

@@
 potential, \begin{equation}{
-\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}},
+\Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
 }\end{equation} where \(a\) is the disc radial scale length, \(b\) is
@@
 (bottom). The Milky Way is at the centre with the disk lying in the
-\(x\)--\(y\) plane. Our selected N-body's \smallperi{} orbit is plotted
-in black, and light green transparent lines represent orbits sampled
-over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
-Sculptor is well-constrained in this potential, and it is unlikely to
-achieve a smaller pericentre than the \smallperi{}
-orbit.}\label{fig:scl_orbits}
+\(x\)--\(y\) plane and the yelow star marks the Sun's location. The
+black line shows our N-body's \smallperi{} orbit and light green
+transparent lines represent orbits sampled over Sculptor observables in
+Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is well-constrained
+in this potential, and it is unlikely to achieve a smaller pericentre
+than the \smallperi{} orbit.}\label{fig:scl_orbits}
 \end{figure}
@@
 \caption[Orbit initial conditions]{The orbital initial conditions for models presented in this work.
-The observables represent the medians from orbital integration used to derive the orbits.
-Instead, the initial position and velocity represent the initialization of the actual N-body model.
-The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects.
+The observables record the intended final position of the N-body model derived from point-mass orbit quantiles.
+Instead, the initial position and velocity represent the initial conditions for the actual N-body model.
+The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects.
 We describe the \texttt{LMC-flyby} orbit in Section \ref{sec:scl_lmc}.
@@
 \midrule
-distance / kpc & 82.6 & 73.1 & 64.6\\
+distance $/$ kpc & 82.6 & 73.1 & 64.6\\
 $\pmra / \masyr$ & 0.134 & 0.137 & -0.158\\
 $\pmdec / \masyr$ & -0.198 & -0.156 & 0.050\\
-LOS velocity / $\kms$ & 111.2 & 111.2 & -245.75\\
+LOS velocity $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
 $t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
@@
 ${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
-$\V_{x\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
-$\V_{y\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
-$\V_{z\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
+$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
+$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
+$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
 \bottomrule
@@
 \(z=0\) mass-concentration relation, this constraint translates into a
-\(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
-galaxy.
+\(\rmax \approx 6\,\kpc\) and \(\rmax \approx 5\,\kpc\) for each galaxy.
 
-The observed velocity dispersion constrains the total mass within the
-stellar half-light radius. From the \citet{wolf+2010} mass estimator,
-the mass contained within the 3D half-light radius \(r_h\) is
+The observed velocity dispersion constrains the inner mass profile. From
+the \citet{wolf+2010} mass estimator, the mass contained within the 3D
+half-light radius \(r_h\) is
 \begin{equation}\protect\phantomsection\label{eq:wolf_mass}{
@@
 evolution for either galaxy. So, the tidal effects on stars should be
-similar for halos with similar velocity dispersions.
+similar for halos with similar velocity dispersions (See Appendix
+\ref{sec:extra_results}).
 
@@
 We record the total luminosity, stellar mass, mass-to-light ratio, dark matter halo $\vmax$ and $\rmax$, and dark matter halo virial mass $M_{200}$ and concentration $c_{\rm NFW}$.
-Uncertainties are the 16-84th percentile derived using Monte-Carlo sampling, assuming 0.035 and 0.1 dex uncertainties in the $\vmax(M_\star)$ and $c(M_{200})$ relations.
+Uncertainties are the 16-84th percentile derived using Monte Carlo sampling, assuming 0.035 and 0.1 dex uncertainties in the $\vmax(M_\star)$ and $c(M_{200})$ relations.
 }
@@
 $M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & $1.9 \times 10^{\pm 0.17}$\\
-$\vmax$ & $31\pm 3\,\kms$ & $27_{-2}^{+3}\,\kms$\\
+$\vmax$ & $31.5\pm 3\,\kms$ & $27_{-2}^{+3}\,\kms$\\
 $\rmax$ & $6 \pm 2$ kpc & $5_{-1}^{+2}$ kpc\\
-$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & $0.3_{-0.1}^{+0.2} \times 10^{10}\,\Mo$\\
+$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ Mo$ & $0.3_{-0.1}^{+0.2} \times 10^{10}\,\Mo$\\
 $c_{\rm NFW}$ & $13_{-3}^{+4}$ & $13.5_{-3}^{+4}$\\
@@
 \pandocbounded{\includegraphics[keepaspectratio]{figures/initial_halo_selection.pdf}}
-\caption[Initial halo parameter choice]{The blue circle indicates our
+\caption[Initial halo parameter choice]{The blue point indicates our
 choice of initial halo parameters for Sculptor and Ursa Minor. The grey
@@
 \citet{ludlow+2016} mass-concentration relation and \citet{fattahi+2018}
-SMHM relation, respectively. The curved orange lines represent the
-velocity dispersion of an exponential stellar component with the
-observed half-light radius of Sculptor and Ursa
-Minor.}\label{fig:scl_halos}
+SMHM relation, respectively. The curved orange lines show the velocity
+dispersion of an exponential stellar component with the observed
+half-light radius of Sculptor and Ursa Minor.}\label{fig:scl_halos}
 \end{figure}
@@
 particle-particle gravitational deflections), the force is tapered below
-a ``softening length.'' Resolution is limited by the number of particles
-and the softening length. Tree codes organize particles into a spatial
-tree, enabling the grouping of the gravitational forces from nearby
-particles. A gravitational tree code substantially reduces the required
-number of force computations, as compared to the exact, direct-summation
-method.
+a \emph{softening length}. Resolution is limited by the number of
+particles and the softening length. Tree codes organize particles into a
+spatial tree, enabling the grouping of the gravitational forces from
+nearby particles. A gravitational tree code substantially reduces the
+required number of force computations, as compared to the exact,
+direct-summation method.
 
@@
 h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
-}\end{equation} where the force is softened by a spline kernel
-\citetext{\citealp[eqs. 70--71
+}\end{equation} for a halo with \(N\) particles. \gadget{} adopts a
+spline kernel softening law \citetext{\citealp[eqs. 70--71
 in][]{springel+yoshida+white2001}; \citealp[but with the spline
@@
 numerical methods are able to approximate well an NFW halo down to a
-given innermost radius that strongly depends on resolution. The larger
-the number of particles, the smaller the radius that is effectively
-``resolved'' in a given simulation. For the Sculptor-like halo shown in
-this figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)), a
-simulation with \(10^7\) particles is needed to resolve the innermost
-100 pc. For reference, the half-light radius of Sculptor is roughly 100
-pc, which means that at least 10 million particles are needed to follow
-faithfully its tidal evolution. Vertical arrows in
-Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
-defined by \citet[their eq. 13]{power+2003} for NFW halos formed in
-cosmological N-body simulations. This radius marks the region where
-collisional effects driven by the finite number of particles used to
-describe the innermost regions of a halo become important over a Hubble
-time in a cosmological simulation. The softening length (from
+resolution-dependent radius. The larger the number of particles, the
+smaller the radius that is effectively ``resolved'' in a given
+simulation. For the Sculptor-like halo shown in this figure (with
+\(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)), a simulation with
+\(10^7\) particles is needed to resolve the innermost 100 pc. For
+reference, the half-light radius of Sculptor is \({\sim}200\,\)pc, which
+means that at least 10 million particles are needed to follow faithfully
+its tidal evolution.
+
+Vertical arrows in Fig.~\ref{fig:numerical_convergence} indicate the
+``convergence radius'' defined by \citet[their eq. 13]{power+2003} for
+NFW halos formed in cosmological N-body simulations. This radius marks
+the region where collisional effects driven by the finite number of
+particles used to describe the innermost regions of a halo become
+important over a Hubble time in a cosmological simulation. Our effective
+resolution is nearly the converged radius. The softening length (from
 Eq.~\ref{eq:softening_length}) is typically a few times smaller than the
@@
 \caption[Numerical convergence of the N-body simulation]{Numerical
-convergence test for circular velocity as a function of log radius for
+convergence test for circular velocity as a function of radius for
 simulations with different total numbers of particles in isolation.
@@
 when the halo is closest to its present-day observed position in the MW
-halo.
+halo.\footnote{Specifically, the snapshot which best agrees (as measured
+  by \(\chi^2\)) with the intended final position (in
+  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
+  uncertainties. We use \(R_h\) as the uncertainty in \(\alpha\) and
+  \(\delta\).}
 
@@
 
-To accurately follow the evolution of a halo, it is important to
-determine the centre of the self-bound halo remnant at each time chosen
-for analysis. We use a shrinking-spheres centre method inspired by
-\citet{power+2003}. First, we start with an initial centre estimate
-using all bound particles from the previous snapshot. Then, we calculate
-the distance of all particles from the centre, remove particles with a
-distance beyond the 0.975 quantile of the centre, and recalculate the
-centre of mass. The procedure is repeated until the selection radius is
-less than \textasciitilde1kpc or fewer than 0.1\% of particles remain.
-After a centre has been chosen, we remove all unbound particles based on
-the \gadget{} calculated potential of the halo. For all future
-timesteps, we consider only particles retained from the previous
-iteration.
+To accurately follow the evolution of a halo, we must determine the
+centre of the self-bound halo remnant at each time chosen for analysis.
+We use a shrinking-spheres centre method inspired by \citet{power+2003}.
+First, we start with an initial centre estimate using all bound
+particles from the previous snapshot. Then, we calculate the distance of
+all particles from the centre, remove particles with a distance beyond
+the 0.975 quantile of the centre, and recalculate the centre of mass.
+The procedure is repeated until the selection radius is less than
+\textasciitilde1kpc or fewer than 0.1\% of particles remain. We then
+remove all unbound particles based on the \gadget{} calculated potential
+of the halo and recalculate the centre. For all future timesteps, we
+consider only particles retained from the previous iteration.
 
 The statistical centring uncertainty for the full resolution (\(10^7\)
-particle) isolation run is of order 0.003 kpc, but fluctuations are
-observed of order \(\sim 0.03\,\kpc\). This is about three times the
-softening length but is less than the numerically converged radial
-scale.
+particle) isolation run is \({\sim}0.003\) kpc, but fluctuations are
+observed \({\sim} 0.03\,\kpc\). This is about three times the softening
+length but is still less than the numerically converged radial scale.
 
@@
 \({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 v^2\). If we
-know the distribution function \(f({\cal E})\) (the phase-space density
-as a function of energy), then we assign a stellar weight to a given
-particle with energy \({\cal E}\) using \begin{equation}{
+know the distribution function\footnote{i.e., the phase-space density of
+  particles. Note that since \(f\) is formally defined in 6D phase
+  space, there is an additional ``density of states'' term in order to
+  calculate the histogram of particles with \({\cal E}\),
+  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
+then we assign a stellar weight to a given particle with energy
+\({\cal E}\) using \begin{equation}{
 P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-}\end{equation} While \(f({\cal E})\) is a phase-space density, the
-differential energy distribution includes an additional \(g({\cal E})\)
-occupation term \citep{BT1987}. We use Eddington inversion to find the
-distribution function, \citep[eq. 4-140b in][]{BT1987} \begin{equation}{
-f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
-}\end{equation} In practice the right, boundary term is zero.\footnote{If,
-  at large \(r\), \(\rho \propto r^{-n}\) with \(n>1\), then \(\Psi\)
-  falls off no steeper than \(\Psi \sim r^{-1}\). So \(d\rho / d\Psi\)
-  decays no less rapidly than \$ r\^{}\{-n+1\}\$ which \(\rightarrow0\)
-  at large \(r\).} We take \(\Psi\) from the underlying assumed analytic
-dark matter potential. \(\rho_\star\) can be calculated from the surface
-density, \(\Sigma_\star\), via the inverse Abel transform.
+}\end{equation} We use Eddington inversion to find the distribution
+function, \begin{equation}{
+f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
+}\end{equation} \citep[eq. 4-140b in][]{BT1987}. In practice the right,
+boundary term is zero.\footnote{If, at large \(r\),
+  \(\rho \propto r^{-n}\) with \(n>1\), then \(\Psi\) falls off no
+  steeper than \(\Psi \sim r^{-1}\). So \(d\rho / d\Psi\) decays no less
+  rapidly than \(r^{-n+1}\) which \(\rightarrow0\) at large \(r\).} We
+take \(\Psi\) from the underlying assumed analytic dark matter
+potential. \(\rho_\star\) can be calculated from the surface density,
+\(\Sigma_\star\), via the inverse Abel transform.
 
@@
 profiles are solid. The green cross marks the present-day half-light
-radius and velocity dispersion, and the black band represents the
-1-sigma mean density of the MW at pericentre across orbits. Initial
-conditions are stable in isolation, and mass is dominated by dark
-matter.}\label{fig:scl_umi_initial_isolation}
+radius and velocity dispersion. To calculate the velocity profiles,
+unbound particles are iteratively removed, recalculating the potential
+at each step assuming spherical
+symmetry.}\label{fig:scl_umi_initial_isolation}
 \end{figure}

### 2025-10-21 thesis/results.tex
thesis text to kim

@@
 the MW on its dark matter and stellar components. Next, we consider how
-accounting for the LMC may affect our conclusions. We then similarly
-analyze UMi, considering in turn the dark matter evolution, stellar
-evolution, and orbital effects of the LMC.
+accounting for the LMC may affect our conclusions. We then analyze UMi,
+considering in turn the dark matter evolution, stellar evolution, and
+orbital effects of the LMC.
 
@@
 initial mass (see Table~\ref{tbl:scl_sim_results}). However, the inner
-structure is not affected, as the Jacobi radius is over 3kpc, outside of
-the initial and final \(\rmax\) (see Table~\ref{tbl:scl_sim_results} and
+structure is not affected, as the Jacobi radius is over \(3\,\kpc\),
+outside of \(\rmax\) (see Table~\ref{tbl:scl_sim_results} and
 Fig.~\ref{fig:scl_tidal_track}). Thus, tides may remove significant
@@
 circular velocity profiles, and blue and orange lines show the dark
-matter and stellar (2D exponential) profiles. The points represent the
-evolution of the maximum circular velocity, and the dotted black line
-shows the tidal track from \citet{EN2021}. To calculate the velocity
-profiles, unbound particles are iteratively removed, recalculating the
-potential at each step assuming spherical
-symmetry.}\label{fig:scl_tidal_track}
+matter and stellar (2D exponential) profiles. The blue points represent
+the evolution of the maximum circular velocity, and the dashed black
+line shows the tidal track from \citet{EN2021}. The grey diagonal line
+marks three times the MW mean enclosed density at pericentre, whose
+intersection with the initial halo is the Jacobi
+radius.}\label{fig:scl_tidal_track}
 \end{figure}
@@
 \midrule
-pericentre & $53\pm3$ & 42\\
-apocentre & $102\pm3$ & 94.4\\
-time of last pericentre / Gyr & $-0.45 \pm 0.2$ & -0.47\\
+pericentre $/$ kpc & $53\pm3$ & 42\\
+apocentre $/$ kpc & $102\pm3$ & 94.4\\
+time of last pericentre $/$ Gyr & $-0.45 \pm 0.2$ & -0.47\\
 number of pericentres & 5–6 & 6\\
-Jacobi radius / kpc & $4.5 \pm 0.3$ & 3.5\\
-Jacobi radius / arcmin & $186\pm12$ & 148\\
-final heliocentric distance / kpc & $83.2\pm2$ & 81.6\\
+Jacobi radius $/$ kpc & $4.5 \pm 0.3$ & 3.5\\
+Jacobi radius $/$ arcmin & $186\pm12$ & 148\\
+final heliocentric distance $/$ kpc & $83.2\pm2$ & 81.6\\
 $\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.695\\
@@
 The Jacobi and break radii further support the conclusion that tidal
-effects should not be apparent in the observed stellar component. As
-calculated for this model (see
+effects should not be apparent. As calculated for this model (see
 Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
-break and Jacobi radii both fall outside of \(\sim 100\) arcminutes for
-either stellar component. Indeed, the stellar component only begins to
-deviate from an exponential profile around this break radius
+break and Jacobi radii both fall outside \(\gtrsim 100\) arcminutes,
+beyond the reach of our data. Indeed, the stellar component only begins
+to deviate from an exponential profile around the break radius
 (Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
 Since no orbits of Scl produce significantly smaller break or Jacobi
-radii, it is unlikely tha any orbit would produce an observable density
+radii, it is unlikely that any orbit would produce an observable density
 excess.
@@
 is consistent with adiabatic expansion due to the reduction of the total
-mass \citep[e.g.,][]{stucker+2023}. In addition, the break and Jacobi
-radii are \(\gtrsim 100\) arcminutes on the sky---tidal signatures would
-be beyond the reach of our data. Altogether, Galactic tides negligibly
-impact Scl's stellar component.
+mass \citep[e.g.,][]{stucker+2023}. Altogether, Galactic tides
+negligibly impact Scl's stellar component.
 
@@
 initial (dotted) and final (solid) stellar density profiles as compared
-to the observed stellar density profile. Arrows mark the half-light
-(\(R_h\)), break, and Jacobi radii
+to the observed stellar density profile (black points). Arrows mark the
+half-light (\(R_h\)), break, and Jacobi radii
 (Eqs.~\ref{eq:r_break}, \ref{eq:r_jacobi})
@@
 $R_{h, f}\,/\,\kpc$ & 0.189 & 0.227\\
-break radius / arcmin & $98$ & $105$\\
-break radius / kpc & 2.3 & 2.5\\
+break radius $/$ arcmin & $98$ & $105$\\
+break radius $/$ kpc & 2.3 & 2.5\\
 \bottomrule
@@
 \midrule
-pericentre & $44\pm 3$ ($29 \pm 2$) & 39 (20)\\
-apocentre & $218 \pm 8$ & –\\
-time of last pericentre / Gyr & $-0.38\pm0.01$ (-0.11) & -0.33 (-0.10)\\
+pericentre $/$ kpc & $44\pm 3$ ($29 \pm 2$) & 39 (20)\\
+apocentre $/$ kpc & $218 \pm 8$ & –\\
+time of last pericentre $/$ Gyr & $-0.38\pm0.01$ ($-0.11$) & $-0.33$ ($-0.10$)\\
 number of pericentres & 1 (1) & 1 (1)\\
-Jacobi radius / kpc & $3.3\pm0.2$ ($3.6\pm0.2$) & 2.8 (2.6)\\
-Jacobi radius / arcmin & $136 \pm 9$ ($159\pm5$) & 132 (121)\\
-final heliocentric distance / kpc & $83.2\pm2$ & 72.9\\
+Jacobi radius $/$ kpc & $3.3\pm0.2$ ($3.6\pm0.2$) & 2.8 (2.6)\\
+Jacobi radius $/$ arcmin & $136 \pm 9$ ($159\pm5$) & 132 (121)\\
+final heliocentric distance $/$ kpc & $83.2\pm2$ & 72.9\\
 $\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.928\\
@@
 
-\subsection{Tidal effects from the
-LMC}\label{tidal-effects-from-the-lmc}
+\subsection{Tidal effects of the LMC}\label{tidal-effects-of-the-lmc}
 
@@
 that the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
-first infall. However, with only 1 pericentre each for the LMC and MW,
+first infall. However, with only one pericentre each for the LMC and MW,
 the combined tides are weaker than for our initial model. In either
@@
 orbit. Dark matter evolution is more dramatic than for
-Scl.}\label{fig:umi_sim_images}
+Sculptor.}\label{fig:umi_sim_images}
 \end{figure}
@@
 \midrule
-pericentre & $37\pm3$ & 30\\
-apocentre & $83 \pm 4$ & 75\\
-time of last pericentre & $-0.96 \pm 0.07$ & -0.80\\
-number of pericentres & 7–8 & 8\\
-Jacobi radius / kpc & $3.7 \pm 0.3$ & 2.9\\
-Jacobi radius / arcmin & $184 \pm 12$ & 156\\
-final heliocentric distance & $70.1 \pm 3.6$ & 64.7\\
-${\vmax}_f / {\vmax}_i$ &  & 0.511\\
-${\rmax}_f / {\rmax}_i$ &  & 0.249\\
-fractional dm final mass &  & 0.035\\
+pericentre $/$ kpc & $37\pm3$ & $30$\\
+apocentre $/$ kpc & $83 \pm 4$ & $75$\\
+time of last pericentre $/$ Gyr & $-0.96 \pm 0.07$ & $-0.80$\\
+number of pericentres & 7–8 & $8$\\
+Jacobi radius $/$ kpc & $3.7 \pm 0.3$ & $2.9$\\
+Jacobi radius $/$ arcmin & $184 \pm 12$ & $156$\\
+final heliocentric distance $/$ kpc & $70.1 \pm 3.6$ & $64.7$\\
+${\vmax}_f / {\vmax}_i$ &  & $0.511$\\
+${\rmax}_f / {\rmax}_i$ &  & $0.249$\\
+fractional final bound mass &  & $0.035$\\
 \bottomrule
@@
 substantial tidal evolution, the Jacobi radius is still large, lying at
-\(\sim 4\,\kpc\), well beyond the final \(\rmax\).
+\({\sim} 3\,\kpc\), well beyond the final \(\rmax\).
 
@@
 Section~\ref{sec:orbit_corrections}, we recover nearly exactly the
-present-day position of Ursa Minor by changing the initial
-position/velocity by \(20\,\kpc\) and \(\sim 9\,\kms\). These
+present-day position of Ursa Minor by changing the initial position and
+velocity by \({\sim} 20\,\kpc\) and \({\sim} 9\,\kms\). These
 adjustments do not significantly affect the qualitative structure or
@@
 present-day density profile. The stellar evolution of this Plummer
-stellar component is similar (Fig.~\ref{fig:umi_plummer_i_f}). However,
-because there are more loosely-bound stars, the Plummer model loses
-nearly 7\% of its initial stellar mass to tides
-(Table~\ref{tbl:umi_sim_stars_results}), and tidal features may be
+stellar component is similar (Fig.~\ref{fig:umi_plummer_i_f}). Because
+there are more loosely-bound stars, the Plummer model loses nearly 7\%
+of its initial stellar mass to tides
+(Table~\ref{tbl:umi_sim_stars_results}). Tidal features may be
 detectable if we measure densities 2 orders of magnitude fainter than
@@
 Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
-the orbit of Ursa Minor. Predominantly, the effect is to increase the
-orbital period, apocentre, and pericentre. Yet, the orbit remains in a
-similar plane and with similar shape. As UMi is on the opposite side of
-the Galaxy of the LMC, and has a closest LMC approach of
+the orbit of Ursa Minor. Predominantly, the orbital period, apocentre,
+and pericentre all increase. Yet, the orbit remains in a similar plane
+and with similar shape to the MW-only case. As UMi is on the opposite
+side of the Galaxy of the LMC and has a closest LMC approach of
 \(\gtrsim 100\,\kpc\), this is not surprising.
@@
 required to disrupt the stellar component depends mainly on the total
-mass inside the stellar half-mass radius, and less on the actual shape
-of the density profile of the dark matter inside that radius.
+mass inside the half-light radius, and less on the inner density profile
+shape.
 
@@
 inadequacies, calculated orbits may diverge significantly from the true
-orbits of satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth
-of the Milky Way and dynamical friction imply that orbits were typically
-less bound in the past. Orbital energy and angular momentum of subhalos
-are not conserved in cosmological N-body simulations. Consequently,
-orbits in analytic potentials may overestimate the pericentre and
-underestimate the maximum tidal stress \citep[although typically not by
-enough to change our
+orbits of satellites \citep[e.g.,][]{dsouza+bell2022}. Dynamical
+friction and the mass-growth of the Milky Way both imply that orbits
+were typically less bound in the past. Orbital energy and angular
+momentum of subhalos are also not conserved in cosmological N-body
+simulations. Consequently, orbits in analytic potentials may
+overestimate the pericentre and underestimate the maximum tidal stress
+\citep[although typically not by enough to change our
 conclusions,][]{santistevan+2023, santistevan+2024}.

### 2025-10-21 thesis/data.tex

@@
 candidate members out to \textasciitilde10 half-light radii from the
-centres of some galaxies (\(R_h\)). Similar recent work has also
-included \citet{pace+li2019}; \citet{battaglia+2022};
-\citet{pace+erkal+li2022}; \citet{qi+2022}.
+centres of some galaxies (\(R_h\)). Similar recent work includes
+\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
+\citet{qi+2022}.
 
-J+24's method uses likelihoods, \({\cal L}\), representing the
-probability density that a star is consistent with either the
-foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
-\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
-of a spatial, PM, and CMD component, \begin{equation}{
+J+24's formulate membership through likelihoods, \({\cal L}\),
+representing the probability density that a star is consistent with
+either the foreground/background, \({\cal L}_{\rm bg}\), or the
+satellite galaxy, \({\cal L}_{\rm sat}\). In either case, the
+likelihoods are the product of a spatial, PM, and CMD component,
+\begin{equation}{
 {\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
@@
 of 0.2, the purity remains high when validated against spectroscopic
-line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
-however, that these purity estimates may be biased. Stars with LOS
-velocities are typically brighter and, as a result, have more precise
-\emph{Gaia} measurements. However, we find that our conclusions are
-unchanged when limiting samples to only the brightest stars. Altogether,
-the J+24 method provides a high-quality, low-contamination sample of
-dwarf galaxy candidate member stars, which we will now investigate in
-further detail.
+line-of-sight (LOS) velocities (\textasciitilde90\%, J+24).\footnote{This
+  would indicate that the J+24 model probabilities are mis-calibrated.
+  However, most LOS surveys of dwarf galaxies select brighter stars
+  (which have better \emph{Gaia} measurements) and likely members,
+  complicating the interpretation of this purity estimate.} However, we
+find that our conclusions are unchanged when limiting samples to only
+the brightest stars. Altogether, the J+24 method provides a
+high-quality, low-contamination sample of dwarf galaxy candidate member
+stars, which we will now investigate in further detail.
 
@@
 show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
-Fornax), as well as the impact of different samples, in tangent-plane
+Fornax), as well as the impact of sample selection, in tangent-plane
 coordinates, the CMD, and proper-motion space. The ``all'' sample
@@
 a radius of \(6R_h\) and less than \(1/10^6\) fall outside
-\(10R_h\)---the mere presence of very distant member stars provides
-evidence for deviations from an exponential profile.
+\(10R_h\)---the mere presence of very distant member stars implies
+likely deviations from an exponential profile.
 
@@
 for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
-three different samples from above: ``all'', ``CMD+PM'', and the
+three different samples from above: ``all,'' ``CMD+PM,'' and the
 ``fiducial'' sample. In each case, all samples coincide towards the
@@
 
-While the density profiles of all samples agree in the inner regions,
-they begin to deviate outside \(6R_h\). Outside this radius, the
-fiducial profile extends 1--2 magnitudes below the ``CMD+PM'' selection
-background. Because the fiducial profile in this region depends
-critically on the spatial likelihood, its shape is vulnerable to the
-assumed density profile (an exponential in J+24). We explore this in
-more detail in Appendix \ref{sec:density_extra}.
+The density profiles of all samples agree in the inner regions. However,
+outside \({\sim} 6R_h\), these profiles would diverge as the
+``fiducial'' sample becomes vulnerable to the assumed spatial likelihood
+(an exponential in J+24). We thus truncate the fiducial profiles near
+the ``CMD+PM'' background to avoid misleading density profiles. We
+explore this in more detail in Appendix \ref{sec:density_extra}.
 
@@
 \caption[Classical dwarf density profiles]{The density profiles of
-Sculptor, Ursa Minor, and other classical dwarfs compared to Exp2D and
-Plummer density profiles. Dwarf galaxies are scaled by the half-light
-radius and density at half-light radius. Sculptor and Ursa Minor have an
-excess of stars in the outer regions (past \(\log R/R_h \sim 0.3\))
-compared with other classical
+Sculptor, Ursa Minor, and other classical dwarfs compared to a 2D
+exponential and Plummer density profiles. Dwarf galaxies are scaled by
+the half-light radius and density at half-light radius. The residuals
+(lower panel) are with respect to a 2D exponential. Sculptor and Ursa
+Minor have an excess of stars in the outer regions (past
+\(\log R/R_h \sim 0.3\)) compared with other classical
 dwarfs.}\label{fig:classical_dwarfs_densities}

### 2025-10-21 thesis/discussion.tex

@@
 satellites, and that recent tides are an unlikely explanation for their
-extended outer structure. We therefore suggest the reason that Scl and
-UMi have extended light profiles is that they host an additional,
-extended population of stars, such as a ``stellar halo.'' With our data,
-other classical dwarfs lack evidence for similar extended populations.
-Thus, Scl and UMi have followed a divergent evolutionary path,
-motivating our discussion of extended density formation pathways here.
+extended outer structure. We therefore suggest Scl and UMi host an
+additional, extended population of stars, such as a ``stellar halo.''
+With our data, other classical dwarfs lack evidence for similar extended
+populations. Thus, Scl and UMi have followed a divergent evolutionary
+path, motivating our following discussion of extended density formation
+pathways here.
 
-In this chapter, we place our results in context with past work. We
-review that work, considering tidal effects on Scl and UMi and
-detections of peculiarities. We then consider possible origins of
-extended stellar populations in these galaxies. Scenarios include
-multi-epoch star formation, past mergers depositing stars into the
-outskirts, and even departures from \LCDM{}. We later discuss how future
-observations and simulations may distinguish among formation scenarios.
-Finally, we summarize the main conclusions of this thesis.
+In this chapter, we compare our results to past work and
+interpretations. We first review studies considering tidal effects on
+Scl and UMi and detections of peculiarities. We then consider possible
+origins of extended stellar populations in these galaxies. Scenarios
+include multi-epoch star formation, past mergers depositing stars into
+the outskirts, and even departures from \LCDM{}. We later discuss how
+future studies may distinguish formation scenarios. Finally, we
+summarize the main conclusions of this thesis.
 
@@
 
-A few studies have modelled Scl or UMi specifically, finding similar
-conclusions. \citet{iorio+2019} applied idealized N-body simulations to
-study tidal effects on Scl. They similarly found weak tidal effects,
-even for a dark-matter-free model. Most recently,
+A few studies have modelled Sculptor or Ursa Minor specifically, finding
+similar conclusions. \citet{iorio+2019} applied idealized N-body
+simulations to study tidal effects on Scl. They similarly found weak
+tidal effects, even for a dark-matter-free model. Most recently,
 \citet{tchiorniy+genina2025} also used idealized simulations tuned to
@@
 Despite these considerations, we still reach similar conclusions: tidal
-effects do not shape Scl or UMi's stellar component. A non-tidal
-explanation likely underlies the origin of these galaxies' outer density
-excess.
+effects do not shape Scl or UMi's stellar component. A non-tidal process
+likely underlies the origin of these galaxies' outer density excess.
 
@@
 
-Besides extended stellar density profiles, Scl and UMi display other
-peculiarities that may hold clues to their formation. Scl and UMi both
-host at least two distinct chemodynamic populations, as revealed through
-their photometric or metallicity-velocity structure\footnote{Other
-  examples of galaxies with multiple populations include Carina
-  \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
+Besides their extended stellar density profiles, Sculptor and Ursa Minor
+display other peculiarities that may hold clues to their formation. Scl
+and UMi both host at least two distinct chemodynamic populations, as
+revealed through their photometric or metallicity-velocity
+structure\footnote{Other examples of galaxies with multiple populations
+  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
+  Fornax
   \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
@@
 excess? In Scl, both populations have scale radii likely smaller than
-the half-light radius. But, a third, more tentatively detected
+the half-light radius. However, a third, more tentatively detected
 population in \citet{arroyo-polonio+2024} would reside within the outer
 density excess. In Ursa Minor, the transition between the metal-rich and
-metal-poor stellar components occurs at \(\sim 30\,\)arcmin
-\citep{pace+2020}, which coincides with the density excess. This
+metal-poor stellar components occurs at \({\sim} 30\,\)arcmin
+\citep{pace+2020}, which coincides with our density excess. This
 metal-poor population may thus be related to the extended stellar
@@
 
-Given their multiple stellar populations and possible inner
-substructure, this hints that Scl and UMi's distinct structure today
-arises from an unusual past. For example, a merger may naturally explain
-both extended outer and multiple stellar populations.
+The multiple chemical populations and possible inner structure of
+Sculptor and Ursa Minor may represent the signatures of the unusual
+assembly histories that produced the dwarfs' extended stellar densities.
 
@@
 such processes would not explain why extended stellar populations appear
-to be non-universal. If an extended population was formed in an older,
-induced burst of star formation, then the star formation history should
+to be non-universal. If an extended population was formed during a
+separate star formation event, then the star formation history should
 contain evidence of a corresponding burst.
@@
 \citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
-the merger can trigger new star formation, forming a younger population
-of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
-to have undergone a major merger, including Tucana II, Andromeda II, and
-Phoenix
+the merger can also trigger new star formation, forming a younger
+population of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are
+suspected to have undergone a major merger, including Tucana II,
+Andromeda II, and Phoenix
 \citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
@@
 
-Just as the Milky Way's halo is believed to be built from many minor
-mergers, dwarf galaxies may build accreted ``stellar halos'' composed of
+Just as the Milky Way's halo is likely built from many minor mergers,
+dwarf galaxies may have built accreted ``stellar halos'' composed of
 even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
@@
 
-\textbf{Dark-matter-free dwarfs.} Tidal dwarfs form in gas-rich tidal
-streams created during the merger of two galaxies
+\textbf{Dark-matter-free dwarfs.} Tidal dwarfs are galaxies that formed
+in gas-rich tidal streams during the merger of two larger galaxies
 \citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
@@
 
-We have reviewed above a number of different, possibly concurrent
-explanations for the formation of extended populations. While we leave
-the nature of Scl and UMi's extended stellar densities an open question,
-we can discuss possible clues to different formation scenarios.
+We have reviewed a number of different, possibly concurrent explanations
+for the formation of extended populations. While we leave the nature of
+Scl and UMi's extended stellar densities an open question, we can
+discuss possible clues to different formation scenarios.
 
-\emph{Precise chemistry}. Chemistry, particularly comparing the inner
-and outer regions, could test if the halo appears to originate from a
-distinct system than the dwarf. This would test if the halo is a
-separate chemical population, may reveal the properties of any merged
-galaxies, or show if the halo is consistent with originating from the
-same galaxy.
+\emph{Precise chemistry}. Chemical abundances, comparing the inner and
+outer regions, can test if the halo was formed within the dwarf or
+accreted. If external in origin, chemistry could reveal the properties
+of any past mergers.
 
@@
 multi-object spectroscopy on large telescopes, such as 4MOST
-\citep{skuladottir+2023} or with the extremely large telescope
+\citep{skuladottir+2023} or the extremely large telescope
 \citep{jagourel+2018}. Mapping the 3D kinematics across a dwarf may
@@
 can derive precise proper motions within select regions of dwarf
-galaxies, able to reconstruct internal structure and anisotropies
+galaxies, reconstructing internal structure and anisotropies
 \citep[e.g.,][]{vitral+2025}. Current wide-field photometric surveys by
-Rubin Observatory \citep{ivezic+2019} and the Euclid mission
+the Rubin Observatory \citep{ivezic+2019} and \emph{Euclid}
 \citep{euclidcollaboration+2025} will trace even fainter features around
@@
 dwarfs with ever-increasing detail. Abundant opportunities await to
-uncover the cosmic origins and history of dwarf galaxies, the building
-blocks of galaxies like ours.
+uncover the cosmic origins of dwarf galaxies, the building blocks of
+galaxies like ours.

### 2025-10-21 thesis/extra_density.tex
thesis updates

@@
 To create a high-quality sample, J+24 select stars initially from Gaia
-within a 2--4 degree region around each satellite with high-quality
-astrometry, reliable photometry, consistent parallaxes, and broadly
+within a 2--4 degree region around each satellite with reliable
+astrometry, sensible photometry, consistent parallaxes, and broadly
 consistent proper motions and colours. Stars are removed if they have
@@
 \centering
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
 \caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
@@
 begins to dominate.
+
+\textbf{Likelihood misspecification}. The J+24 likelihoods may be
+misspecified. Particularly, the CMD model is simple and does not account
+for, e.g., the expected number of stars on various parts of the
+isochrone and assume the galaxy is a single population. In this case,
+the probabilities likely do not reflect true membership probabilities.
 
@@
 To bin stars, we hold fixed J+24's structural parameters and create bins
-of the wider of the width \(\Delta \log R=0.05\) or containing the next
-20 stars. We then derive posterior \(f_{\rm sat}\) distributions with
-MCMC (48 walkers of 1000 steps each, and using the No-U-Turn sampler as
-implemented in Turing.jl). The final density profile is directly derived
-from \(f_{\rm sat}\) and the number of stars in each bin.
+of width \(\Delta \log R=0.05\) or containing the next 20 stars,
+whichever is wider. We then derive posterior \(f_{\rm sat}\)
+distributions with MCMC (48 walkers of 1000 steps each, and using the
+No-U-Turn sampler as implemented in Turing.jl). The final density
+profile is directly derived from \(f_{\rm sat}\) and the number of stars
+in each bin.
 
@@
 overestimate faint densities and confidently derive densities where the
-MCMC model fails to derive a density estimate.
+MCMC model fails to determine the satellite density.
 

### 2025-10-21 thesis/main.tex
julio final edits

@@
 \title{
-    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?
+    The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Innate nature or tidal nurture?
 }
@@
        \@author\\
-       B.Sc. The Ohio State University, 2023\\
+       Bachelor of Science, The Ohio State University, 2023\\
        \vspace*{3\baselineskip}
@@
 \skipline
-   \end{center}
+   {\small
 We acknowledge and respect the \Lekwnen{} (Songhees and X\textsuperscript{w}seps\schwa{}m/
@@
 land continue to this day. 
+}
+   \end{center}
+
 \end{titlepage}
@@
 
+\phantomsection
 \addcontentsline{toc}{chapter}{Supervisory Committee}
@@
 \@author\\
-B.Sc. The Ohio State University, 2023\\
+Bachelor of Science, The Ohio State University, 2023\\
 \vspace*{3\baselineskip}
@@
 
-\noindent Dr. Kim Venn\\
+\noindent Dr. Kim Venn, Departmental Member\\
 Department of Astronomy\\
-
-\skipline
-
-\noindent External Examiner \\
-Outside Department
 
@@
 \addcontentsline{toc}{chapter}{Acknowledgments}
-\input{acknowledgments}
+\input{acknowledgements}
 
@@
 \newpage
+\thispagestyle{empty}
 
-\thispagestyle{empty}
 \renewcommand{\epigraphflush}{center}
-\vfill
+\phantomsection
+\addcontentsline{toc}{chapter}{Epigraph}
+       \vspace*{6\baselineskip}
 \setlength{\epigraphwidth}{0.5\textwidth}
-\epigraph{The sun had fallen beneath the horizon in the east, and the long day was about to end.
+%\epigraph{The sun had fallen beneath the horizon in the east, and the long day was about to end.
 But tomorrow, the sun would rise in the west again, bathing the world in a kinder light.}{Baoshu, translation by Ken Liu}
@@
 \epigraph{
-In the land of my birth, there is a hill, and upon that hill is a rock where I have sat and watched a changing world.
-This rock will not change, I would say.
-I'll pass from this Earth, but the rock will remain ever watchful of the land.
-
-But my feet kicked the dirt beneath me. Were the granules of soil not once like this rock? Is the rock not a collection of their multitudes?
-
-[We are] neither destroyed nor created, but a collection of dust meant to observe a changing world and then return to dust. 
-}{Joe Fisher, {\it Midnight Burger}}
-\vfill
+In the land of my birth, there is a hill.
+And upon that hill is a rock where I have sat and watched a changing world.
+This rock will not change---I would say.
+I'll pass from this Earth but the rock will remain, ever watchful of the land.
+But my feet kicked the dirt beneath me.
+Were the granules of soil not once like this rock?
+Is the rock not a collection of their multitudes?
+\ldots\ And is my path so different?
+Neither destroyed nor created, but a collection of dust meant to observe a changing world and then return to dust.
+}{---Joe Fisher, {\it Midnight Burger}}
 
@@
 \addcontentsline{toc}{chapter}{Research Acknowledgments}
-\input{research_acknowledgments}
+\input{research_acknowledgements}
 

### 2025-10-21 thesis/plots/scl_rv_gradient.jl
minor final rv gradient tweak

@@ -101,7 +101,7 @@ v0 = gsr.radial_velocity
 
 # ╔═╡ d413132c-d6f3-45d4-8a1e-1306b8ad084b
 let
-	fig = Figure(size=(5*72, 5*72))
+	fig = Figure(size=(6*72, 5*72))
 
 	gs = GridLayout(fig[1,1])
 	ax = Axis(gs[1,1],
@@ -109,7 +109,7 @@ let
 		ylabel = L"\eta\ /\ \textrm{arcmin}",
 		xreversed=true,
 		aspect = DataAspect(), 
-			  limits=(-50, 50, -30, 30),
+			  limits=(-50, 50, -50, 50),
 		xticks = -40:20:40,
 		yticks = -40:20:40,
 	)
@@ -183,7 +183,8 @@ let
 	lines!(xs, ys,  color=COLORS[2], label="rolling median")
 
 	annotation!(0, 36, obs_props["R_h"], 35, text=L"R_h")
-	axislegend(unique=true, merge=true, tellwidth=false, position=:lt, backgroundcolor=(:white, 0.5))
+	# axislegend(unique=true, merge=true, tellwidth=false, position=:lt, backgroundcolor=(:white, 0.5))
+	Legend(gs[1, 3], ax, unique=true, merge=true, tellheight=false,halign=:right, valign=-0.3)
 
 	colsize!(gs_lower, 1, Relative(3/3))
 

### 2025-10-21 thesis/plots/umi_rv_gradient.jl

@@ -104,7 +104,7 @@ v0 = gsr.radial_velocity
 
 # ╔═╡ d413132c-d6f3-45d4-8a1e-1306b8ad084b
 let
-	fig = Figure(size=(5*72, 5*72))
+	fig = Figure(size=(6*72, 5*72))
 
 	gs = GridLayout(fig[1,1])
 	ax = Axis(gs[1,1],
@@ -112,7 +112,9 @@ let
 		ylabel = L"\eta\ /\ \textrm{arcmin}",
 		xreversed=true,
 		aspect = DataAspect(), 
-			  limits=(-50, 50, -50, 50)
+			  limits=(-50, 50, -50, 50),
+			  xticks=-80:20:80,
+			  yticks=-80:20:80,
 	)
 
 	arrowscale = 4
@@ -184,7 +186,9 @@ let
 	lines!(xs, ys,  color=COLORS[2], label="rolling median")
 
 	annotation!(0, 36, obs_props["R_h"], -117, text=L"R_h")
-	axislegend(unique=true, merge=true, tellwidth=false, position=:lt, backgroundcolor=(:white, 0.5))
+	#axislegend(unique=true, merge=true, tellwidth=false, position=:lt, backgroundcolor=(:white, 0.5))
+
+	Legend(gs[1,3], ax, unique=true, merge=true, tellheight=false, haligh=:right, valign=-0.3)
 
 	colsize!(gs_lower, 1, Relative(3/3))
 

### 2025-10-21 paper/figures/sync.sh
paper template

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+rsync -av cosma:~/paper/figures/ ./  --exclude *.sh --exclude *.jpg

### 2025-10-21 paper/paper.tex
early chapter drafts for paper

%-----------------------------------------------------------------------
%                                                                 aa.tex
% AA vers. 9.3, LaTeX class for Astronomy & Astrophysics
% Demonstration file
%                                                       (c) EDP Sciences
%-----------------------------------------------------------------------
%
%\documentclass[referee]{aa}    % for a referee version
%\documentclass[onecolumn]{aa}  % for a paper on 1 column  
%\documentclass[longauth]{aa}   % for long lists of authors and/or affiliations. 
                                % This command displays the first eight authors on page 1
                                % and shift the whole list after the references.
                                % Ensure to separate each author with the \and command (see below)
%\documentclass[letter]{aa}     % for the letters
%\documentclass[bibyear]{aa}    % if the references are not structured
                                % according to the author-year natbib style

\documentclass{aa}

\usepackage{graphicx}
\usepackage{txfonts}
\usepackage{lipsum}
\usepackage{subcaption}         % necessary for continued figures, example in section 3
                                % and appendix
\usepackage{lscape}             % to rotate a single page table, example in appendix.
                                % For landscape tables, see the longtable examples.
\usepackage{placeins}           % useful with \FloatBarrier, to keep 
                                % onecolumn floats from drifting to the next section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[]{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
    citecolor=blue,
}

% extra commands
\newcommand{\V}{{\rm v}}
\newcommand{\vmax}{\V_{\rm max}}
\newcommand{\vcirc}{\V_{\rm circ}}
\newcommand{\rmax}{r_{\rm max}}
\newcommand{\kpc}{{\rm kpc}}
\newcommand{\Gyr}{{\rm Gyr}}
\newcommand{\kms}{{\rm km\,s^{-1}}}
\newcommand{\masyr}{{\rm mas\,yr^{-1}}}
\newcommand{\Mo}{M_\odot}

\newcommand{\agama}{{\tt Agama}}
\newcommand{\gadget}{{\tt Gadget-4}}
\newcommand{\smallperi}{{\tt smallperi}}
\newcommand{\LCDM}{$\Lambda$CDM}

%%% citepos command
\makeatletter
\DeclareRobustCommand\citepos
  {\begingroup
   \let\NAT@nmfmt\NAT@posfmt% ...except with a different name format
   \NAT@swafalse\let\NAT@ctype\z@\NAT@partrue
   \@ifstar{\NAT@fulltrue\NAT@citetp}{\NAT@fullfalse\NAT@citetp}}

\let\NAT@orig@nmfmt\NAT@nmfmt
\def\NAT@posfmt#1{\NAT@orig@nmfmt{#1's}}

\makeatother

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% if you use custom commands in your title,
% ensure to check your title when submitting!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{The extended stellar densities of the Sculptor and Ursa Minor dwarf galaxies}

\subtitle{Innate nature or tidal nurture?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% % Please separate each author with the \and command
%
% Please do not include ORCIDs next to author names.
% Only ORCIDs authenticated by individual authors in EDPS
% editorial system will be taken into account.
% ORCIDs included here will be removed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

   \author{D. A. Boyea\inst{1}
        \and others
        }

        \institute{%
            Department of Physics \& Astronomy, University of Victoria, Finnerty Road, Victoria, British Columbia V8A 1A1, Canada
             \email{danielaboyea@gmail.com}
             \thanks{Shows the usage of elements in the institute field}
            \and others\\ }

   \date{Received December XX, 20XX}

% \abstract{}{}{}{}{}
% 5 {} token are mandatory

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Most dwarf spheroidals (dSph) of the Milky Way follow an exponential surface density profile. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We aim to determine if tides may explain the excess of stars in Sculptor and Ursa Minor.} 
  % methods heading (mandatory)
   {We develop N-body models simulating tidal effects in a Milky Way and Large Magellanic Cloud (LMC) potential. }
  % results heading (mandatory)
   {Neighet galaxy is strongly tidally affected. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component barely evolves. UMi loses about 90\% of its dark matter mass but the stellar component would not be observationally affected. }
  % conclusions heading (optional), leave it empty if necessary
   {We conclude that the excess of stars likely reflects a divergent assembly history as compared to other local dSphs, possibly explained by past mergers or star formation bursts. }

   \keywords{giant planet formation --
                $\kappa$-mechanism --
                stability of gas spheres
               }

   \maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Since \citet{faber+lin1983}, exponential density profiles have been commonly used for dwarf
spheroidal galaxies \citealt{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
However, Sculptor and Ursa Minor, among Milky Way satellites, show strong deviations from an exponential profile (figure X), often speculated to be a result of tidal interactions
\citep[e.g.,][]{innanen+papp1979, exkridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003} or sometimes a ``halo''of stars
surrounding the dwarf \citep{westfall+2006}.
Most recently, \citet{sestito+2023a, sestito+2023b} report a
``kink'' in the density profile of each galaxy. They spectroscopically
follow up distant stars, finding members as far as 6 and 12 half-light
radii from the centre of each dwarf. If these dwarfs had exponential
profiles, like Fornax, then these far-outlying stars should be much
rarer.

The Local Group hosts several examples of ongoing tidal disruption. The
Magellanic stream, a massive, gas-rich feature emanating from the
Magellanic clouds, is believed to arise partially from the MW's tides
\citep{putman+1998, diaz+bekki2012, donghia+fox2016}. Other clear
examples of tidal streams include the Sagittarius stream, the Andromeda
Giant Southern stream, and the Tucana III stream
\citep[e.g.,][]{ibata+gilmore+irwin1994, ibata+2001, li+2018}. These
examples illustrate that hierarchical accretion remains an active
process. Interpreting such observations relies on simulations of tidal
disruption.

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.

Idealized simulations predict clear properties of tidally disrupting
dwarf spheroidal galaxies. Tidally stripped stars form \emph{tidal
streams}---stellar tails with a bulk velocity gradient
\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
Most mass loss happens near pericentre, where tides are strongest.
However, the central structure of a dwarf galaxy often remains
undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
NFW halos are also found to be resilient to full tidal disruption
\citep{EP2020}, but cored dark matter halos may disrupt fully and faster
\citep[e.g.,][]{penarrubia+2010, errani+2023a}.
To first order, tidal mass loss peels away the outer layers of a dwarf
galaxy in energy space.
\citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
showed that tidal effects are nearly entirely described as the removal
of particles above a truncation energy \citep[see
also][]{choi+weinberg+katz2009}. \citet{stucker+2023} generalized this
idea, creating a model for adiabatic tidal mass loss in an isotropic
tidal field. Their model explains the resilience of NFW halos against
full tidal disruption and the origin of well-defined ``tidal tracks''
\citep[as observed in][]{PNM2008, green+vandenbosch2019, EN2021}.

With precise orbital constraints and improved models of the Milky Way
potential, recent studies have continued to probe the dynamical
histories of individual dwarf galaxies.
\citet{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024}
both ran simulations tuned to Fornax, showing that this galaxy's stellar
component or globular clusters are likely not affected by tides.
Similarly, \citet{borukhovetskaya+2022a} analyzed Crater II, showing
that the present-day structure is challenging to reconcile with NFW
initial conditions and Galactic tides. Most relevantly,
\citet{iorio+2019} also tailored simulations to Scl, finding weak
Galactic tidal influence.

The \textbf{Jacobi radius} represents the approximate radius where stars
become unbound for a galaxy in a circular orbit around a host
galaxy.\footnote{The Jacobi radius was derived at least as early as
  \citet{laplace1798}. This radius also bears other names, such as the
  Hill radius \citep[from][]{hill1878}. Likely only named after Jacobi
  for the Jacobi integral \citep{jacobi1836}.} Calculated from an
approximation of the location of the \(L_1\) and \(L_2\) Lagrange
points, the Jacobi radius is where the mean density of the dwarf galaxy
is roughly three times the mean interior density of the host galaxy at
pericentre, or
\begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. While strictly valid for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages.

We also use the \textbf{break radius} as defined in
\citet{penarrubia+2009}, marking the outermost radius within which the
dwarf has been able to achieve equilibrium after pericentric passage in
a highly eccentric orbit. The break radius \(r_{\rm break}\) is
proportional to the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\),
\begin{equation}\protect\phantomsection\label{eq:r_break}{
r_{\rm break} = C\,\sigma_{v}\,\Delta t
}\end{equation} where the empirical constant is \(C \approx 0.55\).
\(r_{\rm break}\) describes where the dynamical timescale becomes longer
than the time since the perturbation, i.e., the radius within which the
galaxy is dynamically relaxed.

Building on this body of work, we will use idealized simulations to
understand the severity of tidal effects on Sculptor and Ursa Minor.

\begin{table}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: (1) Muñoz et al. (2018, Sérsic fit), (2) Tran et al. (2022, RR lyrae distance), (3) Alan W. McConnachie and Venn (2020b), (4) Arroyo-Polonio et al. (2024). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\hline
parameter & value & Source\\
\hline
$\alpha / ^\circ$ & $15.0183 \pm 0.0012$ & 1\\
$\delta / ^\circ$ & $-33.7186 \pm 0.0007$ & ”\\
distance modulus & $19.60 \pm 0.05$ & 2\\
distance & $83.2 \pm 2$ kpc & ”\\
$\mu_{\alpha*} / \masyr$  & $0.099 \pm 0.002 \pm 0.017$ & 3\\
$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$  & ”\\
$\V_{\rm los}$ / ${\rm km\,s^{-1}}$ & $111.2 \pm 0.3\ $ & 4\\
$\sigma_v$ / ${\rm km\,s^{-1}}$ & $9.7\pm0.2\ $ & ”\\
$R_h/'$ & $9.79 \pm 0.04$ & 1\\
ellipticity & $0.37 \pm 0.01$ & ”\\
position angle & $94\pm1^\circ$ & ”\\
$M_V$ & $-10.82\pm0.14$ & ”\\
\hline
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018, Sérsic fit), (2) Garofalo et al. (2025, RR lyrae distance), (3) Alan W. McConnachie and Venn (2020a), (4) Pace et al. (2020), average of MMT and Keck results with systematic uncertainty based on the difference between the MMT and Keck means. }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\hline
parameter & value & Source\\
\hline
$\alpha/^\circ$ & $ 227.2420 \pm 0.0045$ & 1\\
$\delta/^\circ$ & $67.2221 \pm 0.0016$ & 1\\
distance modulus & $19.23 \pm 0.11$ & 2\\
distance/kpc & $70.1 \pm 3.6$ & ”\\
$\mu_\alpha*$ / mas yr$^{-1}$  & $-0.124 \pm 0.004 \pm 0.017$ & 3\\
$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ & ”\\
$\V_{\rm los}$ / km s$^{-1}$  &$-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ & 4\\
$\sigma_v$ & $8.6 \pm 0.3$ & 4\\
$R_h$ & $11.62 \pm 0.1$ & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\hline
\end{tabular}
\end{table}

\section{{\it Gaia} density profiles}

Reducing background-contamination is essential to determine faint structures around dwarf galaxies. Before {\it Gaia}, most studies relied on colour-magnitude selections to select stars more likely to be members. Now with {\it Gaia}, proper motions and parallax information can substantially reduce the background of likely stars, producing low-contamination membership lists for dwarf galaxies.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite or foreground/background. By accounting for PM
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component. J+24 detected
candidate members out to \textasciitilde10 half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24).\footnote{This
  would indicate that the J+24 model probabilities are mis-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate.} However, we
find that our conclusions are unchanged when limiting samples to only
the brightest stars. Altogether, the J+24 method provides a
high-quality, low-contamination sample of dwarf galaxy candidate member
stars, which we will now investigate in further detail.

Fig.~\ref{fig:observed_density_profiles} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius, taken
from \citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{Comparing
  density profiles of dwarf galaxies is complicated by variations in the
  effective depth between galaxies. However, the deviations between Scl,
  UMi, and other galaxies are apparent even where the data is complete
  across all dwarfs.} In the outer regions, however, Sculptor and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards,
and at \(\sim 8 R_h\), may reach 2 orders of magnitude.

\begin{figure}
    \centering
    \includegraphics{figures/classical_dwarf_profiles.pdf}
    {\caption{The density profiles of classical dwarfs, using data from \citet{jensen+2024}.}}
    \label{fig:observed_density_profiles}
\end{figure}

\section{Methods}

\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
To convert observed positions and velocities to Galactocentric
coordinates, we use the Astropy v4 Galactocentric frame
\citep{astropycollaboration+2022}. Our Cartesian Galactocentric
coordinates here assume the Galactic centre is at
\([x, y, z] = [0,0,0]\), where \(x\) is the direction from the sun to
the Galactic centre, \(y\) is the direction of the motion of the Local
Standard of Rest, and \(z\) is the direction perpendicular to the
Galactic plane. The coordinate frame is also right-handed, such that the
\(z\)-angular momentum of the sun is negative (since the sun is at
\(x<0\)). In this frame, the solar position is
\([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
km/s
\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
The uncertainties in this reference frame are typically smaller than the
uncertainties on a dwarf galaxy's distance and tangential velocity.

We adopt the Milky Way potential described in \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The Galactic bulge is described by a \citet{hernquist1990} potential,
\begin{equation}{
    \Phi(r) = - \frac{GM}{r + a},
}\end{equation} 
where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
disks are represented with the \citet{miyamoto+nagai1975} cylindrical
potential, 
\begin{equation}{
    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
}\end{equation} 
where \(a\) is the disc radial scale length, \(b\) is
the scale height, and \(M\) is the total mass of the disk. For the thin
disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
The halo is an NFW \citep{NFW1996, 1997} dark matter halo (Eq.~\ref{eq:nfw}) 
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and density. We parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\), or
\(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).

\subsection{Orbits}\label{sec:scl_smallperi}

To explore the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables. The present-day position,
distance modulus, LOS velocity, and proper motions are each sampled from
normal distributions given the reported uncertainties in
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
each sampled position/velocity back in time for 10 Gyr using \agama{}
\citep{agama}. Dynamical friction is not expected to impact orbits
substantially, so we assume a single point-mass particle for the backwards
integration.
Figs.~\ref{fig:scl_orbits},~\ref{fig:umi_orbits} shows 100 samples of these orbits.

To maximize tidal effects, we select an orbit with the \(\sim 6\sigma\)
smallest pericentre among all possible orbits integrated backwards for
\(10\,\Gyr\). We achieve this by taking the median parameters of all
orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc. Given the current observations, it is
unlikely that Sculptor has had a significantly smaller pericentre than
our selected orbit, which we refer to as the \smallperi{} orbit. We take
the first apocentre after a look-back time of 10 Gyr, or at
\(\sim9.1\,\Gyr\), as the initial conditions for our model of Sculptor,
noted in Table~\ref{tbl:orbit_ics}.

We iteratively adjust Ursa Minor's initial
conditions to better match the present-day position. Initially, starting with low-resolution runs, we adjust the
cylindrical actions of the initial orbit by the final difference in
actions at the end of orbital evolution. After the initial actions have
converged (2 iterations), we change the initial action angles by the
final difference in action angles. This method converges within 4
iterations to an orbit agreeing with the observed kinematics of Ursa
Minor. Since Sculptor's orbit is less strongly affected by tides, we do
not carry out this correction for Sculptor.

\begin{table}[t]
\centering
\caption{The orbital initial conditions for models presented in this work. The observables record the intended final position of the N-body model derived from point-mass orbit quantiles. Instead, the initial position and velocity represent the initial conditions for the actual N-body model. The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects. We describe the \texttt{LMC-flyby} orbit in Section \ref{sec:scl_lmc}. }
\label{tbl:orbit_ics}
\begin{tabular}{llll}
\hline
Property & Scl & Scl-\verb|LMC| & UMi\\
\hline
$\vmax / \kms$ & 31 & 25 & 38 \\
$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
$R_{s, \rm exp}$ & 0.10 & 0.10 \\
$R_{s, \rm Plummer}$ & 0.20 & 0.20 \\
$N_\star$ \\
\hline
distance $/$ kpc & 82.6 & 73.1 & 64.6\\
$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
LOS velocity $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
\hline
$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
\hline
\end{tabular}
\end{table}

\subsection{Initial conditions}\label{initial-conditions}

We use \agama{} \citep{agama} to generate the initial N-body dark matter
halo. We assume galaxies are described by a spherical, isotropic NFW
dark matter potential (Eq.~\ref{eq:nfw}). We also assume the stars do
not contribute to the potential. The dark matter density is truncated in
the outer regions by
\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
}\end{equation}
where we adopt \(r_t = 20 r_s\).

We select cosmologically-motivated initial dark matter halos. First, taking the absolute
magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
\citet{woo+courteau+dekel2008}, the total current stellar mass of
Sculptor and Ursa Minor are \(M_\star \approx 3.1 \times 10^6 \Mo\) and
\(M_\star \approx 7 \times 10^5 \Mo\), respectively. Based on the
stellar mass-\(\vmax\) relation \citep[from][]{fattahi+2018}, Sculptor
and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
\(z=0\) mass-concentration relation, this constraint translates into a
\(\rmax \approx 6\,\kpc\) and \(\rmax \approx 5\,\kpc\) for each galaxy.
The cosmological means predict too low a velocity dispersion
(for Scl and
UMi, \(\sigma_{\V, i}  \approx 8.5\,\kms\) and \(8.0\,\kms\) given their observed half-light radius). 
To match the observed \(\sigma_\V\) at the end of tidal evolution, we
choose \(\vmax = 31\,\kms\) and \(\rmax=3.2\,\kpc\) for Scl and
\(\vmax=38\,\kms\) and \(\rmax=4\,\kpc\) for UMi.

\subsection{N-body methods}\label{isolation-runs-and-simulation-parameters}

To ensure that the initial conditions of the simulation are dynamically
relaxed, we run a halo first in isolation using
\gadget{}. Since gravity is scale-free, we use the same isolation run
for all halos and rescale the results to the desired values of size and
mass. We adopt a fiducial value of \(\rmax = 6.0\,\)kpc and
\(\vmax = 31\,\kms\) for the isolation halo based on Sculptor's mean
properties. We run this model for 5 Gyr (about one-half crossing time
\(t_{\rm cross} = 2\pi\,r /\vcirc  \approx 9\,\Gyr\) at
\(r_{200}=36\,\)kpc).

For our simulation parameters, we adopt a softening length of
\begin{equation}\protect\phantomsection\label{eq:softening_length}{
    h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
}\end{equation} 
for a halo with \(N\) particles.
Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).

Next, we evolve the halo in the Galactic potential. We scale the relaxed
snapshot and softening length to match the initial halo, and shift the
snapshot to the initial conditions inferred from the orbital analysis
(see Table~\ref{tbl:orbit_ics}). We then evolve the full N-body NFW
model forward in time in the Galactic potential until the present time,
when the halo is closest to its present-day observed position in the MW
halo.\footnote{Specifically, the snapshot which best agrees (as measured
  by \(\chi^2\)) with the intended final position (in
  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
  uncertainties. We use \(R_h\) as the uncertainty in \(\alpha\) and
  \(\delta\).}

We determine the halo centre using a shrinking-spheres centre method inspired by \citet{power+2003}.
First, we start with an initial centre estimate using all bound
particles from the previous snapshot. Then, we calculate the distance of
all particles from the centre, remove particles with a distance beyond
the 0.975 quantile of the centre, and recalculate the centre of mass.
The procedure is repeated until the selection radius is less than
\textasciitilde1kpc or fewer than 0.1\% of particles remain. We then
remove all unbound particles based on the \gadget{} calculated potential
of the halo and recalculate the centre. For all future timesteps, we
consider only particles retained from the previous iteration.

\subsection{Sculptor and Ursa Minor's initial stellar
components}\label{sec:painting_stars}

We ``paint'' stars onto dark matter particles using the particle-tagging
method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
symmetry. We initially assume stars follow a projected exponential law
(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,\kpc\) for both
galaxies. The tagging method assigns a probability to each dark matter
particle, which is proportional to the ``light-to-mass'' ratio required
to match the assumed stellar light profile. We briefly describe the
procedure next, but refer interested readers to \citet{EP2020}.

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 v^2\). If we
know the distribution function\footnote{i.e., the phase-space density of
  particles. Note that since \(f\) is formally defined in 6D phase
  space, there is an additional ``density of states'' term in order to
  calculate the histogram of particles with \({\cal E}\),
  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
then we assign a stellar weight to a given particle with energy
\({\cal E}\) using \begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation} We use Eddington inversion to find the distribution
function, 
\begin{equation}{
    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
}
\end{equation} 
\citep[eq. 4-140b in][]{BT1987}. In practice the right,
boundary term is zero. We
take \(\Psi\) from the underlying assumed analytic dark matter
potential. \(\rho_\star\) can be calculated from the surface density,
\(\Sigma_\star\), via the inverse Abel transform.

\begin{figure}
    \centering
    \includegraphics{figures/scl_lmc_xyzr_orbits.pdf}
    \caption{
        Orbits of Sculptor with and without the LMC.
    }
    \label{fig:scl_orbits}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/umi_xyzr_orbits.pdf}
    \caption{
        Orbits of Ursa Minor
    }
    \label{fig:umi_orbits}
\end{figure}

\section{Results}

\begin{itemize}
    \item Scl minimally evolves as a result of having one pericentre
    \item The LMC critically changes Scl's tidal history.
    \item UMi loses DM mass but the stellar component, as observed, would not show tidal signatures
    \item Break and Jacobi radii suport these interpretations. Stronger tidal evolution is unlikely under any other orbits in these potentials.
\end{itemize}

\subsection{Sculptor}
\begin{table}[t]
\centering
\caption[Orbits and results for Sculptor in the MW+LMC potential.]{The orbital properties and dark matter evolution for the models including an LMC. Similar to Table \ref{tbl:scl_sim_results} except quantities with respect to the LMC are in parentheses. }
\label{tbl:scl_lmc_sims}
\begin{tabular}{lll}
\hline
Property & random samples & Scl\\
\hline
pericentre $/$ kpc & $44\pm 3$ ($29 \pm 2$) & 39 (20)\\
apocentre $/$ kpc & $218 \pm 8$ & –\\
time of last pericentre $/$ Gyr & $-0.38\pm0.01$ ($-0.11$) & $-0.33$ ($-0.10$)\\
number of pericentres & 1 (1) & 1 (1)\\
Jacobi radius $/$ kpc & $3.3\pm0.2$ ($3.6\pm0.2$) & 2.8 (2.6)\\
Jacobi radius $/$ arcmin & $136 \pm 9$ ($159\pm5$) & 132 (121)\\
final heliocentric distance $/$ kpc & $83.2\pm2$ & 72.9\\
$\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.928\\
$r_\textrm{max, f} / r_\textrm{max, i}$ &  & 0.763\\
fractional final bound mass &  & 0.5402\\
\hline
\end{tabular}
\end{table}

\begin{figure*}
    \centering
    \includegraphics{figures/scl_lmc_sim_images.png}
    \caption{Snapshots of Scl in the LMC and MW potential.}
    \label{fig:scl_sim_images}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics{figures/scl_density_i_f.pdf}
    \caption{Simulated initial and final stellar density profiles for a tidal simulation of Scl in the MW+LMC potential} 
    \label{fig:scl_density_i_f}
\end{figure}

\subsection{Ursa Minor}
\begin{figure*}
    \centering
    \includegraphics{figures/umi_sim_images.png}
    \caption{Snapshots of UMi's tidal evolution in the MW potential}
    \label{fig:umi_sim_images}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics{figures/umi_density_i_f.pdf}
    \caption{Similar to \ref{fig:scl_density_i_f} except for UMi.} 
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

\subsection{Comparison with prior work}\label{comparison-with-prior-work}

A few studies have modelled Sculptor or Ursa Minor specifically, finding
similar conclusions. \citet{iorio+2019} applied idealized N-body
simulations to study tidal effects on Scl. They similarly found weak
tidal effects, even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also used idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal process
likely underlies the origin of these galaxies' outer density excess.

Besides their extended stellar density profiles, Sculptor and Ursa Minor
display other peculiarities that may hold clues to their formation. Scl
and UMi both host at least two distinct chemodynamic populations, as
revealed through their photometric or metallicity-velocity
structure\footnote{Other examples of galaxies with multiple populations
  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
  Fornax
  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
  Sextans
  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
  Andromeda II
  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
younger, higher metallicity, and dynamically colder, whereas the outer
population is older, lower metallicity, and dynamically hotter.
These populations may map onto the observed density excess for Ursa Minor, and \citepos{arroyo-polonio+2024} third population would reside within the density excess.

%Ursa Minor has also shown evidence for possible inner substructure, such
%as stellar or kinematic ``clumps''
%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
%While one clump was re-detected kinematically in \citet{pace+2014},
%\citet{munoz+2018} find no evidence of substructure with modern
%photometry.

\subsection{Discussion and Conclusions}\label{sec:stellar_halos}
In this work, we have investigated the extended, outer density excess
of Sculptor and Ursa Minor and its possible tidal origin. We have shown
that the density profiles are robust, implying that Scl and UMi contain
a true density excess relative to other classical dwarfs.

We then investigated whether tides are a possible explanation. By
modelling each galaxy based on cosmological initial conditions, we
showed that tides do not strongly affect either galaxy. The LMC changes
the orbital history of Scl and UMi, but tides become even weaker in a
combined LMC and MW potential. We conclude that recent tides are
unlikely to shape the outer radial distribution of stars in Scl and UMi.

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why extended stellar populations appear
to be non-universal. If an extended population was formed during a
separate star formation event, then the star formation history should
contain evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
the merger can also trigger new star formation, forming a younger
population of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are
suspected to have undergone a major merger, including Tucana II,
Andromeda II, and Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should harbour at least two or
three populations with distinct origin.

Just as the Milky Way's halo is likely built from many minor mergers,
dwarf galaxies may have built accreted ``stellar halos'' composed of
even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
examples of dwarf-dwarf accretion are known, but a possible stream
around And II or chemical peculiarities around Sextans I may be such
instances \citep{amorisco+evans+vandeven2014, roederer+2023}. In this
scenario, the extended stellar populations would contain chemical
signatures from several distinct ultra-faint dwarf galaxies.

\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
``preprocessed'' by a massive satellite like the LMC
\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
redistributes already-present stellar populations and may mimic a
stellar halo. Key evidence suggestive of preprocessing may include
distant dwarf stars or stellar streams.

\textbf{Dark-matter-free dwarfs.} Tidal dwarfs are galaxies that formed
in gas-rich tidal streams during the merger of two larger galaxies
\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
dark matter halo, tidal dwarfs may be more susceptible to tides, forming
extended density profiles and inflating the velocity dispersion
\citep{casas+2012, yang+2014, wang+2024a}.

Similarly, in Modified Newtonian Dynamics (MOND), galaxies do not
contain dark matter. Instead, the gravitational force law is altered to
explain rotation curves. Dwarfs would similarly experience stronger
tidal effects in MOND \citep{mcgaugh+wolf2010, brada+milgrom2000}.

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features {\bf Ref}---future data would be needed to uncover
signs of disruption to support this hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
dynamically hotter than younger stars due to processes including stellar
feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf
galaxies---extended stellar populations should be more common if these
processes are important.

We have reviewed a number of different, possibly concurrent explanations
for the formation of extended populations. While we leave the nature of
Scl and UMi's extended stellar densities an open question, we can
discuss possible clues to different formation scenarios.
\emph{Precise chemistry}. Chemical abundances, comparing the inner and
outer regions, can test if the halo was formed within the dwarf or
accreted. If external in origin, chemistry could reveal the properties
of any past mergers.
\emph{Detailed kinematics}. For models relying on recent tidal
disruption, kinematic disequilibrium features should be visible. These
would appear as velocity gradients, increasing velocity dispersions,
outward-biased moving stars, or non-phase mixed structures
\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}. In
addition, measurements of the ellipticity and anisotropy would help
verify or alter our understanding of the mass structure and whether
DM-free models may be permissible.
\emph{Deep photometry} may find or rule out signs of dynamical
disequilibrium and tidal tails for tidally susceptible models (e.g.,
\emph{pre-processing} and \emph{DM-free dwarfs}). In addition,
photometry will help constrain the prevalence and nature of extended
features in dwarf galaxies.
\emph{Star formation histories} can be derived through photometry or
chemistry. Evidence of significant star formation episodes or lack
thereof may differentiate scenarios that rely on strong star formation
bursts (e.g., \emph{episodic star formation history, induced star
formation,} and \emph{gas-rich mergers}).

Both upcoming observatories and novel simulation methods will be instrumental to uncovering the possible origins of the extended stellar densities in Sculptor and Ursa Minor.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
\end{acknowledgements}

\bibliographystyle{bibtex/aa.bst} % style aa.bst
\bibliography{paper.bib} % your references Yourfile.bib

\end{document}

### 2025-11-13 paper/paper.tex
paper updates

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{The extended stellar densities of the Sculptor and Ursa Minor dwarf galaxies}
-
-\subtitle{Innate nature or tidal nurture?}
+\title{Could tides explain the extended stellar densities of the Sculptor and Ursa Minor dwarfs?}
 
@@
              \email{danielaboyea@gmail.com}
-             \thanks{Shows the usage of elements in the institute field}
+             %\thanks{Shows the usage of elements in the institute field}
             \and others\\ }
@@
   % context heading (optional)
-  % {} leave it empty if necessary  
-   {Most dwarf spheroidals (dSph) of the Milky Way follow an exponential surface density profile. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
+   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density.
+However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
   % aims heading (mandatory)
-   {We aim to determine if tides may explain the excess of stars in Sculptor and Ursa Minor.} 
+   {We investigate whether tides may produce these extended light profiles.}
   % methods heading (mandatory)
-   {We develop N-body models simulating tidal effects in a Milky Way and Large Magellanic Cloud (LMC) potential. }
+   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
   % results heading (mandatory)
-   {Neighet galaxy is strongly tidally affected.
-Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component barely evolves.
-UMi loses about 90\% of its dark matter mass but the stellar component would not be observationally affected.
+   {Tides affect neither galaxy strongly.
+Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little.
+UMi instead loses about 95\% of its dark matter mass, but the stars would still appear unaffected with current observations.
 }
   % conclusions heading (optional), leave it empty if necessary
-   {We conclude that the excess of stars likely reflects a divergent assembly history as compared to other local dSphs, possibly explained by past mergers or star formation bursts. }
+   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }
 
-   \keywords{giant planet formation --
-                $\kappa$-mechanism --
-                stability of gas spheres
+   \keywords{Galaxies: dwarf --
+       Galaxies: kinematics and dynamics --
+       Local Group
                }
@@
 
-Since \citet{faber+lin1983}, exponential density profiles have been commonly used for dwarf
-spheroidal galaxies \citealt{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
-However, Sculptor and Ursa Minor, among Milky Way satellites, show strong deviations from an exponential profile (figure X), often speculated to be a result of tidal interactions
-\citep[e.g.,][]{innanen+papp1979, exkridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003} or sometimes a ``halo''of stars
-surrounding the dwarf \citep{westfall+2006}.
+Since \citet{faber+lin1983}, exponential surface density profiles have been commonly used to describe dwarf spheroidal (dSph) galaxies \citep[e.g.,][]{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
+However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show clear deviations from an exponential profile (Fig.~\ref{fig:observed_density_profiles}; e.g., \citealt{eskridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003}).
+This excess of stars at large radii was often speculated to be tidal interaction \citep[e.g.][]{innanen+papp1979, walcher+2003} or sometimes a `halo'of stars surrounding the dwarf \citep{westfall+2006}.
 Most recently, \citet{sestito+2023a, sestito+2023b} report a
-``kink'' in the density profile of each galaxy. They spectroscopically
-follow up distant stars, finding members as far as 6 and 12 half-light
+`kink' in the density profile of each galaxy. They spectroscopically
+confirm distant members out to ${\sim}10$ half-light
 radii from the centre of each dwarf. If these dwarfs had exponential
-profiles, like Fornax, then these far-outlying stars should be much
-rarer.
+profiles, then these far-outlying stars should be much
+rarer. We aim to test whether tides may form the extended stellar densities of Scl and UMi.
 
-The Local Group hosts several examples of ongoing tidal disruption. The
-Magellanic stream, a massive, gas-rich feature emanating from the
-Magellanic clouds, is believed to arise partially from the MW's tides
-\citep{putman+1998, diaz+bekki2012, donghia+fox2016}. Other clear
-examples of tidal streams include the Sagittarius stream, the Andromeda
-Giant Southern stream, and the Tucana III stream
-\citep[e.g.,][]{ibata+gilmore+irwin1994, ibata+2001, li+2018}. These
-examples illustrate that hierarchical accretion remains an active
-process. Interpreting such observations relies on simulations of tidal
-disruption.
-
-Cosmological simulations struggle to resolve tidal effects on dwarfs.
-Since many dwarfs are near the resolution limit, they are vulnerable to
-artificial disruption
+Cosmological simulations struggle to resolve tidal effects on dwarfs. Dwarfs are 
+often near the resolution limit and vulnerable to artificial disruption
 \citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
-numerical challenges, idealized simulations model a single subhalo in an
+these challenges, idealized simulations model a single subhalo in an
 analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
-
-Idealized simulations predict clear properties of tidally disrupting
-dwarf spheroidal galaxies. Tidally stripped stars form \emph{tidal
+From these simulations, tidally stripped stars form \emph{tidal
 streams}---stellar tails with a bulk velocity gradient
@@
 However, the central structure of a dwarf galaxy often remains
-undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
-NFW halos are also found to be resilient to full tidal disruption
-\citep{EP2020}, but cored dark matter halos may disrupt fully and faster
-\citep[e.g.,][]{penarrubia+2010, errani+2023a}.
+undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. 
 To first order, tidal mass loss peels away the outer layers of a dwarf
-galaxy in energy space.
-\citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
-showed that tidal effects are nearly entirely described as the removal
-of particles above a truncation energy \citep[see
-also][]{choi+weinberg+katz2009}. \citet{stucker+2023} generalized this
-idea, creating a model for adiabatic tidal mass loss in an isotropic
-tidal field. Their model explains the resilience of NFW halos against
-full tidal disruption and the origin of well-defined ``tidal tracks''
-\citep[as observed in][]{PNM2008, green+vandenbosch2019, EN2021}.
+galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023}).
+The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.
 
-With precise orbital constraints and improved models of the Milky Way
-potential, recent studies have continued to probe the dynamical
-histories of individual dwarf galaxies.
-\citet{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024}
-both ran simulations tuned to Fornax, showing that this galaxy's stellar
-component or globular clusters are likely not affected by tides.
-Similarly, \citet{borukhovetskaya+2022a} analyzed Crater II, showing
-that the present-day structure is challenging to reconcile with NFW
-initial conditions and Galactic tides. Most relevantly,
-\citet{iorio+2019} also tailored simulations to Scl, finding weak
-Galactic tidal influence.
+Improved orbital constraints and Milky Way
+potential models have enabled precise dynamical modeling of dwarf galaxies.
+Simulations tuned to Fornax \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024},
+Scl \citep{iorio+2019}, or MW classicals\footnote{We define classicals to be dwarfs with absolute magnitudes between $-7.7$ and $-14$.} in general \citep{tchiorniy+genina2025} often find tides play a minor role in the stellar evolution of dwarf classicals.
+However, modeling tidal evolution under the Large Magellanic Cloud (LMC) has not yet been considered for most classical dSphs.
 
-The \textbf{Jacobi radius} represents the approximate radius where stars
-become unbound for a galaxy in a circular orbit around a host
-galaxy.\footnote{The Jacobi radius was derived at least as early as
-  \citet{laplace1798}. This radius also bears other names, such as the
-  Hill radius \citep[from][]{hill1878}. Likely only named after Jacobi
-  for the Jacobi integral \citep{jacobi1836}.} Calculated from an
-approximation of the location of the \(L_1\) and \(L_2\) Lagrange
-points, the Jacobi radius is where the mean density of the dwarf galaxy
-is roughly three times the mean interior density of the host galaxy at
-pericentre, or
-\begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
+Two radii provide estimates of where tidal effects occur.
+The \textbf{Jacobi radius}, $r_{\rm J}$. approximates the radius where stars
+become unbound in a galaxy\footnote{ 
+    While strictly valid only for circular orbits, assuming
+\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
+lost during pericentric passages. $r_{\rm J}$ solves
+}
+\begin{equation}\label{eq:r_jacobi}
     3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
-}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) is comparable to
+\end{equation}
+given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
+and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
+If \(r_J\) is comparable to
 the visible extent of a galaxy, we should expect to find clear signs of
-tidal disturbance. While strictly valid for circular orbits, assuming
-\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
-lost during pericentric passages.
-
-We also use the \textbf{break radius} as defined in
-\citet{penarrubia+2009}, marking the outermost radius within which the
-dwarf has been able to achieve equilibrium after pericentric passage in
-a highly eccentric orbit. The break radius \(r_{\rm break}\) is
-proportional to the velocity dispersion, \(\sigma_v\), and time elapsed
-since pericentre, \(\Delta t\),
-\begin{equation}\protect\phantomsection\label{eq:r_break}{
+tidal disturbance. 
+The \textbf{break radius}, as defined in
+\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
+since pericentre, \(\Delta t\), the break radius is
+\begin{equation}\label{eq:r_break}
     r_{\rm break} = C\,\sigma_{v}\,\Delta t
-}\end{equation} where the empirical constant is \(C \approx 0.55\).
-\(r_{\rm break}\) describes where the dynamical timescale becomes longer
-than the time since the perturbation, i.e., the radius within which the
-galaxy is dynamically relaxed.
+\end{equation} 
+where the empirical constant is \(C \approx 0.55\).
 
 Building on this body of work, we will use idealized simulations to
-understand the severity of tidal effects on Sculptor and Ursa Minor.
+understand the severity of tidal effects on Scl and UMi.
+
+\section{{\it Gaia} density profiles}
+
+Measuring the light profile of a resolved galaxy requires careful
+consideration of whether any given star belongs to the system or not.
+Without removing contamination from foreground/background sources, faint
+features may be lost in the noise or be of uncertain association.
+In addition to colours and magnitudes, \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallax and proper motion which improve membership assignment.
+
+Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
+probabilities from \emph{Gaia} data.\footnotemark{} J+24 used a Bayesian framework
+incorporating proper motion, colour--magnitude diagram, and
+spatial information to determine the probability that a given star
+belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
+in particular, J+24 produced low contamination samples of candidate
+member stars out to large distances from a dwarf galaxy. J+24 extended
+the algorithm presented in \citet{MV2020a, MV2020b} by additionally
+including a secondary, extended spatial component, J+24 detected
+candidate members out to ${\sim}10$ half-light radii from the
+centres of some galaxies (\(R_h\)). Similar recent work includes
+\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
+\citet{qi+2022}.
+
+\footnotetext{
+We use
+the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
+galaxy shows evidence for an outer component, the 1-component run
+otherwise. 
+}
+
+For our sample, we adopt a minimum probability of
+\(P_{\rm sat} = 0.2\).
+%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
+Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
+choice of probability threshold has little effect on the resulting sample. Even
+at the probability threshold of 0.2, the purity remains high when validated
+against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
+J+24).\footnotemark{} However, we find that our conclusions are unchanged when
+limiting samples to only the brightest stars. Altogether, the J+24 method
+provides a high-quality, low-contamination sample of dwarf galaxy candidate
+member stars, which we will now investigate in further detail.
+
+\footnotetext{This
+  would indicate that the J+24 model probabilities are miss-calibrated.
+  However, most LOS surveys of dwarf galaxies select brighter stars
+  (which have better \emph{Gaia} measurements) and likely members,
+  complicating the interpretation of this purity estimate.}
+
+Fig.~\ref{fig:observed_density_profiles} compares the density
+profiles of Scl, UMi, and eight other classical dwarf
+galaxies.\footnote{
+    We define classicals to be non-Magellanic dwarfs brighter than
+    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
+    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
+    these dwarfs, we exclude Antlia II, due to the extremely high background,
+    and Sagittarius, which was not included in J+24. 
+}
+All of the classical dwarfs appear to be well
+described by an exponential profile in the inner regions.\footnote{
+    Comparing
+    density profiles of dwarf galaxies is complicated by variations in the
+    effective depth between galaxies. However, the deviations between Scl,
+    UMi, and other galaxies are apparent even where the data is complete
+    across all dwarfs.} 
+In the outer regions, however, Scl and Ursa
+Minor deviate and show a clear outer excess over an exponential law
+(solid black line). These galaxies are better fit by a Plummer law
+(dashed black line). The deviation from an exponential grows outwards,
+and at \(\sim 8 R_h\), may reach 2 orders of magnitude.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/classical_dwarf_profiles.pdf}
+    \caption{
+        The density profiles of
+        Sculptor (orange squares), Ursa Minor (red triangles), and other
+        classical dwarfs (blue lines) compared to 2D exponential (solid black
+        line) and Plummer density profiles (dashed black line). Density
+        profiles are plotted in terms of the logarithm of elliptical radius
+        scaled by the half-light radius from \citet{munoz+2018}, and the
+        logarithmic stellar density scaled by the density at half-light
+        radius. The residuals (lower panel) are with respect to the 2D
+        exponential. Sculptor and Ursa Minor have an excess of stars in the
+        outer regions (past \(\log R/R_h \sim 0.3\)) compared with other
+        classical dwarfs.
+}
+
+    \label{fig:observed_density_profiles}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Methods}
+
+\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
+To convert observed positions and velocities to Galactocentric
+Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
+\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
+the Galactic centre, $y$ is the direction of the motion of the Local
+Standard of Rest, and $z$ is the direction perpendicular to the
+Galactic plane. 
+% The coordinate frame is also right-handed, such that the
+% \(z\)-angular momentum of the sun is negative (since the sun is at
+% \(x<0\)). In this frame, the solar position is
+% \([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
+% \citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
+% velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
+% km/s
+% \citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
+% The uncertainties in this reference frame are typically smaller than the
+% uncertainties on a dwarf galaxy's distance and tangential velocity.
+
+For our MW-only models, we adopt the potential from \citet{EP2020}, which is
+an analytic approximation to that proposed by \citet{mcmillan2011}.
+The Potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
+A NFW density profile is defined as,
+\begin{equation}\label{eq:nfw}{
+\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
+}\end{equation}
+where $r_s$ and $\rho_s$ are the scale mass and scale density.
+We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.
+In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\).
+Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
+\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
+%The Galactic bulge is described by a \citet{hernquist1990} potential,
+%\begin{equation}{
+%    \Phi(r) = - \frac{GM}{r + a},
+%}\end{equation} 
+%where \(a=1.3\,{\rm kpc}\) is the scale radius and
+%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
+%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
+%potential, 
+%\begin{equation}{
+%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
+%}\end{equation} 
+%where \(a\) is the disc radial scale length, \(b\) is
+%the scale height, and \(M\) is the total mass of the disk. For the thin
+%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
+%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
+%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
+
+For our MW and LMC model, we use the \texttt{L3M11} model from
+\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
+multipole approximation of an N-body simulation including a live MW and
+LMC dark matter halo. The potential includes a static MW bulge and disk,
+evolving MW and LMC halos, and the MW reflex motion. In their
+simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
+\(M_{\rm 200}= 98.4\times10^{10}\Mo\), and the LMC a NFW halo with
+\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
+\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
+potential.
+
+\subsection{Orbits}\label{sec:scl_smallperi}
+
+\begin{table}[t]
+\centering
+\caption{The initial conditions for models presented in this work.
+The observables record the intended final position of the N-body model derived from point-mass orbit quantiles.
+Instead, the initial position and velocity represent the initial conditions for the actual N-body model.
+The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects}
+\label{tbl:orbit_ics}
+\begin{tabular}{llll}
+\hline
+Property & Scl: MW& Scl: MW+LMC & UMi: MW\\
+\hline
+$\vmax / \kms$ & 31 & 25 & 38 \\
+$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
+$R_{h, \rm exp}$ & 0.168 & 0.185 & 0.168 \\
+$R_{h, \rm Plummer}$ & 0.20 & 0.20  & 0.20 \\
+\hline
+distance $/$ kpc & 82.6 & 73.1 & 64.6\\
+$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
+$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
+$v_{\rm los}$ $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
+\hline
+$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
+${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
+${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
+${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
+$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
+$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
+$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
+\hline
+\end{tabular}
+\end{table}
+
+To explore the possible orbits of a dwarf galaxy, we perform a Monte
+Carlo sampling of the present-day observables (assuming Normal distributions
+and the values in Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}). We
+integrate
+each sampled position/velocity back in time for 10 Gyr using \agama{}
+\citep{agama}. Dynamical friction is not expected to impact orbits
+substantially, so we assume a single point-mass particle for the backwards
+integration.
+Fig.~\ref{fig:orbits} shows 100 samples of these orbits in both potentials (MW only and MW+LMC), and the orbits of our N-body models.
+
+To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
+smallest pericentre among all possible orbits integrated backwards for
+\(10\,\Gyr\). We achieve this by selecting the median parameters of all
+orbits with a pericentre less than the 0.0027th quantile pericentre,
+yielding a pericentre of 43 kpc. Given the current observations, it is
+unlikely that either has had a significantly smaller pericentre than
+our selected orbit, which we refer to as the \smallperi{} orbit. We take
+the first apocentre after a look-back time of 10 Gyr as the initial conditions for our models of Scl and UMi,
+noted in Table~\ref{tbl:orbit_ics}.
+
+For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties), and only integrate backwards for $2\,\Gyr$.
+
+We iteratively adjust UMi's initial conditions to better match the
+present-day position. Initially, starting with low-resolution runs, we adjust
+the cylindrical actions of the initial orbit by the final difference in actions
+at the end of orbital evolution. After the initial actions have converged (2
+iterations), we change the initial action angles by the final difference in
+action angles (2 more iterations). This method converges to an orbit agreeing
+with the observed kinematics of UMi. Since Scl's orbit is less
+strongly affected by tides, we do not carry out this correction for Scl.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/orbits.pdf}
+    \caption{
+        The orbits of Scl (left) and UMi (right) in the heavy MW-only potential (light blue lines) and the MW+LMC
+        potential (light orange lines). Black lines shows our selected N-body models and
+        the green dashed line shows the LMC trajectory. The top panels show orbits in the $y$--$z$ plane and the botom panels show orbits in terms of Galactocentric radius with time.
+}
+
+    \label{fig:orbits}
+\end{figure}
+
+\subsection{Initial conditions}\label{initial-conditions}
+
+We use \agama{} \citep{agama} to generate the initial N-body dark matter
+halo. We assume galaxies are described by a spherical, isotropic 
+dark matter halo (Eq.~\ref{eq:nfw}). The NFW
+profile truncated in the outer regions by
+\begin{equation}\label{eq:trunc_nfw}
+    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
+\end{equation}
+where we adopt \(r_t = 20 r_s\).
+
+We select cosmologically-motivated initial dark matter halos.
+We derive total stellar masses from \citet{munoz+2018} magnitudes and \citet{woo+courteau+dekel2008} stellar mass-to-light ratios, finding Scl and UMi have stellar masses of $3\times10^6$ and $7\times10^5\,\Mo$.
+Using the stellar mass-\(\vmax\) relation \citep[from][]{fattahi+2018} and mass-concentration relation from \citet{ludlow+2016}, Scl
+and UMi's halos should have \(\vmax \approx 31 \,\kms\), $\rmax\approx6\,\kpc$ and
+\(\vmax \approx 27\,\kms\), $\rmax\approx5\,\kpc$. 
+The cosmological means predict a low velocity dispersion compared to the observed dispersion.
+To match the observed \(\sigma_\V\) at the end of tidal evolution, we
+choose \(\vmax = 31\,\kms\) and \(\rmax=3.2\,\kpc\) for Scl in the MW-only potential, $\vmax=25\,\kms$ and $\rmax=2.5\,\kpc$ for Scl in the MW+LMC potential, and
+\(\vmax=38\,\kms\) and \(\rmax=4\,\kpc\) for UMi.
+
+\subsection{N-body methods}\label{isolation-runs-and-simulation-parameters}
+
+We use \gadget{}, with a \agama{} external potential plugin, for our N-body simulations. 
+We adopt a softening length of
+\begin{equation}\label{eq:softening_length}
+    h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
+\end{equation} 
+for a halo with \(N\) particles.
+We use $10^7$ particles, and adopt a relative accuracy criterion of $\alpha=0.005$ and integration accuracy of $\eta=0.01$.\footnote{i.e., tree nodes open if $M\,l/r^3 < \alpha |a|$ for node mass $M$, sidelength $l$, distance $r$, and particle acceleration $a$; and particles must take time-steps smaller than $\sqrt{2\,\eta\,h_{\rm grav}}$.}
+
+To ensure that the initial conditions of the simulation are dynamically
+relaxed, we run a halo first in isolation using
+\gadget{}. We use the same isolation run
+for all halos (with $\rmax=6$ and $\vmax=31$, ran for $5\,\Gyr$) and rescale the results to the desired initial halo.
+Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).
+
+Next, we evolve the halo in the Galactic potential. We rescale and shift the
+isolation-relaxed snapshot to our initial conditions (see Table~\ref{tbl:orbit_ics}). 
+We then evolve the full N-body NFW model forward in time in the Galactic
+potential until the present time, when the halo is closest to its present-day
+observed position in the MW
+halo.\footnote{Specifically, the snapshot which best agrees (as measured
+  by \(\chi^2\)) with the intended final position (in
+  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
+  uncertainties. We use \(R_h\) as the uncertainty in \(\alpha\) and
+  \(\delta\).}
+
+We determine the halo centre using a shrinking-spheres centre method inspired by \citet{power+2003}.
+First, we start with an initial centre estimate using all bound
+particles from the previous snapshot. Then, we calculate the distance of
+all particles from the centre, remove particles with a distance beyond
+the 0.975 quantile of the centre, and recalculate the centre of mass.
+The procedure is repeated until the selection radius is less than
+\textasciitilde1kpc or fewer than 0.1\% of particles remain. We then
+remove all unbound particles based on the \gadget{} calculated potential
+of the halo and recalculate the centre. For all future timesteps, we
+consider only particles retained from the previous iteration.
+
+\subsection{Sculptor and Ursa Minor's initial stellar
+components}\label{sec:painting_stars}
+
+\begin{figure}
+    \includegraphics{figures/initial_velocity.pdf}
+    \caption{
+        {\bf Top:} The initial circular velocity profiles in terms of radius
+        for Sculptor and Ursa Minor for the MW-only models with exponential
+        stars. We plot the models after 5\,Gyr of evolution in isolation, which
+        are nearly identical to the initial conditions. The vertical grey line
+        marks the Jacobi radius (Eq.~\ref{eq:r_jacobi})        
+        {\bf Bottom:} The initial velocity dispersion profiles for the exponential stellar
+        component, as compared to our velocity sample (binned in ten equal
+        number bins). See Section~\ref{sec:velocity_sample} for a description
+        of our velocity sample.
+    }
+    \label{fig:initial_halos}
+\end{figure}
+
+We `paint' stars onto dark matter particles using the particle-tagging
+method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
+symmetry. We briefly describe the
+procedure next, but refer interested readers to \citet{EP2020}.
+
+If we
+know the distribution function\footnote{i.e., the phase-space density of
+  particles. Note that since \(f\) is formally defined in 6D phase
+  space, there is an additional `density of states' term in order to
+  calculate the histogram of particles with \({\cal E}\),
+  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
+then we assign a stellar weight to a given particle with energy
+\({\cal E}\) using \begin{equation}{
+P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
+}\end{equation}
+We use Eddington inversion to find the distribution
+function, \begin{equation}{
+    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
+}
+\end{equation} 
+where $\rho$ is the stellar or DM density, \(\Psi\) is the potential (normalized to vanish at infinity), and
+\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 \V^2\)  
+\citep[see eq. 4-140b in][]{BT1987}. We
+take \(\Psi\) and $\rho$ from the assumed analytic density profiles. We calculate \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.
+
+We consider both an exponential stellar profile, 
+\begin{equation}\label{eq:exponential_law}
+\Sigma_{\rm exp} = \Sigma_0\exp(-\alpha\,R / \alpha\,R_h).
+\end{equation}
+for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
+We also consider a more extended Plummer law, 
+\begin{equation}{
+\Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
+}\end{equation}
+for half-light radius $R_h$.
+
+Fig.~\ref{fig:initial_halos} shows the resulting initial conditions for the MW-only models before and after 5 Gyr of isolation evolution.
+
+\section{Results}
+
+\subsection{Orbits of Sculptor and Ursa Minor}
+Fig.~\ref{fig:orbits} shows orbits of Scl and UMi in the MW-only and MW+LMC potentials.
+In the MW-only potential, both galaxies follow relatively well-constrained orbits consistent with long-term MW satellites.
+However, including the LMC dramatically modifies Scl's orbit---the apocentre increases from ${\sim}100$ to ${\sim}220\,\kpc$.
+Scl also passes by the LMC 100 Myr ago, making the LMC a potential tidal stressor.
+In contrast, UMi's orbit only slightly increases in characteristic radius, reducing tidal forces overall.\footnote{Note that our MW-only model has a heavier MW mass than the MW+LMC model, but the qualitative effects on each galaxy's orbit are largely similar when using the same MW model in each case.}
+
+The black lines in \ref{fig:orbits} show our selected orbits (black), representing an extreme low-pericentre orbit.
+We adopt these orbits as the most severe observationally-permisible tidal histories of each galaxy.
+
+\subsection{Tidal effects}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/density_i_f.pdf}
+    \caption{
+        The stellar density of each model before (dotted) and after
+        (solid) tidal evolution as compared to the observed density profile
+        (solid points). The left and righ panels show results for Sculptor and
+        Ursa Minor in the MW-only potential. The top and bottom panels show
+        results for exponential and Plummer initial stellar conditions. We mark
+        the half-light ($R_h$), the Jacobi radius, and break radius with arrows. 
+}
+    \label{fig:stellar_model_density}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/scl_lmc_density_i_f.pdf}
+    \caption{
+        Similar to Fig.~\ref{fig:stellar_model_density} except for Scl in the combined MW and LMC potential.
+The left and right panels show results for exponential and Plummer initial stellar profiles, respectively.
+    }
+    \label{fig:stellar_model_density_lmc}
+\end{figure}
+
+Fig.~\ref{fig:stellar_model_density} shows the initial and final stellar
+densities of each galaxy in the MW-only potential compared to observations. The final profile resembles the initial profile
+over the range of the observed density profile.
+UMi experiences the strongest evolution, with the stars expanding slightly and flattening past the break radius.
+But UMi's stars do not change measurably in shape.
+In addition, the break and Jacobi radii fall outside the observed stellar component, indicating that tidal effects should not be observable.
+
+Since Scl and UMi may have been born with more extended stellar components, we
+also consider Plummer initial stars, approximately matching the observed
+density profiles. The Plummer profiles also remain nearly unchanged during
+tidal evolution. If any tidal disturbances are visible, than observations would
+have to probe more than two orders of magnitude fainter features than our data
+reach. Therefore, tides do not substantively affect the observed stellar components of these models.
+
+Fig.~\ref{fig:stellar_model_density_lmc} shows the tidal effects on Scl in the LMC+MW potential.
+While the break radius coincides with the observed
+excess, the Jacobi radius is too large to observe tidal signatures. In
+addition, with a reduced number of pericentres compared to the MW-only case,
+the stellar profiles evolve even less.
+
+Not shown here, cored, anisotropic, and flattened models
+all appear to evolve similarly over time. Given the observed velocity
+dispersions, alternative dark matter structures still require a similar mass
+within $R_h$ at the present day. Reasonable changes to the dark matter
+structure likely do not change our conclusions.
+
+A key caveat is that the long-term orbits of satellites are highly uncertain.
+Analytic MW potentials neglect unknown details including triaxiality, mass
+evolution, and substructure, causing analytic orbits to deviate significantly
+from true satellite orbits \citep[e.g.,]{dsouza+bell2022, santistevan+2023}.
+While we cannot constrain the long-term histories of these dwarfs, our results
+indicate that {\it recent tidal effects} would not be observable in Scl and
+UMi.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion \& Conclusions}
+We investigated whether tides from the Milky Way or LMC may explain the extended stellar densities of Scl and UMi.
+Using idealized N-body simulation on orbits with extreme pericentres, we find that tides are insufficient to influence the stellar light profile of either galaxy.
+We also find the LMC strongly perturbs Scl's orbit, weakining tides further.
+We conclude the observed density excess in Sculptor and Ursa Minor are unlikely to be tidal in origin.
+
+Scl and UMi display other peculiarities that may hold clues to their formation. Both galaxies host at least two distinct chemodynamic populations
+\citep{tolstoy+2004, battaglia+2008, pace+2020}.\footnotemark{} The inner population is
+younger, higher metallicity, and dynamically colder, whereas the outer
+population is older, lower metallicity, and dynamically hotter.
+Fig.~\ref{fig:metallicity_gradient} compares a double-exponential density fit to the observed metallicity and velocity dispersion profiles.
+A double-exponential populations better explains the observed profiles than a single-exponential.
+Our two populations are similar to \citet[MMT]{pace+2020}.
+Though our second population is more extended than in \citet{arroyo-polonio+2024}, their third population may be contained within our outer component.
+
+\footnotetext{Other examples of galaxies with multiple populations
+  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
+  Fornax
+  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
+  Sextans
+  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
+  Andromeda II
+  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
+
+\begin{figure}
+    \includegraphics{figures/scl_umi_fe_h_gradient.pdf}
+
+    \caption{
+        {\bf Top:} Density profiles of Sculptor (top left) and Ursa Minor (top
+        right) with double exponential fits (see
+        appendix~\ref{sec:double_exp_fits}). Blue, pink, and red lines
+        represent the inner, outer, and combined density, respectively.
+        {\bf Middle:} The metallicities of member stars in both galaxies with
+        radius. For Sculptor, we plot stars from
+        APOGEE (green circles), \citet[][blue squares]{tolstoy+2023}, and
+        \citet[][orange stars]{sestito+2023a}. For Ursa Minor, we
+        plot stars from APOGEE (green circles), \citet[blue
+        squares]{pace+2020}, and \citet[red stars]{sestito+2023b}. Black points
+        represent median bins, and the pink line
+        represent a simple mixture model of metallicity, transitioning from
+        higher to lower metallicity based on the relative inner and outer
+        component densities (see Table~\ref{tbl:double_exp_fits}). 
+        {\bf Bottom:} The velocity dispersion profiles of our radial velocity
+        sample (black points) as compared to the velocity dispersion profile
+        for exponential profiles. We calculate the velocity dispersion of the
+        model through the Jean's equations assuming a NFW dark halo of
+        ($\rmax=2.5\,\kpc$, $\vmax=25\,\kms$) for Scl and ($\rmax=5\,\kpc$, $\vmax=27\,\kms$) for UMi.
+    }
+    \label{fig:metallicity_gradient}
+\end{figure}
+
+%Ursa Minor has also shown evidence for possible inner substructure, such
+%as stellar or kinematic `clumps'
+%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
+%While one clump was re-detected kinematically in \citet{pace+2014},
+%\citet{munoz+2018} find no evidence of substructure with modern
+%photometry. 
+  We next consider a variety of possible origins.
+
+\textbf{Episodic star formation.} Star formation may quench and
+reignite, creating successive stellar generations with differing
+distributions. External star formation triggers include tidal
+compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
+gaseous filaments or mergers \citep{genina+2019}, perturbations from dark halos
+\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
+\citep{wright+2019}.\footnote{More common mechanisms, like feedback or
+reionization-driven quenching, may also form multiple stellar
+generations
+\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
+such processes would not explain why extended stellar populations appear
+to be non-universal.} In this case, the star formation history should
+contain evidence of a corresponding burst.
+
+\textbf{Major mergers.} In a dwarf merger, stars from
+the lower-mass galaxy are preferentially dispersed, forming an extended
+stellar component and population gradient
+\citep{benitez-llambay+2016, deason+2022}. A halo may also form of many pre-reionization quenched ultra-faint dwarf galaxies \citep{ricotti+polisensky+cleland2022}. A few local dwarfs are
+suspected to have undergone a major merger, including Tucana II,
+Andromeda II, and Phoenix
+\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
+A galaxy having undergone a major merger should harbour at least two or
+three populations with distinct origin.
+
+\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
+`preprocessed' by a massive satellite like the LMC
+\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
+redistributes already-present stellar populations and may mimic a
+stellar halo. Key evidence suggestive of preprocessing may include
+distant dwarf stars or stellar streams.
+
+\textbf{Dark-matter-free dwarfs.} Dark-matter-free tidal dwarfs 
+%(formed in gas-rich tidal streams during the merger of two larger galaxies, e.g., \citealt{mirabel+dottori+lutz1992, bournaud+duc2006}) 
+and galaxies in Modified Newtonian Dynamics (MOND) would be more susceptible to tides (\citealt{brada+milgrom2000, mcgaugh+wolf2010, casas+2012, yang+2014, wang+2024a}; but see also \citealt[][]{sanchez-salcedo+hernandez2007}).
+In these cases, the high observed velocity dispersion requires evidence of strong tidal disruption, such as velocity gradients or tidal tails
+\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.
+
+\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
+dynamically hotter than younger stars. Dynamical heating processes including
+stellar feedback \citep{stinson+2009, maxwell+2012, el-badry+2016,
+mercado+2021}, sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy
+dark matter interference fringes \citep[e.g.,][]{el-zant+2020,
+duttachowdhury+2023}. However, most of these processes should operate similarly
+across dwarf galaxies, not just Scl and UMi.
+
+While a number of scenarios may form an extended outer stellar density, future
+observations have promise to discern different scenarios. 
+{Precise chemistry}, comparing the inner and
+outer regions, can test if the was accreted and could constrain properties
+of merged satellites. 
+{Detailed kinematics} could constrain the halo shape (ellipticity and anisotropy) and
+if the dwarf shows signs of tidal disruption.
+{Deep photometry} may find or rule out signs of dynamical
+disequilibrium and tidal tails for tidally susceptible models and constrain the prevalence and nature of extended features in dwarf galaxies.
+Finally, {star formation histories} may differentiate scenarios that rely on or create strong star formation bursts.
+Both upcoming observatories and novel simulation methods will be instrumental to uncovering the formation histories of Sculptor and Ursa Minor.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{acknowledgements}
+\end{acknowledgements}
+
+\bibliographystyle{bibtex/aa.bst} % style aa.bst
+\bibliography{paper.bib} % your references Yourfile.bib
+
+\appendix
+
+\section{Additional model results}
+In this section, we present tables of the resulting properties of the dark matter and stellar components for each model.
+
+Tables~\ref{tbl:scl_orbit_props}, \ref{tbl:umi_orbit_props} contain information about the orbital properties of the models and our Monte carlo sampled orbits.
+Table~\ref{tbl:sim_results} contains the initial and final properties of the dark matter and stars for each simulation presented in the main text.
 
@@
 $\alpha / ^\circ$ & $15.0183 \pm 0.0012$ & 1\\
-$\delta / ^\circ$ & $-33.7186 \pm 0.0007$ & ”\\
+$\delta / ^\circ$ & $-33.7186 \pm 0.0007$ & 1\\
 distance modulus & $19.60 \pm 0.05$ & 2\\
-distance & $83.2 \pm 2$ kpc & ”\\
+distance & $83.2 \pm 2$ kpc & 2\\
 $\mu_{\alpha*} / \masyr$  & $0.099 \pm 0.002 \pm 0.017$ & 3\\
-$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$  & ”\\
+$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$  & 3\\
 $\V_{\rm los}$ / ${\rm km\,s^{-1}}$ & $111.2 \pm 0.3\ $ & 4\\
-$\sigma_v$ / ${\rm km\,s^{-1}}$ & $9.7\pm0.2\ $ & ”\\
+$\sigma_v$ / ${\rm km\,s^{-1}}$ & $9.7\pm0.2\ $ & 4\\
 $R_h/'$ & $9.79 \pm 0.04$ & 1\\
-ellipticity & $0.37 \pm 0.01$ & ”\\
-position angle & $94\pm1^\circ$ & ”\\
-$M_V$ & $-10.82\pm0.14$ & ”\\
+ellipticity & $0.37 \pm 0.01$ & 1\\
+position angle & $94\pm1^\circ$ & 1\\
+$M_V$ & $-10.82\pm0.14$ & 1\\
 \hline
@@
 distance modulus & $19.23 \pm 0.11$ & 2\\
-distance/kpc & $70.1 \pm 3.6$ & ”\\
+distance/kpc & $70.1 \pm 3.6$ & 2\\
 $\mu_\alpha*$ / mas yr$^{-1}$  & $-0.124 \pm 0.004 \pm 0.017$ & 3\\
-$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ & ”\\
+$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ & 3\\
 $\V_{\rm los}$ / km s$^{-1}$  &$-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ & 4\\
@@
 
-\section{{\it Gaia} density profiles}
+\begin{table*}[t]
+\centering
+\caption{The orbital properties for Scl.}
+\label{tbl:scl_orbit_props}
+\begin{tabular}{llllllll}
+\hline
+Property & MW orbits & MW N-body & LMC orbits & LMC N-body \\
+\hline
+pericentre $/$ kpc & $53\pm3$ & 42 & $44 \pm 3$ & 39 \\
+LMC pericentre $/$ kpc & -- & -- & $28.7^{+1.6}_{-1.6}$ & 20 \\
+apocentre $/$ kpc & $102\pm3$ & 94 & $218 \pm 8$ & -- \\
+time of last pericentre $/$ Gyr & $-0.45\pm0.2$ & $-0.47$ & $-0.38\pm0.01$ & $-0.33$ \\
+time last peri LMC / Gyr & -- & -- & $-0.111\pm 0.003$ & -0.10 \\
+number of pericentres & 5--6 & 6 & 1 & 1 \\
+final heliocentric distance $/$ kpc & $83.2\pm2$ & 81.6 & --- & 72.9\\
+\hline
+Jacobi radius $/$ kpc & $4.5 \pm 0.3$ & 3.5 & $3.3\pm0.2$  & 2.8 \\
+Jacobi radius $/$ arcmin & $186\pm12$ & 148 & $136 \pm 9$  & 132 \\
+LMC Jacobi radius $/$ kpc & -- & -- &  $3.6\pm0.2$ & 2.6\\
+LMC Jacobi radius $/$ arcmin & -- & -- &  $159\pm5$ & 121\\
+\hline
+break radius $/$ kpc & $-2.45\pm0.11$ & 2.6 & $-2.06\pm0.07$ & 1.82\\
+break radius $/$ arcmin & $-101\pm4$ & 109 & $-85.1\pm2.1$ & 85 \\
+LMC break radius $/$ kpc  & -- & -- & $-0.604^{+0.019}_{-0.019}$ & 0.56\\
+LMC break radius $/$ arcmin & -- & -- & $-25\pm0.6$ & 27\\
+\hline
+\end{tabular}
+\end{table*}
 
-Reducing background-contamination is essential to determine faint structures around dwarf galaxies.
-Before {\it Gaia}, most studies relied on colour-magnitude selections to select stars more likely to be members.
-Now with {\it Gaia}, proper motions and parallax information can substantially reduce the background of likely stars, producing low-contamination membership lists for dwarf galaxies.
-
-Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
-probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
-incorporating proper motion (PM), colour-magnitude diagram (CMD), and
-spatial information to determine the probability that a given star
-belongs to the satellite or foreground/background. By accounting for PM
-in particular, J+24 produced low contamination samples of candidate
-member stars out to large distances from a dwarf galaxy. J+24 extended
-the algorithm presented in \citet{MV2020a, MV2020b} by additionally
-including a secondary, extended spatial component. J+24 detected
-candidate members out to \textasciitilde10 half-light radii from the
-centres of some galaxies (\(R_h\)). Similar recent work includes
-\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
-\citet{qi+2022}.
-
-For our fiducial sample, we adopt a minimum probability of
-\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
-J+24's quality cuts typically only include stars with \(G < 21\). We use
-the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
-galaxy shows evidence for an outer component, the 1-component run
-otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
-1, so the exact choice of probability threshold has little effect on the
-resulting sample. Even at our relatively generous probability threshold
-of 0.2, the purity remains high when validated against spectroscopic
-line-of-sight (LOS) velocities (\textasciitilde90\%, J+24).\footnote{This
-  would indicate that the J+24 model probabilities are mis-calibrated.
-  However, most LOS surveys of dwarf galaxies select brighter stars
-  (which have better \emph{Gaia} measurements) and likely members,
-  complicating the interpretation of this purity estimate.} However, we
-find that our conclusions are unchanged when limiting samples to only
-the brightest stars. Altogether, the J+24 method provides a
-high-quality, low-contamination sample of dwarf galaxy candidate member
-stars, which we will now investigate in further detail.
-
-Fig.~\ref{fig:observed_density_profiles} compares the fiducial density
-profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
-galaxies. Of the classicals, we exclude Antlia II, due to the extremely
-high background, and Sagittarius, which was not included in J+24. The
-density profiles are scaled to match at the half-light radius, taken
-from \citet{munoz+2018}. All of the classical dwarfs appear to be well
-described by an exponential profile in the inner regions.\footnote{Comparing
-  density profiles of dwarf galaxies is complicated by variations in the
-  effective depth between galaxies. However, the deviations between Scl,
-  UMi, and other galaxies are apparent even where the data is complete
-  across all dwarfs.} In the outer regions, however, Sculptor and Ursa
-Minor deviate and show a clear outer excess over an exponential law
-(solid black line). These galaxies are better fit by a Plummer law
-(dashed black line). The deviation from an exponential grows outwards,
-and at \(\sim 8 R_h\), may reach 2 orders of magnitude.
-
-\begin{figure}
+\begin{table*}[t]
 \centering
-    \includegraphics{figures/classical_dwarf_profiles.pdf}
-    {\caption{The density profiles of classical dwarfs, using data from \citet{jensen+2024}.}}
-    \label{fig:observed_density_profiles}
-\end{figure}
-
-\section{Methods}
-
-\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
-To convert observed positions and velocities to Galactocentric
-coordinates, we use the Astropy v4 Galactocentric frame
-\citep{astropycollaboration+2022}. Our Cartesian Galactocentric
-coordinates here assume the Galactic centre is at
-\([x, y, z] = [0,0,0]\), where \(x\) is the direction from the sun to
-the Galactic centre, \(y\) is the direction of the motion of the Local
-Standard of Rest, and \(z\) is the direction perpendicular to the
-Galactic plane. The coordinate frame is also right-handed, such that the
-\(z\)-angular momentum of the sun is negative (since the sun is at
-\(x<0\)). In this frame, the solar position is
-\([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
-\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
-velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
-km/s
-\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
-The uncertainties in this reference frame are typically smaller than the
-uncertainties on a dwarf galaxy's distance and tangential velocity.
-
-We adopt the Milky Way potential described in \citet{EP2020}, which is
-an analytic approximation to that proposed by \citet{mcmillan2011}.
-The Galactic bulge is described by a \citet{hernquist1990} potential,
-\begin{equation}{
-    \Phi(r) = - \frac{GM}{r + a},
-}\end{equation} 
-where \(a=1.3\,{\rm kpc}\) is the scale radius and
-\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
-disks are represented with the \citet{miyamoto+nagai1975} cylindrical
-potential, 
-\begin{equation}{
-    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
-}\end{equation} 
-where \(a\) is the disc radial scale length, \(b\) is
-the scale height, and \(M\) is the total mass of the disk. For the thin
-disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
-\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
-\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
-The halo is an NFW \citep{NFW1996, 1997} dark matter halo (Eq.~\ref{eq:nfw}) 
-\begin{equation}\protect\phantomsection\label{eq:nfw}{
-\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
-}\end{equation}
-where $r_s$ and $\rho_s$ are the scale mass and density. We parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
-In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\), or
-\(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
-
-\subsection{Orbits}\label{sec:scl_smallperi}
-
-To explore the possible orbits of a dwarf galaxy, we perform a Monte
-Carlo sampling of the present-day observables. The present-day position,
-distance modulus, LOS velocity, and proper motions are each sampled from
-normal distributions given the reported uncertainties in
-Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
-each sampled position/velocity back in time for 10 Gyr using \agama{}
-\citep{agama}. Dynamical friction is not expected to impact orbits
-substantially, so we assume a single point-mass particle for the backwards
-integration.
-Figs.~\ref{fig:scl_orbits},~\ref{fig:umi_orbits} shows 100 samples of these orbits.
-
-To maximize tidal effects, we select an orbit with the \(\sim 6\sigma\)
-smallest pericentre among all possible orbits integrated backwards for
-\(10\,\Gyr\). We achieve this by taking the median parameters of all
-orbits with a pericentre less than the 0.0027th quantile pericentre,
-yielding a pericentre of 43 kpc. Given the current observations, it is
-unlikely that Sculptor has had a significantly smaller pericentre than
-our selected orbit, which we refer to as the \smallperi{} orbit. We take
-the first apocentre after a look-back time of 10 Gyr, or at
-\(\sim9.1\,\Gyr\), as the initial conditions for our model of Sculptor,
-noted in Table~\ref{tbl:orbit_ics}.
-
-We iteratively adjust Ursa Minor's initial
-conditions to better match the present-day position. Initially, starting with low-resolution runs, we adjust the
-cylindrical actions of the initial orbit by the final difference in
-actions at the end of orbital evolution. After the initial actions have
-converged (2 iterations), we change the initial action angles by the
-final difference in action angles. This method converges within 4
-iterations to an orbit agreeing with the observed kinematics of Ursa
-Minor. Since Sculptor's orbit is less strongly affected by tides, we do
-not carry out this correction for Sculptor.
+\caption{Similar to Table~\ref{tbl:scl_orbit_props}, except for the orbital properties of UMi.}
+\label{tbl:umi_orbit_props}
+\begin{tabular}{llll}
+\hline
+Property &  MW orbits & MW N-body & LMC orbits\\
+\hline
+pericentre $/$ kpc & $37\pm3$ & 30 & $45.6^{+11.0}_{-8.2}$\\
+LMC pericentre $/$ kpc & -- & -- & $118.1^{+3.7}_{-83.0}$ \\
+apocentre $/$ kpc & $83\pm4$ & 75 &  $131.0^{+300.0}_{-18.0}$\\
+time of last pericentre $/$ Gyr  & $-0.96\pm0.07$ & $-0.80$ & $-2.14^{+0.24}_{-0.29}$ \\
+time of last peri LMC $/$ Gyr  & -- & -- & $-3.43^{+0.45}_{-0.71}$ \\
+number of pericentres & 7--8 & 8 & $3\pm1$ \\
+final heliocentric distance & $70.1\pm3.6$ & 64.7 & $70.1 \pm 3.6$\\
+\hline
+Jacobi radius $/$ kpc & $3.7\pm0.3$ & 2.9 & $5.2^{+1.3}_{-1.0}$ \\
+Jacobi radius $/$ arcmin & $184\pm12$ & 156 & $259.0^{+54.0}_{-50.0}$ \\
+LMC Jacobi radius $/$ kpc & -- & -- &  $22.2^{+0.71}_{-15.0}$ \\
+LMC Jacobi radius $/$ arcmin & -- & -- & $1079.0^{+23.0}_{-730.0}$ \\
+\hline
+break radius $/$ kpc  & $-4.65^{+0.35}_{-0.39}$ & 3.9 &  $10.3^{+1.2}_{-1.5}$\\
+break radius $/$ arcmin &  $-228.0^{+12.0}_{-13.0}$ & 206 & $507.0^{+45.0}_{-53.0}$ \\
+LMC break radius $/$ kpc  & -- & -- & $16.6^{+2.3}_{-3.5}$ \\
+LMC break radius $/$ arcmin & -- & -- &  $814.0^{+83.0}_{-130.0}$\\
+\hline
+\end{tabular}
+\end{table*}
 
@@
 \centering
-\caption{The orbital initial conditions for models presented in this work.
-The observables record the intended final position of the N-body model derived from point-mass orbit quantiles.
-Instead, the initial position and velocity represent the initial conditions for the actual N-body model.
-The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects.
-We describe the \texttt{LMC-flyby} orbit in Section \ref{sec:scl_lmc}.
-}
-\label{tbl:orbit_ics}
+\caption{The simulation properties for Scl and UMi.}
+\label{tbl:sim_results}
 \begin{tabular}{llll}
 \hline
-Property & Scl & Scl-\verb|LMC| & UMi\\
+Property & Scl MW & Scl LMC & UMi MW \\
 \hline
-$\vmax / \kms$ & 31 & 25 & 38 \\
-$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
-$R_{s, \rm exp}$ & 0.10 & 0.10 \\
-$R_{s, \rm Plummer}$ & 0.20 & 0.20 \\
-$N_\star$ \\
+$r_\textrm{max, f} / r_\textrm{max, i}$ & 0.406 & 0.763 & 0.249 \\
+$\V_\textrm{max, f} / \V_\textrm{max, i}$ &  0.695 & 0.928 & 0.511 \\
+$f_{\rm lost, DM}$ & 0.917 & 0.451 & 0.965 \\
 \hline
-distance $/$ kpc & 82.6 & 73.1 & 64.6\\
-$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
-$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
-LOS velocity $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
+$R_{h, \rm exp, i} / \kpc$ &  0.168 & 0.186 & 0.168\\
+$R_{h, \rm exp, f}/\kpc$ &  0.189 & 0.189 & 0.212\\
+$\sigma_{h, \rm exp, i}/\kms$ & 9.8 & 9.0 & 11.0 \\
+$\sigma_{h, \rm exp, f}/\kms$ & 8.8 & 8.8 & 8.8 \\
+$f_{\rm lost, exp}$ & $2.1\times 10^{-6}$ & $5\times10^{-13}$ & 0.0014\\
 \hline
-$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
-${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
-${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
-${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
-$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
-$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
-$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
+$R_{h, \rm Plummer, i}/\kpc$ & 0.202 & 0.201 & 0.201 \\
+$R_{h, \rm Plummer, f}/\kpc$ & 0.227 & 0.205 & 0.257\\
+$\sigma_{h, \rm Plummer, i}/\kms$ & 10.7 & 9.4 & 12.1\\
+$\sigma_{h, \rm Plummer, f}/\kms$ & 9.4 & 9.2 & 9.1\\
+$f_{\rm lost, Plummer}$ & $0.024$ & 0.0013 & 0.068\\
 \hline
@@
 
-\subsection{Initial conditions}\label{initial-conditions}
+% Fig.~\ref{fig:umi_sim_images} shows the dark matter evolution of our UMi
+% model, the most tidally affected model. The dark matter creates large wakes and
+% streams encircling the Galaxy several times, but the inner dark matter
+% structure remains intact (not shown).
 
-We use \agama{} \citep{agama} to generate the initial N-body dark matter
-halo. We assume galaxies are described by a spherical, isotropic NFW
-dark matter potential (Eq.~\ref{eq:nfw}). We also assume the stars do
-not contribute to the potential. The dark matter density is truncated in
-the outer regions by
-\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
-    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
-}\end{equation}
-where we adopt \(r_t = 20 r_s\).
+%\begin{figure}
+%    \centering
+%    \includegraphics[width=\columnwidth]{figures/umi_sim_images.png}
+%    \caption{
+%        Images of Ursa Minor's dark matter over a selection of past apocentres
+%        and the present day. The colours represent the projected DM density in
+%        the $y$-$z$ plane, scaled to be logarithmic over 5 orders of magnitude.
+%        The green dot represents the final expected position of the galaxy, and
+%        the solid and dotted grey curves represent the orbit over one previous
+%        or future radial oscillation, respectively.
+%}
+%    \label{fig:umi_sim_images}
+%\end{figure}
 
-We select cosmologically-motivated initial dark matter halos. First, taking the absolute
-magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
-\citet{woo+courteau+dekel2008}, the total current stellar mass of
-Sculptor and Ursa Minor are \(M_\star \approx 3.1 \times 10^6 \Mo\) and
-\(M_\star \approx 7 \times 10^5 \Mo\), respectively. Based on the
-stellar mass-\(\vmax\) relation \citep[from][]{fattahi+2018}, Sculptor
-and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
-\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
-\(z=0\) mass-concentration relation, this constraint translates into a
-\(\rmax \approx 6\,\kpc\) and \(\rmax \approx 5\,\kpc\) for each galaxy.
-The cosmological means predict too low a velocity dispersion
-(for Scl and
-UMi, \(\sigma_{\V, i}  \approx 8.5\,\kms\) and \(8.0\,\kms\) given their observed half-light radius). 
-To match the observed \(\sigma_\V\) at the end of tidal evolution, we
-choose \(\vmax = 31\,\kms\) and \(\rmax=3.2\,\kpc\) for Scl and
-\(\vmax=38\,\kms\) and \(\rmax=4\,\kpc\) for UMi.
+\section{Fits to density profiles and metallicity gradients}
 
-\subsection{N-body methods}\label{isolation-runs-and-simulation-parameters}
+\subsection{Spectroscopic sample selection} \label{sec:velocity_sample}
+To create a sample of spectroscopic members, we compile LOS velocities from
+several surveys. For
+Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
+\citet{sestito+2023a}; and APOGEE \citep[DR17,][]{abdurrouf+2022}. For
+Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
+\citet{sestito+2023b}; and APOGEE. We then cross-match all catalogues to
+J+24 Gaia stars. If a study did not report \emph{Gaia} DR3 source IDs,
+we match to the nearest star within 1--3 arcseconds. We combine
+measurements of the same star using inverse-variance weighting. To
+reduce likely binaries, we remove stars with significant velocity
+dispersions.\footnote{Specifically, using that
+  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
+  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
+  distribution with \(N-1\) measurements.}
 
-To ensure that the initial conditions of the simulation are dynamically
-relaxed, we run a halo first in isolation using
-\gadget{}. Since gravity is scale-free, we use the same isolation run
-for all halos and rescale the results to the desired values of size and
-mass. We adopt a fiducial value of \(\rmax = 6.0\,\)kpc and
-\(\vmax = 31\,\kms\) for the isolation halo based on Sculptor's mean
-properties. We run this model for 5 Gyr (about one-half crossing time
-\(t_{\rm cross} = 2\pi\,r /\vcirc  \approx 9\,\Gyr\) at
-\(r_{200}=36\,\)kpc).
+We build on J+24's likelihood by adding multiplicative terms in the
+total likelihood for the velocity consistency. We assume that the satellite and
+background \(v_{\rm los, gsr}\) distributions are Gaussian in the
+Galactic Standard of Rest (GSR, i.e., same location as ICRS but
+velocities relative to the galactic centre). For the satellite, we adopt
+a mean and standard deviation based on
+Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}, and, for the
+background, mean \(0\,\kms\) and dispersion
+\(\sigma_{\rm halo} = 100\,\kms\) \citep[e.g.][]{brown+2010}. We select
+stars with velocity-informed satellite membership probabilities of
+greater than 0.2. For Scl, we find 1918 unique members and UMi, 831.
 
-For our simulation parameters, we adopt a softening length of
-\begin{equation}\protect\phantomsection\label{eq:softening_length}{
-    h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
-}\end{equation} 
-for a halo with \(N\) particles.
-Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).
+For UMi, we shifted the velocities of \citet{spencer+2018}
+(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
+a systematic velocity offset. Otherwise, all studies appear to be on a
+similar velocity scale.
 
-Next, we evolve the halo in the Galactic potential. We scale the relaxed
-snapshot and softening length to match the initial halo, and shift the
-snapshot to the initial conditions inferred from the orbital analysis
-(see Table~\ref{tbl:orbit_ics}). We then evolve the full N-body NFW
-model forward in time in the Galactic potential until the present time,
-when the halo is closest to its present-day observed position in the MW
-halo.\footnote{Specifically, the snapshot which best agrees (as measured
-  by \(\chi^2\)) with the intended final position (in
-  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
-  uncertainties. We use \(R_h\) as the uncertainty in \(\alpha\) and
-  \(\delta\).}
+We correct the velocities for the solar motion and the on-sky size of
+the galaxy. We transform the velocities into the GSR and correct for the
+apparent gradient induced by the dwarf's proper motion
+\citep[see,][]{WMO2008, strigari2010}. The
+correction from both effects induces an apparent gradient of about
+$1.3\,{\rm km\,s^{-1}\,deg^{-1}}$ for Sculptor and less for Ursa Minor.
 
-We determine the halo centre using a shrinking-spheres centre method inspired by \citet{power+2003}.
-First, we start with an initial centre estimate using all bound
-particles from the previous snapshot. Then, we calculate the distance of
-all particles from the centre, remove particles with a distance beyond
-the 0.975 quantile of the centre, and recalculate the centre of mass.
-The procedure is repeated until the selection radius is less than
-\textasciitilde1kpc or fewer than 0.1\% of particles remain. We then
-remove all unbound particles based on the \gadget{} calculated potential
-of the halo and recalculate the centre. For all future timesteps, we
-consider only particles retained from the previous iteration.
+\subsection{Density profile fits} \label{sec:double_exp_fits}
 
-\subsection{Sculptor and Ursa Minor's initial stellar
-components}\label{sec:painting_stars}
+Here, we describe our fitting procedure to derive the double exponential fits and metallicity gradients in Fig.~\ref{fig:metallicity_gradient}.
 
-We ``paint'' stars onto dark matter particles using the particle-tagging
-method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
-symmetry. We initially assume stars follow a projected exponential law
-(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,\kpc\) for both
-galaxies. The tagging method assigns a probability to each dark matter
-particle, which is proportional to the ``light-to-mass'' ratio required
-to match the assumed stellar light profile. We briefly describe the
-procedure next, but refer interested readers to \citet{EP2020}.
+For the double exponential fits, we extend \citet{jensen+2024} to fit a double exponential with free inner and outer parameters.
+Specifically, we use a similar likelihood formulation except we hold the proper motion likelihood fixed, and allow the inner and outer density components to include free position angle, ellipticity, scale radius, and fit for the relative total mass of the outer component $f_{\rm outer}$.
+We perform the fits using the No-U Turn Sampler (an adaptive Hamilton-Monte Carlo algorithm), assuming uniform priors for position angle and ellipticity, and log-uniform priors for scale radius.
 
-Let \(\Psi\) be the potential (normalized to vanish at infinity) and
-\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 v^2\). If we
-know the distribution function\footnote{i.e., the phase-space density of
-  particles. Note that since \(f\) is formally defined in 6D phase
-  space, there is an additional ``density of states'' term in order to
-  calculate the histogram of particles with \({\cal E}\),
-  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
-then we assign a stellar weight to a given particle with energy
-\({\cal E}\) using \begin{equation}{
-P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-}\end{equation} We use Eddington inversion to find the distribution
-function, 
-\begin{equation}{
-    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
-}
-\end{equation} 
-\citep[eq. 4-140b in][]{BT1987}. In practice the right,
-boundary term is zero. We
-take \(\Psi\) from the underlying assumed analytic dark matter
-potential. \(\rho_\star\) can be calculated from the surface density,
-\(\Sigma_\star\), via the inverse Abel transform.
+To fit the metallicity gradient, we instead assume that the density profile follows the double exponential fit above, and then fit a mixture model of the metallicity gradient as a sum of two Gaussian distributions.
+The four free parameters of this model are the mean and metallicity dispersion of the inner and outer components.
 
-\begin{figure}
+Table~\ref{tbl:double_exp_fits} contain the derived parameters and quantile uncertainties for both Scl and UMi.
+We find strong evidence for multiple populations with different mean metallicities.
+Our outer population scale radius is smaller than in \citet{jensen+2024} for Scl but is similar for Ursa Minor.
+
+\begin{table}
 \centering
-    \includegraphics{figures/scl_lmc_xyzr_orbits.pdf}
-    \caption{
-        Orbits of Sculptor with and without the LMC.
-    }
-    \label{fig:scl_orbits}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/umi_xyzr_orbits.pdf}
-    \caption{
-        Orbits of Ursa Minor
-    }
-    \label{fig:umi_orbits}
-\end{figure}
-
-\section{Results}
-
-\begin{itemize}
-    \item Scl minimally evolves as a result of having one pericentre
-    \item The LMC critically changes Scl's tidal history.
-    \item UMi loses DM mass but the stellar component, as observed, would not show tidal signatures
-    \item Break and Jacobi radii suport these interpretations. Stronger tidal evolution is unlikely under any other orbits in these potentials.
-\end{itemize}
-
-\subsection{Sculptor}
-\begin{table}[t]
-\centering
-\caption[Orbits and results for Sculptor in the MW+LMC potential.]{The orbital properties and dark matter evolution for the models including an LMC.
-Similar to Table \ref{tbl:scl_sim_results} except quantities with respect to the LMC are in parentheses.
-}
-\label{tbl:scl_lmc_sims}
-\begin{tabular}{lll}
+\caption{The double exponential fit results for Scl and UMi}
+\label{tbl:double_exp_fits}
+\begin{tabular}{lllll}
 \hline
-Property & random samples & Scl\\
+Parameter & Scl  & Scl  & UMi  & UMi  \\
+ & inner & outer & inner & outer \\
 \hline
-pericentre $/$ kpc & $44\pm 3$ ($29 \pm 2$) & 39 (20)\\
-apocentre $/$ kpc & $218 \pm 8$ & –\\
-time of last pericentre $/$ Gyr & $-0.38\pm0.01$ ($-0.11$) & $-0.33$ ($-0.10$)\\
-number of pericentres & 1 (1) & 1 (1)\\
-Jacobi radius $/$ kpc & $3.3\pm0.2$ ($3.6\pm0.2$) & 2.8 (2.6)\\
-Jacobi radius $/$ arcmin & $136 \pm 9$ ($159\pm5$) & 132 (121)\\
-final heliocentric distance $/$ kpc & $83.2\pm2$ & 72.9\\
-$\V_\textrm{max, f} / \V_\textrm{max, i}$ &  & 0.928\\
-$r_\textrm{max, f} / r_\textrm{max, i}$ &  & 0.763\\
-fractional final bound mass &  & 0.5402\\
+$R_h$ & $7.85^{+0.32}_{-0.33}$ & $17.1^{+1.3}_{-1.2}$ & $10.58^{+0.52}_{-0.42}$ & $29.7^{+3.4}_{-3.2}$\\
+$\theta$ & $98.8^{+4.2}_{-4.0}$ & $91.8^{+3.0}_{-3.1}$ & $50.4^{+1.7}_{-1.6}$ & $74.0^{+68.0}_{-45.0}$ \\
+ell & $0.243^{+0.031}_{-0.034}$ & $0.383^{+0.035}_{-0.036}$ & $0.598^{+0.025}_{-0.029}$ &  $0.087^{+0.086}_{-0.059}$ \\
+$f_{\rm outer}$ & --- & $0.315^{+0.058}_{-0.052}$ & --- & $0.249^{+0.043}_{-0.041}$ \\
+\hline
+$\mu_{\rm [Fe/H]}$ &  $-1.671^{+0.021}_{-0.021}$ &  $-1.982^{+0.024}_{-0.024}$ & $-2.147^{+0.02}_{-0.02}$ & $-2.523^{+0.095}_{-0.087}$ \\
+$\sigma_{\rm [Fe/H]}$ & $0.574^{+0.016}_{-0.015}$ & $0.336^{+0.019}_{-0.018}$ & $0.347^{+0.015}_{-0.014}$ & $0.505^{+0.067}_{-0.066}$ \\
 \hline
@@
 
-\begin{figure*}
-    \centering
-    \includegraphics{figures/scl_lmc_sim_images.png}
-    \caption{Snapshots of Scl in the LMC and MW potential.}
-    \label{fig:scl_sim_images}
-\end{figure*}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/scl_density_i_f.pdf}
-    \caption{Simulated initial and final stellar density profiles for a tidal simulation of Scl in the MW+LMC potential} 
-    \label{fig:scl_density_i_f}
-\end{figure}
-
-\subsection{Ursa Minor}
-\begin{figure*}
-    \centering
-    \includegraphics{figures/umi_sim_images.png}
-    \caption{Snapshots of UMi's tidal evolution in the MW potential}
-    \label{fig:umi_sim_images}
-\end{figure*}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/umi_density_i_f.pdf}
-    \caption{Similar to \ref{fig:scl_density_i_f} except for UMi.} 
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
-\subsection{Comparison with prior work}\label{comparison-with-prior-work}
-
-A few studies have modelled Sculptor or Ursa Minor specifically, finding
-similar conclusions. \citet{iorio+2019} applied idealized N-body
-simulations to study tidal effects on Scl. They similarly found weak
-tidal effects, even for a dark-matter-free model. Most recently,
-\citet{tchiorniy+genina2025} also used idealized simulations tuned to
-five classical dwarfs with a focus on the inner density, concluding that
-tides do not strongly affect equilibrium assumptions or the stellar
-component.
-
-Our work extends these models, adding updated structural properties, a
-broader range of orbital histories, and the influence of the LMC.
-Despite these considerations, we still reach similar conclusions: tidal
-effects do not shape Scl or UMi's stellar component. A non-tidal process
-likely underlies the origin of these galaxies' outer density excess.
-
-Besides their extended stellar density profiles, Sculptor and Ursa Minor
-display other peculiarities that may hold clues to their formation. Scl
-and UMi both host at least two distinct chemodynamic populations, as
-revealed through their photometric or metallicity-velocity
-structure\footnote{Other examples of galaxies with multiple populations
-  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
-  Fornax
-  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
-  Sextans
-  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
-  Andromeda II
-  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
-\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
-younger, higher metallicity, and dynamically colder, whereas the outer
-population is older, lower metallicity, and dynamically hotter.
-These populations may map onto the observed density excess for Ursa Minor, and \citepos{arroyo-polonio+2024} third population would reside within the density excess.
-
-%Ursa Minor has also shown evidence for possible inner substructure, such
-%as stellar or kinematic ``clumps''
-%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
-%While one clump was re-detected kinematically in \citet{pace+2014},
-%\citet{munoz+2018} find no evidence of substructure with modern
-%photometry.
-
-\subsection{Discussion and Conclusions}\label{sec:stellar_halos}
-In this work, we have investigated the extended, outer density excess
-of Sculptor and Ursa Minor and its possible tidal origin. We have shown
-that the density profiles are robust, implying that Scl and UMi contain
-a true density excess relative to other classical dwarfs.
-
-We then investigated whether tides are a possible explanation. By
-modelling each galaxy based on cosmological initial conditions, we
-showed that tides do not strongly affect either galaxy. The LMC changes
-the orbital history of Scl and UMi, but tides become even weaker in a
-combined LMC and MW potential. We conclude that recent tides are
-unlikely to shape the outer radial distribution of stars in Scl and UMi.
-
-\textbf{Episodic star formation.} Star formation may quench and
-reignite, creating successive stellar generations with differing
-distributions. External star formation triggers include tidal
-compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
-gaseous filaments \citep{genina+2019}, perturbations from dark halos
-\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
-\citep{wright+2019}. More common mechanisms, like feedback or
-reionization-driven quenching, may also form multiple stellar
-generations
-\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
-such processes would not explain why extended stellar populations appear
-to be non-universal. If an extended population was formed during a
-separate star formation event, then the star formation history should
-contain evidence of a corresponding burst.
-
-\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
-\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
-the lower-mass galaxy are preferentially dispersed, forming an extended
-stellar component and population gradient
-\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
-the merger can also trigger new star formation, forming a younger
-population of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are
-suspected to have undergone a major merger, including Tucana II,
-Andromeda II, and Phoenix
-\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
-A galaxy having undergone a major merger should harbour at least two or
-three populations with distinct origin.
-
-Just as the Milky Way's halo is likely built from many minor mergers,
-dwarf galaxies may have built accreted ``stellar halos'' composed of
-even fainter satellites \citep{ricotti+polisensky+cleland2022}. Few
-examples of dwarf-dwarf accretion are known, but a possible stream
-around And II or chemical peculiarities around Sextans I may be such
-instances \citep{amorisco+evans+vandeven2014, roederer+2023}. In this
-scenario, the extended stellar populations would contain chemical
-signatures from several distinct ultra-faint dwarf galaxies.
-
-\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
-``preprocessed'' by a massive satellite like the LMC
-\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
-redistributes already-present stellar populations and may mimic a
-stellar halo. Key evidence suggestive of preprocessing may include
-distant dwarf stars or stellar streams.
-
-\textbf{Dark-matter-free dwarfs.} Tidal dwarfs are galaxies that formed
-in gas-rich tidal streams during the merger of two larger galaxies
-\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
-dark matter halo, tidal dwarfs may be more susceptible to tides, forming
-extended density profiles and inflating the velocity dispersion
-\citep{casas+2012, yang+2014, wang+2024a}.
-
-Similarly, in Modified Newtonian Dynamics (MOND), galaxies do not
-contain dark matter. Instead, the gravitational force law is altered to
-explain rotation curves. Dwarfs would similarly experience stronger
-tidal effects in MOND \citep{mcgaugh+wolf2010, brada+milgrom2000}.
-
-In both cases, tides more plausibly produce extended density profiles.
-However, recovering the observed velocity dispersions often requires
-ongoing tidal disruption
-\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
-also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
-evidence of such features {\bf Ref}---future data would be needed to uncover
-signs of disruption to support this hypothesis.
-
-\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
-dynamically hotter than younger stars due to processes including stellar
-feedback
-\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
-sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
-matter interference fringes
-\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
-these processes should operate similarly across dwarf
-galaxies---extended stellar populations should be more common if these
-processes are important.
-
-We have reviewed a number of different, possibly concurrent explanations
-for the formation of extended populations. While we leave the nature of
-Scl and UMi's extended stellar densities an open question, we can
-discuss possible clues to different formation scenarios.
-\emph{Precise chemistry}. Chemical abundances, comparing the inner and
-outer regions, can test if the halo was formed within the dwarf or
-accreted. If external in origin, chemistry could reveal the properties
-of any past mergers.
-\emph{Detailed kinematics}. For models relying on recent tidal
-disruption, kinematic disequilibrium features should be visible. These
-would appear as velocity gradients, increasing velocity dispersions,
-outward-biased moving stars, or non-phase mixed structures
-\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}. In
-addition, measurements of the ellipticity and anisotropy would help
-verify or alter our understanding of the mass structure and whether
-DM-free models may be permissible.
-\emph{Deep photometry} may find or rule out signs of dynamical
-disequilibrium and tidal tails for tidally susceptible models (e.g.,
-\emph{pre-processing} and \emph{DM-free dwarfs}). In addition,
-photometry will help constrain the prevalence and nature of extended
-features in dwarf galaxies.
-\emph{Star formation histories} can be derived through photometry or
-chemistry. Evidence of significant star formation episodes or lack
-thereof may differentiate scenarios that rely on strong star formation
-bursts (e.g., \emph{episodic star formation history, induced star
-formation,} and \emph{gas-rich mergers}).
-
-Both upcoming observatories and novel simulation methods will be instrumental to uncovering the possible origins of the extended stellar densities in Sculptor and Ursa Minor.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgements}
-\end{acknowledgements}
-
-\bibliographystyle{bibtex/aa.bst} % style aa.bst
-\bibliography{paper.bib} % your references Yourfile.bib
-
 \end{document}

### 2025-11-13 thesis/acknowledgements.tex
paper updates

@@
 
-To Raphaël, who has helped inspire and direct this project. To everyone
-who has helped provide me opportunities to share work, and share
-wonderful ideas with me. Isabel, Jens, Eduardo, Jorge, David, Justin,
-Eugene, Azi, José, and many others.
+To Raphaël, who has helped inspire and direct this project, and to
+everyone who has shared ideas and opportunities along the way: Isabel,
+Jens, Eduardo, Jorge, David, Justin, Eugene, Azi, José, and many others.
 
-To Jax, who has been a support along the way, guiding my research
-methods and project.
+To Jax for you guidance and personal support for this project.
 
-To all of my new friends I have made along the way. Ruxin, Aviv. To
+To all of my new friends I have made along the way---Ruxin, Aviv. To
 Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud, Anne, Anya,
 Shoshanah. To Akshara and Simon for all your personal and professional
-advice. We have built a wonderful community of students in this
-department. To all the students who have since moved on and show what we
-may be: Bobby, Leo, Ryan, Katie, Fletcher, Ayden, Maheyer, Jess, Dhwani,
-Spencer, Mallory, Omar, Isaac, and Dasha.
+advice. We have built a wonderful graduate community. To all the
+students who have since moved on but still inspire me: Bobby, Leo, Ryan,
+Katie, Fletcher, Ayden, Maheyer, Jess, Dhwani, Spencer, Mallory, Omar,
+Isaac, and Dasha.
 
-And to my friends across UVic, Wyatt, Rosemary, Maggie. To the
-Linguistics extended family and team \emph{adequate}. Melody, Lily,
-Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are all
-wonderful human beings, cooks, and always a joy to spend time with.
+To my friends across UVic: Wyatt, Rosemary, Maggie. To the Linguistics
+extended family and team \emph{adequate}: Melody, Lily, Gillian,
+Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are all wonderful
+human beings, cooks, and always a joy to spend time with.
 
-To my musical friends, Adrienne, Alex for our shared joy in music
+To my musical friends, Adrienne and Alex, for our shared joy in music
 making. And Steve, for showing me the joy in a simple life and beautiful
@@
 
-To my fellow fibre friends and crafters. Leise, Wes, Ashley, Sabrina,
+To my fellow fibre friends and crafters: Leise, Wes, Ashley, Sabrina,
 Sarah. To the graduate student society for continuing to connect us
@@
 To the international community, and my international friends---Marcela,
-Bernardo, Daniel, Victor, Walla, Mehrab, and everyone else I have met.
+Bernardo, Daniel, Victor, Mehrab, and many others.
 
@@
 
-To my long-time friends. Nikola and Eric. Mikey and Anthony. Anya, Kaia,
-Alyssa, Aahliyah, Cameron. May we continue to find each other even as we
-drift apart.
+To my long-time friends---Nikola, Eric, Mikey, Anthony, Anya, Kaia,
+Alyssa, Aahliyah, Cameron---may we continue to stay in touch.
 
 To my parents for their continuous support throughout this project. To
-my grandparents for always being there.
-
-To Nolan and Olivia, always a joy to see each other. To Bridget and
-Gretchen. To my extended family. And to my siblings for always being a
-source of joy.
+my grandparents for always being there. To Nolan and Olivia, always a
+joy to see each other. To Bridget and Gretchen. To my extended family.
+And to my siblings for always being a source of joy.
 

### 2025-11-13 thesis/introduction.tex
paper updates

@@
 \centering
-\caption[Observed Properties of Sculptor]{Observed properties of Sculptor.
+\caption[Observed properties of Sculptor]{Observed properties of Sculptor.
 References are: (1) Muñoz et al.
@@
 \centering
-\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor.
+\caption[Observed properties of Ursa Minor]{Observed properties of Ursa Minor.
 References are: (1) Muñoz et al.

### 2025-11-13 orbits/bootes3/mean_reflex_motion.jl

@@ -56,14 +56,17 @@ icrs = ICRS(obs_props)
 # ╔═╡ d0cf64df-d46b-4776-9b46-d1c69a02cc61
 units = Agama.VASILIEV_UNITS
 
+# ╔═╡ f9075bfd-c7ab-4bf4-8774-7e7074a0c29f
+timerange = (0, 1/T2GYR)
+
 # ╔═╡ 77503fde-9be2-4a7b-bed0-847297be2c76
-orbit = LilGuys.agama_orbit(pot, icrs, timerange=(0, -3/T2GYR), agama_units=units)
+orbit = LilGuys.agama_orbit(pot, icrs, timerange=timerange, agama_units=units)
 
 # ╔═╡ f7c3629d-4339-4e8f-b77b-f9c48ea0bbe4
-orbit_no = LilGuys.agama_orbit(pot_mw_static, icrs, timerange=(0, -3/T2GYR), agama_units=units)
+orbit_no = LilGuys.agama_orbit(pot_mw_static, icrs, timerange=timerange, agama_units=units)
 
 # ╔═╡ 6aea4a07-3405-4e04-a4d4-b31598455043
-orbit_noreflex = LilGuys.agama_orbit(pot_noreflex, icrs, timerange=(0, -3/T2GYR), agama_units=units)
+orbit_noreflex = LilGuys.agama_orbit(pot_noreflex, icrs, timerange=timerange, agama_units=units)
 
 # ╔═╡ 299038ee-d228-4eb9-9a97-708a95668fd2
 acc_tot = Agama.acceleration(pot, orbit.positions, units, t=orbit.times)
@@ -239,6 +242,7 @@ lines(times, L[3, :])
 # ╠═6195c401-2dc7-425b-a082-8bbda16d0ba3
 # ╠═2ccdac03-56fd-4f30-a237-b473435fe321
 # ╠═d0cf64df-d46b-4776-9b46-d1c69a02cc61
+# ╠═f9075bfd-c7ab-4bf4-8774-7e7074a0c29f
 # ╠═77503fde-9be2-4a7b-bed0-847297be2c76
 # ╠═f7c3629d-4339-4e8f-b77b-f9c48ea0bbe4
 # ╠═6aea4a07-3405-4e04-a4d4-b31598455043

### 2025-12-08 paper/paper.tex
minor edits

@@
 Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little.
-UMi instead loses about 95\% of its dark matter mass, but the stars would still appear unaffected with current observations.
+UMi instead loses about 95\% of its dark matter mass, but the stars should appear unaffected.
 }
@@
 Two radii provide estimates of where tidal effects occur.
-The \textbf{Jacobi radius}, $r_{\rm J}$. approximates the radius where stars
-become unbound in a galaxy\footnote{ 
+The \textbf{Jacobi radius}, $r_{\rm J}$, approximates the radius where stars
+become unbound in a galaxy.\footnote{ 
     While strictly valid only for circular orbits, assuming
 \(r_{\rm peri}\) for the host-dwarf distance works as most stars are
-lost during pericentric passages. $r_{\rm J}$ solves
-}
+lost during pericentric passages.} $r_{\rm J}$ solves
 \begin{equation}\label{eq:r_jacobi}
@@
 features may be lost in the noise or be of uncertain association.
-In addition to colours and magnitudes, \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallax and proper motion which improve membership assignment.
+\emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallaxes and proper motions, improving membership by providing key additional information.
 
 Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
-probabilities from \emph{Gaia} data.\footnotemark{} J+24 used a Bayesian framework
+probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
 incorporating proper motion, colour--magnitude diagram, and
@@
 the algorithm presented in \citet{MV2020a, MV2020b} by additionally
-including a secondary, extended spatial component, J+24 detected
+including a secondary, extended spatial component.\footnotemark{} J+24 detected
 candidate members out to ${\sim}10$ half-light radii from the
@@
 against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
-J+24).\footnotemark{} However, we find that our conclusions are unchanged when
-limiting samples to only the brightest stars. Altogether, the J+24 method
+J+24).\footnotemark{} Altogether, the J+24 method
 provides a high-quality, low-contamination sample of dwarf galaxy candidate
@@
   (which have better \emph{Gaia} measurements) and likely members,
-  complicating the interpretation of this purity estimate.}
+  complicating the interpretation of this purity estimate. However, we find that our conclusions are unchanged when
+limiting samples to only the brightest stars. }
 
@@
 (solid black line). These galaxies are better fit by a Plummer law
-(dashed black line). The deviation from an exponential grows outwards,
-and at \(\sim 8 R_h\), may reach 2 orders of magnitude.
+(dashed black line). The deviation from an exponential grows outwards
+and may reach 2 orders of magnitude at $8 R_h$.
 
@@
 an analytic approximation to that proposed by \citet{mcmillan2011}.
-The Potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
+The potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
 A NFW density profile is defined as,
@@
 noted in Table~\ref{tbl:orbit_ics}.
-
-For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties), and only integrate backwards for $2\,\Gyr$.
+For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties). We only integrate this model backwards for $2\,\Gyr$.
 
@@
 halo. We assume galaxies are described by a spherical, isotropic 
-dark matter halo (Eq.~\ref{eq:nfw}). The NFW
+NFW dark matter halo (Eq.~\ref{eq:nfw}). The NFW
 profile truncated in the outer regions by
@@
 To ensure that the initial conditions of the simulation are dynamically
-relaxed, we run a halo first in isolation using
-\gadget{}. We use the same isolation run
+relaxed, we run a halo first in isolation. We use the same isolation run
 for all halos (with $\rmax=6$ and $\vmax=31$, ran for $5\,\Gyr$) and rescale the results to the desired initial halo.
@@
 \citep[see eq. 4-140b in][]{BT1987}. We
-take \(\Psi\) and $\rho$ from the assumed analytic density profiles. We calculate \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.
+take \(\Psi\) and $\rho$ from the assumed analytic density profiles, calculating \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.
 
@@
 for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
-We also consider a more extended Plummer law, 
+We also consider a more extended \citep{plummer1911} law component, 
 \begin{equation}{
@@
 for half-light radius $R_h$. 
-
 Fig.~\ref{fig:initial_halos} shows the resulting initial conditions for the MW-only models before and after 5 Gyr of isolation evolution.
@@
 The black lines in \ref{fig:orbits} show our selected orbits (black), representing an extreme low-pericentre orbit.
-We adopt these orbits as the most severe observationally-permisible tidal histories of each galaxy.
+We adopt these orbits as the most severe observationally-permissible tidal histories of each galaxy.
 
@@
 evolution, and substructure, causing analytic orbits to deviate significantly
-from true satellite orbits \citep[e.g.,]{dsouza+bell2022, santistevan+2023}.
+from true satellite orbits \citep[e.g.,][]{dsouza+bell2022, santistevan+2023}.
 While we cannot constrain the long-term histories of these dwarfs, our results
-indicate that {\it recent tidal effects} would not be observable in Scl and
+indicate that {\it recent tidal effects} are likely not be observable in Scl and
 UMi.

### 2025-12-11 paper/plots/velocity_dispersions.jl
updated paper and thesis presentation plots

@@ -88,6 +88,9 @@ function calc_sigma_v_prof(stars, w = stars.weights)
 	
 	r_bins = DensityEstimators.make_bins(x, (-1, 3), DensityEstimators.bins_equal_number, n=bins)
 
+	r_max = LilGuys.quantile(x, w, 1-1e-6)
+	r_bins = r_bins[r_bins .< r_max]
+	bins = length(r_bins) - 1
 	
 	σs = Vector{Float64}(undef, bins)
 	Ns = Vector{Float64}(undef, bins)
@@ -108,38 +111,7 @@ end
 function plot_σv_prof!(galaxyname, args...; filename="final.fits", R_shift=0, kwargs...)
 	stars = get_stars_final(galaxyname, args..., filename)
 
-
-	x = log10.(stars.r_ell)
-	w = stars.weights
-	v = stars.radial_velocity
-
-	filt = @. !isnan(x)
-
-	x = x[filt]
-	w = w[filt]
-	v = v[filt]
-
-	bins = 80
-	
-	r_bins = DensityEstimators.make_bins(x, (-1, 3), DensityEstimators.bins_equal_number, n=bins)
-
-	
-	σs = Vector{Float64}(undef, bins)
-	err = Vector{Float64}(undef, bins)
-	Ns = Vector{Float64}(undef, bins)
-	
-	for i in 1:bins
-		filt = r_bins[i] .<= x .< r_bins[i+1]
-		σs[i] = calc_σv(x[filt], v[filt], w[filt], r_max=Inf)
-		N = sum(filt)
-		if N > 1
-			err[i] = σs[i] * sqrt(2/(N - 1))
-		else
-			err[i] = NaN
-		end
-		Ns[i] = N
-	end
-
+	r_bins, σs, Ns, Ms = calc_sigma_v_prof(stars)
 
 	lines!(LilGuys.midpoints(r_bins), σs ./ sqrt(10^R_shift); kwargs...)
 
@@ -327,6 +299,45 @@ end
 
 end
 
+# ╔═╡ 2a382105-9db1-48fd-96c5-6c3922d3c4c7
+@savefig "velocity_dispersion_profiles" let
+	fig = Figure()
+
+	ax_scl_sigma = plot_σv_prof(fig[1,1], modelnames["scl_smallperi_2exp"]..., color=COLORS[5], label="2exp")
[…]

### 2025-12-11 presentations/thesis/classical_dwarf_densities.jl

@@ -97,13 +97,15 @@ galaxies = OrderedDict(
 		marker = :dot,
 		label = "Fornax",
 		linestyle=:solid,
-		#linewidth=2
 	),
 	"sculptor" => (
 		color = COLORS[2], 
 		marker=:rect,
 		label = "Sculptor",
 		linestyle=:solid,
+		linewidth = 1.5*theme(:linewidth)[],
+		markersize=1.5*theme(:markersize)[]
+
 		#linewidth=1.75
 	),
 	"ursa_minor" => (
@@ -111,10 +113,16 @@ galaxies = OrderedDict(
 		marker = :star5,
 		label = "Ursa Minor",
 		linestyle=:solid,
+		linewidth = 1.5*theme(:linewidth)[],
+		markersize=1.5*theme(:markersize)[]
+
 		#markersize=8,
 		#linewidth=1.5
 	),
 )
+
+# ╔═╡ b05a2d84-2b5f-4d37-8c45-48e55152c43a
+theme(:linewidth)
 
 # ╔═╡ 552e9438-862f-4710-a7d4-c8d798b5f1aa
 galaxynames = [
@@ -182,7 +190,7 @@ let
 	ax = Axis(fig[1,1], 
 		ylabel = L"log surface density / $\Sigma_h$",
 				xticks = -2:1:1,
-		limits=(-1.5, nothing, -4.5, 1.1)
+		limits=(-1.5, nothing, -3.5, 0.9)
 
 	)
 
@@ -198,7 +206,7 @@ let
 		prof = profiles[galaxy]
 		lines!(prof.log_R, prof.log_Sigma, color=COLORS[1], alpha=0.5, label="classical dwarfs", linestyle=:solid)
 	end
-	axislegend(position=:lb, merge=true, unique=true, patchsize=(48, 6))
+	axislegend(position=:lb, merge=true, unique=true, patchsize=(48, 6), margin=theme(:Legend)[:margin][])
 
 
 	# residuals
@@ -206,7 +214,7 @@ let
 		ylabel = "residual",
 		xlabel = L"log Radius / $R_h$",
 		xticks = -2:1:1,
-		limits=(-1.5, 1.1, -1.5, 1.5),
+		limits=(-1.5, 1.0, -1.5, 1.5),
 		yticks = -2:1:2
 	)
 
@@ -246,7 +254,7 @@ let
[…]

### 2025-12-11 presentations/thesis/scl_density_i_f.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.20.15
+# v0.20.21
 
 using Markdown
 using InteractiveUtils
@@ -30,7 +30,7 @@ R_h = TOML.parsefile(joinpath(ENV["DWARFS_ROOT"], "observations/sculptor/observe
 
 # ╔═╡ 65d3653d-5734-40ec-a057-aaa1e509968a
 prof_expected = let
-	prof = SurfaceDensityProfile(ENV["DWARFS_ROOT"] * "/observations/sculptor/density_profiles/jax_2c_eqw_profile.toml")
+	prof = SurfaceDensityProfile(ENV["DWARFS_ROOT"] * "/observations/sculptor/density_profiles/fiducial_profile.toml")
 		
 	prof =  LilGuys.filter_empty_bins(prof)
 
@@ -83,12 +83,6 @@ function get_r_j(haloname, orbitname; lmc=false)
 	return props["r_J"] / R_h
 end
 
-# ╔═╡ 991fe214-c10c-44a6-a284-0356ef993412
-get_r_j("1e7_V31_r3.2", "orbit_smallperi", )
-
-# ╔═╡ 43c973e6-ad30-4bf8-93b3-4924bc498927
-readdir("/cosma/home/durham/dc-boye1/data/dwarfs/analysis/sculptor/1e7_V31_r3.2/orbit_mean/stars/")
-
 # ╔═╡ f875e0e9-2f05-4420-8f40-76284be58e03
 function get_normalization(prof)
 	return -LilGuys.mean(prof.log_Sigma[1:5])
@@ -127,33 +121,17 @@ function compare_profiles(prof_i, prof, r_b; break_height=-6, norm_shift = 0, r_
 		lines!(prof.log_R, prof.log_Sigma .+ dy, color=COLORS[2], linestyle=:solid,
 				label="final")
 	end
-	
-	if !isnothing(break_height)
-		annotation!(0, 60, log10(r_b), break_height, 
-					color=COLORS[2], 
-					linewidth=theme(:linewidth)[],  
-					style=Ann.Styles.LineArrow(head = Ann.Arrows.Head(length=theme(:markersize)[])),
-				  )
-		
[…]

### 2025-12-11 paper/paper.tex
paper clarity updates

@@
-%-----------------------------------------------------------------------
+%----------------------------------------------------------------------
 %                                                                 aa.tex
@@
 \usepackage{graphicx}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{txfonts}
@@
 \newcommand{\V}{{\rm v}}
-\newcommand{\vmax}{\V_{\rm max}}
-\newcommand{\vcirc}{\V_{\rm circ}}
-\newcommand{\rmax}{r_{\rm max}}
-\newcommand{\kpc}{{\rm kpc}}
-\newcommand{\Gyr}{{\rm Gyr}}
-\newcommand{\kms}{{\rm km\,s^{-1}}}
-\newcommand{\masyr}{{\rm mas\,yr^{-1}}}
-\newcommand{\Mo}{M_\odot}
+\newcommand{\vmax}{\V_\text{max}}
+\newcommand{\vcirc}{\V_\text{circ}}
+\newcommand{\rmax}{r_\text{max}}
+\newcommand{\kpc}{\text{kpc}}
+\newcommand{\Gyr}{\text{Gyr}}
+\newcommand{\kms}{\mathrm{km\,s^{-1}}}
+\newcommand{\masyr}{\mathrm{mas\,yr^{-1}}}
+\newcommand{\Mo}{\mathrm{M}_\odot}
 
@@
 \newcommand{\gadget}{{\tt Gadget-4}}
-\newcommand{\smallperi}{{\tt smallperi}}
 \newcommand{\LCDM}{$\Lambda$CDM}
+
+\newcommand{\red}{\color{Thistle}}
+\newcommand{\dbnote}[1]{{\red (DB: #1)}}
+\newcommand{\dbadd}[1]{{\red #1}}
+\newcommand{\dbstrike}[1]{{\red \sout{#1}}}
 
@@
 
+%Ursa Minor has also shown evidence for possible inner substructure, such
+%as stellar or kinematic `clumps'
+%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
+%While one clump was re-detected kinematically in \citet{pace+2014},
+%\citet{munoz+2018} find no evidence of substructure with modern
+%photometry. 
    \author{D. A. Boyea\inst{1}
@@
   % context heading (optional)
-   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density.
-However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
+   {Most dwarf spheroidals (dSph) of the Milky Way have surface density profiles following an exponential law, with a sharply declining outer density.
+However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
   % aims heading (mandatory)
-   {We investigate whether tides may produce these extended light profiles.}
+   {We investigate how tides may affect the outer density structure of Scl and UMi.}
   % methods heading (mandatory)
-   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
+   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way with an optional time-evolving Large Magellanic Cloud (LMC).
+We explore the range of orbits consistent with observed kinematics and the evolution of cored, flattened, and anisotropic models.
+}
   % results heading (mandatory)
    {Tides affect neither galaxy strongly.
-Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little.
-UMi instead loses about 95\% of its dark matter mass, but the stars should appear unaffected.
+Including the LMC reshapes Scl's orbit, reducing the number of pericentres and Scl's tidal evolution.
+With the LMC, Scl's dark matter evolves little and the stellar component is nearly unchanged.
+UMi loses about 95\% of its dark matter mass, but the stars should also appear unaffected and including the LMC only weakens UMi's tidal history.
 }
   % conclusions heading (optional), leave it empty if necessary
-   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }
+   {The outer stellar density excess in Scl and UMi are unlikely of tidal origin. Instead, these features may be innate, possibly explained by past merger or star formation episodes. }
 
@@
 
+%To first order, tidal mass loss peels away the outer layers of a dwarf
+%galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, amorisco2021, drakos+taylor+benson2022, errani+2022}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023} \dbnote{rapha disagrees here}).
+The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.
+%\citep{oh+lin+aarseth1995, piatek+pryor1995}.
+
+%In idealized simulations, most mass loss occurs near pericentre, mass loss removed loosely-bound material first, and tidally stripped stars form unbound \textit{ streams} with a bulk velocity gradient
+%\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Since \citet{faber+lin1983}, exponential surface density profiles have been commonly used to describe dwarf spheroidal (dSph) galaxies \citep[e.g.,][]{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
-However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show clear deviations from an exponential profile (Fig.~\ref{fig:observed_density_profiles}; e.g., \citealt{eskridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003}).
-This excess of stars at large radii was often speculated to be tidal interaction \citep[e.g.][]{innanen+papp1979, walcher+2003} or sometimes a `halo'of stars surrounding the dwarf \citep{westfall+2006}.
-Most recently, \citet{sestito+2023a, sestito+2023b} report a
-`kink' in the density profile of each galaxy. They spectroscopically
-confirm distant members out to ${\sim}10$ half-light
-radii from the centre of each dwarf. If these dwarfs had exponential
-profiles, then these far-outlying stars should be much
-rarer. We aim to test whether tides may form the extended stellar densities of Scl and UMi.
+The outer structure of dwarf galaxies holds potential clues to their
+evolutionary histories. Extended stellar populations may form from tidal
+interactions, episodic star formation, or past dwarf mergers \citep[see
+discussion in Section~\ref{sec:discussion} and, e.g., ][]{penarrubia+2009,
+genina+2019, deason+2022}.
+If a galaxy hosts such an extended stellar population, determining whether the population arises from tides may provide crucial physical insight.
+Tidally disrupting dwarf galaxies can constrain the Milky Way (MW) potential and dark matter properties on small scales \citep[e.g.,][]{penarrubia+2010, cartonzeng+2024}.
+On the other hand, non-tidal extended populations may reveal a rare glimpse into conditions in the early universe or the hierarchical assembly of dwarf galaxies.
 
-Cosmological simulations struggle to resolve tidal effects on dwarfs. Dwarfs are 
+Most dwarf spheroidals (dSph) are well-described by an exponential
+surface density profile \citep[e.g.,][]{binggeli+sandage+tarenghi1984,
+mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016}.
+However, Scl and UMi have a excess of stars at large radii relative to an exponential (see Section~\ref{sec:gaia}; e.g., \citealt{eskridge1988, IH1995,
+martinez-delgado+2001, walcher+2003, palma+2003}). 
+Using \textit{Gaia} proper-motion informed stellar candidates from \citet[see also Section~\ref{sec:gaia}]{jensen+2024}, \citet{sestito+2023a, sestito+2023b} spectroscopically confirmed distant members out to ${\sim}10$ half-light radii ($R_h$) from the centre of each dwarf.
+If these galaxies were described by exponential profiles, fewer than $1/10^6$ stars should fall outside $10\,R_h$.
+While tidal interactions are suspected to produce these distant stars \citep[e.g.][]{innanen+papp1979, walcher+2003, sestito+2023a}, few works have dynamically modelled Scl and UMi specifically.
+
+Cosmological simulations struggle to resolve tidal effects on dwarfs, which are 
 often near the resolution limit and vulnerable to artificial disruption
-\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
-these challenges, idealized simulations model a single subhalo in an
-analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
-From these simulations, tidally stripped stars form \emph{tidal
-streams}---stellar tails with a bulk velocity gradient
-\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
-Most mass loss happens near pericentre, where tides are strongest.
-However, the central structure of a dwarf galaxy often remains
-undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. 
-To first order, tidal mass loss peels away the outer layers of a dwarf
-galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023}).
-The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.
+\citep[e.g.,][]{PNM2008, vandenbosch+2018, EP2020, santos-santos+2025}. To overcome
+these challenges, idealized simulations model a single subhalo\footnote{A `halo' is a gravitationally self-bound overdensity of dark matter.
+A subhalo is a halo gravitationally bound to a larger halo.} in an
+analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, penarrubia+2010, ogiya+2019, EN2021}.
+Several recent studies have modeled specific galaxies and generally find that classical dwarfs experience weak Galactic tides \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024, tchiorniy+genina2025}.
+Most relevantly, \citet{iorio+2019} simulate Scl, also finding little tidal evolution.
+However, no studies since have fully explored tidal effects on Scl and UMi using updated structural and kinematic properties or accounted for orbital and tidal effects from the Large Magellanic Cloud (LMC).
 
-Improved orbital constraints and Milky Way
-potential models have enabled precise dynamical modeling of dwarf galaxies.
-Simulations tuned to Fornax \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024},
-Scl \citep{iorio+2019}, or MW classicals\footnote{We define classicals to be dwarfs with absolute magnitudes between $-7.7$ and $-14$.} in general \citep{tchiorniy+genina2025} often find tides play a minor role in the stellar evolution of dwarf classicals.
-However, modeling tidal evolution under the Large Magellanic Cloud (LMC) has not yet been considered for most classical dSphs.
+Given recent confirmation of an outer excess of stars in Sculptor and
+Ursa Minor, we revisit whether tides may affect the outskirts of Scl and UMi.
+We aim to model tidal effects with galaxy-tailored, idealized N-body simulations in both a Milky Way and evolving MW+LMC potential.
+In Section~\ref{sec:gaia}, we describe our density profile data.
+Section~\ref{sec:methods} describes our numerical N-body methods and orbital integration.
+Then, Section~\ref{sec:results} discusses the results of our tidal simulations.
+Finally, Section~\ref{sec:discussion} summarizes our results and discusses the implications of extended stellar density profiles.
 
-Two radii provide estimates of where tidal effects occur.
-The \textbf{Jacobi radius}, $r_{\rm J}$, approximates the radius where stars
-become unbound in a galaxy.\footnote{ 
-    While strictly valid only for circular orbits, assuming
-\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
-lost during pericentric passages.} $r_{\rm J}$ solves
-\begin{equation}\label{eq:r_jacobi}
-    3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
-\end{equation}
-given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
-and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
-If \(r_J\) is comparable to
-the visible extent of a galaxy, we should expect to find clear signs of
-tidal disturbance. 
-The \textbf{break radius}, as defined in
-\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
-since pericentre, \(\Delta t\), the break radius is
-\begin{equation}\label{eq:r_break}
-    r_{\rm break} = C\,\sigma_{v}\,\Delta t
-\end{equation} 
-where the empirical constant is \(C \approx 0.55\).
-
-Building on this body of work, we will use idealized simulations to
-understand the severity of tidal effects on Scl and UMi.
-
-\section{{\it Gaia} density profiles}
+\section{\textit{Gaia} density profiles} \label{sec:gaia}
 
@@
 Without removing contamination from foreground/background sources, faint
-features may be lost in the noise or be of uncertain association.
-\emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallaxes and proper motions, improving membership by providing key additional information.
+features may be lost in the noise or be of uncertain association. \textit{Gaia} \citep{gaiacollaboration+2023} now provides stellar parallaxes and proper motions, improving membership precision.
 
 Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
-probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
-incorporating proper motion, colour--magnitude diagram, and
+probabilities from \textit{Gaia} data. J+24 used a Bayesian framework
+incorporating proper motion, colour--magnitude, and
 spatial information to determine the probability that a given star
-belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
+belongs to the satellite, $P_\text{sat}$. By accounting for proper motions
 in particular, J+24 produced low contamination samples of candidate
-member stars out to large distances from a dwarf galaxy. J+24 extended
+member stars. J+24 extended
 the algorithm presented in \citet{MV2020a, MV2020b} by additionally
-including a secondary, extended spatial component.\footnotemark{} J+24 detected
-candidate members out to ${\sim}10$ half-light radii from the
-centres of some galaxies (\(R_h\)). Similar recent work includes
+including a secondary, extended spatial component.\footnotemark{} With this addition, J+24 detected
+candidate members out to ${\sim}10$ half-light radii ($R_h$) from the
+centres of some galaxies . Similar recent work includes
 \citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
@@
 We use
-the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
+the $P_\textrm{sat}$ values from the elliptical 2-component runs if a
 galaxy shows evidence for an outer component, the 1-component run
@@
 For our sample, we adopt a minimum probability of
-\(P_{\rm sat} = 0.2\).
-%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
-Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
+$P_\text{sat} = 0.2$.
+%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with $G < 21$.
+Most stars have $P_\text{sat}$ values which are nearly 0 or 1, so the exact
 choice of probability threshold has little effect on the resulting sample. Even
-at the probability threshold of 0.2, the purity remains high when validated
-against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
-J+24).\footnotemark{} Altogether, the J+24 method
-provides a high-quality, low-contamination sample of dwarf galaxy candidate
-member stars, which we will now investigate in further detail.
-
-\footnotetext{This
-  would indicate that the J+24 model probabilities are miss-calibrated.
-  However, most LOS surveys of dwarf galaxies select brighter stars
-  (which have better \emph{Gaia} measurements) and likely members,
-  complicating the interpretation of this purity estimate. However, we find that our conclusions are unchanged when
-limiting samples to only the brightest stars. }
+at the probability threshold of 0.2, the purity of line-of-sight velocity members remains
+high (${\sim}90\%$, J+24), although many surveys used for validation are also
+of suspected member stars. %Altogether, the J+24 method
+%provides a high-quality, low-contamination sample of dwarf galaxy candidate
+%member stars, which we will now investigate in further detail.
+We truncate density profiles once the apparent background of stars with satellite-consistent proper motions and colour-magnitudes dominates the membership density.
 
@@
 profiles of Scl, UMi, and eight other classical dwarf
-galaxies.\footnote{
-    We define classicals to be non-Magellanic dwarfs brighter than
-    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
-    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
-    these dwarfs, we exclude Antlia II, due to the extremely high background,
-    and Sagittarius, which was not included in J+24. 
-}
-All of the classical dwarfs appear to be well
+galaxies. All classical dwarfs are well
 described by an exponential profile in the inner regions.\footnote{
@@
     across all dwarfs.} 
-In the outer regions, however, Scl and Ursa
-Minor deviate and show a clear outer excess over an exponential law
-(solid black line). These galaxies are better fit by a Plummer law
-(dashed black line). The deviation from an exponential grows outwards
+In the outer regions, however, Scl and UMi deviate and show a clear outer excess over an exponential law. The deviation from an exponential grows outwards
 and may reach 2 orders of magnitude at $8 R_h$. 
+Instead, a Plummer law (dashed black line) better describes Scl and UMi. This paper aims to determine whether tides Scl and UMi's density excess.
 
@@
         The density profiles of
-        Sculptor (orange squares), Ursa Minor (red triangles), and other
+        Sculptor (orange squares), Ursa Minor (green triangles), and other
         classical dwarfs (blue lines) compared to 2D exponential (solid black
-        line) and Plummer density profiles (dashed black line). Density
+        line) and Plummer density profiles (dashed black line). For other
+        classical dwarfs, we include Fornax, Leo I, Leo II, Carina, Draco,
+        Canes Venatici I, Sextans I, and Crater II. We excluding Sagittarius and
+        Antlia II due to high foreground/background. Density
         profiles are plotted in terms of the logarithm of elliptical radius
@@
         logarithmic stellar density scaled by the density at half-light
-        radius. The residuals (lower panel) are with respect to the 2D
-        exponential. Sculptor and Ursa Minor have an excess of stars in the
-        outer regions (past \(\log R/R_h \sim 0.3\)) compared with other
-        classical dwarfs.
+        radius. We use bins of 0.05 dex in $\log R$, and  residuals (lower
+        panel) are with respect to the 2D exponential. 
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
+\section{Methods} \label{sec:methods}
 
 \subsection{Milky Way potential and reference frame}\label{milky-way-potential}
+%To convert observed positions and velocities to Galactocentric
+%Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
+%\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
+%the Galactic centre, $y$ is the direction of the motion of the Local
+%Standard of Rest, and $z$ is the direction perpendicular to the
+%Galactic plane. 
 To convert observed positions and velocities to Galactocentric
-Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
-\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
-the Galactic centre, $y$ is the direction of the motion of the Local
+coordinates, we use the Astropy v4 right-handed Cartesian Galactocentric frame
+\citep{astropycollaboration+2022}. In this frame, $x$ is the direction from the sun to the Galactic centre, $y$ is the direction of the motion of the Local
 Standard of Rest, and $z$ is the direction perpendicular to the
-Galactic plane. 
-% The coordinate frame is also right-handed, such that the
-% \(z\)-angular momentum of the sun is negative (since the sun is at
-% \(x<0\)). In this frame, the solar position is
-% \([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
-% \citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
-% velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
-% km/s
-% \citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
-% The uncertainties in this reference frame are typically smaller than the
-% uncertainties on a dwarf galaxy's distance and tangential velocity.
+Galactic plane. The solar position is
+$[-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc$
+\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
+velocity is $\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\ \kms$
+\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
+The uncertainties on this reference frame are typically smaller than the
+uncertainties on a dwarf galaxy's distance and tangential velocity.
 
-For our MW-only models, we adopt the potential from \citet{EP2020}, which is
-an analytic approximation to that proposed by \citet{mcmillan2011}.
+For our MW-only models, we adopt the potential from \citet{EP2020}, 
+a re-parameterization of the \citet{mcmillan2011} Galactic potential.
 The potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
@@
 We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.
-In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\).
-Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
+In terms of these parameters, $\rmax = 43.7\,$kpc and $\vmax = 191\,\kms$.
+Equivalently, the dark matter halo has a virial mass\footnotemark{} $M_{200} = 126.6\times 10^{10}\,\Mo$ and $r_s=20.2\,\kpc$.
 \footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
@@
 %}\end{equation} 
-%where \(a=1.3\,{\rm kpc}\) is the scale radius and
-%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
+%where $a=1.3${\rm kpc}\) is the scale radius and
+%$M=2.1 \times 10^{10}\,\Mo$ is the total mass. The thin and thick
 %disks are represented with the \citet{miyamoto+nagai1975} cylindrical
@@
 %}\end{equation} 
-%where \(a\) is the disc radial scale length, \(b\) is
-%the scale height, and \(M\) is the total mass of the disk. For the thin
-%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
-%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
-%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
+%where $a$ is the disc radial scale length, $b$ is
+%the scale height, and $M$ is the total mass of the disk. For the thin
+%disk, $a=3.944\,$kpc, $b=0.311\,$kpc, and
+%$M=3.944\times10^{10}\,$M$_\odot$. For the thick disk,
+%$a=4.4\,$kpc, $b=0.92\,$kpc, and $M=2\times10^{10}$\)M$_\odot$.
 
@@
 evolving MW and LMC halos, and the MW reflex motion. In their
-simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
-\(M_{\rm 200}= 98.4\times10^{10}\Mo\), and the LMC a NFW halo with
-\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
+simulation, the MW was initially a NFW halo with $r_s=16.5\,$kpc and
+$M_{200}= 98.4\times10^{10}\Mo$, and the LMC a NFW halo with
+$r_s=11.7$ and $M_{200} = 24.6 \times 10^{10} \Mo$. The total
 \texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
@@
 \subsection{Orbits}\label{sec:scl_smallperi}
-
-\begin{table}[t]
-\centering
-\caption{The initial conditions for models presented in this work.
-The observables record the intended final position of the N-body model derived from point-mass orbit quantiles.
-Instead, the initial position and velocity represent the initial conditions for the actual N-body model.
-The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects}
-\label{tbl:orbit_ics}
-\begin{tabular}{llll}
-\hline
-Property & Scl: MW& Scl: MW+LMC & UMi: MW\\
-\hline
-$\vmax / \kms$ & 31 & 25 & 38 \\
-$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
-$R_{h, \rm exp}$ & 0.168 & 0.185 & 0.168 \\
-$R_{h, \rm Plummer}$ & 0.20 & 0.20  & 0.20 \\
-\hline
-distance $/$ kpc & 82.6 & 73.1 & 64.6\\
-$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
-$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
-$v_{\rm los}$ $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
-\hline
-$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
-${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
-${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
-${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
-$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
-$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
-$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
-\hline
-\end{tabular}
-\end{table}
-
-To explore the possible orbits of a dwarf galaxy, we perform a Monte
-Carlo sampling of the present-day observables (assuming Normal distributions
-and the values in Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}). We
-integrate
-each sampled position/velocity back in time for 10 Gyr using \agama{}
-\citep{agama}. Dynamical friction is not expected to impact orbits
-substantially, so we assume a single point-mass particle for the backwards
-integration.
-Fig.~\ref{fig:orbits} shows 100 samples of these orbits in both potentials (MW only and MW+LMC), and the orbits of our N-body models.
-
-To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
-smallest pericentre among all possible orbits integrated backwards for
-\(10\,\Gyr\). We achieve this by selecting the median parameters of all
-orbits with a pericentre less than the 0.0027th quantile pericentre,
-yielding a pericentre of 43 kpc. Given the current observations, it is
-unlikely that either has had a significantly smaller pericentre than
-our selected orbit, which we refer to as the \smallperi{} orbit. We take
-the first apocentre after a look-back time of 10 Gyr as the initial conditions for our models of Scl and UMi,
-noted in Table~\ref{tbl:orbit_ics}.
-For the Scl MW+LMC model, we select, instead, the $6\sigma$ smallest pericentre with respect to the LMC, (by doubling the observed uncertainties). We only integrate this model backwards for $2\,\Gyr$.
-
-We iteratively adjust UMi's initial conditions to better match the
-present-day position. Initially, starting with low-resolution runs, we adjust
-the cylindrical actions of the initial orbit by the final difference in actions
-at the end of orbital evolution. After the initial actions have converged (2
-iterations), we change the initial action angles by the final difference in
-action angles (2 more iterations). This method converges to an orbit agreeing
-with the observed kinematics of UMi. Since Scl's orbit is less
-strongly affected by tides, we do not carry out this correction for Scl.
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/orbits.pdf}
-    \caption{
-        The orbits of Scl (left) and UMi (right) in the heavy MW-only potential (light blue lines) and the MW+LMC
-        potential (light orange lines). Black lines shows our selected N-body models and
-        the green dashed line shows the LMC trajectory. The top panels show orbits in the $y$--$z$ plane and the botom panels show orbits in terms of Galactocentric radius with time.
-}
-
-    \label{fig:orbits}
-\end{figure}
-
-\subsection{Initial conditions}\label{initial-conditions}
-
-We use \agama{} \citep{agama} to generate the initial N-body dark matter
-halo. We assume galaxies are described by a spherical, isotropic 
-NFW dark matter halo (Eq.~\ref{eq:nfw}). The NFW
-profile truncated in the outer regions by
-\begin{equation}\label{eq:trunc_nfw}
-    \rho_{\rm tNFW}(r) = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(r),
-\end{equation}
-where we adopt \(r_t = 20 r_s\).
-
-We select cosmologically-motivated initial dark matter halos.
-We derive total stellar masses from \citet{munoz+2018} magnitudes and \citet{woo+courteau+dekel2008} stellar mass-to-light ratios, finding Scl and UMi have stellar masses of $3\times10^6$ and $7\times10^5\,\Mo$.
-Using the stellar mass-\(\vmax\) relation \citep[from][]{fattahi+2018} and mass-concentration relation from \citet{ludlow+2016}, Scl
-and UMi's halos should have \(\vmax \approx 31 \,\kms\), $\rmax\approx6\,\kpc$ and
-\(\vmax \approx 27\,\kms\), $\rmax\approx5\,\kpc$. 
-The cosmological means predict a low velocity dispersion compared to the observed dispersion.
-To match the observed \(\sigma_\V\) at the end of tidal evolution, we
-choose \(\vmax = 31\,\kms\) and \(\rmax=3.2\,\kpc\) for Scl in the MW-only potential, $\vmax=25\,\kms$ and $\rmax=2.5\,\kpc$ for Scl in the MW+LMC potential, and
-\(\vmax=38\,\kms\) and \(\rmax=4\,\kpc\) for UMi.
-
-\subsection{N-body methods}\label{isolation-runs-and-simulation-parameters}
-
-We use \gadget{}, with a \agama{} external potential plugin, for our N-body simulations. 
-We adopt a softening length of
-\begin{equation}\label{eq:softening_length}
-    h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
-\end{equation} 
-for a halo with \(N\) particles.
-We use $10^7$ particles, and adopt a relative accuracy criterion of $\alpha=0.005$ and integration accuracy of $\eta=0.01$.\footnote{i.e., tree nodes open if $M\,l/r^3 < \alpha |a|$ for node mass $M$, sidelength $l$, distance $r$, and particle acceleration $a$; and particles must take time-steps smaller than $\sqrt{2\,\eta\,h_{\rm grav}}$.}
-
-To ensure that the initial conditions of the simulation are dynamically
-relaxed, we run a halo first in isolation. We use the same isolation run
-for all halos (with $\rmax=6$ and $\vmax=31$, ran for $5\,\Gyr$) and rescale the results to the desired initial halo.
-Our isolation halo is converged down to a radius $r\approx 100\,{\rm pc}$ (see also the convergence radius from \citealt{power+2003}).
-
-Next, we evolve the halo in the Galactic potential. We rescale and shift the
-isolation-relaxed snapshot to our initial conditions (see Table~\ref{tbl:orbit_ics}). 
-We then evolve the full N-body NFW model forward in time in the Galactic
-potential until the present time, when the halo is closest to its present-day
-observed position in the MW
-halo.\footnote{Specifically, the snapshot which best agrees (as measured
-  by \(\chi^2\)) with the intended final position (in
-  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
-  uncertainties. We use \(R_h\) as the uncertainty in \(\alpha\) and
-  \(\delta\).}
-
-We determine the halo centre using a shrinking-spheres centre method inspired by \citet{power+2003}.
-First, we start with an initial centre estimate using all bound
-particles from the previous snapshot. Then, we calculate the distance of
-all particles from the centre, remove particles with a distance beyond
-the 0.975 quantile of the centre, and recalculate the centre of mass.
-The procedure is repeated until the selection radius is less than
-\textasciitilde1kpc or fewer than 0.1\% of particles remain. We then
-remove all unbound particles based on the \gadget{} calculated potential
-of the halo and recalculate the centre. For all future timesteps, we
-consider only particles retained from the previous iteration.
-
-\subsection{Sculptor and Ursa Minor's initial stellar
-components}\label{sec:painting_stars}
-
-\begin{figure}
-    \includegraphics{figures/initial_velocity.pdf}
-    \caption{
-        {\bf Top:} The initial circular velocity profiles in terms of radius
-        for Sculptor and Ursa Minor for the MW-only models with exponential
-        stars. We plot the models after 5\,Gyr of evolution in isolation, which
-        are nearly identical to the initial conditions. The vertical grey line
-        marks the Jacobi radius (Eq.~\ref{eq:r_jacobi})        
-        {\bf Bottom:} The initial velocity dispersion profiles for the exponential stellar
-        component, as compared to our velocity sample (binned in ten equal
-        number bins). See Section~\ref{sec:velocity_sample} for a description
-        of our velocity sample.
-    }
-    \label{fig:initial_halos}
-\end{figure}
-
-We `paint' stars onto dark matter particles using the particle-tagging
-method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
-symmetry. We briefly describe the
-procedure next, but refer interested readers to \citet{EP2020}.
-
-If we
-know the distribution function\footnote{i.e., the phase-space density of
-  particles. Note that since \(f\) is formally defined in 6D phase
-  space, there is an additional `density of states' term in order to
-  calculate the histogram of particles with \({\cal E}\),
-  \(dM/d{\cal E}\) \citep[section 4.4.5,][]{BT1987}.} \(f({\cal E})\),
-then we assign a stellar weight to a given particle with energy
-\({\cal E}\) using \begin{equation}{
-P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
-}\end{equation}
-We use Eddington inversion to find the distribution
-function, \begin{equation}{
-    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
-}
-\end{equation} 
-where $\rho$ is the stellar or DM density, \(\Psi\) is the potential (normalized to vanish at infinity), and
-\({\cal E}\) the binding energy, \({\cal E} = \Psi - 1/2 \V^2\)  
-\citep[see eq. 4-140b in][]{BT1987}. We
-take \(\Psi\) and $\rho$ from the assumed analytic density profiles, calculating \(\rho_\star\) from \(\Sigma_\star\) via the inverse Abel transform.
-
-We consider both an exponential stellar profile, 
-\begin{equation}\label{eq:exponential_law}
-\Sigma_{\rm exp} = \Sigma_0\exp(-\alpha\,R / \alpha\,R_h).
-\end{equation}
-for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
-We also consider a more extended \citep{plummer1911} law component, 
-\begin{equation}{
-\Sigma_{\rm Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2},
-}\end{equation}
-for half-light radius $R_h$. 
-Fig.~\ref{fig:initial_halos} shows the resulting initial conditions for the MW-only models before and after 5 Gyr of isolation evolution.
-
-\section{Results}
-
-\subsection{Orbits of Sculptor and Ursa Minor}
-Fig.~\ref{fig:orbits} shows orbits of Scl and UMi in the MW-only and MW+LMC potentials.
-In the MW-only potential, both galaxies follow relatively well-constrained orbits consistent with long-term MW satellites.
-However, including the LMC dramatically modifies Scl's orbit---the apocentre increases from ${\sim}100$ to ${\sim}220\,\kpc$.
-Scl also passes by the LMC 100 Myr ago, making the LMC a potential tidal stressor.
-In contrast, UMi's orbit only slightly increases in characteristic radius, reducing tidal forces overall.\footnote{Note that our MW-only model has a heavier MW mass than the MW+LMC model, but the qualitative effects on each galaxy's orbit are largely similar when using the same MW model in each case.}
-
-The black lines in \ref{fig:orbits} show our selected orbits (black), representing an extreme low-pericentre orbit.
-We adopt these orbits as the most severe observationally-permissible tidal histories of each galaxy.
-
-\subsection{Tidal effects}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/density_i_f.pdf}
-    \caption{
-        The stellar density of each model before (dotted) and after
-        (solid) tidal evolution as compared to the observed density profile
-        (solid points). The left and righ panels show results for Sculptor and
-        Ursa Minor in the MW-only potential. The top and bottom panels show
-        results for exponential and Plummer initial stellar conditions. We mark
-        the half-light ($R_h$), the Jacobi radius, and break radius with arrows. 
-}
-    \label{fig:stellar_model_density}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{figures/scl_lmc_density_i_f.pdf}
-    \caption{
-        Similar to Fig.~\ref{fig:stellar_model_density} except for Scl in the combined MW and LMC potential.
-The left and right panels show results for exponential and Plummer initial stellar profiles, respectively.
-    }
-    \label{fig:stellar_model_density_lmc}
-\end{figure}
-
-Fig.~\ref{fig:stellar_model_density} shows the initial and final stellar
-densities of each galaxy in the MW-only potential compared to observations. The final profile resembles the initial profile
-over the range of the observed density profile.
-UMi experiences the strongest evolution, with the stars expanding slightly and flattening past the break radius.
-But UMi's stars do not change measurably in shape.
-In addition, the break and Jacobi radii fall outside the observed stellar component, indicating that tidal effects should not be observable.
-
-Since Scl and UMi may have been born with more extended stellar components, we
-also consider Plummer initial stars, approximately matching the observed
-density profiles. The Plummer profiles also remain nearly unchanged during
-tidal evolution. If any tidal disturbances are visible, than observations would
-have to probe more than two orders of magnitude fainter features than our data
-reach. Therefore, tides do not substantively affect the observed stellar components of these models.
-
-Fig.~\ref{fig:stellar_model_density_lmc} shows the tidal effects on Scl in the LMC+MW potential.
-While the break radius coincides with the observed
-excess, the Jacobi radius is too large to observe tidal signatures. In
-addition, with a reduced number of pericentres compared to the MW-only case,
-the stellar profiles evolve even less.
-
-Not shown here, cored, anisotropic, and flattened models
-all appear to evolve similarly over time. Given the observed velocity
-dispersions, alternative dark matter structures still require a similar mass
-within $R_h$ at the present day. Reasonable changes to the dark matter
-structure likely do not change our conclusions.
-
-A key caveat is that the long-term orbits of satellites are highly uncertain.
-Analytic MW potentials neglect unknown details including triaxiality, mass
-evolution, and substructure, causing analytic orbits to deviate significantly
-from true satellite orbits \citep[e.g.,][]{dsouza+bell2022, santistevan+2023}.
-While we cannot constrain the long-term histories of these dwarfs, our results
-indicate that {\it recent tidal effects} are likely not be observable in Scl and
-UMi.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion \& Conclusions}
-We investigated whether tides from the Milky Way or LMC may explain the extended stellar densities of Scl and UMi.
-Using idealized N-body simulation on orbits with extreme pericentres, we find that tides are insufficient to influence the stellar light profile of either galaxy.
-We also find the LMC strongly perturbs Scl's orbit, weakining tides further.
-We conclude the observed density excess in Sculptor and Ursa Minor are unlikely to be tidal in origin.
-
-Scl and UMi display other peculiarities that may hold clues to their formation. Both galaxies host at least two distinct chemodynamic populations
-\citep{tolstoy+2004, battaglia+2008, pace+2020}.\footnotemark{} The inner population is
-younger, higher metallicity, and dynamically colder, whereas the outer
-population is older, lower metallicity, and dynamically hotter.
-Fig.~\ref{fig:metallicity_gradient} compares a double-exponential density fit to the observed metallicity and velocity dispersion profiles.
-A double-exponential populations better explains the observed profiles than a single-exponential.
-Our two populations are similar to \citet[MMT]{pace+2020}.
-Though our second population is more extended than in \citet{arroyo-polonio+2024}, their third population may be contained within our outer component.
-
-\footnotetext{Other examples of galaxies with multiple populations
-  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
-  Fornax
-  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
-  Sextans
-  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
-  Andromeda II
-  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
-
-\begin{figure}
-    \includegraphics{figures/scl_umi_fe_h_gradient.pdf}
-
-    \caption{
-        {\bf Top:} Density profiles of Sculptor (top left) and Ursa Minor (top
-        right) with double exponential fits (see
-        appendix~\ref{sec:double_exp_fits}). Blue, pink, and red lines
-        represent the inner, outer, and combined density, respectively.
-        {\bf Middle:} The metallicities of member stars in both galaxies with
-        radius. For Sculptor, we plot stars from
-        APOGEE (green circles), \citet[][blue squares]{tolstoy+2023}, and
-        \citet[][orange stars]{sestito+2023a}. For Ursa Minor, we
-        plot stars from APOGEE (green circles), \citet[blue
-        squares]{pace+2020}, and \citet[red stars]{sestito+2023b}. Black points
-        represent median bins, and the pink line
-        represent a simple mixture model of metallicity, transitioning from
-        higher to lower metallicity based on the relative inner and outer
-        component densities (see Table~\ref{tbl:double_exp_fits}). 
-        {\bf Bottom:} The velocity dispersion profiles of our radial velocity
-        sample (black points) as compared to the velocity dispersion profile
-        for exponential profiles. We calculate the velocity dispersion of the
-        model through the Jean's equations assuming a NFW dark halo of
-        ($\rmax=2.5\,\kpc$, $\vmax=25\,\kms$) for Scl and ($\rmax=5\,\kpc$, $\vmax=27\,\kms$) for UMi.
-    }
-    \label{fig:metallicity_gradient}
-\end{figure}
-
-%Ursa Minor has also shown evidence for possible inner substructure, such
-%as stellar or kinematic `clumps'
-%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
-%While one clump was re-detected kinematically in \citet{pace+2014},
-%\citet{munoz+2018} find no evidence of substructure with modern
-%photometry. 
-  We next consider a variety of possible origins.
-
-\textbf{Episodic star formation.} Star formation may quench and
-reignite, creating successive stellar generations with differing
-distributions. External star formation triggers include tidal
-compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
-gaseous filaments or mergers \citep{genina+2019}, perturbations from dark halos
-\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
-\citep{wright+2019}.\footnote{More common mechanisms, like feedback or
-reionization-driven quenching, may also form multiple stellar
-generations
-\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
-such processes would not explain why extended stellar populations appear
-to be non-universal.} In this case, the star formation history should
-contain evidence of a corresponding burst.
-
-\textbf{Major mergers.} In a dwarf merger, stars from
-the lower-mass galaxy are preferentially dispersed, forming an extended
-stellar component and population gradient
-\citep{benitez-llambay+2016, deason+2022}. A halo may also form of many pre-reionization quenched ultra-faint dwarf galaxies \citep{ricotti+polisensky+cleland2022}. A few local dwarfs are
-suspected to have undergone a major merger, including Tucana II,
-Andromeda II, and Phoenix
-\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
-A galaxy having undergone a major merger should harbour at least two or
-three populations with distinct origin.
-
-\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
-`preprocessed' by a massive satellite like the LMC
-\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
-redistributes already-present stellar populations and may mimic a
-stellar halo. Key evidence suggestive of preprocessing may include
-distant dwarf stars or stellar streams.
-
-\textbf{Dark-matter-free dwarfs.} Dark-matter-free tidal dwarfs 
-%(formed in gas-rich tidal streams during the merger of two larger galaxies, e.g., \citealt{mirabel+dottori+lutz1992, bournaud+duc2006}) 
-and galaxies in Modified Newtonian Dynamics (MOND) would be more susceptible to tides (\citealt{brada+milgrom2000, mcgaugh+wolf2010, casas+2012, yang+2014, wang+2024a}; but see also \citealt[][]{sanchez-salcedo+hernandez2007}).
-In these cases, the high observed velocity dispersion requires evidence of strong tidal disruption, such as velocity gradients or tidal tails
-\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.
-
-\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
-dynamically hotter than younger stars. Dynamical heating processes including
-stellar feedback \citep{stinson+2009, maxwell+2012, el-badry+2016,
-mercado+2021}, sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy
-dark matter interference fringes \citep[e.g.,][]{el-zant+2020,
-duttachowdhury+2023}. However, most of these processes should operate similarly
-across dwarf galaxies, not just Scl and UMi.
-
-While a number of scenarios may form an extended outer stellar density, future
-observations have promise to discern different scenarios. 
-{Precise chemistry}, comparing the inner and
-outer regions, can test if the was accreted and could constrain properties
-of merged satellites. 
-{Detailed kinematics} could constrain the halo shape (ellipticity and anisotropy) and
-if the dwarf shows signs of tidal disruption.
-{Deep photometry} may find or rule out signs of dynamical
-disequilibrium and tidal tails for tidally susceptible models and constrain the prevalence and nature of extended features in dwarf galaxies.
-Finally, {star formation histories} may differentiate scenarios that rely on or create strong star formation bursts.
-Both upcoming observatories and novel simulation methods will be instrumental to uncovering the formation histories of Sculptor and Ursa Minor.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgements}
-\end{acknowledgements}
-
-\bibliographystyle{bibtex/aa.bst} % style aa.bst
-\bibliography{paper.bib} % your references Yourfile.bib
-
-\appendix
-
-\section{Additional model results}
-In this section, we present tables of the resulting properties of the dark matter and stellar components for each model.
-
-Tables~\ref{tbl:scl_orbit_props}, \ref{tbl:umi_orbit_props} contain information about the orbital properties of the models and our Monte carlo sampled orbits.
-Table~\ref{tbl:sim_results} contains the initial and final properties of the dark matter and stars for each simulation presented in the main text.
 
@@
 $\mu_{\alpha*} / \masyr$  & $0.099 \pm 0.002 \pm 0.017$ & 3\\
-$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$  & 3\\
-$\V_{\rm los}$ / ${\rm km\,s^{-1}}$ & $111.2 \pm 0.3\ $ & 4\\
-$\sigma_v$ / ${\rm km\,s^{-1}}$ & $9.7\pm0.2\ $ & 4\\
+$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_\text{stat} \pm 0.017_\text{sys}$  & 3\\
+$\V_\text{los} / \kms$ & $111.2 \pm 0.3\ $ & 4\\
+$\sigma_v / \kms$ & $9.7\pm0.2\ $ & 4\\
 $R_h/'$ & $9.79 \pm 0.04$ & 1\\
@@
 $\mu_\alpha*$ / mas yr$^{-1}$  & $-0.124 \pm 0.004 \pm 0.017$ & 3\\
-$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ & 3\\
-$\V_{\rm los}$ / km s$^{-1}$  &$-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ & 4\\
+$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_\text{stat} \pm 0.017_\text{sys}$ & 3\\
+$\V_\text{los}$ / km s$^{-1}$  &$-245.9 \pm 0.3_\text{stat} \pm 1_\text{sys}$ & 4\\
 $\sigma_v$ & $8.6 \pm 0.3$ & 4\\
@@
 \end{table}
+
+\begin{table}[t]
+\centering
+\caption{The initial conditions for models presented in this work.
+The observables record the intended final position of the N-body model derived from point-mass orbit quantiles.
+Instead, the initial position and velocity represent the initial conditions for the actual N-body model.
+}
+\label{tbl:orbit_ics}
+\begin{tabular}{llll}
+\hline
+Property & Scl: MW& Scl: MW+LMC & UMi: MW\\
+\hline
+$\vmax / \kms$ & 31 & 25 & 38 \\
+$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
+$R_{h, \rm exp}$ & 0.168 & 0.185 & 0.168 \\
+%$R_{h, \rm outer}$ & 0.20 & 0.20  & 0.20 \\
+\hline
+distance $/$ kpc & 82.6 & 73.1 & 64.6\\
+$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
+$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
+$v_\text{los}$ $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
+\hline
+$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
+${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
+${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
+${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
+$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
+$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
+$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
+\hline
+\end{tabular}
+\end{table}
+
+To explore the possible orbits of a dwarf galaxy, we perform a Monte Carlo
+sampling of the present-day observables.
+From the position and velocities and their uncertainties in Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}) for Scl and UMi, we draw coordinates assuming Normal distributions and not accounting for covariance.
+We then integrate each sampled position/velocity back in time for 10 Gyr using \agama{}
+\citep{agama}. Dynamical friction is not expected to impact orbits
+substantially, so we assume a single point-mass particle for the backwards
+integration.
+
+To maximize tidal effects, we select an orbit with the $\sim 3\sigma$
+smallest pericentre among all possible orbits integrated backwards for
+$10\,\Gyr$. We achieve this by selecting the median parameters of all
+orbits with a pericentre less than the 0.0027th quantile pericentre,
+yielding a pericentre of 43 kpc. Given the current observations, it is
+unlikely that either has had a significantly smaller pericentre than
+our selected orbit. We take
+the first apocentre after a look-back time of 10 Gyr as the initial conditions for our models of Scl and UMi,
+noted in Table~\ref{tbl:orbit_ics}.
+For the Scl MW+LMC model, we select instead the $6\sigma$ smallest pericentre with respect to the LMC (by doubling the observed uncertainties) because we do not account for uncertainties of the LMC trajectory.
+We only integrate this model backwards for $2\,\Gyr$.
+
+UMi's N-body model deviates from a point particle orbit due to dynamical self-friction \citep[e.g.,][]{white1983, miller+2020}.
+To better match the present-day position and velocity, we iteratively adjust UMi's initial conditions.
+Initially, starting with low-resolution runs, we shift
+the cylindrical actions of the initial orbit by the final difference in actions
+after tidal evolution. Once the initial actions have converged (2
+iterations), we change the initial action angles by the final difference in
+action angles (2 more iterations). This method converges to an orbit agreeing
+with the observed kinematics of UMi. Since Scl's orbit is less
+strongly affected by tides, we do not carry out this correction for Scl.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/orbits.pdf}
+    \caption{
+        The orbits of Scl (left) and UMi (right) in the heavy MW-only potential (light blue lines) and the MW+LMC
+        potential (light orange lines). Black lines shows our selected N-body models and
+        the green dashed line shows the LMC trajectory. The top panels show orbits in the $y$--$z$ plane and the bottom panels show orbits in terms of Galactocentric radius with time.
+}
+
+    \label{fig:orbits}
+\end{figure}
+
+\subsection{Initial conditions}\label{initial-conditions}
+
+We use \agama{} \citep{agama} to generate the initial N-body dark matter
+halo. We assume galaxies are described by a spherical, isotropic 
+NFW dark matter halo (Eq.~\ref{eq:nfw}). We truncate the NFW profile with
+\begin{equation}\label{eq:trunc_nfw}
+    \rho_\text{tNFW}(r) = e^{-(r/r_t)^3}\ \rho_\text{NFW}(r),
+\end{equation}
+where we set the truncation radius $r_t = 20 r_s$.
+
+We select cosmologically-motivated initial dark matter halos.
+We derive total stellar masses from \citet{munoz+2018} magnitudes and \citet{woo+courteau+dekel2008} stellar mass-to-light ratios, finding Scl and UMi have stellar masses of $3\times10^6$ and $7\times10^5\,\Mo$.
+Using the stellar mass--$\vmax$ relation from \citet{fattahi+2018} and mass-concentration relation from \citet{ludlow+2016}, Scl
+and UMi's halos should have $\vmax \approx 31 \,\kms$, $\rmax\approx6\,\kpc$ and
+$\vmax \approx 27\,\kms$, $\rmax\approx5\,\kpc$, respectively. 
+The cosmological means predict too low a velocity dispersion compared to the observed value.
+To better match the observed velocity dispersion after tidal evolution, we
+choose $\vmax = 31\,\kms$ and $\rmax=3.2\,\kpc$ for Scl in the MW-only potential, $\vmax=25\,\kms$ and $\rmax=2.5\,\kpc$ for Scl in the MW+LMC potential, and
+$\vmax=38\,\kms$ and $\rmax=4\,\kpc$ for UMi. Our initial dark matter halos are within the 2--3$\sigma$ scatter of the cosmological trends in \citet{ludlow+2016, fattahi+2018}.
+
+\subsection{N-body methods}\label{isolation-runs-and-simulation-parameters}
+
+We use \gadget{}, with a \agama{} external potential plugin, for our N-body simulations. 
+We adopt a softening length of
+\begin{equation}\label{eq:softening_length}
+    h_\text{grav} = 0.014\,\text{kpc}\left(\frac{r_\text{max}}{6.0\,\text{kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2},
+\end{equation} 
+for a halo with $N$ particles. While our softening choice is about 3 times smaller than \citep{power+2003}, we find this choice best balances convergence and computational efficiency.
+We use $10^7$ particles, a relative accuracy criterion of $\alpha=0.005$, and an integration accuracy of $\eta=0.01$.\footnote{i.e., tree nodes open if $M\,l/r^3 < \alpha |a|$ for node mass $M$, side-length $l$, distance $r$, and particle acceleration $a$; and particles must take time-steps smaller than $\sqrt{2\,\eta\,h_\text{grav}}$.}
+
+To ensure that the initial conditions of the simulation are dynamically
+relaxed, we run a halo first in isolation. We use the same isolation run
+for all models and rescale the results to the desired initial halo. Our isolation halo was run for 5 Gyr assuming $\rmax=6\,\kpc$ and $\vmax=31\,\kms$ based on Scl's cosmological-mean halo.
+This model is converged down to a radius $r\approx 150\,\text{pc}$, where the enclosed mass changes by less than 2\% over $5\,\Gyr$ (see also the convergence radius from \citealt{power+2003}).
+
+Next, we evolve the halo in the Galactic potential. We rescale and shift the
+isolation-relaxed halo to our initial conditions (see Table~\ref{tbl:orbit_ics}). 
+We then evolve the full N-body NFW model in the Galactic
+potential until the present time, when the halo is closest to its present-day
+observed position in the MW
+halo.\footnote{Specifically, the snapshot which best agrees (as measured
+  by $\chi^2$) with the intended final position (in
+  Table~\ref{tbl:orbit_ics}) and the observed position and velocity
+  uncertainties. We use $R_h$ as the uncertainty in $\alpha$ and
+  $\delta$.}
+
+We determine the halo centre using a shrinking-spheres centre method inspired by \citet{power+2003}.
+First, we start with an initial centre estimate based on the previous centre and acceleration of the last snapshot. Then, we calculate the distance of
+all particles from the centre, remove particles with a distance beyond
+the 0.95 quantile of the centre, and recalculate the centre of mass.
+The procedure is repeated until the selection radius is less than
+1\,kpc. We then
+remove all unbound particles based on the \gadget{} calculated potential
+of the halo and recalculate the centre. For all future time-steps, we
+consider only particles retained from the previous iteration.
+
+\subsection{Sculptor and Ursa Minor's initial stellar
+components}\label{sec:painting_stars}
+
+We `paint' stars onto dark matter particles using the particle-tagging
+method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
+symmetry. We briefly describe the
+procedure next, but refer interested readers to \citet{EP2020}.
+
+If we
+know the distribution function\footnote{i.e., the phase-space density of
+  particles. Note that since $f$ is formally defined in 6D phase
+  space, there is an additional `density of states' term in order to
+  calculate the histogram of particles with ${\cal E}$,
+  $dM/d{\cal E}$ \citep[section 4.4.5,][]{BT1987}.} $f({\cal E})$,
+then we assign a stellar weight to a given particle with energy
+${\cal E}$ using \begin{equation}{
+P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_\text{halo}({\cal E})}.
+}\end{equation}
+We use Eddington inversion to find the distribution
+function, \begin{equation}{
+    f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right)
+}
+\end{equation} 
+where $\rho$ is the stellar or dark matter density, $\Psi$ is the negative relative potential (normalized to vanish at infinity), and
+${\cal E}$ the binding energy, ${\cal E} = \Psi - 1/2 \V^2$  
+\citep[see eq. 4-140b in][]{BT1987}. We
+take $\Psi$ and $\rho$ from the assumed analytic density profiles, calculating $\rho_\star$ from $\Sigma_\star$ via the inverse Abel transform.
+
+We consider initially a 2D-exponential stellar profile, 
+\begin{equation}\label{eq:exponential_law}
+\Sigma_\text{exp} = \Sigma_0\exp(-\alpha\,R / \alpha\,R_h),
+\end{equation}
+for central surface density $\Sigma_0$, half-light radius $R_h$, and $\alpha \approx 1.6783$.
+We also consider a more extended \citep{plummer1911} law component, 
+\begin{equation}
+\Sigma_\text{Pl} = \frac{\Sigma_0}{(1 + (R/R_h)^2)^2}.
+\end{equation}
+
+Fig.~\ref{fig:initial_halos} shows the initial circular velocity profiles of the dark matter and stars after evolution in isolation for both galaxies.
+Not shown, the profiles are nearly unchanged throughout this evolution.
+In addition, the stellar component is about $10\times$ more compact than the dark matter.
+The mass is dominated by dark matter across the galaxy---assuming stars do not contribute to the potential is reasonable.
+
+\begin{figure}
+    \includegraphics{figures/initial_velocity_nosigma.pdf}
+    \caption{
+       The initial circular velocity profiles in terms of radius
+        for Sculptor and Ursa Minor for the MW-only models with exponential
+        stars. We plot the models after 5\,Gyr of evolution in isolation, which
+        are nearly identical to the initial conditions. The vertical grey line
+        marks the Jacobi radius (Eq.~\ref{eq:r_jacobi}), and the green dot marks
+        the observed velocity dispersion and half-light radius.        
+    }
+    \label{fig:initial_halos}
+\end{figure}
+
+\section{Results} \label{sec:results}
+
+\subsection{Orbits of Sculptor and Ursa Minor}
+Fig.~\ref{fig:orbits} shows orbits of Scl and UMi in the MW-only and MW+LMC potentials.
+In the MW-only potential, both galaxies follow relatively well-constrained orbits consistent with long-term MW satellites.
+Scl experiences three pericentres over the past 5 Gyr of $53\pm 3\,\kpc$.
+Over the same time, UMi experiences four smaller pericentres of $37\pm3\,\kpc$.
+However, including the LMC dramatically modifies Scl's orbit---the apocentre increases from ${\sim}100$ to ${\sim}220\,\kpc$.
+Scl also passes by the LMC 100 Myr ago, making the LMC a potential tidal stressor.
+In contrast, UMi's orbit only increases in pericentre to $\approx 45\,\kpc$ when including the LMC, reducing tidal forces overall.\footnote{Note that our MW-only model has a heavier MW mass than the MW+LMC model, but the qualitative effects on each galaxy's orbit are largely similar when using the same MW model in each case.}
+
+The black lines in Fig.~\ref{fig:orbits} show our selected N-body model orbits for Scl and UMi. We adopt these orbits as representations of an extreme observationally-consistent tidal history.
+
+\subsection{Tidal effects}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=9cm]{figures/umi_zoom_image.png}
+    \caption{
+        The final dark matter and stellar distribution for the Ursa Minor MW only model in the $y$--$z$ (near-orbital) plane.
+Dark matter is in purple with stars in white, each colour-range spanning 5 orders of magnitude from the peak value.
+The grey dotted line shows the past orbit and the insets zoom into the grey box in the centre.
+    }
+    \label{fig:umi_sim_image}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/density_i_f.pdf}
+    \caption{
+        The stellar density of each model before (dotted) and after
+        (solid) tidal evolution as compared to the observed density profile
+        (solid points). The top panels show 2D exponential initial stellar
+        profiles and the bottom show Plummer . We mark the
+        half-light ($R_h$), the Jacobi radius, and break
+        radius with arrows. 
+}
+    \label{fig:stellar_model_density}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/scl_lmc_density_i_f.pdf}
+    \caption{
+        Similar to Fig.~\ref{fig:stellar_model_density} except for Scl in the combined MW and LMC potential.
+The left and right panels show results for exponential and Plummer initial stellar profiles, respectively.
+    }
+    \label{fig:stellar_model_density_lmc}
+\end{figure}
+
+UMi loses the most total mass, with a smaller pericentre than Scl.
+Fig.~\ref{fig:umi_sim_image} shows the final distribution of dark matter and stars for Ursa Minor.
+More than 95\% of UMi's dark matter mass is lost.
+However, the stellar component remains almost entirely bound and spherical, only expanding slightly over time due to the reduction in the galaxy's total mass.
+
+Fig.~\ref{fig:stellar_model_density} shows the initial and final stellar
+densities of each galaxy in the MW-only potential compared to observations. The final profile resembles the initial profile
+over the range of the observed density profile.
+UMi experiences the strongest evolution, with the stars expanding slightly and flattening past 100 arcminutes.
+In either case, such density perturbations are unlikely to be observable in the near future.
+
+Since Scl and UMi may have been born with more extended stellar components, we
+also consider Plummer initial stars, approximately matching the observed
+density profiles. The Plummer profiles also remain nearly unchanged during
+tidal evolution (see Fig.~\ref{fig:stellar_model_density}). If any tidal disturbances are visible, than observations would
+have to probe more than two orders of magnitude fainter features than our data
+reach.
+
+Two radii provide estimates of where tidal effects occur.
+The \textbf{Jacobi radius}, $r_\text{J}$, approximates the radius where stars
+become unbound in a galaxy.\footnote{ 
+    While strictly valid only for circular orbits, assuming
+$r_\text{peri}$ for the host-dwarf distance works as most stars are
+lost during pericentric passages.} $r_\text{J}$ solves
+\begin{equation}\label{eq:r_jacobi}
+    3\bar \rho_\text{MW}(r_\text{peri}) \approx \bar \rho_\text{dwarf}(r_J),
+\end{equation}
+given the mean enclosed  density of the host at pericentre $\bar \rho_\text{MW}(r_\text{peri})$,
+and the mean enclosed satellite density $\bar \rho_\text{dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
+If $r_J$ is comparable to
+the visible extent of a galaxy, we should expect to find clear signs of
+tidal disturbance. 
+The \textbf{break radius}, as defined in
+\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, $\sigma_v$, and time elapsed
+since pericentre, $\Delta t$, the break radius is
+\begin{equation}\label{eq:r_break}
+    r_\text{break} = C\,\sigma_{v}\,\Delta t
+\end{equation} 
+where the empirical constant is $C \approx 0.55$.
+
+Based on both the Jacobi and break radii as marked in Fig.~\ref{fig:stellar_model_density}, tidal effects should not be apparent until past 100 arcminutes in either galaxy.
+This radius falls outside the range of the observed density profile, and are typically larger for less extreme orbits.
+Given these estimates, it would be unexpected to observe any tidal effects without substantially deeper data.
+Therefore, tides do not substantively affect the observed stellar components of these models.
+
+Fig.~\ref{fig:stellar_model_density_lmc} shows the tidal effects on Scl in the LMC+MW potential.
+While the break radius coincides with the observed
+excess, the Jacobi radius is too large to observe tidal signatures. In
+addition, with a reduced number of pericentres compared to the MW-only case,
+the stellar profiles evolve even less.
+Our selected LMC model is already fairly massive $M_{200} \approx 3\times10^{11}\,\Mo$ -- exploring other LMC orbit models (not shown) also supports our conclusion.
+We conclude that accounting for the LMC potential only weakens recent tidal forces on Scl.
+
+In Appendix~\ref{sec:extra_models}, we test cored, anisotropic, and flattened models, finding that these halos evolve similarly over time.
+Given the observed velocity dispersions, alternative dark matter structures still require a similar mass within $R_h$ at the present day.
+Reasonable changes to the dark matter
+structure likely do not change our conclusions.
+
+A key caveat is that the long-term orbits of satellites are highly uncertain.
+Analytic MW potentials neglect unknown details including triaxiality, mass
+evolution, and substructure, causing orbits in analytic potentials to deviate
+significantly from true satellite orbits in cosmological simulations
+\citep[e.g.,][]{dsouza+bell2022, santistevan+2023}.
+Indeed, UMi may have even been bound to the LMC more than $5\,\Gyr$ ago \citep{vasiliev2024}. 
+While we cannot constrain the long-term histories of these dwarfs, our results
+indicate that \textit{recent tidal effects} are likely not be observable in Scl and UMi.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion \& Conclusions} \label{sec:discussion}
+We investigated whether tides from the Milky Way or LMC may explain the extended stellar densities of Scl and UMi.
+Using idealized N-body simulation on orbits with extreme pericentres, we find that tides are insufficient to influence the stellar light profile of either galaxy.
+We also find the LMC strongly perturbs Scl's orbit, but accounting for the LMC only weakens the total tidal forces on Scl and UMi.
+We conclude the observed density excess in Sculptor and Ursa Minor are unlikely to be tidal in origin.
+
+Scl and UMi display other peculiarities that may hold clues to their formation. Both galaxies host at least two distinct chemodynamic populations
+\citep{tolstoy+2004, battaglia+2008, pace+2020}.\footnotemark{} The inner population is
+younger, higher metallicity, and dynamically colder, whereas the outer
+population is older, lower metallicity, and dynamically hotter.
+Such distinct populations could indicate past accretion events or multiple generations of star formation, processes which may also create extended outer stellar densities.
+
+Whether Scl and UMi's have multiple populations corresponding to the observed density excess is unclear with present data.
+Fig.~\ref{fig:metallicity_gradient} compares the best fitting double-exponential profile (from Appendix \ref{sec:double_exp_fits}) to the observed density, velocity dispersion, and metallicity profiles.
+We model the metallicity profile as a mixture of two single-metallicity populations corresponding to each exponential component, and the velocity profile results from the density profile embedded in a NFW dark matter halo.
+With two components, the model approximately matches the metallicity gradient, and better reproduces the observed flat or rising velocity dispersion profile. 
+Our two populations in UMi are similar in radii to \citet[MMT]{pace+2020}.
+For Scl, though our second population is more extended than in \citet{arroyo-polonio+2024}, their third, outermost population may partially comprise our outer component.
+
+\footnotetext{Other examples of galaxies with multiple populations
+  include Carina \citep{battaglia+2012, fabrizio+2016, kordopatis+2016},
+  Fornax
+  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
+  Sextans
+  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
+  Andromeda II
+  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
+
+\begin{figure}
+    \includegraphics{figures/scl_umi_fe_h_gradient.pdf}
+
+    \caption{
+        \textbf{Top:} Density profiles of Sculptor (top left) and Ursa Minor (top
+        right) with double exponential fits (see
+        appendix~\ref{sec:double_exp_fits}). Blue, pink, and red lines
+        represent the inner, outer, and combined density, respectively.
+        \textbf{Middle} The metallicities of member stars in both galaxies with
+        radius. For Sculptor, we plot stars from
+        APOGEE (green circles), \citet[][blue squares]{tolstoy+2023}, and
+        \citet[][orange stars]{sestito+2023a}. For Ursa Minor, we
+        plot stars from APOGEE (green circles), \citet[blue
+        squares]{pace+2020}, and \citet[red stars]{sestito+2023b}. Black points
+        represent median bins, and the pink line
+        represent a simple mixture model of metallicity, transitioning from
+        higher to lower metallicity based on the relative inner and outer
+        component densities (see Table~\ref{tbl:double_exp_fits}). 
+        \textbf{Bottom:} The velocity dispersion profiles of our radial velocity
+        sample (black points) as compared to the velocity dispersion profile
+        for exponential profiles. We calculate the velocity dispersion of the
+        model through the Jean's equations assuming a NFW dark halo of
+        ($\rmax=2.5\,\kpc$, $\vmax=25\,\kms$) for Scl and ($\rmax=5\,\kpc$, $\vmax=27\,\kms$) for UMi.
+    }
+    \label{fig:metallicity_gradient}
+\end{figure}
+
+  We next consider other possible origins for the outer density strucutre of Scl and UMi.
+
+\textbf{Episodic star formation.} Star formation may quench and
+reignite, creating successive stellar generations with differing
+distributions. External star formation triggers include tidal
+compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
+gaseous filaments or mergers \citep{genina+2019}, perturbations from dark halos
+\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
+\citep{wright+2019}.\footnote{More common mechanisms, like feedback or
+reionization-driven quenching, may also form multiple stellar
+generations
+\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
+such processes would not explain why extended stellar populations appear
+to be non-universal.} In this case, the star formation history should
+contain evidence of a corresponding burst.
+
+\textbf{Major mergers.} In a dwarf merger, stars from
+the lower-mass galaxy are preferentially dispersed, forming an extended
+stellar component and population gradient
+\citep{benitez-llambay+2016, deason+2022}. A halo may also form from many accreted ultra-faint dwarf galaxies \citep{ricotti+polisensky+cleland2022}. A few local dwarfs are
+suspected to have undergone a major merger, including Tucana II,
+Andromeda II, and Phoenix
+\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
+If a galaxy has accreted other systems, it should harbour multiple stellar populations with distinct origin.
+
+\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
+`preprocessed' by a massive satellite like the LMC
+\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
+redistributes already-present stellar populations and may mimic a
+stellar halo. Key evidence suggestive of preprocessing may include
+distant dwarf stars or stellar streams.
+
+\textbf{Dark-matter-free dwarfs.} Dark-matter-free tidal dwarfs 
+%(formed in gas-rich tidal streams during the merger of two larger galaxies, e.g., \citealt{mirabel+dottori+lutz1992, bournaud+duc2006}) 
+and galaxies in Modified Newtonian Dynamics (MOND) would be more susceptible to tides (\citealt{brada+milgrom2000, mcgaugh+wolf2010, casas+2012, yang+2014, wang+2024a}; but see also \citealt[][]{sanchez-salcedo+hernandez2007}).
+However, the high observed velocity dispersion of many dSph requires active tidal disruption, which should be evident as velocity gradients or tidal tails
+\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.
+
+\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
+dynamically hotter than younger stars. Dynamical heating processes including
+stellar feedback \citep{stinson+2009, maxwell+2012, el-badry+2016,
+mercado+2021}, sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy
+dark matter interference fringes \citep[e.g.,][]{el-zant+2020,
+duttachowdhury+2023}. However, most of these processes should operate similarly
+across dwarf galaxies, failing to explain why Scl and UMi have unusual density profiles.
+
+While a number of scenarios may form an extended outer stellar density, future
+observations have promise to discern different scenarios. 
+{Precise chemistry} can test if an extended stellar population was accreted and could constrain properties of merged satellites or the star formation history. 
+{Detailed kinematics} could constrain the halo shape (ellipticity and anisotropy) and
+if the dwarf shows signs of tidal disruption.
+{Deep photometry} may find or rule out signs of dynamical
+disequilibrium and tidal tails for tidally susceptible models and constrain the prevalence and nature of extended features in dwarf galaxies.
+Finally, {star formation histories} may differentiate scenarios that rely on or create strong star formation bursts.
+Both upcoming observatories and novel simulation methods will be instrumental to uncovering the formation histories of Sculptor and Ursa Minor.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{acknowledgements}
+\end{acknowledgements}
+
+\bibliographystyle{bibtex/aa.bst} % style aa.bst
+\bibliography{paper.bib} % your references Yourfile.bib
+
+\appendix
+
+\section{Additional model results} \label{sec:extra_models}
+In this section, we discuss additional details of each model, and models not shown in the main paper.
+
+Tables~\ref{tbl:scl_orbit_props}, \ref{tbl:umi_orbit_props} contain information about the orbital properties of the models and our Monte carlo sampled orbits.
+Table~\ref{tbl:sim_results} contains the initial and final properties of the dark matter and stars for each simulation presented in the main text.
+
+Not shown in the main text, Fig.~\ref{fig:sim_velocity_profiles} shows the velocity dispersion profiles for the MW-only Scl and UMi models.
+While there is an increasing velocity begining near the break radius (due to unbound stars), the velocity dispersion profiles are otherwise similar to the predicted profiles shown in Fig.~\ref{fig:metallicity_gradient} but noisier.
+As such, the internal kinematics of a multiple population galaxy are likely unaffected by tidal evolution.
+
+We additionally explore a cored, anisotropic, and flattened model for the intermediate Scl model in the MW-only potential.
+
+Fig.~\ref{fig:extra_tidal_tracks}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/velocity_dispersion_profiles.pdf}
+    \caption{The velocity dispersion profiles of Scl and UMi models in the MW-only potential model}
+    \label{fig:sim_velocity_profiles}
+\end{figure}
 
@@
 $\V_\textrm{max, f} / \V_\textrm{max, i}$ &  0.695 & 0.928 & 0.511 \\
-$f_{\rm lost, DM}$ & 0.917 & 0.451 & 0.965 \\
+$f_\text{lost, DM}$ & 0.917 & 0.451 & 0.965 \\
 \hline
@@
 $\sigma_{h, \rm exp, f}/\kms$ & 8.8 & 8.8 & 8.8 \\
-$f_{\rm lost, exp}$ & $2.1\times 10^{-6}$ & $5\times10^{-13}$ & 0.0014\\
+$f_\text{lost, exp}$ & $2.1\times 10^{-6}$ & $5\times10^{-13}$ & 0.0014\\
 \hline
@@
 $\sigma_{h, \rm Plummer, f}/\kms$ & 9.4 & 9.2 & 9.1\\
-$f_{\rm lost, Plummer}$ & $0.024$ & 0.0013 & 0.068\\
+$f_\text{lost, Plummer}$ & $0.024$ & 0.0013 & 0.068\\
 \hline
@@
 
-% Fig.~\ref{fig:umi_sim_images} shows the dark matter evolution of our UMi
-% model, the most tidally affected model. The dark matter creates large wakes and
-% streams encircling the Galaxy several times, but the inner dark matter
-% structure remains intact (not shown).
+\begin{figure}
+    \includegraphics{figures/extra_tidal_tracks.pdf}
 
-%\begin{figure}
-%    \centering
-%    \includegraphics[width=\columnwidth]{figures/umi_sim_images.png}
-%    \caption{
-%        Images of Ursa Minor's dark matter over a selection of past apocentres
-%        and the present day. The colours represent the projected DM density in
-%        the $y$-$z$ plane, scaled to be logarithmic over 5 orders of magnitude.
-%        The green dot represents the final expected position of the galaxy, and
-%        the solid and dotted grey curves represent the orbit over one previous
-%        or future radial oscillation, respectively.
-%}
-%    \label{fig:umi_sim_images}
-%\end{figure}
+    \label{fig:extra_tracks}
+    \caption{The tidal tracks for Scl models in the MW-only potential with various structural parameters.
+While different structural parameters may evolve further along the tidal track, none completely disrupts and all evolve to a similar final state (except the cored model).}
+\end{figure}
 
@@
 \subsection{Spectroscopic sample selection} \label{sec:velocity_sample}
-To create a sample of spectroscopic members, we compile LOS velocities from
+To create a sample of spectroscopic members, we compile line-of-sight (LOS) velocities from
 several surveys. For
@@
 \citet{sestito+2023b}; and APOGEE. We then cross-match all catalogues to
-J+24 Gaia stars. If a study did not report \emph{Gaia} DR3 source IDs,
+J+24 Gaia stars. If a study did not report \textit{Gaia} DR3 source IDs,
 we match to the nearest star within 1--3 arcseconds. We combine
@@
 dispersions.\footnote{Specifically, using that
-  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
-  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
-  distribution with \(N-1\) measurements.}
+  $\chi^2=\frac{s^2}{\delta \bar v^2}$, we remove stars with a
+  $\chi^2$ larger than the 99.9th percentile of the $\chi^2$
+  distribution with $N-1$ measurements.}
 
@@
 total likelihood for the velocity consistency. We assume that the satellite and
-background \(v_{\rm los, gsr}\) distributions are Gaussian in the
+background $v_\text{los, gsr}$ distributions are Gaussian in the
 Galactic Standard of Rest (GSR, i.e., same location as ICRS but
@@
 Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}, and, for the
-background, mean \(0\,\kms\) and dispersion
-\(\sigma_{\rm halo} = 100\,\kms\) \citep[e.g.][]{brown+2010}. We select
+background, mean $0\,\kms$ and dispersion
+$\sigma_\text{halo} = 100\,\kms$ \citep[e.g.][]{brown+2010}. We select
 stars with velocity-informed satellite membership probabilities of
@@
 For UMi, we shifted the velocities of \citet{spencer+2018}
-(\(-1.1\,\kms\)) and \citet{pace+2020} (\(+1.1\,\kms\) ) to account for
+($-1.1\,\kms$) and \citet{pace+2020} ($+1.1\,\kms$ ) to account for
 a systematic velocity offset. Otherwise, all studies appear to be on a
@@
 correction from both effects induces an apparent gradient of about
-$1.3\,{\rm km\,s^{-1}\,deg^{-1}}$ for Sculptor and less for Ursa Minor.
+$1.3\,\mathrm{km\,s^{-1}\,deg^{-1}}$ for Sculptor and less for Ursa Minor.
+
+Using the spectroscopic samples, we derive the velocity dispersion profiles in the GSR frame through a bayesian fit to bins of width XXX in logarithmic radius.
+\dbnote{complete}
 
@@
 For the double exponential fits, we extend \citet{jensen+2024} to fit a double exponential with free inner and outer parameters.
-Specifically, we use a similar likelihood formulation except we hold the proper motion likelihood fixed, and allow the inner and outer density components to include free position angle, ellipticity, scale radius, and fit for the relative total mass of the outer component $f_{\rm outer}$.
+Specifically, we use a similar likelihood formulation except we hold the proper motion likelihood fixed, and allow the inner and outer density components to include free position angle, ellipticity, scale radius, and fit for the relative total mass of the outer component $f_\text{outer}$.
 We perform the fits using the No-U Turn Sampler (an adaptive Hamilton-Monte Carlo algorithm), assuming uniform priors for position angle and ellipticity, and log-uniform priors for scale radius.
@@
 
-\begin{table}
+\begin{table*}
 \centering
@@
 \hline
-$R_h$ & $7.85^{+0.32}_{-0.33}$ & $17.1^{+1.3}_{-1.2}$ & $10.58^{+0.52}_{-0.42}$ & $29.7^{+3.4}_{-3.2}$\\
+$R_h$ / arcmin & $7.85^{+0.32}_{-0.33}$ & $17.1^{+1.3}_{-1.2}$ & $10.58^{+0.52}_{-0.42}$ & $29.7^{+3.4}_{-3.2}$\\
 $\theta$ & $98.8^{+4.2}_{-4.0}$ & $91.8^{+3.0}_{-3.1}$ & $50.4^{+1.7}_{-1.6}$ & $74.0^{+68.0}_{-45.0}$ \\
 ell & $0.243^{+0.031}_{-0.034}$ & $0.383^{+0.035}_{-0.036}$ & $0.598^{+0.025}_{-0.029}$ &  $0.087^{+0.086}_{-0.059}$ \\
-$f_{\rm outer}$ & --- & $0.315^{+0.058}_{-0.052}$ & --- & $0.249^{+0.043}_{-0.041}$ \\
+$f_\text{outer}$ & --- & $0.315^{+0.058}_{-0.052}$ & --- & $0.249^{+0.043}_{-0.041}$ \\
 \hline
-$\mu_{\rm [Fe/H]}$ &  $-1.671^{+0.021}_{-0.021}$ &  $-1.982^{+0.024}_{-0.024}$ & $-2.147^{+0.02}_{-0.02}$ & $-2.523^{+0.095}_{-0.087}$ \\
-$\sigma_{\rm [Fe/H]}$ & $0.574^{+0.016}_{-0.015}$ & $0.336^{+0.019}_{-0.018}$ & $0.347^{+0.015}_{-0.014}$ & $0.505^{+0.067}_{-0.066}$ \\
+$\mu_\text{[Fe/H]}$ &  $-1.671^{+0.021}_{-0.021}$ &  $-1.982^{+0.024}_{-0.024}$ & $-2.147^{+0.02}_{-0.02}$ & $-2.523^{+0.095}_{-0.087}$ \\
+$\sigma_\text{[Fe/H]}$ & $0.574^{+0.016}_{-0.015}$ & $0.336^{+0.019}_{-0.018}$ & $0.347^{+0.015}_{-0.014}$ & $0.505^{+0.067}_{-0.066}$ \\
 \hline
 \end{tabular}
-\end{table}
+\end{table*}
 

### 2025-12-11 notebooks/saga.jl

@@ -1,0 +1,172 @@
+### A Pluto.jl notebook ###
+# v0.20.20
+
+using Markdown
+using InteractiveUtils
+
+# ╔═╡ 8210d650-c4a3-11f0-ba24-efea8c8c5414
+begin
+	import Pkg; Pkg.activate()
+
+	using LilGuys
+	using Arya
+	using CairoMakie
+	using DataFrames, CSV
+end
+
+# ╔═╡ 478f92ef-0e95-4d0f-a9ac-8104280902d6
+CairoMakie.activate!(type=:png)
+
+# ╔═╡ e320a3f6-49e5-47ad-8698-117c7b3a505a
+saga = CSV.read("saga.txt", DataFrame)
+
+# ╔═╡ c72b1f54-b60f-4d51-8e65-fa51d33cf296
+names(saga)
+
+# ╔═╡ 681962ca-bf54-4e53-8c77-dc3c64224a13
+saga.Object
+
+# ╔═╡ fe297631-d40c-434c-94f0-bc554c4c8de7
+scatter(saga[!, "log g"], saga.Teff)
+
+# ╔═╡ da41bcb1-e867-426f-8b31-c092af016417
+function get_abund_ratio(df, col)
+	ele1, ele2 = split(col, "/")
+	if ele2 == "H"
+		xs = df[!, "[$ele1/H]"]
+	else
+		xs = df[!, "[$ele1/H]"] .- df[!, "[$ele2/H]"]
+	end
+
+	return xs
+end
+
+# ╔═╡ baf0d831-80ed-44e1-8a56-09cb016e557d
+function plot_abund!(df, x, y; color=nothing, kwargs...)
+	xs = get_abund_ratio(df, x)
+	ys = get_abund_ratio(df, y)
+
+
+	filt = .!ismissing.(xs)
+	filt .&= .!ismissing.(ys)
+	if (color isa String) && (length(split(color, "/")) == 2)
+		color = get_abund_ratio(df, color)
+		filt .&= .!ismissing.(color)
+		color = color[filt]
+	end
+	
+	scatter!(xs[filt], ys[filt]; color=color, kwargs...)
+
+end
+
[…]
